### 2003-01-31 trunk/vcd/paper/dsc2003.tex
first version of DSC paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@189 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\documentclass[10pt]{article}
\usepackage{color, natbib, graphicx}
\usepackage{dsc2003}

\begin{document}

\title{Visualizing the independence problem using extended association and
mosaic plots}
\author{David Meyer
\thanks{Institute of Statistics and Probability Theory, Vienna University of Technology}
\And
Achim Zeileis 
\thanks{Institute of Statistics and Probability Theory, Vienna
University of Technology}
\And
Kurt Hornik
\thanks{Institute of Statistics, Vienna
University of Economics and Business Administration}
}

\maketitle

\section{Introduction}

Statistical models built for the analysis of multivariate data,
quickly become complex with increasing dimensionality.
One idea of visualization techniques is to
use the human visual system to detect structures in the data that
possibly are not obvious from solely numeric output (e.g., test
statistics).
The {\sf R} package `\texttt{vcd}'---based on the Book `Visualizing Categorical
Data' \cite[][]{vcd:Friendly:2000}---includes methods for the (mostly
graphical) exploration of categorical data, such as:
\begin{itemize}
 \item fitting and graphing of discrete data,
 \item plots for the independence and symmetry problems, and
 \item visualization techniques for log-linear models.
\end{itemize}
In this talk, we focus on the visualization of the independence
problem, typically analyzed using a table of relative frequencies $\pi_{ij\dots}$
with two or more dimensions. In the setting of bivariate problems, the
null hypotheses is $\pi_{ij} = \pi_{i+}\pi_{+j}$. With additional variables
(often used to stratify the data), more independence models can be
envisaged, including the null hypotheses of:
\begin{itemize}
 \item total independence:
  $\pi_{ijk\dots} = \pi_{i++\dots}\pi_{+j+\dots}\pi_{++k\dots}\dots$
 \item conditional independence:
  $\pi_{ijk\dots} = \pi_{i|k\dots}\pi_{j|k\dots}$
 \item joint independence:
  $\pi_{ijk\dots} = \pi_{ij+\dots}\pi_{++k\dots}$
\end{itemize}
Classical non-graphical methods for these problems include the Chi-square
test, Fisher's exact test (for fourfold tables), the
Cochran-Mantel-Haenzel test (for $2 \times 2 \times
n$-tables), and the analysis of log-linear models for more
complex settings.

\section{Association Plots}

Association plots \cite[][]{vcd:Cohen:1980} are inspired by the classical Chi-squared test of
independence for two categorical variables, which---based on a $m
\times n$-table of observed frequencies $o_{ij}$---is performed using the test
statistic

\[\chi^2 = \sum_{i}^{m}\sum_{j}^{n}r_{ij}^2\]

\noindent where the \emph{Pearson-residuals} $r_{ij}$ are the standardized deviations
of the observed from the expected frequencies:

\[r_{ij} = \frac{o_{ij} - e_{ij}}{\sqrt{e_{ij}}}\]

\noindent Under the null of independence, $e_{ij} = o_{i+}o_{+j}/o_{++}$, and $\chi^2$
follows a Chi-squared distribution with $(m-1)(n-1)$ degrees of
freedom.

The association plot precisely visualizes the table of
Pearson-residuals (see Figure \ref{fig:assocbase}):
each cell is represented by a rectangle that has (signed) height
proportional to the corresponding Pearson-residual $r_{ij}$ and width proportional to the
square root of the expected counts $\sqrt{e_{ij}}$. Thus the area is proportional to
the raw residuals $o_{ij} - e_{ij}$. The sign of the residual is redundantly
coded by orientation and color of the corresponding rectangle.

\begin{figure*}[htbp]
  \begin{center}
    \includegraphics[width = 10cm]{assocbase}
    \includegraphics[width = 10cm]{assocnew}
    \caption{`Classical' (top) and extended (bottom) association plot
    for the 1995 home and away goals of the Deutsche Bundesliga. The
    shading gives more information on the importance of the Pearson residuals.}
    \label{fig:assocbase}
  \end{center}
\end{figure*}

% [1 graphisches Beispiel; HairEyeColor? Fussballdaten?]

\section{Mosaic Plots}
Mosaic Plots can be seen as an extension of grouped
bar charts, where width and heights of the bars show the relative
frequencies of the two variables: a mosaic plot simply
consists of a collection of tiles whose
sizes are proportional to the observed cell frequencies (see Figure \ref{fig:mosaicbase}).

\begin{figure*}[htbp]
  \begin{center}
    \includegraphics[width = 10cm]{mosaicbase}
    \caption{Mosaic plot (with Friendly extensions)
    for the 1995 home and away goals of the Deutsche Bundesliga. As in
    extended association plots, the shading visualizes the deviations
    (residuals) from the independence model.}
    \label{fig:mosaicbase}
  \end{center}
\end{figure*}

Sequential horizontal and vertical recursive splits are used to
visualize the frequencies of more than 2 variables, each new variable
conditional to the previously entered variables. A first extension by
\cite[][]{vcd:Friendly:1994} uses a color coding of the tiles to visualize deviations
(residuals) from a given log-linear model fitted to the table, that is, from
the expected frequencies under arbitrary (independence)
hypotheses---the color of each tile depending on the sign, and the
brightness being proportional to the corresponding residual.

% [1 graphisches Beispiel mit selben Daten]

\section{Extensions}
Our extensions to these two visualization methods cover three topics:
improvement of color coding, visualization of the overall
test of independence, and flexible specification of graphical
parameters.
The color schemes used by Michael Friendly in his implementations are
rather poor, due to limitations of the SAS software used that only
implements a discrete color palette in the `RGB' color space that does not, e.g., 
allow for copier proofness and homogeneous saturations over different colors. Our
implementation uses continuous color schemes in the device-independent
`HCL' (Hue-Chroma-Luminance) color space. The continuous scheme is
more appropriate to visualize the importance of the Pearson residuals
as any possible discretization of the scale.

In addition to the `local' independence information (shading of the
residuals), we also offer to simultaneously visualize the overall independence
model by adding white stripes to the rectangles as long as a specified test does not
reject the null---the colors, then, appear `lighter' than without stripes.
The user currently can choose the test---either the classical
Chi-Square test, or the maximum-residual test. The latter uses a
(sampled) conditional exact distribution of the maximal Pearson
residual (compared to the sum of squared residuals used for the
Chi-squared test). Using this maximum-test assures consistency between the `local' and
`global' independence information: when the overall model is
significant, the plot indeed visualizes the `guilty' residuals.

Finally, one might wish to use graphical parameters varying from tile
to tile to visualize additional information.
Our implementation offers full control over 
each element by using a call-back function
invoked each time a tile is plotted and that is given the position
information (that is, the index in the visualized table).

\section{Conclusion}
Both methods described (association plot and mosaic plot) are used to
visualize the independence problem. Our extensions include better
color schemes, visualization of the overall independence model, and
more flexible manipulation of the graphical parameters.
Future Work will include extensions for more than two dimensions
(such as conditional plots or trellis layouts), and improved
labelling.

\bibliographystyle{apalike}
\bibliography{vcd}

\end{document}
% LocalWords:  proofness

### 2003-03-08 trunk/vcd/paper/draft.Rnw
Draft for vcd paper added. dsc2003.* is now depracated and will probably be removed. Draft is now an Sweave file and therefore the graphics were removed from the cvs.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@217 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\documentclass[10pt]{article}
\usepackage{color, natbib, graphicx}
\usepackage{dsc2003}

\setlength{\parskip}{0.5ex plus0.1ex minus0.1ex}
\setlength{\parindent}{0em}

\bibpunct{(}{)}{;}{a}{}{,}

\begin{document}

\SweaveOpts{engine=R,eps=FALSE}

<<preliminaries,echo=FALSE,results=hide>>=
library(vcd)
data(Bundesliga)
goals <- xtabs(~ AwayGoals + HomeGoals, data = Bundesliga, subset = Year == 1995)
data(HairEyeColor)
haireye <- margin.table(HairEyeColor, 1:2)
data(UCBAdmissions)
@

\title{Visualizing the Independence Problem Using Extended Association and
Mosaic Plots}
\author{\hfill David Meyer$^\dag$ \hfill Achim Zeileis$^\dag$ \hfill Kurt Hornik$^\ddag$ \hfill \hfill \\[0.2cm]
\dag {\it \small Institut f\"ur Statistik \& Wahrscheinlichkeitstheorie,
 Technische Universit\"at Wien}\\
\ddag {\it \small Institut f\"ur Statistik, Wirtschaftsuniversit\"at Wien}
}

\maketitle

\begin{abstract}
Two visualization techniques for the independence problem in 2-way
contingency tables---assocation and mosaic plots---are extended in two directions:
\begin{enumerate}
\item The visualization is enhanced by displaying the significance of an appropriate
test for independence and by using better color schemes.
\item The implementation in the \textsf{R} system is improved using a more modular
design and allowing for more flexible specification of plotting parameters.
\end{enumerate}
Furthermore, some enhancements for displays of various types of independence in
general multi-way tables are suggested.\footnote{Please note that all of this is
work in progress and neither theoretically nor computationally fully sound.}
\end{abstract}

\section{Introduction} \label{sec:introduction}

Statistical models built for the analysis of multivariate data
quickly become complex with increasing dimensionality.
One idea of visualization techniques is to
use the human visual system to detect structures in the data that
possibly are not obvious from solely numeric output (e.g., test
statistics).
The \textsf{R} package \texttt{vcd}---inspired by the Book ``Visualizing Categorical
Data'' \citep{vcd:Friendly:2000}---includes methods for the (mostly
graphical) exploration of categorical data, such as:
\begin{itemize}
 \item fitting and graphing of discrete distributions,
 \item plots and tests for independence and symmetry problems,
 \item visualization techniques for log-linear models.
\end{itemize}
Here, we focus on the visualization of the independence
problem, in particular in 2-way tables.

\section{Tests for independence in 2-way tables} \label{sec:tests}

We consider a 2-way contingency table with cell frequencies $(n_{ij})$ for
$i = 1, \dots, I$ and $j = 1, \dots, J$ and row and column sums
$n_{i+} = \sum_i n_{ij}$ and $n_{+j} = \sum_j n_{ij}$ respectively. For
convenience the number of observations is denoted $n = n_{++}$.

Given an underlying distribution with theoretical cell probabilities
$\pi_{ij}$ the null hypothesis of independence of the two categorical
variables can be formulated as
\begin{equation} \label{eq:H0}
H_0: \; \pi_{ij} \quad = \quad \pi_{i+} \pi_{+j}.
\end{equation}

The expected cell frequencies in this model are
$\hat n_{ij} = n_{i+}n_{+j}/n$. The best known and most used way
to measure the discrepancy
between observed and expected values are the Pearson residuals
\begin{equation} \label{eq:Pearson}
r_{ij} = \frac{n_{ij} - \hat n_{ij}}{\sqrt{\hat n_{ij}}}.
\end{equation}
Therefore, a rather intuitive idea is to reject the null hypothesis when
there are residuals which are too extreme, i.e., not close enough to
zero. The most convenient way to aggregate the $I \times J$ residuals
to one test statistic is their squared sum
\begin{equation} \label{eq:ChiSquare}
X^2 = \sum_{i, j}r_{ij}^2,
\end{equation}
because it is known to have a limiting $\chi^2$ distribution with $(I-1)(J-1)$
degrees of freedom under the null hypothesis.
This is the well-known $\chi^2$ test for independence in 2-way tables.
%% reference needed

But this is not the only plausible way of aggregation of the Pearson
residuals. There are many thinkable functionals $\lambda(\cdot)$ which
lead to reasonable test statistics $\lambda(r_{ij})$, the sum of squares
is just one of them. Another functional which is very helpful for identifying
the cells which cause the potential dependence is the maximum of the absolute
values
\begin{equation} \label{eq:MaxAbs}
M = \max_{i, j} |r_{ij}|.
\end{equation}
Given a critical value $c_\alpha$ for this test statistic all residuals whose
absolute value exceeds $c_\alpha$ violate the hypothesis of indendence at level
$\alpha$ \cite[][ch. 7]{vcd:Mazanec+Strasser:2000}. Thus, the culprits causing
the dependence can be easily identified. The critical values $c_\alpha$ can be
derived from the conditional permutation distribution of the table $(n_{ij})$.
Of course, the $\chi^2$ test \ref{eq:ChiSquare} can also be carried out as
a conditional test instead of using the unconditional asymptotic distribution.

Other classical tests for the hypothesis of independence include
Fisher's exact test, asymptotic tests of the odds ratio
and the Mantel-Haenzel test.

\section{Visualization techniques for 2-way tables} \label{sec:visualization}

The two best known visualization techniques for independence in 2-way tables are
association plots and mosaic plots. Both are suitable to bring out departure of an
observed table $(n_{ij})$ from the expected table $(\hat n_{ij})$ in a graphical way.
The former focuses on the visualization of the Pearson residuals $r_{ij}$ (under
independence) while the latter primarily displays the observed frequencies
$n_{ij}$.

\subsection{Association plots}

Association plots \citep{vcd:Cohen:1980} visualize the table of Pearson residuals:
each cell is represented by a rectangle that has (signed) height
proportional to the corresponding Pearson-residual $r_{ij}$ and width proportional to the
square root of the expected counts $\sqrt{\hat n_{ij}}$. Thus the area is proportional to
the raw residuals $n_{ij} - \hat n_{ij}$. The sign of the residual is redundantly
coded by orientation and color of the corresponding rectangle.

Figure \ref{fig:goals-assoc-classic} shows the association plot for the table of
home and away goals for all 306 games in the 1995/6 season of the German soccer league
Bundesliga \citep{vcd:Knorr-Held:1999}. In particular, it can be seen that there are
fewer games ending 1:0 and more ending 1:1 than would be expected under independence.
But it is absolutely unclear whether this difference is significant or not.

\begin{figure}[htbp]
\begin{center}
<<goals-assoc-classic,fig=TRUE,echo=FALSE,height=5,width=5>>=
grid.assocplot(goals, rot = 0)
@
\caption{Association plot for the Bundesliga data.}
\label{fig:goals-assoc-classic}
\end{center}
\end{figure}

\subsection{Mosaic plots}

Mosaic plots can be seen as an extension of grouped
bar charts, where width and height of the bars show the relative
frequencies of the two variables: a mosaic plot simply
consists of a collection of tiles whose
sizes are proportional to the observed cell frequencies
(see Figure \ref{fig:goals-mosaic-classic}).

\begin{figure}[htbp]
\begin{center}
<<goals-mosaic-classic,fig=TRUE,echo=FALSE,height=5,width=5>>=
mosaicplot(t(goals), shade = TRUE, main = "")
@
\caption{Friendly mosaic plot for the Bundesliga data.}
\label{fig:goals-mosaic-classic}
\end{center}
\end{figure}

Sequential horizontal and vertical recursive splits are used to
visualize the frequencies of more than 2 variables, each new variable
conditional to the previously entered variables. A first extension by
\cite{vcd:Friendly:1994} uses a color coding of the tiles to visualize deviations
(residuals) from a given log-linear model fitted to the table, that is, from
the expected frequencies under arbitrary independence hypotheses. In this
extension, the sign of the residuals is coded by solid and dashed line
rectangles respectively. Furthermore, residuals that exceed an absolute
value of 2 are shaded light blue and red respectively, those that even
exceed an absolute value of 4 are shaded with full saturation (in an
HSV color scheme). The heuristic idea behind this shading is that
cells with a large residual might cause the maximum test (\ref{eq:MaxAbs})
to reject the hypothesis of independence. But in general it is unknown to
which significance level $\alpha$ the values $c_\alpha = 2$ and $4$
correspond.

Figure \ref{fig:goals-mosaic-classic} shows the Friendly mosaic plot for
the same Bundesliga data as in Figure \ref{fig:goals-assoc-classic}.
In this plot the residuals
are visualized by the color shading of the cells. Again, the cells for
the 1:0 and 1:1 results are highlighted, furthermore there seem to be
too many 6:3 results as would be plausible under independence. Whereas
the association plot in Figure \ref{fig:goals-assoc-classic} contains
no information about the result of a test for independence, the shading
in this plot conveys the impression that there might indeed be
significant departure from independence (which would indicate that there
is not only {\em no} home field advantage in the German Bundesliga but even
an advantage for the away team). However, it is not clear if there
is indeed evidence for this or not.

\section{Extensions} \label{sec:extensions}

In this section we suggest some improvements and extensions which can be
applied to both visualization techniques introduced in Section
\ref{sec:visualization}. Firstly, the visual enhancements
\begin{itemize}
\item combining visualization of and testing for independence
\item better color schemes
\end{itemize}
are addressed. Secondly, the implementation enhancements
\begin{itemize}
\item more modular implementation using \texttt{grid}
\item more flexible plotting parameter specification
\end{itemize}
are described. None of these should be thought of as finished work
but rather as a collection of ideas which still need thorough
refinement and proper implementation.

\subsection{Visualization enhancements}

The extensions of \cite{vcd:Friendly:1994} to mosaic plots provide
substantial improvement of the original mosaic plots and lifted them
from a plot for contingency tables to a visualization technique for
independence problems. However, it has two major drawbacks: Firstly,
the significance level for the hard-coded critical values 2 and 4 is
usually unknown. Secondly, the HSV colors for (light) blue and red
were chosen due to the restrictions of the SAS software (in which
the plots were first implemented), they are not device-independent and
do not provide homogeneous saturations over different colors
and copier proofness. We suggest some ways to overcome these drawbacks.

As pointed out in Section \ref{sec:tests} the critical values for
the maximum statistic $M$ from (\ref{eq:MaxAbs}) can be derived
from the conditional permutation distribution. Instead of the hard-coded
values 2 and 4 the particular critical values, e.g., at the level
$\alpha = 0.9$ and $0.95$, for the table to be visualized could be
used, thus highlighting exactly those residuals who cause a (potential)
dependency within the table. At the moment these critical values
are derived in \texttt{vcd} by simulation of the underlying distribution.

As an alternative to the HSV colors the device-independent
HCL (Hue-Chroma-Luminance) color space \citep{vcd:Ihaka:2003} could be
used. The resulting association plot (using a critical value for
$\alpha = 0.95$) for the Bundesliga data can be seen in
Figure \ref{fig:goals-assoc-signif}. It indicates very clearly that
there is no evidence for rejecting the hypothesis of independence
in this table. Thus, there seems to be neither an advantage for
the home team nor for the away team.

\begin{figure}[htbp]
\begin{center}
<<goals-assoc-signif,fig=TRUE,echo=FALSE,height=5,width=5>>=
grid.assocplot(goals, color = "signif", rot = 0)
@
\caption{Extended association plots for the Bundesliga data.}
\label{fig:goals-assoc-signif}
\end{center}
\end{figure}

To illustrate what a `postive' example looks like we give the association plot
for the famous Hair-Eye-Color data set cross-tabulating the hair and eye
color of 592 statistics students. It is easy to see the hypothesis
of independence is rejected at 95\% level by the maximum test (\ref{eq:MaxAbs})
due to higher frequencies of student with hair and eye color
black/brown, blond/blue and red/green and lower frequencies of
blond/brown and black/blue than would be plausible under independence.

\begin{figure}[htbp]
\begin{center}
<<haireye-assoc-signif,fig=TRUE,echo=FALSE,height=5,width=5>>=
grid.assocplot(haireye, color = "signif")
@
\caption{Extended association plots for the Hair-Eye-Color data.}
\label{fig:haireye-assoc-signif}
\end{center}
\end{figure}

We have seen that this approach performs very well when the test statistic
which should be visualized is the maximum statistic (\ref{eq:MaxAbs}) but
it is difficult to extend this approach to general tests for independence,
in particular in the HCL space, where the range of chroma and luminance are
not independent. HSV space on the other hand has the advantage, that all
three dimensions hue, saturation and value have the range $[0,1]$ and can
be varied independently. Therefore, a different approach than the one
described above would be to use the hue for the sign of the residuals as
before, the saturation for the absolute size of the residuals and the
value as an indicator for the significance of some test statistic, only
using the full value when the overall test for independence rejects
the null hypothesis. The resulting mosaic plots under this paradigm
are depicted in Figure \ref{fig:both-mosaic-signif}. Again, it can
be clearly seen that despite some large residuals there is no
evidence against independence for the Bundesliga, but that there the
null hypothesis has to be rejected for the Hair-Eye-Color data.

\begin{figure}[htbp]
\begin{center}
<<goals-mosaic-signif,fig=TRUE,echo=FALSE,height=5,width=5>>=
grid.mosaicplot(goals, main = NULL)
@
<<haireye-mosaic-signif,fig=TRUE,echo=FALSE,height=5,width=5>>=
grid.mosaicplot(haireye, main = NULL)
@
\caption{Extended mosaic plots for the Bundesliga and Hair-Eye-Color data.}
\label{fig:both-mosaic-signif}
\end{center}
\end{figure}

\subsection{Implementation enhancements}

The current implementation of association and mosaic plots in \textsf{R} suffers
from two main disadvantages: Firstly, it is not easy to recycle the
plots in conditioning plots or pairs plots (like mosaic matrices) as
they have been implemented using \textsf{R}'s base graphics engine.
The new implementation was written from scratch in \texttt{grid} offering
much more versatility amongst some other minor advantages and convenient
improvements. Secondly, the graphics parameters of the rectangles in
association and mosaic plots (like in almost all standard \textsf{R}
plots), like color and line type, cannot be specified for each cell
by the user. To overcome this, the user can specify in the current
implementation in \texttt{vcd} either arrays of graphics parameters
of the same dimensionality as the object that is plotted or a function
which computes these graphics parameters based on the original table
and its Pearson residuals. Functions are provided for the shading schemes
described in the previous section.

\section{Multi-way tables}

Independence problems do not only occur in 2-way tables, although that
is an important special case, but they are also important in tables of
higher dimensionality and can follow much more complex patterns.
These are again defined based on the underlying table of theoretical cell probabilities
$(\pi_{ij\dots})$ with more than two dimensions. Models of interest include the
null hypotheses of:
\begin{itemize}
 \item total independence:
  $\pi_{ijk\dots} = \pi_{i++\dots}\pi_{+j+\dots}\pi_{++k\dots}\dots$
 \item conditional independence:
  $\pi_{ijk\dots} = \pi_{i|k\dots}\pi_{j|k\dots}$
 \item joint independence:
  $\pi_{ijk\dots} = \pi_{ij+\dots}\pi_{++k\dots}$
\end{itemize}
Classical non-graphical methods for these problems include the Chi-square
test, Fisher's exact test, the Cochran-Mantel-Haenzel test (for $2 \times 2 \times
K$-tables), and the analysis of log-linear models for more
complex settings.

Two natural ways to use the visualization techniques described in the previous
sections would be to use (Trellis-like) conditioned plots or pairs plots
(like mosaic matrices) to visualize these more complex patterns of
independence.

Two ideas for the problem of conditional independence are briefly outlined
here and illustrated using the famous admissions data of the University of
California at Berkley (UCB). In this data the question whether there is sex
discrimination at the UCB leads to the result that although women seem to
be disadvantaged at the aggregated level there is no sex discrimination
conditioned on the department---with the very exception of one department
in which women are {\em more} likely to be admitted than would be plausible
under independence. Exactly this is illustrated in the conditioned
assocation plot in Figure \ref{fig:UCB-assoc-signif}.

\begin{figure}[htbp]
\begin{center}
<<UCB-assoc-signif,fig=TRUE,echo=FALSE,height=5,width=5>>=
tabplot(UCBAdmissions, color = "signif")
@
\caption{Extended conditioned association plots for the UCB Admissions data.}
\label{fig:UCB-assoc-signif}
\end{center}
\end{figure}

Similarly, the same data can be visualized using a mosaic matrix where
a conditional independence model is fitted in each plot (see Figure
\ref{fig:UCB-mosaic-signif}).

\begin{figure}[htbp]
\begin{center}
<<UCB-mosaic-signif,fig=TRUE,echo=FALSE,height=5,width=5>>=
grid.mosaicpairs(UCBAdmissions, type = "conditional")
@
\caption{Extended conditioned association plots for the UCB Admissions data.}
\label{fig:UCB-mosaic-signif}
\end{center}
\end{figure}

\section{Conclusion}
We suggest a set of enhancements for visualizing the independence
problem in 2-way tables with some outlook to multi-way tables. The
extensions aim at improving the visualization by displaying both
the size of the residuals and the significance of a test for
independence and by using better color schemes. Furthermore, a new
implementation is outlined based on the graphics package \texttt{grid}
which provides more modular design and more flexible specification
of graphical parameters.

\bibliographystyle{asa}
\bibliography{vcd}

\end{document}
% LocalWords:  proofness

### 2003-03-08 trunk/vcd/paper/draft.Rnw
*** empty log message ***
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@218 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
*** empty log message ***
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@219 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \begin{center}
-<<goals-assoc-classic,fig=TRUE,echo=FALSE,height=5,width=5>>=
-grid.assocplot(goals, rot = 0)
+<<goals-assoc-classic,fig=TRUE,echo=FALSE,height=6,width=6>>=
+grid.assocplot(goals, rot = 0, gp = gp.binary)
 @
@@
 \begin{center}
-<<goals-mosaic-classic,fig=TRUE,echo=FALSE,height=5,width=5>>=
+<<goals-mosaic-classic,fig=TRUE,echo=FALSE,height=6,width=6>>=
 mosaicplot(t(goals), shade = TRUE, main = "")
@@
 \begin{center}
-<<goals-assoc-signif,fig=TRUE,echo=FALSE,height=5,width=5>>=
-grid.assocplot(goals, color = "signif", rot = 0)
+<<goals-assoc-signif,fig=TRUE,echo=FALSE,height=6,width=6>>=
+grid.assocplot(goals, rot = 0)
 @
@@
 \begin{center}
-<<haireye-assoc-signif,fig=TRUE,echo=FALSE,height=5,width=5>>=
-grid.assocplot(haireye, color = "signif")
+<<haireye-assoc-signif,fig=TRUE,echo=FALSE,height=6,width=6>>=
+grid.assocplot(haireye)
 @
@@
 \begin{center}
-<<goals-mosaic-signif,fig=TRUE,echo=FALSE,height=5,width=5>>=
+<<goals-mosaic-signif,fig=TRUE,echo=FALSE,height=6,width=6>>=
 grid.mosaicplot(goals, main = NULL)
 @
-<<haireye-mosaic-signif,fig=TRUE,echo=FALSE,height=5,width=5>>=
+<<haireye-mosaic-signif,fig=TRUE,echo=FALSE,height=6,width=6>>=
 grid.mosaicplot(haireye, main = NULL)
@@
 \begin{center}
-<<UCB-assoc-signif,fig=TRUE,echo=FALSE,height=5,width=5>>=
-tabplot(UCBAdmissions, color = "signif")
+<<UCB-assoc-signif,fig=TRUE,echo=FALSE,height=10,width=10>>=
+tabplot(UCBAdmissions)
 @
@@
 \begin{center}
-<<UCB-mosaic-signif,fig=TRUE,echo=FALSE,height=5,width=5>>=
-grid.mosaicpairs(UCBAdmissions, type = "conditional")
+<<UCB-mosaic-signif,fig=TRUE,echo=FALSE,height=8,width=8>>=
+grid.mosaicpairs(UCBAdmissions, type = "conditional", main = NULL)
 @

### 2003-03-08 trunk/vcd/R/hcl.R
*** empty log message ***
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@220 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -29,9 +29,11 @@
 #    gamma     - The device gamma.  PC CRTs have a gamma of about
 #                2.2, Macs about 1.5 (from memory) and SGI displays
 #                have a gamma of 1.
+#
+#    verbose   - give warnings if values are out of gamut
 
 hcl =
-  function(hue, chroma = 35, luminance = 85, correct = FALSE, gamma = 2.2)
+  function(hue, chroma = 35, luminance = 85, correct = FALSE, gamma = 2.2, verbose = TRUE)
   {
     HEXDIGITS = c("0", "1", "2", "3", "4", "5", "6", "7",
                   "8", "9", "A", "B", "C", "D", "E", "F")
@@ -81,7 +83,7 @@ hcl =
     #  Convert to hexadecimal strings
 
     if (any(RGB < 0 | RGB > 255, na.rm=TRUE)) {
-      warning("out of gammut RGB values")
+      if(verbose) warning("out of gammut RGB values")
       if (correct)
         RGB = pmax(pmin(RGB,  255), 0)
       else

### 2003-03-08 trunk/vcd/R/grid.assocplot.R
*** empty log message ***
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@220 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -4,19 +4,12 @@ grid.assocplot <- function(x, ...)
 }
 
 grid.assocplot.default <- function(x, xlab = NULL, ylab = NULL, main = NULL,
-                           model.formula = NULL, transpose = FALSE, labels = TRUE,
-			   color.type = "Friendly", legend = TRUE, colbins = NULL,
+                           model.formula = NULL, labels = TRUE, fontsize = 12,
+			   gp = gp.signif, legend = TRUE, colbins = NULL,
                            scale = 0.4, rot = 90, check.overlap = TRUE, axis.labels = TRUE,
 			   show.grid = FALSE, panel = FALSE, margins = c(2, 4, 5, 2))
 {
   require(grid)
-
-  if(is.null(dimnames(x))) dimnames(x) <- list(rep("", ncol(x)), rep("", nrow(x)))
-  if(is.null(names(dimnames(x)))) {
-    names(dimnames(x)) <- c("Y", "X")
-    if(is.null(xlab)) xlab <- ""
-    if(is.null(ylab)) ylab <- ""
-  }
 
   ## This could be used to fit more complex models:
   ## require(MASS)
@@ -34,15 +27,6 @@ grid.assocplot.default <- function(x, xlab = NULL, ylab = NULL, main = NULL,
   sexpctd <- sqrt(expctd)
   res <- (x - expctd) / sexpctd
   ##res[is.na(res)] <- 0
-
-  ## Transposing - why??
-  if(transpose)
-  {
-    expctd <- t(expctd)
-    x <- t(x)
-    sexpctd <- t(sexpctd)
-    res <- t(res)
-  }
 
   if(is.null(ylab)) ylab <- names(dimnames(x))[1]
   if(is.null(xlab)) xlab <- names(dimnames(x))[2]
@@ -62,12 +46,16 @@ grid.assocplot.default <- function(x, xlab = NULL, ylab = NULL, main = NULL,
   y.pos.matrix <- matrix(rep(rev(y.pos), ncols), ncol = ncols)
   cell.height <- res
   cell.width <- sexpctd
-  cell.col <- colorscheme(res, x, type = color.type)
+  if(is.function(gp)) {
+    gpfun <- gp
+    gp <- gpfun(x, res)
+  } else legend <- FALSE
+
 
   if(!panel) grid.newpage()
   if(!is.null(main)) {
     margins[3] <- margins[3] + 2
-    grid.text(main, y = unit(1, "npc") - unit(2, "lines"), gp = gpar(fontsize = 20))
[…]

### 2003-03-08 trunk/vcd/R/grid.assocplot.R
*** empty log message ***
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@221 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -186,8 +186,14 @@ gp.shading <- function(observed, residuals, hue = c(2/3, 0), colbins = c(2, 4),
   colbins <- sort(colbins)
   hue <- ifelse(res > 0, hue[1], hue[2])
   saturation <- rep(0, length(res)) + ifelse(abs(res) > colbins[1], 0.5, 0) + ifelse(abs(res) > colbins[2], 0.5, 0)
-  if(is.null(test)) value <- 1
-  else (test(observed)$p.value < (1-level))*0.5 + 0.5
+  if(is.null(test)) {
+    value <- 1
+    p.value <- NULL
+  } else {
+    p.value <- test(observed)$p.value
+    value <- (p.value < (1-level))*0.5 + 0.5
+  }
+
   col <- hsv(hue, saturation, value)
   dim(col) <- dim(residuals)
   colbins <- sort(c(range(res), colbins, -colbins, 0))
@@ -196,7 +202,7 @@ gp.shading <- function(observed, residuals, hue = c(2/3, 0), colbins = c(2, 4),
   lty <- ifelse(residuals > 0, lty[1], lty[2])
   ltybins <- range(res)
   if(diff(sign(ltybins)) > 0) ltybins <- sort(c(0, ltybins))
-  rval <- list(fill = col, lty = lty, colbins = colbins, ltybins = ltybins)
+  rval <- list(fill = col, lty = lty, colbins = colbins, ltybins = ltybins, p.value = p.value)
   class(rval) <- c("vcd.gpar", "gpar")
   return(rval)
 }

### 2003-03-11 trunk/vcd/paper/draft.Rnw
*** empty log message ***
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@222 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \SweaveOpts{engine=R,eps=FALSE}
+\setkeys{Gin}{width=0.8\textwidth}
 
@@
 \end{equation}
-Given a critical value $c_\alpha$ for this test statistic all residuals whose
+Given a critical value $c_\alpha$ for this test statistic, all residuals whose
 absolute value exceeds $c_\alpha$ violate the hypothesis of indendence at level
 $\alpha$ \cite[][ch. 7]{vcd:Mazanec+Strasser:2000}. Thus, the culprits causing
-the dependence can be easily identified. The critical values $c_\alpha$ can be
-derived from the conditional permutation distribution of the table $(n_{ij})$.
-Of course, the $\chi^2$ test \ref{eq:ChiSquare} can also be carried out as
-a conditional test instead of using the unconditional asymptotic distribution.
+the dependence can easily be identified. The critical values $c_\alpha$ can be
+derived from the conditional permutation distribution of the table $(n_{ij})$
+which is the distributation generated by all possible tables with the same
+row and column sums. Of course, the $\chi^2$ test (\ref{eq:ChiSquare}) can also be
+carried out as a conditional test instead of using the unconditional asymptotic
+distribution.
 
@@
 
-The two best known visualization techniques for independence in 2-way tables are
+The two best-known visualization techniques for independence in 2-way tables are
 association plots and mosaic plots. Both are suitable to bring out departure of an
@@
 
+\setkeys{Gin}{width=0.7\textwidth}
 \begin{figure}[htbp]
 \begin{center}
-<<goals-assoc-classic,fig=TRUE,echo=FALSE,height=6,width=6>>=
+<<goals-assoc-classic,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
 grid.assocplot(goals, rot = 0, gp = gp.binary)
@@
 \begin{center}
-<<goals-mosaic-classic,fig=TRUE,echo=FALSE,height=6,width=6>>=
+<<goals-mosaic-classic,fig=TRUE,echo=FALSE,height=6,width=7>>=
 mosaicplot(t(goals), shade = TRUE, main = "")
@@
 values 2 and 4 the particular critical values, e.g., at the level
-$\alpha = 0.9$ and $0.95$, for the table to be visualized could be
-used, thus highlighting exactly those residuals who cause a (potential)
-dependency within the table. At the moment these critical values
+$\alpha = 0.9$ and $0.99$, for the table to be visualized could be
+used. Hence, exactly those residuals causing the (potential)
+dependency within the table are highlighted. At the moment these critical values
 are derived in \texttt{vcd} by simulation of the underlying distribution.
@@
 HCL (Hue-Chroma-Luminance) color space \citep{vcd:Ihaka:2003} could be
-used. The resulting association plot (using a critical value for
-$\alpha = 0.95$) for the Bundesliga data can be seen in
+used. The resulting association plot (using a critical values for
+$\alpha = 0.9$ and $0.99$) for the Bundesliga data can be seen in
 Figure \ref{fig:goals-assoc-signif}. It indicates very clearly that
@@
 
+\setkeys{Gin}{width=0.8\textwidth}
 \begin{figure}[htbp]
@@
 To illustrate what a `postive' example looks like we give the association plot
-for the famous Hair-Eye-Color data set cross-tabulating the hair and eye
-color of 592 statistics students. It is easy to see the hypothesis
-of independence is rejected at 95\% level by the maximum test (\ref{eq:MaxAbs})
-due to higher frequencies of student with hair and eye color
-black/brown, blond/blue and red/green and lower frequencies of
-blond/brown and black/blue than would be plausible under independence.
+for the wll-known Hair-Eye-Color data set which cross-tabulates the hair and eye
+color of 592 statistics students. It is easy to see that the hypothesis
+of independence is rejected at 99\% level by the maximum test (\ref{eq:MaxAbs})
+due to higher frequencies of students with hair and eye color
+black/brown and blond/blue and lower frequencies of
+blond/brown and black/blue than would be plausible under independence. Furthermore,
+it can be seen that the residuals corresponding to the red/green and blond/hazel
+cells are also significant at the 90\% level.
 
@@
 it is difficult to extend this approach to general tests for independence,
-in particular in the HCL space, where the range of chroma and luminance are
-not independent. HSV space on the other hand has the advantage, that all
+in particular in the HCL space, where the ranges of chroma and luminance are
+not independent. The HSV space on the other hand has the advantage that all
 three dimensions hue, saturation and value have the range $[0,1]$ and can
@@
 are depicted in Figure \ref{fig:both-mosaic-signif}. Again, it can
-be clearly seen that despite some large residuals there is no
-evidence against independence for the Bundesliga, but that there the
+clearly be seen that despite some large residuals there is no
+evidence against independence for the Bundesliga data, but that the
 null hypothesis has to be rejected for the Hair-Eye-Color data.
@@
 plots), like color and line type, cannot be specified for each cell
-by the user. To overcome this, the user can specify in the current
-implementation in \texttt{vcd} either arrays of graphics parameters
+by the user. To overcome this the current implementation in \texttt{vcd}
+allows the user to specify either arrays of graphics parameters
 of the same dimensionality as the object that is plotted or a function
@@
 and its Pearson residuals. Functions are provided for the shading schemes
-described in the previous section.
+described in the previous section. Without this feature we would not be
+able to specify the shading schemes described in the previous section
+in such a flexible way.
 
@@
 Two natural ways to use the visualization techniques described in the previous
-sections would be to use (Trellis-like) conditioned plots or pairs plots
+sections would be to use (Trellis-like) conditioning plots or pairs plots
 (like mosaic matrices) to visualize these more complex patterns of

### 2003-03-11 trunk/vcd/R/assocplot.R

@@ -179,7 +179,7 @@ pearson.test <- function (x, n = 1000, FUN = function(x) max(abs(x)),
   FUNPearson <- function(x) FUN(Pearson(x))
   dist <- sapply(r2dtable(n, rowTotals, colTotals), FUNPearson)
   STATISTIC <- FUNPearson(x)
-  pdist <- function(x) mean(dist <= x)
+  pdist <- function(x) sapply(x, function(y) mean(dist <= y))
   qdist <- function(p) quantile(dist, p)
   PVAL <- switch(alternative,
                  greater = mean(dist >= STATISTIC),

### 2003-03-11 trunk/vcd/R/grid.assocplot.R
*** empty log message ***
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@222 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -5,7 +5,7 @@ grid.assocplot <- function(x, ...)
 
 grid.assocplot.default <- function(x, xlab = NULL, ylab = NULL, main = NULL,
                            model.formula = NULL, labels = TRUE, fontsize = 12,
-			   gp = gp.signif, legend = TRUE, colbins = NULL,
+			   gp = gp.signif, legend = TRUE, col.bins = NULL,
                            scale = 0.4, rot = 90, check.overlap = TRUE, axis.labels = TRUE,
 			   show.grid = FALSE, panel = FALSE, margins = c(2, 4, 5, 2))
 {
@@ -97,28 +97,28 @@ grid.assocplot.default <- function(x, xlab = NULL, ylab = NULL, main = NULL,
                   yscale = range(res), default.unit = "npc",
                   height = 0.8, width = 0.2))
 
-    colbins <- gp$colbins
-    if(is.null(colbins)) colbins <- 200
-    if(length(colbins) == 1) {
-      colbins <- floor(colbins) - 1
-      colbins <- min(res) + diff(range(res)) * ((0:colbins)/colbins)
+    if(!is.null(gp$col.legend)) {
+      col.legend <- gp$col.legend
+    } else {
+      col.bins <- gp$col.bins
+      if(is.null(col.bins)) col.bins <- min(res) + diff(range(res)) * ((0:200)/200)
+      y.pos <- col.bins[-length(col.bins)]
+      y.height <- diff(col.bins)
+      y.col <- gpfun(x, y.pos + 0.5*y.height)$fill
+      col.legend <- list(pos = y.pos, height = y.height, col = y.col)
     }
-    ltybins <- gp$ltybins
 
-    y.pos <- colbins[-length(colbins)]
-    y.height <- diff(colbins)
-    y.col <- gpfun(x, y.pos + 0.5*y.height)$fill
-
-    grid.rect(x = unit(rep(0.5, length(y.pos)), "npc"), y = y.pos,
-              height = y.height, default.unit = "native",
-              gp = gpar(fill = y.col, col = NULL),
+    grid.rect(x = unit(rep(0.5, length(col.legend$pos)), "npc"), y = col.legend$pos,
+              height = col.legend$height, default.unit = "native",
[…]

### 2003-03-12 trunk/vcd/paper/draft.Rnw
KH comments added together with correct Friendly citations
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@230 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 data(HairEyeColor)
-haireye <- margin.table(HairEyeColor, 1:2)
+haireye <- xtabs(Freq ~ Hair + Eye, data = as.data.frame(HairEyeColor), subset = Sex == "Female")
 data(UCBAdmissions)
@@
 
-\title{Visualizing the Independence Problem Using Extended Association and
-Mosaic Plots}
+\title{Visualizing Independence Using Extended Association and Mosaic Plots}
 \author{\hfill David Meyer$^\dag$ \hfill Achim Zeileis$^\dag$ \hfill Kurt Hornik$^\ddag$ \hfill \hfill \\[0.2cm]
@@
 \item The visualization is enhanced by displaying the significance of an appropriate
-test for independence and by using better color schemes.
+test for independence and by using improved color schemes.
 \item The implementation in the \textsf{R} system is improved using a more modular
@@
 Furthermore, some enhancements for displays of various types of independence in
-general multi-way tables are suggested.\footnote{Please note that all of this is
-work in progress and neither theoretically nor computationally fully sound.}
+general multi-way tables are suggested.
 \end{abstract}
@@
 The expected cell frequencies in this model are
-$\hat n_{ij} = n_{i+}n_{+j}/n$. The best known and most used way
-to measure the discrepancy
+$\hat n_{ij} = n_{i+}n_{+j}/n$. The best known and most used
+measure of discrepancy
 between observed and expected values are the Pearson residuals
 \begin{equation} \label{eq:Pearson}
-r_{ij} = \frac{n_{ij} - \hat n_{ij}}{\sqrt{\hat n_{ij}}}.
+r_{ij} \quad = \quad \frac{n_{ij} - \hat n_{ij}}{\sqrt{\hat n_{ij}}}.
 \end{equation}
@@
 \begin{equation} \label{eq:ChiSquare}
-X^2 = \sum_{i, j}r_{ij}^2,
+X^2 \quad = \quad \sum_{i, j}r_{ij}^2,
 \end{equation}
@@
 But this is not the only plausible way of aggregation of the Pearson
-residuals. There are many thinkable functionals $\lambda(\cdot)$ which
+residuals. There are many conceivable functionals $\lambda(\cdot)$ which
 lead to reasonable test statistics $\lambda(r_{ij})$, the sum of squares
 is just one of them. Another functional which is very helpful for identifying
-the cells which cause the potential dependence is the maximum of the absolute
+the cells which cause the dependence (if any) is the maximum of the absolute
 values
 \begin{equation} \label{eq:MaxAbs}
-M = \max_{i, j} |r_{ij}|.
+M \quad = \quad \max_{i, j} |r_{ij}|.
 \end{equation}
@@
 absolute value exceeds $c_\alpha$ violate the hypothesis of indendence at level
-$\alpha$ \cite[][ch. 7]{vcd:Mazanec+Strasser:2000}. Thus, the culprits causing
-the dependence can easily be identified. The critical values $c_\alpha$ can be
-derived from the conditional permutation distribution of the table $(n_{ij})$
-which is the distributation generated by all possible tables with the same
-row and column sums. Of course, the $\chi^2$ test (\ref{eq:ChiSquare}) can also be
-carried out as a conditional test instead of using the unconditional asymptotic
-distribution.
+$\alpha$ \cite[][ch. 7]{vcd:Mazanec+Strasser:2000}. Thus, the interesting cells causing
+the dependence can easily be identified.
 
-Other classical tests for the hypothesis of independence include
-Fisher's exact test, asymptotic tests of the odds ratio
-and the Mantel-Haenzel test.
+Furthermore, the main reason for using the unconditional limiting distribution
+for the $\chi^2$ statistic (\ref{eq:ChiSquare}) was the closed form result
+for the distribution. Recently, with the improving perfomance of computers,
+performing permutation tests---either by simulation or by computation of the
+permutation distribution---became more and more popular. In particular for
+the independence hypothesis (\ref{eq:H0}) using a permutation test is very intuitive
+due to the permutation invariance (given row and column sums) of this problem.
+By employing this approach the permutation distribution of statistics of the
+type $\lambda(r_{ij})$, i.e., also of the $\chi^2$ statistic, can be derived.
+
+Other classical tests for the hypothesis of independence  which do not fit in
+the Pearson residual-based framework described aboveinclude Fisher's exact test,
+asymptotic tests of the odds ratio (for $2\times2$ tables) and the Mantel-Haenzel
+test.
 
@@
 
-The two best-known visualization techniques for independence in 2-way tables are
+The two best known visualization techniques for independence in 2-way tables are
 association plots and mosaic plots. Both are suitable to bring out departure of an
@@
 each cell is represented by a rectangle that has (signed) height
-proportional to the corresponding Pearson-residual $r_{ij}$ and width proportional to the
+proportional to the corresponding Pearson residual $r_{ij}$ and width proportional to the
 square root of the expected counts $\sqrt{\hat n_{ij}}$. Thus the area is proportional to
 the raw residuals $n_{ij} - \hat n_{ij}$. The sign of the residual is redundantly
-coded by orientation and color of the corresponding rectangle.
+coded by position relative to the baseline and color of the corresponding rectangle.
 
@@
 Bundesliga \citep{vcd:Knorr-Held:1999}. In particular, it can be seen that there are
-fewer games ending 1:0 and more ending 1:1 than would be expected under independence.
-But it is absolutely unclear whether this difference is significant or not.
+fewer games ending 1-0 and more ending 1-1 than would be expected under independence.
+But this does not yet imply the significance of some test statistic.
 
@@
 (residuals) from a given log-linear model fitted to the table, that is, from
-the expected frequencies under arbitrary independence hypotheses. In this
-extension, the sign of the residuals is coded by solid and dashed line
-rectangles respectively. Furthermore, residuals that exceed an absolute
+the expected frequencies under independence (which does not only work in 2-way
+tables but also log-linear models for multi-way tables). In this
+extension, positive and negative signs of the residuals are coded by solid and dashed line
+rectangles respectively. Furthermore, residuals exceeding an absolute
 value of 2 are shaded light blue and red respectively, those that even
-exceed an absolute value of 4 are shaded with full saturation (in an
-HSV color scheme). The heuristic idea behind this shading is that
-cells with a large residual might cause the maximum test (\ref{eq:MaxAbs})
-to reject the hypothesis of independence. But in general it is unknown to
-which significance level $\alpha$ the values $c_\alpha = 2$ and $4$
-correspond.
+exceed an absolute value of 4 are shaded with full saturation. The heuristic
+behind this shading is that as
+the Pearson residuals are approximately $\mathcal{N}(0,1)$ distributed
+values, which implies that the highlighted cells are those with
+residuals individually significant at approximately the 5\% and 0.01\%
+level. But the main purpose of the shading is to visualize the pattern of
+deviation from independence \citep{vcd:Friendly:2000}. In particular,
+it does not provide a visualization of the maximum test (\ref{eq:MaxAbs})
+or in other words: it is in general unknown to which significance level $\alpha$ the
+values $c_\alpha = 2$ and $4$ correspond for the table being visualized.
 
@@
 are visualized by the color shading of the cells. Again, the cells for
-the 1:0 and 1:1 results are highlighted, furthermore there seem to be
-too many 6:3 results as would be plausible under independence. Whereas
+the 1-0 and 1-1 results are highlighted, furthermore there seem to be
+too many 6-3 results as would be plausible under independence. Whereas
 the association plot in Figure \ref{fig:goals-assoc-classic} contains
 no information about the result of a test for independence, the shading
-in this plot conveys the impression that there might indeed be
-significant departure from independence (which would indicate that there
-is not only {\em no} home field advantage in the German Bundesliga but even
+in this plot conveys the impression that there might be
+significant dependence (which would indicate that there
+is not only {\em no} home field advantage in the German Bundesliga in 1995/6 but even
 an advantage for the away team). However, it is not clear if there
@@
 applied to both visualization techniques introduced in Section
-\ref{sec:visualization}. Firstly, the visual enhancements
+\ref{sec:visualization}. First, the visual enhancements
 \begin{itemize}
@@
 \end{itemize}
-are addressed. Secondly, the implementation enhancements
+are addressed. Second, the implementation enhancements
 \begin{itemize}
@@
 The extensions of \cite{vcd:Friendly:1994} to mosaic plots provide
-substantial improvement of the original mosaic plots and lifted them
+substantial improvement of the original mosaic plots and enhanced them
 from a plot for contingency tables to a visualization technique for
-independence problems. However, it has two major drawbacks: Firstly,
+log-linear models (and therefore also for certain independence problems).
+However, it has two major drawbacks: First,
 the significance level for the hard-coded critical values 2 and 4 is
-usually unknown. Secondly, the HSV colors for (light) blue and red
-were chosen due to the restrictions of the SAS software (in which
-the plots were first implemented), they are not device-independent and
+usually unknown. Second, the colors for (light) blue and red
+used in the original \textsf{SAS} implementation and also in the base implementation
+in \textsf{R} are not device-independent and
 do not provide homogeneous saturations over different colors
@@
 the maximum statistic $M$ from (\ref{eq:MaxAbs}) can be derived
-from the conditional permutation distribution. Instead of the hard-coded
+from the permutation distribution. Instead of the hard-coded
 values 2 and 4 the particular critical values, e.g., at the level
-$\alpha = 0.9$ and $0.99$, for the table to be visualized could be
+$\alpha = 0.1$ and $0.01$, for the table to be visualized could be
 used. Hence, exactly those residuals causing the (potential)
@@
 
-As an alternative to the HSV colors the device-independent
+The original implementation of mosaic plots in \textsf{SAS}
+used (by default) colors in the HLS (Hue-Luminance-Saturation) space
+with hues for blue and red, full saturation and varying luminance
+for the lighter colors. The current \textsf{R} implementation is
+based on colors in the HSV (Hue-Saturation-Value) space where
+decreasing the saturation from 1 towards 0---given the hue for
+blue or red and full saturation---generates very similar colors
+as increasing the luminance for HSL colors from 0.5 towars 1.
+As an alternative to those HLS and HSV colors the device-independent
 HCL (Hue-Chroma-Luminance) color space \citep{vcd:Ihaka:2003} could be
-used. The resulting association plot (using a critical values for
-$\alpha = 0.9$ and $0.99$) for the Bundesliga data can be seen in
+used also providing homogeneous saturations over different colors and copier
+proofness. The resulting association plot (using a critical values for
+$\alpha = 0.1$ and $0.01$) for the Bundesliga data can be seen in
 Figure \ref{fig:goals-assoc-signif}. It indicates very clearly that
@@
 To illustrate what a `postive' example looks like we give the association plot
-for the wll-known Hair-Eye-Color data set which cross-tabulates the hair and eye
-color of 592 statistics students. It is easy to see that the hypothesis
-of independence is rejected at 99\% level by the maximum test (\ref{eq:MaxAbs})
+for the well-known Hair-Eye-Color data set which cross-tabulates the hair and eye
+color and sex of 592 statistics students: Figure \ref{fig:haireye-assoc-signif}
+shows the association plot for the 328 female students.. It is easy to see that
+the hypothesis of independence is rejected at 1\% level by the maximum test (\ref{eq:MaxAbs})
 due to higher frequencies of students with hair and eye color
 black/brown and blond/blue and lower frequencies of
-blond/brown and black/blue than would be plausible under independence. Furthermore,
-it can be seen that the residuals corresponding to the red/green and blond/hazel
-cells are also significant at the 90\% level.
+blond/brown and brown/blue than would be plausible under independence. Furthermore,
+it can be seen that the residual corresponding to the black/blue
+cell is also significant at the 10\% level.
 
@@
 @
-\caption{Extended association plots for the Hair-Eye-Color data.}
+\caption{Extended association plots for the Hair-Eye-Color data (female students).}
 \label{fig:haireye-assoc-signif}
@@
 <<goals-mosaic-signif,fig=TRUE,echo=FALSE,height=6,width=6>>=
-grid.mosaicplot(goals, main = NULL)
+grid.mosaicplot(t(goals), main = NULL)
 @
 <<haireye-mosaic-signif,fig=TRUE,echo=FALSE,height=6,width=6>>=
-grid.mosaicplot(haireye, main = NULL)
+grid.mosaicplot(t(haireye), main = NULL)
 @
-\caption{Extended mosaic plots for the Bundesliga and Hair-Eye-Color data.}
+\caption{Extended mosaic plots for the Bundesliga and Hair-Eye-Color data (female students).}
 \label{fig:both-mosaic-signif}
@@
 The current implementation of association and mosaic plots in \textsf{R} suffers
-from two main disadvantages: Firstly, it is not easy to recycle the
+from two main disadvantages: First, it is not easy to recycle the
 plots in conditioning plots or pairs plots (like mosaic matrices) as
@@
 much more versatility amongst some other minor advantages and convenient
-improvements. Secondly, the graphics parameters of the rectangles in
+improvements. Second, the graphics parameters of the rectangles in
 association and mosaic plots (like in almost all standard \textsf{R}
@@
 allows the user to specify either arrays of graphics parameters
-of the same dimensionality as the object that is plotted or a function
+of the same dimensionality as the object being plotted or a function
 which computes these graphics parameters based on the original table
@@
 \end{itemize}
-Classical non-graphical methods for these problems include the Chi-square
+Classical non-graphical methods for these problems include the $\chi^2$
 test, Fisher's exact test, the Cochran-Mantel-Haenzel test (for $2 \times 2 \times

### 2003-03-12 trunk/vcd/paper/draft.Rnw
*** empty log message ***
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@232 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 <<preliminaries,echo=FALSE,results=hide>>=
+set.seed(12345)
 library(vcd)
@@
 \end{itemize}
-Here, we focus on the visualization of the independence
-problem, in particular in 2-way tables.
+Here, we focus on the visualization of independence,
+in particular in 2-way tables.
 
@@
 Given an underlying distribution with theoretical cell probabilities
-$\pi_{ij}$ the null hypothesis of independence of the two categorical
+$\pi_{ij}$, the null hypothesis of independence of the two categorical
 variables can be formulated as
@@
 \end{equation}
-because it is known to have a limiting $\chi^2$ distribution with $(I-1)(J-1)$
+because this is known to have a limiting $\chi^2$ distribution with $(I-1)(J-1)$
 degrees of freedom under the null hypothesis.
@@
 lead to reasonable test statistics $\lambda(r_{ij})$, the sum of squares
-is just one of them. Another functional which is very helpful for identifying
+is just one of them. Another functional suitable for identifying
 the cells which cause the dependence (if any) is the maximum of the absolute
@@
 permutation distribution---became more and more popular. In particular for
-the independence hypothesis (\ref{eq:H0}) using a permutation test is very intuitive
+the independence hypothesis (\ref{eq:H0}), using a permutation test is very intuitive
 due to the permutation invariance (given row and column sums) of this problem.
-By employing this approach the permutation distribution of statistics of the
-type $\lambda(r_{ij})$, i.e., also of the $\chi^2$ statistic, can be derived.
+By employing this approach the permutation distribution of statistics of
+type $\lambda(r_{ij})$ (also including the $\chi^2$ statistic) can be derived.
 
-Other classical tests for the hypothesis of independence  which do not fit in
-the Pearson residual-based framework described aboveinclude Fisher's exact test,
+Other classical tests for the hypothesis of independence  not fitting in
+the Pearson residual-based framework described above include Fisher's exact test,
 asymptotic tests of the odds ratio (for $2\times2$ tables) and the Mantel-Haenzel
@@
 proportional to the corresponding Pearson residual $r_{ij}$ and width proportional to the
-square root of the expected counts $\sqrt{\hat n_{ij}}$. Thus the area is proportional to
+square root of the expected counts $\sqrt{\hat n_{ij}}$. Thus, the area is proportional to
 the raw residuals $n_{ij} - \hat n_{ij}$. The sign of the residual is redundantly
-coded by position relative to the baseline and color of the corresponding rectangle.
+coded by the rectangle's color and its position relative to the baseline.
 
@@
 Mosaic plots can be seen as an extension of grouped
-bar charts, where width and height of the bars show the relative
+bar charts where width and height of the bars show the relative
 frequencies of the two variables: a mosaic plot simply
@@
 Sequential horizontal and vertical recursive splits are used to
-visualize the frequencies of more than 2 variables, each new variable
+visualize the frequencies of more than two variables, each new variable
 conditional to the previously entered variables. A first extension by
@@
 (residuals) from a given log-linear model fitted to the table, that is, from
-the expected frequencies under independence (which does not only work in 2-way
-tables but also log-linear models for multi-way tables). In this
-extension, positive and negative signs of the residuals are coded by solid and dashed line
-rectangles respectively. Furthermore, residuals exceeding an absolute
+the expected frequencies under independence. This approach does not only work in 2-way
+tables but also in log-linear models fitted to multi-way tables. In this
+extension, positive and negative signs of the residuals are coded by rectangles with
+solid and dashed borders respectively. Furthermore, residuals exceeding an absolute
 value of 2 are shaded light blue and red respectively, those that even
 exceed an absolute value of 4 are shaded with full saturation. The heuristic
-behind this shading is that as
-the Pearson residuals are approximately $\mathcal{N}(0,1)$ distributed
-values, which implies that the highlighted cells are those with
-residuals individually significant at approximately the 5\% and 0.01\%
-level. But the main purpose of the shading is to visualize the pattern of
-deviation from independence \citep{vcd:Friendly:2000}. In particular,
-it does not provide a visualization of the maximum test (\ref{eq:MaxAbs})
-or in other words: it is in general unknown to which significance level $\alpha$ the
+behind this shading is that
+the Pearson residuals are approximately standard normal which implies
+that the highlighted cells are those with
+residuals {\em individually} significant at approximately the 5\% and 0.01\%
+level. But the main purpose of the shading is not to visualize significance but the pattern of
+deviation from independence \citep[p. 109]{vcd:Friendly:2000}. In particular,
+this shading does not provide a visualization of the maximum test (\ref{eq:MaxAbs})
+as in general it is unknown to which significance level $\alpha$ the
 values $c_\alpha = 2$ and $4$ correspond for the table being visualized.
@@
 the same Bundesliga data as in Figure \ref{fig:goals-assoc-classic}.
-In this plot the residuals
+In this plot, the residuals
 are visualized by the color shading of the cells. Again, the cells for
-the 1-0 and 1-1 results are highlighted, furthermore there seem to be
-too many 6-3 results as would be plausible under independence. Whereas
+the 1-0 and 1-1 results are highlighted and furthermore, there seem to be
+too many 6-3 results than would be plausible under independence. Whereas
 the association plot in Figure \ref{fig:goals-assoc-classic} contains
@@
 is not only {\em no} home field advantage in the German Bundesliga in 1995/6 but even
-an advantage for the away team). However, it is not clear if there
-is indeed evidence for this or not.
+an advantage for the away team). However, it is unclear whether there
+is evidence for this or not.
 
@@
 
-In this section we suggest some improvements and extensions which can be
+In this section, we suggest some improvements and extensions which can be
 applied to both visualization techniques introduced in Section
@@
 \item combining visualization of and testing for independence
-\item better color schemes
+\item improved color schemes
 \end{itemize}
@@
 
-As pointed out in Section \ref{sec:tests} the critical values for
+As pointed out in Section \ref{sec:tests}, the critical values for
 the maximum statistic $M$ from (\ref{eq:MaxAbs}) can be derived
 from the permutation distribution. Instead of the hard-coded
-values 2 and 4 the particular critical values, e.g., at the level
+values 2 and 4, the particular critical values, e.g., at the levels
 $\alpha = 0.1$ and $0.01$, for the table to be visualized could be
 used. Hence, exactly those residuals causing the (potential)
-dependency within the table are highlighted. At the moment these critical values
+dependency within the table are highlighted. At the moment, these critical values
 are derived in \texttt{vcd} by simulation of the underlying distribution.
@@
 The original implementation of mosaic plots in \textsf{SAS}
-used (by default) colors in the HLS (Hue-Luminance-Saturation) space
-with hues for blue and red, full saturation and varying luminance
-for the lighter colors. The current \textsf{R} implementation is
-based on colors in the HSV (Hue-Saturation-Value) space where
+uses (by default) colors in the HLS (Hue--Luminance--Saturation) color space
+with hues for blue and red, full saturation, and varying luminance
+for lighter colors. The current \textsf{R} implementation is
+based on colors in the HSV (Hue--Saturation--Value) space where
 decreasing the saturation from 1 towards 0---given the hue for
 blue or red and full saturation---generates very similar colors
-as increasing the luminance for HSL colors from 0.5 towars 1.
-As an alternative to those HLS and HSV colors the device-independent
-HCL (Hue-Chroma-Luminance) color space \citep{vcd:Ihaka:2003} could be
-used also providing homogeneous saturations over different colors and copier
-proofness. The resulting association plot (using a critical values for
+as increasing the luminance for HSL colors from 0.5 towards 1.
+As an alternative to those HLS and HSV colors, the device-independent
+HCL (Hue--Chroma--Luminance) color space \citep{vcd:Ihaka:2003} could be
+used which also provides homogeneous saturations over different colors and copier
+proofness. The resulting association plot (using critical values for
 $\alpha = 0.1$ and $0.01$) for the Bundesliga data can be seen in
@@
 color and sex of 592 statistics students: Figure \ref{fig:haireye-assoc-signif}
-shows the association plot for the 328 female students.. It is easy to see that
+shows the association plot for the 328 female students. It is easy to see that
 the hypothesis of independence is rejected at 1\% level by the maximum test (\ref{eq:MaxAbs})
@@
 blond/brown and brown/blue than would be plausible under independence. Furthermore,
-it can be seen that the residual corresponding to the black/blue
-cell is also significant at the 10\% level.
+it can also be seen that the residual corresponding to the black/blue
+cell is significant at the 10\% level.
 
@@
 We have seen that this approach performs very well when the test statistic
-which should be visualized is the maximum statistic (\ref{eq:MaxAbs}) but
+which should be visualized is the maximum statistic (\ref{eq:MaxAbs}). However,
 it is difficult to extend this approach to general tests for independence,
 in particular in the HCL space, where the ranges of chroma and luminance are
-not independent. The HSV space on the other hand has the advantage that all
+not independent. In contrast the HSV space has the advantage that all
 three dimensions hue, saturation and value have the range $[0,1]$ and can
 be varied independently. Therefore, a different approach than the one
-described above would be to use the hue for the sign of the residuals as
-before, the saturation for the absolute size of the residuals and the
-value as an indicator for the significance of some test statistic, only
+described above could be to use the hue for the sign of the residuals as
+before, the saturation for the absolute size of the residuals, and the
+value as an indicator for the significance of some test statistic (only
 using the full value when the overall test for independence rejects
-the null hypothesis. The resulting mosaic plots under this paradigm
+the null hypothesis). The resulting mosaic plots under this paradigm
 are depicted in Figure \ref{fig:both-mosaic-signif}. Again, it can
@@
 plots in conditioning plots or pairs plots (like mosaic matrices) as
-they have been implemented using \textsf{R}'s base graphics engine.
+they have been implemented using \textsf{R}'s base graphics engine where in
+general plotting to relative coordinates is not supported.
 The new implementation was written from scratch in \texttt{grid} offering
-much more versatility amongst some other minor advantages and convenient
+much more versatility amongst some minor advantages and convenient
 improvements. Second, the graphics parameters of the rectangles in
 association and mosaic plots (like in almost all standard \textsf{R}
-plots), like color and line type, cannot be specified for each cell
+plots), such as color and line type, cannot be specified for each cell
 by the user. To overcome this the current implementation in \texttt{vcd}
@@
 
-Two natural ways to use the visualization techniques described in the previous
+As an example, two natural ways to use the visualization techniques described in the previous
 sections would be to use (Trellis-like) conditioning plots or pairs plots
@@
 here and illustrated using the famous admissions data of the University of
-California at Berkley (UCB). In this data the question whether there is sex
+California at Berkley (UCB). In this data, the question whether there is sex
 discrimination at the UCB leads to the result that although women seem to
@@
 in which women are {\em more} likely to be admitted than would be plausible
-under independence. Exactly this is illustrated in the conditioned
+under independence. Exactly this is illustrated in the conditioning
 assocation plot in Figure \ref{fig:UCB-assoc-signif}.

### 2003-03-13 trunk/vcd/paper/draft.Rnw
*** empty log message ***
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@233 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-We consider a 2-way contingency table with cell frequencies $(n_{ij})$ for
+We consider a 2-way contingency table with cell frequencies $\{n_{ij}\}$ for
 $i = 1, \dots, I$ and $j = 1, \dots, J$ and row and column sums
@@
 residuals. There are many conceivable functionals $\lambda(\cdot)$ which
-lead to reasonable test statistics $\lambda(r_{ij})$, the sum of squares
+lead to reasonable test statistics $\lambda(\{r_{ij}\})$, the sum of squares
 is just one of them. Another functional suitable for identifying
@@
 By employing this approach the permutation distribution of statistics of
-type $\lambda(r_{ij})$ (also including the $\chi^2$ statistic) can be derived.
+type $\lambda(\{r_{ij}\})$ (also including the $\chi^2$ statistic) can be derived.
 
@@
 home and away goals for all 306 games in the 1995/6 season of the German soccer league
-Bundesliga \citep{vcd:Knorr-Held:1999}. In particular, it can be seen that there are
-fewer games ending 1-0 and more ending 1-1 than would be expected under independence.
+Bundesliga \citep[the data is available in the package \texttt{vcd}]{vcd:Knorr-Held:1999}.
+In particular, it can be seen that there are
+fewer games ending 1--0 and more ending 1--1 than would be expected under independence.
 But this does not yet imply the significance of some test statistic.
@@
 that the highlighted cells are those with
-residuals {\em individually} significant at approximately the 5\% and 0.01\%
-level. But the main purpose of the shading is not to visualize significance but the pattern of
-deviation from independence \citep[p. 109]{vcd:Friendly:2000}. In particular,
+residuals \emph{individually} significant at approximately the 5\% and 0.01\%
+level. But the main purpose of the shading is not to visualize significance but the \emph{pattern} of
+deviation from independence \citep[p.~109]{vcd:Friendly:2000}. In particular,
 this shading does not provide a visualization of the maximum test (\ref{eq:MaxAbs})
@@
 are visualized by the color shading of the cells. Again, the cells for
-the 1-0 and 1-1 results are highlighted and furthermore, there seem to be
-too many 6-3 results than would be plausible under independence. Whereas
+the 1--0 and 1--1 results are highlighted and furthermore, there seem to be
+too many 6--3 results than would be plausible under independence. Whereas
 the association plot in Figure \ref{fig:goals-assoc-classic} contains
@@
 usually unknown. Second, the colors for (light) blue and red
-used in the original \textsf{SAS} implementation and also in the base implementation
+used in the Michael Friendly's \textsf{SAS} implementation and also in the base implementation
 in \textsf{R} are not device-independent and
@@
 
-The original implementation of mosaic plots in \textsf{SAS}
+The implementation of Friendly mosaic plots in \textsf{SAS}
 uses (by default) colors in the HLS (Hue--Luminance--Saturation) color space
@@
 based on colors in the HSV (Hue--Saturation--Value) space where
-decreasing the saturation from 1 towards 0---given the hue for
-blue or red and full saturation---generates very similar colors
+decreasing the saturation from 1 towards 0
+%% ---given the hue for blue or red and full saturation---
+generates very similar colors
 as increasing the luminance for HSL colors from 0.5 towards 1.
@@
 To illustrate what a `postive' example looks like we give the association plot
-for the well-known Hair-Eye-Color data set which cross-tabulates the hair and eye
+for the well-known Hair-Eye-Color data set (avilable in base \textsf{R}) which cross-tabulates the hair and eye
 color and sex of 592 statistics students: Figure \ref{fig:haireye-assoc-signif}
@@
 here and illustrated using the famous admissions data of the University of
-California at Berkley (UCB). In this data, the question whether there is sex
+California at Berkley (UCB) which is available in base \textsf{R}.
+In this data, the question whether there is sex
 discrimination at the UCB leads to the result that although women seem to

### 2003-04-04 trunk/vcd/paper/draft.Rnw
*** empty log message ***
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@234 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
fix to pass R CMD check
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@244 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \documentclass[10pt]{article}
-\usepackage{color, natbib, graphicx}
-\usepackage{dsc2003}
+\usepackage{Sweave, graphicx}
+\usepackage{Z}
 
@@
 \title{Visualizing Independence Using Extended Association and Mosaic Plots}
-\author{\hfill David Meyer$^\dag$ \hfill Achim Zeileis$^\dag$ \hfill Kurt Hornik$^\ddag$ \hfill \hfill \\[0.2cm]
+\author{\hfill Achim Zeileis$^\dag$ \hfill David Meyer$^\dag$ \hfill Kurt Hornik$^\ddag$ \hfill \hfill \\[0.2cm]
 \dag {\it \small Institut f\"ur Statistik \& Wahrscheinlichkeitstheorie,
@@
 }
-
+\date{}
 \maketitle
@@
 <<UCB-mosaic-signif,fig=TRUE,echo=FALSE,height=8,width=8>>=
-grid.mosaicpairs(UCBAdmissions, type = "conditional", main = NULL)
+grid.mosaicpairs(UCBAdmissions, type = "conditional", gp = gp.shading, main = NULL)
 @

### 2003-04-09 trunk/vcd/inst/doc/dsc2003-independence.Rnw
*** empty log message ***
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@249 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2003-04-04 trunk/vcd/paper/draft.Rnw
@@
 \SweaveOpts{engine=R,eps=FALSE}
+%\VignetteIndexEntry{Visualizing Independence Using Extended Association and Mosaic Plots}
+%\VignetteDepends{vcd,MASS,grid}
+%\VignetteKeywords{mosaic plot, association plot, independence, HCL colors}
+%\VignettePackage{vcd}
+
 \setkeys{Gin}{width=0.8\textwidth}
@@
 <<UCB-mosaic-signif,fig=TRUE,echo=FALSE,height=8,width=8>>=
-grid.mosaicpairs(UCBAdmissions, type = "conditional", gp = gp.shading, main = NULL)
+##grid.mosaicpairs(UCBAdmissions, type = "conditional", main = NULL)
+grid.mosaicpairs(UCBAdmissions, type = "conditional", permute = FALSE)
 @

### 2003-05-22 trunk/vcd/paper/dsc2003.Rnw
paper for dsc proceedings added (old file dsc2003.tex deleted).
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@260 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
draft for dsc2003 proceedings paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@261 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2003-03-13 trunk/vcd/paper/draft.Rnw
@@
 \documentclass[10pt]{article}
-\usepackage{color, natbib, graphicx}
 \usepackage{dsc2003}
+\usepackage{Sweave, graphicx}
+\usepackage{Z}
 
@@
 
-\SweaveOpts{engine=R,eps=FALSE}
+\SweaveOpts{engine=R}
 \setkeys{Gin}{width=0.8\textwidth}
@@
 
-\title{Visualizing Independence Using Extended Association and Mosaic Plots}
+\title{Visualizing Independence Using Extended Association Plots}
 \author{\hfill David Meyer$^\dag$ \hfill Achim Zeileis$^\dag$ \hfill Kurt Hornik$^\ddag$ \hfill \hfill \\[0.2cm]
@@
 \begin{abstract}
-Two visualization techniques for the independence problem in 2-way
-contingency tables---assocation and mosaic plots---are extended in two directions:
+Association plots---a visualization technique for the independence problem in 2-way
+contingency tables---are extended in three directions:
 \begin{enumerate}
-\item The visualization is enhanced by displaying the significance of an appropriate
-test for independence and by using improved color schemes.
+ \item The visualization is enhanced by using colors for the
+  importance of the residuals.
+ \item Two extensions for the use of multi-way tables are proposed.
  \item The implementation in the \textsf{R} system is improved using a more modular
-design and allowing for more flexible specification of plotting parameters.
+  design and allowing a more flexible specification of plotting parameters.
 \end{enumerate}
-Furthermore, some enhancements for displays of various types of independence in
-general multi-way tables are suggested.
 \end{abstract}
@@
 
-Statistical models built for the analysis of multivariate data
-quickly become complex with increasing dimensionality.
-One idea of visualization techniques is to
-use the human visual system to detect structures in the data that
-possibly are not obvious from solely numeric output (e.g., test
-statistics).
-The \textsf{R} package \texttt{vcd}---inspired by the Book ``Visualizing Categorical
-Data'' \citep{vcd:Friendly:2000}---includes methods for the (mostly
-graphical) exploration of categorical data, such as:
-\begin{itemize}
- \item fitting and graphing of discrete distributions,
- \item plots and tests for independence and symmetry problems,
- \item visualization techniques for log-linear models.
-\end{itemize}
-Here, we focus on the visualization of independence,
-in particular in 2-way tables.
+Given two categorical variables, or, equivalently, their contingency
+table, one is often interested in investigating whether they are 
+independent or not. Usually, the $\chi^2$ test statistic---the sum of the
+squared Pearson residuals---is used as a measure of association.
+Now, whenever the statistic is
+significant, one might ask for a more detailed analysis on the basis
+of the residuals themselves.
+This task can greatly be sustained by the use of graphical
+methods such as the association plot, for which we propose some
+extensions: the use of colors for the shading of the residuals, two
+possible extensions for multi-way tables, and a new, more flexible
+implementation using {\sf R}'s alternative graphical system,
+\texttt{grid}. The new functionality is provided by the {\sf R} package
+\texttt{vcd}, inspired by the book `Visualizing Categorical Data'
+\citep{vcd:Friendly:2000}.
 
-\section{Tests for independence in 2-way tables} \label{sec:tests}
+\section{Association plots for 2-way tables} \label{sec:assocplots}
 
-We consider a 2-way contingency table with cell frequencies $\{n_{ij}\}$ for
+To assess independence of 2 categorical variables, usually
+a 2-way contingency table is considered with cell frequencies $\{n_{ij}\}$ for
 $i = 1, \dots, I$ and $j = 1, \dots, J$ and row and column sums
 $n_{i+} = \sum_i n_{ij}$ and $n_{+j} = \sum_j n_{ij}$ respectively. For
-convenience the number of observations is denoted $n = n_{++}$.
+convenience, the number of observations is denoted $n = n_{++}$.
 
@@
 This is the well-known $\chi^2$ test for independence in 2-way tables. 
-%% reference needed
-
-But this is not the only plausible way of aggregation of the Pearson
-residuals. There are many conceivable functionals $\lambda(\cdot)$ which
-lead to reasonable test statistics $\lambda(\{r_{ij}\})$, the sum of squares
-is just one of them. Another functional suitable for identifying
-the cells which cause the dependence (if any) is the maximum of the absolute
-values
-\begin{equation} \label{eq:MaxAbs}
-M \quad = \quad \max_{i, j} |r_{ij}|.
-\end{equation}
-Given a critical value $c_\alpha$ for this test statistic, all residuals whose
-absolute value exceeds $c_\alpha$ violate the hypothesis of indendence at level
-$\alpha$ \cite[][ch. 7]{vcd:Mazanec+Strasser:2000}. Thus, the interesting cells causing
-the dependence can easily be identified.
-
-Furthermore, the main reason for using the unconditional limiting distribution
-for the $\chi^2$ statistic (\ref{eq:ChiSquare}) was the closed form result
-for the distribution. Recently, with the improving perfomance of computers,
-performing permutation tests---either by simulation or by computation of the
-permutation distribution---became more and more popular. In particular for
-the independence hypothesis (\ref{eq:H0}), using a permutation test is very intuitive
-due to the permutation invariance (given row and column sums) of this problem.
-By employing this approach the permutation distribution of statistics of
-type $\lambda(\{r_{ij}\})$ (also including the $\chi^2$ statistic) can be derived.
-
-Other classical tests for the hypothesis of independence  not fitting in
-the Pearson residual-based framework described above include Fisher's exact test,
-asymptotic tests of the odds ratio (for $2\times2$ tables) and the Mantel-Haenzel
-test.
-
-\section{Visualization techniques for 2-way tables} \label{sec:visualization}
-
-The two best known visualization techniques for independence in 2-way tables are
-association plots and mosaic plots. Both are suitable to bring out departure of an
-observed table $(n_{ij})$ from the expected table $(\hat n_{ij})$ in a graphical way.
-The former focuses on the visualization of the Pearson residuals $r_{ij}$ (under
-independence) while the latter primarily displays the observed frequencies
-$n_{ij}$.
-
-\subsection{Association plots}
+Now, when the $chi^2$ test statistic turns out to be significant for some data, 
+it seems natural to go back to its components, i.e. the  residuals, for a more detailed analysis.
 
@@
 
-Figure \ref{fig:goals-assoc-classic} shows the association plot for the table of
-home and away goals for all 306 games in the 1995/6 season of the German soccer league
-Bundesliga \citep[the data is available in the package \texttt{vcd}]{vcd:Knorr-Held:1999}.
-In particular, it can be seen that there are
-fewer games ending 1--0 and more ending 1--1 than would be expected under independence.
-But this does not yet imply the significance of some test statistic.
+Figure \ref{fig:assoc-classic1} shows the association plot
+(produced by the implementation in {\sf R} `base') for the
+variables `Hair' and `Eye' of the well-known `HairEyeColor' data
+set (Hair and Eye colors of 264 male and 328 female statistics students). 
+The biggest tiles are easy to depict: e.g., the tiles for 
+brown and blue eyes, given blond hair, seem to represent important residuals. 
+But for the others, their classification in important and less important 
+categories seems not obvious. However, with some more information
+visualized, this task could be facilitated.
 
@@
 \begin{center}
-<<goals-assoc-classic,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-grid.assocplot(goals, rot = 0, gp = gp.binary)
+<<assoc-classic1,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
+base::assocplot(t(HairEyeColor[,,2]))
 @
-\caption{Association plot for the Bundesliga data.}
-\label{fig:goals-assoc-classic}
+\caption{Association plot for `Hair' and `Eye' colors of female students.}
+\label{fig:assoc-classic1}
 \end{center}
@@
 
-\subsection{Mosaic plots}
+\section{Improved visualization through residual shading}
 
-Mosaic plots can be seen as an extension of grouped
-bar charts where width and height of the bars show the relative
-frequencies of the two variables: a mosaic plot simply
-consists of a collection of tiles whose
-sizes are proportional to the observed cell frequencies
-(see Figure \ref{fig:goals-mosaic-classic}).
-
-\begin{figure}[htbp]
-\begin{center}
-<<goals-mosaic-classic,fig=TRUE,echo=FALSE,height=6,width=7>>=
-mosaicplot(t(goals), shade = TRUE, main = "")
-@
-\caption{Friendly mosaic plot for the Bundesliga data.}
-\label{fig:goals-mosaic-classic}
-\end{center}
-\end{figure}
-
-Sequential horizontal and vertical recursive splits are used to
-visualize the frequencies of more than two variables, each new variable
-conditional to the previously entered variables. A first extension by
-\cite{vcd:Friendly:1994} uses a color coding of the tiles to visualize deviations
+We propose to use a color-shading similar to Michael Friendly's
+enhancements of the mosaic plot \cite[a graphical tool for the 
+visualization of the observed frequencies in a contingency table; 
+see][]{vcd:Hartigan+Kleiner:1984}:
+this extension \citep{vcd:Friendly:1994} 
+uses a color coding of the tiles to visualize deviations
 (residuals) from a given log-linear model fitted to the table, that is, from
-the expected frequencies under independence. This approach does not only work in 2-way
-tables but also in log-linear models fitted to multi-way tables. In this
-extension, positive and negative signs of the residuals are coded by rectangles with
-solid and dashed borders respectively. Furthermore, residuals exceeding an absolute
+the expected frequencies under some independence hypothesis. 
+Positive and negative signs of the residuals are coded by rectangles with
+solid and dashed borders respectively. Furthermore, residuals can be
+classified according to specified shading levels: per default,
+residuals exceeding an absolute
 value of 2 are shaded light blue and red respectively, those that even
@@
 residuals \emph{individually} significant at approximately the 5\% and 0.01\%
-level. But the main purpose of the shading is not to visualize significance but the \emph{pattern} of
-deviation from independence \citep[p.~109]{vcd:Friendly:2000}. In particular,
-this shading does not provide a visualization of the maximum test (\ref{eq:MaxAbs})
-as in general it is unknown to which significance level $\alpha$ the
-values $c_\alpha = 2$ and $4$ correspond for the table being visualized.
+level. But the main purpose of the shading is not to visualize
+significance but the \emph{pattern} of
+deviation from independence \citep[p.~109]{vcd:Friendly:2000}.
 
-Figure \ref{fig:goals-mosaic-classic} shows the Friendly mosaic plot for
-the same Bundesliga data as in Figure \ref{fig:goals-assoc-classic}.
-In this plot, the residuals
-are visualized by the color shading of the cells. Again, the cells for
-the 1--0 and 1--1 results are highlighted and furthermore, there seem to be
-too many 6--3 results than would be plausible under independence. Whereas
-the association plot in Figure \ref{fig:goals-assoc-classic} contains
-no information about the result of a test for independence, the shading
-in this plot conveys the impression that there might be
-significant dependence (which would indicate that there
-is not only {\em no} home field advantage in the German Bundesliga in 1995/6 but even
-an advantage for the away team). However, it is unclear whether there
-is evidence for this or not.
+Applying this color scheme to association plots similarly facilitates
+the detection of (in)dependence patterns. Consider the `HairEyeColor' example
+introduced in the previous section: Figure \ref{fig:assoc-improved1}
+shows the same residuals than Figure \label{fig:assoc-classic1}, but
+now, the color shading help us to sort the residuals into the 
+three categories: important--less important--unimportant and thus
+emphasizes the associational pattern in the underlying data.
 
-\section{Extensions} \label{sec:extensions}
-
-In this section, we suggest some improvements and extensions which can be
-applied to both visualization techniques introduced in Section
-\ref{sec:visualization}. First, the visual enhancements
-\begin{itemize}
-\item combining visualization of and testing for independence
-\item improved color schemes
-\end{itemize}
-are addressed. Second, the implementation enhancements
-\begin{itemize}
-\item more modular implementation using \texttt{grid}
-\item more flexible plotting parameter specification
-\end{itemize}
-are described. None of these should be thought of as finished work
-but rather as a collection of ideas which still need thorough
-refinement and proper implementation.
-
-\subsection{Visualization enhancements}
-
-The extensions of \cite{vcd:Friendly:1994} to mosaic plots provide
-substantial improvement of the original mosaic plots and enhanced them
-from a plot for contingency tables to a visualization technique for
-log-linear models (and therefore also for certain independence problems).
-However, it has two major drawbacks: First,
-the significance level for the hard-coded critical values 2 and 4 is
-usually unknown. Second, the colors for (light) blue and red
-used in the Michael Friendly's \textsf{SAS} implementation and also in the base implementation
-in \textsf{R} are not device-independent and
-do not provide homogeneous saturations over different colors
-and copier proofness. We suggest some ways to overcome these drawbacks.
-
-As pointed out in Section \ref{sec:tests}, the critical values for
-the maximum statistic $M$ from (\ref{eq:MaxAbs}) can be derived
-from the permutation distribution. Instead of the hard-coded
-values 2 and 4, the particular critical values, e.g., at the levels
-$\alpha = 0.1$ and $0.01$, for the table to be visualized could be
-used. Hence, exactly those residuals causing the (potential)
-dependency within the table are highlighted. At the moment, these critical values
-are derived in \texttt{vcd} by simulation of the underlying distribution.
-
-The implementation of Friendly mosaic plots in \textsf{SAS}
-uses (by default) colors in the HLS (Hue--Luminance--Saturation) color space
-with hues for blue and red, full saturation, and varying luminance
-for lighter colors. The current \textsf{R} implementation is
-based on colors in the HSV (Hue--Saturation--Value) space where
-decreasing the saturation from 1 towards 0
-%% ---given the hue for blue or red and full saturation---
-generates very similar colors
-as increasing the luminance for HSL colors from 0.5 towards 1.
-As an alternative to those HLS and HSV colors, the device-independent
-HCL (Hue--Chroma--Luminance) color space \citep{vcd:Ihaka:2003} could be
-used which also provides homogeneous saturations over different colors and copier
-proofness. The resulting association plot (using critical values for
-$\alpha = 0.1$ and $0.01$) for the Bundesliga data can be seen in
-Figure \ref{fig:goals-assoc-signif}. It indicates very clearly that
-there is no evidence for rejecting the hypothesis of independence
-in this table. Thus, there seems to be neither an advantage for
-the home team nor for the away team.
-
-\setkeys{Gin}{width=0.8\textwidth}
+\setkeys{Gin}{width=0.7\textwidth}
 \begin{figure}[htbp]
 \begin{center}
-<<goals-assoc-signif,fig=TRUE,echo=FALSE,height=6,width=6>>=
-grid.assocplot(goals, rot = 0)
+<<assoc-improved1,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
+grid.assocplot(HairEyeColor[,,2], gp = gp.shading)
 @
-\caption{Extended association plots for the Bundesliga data.}
-\label{fig:goals-assoc-signif}
+\caption{Extended association plot for `Hair' and `Eye' colors of female students.}
+\label{fig:assoc-improved1}
 \end{center}
@@
 
-To illustrate what a `postive' example looks like we give the association plot
-for the well-known Hair-Eye-Color data set (avilable in base \textsf{R}) which cross-tabulates the hair and eye
-color and sex of 592 statistics students: Figure \ref{fig:haireye-assoc-signif}
-shows the association plot for the 328 female students. It is easy to see that
-the hypothesis of independence is rejected at 1\% level by the maximum test (\ref{eq:MaxAbs})
-due to higher frequencies of students with hair and eye color
-black/brown and blond/blue and lower frequencies of
-blond/brown and brown/blue than would be plausible under independence. Furthermore,
-it can also be seen that the residual corresponding to the black/blue
-cell is significant at the 10\% level.
+Another example is given in Figure \ref{fig:assoc-classic2}:
+this time, we cross-tabulate the variables `Eye' and `Sex'.
+The big tiles for brown Eyes suggest a strong asymmetry between
+male and female students, and hence rejection of the independence
+hypothesis. However, the $\chi^2$ test statistic is not significant at
+a 5\% level: in this example, its value is 7.28, and the corresponding
+$P$ value approximately $0.063$. Apparently, the relative size of the
+tiles is insufficient to assess the ``strength'' of association
+between two variables; again, information on the absolute size of the
+residuals would be useful.
+Using the extended shading (Figure \ref{fig:assoc-improved2}), we see at first glance
+that \emph{no} residual is bigger than 2 (respectively smaller than -2 for
+negative ones) because simply no tile is shaded. In fact, as we can depict from
+the legend, the overall range of the residuals is approximately $[-1.5;1.5]$.
+
+\setkeys{Gin}{width=0.7\textwidth}
+\begin{figure}[htbp]
+\begin{center}
+<<assoc-classic2,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
+base::assocplot(margin.table(HairEyeColor, 2:3))
+@
+\caption{Association plot for `Hair' and `Sex' in the `HairEyeColor' data.}
+\label{fig:assoc-classic2}
+\end{center}
+\end{figure}
+
+\setkeys{Gin}{width=0.7\textwidth}
+\begin{figure}[htbp]
+\begin{center}
+<<assoc-improved,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
+grid.assocplot(t(margin.table(HairEyeColor, 2:3)), gp = gp.shading)
+@
+\caption{Extended association plot for `Hair' and `Sex' in the `HairEyeColor' data.}
+\label{fig:assoc-improved2}
+\end{center}
+\end{figure}
+
+\section{Extension to multi-way tables}
+
+Independence problems do not only occur in 2-way tables, but 
+are also important in tables of higher dimensionality where they
+can follow much more complex patterns.
+These are again defined based on the underlying table of theoretical cell probabilities
+$(\pi_{ij\dots})$ with more than two dimensions. Apart from the
+analysis of all pairwise associations, models of interest include,
+e.g., the null hypotheses of conditional independence:
+
+\[\pi_{ijk\dots} = \pi_{i|k\dots}\pi_{j|k\dots}\]
+
+A quick overview of the mutual independence structure of a multi-way table
+can be given by a matrix of plots where each cell contains an
+association plot of the corresponding row and column variables. Such an
+\emph{association pairsplot} allows us to
+quickly scan the association plots for all variable combinations. As
+an example, we again use the HairEyeColor data (see Figure
+\ref{fig:HEC-pairs}): only the cells corresponding to the variables
+`Hair' and `Eye' contain an ``interesting'' association plot with
+colored residuals.
 
@@
 \begin{center}
-<<haireye-assoc-signif,fig=TRUE,echo=FALSE,height=6,width=6>>=
-grid.assocplot(haireye)
+<<UCB-mosaic-signif,fig=TRUE,echo=FALSE,height=8,width=8>>=
+ panel.assocplot <- function (x, i, j, ..., axes, abbreviate) grid.assocplot(margin.table(x, c(i,j)), ..., panel = T, axis.labels = F, labels = F, main = NULL) 
+grid.mosaicpairs(aperm(HairEyeColor), panel.upper= panel.assocplot, panel.lower = panel.assocplot, gp = gp.shading, main = NULL)
 @
-\caption{Extended association plots for the Hair-Eye-Color data (female students).}
-\label{fig:haireye-assoc-signif}
+\caption{Extended association pairs for the `HairEyeColor' data.}
+\label{fig:HEC-pairs}
 \end{center}
@@
 
-We have seen that this approach performs very well when the test statistic
-which should be visualized is the maximum statistic (\ref{eq:MaxAbs}). However,
-it is difficult to extend this approach to general tests for independence,
-in particular in the HCL space, where the ranges of chroma and luminance are
-not independent. In contrast the HSV space has the advantage that all
-three dimensions hue, saturation and value have the range $[0,1]$ and can
-be varied independently. Therefore, a different approach than the one
-described above could be to use the hue for the sign of the residuals as
-before, the saturation for the absolute size of the residuals, and the
-value as an indicator for the significance of some test statistic (only
-using the full value when the overall test for independence rejects
-the null hypothesis). The resulting mosaic plots under this paradigm
-are depicted in Figure \ref{fig:both-mosaic-signif}. Again, it can
-clearly be seen that despite some large residuals there is no
-evidence against independence for the Bundesliga data, but that the
-null hypothesis has to be rejected for the Hair-Eye-Color data.
+To visualize conditional independence, we could use conditioning
+plots: like in trellis displays, we produce a separate plot
+for each level of a conditioning variable (in the case of two
+or more conditioning variables, for each level combination). 
+This can be illustrated using the famous admissions data of the University of
+California at Berkley (UCB) which is available in base \textsf{R}.
+In this data, the question whether there is sex
+discrimination at the UCB leads to the result that although women seem to
+be disadvantaged at the aggregated level there is no sex discrimination
+in the department strata---with the very exception of one department
+in which women are {\em more} likely to be admitted than would be plausible
+under independence. Exactly this is illustrated in the conditioning
+association plot in Figure \ref{fig:UCB-assoc-signif}.
 
@@
 \begin{center}
-<<goals-mosaic-signif,fig=TRUE,echo=FALSE,height=6,width=6>>=
-grid.mosaicplot(t(goals), main = NULL)
+<<UCB-assoc-signif,fig=TRUE,echo=FALSE,height=10,width=10>>=
+tabplot(UCBAdmissions, gp = gp.shading)
 @
-<<haireye-mosaic-signif,fig=TRUE,echo=FALSE,height=6,width=6>>=
-grid.mosaicplot(t(haireye), main = NULL)
-@
-\caption{Extended mosaic plots for the Bundesliga and Hair-Eye-Color data (female students).}
-\label{fig:both-mosaic-signif}
+\caption{Extended conditioning association plots for the `UCB Admissions' data.}
+\label{fig:UCB-assoc-signif}
 \end{center}
@@
 
-\subsection{Implementation enhancements}
+\section{Implementation enhancements}
 
-The current implementation of association and mosaic plots in \textsf{R} suffers
+The current implementation of association plots in \textsf{R} suffers
 from two main disadvantages: First, it is not easy to recycle the
-plots in conditioning plots or pairs plots (like mosaic matrices) as
+plots in conditioning plots or pairs plots as
 they have been implemented using \textsf{R}'s base graphics engine where in
@@
 much more versatility amongst some minor advantages and convenient
-improvements. Second, the graphics parameters of the rectangles in
-association and mosaic plots (like in almost all standard \textsf{R}
+improvements. Second, the graphics parameters of the rectangles in the
+association plots (like in almost all standard \textsf{R}
 plots), such as color and line type, cannot be specified for each cell
-by the user. To overcome this the current implementation in \texttt{vcd}
+by the user. To overcome this, the current implementation in \texttt{vcd}
 allows the user to specify either arrays of graphics parameters
@@
 which computes these graphics parameters based on the original table
-and its Pearson residuals. Functions are provided for the shading schemes
-described in the previous section. Without this feature we would not be
-able to specify the shading schemes described in the previous section
-in such a flexible way.
-
-\section{Multi-way tables}
-
-Independence problems do not only occur in 2-way tables, although that
-is an important special case, but they are also important in tables of
-higher dimensionality and can follow much more complex patterns.
-These are again defined based on the underlying table of theoretical cell probabilities
-$(\pi_{ij\dots})$ with more than two dimensions. Models of interest include the
-null hypotheses of:
-\begin{itemize}
- \item total independence:
-  $\pi_{ijk\dots} = \pi_{i++\dots}\pi_{+j+\dots}\pi_{++k\dots}\dots$
- \item conditional independence:
-  $\pi_{ijk\dots} = \pi_{i|k\dots}\pi_{j|k\dots}$
- \item joint independence:
-  $\pi_{ijk\dots} = \pi_{ij+\dots}\pi_{++k\dots}$
-\end{itemize}
-Classical non-graphical methods for these problems include the $\chi^2$
-test, Fisher's exact test, the Cochran-Mantel-Haenzel test (for $2 \times 2 \times
-K$-tables), and the analysis of log-linear models for more
-complex settings.
-
-As an example, two natural ways to use the visualization techniques described in the previous
-sections would be to use (Trellis-like) conditioning plots or pairs plots
-(like mosaic matrices) to visualize these more complex patterns of
-independence.
-
-Two ideas for the problem of conditional independence are briefly outlined
-here and illustrated using the famous admissions data of the University of
-California at Berkley (UCB) which is available in base \textsf{R}.
-In this data, the question whether there is sex
-discrimination at the UCB leads to the result that although women seem to
-be disadvantaged at the aggregated level there is no sex discrimination
-conditioned on the department---with the very exception of one department
-in which women are {\em more} likely to be admitted than would be plausible
-under independence. Exactly this is illustrated in the conditioning
-assocation plot in Figure \ref{fig:UCB-assoc-signif}.
-
-\begin{figure}[htbp]
-\begin{center}
-<<UCB-assoc-signif,fig=TRUE,echo=FALSE,height=10,width=10>>=
-tabplot(UCBAdmissions)
-@
-\caption{Extended conditioned association plots for the UCB Admissions data.}
-\label{fig:UCB-assoc-signif}
-\end{center}
-\end{figure}
-
-Similarly, the same data can be visualized using a mosaic matrix where
-a conditional independence model is fitted in each plot (see Figure
-\ref{fig:UCB-mosaic-signif}).
-
-\begin{figure}[htbp]
-\begin{center}
-<<UCB-mosaic-signif,fig=TRUE,echo=FALSE,height=8,width=8>>=
-grid.mosaicpairs(UCBAdmissions, type = "conditional", main = NULL)
-@
-\caption{Extended conditioned association plots for the UCB Admissions data.}
-\label{fig:UCB-mosaic-signif}
-\end{center}
-\end{figure}
+and its Pearson residuals. Functions are provided for shading schemes
+like the one described in the previous section.
 
@@
 We suggest a set of enhancements for visualizing the independence
-problem in 2-way tables with some outlook to multi-way tables. The
-extensions aim at improving the visualization by displaying both
-the size of the residuals and the significance of a test for
-independence and by using better color schemes. Furthermore, a new
+problem in 2-way tables using association plots. The
+extensions aim at improving the visualization by displaying
+the size of the residuals, and by extending the concept of association
+plots to multi-way tables. Furthermore, a new
 implementation is outlined based on the graphics package \texttt{grid}
@@
 of graphical parameters.
+
+Our work is still not at its end, though: for example, the plots for
+multi-way tables currently do not use the same scale for all subplots,
+and thus residuals of different subplots are not
+comparable. Furthermore, the current color scheme only allows the
+detection of ``patterns'' of independence. It would be nice, however, 
+that the shading levels were
+chosen such that a true visual test of independence can be performed,
+that is, such that the presence of ``significant'' residuals lead to a
+rejection of the hypothesis of independence. Current work includes the
+development of such color schemes, based on an alternative test
+statistic---the maximum of the absolute values of the pearson residuals.
 

### 2003-05-27 trunk/vcd/paper/dsc2003.Rnw
several fixes
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@262 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 Figure \ref{fig:assoc-classic1} shows the association plot
-(produced by the implementation in {\sf R} `base') for the
+(produced by the implementation in \texttt{base} {\sf R}) for the
 variables `Hair' and `Eye' of the well-known `HairEyeColor' data
 set (Hair and Eye colors of 264 male and 328 female statistics students). 
-The biggest tiles are easy to depict: e.g., the tiles for 
+The biggest tiles are easily made out: e.g., the tiles for 
 brown and blue eyes, given blond hair, seem to represent important residuals. 
 But for the others, their classification in important and less important 
-categories seems not obvious. However, with some more information
-visualized, this task could be facilitated.
+categories seems not obvious.
 
@@
 
+\setkeys{Gin}{width=0.7\textwidth}
+\begin{figure}[htbp]
+\begin{center}
+<<assoc-improved1,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
+grid.assocplot(HairEyeColor[,,2], gp = gp.shading)
+@
+\caption{Extended association plot for `Hair' and `Eye' colors of female students.}
+\label{fig:assoc-improved1}
+\end{center}
+\end{figure}
+
 \section{Improved visualization through residual shading}
 
-We propose to use a color-shading similar to Michael Friendly's
-enhancements of the mosaic plot \cite[a graphical tool for the 
+We propose to use a color-shading similar to
+enhancements of \cite{vcd:Friendly:1994} for the mosaic plot 
+\cite[a graphical tool for the 
 visualization of the observed frequencies in a contingency table; 
 see][]{vcd:Hartigan+Kleiner:1984}:
-this extension \citep{vcd:Friendly:1994} 
-uses a color coding of the tiles to visualize deviations
+this extension 
+uses a color coding of the mosaic tiles to visualize deviations
 (residuals) from a given log-linear model fitted to the table, that is, from
-the expected frequencies under some independence hypothesis. 
+the expected frequencies under independence.
 Positive and negative signs of the residuals are coded by rectangles with
@@
 
-Applying this color scheme to association plots similarly facilitates
-the detection of (in)dependence patterns. Consider the `HairEyeColor' example
+This color scheme similarly facilitates
+the detection of (in)dependence patterns in association plots. 
+Consider the `HairEyeColor' example
 introduced in the previous section: Figure \ref{fig:assoc-improved1}
-shows the same residuals than Figure \label{fig:assoc-classic1}, but
-now, the color shading help us to sort the residuals into the 
+shows the same residuals as Figure \ref{fig:assoc-classic1}, but
+now, the color shading helps us to sort the residuals into the 
 three categories: important--less important--unimportant and thus
@@
 
-\setkeys{Gin}{width=0.7\textwidth}
-\begin{figure}[htbp]
-\begin{center}
-<<assoc-improved1,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-grid.assocplot(HairEyeColor[,,2], gp = gp.shading)
-@
-\caption{Extended association plot for `Hair' and `Eye' colors of female students.}
-\label{fig:assoc-improved1}
-\end{center}
-\end{figure}
-
 Another example is given in Figure \ref{fig:assoc-classic2}:
 this time, we cross-tabulate the variables `Eye' and `Sex'.
-The big tiles for brown Eyes suggest a strong asymmetry between
-male and female students, and hence rejection of the independence
-hypothesis. However, the $\chi^2$ test statistic is not significant at
+The big tiles for brown Eyes suggest convey the impression of an asymmetry between
+male and female students, and hence that hypothesis of independence is
+to be rejected. However, the $\chi^2$ test statistic is not significant at
 a 5\% level: in this example, its value is 7.28, and the corresponding
-$P$ value approximately $0.063$. Apparently, the relative size of the
+$p$ value approximately $0.063$. Apparently, the relative size of the
 tiles is insufficient to assess the ``strength'' of association
@@
 Using the extended shading (Figure \ref{fig:assoc-improved2}), we see at first glance
-that \emph{no} residual is bigger than 2 (respectively smaller than -2 for
-negative ones) because simply no tile is shaded. In fact, as we can depict from
-the legend, the overall range of the residuals is approximately $[-1.5;1.5]$.
+that \emph{no} residual is bigger than 2 (or smaller than -2 for
+negative ones, respectively) 
+because simply no tile is shaded. In fact, as we can depict from
+the legend, the overall range of the residuals is approximately $[-1.5,1.5]$.
 
@@
 @
-\caption{Association plot for `Hair' and `Sex' in the `HairEyeColor' data.}
+\caption{Association plot for `Eye' and `Sex' in the `HairEyeColor' data.}
 \label{fig:assoc-classic2}
@@
 @
-\caption{Extended association plot for `Hair' and `Sex' in the `HairEyeColor' data.}
+\caption{Extended association plot for `Eye' and `Sex' in the `HairEyeColor' data.}
 \label{fig:assoc-improved2}
@@
 
-A quick overview of the mutual independence structure of a multi-way table
+A quick overview of the mutual independence structure of a multi-way
+table (i.e., all pairwise \textit{marginal} associations)
 can be given by a matrix of plots where each cell contains an
@@
 @
-\caption{Extended association pairs for the `HairEyeColor' data.}
+\caption{Extended association pairs for the `HairEyeColor' data,
+  visualizing all pairwise marginal associations.}
 \label{fig:HEC-pairs}
@@
 
-To visualize conditional independence, we could use conditioning
+To visualize \textit{conditional} independence, we could use conditioning
 plots: like in trellis displays, we produce a separate plot
@@
 and its Pearson residuals. Functions are provided for shading schemes
-like the one described in the previous section.
+like those described in the previous sections.
 

### 2003-05-28 trunk/vcd/paper/dsc2003.Rnw
small fixes.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@263 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 This is the well-known $\chi^2$ test for independence in 2-way tables. 
-Now, when the $chi^2$ test statistic turns out to be significant for some data, 
+Now, when the $\chi^2$ test statistic turns out to be significant for some data, 
 it seems natural to go back to its components, i.e. the  residuals, for a more detailed analysis.
@@
 this time, we cross-tabulate the variables `Eye' and `Sex'.
-The big tiles for brown Eyes suggest convey the impression of an asymmetry between
+For example, the big tiles for blond hair convey the impression of an asymmetry between
 male and female students, and hence that hypothesis of independence is
 to be rejected. However, the $\chi^2$ test statistic is not significant at
-a 5\% level: in this example, its value is 7.28, and the corresponding
-$p$ value approximately $0.063$. Apparently, the relative size of the
+a 5\% level: in this example, its value is 6.221, and the corresponding
+$p$ value approximately $0.1013$. Apparently, the relative size of the
 tiles is insufficient to assess the ``strength'' of association
@@
 because simply no tile is shaded. In fact, as we can depict from
-the legend, the overall range of the residuals is approximately $[-1.5,1.5]$.
+the legend, the overall range of the residuals is smaller than $[-1.5,1.5]$.
 
@@
 <<assoc-classic2,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-base::assocplot(margin.table(HairEyeColor, 2:3))
+base::assocplot(margin.table(HairEyeColor, c(1,3)))
 @
-\caption{Association plot for `Eye' and `Sex' in the `HairEyeColor' data.}
+\caption{Association plot for `Hair' and `Sex' in the `HairEyeColor' data.}
 \label{fig:assoc-classic2}
@@
 <<assoc-improved,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-grid.assocplot(t(margin.table(HairEyeColor, 2:3)), gp = gp.shading)
+grid.assocplot(t(margin.table(HairEyeColor, c(1,3))), gp = gp.shading)
 @
-\caption{Extended association plot for `Eye' and `Sex' in the `HairEyeColor' data.}
+\caption{Extended association plot for `Hair' and `Sex' in the `HairEyeColor' data.}
 \label{fig:assoc-improved2}
@@
 
+A general approach for visualizing such kind of hypotheses are
+extended mosaic plots. In this section, we give two examples for the
+applicability of association plots to multi-way tables.
 A quick overview of the mutual independence structure of a multi-way
@@
 improvements. Second, the graphics parameters of the rectangles in the
-association plots (like in almost all standard \textsf{R}
-plots), such as color and line type, cannot be specified for each cell
+association plots, such as color and line type, cannot be specified for each cell
 by the user. To overcome this, the current implementation in \texttt{vcd}
@@
 of graphical parameters.
-
 Our work is still not at its end, though: for example, the plots for

### 2003-06-10 trunk/vcd/paper/dsc2003.Rnw
editorial changes.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@266 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
  \item Two extensions for the use of multi-way tables are proposed.
- \item The implementation in the \textsf{R} system is improved using a more modular
+ \item The implementation currently available in \textsf{R} is improved using a more modular
   design and allowing a more flexible specification of plotting parameters.
@@
 Given two categorical variables, or, equivalently, their contingency
-table, one is often interested in investigating whether they are 
+table, one is often interested in whether they are 
 independent or not. Usually, the $\chi^2$ test statistic---the sum of the
@@
 implementation using {\sf R}'s alternative graphical system,
-\texttt{grid}. The new functionality is provided by the {\sf R} package
+\texttt{grid} \cite{vcd:Murrell:2002}. The new functionality is provided by the {\sf R} package
 \texttt{vcd}, inspired by the book `Visualizing Categorical Data'
@@
 $i = 1, \dots, I$ and $j = 1, \dots, J$ and row and column sums
-$n_{i+} = \sum_i n_{ij}$ and $n_{+j} = \sum_j n_{ij}$ respectively. For
+$n_{i+} = \sum_i n_{ij}$ and $n_{+j} = \sum_j n_{ij}$, respectively. For
 convenience, the number of observations is denoted $n = n_{++}$.
@@
 zero. The most convenient way to aggregate the $I \times J$ residuals
-to one test statistic is their squared sum
+into one test statistic is their squared sum
 \begin{equation} \label{eq:ChiSquare}
@@
 \end{equation}
-because this is known to have a limiting $\chi^2$ distribution with $(I-1)(J-1)$
+which is known to have a limiting $\chi^2$ distribution with $(I-1)(J-1)$
 degrees of freedom under the null hypothesis.
@@
 Now, when the $\chi^2$ test statistic turns out to be significant for some data, 
-it seems natural to go back to its components, i.e. the  residuals, for a more detailed analysis.
+it seems natural to go back to its components, i.e., the  residuals, for a more detailed analysis.
 
@@
 square root of the expected counts $\sqrt{\hat n_{ij}}$. Thus, the area is proportional to
-the raw residuals $n_{ij} - \hat n_{ij}$. The sign of the residual is redundantly
+the raw residuals $n_{ij} - \hat n_{ij}$. In available
+implementations, the sign of the residual is typically redundantly
 coded by the rectangle's color and its position relative to the baseline.
@@
 variables `Hair' and `Eye' of the well-known `HairEyeColor' data
-set (Hair and Eye colors of 264 male and 328 female statistics students). 
+set \cite[Hair and Eye colors of 264 male and 328 female statistics
+students; see][]{vcd:Snee:1974}. 
 The biggest tiles are easily made out: e.g., the tiles for 
 brown and blue eyes, given blond hair, seem to represent important residuals. 
-But for the others, their classification in important and less important 
-categories seems not obvious.
+But assessing the ``importance'' of the others is far from obvious.
 
@@
 \end{figure}
-
+\pagebreak
 \section{Improved visualization through residual shading}
@@
 uses a color coding of the mosaic tiles to visualize deviations
-(residuals) from a given log-linear model fitted to the table, that is, from
-the expected frequencies under independence.
+(residuals) from a given log-linear model fitted to the table.
 Positive and negative signs of the residuals are coded by rectangles with
@@
 Using the extended shading (Figure \ref{fig:assoc-improved2}), we see at first glance
-that \emph{no} residual is bigger than 2 (or smaller than -2 for
+that \emph{no} residual is bigger than 2 (or smaller than $-2$ for
 negative ones, respectively) 
-because simply no tile is shaded. In fact, as we can depict from
+because simply no tile is shaded. In fact, as we can see from
 the legend, the overall range of the residuals is smaller than $[-1.5,1.5]$.
@@
 analysis of all pairwise associations, models of interest include,
-e.g., the null hypotheses of conditional independence:
-
-\[\pi_{ijk\dots} = \pi_{i|k\dots}\pi_{j|k\dots}\]
-
+e.g., the null hypotheses of conditional independence (for example the
+model: $\pi_{ijk} = \pi_{i|k}\pi_{j|k}$).
 A general approach for visualizing such kind of hypotheses are
@@
 association plot of the corresponding row and column variables. Such an
-\emph{association pairsplot} allows us to
+\emph{association pairs plot} allows us to
 quickly scan the association plots for all variable combinations. As
@@
 This can be illustrated using the famous admissions data of the University of
-California at Berkley (UCB) which is available in base \textsf{R}.
-In this data, the question whether there is sex
+California at Berkley (UCB)---see \cite{vcd:Bickel+Hammel+O'connell:1975}---which is available in base \textsf{R}.
+In this data, the question whether there is gender
 discrimination at the UCB leads to the result that although women seem to
-be disadvantaged at the aggregated level there is no sex discrimination
+be disadvantaged at the aggregated level there is no gender discrimination
 in the department strata---with the very exception of one department
@@
 We suggest a set of enhancements for visualizing the independence
-problem in 2-way tables using association plots. The
+problem in contingency tables using association plots. The
 extensions aim at improving the visualization by displaying
@@
 detection of ``patterns'' of independence. It would be nice, however, 
-that the shading levels were
-chosen such that a true visual test of independence can be performed,
-that is, such that the presence of ``significant'' residuals lead to a
+if the shading levels were
+chosen in a way that a true visual test of independence can be performed,
+that is, that the presence of ``significant'' residuals leads to a
 rejection of the hypothesis of independence. Current work includes the
 development of such color schemes, based on an alternative test
-statistic---the maximum of the absolute values of the pearson residuals.
+statistic---the maximum of the absolute values of the Pearson residuals.
 

### 2003-06-10 trunk/vcd/paper/dsc2003.Rnw
more editorial changes.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@267 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
   importance of the residuals.
- \item Two extensions for the use of multi-way tables are proposed.
+ \item Two methods for the use of multi-way tables are proposed.
  \item The implementation currently available in \textsf{R} is improved using a more modular
@@
 extensions: the use of colors for the shading of the residuals, two
-possible extensions for multi-way tables, and a new, more flexible
-implementation using {\sf R}'s alternative graphical system,
-\texttt{grid} \cite{vcd:Murrell:2002}. The new functionality is provided by the {\sf R} package
+possible methods for multi-way tables, and a new, more flexible
+implementation using {\sf R}'s alternative graphical system
+\texttt{grid} \citep{vcd:Murrell:2002}. The new functionality is provided 
+by the {\sf R} package
 \texttt{vcd}, inspired by the book `Visualizing Categorical Data'
@@
 Figure \ref{fig:assoc-classic1} shows the association plot
-(produced by the implementation in \texttt{base} {\sf R}) for the
+(produced by the implementation in base {\sf R}) for the
 variables `Hair' and `Eye' of the well-known `HairEyeColor' data
@@
 
-We propose to use a color-shading similar to
+We propose to use a color-shading similar to the
 enhancements of \cite{vcd:Friendly:1994} for the mosaic plot 
@@
 a 5\% level: in this example, its value is 6.221, and the corresponding
-$p$ value approximately $0.1013$. Apparently, the relative size of the
+$p$ value is approximately $0.101$. Apparently, the relative size of the
 tiles is insufficient to assess the ``strength'' of association
@@
 This can be illustrated using the famous admissions data of the University of
-California at Berkley (UCB)---see \cite{vcd:Bickel+Hammel+O'connell:1975}---which is available in base \textsf{R}.
+California at Berkley \citep{vcd:Bickel+Hammel+O'connell:1975} 
+which is available in base \textsf{R}.
 In this data, the question whether there is gender
-discrimination at the UCB leads to the result that although women seem to
+discrimination at the university leads to the result that although women seem to
 be disadvantaged at the aggregated level there is no gender discrimination
@@
 
+\section*{Addresses}
+
+David Meyer\\
+Achim Zeileis\\
+E-mail: \texttt{\{David.Meyer,Achim.Zeileis\}@ci.tuwien.ac.at}\\
+Institut f\"ur Statistik und Wahrscheinlichkeitstheorie\\
+Technische Universit\"at Wien\\
+Wiedner Hauptstra{\ss}e 8--10/1071\\
+1040 Wien\\
+\\
+Kurt Hornik\\
+E-mail: \texttt{Kurt.Hornik@wu-wien.ac.at}\\
+Institut f\"ur Statistik\\
+Wirtschaftsuniversit\"at Wien\\
+Augasse 2--6\\
+1090 Wien\\
+
 \end{document}

### 2003-07-08 trunk/vcd/paper/draft.Rnw
departures
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@268 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 The two best known visualization techniques for independence in 2-way tables are
-association plots and mosaic plots. Both are suitable to bring out departure of an
-observed table $(n_{ij})$ from the expected table $(\hat n_{ij})$ in a graphical way.
+association plots and mosaic plots. Both are suitable to bring out
+departures of an observed table $(n_{ij})$ from the expected table $(\hat n_{ij})$ in a graphical way.
 The former focuses on the visualization of the Pearson residuals $r_{ij}$ (under

### 2003-07-08 trunk/vcd/R/grid.mosaicplot.R
incorporated changes for Reisensburg 2003
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@270 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -13,7 +13,7 @@ panel.barplot <- function(x, color = "blue", fontsize = 20, dimnames, ...) {
   pop.viewport(1)
   grid.text(names(dimnames(x)), y = 1, just = c("center", "top"),
             gp = gpar(fontsize = fontsize))
-  
+
 }
 
 panel.mosaicplot <- function(x, i, j, type, legend = FALSE, axes = TRUE,
@@ -49,7 +49,7 @@ grid.mosaicpairs <- function(x, main = NULL,
                              type.upper = NULL,
                              type.lower = NULL,
                              newpage = TRUE,
-                             
+
                              space = 0.1,
                              legend = FALSE,
                              axes = FALSE,
@@ -74,7 +74,7 @@ grid.mosaicpairs <- function(x, main = NULL,
   else
     match.arg(type.lower, type)
   if (permute) x <- aperm(x)
-  
+
   d <- length(dim(x))
   l <- grid.layout(d, d)
   push.viewport(viewport(width = unit(1, "snpc"), height = unit(1, "snpc")))
@@ -83,7 +83,7 @@ grid.mosaicpairs <- function(x, main = NULL,
                          y = 0, just = "bottom"))
   grid.text(main, y = unit(1.05, "npc"), gp = gpar(fontsize = 20))
 
-  
+
   for (i in 1:d)
     for(j in 1:d) {
       push.viewport(viewport(layout.pos.col = i, layout.pos.row = j))
@@ -95,7 +95,7 @@ grid.mosaicpairs <- function(x, main = NULL,
       else if (i < j)
         panel.lower(x, i, j, type.lower, legend = legend, axes = axes,
                     margins = panel.margins, abbreviate = abbreviate, gp = gp, ...)
-      else 
+      else
         panel.diag(margin.table(x, i), fontsize = diag.fontsize,
                    dimnames = diag.dimnames, ...)
 
@@ -141,13 +141,13 @@ legend.block <- function(res,
                          gp = gp.shading,
                          fontsize = 12,
                          panel = FALSE,
-                         space = 0.6, 
+                         space = 2,        #Z# was: from left in "native", now: from right in "lines"
                          text = "Pearson\nresiduals:") {
-  
+
   push.viewport(viewport(layout.pos.row = 1, layout.pos.col = 2))
[…]

### 2003-07-08 trunk/vcd/R/grid.assocplot.R
incorporated changes for Reisensburg 2003
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@270 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -5,20 +5,15 @@ grid.assocplot <- function(x, ...)
 
 grid.assocplot.default <- function(x, xlab = NULL, ylab = NULL, main = NULL,
                            model.formula = NULL, labels = TRUE, fontsize = 12,
-			   gp = gp.signif, legend = TRUE, col.bins = NULL,
+                           gp = gp.max, legend = TRUE,
                            scale = 0.4, rot = 90, check.overlap = TRUE, axis.labels = TRUE,
-			   show.grid = FALSE, panel = FALSE, margins = c(2, 4, 5, 2))
+                           show.grid = FALSE, panel = FALSE, margins = c(1, 1, 1, 1))
 {
-  require(grid)
-
-  ## This could be used to fit more complex models:
-  ## require(MASS)
-  ## if(is.null(model.formula))
-  ##   model.formula <- as.formula(paste("~ ", paste(names(dimnames(x)), collapse = " + "), sep = " "))
-  ## x.fit <- loglm(model.formula, data = x, fit = TRUE)
-  ## expctd <- fitted(x.fit)
-  ## res <- residuals(x.fit, type = "pearson")
-
+  require(grid) #Z# require globally?
+  if (length(dim(x)) != 2)
+      stop("x must be a 2-d contingency table")
+  if (any(x < 0) || any(is.na(x)))
+      stop("all entries of x must be nonnegative and finite")
 
   ## independence model fitting
   rowTotals <- rowSums(x)
@@ -26,7 +21,6 @@ grid.assocplot.default <- function(x, xlab = NULL, ylab = NULL, main = NULL,
   expctd <- rowTotals %o% colTotals / sum(colTotals)
   sexpctd <- sqrt(expctd)
   res <- (x - expctd) / sexpctd
-  ##res[is.na(res)] <- 0
 
   if(is.null(ylab)) ylab <- names(dimnames(x))[1]
   if(is.null(xlab)) xlab <- names(dimnames(x))[2]
@@ -46,29 +40,45 @@ grid.assocplot.default <- function(x, xlab = NULL, ylab = NULL, main = NULL,
   y.pos.matrix <- matrix(rep(rev(y.pos), ncols), ncol = ncols)
   cell.height <- res
   cell.width <- sexpctd
+
   if(is.function(gp)) {
     gpfun <- gp
     gp <- gpfun(x, res)
-  } else legend <- FALSE
+  } else {
[…]

### 2003-07-08 trunk/vcd/R/hcl.R

@@ -10,7 +10,7 @@
 #
 #  These values define a point in CIELUV using polar coordinates.
 #
-#    hue       - The angle to the color in degrees. 
+#    hue       - The angle to the color in degrees.
 #                Roughly speaking red=0, 120=green, 240=blue
 #
 #    chroma    - The ``colorfulness'' of the color (rather like
@@ -32,74 +32,72 @@
 #
 #    verbose   - give warnings if values are out of gamut
 
-hcl =
-  function(hue, chroma = 35, luminance = 85, correct = FALSE, gamma = 2.2, verbose = TRUE)
-  {
-    HEXDIGITS = c("0", "1", "2", "3", "4", "5", "6", "7",
-                  "8", "9", "A", "B", "C", "D", "E", "F")
-    gtrans =
-      function(u, gamma = 2.4)
-      ifelse(u > 0.00304, 1.055 * u^(1 / gamma) - 0.055, 12.92 * u)
+hcl <- function(hue, chroma = 35, luminance = 85, correct = FALSE, gamma = 2.2, verbose = TRUE)
+{
+  HEXDIGITS = c("0", "1", "2", "3", "4", "5", "6", "7",
+                "8", "9", "A", "B", "C", "D", "E", "F")
+  gtrans =
+    function(u, gamma = 2.4)
+    ifelse(u > 0.00304, 1.055 * u^(1 / gamma) - 0.055, 12.92 * u)
 
-    #  Assume a D65 whitepoint with luminance 100.
-    #  Ultimately, this should be a parameter.
-    #  These are the CIE XYZ values.
+  #  Assume a D65 whitepoint with luminance 100.
+  #  Ultimately, this should be a parameter.
+  #  These are the CIE XYZ values.
 
-    XN =  95.047
-    YN = 100.000
-    ZN = 108.883
+  XN =  95.047
+  YN = 100.000
+  ZN = 108.883
 
-    #  uN and vN are the corresponding LUV chromaticities
+  #  uN and vN are the corresponding LUV chromaticities
 
-    tmp = XN + YN + ZN
-    xN = XN / tmp
-    yN = YN / tmp
-    uN = 2 * xN /(6 * yN - xN + 1.5)
-    vN = 4.5 * yN / (6 * yN - xN + 1.5)
+  tmp = XN + YN + ZN
+  xN = XN / tmp
[…]

### 2003-07-08 trunk/vcd/R/shadings.R

@@ -1,0 +1,165 @@
+gp.HSVshading <- function(observed, residuals,
+                          hue = c(2/3, 0), saturation = c(1, 0), value = c(1, 0.5),
+                          interpolate = c(2, 4), lty = 1:2,
+                          test = chisq.test, level = 0.95)
+{
+  res <- as.vector(residuals)
+
+  my.h <- rep(hue, length.out = 2)        ## positive and negative hue
+  my.s <- rep(saturation, length.out = 2) ## maximum and minimum saturation
+  my.v <- rep(value, length.out = 2)      ## significant and non-significant value
+  lty <- rep(lty, length.out = 2)         ## positive and negative lty
+
+  if(is.function(interpolate)) {
+    col.bins <- min(res) + diff(range(res)) * ((0:200)/200)
+  } else {
+    cb <- col.bins <- sort(interpolate)
+    require(stepfun)  #Z# require globally?
+    interpolate <- function(x) stepfun(cb,  seq(my.s[2], my.s[1], length = length(cb) + 1))(abs(x))
+  }
+
+  hue <- ifelse(res > 0, my.h[1], my.h[2])
+  saturation <- pmax(pmin(interpolate(res), 1), 0)
+  if(is.null(test)) {
+    value <- my.v[1]
+    p.value <- NULL
+  } else {
+    p.value <- test(observed)$p.value
+    value <- ifelse(p.value < (1-level), my.v[1], my.v[2])
+  }
+
+  col <- hsv(hue, saturation, value)
+  dim(col) <- dim(residuals)
+
+  ## col.bins
+  col.bins <- sort(c(range(res), col.bins, -col.bins, 0))
+  col.bins <- col.bins[col.bins <= max(res) & col.bins >= min(res)]
+
+  ## col.legend
+  y.pos <- col.bins[-length(col.bins)]
+  y.height <- diff(col.bins)
+  res2 <- y.pos + 0.5*y.height
+  hue2 <- ifelse(res2 > 0, my.h[1], my.h[2])
+  saturation2 <- pmax(pmin(interpolate(res2), 1), 0)
+  y.col <- hsv(hue2, saturation2, value)
+  legend.col <- list(pos = y.pos, height = y.height, col = y.col)
+
+  ## lty and lty.bins
+  lty <- ifelse(residuals > 0, lty[1], lty[2])
+  lty.bins <- sort(c(range(res), 0))
+
[…]

### 2003-07-08 trunk/vcd/R/tabplot.R

@@ -1,0 +1,92 @@
+tabplot <- function(x, panel = function(x, ...) grid.assocplot(x, panel = TRUE, legend = FALSE, ...),
+                    margins = rep(1,4), ...)
+{
+  UseMethod("tabplot")
+}
+
+tabplot.formula <- function(formula, panel = function(x, ...) grid.assocplot(x, panel = TRUE, legend = FALSE, ...),
+                    margins = rep(1,4), data, ...)
+{
+  formula <- as.character(formula)
+  formula <- paste(paste(c("Freq", formula[1:2]), collapse = " "), formula[3], sep = " + ")
+  formula <- gsub("\\|", "+", formula)
+  formula <- gsub("\\*", "+", formula)
+  x <- xtabs(as.formula(formula), data = as.data.frame(data))
+  tabplot(x, panel = panel, margins = margins, ...)
+}
+
+
+tabplot.default <- function(x, panel = function(x, ...) grid.assocplot(x, panel = TRUE, legend = FALSE, axis.labels = FALSE, fontsize = fontsize, ...),
+                    margins = rep(1, 4), fontsize = 12, ...)
+{
+  grid.newpage()
+
+  if(length(dim(x)) <= 2) panel(x, ...) ## no conditioning variables
+  else {
+
+  axis.names <- names(dimnames(x))[1:2]
+  condition <- dimnames(x)[-(1:2)]
+  condvars <- names(condition)
+  ncond <- length(condvars)
+  nlevels <- sapply(condition, length)
+  nplots <- prod(nlevels)
+  condition <- as.matrix(sapply(expand.grid(condition), as.character))
+
+  ## compute layout
+  layout <- c(1,1,1) ## rows, cols, pages
+  if(ncond == 1) {
+    layout[2] <- ceiling(sqrt(floor(nlevels)))
+    layout[1] <- ceiling(nlevels/layout[2])
+    layout <- expand.grid(lapply(layout, function(x) 1:x))[1:nplots,]
+  }
+  else {
+    layout[1] <- nlevels[1]
+    layout[2] <- nlevels[2]
+    if(ncond > 3) layout[3] <- nplots/prod(nlevels[1:2])
+    if(layout[3] > 1) stop("multiple pages not supported yet")
+    layout <- expand.grid(lapply(layout, function(x) 1:x))
+  }
+
+  nr <- max(layout[,1])
+  nc <- max(layout[,2])
[…]

### 2003-07-08 trunk/vcd/paper/dsc2003.Rnw
fixes for dsc-papers
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@271 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 visualization of the observed frequencies in a contingency table; 
-see][]{vcd:Hartigan+Kleiner:1984}:
-this extension 
-uses a color coding of the mosaic tiles to visualize deviations
+see][]{vcd:Hartigan+Kleiner:1981,vcd:Hartigan+Kleiner:1984}:
+this extension uses a color coding of the mosaic tiles to visualize deviations
 (residuals) from a given log-linear model fitted to the table.
@@
 significance but the \emph{pattern} of
-deviation from independence \citep[p.~109]{vcd:Friendly:2000}.
+deviation from independence \citep[p.~109]{vcd:Friendly:2000}. A different
+approach for visualizing the residuals of log-linear models is
+used by \cite{vcd:Theus+Laurer:1999} who propose to use relative
+residuals, i.e., to scale all residuals according to the absolute sum
+of residuals. This is necessary when shading
+cells in a mosaic plot for expected frequencies, but offers no
+advantages when the residuals themselves are visualized.
 

### 2003-07-08 trunk/vcd/paper/dsc2003.Rnw
made changes necessary for the package, vignette and papers to pass R CMD check; vignette should probably be removed for release
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@273 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \usepackage{dsc2003}
-\usepackage{Sweave, graphicx}
 \usepackage{Z}
@@
 <<assoc-improved1,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-grid.assocplot(HairEyeColor[,,2], gp = gp.shading)
+grid.assocplot(HairEyeColor[,,2], gp = gp.Friendly)
 @
@@
 <<assoc-improved,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-grid.assocplot(t(margin.table(HairEyeColor, c(1,3))), gp = gp.shading)
+grid.assocplot(t(margin.table(HairEyeColor, c(1,3))), gp = gp.Friendly)
 @
@@
  panel.assocplot <- function (x, i, j, ..., axes, abbreviate) grid.assocplot(margin.table(x, c(i,j)), ..., panel = T, axis.labels = F, labels = F, main = NULL)
-grid.mosaicpairs(aperm(HairEyeColor), panel.upper= panel.assocplot, panel.lower = panel.assocplot, gp = gp.shading, main = NULL)
+grid.mosaicpairs(aperm(HairEyeColor), panel.upper= panel.assocplot, panel.lower = panel.assocplot, gp = gp.Friendly, main = NULL)
 @
@@
 <<UCB-assoc-signif,fig=TRUE,echo=FALSE,height=10,width=10>>=
-tabplot(UCBAdmissions, gp = gp.shading)
+tabplot(UCBAdmissions, gp = gp.Friendly)
 @

### 2003-07-08 trunk/vcd/R/grid.mosaicplot.R

@@ -18,7 +18,7 @@ panel.barplot <- function(x, color = "blue", fontsize = 20, dimnames, ...) {
 
 panel.mosaicplot <- function(x, i, j, type, legend = FALSE, axes = TRUE,
                              margins = c(0, 0, 0, 0), abbreviate = FALSE,
-                             gp = gp.shading, ...) {
+                             gp = gp.Friendly, ...) {
   index <- 1:length(dim(x))
   rest <- index[!index %in% c(i, j)]
   grid.mosaicplot(x = margin.table(x,
@@ -58,7 +58,7 @@ grid.mosaicpairs <- function(x, main = NULL,
                              panel.margins = c(0, 0, 0, 0),
                              diag.fontsize = 20,
                              diag.dimnames = TRUE,
-                             gp = gp.shading,
+                             gp = gp.Friendly,
                              permute = TRUE,
                              ...)
 {
@@ -138,7 +138,7 @@ function(formula, data = NULL, ...,
 }
 
 legend.block <- function(res,
-                         gp = gp.shading,
+                         gp = gp.Friendly,
                          fontsize = 12,
                          panel = FALSE,
                          space = 2,        #Z# was: from left in "native", now: from right in "lines"
@@ -196,7 +196,7 @@ grid.mosaicplot.default <-
            type = c("Pearson", "deviance", "FT"),
            shade = TRUE,
            legend = TRUE,
-           gp = gp.shading,
+           gp = gp.Friendly,
            fontsize = 12,
 
            margins = c(1, 1, 1, 1),

### 2003-07-08 trunk/vcd/R/shadings.R

@@ -127,7 +127,7 @@ gp.HCLshading <- function(observed, residuals,
 gp.Friendly <- function(observed, residuals,
                         hue = c(2/3, 0), lty = 1:2, interpolate = c(2, 4))
 {
-  gp.HSVshading(observed, residuals, hue = hue, value = 1, interpolate = col.bins, test = NULL)
+  gp.HSVshading(observed, residuals, hue = hue, value = 1, interpolate = interpolate, test = NULL)
 }
 
 gp.max <- function(observed, residuals,

### 2003-07-08 trunk/vcd/paper/draft.Rnw
made changes necessary for the package, vignette and papers to pass R CMD check; vignette should probably be removed for release
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@273 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \documentclass[10pt]{article}
-\usepackage{Sweave, graphicx}
 \usepackage{Z}
@@
 <<goals-mosaic-signif,fig=TRUE,echo=FALSE,height=6,width=6>>=
-grid.mosaicplot(t(goals), main = NULL)
+grid.mosaicplot(goals, main = NULL, gp = gp.HCLshading)
 @
 <<haireye-mosaic-signif,fig=TRUE,echo=FALSE,height=6,width=6>>=
-grid.mosaicplot(t(haireye), main = NULL)
+grid.mosaicplot(haireye, main = NULL, gp = gp.HCLshading)
 @
@@
 ##grid.mosaicpairs(UCBAdmissions, type = "conditional", main = NULL)
-grid.mosaicpairs(UCBAdmissions, type = "conditional", permute = FALSE)
+grid.mosaicpairs(UCBAdmissions, type = "conditional", permute = FALSE, gp = gp.HCLshading)
 @

### 2003-07-08 trunk/vcd/R/grid.fourfoldplot.R

@@ -35,6 +35,8 @@ function(x, color = c("#99CCFF","#6699CC","#FF5050","#6060A0", "#FF0000", "#0000
     ##   in each step by shifting the origin so that we can always plot
     ##   as detailed above.
 
+    require(grid)
+
     if(!is.array(x))
         stop("x must be an array")
     if(length(dim(x)) == 2) {

### 2003-07-08 trunk/vcd/demo/mosaic.R

@@ -138,7 +138,7 @@ grid.mosaicpairs(UCBAdmissions)
 
 grid.mosaicpairs(UCBAdmissions, type="conditional")
 
-grid.mosaicpairs(UCBAdmissions, type="pairwise", gp = gp.signif)
+grid.mosaicpairs(UCBAdmissions, type="pairwise", gp = gp.max)
 
 
 

### 2003-07-08 trunk/vcd/R/shadings.R
fixed binary shading
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@274 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -35,7 +35,7 @@ gp.HSVshading <- function(observed, residuals,
   col.bins <- sort(c(range(res), col.bins, -col.bins, 0))
   col.bins <- col.bins[col.bins <= max(res) & col.bins >= min(res)]
 
-  ## col.legend
+  ## legend.col
   y.pos <- col.bins[-length(col.bins)]
   y.height <- diff(col.bins)
   res2 <- y.pos + 0.5*y.height
@@ -98,7 +98,7 @@ gp.HCLshading <- function(observed, residuals,
   col.bins <- sort(c(range(res), col.bins, -col.bins, 0))
   col.bins <- col.bins[col.bins <= max(res) & col.bins >= min(res)]
 
-  ## col.legend
+  ## legend.col
   y.pos <- col.bins[-length(col.bins)]
   y.height <- diff(col.bins)
   res2 <- y.pos + 0.5*y.height
@@ -152,9 +152,9 @@ gp.binary <- function(observed, residuals, col = 1:2)
   y.height <- diff(col.bins)
   res2 <- y.pos + 0.5*y.height
   y.col <- ifelse(res2 > 0, col[1], col[2])
-  col.legend <- list(pos = y.pos, height = y.height, col = y.col)
+  legend.col <- list(pos = y.pos, height = y.height, col = y.col)
 
-  rval <- list(fill = fill, col.bins = col.bins, col.legend = col.legend)
+  rval <- list(fill = fill, legend.col = legend.col)
   class(rval) <- c("vcd.gpar", "gpar")
   return(rval)
 }

### 2003-08-08 trunk/vcd/paper/dsc2003.Rnw
small bug fixes & final paper update for DSC paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@276 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2003-06-10 trunk/vcd/paper/dsc2003.Rnw
@@
-\documentclass[10pt]{article}
+\documentclass[10pt,a4paper]{article}
 \usepackage{dsc2003}
-\usepackage{Sweave, graphicx}
-\usepackage{Z}
-
-\setlength{\parskip}{0.5ex plus0.1ex minus0.1ex}
-\setlength{\parindent}{0em}
-
-\bibpunct{(}{)}{;}{a}{}{,}
+\usepackage[authoryear,round,longnamesfirst]{natbib}
+%% don't need: \usepackage{vcd}
 
@@
 
-\SweaveOpts{engine=R}
+\SweaveOpts{engine=R,eps=FALSE}
 \setkeys{Gin}{width=0.8\textwidth}
@@
 data(UCBAdmissions)
+
+gp.old <- function(x, y) gp.Friendly(x, y, lty = 1)
 @
@@
 \title{Visualizing Independence Using Extended Association Plots}
-\author{\hfill David Meyer$^\dag$ \hfill Achim Zeileis$^\dag$ \hfill Kurt Hornik$^\ddag$ \hfill \hfill \\[0.2cm]
-\dag {\it \small Institut f\"ur Statistik \& Wahrscheinlichkeitstheorie,
- Technische Universit\"at Wien}\\
-\ddag {\it \small Institut f\"ur Statistik, Wirtschaftsuniversit\"at Wien}
-}
+\author{David Meyer \And Achim Zeileis \And Kurt Hornik}
 \date{}
@@
   importance of the residuals.
- \item Two methods for the use of multi-way tables are proposed.
  \item The implementation currently available in \textsf{R} is improved using a more modular
   design and allowing a more flexible specification of plotting parameters.
+ \item Two methods for the use of multi-way tables are proposed.
 \end{enumerate}
@@
 This task can greatly be sustained by the use of graphical
-methods such as the association plot, for which we propose some
+methods such as the mosaic display
+\citep{vcd:Hartigan+Kleiner:1981,vcd:Friendly:1994,vcd:Theus+Lauer:1999,vcd:Hofmann:2003}
+or the association plot \citep{vcd:Cohen:1980,vcd:Friendly:2000}.
+For the latter we propose some
 extensions: the use of colors for the shading of the residuals, two
@@
 \citep{vcd:Friendly:2000}.
+
+\pagebreak
 
@@
 <<assoc-improved1,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-grid.assocplot(HairEyeColor[,,2], gp = gp.shading)
+grid.assocplot(HairEyeColor[,,2], gp = gp.old)
 @
@@
 visualization of the observed frequencies in a contingency table;
-see][]{vcd:Hartigan+Kleiner:1984}:
-this extension 
-uses a color coding of the mosaic tiles to visualize deviations
-(residuals) from a given log-linear model fitted to the table.
-Positive and negative signs of the residuals are coded by rectangles with
-solid and dashed borders respectively. Furthermore, residuals can be
-classified according to specified shading levels: per default,
-residuals exceeding an absolute
-value of 2 are shaded light blue and red respectively, those that even
+see][]{vcd:Hartigan+Kleiner:1981,vcd:Hartigan+Kleiner:1984}.
+This extension uses a color coding of the mosaic tiles to visualize deviations
+(residuals) from independence. The sign and size of residuals are
+coded by color shadings of the corresponding rectangles: by default,
+residuals with an absolute value smaller than 2 are white, positive
+and negative residuals exceeding an absolute
+value of 2 are shaded light blue and red respectively, and those that even
 exceed an absolute value of 4 are shaded with full saturation. The heuristic
@@
 <<assoc-improved,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-grid.assocplot(t(margin.table(HairEyeColor, c(1,3))), gp = gp.shading)
+grid.assocplot(t(margin.table(HairEyeColor, c(1,3))), gp = gp.old)
 @
@@
 \end{figure}
+
+\section{Implementation enhancements}
+
+The current implementation of association plots in \textsf{R} suffers
+from two main disadvantages: First, it is not easy to recycle the
+plots in conditioning plots or pairs plots as they have been implemented
+using \textsf{R}'s base graphics engine where in general plotting to relative
+coordinates is not supported. Second, the graphics parameters of the rectangles in the
+association plots, such as color and line type, cannot be specified for each cell
+by the user. To overcome this, a new implementation was written from scratch
+in \texttt{grid} offering much more versatility amongst some minor advantages
+and convenient improvements. It allows the user to specify either 2-way tables of graphics parameters,
+i.e., arrays of the same dimensionality as the object being plotted, or a function
+which computes these graphics parameters based on the original table
+and its Pearson residuals. This makes it easy for users to implement their
+own shading schemes; \texttt{vcd} provides functions for shading schemes
+like those described in the previous sections. How association plots can
+be recycled using the new \texttt{grid} implementation is described in the next
+section.
 
@@
 <<UCB-mosaic-signif,fig=TRUE,echo=FALSE,height=8,width=8>>=
- panel.assocplot <- function (x, i, j, ..., axes, abbreviate) grid.assocplot(margin.table(x, c(i,j)), ..., panel = T, axis.labels = F, labels = F, main = NULL) 
-grid.mosaicpairs(aperm(HairEyeColor), panel.upper= panel.assocplot, panel.lower = panel.assocplot, gp = gp.shading, main = NULL)
+panel.assocplot <- function (x, i, j, ..., axes, abbreviate)
+  grid.assocplot(t(margin.table(x, c(i,j))), ..., panel = TRUE, axis.labels = FALSE, labels = FALSE, main = NULL)
+grid.mosaicpairs(aperm(HairEyeColor), panel.upper= panel.assocplot, panel.lower = panel.assocplot, gp = gp.old, main = NULL)
 @
@@
 <<UCB-assoc-signif,fig=TRUE,echo=FALSE,height=10,width=10>>=
-tabplot(UCBAdmissions, gp = gp.shading)
+tabplot(UCBAdmissions, gp = gp.old)
 @
@@
 
-\section{Implementation enhancements}
-
-The current implementation of association plots in \textsf{R} suffers
-from two main disadvantages: First, it is not easy to recycle the
-plots in conditioning plots or pairs plots as
-they have been implemented using \textsf{R}'s base graphics engine where in
-general plotting to relative coordinates is not supported.
-The new implementation was written from scratch in \texttt{grid} offering
-much more versatility amongst some minor advantages and convenient
-improvements. Second, the graphics parameters of the rectangles in the
-association plots, such as color and line type, cannot be specified for each cell
-by the user. To overcome this, the current implementation in \texttt{vcd}
-allows the user to specify either arrays of graphics parameters
-of the same dimensionality as the object being plotted or a function
-which computes these graphics parameters based on the original table
-and its Pearson residuals. Functions are provided for shading schemes
-like those described in the previous sections.
-
 \section{Conclusion}
@@
 problem in contingency tables using association plots. The
-extensions aim at improving the visualization by displaying
-the size of the residuals, and by extending the concept of association
-plots to multi-way tables. Furthermore, a new
-implementation is outlined based on the graphics package \texttt{grid}
-which provides more modular design and more flexible specification
-of graphical parameters.
+extensions aim at improving (a) the visualization by displaying
+the size of the residuals and (b) the implementation based on the
+graphics package \texttt{grid} which provides more modular design
+and more flexible specification of graphical parameters. Furthermore,
+some extensions of association plots to multi-way tables are outlined.
 Our work is still not at its end, though: for example, the plots for
@@
 
-\bibliographystyle{asa}
+\bibliographystyle{abbrvnat}
 \bibliography{vcd}
 
-\section*{Addresses}
+\subsection*{Affiliation}
 
-David Meyer\\
-Achim Zeileis\\
-E-mail: \texttt{\{David.Meyer,Achim.Zeileis\}@ci.tuwien.ac.at}\\
+David Meyer, Achim Zeileis\\
 Institut f\"ur Statistik und Wahrscheinlichkeitstheorie\\
@@
 Wiedner Hauptstra{\ss}e 8--10/1071\\
-1040 Wien\\
+A-1040 Wien, Austria\\
+E-mail: \email{David.Meyer@ci.tuwien.ac.at}, \email{Achim.Zeileis@ci.tuwien.ac.at}\\
 \\
 Kurt Hornik\\
-E-mail: \texttt{Kurt.Hornik@wu-wien.ac.at}\\
 Institut f\"ur Statistik\\
@@
 Augasse 2--6\\
-1090 Wien\\
+A-1090 Wien, Austria\\
+E-mail: \email{Kurt.Hornik@wu-wien.ac.at}\\
 
 \end{document}
-% LocalWords:  proofness

### 2003-08-08 trunk/vcd/R/grid.mosaicplot.R
small bug fixes & final paper update for DSC paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@276 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -262,7 +262,7 @@ grid.mosaicplot.default <-
                             ifelse(x > E, tmp, -tmp)
                           },
                           FT = sqrt(x) + sqrt(x + 1) - sqrt(4 * E + 1))
-      if(freq.type == "expected") x <- E #Z# caution: this is not not very clever
+      if(match.arg(freq.type) == "expected") x <- E #Z# caution: this is not not very clever
     }
 
     ## colors and legend

### 2003-08-08 trunk/vcd/R/shadings.R

@@ -127,7 +127,7 @@ gp.HCLshading <- function(observed, residuals,
 gp.Friendly <- function(observed, residuals,
                         hue = c(2/3, 0), lty = 1:2, interpolate = c(2, 4))
 {
-  gp.HSVshading(observed, residuals, hue = hue, value = 1, interpolate = interpolate, test = NULL)
+  gp.HSVshading(observed, residuals, hue = hue, value = 1, lty = lty, interpolate = interpolate, test = NULL)
 }
 
 gp.max <- function(observed, residuals,

### 2004-08-13 trunk/vcd/paper/JCGS.Rnw
started paper for JCGS
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@296 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2003-07-08 trunk/vcd/paper/draft.Rnw
@@
-\documentclass[10pt]{article}
-\usepackage{Sweave, graphicx}
-\usepackage{Z}
+\documentclass[10pt,a4paper]{article}
 
-\setlength{\parskip}{0.5ex plus0.1ex minus0.1ex}
-\setlength{\parindent}{0em}
+%% Packages:
+\usepackage{graphicx,color,a4wide,thumbpdf,hyperref}
 
+%% \usepackage{Sweave} is essentially
+\RequirePackage[T1]{fontenc}
+\RequirePackage{ae,fancyvrb}
+\DefineVerbatimEnvironment{Sinput}{Verbatim}{fontshape=sl}
+\DefineVerbatimEnvironment{Soutput}{Verbatim}{}
+\DefineVerbatimEnvironment{Scode}{Verbatim}{fontshape=sl}
+\newenvironment{Schunk}{}{}
+
+%% BibTeX
+\usepackage[authoryear,round,longnamesfirst]{natbib}
+\bibliographystyle{asa}
 \bibpunct{(}{)}{;}{a}{}{,}
 
+%% new commands
+\let\code=\texttt
+\let\proglang=\textsf
+\newcommand{\pkg}[1]{{\normalfont\fontseries{b}\selectfont #1}}
+\newcommand{\given}{\, | \,}
+
+%% Colors
+\definecolor{Red}{rgb}{0.7,0,0}
+\definecolor{Blue}{rgb}{0,0,0.8}
+\definecolor{hellgrau}{rgb}{0.55,0.55,0.55}
+
+%% mimic JCGS style
+\renewcommand{\section}{\secdef \mysec \mysecnn}
+\newcommand{\mysec}[2][default]{\vspace{1.3\baselineskip}%
+  \pdfbookmark[1]{#1}{Section.\thesection.#1}%
+  \refstepcounter{section}%
+  {\begin{center} \large \bf \thesection. \uppercase{#1} \end{center}}%
+  \vspace{0.7\baselineskip}}
+\newcommand{\mysecnn}[1]{\vspace{1.3\baselineskip}%
+  {\begin{center}\large \bf \uppercase{#1} \end{center}}%
+  \vspace{0.7\baselineskip}}
+
+\renewcommand{\subsection}{\secdef \mysubsec \mysubsecnn}
+\newcommand{\mysubsec}[2][default]{\vspace{1.2\baselineskip}%
+  \pdfbookmark[2]{#1}{Subsection.\thesubsection.#1}%
+  \refstepcounter{subsection}%
+  \noindent%
+  {\large \textbf{\thesubsection{} {\sc#1}}} \vspace{.7\baselineskip}}
+\newcommand{\mysubsecnn}[1]{\vspace{1.2\baselineskip}%
+  \noindent%
+  {\large \textbf{{\sc #1}}} \vspace{.7\baselineskip}}
+\renewcommand{\abstractname}{}
+\renewcommand{\refname}{REFERENCES}
+
+%% Hyperref
+\hypersetup{%
+  pdftitle = {Visualizing (Conditional) Independence Using Extended Association and Mosaic Plots},
+  pdfsubject = {Visualizing (Conditional) Independence},
+  pdfkeywords = {contingency tables, association plots, mosaic plots, conditional
+    inference, permutation tests},
+  pdfauthor = {Achim Zeileis, David Meyer, Kurt Hornik},
+  %% change colorlinks to false for pretty printing
+  colorlinks = {true},
+  linkcolor = {Blue},
+  citecolor = {Blue},
+  urlcolor = {Red},
+  hyperindex = {true},
+  linktocpage = {true},
+}
+
 \begin{document}
+
+\begin{center}
+\textbf{\LARGE Visualizing (Conditional) Independence Using Extended Association and Mosaic Plots}
+
+\vspace{0.5cm}
+{\large Achim \textsc{Zeileis}, David \textsc{Meyer}, and Kurt \textsc{Hornik}}
+
+\vspace{0.3cm}
+\textit{\large Wirtschaftsuniversit\"at Wien, Augasse 2--6, 1090 Wien, Austria}\\[0.14cm]
+{\large \textit{E-mail:} \texttt{Firstname.Lastname@wu-wien.ac.at}}
+\end{center}
 
@@
 library(vcd)
+
 data(Bundesliga)
 goals <- xtabs(~ AwayGoals + HomeGoals, data = Bundesliga, subset = Year == 1995)
+
 data(HairEyeColor)
 haireye <- xtabs(Freq ~ Hair + Eye, data = as.data.frame(HairEyeColor), subset = Sex == "Female")
+
 data(UCBAdmissions)
+ucb <- UCBAdmissions
+dimnames(ucb)[[1]] <- c("Admit", "Reject")
+
+data(Arthritis)
+art <- xtabs(~ Improved + Treatment, data = Arthritis)
 @
 
-\title{Visualizing Independence Using Extended Association and Mosaic Plots}
-\author{\hfill Achim Zeileis$^\dag$ \hfill David Meyer$^\dag$ \hfill Kurt Hornik$^\ddag$ \hfill \hfill \\[0.2cm]
-\dag {\it \small Institut f\"ur Statistik \& Wahrscheinlichkeitstheorie,
- Technische Universit\"at Wien}\\
-\ddag {\it \small Institut f\"ur Statistik, Wirtschaftsuniversit\"at Wien}
+<<auxfunctions,echo=FALSE,results=hide>>=
+he.test <- pearson.test(haireye, return = TRUE)
+cb <- he.test$qdist(c(0.9, 0.99))
+
+gp.shading <- function(x, y) gp.HSVshading(x, y, lty = 1, test = NULL)
+shadeHSVchisq <- function(x, y) gp.HSVshading(x, y, lty = 1)
+shadeHSVmax <- function(x, y) gp.HSVshading(x, y, interpolate = cb, lty = 1)
+shadeHCLmax <- function(x, y) gp.HCLshading(x, y, interpolate = cb, lty = 1)
+shadeHSVmax2 <- function(x, y) gp.HSVshading(x, y, interpolate = c(5,10), test = NULL)
+shadeHCLchisq <- function(x, y) gp.HCLshading(x, y, lty = 1)
+
+hue.slice <- function(hue, grid.n = 100, type = c("HCL", "HSV"), plot = TRUE, correct = FALSE)
+{
+  type <- match.arg(type)
+  if(type == "HCL") {
+    chroma = seq(0, 100, length = grid.n)
+    luminance = seq(0, 100, length = grid.n)
+    nc <- length(chroma)
+    nl <- length(luminance)
+    color.slice <- outer(chroma, luminance, function(y, x) hcl(hue, x, y, verbose = FALSE, correct = correct))
+    xlab <- "chroma"
+    ylab <- "luminance"
+    main <- paste("hue =", round(hue, digits = 0))
+  } else {
+    chroma = seq(0, 1, length = grid.n)
+    luminance = seq(0, 1, length = grid.n)
+    nc <- length(chroma)
+    nl <- length(luminance)
+    color.slice <- outer(chroma, luminance, function(y, x) hsv(hue, x, y))
+    xlab <- "saturation"
+    ylab <- "value"
+    main <- paste("hue =", round(hue, digits = 3))
   }
-\date{}
-\maketitle
+  if(plot) {
+    plot(0.5, 0.5, xlim = range(chroma), ylim = range(luminance), type = "n", axes = FALSE,
+         xlab = xlab, ylab = ylab, yaxs = "i", xaxs = "i", main = main)
+    for(i in 1:(nc-1)) {
+      rect(chroma[i], luminance[-nl], chroma[i] + 100/(nc-1), luminance[-1], border = color.slice[,i+1], col = color.slice[,i+1])
+    }
+    axis(1)
+    axis(2)
+    box()
+  }
+  colnames(color.slice) <- chroma
+  rownames(color.slice) <- luminance
+  attr(color.slice, "type") <- type
+  class(color.slice) <- "slice"
+  invisible(color.slice)
+}
+
+plot.slice <- function(color.slice)
+{
+  nc <- ncol(color.slice)
+  nl <- nrow(color.slice)
+  chroma <- as.numeric(colnames(color.slice))
+  luminance <- as.numeric(rownames(color.slice))
+  if(attr(color.slice, "type") == "HCL") {
+    xlab <- "chroma"
+    ylab <- "luminance"
+  } else {
+    xlab <- "saturation"
+    ylab <- "value"
+  }
+  plot(0.5, 0.5, xlim = range(chroma), ylim = range(luminance), type = "n", axes = FALSE,
+       xlab = xlab, ylab = ylab, yaxs = "i", xaxs = "i", main = deparse(substitute(color.slice)))
+  for(i in 1:(nc-1)) {
+    rect(chroma[i], luminance[-nl], chroma[i] + 100/(nc-1), luminance[-1], border = color.slice[,i+1], col = color.slice[,i+1])
+  }
+  axis(1)
+  axis(2)
+  box()
+}
+@
 
 \begin{abstract}
-Two visualization techniques for the independence problem in 2-way
-contingency tables---assocation and mosaic plots---are extended in two directions:
-\begin{enumerate}
-\item The visualization is enhanced by displaying the significance of an appropriate
-test for independence and by using improved color schemes.
-\item The implementation in the \textsf{R} system is improved using a more modular
-design and allowing for more flexible specification of plotting parameters.
-\end{enumerate}
-Furthermore, some enhancements for displays of various types of independence in
-general multi-way tables are suggested.
+Story starts out not from exploratory analysis with \proglang{R}
+but from the problem of visualizing independence. Several
+exploratory approaches but not combined with significance
+testing.
+
+Paper could be structured in the following way
+\begin{itemize}
+  \item Introduction
+  \item Tests
+  \item Data sets
+  \item Visualization Techniques
+    \begin{itemize}
+      \item Mosaic displays
+      \item Association plots
+      \item Residuals-based shading
+    \end{itemize}
+  \item Extensions
+    \begin{itemize}
+      \item max-shading
+      \item HCL colors
+      \item conditional independence (coplots)
+      \item \textit{not the implementation, just mention in summary}
+    \end{itemize}
+  \item Summary
+\end{itemize}
+
+The \proglang{R} side then has to be described in another paper,
+probably for the Journal of Statistical Software.\\[0.3cm]
+\textbf{Key Words:} Association plots; Conditional inference; 
+  Contingency tables; Mosaic plots; Permutation tests.
 \end{abstract}
@@
 
-\section{Tests for independence in 2-way tables} \label{sec:tests}
+\section{Tests for Independence in 2-way Tables} \label{sec:tests}
 
@@
 
-\section{Visualization techniques for 2-way tables} \label{sec:visualization}
+\section{Visualization Techniques for 2-way Tables} \label{sec:visualization}
 
@@
 
-\subsection{Association plots}
+\subsection{Association Plots}
 
@@
 
-\setkeys{Gin}{width=0.7\textwidth}
-\begin{figure}[htbp]
-\begin{center}
-<<goals-assoc-classic,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-grid.assocplot(goals, rot = 0, gp = gp.binary)
-@
-\caption{Association plot for the Bundesliga data.}
-\label{fig:goals-assoc-classic}
-\end{center}
-\end{figure}
-
-\subsection{Mosaic plots}
+\subsection{Mosaic Plots}
 
@@
 (see Figure \ref{fig:goals-mosaic-classic}).
-
-\begin{figure}[htbp]
-\begin{center}
-<<goals-mosaic-classic,fig=TRUE,echo=FALSE,height=6,width=7>>=
-mosaicplot(t(goals), shade = TRUE, main = "")
-@
-\caption{Friendly mosaic plot for the Bundesliga data.}
-\label{fig:goals-mosaic-classic}
-\end{center}
-\end{figure}
 
@@
 
-\subsection{Visualization enhancements}
+\subsection{Visualization Enhancements}
 
@@
 
-\setkeys{Gin}{width=0.8\textwidth}
-\begin{figure}[htbp]
-\begin{center}
-<<goals-assoc-signif,fig=TRUE,echo=FALSE,height=6,width=6>>=
-grid.assocplot(goals, rot = 0)
-@
-\caption{Extended association plots for the Bundesliga data.}
-\label{fig:goals-assoc-signif}
-\end{center}
-\end{figure}
-
 To illustrate what a `postive' example looks like we give the association plot
 for the well-known Hair-Eye-Color data set (avilable in base \textsf{R}) which cross-tabulates the hair and eye
-color and sex of 592 statistics students: Figure \ref{fig:haireye-assoc-signif}
+color and gender of 592 statistics students: Figure \ref{fig:haireye-assoc-signif}
 shows the association plot for the 328 female students. It is easy to see that
@@
 cell is significant at the 10\% level.
-
-\begin{figure}[htbp]
-\begin{center}
-<<haireye-assoc-signif,fig=TRUE,echo=FALSE,height=6,width=6>>=
-grid.assocplot(haireye)
-@
-\caption{Extended association plots for the Hair-Eye-Color data (female students).}
-\label{fig:haireye-assoc-signif}
-\end{center}
-\end{figure}
 
@@
 
-\begin{figure}[htbp]
-\begin{center}
-<<goals-mosaic-signif,fig=TRUE,echo=FALSE,height=6,width=6>>=
-grid.mosaicplot(t(goals), main = NULL)
-@
-<<haireye-mosaic-signif,fig=TRUE,echo=FALSE,height=6,width=6>>=
-grid.mosaicplot(t(haireye), main = NULL)
-@
-\caption{Extended mosaic plots for the Bundesliga and Hair-Eye-Color data (female students).}
-\label{fig:both-mosaic-signif}
-\end{center}
-\end{figure}
-
-\subsection{Implementation enhancements}
+\subsection{Implementation Enhancements}
 
@@
 
-\section{Multi-way tables}
+\section{Multi-way Tables}
 
@@
 California at Berkley (UCB) which is available in base \textsf{R}.
-In this data, the question whether there is sex
+In this data, the question whether there is gender
 discrimination at the UCB leads to the result that although women seem to
-be disadvantaged at the aggregated level there is no sex discrimination
+be disadvantaged at the aggregated level there is no gender discrimination
 conditioned on the department---with the very exception of one department
@@
 
-\begin{figure}[htbp]
-\begin{center}
-<<UCB-assoc-signif,fig=TRUE,echo=FALSE,height=10,width=10>>=
-tabplot(UCBAdmissions)
-@
-\caption{Extended conditioned association plots for the UCB Admissions data.}
-\label{fig:UCB-assoc-signif}
-\end{center}
-\end{figure}
-
 Similarly, the same data can be visualized using a mosaic matrix where
@@
 
-\begin{figure}[htbp]
-\begin{center}
-<<UCB-mosaic-signif,fig=TRUE,echo=FALSE,height=8,width=8>>=
-##grid.mosaicpairs(UCBAdmissions, type = "conditional", main = NULL)
-grid.mosaicpairs(UCBAdmissions, type = "conditional", permute = FALSE)
-@
-\caption{Extended conditioned association plots for the UCB Admissions data.}
-\label{fig:UCB-mosaic-signif}
-\end{center}
-\end{figure}
-
-\section{Conclusion}
+\section{Conclusions}
 We suggest a set of enhancements for visualizing the independence
@@
 
-\bibliographystyle{asa}
+\setkeys{Gin}{width=0.7\textwidth}
+\begin{figure}[htbp]
+\begin{center}
+<<goals-assoc-classic,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
+grid.assocplot(goals, rot = 0, gp = gp.binary)
+@
+<<goals-mosaic-classic,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
+mosaicplot(t(goals), shade = TRUE, main = "")
+@
+\caption{Classic association and mosaic plot for the Bundesliga data.}
+\label{fig:goals-assoc-classic}
+\end{center}
+\end{figure}
+
+\begin{figure}[htbp]
+\begin{center}
+<<goals-mosaic,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
+grid.mosaicplot(goals, gp = gp.max)
+@
+<<art-mosaic,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
+grid.mosaicplot(art, gp = gp.max)
+@
+\caption{Extended mosaic plots for the Bundesliga and Arthritis data sets.}
+\label{fig:assoc}
+\end{center}
+\end{figure}
+
+\setkeys{Gin}{width=0.8\textwidth}
+\begin{figure}[htbp]
+\begin{center}
+<<Friendly-hue,fig=TRUE,echo=FALSE,height=4.8,width=8>>=
+plot(0.5, 0.5, xlim = c(0, 1), ylim = c(0, 1), type = "n", axes = FALSE, xlab = "", ylab = "", yaxs = "i")
+dummy <- 0:100/100
+dummy.col <- hsv(dummy[-101] + 1/200, 1, 1)
+rect(dummy[-101], 0.6, dummy[-1], 1, border = dummy.col, col = dummy.col)
+rect(0, 0.6, 1, 1)
+axis(3)
+mtext("hue", line = 2, cex = 1.5)
+text(1, 0.45, "saturation = 1\nvalue = 1", pos = 2, cex = 1.2)
+lines(c(0.1, 0), c(0.3, 0.6), lty = 2)
+lines(c(2/3, 2/3), c(0.3, 0.6), lty = 2)
+rect(0, 0.1, 0.2, 0.3, col = hsv(0, 1, 1))
+rect(2/3 - 0.1, 0.1, 2/3 + 0.1, 0.3, col = hsv(2/3, 1, 1))
+text(0.1, 0.1, expression(r[ij] < 0), pos = 1, cex = 1.5)
+text(2/3, 0.1, expression(r[ij] > 0), pos = 1, cex = 1.5)
+@
+
+<<Friendly-sat,fig=TRUE,echo=FALSE,height=4.8,width=8>>=
+plot(0.5, 0.5, xlim = c(0, 1), ylim = c(0, 1), type = "n", axes = FALSE, xlab = "", ylab = "", yaxs = "i")
+dummy <- 0:100/100
+dummy.col <- hsv(2/3, dummy[-101] + 1/200, 1)
+rect(dummy[-101], 0.8, dummy[-1], 1, border = dummy.col, col = dummy.col)
+dummy.col <- hsv(0, dummy[-101] + 1/200, 1)
+rect(dummy[-101], 0.6, dummy[-1], 0.8, border = dummy.col, col = dummy.col)
+rect(0, 0.8, 1, 1)
+rect(0, 0.6, 1, 0.8)
+axis(3)
+mtext("saturation", line = 2, cex = 1.5)
+mtext(c("h = 0", "h = 2/3"), side = 2, at = c(0.7, 0.9), las = 1, cex = 1.2)
+text(0.8, 0.5, "value = 1", pos = 2, cex = 1.2)
+lines(c(0.1, 0), c(0.3, 0.6), lty = 2)
+lines(c(0.5, 0.5), c(0.3, 0.6), lty = 2)
+lines(c(0.9, 1), c(0.3, 0.6), lty = 2)
+rect(0, 0.2, 0.2, 0.3, col = hsv(2/3, 0, 1))
+rect(0, 0.1, 0.2, 0.2, col = hsv(0, 0, 1))
+rect(0.4, 0.2, 0.6, 0.3, col = hsv(2/3, 0.5, 1))
+rect(0.4, 0.1, 0.6, 0.2, col = hsv(0, 0.5, 1))
+rect(0.8, 0.2, 1, 0.3, col = hsv(2/3, 1, 1))
+rect(0.8, 0.1, 1, 0.2, col = hsv(0, 1, 1))
+text(0.1, 0.1, expression(abs(r[ij]) < 2), pos = 1, cex = 1.5)
+text(0.5, 0.1, expression(2 < abs(r[ij]) < 4), pos = 1, cex = 1.5)
+text(0.9, 0.1, expression(abs(r[ij]) > 4), pos = 1, cex = 1.5)
+@
+
+<<Friendly-val,fig=TRUE,echo=FALSE,height=4.8,width=8>>=
+plot(0.5, 0.5, xlim = c(0, 1), ylim = c(0, 1), type = "n", axes = FALSE, xlab = "", ylab = "", yaxs = "i")
+dummy <- 0:100/100
+dummy.col <- hsv(2/3, 1, dummy[-101] + 1/200)
+rect(dummy[-101], 0.8, dummy[-1], 1, border = dummy.col, col = dummy.col)
+dummy.col <- hsv(0, 1, dummy[-101] + 1/200)
+rect(dummy[-101], 0.6, dummy[-1], 0.8, border = dummy.col, col = dummy.col)
+rect(0, 0.8, 1, 1)
+rect(0, 0.6, 1, 0.8)
+axis(3)
+mtext("value", line = 2, cex = 1.5)
+mtext(c("h = 0", "h = 2/3"), side = 2, at = c(0.7, 0.9), las = 1, cex = 1.2)
+text(0, 0.5, paste("saturation =", 1), pos = 4, cex = 1.2)
+lines(c(0.85, 1), c(0.3, 0.6), lty = 2)
+rect(0.7, 0.2, 0.8, 0.3, col = hsv(2/3, 0, 1))
+rect(0.8, 0.2, 0.9, 0.3, col = hsv(2/3, 0.5, 1))
+rect(0.9, 0.2, 1, 0.3, col = hsv(2/3, 1, 1))
+rect(0.7, 0.1, 0.8, 0.2, col = hsv(0, 0, 1))
+rect(0.8, 0.1, 0.9, 0.2, col = hsv(0, 0.5, 1))
+rect(0.9, 0.1, 1, 0.2, col = hsv(0, 1, 1))
+lines(c(0.5, 0.5), c(0.3, 0.6), lty = 2)
+lines(c(0.85, 1), c(0.3, 0.6), lty = 2)
+rect(0.35, 0.2, 0.45, 0.3, col = hsv(2/3, 0, 0.5))
+rect(0.45, 0.2, 0.55, 0.3, col = hsv(2/3, 0.5, 0.5))
+rect(0.55, 0.2, 0.65, 0.3, col = hsv(2/3, 1, 0.5))
+rect(0.35, 0.1, 0.45, 0.2, col = hsv(0, 0, 0.5))
+rect(0.45, 0.1, 0.55, 0.2, col = hsv(0, 0.5, 0.5))
+rect(0.55, 0.1, 0.65, 0.2, col = hsv(0, 1, 0.5))
+rect(0.7, 0.2, 0.8, 0.3, col = hsv(2/3, 0, 1))
+rect(0.8, 0.2, 0.9, 0.3, col = hsv(2/3, 0.5, 1))
+rect(0.9, 0.2, 1, 0.3, col = hsv(2/3, 1, 1))
+rect(0.7, 0.1, 0.8, 0.2, col = hsv(0, 0, 1))
+rect(0.8, 0.1, 0.9, 0.2, col = hsv(0, 0.5, 1))
+rect(0.9, 0.1, 1, 0.2, col = hsv(0, 1, 1))
+text(0.5, 0.1, "non-significant", pos = 1, cex = 1.5)
+text(0.85, 0.1, "significant", pos = 1, cex = 1.5)
+@
+
+\caption{Extended Friendly shading in HSV space.}
+\label{fig:Friendly-HSV}
+\end{center}
+\end{figure}
+
+\begin{figure}[htbp]
+\begin{center}
+<<Friendly-HSV,fig=TRUE,echo=FALSE,height=7,width=8>>=
+## generate colors
+hue23 <- hue.slice(2/3, grid.n = 101, plot = FALSE, type = "HSV")
+hue0 <- hue.slice(0, grid.n = 101, plot = FALSE, type = "HSV")
+mysaturation <- as.numeric(colnames(hue23))
+value <- as.numeric(rownames(hue23))
+
+## select those with value >= 0.5
+hue23 <- hue23[value >= .5, ]
+hue0 <- hue0[value >= .5, ]
+value <- value[value >= .5]
+nl <- nrow(hue23)
+nc <- ncol(hue23)
+
+## plot 2 slides from HSV space
+plot(0.5, 0.5, xlim = c(-1, 1), ylim = c(0, 1), type = "n", axes = FALSE,
+       xlab = "", ylab = "", yaxs = "i", xaxs = "i", main = "")
+for(i in 1:(nc-1)) {
+  rect(mysaturation[i], value[-nl], mysaturation[i] + 1/(nc-1), value[-1], border = hue23[,i+1], col = hue23[,i+1])
+}
+for(i in 1:(nc-1)) {
+  rect(-mysaturation[i], value[-nl], -(mysaturation[i] + 1/(nc-1)), value[-1], border = hue0[,i+1], col = hue0[,i+1])
+}
+axis(2, at = c(50, 75, 100)/100, labels = c(0.5, 0.75, 1))
+axis(4, at = c(50, 75, 100)/100, labels = c(0.5, 0.75, 1))
+axis(3, at = -4:4*.25, labels=c(4:0*.25, 1:4*.25))
+mtext(c("hue = 0", "hue = 2/3"), side = 3, at = c(-.5, .5), line = 3, cex = 1.2)
+mtext("saturation", side = 3, at = 0, line = 2)
+mtext("value", side = 2, at = .75, line = 2)
+mtext("value", side = 4, at = .75, line = 2)
+lines(c(-1, 1), c(.5, .5))
+
+## significant colors
+rect(-1, 0.95, -.90, 1, col = hsv(0, 1, 1))
+rect(-0.45, 0.95, -.55, 1, col = hsv(0, 0.5, 1))
+rect(-.05, .95, .05, 1, col = hsv(2/3, 0, 1))
+rect(0.45, 0.95, .55, 1, col = hsv(2/3, 0.5, 1))
+rect(.90, .95, 1, 1, col = hsv(2/3, 1, 1))
+
+text(-1, .33, "significant", pos = 4, cex = 1.2)
+rect(-1, .20, -.80, .30, col = hsv(0, 1, 1))
+rect(-.40, .20, -0.6, .30, col = hsv(0, 0.5, 1))
+rect(-.20, .20, 0, .30, col = hsv(0, 0, 1))
+rect(0, .20, .20, .30, col = hsv(2/3, 0, 1))
+rect(0.4, .20, .60, .30, col = hsv(2/3, .5, 1))
+rect(.80, .20, 1, .30, col = hsv(2/3, 1, 1))
+
+lines(c(-.9, -.55), c(0.975, .975), lty = 2)
+lines(c(-.45, -.05), c(0.975, .975), lty = 2)
+lines(c(.45, .05), c(0.975, .975), lty = 2)
+lines(c(.9, .55), c(0.975, .975), lty = 2)
+
+## non-significant colors
+rect(-1, 0.5, -.90, 0.55, col = hsv(0, 1, 0.5))
+rect(-0.45, 0.5, -.55, 0.55, col = hsv(0, 0.5, 0.5))
+rect(-.05, .5, .05, 0.55, col = hsv(2/3, 0, 0.5))
+rect(0.45, 0.5, .55, 0.55, col = hsv(2/3, 0.5, 0.5))
+rect(.90, .5, 1, 0.55, col = hsv(2/3, 1, 0.5))
+
+text(-1, .13, "non-significant", pos = 4, cex = 1.2)
+rect(-1, 0, -.80, .10, col = hsv(0, 1, 0.5))
+rect(-.60, 0, -.4, .10, col = hsv(0, 0.5, 0.5))
+rect(-.20, 0, 0, .10, col = hsv(0, 0, 0.5))
+rect(0, 0, .20, .10, col = hsv(2/3, 0, 0.5))
+rect(0.4, 0, .60, .1, col = hsv(2/3, .5, 0.5))
+rect(.80, 0, 1, .10, col = hsv(2/3, 1, 0.5))
+
+lines(c(-.9, -.55), c(0.525, .525), lty = 2)
+lines(c(-.45, -.05), c(0.525, .525), lty = 2)
+lines(c(.45, .05), c(0.525, .525), lty = 2)
+lines(c(.9, .55), c(0.525, .525), lty = 2)
+@
+
+<<HCL-space,fig=TRUE,echo=FALSE,height=7,width=8>>=
+## generate colors
+hue260 <- hue.slice(260, grid.n = 101, plot = FALSE)
+hue360 <- hue.slice(360, grid.n = 101, plot = FALSE)
+mychroma <- as.numeric(colnames(hue260))
+luminance <- as.numeric(rownames(hue260))
+
+## select those with lumincance >= 50
+hue260 <- hue260[luminance >= 50, ]
+hue360 <- hue360[luminance >= 50, ]
+luminance <- luminance[luminance >= 50]
+nc <- ncol(hue260)
+nl <- nrow(hue260)
+
+## plot 2 slides from HCL space
+plot(0.5, 0.5, xlim = c(-100, 100), ylim = c(0, 100), type = "n", axes = FALSE,
+       xlab = "", ylab = "", yaxs = "i", xaxs = "i", main = "")
+for(i in 1:(nc-1)) {
+  rect(mychroma[i], luminance[-nl], mychroma[i] + 100/(nc-1), luminance[-1], border = hue260[,i+1], col = hue260[,i+1])
+}
+for(i in 1:(nc-1)) {
+  rect(-mychroma[i], luminance[-nl], -(mychroma[i] + 100/(nc-1)), luminance[-1], border = hue360[,i+1], col = hue360[,i+1])
+}
+axis(2, at = c(50, 75, 100), labels = c(50, 75, 100))
+axis(4, at = c(50, 75, 100), labels = c(50, 75, 100))
+axis(3, at = -4:4*25, labels=c(4:0*25, 1:4*25))
+mtext(c("hue = 0", "hue = 260"), side = 3, at = c(-50, 50), line = 3, cex = 1.2)
+mtext("chroma", side = 3, at = 0, line = 2)
+mtext("luminance", side = 2, at = 75, line = 2)
+mtext("luminance", side = 4, at = 75, line = 2)
+lines(c(-100, 100), c(50, 50))
+
+## significant colors
+rect(-100, 47.5, -90, 52.5, col = hcl(0, 100, 50))
+rect(-55, 67.5, -45, 72.5, col = hcl(0, 50, 70))
+rect(-5, 95, 5, 100, col = hcl(260, 0, 100))       ## grey vs. white
+rect(-5, 87.5, 5, 92.5, col = hcl(260, 0, 90))     ## grey vs. white
+rect(45, 67.5, 55, 72.5, col = hcl(260, 50, 70))
+rect(90, 47.5, 100, 52.5, col = hcl(260, 100, 50))
+
+text(-100, 33, "significant", pos = 4, cex = 1.2)
+rect(-100, 20, -80, 30, col = hcl(0, 100, 50))
+rect(-60, 20, -40, 30, col = hcl(0, 50, 70))
+rect(-20, 20, 0, 30, col = hcl(0, 0, 90))       
+rect(0, 20, 20, 30, col = hcl(260, 0, 90))
+#white# rect(-20, 20, 0, 30, col = hcl(0, 0, 100))
+#white# rect(0, 20, 20, 30, col = hcl(260, 0, 100))
+rect(40, 20, 60, 30, col = hcl(260, 50, 70))
+rect(80, 20, 100, 30, col = hcl(260, 100, 50))
+
+lines(c(-45, -5), c(72.5, 87.5), lty = 2)
+lines(c(45, 5), c(72.5, 87.5), lty = 2)
+lines(c(-95, -55), c(52.5, 67.5), lty = 2)
+lines(c(95, 55), c(52.5, 67.5), lty = 2)
+
+## non-significant colors
+rect(-25, 47.5, -15, 52.5, col = hcl(0, 20, 50))
+rect(-15, 67.5, -5, 72.5, col = hcl(0, 10, 70))
+rect(5, 67.5, 15, 72.5, col = hcl(260, 10, 70))
+rect(25, 47.5, 15, 52.5, col = hcl(260, 20, 50))
+
+text(-100, 13, "non-significant", pos = 4, cex = 1.2)
+rect(-60, 0, -40, 10, col = hcl(0, 20, 50))
+rect(-40, 0, -20, 10, col = hcl(0, 10, 70))
+rect(-20, 0, 0, 10, col = hcl(0, 0, 90))
+rect(0, 0, 20, 10, col = hcl(260, 0, 90))
+rect(20, 0, 40, 10, col = hcl(260, 10, 70))
+rect(40, 0, 60, 10, col = hcl(260, 20, 50))
+
+lines(c(-18.75, -11.25), c(52.5, 67.5), lty = 2)
+lines(c(-8.75, -1.25), c(72.5, 87.5), lty = 2)
+lines(c(18.75, 11.75), c(52.5, 67.5), lty = 2)
+lines(c(8.75, 1.25), c(72.5, 87.5), lty = 2)
+@
+\caption{Extended shading in HSV and HCL space.}
+\label{fig:ExtHSVHCL}
+\end{center}
+\end{figure}
+
+\setkeys{Gin}{width=\textwidth}
+\begin{figure}[htbp]
+\begin{center}
+<<UCB-mosaic1,fig=TRUE,echo=FALSE,height=4,width=8>>=
+fm <- loglm(~(Gender + Admit)*Dept, data = UCBAdmissions, fitted = TRUE)
+res.fm <- (UCBAdmissions - fitted(fm))/sqrt(fitted(fm))
+dim(res.fm) <- dim(UCBAdmissions)
+my.gp <- gp.HCLshading(UCBAdmissions, aperm(res.fm))
+my.gp$p.value <- summary(fm)$tests[2,3]
+my.gp$lty <- c(1,1)
+grid.mosaicplot(ucb, gp = my.gp, margin = ~ (Gender + Admit)*Dept, space = c(0.1, 0.01, 0))
+@
+\caption{UCB admission with mosaic display.}
+\label{fig:UCB-mosaic1}
+\end{center}
+\end{figure}
+
+\begin{figure}[htbp]
+\begin{center}
+<<UCB-mosaic2,fig=TRUE,echo=FALSE,height=6,width=8>>=
+mygp.max <- function(x,y) gp.max(x, y, level = (1 - c(0.1, 0.01)/6))
+tabplot(margin.table(ucb, c(2,1,3)), panel = function(x, ...) grid.mosaicplot(x, panel = TRUE, legend = FALSE, labels = FALSE, gp = mygp.max, space = 0.02, ...))
+@
+\caption{UCB admission with conditioned mosaic display.}
+\label{fig:UCB-mosaic2}
+\end{center}
+\end{figure}
+
+\begin{figure}[htbp]
+\begin{center}
+<<UCB-assoc,fig=TRUE,echo=FALSE,height=8,width=8>>=
+tabplot(margin.table(ucb, c(2,1,3)), gp = mygp.max)
+@
+\caption{UCB admission with conditioned association plot.}
+\label{fig:UCB-assoc}
+\end{center}
+\end{figure}
+
 \bibliography{vcd}
@@
 \end{document}
-% LocalWords:  proofness

### 2004-08-13 trunk/vcd/paper/JCGS.Rnw
*** empty log message ***
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@297 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \begin{center}
-<<goals-mosaic,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
+<<haireye-Friendly,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
+grid.mosaicplot(haireye, gp = gp.Friendly)
+@
+<<haireye-max,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
+grid.mosaicplot(haireye, gp = gp.max)
+@
+\caption{Mosaic plots for the Hair-Eye-Color data.}
+\label{fig:haireye}
+\end{center}
+\end{figure}
+
+\begin{figure}[htbp]
+\begin{center}
+<<goals-Friendly,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
+grid.mosaicplot(goals, gp = gp.Friendly)
+@
+<<goals-max,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
 grid.mosaicplot(goals, gp = gp.max)
 @
-<<art-mosaic,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
+\caption{Mosaic plots for the Bundesliga data.}
+\label{fig:goals}
+\end{center}
+\end{figure}
+
+\begin{figure}[htbp]
+\begin{center}
+<<art-Friendly,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
+grid.mosaicplot(art, gp = gp.Friendly)
+@
+<<art-max,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
 grid.mosaicplot(art, gp = gp.max)
 @
-\caption{Extended mosaic plots for the Bundesliga and Arthritis data sets.}
-\label{fig:assoc}
+\caption{Mosaic plots for the Arthritis data.}
+\label{fig:art}
 \end{center}
@@
 mygp.max <- function(x,y) gp.max(x, y, level = (1 - c(0.1, 0.01)/6))
-tabplot(margin.table(ucb, c(2,1,3)), panel = function(x, ...) grid.mosaicplot(x, panel = TRUE, legend = FALSE, labels = FALSE, gp = mygp.max, space = 0.02, ...))
+tabplot(ucb, panel = function(x, ...) grid.mosaicplot(x, panel = TRUE, legend = FALSE, labels = FALSE, gp = mygp.max, space = 0.02, ...))
 @
@@
 
+\nocite{*}
 \bibliography{vcd}

### 2004-08-13 trunk/vcd/R/shadings.R

@@ -14,7 +14,6 @@ gp.HSVshading <- function(observed, residuals,
     col.bins <- min(res) + diff(range(res)) * ((0:200)/200)
   } else {
     cb <- col.bins <- sort(interpolate)
-    require(stepfun)  #Z# require globally?
     interpolate <- function(x) stepfun(cb,  seq(my.s[2], my.s[1], length = length(cb) + 1))(abs(x))
   }
 
@@ -76,7 +75,6 @@ gp.HCLshading <- function(observed, residuals,
     col.bins <- min(res) + diff(range(res)) * ((0:200)/200)
   } else {
     cb <- col.bins <- sort(interpolate)
-    require(stepfun)  #Z# require globally?
     interpolate <- function(x) stepfun(cb,  seq(0, 1, length = length(cb) + 1))(abs(x))
   }
 
@@ -125,9 +123,12 @@ gp.HCLshading <- function(observed, residuals,
 }
 
 gp.Friendly <- function(observed, residuals,
-                        hue = c(2/3, 0), lty = 1:2, interpolate = c(2, 4))
+                        hue = c(2/3, 0), lty = 1:2, interpolate = c(2, 4), fuzz = 0.05)
 {
-  gp.HSVshading(observed, residuals, hue = hue, value = 1, lty = lty, interpolate = interpolate, test = NULL)
+  rval <- gp.HSVshading(observed, residuals, hue = hue, value = 1, lty = lty, interpolate = interpolate, test = NULL)
+  rval$col <- ifelse(rval$lty > 1, hsv(hue[2], 1, 1), hsv(hue[1], 1, 1)) ## as in VCD book
+  rval$col <- ifelse(abs(residuals) > fuzz, rval$col, hsv(0, 1, 0))      ## with FUZZ
+  return(rval)
 }
 
 gp.max <- function(observed, residuals,

### 2004-08-18 trunk/vcd/R/shadings.R
Added several 3-way and 4-way tables from Andersen (1991)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@298 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -126,8 +126,9 @@ gp.Friendly <- function(observed, residuals,
                         hue = c(2/3, 0), lty = 1:2, interpolate = c(2, 4), fuzz = 0.05)
 {
   rval <- gp.HSVshading(observed, residuals, hue = hue, value = 1, lty = lty, interpolate = interpolate, test = NULL)
-  rval$col <- ifelse(rval$lty > 1, hsv(hue[2], 1, 1), hsv(hue[1], 1, 1)) ## as in VCD book
-  rval$col <- ifelse(abs(residuals) > fuzz, rval$col, hsv(0, 1, 0))      ## with FUZZ
+  #Z# rval$col <- ifelse(rval$lty > 1, hsv(hue[2], 1, 1), hsv(hue[1], 1, 1)) ## as in VCD book
+  #Z# rval$col <- ifelse(abs(residuals) > fuzz, rval$col, hsv(0, 1, 0))      ## with FUZZ
+  #Z# rval$lty <- ifelse(abs(residuals) > fuzz, rval$lty, 1)                 ## with FUZZ
   return(rval)
 }
 

### 2004-08-24 trunk/vcd/paper/JCGS.Rnw
Added several 3-way and 4-way tables from Andersen (1991)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@298 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
*** empty log message ***
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@301 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \hypersetup{%
-  pdftitle = {Visualizing (Conditional) Independence Using Extended Association and Mosaic Plots},
-  pdfsubject = {Visualizing (Conditional) Independence},
-  pdfkeywords = {contingency tables, association plots, mosaic plots, conditional
-    inference, permutation tests},
+  pdftitle = {Residual-based Shadings for Visualizing (Conditional) Independence},
+  pdfsubject = {Residual-based Shadings},
+  pdfkeywords = {contingency tables, association plots, mosaic plots, HCL colors, 
+    HSV colors, conditional inference, permutation tests},
   pdfauthor = {Achim Zeileis, David Meyer, Kurt Hornik},
@@
 \begin{center}
-\textbf{\LARGE Visualizing (Conditional) Independence Using Extended Association and Mosaic Plots}
+\textbf{\LARGE Residual-based Shadings for Visualizing (Conditional) Independence}
 
@@
 
-data(Bundesliga)
-goals <- xtabs(~ AwayGoals + HomeGoals, data = Bundesliga, subset = Year == 1995)
+data(Arthritis)
+art <- xtabs(~ Improved + Treatment, data = Arthritis, subset = Sex == "Female")
 
-data(HairEyeColor)
-haireye <- xtabs(Freq ~ Hair + Eye, data = as.data.frame(HairEyeColor), subset = Sex == "Female")
-
-data(UCBAdmissions)
-ucb <- UCBAdmissions
-dimnames(ucb)[[1]] <- c("Admit", "Reject")
-
-data(Arthritis)
-art <- xtabs(~ Improved + Treatment, data = Arthritis)
+data(Employment)
+emp <- Employment
+emp1 <- co.table(Employment, "LayoffCause")[["Replaced"]]
 @
@@
 <<auxfunctions,echo=FALSE,results=hide>>=
-he.test <- pearson.test(haireye, return = TRUE)
-cb <- he.test$qdist(c(0.9, 0.99))
-
-gp.shading <- function(x, y) gp.HSVshading(x, y, lty = 1, test = NULL)
-shadeHSVchisq <- function(x, y) gp.HSVshading(x, y, lty = 1)
-shadeHSVmax <- function(x, y) gp.HSVshading(x, y, interpolate = cb, lty = 1)
-shadeHCLmax <- function(x, y) gp.HCLshading(x, y, interpolate = cb, lty = 1)
-shadeHSVmax2 <- function(x, y) gp.HSVshading(x, y, interpolate = c(5,10), test = NULL)
-shadeHCLchisq <- function(x, y) gp.HCLshading(x, y, lty = 1)
+pmos.Friendly <- function(x, ...) grid.mosaicplot(x, panel = TRUE, legend = FALSE,
+  labels = FALSE, space = 0.02, gp = gp.Friendly, ...)
+pmos.max <- function(x, ...) grid.mosaicplot(x, panel = TRUE, legend = FALSE,
+  labels = FALSE, space = 0.02, gp = gp.max, ...)
 
@@
 \textbf{Key Words:} Association plots; Conditional inference; 
-  Contingency tables; Mosaic plots; Permutation tests.
+  Contingency tables; HCL colors; HSV colors; Mosaic plots; Permutation tests.
 \end{abstract}
@@
 <<goals-assoc-classic,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-grid.assocplot(goals, rot = 0, gp = gp.binary)
+grid.assocplot(emp1, gp = gp.binary)
 @
 <<goals-mosaic-classic,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-mosaicplot(t(goals), shade = TRUE, main = "")
+mosaicplot(t(emp1), shade = TRUE, main = "")
 @
-\caption{Classic association and mosaic plot for the Bundesliga data.}
-\label{fig:goals-assoc-classic}
+\caption{Classic association and mosaic plot for the ``Replaced'' subset of the Employment data.}
+\label{fig:classic}
 \end{center}
@@
 \begin{center}
-<<haireye-Friendly,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-grid.mosaicplot(haireye, gp = gp.Friendly)
+<<emp1-Friendly,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
+grid.mosaicplot(emp1, gp = gp.Friendly)
 @
-<<haireye-max,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-grid.mosaicplot(haireye, gp = gp.max)
+<<emp1-max,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
+grid.mosaicplot(emp1, gp = gp.max)
 @
-\caption{Mosaic plots for the Hair-Eye-Color data.}
-\label{fig:haireye}
-\end{center}
-\end{figure}
-
-\begin{figure}[htbp]
-\begin{center}
-<<goals-Friendly,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-grid.mosaicplot(goals, gp = gp.Friendly)
-@
-<<goals-max,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-grid.mosaicplot(goals, gp = gp.max)
-@
-\caption{Mosaic plots for the Bundesliga data.}
-\label{fig:goals}
+\caption{Mosaic plots for the ``Replaced'' subset of the Employment data.}
+\label{fig:emp1}
 \end{center}
@@
 \begin{center}
-<<UCB-mosaic1,fig=TRUE,echo=FALSE,height=4,width=8>>=
-fm <- loglm(~(Gender + Admit)*Dept, data = UCBAdmissions, fitted = TRUE)
-res.fm <- (UCBAdmissions - fitted(fm))/sqrt(fitted(fm))
-dim(res.fm) <- dim(UCBAdmissions)
-my.gp <- gp.HCLshading(UCBAdmissions, aperm(res.fm))
+<<UCB-mosaic1,fig=TRUE,echo=FALSE,height=6,width=8>>=
+fm <- loglm(~(EmploymentStatus+EmploymentLength)*LayoffCause, data = emp, fitted = TRUE)
+my.gp <- gp.HCLshading(emp, residuals(fm, type = "pearson"))
 my.gp$p.value <- summary(fm)$tests[2,3]
 my.gp$lty <- c(1,1)
-grid.mosaicplot(ucb, gp = my.gp, margin = ~ (Gender + Admit)*Dept, space = c(0.1, 0.01, 0))
+grid.mosaicplot(emp, gp = my.gp, margin = ~(EmploymentStatus+EmploymentLength)*LayoffCause,
+  space = c(0.1, 0.01, 0), dir = "v")
 @
-\caption{UCB admission with mosaic display.}
-\label{fig:UCB-mosaic1}
+\caption{Employment data with mosaic display.}
+\label{fig:emp-mosaic1}
 \end{center}
@@
 \begin{center}
-<<UCB-mosaic2,fig=TRUE,echo=FALSE,height=6,width=8>>=
-mygp.max <- function(x,y) gp.max(x, y, level = (1 - c(0.1, 0.01)/6))
-tabplot(ucb, panel = function(x, ...) grid.mosaicplot(x, panel = TRUE, legend = FALSE, labels = FALSE, gp = mygp.max, space = 0.02, ...))
+<<UCB-mosaic2,fig=TRUE,echo=FALSE,height=4,width=8>>=
+tabplot(emp, panel = pmos.max)
 @
-\caption{UCB admission with conditioned mosaic display.}
-\label{fig:UCB-mosaic2}
+\caption{Employment data with conditioned mosaic display.}
+\label{fig:emp-mosaic2}
 \end{center}
@@
 \begin{center}
-<<UCB-assoc,fig=TRUE,echo=FALSE,height=8,width=8>>=
-tabplot(margin.table(ucb, c(2,1,3)), gp = mygp.max)
+<<UCB-assoc,fig=TRUE,echo=FALSE,height=4,width=8>>=
+tabplot(emp)
 @
-\caption{UCB admission with conditioned association plot.}
-\label{fig:UCB-assoc}
+\caption{Employment data with conditioned association plot.}
+\label{fig:emp-assoc}
 \end{center}

### 2004-10-26 trunk/vcd/inst/doc/dsc2003-independence.Rnw
fix for namespace, vignette, and tabplot to enable build.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@306 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 <<goals-assoc-classic,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-grid.assocplot(goals, rot = 0, gp = gp.binary)
+assoc(goals, shade = TRUE, gp = gp.binary())
 @
@@
 <<goals-mosaic-classic,fig=TRUE,echo=FALSE,height=6,width=7>>=
-mosaicplot(t(goals), shade = TRUE, main = "")
+mosaic(goals, shade = TRUE)
 @
@@
 <<goals-assoc-signif,fig=TRUE,echo=FALSE,height=6,width=6>>=
-grid.assocplot(goals, rot = 0)
+assoc(goals)
 @
@@
 <<haireye-assoc-signif,fig=TRUE,echo=FALSE,height=6,width=6>>=
-grid.assocplot(haireye)
+assoc(haireye)
 @
@@
 <<goals-mosaic-signif,fig=TRUE,echo=FALSE,height=6,width=6>>=
-grid.mosaicplot(t(goals), main = NULL)
+mosaic(goals)
 @
 <<haireye-mosaic-signif,fig=TRUE,echo=FALSE,height=6,width=6>>=
-grid.mosaicplot(t(haireye), main = NULL)
+mosaic(haireye)
 @
@@
 ##grid.mosaicpairs(UCBAdmissions, type = "conditional", main = NULL)
-grid.mosaicpairs(UCBAdmissions, type = "conditional", permute = FALSE)
+pairs(UCBAdmissions, type = "conditional")
 @

### 2004-10-26 trunk/vcd/R/agreementplot.R

@@ -162,3 +162,246 @@ function (formula, data = NULL, ..., subset)
 
 
 
+Kappa <- function (x, weights = c("Equal-Spacing", "Fleiss-Cohen"))
+{
+  if (is.character(weights))
+      weights <- match.arg(weights)
+
+  d  <- diag(x)
+  n  <- sum(x)
+  nc <- ncol(x)
+  colFreqs <- colSums(x)/n
+  rowFreqs <- rowSums(x)/n
+  
+  ## Kappa
+  kappa <- function (po, pc)
+    (po - pc) / (1 - pc)
+  std  <- function (po, pc, W = 1)
+    sqrt(sum(W * W * po * (1 - po)) / crossprod(1 - pc) / n)
+    
+  ## unweighted
+  po <- sum(d) / n
+  pc <- crossprod(colFreqs, rowFreqs)
+  k <- kappa(po, pc)
+  s <- std(po, pc)
+  
+  ## weighted 
+  W <- if (is.matrix(weights))
+    weights
+  else if (weights == "Equal-Spacing")
+    1 - abs(outer(1:nc, 1:nc, "-")) / (nc - 1)
+  else
+    1 - (abs(outer(1:nc, 1:nc, "-")) / (nc - 1))^2
+  pow <- sum(W * x) / n
+  pcw <- sum(W * colFreqs %o% rowFreqs)
+  kw <- kappa(pow, pcw)
+  sw <- std(x / n, 1 - pcw, W)
+
+  structure(
+            list(Unweighted = c(
+                   value = k,
+                   ASE   = s
+                   ),
+                 Weighted = c(
+                   value = kw,
+                   ASE   = sw
+                   ),
+                 Weights = W
+                 ),
+            class = "Kappa"
+       )
+}
+
+print.Kappa <- function (x, ...) {
+  tab <- rbind(x$Unweighted, x$Weighted)
+  rownames(tab) <- names(x)[1:2]
+  print(tab, ...)
+  invisible(x)
+}
+
+summary.Kappa <- function (object, ...)
+  structure(object, class = "summary.Kappa")
+
+print.summary.Kappa <- function (x, ...) {
+  print.Kappa(x, ...)
+  cat("\nWeights:\n")
+  print(x$Weights, ...)
+  invisible(x)
+}
+
+confint.Kappa <- function(object, parm, level = 0.95, ...) {
+  q <- qnorm((1 + level) / 2)
[…]

### 2004-10-26 trunk/vcd/R/tabplot.R
fix for namespace, vignette, and tabplot to enable build.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@306 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -1,10 +1,10 @@
-tabplot <- function(x, panel = function(x, ...) grid.assocplot(x, panel = TRUE, legend = FALSE, ...),
+tabplot <- function(x, panel = function(x, ...) assoc(x, newpage = FALSE, ...),
                     margins = rep(1,4), ...)
 {
   UseMethod("tabplot")
 }
 
-tabplot.formula <- function(formula, panel = function(x, ...) grid.assocplot(x, panel = TRUE, legend = FALSE, ...),
+tabplot.formula <- function(formula, panel = function(x, ...) assoc(x, newpage = FALSE, ...),
                     margins = rep(1,4), data, ...)
 {
   formula <- as.character(formula)
@@ -16,8 +16,7 @@ tabplot.formula <- function(formula, panel = function(x, ...) grid.assocplot(x, 
 }
 
 
-tabplot.default <- function(x, panel = function(x, ...) grid.assocplot(x, panel = TRUE, legend = FALSE, axis.labels = FALSE, fontsize = fontsize, ...),
-                    margins = rep(1, 4), fontsize = 12, ...)
+tabplot.default <- function(x, panel = function(x, ...) assoc(x, newpage = FALSE, labels = NULL, ...), margins = rep(1, 4), fontsize = 12, ...)
 {
   grid.newpage()
 

### 2005-07-12 trunk/vcd/paper/JCGS.Rnw
started changes for JCGS paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@392 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 data(Arthritis)
-art <- xtabs(~ Improved + Treatment, data = Arthritis, subset = Sex == "Female")
+art <- xtabs(~ Treatment + Improved, data = Arthritis, subset = Sex == "Female")
 
 data(Employment)
-emp <- Employment
-emp1 <- co.table(Employment, "LayoffCause")[["Replaced"]]
+emp <- aperm(Employment)
+emp1 <- co_table(Employment, "LayoffCause")[["Replaced"]]
 @
@@
 <<auxfunctions,echo=FALSE,results=hide>>=
-pmos.Friendly <- function(x, ...) grid.mosaicplot(x, panel = TRUE, legend = FALSE,
-  labels = FALSE, space = 0.02, gp = gp.Friendly, ...)
-pmos.max <- function(x, ...) grid.mosaicplot(x, panel = TRUE, legend = FALSE,
-  labels = FALSE, space = 0.02, gp = gp.max, ...)
+## pmos.Friendly <- function(x, ...) grid.mosaicplot(x, panel = TRUE, legend = FALSE,
+##   labels = FALSE, space = 0.02, gp = gp.Friendly, ...)
+## pmos.max <- function(x, ...) grid.mosaicplot(x, panel = TRUE, legend = FALSE,
+##   labels = FALSE, space = 0.02, gp = gp.max, ...)
 
-hue.slice <- function(hue, grid.n = 100, type = c("HCL", "HSV"), plot = TRUE, correct = FALSE)
+hue.slice <- function(hue, grid.n = 100, type = c("HCL", "HSV"), plot = TRUE, fixup = FALSE)
 {
@@
     nl <- length(luminance)
-    color.slice <- outer(chroma, luminance, function(y, x) hcl(hue, x, y, verbose = FALSE, correct = correct))
+    color.slice <- outer(chroma, luminance, function(y, x) hcl(hue, x, y, fixup = fixup))
     xlab <- "chroma"
@@
 \begin{abstract}
-Story starts out not from exploratory analysis with \proglang{R}
-but from the problem of visualizing independence. Several
-exploratory approaches but not combined with significance
-testing.
 
-Paper could be structured in the following way
-\begin{itemize}
-  \item Introduction
-  \item Tests
-  \item Data sets
-  \item Visualization Techniques
-    \begin{itemize}
-      \item Mosaic displays
-      \item Association plots
-      \item Residuals-based shading
-    \end{itemize}
-  \item Extensions
-    \begin{itemize}
-      \item max-shading
-      \item HCL colors
-      \item conditional independence (coplots)
-      \item \textit{not the implementation, just mention in summary}
-    \end{itemize}
-  \item Summary
-\end{itemize}
+Visualizing independence by association and mosaic plots.
 
-The \proglang{R} side then has to be described in another paper,
-probably for the Journal of Statistical Software.\\[0.3cm]
+Enhancements: diverging color palette based on HCL color space
+combined with visualization of the result of a significance test.
+
+Extensions to visualizing conditional independence.\\[0.3cm]
+
 \textbf{Key Words:} Association plots; Conditional inference; 
@@
 
-Statistical models built for the analysis of multivariate data
-quickly become complex with increasing dimensionality.
-One idea of visualization techniques is to
-use the human visual system to detect structures in the data that
-possibly are not obvious from solely numeric output (e.g., test
-statistics).
-The \textsf{R} package \texttt{vcd}---inspired by the Book ``Visualizing Categorical
-Data'' \citep{vcd:Friendly:2000}---includes methods for the (mostly
-graphical) exploration of categorical data, such as:
-\begin{itemize}
- \item fitting and graphing of discrete distributions,
- \item plots and tests for independence and symmetry problems,
- \item visualization techniques for log-linear models.
-\end{itemize}
-Here, we focus on the visualization of independence,
-in particular in 2-way tables.
+independence in 2-way tables
 
-\section{Tests for Independence in 2-way Tables} \label{sec:tests}
+association and mosaic displays
+
+colors and color spaces
+
+\section{Independence in 2-way Tables}
+\label{sec:2way}
+
+\subsection{Tests}
+\label{sec:tests}
 
@@
 
-\section{Visualization Techniques for 2-way Tables} \label{sec:visualization}
+\section{Visualizations} \label{sec:visualization}
+
+ONLY SHAPE! not color
 
@@
 
-\subsection{Association Plots}
-
-Association plots \citep{vcd:Cohen:1980} visualize the table of Pearson residuals:
+\textbf{Association Plots}
+\citep{vcd:Cohen:1980} visualize the table of Pearson residuals:
 each cell is represented by a rectangle that has (signed) height
@@
 \subsection{Mosaic Plots}
-
-Mosaic plots can be seen as an extension of grouped
+\citep{vcd:Hartigan+Kleiner:1981} can be seen as an extension of grouped
 bar charts where width and height of the bars show the relative
@@
 
-Figure \ref{fig:goals-mosaic-classic} shows the Friendly mosaic plot for
-the same Bundesliga data as in Figure \ref{fig:goals-assoc-classic}.
-In this plot, the residuals
-are visualized by the color shading of the cells. Again, the cells for
-the 1--0 and 1--1 results are highlighted and furthermore, there seem to be
-too many 6--3 results than would be plausible under independence. Whereas
-the association plot in Figure \ref{fig:goals-assoc-classic} contains
-no information about the result of a test for independence, the shading
-in this plot conveys the impression that there might be
-significant dependence (which would indicate that there
-is not only {\em no} home field advantage in the German Bundesliga in 1995/6 but even
-an advantage for the away team). However, it is unclear whether there
-is evidence for this or not.
+\section{Residual-based Shadings} \label{sec:shadings}
 
-\section{Extensions} \label{sec:extensions}
+remarks: Friendly 1994
 
-In this section, we suggest some improvements and extensions which can be
-applied to both visualization techniques introduced in Section
-\ref{sec:visualization}. First, the visual enhancements
-\begin{itemize}
-\item combining visualization of and testing for independence
-\item improved color schemes
-\end{itemize}
-are addressed. Second, the implementation enhancements
-\begin{itemize}
-\item more modular implementation using \texttt{grid}
-\item more flexible plotting parameter specification
-\end{itemize}
-are described. None of these should be thought of as finished work
-but rather as a collection of ideas which still need thorough
-refinement and proper implementation.
+\subsection{Colors}
 
-\subsection{Visualization Enhancements}
+\subsection{Significance}
 
@@
 
-\subsection{Implementation Enhancements}
-
-The current implementation of association and mosaic plots in \textsf{R} suffers
-from two main disadvantages: First, it is not easy to recycle the
-plots in conditioning plots or pairs plots (like mosaic matrices) as
-they have been implemented using \textsf{R}'s base graphics engine where in
-general plotting to relative coordinates is not supported.
-The new implementation was written from scratch in \texttt{grid} offering
-much more versatility amongst some minor advantages and convenient
-improvements. Second, the graphics parameters of the rectangles in
-association and mosaic plots (like in almost all standard \textsf{R}
-plots), such as color and line type, cannot be specified for each cell
-by the user. To overcome this the current implementation in \texttt{vcd}
-allows the user to specify either arrays of graphics parameters
-of the same dimensionality as the object being plotted or a function
-which computes these graphics parameters based on the original table
-and its Pearson residuals. Functions are provided for the shading schemes
-described in the previous section. Without this feature we would not be
-able to specify the shading schemes described in the previous section
-in such a flexible way.
-
-\section{Multi-way Tables}
+\section{Extensions to Conditional Independence}
 
@@
 $(\pi_{ij\dots})$ with more than two dimensions. Models of interest include the
-null hypotheses of:
-\begin{itemize}
- \item total independence:
-  $\pi_{ijk\dots} = \pi_{i++\dots}\pi_{+j+\dots}\pi_{++k\dots}\dots$
- \item conditional independence:
-  $\pi_{ijk\dots} = \pi_{i|k\dots}\pi_{j|k\dots}$
- \item joint independence:
-  $\pi_{ijk\dots} = \pi_{ij+\dots}\pi_{++k\dots}$
-\end{itemize}
+null hypotheses of conditional independence:
+  \[ \pi_{ijk\dots} = \pi_{i|k\dots}\pi_{j|k\dots} \]
+
 Classical non-graphical methods for these problems include the $\chi^2$
@@
 \section{Conclusions}
-We suggest a set of enhancements for visualizing the independence
-problem in 2-way tables with some outlook to multi-way tables. The
-extensions aim at improving the visualization by displaying both
-the size of the residuals and the significance of a test for
-independence and by using better color schemes. Furthermore, a new
-implementation is outlined based on the graphics package \texttt{grid}
-which provides more modular design and more flexible specification
-of graphical parameters.
 
@@
 <<goals-assoc-classic,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-grid.assocplot(emp1, gp = gp.binary)
+assoc(emp1)
 @
 <<goals-mosaic-classic,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-mosaicplot(t(emp1), shade = TRUE, main = "")
+mosaic(emp1, split_vertical = TRUE)
 @
@@
 \begin{center}
-<<emp1-Friendly,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-grid.mosaicplot(emp1, gp = gp.Friendly)
-@
-<<emp1-max,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-grid.mosaicplot(emp1, gp = gp.max)
-@
-\caption{Mosaic plots for the ``Replaced'' subset of the Employment data.}
-\label{fig:emp1}
-\end{center}
-\end{figure}
-
-\begin{figure}[htbp]
-\begin{center}
 <<art-Friendly,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-grid.mosaicplot(art, gp = gp.Friendly)
+mosaic(art, gp = shading_Friendly)
 @
 <<art-max,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-grid.mosaicplot(art, gp = gp.max)
+mosaic(art, gp = shading_max)
 @
@@
 \label{fig:art}
-\end{center}
-\end{figure}
-
-\setkeys{Gin}{width=0.8\textwidth}
-\begin{figure}[htbp]
-\begin{center}
-<<Friendly-hue,fig=TRUE,echo=FALSE,height=4.8,width=8>>=
-plot(0.5, 0.5, xlim = c(0, 1), ylim = c(0, 1), type = "n", axes = FALSE, xlab = "", ylab = "", yaxs = "i")
-dummy <- 0:100/100
-dummy.col <- hsv(dummy[-101] + 1/200, 1, 1)
-rect(dummy[-101], 0.6, dummy[-1], 1, border = dummy.col, col = dummy.col)
-rect(0, 0.6, 1, 1)
-axis(3)
-mtext("hue", line = 2, cex = 1.5)
-text(1, 0.45, "saturation = 1\nvalue = 1", pos = 2, cex = 1.2)
-lines(c(0.1, 0), c(0.3, 0.6), lty = 2)
-lines(c(2/3, 2/3), c(0.3, 0.6), lty = 2)
-rect(0, 0.1, 0.2, 0.3, col = hsv(0, 1, 1))
-rect(2/3 - 0.1, 0.1, 2/3 + 0.1, 0.3, col = hsv(2/3, 1, 1))
-text(0.1, 0.1, expression(r[ij] < 0), pos = 1, cex = 1.5)
-text(2/3, 0.1, expression(r[ij] > 0), pos = 1, cex = 1.5)
-@
-
-<<Friendly-sat,fig=TRUE,echo=FALSE,height=4.8,width=8>>=
-plot(0.5, 0.5, xlim = c(0, 1), ylim = c(0, 1), type = "n", axes = FALSE, xlab = "", ylab = "", yaxs = "i")
-dummy <- 0:100/100
-dummy.col <- hsv(2/3, dummy[-101] + 1/200, 1)
-rect(dummy[-101], 0.8, dummy[-1], 1, border = dummy.col, col = dummy.col)
-dummy.col <- hsv(0, dummy[-101] + 1/200, 1)
-rect(dummy[-101], 0.6, dummy[-1], 0.8, border = dummy.col, col = dummy.col)
-rect(0, 0.8, 1, 1)
-rect(0, 0.6, 1, 0.8)
-axis(3)
-mtext("saturation", line = 2, cex = 1.5)
-mtext(c("h = 0", "h = 2/3"), side = 2, at = c(0.7, 0.9), las = 1, cex = 1.2)
-text(0.8, 0.5, "value = 1", pos = 2, cex = 1.2)
-lines(c(0.1, 0), c(0.3, 0.6), lty = 2)
-lines(c(0.5, 0.5), c(0.3, 0.6), lty = 2)
-lines(c(0.9, 1), c(0.3, 0.6), lty = 2)
-rect(0, 0.2, 0.2, 0.3, col = hsv(2/3, 0, 1))
-rect(0, 0.1, 0.2, 0.2, col = hsv(0, 0, 1))
-rect(0.4, 0.2, 0.6, 0.3, col = hsv(2/3, 0.5, 1))
-rect(0.4, 0.1, 0.6, 0.2, col = hsv(0, 0.5, 1))
-rect(0.8, 0.2, 1, 0.3, col = hsv(2/3, 1, 1))
-rect(0.8, 0.1, 1, 0.2, col = hsv(0, 1, 1))
-text(0.1, 0.1, expression(abs(r[ij]) < 2), pos = 1, cex = 1.5)
-text(0.5, 0.1, expression(2 < abs(r[ij]) < 4), pos = 1, cex = 1.5)
-text(0.9, 0.1, expression(abs(r[ij]) > 4), pos = 1, cex = 1.5)
-@
-
-<<Friendly-val,fig=TRUE,echo=FALSE,height=4.8,width=8>>=
-plot(0.5, 0.5, xlim = c(0, 1), ylim = c(0, 1), type = "n", axes = FALSE, xlab = "", ylab = "", yaxs = "i")
-dummy <- 0:100/100
-dummy.col <- hsv(2/3, 1, dummy[-101] + 1/200)
-rect(dummy[-101], 0.8, dummy[-1], 1, border = dummy.col, col = dummy.col)
-dummy.col <- hsv(0, 1, dummy[-101] + 1/200)
-rect(dummy[-101], 0.6, dummy[-1], 0.8, border = dummy.col, col = dummy.col)
-rect(0, 0.8, 1, 1)
-rect(0, 0.6, 1, 0.8)
-axis(3)
-mtext("value", line = 2, cex = 1.5)
-mtext(c("h = 0", "h = 2/3"), side = 2, at = c(0.7, 0.9), las = 1, cex = 1.2)
-text(0, 0.5, paste("saturation =", 1), pos = 4, cex = 1.2)
-lines(c(0.85, 1), c(0.3, 0.6), lty = 2)
-rect(0.7, 0.2, 0.8, 0.3, col = hsv(2/3, 0, 1))
-rect(0.8, 0.2, 0.9, 0.3, col = hsv(2/3, 0.5, 1))
-rect(0.9, 0.2, 1, 0.3, col = hsv(2/3, 1, 1))
-rect(0.7, 0.1, 0.8, 0.2, col = hsv(0, 0, 1))
-rect(0.8, 0.1, 0.9, 0.2, col = hsv(0, 0.5, 1))
-rect(0.9, 0.1, 1, 0.2, col = hsv(0, 1, 1))
-lines(c(0.5, 0.5), c(0.3, 0.6), lty = 2)
-lines(c(0.85, 1), c(0.3, 0.6), lty = 2)
-rect(0.35, 0.2, 0.45, 0.3, col = hsv(2/3, 0, 0.5))
-rect(0.45, 0.2, 0.55, 0.3, col = hsv(2/3, 0.5, 0.5))
-rect(0.55, 0.2, 0.65, 0.3, col = hsv(2/3, 1, 0.5))
-rect(0.35, 0.1, 0.45, 0.2, col = hsv(0, 0, 0.5))
-rect(0.45, 0.1, 0.55, 0.2, col = hsv(0, 0.5, 0.5))
-rect(0.55, 0.1, 0.65, 0.2, col = hsv(0, 1, 0.5))
-rect(0.7, 0.2, 0.8, 0.3, col = hsv(2/3, 0, 1))
-rect(0.8, 0.2, 0.9, 0.3, col = hsv(2/3, 0.5, 1))
-rect(0.9, 0.2, 1, 0.3, col = hsv(2/3, 1, 1))
-rect(0.7, 0.1, 0.8, 0.2, col = hsv(0, 0, 1))
-rect(0.8, 0.1, 0.9, 0.2, col = hsv(0, 0.5, 1))
-rect(0.9, 0.1, 1, 0.2, col = hsv(0, 1, 1))
-text(0.5, 0.1, "non-significant", pos = 1, cex = 1.5)
-text(0.85, 0.1, "significant", pos = 1, cex = 1.5)
-@
-
-\caption{Extended Friendly shading in HSV space.}
-\label{fig:Friendly-HSV}
 \end{center}
@@
 <<UCB-mosaic1,fig=TRUE,echo=FALSE,height=6,width=8>>=
-fm <- loglm(~(EmploymentStatus+EmploymentLength)*LayoffCause, data = emp, fitted = TRUE)
-my.gp <- gp.HCLshading(emp, residuals(fm, type = "pearson"))
-my.gp$p.value <- summary(fm)$tests[2,3]
-my.gp$lty <- c(1,1)
-grid.mosaicplot(emp, gp = my.gp, margin = ~(EmploymentStatus+EmploymentLength)*LayoffCause,
-  space = c(0.1, 0.01, 0), dir = "v")
+## fm <- loglm(~(EmploymentStatus+EmploymentLength)*LayoffCause, data = emp, fitted = TRUE)
+## my.gp <- gp.HCLshading(emp, residuals(fm, type = "pearson"))
+## my.gp$p.value <- summary(fm)$tests[2,3]
+## my.gp$lty <- c(1,1)
+## grid.mosaicplot(emp, gp = my.gp, margin = ~(EmploymentStatus+EmploymentLength)*LayoffCause,
+##   space = c(0.1, 0.01, 0), dir = "v")
+mosaic(emp, expected = ~(EmploymentStatus+EmploymentLength)*LayoffCause)
 @
@@
 
+\nocite{*}
+\bibliography{vcd}
+
+\end{document}
+
+\setkeys{Gin}{width=0.8\textwidth}
 \begin{figure}[htbp]
 \begin{center}
-<<UCB-mosaic2,fig=TRUE,echo=FALSE,height=4,width=8>>=
+<<Friendly-hue,fig=TRUE,echo=FALSE,height=4.8,width=8>>
+plot(0.5, 0.5, xlim = c(0, 1), ylim = c(0, 1), type = "n", axes = FALSE, xlab = "", ylab = "", yaxs = "i")
+dummy <- 0:100/100
+dummy.col <- hsv(dummy[-101] + 1/200, 1, 1)
+rect(dummy[-101], 0.6, dummy[-1], 1, border = dummy.col, col = dummy.col)
+rect(0, 0.6, 1, 1)
+axis(3)
+mtext("hue", line = 2, cex = 1.5)
+text(1, 0.45, "saturation = 1\nvalue = 1", pos = 2, cex = 1.2)
+lines(c(0.1, 0), c(0.3, 0.6), lty = 2)
+lines(c(2/3, 2/3), c(0.3, 0.6), lty = 2)
+rect(0, 0.1, 0.2, 0.3, col = hsv(0, 1, 1))
+rect(2/3 - 0.1, 0.1, 2/3 + 0.1, 0.3, col = hsv(2/3, 1, 1))
+text(0.1, 0.1, expression(r[ij] < 0), pos = 1, cex = 1.5)
+text(2/3, 0.1, expression(r[ij] > 0), pos = 1, cex = 1.5)
+@
+
+<<Friendly-sat,fig=TRUE,echo=FALSE,height=4.8,width=8>>
+plot(0.5, 0.5, xlim = c(0, 1), ylim = c(0, 1), type = "n", axes = FALSE, xlab = "", ylab = "", yaxs = "i")
+dummy <- 0:100/100
+dummy.col <- hsv(2/3, dummy[-101] + 1/200, 1)
+rect(dummy[-101], 0.8, dummy[-1], 1, border = dummy.col, col = dummy.col)
+dummy.col <- hsv(0, dummy[-101] + 1/200, 1)
+rect(dummy[-101], 0.6, dummy[-1], 0.8, border = dummy.col, col = dummy.col)
+rect(0, 0.8, 1, 1)
+rect(0, 0.6, 1, 0.8)
+axis(3)
+mtext("saturation", line = 2, cex = 1.5)
+mtext(c("h = 0", "h = 2/3"), side = 2, at = c(0.7, 0.9), las = 1, cex = 1.2)
+text(0.8, 0.5, "value = 1", pos = 2, cex = 1.2)
+lines(c(0.1, 0), c(0.3, 0.6), lty = 2)
+lines(c(0.5, 0.5), c(0.3, 0.6), lty = 2)
+lines(c(0.9, 1), c(0.3, 0.6), lty = 2)
+rect(0, 0.2, 0.2, 0.3, col = hsv(2/3, 0, 1))
+rect(0, 0.1, 0.2, 0.2, col = hsv(0, 0, 1))
+rect(0.4, 0.2, 0.6, 0.3, col = hsv(2/3, 0.5, 1))
+rect(0.4, 0.1, 0.6, 0.2, col = hsv(0, 0.5, 1))
+rect(0.8, 0.2, 1, 0.3, col = hsv(2/3, 1, 1))
+rect(0.8, 0.1, 1, 0.2, col = hsv(0, 1, 1))
+text(0.1, 0.1, expression(abs(r[ij]) < 2), pos = 1, cex = 1.5)
+text(0.5, 0.1, expression((2 < abs(r[ij])) < 4), pos = 1, cex = 1.5)
+text(0.9, 0.1, expression(abs(r[ij]) > 4), pos = 1, cex = 1.5)
+@
+
+<<Friendly-val,fig=TRUE,echo=FALSE,height=4.8,width=8>>
+plot(0.5, 0.5, xlim = c(0, 1), ylim = c(0, 1), type = "n", axes = FALSE, xlab = "", ylab = "", yaxs = "i")
+dummy <- 0:100/100
+dummy.col <- hsv(2/3, 1, dummy[-101] + 1/200)
+rect(dummy[-101], 0.8, dummy[-1], 1, border = dummy.col, col = dummy.col)
+dummy.col <- hsv(0, 1, dummy[-101] + 1/200)
+rect(dummy[-101], 0.6, dummy[-1], 0.8, border = dummy.col, col = dummy.col)
+rect(0, 0.8, 1, 1)
+rect(0, 0.6, 1, 0.8)
+axis(3)
+mtext("value", line = 2, cex = 1.5)
+mtext(c("h = 0", "h = 2/3"), side = 2, at = c(0.7, 0.9), las = 1, cex = 1.2)
+text(0, 0.5, paste("saturation =", 1), pos = 4, cex = 1.2)
+lines(c(0.85, 1), c(0.3, 0.6), lty = 2)
+rect(0.7, 0.2, 0.8, 0.3, col = hsv(2/3, 0, 1))
+rect(0.8, 0.2, 0.9, 0.3, col = hsv(2/3, 0.5, 1))
+rect(0.9, 0.2, 1, 0.3, col = hsv(2/3, 1, 1))
+rect(0.7, 0.1, 0.8, 0.2, col = hsv(0, 0, 1))
+rect(0.8, 0.1, 0.9, 0.2, col = hsv(0, 0.5, 1))
+rect(0.9, 0.1, 1, 0.2, col = hsv(0, 1, 1))
+lines(c(0.5, 0.5), c(0.3, 0.6), lty = 2)
+lines(c(0.85, 1), c(0.3, 0.6), lty = 2)
+rect(0.35, 0.2, 0.45, 0.3, col = hsv(2/3, 0, 0.5))
+rect(0.45, 0.2, 0.55, 0.3, col = hsv(2/3, 0.5, 0.5))
+rect(0.55, 0.2, 0.65, 0.3, col = hsv(2/3, 1, 0.5))
+rect(0.35, 0.1, 0.45, 0.2, col = hsv(0, 0, 0.5))
+rect(0.45, 0.1, 0.55, 0.2, col = hsv(0, 0.5, 0.5))
+rect(0.55, 0.1, 0.65, 0.2, col = hsv(0, 1, 0.5))
+rect(0.7, 0.2, 0.8, 0.3, col = hsv(2/3, 0, 1))
+rect(0.8, 0.2, 0.9, 0.3, col = hsv(2/3, 0.5, 1))
+rect(0.9, 0.2, 1, 0.3, col = hsv(2/3, 1, 1))
+rect(0.7, 0.1, 0.8, 0.2, col = hsv(0, 0, 1))
+rect(0.8, 0.1, 0.9, 0.2, col = hsv(0, 0.5, 1))
+rect(0.9, 0.1, 1, 0.2, col = hsv(0, 1, 1))
+text(0.5, 0.1, "non-significant", pos = 1, cex = 1.5)
+text(0.85, 0.1, "significant", pos = 1, cex = 1.5)
+@
+
+\caption{Extended Friendly shading in HSV space.}
+\label{fig:Friendly-HSV}
+\end{center}
+\end{figure}
+
+\begin{figure}[htbp]
+\begin{center}
+<<emp1-Friendly,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>
+mosaic(emp1, gp = shading_Friendly)
+@
+<<emp1-max,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>
+mosaic(emp1, gp = shading_max)
+@
+\caption{Mosaic plots for the ``Replaced'' subset of the Employment data.}
+\label{fig:emp1}
+\end{center}
+\end{figure}
+
+\begin{figure}[htbp]
+\begin{center}
+<<UCB-mosaic2,fig=TRUE,echo=FALSE,height=4,width=8>>
 tabplot(emp, panel = pmos.max)
@@
 \begin{center}
-<<UCB-assoc,fig=TRUE,echo=FALSE,height=4,width=8>>=
+<<UCB-assoc,fig=TRUE,echo=FALSE,height=4,width=8>>
 tabplot(emp)
@@
 \end{figure}
-
-\nocite{*}
-\bibliography{vcd}
-
-\end{document}

### 2005-07-18 trunk/vcd/paper/JCGS.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@394 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
+somewhere:
+mention Augsburg stuff, visualization of log-linear models via mosaics
+\citep{vcd:Theus+Lauer:1999,vcd:Hofmann:2003,vcd:Hofmann:2001}.
+
+Association plots as residual plots for log-linear models
+\citep{vcd:Meyer+Zeileis+Hornik:2003}, especially
+in coplot or trellis layout.
+
+Arthritis data \citep{vcd:Koch+Edwards:1988}
+
+UCB admissions data \citep{vcd:Bickel+Hammel+O'Connell:1975}
+
 \section{Independence in 2-way Tables}
@@
 
-Figure \ref{fig:goals-assoc-classic} shows the association plot for the table of
-home and away goals for all 306 games in the 1995/6 season of the German soccer league
-Bundesliga \citep[the data is available in the package \texttt{vcd}]{vcd:Knorr-Held:1999}.
-In particular, it can be seen that there are
-fewer games ending 1--0 and more ending 1--1 than would be expected under independence.
-But this does not yet imply the significance of some test statistic.
+%% Figure \ref{fig:goals-assoc-classic} shows the association plot for the table of
+%% home and away goals for all 306 games in the 1995/6 season of the German soccer league
+%% Bundesliga \citep[the data is available in the package \texttt{vcd}]{vcd:Knorr-Held:1999}.
+%% In particular, it can be seen that there are
+%% fewer games ending 1--0 and more ending 1--1 than would be expected under independence.
+%% But this does not yet imply the significance of some test statistic.
 
-\subsection{Mosaic Plots}
+\textbf{Mosaic Plots}
 \citep{vcd:Hartigan+Kleiner:1981} can be seen as an extension of grouped
@@
 
+currently: chosen based on HSV or HLS color spaces,
+hue (red and blue) codes sign of residuals, and saturation the absolute size
+in three steps, value currently unused (always set to full value)
+
+disadvantages:
+  - perceptually not uniform
+    because the three perceptual dimensions (hue, lightness, saturation)
+    are not properly mapped to the three dimensions of the color space
+    and hence are confounded \citep{vcd:Brewer:1999}    
+    in particular, saturation is not uniform across different hues.
+
+  - can lead to color-caused optical illusions in statistical graphs
+    \citep{vcd:Cleveland+McGill:1983}
+
+  - flashy fully saturated HSV colors are good for drawing attention to a plot,
+    but hard to look at for a longer time, hence mosaic displays using these 
+    colors are harder to interpret
+
+  - white is not very suitable as a neutral color, grey conveys neutrality or 
+    uninterestingness much better
+
+Alternatives:
+  - \cite{vcd:Munsell:1905} introduced color notation for balanced colors
+  - Commission Internationale de l'\'Eclairage (CIE) introduced two perceptually
+    based color spaces CIELAB and CIELUV where the latter is preferred for 
+    emissive color technologies such as computer displays.
+  - \pkg{ColorBrewer.org} \citep{vcd:Harrower+Brewer:2003} is an online tool
+    for selcting different sets of colors, based on similar ideas. It includes
+    qualitative, sequential and diverging palettes.
+  - \cite{vcd:Ihaka:2003} discusses how CIELUV colors can be used for choosing
+    colors for statistical graphics such as barplots. He recommends to use
+    polar coordinates in the CIELUV space---called HCL space, using hue, chroma
+    and luminance---and suggests a strategy for choosing qualitative schemes based
+    on these.
+
+We discuss how similar ideas can be used for deriving a diverging HCL scheme that 
+is suitable for visualizing residuals in association and mosaic displays.
+
+HCL
+  - irregular double cone, i.e., chroma/luminance plane looks rather different
+    for different hues
+  - choose two hues that look similar in the chroma/luminance plane to code
+    the sign of the residuals. Also use red and blue at full chroma for large
+    residuals and use a light gray (instead of a white at the top of the luminance
+    scale) as the neutral color for small residuals, interpolate between.
+  - for non-significant results, clearly reduce the amount of color, i.e. the maximum
+    chroma used.
+
+Advantages
+  - based on a perceptually uniform color space
+  - device independent
+
 \subsection{Significance}
@@
 generates very similar colors
-as increasing the luminance for HSL colors from 0.5 towards 1.
+as increasing the luminance for HLS colors from 0.5 towards 1.
 As an alternative to those HLS and HSV colors, the device-independent
@@
 \section{Extensions to Conditional Independence}
+
+\cite{vcd:Friendly:1999} discusses grouping similar to coplots (conditioning
+plots) \citep{vcd:Cleveland:1993} that lead to trellis graphics
+(Becker, R.A., Cleveland, W.S. and Shyu, M. ``The Visual Design and
+Control of Trellis Display'', Journal of Computational and
+Graphical Statistics).
 
@@
 
-\nocite{*}
 \bibliography{vcd}

### 2005-07-20 trunk/vcd/paper/JCGS.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@402 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-<<preliminaries,echo=FALSE,results=hide>>=
-set.seed(12345)
+<<load_data,echo=FALSE,results=hide>>=
 library(vcd)
@@
 
-data(Employment)
-emp <- aperm(Employment)
-emp1 <- co_table(Employment, "LayoffCause")[["Replaced"]]
+data(UCBAdmissions)
+ucb <- aperm(UCBAdmissions)
 @
 
-<<auxfunctions,echo=FALSE,results=hide>>=
-## pmos.Friendly <- function(x, ...) grid.mosaicplot(x, panel = TRUE, legend = FALSE,
-##   labels = FALSE, space = 0.02, gp = gp.Friendly, ...)
-## pmos.max <- function(x, ...) grid.mosaicplot(x, panel = TRUE, legend = FALSE,
-##   labels = FALSE, space = 0.02, gp = gp.max, ...)
-
+<<auxiliary_functions,echo=FALSE,results=hide>>=
 hue.slice <- function(hue, grid.n = 100, type = c("HCL", "HSV"), plot = TRUE, fixup = FALSE)
@@
 
-\section{Visualizations} \label{sec:visualization}
+\subsection{Visualizations} \label{sec:visualization}
 
@@
 values $c_\alpha = 2$ and $4$ correspond for the table being visualized.
+
+\setkeys{Gin}{width=\textwidth}
+\begin{figure}[htbp]
+\begin{center}
+<<classic_plots,fig=TRUE,echo=FALSE,height=6,width=11,results=hide>>=
+grid.newpage()
+pushViewport(viewport(layout = grid.layout(1, 2)))
+pushViewport(viewport(layout.pos.col=1, layout.pos.row=1))
+mosaic(art, newpage = FALSE, margins = c(2.5, 4, 2.5, 3))
+popViewport()
+pushViewport(viewport(layout.pos.col=2, layout.pos.row=1))
+assoc(art, newpage = FALSE, margins = c(5, 2, 5, 4))
+popViewport(2)
+@
+\caption{Classic mosaic and association plot for the Arthritis data.}
+\label{fig:classic}
+\end{center}
+\end{figure}
 
@@
 
-\subsection{Significance}
-
-The extensions of \cite{vcd:Friendly:1994} to mosaic plots provide
-substantial improvement of the original mosaic plots and enhanced them
-from a plot for contingency tables to a visualization technique for
-log-linear models (and therefore also for certain independence problems).
-However, it has two major drawbacks: First,
-the significance level for the hard-coded critical values 2 and 4 is
-usually unknown. Second, the colors for (light) blue and red
-used in the Michael Friendly's \textsf{SAS} implementation and also in the base implementation
-in \textsf{R} are not device-independent and
-do not provide homogeneous saturations over different colors
-and copier proofness. We suggest some ways to overcome these drawbacks.
-
-As pointed out in Section \ref{sec:tests}, the critical values for
-the maximum statistic $M$ from (\ref{eq:MaxAbs}) can be derived
-from the permutation distribution. Instead of the hard-coded
-values 2 and 4, the particular critical values, e.g., at the levels
-$\alpha = 0.1$ and $0.01$, for the table to be visualized could be
-used. Hence, exactly those residuals causing the (potential)
-dependency within the table are highlighted. At the moment, these critical values
-are derived in \texttt{vcd} by simulation of the underlying distribution.
-
-The implementation of Friendly mosaic plots in \textsf{SAS}
-uses (by default) colors in the HLS (Hue--Luminance--Saturation) color space
-with hues for blue and red, full saturation, and varying luminance
-for lighter colors. The current \textsf{R} implementation is
-based on colors in the HSV (Hue--Saturation--Value) space where
-decreasing the saturation from 1 towards 0
-%% ---given the hue for blue or red and full saturation---
-generates very similar colors
-as increasing the luminance for HLS colors from 0.5 towards 1.
-As an alternative to those HLS and HSV colors, the device-independent
-HCL (Hue--Chroma--Luminance) color space \citep{vcd:Ihaka:2003} could be
-used which also provides homogeneous saturations over different colors and copier
-proofness. The resulting association plot (using critical values for
-$\alpha = 0.1$ and $0.01$) for the Bundesliga data can be seen in
-Figure \ref{fig:goals-assoc-signif}. It indicates very clearly that
-there is no evidence for rejecting the hypothesis of independence
-in this table. Thus, there seems to be neither an advantage for
-the home team nor for the away team.
-
-To illustrate what a `postive' example looks like we give the association plot
-for the well-known Hair-Eye-Color data set (avilable in base \textsf{R}) which cross-tabulates the hair and eye
-color and gender of 592 statistics students: Figure \ref{fig:haireye-assoc-signif}
-shows the association plot for the 328 female students. It is easy to see that
-the hypothesis of independence is rejected at 1\% level by the maximum test (\ref{eq:MaxAbs})
-due to higher frequencies of students with hair and eye color
-black/brown and blond/blue and lower frequencies of
-blond/brown and brown/blue than would be plausible under independence. Furthermore,
-it can also be seen that the residual corresponding to the black/blue
-cell is significant at the 10\% level.
-
-We have seen that this approach performs very well when the test statistic
-which should be visualized is the maximum statistic (\ref{eq:MaxAbs}). However,
-it is difficult to extend this approach to general tests for independence,
-in particular in the HCL space, where the ranges of chroma and luminance are
-not independent. In contrast the HSV space has the advantage that all
-three dimensions hue, saturation and value have the range $[0,1]$ and can
-be varied independently. Therefore, a different approach than the one
-described above could be to use the hue for the sign of the residuals as
-before, the saturation for the absolute size of the residuals, and the
-value as an indicator for the significance of some test statistic (only
-using the full value when the overall test for independence rejects
-the null hypothesis). The resulting mosaic plots under this paradigm
-are depicted in Figure \ref{fig:both-mosaic-signif}. Again, it can
-clearly be seen that despite some large residuals there is no
-evidence against independence for the Bundesliga data, but that the
-null hypothesis has to be rejected for the Hair-Eye-Color data.
-
-\section{Extensions to Conditional Independence}
-
-\cite{vcd:Friendly:1999} discusses grouping similar to coplots (conditioning
-plots) \citep{vcd:Cleveland:1993} that lead to trellis graphics
-(Becker, R.A., Cleveland, W.S. and Shyu, M. ``The Visual Design and
-Control of Trellis Display'', Journal of Computational and
-Graphical Statistics).
-
-Independence problems do not only occur in 2-way tables, although that
-is an important special case, but they are also important in tables of
-higher dimensionality and can follow much more complex patterns.
-These are again defined based on the underlying table of theoretical cell probabilities
-$(\pi_{ij\dots})$ with more than two dimensions. Models of interest include the
-null hypotheses of conditional independence:
-  \[ \pi_{ijk\dots} = \pi_{i|k\dots}\pi_{j|k\dots} \]
-
-Classical non-graphical methods for these problems include the $\chi^2$
-test, Fisher's exact test, the Cochran-Mantel-Haenzel test (for $2 \times 2 \times
-K$-tables), and the analysis of log-linear models for more
-complex settings.
-
-As an example, two natural ways to use the visualization techniques described in the previous
-sections would be to use (Trellis-like) conditioning plots or pairs plots
-(like mosaic matrices) to visualize these more complex patterns of
-independence.
-
-Two ideas for the problem of conditional independence are briefly outlined
-here and illustrated using the famous admissions data of the University of
-California at Berkley (UCB) which is available in base \textsf{R}.
-In this data, the question whether there is gender
-discrimination at the UCB leads to the result that although women seem to
-be disadvantaged at the aggregated level there is no gender discrimination
-conditioned on the department---with the very exception of one department
-in which women are {\em more} likely to be admitted than would be plausible
-under independence. Exactly this is illustrated in the conditioning
-assocation plot in Figure \ref{fig:UCB-assoc-signif}.
-
-Similarly, the same data can be visualized using a mosaic matrix where
-a conditional independence model is fitted in each plot (see Figure
-\ref{fig:UCB-mosaic-signif}).
-
-\section{Conclusions}
-
-\setkeys{Gin}{width=0.7\textwidth}
-\begin{figure}[htbp]
+\setkeys{Gin}{width=.8\textwidth}
+\begin{figure}[p]
 \begin{center}
-<<goals-assoc-classic,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-assoc(emp1)
-@
-<<goals-mosaic-classic,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-mosaic(emp1, split_vertical = TRUE)
-@
-\caption{Classic association and mosaic plot for the ``Replaced'' subset of the Employment data.}
-\label{fig:classic}
-\end{center}
-\end{figure}
-
-\begin{figure}[htbp]
-\begin{center}
-<<art-Friendly,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-mosaic(art, gp = shading_Friendly)
-@
-<<art-max,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-mosaic(art, gp = shading_max)
-@
-\caption{Mosaic plots for the Arthritis data.}
-\label{fig:art}
-\end{center}
-\end{figure}
-
-\begin{figure}[htbp]
-\begin{center}
-<<Friendly-HSV,fig=TRUE,echo=FALSE,height=7,width=8>>=
+<<shading_HSV,fig=TRUE,echo=FALSE,height=7,width=8>>=
 ## generate colors
@@
 
-<<HCL-space,fig=TRUE,echo=FALSE,height=7,width=8>>=
+<<shading_HCL,fig=TRUE,echo=FALSE,height=7,width=8>>=
 ## generate colors
@@
 \caption{Extended shading in HSV and HCL space.}
-\label{fig:ExtHSVHCL}
+\label{fig:shadingHSVHCL}
 \end{center}
 \end{figure}
+
+\subsection{Significance}
+
+The extensions of \cite{vcd:Friendly:1994} to mosaic plots provide
+substantial improvement of the original mosaic plots and enhanced them
+from a plot for contingency tables to a visualization technique for
+log-linear models (and therefore also for certain independence problems).
+However, it has two major drawbacks: First,
+the significance level for the hard-coded critical values 2 and 4 is
+usually unknown. Second, the colors for (light) blue and red
+used in the Michael Friendly's \textsf{SAS} implementation and also in the base implementation
+in \textsf{R} are not device-independent and
+do not provide homogeneous saturations over different colors
+and copier proofness. We suggest some ways to overcome these drawbacks.
+
+As pointed out in Section \ref{sec:tests}, the critical values for
+the maximum statistic $M$ from (\ref{eq:MaxAbs}) can be derived
+from the permutation distribution. Instead of the hard-coded
+values 2 and 4, the particular critical values, e.g., at the levels
+$\alpha = 0.1$ and $0.01$, for the table to be visualized could be
+used. Hence, exactly those residuals causing the (potential)
+dependency within the table are highlighted. At the moment, these critical values
+are derived in \texttt{vcd} by simulation of the underlying distribution.
+
+The implementation of Friendly mosaic plots in \textsf{SAS}
+uses (by default) colors in the HLS (Hue--Luminance--Saturation) color space
+with hues for blue and red, full saturation, and varying luminance
+for lighter colors. The current \textsf{R} implementation is
+based on colors in the HSV (Hue--Saturation--Value) space where
+decreasing the saturation from 1 towards 0
+%% ---given the hue for blue or red and full saturation---
+generates very similar colors
+as increasing the luminance for HLS colors from 0.5 towards 1.
+As an alternative to those HLS and HSV colors, the device-independent
+HCL (Hue--Chroma--Luminance) color space \citep{vcd:Ihaka:2003} could be
+used which also provides homogeneous saturations over different colors and copier
+proofness. The resulting association plot (using critical values for
+$\alpha = 0.1$ and $0.01$) for the Bundesliga data can be seen in
+Figure \ref{fig:goals-assoc-signif}. It indicates very clearly that
+there is no evidence for rejecting the hypothesis of independence
+in this table. Thus, there seems to be neither an advantage for
+the home team nor for the away team.
+
+To illustrate what a `postive' example looks like we give the association plot
+for the well-known Hair-Eye-Color data set (avilable in base \textsf{R}) which cross-tabulates the hair and eye
+color and gender of 592 statistics students: Figure \ref{fig:haireye-assoc-signif}
+shows the association plot for the 328 female students. It is easy to see that
+the hypothesis of independence is rejected at 1\% level by the maximum test (\ref{eq:MaxAbs})
+due to higher frequencies of students with hair and eye color
+black/brown and blond/blue and lower frequencies of
+blond/brown and brown/blue than would be plausible under independence. Furthermore,
+it can also be seen that the residual corresponding to the black/blue
+cell is significant at the 10\% level.
+
+We have seen that this approach performs very well when the test statistic
+which should be visualized is the maximum statistic (\ref{eq:MaxAbs}). However,
+it is difficult to extend this approach to general tests for independence,
+in particular in the HCL space, where the ranges of chroma and luminance are
+not independent. In contrast the HSV space has the advantage that all
+three dimensions hue, saturation and value have the range $[0,1]$ and can
+be varied independently. Therefore, a different approach than the one
+described above could be to use the hue for the sign of the residuals as
+before, the saturation for the absolute size of the residuals, and the
+value as an indicator for the significance of some test statistic (only
+using the full value when the overall test for independence rejects
+the null hypothesis). The resulting mosaic plots under this paradigm
+are depicted in Figure \ref{fig:both-mosaic-signif}. Again, it can
+clearly be seen that despite some large residuals there is no
+evidence against independence for the Bundesliga data, but that the
+null hypothesis has to be rejected for the Hair-Eye-Color data.
 
@@
 \begin{center}
-<<UCB-mosaic1,fig=TRUE,echo=FALSE,height=6,width=8>>=
-## fm <- loglm(~(EmploymentStatus+EmploymentLength)*LayoffCause, data = emp, fitted = TRUE)
-## my.gp <- gp.HCLshading(emp, residuals(fm, type = "pearson"))
-## my.gp$p.value <- summary(fm)$tests[2,3]
-## my.gp$lty <- c(1,1)
-## grid.mosaicplot(emp, gp = my.gp, margin = ~(EmploymentStatus+EmploymentLength)*LayoffCause,
-##   space = c(0.1, 0.01, 0), dir = "v")
-mosaic(emp, expected = ~(EmploymentStatus+EmploymentLength)*LayoffCause)
+<<shadings,fig=TRUE,echo=FALSE,height=6,width=11,results=hide>>=
+grid.newpage()
+pushViewport(viewport(layout = grid.layout(1, 2)))
+pushViewport(viewport(layout.pos.col=1, layout.pos.row=1))
+mosaic(art, newpage = FALSE, gp = shading_Friendly, margins = c(2.5, 4, 2.5, 3))
+popViewport()
+pushViewport(viewport(layout.pos.col=2, layout.pos.row=1))
+set.seed(1234)
+mosaic(art, newpage = FALSE, gp = shading_max, margins = c(2.5, 3, 2.5, 4))
+popViewport(2)
 @
-\caption{Employment data with mosaic display.}
+\caption{Mosaic plot for the Arthritis data with Friendly shading (left) and extended shading (right).}
+\label{fig:shadings}
+\end{center}
+\end{figure}
+
+\section{Extensions to Conditional Independence}
+
+\cite{vcd:Friendly:1999} discusses grouping similar to coplots (conditioning
+plots) \citep{vcd:Cleveland:1993} that lead to trellis graphics
+(Becker, R.A., Cleveland, W.S. and Shyu, M. ``The Visual Design and
+Control of Trellis Display'', Journal of Computational and
+Graphical Statistics).
+
+Independence problems do not only occur in 2-way tables, although that
+is an important special case, but they are also important in tables of
+higher dimensionality and can follow much more complex patterns.
+These are again defined based on the underlying table of theoretical cell probabilities
+$(\pi_{ij\dots})$ with more than two dimensions. Models of interest include the
+null hypotheses of conditional independence:
+  \[ \pi_{ijk\dots} = \pi_{i|k\dots}\pi_{j|k\dots} \]
+
+Classical non-graphical methods for these problems include the $\chi^2$
+test, Fisher's exact test, the Cochran-Mantel-Haenzel test (for $2 \times 2 \times
+K$-tables), and the analysis of log-linear models for more
+complex settings.
+
+As an example, two natural ways to use the visualization techniques described in the previous
+sections would be to use (Trellis-like) conditioning plots or pairs plots
+(like mosaic matrices) to visualize these more complex patterns of
+independence.
+
+Two ideas for the problem of conditional independence are briefly outlined
+here and illustrated using the famous admissions data of the University of
+California at Berkley (UCB) which is available in base \textsf{R}.
+In this data, the question whether there is gender
+discrimination at the UCB leads to the result that although women seem to
+be disadvantaged at the aggregated level there is no gender discrimination
+conditioned on the department---with the very exception of one department
+in which women are {\em more} likely to be admitted than would be plausible
+under independence. Exactly this is illustrated in the conditioning
+assocation plot in Figure \ref{fig:UCB-assoc-signif}.
+
+Similarly, the same data can be visualized using a mosaic matrix where
+a conditional independence model is fitted in each plot (see Figure
+\ref{fig:UCB-mosaic-signif}).
+
+\setkeys{Gin}{width=\textwidth}
+\begin{figure}[htbp]
+\begin{center}
+<<conditional_assoc,fig=TRUE,echo=FALSE,height=6,width=9>>=
+set.seed(1234)
+cotabplot(~ Admit + Gender | Dept, data = ucb, panel = cotab_coindep, type = "assoc")
+@
+\caption{Conditional association plot for UCB admissions.}
 \label{fig:emp-mosaic1}
@@
 \end{figure}
+
+\section{Conclusions}
 
@@
 
+%% for reproducibility, print also
+
+<<graphics>>
+## load packages and data
+library(vcd)
+data(Arthritis)
+art <- xtabs(~ Treatment + Improved, data = Arthritis, subset = Sex == "Female")
+data(UCBAdmissions)
+ucb <- aperm(UCBAdmissions)
+
+## plain classic plots
+mosaic(art)
+assoc(art)
+
+## new max shading with HCL colors
+mosaic(art, gp = shading_Friendly)
+set.seed(1234)
+mosaic(art, gp = shading_max)
+
+## conditional independence
+set.seed(1234)
+cotabplot(~ Admit + Gender | Dept, data = ucb, panel = cotab_coindep, type = "assoc")
+## which uses the double max functional
+set.seed(1234)
+coindep_test(ucb, "Dept")
+
+## alternatively use Peason functional
+## via simulation of conditional distribution:
+set.seed(1234)
+coindep_test(ucb, "Dept", indepfun = function(x) sum(x^2), aggfun = sum)
+## via asymptotic distribution:
+fm <- loglm(~ Dept * (Gender + Admit), data = ucb)
+fm
+plot(fm, labeling_args = list(abbreviate = c(Admit = 2)))
+@
+
+%% deprecated stuff
+
+%% employment data
+<<employment>>
+## data
+data(Employment)
+emp <- aperm(Employment)
+emp1 <- co_table(Employment, "LayoffCause")[["Replaced"]]
+
+## plain
+assoc(emp1)
+mosaic(emp1)
+
+## new shadings
+mosaic(emp1, gp = shading_Friendly)
+mosaic(emp1, gp = shading_max)
+
+## conditional
+mosaic(emp, expected = ~(EmploymentStatus+EmploymentLength)*LayoffCause)
+cotabplot(emp, "LayoffCause", split_vertical = TRUE)
+cotabplot(emp, panel = cotab_coindep, type = "assoc")
+cotabplot(emp, "LayoffCause", panel = cotab_coindep, type = "assoc", split_vertical = TRUE)
+@
+
+%% Friendly shading in detail
 \setkeys{Gin}{width=0.8\textwidth}
@@
 \end{figure}
-
-\begin{figure}[htbp]
-\begin{center}
-<<emp1-Friendly,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>
-mosaic(emp1, gp = shading_Friendly)
-@
-<<emp1-max,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>
-mosaic(emp1, gp = shading_max)
-@
-\caption{Mosaic plots for the ``Replaced'' subset of the Employment data.}
-\label{fig:emp1}
-\end{center}
-\end{figure}
-
-\begin{figure}[htbp]
-\begin{center}
-<<UCB-mosaic2,fig=TRUE,echo=FALSE,height=4,width=8>>
-tabplot(emp, panel = pmos.max)
-@
-\caption{Employment data with conditioned mosaic display.}
-\label{fig:emp-mosaic2}
-\end{center}
-\end{figure}
-
-\begin{figure}[htbp]
-\begin{center}
-<<UCB-assoc,fig=TRUE,echo=FALSE,height=4,width=8>>
-tabplot(emp)
-@
-\caption{Employment data with conditioned association plot.}
-\label{fig:emp-assoc}
-\end{center}
-\end{figure}

### 2005-07-29 trunk/vcd/inst/doc/strucplot.Rnw
Vignette added.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@416 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\documentclass[a4paper]{article}
\usepackage{hyperref, graphicx, color, alltt, a4wide}
\usepackage{Sweave}
\usepackage[round]{natbib}
\definecolor{Red}{rgb}{0.7,0,0}
\definecolor{Blue}{rgb}{0,0,0.8}
\definecolor{hellgrau}{rgb}{0.55,0.55,0.55}
\newcommand{\pkg}[1]{{\normalfont\fontseries{b}\selectfont #1}}
\newcommand{\code}[1]{\texttt{#1}}
\newenvironment{smallexample}{\begin{alltt}}{\end{alltt}}

\begin{document}

%\VignetteIndexEntry{strucplots---Visualizing higher-dimensional contigency tables in vcd}
%\VignetteDepends{vcd}
%\VignetteKeywords{graphics, categorical data}
%\VignettePackage{vcd}

\SweaveOpts{engine=R,eps=TRUE}
\setkeys{Gin}{width=0.8\textwidth}

<<preliminaries,echo=FALSE,results=hide>>=
set.seed(1071)
library(vcd)
data(Titanic)
@

\title{strucplots---Visualizing higher-dimensional Contingency Tables in \pkg{vcd}}
\author{by David Meyer\\
  Wirtschaftsuniversit\"at Wien, Austria\\
\url{David.Meyer@R-project.org}
}
\maketitle
\sloppy

The \code{strucplot} framework in \pkg{vcd} can be used to visualize
higher-dimensional contingency tables and integrates techniques such as
mosaic displays and association plots. The main idea is to visualize
the tables' cells arranged in rectangular form. To allow for more than
two dimensions, the plot area is constructed using recursive
conditional splits, given the tiles representing the levels of the
previous dimension.
This principle defines a graph `family' with still many parameters
left such as:

\begin{itemize}
\item the split direction for each dimension
\item the content of the tiles
\item the graphical parameters of the tiles' content
\item the spacing between the tiles
\item the labeling of the tiles
\end{itemize}

\noindent The document at hand gives an introduction to the framework,
whereas labeling and shading issues are described in separate vignettes.

\section{Framework Overview}

The strucplot framework is highly modularized; Figure \ref{fig:struc}
shows the hierarchical relationship between the various components. 
The core function is \code{strucplot()} which processes the parameters,
sets up the graphical layout using grid viewports (see Figure \ref{fig:layout}),
and coordinates panel, labeling, shading, and spacing functions to produce 
the plot. The actual workhorse is the chosen panel function: currently, 
low-level panel functions for association plots (\code{struc\_assoc}) and 
mosaic plots (\code{struc\_mosaic}) are available. In addition, there are
high-level wrapper functions such as \code{mosaic()} and \code{assoc()} 
to \code{strucplot()} that conveniently set the 
parameters---it is these high-level functions that the user normally would like to use, 
although \code{strucplot()} is fully functional on its own. \code{mosaic()} and 
\code{assoc} are generic functions: for both, there exist a formula method and a default 
method. Other functions, such as \code{doubledecker()}, \code{pairs()}, and 
\code{cotabplot}, are on even higher level and make use of 
code{mosaic()} and \code{assoc()} to produce plots.

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{struc}
  \caption{Strucplot components.}
  \label{fig:struc}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics{vcdViewport}
  \caption{Viewport layout for strucplots with their names. [A] =
    ``cornerTL'', [B] = ``cornerTR'', [C] = ``cornerBL'', [D] = ``cornerBR''.}
  \label{fig:layout}
\end{figure}

\section{Mosaic and Association Plots}

As an example, consider the famous `Survival on the Titanic' 
data containing two binary variables (survival to the disaster and
gender), as well as two categorical variables (class and age). The
`flattened' contincency table can be obtained using the
\code{structable()} function (quite similar to \code{ftable()}, but
allowing the specification of split directions as in mosaic():

<<structable,fig=FALSE,echo=TRUE,height=6,width=7,results=verbatim>>=
structable(Titanic)
@

Let us
first visualize the contingency table by the means of a mosaic plot
\citep{vcd:Hartigan+Kleiner:1984} which is basically 
an area-proportional visualization of (typically observed) frequencies, composed
of tiles (corresponding to the cells) created by recursive
vertical and horizontal splits of a square. Thus, the area of each tile
is proportional to the corresponding cell entry \emph{given} the
dimensions of previous splits (see Figure \ref{fig:observed}):
\begin{Sinput}
> mosaic(Titanic, type = "expected", main = TRUE, sub = "Expected values")
\end{Sinput}

\noindent The small bullets indicate zero entries in the corresponding cell.

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<Observed,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(Titanic, main = TRUE)
@
\caption{Mosaic plot for the Titanic data.}
\label{fig:observed}
\end{center}
\end{figure}

\noindent  It is also possible to visualize the expected values instead of the
observed values (see Figure \ref{fig:expected}):
\begin{Sinput}
> mosaic(Titanic, type = "expected", main = TRUE, sub = "Expected values")
\end{Sinput}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]

\begin{center}
<<Expected,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(Titanic, type = "expected", main = TRUE, sub = "Expected values")
@
\caption{Mosaic plot for the Titanic data (expected values).}
\label{fig:expected}
\end{center}
\end{figure}

\noindent The Pearson residuals (standardized deviations of observed from expected values) 
are preferably visualized using association
plots \citep{vcd:Cohen:1980}. In contrast to \code{assocplot()}, \pkg{vcd}'s \code{assoc()}
function scales to more than two variables (see Figure \ref{fig:residuals}):
\begin{Sinput}
> assoc(Titanic, compress = FALSE)
\end{Sinput}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<Residuals,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
assoc(Titanic, compress = FALSE)
@
\caption{Association plot for the Titanic data.}
\label{fig:residuals}
\end{center}
\end{figure}

\noindent The \code{compress} argument keeps distances between tiles
equal for better comparison.

So far, we have visualized the full table. Parts of the data can be
selected either using \code{margin.table()} on the input data, or the
formula interfaces. The following example compares the age
distribution among classes, given gender, and also shows how to 
use grid's viewport framework to put two plots besides 
each other:
\begin{Sinput}
> pushViewport(viewport(layout = grid.layout(ncol = 2)))

> pushViewport(viewport(layout.pos.col = 1))
> mosaic(~ Class + Age, data = Titanic[,"Male",,], 
+        margin = c(bottom = 0), sub = "Male", newpage = FALSE)
> popViewport()

> pushViewport(viewport(layout.pos.col = 2))
> mosaic(~ Class + Age, data = Titanic[,"Female",,], 
+        margin = c(bottom = 0), sub = "Female", newpage = FALSE)
> popViewport(2)
\end{Sinput}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<Variables,fig=TRUE,echo=FALSE,height=6,width=16,results=hide>>=
pushViewport(viewport(layout = grid.layout(ncol = 2)))

pushViewport(viewport(layout.pos.col = 1))
mosaic(~ Class + Age, data = Titanic[,"Male",,], 
       margin = c(bottom = 0), sub = "Male", newpage = FALSE,
       labeling_args = list(gp_labels = gpar(fontsize = 24), 
                            gp_varnames = gpar(fontsize = 24)),
       title_gp = gpar(fontsize = 35))
popViewport()

pushViewport(viewport(layout.pos.col = 2))
mosaic(~ Class + Age, data = Titanic[,"Female",,], 
       margin = c(bottom = 0), sub = "Female", newpage = FALSE,
       labeling_args = list(gp_labels = gpar(fontsize = 24), 
                            gp_varnames = gpar(fontsize = 24)),
       title_gp = gpar(fontsize = 35))
popViewport(2)
@
\caption{Age distribution among classes, given gender.}
\label{fig:parttable}
\end{center}
\end{figure}

\noindent note the use of the \code{margin} argument: it takes a
vector with up to four values whose unnamed components are recycled,
but ``overruled'' by the named arguments. Thus, in the example, only
the bottom margin is set to 0, leaving the default size of 2 lines
unchanged for the other margins. This idea applies to almost all
vectorized arguments in the strucplot framework (with
\code{split\_vertical} as a prominent exception).

Since mosaic displays are ``conditional plots'' by definition, we can
also use one single mosaic for stratified plots.
The formula interface of \code{mosaic()} allows the
specification of conditioning variables (see Figure \ref{fig:conditioning}):

\begin{Sinput}
> mosaic(~ Class + Age | Sex, data = Titanic, split = TRUE)
\end{Sinput}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]

\begin{center}
<<Conditioning,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(~ Class + Age | Sex, data = Titanic, split = TRUE)
@
\caption{Mosaic plot for the Titanic data (expected values).}
\label{fig:conditioning}
\end{center}
\end{figure}

\noindent Note, however, that the plots in the ``pseudo-strata'' are
distorted since they are not corrected for the marginal
distribution(s) of the conditioning variables. Indeed,
the \code{cotabplot()} function would do a much better job here (see
next section).

For both mosaic plots and association plots, 
the splitting of the tiles can be controlled using the
\code{split\_vertical} argument (default: alternating splits starting
with a vertical one). Sensible recycling rules apply. 
For compatibility with \code{mosaicplot()}, the \code{mosaic()} function
also allows the use of a \code{"direction"} argument taking a vector of \code{"h"} and
\code{"v"} characters (see Figures \ref{fig:split1} and \ref{fig:split2}):
\begin{Sinput}
> mosaic(Titanic, direction = "h")
> mosaic(Titanic, split_vertical = c(TRUE, TRUE, TRUE, FALSE))
\end{Sinput}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<split1,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(Titanic, direction = "h")
@
\caption{Mosaic plot for the Titanic data---alternative splitting (1).}
\label{fig:split1}
\end{center}
\end{figure}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<split2,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(Titanic, split_vertical = c(TRUE, TRUE, TRUE, FALSE),
                labeling_args = list(abbreviate = c(Sex = TRUE, Age = TRUE)))
@
\caption{Mosaic plot for the Titanic data---alternative splitting (2).}
\label{fig:split2}
\end{center}
\end{figure}

\section{Higher-level functions}

The last plot pretty much resembles a doubledecker plot, except for
the spacing. Since spacing is also modular in the strucplot framework,
the \code{doubledecker()} function was straightforward to implement:
it really is just a wrapper for \code{mosaic()}, setting the right
defaults. Figure \ref{fig:doubledecker} shows a doubledecker plot of
the Titanic data, explaining the probability of Survival by Age, given
Sex, given Class:
\begin{Sinput}
> doubledecker(Titanic)
> ## equivalent to:
> doubledecker(Survived ~ Class + Sex + Age, data = Titanic)
\end{Sinput}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<doubledecker,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
doubledecker(Titanic)
@
\caption{Doubledecker plot for the Titanic data.}
\label{fig:doubledecker}
\end{center}
\end{figure}

Another high-level function is \code{pairs.table()}, that is, the S3-method for objects
of class "table" of the \code{pairs()} generic. This function produces a matrix 
of strucplots (currently, association or mosaic plots) in the off-diagonal cells, and
the variable names (with, optionally, univariate statistics) in the diagonal cells. Figure
\ref{fig:pairs} shows mosaic plots visualizing mutual independence in the lower triangle,
association plots for the same in the upper triangle, and bar charts in the diagonal.

\begin{Sinput}
> pairs(Titanic, upper_panel = pairs_assoc)
\end{Sinput}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<pairs,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
pairs(Titanic, upper_panel = pairs_assoc)
@
\caption{Pairs plot for the Titanic data.}
\label{fig:pairs}
\end{center}
\end{figure}

\noindent Each cell's row and column define two variables $X$ and $Y$ used for the
specification of four different types of independence: `\code{pairwise}',
     `\code{total}', `\code{conditional}' and `\code{joint}'. The pairwise mosaic
     matrix shows bivariate marginal relations between $X$ and $Y$, collapsed over all
     other variables. The total independence mosaic matrix shows mosaic
     plots for mutual independence, i.e., for marginal and conditional
     independence among all pairs of variables. The conditional
     independence mosaic matrix shows mosaic plots for marginal
     independence of $X$ and $Y$ given all other variables. The joint independence
     mosaic matrix shows mosaic plots for joint independence of all
     pairs ($X$, $Y$) of variables from the others.

Since the matrix is symmetric, the upper and lower part can be used to display different 
types of independence models, or different strucplots (currently, only association and 
mosaic plots are supported). The available panel functions (\code{pairs\_assoc()} and 
\code{pairs\_mosaic()}) are just simple wrappers to \code{assoc()} and 
\code{mosaic()}. Obviously, seeing patterns in strucplot matrices 
becomes increasingly difficult with
higher dimensionality. Typically, therefore, this plot is used with a suitable 
residual-based shading (described in a separate vignette).

Another useful high-level plot is produced by \code{cotabplot()}: this function can be
used for plotting stratified strucplots in a lattice-like display. Whereas \code{pairs()} 
is conditioning on variables, \code{cotabplot()} is conditioning on variable \emph{levels}.
The plot in Figure \ref{fig:cotabplot} shows the Survival rate among the classes, given
all four level combinations of Sex and Age.

\begin{Sinput}
> cotabplot(~ Class + Survived | Sex + Age, data = Titanic)
\end{Sinput}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<cotabplot,fig=TRUE,echo=FALSE,height=6,width=12,results=hide>>=
cotabplot(~ Class + Survived | Sex + Age, data = Titanic)
@
\caption{cotabplot for the Titanic data.}
\label{fig:cotabplot}
\end{center}
\end{figure}

\section{Technical Notes on Panel Functions}

All strucplot components (including the panel functions) 
are parameterized through the use of generating functions, that is,
functions with all parameters in their signature returning a function
to be called by \code{strucplot()}. This returned function has access
to all parameters passed to the generating function thanks to 
lexical scope. The idea is to use module-specific `code
factories' that generate parameterized workhorses rather than
struggling with monolithic functions with endless parameter lists. As
a simple example, consider \code{struc\_mosaic()} and \code{struc\_assoc},
the generating functions for the mosaic and the association plots:

\begin{smallexample}
struc_mosaic <- function(zero_size = 0.5)
  function(residuals, observed, expected = NULL, spacing, gp, split_vertical) {
  \dots
  }

struc_assoc <- function(compress = TRUE, xlim = NULL, ylim = NULL,
                        yspace = unit(0.5, "lines"), xscale = 0.9)
  function(residuals, observed = NULL, expected, spacing, gp, split_vertical) {
  \dots
  }
\end{smallexample}

\noindent Both functions have specific parameters in their signature,
but both return a function with a standardized interface that
subsequently is called by \code{strucplot} to produce the actual
plot. As can be seen, the only information \code{strucplot()} passes
to the panel functions are residuals, observed and expected
values, a structure containing the between-tile spacing information,
an object of class \code{"gpar"} for the tiles' graphical parameters,
and a logical vector with the splitting information for each
dimension. The workhorse function, of course, is free to use only
part or all of this information. High-level interface functions such
as \code{mosaic()} and \code{assoc{}} call \code{strucplot()} by
specifying the corresponding function along with the specific
parameters. Now, there a several ways of doing the same thing, as
illustrated by the following equivalent calls:

\begin{Sinput}
> strucplot(Titanic, panel = struc_mosaic(zero_size = 1))
> strucplot(Titanic, panel = struc_mosaic, panel_args = list(zero_size = 1))
> strucplot(Titanic, panel = myworkhorse)
\end{Sinput}

\noindent where \code{myworkhorse} could be, e.g., a function with
hard-wired parameters, but with an interface complying with the
workhorse signature. To enable \code{strucplot()} to distinguish
between generating functions and workhorse functions, the generating
functions must be of class \code{"panel\_generator"}.

All panel functions are supposed to produce conditional hierarchical
plots by the means of nested viewports, corresponding to the provided splitting
information. Thus, at the end of the plotting, each tile is associated with
a particular viewport. Each of those viewports has to be
conventionally named, enabling other strucplot modules, in particular the
labeling functions, to access specific tiles after they have been
plotted. Note that the viewport tree is popped by default. The
following example shows how to access parts of the plot after it has
been drawn (see Figure \ref{fig:afterplot}):

\begin{Sinput}
> mosaic(~ Sex + Class, data = Titanic, pop = FALSE)
> current.vpTree() ## output not shown

> seekViewport("cell..Sex.Male")
> grid.rect(gp = gpar(col = "red", lwd = 4))

> seekViewport("cell..Sex.Male..Class.Crew")
> grid.circle(r = 0.2, gp = gpar(fill = "cyan"))
\end{Sinput}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<afterplot,fig=TRUE,echo=FALSE,height=6,width=12,results=hide>>=
mosaic(~ Sex + Class, data = Titanic, pop = FALSE)

seekViewport("cell..Sex.Male")
grid.rect(gp = gpar(col = "red", lwd = 4))

seekViewport("cell..Sex.Male..Class.Crew")
grid.circle(r = 0.2, gp = gpar(fill = "cyan"))
@
\caption{Adding elements to a mosaic plot after drawing.}
\label{fig:afterplot}
\end{center}
\end{figure}

\noindent The naming convention for the viewports is:
\code{cell..\emph{[Variable 1]}.\emph{[Level 1]}..\emph{[Variable 2]}.\emph{[Level 2]}} \dots. Clearly, these names depend on the splitting.

In addition to the viewports, the main graphical elements get names
following a similar construction method. This allows to change
graphical parameters of plot elements after the plotting (see Figure \ref{fig:changeplot}):

\begin{Sinput}
> assoc(Titanic)
> getNames() ## output not shown
> grid.edit("rect..Class.1st..Age.Adult..Sex.Female..Survived.Yes", 
            gp = gpar(fill = "red"))
\end{Sinput}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<changeplot,fig=TRUE,echo=FALSE,height=6,width=12,results=hide>>=
assoc(Titanic)
getNames() ## output not shown
grid.edit("rect..Class.1st..Age.Adult..Sex.Female..Survived.Yes", 
          gp = gpar(fill = "red"))
@
\caption{Changing graphical parameters of elements after drawing.}
\label{fig:changeplot}
\end{center}
\end{figure}

\begin{thebibliography}{}

\bibitem[Cohen, 1980]{vcd:Cohen:1980}
Cohen, A. (1980).
\newblock On the graphical display of the significant components in a two-way
  contingency table.
\newblock {\em Communications in Statistics---Theory and Methods},
  A9:1025--1041.

\bibitem[Hartigan and Kleiner, 1984]{vcd:Hartigan+Kleiner:1984}
Hartigan, J. and Kleiner, B. (1984).
\newblock A mosaic of television ratings.
\newblock {\em The American Statistician}, 38:32--35.

\end{thebibliography}

\end{document}

### 2005-08-01 trunk/vcd/inst/doc/strucplot.Rnw
Now really fixed the conditioning formula in mosaic.formula.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@417 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 @
-\caption{Mosaic plot for the Titanic data (expected values).}
+\caption{Mosaic plot for conditional independence structures.}
 \label{fig:conditioning}
@@
 
-\noindent Note, however, that the plots in the ``pseudo-strata'' are
+\noindent The effect of using this kind of formula is that
+conditioning variables are permuted before the conditioned variables
+in the table, and that \code{spacing\_conditional()} is used to better
+distinguish conditioning from conditioned dimensions.
+Note, however, that the plots in the ``pseudo-strata'' are
 distorted since they are not corrected for the marginal

### 2005-08-01 trunk/vcd/R/mosaic.R

@@ -35,7 +35,7 @@ function(formula, data = NULL, ..., main = NULL)
       dat <- margin.table(dat, ind)
     }
     mosaic.default(dat, main = main,
-                   condvars = if (is.null(condind)) NULL else length(condind), ...)
+                   condvars = if (is.null(condind)) NULL else match(condnames, names(dimnames(dat))), ...)
   } else {
     tab <- if ("Freq" %in% colnames(data))
       xtabs(formula(paste("Freq~", paste(c(condnames, varnames), collapse = "+"))),

### 2005-08-01 trunk/vcd/inst/doc/strucplot.Rnw
plots enhanced
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@419 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-\setkeys{Gin}{width=0.7\textwidth}
+\setkeys{Gin}{width=\textwidth}
 \begin{figure}[h]
@@
 \begin{center}
-<<afterplot,fig=TRUE,echo=FALSE,height=6,width=12,results=hide>>=
+<<afterplot,fig=TRUE,echo=FALSE,height=6,width=6,results=hide>>=
 mosaic(~ Sex + Class, data = Titanic, pop = FALSE)
@@
 \begin{center}
-<<changeplot,fig=TRUE,echo=FALSE,height=6,width=12,results=hide>>=
+<<changeplot,fig=TRUE,echo=FALSE,height=6,width=6,results=hide>>=
 assoc(Titanic)
-getNames() ## output not shown
-grid.edit("rect..Class.1st..Age.Adult..Sex.Female..Survived.Yes", 
-          gp = gpar(fill = "red"))
+x = Titanic
+x[] = NA
+x["1st", "Female", "Adult", "Yes"] = "Red"
+assoc(Titanic, gp = gpar(fill = x))
 @

### 2005-08-01 trunk/vcd/R/labeling.R
labeling_text changed to labeling_border
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@420 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -32,7 +32,7 @@ labeling_list <- function(gp = gpar(),
     if (is.table(d))
       d <- dimnames(d)
     ld <- length(d)
-    labeling_text(labels = FALSE, varnames = varnames)(d, split_vertical, condvars)
+    labeling_border(labels = FALSE, varnames = varnames)(d, split_vertical, condvars)
     seekViewport("marginBottom")
     pos <- unit(switch(pos, left = 0, center = 0.5, 1) / cols, "npc")
     ind <- split(seq(ld), rep.int(seq(cols), ceiling(ld / cols))[seq(ld)])
@@ -57,7 +57,7 @@ labeling_conditional <- function(...)
       d <- dimnames(d)
     v <- rep.int(TRUE, length(d))
     v[seq(condvars)] <- FALSE
-    labeling_text(labels = !v, ...)(d, split_vertical, condvars)
+    labeling_border(labels = !v, ...)(d, split_vertical, condvars)
     labeling_cells(labels = v, ...)(d, split_vertical, condvars)
   }
 class(labeling_conditional) <- "panel_generator"
@@ -132,7 +132,7 @@ labeling_cells <- function(labels = TRUE, varnames = TRUE,
 }
 class(labeling_cells) <- "panel_generator"
 
-labeling_text <- function(labels = TRUE, varnames = labels,
+labeling_border <- function(labels = TRUE, varnames = labels,
                           tl_labels = NULL, tl_varnames = NULL, 
                           gp_labels = gpar(fontsize = 12),
                           gp_varnames = gpar(fontsize = 12, fontface = 2),
@@ -461,14 +461,14 @@ labeling_text <- function(labels = TRUE, varnames = labels,
     split()
     
   }
-class(labeling_text) <- "panel_generator"
+class(labeling_border) <- "panel_generator"
 
 labeling_doubledecker <- function(lab_pos = c("bottom", "top"), ...) {
   lab_pos <- match.arg(lab_pos)
   function(d, split_vertical, condvars) {
     if (is.table(d))
       d <- dimnames(d)
-    labeling_text(boxes = c(rep.int(TRUE, length(d) - 1), FALSE),
+    labeling_border(boxes = c(rep.int(TRUE, length(d) - 1), FALSE),
                   clip = c(rep.int(TRUE, length(d) - 1), FALSE),
                   labbl_varnames = FALSE,
                   rot_labels = rep.int(0, 4),
[…]

### 2005-08-01 trunk/vcd/R/strucplot.R

@@ -18,7 +18,7 @@ strucplot <- function(## main parameters
                       spacing_args = list(),
                       gp = NULL,
 		      gp_args = list(),   
-                      labeling = labeling_text,
+                      labeling = labeling_border,
                       labeling_args = list(),
                       panel = struc_mosaic,
                       panel_args = list(),

### 2005-08-01 trunk/vcd/inst/doc/labeling.Rnw
Vignette Nr. 2 added. Temp files moved to pics subdir, which has been added to .Rbuildignore
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@421 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\documentclass[a4paper]{article}
\usepackage{hyperref, graphicx, color, alltt, a4wide}
\usepackage{Sweave}
\usepackage[round]{natbib}
\definecolor{Red}{rgb}{0.7,0,0}
\definecolor{Blue}{rgb}{0,0,0.8}
\definecolor{hellgrau}{rgb}{0.55,0.55,0.55}
\newcommand{\pkg}[1]{{\normalfont\fontseries{b}\selectfont #1}}
\newcommand{\code}[1]{\texttt{#1}}
\newenvironment{smallexample}{\begin{alltt}}{\end{alltt}}

\begin{document}

%\VignetteIndexEntry{Labeling in Strucplots}
%\VignetteDepends{vcd}
%\VignetteKeywords{graphics, categorical data}
%\VignettePackage{vcd}

\SweaveOpts{engine=R,eps=TRUE,prefix.string=pics/labeling}
\setkeys{Gin}{width=0.8\textwidth}

<<preliminaries,echo=FALSE,results=hide>>=
set.seed(1071)
library(vcd)
data(Titanic)
@

\title{Labeling in Stucplots}
\author{by David Meyer\\
  Wirtschaftsuniversit\"at Wien, Austria\\
\url{David.Meyer@R-project.org}
}
\maketitle
\sloppy

One of the major enhancements in vcd is the labeling of strucplots which now
has become much more powerful and flexible. Like the panel, the
shading, the spacing, and the legend, labeling is now completely modular.
The user supplies either a labeling function, or, alternatively, 
a generating function that parameterizes a labeling function, 
to \code{strucplot()} which then draws the labels. The labeling is 
well-separated from the actual plotting that occurs in the panel functions;
it actually only relies on the viewport tree produced by them, and the
`dimnames' attribute of the visualized table. The drawing of the
labels happens after the panel function has drawn the actual plot.
Thus, it is possible to supply one's own labeling function, or
to combine some of the basic functions to produce a more complex labeling.
As an example, consider \code{labeling\_list()}:

\begin{Sinput}
labeling_list <- function(gp = gpar(), just = "left", pos = "left", lsep = ": ", 
                          sep = " ", offset = unit(c(2, 2), "lines"),
                          varnames = TRUE, cols = 2, ...) 
  function(d, split_vertical, condvars) {
    \dots
  }
\end{Sinput}

\code{labeling\_list()} is a generating function that takes all parameters
as arguments, and returns the actual labeling function which takes
only three arguments: the dimnames structure of the table, the logical
vector with the splitting information, and the number of conditioning
variables (if any). The labeling function has access to the arguments
of the generating function because of lexical scoping. In the
following, we describe the three basic modules
(\code{labeling\_text()}, \code{labeling\_list()}, and \code{labeling\_cells()}) and
derived functions that build upon them.

\section{Labels in the borders: \code{labeling\_text()}}

\code{labeling\_text()} is the default for all strucplot displays. It
plots labels in the borders similar to the \code{mosaicplot()}
function, but is much more flexible: it is not limited
to 4 dimensions, and the positioning and graphical parameters 
of levels and variable names are quite customizable. In addition,
the problem of overlapping labels can be handled in several ways.

As an example, consider the `Survival on the Titanic' data, consisting
of 4 categorical variables: Survival, Gender (``Sex''), Age, and
Crew. By default, the variable names and levels are plotted `around' the
plot in a counter-clockwise way (see Figure \ref{fig:defaults}):

\begin{Sinput}
> mosaic(Titanic)
\end{Sinput}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<default,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(Titanic)
@
\caption{Mosaic plot for the Titanic data.}
\label{fig:defaults}
\end{center}
\end{figure}

\noindent Note that the last two levels of the `Survived' variable do
overlap. This can be addressed in several ways. The `brute force'
method is to enable clipping for the `Survivcal' dimension (see Figure \ref{fig:clipping}):

\begin{Sinput}
> mosaic(Titanic, labeling_args = list(clip = c(Survived = TRUE)))
\end{Sinput}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<clipping,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(Titanic, labeling_args = list(clip = c(Survived = TRUE)))
@
\caption{The effect of clipping.}
\label{fig:clipping}
\end{center}
\end{figure}

\noindent A more sensible solution is to abbreviate the `Survival'
levels (see Figure \ref{fig:abbreviating}):

\begin{Sinput}
> mosaic(Titanic, labeling_args = list(abbreviate = c(Survived = TRUE)))
\end{Sinput}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<abbreviating,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(Titanic, labeling_args = list(abbreviate = c(Survived = TRUE)))
@
\caption{Abbreviating.}
\label{fig:abbreviating}
\end{center}
\end{figure}

\noindent The \code{abbreviate} argument actually takes a vector of
integers indicating the number of characters the levels should be
abbreviated to, but logical values are equivalent to 1. Another
possibility is to rotate the levels (see Figure \ref{fig:rotating}):

\begin{Sinput}
> mosaic(Titanic, labeling_args = list(rot_labels = c(bottom = 45)))
\end{Sinput}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<rotate,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(Titanic, labeling_args = list(rot_labels = c(bottom = 45)))
@
\caption{Rotating labels.}
\label{fig:rotating}
\end{center}
\end{figure}

\noindent Finally, we could also inhibit the output of repeated levels
for the `Survival' dimension (see Figure \ref{fig:repeat}):

\begin{Sinput}
> mosaic(Titanic, labeling_args = list(rep = c(Survived = FALSE)))
\end{Sinput}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<repeat,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(Titanic, labeling_args = list(rep = c(Survived = FALSE)))
@
\caption{Inhibiting the repetition of levels.}
\label{fig:repeat}
\end{center}
\end{figure}

We proceed with a few more `cosmetic' changes. 
A first simple, but effectful modification is to position
all labels and variables left-aligned, which is
also achieved by using \code{labeling\_left()} (see Figure \ref{fig:left}):

\begin{Sinput}
> mosaic(Titanic, labeling_args = list(pos_varnames = "left", pos_labels = "left", just_labels = "left", rep = FALSE))
\end{Sinput}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<left,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(Titanic, labeling_args = list(pos_varnames = "left", pos_labels = "left", just_labels = "left", rep = FALSE))
@
\caption{Left-aligning.}
\label{fig:left}
\end{center}
\end{figure}

\noindent Next, we show how to put all levels to the
bottom and right margins, and all variable names to the top and left
margins (see Figure \ref{fig:margins}):

\begin{Sinput}
> mosaic(Titanic, labeling_args = list(tl_labels = FALSE, tl_varnames = TRUE))
\end{Sinput}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<margins,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(Titanic, labeling_args = list(tl_labels = FALSE, tl_varnames = TRUE))
@
\caption{Changes in the margins.}
\label{fig:margins}
\end{center}
\end{figure}

\noindent Now, we will add boxes to the labels and additionally
enable clipping (see Figure \ref{fig:boxes}):

\begin{Sinput}
> mosaic(Titanic, labeling_args = list(tl_labels = FALSE, tl_varnames = TRUE, boxes = TRUE, clip = TRUE))
\end{Sinput}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<boxes,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(Titanic, labeling_args = list(tl_labels = FALSE, tl_varnames = TRUE, boxes = TRUE, clip = TRUE))
@
\caption{Boxes and Clipping.}
\label{fig:boxes}
\end{center}
\end{figure}

\noindent This is pretty close to what the \code{labels\_cboxed()}
wrapper does. Another variant is to put the variable names into the
same line as the levels (see Figure \ref{fig:labbl}):

\begin{Sinput}
> mosaic(Titanic, labeling_args = list(tl_labels = TRUE, boxes = TRUE, clip = TRUE, labbl_varnames = TRUE), margins = c(left = 4, right = 1, 3))
\end{Sinput}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<labbll,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(Titanic, labeling_args = list(tl_labels = TRUE, boxes = TRUE, clip = TRUE, labbl_varnames = TRUE), margins = c(left = 4, right = 1, 3))
@
\caption{Variable names beneath levels.}
\label{fig:labbl}
\end{center}
\end{figure}

\noindent This is basically what \code{labeling\_lboxed()} produces,
and a similar design is used by the \code{doubledecker()}.

\section{Labels in the cells: \code{labeling\_cells()}}

This labeling draws both variable names and levels in the
cells. As an example, we use the `PreSex' data on pre- and
extramarital sex and divorce (see Figure \ref{fig:cell}):

\begin{Sinput}
> mosaic(~ MaritalStatus + Gender, data = PreSex, labeling = labeling_cells)
\end{Sinput}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<cell,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(~ MaritalStatus + Gender, data = PreSex, labeling = labeling_cells)
@
\caption{Cell labeling for the `PreSex' data.}
\label{fig:cell}
\end{center}
\end{figure}

\noindent In the case of narrow cells, the labels can be abbreviated 
(see Figure \ref{fig:cell2}):

\begin{Sinput}
> mosaic(~ PremaritalSex + ExtramaritalSex, data = PreSex, labeling = labeling_cells(abbreviate_labels = TRUE, abbreviate_varnames = TRUE, clip = F))
\end{Sinput}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<cell2,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(~ PremaritalSex + ExtramaritalSex, data = PreSex, labeling = labeling_cells(abbreviate_labels = TRUE, abbreviate_varnames = TRUE, clip = F))
@
\caption{Cell labeling for the `PreSex' data, labels abbreviated.}
\label{fig:cell2}
\end{center}
\end{figure}

\noindent For some data, it might be convenient to combine cell
labeling with border labeling as done by \code{labels\_conditional()}
(see Figure \ref{fig:conditional}):

\begin{Sinput}
> mosaic(~ PremaritalSex + ExtramaritalSex | MaritalStatus + Gender, data = PreSex, labeling = labeling_conditional(abbreviate_varnames = TRUE, abbreviate_labels = TRUE))
\end{Sinput}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<conditional,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(~ PremaritalSex + ExtramaritalSex | MaritalStatus + Gender, data = PreSex, labeling = labeling_conditional(abbreviate_varnames = TRUE, abbreviate_labels = TRUE))
@
\caption{Conditional labeling for the `PreSex' data, labels abbreviated.}
\label{fig:conditional}
\end{center}
\end{figure}

\noindent Additionally, the cell labeling allows the user to add
arbitrary text to the cells by supplying a character array in the same
shape than the data array to the \code{text} argument (cells with missing values
are ignored). In the
following example, this is used to add all observed values greater
than 5 to the cells after the mosaic for the `Titanic' data has been plotted
(see Figure \ref{fig:text}):

\begin{Sinput}
> mosaic(Titanic, pop = FALSE)

> tab <- ifelse(Titanic < 6, NA, Titanic)
> labeling_cells(text = tab, clip = FALSE)(Titanic)
\end{Sinput}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<text,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(Titanic, pop = FALSE)

tab <- ifelse(Titanic < 6, NA, Titanic)
labeling_cells(text = tab, clip = FALSE)(Titanic)
@
\caption{User-supplied Text added to a mosaic display of the `Titanic' data.}
\label{fig:text}
\end{center}
\end{figure}

\section{A simple list of labels:\code{labels\_list()}}

If problems with overlapping labels cannot satisfactorily resolved,
the last remedy could be to simply list the levels below the plot
(see Figure \ref{fig:list}):

\begin{Sinput}
> mosaic(Titanic, labeling = labeling_list, margins = c(bottom = 5))
\end{Sinput}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<list,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(Titanic, labeling = labeling_list, margins = c(bottom = 5))
@
\caption{Labels indicated below the plot.}
\label{fig:list}
\end{center}
\end{figure}

\noindent The number of columns can be specified.

\end{document}

### 2005-08-02 trunk/vcd/inst/doc/labeling.Rnw
Vignettes enhanced. labeling_text.Rd renamed to labeling_border.Rd
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@422 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 `dimnames' attribute of the visualized table. The drawing of the
-labels happens after the panel function has drawn the actual plot.
+labels happens after the actual plot has been drawn by the panel function.
 Thus, it is possible to supply one's own labeling function, or
@@
 \noindent Note that the last two levels of the `Survived' variable do
-overlap. This can be addressed in several ways. The `brute force'
+overlap. This issue can be addressed in several ways. The `brute force'
 method is to enable clipping for the `Survivcal' dimension (see Figure \ref{fig:clipping}):
@@
 
-\noindent A more sensible solution is to abbreviate the `Survival'
+\noindent The \code{clip} parameter is passed to the labeling function
+via the \code{labeling\_args} argument which takes a list of
+parameters. \code{clip} itself takes a vector of logicals (one for
+each dimension). Almost all vectorized arguments in the strucplot
+framework can be abbreviated in the following way: unnamed components
+(or the defaults, if there are none) are recycled as needed, but
+overridden by the named components. Here, the default is \code{FALSE},
+and therefore clipping is enabled only for the \code{Survived} dimension.
+A more sensible solution to the overlap problem is to abbreviate the `Survival'
 levels (see Figure \ref{fig:abbreviating}):
@@
 integers indicating the number of characters the levels should be
-abbreviated to, but logical values are equivalent to 1. Another
+abbreviated to, but logical values are equivalent to 1. Abbreviation
+is performed using {\sf R}'s \code{abbreviate()} function. Another
 possibility is to rotate the levels (see Figure \ref{fig:rotating}):
@@
 
-We proceed with a few more `cosmetic' changes. 
+We now proceed with a few more `cosmetic' features.
 A first simple, but effectful modification is to position
-all labels and variables left-aligned, which is
-also achieved by using \code{labeling\_left()} (see Figure \ref{fig:left}):
+all labels and variables left-aligned, which can
+also be achieved by using \code{labeling\_left()} (see Figure \ref{fig:left}):
 
 \begin{Sinput}
-> mosaic(Titanic, labeling_args = list(pos_varnames = "left", pos_labels = "left", just_labels = "left", rep = FALSE))
+> mosaic(Titanic, labeling_args = list(pos_varnames = "left", pos_labels = "left", 
+         just_labels = "left", rep = FALSE))
 \end{Sinput}
@@
 <<left,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-mosaic(Titanic, labeling_args = list(pos_varnames = "left", pos_labels = "left", just_labels = "left", rep = FALSE))
+mosaic(Titanic, labeling_args = list(pos_varnames = "left", pos_labels = "left", 
+       just_labels = "left", rep = FALSE))
 @
@@
 
-\noindent Next, we show how to put all levels to the
+\noindent Note that obviously we need to change the justification to \code{"left"} as well. Next, we show how to put all levels to the
 bottom and right margins, and all variable names to the top and left
@@
 
-\noindent Now, we will add boxes to the labels and additionally
+\noindent The \code{"tl\_foo"} (``top left'') arguments are \code{TRUE} by default. 
+Now, we will add boxes to the labels and additionally
 enable clipping (see Figure \ref{fig:boxes}):
@@
 \begin{Sinput}
-> mosaic(Titanic, labeling_args = list(tl_labels = FALSE, tl_varnames = TRUE, boxes = TRUE, clip = TRUE))
+> mosaic(Titanic, labeling_args = list(tl_labels = FALSE, tl_varnames = TRUE, 
+         boxes = TRUE, clip = TRUE))
 \end{Sinput}
@@
 <<boxes,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-mosaic(Titanic, labeling_args = list(tl_labels = FALSE, tl_varnames = TRUE, boxes = TRUE, clip = TRUE))
+mosaic(Titanic, labeling_args = list(tl_labels = FALSE, tl_varnames = TRUE, 
+       boxes = TRUE, clip = TRUE))
 @
@@
 
-\noindent This is pretty close to what the \code{labels\_cboxed()}
+\noindent The values to \code{boxes} and \code{clip} are recycled for
+all dimensions. The result is pretty close to what the \code{labels\_cboxed()}
 wrapper does. Another variant is to put the variable names into the
@@
 \begin{Sinput}
-> mosaic(Titanic, labeling_args = list(tl_labels = TRUE, boxes = TRUE, clip = TRUE, labbl_varnames = TRUE), margins = c(left = 4, right = 1, 3))
+> mosaic(Titanic, labeling_args = list(tl_labels = TRUE, boxes = TRUE, 
+         clip = TRUE, labbl_varnames = TRUE), margins = c(left = 4, right = 1, 3))
 \end{Sinput}
@@
 <<labbll,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-mosaic(Titanic, labeling_args = list(tl_labels = TRUE, boxes = TRUE, clip = TRUE, labbl_varnames = TRUE), margins = c(left = 4, right = 1, 3))
+mosaic(Titanic, labeling_args = list(tl_labels = TRUE, boxes = TRUE, clip = TRUE, 
+       labbl_varnames = TRUE), margins = c(left = 4, right = 1, 3))
 @
@@
 
-\noindent This is basically what \code{labeling\_lboxed()} produces,
+\noindent \code{labbl\_varnames} (``variable names to the bottom/left of the
+labels'') is a vector of logicals indicating the side for the
+variable names. The resulting layout is basically what \code{labeling\_lboxed()} produces,
 and a similar design is used by the \code{doubledecker()}.
@@
 
-\noindent In the case of narrow cells, the labels can be abbreviated 
-(see Figure \ref{fig:cell2}):
+\noindent In the case of narrow cells, it might be useful to
+abbreviate labels and/or variable names and turn off clipping (see Figure \ref{fig:cell2}):
 
 \begin{Sinput}
-> mosaic(~ PremaritalSex + ExtramaritalSex, data = PreSex, labeling = labeling_cells(abbreviate_labels = TRUE, abbreviate_varnames = TRUE, clip = F))
+> mosaic(~ PremaritalSex + ExtramaritalSex, data = PreSex, 
+         labeling = labeling_cells(abbreviate_labels = TRUE, 
+         abbreviate_varnames = TRUE, clip = F))
 \end{Sinput}
@@
 \begin{Sinput}
-> mosaic(~ PremaritalSex + ExtramaritalSex | MaritalStatus + Gender, data = PreSex, labeling = labeling_conditional(abbreviate_varnames = TRUE, abbreviate_labels = TRUE))
+> mosaic(~ PremaritalSex + ExtramaritalSex | MaritalStatus + Gender, data = PreSex, 
+         labeling = labeling_conditional(abbreviate_varnames = TRUE, 
+         abbreviate_labels = TRUE))
 \end{Sinput}
@@
 shape than the data array to the \code{text} argument (cells with missing values
-are ignored). In the
-following example, this is used to add all observed values greater
+are ignored). In the following example, this is used to add all observed values greater
 than 5 to the cells after the mosaic for the `Titanic' data has been plotted

### 2005-08-02 trunk/vcd/paper/JCGS.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@423 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \begin{center}
-\textbf{\LARGE Residual-based Shadings for Visualizing (Conditional) Independence}
+\textbf{\LARGE Residual-based Shadings for Visualizing\\ (Conditional) Independence}
 
@@
 $i = 1, \dots, I$ and $j = 1, \dots, J$ and row and column sums
-$n_{i+} = \sum_i n_{ij}$ and $n_{+j} = \sum_j n_{ij}$ respectively. For
-convenience the number of observations is denoted $n = n_{++}$.
-
+$n_{i+} = \sum_i n_{ij}$ and $n_{+j} = \sum_j n_{ij}$ respectively.
+%% For convenience the number of observations is denoted $n = n_{++}$.
 Given an underlying distribution with theoretical cell probabilities
@@
 measure of discrepancy
-between observed and expected values are the Pearson residuals
+between observed and expected values are probably the Pearson residuals
 \begin{equation} \label{eq:Pearson}
@@
 values $c_\alpha = 2$ and $4$ correspond for the table being visualized.
+
+\begin{table}[htb]
+\begin{center}
+\begin{tabular}{rr|rrr}
+                   &         & \multicolumn{3}{|l}{\textbf{Improved}} \\
+                   &         & None & Some & Marked \\ \hline
+\textbf{Treatment} & Placebo &   19 &    7 &      6 \\
+                   & Treated &    6 &    5 &     16
+\end{tabular}
+\caption{\label{tab:art} Treatment and improvement among 59 female
+patients with rheumatoid arthritis.}
+\end{center}
+\end{table}
 

### 2005-08-02 trunk/vcd/R/cotabplot.R
by default use the same xlim for conditional assoc plots
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@424 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -164,14 +164,17 @@ cotab_mosaic <- function(x = NULL, condvars = NULL, ...) {
 }
 class(cotab_mosaic) <- "panel_generator"
 
-cotab_assoc <- function(x = NULL, condvars = NULL, ylim = NULL, ...) {
-  if(is.null(ylim) & !is.null(x))
-    ylim <- range(residuals(coindep_test(x, condvars, n = 1)))
+cotab_assoc <- function(x = NULL, condvars = NULL, xlim = TRUE, ylim = NULL, ...) {
+  if(!is.null(x)) {
+    fm <- coindep_test(x, condvars, n = 1)
+    if(!is.null(xlim) && xlim) xlim <- c(-0.5, 0.5) * max(sqrt(fitted(fm)))
+    if(is.null(ylim)) ylim <- range(max(residuals(fm)))
+  }
   
   function(x, condlevels) {
     if(is.null(condlevels)) assoc(x, newpage = FALSE, pop = TRUE, ylim = ylim, ...)
       else assoc(co_table(x, names(condlevels))[[paste(condlevels, collapse = ".")]],
-                  newpage = FALSE, pop = TRUE, ylim = ylim, ...)
+                  newpage = FALSE, pop = TRUE, xlim = xlim, ylim = ylim, ...)
   }
 }
 class(cotab_assoc) <- "panel_generator"
@@ -180,7 +183,7 @@ cotab_coindep <- function(x, condvars,
   test = c("max", "Chisq"), level = NULL, n = 1000,
   h = NULL, c = NULL, l = NULL, lty = 1,
   type = c("mosaic", "assoc"),
-  legend = FALSE, split_vertical = TRUE, ylim = NULL, ...)
+  legend = FALSE, xlim = TRUE, ylim = NULL, ...)
 {
   if(is.null(condvars))
     stop("at least one conditioning variable is required")
@@ -246,22 +249,24 @@ cotab_coindep <- function(x, condvars,
     rval <- function(x, condlevels) {
     if(is.null(condlevels))
       mosaic(x, newpage = FALSE, pop = TRUE,
-             gp = gpfun, legend = legend, split_vertical = split_vertical, ...)
+             gp = gpfun, legend = legend, ...)
     else
       mosaic(co_table(x, names(condlevels))[[paste(condlevels, collapse = ".")]],
              newpage = FALSE, pop = TRUE,
-	     gp = gpfun, legend = legend, split_vertical = split_vertical, ...)
[…]

### 2005-08-02 trunk/vcd/R/shadings.R
properly included now the "fuzz" for the borders in Friendly shading
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@425 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -5,7 +5,8 @@
 
 shading_hsv <- function(observed, residuals = NULL, expected = NULL, df = NULL,
   h = c(2/3, 0), s = c(1, 0), v = c(1, 0.5),
-  interpolate = c(2, 4), lty = 1, p.value = NULL, level = 0.95, ...)
+  interpolate = c(2, 4), lty = 1, eps = NULL,
+  p.value = NULL, level = 0.95, ...)
 {
   ## get h/s/v and lty
   my.h <- rep(h, length.out = 2)  ## positive and negative hue
@@ -63,15 +64,23 @@ shading_hsv <- function(observed, residuals = NULL, expected = NULL, df = NULL,
   rval <- function(x) {
     res <- as.vector(x)
 
-    col <- hsv(ifelse(res > 0, my.h[1], my.h[2]),
-               pmax(pmin(interpolate(abs(res)), 1), 0),
-	       v, ...)
+    fill <- hsv(ifelse(res > 0, my.h[1], my.h[2]),
+                pmax(pmin(interpolate(abs(res)), 1), 0),
+	        v, ...)
+    dim(fill) <- dim(x)
+    
+    col <- rep.int("black", length(res))
+    if(!is.null(eps)) {
+      eps <- abs(eps)
+      col[res > eps] <- hsv(my.h[1], 1, v, ...)
+      col[res < -eps] <- hsv(my.h[2], 1, v, ...)      
+    }
     dim(col) <- dim(x)
     
     lty <- ifelse(x > 0, lty[1], lty[2])    
     dim(lty) <- dim(x)
 
-    return(structure(list(fill = col, lty = lty), class = "gpar"))
+    return(structure(list(col = col, fill = fill, lty = lty), class = "gpar"))
   }
   attr(rval, "legend") <- legend
   attr(rval, "p.value") <- p.value
@@ -82,8 +91,8 @@ class(shading_hsv) <- "panel_generator"
 
 shading_hcl <- function(observed, residuals = NULL, expected = NULL, df = NULL,
   h = NULL, c = NULL, l = NULL,
-  interpolate = c(2, 4), lty = 1, p.value = NULL, level = 0.95,
-  ...)
+  interpolate = c(2, 4), lty = 1, eps = NULL, 
+  p.value = NULL, level = 0.95, ...)
 {
   ## set defaults
   if(is.null(h)) h <- c(260, 0)
@@ -147,16 +156,24 @@ shading_hcl <- function(observed, residuals = NULL, expected = NULL, df = NULL,
[…]

### 2005-08-02 trunk/vcd/paper/JCGS.Rnw
properly included now the "fuzz" for the borders in Friendly shading
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@425 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2005-07-20 trunk/vcd/paper/JCGS.Rnw
@@
 \begin{center}
-\textbf{\LARGE Residual-based Shadings for Visualizing (Conditional) Independence}
+\textbf{\LARGE Residual-based Shadings for Visualizing\\[.15cm] (Conditional) Independence}
 
@@
 
-\vspace{0.3cm}
-\textit{\large Wirtschaftsuniversit\"at Wien, Augasse 2--6, 1090 Wien, Austria}\\[0.14cm]
-{\large \textit{E-mail:} \texttt{Firstname.Lastname@wu-wien.ac.at}}
+%% \vspace{0.3cm}
+%% \textit{\large Wirtschaftsuniversit\"at Wien, Augasse 2--6, 1090 Wien, Austria}\\[0.14cm]
+%% {\large \textit{E-mail:} \texttt{Firstname.Lastname@wu-wien.ac.at}}
 \end{center}
@@
 ucb <- aperm(UCBAdmissions)
+
+myseed <- 1071
+@
+
+<<inference,echo=FALSE,results=hide>>=
+if(FALSE) {
+  set.seed(myseed)
+  art_max <- coindep_test(art, n = 5000)
+  set.seed(myseed)
+  art_chisq <- coindep_test(art, n = 5000, indepfun = function(x) sum(x^2))
+
+  set.seed(myseed)
+  art_gp <- shading_max(art, n = 5000)
+
+  set.seed(myseed)
+  ucb_doublemax <- coindep_test(ucb, "Dept", n = 5000)
+  set.seed(myseed)
+  ucb_maxChisq <- coindep_test(ucb, "Dept", n = 5000, indepfun = function(x) sum(x^2))
+  set.seed(myseed)
+  ucb_sumChisq <- coindep_test(ucb, "Dept", n = 5000, indepfun = function(x) sum(x^2), aggfun = sum)
+
+  set.seed(myseed)
+  ucb_cotab <- cotab_coindep(ucb, condvars = "Dept", type = "assoc", n = 5000)
+
+  save(art_max, art_chisq, art_gp,
+       ucb_doublemax, ucb_maxChisq, ucb_sumChisq, ucb_cotab,
+       file = "JCGS-inference.rda")
+} else {
+  load("JCGS-inference.rda")
+}
 @
@@
 <<auxiliary_functions,echo=FALSE,results=hide>>=
+stat <- function(x, digits = 3) as.vector(round(x$statistic, digits = digits))
+pval <- function(x, digits = 6) {
+  oop <- options()
+  on.exit(oop)
+  options(scipen = 1000)
+  as.vector(round(x$p.value, digits = digits))
+}
+
 hue.slice <- function(hue, grid.n = 100, type = c("HCL", "HSV"), plot = TRUE, fixup = FALSE)
@@
 }
-
-plot.slice <- function(color.slice)
-{
-  nc <- ncol(color.slice)
-  nl <- nrow(color.slice)
-  chroma <- as.numeric(colnames(color.slice))
-  luminance <- as.numeric(rownames(color.slice))
-  if(attr(color.slice, "type") == "HCL") {
-    xlab <- "chroma"
-    ylab <- "luminance"
-  } else {
-    xlab <- "saturation"
-    ylab <- "value"
-  }
-  plot(0.5, 0.5, xlim = range(chroma), ylim = range(luminance), type = "n", axes = FALSE,
-       xlab = xlab, ylab = ylab, yaxs = "i", xaxs = "i", main = deparse(substitute(color.slice)))
-  for(i in 1:(nc-1)) {
-    rect(chroma[i], luminance[-nl], chroma[i] + 100/(nc-1), luminance[-1], border = color.slice[,i+1], col = color.slice[,i+1])
-  }
-  axis(1)
-  axis(2)
-  box()
-}
 @
@@
 \begin{abstract}
-
 Visualizing independence by association and mosaic plots.
@@
 
-Arthritis data \citep{vcd:Koch+Edwards:1988}
-
-UCB admissions data \citep{vcd:Bickel+Hammel+O'Connell:1975}
-
 \section{Independence in 2-way Tables}
 \label{sec:2way}
+
+In this section, the basic tools for testing and visualizing 
+independence in 2-way tables are briefly reviewed. 
+For illustration, a data set about treatment and improvement of
+patients with rheumatoid arthritis from \cite{vcd:Koch+Edwards:1988}
+is used. The data set is also discussed in \cite{vcd:Friendly:2000}
+and the subset of the $59$ female patients from the study is given in
+Table~\ref{tab:art}.
 
@@
 
-We consider a 2-way contingency table with cell frequencies $\{n_{ij}\}$ for
-$i = 1, \dots, I$ and $j = 1, \dots, J$ and row and column sums
-$n_{i+} = \sum_i n_{ij}$ and $n_{+j} = \sum_j n_{ij}$ respectively. For
-convenience the number of observations is denoted $n = n_{++}$.
-
+To fix notations, we consider a 2-way contingency table with cell
+frequencies $\{n_{ij}\}$ for $i = 1, \dots, I$ and $j = 1, \dots, J$
+and row and column sums $n_{i+} = \sum_i n_{ij}$ and $n_{+j} = \sum_j n_{ij}$
+respectively.
+For convenience the number of observations is denoted $n = n_{++}$.
 Given an underlying distribution with theoretical cell probabilities
@@
 The expected cell frequencies in this model are
-$\hat n_{ij} = n_{i+}n_{+j}/n$. The best known and most used
-measure of discrepancy
-between observed and expected values are the Pearson residuals
+$\hat n_{ij} = n_{i+}n_{+j}/n$. The probably best known and most used
+measure of discrepancy between observed and expected values are the Pearson residuals
 \begin{equation} \label{eq:Pearson}
@@
 \end{equation}
-Therefore, a rather intuitive idea is to reject the null hypothesis when
-there are residuals which are too extreme, i.e., not close enough to
-zero. The most convenient way to aggregate the $I \times J$ residuals
+The most convenient way to aggregate the $I \times J$ residuals
 to one test statistic is their squared sum
@@
 \end{equation}
-because this is known to have a limiting $\chi^2$ distribution with $(I-1)(J-1)$
+because this is known to have an unconditional limiting $\chi^2$ distribution with $(I-1)(J-1)$
 degrees of freedom under the null hypothesis.
-This is the well-known $\chi^2$ test for independence in 2-way tables.
-%% reference needed
+This is the well-known $\chi^2$ test which is typically introduced at the very beginning
+of the chapter about independence in 2-way tables in statistics textbooks
+\citep[see e.g.,][]{vcd:Agresti:2002}.
 
-But this is not the only plausible way of aggregation of the Pearson
-residuals. There are many conceivable functionals $\lambda(\cdot)$ which
-lead to reasonable test statistics $\lambda(\{r_{ij}\})$, the sum of squares
-is just one of them. Another functional suitable for identifying
-the cells which cause the dependence (if any) is the maximum of the absolute
-values
+\begin{table}[t!]
+\begin{center}
+\caption{\label{tab:art} Treatment and improvement among $59$ patients
+with rheumatoid arthritis.}
+\begin{tabular}{rr|rrr}
+                   &         & \multicolumn{3}{|l}{\textbf{Improved}} \\
+                   &         & None & Some & Marked \\ \hline
+\textbf{Treatment} & Placebo &   19 &    7 &      6 \\
+                   & Treated &    6 &    5 &     16
+\end{tabular}
+\end{center}
+\end{table}
+
+But the sum of squares is not the only plausible way of capturing deviations
+from $0$ in the Pearson residuals.
+There are many other conceivable functionals $\lambda(\cdot)$ which
+lead to reasonable test statistics $\lambda(\{r_{ij}\})$: one which is particularly
+suitable for identifying the cells which cause the dependence (if any) is
+the maximum of the absolute values
 \begin{equation} \label{eq:MaxAbs}
@@
 Given a critical value $c_\alpha$ for this test statistic, all residuals whose
-absolute value exceeds $c_\alpha$ violate the hypothesis of indendence at level
-$\alpha$ \cite[][ch. 7]{vcd:Mazanec+Strasser:2000}. Thus, the interesting cells causing
+absolute value exceeds $c_\alpha$ violate the hypothesis of indendence at significance level
+$\alpha$ \cite[][Ch.~7]{vcd:Mazanec+Strasser:2000}. Thus, the interesting cells causing
 the dependence can easily be identified.
 
-Furthermore, the main reason for using the unconditional limiting distribution
-for the $\chi^2$ statistic (\ref{eq:ChiSquare}) was the closed form result
+Furthermore, an important reason for using the unconditional limiting distribution
+for the $X^2$ statistic (\ref{eq:ChiSquare}) was the closed form result
 for the distribution. Recently, with the improving perfomance of computers,
-performing permutation tests---either by simulation or by computation of the
-permutation distribution---became more and more popular. In particular for
-the independence hypothesis (\ref{eq:H0}), using a permutation test is very intuitive
-due to the permutation invariance (given row and column sums) of this problem.
-By employing this approach the permutation distribution of statistics of
-type $\lambda(\{r_{ij}\})$ (also including the $\chi^2$ statistic) can be derived.
+conditional inference (or permutation tests)---carried out either by simulation
+or by computation of the (asymptotic) permutation distribution---have been receiving
+increasing attention \citep{vcd:Pesarin:2001,vcd:Strasser+Weber:1999}. 
+For testing the independence hypothesis (\ref{eq:H0}), using a permutation test is particularly
+intuitive due to the permutation invariance (given row and column sums) of this problem.
+Consequently, all results in this paper are based on conditional inference performed
+by simulating from the permutation distribution of test statistics of type
+$\lambda(\{r_{ij}\})$.
 
-Other classical tests for the hypothesis of independence  not fitting in
-the Pearson residual-based framework described above include Fisher's exact test,
-asymptotic tests of the odds ratio (for $2\times2$ tables) and the Mantel-Haenzel
-test.
+Other measures of discrepancy (e.g., deviance residuals) could, of course, also be used
+instead of the the Pearson residuals $\{r_{ij}\}$. But as the ideas discussed here
+extend straightforwardly to that situation, we do not discuss it in detail.
+
+For the arthritis data from Table~\ref{tab:art}, both tests indicate a clearly significant
+dependence of improvement on treatment: the sum-of-squares statistic from Equation~(\ref{eq:ChiSquare})
+is $X^2 = \Sexpr{stat(art_chisq)}$ with a $p$ value of $p = \Sexpr{pval(art_chisq)}$,
+and the maximum statistion from Equation~(\ref{eq:MaxAbs}) is $M = \Sexpr{stat(art_max)}$
+with a $p$ value of $p = \Sexpr{pval(art_max)}$. Both $p$ values have been computed
+by drawing a sample of size $5000$ from the permutation distribution under independence.
 
@@
 
-ONLY SHAPE! not color
-
-The two best known visualization techniques for independence in 2-way tables are
-association plots and mosaic plots. Both are suitable to bring out
-departures of an observed table $(n_{ij})$ from the expected table $(\hat n_{ij})$ in a graphical way.
-The former focuses on the visualization of the Pearson residuals $r_{ij}$ (under
-independence) while the latter primarily displays the observed frequencies
+Two well-established visualization techniques for independence in 2-way tables are
+mosaic plots and association plots. Both are suitable to bring out
+departures of an observed table $\{n_{ij}\}$ from the expected table $\{\hat n_{ij}\}$
+in a graphical way. The latter focuses on the visualization of the Pearson residuals
+$r_{ij}$ (under independence) while the latter primarily displays the observed frequencies
 $n_{ij}$.
 
-\textbf{Association Plots}
+\emph{Mosaic plots}
+\citep{vcd:Hartigan+Kleiner:1981} can be seen as an extension of grouped
+bar charts where width and height of the bars show the relative
+frequencies of the two variables: a mosaic plot simply
+consists of a collection of tiles whose
+sizes are proportional to the observed cell frequencies as shown
+in the left panel of Figure~\ref{fig:classic}. A rectangle corresponding
+to $100$ percent of the observations is first split horizontally with respect to 
+the treatment frequencies and then vertically with respect to the conditional
+improvement frequencies. This shows that there have been more placebo than treated 
+patients with no improvement and vice versa for marked improvement.
+This strategy of splitting with respect to conditional frequencies given all
+previous variables can also directly be used for visualizing multi-way tables
+\citep[see][for an overview of how to construct and read mosaic displays]{vcd:Hofmann:2003}.
+
+\emph{Association plots}
 \citep{vcd:Cohen:1980} visualize the table of Pearson residuals:
@@
 square root of the expected counts $\sqrt{\hat n_{ij}}$. Thus, the area is proportional to
-the raw residuals $n_{ij} - \hat n_{ij}$. The sign of the residual is redundantly
-coded by the rectangle's color and its position relative to the baseline.
-
-%% Figure \ref{fig:goals-assoc-classic} shows the association plot for the table of
-%% home and away goals for all 306 games in the 1995/6 season of the German soccer league
-%% Bundesliga \citep[the data is available in the package \texttt{vcd}]{vcd:Knorr-Held:1999}.
-%% In particular, it can be seen that there are
-%% fewer games ending 1--0 and more ending 1--1 than would be expected under independence.
-%% But this does not yet imply the significance of some test statistic.
-
-\textbf{Mosaic Plots}
-\citep{vcd:Hartigan+Kleiner:1981} can be seen as an extension of grouped
-bar charts where width and height of the bars show the relative
-frequencies of the two variables: a mosaic plot simply
-consists of a collection of tiles whose
-sizes are proportional to the observed cell frequencies
-(see Figure \ref{fig:goals-mosaic-classic}).
-
-Sequential horizontal and vertical recursive splits are used to
-visualize the frequencies of more than two variables, each new variable
-conditional to the previously entered variables. A first extension by
-\cite{vcd:Friendly:1994} uses a color coding of the tiles to visualize deviations
-(residuals) from a given log-linear model fitted to the table, that is, from
-the expected frequencies under independence. This approach does not only work in 2-way
-tables but also in log-linear models fitted to multi-way tables. In this
-extension, positive and negative signs of the residuals are coded by rectangles with
-solid and dashed borders respectively. Furthermore, residuals exceeding an absolute
-value of 2 are shaded light blue and red respectively, those that even
-exceed an absolute value of 4 are shaded with full saturation. The heuristic
-behind this shading is that
-the Pearson residuals are approximately standard normal which implies
-that the highlighted cells are those with
-residuals \emph{individually} significant at approximately the 5\% and 0.01\%
-level. But the main purpose of the shading is not to visualize significance but the \emph{pattern} of
-deviation from independence \citep[p.~109]{vcd:Friendly:2000}. In particular,
-this shading does not provide a visualization of the maximum test (\ref{eq:MaxAbs})
-as in general it is unknown to which significance level $\alpha$ the
-values $c_\alpha = 2$ and $4$ correspond for the table being visualized.
+the raw residuals $n_{ij} - \hat n_{ij}$. The association plot for the arthritis
+data is shown in the right panel of Figure~\ref{fig:classic} which leads to the
+same interpretation as the mosaic plot: there are more placebo patients with no 
+improvement and fewer with marked improvement than expected under independence---vice
+versa for the treated patiens.
 
@@
 remarks: Friendly 1994
+
+A first extension by
+\cite{vcd:Friendly:1994} uses a color coding of the tiles to visualize deviations
+(residuals) from a given log-linear model fitted to the table, that is, from
+the expected frequencies under independence. This approach does not only work in 2-way
+tables but also in log-linear models fitted to multi-way tables. In this
+extension, positive and negative signs of the residuals are coded by rectangles with
+solid and dashed borders respectively. Furthermore, residuals exceeding an absolute
+value of 2 are shaded light blue and red respectively, those that even
+exceed an absolute value of 4 are shaded with full saturation. The heuristic
+behind this shading is that
+the Pearson residuals are approximately standard normal which implies
+that the highlighted cells are those with
+residuals \emph{individually} significant at approximately the 5\% and 0.01\%
+level. But the main purpose of the shading is not to visualize significance but the \emph{pattern} of
+deviation from independence \citep[p.~109]{vcd:Friendly:2000}. In particular,
+this shading does not provide a visualization of the maximum test (\ref{eq:MaxAbs})
+as in general it is unknown to which significance level $\alpha$ the
+values $c_\alpha = 2$ and $4$ correspond for the table being visualized.
 
@@
 hue0 <- hue.slice(0, grid.n = 101, plot = FALSE, type = "HSV")
-mysaturation <- as.numeric(colnames(hue23))
+saturation <- as.numeric(colnames(hue23))
 value <- as.numeric(rownames(hue23))
@@
 for(i in 1:(nc-1)) {
-  rect(mysaturation[i], value[-nl], mysaturation[i] + 1/(nc-1), value[-1], border = hue23[,i+1], col = hue23[,i+1])
+  rect(saturation[i], value[-nl], saturation[i] + 1/(nc-1), value[-1], border = hue23[,i+1], col = hue23[,i+1])
 }
 for(i in 1:(nc-1)) {
-  rect(-mysaturation[i], value[-nl], -(mysaturation[i] + 1/(nc-1)), value[-1], border = hue0[,i+1], col = hue0[,i+1])
+  rect(-saturation[i], value[-nl], -(saturation[i] + 1/(nc-1)), value[-1], border = hue0[,i+1], col = hue0[,i+1])
 }
@@
 pushViewport(viewport(layout.pos.col=1, layout.pos.row=1))
-mosaic(art, newpage = FALSE, gp = shading_Friendly, margins = c(2.5, 4, 2.5, 3))
+mosaic(art, newpage = FALSE, gp = shading_Friendly, margins = c(2.5, 4, 2.5, 3), gp_args = list(eps = NULL))
 popViewport()
 pushViewport(viewport(layout.pos.col=2, layout.pos.row=1))
-set.seed(1234)
-mosaic(art, newpage = FALSE, gp = shading_max, margins = c(2.5, 3, 2.5, 4))
+mosaic(art, newpage = FALSE, gp = art_gp, margins = c(2.5, 3, 2.5, 4))
 popViewport(2)
@@
 \section{Extensions to Conditional Independence}
+
+UCB admissions data \citep{vcd:Bickel+Hammel+O'Connell:1975}
+double max: \Sexpr{stat(ucb_doublemax)}, p-value = \Sexpr{pval(ucb_doublemax)}
+max Chisq: \Sexpr{stat(ucb_maxChisq)}, p-value = \Sexpr{pval(ucb_maxChisq)}
+sum Chisq: \Sexpr{stat(ucb_sumChisq)}, p-value = \Sexpr{pval(ucb_sumChisq)}
+simulated from $5000$ repetitions
 
@@
 <<conditional_assoc,fig=TRUE,echo=FALSE,height=6,width=9>>=
-set.seed(1234)
-cotabplot(~ Admit + Gender | Dept, data = ucb, panel = cotab_coindep, type = "assoc")
+cotabplot(~ Gender + Admit | Dept, data = ucb, panel = ucb_cotab)
 @

### 2005-08-03 trunk/vcd/paper/JCGS.Rnw
essentially finished Section 2, started Section 3
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@426 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-remarks: Friendly 1994
+Colors are commonly used to enhance mosaic and association plots. To integrate
+a visualization of the Pearson residuals $\{r_{ij}\}$ into the mosaic display---which
+in its `raw' version only visualizes the observed frequencies $\{n_{ij}\}$---\cite{vcd:Friendly:1994}
+suggested a residual-based shading for the mosaic tiles that can also be applied
+to the rectangles in association plots \citep{vcd:Meyer+Zeileis+Hornik:2003}. 
+In this section, we first briefly review the \cite{vcd:Friendly:1994} shading, 
+before we suggest different colors and a combination of visualization and significance
+testing.
+
+\subsection{Friendly Shading}
 
@@
 level. But the main purpose of the shading is not to visualize significance but the \emph{pattern} of
-deviation from independence \citep[p.~109]{vcd:Friendly:2000}. In particular,
-this shading does not provide a visualization of the maximum test (\ref{eq:MaxAbs})
-as in general it is unknown to which significance level $\alpha$ the
-values $c_\alpha = 2$ and $4$ correspond for the table being visualized.
+deviation from independence \citep[p.~109]{vcd:Friendly:2000}.
+
+The extensions of \cite{vcd:Friendly:1994} to mosaic plots provide
+substantial improvement of the original mosaic plots and enhanced them
+from a plot for contingency tables to a visualization technique for
+log-linear models (and therefore also for certain independence problems).
+However, it has two major drawbacks: First,
+the significance level for the hard-coded critical values 2 and 4 is
+usually unknown. Second, the colors for (light) blue and red
+used in the Michael Friendly's \textsf{SAS} implementation and also in the base implementation
+in \textsf{R} are not device-independent and
+do not provide homogeneous saturations over different colors
+and copier proofness. We suggest some ways to overcome these drawbacks.
+
+The implementation of Friendly mosaic plots in \textsf{SAS}
+uses (by default) colors in the HLS (Hue--Luminance--Saturation) color space
+with hues for blue and red, full saturation, and varying luminance
+for lighter colors. The current \textsf{R} implementation is
+based on colors in the HSV (Hue--Saturation--Value) space where
+decreasing the saturation from 1 towards 0
+%% ---given the hue for blue or red and full saturation---
+generates very similar colors
+as increasing the luminance for HLS colors from 0.5 towards 1.
+As an alternative to those HLS and HSV colors, the device-independent
+HCL (Hue--Chroma--Luminance) color space \citep{vcd:Ihaka:2003} could be
+used which also provides homogeneous saturations over different colors and copier
+proofness. The resulting association plot (using critical values for
+$\alpha = 0.1$ and $0.01$) for the Bundesliga data can be seen in
+Figure \ref{fig:goals-assoc-signif}. It indicates very clearly that
+there is no evidence for rejecting the hypothesis of independence
+in this table. Thus, there seems to be neither an advantage for
+the home team nor for the away team.
 
@@
 
-The extensions of \cite{vcd:Friendly:1994} to mosaic plots provide
-substantial improvement of the original mosaic plots and enhanced them
-from a plot for contingency tables to a visualization technique for
-log-linear models (and therefore also for certain independence problems).
-However, it has two major drawbacks: First,
-the significance level for the hard-coded critical values 2 and 4 is
-usually unknown. Second, the colors for (light) blue and red
-used in the Michael Friendly's \textsf{SAS} implementation and also in the base implementation
-in \textsf{R} are not device-independent and
-do not provide homogeneous saturations over different colors
-and copier proofness. We suggest some ways to overcome these drawbacks.
+In particular,
+the Friendly shading does not provide a visualization of the maximum test (\ref{eq:MaxAbs})
+as in general it is unknown to which significance level $\alpha$ the
+values $c_\alpha = 2$ and $4$ correspond for the table being visualized.
 
@@
 are derived in \texttt{vcd} by simulation of the underlying distribution.
-
-The implementation of Friendly mosaic plots in \textsf{SAS}
-uses (by default) colors in the HLS (Hue--Luminance--Saturation) color space
-with hues for blue and red, full saturation, and varying luminance
-for lighter colors. The current \textsf{R} implementation is
-based on colors in the HSV (Hue--Saturation--Value) space where
-decreasing the saturation from 1 towards 0
-%% ---given the hue for blue or red and full saturation---
-generates very similar colors
-as increasing the luminance for HLS colors from 0.5 towards 1.
-As an alternative to those HLS and HSV colors, the device-independent
-HCL (Hue--Chroma--Luminance) color space \citep{vcd:Ihaka:2003} could be
-used which also provides homogeneous saturations over different colors and copier
-proofness. The resulting association plot (using critical values for
-$\alpha = 0.1$ and $0.01$) for the Bundesliga data can be seen in
-Figure \ref{fig:goals-assoc-signif}. It indicates very clearly that
-there is no evidence for rejecting the hypothesis of independence
-in this table. Thus, there seems to be neither an advantage for
-the home team nor for the away team.
 
@@
 pushViewport(viewport(layout.pos.col=1, layout.pos.row=1))
-mosaic(art, newpage = FALSE, gp = shading_Friendly, margins = c(2.5, 4, 2.5, 3), gp_args = list(eps = NULL))
+mosaic(art, newpage = FALSE, gp = shading_Friendly, margins = c(2.5, 4, 2.5, 3), gp_args = list(eps = NULL, lty = 1))
 popViewport()
@@
 plots) \citep{vcd:Cleveland:1993} that lead to trellis graphics
-(Becker, R.A., Cleveland, W.S. and Shyu, M. ``The Visual Design and
-Control of Trellis Display'', Journal of Computational and
-Graphical Statistics).
+\citep{vcd:Becker+Cleveland+Shyu:1996}.
 

### 2005-08-03 trunk/vcd/paper/JCGS.Rnw
finished Sections 3.1 and 3.2
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@427 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 colors and color spaces
+    because the three perceptual dimensions (hue, lightness, saturation)
+    are not properly mapped to the three dimensions of the color space
+    and hence are confounded \citep{vcd:Brewer:1999}
 
@@
 before we suggest different colors and a combination of visualization and significance
-testing.
+testing to extend these residual-based shadings.
 
@@
 
-A first extension by
-\cite{vcd:Friendly:1994} uses a color coding of the tiles to visualize deviations
-(residuals) from a given log-linear model fitted to the table, that is, from
-the expected frequencies under independence. This approach does not only work in 2-way
-tables but also in log-linear models fitted to multi-way tables. In this
-extension, positive and negative signs of the residuals are coded by rectangles with
-solid and dashed borders respectively. Furthermore, residuals exceeding an absolute
-value of 2 are shaded light blue and red respectively, those that even
-exceed an absolute value of 4 are shaded with full saturation. The heuristic
-behind this shading is that
-the Pearson residuals are approximately standard normal which implies
-that the highlighted cells are those with
-residuals \emph{individually} significant at approximately the 5\% and 0.01\%
+The extensions of \cite{vcd:Friendly:1994} to mosaic plots provide a
+substantial improvement of the original mosaic plots and enhanced them
+from a plot for contingency tables to a visualization technique for
+log-linear models---and thus also for independence problems including
+2-way tables as the simplest special case.
+
+The idea is to use a color coding for the mosaic tiles that visualizes the
+sign and absolute size of each residual $r_{ij}$: Cells corresponding to small
+residuals ($|r_{ij}| < 2$) are shaded white. Cells with medium sized residuals
+($2 \le |r_{ij}| < 4$) are shaded in light blue and light red for positive negative
+residuals respectively. Cells with large residuals ($|r_{ij}| \ge 4$) are shaded
+with a fully saturated blue and red respectively. Mosaic plots enhanced by this 
+shading can thus also bring out departures from independence (or other log-liner 
+models in multi-way tables) graphically and visualize patterns of dependence.
+The heuristic for choosing the cut offs $2$ and $4$ is that the Pearson residuals
+are approximately standard normal which implies that the highlighted cells are those with
+residuals \emph{individually} significant at approximately the $\alpha = 0.05$ and $\alpha = 0.0001$
 level. But the main purpose of the shading is not to visualize significance but the \emph{pattern} of
@@
 
-The extensions of \cite{vcd:Friendly:1994} to mosaic plots provide
-substantial improvement of the original mosaic plots and enhanced them
-from a plot for contingency tables to a visualization technique for
-log-linear models (and therefore also for certain independence problems).
-However, it has two major drawbacks: First,
-the significance level for the hard-coded critical values 2 and 4 is
-usually unknown. Second, the colors for (light) blue and red
-used in the Michael Friendly's \textsf{SAS} implementation and also in the base implementation
-in \textsf{R} are not device-independent and
-do not provide homogeneous saturations over different colors
-and copier proofness. We suggest some ways to overcome these drawbacks.
-
-The implementation of Friendly mosaic plots in \textsf{SAS}
-uses (by default) colors in the HLS (Hue--Luminance--Saturation) color space
-with hues for blue and red, full saturation, and varying luminance
-for lighter colors. The current \textsf{R} implementation is
-based on colors in the HSV (Hue--Saturation--Value) space where
-decreasing the saturation from 1 towards 0
-%% ---given the hue for blue or red and full saturation---
-generates very similar colors
-as increasing the luminance for HLS colors from 0.5 towards 1.
-As an alternative to those HLS and HSV colors, the device-independent
-HCL (Hue--Chroma--Luminance) color space \citep{vcd:Ihaka:2003} could be
-used which also provides homogeneous saturations over different colors and copier
-proofness. The resulting association plot (using critical values for
-$\alpha = 0.1$ and $0.01$) for the Bundesliga data can be seen in
-Figure \ref{fig:goals-assoc-signif}. It indicates very clearly that
-there is no evidence for rejecting the hypothesis of independence
-in this table. Thus, there seems to be neither an advantage for
-the home team nor for the away team.
+In addition to the shading of the rectangles themselves, the Friendly shading also
+encompasses a choice of line type and line color of the borders of the rectangles
+with similar ideas as described above. As both mosaic and association plots are
+area-proportional visualization techniques, we focus on area shadings and always 
+use solid black borders throughout this paper, but the extensions suggested in the
+following could also be applied to control line type and color.
 
@@
 
-currently: chosen based on HSV or HLS color spaces,
-hue (red and blue) codes sign of residuals, and saturation the absolute size
-in three steps, value currently unused (always set to full value)
+The way the (light) blue and red colors are chosen differs somewhat between
+various implementations of Friendly mosaic plots: In his original 
+\proglang{SAS} implementation \citep[see][]{vcd:Friendly:2000}, Michael Friendly
+uses colors from a palette based on HLS (Hue--Luminance--Saturation) color space.
+The implementation in the \proglang{R} system for statistical computing
+\citep[\url{http://www.R-project.org/}]{vcd:R:2005} employs colors from
+HSV (Hue--Saturation--Value) space. The latter is a very common implementation of
+colors in many computer packages and makes the generation of the Friendly shading
+very simple.
 
-disadvantages:
-  - perceptually not uniform
-    because the three perceptual dimensions (hue, lightness, saturation)
-    are not properly mapped to the three dimensions of the color space
-    and hence are confounded \citep{vcd:Brewer:1999}    
-    in particular, saturation is not uniform across different hues.
+The HSV space originally looks like a regular cone with black at its
+peak (zero value) and full color wheels for different saturations at the other end,
+centered around a white (full value). Type and amount of color are controlled by hue
+and saturation respectively. Typically, polar coordinates that all range in the 
+unit interval are used in this space, such that it looks like a regular cylinder.
+For generating colors in the Friendly shading, the following strategy is used:
+The hue $h$ codes the sign of the residuals---$h = 0$ (red hue) is used for negative residuals,
+$h = 2/3$ (blue hue) for positive residuals. The absolute size of the residuals
+is then coded by the saturation $s$ which is set to $0$, $0.5$ and $1$ respectively
+for small/medium/large residuals respectively. The value is always fixed at $v = 1$.
+This is also depicted in the upper panel of Figure~\ref{fig:shadingHSVHCL} which shows
+to slides from the HSV cylinder side to side. The plot shows the saturation/value plane
+for the given hues $h = 0$ and $h = 2/3$. The significant color palette shows the
+colors used for the Friendly shading when the residuals are increasing from left to right.
+The set of non-significant colors will be explained in the next section.
 
-  - can lead to color-caused optical illusions in statistical graphs
-    \citep{vcd:Cleveland+McGill:1983}
+Although this HSV-based shading is already very useful for enhancing mosaic and
+association plots and although HSV is very commonly available implementation of
+color spaces, HSV colors in general and the Friendly shading in particular have
+a number of disadvantages. Most importantly, HSV colors are not perceptually
+uniform because the three HSV dimensions map only poorly to the three
+perceptual dimensions of the human visual system  \citep{vcd:Brewer:1999,vcd:Ihaka:2003}.
+Consequently, the the HSV dimensions are confounded, e.g., saturation
+is not uniform across different hues. A fully saturated blue $(2/3, 1, 1)$ is
+much darker as a fully saturated red $(0, 1, 1)$ or green $(1/3, 1, 1)$.
+This makes it more difficult for the human eye to judge the size of shaded areas
+and can therefore lead to color-caused optical illusions when used in statistical graphs
+\citep{vcd:Cleveland+McGill:1983}. Furthermore, flashy fully saturated HSV colors are
+good for drawing attention to a plot, but hard to look at for a longer time \citep{vcd:Ihaka:2003}
+which makes HSV-shaded graphics harder to interpret. Finally, white is employed as
+the neutral color for small residuals in the Friendly shading, but typically
+grey is found to convey neutrality or uninterestingness much better than white does
+\citep{vcd:Brewer:1999}.
 
-  - flashy fully saturated HSV colors are good for drawing attention to a plot,
-    but hard to look at for a longer time, hence mosaic displays using these 
-    colors are harder to interpret
+Alternative ways to choose colors have been available for a long time, but have
+been only slowly adopted for implementations of colors in computer packages in 
+general and for shading in statistical graphs in particular. The idea of using
+perceptually based colors that are `in harmony' go back to \cite{vcd:Munsell:1905} 
+who introduced a color notation for balanced colors. Based on similar principles, Cynthia
+Brewer and co-workers suggested different types of palettes (qualitative/sequential/diverging)
+and provided the online tool \pkg{ColorBrewer.org} \citep{vcd:Harrower+Brewer:2003} for
+selecting an appropriate palette for a specific problem. Furthermore, the
+Commission Internationale de l'\'Eclairage (CIE) introduced the two perceptually
+based color spaces CIELAB and CIELUV where the latter is typically preferred for 
+emissive color technologies such as computer displays. \cite{vcd:Ihaka:2003} discusses
+how CIELUV colors can be used for choosing colors for statistical graphics such as barplots.
+By taking polar coordinates in CIELUV space, it is called HCL (Hue--Chroma--Luminance) space
+and qualitative palettes can easily be chosen by using a range of hues for fixed values
+of chroma and luminance. Such colors are always balanced towards the same grey 
+and thus do not have the problem of varying saturations as for the HSV colors.
 
-  - white is not very suitable as a neutral color, grey conveys neutrality or 
-    uninterestingness much better
+Here, we discuss how similar ideas can be used for deriving a diverging HCL palettes that 
+provide a suitable translation of the ideas from the Friendly shading to perceptually
+uniform HCL colors. 
+The HCL space looks like a double cone with black (zero luminance) at one end and white
+at the other (full luminance). In its middle, there is a full color wheel for different
+values of chroma (that controls the colorfulness). Unfortunately, the HCL space is not completely
+regular as the HSV space is and consequently its dimensions cannot be standardized to 
+unit intervals: the hue ranges in $[0, 360]$ degrees and chroma and luminance both in $[0, 100]$
+percent. But not all combinations $(h, c, l)$ yield valid HCL colors and the admissable combinations
+of $c$ and $l$ vary across different hues $h$. But for the problem of constructing
+a diverging palette, this problem can easily be overcome as we just need two different
+hues (a `negative' and a `positive' hue) and hence choose two hues that correspond to 
+similar shapes in the chroma/luminance plane. The lower panel of Figure~\ref{fig:shadingHSVHCL}
+shows two such planes side to side for the hues $h = 0$ and $h = 260$. To obtain a 
+sequence of colors with the same properties as the Friendly shading, the palette starts
+at a fully saturated red $(0, 100, 50)$, goes via a neutral color, ends at a fully saturated
+blue $(260, 100, 50)$, and uses linear interpolation in between. Instead of using white
+$(0, 0, 100)$ as the neutral color, a light grey $(0, 0, 90)$ is employed as motivated above.
 
-Alternatives:
-  - \cite{vcd:Munsell:1905} introduced color notation for balanced colors
-  - Commission Internationale de l'\'Eclairage (CIE) introduced two perceptually
-    based color spaces CIELAB and CIELUV where the latter is preferred for 
-    emissive color technologies such as computer displays.
-  - \pkg{ColorBrewer.org} \citep{vcd:Harrower+Brewer:2003} is an online tool
-    for selcting different sets of colors, based on similar ideas. It includes
-    qualitative, sequential and diverging palettes.
-  - \cite{vcd:Ihaka:2003} discusses how CIELUV colors can be used for choosing
-    colors for statistical graphics such as barplots. He recommends to use
-    polar coordinates in the CIELUV space---called HCL space, using hue, chroma
-    and luminance---and suggests a strategy for choosing qualitative schemes based
-    on these.
+This diverging palette (see the `significant' colors in Figure~\ref{fig:shadingHSVHCL},
+the `non-significant' colors are again discussed in the following section)
+uses both chroma, i.e., the colorfulness, and luminance, i.e., 
+the amount of grey, to code the absolute size of the quantity visualized, i.e., the
+Pearson residuals when applied to the independence problem. By changing the neutral
+color or by changing the maximum chroma respectively, this can be changed to using only
+chroma or luminance for this purpose, but using both seems to be a very effective way
+of visualization.
 
-We discuss how similar ideas can be used for deriving a diverging HCL scheme that 
-is suitable for visualizing residuals in association and mosaic displays.
-
-HCL
-  - irregular double cone, i.e., chroma/luminance plane looks rather different
-    for different hues
-  - choose two hues that look similar in the chroma/luminance plane to code
-    the sign of the residuals. Also use red and blue at full chroma for large
-    residuals and use a light gray (instead of a white at the top of the luminance
-    scale) as the neutral color for small residuals, interpolate between.
-  - for non-significant results, clearly reduce the amount of color, i.e. the maximum
-    chroma used.
-
-Advantages
-  - based on a perceptually uniform color space
-  - device independent
+The advantages of using this HCL shading scheme instead of the HSV scheme is that
+it is based on a perceptually based color space that provides uniform saturations
+for different hues and is device inependent.
 
@@
 }
-axis(2, at = c(50, 75, 100), labels = c(50, 75, 100))
-axis(4, at = c(50, 75, 100), labels = c(50, 75, 100))
+axis(2, at = c(50, 70, 90, 100), labels = c(50, 70, 90, 100))
+axis(4, at = c(50, 70, 90, 100), labels = c(50, 70, 90, 100))
 axis(3, at = -4:4*25, labels=c(4:0*25, 1:4*25))
@@
 @
-\caption{Extended shading in HSV and HCL space.}
+\caption{Extended shading in HSV (upper) and HCL space (lower).}
 \label{fig:shadingHSVHCL}
@@
 \subsection{Significance}
+
+However, it has two major drawbacks: First,
+the significance level for the hard-coded critical values 2 and 4 is
+usually unknown.
+
+  - for non-significant results, clearly reduce the amount of color, i.e. the maximum
+    chroma used.
 

### 2005-08-03 trunk/vcd/paper/JCGS.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@428 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-However, it has two major drawbacks: First,
-the significance level for the hard-coded critical values 2 and 4 is
-usually unknown.
+The shading scheme of \cite{vcd:Friendly:1994} was suggested to visualize
+the pattern of dependence in contingency tables, as discussed above, but
+the presence (or absence) of colors in a plot also always conveys an
+impression of interestingness (or uninterstingness respectively). That is,
+statisticians might be tempted to interpret the absence of color in a plot
+as a clue that there is no significant departure from independence. Or vice
+versa, colored cells would convey the impression that there is significant
+dependence. Currently, both is not true as can be seen in the left panel
+of Figure~\ref{fig:shadings} which shows the mosaic display for the arthritis
+data with Friendly shading. Although there is significant dependence, no 
+residual exceeds an absolute value of $2$ and hence no cell is colored.
+Of course, it can be argued that the shading was not designed for this purpose
+and that different cut offs than $2$ and $4$ should be used here. But then
+again, it would be nice if these cut offs could be chosen automatically in a
+data-driven fashion which is what we do below.
 
-  - for non-significant results, clearly reduce the amount of color, i.e. the maximum
-    chroma used.
+The Friendly shading can be interpreted to be a visualization of the maximum
+statistic $M$ (\ref{eq:MaxAbs}) which always employs the critical values $c_\alpha$
+of $2$ and $4$ respectively. But the problem is that it is not clear to which
+significance levels $\alpha$ these critical values correspond because the distribution
+of $M$ depends on the underlying contingency table. The natural solution to this
+problem is to compute the critical values from the distribution of $M$ in a data-driven
+way (i.e., for the table visualized) and use these instead of the hard-coded values
+$2$ and $4$. In the right panel of Figure~\ref{fig:shadings} this is done for
+the arthritis data by employing the critical values
+$\Sexpr{round(as.vector(art_max$qdist(0.9)), digits = 2)}$ at level $\alpha = 0.1$ and
+$\Sexpr{round(as.vector(art_max$qdist(0.99)), digits = 2)}$ at level $\alpha = 0.01$
+(and using the diverging HCL palette). By using these cut offs, the presence of color
+in the plot is equivalent to significance (of the maximum statistic $M$) at 
+level $\alpha = 0.1$ and $\alpha = 0.01$ respectively and exactly the cells which `cause'
+the dependence are highlighted. For the arthritis data, these are in particular the 
+cells in the last column that signal that there are significantly more
+treated patients and fewer placebo patients with marked improvement than would be expected
+under independence.
 
-In particular,
-the Friendly shading does not provide a visualization of the maximum test (\ref{eq:MaxAbs})
-as in general it is unknown to which significance level $\alpha$ the
-values $c_\alpha = 2$ and $4$ correspond for the table being visualized.
+The significance $alpha = 0.1$ and $0.01$ are chosen because this leads to displays where
+fully colored cells are clearly significant (i.e., with a $p$ value below $0.01$),
+cells without color are clearly non-significant (i.e., with a $p$ value above $0.1$), and
+cells in between can be considered to be weakly significant (i.e., $0.1 \le p < 0.01$).
+Of course, users could choose any other set of significance levels they feel comfortable
+with, e.g., only a single cutoff at $\alpha = 0.05$ or three cut offs at $0.1$, $0.05$ and $0.01$
+etc. Another option could be to use a continuous shading where the $p$ value corresponding
+to a cell controls the interpolation between the neutral and the full color. However, our
+experience is that this typically results in too much color in a plot which in turn tends
+to conceal the important cells and over-emphasizes the unimportant ones. Hence, we stick
+to a shading with fewer colors that are easier to interpret.
 
-As pointed out in Section \ref{sec:tests}, the critical values for
-the maximum statistic $M$ from (\ref{eq:MaxAbs}) can be derived
-from the permutation distribution. Instead of the hard-coded
-values 2 and 4, the particular critical values, e.g., at the levels
-$\alpha = 0.1$ and $0.01$, for the table to be visualized could be
-used. Hence, exactly those residuals causing the (potential)
-dependency within the table are highlighted. At the moment, these critical values
-are derived in \texttt{vcd} by simulation of the underlying distribution.
+This extended shading is already very flexible and combines visualization and inference,
+but it should only be applied when using the maximum statistic is appropriate. For 
+the sum-of-squares statistic $X^2$ (and all other functionals $\lambda(\cdot)$), this
+shading cannot be used.
 
-To illustrate what a `postive' example looks like we give the association plot
-for the well-known Hair-Eye-Color data set (avilable in base \textsf{R}) which cross-tabulates the hair and eye
-color and gender of 592 statistics students: Figure \ref{fig:haireye-assoc-signif}
-shows the association plot for the 328 female students. It is easy to see that
-the hypothesis of independence is rejected at 1\% level by the maximum test (\ref{eq:MaxAbs})
-due to higher frequencies of students with hair and eye color
-black/brown and blond/blue and lower frequencies of
-blond/brown and brown/blue than would be plausible under independence. Furthermore,
-it can also be seen that the residual corresponding to the black/blue
-cell is significant at the 10\% level.
+use pre-determined (or manually adjusted) cut offs, but visualize the significance
+by the amount of color in the plot, i.e., use only little color if the corresponding
+test result is non-significant.
+
+For HCL, the amount of color can easily be controlled by varying the maximum
+chroma value. So far $100$, now $20$.
+
+For HSV, a similar effect can be achieved by varying the value dimension which
+was not used up to now. Setting value to $v = 0.5$ a darker palette of colors can be
+obtained.
 

### 2005-08-04 trunk/vcd/paper/JCGS.Rnw
section 3 finished
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@430 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-\textbf{Key Words:} Association plots; Conditional inference; 
+\noindent \textbf{Key Words:} Association plots; Conditional inference; 
 Contingency tables; HCL colors; HSV colors; Mosaic plots; Permutation tests.
@@
 Furthermore, an important reason for using the unconditional limiting distribution
-for the $X^2$ statistic (\ref{eq:ChiSquare}) was the closed form result
+for the $X^2$ statistic from (\ref{eq:ChiSquare}) was the closed form result
 for the distribution. Recently, with the improving perfomance of computers,
@@
 increasing attention \citep{vcd:Pesarin:2001,vcd:Strasser+Weber:1999}. 
-For testing the independence hypothesis (\ref{eq:H0}), using a permutation test is particularly
+For testing the independence hypothesis from (\ref{eq:H0}), using a permutation test is particularly
 intuitive due to the permutation invariance (given row and column sums) of this problem.
@@
 For the arthritis data from Table~\ref{tab:art}, both tests indicate a clearly significant
-dependence of improvement on treatment: the sum-of-squares statistic from Equation~(\ref{eq:ChiSquare})
+dependence of improvement on treatment: the sum-of-squares statistic from (\ref{eq:ChiSquare})
 is $X^2 = \Sexpr{stat(art_chisq)}$ with a $p$ value of $p = \Sexpr{pval(art_chisq)}$,
-and the maximum statistion from Equation~(\ref{eq:MaxAbs}) is $M = \Sexpr{stat(art_max)}$
+and the maximum statistion from (\ref{eq:MaxAbs}) is $M = \Sexpr{stat(art_max)}$
 with a $p$ value of $p = \Sexpr{pval(art_max)}$. Both $p$ values have been computed
-by drawing a sample of size $5000$ from the permutation distribution under independence.
+from a sample of size $5000$ from the permutation distribution under independence
+generated via sampling tables with the same row and column sums $n_{i+}$ and $n_{+j}$
+using the \cite{vcd:Patefield:1981} algorithm and computing the maximum statistic for 
+each of these tables.
 
@@
 The Friendly shading can be interpreted to be a visualization of the maximum
-statistic $M$ (\ref{eq:MaxAbs}) which always employs the critical values $c_\alpha$
+statistic $M$ from (\ref{eq:MaxAbs}) which always employs the critical values $c_\alpha$
 of $2$ and $4$ respectively. But the problem is that it is not clear to which
@@
 $\Sexpr{round(as.vector(art_max$qdist(0.99)), digits = 2)}$ at level $\alpha = 0.01$
-(and using the diverging HCL palette). By using these cut offs, the presence of color
+(and using the diverging HCL palette) derived from the permutation distribution of $M$ for 
+the arthritis data as described in Section~\ref{sec:tests}. By using these cut offs, the presence of color
 in the plot is equivalent to significance (of the maximum statistic $M$) at 
@@
 
-The significance $alpha = 0.1$ and $0.01$ are chosen because this leads to displays where
+The significance $\alpha = 0.1$ and $0.01$ are chosen because this leads to displays where
 fully colored cells are clearly significant (i.e., with a $p$ value below $0.01$),
@@
 
-This extended shading is already very flexible and combines visualization and inference,
-but it should only be applied when using the maximum statistic is appropriate. For 
-the sum-of-squares statistic $X^2$ (and all other functionals $\lambda(\cdot)$), this
-shading cannot be used.
-
-use pre-determined (or manually adjusted) cut offs, but visualize the significance
-by the amount of color in the plot, i.e., use only little color if the corresponding
-test result is non-significant.
-
-For HCL, the amount of color can easily be controlled by varying the maximum
-chroma value. So far $100$, now $20$.
-
-For HSV, a similar effect can be achieved by varying the value dimension which
-was not used up to now. Setting value to $v = 0.5$ a darker palette of colors can be
-obtained.
-
-We have seen that this approach performs very well when the test statistic
-which should be visualized is the maximum statistic (\ref{eq:MaxAbs}). However,
-it is difficult to extend this approach to general tests for independence,
-in particular in the HCL space, where the ranges of chroma and luminance are
-not independent. In contrast the HSV space has the advantage that all
-three dimensions hue, saturation and value have the range $[0,1]$ and can
-be varied independently. Therefore, a different approach than the one
-described above could be to use the hue for the sign of the residuals as
-before, the saturation for the absolute size of the residuals, and the
-value as an indicator for the significance of some test statistic (only
-using the full value when the overall test for independence rejects
-the null hypothesis). The resulting mosaic plots under this paradigm
-are depicted in Figure \ref{fig:both-mosaic-signif}. Again, it can
-clearly be seen that despite some large residuals there is no
-evidence against independence for the Bundesliga data, but that the
-null hypothesis has to be rejected for the Hair-Eye-Color data.
+This extended shading is already very flexible and combines visualization and inference.
+However, it can only be applied when employing the maximum statistic because it is the only
+aggregation functional $\lambda(\cdot)$ where a single large residual $|r_{ij}|$ exceeding
+its critical value (independent from the values of the remaining residuals) is equivalent
+to the a significant value of the whole test statistic $\lambda(\{r_{ij}\})$.\footnote{This
+can easily be seen from the following consideration: To obtain a level $\alpha$ test, where 
+any residual $|r_{ij}|$ exceeding the critical value $c_\alpha$ gives a significant test
+result, the equation $\alpha = P(\mbox{any } |r_{ij}| > c_\alpha)$ must hold. This is equivalent
+to $1 - \alpha = P(\mbox{all } |r_{ij}| \le c_alpha) = P(\max_{i,j} |r_{ij}| \le c_alpha)$.}
+Typically, applying the maximum statistic is feasible and also appropriate for exploratory 
+analysis, but it would be desirable to also have a residual-based shading that can incorporate
+visualization of significance when the sum-of-squares statistic $X^2$ (or any other functional
+$\lambda(\cdot)$) is used. For the reasons discussed above, it is not possible to achieve this by shading
+individual cells differently but can only be realized by using different colors for the
+whole table. A outlined before, colorfulness is often intuitively matched with interestingness,
+therefore a rather natural idea is to use the fully colored palette only when the corresponding
+test is significant and to use a less colorful palette if not. For the HCL scheme, the
+amount of color can easily be controlled by varying the maximum chroma value used. For the full
+colors, the maximum chroma was set to $100$ as shown in Figure~\ref{fig:shadingHSVHCL} and 
+is now reduced to $20$ for the non-significant palette. This palette still codes the absolute
+size of the residuals by luminance (i.e., the amount of grey), uses the same neutral grey
+for small `un-interesting' residuals, codes positive and negative residuals by different hues,
+but gives less emphasis to the pattern by making the plot less colorful. A similar effect
+can be obtained in the HSV space if the value is reduced from $1$ to $0.5$ for non-significant 
+palettes (see Figure~\ref{fig:shadingHSVHCL}). But as for the significant colors,
+the dimensions used for creating this palette are confounded and hence the HCL scheme is clearly
+preferable.
 
@@
 
+We have presented two approaches for combining visualization and significance testing
+via residual-based shadings. The first one alwyas uses the fully colored palette but relies on
+data-driven cut offs such that the presence of color is equivalent to significance of the
+associated maximum test. The second approach uses pre-defined cut offs (such as $2$ and $4$) but
+codes the result of the significance test by using full colors only for significant results and
+the same type of palette but with a reduced amount of color for non-significant results.
+
 \bibliography{vcd}
@@
 library(vcd)
-data(Arthritis)
 art <- xtabs(~ Treatment + Improved, data = Arthritis, subset = Sex == "Female")
-data(UCBAdmissions)
 ucb <- aperm(UCBAdmissions)
@@
 ## new max shading with HCL colors
-mosaic(art, gp = shading_Friendly)
-set.seed(1234)
+mosaic(art, gp = shading_Friendly(lty = 1, eps = NULL))
+set.seed(1071)
 mosaic(art, gp = shading_max)
@@
 ## conditional independence
-set.seed(1234)
+set.seed(1071)
 cotabplot(~ Admit + Gender | Dept, data = ucb, panel = cotab_coindep, type = "assoc")
 ## which uses the double max functional
-set.seed(1234)
+set.seed(1071)
 coindep_test(ucb, "Dept")
@@
 ## via simulation of conditional distribution:
-set.seed(1234)
+set.seed(1071)
 coindep_test(ucb, "Dept", indepfun = function(x) sum(x^2), aggfun = sum)

### 2005-08-04 trunk/vcd/inst/doc/strucplot.Rnw
Vignette Nr. 2 added. Temp files moved to pics subdir, which has been added to .Rbuildignore
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@421 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
sieveplot added to the strucplot framework; old function removed. line_col argument added to shading functions for default line color. dimnames added to independence_table.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@431 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-\SweaveOpts{engine=R,eps=TRUE}
+\SweaveOpts{engine=R,eps=FALSE,prefix.string=pics/strucplot}
 \setkeys{Gin}{width=0.8\textwidth}
@@
 higher-dimensional contingency tables and integrates techniques such as
-mosaic displays and association plots. The main idea is to visualize
+mosaic displays, association plots, and sieve plots. The main idea is to visualize
 the tables' cells arranged in rectangular form. To allow for more than
@@
 the plot. The actual workhorse is the chosen panel function: currently, 
-low-level panel functions for association plots (\code{struc\_assoc}) and 
-mosaic plots (\code{struc\_mosaic}) are available. In addition, there are
-high-level wrapper functions such as \code{mosaic()} and \code{assoc()} 
-to \code{strucplot()} that conveniently set the 
+low-level panel functions for association plots (\code{struc\_assoc}), 
+mosaic plots (\code{struc\_mosaic}), and sieve plots (\code{struc\_sieve})
+are available. In addition, there are
+high-level wrapper functions such as \code{mosaic()}, \code{assoc()},
+and \code{sieve()} to \code{strucplot()} that conveniently set the 
 parameters---it is these high-level functions that the user normally would like to use, 
-although \code{strucplot()} is fully functional on its own. \code{mosaic()} and 
-\code{assoc} are generic functions: for both, there exist a formula method and a default 
+although \code{strucplot()} is fully functional on its own. \code{mosaic()}, 
+\code{assoc}, and \code{sieve()} are generic functions: 
+for both, there exist a formula method and a default 
 method. Other functions, such as \code{doubledecker()}, \code{pairs()}, and 
 \code{cotabplot}, are on even higher level and make use of 
-code{mosaic()} and \code{assoc()} to produce plots.
+code{mosaic()}, \code{assoc()}, and \code{sieve()} to produce plots.
 
@@
 \begin{figure}[h]
-
 \begin{center}
@@
 
-\noindent The Pearson residuals (standardized deviations of observed from expected values) 
+\noindent In order to directly compare observed and expected values,
+we can use a sieve plot (see Figure \ref{fig:sieve}):
+\begin{Sinput}
+> sieve(Titanic)
+\end{Sinput}
+
+\setkeys{Gin}{width=0.7\textwidth}
+\begin{figure}[h]
+\begin{center}
+<<sieve,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
+sieve(Titanic)
+@
+\caption{Sieve plot for the Titanic data visualizing simultaneously
+  observed and expected values.}
+\label{fig:sieve}
+\end{center}
+\end{figure}
+
+\noindent The next step is to directly inspect the residuals. 
+The Pearson residuals (standardized deviations of observed from expected values) 
 are preferably visualized using association
@@
 
-\setkeys{Gin}{width=0.7\textwidth}
+\setkeys{Gin}{width=0.6\textwidth}
 \begin{figure}[h]

### 2005-08-04 trunk/vcd/R/cotabplot.R

@@ -164,6 +164,15 @@ cotab_mosaic <- function(x = NULL, condvars = NULL, ...) {
 }
 class(cotab_mosaic) <- "panel_generator"
 
+cotab_sieve <- function(x = NULL, condvars = NULL, ...) {
+  function(x, condlevels) {
+    if(is.null(condlevels)) sieve(x, newpage = FALSE, pop = TRUE, ...)
+      else sieve(co_table(x, names(condlevels))[[paste(condlevels, collapse = ".")]],
+                 newpage = FALSE, pop = TRUE, ...)
+  }
+}
+class(cotab_sieve) <- "panel_generator"
+
 cotab_assoc <- function(x = NULL, condvars = NULL, ylim = NULL, ...) {
   if(!is.null(x)) {
     fm <- coindep_test(x, condvars, n = 1)

### 2005-08-04 trunk/vcd/R/shadings.R

@@ -5,7 +5,7 @@
 
 shading_hsv <- function(observed, residuals = NULL, expected = NULL, df = NULL,
   h = c(2/3, 0), s = c(1, 0), v = c(1, 0.5),
-  interpolate = c(2, 4), lty = 1, eps = NULL,
+  interpolate = c(2, 4), lty = 1, eps = NULL, line_col = "black",
   p.value = NULL, level = 0.95, ...)
 {
   ## get h/s/v and lty
@@ -69,7 +69,7 @@ shading_hsv <- function(observed, residuals = NULL, expected = NULL, df = NULL,
 	        v, ...)
     dim(fill) <- dim(x)
     
-    col <- rep.int("black", length(res))
+    col <- rep(line_col, length.out = length(res))
     if(!is.null(eps)) {
       eps <- abs(eps)
       col[res > eps] <- hsv(my.h[1], 1, v, ...)
@@ -91,7 +91,7 @@ class(shading_hsv) <- "panel_generator"
 
 shading_hcl <- function(observed, residuals = NULL, expected = NULL, df = NULL,
   h = NULL, c = NULL, l = NULL,
-  interpolate = c(2, 4), lty = 1, eps = NULL, 
+  interpolate = c(2, 4), lty = 1, eps = NULL, line_col = "black",
   p.value = NULL, level = 0.95, ...)
 {
   ## set defaults
@@ -162,7 +162,7 @@ shading_hcl <- function(observed, residuals = NULL, expected = NULL, df = NULL,
 	        ...)
     dim(fill) <- dim(x)
     
-    col <- rep.int("black", length(res))
+    col <- rep(line_col, length.out = length(res))
     if(!is.null(eps)) {
       eps <- abs(eps)
       col[res > eps] <- hcl(my.h[1], max.c, my.l[2], ...)
@@ -182,16 +182,16 @@ shading_hcl <- function(observed, residuals = NULL, expected = NULL, df = NULL,
 class(shading_hcl) <- "panel_generator"
 
 shading_Friendly <- function(observed = NULL, residuals = NULL, expected = NULL, df = NULL,
-  h = c(2/3, 0), lty = 1:2, interpolate = c(2, 4), eps = 0.01, ...)
+  h = c(2/3, 0), lty = 1:2, interpolate = c(2, 4), eps = 0.01, line_col = "black", ...)
 {
   shading_hsv(observed = NULL, residuals = NULL, expected = NULL, df = NULL,
[…]

### 2005-08-04 trunk/vcd/R/sieve.R

@@ -1,10 +1,10 @@
 ###########################################################
-## mosaicplot
+## sieveplot
 
-mosaic <- function(x, ...)
-  UseMethod("mosaic")
+sieve <- function(x, ...)
+  UseMethod("sieve")
 
-mosaic.formula <-
+sieve.formula <-
 function(formula, data = NULL, ..., main = NULL, subset = NULL)
 {
   if (is.logical(main) && main)
@@ -34,7 +34,7 @@ function(formula, data = NULL, ..., main = NULL, subset = NULL)
       }
       dat <- margin.table(dat, ind)
     }
-    mosaic.default(dat, main = main,
+    sieve.default(dat, main = main,
                    condvars = if (is.null(condind)) NULL else match(condnames, names(dimnames(dat))), ...)
   } else {
     tab <- if ("Freq" %in% colnames(data))
@@ -44,16 +44,24 @@ function(formula, data = NULL, ..., main = NULL, subset = NULL)
       xtabs(formula(paste("~", paste(c(condnames, varnames), collapse = "+"))),
             data = data, subset = subset)
     
-    mosaic.default(tab, main = main, ...)
+    sieve.default(tab, main = main, ...)
   }
 }
 
-mosaic.default <- function(x, condvars = NULL,
-                           split_vertical = FALSE, direction = NULL,
-                           spacing = NULL, spacing_args = list(),
-                           zero_size = 0.5, main = NULL, ...) {
+sieve.default <- function(x, condvars = NULL,
+                          gp = NULL,
+                          shade = FALSE, legend = FALSE,
+                          split_vertical = FALSE, direction = NULL,
+                          spacing = NULL, spacing_args = list(),
+                          sievetype = c("observed","expected"), main = NULL, ...) {
   if (is.logical(main) && main)
     main <- deparse(substitute(x))
+  sievetype = match.arg(sievetype)
+  if (is.logical(shade) && shade && is.null(gp))
+    gp <- if (sievetype == "observed")
+      shading_Friendly(interpolate = 0, lty = c("longdash", "solid"))
+    else
+      shading_Friendly(interpolate = 0, line_col = "darkgray", eps = Inf, lty = "dotted")
 
   dl <- length(dim(x))
   if (!is.null(condvars)) {
@@ -78,23 +86,28 @@ mosaic.default <- function(x, condvars = NULL,
 
   strucplot(x,
             condvars = if (is.null(condvars)) NULL else length(condvars),
[…]

### 2005-08-04 trunk/vcd/demo/twoway.R

@@ -19,9 +19,11 @@
   fourfold(UCBAdmissions, extended = FALSE)
   fourfold(UCBAdmissions) ## extended plots
 
-  #tabplot(UCBAdmissions,
-  #        panel = function(x, ...) fourfold(x, panel = TRUE, ...)
-  #        )
+  ## using cotabplot
+  cotabplot(UCBAdmissions, panel = function(x, condlevels, ...)
+              fourfold(co_table(x, names(condlevels))[[paste(condlevels, collapse = ".")]],
+                       newpage = F, ...)
+            )
 
   ### Coal Miners Lung Data ###
   #############################
@@ -53,35 +55,39 @@
   ## aggregate over `sex':
   (tab <- margin.table(HairEyeColor, 1:2))
   ## plot expected values:
-  sieveplot(t(tab), type = "expected", values = "both")
+  sieve(t(tab), sievetype = "expected")
 
   ## plot sieve diagram:
-  sieveplot(t(tab))
+  sieve(t(tab))
 
   ### Visual Acuity ###
   #####################
   data(VisualAcuity)
-  sieveplot(Freq ~ right + left,
-                 data = VisualAcuity,
-                 subset = gender == "female",
-                 reverse_y = FALSE,
-                 main = "Unaided distant vision data",
-                 xlab = "Left Eye Grade",
-                 ylab = "Right Eye Grade")
+  attach(VisualAcuity)
+  sieve(Freq ~ right + left,
+        data = VisualAcuity,
+        subset = gender == "female",
+        main = "Unaided distant vision data")
+#        xlab = "Left Eye Grade",
+#        ylab = "Right Eye Grade")
+  detach(VisualAcuity)
   
   ### Berkeley Admission ###
   ##########################
 
-  ## -> Larger tables: Cross factors
+  ## -> Larger tables: e.g., Cross factors
   ### Cross Gender and Admission
   data(UCBAdmissions)
 
   (tab <- xtabs(Freq ~ Dept + I(Gender : Admit), data = UCBAdmissions))
-  sieveplot(tab, reverse_y = FALSE,
-                 xlab = "Gender:Admission",
-                 ylab = "Department",
-                 main = "Berkeley Admissions Data"
-                 )
+  sieve(tab, 
+#        xlab = "Gender:Admission",
+#        ylab = "Department",
+        main = "Berkeley Admissions Data"
+        )
+
+  ## or use extended sieve plots:
+  sieve(UCBAdmissions)
 
   ######################
   ## Association Plot ##

### 2005-08-04 trunk/vcd/R/cotabplot.R
sections 4 and 5 finished, only intro and abstract missing
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@435 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -176,7 +176,7 @@ class(cotab_sieve) <- "panel_generator"
 cotab_assoc <- function(x = NULL, condvars = NULL, ylim = NULL, ...) {
   if(!is.null(x)) {
     fm <- coindep_test(x, condvars, n = 1)
-    if(is.null(ylim)) ylim <- range(max(residuals(fm)))
+    if(is.null(ylim)) ylim <- range(residuals(fm))
   }
   
   function(x, condlevels) {

### 2005-08-04 trunk/vcd/paper/JCGS.Rnw
sections 4 and 5 finished, only intro and abstract missing
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@435 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
   set.seed(myseed)
-  ucb_cotab <- cotab_coindep(ucb, condvars = "Dept", type = "assoc", n = 5000)
+  ucb_cotab <- cotab_coindep(ucb, condvars = "Dept", type = "assoc", n = 5000, margins = c(3, 1, 1, 3))
 
@@
 \noindent \textbf{Key Words:} Association plots; Conditional inference; 
-Contingency tables; HCL colors; HSV colors; Mosaic plots; Permutation tests.
+Contingency tables; HCL colors; HSV colors; Mosaic plots.
 \end{abstract}
@@
 respectively.
-For convenience the number of observations is denoted $n = n_{++}$.
 Given an underlying distribution with theoretical cell probabilities
@@
 The expected cell frequencies in this model are
-$\hat n_{ij} = n_{i+}n_{+j}/n$. The probably best known and most used
+$\hat n_{ij} = n_{i+}n_{+j}/n_{++}$. The probably best known and most used
 measure of discrepancy between observed and expected values are the Pearson residuals
@@
 But the sum of squares is not the only plausible way of capturing deviations
-from $0$ in the Pearson residuals.
+from $0$ in the residuals.
 There are many other conceivable functionals $\lambda(\cdot)$ which
@@
 absolute value exceeds $c_\alpha$ violate the hypothesis of indendence at significance level
-$\alpha$ \cite[][Ch.~7]{vcd:Mazanec+Strasser:2000}. Thus, the interesting cells causing
+$\alpha$ \cite[][ch.~7]{vcd:Mazanec+Strasser:2000}. Thus, the interesting cells causing
 the dependence can easily be identified.
@@
 Consequently, all results in this paper are based on conditional inference performed
-by simulating from the permutation distribution of test statistics of type
+by simulating the permutation distribution of test statistics of type
 $\lambda(\{r_{ij}\})$.
@@
 instead of the the Pearson residuals $\{r_{ij}\}$. But as the ideas discussed here
-extend straightforwardly to that situation, we do not discuss it in detail.
+extend straightforwardly to that situation, we do not go into detail about this.
 
@@
 in a graphical way. The latter focuses on the visualization of the Pearson residuals
-$r_{ij}$ (under independence) while the latter primarily displays the observed frequencies
+$r_{ij}$ (under independence) while the former primarily displays the observed frequencies
 $n_{ij}$.
@@
 Colors are commonly used to enhance mosaic and association plots. To integrate
-a visualization of the Pearson residuals $\{r_{ij}\}$ into the mosaic display---which
+a visualization of the residuals $\{r_{ij}\}$ into the mosaic display---which
 in its `raw' version only visualizes the observed frequencies $\{n_{ij}\}$---\cite{vcd:Friendly:1994}
@@
 residuals ($|r_{ij}| < 2$) are shaded white. Cells with medium sized residuals
-($2 \le |r_{ij}| < 4$) are shaded in light blue and light red for positive negative
+($2 \le |r_{ij}| < 4$) are shaded light blue and light red for positive and negative
 residuals respectively. Cells with large residuals ($|r_{ij}| \ge 4$) are shaded
 with a fully saturated blue and red respectively. Mosaic plots enhanced by this 
-shading can thus also bring out departures from independence (or other log-liner 
+shading can thus also bring out departures from independence (or other log-linear 
 models in multi-way tables) graphically and visualize patterns of dependence.
@@
 is then coded by the saturation $s$ which is set to $0$, $0.5$ and $1$ respectively
-for small/medium/large residuals respectively. The value is always fixed at $v = 1$.
+for small/medium/large residuals. The value is always fixed at $v = 1$.
 This is also depicted in the upper panel of Figure~\ref{fig:shadingHSVHCL} which shows
-to slides from the HSV cylinder side to side. The plot shows the saturation/value plane
+two slides from the HSV cylinder side by side. The plot shows the saturation/value plane
 for the given hues $h = 0$ and $h = 2/3$. The significant color palette shows the
@@
 perceptual dimensions of the human visual system  \citep{vcd:Brewer:1999,vcd:Ihaka:2003}.
-Consequently, the the HSV dimensions are confounded, e.g., saturation
+Consequently, the HSV dimensions are confounded, e.g., saturation
 is not uniform across different hues. A fully saturated blue $(2/3, 1, 1)$ is
-much darker as a fully saturated red $(0, 1, 1)$ or green $(1/3, 1, 1)$.
+perceived to be much darker than a fully saturated red $(0, 1, 1)$ or green $(1/3, 1, 1)$.
 This makes it more difficult for the human eye to judge the size of shaded areas
@@
 the neutral color for small residuals in the Friendly shading, but typically
-grey is found to convey neutrality or uninterestingness much better than white does
+grey is found to convey neutrality or un-interestingness much better than white
 \citep{vcd:Brewer:1999}.
@@
 emissive color technologies such as computer displays. \cite{vcd:Ihaka:2003} discusses
-how CIELUV colors can be used for choosing colors for statistical graphics such as barplots.
+how CIELUV colors can be used for choosing qualitative palettes for statistical graphics such as barplots.
 By taking polar coordinates in CIELUV space, it is called HCL (Hue--Chroma--Luminance) space
@@
 of chroma and luminance. Such colors are always balanced towards the same grey 
-and thus do not have the problem of varying saturations as for the HSV colors.
-
-Here, we discuss how similar ideas can be used for deriving a diverging HCL palettes that 
+and thus do not have the problem of varying saturations as the HSV colors.
+
+Here, we discuss how similar ideas can be used for deriving diverging HCL palettes that 
 provide a suitable translation of the ideas from the Friendly shading to perceptually
@@
 values of chroma (that controls the colorfulness). Unfortunately, the HCL space is not completely
-regular as the HSV space is and consequently its dimensions cannot be standardized to 
-unit intervals: the hue ranges in $[0, 360]$ degrees and chroma and luminance both in $[0, 100]$
+regular as the HSV space is and consequently it cannot be standardized in the same way as the
+HSV space. Its dimensions are usually given by a
+hue ranging in $[0, 360]$ degrees and chroma and luminance ranging in $[0, 100]$
 percent. But not all combinations $(h, c, l)$ yield valid HCL colors and the admissable combinations
@@
 a diverging palette, this problem can easily be overcome as we just need two different
-hues (a `negative' and a `positive' hue) and hence choose two hues that correspond to 
+hues (a `negative' and a `positive' hue) and hence we can choose two hues that correspond to 
 similar shapes in the chroma/luminance plane. The lower panel of Figure~\ref{fig:shadingHSVHCL}
-shows two such planes side to side for the hues $h = 0$ and $h = 260$. To obtain a 
+shows two such planes side by side for the hues $h = 0$ and $h = 260$. To obtain a 
 sequence of colors with the same properties as the Friendly shading, the palette starts
@@
 the amount of grey, to code the absolute size of the quantity visualized, i.e., the
-Pearson residuals when applied to the independence problem. By changing the neutral
+residuals $r_{ij}$ when applied to the independence problem. By changing the neutral
 color or by changing the maximum chroma respectively, this can be changed to using only
@@
 of visualization.
-
-The advantages of using this HCL shading scheme instead of the HSV scheme is that
-it is based on a perceptually based color space that provides uniform saturations
-for different hues and is device inependent.
 
@@
 the presence (or absence) of colors in a plot also always conveys an
-impression of interestingness (or uninterstingness respectively). That is,
+impression of interestingness (or un-interstingness respectively). That is,
 statisticians might be tempted to interpret the absence of color in a plot
@@
 statistic $M$ from (\ref{eq:MaxAbs}) which always employs the critical values $c_\alpha$
-of $2$ and $4$ respectively. But the problem is that it is not clear to which
+$2$ and $4$. But the problem is that it is not clear to which
 significance levels $\alpha$ these critical values correspond because the distribution
@@
 
-The significance $\alpha = 0.1$ and $0.01$ are chosen because this leads to displays where
+The significance levels $\alpha = 0.1$ and $0.01$ are chosen because this leads to displays where
 fully colored cells are clearly significant (i.e., with a $p$ value below $0.01$),
@@
 to conceal the important cells and over-emphasizes the unimportant ones. Hence, we stick
-to a shading with fewer colors that are easier to interpret.
-
-This extended shading is already very flexible and combines visualization and inference.
+to a discrete shading with few colors that is easier to interpret.
+
+This maximum shading is already very flexible and combines visualization and inference.
 However, it can only be applied when employing the maximum statistic because it is the only
 aggregation functional $\lambda(\cdot)$ where a single large residual $|r_{ij}|$ exceeding
-its critical value (independent from the values of the remaining residuals) is equivalent
-to the a significant value of the whole test statistic $\lambda(\{r_{ij}\})$.\footnote{This
+its critical value is equivalent
+to a significant value of the whole test statistic $\lambda(\{r_{ij}\})$.\footnote{This
 can easily be seen from the following consideration: To obtain a level $\alpha$ test, where 
@@
 result, the equation $\alpha = P(\mbox{any } |r_{ij}| > c_\alpha)$ must hold. This is equivalent
-to $1 - \alpha = P(\mbox{all } |r_{ij}| \le c_alpha) = P(\max_{i,j} |r_{ij}| \le c_alpha)$.}
+to $1 - \alpha = P(\mbox{all } |r_{ij}| \le c_\alpha) = P(\max_{i,j} |r_{ij}| \le c_\alpha)$.}
 Typically, applying the maximum statistic is feasible and also appropriate for exploratory 
@@
 individual cells differently but can only be realized by using different colors for the
-whole table. A outlined before, colorfulness is often intuitively matched with interestingness,
+whole table. A outlined before, colorfulness is intuitively matched with interestingness,
 therefore a rather natural idea is to use the fully colored palette only when the corresponding
 test is significant and to use a less colorful palette if not. For the HCL scheme, the
-amount of color can easily be controlled by varying the maximum chroma value used. For the full
+amount of color can conveniently be controlled by varying the maximum chroma value used. For the full
 colors, the maximum chroma was set to $100$ as shown in Figure~\ref{fig:shadingHSVHCL} and 
@@
 @
-\caption{Mosaic plot for the Arthritis data with Friendly shading (left) and extended shading (right).}
+\caption{Mosaic plot for the Arthritis data with Friendly shading (left) and maximum shading (right).}
 \label{fig:shadings}
@@
 
-\section{Extensions to Conditional Independence}
-
-UCB admissions data \citep{vcd:Bickel+Hammel+O'Connell:1975}
-double max: \Sexpr{stat(ucb_doublemax)}, p-value = \Sexpr{pval(ucb_doublemax)}
-max Chisq: \Sexpr{stat(ucb_maxChisq)}, p-value = \Sexpr{pval(ucb_maxChisq)}
-sum Chisq: \Sexpr{stat(ucb_sumChisq)}, p-value = \Sexpr{pval(ucb_sumChisq)}
-simulated from $5000$ repetitions
-
-\cite{vcd:Friendly:1999} discusses grouping similar to coplots (conditioning
-plots) \citep{vcd:Cleveland:1993} that lead to trellis graphics
-\citep{vcd:Becker+Cleveland+Shyu:1996}.
-
-Independence problems do not only occur in 2-way tables, although that
-is an important special case, but they are also important in tables of
-higher dimensionality and can follow much more complex patterns.
-These are again defined based on the underlying table of theoretical cell probabilities
-$(\pi_{ij\dots})$ with more than two dimensions. Models of interest include the
-null hypotheses of conditional independence:
-  \[ \pi_{ijk\dots} = \pi_{i|k\dots}\pi_{j|k\dots} \]
-
-Classical non-graphical methods for these problems include the $\chi^2$
-test, Fisher's exact test, the Cochran-Mantel-Haenzel test (for $2 \times 2 \times
-K$-tables), and the analysis of log-linear models for more
-complex settings.
-
-As an example, two natural ways to use the visualization techniques described in the previous
-sections would be to use (Trellis-like) conditioning plots or pairs plots
-(like mosaic matrices) to visualize these more complex patterns of
-independence.
-
-Two ideas for the problem of conditional independence are briefly outlined
-here and illustrated using the famous admissions data of the University of
-California at Berkley (UCB) which is available in base \textsf{R}.
-In this data, the question whether there is gender
-discrimination at the UCB leads to the result that although women seem to
-be disadvantaged at the aggregated level there is no gender discrimination
-conditioned on the department---with the very exception of one department
-in which women are {\em more} likely to be admitted than would be plausible
-under independence. Exactly this is illustrated in the conditioning
-assocation plot in Figure \ref{fig:UCB-assoc-signif}.
-
-Similarly, the same data can be visualized using a mosaic matrix where
-a conditional independence model is fitted in each plot (see Figure
-\ref{fig:UCB-mosaic-signif}).
+\section{Extension to Conditional Independence}
+
+To introduce the new residual-based shadings without too much overhead in
+Section~\ref{sec:shadings}, we have only considered the independence problem
+in 2-way tables up to now. But, in fact, none of the ideas described are really
+confined to the 2-dimensional case as will be outlined in this section.
+
+Mosaic displays have been emphasized in the literature to be an excellent means
+of visualization for log-linear models \citep{vcd:Friendly:1999,vcd:Theus+Lauer:1999}.
+Problems of interest include complete, joint or conditional independence for all of 
+which the tables of expected values and residuals (again Pearson or deviance) can
+be computed. \citep{vcd:Friendly:1994,vcd:Friendly:1999} shows that his residual-based 
+shading scheme can directly be applied to these more complex independence models.
+For inference, the most commonly used aggregation functional for the residuals
+is again the sum of squares yielding the associated Pearson or likelihood
+ratio statistic respectively \citep{vcd:Agresti:2002}. Hence, our residual-based
+shading with different HCL palettes for significant and non-significant results
+can directly be applied to these models. For the maximum shading with
+data-driven cut offs, even more flexibility is possible than in the 2-way case
+when the structure of the independence problem is exploited which will be discussed 
+below for conditional independence
+
+Association plots are not commonly used for contingency tables with more than two 
+margins, although there is nothing in the definition that would prevent application
+in higher dimensions. However, as argued for the mosaic plots by \cite{vcd:Friendly:1999}
+and \cite{vcd:Theus+Lauer:1999}, it becomes increasingly
+important to choose a good layout as the number of variables grows.
+
+To show how the structure of the independence problem can be exploited for 
+choosing appropriate critical values in the maximum shading and for selecting 
+a suitable layout of mosaic or association plots, we consider the conditional
+independence problem. To keep the notational overhead low again, only the 3-way
+case is spelt out, but the same ideas extend fairly straightforwardly to conditional
+independence in higher dimensions. In a 3-way table $\{n_{ijk}\}$ with underlying
+theoretical probability distribution $\{p_{ijk}\}$ the conditional independence
+problem can be formulated as
+\begin{equation} \label{eq:H0ci}
+H_0: \; \pi_{ij|k} \quad = \quad \pi_{i+|k} \pi_{+j|k}.
+\end{equation}
+where $\pi_{ij|k}$ are the conditional probabilities given the stratum $k$
+with $k = 1, \dots, K$. Under the assumption of conditional independence,
+we can again compute expected frequencies $\{\hat n_{ijk}\}$ and the corresponding 
+residuals $\{r_{ijk}\}$. To test the conditional independence hypothesis, usually
+the sum-of-squares statistic is used
+\begin{equation} \label{eq:sumChisq}
+\sum_{i, j, k} r_{ijk}^2 \quad = \quad \sum_k X_k^2,
+\end{equation}
+which is simply the sum of the individual sum-of-squares statistics $X_k^2$ in each stratum $k$.
+Alternatively, a maximum statistic similar to that from Equation~(\ref{eq:MaxAbs})
+can be constructed
+\begin{equation} \label{eq:doublemax}
+\max_{i, j, k} |r_{ijk}| \quad = \quad \max_k M_k.
+\end{equation}
+As in the 2-way case, this allows for identification of the cells that are responsible
+for the deviation from conditional independence (if any). If it is not so much of interest
+in which \emph{cell} but only in which \emph{stratum} the deviation occures, then
+it would be natural to use 
+\begin{equation} \label{eq:maxChisq}
+\max_k \sum_{i,j} r_{ijk}^2 \quad = \quad \max_k X_k^2.
+\end{equation}
+Given a critical value for this statistic, all strata $k$ whose associated sum-of-squares
+statistics $X_k^2$ exceed the critical value are in conflict with the hypothesis of 
+conditional independence. All these statistics are of type
+\begin{equation} \label{eq:coindep_test}
+\lambda_\mathrm{agg} (\lambda_\mathrm{indep}(r_{ijk}))
+\end{equation}
+where $\lambda_\mathrm{indep}$ is a functional for assessing independence in stratum $k$
+and $\lambda_\mathrm{agg}$ is a functional for aggregating over the $k = 1, \dots, K$ strata.
+If the maximum is used for the latter, then identification of the strata causing the
+independence is possible. If additionally $\lambda_\mathrm{indep}$ is the maximum,
+the corresponding cells can also be identified. Hence, the double maximum
+statistic from~(\ref{eq:doublemax}) is the only functional allowing for detection of
+both the strata and the cells which caused the independence.
+
+But the main purpose of the formulation of the different test statistics is not so much 
+inference but their applicability to diagnostic plots via residual-based shadings. As stated
+at the beginning of this section, it is possible for all aggregation functionals to simply use either
+the significant or the non-significant shading for all cells in the contingency table---this
+strategy would have to be used for the sum-of-squares statistic from~(\ref{eq:sumChisq}). If
+$\lambda_\mathrm{agg}$ is the maximum as in~(\ref{eq:maxChisq}),
+then the significant palette would only be used
+in those strata in conflict with the hypothesis of conditional independence whereas the
+non-significant palette would be used for the remaining strata. Finally, if both $\lambda_\mathrm{agg}$
+and $\lambda_\mathrm{indep}$ are the maximum, then the same strategy as in Section~\ref{sec:shadings}
+can be pursued, i.e., only the significant palette is used but with data-driven cut offs
+derived from the distribution of the double maximum statistic from~(\ref{eq:doublemax}).
+
+To arrange the shaded rectangles of the association or the mosaic plot respectively,
+the most intuitive approach is to use the same conditioning in the display that
+was also used for conditioning in the model. For this situation, \cite{vcd:Friendly:1999}
+discusses a grouping similar to coplots \citep[conditioning plots, see]{vcd:Cleveland:1993}
+that lead to trellis graphics \citep{vcd:Becker+Cleveland+Shyu:1996}. Thus, a
+natural visualization of such an independence model would be a trellis-like coplot
+where each stratum $k$ could be visualized by an association or mosaic display.
+This has also the advantage that only the conditional independence problem but
+not the conditioning distribution over $k = 1, \dots, K$ is visualized which
+could obscure departures from conditional independence if the number of observations
+in each stratum $n_{++k}$ are very different.
+
+For illustration, the famous admissions data of the University of
+California at Berkley (UCB) from \cite{vcd:Bickel+Hammel+O'Connell:1975}
+are employed. Testing the conditional independence of admission and gender
+given the department leads to the well-known result that there is no 
+gender discrimination in five of six departments, but that women are more
+likely to be admitted than men in the first department. This departure
+from conditional independence is picked up by all three statistics formulated
+above: the sum-of-squares statistic from~(\ref{eq:sumChisq}) is
+$\sum_k X_k^2 = \Sexpr{stat(ucb_sumChisq)}$ ($p = \Sexpr{pval(ucb_sumChisq)}$),
+the maximum sum-of-squares statistic from~(\ref{eq:maxChisq}) is
+$\max_k X_k^2 = \Sexpr{stat(ucb_maxChisq)}$ ($p = \Sexpr{pval(ucb_maxChisq)}$),
+and the double maximum statistic from~(\ref{eq:doublemax}) is
+$\max_k M_k = \Sexpr{stat(ucb_doublemax)}$ ($p = \Sexpr{pval(ucb_doublemax)}$).
+The $p$ values are again computed by drawing $5000$ samples from the corresponding
+permutation distribution. The latter two test statistics both indicate the 
+first department as the source of departure and the maximum statistic additionally
+identifies the deviations in the female applicants group. This result
+is also visualized by the conditional association plot in a trellis layout based
+on the double maximum shading in Figure~\ref{fig:ucb}. This can be
+interpreted as a diagnostic residual plot for the associated log-linear model
+because it shows that there are significant departures from the model in department
+A but not in departments B to F where all residuals are very small.
 
@@
 @
-\caption{Conditional association plot for UCB admissions.}
-\label{fig:emp-mosaic1}
+\caption{Conditional association plot for UCB admissions with double maximum shading.}
+\label{fig:ucb}
 \end{center}
@@
 
-We have presented two approaches for combining visualization and significance testing
-via residual-based shadings. The first one alwyas uses the fully colored palette but relies on
+Various strategies for constructing residual-based shadings for visualizing
+(conditional) independence in contingency tables via mosaic and association plots
+are discussed. The shading
+of \cite{vcd:Friendly:1994} is extended in two directions: the use of perceptually
+uniform HCL colors and the combination of visualization and significance testing.
+To achieve the former, a general guideline for constructing diverging palettes
+in HCL space is introduced. The advantages of using this HCL shading scheme
+instead of an HSV scheme are that the colors from this perceptually based color space
+are device inependent and provide uniform saturations over different hues.
+Furthermore, the colorfulness in this shading can be controlled independently from
+the other two dimensions.
+
+To combine visualization and significance testing, two approaches are presented:
+The first approach always uses a fully colored palette but relies on
 data-driven cut offs such that the presence of color is equivalent to significance of the
@@
 
+Both results can not only be applied to the simple independence problem in 2-way
+contingency tables, but also to arbitrary inpendence models fitted via log-linear models
+in higher dimensions. In addition, it might be possible to exploit the structure of a given 
+independence problem to achieve better visualizations which is illustrated
+for the conditional independence problem.
+
+All associated significance tests are carried out using a conditional inference approach
+instead of relying on unconditional asymptotic results.
+
+\section*{Computational Details}
+
+The results in this paper were obtained using
+\proglang{R}~\Sexpr{paste(R.Version()[6:7], collapse = ".")} \citep{vcd:R:2005}
+and the packages
+\pkg{vcd}~\Sexpr{packageDescription("vcd")$Version} \citep{vcd:Meyer+Zeileis+Hornik:2005},
+{\pkg{MASS}~\Sexpr{packageDescription("MASS")$Version}} \citep{vcd:Venables+Ripley:2002},
+{\pkg{grid}~{\Sexpr{packageDescription("grid")$Version}}} \citep{vcd:Murrell:2002} and
+{\pkg{colorspace}~{\Sexpr{packageDescription("colorspace")$Version}}} \citep{vcd:Ihaka:2004}.
+
 \bibliography{vcd}
 
-\end{document}
-
-%% for reproducibility, print also
-
-<<graphics>>
+\newpage
+
+\begin{appendix}
+
+\section{Replication of Results in R}
+
+\begin{verbatim}
+## random seed
+rseed <- 1071
+
 ## load packages and data
@@
 
-## plain classic plots
+## Figure 1
 mosaic(art)
@@
 
-## new max shading with HCL colors
+## Figure 3
 mosaic(art, gp = shading_Friendly(lty = 1, eps = NULL))
-set.seed(1071)
+set.seed(rseed)
 mosaic(art, gp = shading_max)
 
-## conditional independence
-set.seed(1071)
+## Figure 4
+set.seed(rseed)
 cotabplot(~ Admit + Gender | Dept, data = ucb, panel = cotab_coindep, type = "assoc")
-## which uses the double max functional
-set.seed(1071)
-coindep_test(ucb, "Dept")
-
-## alternatively use Peason functional
-## via simulation of conditional distribution:
-set.seed(1071)
-coindep_test(ucb, "Dept", indepfun = function(x) sum(x^2), aggfun = sum)
-## via asymptotic distribution:
-fm <- loglm(~ Dept * (Gender + Admit), data = ucb)
-fm
-plot(fm, labeling_args = list(abbreviate = c(Admit = 2)))
-@
+
+## Inference
+## X^2 (3)
+set.seed(rseed)
+coindep_test(art, n = 5000)
+## M (4)
+set.seed(rseed)
+coindep_test(art, n = 5000, indepfun = function(x) sum(x^2))
+
+## sum X^2 (6)
+set.seed(rseed)
+coindep_test(ucb, "Dept", n = 5000, indepfun = function(x) sum(x^2), aggfun = sum)
+## max M (7)
+set.seed(rseed)
+coindep_test(ucb, "Dept", n = 5000)
+## max X^2 (8)
+set.seed(rseed)
+coindep_test(ucb, "Dept", n = 5000, indepfun = function(x) sum(x^2))
+\end{verbatim}
+
+\end{appendix}
+
+\end{document}
 

### 2005-08-05 trunk/vcd/R/coindep_test.R
added docs for cotabplot
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@436 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -1,7 +1,7 @@
 coindep_test <- function(x, margin = NULL, n = 1000,
   indepfun = function(x) max(abs(x)), aggfun = max,
   alternative = c("greater", "less"),
-  pearson = TRUE, return_distribution = TRUE)
+  pearson = TRUE)
 {
   DNAME <- deparse(substitute(x))
   alternative <- match.arg(alternative)

### 2005-08-07 trunk/vcd/inst/doc/shading.Rnw
New shading vignette added.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@437 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\documentclass[a4paper]{article}
\usepackage{hyperref, graphicx, color, alltt, a4wide}
\usepackage{Sweave}
\usepackage[round]{natbib}
\definecolor{Red}{rgb}{0.7,0,0}
\definecolor{Blue}{rgb}{0,0,0.8}
\definecolor{hellgrau}{rgb}{0.55,0.55,0.55}
\newcommand{\pkg}[1]{{\normalfont\fontseries{b}\selectfont #1}}
\newcommand{\code}[1]{\texttt{#1}}
\newenvironment{smallexample}{\begin{alltt}}{\end{alltt}}

\begin{document}

%\VignetteIndexEntry{Colors and residual-based Shadings in Strucplots}
%\VignetteDepends{vcd}
%\VignetteKeywords{graphics, categorical data}
%\VignettePackage{vcd}

\SweaveOpts{engine=R,eps=TRUE}
\setkeys{Gin}{width=0.8\textwidth}

<<preliminaries,echo=FALSE,results=hide>>=
set.seed(1071)
library(vcd)
data(Titanic)
@

\title{Colors and residual-based Shadings in Strucplots}
\author{by David Meyer\\
  Wirtschaftsuniversit\"at Wien, Austria\\
\url{David.Meyer@R-project.org}
}
\maketitle
\sloppy

Unlike other R graphics functions, strucplots allow almost full
control over the graphical parameters of all plot elements. In
particular, the user can \emph{independently} modify the appearance of the 
tiles of association plots and mosaic plots. Built on top of this functionality, the
framework supplies a set of shading functions that choose colors
appropriate to the visualization of independence models fit to the
visualized data.
The tiles' graphical parameters are set using the \code{gp} argument
of the strucplot functions. This argument basically expects an object
of class \code{gpar} whose components have the same shape (length and dimensionality) 
than the data table (see Section \ref{sec:gp}). For convenience, however, the user can also
supply a function that computes such an object given a vector of
residuals, or, alternatively, a generating function that takes
parameters and returns such a function (see Section \ref{sec:shading}).

\section{Specifying graphical parameters in strucplots}
\label{sec:gp}

As an example, consider the Admissions to the University of California
in Berkely (UCB) data. In the table collapsed over departments, we
would like to highlight the (incidentally wrong) impression that there
were to much male students accepted compared to the presumably
discriminated female students (see Figure \ref{fig:ucb}):

<<ucb,fig=FALSE,echo=TRUE,height=6,width=7,results=verbatim>>=
ucb <- margin.table(UCBAdmissions, 1:2)
ucb

fill_colors <- matrix(c("dark cyan","gray","gray","dark magenta"), ncol = 2)
fill_colors

mosaic(ucb, gp = gpar(fill = fill_colors, col = 0))
@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<ucbfig,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(ucb, gp = gpar(fill = fill_colors, col = 0))
@
\caption{Mosaic plot for the UCBAdmissions data with highlighted cells.}
\label{fig:ucb}
\end{center}
\end{figure}

\noindent As the example shows, we therefore create a fourfold table with appropriate colors and supply them to the \code{fill} component of the \code{gpar} object passed to the \code{gp} argument of \code{mosaic()}. In addition, we use the \code{col} component to turn off the borders for cosmetic reasons.

If the parameters specified in the \code{gpar} object are incomplete,
strucplot functions will recycle them along the last splitting
dimension. In the following example based on the Survival on the
Titanic data, we will highlight all cells corresponding to survived
passengers (see Figure \ref{fig:recycling}):

\begin{Sinput}
> mosaic(Titanic, gp = gpar(fill = c("gray","dark magenta")))
\end{Sinput}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<recycling,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(Titanic, gp = gpar(fill = c("gray","dark magenta")))
@
\caption{Recycling of parameters.}
\label{fig:recycling}
\end{center}
\end{figure}

\section{Using residual-based shadings}
\label{sec:shading}

This new flexible way of specifying graphical parameters is the basis for a suite of
shading functions that modify the tiles' appearance with respect to a vector of 
residuals, resulting from deviations of the observed from the expected
values under a given independence model. 
The idea is to visualize at least the size of the residuals, but some
shadings, additionally, indicate overall significance. One particular
shading, the maximum shading, even allows to identify those cells that
cause the rejection of the independence hypothesis.

All shading functions are parameterized be the means of generating
functions that return the actual shading functions. As an example,
consider the \code{shading\_binary()} function that only visualizes the
sign of the residuals:

\begin{Sinput}
shading_binary <- function(observed = NULL, residuals = NULL, expected = NULL, 
                           df = NULL, col = hcl(c(260, 0), 50, 70))
{
\dots
}
\end{Sinput}

\noindent Each generating function, passed to a strucplot functions' \code{gp}
argument, takes a few arguments that describe the model
(\code{observed}, \code{expected}, \code{residuals}, and \code{df}), and a
number of parameters that specify the appearance (in the case of
\code{shading\_binary()} only \code{col}, a vector of length two, 
that specifies the two colors corresponding to negative and 
positive residuals, respectively). The
generating function then return the actual shading function taking 
only one argument: a vector of residuals, and returning an object
of class \code{gpar} that contains the graphical parameters for all
tiles corresponding to the residuals. In the following example, positive 
and negative residuals of an hypothesized independence model shall be 
visualized by different colors (see Figure \ref{fig:binary}):

\begin{Sinput}
> obs <- margin.table(UCBAdmissions, 1:2)
> exp <- independence_table(obs)
> res <- (obs - exp) / sqrt(exp)
> res
          Gender
Admit           Male    Female
  Admitted  4.784093 -5.793466
  Rejected -3.807325  4.610614

> gpobj <- shading_binary()(res)
> gpobj
$fill
     [,1]      [,2]     
[1,] "#96A2DF" "#E18E9E"
[2,] "#E18E9E" "#96A2DF"

> mosaic(obs, gp = gpobj)
\end{Sinput}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<binary,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
obs <- margin.table(UCBAdmissions, 1:2)
exp <- independence_table(obs)
res <- (obs - exp) / sqrt(exp)
gpobj <- shading_binary()(res)
mosaic(obs, gp = gpobj)
@
\caption{Binary shading.}
\label{fig:binary}
\end{center}
\end{figure}

\noindent The shading-generating functions available
in \pkg{vcd} include model-fitting, and hence the {\sf R} code
corresponding to the example above in fact reduces to:

\begin{Sinput}
> mosaic(obs, gp = shading_binary)
\end{Sinput}

\section{An overview of the shading functions in \pkg{vcd}}

Currently, there are three basic shadings available
(\code{shading\_hcl(), shading\_hsv(), and shading\_binary()}, as well
as two derived functions (\code{shading\_Friendly()} building upon
\code{shading\_hsv()} and \code{shading\_max()} building upon \code{shading\_hcl}).
\code{shading\_binary()} has been explained in the previous
paragraph. \code{shading\_hsv()} and \code{shading\_hcl()} are quite
similar in their functionality, but use different color spaces: the
Hue-Saturation-Value (HSV) and the Hue-Chroma-Luminance (HCL) scheme,
respectively. We will focus on the latter, since the HCL space is
preferable to the HSV space in many respects (e.g., uniform saturations over
different hues, device independentness, \dots).

In the HCL space, colors are specified in three dimensions: hue, chroma (`colorfulness', saturation), and luminance (`brightness', amount of gray). These three dimensions are used by \code{shading\_hcl()} to visualize information about the residuals and the underlying independence model. A residuals' hue indicates its sign (by default blue and red for positive and negative ones, respectively). An overall full or poor chroma indicates significant or non-significant results, respectively. Finally, the luminance of a residual is set according to its size.

As an example, we will visualize the (in)dependence of hair and eye
color in the HairEyeColor data set (see Figure \ref{fig:haireye})

\begin{Sinput}
> haireye <- margin.table(HairEyeColor, 1:2)
> mosaic(haireye, shade = TRUE)
\end{Sinput}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<haireye,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
haireye <- margin.table(HairEyeColor, 1:2)
mosaic(haireye, shade = TRUE)
@
\caption{Shaded residuals in the HairEyeColor data set---two cut-off points}
\label{fig:haireye}
\end{center}
\end{figure}

\noindent The default shading scheme is \code{shading\_hcl()}, so it
suffices to turn shading on to bring colors to the mosaic plot. Large
positive residuals (greater than $4$) can be found for brown eyes/black hair and blue
eyes/blond hair, they are colored in saturated red. 
On the other side, there is a large negative residual
(less than $-4$) for brown eyes/blond hair, colored deep blue. There are also three
medium-sized positive (negative) residuals between 2 and 4 ($-2$ and
$-4$): the colors for them are less saturated. 
The heuristic for choosing the cut-off points $2$ and $4$ is that the Pearson residuals
are approximately standard normal which implies that the highlighted cells are those with
residuals \emph{individually} significant at approximately the $\alpha
= 0.05$ and $\alpha = 0.0001$ level. These default cut-off points can
be changed to alternative values using the \code{interpolate}
argument (see Figure \ref{fig:interpolate}):

\begin{Sinput}
> mosaic(haireye, shade = TRUE, 
         gp_args = list(interpolate = 1:4, h = c(100, 200)))
\end{Sinput}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<interpolate,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(haireye, shade = TRUE, 
       gp_args = list(interpolate = 1:4, h = c(100, 200)))
@
\caption{Shaded residuals in the HairEyeColor data set---four cut-off points}
\label{fig:interpolate}
\end{center}
\end{figure}

\noindent Note that we also changed the hue values for purely cosmetic
reasons. The third remaining dimension, the luminance, is used for
visualizing the significance of a test statistic. The $p$ value can
be supplied to the \code{p.value} argument. If none is specified, it
is computed from a Chi-squared distribution with \code{df} degrees of
freedom. The \code{level} argument is used to specify the confidence
level:  if \code{p.value} is smaller than \code{1 - level}, bright colors are used, 
otherwise dark colors are employed. These colors are computed using
the two luminance values supplied to the \code{l} argument. The
following example shows the relationship of home goals and away
goals of the German Bundesliga in 1995: although there are two
``larger'' residuals (one greater than 2, one less then $-2$), the
Chi-squared test does not reject the null hypothesis of
independence. Consequently, the colors appear dark (see Figure \ref{fig:bundesliga}):

\begin{Sinput}
> bl <- xtabs(~ HomeGoals + AwayGoals, data = Bundesliga, subset = Year == 1995)
> mosaic(bl, shade = TRUE)
\end{Sinput}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<bundesliga,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
bl <- xtabs(~ HomeGoals + AwayGoals, data = Bundesliga, subset = Year == 1995)
mosaic(bl, shade = TRUE)
@
\caption{Non-significant Chi-squared test using part of the Bundesliga data.}
\label{fig:bundesliga}
\end{center}
\end{figure}

\noindent The \code{interpolate} argument also accepts a function that
automatically chooses the cut-off points in a data-driven way. 
This functionality is used by \code{shading\_max()} that uses the
maximum statistic to visualize significant \emph{cells} causing the
rejection of the independence hypothesis. The
\code{level} argument of \code{shading\_max()} 
then can be used to specify several confidence
levels from which the corresponding cut-off points are computed. 
By default, two cut-off points are computed corresponding 
to confidence levels of $90\%$ and $99\%$, respectively.
In the following example, we investigate the effect of a new treatment
for rheumatoid arthritis on a group of female patients using the
maximum shading (see Figure \ref{fig:maximum}):

\begin{Sinput}
> art <- xtabs(~ Improved + Treatment, data = Arthritis, subset = Sex == "Female")
> mosaic(art, shade = TRUE, gp = shading_max)
\end{Sinput}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<arthritis,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
art <- xtabs(~ Improved + Treatment, data = Arthritis, subset = Sex == "Female")
mosaic(art, gp = shading_max)
@
\caption{Significant maximum test on female patients of the Arthritis
  data.}
\label{fig:maximum}
\end{center}
\end{figure}

\noindent The maximum test is significant although the residuals are
all in the $[-2,2]$ interval, as the cut-off points are even smaller.
The \code{shading\_hcl} function with default cut-off points would
not have shown any color. In addition, since the test statistic is the
maximum of the absolute pearson residuals, \emph{every} colored
residual violates the hypotheses of independence, and thus, the
`culprits' can immediately be identified.

The last shading function is \code{shading\_Friendly()}. This shading
mimics the shading introduced by \cite{vcd:Friendly:2000}. This
shading corresponds to the HSV shading with default cut-off points (2 and 4), and in
addition uses the border color and line type to visualize the
residuals' sign. The following example again visualizes the Bundesliga
data from above, this time using the Friendly scheme and, in addition, an
alternative legend:

\begin{Sinput}
> mosaic(bl, gp = shading_Friendly, legend = legend_fixed, zero_size = 0)
\end{Sinput}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<friendly,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(bl, gp = shading_Friendly, legend = legend_fixed, zero_size = 0)
@
\caption{The Bundesliga data for 1995 using the Friendly shading.}
\label{fig:friendly}
\end{center}
\end{figure}

\noindent (The bullets indicating zero observed values are removed here
since this feature does not exist in the original SAS implementation
of the Friendly mosaic plots).

\section{Creating new shading functions}

The example with four cut-off points (Figure \ref{fig:interpolate})
might suggest that a continuous shading simply visualizing the complete
range of confidence levels might be useful. The following code shows
how such a shading function might be constructed:

\begin{Sinput}
shading_continuous <- function(observed = NULL, residuals = NULL,
                               expected = NULL, df = NULL, n = 5000, ...)
{
  ## function works only for two-way tables
  stopifnot(length(dim(observed)) == 2)

  ## sample permutation distribution of the maximum statistic
  obs.test <- coindep_test(observed, n = n)

  ## Use distribution to map residuals to [0,1] range
  ipol <- function(x) abs(obs.test$pdist(x))

  ## call HCL shading 
  shading_hcl(observed = NULL, residuals = NULL, expected = NULL,
              df = NULL, interpolate = ipol, p.value = obs.test$p.value, ...)
}
class(shading_continuous) <- "panel_generator"

mosaic(haireye, gp = shading_continuous)
\end{Sinput}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<continuous,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
shading_continuous <- function(observed = NULL, residuals = NULL,
                               expected = NULL, df = NULL, n = 5000, ...)
{
  ## function works only for two-way tables
  stopifnot(length(dim(observed)) == 2)

  ## sample permutation distribution of the maximum statistic
  obs.test <- coindep_test(observed, n = n)

  ## Use distribution to map residuals to [0,1] range
  ipol <- function(x) abs(obs.test$pdist(x))

  ## call HCL shading 
  shading_hcl(observed = NULL, residuals = NULL, expected = NULL,
              df = NULL, interpolate = ipol, p.value = obs.test$p.value, ...)
}
class(shading_continuous) <- "panel_generator"

mosaic(haireye, gp = shading_continuous)
@
\caption{The HairEyeColor data with continuous shading.}
\label{fig:continuous}
\end{center}
\end{figure}

\noindent We first compute the permutation distribution for the
maximum statistic using \code{coindep\_test}, and then simply use it
as the interpolation function passed to \code{shading\_hcl}. The result
using the HairEyeColor data, however, is disappointing: too much color
makes it difficult to interpret the image, and the subtle color
differences are hard to catch. Therefore, we only included shadings
with discrete cut-off points.

\begin{thebibliography}{}

\bibitem[Friendly, 2000]{vcd:Friendly:2000}
Friendly, M. (2000).
\newblock {\em Visualizing Categorical Data}.
\newblock SAS Insitute, Carey, NC.

\end{thebibliography}

\end{document}

### 2005-08-07 trunk/vcd/R/shadings.R

@@ -48,6 +48,7 @@ shading_hsv <- function(observed, residuals = NULL, expected = NULL, df = NULL,
   }
 
   ## store color and lty information for legend
+  legend <- NULL
   if(!is.null(col.bins)) {
     res2 <- col.bins
     res2 <- c(head(res2, 1) - 1, res2[-1] - diff(res2)/2, tail(res2, 1) + 1)
@@ -139,6 +140,7 @@ shading_hcl <- function(observed, residuals = NULL, expected = NULL, df = NULL,
   }
 
   ## store color and lty information for legend
+  legend <- NULL
   if(!is.null(col.bins)) {
     res2 <- col.bins
     res2 <- c(head(res2, 1) - 1, res2[-1] - diff(res2)/2, tail(res2, 1) + 1)

### 2005-08-10 trunk/vcd/paper/JCGS.Rnw
added docs for cotabplot
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@436 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
added *easy* comments/modifications of KH and DM
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@438 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 <<load_data,echo=FALSE,results=hide>>=
-library(vcd)
-
-data(Arthritis)
+library("vcd")
+
+data("Arthritis")
 art <- xtabs(~ Treatment + Improved, data = Arthritis, subset = Sex == "Female")
 
-data(UCBAdmissions)
+data("UCBAdmissions")
 ucb <- aperm(UCBAdmissions)
@@
 To fix notations, we consider a 2-way contingency table with cell
-frequencies $\{n_{ij}\}$ for $i = 1, \dots, I$ and $j = 1, \dots, J$
-and row and column sums $n_{i+} = \sum_i n_{ij}$ and $n_{+j} = \sum_j n_{ij}$
+frequencies $[n_{ij}]$ for $i = 1, \dots, I$ and $j = 1, \dots, J$
+and row and column sums $n_{i+} = \sum_i n_{ij}$ and $n_{+j} = \sum_j n_{ij}$,
 respectively.
@@
 with rheumatoid arthritis.}
+
 \begin{tabular}{rr|rrr}
+\\
                    &         & \multicolumn{3}{|l}{\textbf{Improved}} \\
@@
 \end{tabular}
+
 \end{center}
@@
 There are many other conceivable functionals $\lambda(\cdot)$ which
-lead to reasonable test statistics $\lambda(\{r_{ij}\})$: one which is particularly
+lead to reasonable test statistics $\lambda([r_{ij}])$: one which is particularly
 suitable for identifying the cells which cause the dependence (if any) is
@@
 by simulating the permutation distribution of test statistics of type
-$\lambda(\{r_{ij}\})$.
+$\lambda([r_{ij}])$.
 
 Other measures of discrepancy (e.g., deviance residuals) could, of course, also be used
-instead of the the Pearson residuals $\{r_{ij}\}$. But as the ideas discussed here
+instead of the the Pearson residuals $[r_{ij}]$. But as the ideas discussed here
 extend straightforwardly to that situation, we do not go into detail about this.
@@
 is $X^2 = \Sexpr{stat(art_chisq)}$ with a $p$ value of $p = \Sexpr{pval(art_chisq)}$,
-and the maximum statistion from (\ref{eq:MaxAbs}) is $M = \Sexpr{stat(art_max)}$
-with a $p$ value of $p = \Sexpr{pval(art_max)}$. Both $p$ values have been computed
+and the maximum statistic from (\ref{eq:MaxAbs}) is $M = \Sexpr{stat(art_max)}$
+with $p = \Sexpr{pval(art_max)}$. Both $p$ values have been computed
 from a sample of size $5000$ from the permutation distribution under independence
@@
 mosaic plots and association plots. Both are suitable to bring out
-departures of an observed table $\{n_{ij}\}$ from the expected table $\{\hat n_{ij}\}$
+departures of an observed table $[n_{ij}]$ from the expected table $[\hat n_{ij}]$
 in a graphical way. The latter focuses on the visualization of the Pearson residuals
@@
 frequencies of the two variables: a mosaic plot simply
-consists of a collection of tiles whose
-sizes are proportional to the observed cell frequencies as shown
+consists of a collection of tiles with areas
+proportional to the observed cell frequencies as shown
 in the left panel of Figure~\ref{fig:classic}. A rectangle corresponding
@@
 @
-\caption{Classic mosaic and association plot for the Arthritis data.}
+\caption{Classic mosaic and association plot for the arthritis data.}
 \label{fig:classic}
@@
 Colors are commonly used to enhance mosaic and association plots. To integrate
-a visualization of the residuals $\{r_{ij}\}$ into the mosaic display---which
-in its `raw' version only visualizes the observed frequencies $\{n_{ij}\}$---\cite{vcd:Friendly:1994}
+a visualization of the residuals $[r_{ij}]$ into the mosaic display---which
+in its `raw' version only visualizes the observed frequencies $[n_{ij}]$---\cite{vcd:Friendly:1994}
 suggested a residual-based shading for the mosaic tiles that can also be applied
@@
 
-\subsection{Friendly Shading}
+\subsection{Friendly Shading} \label{sec:Friendly}
 
 The extensions of \cite{vcd:Friendly:1994} to mosaic plots provide a
-substantial improvement of the original mosaic plots and enhanced them
+substantial improvement of the original mosaic plots enhancing them
 from a plot for contingency tables to a visualization technique for
 log-linear models---and thus also for independence problems including
-2-way tables as the simplest special case.
+2-way tables as the simplest case.
 
@@
 ($2 \le |r_{ij}| < 4$) are shaded light blue and light red for positive and negative
-residuals respectively. Cells with large residuals ($|r_{ij}| \ge 4$) are shaded
-with a fully saturated blue and red respectively. Mosaic plots enhanced by this 
+residuals, respectively. Cells with large residuals ($|r_{ij}| \ge 4$) are shaded
+with a fully saturated blue and red, respectively. Mosaic plots enhanced by this 
 shading can thus also bring out departures from independence (or other log-linear 
@@
 residuals \emph{individually} significant at approximately the $\alpha = 0.05$ and $\alpha = 0.0001$
-level. But the main purpose of the shading is not to visualize significance but the \emph{pattern} of
+levels. However, the main purpose of the shading is not to visualize significance but the \emph{pattern} of
 deviation from independence \citep[p.~109]{vcd:Friendly:2000}.
@@
 
-\subsection{Colors}
+\subsection{Colors} \label{sec:colors}
 
@@
 uses colors from a palette based on HLS (Hue--Luminance--Saturation) color space.
-The implementation in the \proglang{R} system for statistical computing
-\citep[\url{http://www.R-project.org/}]{vcd:R:2005} employs colors from
+The implementation in the \proglang{R} system for statistical computing and graphics
+\citep{vcd:R:2005} employs colors from
 HSV (Hue--Saturation--Value) space. The latter is a very common implementation of
@@
 
-The HSV space originally looks like a regular cone with black at its
+The HSV space originally looks like a cone with black at its
 peak (zero value) and full color wheels for different saturations at the other end,
 centered around a white (full value). Type and amount of color are controlled by hue
-and saturation respectively. Typically, polar coordinates that all range in the 
-unit interval are used in this space, such that it looks like a regular cylinder.
+and saturation, respectively. Typically, polar coordinates $(h, s, v)$ rescaled to the 
+unit interval are used in this space, giving it the appearance of a cylinder.
 For generating colors in the Friendly shading, the following strategy is used:
@@
 $h = 2/3$ (blue hue) for positive residuals. The absolute size of the residuals
-is then coded by the saturation $s$ which is set to $0$, $0.5$ and $1$ respectively
+is then coded by the saturation $s$ which is set to $0$, $0.5$ and $1$, respectively,
 for small/medium/large residuals. The value is always fixed at $v = 1$.
@@
 colors used for the Friendly shading when the residuals are increasing from left to right.
-The set of non-significant colors will be explained in the next section.
+The set of non-significant colors will be explained in Section~\ref{sec:significance}.
 
 Although this HSV-based shading is already very useful for enhancing mosaic and
-association plots and although HSV is very commonly available implementation of
+association plots and although HSV is a very commonly available implementation of
 color spaces, HSV colors in general and the Friendly shading in particular have
@@
 uniform HCL colors. 
-The HCL space looks like a double cone with black (zero luminance) at one end and white
+The HCL space looks like a distorted double cone with black (zero luminance) at one end and white
 at the other (full luminance). In its middle, there is a full color wheel for different
-values of chroma (that controls the colorfulness). Unfortunately, the HCL space is not completely
-regular as the HSV space is and consequently it cannot be standardized in the same way as the
+values of chroma (that controls the colorfulness). Unfortunately, the HCL space is not as
+regular as the HSV space and consequently cannot be standardized in the same way as the
 HSV space. Its dimensions are usually given by a
@@
 
-This diverging palette (see the `significant' colors in Figure~\ref{fig:shadingHSVHCL},
-the `non-significant' colors are again discussed in the following section)
-uses both chroma, i.e., the colorfulness, and luminance, i.e., 
-the amount of grey, to code the absolute size of the quantity visualized, i.e., the
-residuals $r_{ij}$ when applied to the independence problem. By changing the neutral
-color or by changing the maximum chroma respectively, this can be changed to using only
+This diverging palette (see the `significant' colors in Figure~\ref{fig:shadingHSVHCL})
+uses both chroma---i.e., the colorfulness---and luminance---i.e., 
+the amount of grey---to code the absolute size of the quantity visualized---i.e., the
+residuals $r_{ij}$---when applied to the independence problem. By changing the neutral
+color or by changing the maximum chroma, respectively, this can be changed to using only
 chroma or luminance for this purpose, but using both seems to be a very effective way
@@
 
-\subsection{Significance}
+\subsection{Significance} \label{sec:significance}
 
@@
 the presence (or absence) of colors in a plot also always conveys an
-impression of interestingness (or un-interstingness respectively). That is,
-statisticians might be tempted to interpret the absence of color in a plot
+impression of interestingness (or un-interstingness, respectively). That is,
+viewers might be tempted to interpret the absence of color in a plot
 as a clue that there is no significant departure from independence. Or vice
@@
 in the plot is equivalent to significance (of the maximum statistic $M$) at 
-level $\alpha = 0.1$ and $\alpha = 0.01$ respectively and exactly the cells which `cause'
+level $\alpha = 0.1$ and $\alpha = 0.01$, respectively, and exactly the cells which `cause'
 the dependence are highlighted. For the arthritis data, these are in particular the 
@@
 
-The significance levels $\alpha = 0.1$ and $0.01$ are chosen because this leads to displays where
-fully colored cells are clearly significant (i.e., with a $p$ value below $0.01$),
-cells without color are clearly non-significant (i.e., with a $p$ value above $0.1$), and
-cells in between can be considered to be weakly significant (i.e., $0.1 \le p < 0.01$).
+The significance levels $\alpha = 0.1$ and $\alpha = 0.01$ are chosen because this
+leads to displays where fully colored cells are clearly significant ($p < 0.01$),
+cells without color are clearly non-significant ($p > 0.1$), and
+cells in between can be considered to be weakly significant ($0.01 \le p \le 0.1$).
 Of course, users could choose any other set of significance levels they feel comfortable
-with, e.g., only a single cutoff at $\alpha = 0.05$ or three cut offs at $0.1$, $0.05$ and $0.01$
+with, e.g., only a single cut off at $\alpha = 0.05$ or three cut offs at $0.1$, $0.05$ and $0.01$
 etc. Another option could be to use a continuous shading where the $p$ value corresponding
-to a cell controls the interpolation between the neutral and the full color. However, our
-experience is that this typically results in too much color in a plot which in turn tends
-to conceal the important cells and over-emphasizes the unimportant ones. Hence, we stick
-to a discrete shading with few colors that is easier to interpret.
+to a cell controls the interpolation between the neutral and the full color. However,
+this typically results in too much color in the plot which in turn tends
+to conceal the important cells and over-emphasize the unimportant ones. Hence,
+a discrete shading with few colors is much easier to interpret.
 
@@
 its critical value is equivalent
-to a significant value of the whole test statistic $\lambda(\{r_{ij}\})$.\footnote{This
+to a significant value of the whole test statistic $\lambda([r_{ij}])$.\footnote{This
 can easily be seen from the following consideration: To obtain a level $\alpha$ test, where 
@@
 individual cells differently but can only be realized by using different colors for the
-whole table. A outlined before, colorfulness is intuitively matched with interestingness,
+whole table. As outlined before, colorfulness is intuitively matched with interestingness,
 therefore a rather natural idea is to use the fully colored palette only when the corresponding
@@
 colors, the maximum chroma was set to $100$ as shown in Figure~\ref{fig:shadingHSVHCL} and 
-is now reduced to $20$ for the non-significant palette. This palette still codes the absolute
+is reduced to $20$ for the non-significant palette. This palette still codes the absolute
 size of the residuals by luminance (i.e., the amount of grey), uses the same neutral grey
@@
 @
-\caption{Mosaic plot for the Arthritis data with Friendly shading (left) and maximum shading (right).}
+\caption{Mosaic plot for the arthritis data with Friendly shading (left) and maximum shading (right).}
 \label{fig:shadings}
@@
 Section~\ref{sec:shadings}, we have only considered the independence problem
-in 2-way tables up to now. But, in fact, none of the ideas described are really
+in 2-way tables. But, in fact, none of the ideas described are really
 confined to the 2-dimensional case as will be outlined in this section.
@@
 which the tables of expected values and residuals (again Pearson or deviance) can
-be computed. \citep{vcd:Friendly:1994,vcd:Friendly:1999} shows that his residual-based 
+be computed. \cite{vcd:Friendly:1994,vcd:Friendly:1999} shows that his residual-based 
 shading scheme can directly be applied to these more complex independence models.
@@
 is again the sum of squares yielding the associated Pearson or likelihood
-ratio statistic respectively \citep{vcd:Agresti:2002}. Hence, our residual-based
+ratio statistic, respectively \citep{vcd:Agresti:2002}. Hence, our residual-based
 shading with different HCL palettes for significant and non-significant results
@@
 case is spelt out, but the same ideas extend fairly straightforwardly to conditional
-independence in higher dimensions. In a 3-way table $\{n_{ijk}\}$ with underlying
-theoretical probability distribution $\{p_{ijk}\}$ the conditional independence
+independence in higher dimensions. In a 3-way table $[n_{ijk}]$ with underlying
+theoretical probability distribution $[p_{ijk}]$ the conditional independence
 problem can be formulated as
@@
 with $k = 1, \dots, K$. Under the assumption of conditional independence,
-we can again compute expected frequencies $\{\hat n_{ijk}\}$ and the corresponding 
-residuals $\{r_{ijk}\}$. To test the conditional independence hypothesis, usually
+we can again compute expected frequencies $[\hat n_{ijk}]$ and the corresponding 
+residuals $[r_{ijk}]$. To test the conditional independence hypothesis, usually
 the sum-of-squares statistic is used
@@
 which is simply the sum of the individual sum-of-squares statistics $X_k^2$ in each stratum $k$.
+
 Alternatively, a maximum statistic similar to that from Equation~(\ref{eq:MaxAbs})
@@
 for the deviation from conditional independence (if any). If it is not so much of interest
-in which \emph{cell} but only in which \emph{stratum} the deviation occures, then
+in which \emph{cell} but only in which \emph{stratum} the deviation occurs, then
 it would be natural to use 
@@
 statistics $X_k^2$ exceed the critical value are in conflict with the hypothesis of 
-conditional independence. All these statistics are of type
+conditional independence.
+
+All these statistics are of type
 \begin{equation} \label{eq:coindep_test}
-\lambda_\mathrm{agg} (\lambda_\mathrm{indep}(r_{ijk}))
+\lambda_\mathrm{agg} (\lambda_\mathrm{indep}(r_{ijk})),
 \end{equation}
@@
 the corresponding cells can also be identified. Hence, the double maximum
-statistic from~(\ref{eq:doublemax}) is the only functional allowing for detection of
-both the strata and the cells which caused the independence.
+statistic from Equation~(\ref{eq:doublemax}) is the only functional allowing for detection of
+both the strata and the cells which caused the dependence.
 
 But the main purpose of the formulation of the different test statistics is not so much 
-inference but their applicability to diagnostic plots via residual-based shadings. As stated
-at the beginning of this section, it is possible for all aggregation functionals to simply use either
+inference but their applicability to diagnostic plots via residual-based shadings. As already
+discussed, it is possible for all aggregation functionals to simply use either
 the significant or the non-significant shading for all cells in the contingency table---this
-strategy would have to be used for the sum-of-squares statistic from~(\ref{eq:sumChisq}). If
-$\lambda_\mathrm{agg}$ is the maximum as in~(\ref{eq:maxChisq}),
+strategy would have to be used for the sum-of-squares statistic from Equation~(\ref{eq:sumChisq}). If
+$\lambda_\mathrm{agg}$ is the maximum as in Equation~(\ref{eq:maxChisq}),
 then the significant palette would only be used
@@
 can be pursued, i.e., only the significant palette is used but with data-driven cut offs
-derived from the distribution of the double maximum statistic from~(\ref{eq:doublemax}).
-
-To arrange the shaded rectangles of the association or the mosaic plot respectively,
+derived from the distribution of the double maximum statistic from Equation~(\ref{eq:doublemax}).
+
+To arrange the shaded rectangles of the association or the mosaic plot, respectively,
 the most intuitive approach is to use the same conditioning in the display that
 was also used for conditioning in the model. For this situation, \cite{vcd:Friendly:1999}
-discusses a grouping similar to coplots \citep[conditioning plots, see]{vcd:Cleveland:1993}
+discusses a grouping similar to coplots \citep[conditioning plots, see][]{vcd:Cleveland:1993}
 that lead to trellis graphics \citep{vcd:Becker+Cleveland+Shyu:1996}. Thus, a
@@
 from conditional independence is picked up by all three statistics formulated
-above: the sum-of-squares statistic from~(\ref{eq:sumChisq}) is
+above: the sum-of-squares statistic from Equation~(\ref{eq:sumChisq}) is
 $\sum_k X_k^2 = \Sexpr{stat(ucb_sumChisq)}$ ($p = \Sexpr{pval(ucb_sumChisq)}$),
-the maximum sum-of-squares statistic from~(\ref{eq:maxChisq}) is
+the maximum sum-of-squares statistic from Equation~(\ref{eq:maxChisq}) is
 $\max_k X_k^2 = \Sexpr{stat(ucb_maxChisq)}$ ($p = \Sexpr{pval(ucb_maxChisq)}$),
-and the double maximum statistic from~(\ref{eq:doublemax}) is
+and the double maximum statistic from Equation~(\ref{eq:doublemax}) is
 $\max_k M_k = \Sexpr{stat(ucb_doublemax)}$ ($p = \Sexpr{pval(ucb_doublemax)}$).
@@
 instead of an HSV scheme are that the colors from this perceptually based color space
-are device inependent and provide uniform saturations over different hues.
+are device independent and provide uniform saturations over different hues.
 Furthermore, the colorfulness in this shading can be controlled independently from
@@
 The results in this paper were obtained using
-\proglang{R}~\Sexpr{paste(R.Version()[6:7], collapse = ".")} \citep{vcd:R:2005}
-and the packages
+\proglang{R}~\Sexpr{paste(R.Version()[6:7], collapse = ".")}
+\citep[\mbox{\url{http://www.R-project.org/}}]{vcd:R:2005} and the packages
 \pkg{vcd}~\Sexpr{packageDescription("vcd")$Version} \citep{vcd:Meyer+Zeileis+Hornik:2005},
@@
 ## load packages and data
-library(vcd)
+library("vcd")
 art <- xtabs(~ Treatment + Improved, data = Arthritis, subset = Sex == "Female")
@@
 set.seed(rseed)
-cotabplot(~ Admit + Gender | Dept, data = ucb, panel = cotab_coindep, type = "assoc")
+cotabplot(~ Admit + Gender | Dept, data = ucb, panel = cotab_coindep,
+          n = 5000, type = "assoc")
 
@@
 ## data
-data(Employment)
+data("Employment")
 emp <- aperm(Employment)

### 2005-08-10 trunk/vcd/R/assoc.R

@@ -153,7 +153,7 @@ struc_assoc <- function(compress = TRUE, xlim = NULL, ylim = NULL,
                     )
     for (i in seq(along = mnames)) {
       seekViewport(paste("cell", mnames[i], sep = ".."))
-      grid.lines(y = unit(0, "native"), gp = gpar(lty = 5))
+      grid.lines(y = unit(0, "native"), gp = gpar(lty = 3))
       grid.rect(y = 0, x = 0,
                 height = residuals[i],
                 width = xscale * unit(sqrt(expected[i]), "native"),

### 2005-08-11 trunk/vcd/paper/JCGS.Rnw
Added more color references.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@439 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 degrees of freedom under the null hypothesis.
-This is the well-known $\chi^2$ test which is typically introduced at the very beginning
-of the chapter about independence in 2-way tables in statistics textbooks
+This is the well-known $\chi^2$ test which is typically introduced first in statistics
+textbooks if they address the independence problem in 2-way tables
 \citep[see e.g.,][]{vcd:Agresti:2002}.
@@
 Furthermore, an important reason for using the unconditional limiting distribution
-for the $X^2$ statistic from (\ref{eq:ChiSquare}) was the closed form result
+for the $X^2$ statistic from Equation~(\ref{eq:ChiSquare}) was the closed form result
 for the distribution. Recently, with the improving perfomance of computers,
@@
 or by computation of the (asymptotic) permutation distribution---have been receiving
-increasing attention \citep{vcd:Pesarin:2001,vcd:Strasser+Weber:1999}. 
-For testing the independence hypothesis from (\ref{eq:H0}), using a permutation test is particularly
+increasing attention \citep[e.g.,][]{vcd:Ernst:2004,vcd:Pesarin:2001,vcd:Strasser+Weber:1999}. 
+For testing the independence hypothesis from Equation~(\ref{eq:H0}), using a permutation test is particularly
 intuitive due to the permutation invariance (given row and column sums) of this problem.
@@
 For the arthritis data from Table~\ref{tab:art}, both tests indicate a clearly significant
-dependence of improvement on treatment: the sum-of-squares statistic from (\ref{eq:ChiSquare})
+dependence of improvement on treatment: the sum-of-squares statistic from Equation~(\ref{eq:ChiSquare})
 is $X^2 = \Sexpr{stat(art_chisq)}$ with a $p$ value of $p = \Sexpr{pval(art_chisq)}$,
-and the maximum statistic from (\ref{eq:MaxAbs}) is $M = \Sexpr{stat(art_max)}$
+and the maximum statistic from Equation~(\ref{eq:MaxAbs}) is $M = \Sexpr{stat(art_max)}$
 with $p = \Sexpr{pval(art_max)}$. Both $p$ values have been computed
@@
 The implementation in the \proglang{R} system for statistical computing and graphics
-\citep{vcd:R:2005} employs colors from
-HSV (Hue--Saturation--Value) space. The latter is a very common implementation of
-colors in many computer packages and makes the generation of the Friendly shading
-very simple.
+\citep{vcd:R:2005} employs colors from HSV (Hue--Saturation--Value) space. 
+Both spaces are rather similar transformations of RGB (Red--Green--Blue) space
+\citep{vcd:Brewer:1999,vcd:Poynton:2000} and are very common implementations of
+colors in many computer packages \citep{vcd:Moretti+Lyons:2002} making the generation
+of the Friendly shading very simple. As both spaces can generate the
+same colors, we only discuss HSV space in the following.
 
@@
 selecting an appropriate palette for a specific problem. Furthermore, the
-Commission Internationale de l'\'Eclairage (CIE) introduced the two perceptually
+\citet[CIE,][]{vcd:CIE:2004} introduced the two perceptually
 based color spaces CIELAB and CIELUV where the latter is typically preferred for 
@@
 The Friendly shading can be interpreted to be a visualization of the maximum
-statistic $M$ from (\ref{eq:MaxAbs}) which always employs the critical values $c_\alpha$
+statistic $M$ from Equation~(\ref{eq:MaxAbs}) which always employs the critical values $c_\alpha$
 $2$ and $4$. But the problem is that it is not clear to which

### 2005-08-12 trunk/vcd/paper/JCGS.Rnw
KH and DM changes included, abstract added, only intro missing
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@441 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
   pdfsubject = {Residual-based Shadings},
-  pdfkeywords = {contingency tables, association plots, mosaic plots, HCL colors, 
-    HSV colors, conditional inference, permutation tests},
+  pdfkeywords = {association plots, conditional inference, contingency tables, HCL colors, HSV colors, mosaic plots},
   pdfauthor = {Achim Zeileis, David Meyer, Kurt Hornik},
@@
 
-hue.slice <- function(hue, grid.n = 100, type = c("HCL", "HSV"), plot = TRUE, fixup = FALSE)
+hue.slice <- function(hue, grid.n = 101, type = c("HCL", "HSV"), plot = TRUE, fixup = FALSE)
 {
@@
 \begin{abstract}
-Visualizing independence by association and mosaic plots.
-
-Enhancements: diverging color palette based on HCL color space
-combined with visualization of the result of a significance test.
-
-Extensions to visualizing conditional independence.\\[0.3cm]
+Residual-based shadings for enhancing mosaic and association plots
+to visualize independence models for contingency tables are extended
+in two directions: (a)~perceptually uniform HCL colors are used and (b)~the
+result of an associated significance test is coded by the presence of
+color in the visualization. For obtaining (a), a general strategy for
+deriving diverging palettes in the perceptually-based HCL space is suggested. As for (b),
+cut offs that control the appearance of color are computed in a data-driven
+way based on the conditional permutation distribution of maximum-type
+test statistics. The shadings are first established for the case of independence
+in 2-way tables and then extended to more general independence models
+for multi-way tables, including in particular the conditional independence
+problem.\\[0.3cm]
 
@@
 \section{Introduction} \label{sec:introduction}
+
+explore independence structure in contingency tables by log-linear models and visually
+by mosaic displays
 
@@
 \citep{vcd:Theus+Lauer:1999,vcd:Hofmann:2003,vcd:Hofmann:2001}.
-
-Association plots as residual plots for log-linear models
-\citep{vcd:Meyer+Zeileis+Hornik:2003}, especially
-in coplot or trellis layout.
 
@@
 
-But the sum of squares is not the only plausible way of capturing deviations
+However, the sum of squares is not the only plausible way of capturing deviations
 from $0$ in the residuals.
@@
 lead to reasonable test statistics $\lambda([r_{ij}])$: one which is particularly
-suitable for identifying the cells which cause the dependence (if any) is
+suitable for identifying the cells responsible for the dependence (if any) is
 the maximum of the absolute values
@@
 absolute value exceeds $c_\alpha$ violate the hypothesis of indendence at significance level
-$\alpha$ \cite[][ch.~7]{vcd:Mazanec+Strasser:2000}. Thus, the interesting cells causing
-the dependence can easily be identified.
+$\alpha$ \cite[][ch.~7]{vcd:Mazanec+Strasser:2000}. Thus, the interesting cells responsible
+for the dependence can easily be identified.
 
@@
 
-Other measures of discrepancy (e.g., deviance residuals) could, of course, also be used
-instead of the the Pearson residuals $[r_{ij}]$. But as the ideas discussed here
-extend straightforwardly to that situation, we do not go into detail about this.
+Note, that virtually all ideas discussed in this paper also extend straightforwardly to the situation
+where other measures of discrepancy (such as, e.g., deviance residuals) are used
+instead of the Pearson residuals $[r_{ij}]$.
 
@@
 from a plot for contingency tables to a visualization technique for
-log-linear models---and thus also for independence problems including
+log-linear models and their residuals---and thus also for independence problems including
 2-way tables as the simplest case.
@@
 \proglang{SAS} implementation \citep[see][]{vcd:Friendly:2000}, Michael Friendly
-uses colors from a palette based on HLS (Hue--Luminance--Saturation) color space.
+uses colors from a palette based on HLS (Hue-Luminance-Saturation) color space.
 The implementation in the \proglang{R} system for statistical computing and graphics
-\citep{vcd:R:2005} employs colors from HSV (Hue--Saturation--Value) space. 
-Both spaces are rather similar transformations of RGB (Red--Green--Blue) space
+\citep{vcd:R:2005} employs colors from HSV (Hue-Saturation-Value) space. 
+Both spaces are rather similar transformations of RGB (Red-Green-Blue) space
 \citep{vcd:Brewer:1999,vcd:Poynton:2000} and are very common implementations of
@@
 
-The HSV space originally looks like a cone with black at its
-peak (zero value) and full color wheels for different saturations at the other end,
+The HSV space originally looks like a cone\footnote{The wikipedia gives a very useful
+overview at \url{http://en.wikipedia.org/wiki/HSV_color_space} which also provides
+links to comparisons with other color spaces discussed in this paper.}
+with black at its peak (zero value) and full color wheels for different saturations at the other end,
 centered around a white (full value). Type and amount of color are controlled by hue
@@
 This is also depicted in the upper panel of Figure~\ref{fig:shadingHSVHCL} which shows
-two slides from the HSV cylinder side by side. The plot shows the saturation/value plane
-for the given hues $h = 0$ and $h = 2/3$. The significant color palette shows the
+the saturation/value plane for the given hues $h = 0$ and $h = 2/3$. 
+The significant color palette shows the
 colors used for the Friendly shading when the residuals are increasing from left to right.
@@
 general and for shading in statistical graphs in particular. The idea of using
-perceptually based colors that are `in harmony' go back to \cite{vcd:Munsell:1905} 
+perceptually-based colors that are `in harmony' go back to \cite{vcd:Munsell:1905} 
 who introduced a color notation for balanced colors. Based on similar principles, Cynthia
@@
 how CIELUV colors can be used for choosing qualitative palettes for statistical graphics such as barplots.
-By taking polar coordinates in CIELUV space, it is called HCL (Hue--Chroma--Luminance) space
+By taking polar coordinates in CIELUV space, it is called HCL (Hue-Chroma-Luminance) space
 and qualitative palettes can easily be chosen by using a range of hues for fixed values
@@
 hue ranging in $[0, 360]$ degrees and chroma and luminance ranging in $[0, 100]$
-percent. But not all combinations $(h, c, l)$ yield valid HCL colors and the admissable combinations
-of $c$ and $l$ vary across different hues $h$. But for the problem of constructing
+percent. However, not all combinations $(h, c, l)$ yield valid HCL colors and the admissable combinations
+of $c$ and $l$ vary across different hues $h$. For the task of constructing
 a diverging palette, this problem can easily be overcome as we just need two different
@@
 similar shapes in the chroma/luminance plane. The lower panel of Figure~\ref{fig:shadingHSVHCL}
-shows two such planes side by side for the hues $h = 0$ and $h = 260$. To obtain a 
+shows two such planes side by side for the hues $h = 0$ and $h = 260$.\footnote{The hue
+$h = 260$ is chosen (rather than $h = 240$ which would correspond to $h = 2/3$ in HSV space)
+because its chroma/luminance plane is most similar to that of $h = 0$.}
+To obtain a 
 sequence of colors with the same properties as the Friendly shading, the palette starts
@@
 @
-\caption{Extended shading in HSV (upper) and HCL space (lower).}
+\caption{Residual-based shadings in HSV (upper) and HCL space (lower).}
 \label{fig:shadingHSVHCL}
@@
 Of course, it can be argued that the shading was not designed for this purpose
-and that different cut offs than $2$ and $4$ should be used here. But then
-again, it would be nice if these cut offs could be chosen automatically in a
-data-driven fashion which is what we do below.
+and that different cut offs than $2$ and $4$ should be used here. However,
+in this situation, it would be nice if such cut offs could be chosen automatically in a
+data-driven way. Strategies for this are derived in the following.
 
@@
 statistic $M$ from Equation~(\ref{eq:MaxAbs}) which always employs the critical values $c_\alpha$
-$2$ and $4$. But the problem is that it is not clear to which
+$2$ and $4$. The problem is that it is not clear to which
 significance levels $\alpha$ these critical values correspond because the distribution
@@
 in the plot is equivalent to significance (of the maximum statistic $M$) at 
-level $\alpha = 0.1$ and $\alpha = 0.01$, respectively, and exactly the cells which `cause'
-the dependence are highlighted. For the arthritis data, these are in particular the 
+level $\alpha = 0.1$ and $\alpha = 0.01$, respectively, and exactly the cells which
+violate the independence hypothesis are highlighted. For the arthritis data, these are in particular the 
 cells in the last column that signal that there are significantly more
@@
 can be obtained in the HSV space if the value is reduced from $1$ to $0.5$ for non-significant 
-palettes (see Figure~\ref{fig:shadingHSVHCL}). But as for the significant colors,
+palettes (see Figure~\ref{fig:shadingHSVHCL}). As for the significant colors,
 the dimensions used for creating this palette are confounded and hence the HCL scheme is clearly
@@
 Section~\ref{sec:shadings}, we have only considered the independence problem
-in 2-way tables. But, in fact, none of the ideas described are really
-confined to the 2-dimensional case as will be outlined in this section.
+in 2-way tables. In this section, generalizations of these ideas to independence
+problems for multi-way tables are outlined.
 
@@
 is again the sum of squares yielding the associated Pearson or likelihood
-ratio statistic, respectively \citep{vcd:Agresti:2002}. Hence, our residual-based
-shading with different HCL palettes for significant and non-significant results
-can directly be applied to these models. For the maximum shading with
-data-driven cut offs, even more flexibility is possible than in the 2-way case
-when the structure of the independence problem is exploited which will be discussed 
-below for conditional independence
+ratio statistic, respectively \citep{vcd:Agresti:2002}.
+
+As these independence models for multi-way tables also provide the structure
+required for the residual-based HCL shadings derived in Section~\ref{sec:shadings},
+both the maximum shading and the sum-of-squares shading can straightforwardly be applied.
+However, such independence models often additionally provide further structure that allows
+decomposition of the overall model into smaller independence problems which can be exploited
+both for choosing appropriate data-driven cut-offs in the shading and for selecting 
+a suitable layout of mosaic or association plots. Specific strategies for the conditional
+independence problem are derived in the following.
 
@@
 
-To show how the structure of the independence problem can be exploited for 
-choosing appropriate critical values in the maximum shading and for selecting 
-a suitable layout of mosaic or association plots, we consider the conditional
-independence problem. To keep the notational overhead low again, only the 3-way
-case is spelt out, but the same ideas extend fairly straightforwardly to conditional
-independence in higher dimensions. In a 3-way table $[n_{ijk}]$ with underlying
-theoretical probability distribution $[p_{ijk}]$ the conditional independence
-problem can be formulated as
+How the structure of the independence problem can be exploited for 
+selecting a suitable layout and shading for mosaic and association plots,
+is exemplified with the conditional independence problem in 3-way tables.
+For a table $[n_{ijk}]$ with underlying
+theoretical probability distribution $[p_{ijk}]$, this can be formulated as
 \begin{equation} \label{eq:H0ci}
@@
 \begin{equation} \label{eq:coindep_test}
-\lambda_\mathrm{agg} (\lambda_\mathrm{indep}(r_{ijk})),
+\lambda_\mathrm{agg} (\lambda_\mathrm{indep}([r_{ijk}])),
 \end{equation}
@@
 and $\lambda_\mathrm{agg}$ is a functional for aggregating over the $k = 1, \dots, K$ strata.
-If the maximum is used for the latter, then identification of the strata causing the
-independence is possible. If additionally $\lambda_\mathrm{indep}$ is the maximum,
+If the maximum is used for the latter, then identification of the strata responsible for the
+dependence is possible. If additionally $\lambda_\mathrm{indep}$ is the maximum,
 the corresponding cells can also be identified. Hence, the double maximum
 statistic from Equation~(\ref{eq:doublemax}) is the only functional allowing for detection of
-both the strata and the cells which caused the dependence.
-
-But the main purpose of the formulation of the different test statistics is not so much 
+both the strata and the cells violating the conditional independence hypothesis.
+
+However, the main purpose of the formulation of the different test statistics is not so much 
 inference but their applicability to diagnostic plots via residual-based shadings. As already
@@
 For illustration, the famous admissions data of the University of
-California at Berkley (UCB) from \cite{vcd:Bickel+Hammel+O'Connell:1975}
+California, Berkley, (UCB) from \cite{vcd:Bickel+Hammel+O'Connell:1975}
 are employed. Testing the conditional independence of admission and gender
@@
 in HCL space is introduced. The advantages of using this HCL shading scheme
-instead of an HSV scheme are that the colors from this perceptually based color space
-are device independent and provide uniform saturations over different hues.
-Furthermore, the colorfulness in this shading can be controlled independently from
-the other two dimensions.
+instead of an HSV scheme are that the colors from this perceptually-based color space
+provide uniform saturations over different hues and that the colorfulness in this
+shading can be controlled independently from the other two dimensions.
 
@@
 codes the result of the significance test by using full colors only for significant results and
-the same type of palette but with a reduced amount of color for non-significant results.
+the same type of palette with a reduced amount of color for non-significant results.
 
@@
 set.seed(rseed)
-coindep_test(art, n = 5000)
+coindep_test(art, n = 5000, indepfun = function(x) sum(x^2))
 ## M (4)
 set.seed(rseed)
-coindep_test(art, n = 5000, indepfun = function(x) sum(x^2))
+coindep_test(art, n = 5000)
 

### 2005-08-19 trunk/vcd/paper/JCGS.Rnw
comments by DM included
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@442 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-explore independence structure in contingency tables by log-linear models and visually
-by mosaic displays
-
-independence in 2-way tables
-
-association and mosaic displays
-
-colors and color spaces
-    because the three perceptual dimensions (hue, lightness, saturation)
-    are not properly mapped to the three dimensions of the color space
-    and hence are confounded \citep{vcd:Brewer:1999}
-
-somewhere:
-mention Augsburg stuff, visualization of log-linear models via mosaics
-\citep{vcd:Theus+Lauer:1999,vcd:Hofmann:2003,vcd:Hofmann:2001}.
+Relationships between categorical variables are typically
+analyzed based on the underlying contingency tables that can be explored
+for (in)dependence. Two standard methods from the statistical tool box
+are log-linear models \citep[see, e.g.,][]{vcd:Agresti:2002} for modeling
+independence and mosaic plots \citep{vcd:Hartigan+Kleiner:1981} for 
+bringing it out graphically. Both methods can also be combined such that
+a certain mosaic plot visualizes a particular log-linear model by 
+controlling splitting order and direction and shading of the tiles
+in the mosaic display
+\citep{vcd:Friendly:1994,vcd:Friendly:1999,vcd:Theus+Lauer:1999,vcd:Hofmann:2001}.
+In particular, \cite{vcd:Friendly:1994} suggested a shading strategy
+based on the residuals (typically, Pearson or deviance residuals) of the associated
+log-linear model that elevates the mosaic plot from a display for frequencies
+in a contingency table to a visualization technique that encompasses both, observed
+frequencies and residuals. This shading allows for judging the quality of a model
+fit and spotting dependence patterns that have not been accounted for by the model.
+Other techniques for visualizing independence in contingency tables such as
+association plots \citep{vcd:Cohen:1980} can also be enhanced by using this
+residual-based shading.
+
+In this paper, the shading of \cite{vcd:Friendly:1994} is extended in two directions:
+usage of perceptually-based HCL (Hue-Chroma-Luminance) colors and combination of
+visualization and significance
+testing. Friendly's shading is typically implemented in statistical software using color
+spaces whose dimensions are only poorly mapped to the perceptual dimensions of the
+human visual system \citep{vcd:Brewer:1999,vcd:Ihaka:2003} which makes it more difficult
+to properly read and interpret the corresponding plots. For overcoming this problem,
+a general strategy for constructing diverging palettes in the perceptually-based HCL
+space \citep{vcd:Ihaka:2003} is derived. To couple the visualization and the independence
+model of a contingency table more tightly than in previous approaches, the residual-based
+shading is extended such that presence of color in the display is equivalent to significance
+of the associated independence test. This is achieved by using data-driven cut offs
+for the appearance of color computed from the conditional permutation distribution
+\citep{vcd:Ernst:2004,vcd:Pesarin:2001} of maximum-type test statistics. The resulting
+residual-based shadings are illustrated both for mosaic and association plots 
+using real world data sets.
+
+The remainder of the paper is structured as follows: Section~\ref{sec:2way} gives a
+brief introduction to significance tests and visualization techniques for the
+independence problem in 2-way contingency tables. Based on this, Section~\ref{sec:shadings}
+introduces the extended residual-based shadings using perceptually uniform HCL colors
+and combining visualization and significance testing. The results are generalized to
+multi-way tables in Section~\ref{sec:3way} with a more detailed discussion of adapting
+them to conditional independence. Section~\ref{sec:conclusions} summarizes the paper
+and gives some concluding remarks.
 
@@
 Table~\ref{tab:art}.
+
+\begin{table}[b!]
+\begin{center}
+\caption{\label{tab:art} Treatment and improvement among $59$ patients
+with rheumatoid arthritis.}
+
+\begin{tabular}{rr|rrr}
+\\
+                   &         & \multicolumn{3}{|l}{\textbf{Improved}} \\
+                   &         & None & Some & Marked \\ \hline
+\textbf{Treatment} & Placebo &   19 &    7 &      6 \\
+                   & Treated &    6 &    5 &     16
+\end{tabular}
+
+\end{center}
+\end{table}
 
@@
 textbooks if they address the independence problem in 2-way tables
-\citep[see e.g.,][]{vcd:Agresti:2002}.
-
-\begin{table}[t!]
-\begin{center}
-\caption{\label{tab:art} Treatment and improvement among $59$ patients
-with rheumatoid arthritis.}
-
-\begin{tabular}{rr|rrr}
-\\
-                   &         & \multicolumn{3}{|l}{\textbf{Improved}} \\
-                   &         & None & Some & Marked \\ \hline
-\textbf{Treatment} & Placebo &   19 &    7 &      6 \\
-                   & Treated &    6 &    5 &     16
-\end{tabular}
-
-\end{center}
-\end{table}
+\citep[see, e.g.,][]{vcd:Agresti:2002}.
 
@@
 
-\emph{Mosaic plots}
-\citep{vcd:Hartigan+Kleiner:1981} can be seen as an extension of grouped
-bar charts where width and height of the bars show the relative
-frequencies of the two variables: a mosaic plot simply
-consists of a collection of tiles with areas
-proportional to the observed cell frequencies as shown
-in the left panel of Figure~\ref{fig:classic}. A rectangle corresponding
-to $100$ percent of the observations is first split horizontally with respect to 
-the treatment frequencies and then vertically with respect to the conditional
-improvement frequencies. This shows that there have been more placebo than treated 
-patients with no improvement and vice versa for marked improvement.
-This strategy of splitting with respect to conditional frequencies given all
-previous variables can also directly be used for visualizing multi-way tables
-\citep[see][for an overview of how to construct and read mosaic displays]{vcd:Hofmann:2003}.
-
-\emph{Association plots}
-\citep{vcd:Cohen:1980} visualize the table of Pearson residuals:
-each cell is represented by a rectangle that has (signed) height
-proportional to the corresponding Pearson residual $r_{ij}$ and width proportional to the
-square root of the expected counts $\sqrt{\hat n_{ij}}$. Thus, the area is proportional to
-the raw residuals $n_{ij} - \hat n_{ij}$. The association plot for the arthritis
-data is shown in the right panel of Figure~\ref{fig:classic} which leads to the
-same interpretation as the mosaic plot: there are more placebo patients with no 
-improvement and fewer with marked improvement than expected under independence---vice
-versa for the treated patiens.
-
 \setkeys{Gin}{width=\textwidth}
-\begin{figure}[htbp]
+\begin{figure}[b!]
 \begin{center}
@@
 \end{figure}
+
+\emph{Mosaic plots}
+\citep{vcd:Hartigan+Kleiner:1981} can be seen as an extension of grouped
+bar charts where width and height of the bars show the relative
+frequencies of the two variables: a mosaic plot simply
+consists of a collection of tiles with areas
+proportional to the observed cell frequencies as shown
+in the left panel of Figure~\ref{fig:classic}. A rectangle corresponding
+to $100$ percent of the observations is first split horizontally with respect to 
+the treatment frequencies and then vertically with respect to the conditional
+improvement frequencies. This shows that there have been more placebo than treated 
+patients with no improvement and vice versa for marked improvement.
+This strategy of splitting with respect to conditional frequencies given all
+previous variables can also directly be used for visualizing multi-way tables
+\citep[see][for an overview of how to construct and read mosaic displays]{vcd:Hofmann:2003}.
+
+\emph{Association plots}
+\citep{vcd:Cohen:1980} visualize the table of Pearson residuals:
+each cell is represented by a rectangle that has (signed) height
+proportional to the corresponding Pearson residual $r_{ij}$ and width proportional to the
+square root of the expected counts $\sqrt{\hat n_{ij}}$. Thus, the area is proportional to
+the raw residuals $n_{ij} - \hat n_{ij}$. The association plot for the arthritis
+data is shown in the right panel of Figure~\ref{fig:classic} which leads to the
+same interpretation as the mosaic plot: there are more placebo patients with no 
+improvement and fewer with marked improvement than expected under independence---vice
+versa for the treated patiens.
 
@@
 The set of non-significant colors will be explained in Section~\ref{sec:significance}.
+
+\setkeys{Gin}{width=.8\textwidth}
+\begin{figure}[p]
+\begin{center}
+<<shading_HSV,fig=TRUE,echo=FALSE,height=7,width=8>>=
+## generate colors
+hue23 <- hue.slice(2/3, grid.n = 101, plot = FALSE, type = "HSV")
+hue0 <- hue.slice(0, grid.n = 101, plot = FALSE, type = "HSV")
+saturation <- as.numeric(colnames(hue23))
+value <- as.numeric(rownames(hue23))
+
+## select those with value >= 0.5
+hue23 <- hue23[value >= .5, ]
+hue0 <- hue0[value >= .5, ]
+value <- value[value >= .5]
+nl <- nrow(hue23)
+nc <- ncol(hue23)
+
+## plot 2 slides from HSV space
+plot(0.5, 0.5, xlim = c(-1, 1), ylim = c(0, 1), type = "n", axes = FALSE,
+       xlab = "", ylab = "", yaxs = "i", xaxs = "i", main = "")
+for(i in 1:(nc-1)) {
+  rect(saturation[i], value[-nl], saturation[i] + 1/(nc-1), value[-1], border = hue23[,i+1], col = hue23[,i+1])
+}
+for(i in 1:(nc-1)) {
+  rect(-saturation[i], value[-nl], -(saturation[i] + 1/(nc-1)), value[-1], border = hue0[,i+1], col = hue0[,i+1])
+}
+axis(2, at = c(50, 75, 100)/100, labels = c(0.5, 0.75, 1))
+axis(4, at = c(50, 75, 100)/100, labels = c(0.5, 0.75, 1))
+axis(3, at = -4:4*.25, labels=c(4:0*.25, 1:4*.25))
+mtext(c("hue = 0", "hue = 2/3"), side = 3, at = c(-.5, .5), line = 3, cex = 1.2)
+mtext("saturation", side = 3, at = 0, line = 2)
+mtext("value", side = 2, at = .75, line = 2)
+mtext("value", side = 4, at = .75, line = 2)
+lines(c(-1, 1), c(.5, .5))
+
+## significant colors
+rect(-1, 0.95, -.90, 1, col = hsv(0, 1, 1))
+rect(-0.45, 0.95, -.55, 1, col = hsv(0, 0.5, 1))
+rect(-.05, .95, .05, 1, col = hsv(2/3, 0, 1))
+rect(0.45, 0.95, .55, 1, col = hsv(2/3, 0.5, 1))
+rect(.90, .95, 1, 1, col = hsv(2/3, 1, 1))
+
+text(-1, .33, "significant", pos = 4, cex = 1.2)
+rect(-1, .20, -.80, .30, col = hsv(0, 1, 1))
+rect(-.40, .20, -0.6, .30, col = hsv(0, 0.5, 1))
+rect(-.20, .20, 0, .30, col = hsv(0, 0, 1))
+rect(0, .20, .20, .30, col = hsv(2/3, 0, 1))
+rect(0.4, .20, .60, .30, col = hsv(2/3, .5, 1))
+rect(.80, .20, 1, .30, col = hsv(2/3, 1, 1))
+
+lines(c(-.9, -.55), c(0.975, .975), lty = 2)
+lines(c(-.45, -.05), c(0.975, .975), lty = 2)
+lines(c(.45, .05), c(0.975, .975), lty = 2)
+lines(c(.9, .55), c(0.975, .975), lty = 2)
+
+## non-significant colors
+rect(-1, 0.5, -.90, 0.55, col = hsv(0, 1, 0.5))
+rect(-0.45, 0.5, -.55, 0.55, col = hsv(0, 0.5, 0.5))
+rect(-.05, .5, .05, 0.55, col = hsv(2/3, 0, 0.5))
+rect(0.45, 0.5, .55, 0.55, col = hsv(2/3, 0.5, 0.5))
+rect(.90, .5, 1, 0.55, col = hsv(2/3, 1, 0.5))
+
+text(-1, .13, "non-significant", pos = 4, cex = 1.2)
+rect(-1, 0, -.80, .10, col = hsv(0, 1, 0.5))
+rect(-.60, 0, -.4, .10, col = hsv(0, 0.5, 0.5))
+rect(-.20, 0, 0, .10, col = hsv(0, 0, 0.5))
+rect(0, 0, .20, .10, col = hsv(2/3, 0, 0.5))
+rect(0.4, 0, .60, .1, col = hsv(2/3, .5, 0.5))
+rect(.80, 0, 1, .10, col = hsv(2/3, 1, 0.5))
+
+lines(c(-.9, -.55), c(0.525, .525), lty = 2)
+lines(c(-.45, -.05), c(0.525, .525), lty = 2)
+lines(c(.45, .05), c(0.525, .525), lty = 2)
+lines(c(.9, .55), c(0.525, .525), lty = 2)
+@
+
+<<shading_HCL,fig=TRUE,echo=FALSE,height=7,width=8>>=
+## generate colors
+hue260 <- hue.slice(260, grid.n = 101, plot = FALSE)
+hue360 <- hue.slice(360, grid.n = 101, plot = FALSE)
+mychroma <- as.numeric(colnames(hue260))
+luminance <- as.numeric(rownames(hue260))
+
+## select those with lumincance >= 50
+hue260 <- hue260[luminance >= 50, ]
+hue360 <- hue360[luminance >= 50, ]
+luminance <- luminance[luminance >= 50]
+nc <- ncol(hue260)
+nl <- nrow(hue260)
+
+## plot 2 slides from HCL space
+plot(0.5, 0.5, xlim = c(-100, 100), ylim = c(0, 100), type = "n", axes = FALSE,
+       xlab = "", ylab = "", yaxs = "i", xaxs = "i", main = "")
+for(i in 1:(nc-1)) {
+  rect(mychroma[i], luminance[-nl], mychroma[i] + 100/(nc-1), luminance[-1], border = hue260[,i+1], col = hue260[,i+1])
+}
+for(i in 1:(nc-1)) {
+  rect(-mychroma[i], luminance[-nl], -(mychroma[i] + 100/(nc-1)), luminance[-1], border = hue360[,i+1], col = hue360[,i+1])
+}
+axis(2, at = c(50, 70, 90, 100), labels = c(50, 70, 90, 100))
+axis(4, at = c(50, 70, 90, 100), labels = c(50, 70, 90, 100))
+axis(3, at = -4:4*25, labels=c(4:0*25, 1:4*25))
+mtext(c("hue = 0", "hue = 260"), side = 3, at = c(-50, 50), line = 3, cex = 1.2)
+mtext("chroma", side = 3, at = 0, line = 2)
+mtext("luminance", side = 2, at = 75, line = 2)
+mtext("luminance", side = 4, at = 75, line = 2)
+lines(c(-100, 100), c(50, 50))
+
+## significant colors
+rect(-100, 47.5, -90, 52.5, col = hcl(0, 100, 50))
+rect(-55, 67.5, -45, 72.5, col = hcl(0, 50, 70))
+rect(-5, 95, 5, 100, col = hcl(260, 0, 100))       ## grey vs. white
+rect(-5, 87.5, 5, 92.5, col = hcl(260, 0, 90))     ## grey vs. white
+rect(45, 67.5, 55, 72.5, col = hcl(260, 50, 70))
+rect(90, 47.5, 100, 52.5, col = hcl(260, 100, 50))
+
+text(-100, 33, "significant", pos = 4, cex = 1.2)
+rect(-100, 20, -80, 30, col = hcl(0, 100, 50))
+rect(-60, 20, -40, 30, col = hcl(0, 50, 70))
+rect(-20, 20, 0, 30, col = hcl(0, 0, 90))       
+rect(0, 20, 20, 30, col = hcl(260, 0, 90))
+#white# rect(-20, 20, 0, 30, col = hcl(0, 0, 100))
+#white# rect(0, 20, 20, 30, col = hcl(260, 0, 100))
+rect(40, 20, 60, 30, col = hcl(260, 50, 70))
+rect(80, 20, 100, 30, col = hcl(260, 100, 50))
+
+lines(c(-45, -5), c(72.5, 87.5), lty = 2)
+lines(c(45, 5), c(72.5, 87.5), lty = 2)
+lines(c(-95, -55), c(52.5, 67.5), lty = 2)
+lines(c(95, 55), c(52.5, 67.5), lty = 2)
+
+## non-significant colors
+rect(-25, 47.5, -15, 52.5, col = hcl(0, 20, 50))
+rect(-15, 67.5, -5, 72.5, col = hcl(0, 10, 70))
+rect(5, 67.5, 15, 72.5, col = hcl(260, 10, 70))
+rect(25, 47.5, 15, 52.5, col = hcl(260, 20, 50))
+
+text(-100, 13, "non-significant", pos = 4, cex = 1.2)
+rect(-60, 0, -40, 10, col = hcl(0, 20, 50))
+rect(-40, 0, -20, 10, col = hcl(0, 10, 70))
+rect(-20, 0, 0, 10, col = hcl(0, 0, 90))
+rect(0, 0, 20, 10, col = hcl(260, 0, 90))
+rect(20, 0, 40, 10, col = hcl(260, 10, 70))
+rect(40, 0, 60, 10, col = hcl(260, 20, 50))
+
+lines(c(-18.75, -11.25), c(52.5, 67.5), lty = 2)
+lines(c(-8.75, -1.25), c(72.5, 87.5), lty = 2)
+lines(c(18.75, 11.75), c(52.5, 67.5), lty = 2)
+lines(c(8.75, 1.25), c(72.5, 87.5), lty = 2)
+@
+\caption{Residual-based shadings in HSV (upper) and HCL space (lower).}
+\label{fig:shadingHSVHCL}
+\end{center}
+\end{figure}
 
@@
 of visualization.
-
-\setkeys{Gin}{width=.8\textwidth}
-\begin{figure}[p]
-\begin{center}
-<<shading_HSV,fig=TRUE,echo=FALSE,height=7,width=8>>=
-## generate colors
-hue23 <- hue.slice(2/3, grid.n = 101, plot = FALSE, type = "HSV")
-hue0 <- hue.slice(0, grid.n = 101, plot = FALSE, type = "HSV")
-saturation <- as.numeric(colnames(hue23))
-value <- as.numeric(rownames(hue23))
-
-## select those with value >= 0.5
-hue23 <- hue23[value >= .5, ]
-hue0 <- hue0[value >= .5, ]
-value <- value[value >= .5]
-nl <- nrow(hue23)
-nc <- ncol(hue23)
-
-## plot 2 slides from HSV space
-plot(0.5, 0.5, xlim = c(-1, 1), ylim = c(0, 1), type = "n", axes = FALSE,
-       xlab = "", ylab = "", yaxs = "i", xaxs = "i", main = "")
-for(i in 1:(nc-1)) {
-  rect(saturation[i], value[-nl], saturation[i] + 1/(nc-1), value[-1], border = hue23[,i+1], col = hue23[,i+1])
-}
-for(i in 1:(nc-1)) {
-  rect(-saturation[i], value[-nl], -(saturation[i] + 1/(nc-1)), value[-1], border = hue0[,i+1], col = hue0[,i+1])
-}
-axis(2, at = c(50, 75, 100)/100, labels = c(0.5, 0.75, 1))
-axis(4, at = c(50, 75, 100)/100, labels = c(0.5, 0.75, 1))
-axis(3, at = -4:4*.25, labels=c(4:0*.25, 1:4*.25))
-mtext(c("hue = 0", "hue = 2/3"), side = 3, at = c(-.5, .5), line = 3, cex = 1.2)
-mtext("saturation", side = 3, at = 0, line = 2)
-mtext("value", side = 2, at = .75, line = 2)
-mtext("value", side = 4, at = .75, line = 2)
-lines(c(-1, 1), c(.5, .5))
-
-## significant colors
-rect(-1, 0.95, -.90, 1, col = hsv(0, 1, 1))
-rect(-0.45, 0.95, -.55, 1, col = hsv(0, 0.5, 1))
-rect(-.05, .95, .05, 1, col = hsv(2/3, 0, 1))
-rect(0.45, 0.95, .55, 1, col = hsv(2/3, 0.5, 1))
-rect(.90, .95, 1, 1, col = hsv(2/3, 1, 1))
-
-text(-1, .33, "significant", pos = 4, cex = 1.2)
-rect(-1, .20, -.80, .30, col = hsv(0, 1, 1))
-rect(-.40, .20, -0.6, .30, col = hsv(0, 0.5, 1))
-rect(-.20, .20, 0, .30, col = hsv(0, 0, 1))
-rect(0, .20, .20, .30, col = hsv(2/3, 0, 1))
-rect(0.4, .20, .60, .30, col = hsv(2/3, .5, 1))
-rect(.80, .20, 1, .30, col = hsv(2/3, 1, 1))
-
-lines(c(-.9, -.55), c(0.975, .975), lty = 2)
-lines(c(-.45, -.05), c(0.975, .975), lty = 2)
-lines(c(.45, .05), c(0.975, .975), lty = 2)
-lines(c(.9, .55), c(0.975, .975), lty = 2)
-
-## non-significant colors
-rect(-1, 0.5, -.90, 0.55, col = hsv(0, 1, 0.5))
-rect(-0.45, 0.5, -.55, 0.55, col = hsv(0, 0.5, 0.5))
-rect(-.05, .5, .05, 0.55, col = hsv(2/3, 0, 0.5))
-rect(0.45, 0.5, .55, 0.55, col = hsv(2/3, 0.5, 0.5))
-rect(.90, .5, 1, 0.55, col = hsv(2/3, 1, 0.5))
-
-text(-1, .13, "non-significant", pos = 4, cex = 1.2)
-rect(-1, 0, -.80, .10, col = hsv(0, 1, 0.5))
-rect(-.60, 0, -.4, .10, col = hsv(0, 0.5, 0.5))
-rect(-.20, 0, 0, .10, col = hsv(0, 0, 0.5))
-rect(0, 0, .20, .10, col = hsv(2/3, 0, 0.5))
-rect(0.4, 0, .60, .1, col = hsv(2/3, .5, 0.5))
-rect(.80, 0, 1, .10, col = hsv(2/3, 1, 0.5))
-
-lines(c(-.9, -.55), c(0.525, .525), lty = 2)
-lines(c(-.45, -.05), c(0.525, .525), lty = 2)
-lines(c(.45, .05), c(0.525, .525), lty = 2)
-lines(c(.9, .55), c(0.525, .525), lty = 2)
-@
-
-<<shading_HCL,fig=TRUE,echo=FALSE,height=7,width=8>>=
-## generate colors
-hue260 <- hue.slice(260, grid.n = 101, plot = FALSE)
-hue360 <- hue.slice(360, grid.n = 101, plot = FALSE)
-mychroma <- as.numeric(colnames(hue260))
-luminance <- as.numeric(rownames(hue260))
-
-## select those with lumincance >= 50
-hue260 <- hue260[luminance >= 50, ]
-hue360 <- hue360[luminance >= 50, ]
-luminance <- luminance[luminance >= 50]
-nc <- ncol(hue260)
-nl <- nrow(hue260)
-
-## plot 2 slides from HCL space
-plot(0.5, 0.5, xlim = c(-100, 100), ylim = c(0, 100), type = "n", axes = FALSE,
-       xlab = "", ylab = "", yaxs = "i", xaxs = "i", main = "")
-for(i in 1:(nc-1)) {
-  rect(mychroma[i], luminance[-nl], mychroma[i] + 100/(nc-1), luminance[-1], border = hue260[,i+1], col = hue260[,i+1])
-}
-for(i in 1:(nc-1)) {
-  rect(-mychroma[i], luminance[-nl], -(mychroma[i] + 100/(nc-1)), luminance[-1], border = hue360[,i+1], col = hue360[,i+1])
-}
-axis(2, at = c(50, 70, 90, 100), labels = c(50, 70, 90, 100))
-axis(4, at = c(50, 70, 90, 100), labels = c(50, 70, 90, 100))
-axis(3, at = -4:4*25, labels=c(4:0*25, 1:4*25))
-mtext(c("hue = 0", "hue = 260"), side = 3, at = c(-50, 50), line = 3, cex = 1.2)
-mtext("chroma", side = 3, at = 0, line = 2)
-mtext("luminance", side = 2, at = 75, line = 2)
-mtext("luminance", side = 4, at = 75, line = 2)
-lines(c(-100, 100), c(50, 50))
-
-## significant colors
-rect(-100, 47.5, -90, 52.5, col = hcl(0, 100, 50))
-rect(-55, 67.5, -45, 72.5, col = hcl(0, 50, 70))
-rect(-5, 95, 5, 100, col = hcl(260, 0, 100))       ## grey vs. white
-rect(-5, 87.5, 5, 92.5, col = hcl(260, 0, 90))     ## grey vs. white
-rect(45, 67.5, 55, 72.5, col = hcl(260, 50, 70))
-rect(90, 47.5, 100, 52.5, col = hcl(260, 100, 50))
-
-text(-100, 33, "significant", pos = 4, cex = 1.2)
-rect(-100, 20, -80, 30, col = hcl(0, 100, 50))
-rect(-60, 20, -40, 30, col = hcl(0, 50, 70))
-rect(-20, 20, 0, 30, col = hcl(0, 0, 90))       
-rect(0, 20, 20, 30, col = hcl(260, 0, 90))
-#white# rect(-20, 20, 0, 30, col = hcl(0, 0, 100))
-#white# rect(0, 20, 20, 30, col = hcl(260, 0, 100))
-rect(40, 20, 60, 30, col = hcl(260, 50, 70))
-rect(80, 20, 100, 30, col = hcl(260, 100, 50))
-
-lines(c(-45, -5), c(72.5, 87.5), lty = 2)
-lines(c(45, 5), c(72.5, 87.5), lty = 2)
-lines(c(-95, -55), c(52.5, 67.5), lty = 2)
-lines(c(95, 55), c(52.5, 67.5), lty = 2)
-
-## non-significant colors
-rect(-25, 47.5, -15, 52.5, col = hcl(0, 20, 50))
-rect(-15, 67.5, -5, 72.5, col = hcl(0, 10, 70))
-rect(5, 67.5, 15, 72.5, col = hcl(260, 10, 70))
-rect(25, 47.5, 15, 52.5, col = hcl(260, 20, 50))
-
-text(-100, 13, "non-significant", pos = 4, cex = 1.2)
-rect(-60, 0, -40, 10, col = hcl(0, 20, 50))
-rect(-40, 0, -20, 10, col = hcl(0, 10, 70))
-rect(-20, 0, 0, 10, col = hcl(0, 0, 90))
-rect(0, 0, 20, 10, col = hcl(260, 0, 90))
-rect(20, 0, 40, 10, col = hcl(260, 10, 70))
-rect(40, 0, 60, 10, col = hcl(260, 20, 50))
-
-lines(c(-18.75, -11.25), c(52.5, 67.5), lty = 2)
-lines(c(-8.75, -1.25), c(72.5, 87.5), lty = 2)
-lines(c(18.75, 11.75), c(52.5, 67.5), lty = 2)
-lines(c(8.75, 1.25), c(72.5, 87.5), lty = 2)
-@
-\caption{Residual-based shadings in HSV (upper) and HCL space (lower).}
-\label{fig:shadingHSVHCL}
-\end{center}
-\end{figure}
 
@@
 
-\section{Extension to Conditional Independence}
+\clearpage
+
+\section{Extension to Conditional Independence} \label{sec:3way}
 
@@
 
-\section{Conclusions}
+\section{Conclusions} \label{sec:conclusions}
 
@@
 shading can be controlled independently from the other two dimensions.
-
+%
 To combine visualization and significance testing, two approaches are presented:
@@
 the same type of palette with a reduced amount of color for non-significant results.
-
+%
 Both results can not only be applied to the simple independence problem in 2-way
@@
 for the conditional independence problem.
-
+%
 All associated significance tests are carried out using a conditional inference approach

### 2005-08-24 trunk/vcd/inst/doc/labeling.Rnw
sieveplot added to the strucplot framework; old function removed. line_col argument added to shading functions for default line color. dimnames added to independence_table.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@431 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
New shading vignette added.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@437 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
viewport names changed to "obj:dim1=level1,...,dimn=leveln" syntax.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@444 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \newcommand{\pkg}[1]{{\normalfont\fontseries{b}\selectfont #1}}
-\newcommand{\code}[1]{\texttt{#1}}
+\newcommand{\var}[1]{\textit{\texttt{#1}}}
 \newenvironment{smallexample}{\begin{alltt}}{\end{alltt}}
+
+%% \code without `-' ligatures
+{\catcode`\-=\active%
+  \global\def\code{\bgroup%
+    \catcode`\-=\active \let-\codedash%
+    \Rd@code}}
+\def\codedash{-\discretionary{}{}{}}
+\def\Rd@code#1{\texttt{#1}\egroup}
 
@@
 
-\SweaveOpts{engine=R,eps=TRUE,prefix.string=pics/labeling}
-\setkeys{Gin}{width=0.8\textwidth}
+\SweaveOpts{engine=R,eps=TRUE,height=6,width=7,results=hide,fig=FALSE,echo=TRUE}
+\setkeys{Gin}{width=0.7\textwidth}
 
-<<preliminaries,echo=FALSE,results=hide>>=
+<<preliminaries,echo=FALSE>>=
+options(width=70)
 set.seed(1071)
@@
 
-\title{Labeling in Stucplots}
+\title{Labeling in Strucplots}
 \author{by David Meyer\\
@@
 
-One of the major enhancements in vcd is the labeling of strucplots which now
+One of the major enhancements in \pkg{vcd} compared to
+\code{mosaicplot()} and \code{assocplot()} in
+base {\sf R} is the labeling of strucplots which now
 has become much more powerful and flexible. Like the panel, the
@@
 variables (if any). The labeling function has access to the arguments
-of the generating function because of lexical scoping. In the
+of the generating function because in {\sf R}, nested functions are
+lexically scoped. In the
 following, we describe the three basic modules
@@
 
-\section{Labels in the borders: \code{labeling\_text()}}
+\section{Labels in the borders: \texttt{labeling\_text()}}
 
@@
 plots labels in the borders similar to the \code{mosaicplot()}
-function, but is much more flexible: it is not limited
+function in base {\sf R}, but is much more flexible: it is not limited
 to 4 dimensions, and the positioning and graphical parameters 
@@
 
-As an example, consider the `Survival on the Titanic' data, consisting
-of 4 categorical variables: Survival, Gender (``Sex''), Age, and
+As an example, consider the `Titanic' data, consisting
+of 4 categorical variables: Survived, Sex (i.e., `Gender'), Age, and
 Crew. By default, the variable names and levels are plotted `around' the
@@
 
-\begin{Sinput}
-> mosaic(Titanic)
-\end{Sinput}
+<<default>>=
+mosaic(Titanic)
+@
 
-\setkeys{Gin}{width=0.7\textwidth}
 \begin{figure}[h]
 \begin{center}
-<<default,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-mosaic(Titanic)
+<<defaultfig,fig=TRUE,echo=FALSE>>=
+<<default>>
 @
@@
 \noindent Note that the last two levels of the `Survived' variable do
-overlap. This issue can be addressed in several ways. The `brute force'
-method is to enable clipping for the `Survivcal' dimension (see Figure \ref{fig:clipping}):
+overlap, as well as some `Adult' and `Child' labels of the `Age' Variable.
+This issue can be addressed in several ways. The `brute force'
+method is to enable clipping for these dimensions (see Figure \ref{fig:clipping}):
 
-\begin{Sinput}
-> mosaic(Titanic, labeling_args = list(clip = c(Survived = TRUE)))
-\end{Sinput}
+<<clipping>>=
+mosaic(Titanic, labeling_args = list(clip = c(Survived = TRUE, Age = TRUE)))
+@
 
-\setkeys{Gin}{width=0.7\textwidth}
 \begin{figure}[h]
 \begin{center}
-<<clipping,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-mosaic(Titanic, labeling_args = list(clip = c(Survived = TRUE)))
+<<clippingfig,fig=TRUE,echo=FALSE>>=
+<<clipping>>
 @
@@
 overridden by the named components. Here, the default is \code{FALSE},
-and therefore clipping is enabled only for the \code{Survived} dimension.
-A more sensible solution to the overlap problem is to abbreviate the `Survival'
+and therefore clipping is enabled only for the `Survived' and `Age' dimensions.
+A more sensible solution to the overlap problem is to abbreviate the 
 levels (see Figure \ref{fig:abbreviating}):
 
-\begin{Sinput}
-> mosaic(Titanic, labeling_args = list(abbreviate = c(Survived = TRUE)))
-\end{Sinput}
+<<abbreviating>>=
+mosaic(Titanic, labeling_args = list(abbreviate = c(Survived = TRUE, Age = 4)))
+@
 
-\setkeys{Gin}{width=0.7\textwidth}
 \begin{figure}[h]
 \begin{center}
-<<abbreviating,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-mosaic(Titanic, labeling_args = list(abbreviate = c(Survived = TRUE)))
+<<abbreviatingfig,fig=TRUE,echo=FALSE>>=
+<<abbreviating>>
 @
@@
 
-\noindent The \code{abbreviate} argument actually takes a vector of
-integers indicating the number of characters the levels should be
-abbreviated to, but logical values are equivalent to 1. Abbreviation
+\noindent The \code{abbreviate} argument takes a vector of
+integers indicating the number of significant characters the levels should be
+abbreviated to (\code{TRUE} is interpreted as 1, obviously). Abbreviation
 is performed using {\sf R}'s \code{abbreviate()} function. Another
@@
 
-\begin{Sinput}
-> mosaic(Titanic, labeling_args = list(rot_labels = c(bottom = 45)))
-\end{Sinput}
+<<rotate>>=
+mosaic(Titanic, labeling_args = list(rot_labels = c(bottom = 90, right = 0)),
+                margins = c(right = 6, bottom = 4))
+@
 
-\setkeys{Gin}{width=0.7\textwidth}
 \begin{figure}[h]
 \begin{center}
-<<rotate,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-mosaic(Titanic, labeling_args = list(rot_labels = c(bottom = 45)))
+<<rotatefig,fig=TRUE,echo=FALSE>>=
+<<rotate>>
 @
@@
 \noindent Finally, we could also inhibit the output of repeated levels 
-for the `Survival' dimension (see Figure \ref{fig:repeat}):
+(see Figure \ref{fig:repeat}):
 
-\begin{Sinput}
-> mosaic(Titanic, labeling_args = list(rep = c(Survived = FALSE)))
-\end{Sinput}
+<<repeat>>=
+mosaic(Titanic, labeling_args = list(rep = c(Survived = FALSE, Age = FALSE)))
+@
 
-\setkeys{Gin}{width=0.7\textwidth}
 \begin{figure}[h]
 \begin{center}
-<<repeat,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-mosaic(Titanic, labeling_args = list(rep = c(Survived = FALSE)))
+<<repeatfig,fig=TRUE,echo=FALSE>>=
+<<repeat>>
 @
@@
 
-\begin{Sinput}
-> mosaic(Titanic, labeling_args = list(pos_varnames = "left", pos_labels = "left", 
+<<left>>=
+mosaic(Titanic, labeling_args = list(pos_varnames = "left", pos_labels = "left", 
        just_labels = "left", rep = FALSE))
-\end{Sinput}
+@
 
-\setkeys{Gin}{width=0.7\textwidth}
 \begin{figure}[h]
 \begin{center}
-<<left,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-mosaic(Titanic, labeling_args = list(pos_varnames = "left", pos_labels = "left", 
-       just_labels = "left", rep = FALSE))
+<<leftfig,fig=TRUE,echo=FALSE>>=
+<<left>>
 @
@@
 
-\begin{Sinput}
-> mosaic(Titanic, labeling_args = list(tl_labels = FALSE, tl_varnames = TRUE))
-\end{Sinput}
+<<margins>>=
+mosaic(Titanic, labeling_args = list(tl_labels = FALSE, tl_varnames = TRUE, abbreviate = c(Survived = 1, Age = 4)))
+@
 
-\setkeys{Gin}{width=0.7\textwidth}
 \begin{figure}[h]
 \begin{center}
-<<margins,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-mosaic(Titanic, labeling_args = list(tl_labels = FALSE, tl_varnames = TRUE))
+<<marginsfig,fig=TRUE,echo=FALSE>>=
+<<margins>>
 @
@@
 
-\noindent The \code{"tl\_foo"} (``top left'') arguments are \code{TRUE} by default. 
+\noindent The \var{tl\_foo} (``top left'') arguments are \code{TRUE} by default. 
 Now, we will add boxes to the labels and additionally
@@
 
-\begin{Sinput}
-> mosaic(Titanic, labeling_args = list(tl_labels = FALSE, tl_varnames = TRUE, 
+<<boxes>>=
+mosaic(Titanic, labeling_args = list(tl_labels = FALSE, tl_varnames = TRUE, 
        boxes = TRUE, clip = TRUE))
-\end{Sinput}
+@
 
-\setkeys{Gin}{width=0.7\textwidth}
 \begin{figure}[h]
 \begin{center}
-<<boxes,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-mosaic(Titanic, labeling_args = list(tl_labels = FALSE, tl_varnames = TRUE, 
-       boxes = TRUE, clip = TRUE))
+<<boxesfig,fig=TRUE,echo=FALSE>>=
+<<boxes>>
 @
@@
 
-\begin{Sinput}
-> mosaic(Titanic, labeling_args = list(tl_labels = TRUE, boxes = TRUE, 
-         clip = TRUE, labbl_varnames = TRUE), margins = c(left = 4, right = 1, 3))
-\end{Sinput}
+<<labbll>>=
+mosaic(Titanic, labeling_args = list(tl_labels = TRUE, boxes = TRUE, clip = TRUE, 
+       labbl_varnames = TRUE), margins = c(left = 4, right = 1, 3))
+@
 
-\setkeys{Gin}{width=0.7\textwidth}
 \begin{figure}[h]
 \begin{center}
-<<labbll,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-mosaic(Titanic, labeling_args = list(tl_labels = TRUE, boxes = TRUE, clip = TRUE, 
-       labbl_varnames = TRUE), margins = c(left = 4, right = 1, 3))
+<<labbllfig,fig=TRUE,echo=FALSE>>=
+<<labbll>>
 @
@@
 
-\section{Labels in the cells: \code{labeling\_cells()}}
+\section{Labels in the cells: \texttt{labeling\_cells()}}
 
@@
 
-\begin{Sinput}
-> mosaic(~ MaritalStatus + Gender, data = PreSex, labeling = labeling_cells)
-\end{Sinput}
+<<cell>>=
+mosaic(~ MaritalStatus + Gender, data = PreSex, labeling = labeling_cells)
+@
 
-\setkeys{Gin}{width=0.7\textwidth}
 \begin{figure}[h]
 \begin{center}
-<<cell,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-mosaic(~ MaritalStatus + Gender, data = PreSex, labeling = labeling_cells)
+<<cellfig,fig=TRUE,echo=FALSE>>=
+<<cell>>
 @
@@
 
-\begin{Sinput}
-> mosaic(~ PremaritalSex + ExtramaritalSex, data = PreSex, 
-         labeling = labeling_cells(abbreviate_labels = TRUE, 
-         abbreviate_varnames = TRUE, clip = F))
-\end{Sinput}
+<<cell2>>=
+mosaic(~ PremaritalSex + ExtramaritalSex, data = PreSex, 
+labeling = labeling_cells(abbreviate_labels = TRUE, abbreviate_varnames = TRUE, clip = FALSE))
+@
 
-\setkeys{Gin}{width=0.7\textwidth}
 \begin{figure}[h]
 \begin{center}
-<<cell2,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-mosaic(~ PremaritalSex + ExtramaritalSex, data = PreSex, labeling = labeling_cells(abbreviate_labels = TRUE, abbreviate_varnames = TRUE, clip = F))
+<<cell2fig,fig=TRUE,echo=FALSE>>=
+<<cell2>>
 @
@@
 
-\begin{Sinput}
-> mosaic(~ PremaritalSex + ExtramaritalSex | MaritalStatus + Gender, data = PreSex, 
-         labeling = labeling_conditional(abbreviate_varnames = TRUE, 
-         abbreviate_labels = TRUE))
-\end{Sinput}
+<<conditional>>=
+mosaic(~ PremaritalSex + ExtramaritalSex | MaritalStatus + Gender, data = PreSex, labeling = labeling_conditional(abbreviate_varnames = TRUE, abbreviate_labels = TRUE))
+@
 
-\setkeys{Gin}{width=0.7\textwidth}
 \begin{figure}[h]
 \begin{center}
-<<conditional,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-mosaic(~ PremaritalSex + ExtramaritalSex | MaritalStatus + Gender, data = PreSex, labeling = labeling_conditional(abbreviate_varnames = TRUE, abbreviate_labels = TRUE))
+<<conditionalfig,fig=TRUE,echo=FALSE>>=
+<<conditional>>
 @
@@
 
-\begin{Sinput}
-> mosaic(Titanic, pop = FALSE)
-
-> tab <- ifelse(Titanic < 6, NA, Titanic)
-> labeling_cells(text = tab, clip = FALSE)(Titanic)
-\end{Sinput}
-
-\setkeys{Gin}{width=0.7\textwidth}
-\begin{figure}[h]
-\begin{center}
-<<text,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-mosaic(Titanic, pop = FALSE)
+<<text>>=
+mosaic(Titanic, pop = FALSE, labeling_args = list(abbreviate = c(Survived = 1, Age = 4)))
 
@@
 labeling_cells(text = tab, clip = FALSE)(Titanic)
+@
+
+\begin{figure}[h]
+\begin{center}
+<<textfig,fig=TRUE,echo=FALSE>>=
+<<text>>
 @
@@
 
-\section{A simple list of labels:\code{labels\_list()}}
+\section{A simple list of labels:\texttt{labels\_list()}}
 
@@
 
-\begin{Sinput}
-> mosaic(Titanic, labeling = labeling_list, margins = c(bottom = 5))
-\end{Sinput}
+<<list>>=
+mosaic(Titanic, labeling = labeling_list, margins = c(bottom = 5))
+@
 
-\setkeys{Gin}{width=0.7\textwidth}
 \begin{figure}[h]
 \begin{center}
-<<list,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-mosaic(Titanic, labeling = labeling_list, margins = c(bottom = 5))
+<<listfig,fig=TRUE,echo=FALSE>>=
+<<list>>
 @

### 2005-08-24 trunk/vcd/inst/doc/strucplot.Rnw
New shading vignette added.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@437 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
viewport names changed to "obj:dim1=level1,...,dimn=leveln" syntax.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@444 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-\SweaveOpts{engine=R,eps=FALSE,prefix.string=pics/strucplot}
-\setkeys{Gin}{width=0.8\textwidth}
+\SweaveOpts{engine=R,eps=TRUE,height=6,width=7,results=hide,fig=FALSE,echo=TRUE}
+\setkeys{Gin}{width=0.7\textwidth}
 
@@
 
-<<structable,fig=FALSE,echo=TRUE,height=6,width=7,results=verbatim>>=
+<<structable,results=verbatim>>=
 structable(Titanic)
@@
 
-Let us
-first visualize the contingency table by the means of a mosaic plot
+Let us first visualize the contingency table by the means of a mosaic plot
 \citep{vcd:Hartigan+Kleiner:1984} which is basically 
@@
 dimensions of previous splits (see Figure \ref{fig:observed}):
-\begin{Sinput}
-> mosaic(Titanic, type = "expected", main = TRUE, sub = "Expected values")
-\end{Sinput}
+
+<<Observed>>=
+mosaic(Titanic, main = TRUE)
+@
 
@@
 
-\setkeys{Gin}{width=0.7\textwidth}
 \begin{figure}[h]
 \begin{center}
-<<Observed,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-mosaic(Titanic, main = TRUE)
+<<Observedfig,fig=TRUE,echo=FALSE>>=
+<<Observed>>
 @
@@
 observed values (see Figure \ref{fig:expected}):
-\begin{Sinput}
-> mosaic(Titanic, type = "expected", main = TRUE, sub = "Expected values")
-\end{Sinput}
+
+<<Expected>>=
+mosaic(Titanic, type = "expected", main = TRUE, sub = "Expected values")
+@
 
@@
 \begin{center}
-<<Expected,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-mosaic(Titanic, type = "expected", main = TRUE, sub = "Expected values")
+<<Expectedfig,fig=TRUE,echo=FALSE>>=
+<<Expected>>
 @
@@
 we can use a sieve plot (see Figure \ref{fig:sieve}):
-\begin{Sinput}
-> sieve(Titanic)
-\end{Sinput}
 
-\setkeys{Gin}{width=0.7\textwidth}
+<<sieve>>=
+sieve(Titanic)
+@
+
 \begin{figure}[h]
 \begin{center}
-<<sieve,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-sieve(Titanic)
+<<sievefig,fig=TRUE,echo=FALSE>>=
+<<sieve>>
 @
@@
 function scales to more than two variables (see Figure \ref{fig:residuals}):
-\begin{Sinput}
-> assoc(Titanic, compress = FALSE)
-\end{Sinput}
 
-\setkeys{Gin}{width=0.7\textwidth}
+<<Residuals>>=
+assoc(Titanic, compress = FALSE)
+@
+
 \begin{figure}[h]
 \begin{center}
-<<Residuals,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-assoc(Titanic, compress = FALSE)
+<<Residualsfig,fig=TRUE,echo=FALSE>>=
+<<Residuals>>
 @
@@
 each other:
-\begin{Sinput}
-> pushViewport(viewport(layout = grid.layout(ncol = 2)))
 
-> pushViewport(viewport(layout.pos.col = 1))
-> mosaic(~ Class + Age, data = Titanic[,"Male",,], 
-+        margin = c(bottom = 0), sub = "Male", newpage = FALSE)
-> popViewport()
+<<Variables>>=
+pushViewport(viewport(layout = grid.layout(ncol = 2)))
 
-> pushViewport(viewport(layout.pos.col = 2))
-> mosaic(~ Class + Age, data = Titanic[,"Female",,], 
-+        margin = c(bottom = 0), sub = "Female", newpage = FALSE)
-> popViewport(2)
-\end{Sinput}
+pushViewport(viewport(layout.pos.col = 1))
+mosaic(~ Class + Age, data = Titanic[,"Male",,], 
+       margin = c(bottom = 0, 0), sub = "Male", newpage = FALSE)
+popViewport()
+
+pushViewport(viewport(layout.pos.col = 2))
+mosaic(~ Class + Age, data = Titanic[,"Female",,], 
+       margin = c(bottom = 0, 0), sub = "Female", newpage = FALSE)
+popViewport(2)
+@
 
@@
 \begin{center}
-<<Variables,fig=TRUE,echo=FALSE,height=6,width=16,results=hide>>=
-pushViewport(viewport(layout = grid.layout(ncol = 2)))
-
-pushViewport(viewport(layout.pos.col = 1))
-mosaic(~ Class + Age, data = Titanic[,"Male",,], 
-       margin = c(bottom = 0), sub = "Male", newpage = FALSE,
-       labeling_args = list(gp_labels = gpar(fontsize = 24), 
-                            gp_varnames = gpar(fontsize = 24)),
-       title_gp = gpar(fontsize = 35))
-popViewport()
-
-pushViewport(viewport(layout.pos.col = 2))
-mosaic(~ Class + Age, data = Titanic[,"Female",,], 
-       margin = c(bottom = 0), sub = "Female", newpage = FALSE,
-       labeling_args = list(gp_labels = gpar(fontsize = 24), 
-                            gp_varnames = gpar(fontsize = 24)),
-       title_gp = gpar(fontsize = 35))
-popViewport(2)
+<<Variablesfig,fig=TRUE,echo=FALSE,height=6,width=6>>=
+<<Variables>>
 @
@@
 \end{figure}
+\setkeys{Gin}{width=0.7\textwidth}
 
@@
 
-\begin{Sinput}
-> mosaic(~ Class + Age | Sex, data = Titanic, split = TRUE)
-\end{Sinput}
+<<Conditioning>>=
+mosaic(~ Class + Age | Sex, data = Titanic, split = TRUE)
+@
 
-\setkeys{Gin}{width=0.7\textwidth}
 \begin{figure}[h]
-
 \begin{center}
-<<Conditioning,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-mosaic(~ Class + Age | Sex, data = Titanic, split = TRUE)
+<<Conditioningfig,fig=TRUE,echo=FALSE>>=
+<<Conditioning>>
 @
@@
 \code{"v"} characters (see Figures \ref{fig:split1} and \ref{fig:split2}):
-\begin{Sinput}
-> mosaic(Titanic, direction = "h")
-> mosaic(Titanic, split_vertical = c(TRUE, TRUE, TRUE, FALSE))
-\end{Sinput}
 
-\setkeys{Gin}{width=0.7\textwidth}
+<<split1>>=
+mosaic(Titanic, direction = "h")
+@
+
+<<split2>>=
+mosaic(Titanic, split_vertical = c(TRUE, TRUE, TRUE, FALSE),
+                labeling_args = list(abbreviate = c(Sex = TRUE, Age = TRUE)))
+@
+
 \begin{figure}[h]
 \begin{center}
-<<split1,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-mosaic(Titanic, direction = "h")
+<<split1fig,fig=TRUE,echo=FALSE>>=
+<<split1>>
 @
@@
 
-\setkeys{Gin}{width=0.7\textwidth}
 \begin{figure}[h]
 \begin{center}
-<<split2,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-mosaic(Titanic, split_vertical = c(TRUE, TRUE, TRUE, FALSE),
-                labeling_args = list(abbreviate = c(Sex = TRUE, Age = TRUE)))
+<<split2fig,fig=TRUE,echo=FALSE>>=
+<<split2>>
 @
@@
 Sex, given Class:
-\begin{Sinput}
-> doubledecker(Titanic)
-> ## equivalent to:
-> doubledecker(Survived ~ Class + Sex + Age, data = Titanic)
-\end{Sinput}
 
-\setkeys{Gin}{width=0.7\textwidth}
+<<doubledecker1>>=
+doubledecker(Titanic)
+@
+
+\noindent equivalent to:
+
+<<doubledecker2>>=
+doubledecker(Survived ~ Class + Sex + Age, data = Titanic)
+@
+
 \begin{figure}[h]
 \begin{center}
-<<doubledecker,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-doubledecker(Titanic)
+<<doubledeckerfig,fig=TRUE,echo=FALSE>>=
+<<doubledecker1>>
 @
@@
 
-\begin{Sinput}
-> pairs(Titanic, upper_panel = pairs_assoc)
-\end{Sinput}
+<<pairs>>=
+pairs(Titanic, upper_panel = pairs_assoc)
+@
 
@@
 \begin{center}
-<<pairs,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-pairs(Titanic, upper_panel = pairs_assoc)
+<<pairsfig,fig=TRUE,echo=FALSE>>=
+<<pairs>>
 @
@@
 
-\begin{Sinput}
-> cotabplot(~ Class + Survived | Sex + Age, data = Titanic)
-\end{Sinput}
+<<cotabplot>>=
+cotabplot(~ Class + Survived | Sex + Age, data = Titanic)
+@
 
@@
 \begin{center}
-<<cotabplot,fig=TRUE,echo=FALSE,height=6,width=12,results=hide>>=
-cotabplot(~ Class + Survived | Sex + Age, data = Titanic)
+<<cotabplotfig,fig=TRUE,echo=FALSE,height=6,width=6>>=
+<<cotabplot>>
 @
@@
 to be called by \code{strucplot()}. This returned function has access
-to all parameters passed to the generating function thanks to 
-lexical scope. The idea is to use module-specific `code
+to all parameters passed to the generating function because in {\sf
+  R}, nested functions are lexically scoped. The idea is to use module-specific `code
 factories' that generate parameterized workhorses rather than
@@
 
-\begin{Sinput}
-> strucplot(Titanic, panel = struc_mosaic(zero_size = 1))
-> strucplot(Titanic, panel = struc_mosaic, panel_args = list(zero_size = 1))
-> strucplot(Titanic, panel = myworkhorse)
-\end{Sinput}
+<<panelfuns>>=
+strucplot(Titanic, panel = struc_mosaic(zero_size = 1))
+strucplot(Titanic, panel = struc_mosaic, panel_args = list(zero_size = 1))
+@ 
+<<panelfuns2,eval=FALSE>>=
+strucplot(Titanic, panel = myworkhorse)
+@
 
@@
 
-\begin{Sinput}
-> mosaic(~ Sex + Class, data = Titanic, pop = FALSE)
-> current.vpTree() ## output not shown
+<<viewportnames>>=
+mosaic(~ Sex + Class, data = Titanic, pop = FALSE)
+current.vpTree() ## output not shown
 
-> seekViewport("cell..Sex.Male")
-> grid.rect(gp = gpar(col = "red", lwd = 4))
+seekViewport("cell:Sex=Male")
+grid.rect(gp = gpar(col = "red", lwd = 4))
 
-> seekViewport("cell..Sex.Male..Class.Crew")
-> grid.circle(r = 0.2, gp = gpar(fill = "cyan"))
-\end{Sinput}
+seekViewport("cell:Sex=Male,Class=Crew")
+grid.circle(r = 0.2, gp = gpar(fill = "cyan"))
+@
 
@@
 \begin{center}
-<<afterplot,fig=TRUE,echo=FALSE,height=6,width=6,results=hide>>=
-mosaic(~ Sex + Class, data = Titanic, pop = FALSE)
-
-seekViewport("cell..Sex.Male")
-grid.rect(gp = gpar(col = "red", lwd = 4))
-
-seekViewport("cell..Sex.Male..Class.Crew")
-grid.circle(r = 0.2, gp = gpar(fill = "cyan"))
+<<viewportnamesfig,fig=TRUE,echo=FALSE>>=
+<<viewportnames>>
 @
@@
 
-\begin{Sinput}
-> assoc(Titanic)
-> getNames() ## output not shown
-> grid.edit("rect..Class.1st..Age.Adult..Sex.Female..Survived.Yes", 
+<<changeplot>>=
+assoc(Titanic)
+getNames() ## output not shown
+grid.edit("rect:Class=1st,Age=Adult,Sex=Female,Survived=Yes", 
             gp = gpar(fill = "red"))
-\end{Sinput}
+@
 
-\setkeys{Gin}{width=0.6\textwidth}
 \begin{figure}[h]
 \begin{center}
-<<changeplot,fig=TRUE,echo=FALSE,height=6,width=6,results=hide>>=
+<<changeplotfig,fig=TRUE,echo=FALSE>>=
 assoc(Titanic)

### 2005-08-24 trunk/vcd/R/assoc.R

@@ -120,10 +120,10 @@ struc_assoc <- function(compress = TRUE, xlim = NULL, ylim = NULL,
       else
         grid.layout(nrow = 2 * d - 1, heights = dist[idx])
       vproot <- viewport(layout.pos.col = col, layout.pos.row = row,
-                         layout = layout, name = name)
+                         layout = layout, name = substr(name, 1, nchar(name) - 1))
 
       ## next level: either create further splits, or final viewports
-      name <- paste(name, "", dnn[i], dn[[i]], sep = ".")
+      name <- paste(name, dnn[i], "=", dn[[i]], ",", sep = "")
       rows <- cols <- rep.int(1, d)
       if (v) cols <- 2 * 1:d - 1 else rows <- 2 * 1:d - 1
 
@@ -135,11 +135,13 @@ struc_assoc <- function(compress = TRUE, xlim = NULL, ylim = NULL,
       } else {
         if (v)
           function(m) viewport(layout.pos.col = cols[m], layout.pos.row = rows[m],
-                               name = name[m], yscale = res[,1],
+                               name = substr(name[m], 1, nchar(name[m]) - 1),
+                               yscale = res[,1],
                                xscale = sexp[,m], default.units = "null")
         else
           function(m) viewport(layout.pos.col = cols[m], layout.pos.row = rows[m],
-                               name = name[m], yscale = res[,m],
+                               name = substr(name[m], 1, nchar(name[m]) - 1),
+                               yscale = res[,m],
                                xscale = sexp[,1], default.units = "null")
       }
       vpleaves <- structure(lapply(1:d, f), class = c("vpList", "viewport"))
@@ -148,15 +150,15 @@ struc_assoc <- function(compress = TRUE, xlim = NULL, ylim = NULL,
     }
 
     ## start spltting on top, creates viewport-tree
-    pushViewport(split(ylim, xlim, i = 1, name = "cell", row = 1, col = 1))
+    pushViewport(split(ylim, xlim, i = 1, name = "cell:", row = 1, col = 1))
 
     ## draw tiles
     mnames <- paste(apply(expand.grid(dn), 1,
-                          function(i) paste(dnn, i, collapse="..", sep = ".")
+                          function(i) paste(dnn, i, collapse = ",", sep = "=")
                           )
                     )
     for (i in seq(along = mnames)) {
-      seekViewport(paste("cell", mnames[i], sep = ".."))
[…]

### 2005-08-24 trunk/vcd/R/mosaic.R

@@ -108,17 +108,17 @@ struc_mosaic <- function(zero_size = 0.5)
       else
         grid.layout(nrow = 2 * d - 1, heights = dist[idx])
       vproot <- viewport(layout.pos.col = col, layout.pos.row = row,
-                         layout = layout, name = name)
+                         layout = layout, name = substr(name, 1, nchar(name) - 1))
       
       ## next level: either create further splits, or final viewports
-      name <- paste(name, "", dnn[i], dn[[i]], sep = ".")
+      name <- paste(name, dnn[i], "=", dn[[i]], ",", sep = "")
       row <- col <- rep.int(1, d)
       if (v) col <- 2 * 1:d - 1 else row <- 2 * 1:d - 1
       f <- if (i < dl) 
         function(m) split(cotab[[m]], i + 1, name[m], row[m], col[m])
       else
         function(m) viewport(layout.pos.col = col[m], layout.pos.row = row[m],
-                             name = name[m])
+                             name = substr(name[m], 1, nchar(name[m]) - 1))
       vpleaves <- structure(lapply(1:d, f), class = c("vpList", "viewport"))
 
       vpTree(vproot, vpleaves)
@@ -126,28 +126,28 @@ struc_mosaic <- function(zero_size = 0.5)
 
     ## start spltting on top, creates viewport-tree
     pushViewport(split(observed + .Machine$double.eps,
-                       i = 1, name = "cell", row = 1, col = 1))
+                       i = 1, name = "cell:", row = 1, col = 1))
 
     ## draw rectangles
     mnames <-  apply(expand.grid(dn), 1,
-                     function(i) paste(dnn, i, collapse="..", sep = ".")
+                     function(i) paste(dnn, i, collapse=",", sep = "=")
                      )
     zeros <- observed <= .Machine$double.eps
 
     for (i in seq(along = mnames)) {
-      seekViewport(paste("cell", mnames[i], sep = ".."))
+      seekViewport(paste("cell:", mnames[i], sep = ""))
       gpobj <- structure(lapply(gp, function(x) x[i]), class = "gpar")
       if (!zeros[i]) {
-        grid.rect(gp = gpobj, name = paste("rect", mnames[i], sep = ".."))
+        grid.rect(gp = gpobj, name = paste("rect:", mnames[i], sep = ""))
       } else { 
         grid.lines(x = 0.5, gp = gpobj)
         grid.lines(y = 0.5, gp = gpobj)
         if (zero_size > 0) {
[…]

### 2005-08-24 trunk/vcd/R/strucplot.R

@@ -209,14 +209,14 @@ vcdViewport <- function(mar = rep.int(2.5, 4),
   vpPlot <- vpStack(viewport(layout.pos.col = 2, layout.pos.row = 2),
                     viewport(width = 1, height = 1, name = "plot",
                              default.units = if (keep_aspect_ratio) "snpc" else "npc"))
-  vpMarginBottom <- viewport(layout.pos.col = 2, layout.pos.row = 3, name = "marginBottom")
-  vpMarginLeft <- viewport(layout.pos.col = 1, layout.pos.row = 2, name = "marginLeft")
-  vpMarginTop <- viewport(layout.pos.col = 2, layout.pos.row = 1, name = "marginTop")
-  vpMarginRight <- viewport(layout.pos.col = 3, layout.pos.row = 2, name = "marginRight")
-  vpCornerTL <- viewport(layout.pos.col = 1, layout.pos.row = 1, name = "cornerTL")
-  vpCornerTR <- viewport(layout.pos.col = 3, layout.pos.row = 1, name = "cornerTR")
-  vpCornerBL <- viewport(layout.pos.col = 1, layout.pos.row = 3, name = "cornerBL")
-  vpCornerBR <- viewport(layout.pos.col = 3, layout.pos.row = 3, name = "cornerBR")
+  vpMarginBottom <- viewport(layout.pos.col = 2, layout.pos.row = 3, name = "margin_bottom")
+  vpMarginLeft <- viewport(layout.pos.col = 1, layout.pos.row = 2, name = "margin_left")
+  vpMarginTop <- viewport(layout.pos.col = 2, layout.pos.row = 1, name = "margin_top")
+  vpMarginRight <- viewport(layout.pos.col = 3, layout.pos.row = 2, name = "margin_right")
+  vpCornerTL <- viewport(layout.pos.col = 1, layout.pos.row = 1, name = "corner_top_left")
+  vpCornerTR <- viewport(layout.pos.col = 3, layout.pos.row = 1, name = "corner_top_right")
+  vpCornerBL <- viewport(layout.pos.col = 1, layout.pos.row = 3, name = "corner_bottom_left")
+  vpCornerBR <- viewport(layout.pos.col = 3, layout.pos.row = 3, name = "corner_bottom_right")
 
   if (legend) {
     vpLegend <- viewport(layout.pos.col = 4, layout.pos.row = 2, name = "legend")

### 2005-08-24 trunk/vcd/inst/doc/shading.Rnw
viewport names changed to "obj:dim1=level1,...,dimn=leveln" syntax.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@444 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-\SweaveOpts{engine=R,eps=TRUE}
-\setkeys{Gin}{width=0.8\textwidth}
+\SweaveOpts{engine=R,eps=TRUE,height=6,width=7,results=hide,fig=FALSE,echo=TRUE}
+\setkeys{Gin}{width=0.7\textwidth}
 
 <<preliminaries,echo=FALSE,results=hide>>=
+options(width=90)
 set.seed(1071)
@@
 
-As an example, consider the Admissions to the University of California
-in Berkely (UCB) data. In the table collapsed over departments, we
+As an example, consider the `Admissions to the University of California
+in Berkely (UCB)' data. In the table collapsed over departments, we
 would like to highlight the (incidentally wrong) impression that there
@@
 
-<<ucb,fig=FALSE,echo=TRUE,height=6,width=7,results=verbatim>>=
-ucb <- margin.table(UCBAdmissions, 1:2)
-ucb
+<<ucb,results=verbatim>>=
+(ucb <- margin.table(UCBAdmissions, 1:2))
 
-fill_colors <- matrix(c("dark cyan","gray","gray","dark magenta"), ncol = 2)
-fill_colors
+(fill_colors <- matrix(c("dark cyan","gray","gray","dark magenta"), ncol = 2))
 
@@
 
-\setkeys{Gin}{width=0.7\textwidth}
 \begin{figure}[h]
 \begin{center}
-<<ucbfig,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-mosaic(ucb, gp = gpar(fill = fill_colors, col = 0))
+<<ucbfig,fig=TRUE,echo=FALSE>>=
+<<ucb>>
 @
@@
 
-\noindent As the example shows, we therefore create a fourfold table with appropriate colors and supply them to the \code{fill} component of the \code{gpar} object passed to the \code{gp} argument of \code{mosaic()}.
-In addition, we use the \code{col} component to turn off the borders for cosmetic reasons.
+\noindent As the example shows, we therefore create a fourfold table 
+with appropriate colors and supply them to the \code{fill} component 
+of the \code{gpar} object passed to the \code{gp} argument of \code{mosaic()}. 
+In addition, we use the \code{col} component to turn off the borders 
+for cosmetic reasons.
 
@@
 strucplot functions will recycle them along the last splitting
-dimension. In the following example based on the Survival on the
-Titanic data, we will highlight all cells corresponding to survived
+dimension. In the following example based on the `Survival on the
+Titanic' data, we will highlight all cells corresponding to survived
 passengers (see Figure \ref{fig:recycling}):
 
-\begin{Sinput}
-> mosaic(Titanic, gp = gpar(fill = c("gray","dark magenta")))
-\end{Sinput}
+<<recycling>>=
+mosaic(Titanic, gp = gpar(fill = c("gray","dark magenta")))
+@
 
-\setkeys{Gin}{width=0.7\textwidth}
 \begin{figure}[h]
 \begin{center}
-<<recycling,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-mosaic(Titanic, gp = gpar(fill = c("gray","dark magenta")))
+<<recyclingfig,fig=TRUE,echo=FALSE>>=
+<<recycling>>
 @
@@
 
-\begin{Sinput}
-> obs <- margin.table(UCBAdmissions, 1:2)
-> exp <- independence_table(obs)
-> res <- (obs - exp) / sqrt(exp)
-> res
-          Gender
-Admit           Male    Female
-  Admitted  4.784093 -5.793466
-  Rejected -3.807325  4.610614
+<<binary,results=verbatim>>=
+obs <- margin.table(UCBAdmissions, 1:2)
+exp <- independence_table(obs)
+(res <- (obs - exp) / sqrt(exp))
 
-> gpobj <- shading_binary()(res)
-> gpobj
-$fill
-     [,1]      [,2]     
-[1,] "#96A2DF" "#E18E9E"
-[2,] "#E18E9E" "#96A2DF"
+(gpobj <- shading_binary()(res))
 
-> mosaic(obs, gp = gpobj)
-\end{Sinput}
+mosaic(obs, gp = gpobj)
+@
 
-\setkeys{Gin}{width=0.7\textwidth}
 \begin{figure}[h]
 \begin{center}
-<<binary,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-obs <- margin.table(UCBAdmissions, 1:2)
-exp <- independence_table(obs)
-res <- (obs - exp) / sqrt(exp)
-gpobj <- shading_binary()(res)
-mosaic(obs, gp = gpobj)
+<<binaryfig,fig=TRUE,echo=FALSE>>=
+<<binary>>
 @
@@
 
-\begin{Sinput}
-> mosaic(obs, gp = shading_binary)
-\end{Sinput}
+<<binaryfun>>=
+mosaic(obs, gp = shading_binary)
+@
 
@@
 As an example, we will visualize the (in)dependence of hair and eye
-color in the HairEyeColor data set (see Figure \ref{fig:haireye})
+color in the `Hair and Eye Color' data set (see Figure \ref{fig:haireye})
 
-\begin{Sinput}
-> haireye <- margin.table(HairEyeColor, 1:2)
-> mosaic(haireye, shade = TRUE)
-\end{Sinput}
+<<haireye>>=
+haireye <- margin.table(HairEyeColor, 1:2)
+mosaic(haireye, shade = TRUE)
+@
 
-\setkeys{Gin}{width=0.7\textwidth}
 \begin{figure}[h]
 \begin{center}
-<<haireye,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-haireye <- margin.table(HairEyeColor, 1:2)
-mosaic(haireye, shade = TRUE)
+<<haireyefig,fig=TRUE,echo=FALSE>>=
+<<haireye>>
 @
@@
 
-\begin{Sinput}
-> mosaic(haireye, shade = TRUE, 
+<<interpolate>>=
+mosaic(haireye, shade = TRUE, 
        gp_args = list(interpolate = 1:4, h = c(100, 200)))
-\end{Sinput}
+@
 
-\setkeys{Gin}{width=0.7\textwidth}
 \begin{figure}[h]
 \begin{center}
-<<interpolate,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-mosaic(haireye, shade = TRUE, 
-       gp_args = list(interpolate = 1:4, h = c(100, 200)))
+<<interpolatefig,fig=TRUE,echo=FALSE>>=
+<<interpolate>>
 @
@@
 the two luminance values supplied to the \code{l} argument. The
-following example shows the relationship of home goals and away
-goals of the German Bundesliga in 1995: although there are two
+following example using the `Bundesliga' data 
+shows the relationship of home goals and away
+goals of the German football league in 1995: although there are two
 ``larger'' residuals (one greater than 2, one less then $-2$), the
@@
 
-\begin{Sinput}
-> bl <- xtabs(~ HomeGoals + AwayGoals, data = Bundesliga, subset = Year == 1995)
-> mosaic(bl, shade = TRUE)
-\end{Sinput}
+<<bundesliga>>=
+bl <- xtabs(~ HomeGoals + AwayGoals, data = Bundesliga, subset = Year == 1995)
+mosaic(bl, shade = TRUE)
+@
 
-\setkeys{Gin}{width=0.7\textwidth}
 \begin{figure}[h]
 \begin{center}
-<<bundesliga,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-bl <- xtabs(~ HomeGoals + AwayGoals, data = Bundesliga, subset = Year == 1995)
-mosaic(bl, shade = TRUE)
+<<bundesligafig,fig=TRUE,echo=FALSE>>=
+<<bundesliga>>
 @
@@
 
-\begin{Sinput}
-> art <- xtabs(~ Improved + Treatment, data = Arthritis, subset = Sex == "Female")
-> mosaic(art, shade = TRUE, gp = shading_max)
-\end{Sinput}
+<<arthritis>>=
+art <- xtabs(~ Improved + Treatment, data = Arthritis, subset = Sex == "Female")
+mosaic(art, gp = shading_max)
+@
 
-\setkeys{Gin}{width=0.7\textwidth}
 \begin{figure}[h]
 \begin{center}
-<<arthritis,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-art <- xtabs(~ Improved + Treatment, data = Arthritis, subset = Sex == "Female")
-mosaic(art, gp = shading_max)
+<<arthritisfig,fig=TRUE,echo=FALSE>>=
+<<arthritis>>
 @
@@
 
-\begin{Sinput}
-> mosaic(bl, gp = shading_Friendly, legend = legend_fixed, zero_size = 0)
-\end{Sinput}
+<<friendly>>=
+mosaic(bl, gp = shading_Friendly, legend = legend_fixed, zero_size = 0)
+@
 
-\setkeys{Gin}{width=0.7\textwidth}
 \begin{figure}[h]
 \begin{center}
-<<friendly,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-mosaic(bl, gp = shading_Friendly, legend = legend_fixed, zero_size = 0)
+<<friendlyfig,fig=TRUE,echo=FALSE>>=
+<<friendly>>
 @
@@
 
-\begin{Sinput}
+<<continuous>>=
 shading_continuous <- function(observed = NULL, residuals = NULL,
@@
 mosaic(haireye, gp = shading_continuous)
-\end{Sinput}
+@
 
-\setkeys{Gin}{width=0.7\textwidth}
 \begin{figure}[h]
 \begin{center}
-<<continuous,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-shading_continuous <- function(observed = NULL, residuals = NULL,
-                               expected = NULL, df = NULL, n = 5000, ...)
-{
-  ## function works only for two-way tables
-  stopifnot(length(dim(observed)) == 2)
-
-  ## sample permutation distribution of the maximum statistic
-  obs.test <- coindep_test(observed, n = n)
-
-  ## Use distribution to map residuals to [0,1] range
-  ipol <- function(x) abs(obs.test$pdist(x))
-
-  ## call HCL shading 
-  shading_hcl(observed = NULL, residuals = NULL, expected = NULL,
-              df = NULL, interpolate = ipol, p.value = obs.test$p.value, ...)
-}
-class(shading_continuous) <- "panel_generator"
-
-mosaic(haireye, gp = shading_continuous)
+<<continuousfig,fig=TRUE,echo=FALSE>>=
+<<continuous>>
 @

### 2005-08-24 trunk/vcd/R/labeling.R

@@ -33,7 +33,7 @@ labeling_list <- function(gp = gpar(),
       d <- dimnames(d)
     ld <- length(d)
     labeling_border(labels = FALSE, varnames = varnames)(d, split_vertical, condvars)
-    seekViewport("marginBottom")
+    seekViewport("margin_bottom")
     pos <- unit(switch(pos, left = 0, center = 0.5, 1) / cols, "npc")
     ind <- split(seq(ld), rep.int(seq(cols), ceiling(ld / cols))[seq(ld)])
     
@@ -97,8 +97,8 @@ labeling_cells <- function(labels = TRUE, varnames = TRUE,
       for (labind in seq(along = n)) {
         lab <- c(labs, n[labind])
         names(lab) <- names(d)[1:vind]
-        mlab <- paste("cell", paste(dn[1:vind], lab, sep = ".", collapse = ".."),
-                      sep = "..")
+        mlab <- paste("cell:", paste(dn[1:vind], lab, sep = "=", collapse = ","),
+                      sep = "")
 
         if (vind < ld)
           split(vind + 1, lab)
@@ -369,7 +369,7 @@ labeling_border <- function(labels = TRUE, varnames = labels,
     }
 
     ## draw labels
-    split <- function(vind = 1, root = "cell",
+    split <- function(vind = 1, root = "cell:",
                       left = TRUE, right = TRUE, top = TRUE, bottom = TRUE) {
       n <- d[[vind]]
       vl <- length(n)
@@ -378,7 +378,7 @@ labeling_border <- function(labels = TRUE, varnames = labels,
       if (!sp) labseq <- rev(labseq)
       
       for (labind in labseq) {
-        mlab <- paste(root, "", dn[vind], n[labind], sep = ".")
+        mlab <- paste(root, dn[vind], "=", n[labind], sep = "")
         if (labels[vind] && (rep[vind] || !printed[[vind]][labind])) {
           lab <- if (!is.null(set_labels) && !is.null(set_labels[[dn[vind]]]))
             set_labels[[dn[vind]]][labind]
@@ -485,7 +485,7 @@ labeling_border <- function(labels = TRUE, varnames = labels,
           }
         }
         
-        if (vind < ld) Recall(vind + 1, mlab,
+        if (vind < ld) Recall(vind + 1, paste(mlab, ",", sep = ""),
                               if (sp) left && labind == 1 else left,
                               if (sp) right && labind == vl else right,
[…]

### 2005-08-24 trunk/vcd/R/sieve.R

@@ -120,10 +120,10 @@ struc_sieve <- function(sievetype = c("observed", "expected")) {
       else
         grid.layout(nrow = 2 * d - 1, heights = dist[idx])
       vproot <- viewport(layout.pos.col = col, layout.pos.row = row,
-                         layout = layout, name = name)
+                         layout = layout, name = substr(name, 1, nchar(name) - 1))
       
       ## next level: either create further splits, or final viewports
-      name <- paste(name, "", dnn[i], dn[[i]], sep = ".")
+      name <- paste(name, dnn[i], "=", dn[[i]], ",", sep = "")
       row <- col <- rep.int(1, d)
       if (v) col <- 2 * 1:d - 1 else row <- 2 * 1:d - 1
       proptab <- function(x) x / max(sum(x), 1)
@@ -139,12 +139,12 @@ struc_sieve <- function(sievetype = c("observed", "expected")) {
       } else {
         if (v)
           function(m) viewport(layout.pos.col = col[m], layout.pos.row = row[m],
-                               name = name[m],
+                               name = substr(name[m], 1, nchar(name[m]) - 1),
                                yscale = c(0, rowmargin),
                                xscale = c(0, colmargin * proptab(margin)[m]))
         else
           function(m) viewport(layout.pos.col = col[m], layout.pos.row = row[m],
-                               name = name[m],
+                               name = substr(name[m], 1, nchar(name[m]) - 1),
                                yscale = c(0, rowmargin * proptab(margin)[m]),
                                xscale = c(0, colmargin))
       }
@@ -155,16 +155,16 @@ struc_sieve <- function(sievetype = c("observed", "expected")) {
 
     ## start splitting on top, creates viewport-tree
     pushViewport(split(expected + .Machine$double.eps,
-                       i = 1, name = "cell", row = 1, col = 1,
+                       i = 1, name = "cell:", row = 1, col = 1,
                        rowmargin = n, colmargin = n))
 
     ## draw rectangles
     mnames <- apply(expand.grid(dn), 1,
-                    function(i) paste(dnn, i, collapse="..", sep = ".")
+                    function(i) paste(dnn, i, collapse=",", sep = "=")
                     )
     
     for (i in seq(along = mnames)) {
-      seekViewport(paste("cell", mnames[i], sep = ".."))
+      seekViewport(paste("cell:", mnames[i], sep = ""))
[…]

### 2005-08-25 trunk/vcd/paper/JCGS.Rnw
JCGS paper finished
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@446 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \usepackage{graphicx,color,a4wide,thumbpdf,hyperref}
-
-%% \usepackage{Sweave} is essentially
-\RequirePackage[T1]{fontenc}
-\RequirePackage{ae,fancyvrb}
-\DefineVerbatimEnvironment{Sinput}{Verbatim}{fontshape=sl}
-\DefineVerbatimEnvironment{Soutput}{Verbatim}{}
-\DefineVerbatimEnvironment{Scode}{Verbatim}{fontshape=sl}
-\newenvironment{Schunk}{}{}
+%% need no \usepackage{Sweave}
 
@@
   %% change colorlinks to false for pretty printing
-  colorlinks = {true},
+  colorlinks = {false},
   linkcolor = {Blue},
@@
 in two directions: (a)~perceptually uniform HCL colors are used and (b)~the
-result of an associated significance test is coded by the presence of
+result of an associated significance test is coded by the appearance of
 color in the visualization. For obtaining (a), a general strategy for
@@
 in 2-way tables and then extended to more general independence models
-for multi-way tables, including in particular the conditional independence
-problem.\\[0.3cm]
+for multi-way tables, including in particular conditional independence
+problems.\\[0.3cm]
 
@@
 are log-linear models \citep[see, e.g.,][]{vcd:Agresti:2002} for modeling
-independence and mosaic plots \citep{vcd:Hartigan+Kleiner:1981} for 
-bringing it out graphically. Both methods can also be combined such that
+(in)dependence structures and mosaic plots \citep{vcd:Hartigan+Kleiner:1981} for 
+bringing them out graphically. Both methods can also be combined such that
 a certain mosaic plot visualizes a particular log-linear model by 
@@
 model of a contingency table more tightly than in previous approaches, the residual-based
-shading is extended such that presence of color in the display is equivalent to significance
+shading is extended such that appearance of color in the display is equivalent to significance
 of the associated independence test. This is achieved by using data-driven cut offs
@@
 multi-way tables in Section~\ref{sec:3way} with a more detailed discussion of adapting
-them to conditional independence. Section~\ref{sec:conclusions} summarizes the paper
+them to conditional independence problems. Section~\ref{sec:conclusions} summarizes the paper
 and gives some concluding remarks.
@@
 This is the well-known $\chi^2$ test which is typically introduced first in statistics
-textbooks if they address the independence problem in 2-way tables
+textbooks when addressing the independence problem in 2-way tables
 \citep[see, e.g.,][]{vcd:Agresti:2002}.
@@
 However, the sum of squares is not the only plausible way of capturing deviations
-from $0$ in the residuals.
+from zero in the residuals.
 There are many other conceivable functionals $\lambda(\cdot)$ which
 lead to reasonable test statistics $\lambda([r_{ij}])$: one which is particularly
-suitable for identifying the cells responsible for the dependence (if any) is
-the maximum of the absolute values
+suitable for identifying the cells responsible for the `dependence' (i.e., significant
+departure from independence), if any, is the maximum of the absolute values
 \begin{equation} \label{eq:MaxAbs}
@@
 Given a critical value $c_\alpha$ for this test statistic, all residuals whose
-absolute value exceeds $c_\alpha$ violate the hypothesis of indendence at significance level
+absolute value exceeds $c_\alpha$ violate the null hypothesis of indendence at significance level
 $\alpha$ \cite[][ch.~7]{vcd:Mazanec+Strasser:2000}. Thus, the interesting cells responsible
@@
 Furthermore, an important reason for using the unconditional limiting distribution
-for the $X^2$ statistic from Equation~(\ref{eq:ChiSquare}) was the closed form result
+for the $X^2$ statistic from Equation~\ref{eq:ChiSquare} was the closed form result
 for the distribution. Recently, with the improving perfomance of computers,
@@
 increasing attention \citep[e.g.,][]{vcd:Ernst:2004,vcd:Pesarin:2001,vcd:Strasser+Weber:1999}. 
-For testing the independence hypothesis from Equation~(\ref{eq:H0}), using a permutation test is particularly
+For testing the independence hypothesis from Equation~\ref{eq:H0}, using a permutation test is particularly
 intuitive due to the permutation invariance (given row and column sums) of this problem.
@@
 For the arthritis data from Table~\ref{tab:art}, both tests indicate a clearly significant
-dependence of improvement on treatment: the sum-of-squares statistic from Equation~(\ref{eq:ChiSquare})
+dependence of improvement on treatment: the sum-of-squares statistic from Equation~\ref{eq:ChiSquare}
 is $X^2 = \Sexpr{stat(art_chisq)}$ with a $p$ value of $p = \Sexpr{pval(art_chisq)}$,
-and the maximum statistic from Equation~(\ref{eq:MaxAbs}) is $M = \Sexpr{stat(art_max)}$
+and the maximum statistic from Equation~\ref{eq:MaxAbs} is $M = \Sexpr{stat(art_max)}$
 with $p = \Sexpr{pval(art_max)}$. Both $p$ values have been computed
@@
 
+\clearpage
+
 \subsection{Colors} \label{sec:colors}
@@
 uses colors from a palette based on HLS (Hue-Luminance-Saturation) color space.
-The implementation in the \proglang{R} system for statistical computing and graphics
+The implementation in the standard packages of the
+\proglang{R} system for statistical computing and graphics
 \citep{vcd:R:2005} employs colors from HSV (Hue-Saturation-Value) space. 
@@
 The Friendly shading can be interpreted to be a visualization of the maximum
-statistic $M$ from Equation~(\ref{eq:MaxAbs}) which always employs the critical values $c_\alpha$
-$2$ and $4$. The problem is that it is not clear to which
+statistic $M$ from Equation~\ref{eq:MaxAbs} which always employs the critical values $c_\alpha$
+$2$ and $4$. However, it is not clear to which
 significance levels $\alpha$ these critical values correspond because the distribution
@@
 a discrete shading with few colors is much easier to interpret.
+
+\setkeys{Gin}{width=\textwidth}
+\begin{figure}[t!]
+\begin{center}
+<<shadings,fig=TRUE,echo=FALSE,height=6,width=11,results=hide>>=
+grid.newpage()
+pushViewport(viewport(layout = grid.layout(1, 2)))
+pushViewport(viewport(layout.pos.col=1, layout.pos.row=1))
+mosaic(art, newpage = FALSE, gp = shading_Friendly, margins = c(2.5, 4, 2.5, 3), gp_args = list(eps = NULL, lty = 1))
+popViewport()
+pushViewport(viewport(layout.pos.col=2, layout.pos.row=1))
+mosaic(art, newpage = FALSE, gp = art_gp, margins = c(2.5, 3, 2.5, 4))
+popViewport(2)
+@
+\caption{Mosaic plot for the arthritis data with Friendly shading (left) and maximum shading (right).}
+\label{fig:shadings}
+\end{center}
+\end{figure}
 
@@
 
-\setkeys{Gin}{width=\textwidth}
-\begin{figure}[htbp]
-\begin{center}
-<<shadings,fig=TRUE,echo=FALSE,height=6,width=11,results=hide>>=
-grid.newpage()
-pushViewport(viewport(layout = grid.layout(1, 2)))
-pushViewport(viewport(layout.pos.col=1, layout.pos.row=1))
-mosaic(art, newpage = FALSE, gp = shading_Friendly, margins = c(2.5, 4, 2.5, 3), gp_args = list(eps = NULL, lty = 1))
-popViewport()
-pushViewport(viewport(layout.pos.col=2, layout.pos.row=1))
-mosaic(art, newpage = FALSE, gp = art_gp, margins = c(2.5, 3, 2.5, 4))
-popViewport(2)
-@
-\caption{Mosaic plot for the arthritis data with Friendly shading (left) and maximum shading (right).}
-\label{fig:shadings}
-\end{center}
-\end{figure}
-
-\clearpage
-
-\section{Extension to Conditional Independence} \label{sec:3way}
+\section{Extensions} \label{sec:3way}
 
@@
 a suitable layout of mosaic or association plots. Specific strategies for the conditional
-independence problem are derived in the following.
+independence problem in 3-way tables are derived in the following.
 
@@
 
-Alternatively, a maximum statistic similar to that from Equation~(\ref{eq:MaxAbs})
+Alternatively, a maximum statistic similar to that from Equation~\ref{eq:MaxAbs}
 can be constructed
@@
 the corresponding cells can also be identified. Hence, the double maximum
-statistic from Equation~(\ref{eq:doublemax}) is the only functional allowing for detection of
+statistic from Equation~\ref{eq:doublemax} is the only functional allowing for detection of
 both the strata and the cells violating the conditional independence hypothesis.
@@
 the significant or the non-significant shading for all cells in the contingency table---this
-strategy would have to be used for the sum-of-squares statistic from Equation~(\ref{eq:sumChisq}). If
-$\lambda_\mathrm{agg}$ is the maximum as in Equation~(\ref{eq:maxChisq}),
+strategy would have to be used for the sum-of-squares statistic from Equation~\ref{eq:sumChisq}. If
+$\lambda_\mathrm{agg}$ is the maximum as in Equation~\ref{eq:maxChisq},
 then the significant palette would only be used
@@
 can be pursued, i.e., only the significant palette is used but with data-driven cut offs
-derived from the distribution of the double maximum statistic from Equation~(\ref{eq:doublemax}).
+derived from the distribution of the double maximum statistic from Equation~\ref{eq:doublemax}.
 
@@
 from conditional independence is picked up by all three statistics formulated
-above: the sum-of-squares statistic from Equation~(\ref{eq:sumChisq}) is
+above: the sum-of-squares statistic from Equation~\ref{eq:sumChisq} is
 $\sum_k X_k^2 = \Sexpr{stat(ucb_sumChisq)}$ ($p = \Sexpr{pval(ucb_sumChisq)}$),
-the maximum sum-of-squares statistic from Equation~(\ref{eq:maxChisq}) is
+the maximum sum-of-squares statistic from Equation~\ref{eq:maxChisq} is
 $\max_k X_k^2 = \Sexpr{stat(ucb_maxChisq)}$ ($p = \Sexpr{pval(ucb_maxChisq)}$),
-and the double maximum statistic from Equation~(\ref{eq:doublemax}) is
+and the double maximum statistic from Equation~\ref{eq:doublemax} is
 $\max_k M_k = \Sexpr{stat(ucb_doublemax)}$ ($p = \Sexpr{pval(ucb_doublemax)}$).
@@
 {\pkg{MASS}~\Sexpr{packageDescription("MASS")$Version}} \citep{vcd:Venables+Ripley:2002},
-{\pkg{grid}~{\Sexpr{packageDescription("grid")$Version}}} \citep{vcd:Murrell:2002} and
+{\pkg{grid}~{\Sexpr{packageDescription("grid")$Version}}} \citep[see][]{vcd:Murrell:2002} and
 {\pkg{colorspace}~{\Sexpr{packageDescription("colorspace")$Version}}} \citep{vcd:Ihaka:2004}.
+
+\clearpage
 

### 2005-08-30 trunk/vcd/inst/doc/strucplot.Rnw
mosaic,assoc,and sieve now also accept structable objects. as.table.structable and plot.structable added.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@452 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 <<changeplot>>=
-assoc(Titanic)
+assoc(Titanic, pop = FALSE)
 getNames() ## output not shown
-grid.edit("rect:Class=1st,Age=Adult,Sex=Female,Survived=Yes", 
+grid.edit("rect:Class=1st,Sex=Female,Age=Adult,Survived=Yes", 
             gp = gpar(fill = "red"))

### 2005-08-30 trunk/vcd/R/mosaic.R

@@ -54,7 +54,9 @@ mosaic.default <- function(x, condvars = NULL,
                            zero_size = 0.5, main = NULL, ...) {
   if (is.logical(main) && main)
     main <- deparse(substitute(x))
-
+  if (inherits(x, "structable"))
+    x <- as.table(x)
+  
   dl <- length(dim(x))
   if (!is.null(condvars)) {
     if (is.character(condvars))

### 2005-08-30 trunk/vcd/R/sieve.R

@@ -62,6 +62,9 @@ sieve.default <- function(x, condvars = NULL, gp = NULL,
     else
       shading_Friendly(interpolate = 0, line_col = "darkgray", eps = Inf, lty = "dotted")
 
+  if (inherits(x, "structable"))
+    x <- as.table(x)
+  
   dl <- length(dim(x))
   if (!is.null(condvars)) {
     if (is.character(condvars))

### 2005-08-30 trunk/vcd/inst/doc/labeling.Rnw
Try better to have \code not hyphenate.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@445 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
viewport name example in strucplot() fixed.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@451 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
mosaic,assoc,and sieve now also accept structable objects. as.table.structable and plot.structable added.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@452 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \newcommand{\var}[1]{\textit{\texttt{#1}}}
-\newenvironment{smallexample}{\begin{alltt}}{\end{alltt}}
+\newcommand{\email}[1]{\href{mailto:#1}{\normalfont\texttt{#1}}}
 
 %% \code without `-' ligatures
+\def\nohyphenation{\hyphenchar\font=-1 \aftergroup\restorehyphenation}
+\def\restorehyphenation{\hyphenchar\font=`-}
 {\catcode`\-=\active%
@@
 \def\codedash{-\discretionary{}{}{}}
-\def\Rd@code#1{\texttt{#1}\egroup}
+\def\Rd@code#1{\texttt{\nohyphenation#1}\egroup}
+
+\newcommand{\codefun}[1]{\code{#1()}}
 
@@
   Wirtschaftsuniversit\"at Wien, Austria\\
-\url{David.Meyer@R-project.org}
+\email{David.Meyer@R-project.org}
 }
@@
 One of the major enhancements in \pkg{vcd} compared to
-\code{mosaicplot()} and \code{assocplot()} in
-base {\sf R} is the labeling of strucplots which now
+\codefun{mosaicplot} and \codefun{assocplot} in
+base {\sf R} is the labeling in the strucplot framework which now
 has become much more powerful and flexible. Like the panel, the
@@
 a generating function that parameterizes a labeling function, 
-to \code{strucplot()} which then draws the labels. The labeling is 
+to \codefun{strucplot} which then draws the labels (generating
+functions are discussed in a separate vignette). The labeling is 
 well-separated from the actual plotting that occurs in the panel functions;
@@
 to combine some of the basic functions to produce a more complex labeling.
-As an example, consider \code{labeling\_list()}:
+% As an example, consider \codefun{labeling\_list}:
 
-\begin{Sinput}
-labeling_list <- function(gp = gpar(), just = "left", pos = "left", lsep = ": ", 
-                          sep = " ", offset = unit(c(2, 2), "lines"),
-                          varnames = TRUE, cols = 2, ...) 
-  function(d, split_vertical, condvars) {
-    \dots
-  }
-\end{Sinput}
+% \begin{Sinput}
+% labeling_list <- function(gp = gpar(), just = "left", pos = "left", lsep = ": ", 
+%                           sep = " ", offset = unit(c(2, 2), "lines"),
+%                           varnames = TRUE, cols = 2, ...) 
+%   function(d, split_vertical, condvars) {
+%     \dots
+%   }
+% \end{Sinput}
 
-\code{labeling\_list()} is a generating function that takes all parameters
-as arguments, and returns the actual labeling function which takes
-only three arguments: the dimnames structure of the table, the logical
-vector with the splitting information, and the number of conditioning
-variables (if any). The labeling function has access to the arguments
-of the generating function because in {\sf R}, nested functions are
-lexically scoped. In the
-following, we describe the three basic modules
-(\code{labeling\_text()}, \code{labeling\_list()}, and \code{labeling\_cells()}) and
+% \codefun{labeling\_list} is a generating function that takes all parameters
+% as arguments, and returns the actual labeling function which takes
+% only three arguments: the dimnames structure of the table, the logical
+% vector with the splitting information, and the number of conditioning
+% variables (if any). The labeling function has access to the arguments
+% of the generating function because in {\sf R}, nested functions are
+% lexically scoped. 
+In the following, we describe the three basic modules
+(\codefun{labeling\_text}, \codefun{labeling\_list}, and \codefun{labeling\_cells}) and
 derived functions that build upon them.
@@
 
-\code{labeling\_text()} is the default for all strucplot displays. It
-plots labels in the borders similar to the \code{mosaicplot()}
+\codefun{labeling\_text} is the default for all strucplot displays. It
+plots labels in the borders similar to the \codefun{mosaicplot}
 function in base {\sf R}, but is much more flexible: it is not limited
@@
 
-\begin{figure}[h]
+\begin{figure}[p]
 \begin{center}
@@
 abbreviated to (\code{TRUE} is interpreted as 1, obviously). Abbreviation
-is performed using {\sf R}'s \code{abbreviate()} function. Another
+is performed using the \codefun{abbreviate} function in base {\sf R}. Another
 possibility is to rotate the levels (see Figure \ref{fig:rotating}):
@@
 all labels and variables left-aligned, which can
-also be achieved by using \code{labeling\_left()} (see Figure \ref{fig:left}):
+also be achieved by using \codefun{labeling\_left} (see Figure \ref{fig:left}):
 
@@
 \noindent The values to \code{boxes} and \code{clip} are recycled for
-all dimensions. The result is pretty close to what the \code{labels\_cboxed()}
+all dimensions. The result is pretty close to what the \codefun{labels\_cboxed}
 wrapper does. Another variant is to put the variable names into the
@@
 labels'') is a vector of logicals indicating the side for the
-variable names. The resulting layout is basically what \code{labeling\_lboxed()} produces,
-and a similar design is used by the \code{doubledecker()}.
+variable names. The resulting layout is basically what \codefun{labeling\_lboxed} produces,
+and a similar design is used by the \codefun{doubledecker}.
 
@@
 \noindent For some data, it might be convenient to combine cell
-labeling with border labeling as done by \code{labels\_conditional()}
+labeling with border labeling as done by \codefun{labels\_conditional}
 (see Figure \ref{fig:conditional}):
@@
 
-\section{A simple list of labels:\texttt{labels\_list()}}
+\section{A simple list of labels: \texttt{labels\_list()}}
 

### 2005-08-30 trunk/vcd/R/assoc.R

@@ -30,11 +30,14 @@ assoc.default <- function(x,
   if (is.logical(main) && main)
     main <- deparse(substitute(x))
 
-  if (!inherits(x, "ftable")) {
-    if (is.null(row_vars) && is.null(col_vars) && is.table(x))
-      row_vars <- names(dimnames(x))[seq(1, length(dim(x)), by = 2)]
-    x <- ftable(x, row.vars = row_vars, col.vars = col_vars)
-  }
+  if (!inherits(x, "ftable"))
+    x <- structable(x)
+
+#     {
+#     if (is.null(row_vars) && is.null(col_vars) && is.table(x))
+#       row_vars <- names(dimnames(x))[seq(1, length(dim(x)), by = 2)]
+#     x <- ftable(x, row.vars = row_vars, col.vars = col_vars)
+#   }
 
   tab <- as.table(x)
   dl <- length(dim(tab))
@@ -47,6 +50,8 @@ assoc.default <- function(x,
   spacing <- spacing(dim(tab), condvars = which(cond))
 
   ## splitting arguments
+  if (is.null(split_vertical))
+    split_vertical <- attr(x, "split_vertical")
   if (is.null(split_vertical)) {
     split_vertical <- rep(FALSE, dl)
     names(split_vertical) <- names(dimnames(tab))

### 2005-09-18 trunk/vcd/inst/doc/labeling.Rnw
- vignettes revised - bugs in structable indexing fixed - mosaic, sieve, and assoc made structable aware - panel_generator class -> generating_function - offset_labels and offset_varnames added to labeling_border - spacing_highlighting replaces spacing_doubledecker
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@457 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \newcommand{\email}[1]{\href{mailto:#1}{\normalfont\texttt{#1}}}
+\newcommand{\data}[1]{`\texttt{#1}'}
 
@@
 
-\title{Labeling in Strucplots}
+\title{Labeling in the Strucplots Framework}
 \author{by David Meyer\\
@@
 
-One of the major enhancements in \pkg{vcd} compared to
+One of the major enhancements in package \pkg{vcd} compared to
 \codefun{mosaicplot} and \codefun{assocplot} in
-base {\sf R} is the labeling in the strucplot framework which now
-has become much more powerful and flexible. Like the panel, the
-shading, the spacing, and the legend, labeling is now completely modular.
+base {\sf R} is the labeling in the strucplot framework which 
+offers much more features and flexibility. Like the core, 
+shading, spacing, and legend modules, labeling is now completely
+modular (see the introducing vignette for an overview on
+the strucplot framework).
 The user supplies either a labeling function, or, alternatively, 
 a generating function that parameterizes a labeling function, 
-to \codefun{strucplot} which then draws the labels (generating
-functions are discussed in a separate vignette). The labeling is 
-well-separated from the actual plotting that occurs in the panel functions;
-it actually only relies on the viewport tree produced by them, and the
-`dimnames' attribute of the visualized table. The drawing of the
-labels happens after the actual plot has been drawn by the panel function.
+to \codefun{strucplot} which then draws the labels (the idea behind generating
+functions is discussed in detail in the vignette on colors and shadings). 
+Labeling is well-separated from the actual plotting that occurs in the
+low-level core functions. It only relies on the viewport tree produced by them, and the
+`\texttt{dimnames}' attribute of the visualized table. The drawing of the
+labels happens after the actual plot has been drawn by the core function.
 Thus, it is possible to supply one's own labeling function, or
 to combine some of the basic functions to produce a more complex labeling.
-% As an example, consider \codefun{labeling\_list}:
-
-% \begin{Sinput}
-% labeling_list <- function(gp = gpar(), just = "left", pos = "left", lsep = ": ", 
-%                           sep = " ", offset = unit(c(2, 2), "lines"),
-%                           varnames = TRUE, cols = 2, ...) 
-%   function(d, split_vertical, condvars) {
-%     \dots
-%   }
-% \end{Sinput}
-
-% \codefun{labeling\_list} is a generating function that takes all parameters
-% as arguments, and returns the actual labeling function which takes
-% only three arguments: the dimnames structure of the table, the logical
-% vector with the splitting information, and the number of conditioning
-% variables (if any). The labeling function has access to the arguments
-% of the generating function because in {\sf R}, nested functions are
-% lexically scoped. 
 In the following, we describe the three basic modules
@@
 
-As an example, consider the `Titanic' data, consisting
+As an example, consider the \data{Titanic} data, consisting
 of 4 categorical variables: Survived, Sex (i.e., `Gender'), Age, and
@@
 @
-\caption{Mosaic plot for the Titanic data.}
+\caption{Mosaic plot for the \data{Titanic} data.}
 \label{fig:defaults}
@@
 parameters. \code{clip} itself takes a vector of logicals (one for
-each dimension). Almost all vectorized arguments in the strucplot
+each dimension). 
+% as mentioned before
+Almost all vectorized arguments in the strucplot
 framework can be abbreviated in the following way: unnamed components
@@
 <<rotate>>=
-mosaic(Titanic, labeling_args = list(rot_labels = c(bottom = 90, right = 0)),
-                margins = c(right = 6, bottom = 4))
+mosaic(Titanic, labeling_args = list(rot_labels = c(bottom = 90, right = 0), 
+offset_varnames = c(right = 1), offset_labels = c(right = 0.3)), margins = c(right = 4, bottom = 3))
 @
@@
 variable names. The resulting layout is basically what \codefun{labeling\_lboxed} produces,
-and a similar design is used by the \codefun{doubledecker}.
+and a similar design is used by the \codefun{doubledecker} function.
 
@@
 This labeling draws both variable names and levels in the
-cells. As an example, we use the `PreSex' data on pre- and
+cells. As an example, we use the \data{PreSex} data on pre- and
 extramarital sex and divorce (see Figure \ref{fig:cell}):
@@
 @
-\caption{Cell labeling for the `PreSex' data.}
+\caption{Cell labeling for the \data{PreSex} data.}
 \label{fig:cell}
@@
 @
-\caption{Cell labeling for the `PreSex' data, labels abbreviated.}
+\caption{Cell labeling for the \data{PreSex} data, labels abbreviated.}
 \label{fig:cell2}
@@
 @
-\caption{Conditional labeling for the `PreSex' data, labels abbreviated.}
+\caption{Conditional labeling for the \data{PreSex}, labels abbreviated.}
 \label{fig:conditional}
@@
 shape than the data array to the \code{text} argument (cells with missing values
-are ignored). In the following example, this is used to add all observed values greater
-than 5 to the cells after the mosaic for the `Titanic' data has been plotted
-(see Figure \ref{fig:text}):
+are ignored). In the following example using the `Titanic' data, 
+this is used to add all observed values greater
+than 5 to the cells after the mosaic has been plotted (see Figure \ref{fig:text}):
 
 <<text>>=
-mosaic(Titanic, pop = FALSE, labeling_args = list(abbreviate = c(Survived = 1, Age = 4)))
+mosaic(Titanic, labeling_args = list(abbreviate = c(Survived = 1, Age = 4)), pop = FALSE)
 
@@
 @
-\caption{User-supplied Text added to a mosaic display of the `Titanic' data.}
+\caption{User-supplied Text added to a mosaic display of the \data{Titanic} data.}
 \label{fig:text}
@@
 
-\section{A simple list of labels: \texttt{labels\_list()}}
+\section{A simple list of labels: \texttt{labeling\_list()}}
 

### 2005-09-18 trunk/vcd/inst/doc/shading.Rnw
- vignettes revised - bugs in structable indexing fixed - mosaic, sieve, and assoc made structable aware - panel_generator class -> generating_function - offset_labels and offset_varnames added to labeling_border - spacing_highlighting replaces spacing_doubledecker
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@457 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \newenvironment{smallexample}{\begin{alltt}}{\end{alltt}}
+\newcommand{\var}[1]{\textit{\texttt{#1}}}
+\newcommand{\email}[1]{\href{mailto:#1}{\normalfont\texttt{#1}}}
+\newcommand{\data}[1]{`\texttt{#1}'}
+
+%% \code without `-' ligatures
+\def\nohyphenation{\hyphenchar\font=-1 \aftergroup\restorehyphenation}
+\def\restorehyphenation{\hyphenchar\font=`-}
+{\catcode`\-=\active%
+  \global\def\code{\bgroup%
+    \catcode`\-=\active \let-\codedash%
+    \Rd@code}}
+\def\codedash{-\discretionary{}{}{}}
+\def\Rd@code#1{\texttt{\nohyphenation#1}\egroup}
+
+\newcommand{\codefun}[1]{\code{#1()}}
 
@@
 
-\title{Colors and residual-based Shadings in Strucplots}
+\title{Colors and Residual-based Shadings in Strucplots}
 \author{by David Meyer\\
   Wirtschaftsuniversit\"at Wien, Austria\\
-\url{David.Meyer@R-project.org}
+\email{David.Meyer@R-project.org}
 }
@@
 
-Unlike other R graphics functions, strucplots allow almost full
-control over the graphical parameters of all plot elements. In
-particular, the user can \emph{independently} modify the appearance of the 
-tiles of association plots and mosaic plots. Built on top of this functionality, the
-framework supplies a set of shading functions that choose colors
-appropriate to the visualization of independence models fit to the
-visualized data.
+Unlike other graphics functions in base {\sf R}, the strucplot framework 
+allows almost full control over the graphical parameters of all plot elements. In
+particular, the user can \emph{individually} modify the graphical appearance of  
+tiles in association plots, mosaic plots, and sieve plots. 
+Built on top of this functionality, the
+framework supplies a set of shading functions choosing colors
+appropriate for the visualization of log-linear models.
 The tiles' graphical parameters are set using the \code{gp} argument
 of the strucplot functions. This argument basically expects an object
-of class \code{gpar} whose components have the same shape (length and dimensionality) 
-than the data table (see Section \ref{sec:gp}). For convenience, however, the user can also
+of class \code{gpar} whose components are arrays of the same shape 
+(length and dimensionality) than the data table (see Section \ref{sec:gp}). 
+For convenience, however, the user can also
 supply a function that computes such an object given a vector of
 residuals, or, alternatively, a generating function that takes
-parameters and returns such a function (see Section \ref{sec:shading}).
+parameters and returns such a function (see Section
+\ref{sec:shading}). We provide several shading functions, including
+support for both the HSV and the HCL color space, and the
+visualization of significance tests (see Section
+\ref{sec:overview}).
 
-\section{Specifying graphical parameters in strucplots}
+\section{Specifying graphical parameters of strucplot displays}
 \label{sec:gp}
 
-As an example, consider the `Admissions to the University of California
-in Berkely (UCB)' data. In the table collapsed over departments, we
+As an example, consider the \data{UCBAdmissions} data. 
+In the table collapsed over departments, we
 would like to highlight the (incidentally wrong) impression that there
@@
 @
-\caption{Mosaic plot for the UCBAdmissions data with highlighted cells.}
+
+\caption{Mosaic plot for the \data{UCBAdmissions} data with highlighted cells.}
 \label{fig:ucb}
@@
 
-\noindent As the example shows, we therefore create a fourfold table 
-with appropriate colors and supply them to the \code{fill} component 
-of the \code{gpar} object passed to the \code{gp} argument of \code{mosaic()}. 
-In addition, we use the \code{col} component to turn off the borders 
-for cosmetic reasons.
+\noindent As the example shows, we create a fourfold table 
+with appropriate colors (dark cyan for admitted male students and dark
+magenta for rejected female students) and supply them to the \code{fill} component 
+of the \code{gpar} object passed to the \code{gp} argument of \codefun{mosaic}. 
+For optical clarity, we addtionally hide the tiles' borders by setting the \code{col} 
+component to 0 (white).
 
@@
 strucplot functions will recycle them along the last splitting
-dimension. In the following example based on the `Survival on the
-Titanic' data, we will highlight all cells corresponding to survived
+dimension. In the following example based on the \data{Titanic} data, 
+we will highlight all cells corresponding to survived
 passengers (see Figure \ref{fig:recycling}):
@@
 <<recycling>>=
-mosaic(Titanic, gp = gpar(fill = c("gray","dark magenta")))
+mosaic(Titanic, gp = gpar(fill = c("gray","dark magenta")), 
+                spacing = spacing_highlighting,
+                labeling_args = list(abbreviate = c(Age = 3), rep = c(Survived = FALSE))
+      )
 @
+
+\noindent Note that \codefun{spacing\_highlighting} sets the spaces
+between tiles in the last dimension to 0. The \code{labeling\_args} 
+argument ensures that labels do not overlapp (see the separate
+vignette on labeling for more information).
 
@@
 @
-\caption{Recycling of parameters.}
+\caption{Recycling of parameters, used for highlighting the survived
+  passengers in the \data{Titanic} data.}
 \label{fig:recycling}
@@
 
-\section{Using residual-based shadings}
+\section{Customizing residual-based shadings}
 \label{sec:shading}
 
-This new flexible way of specifying graphical parameters is the basis for a suite of
+This flexible way of specifying graphical parameters is the basis for a suite of
 shading functions that modify the tiles' appearance with respect to a vector of 
 residuals, resulting from deviations of the observed from the expected
-values under a given independence model. 
+values under a given log-linear model. 
 The idea is to visualize at least the size of the residuals, but some
@@
 shading, the maximum shading, even allows to identify those cells that
-cause the rejection of the independence hypothesis.
+cause the rejection of the null hypothesis.
 
-All shading functions are parameterized be the means of generating
-functions that return the actual shading functions. As an example,
-consider the \code{shading\_binary()} function that only visualizes the
-sign of the residuals:
+Conceptually, the strucplot framework offers three alternatives to
+add residual-based shading to plots:
 
-\begin{Sinput}
-shading_binary <- function(observed = NULL, residuals = NULL, expected = NULL, 
-                           df = NULL, col = hcl(c(260, 0), 50, 70))
-{
-\dots
-}
-\end{Sinput}
+\begin{enumerate}
+\item Pre-computing the graphical parameters (e.g., fill colors),
+  encapsulating them into an object of class
+  \code{"gpar"} as demonstrated in the previous section, and
+  passing this object to the \code{gp} argument.
+\item Providing a function to the \code{gp} argument that 
+  takes residuals as input and returns an object as described in 1.
+\item Providing a \emph{generating} function taking parameters and returning
+  a function as described in 2. The parameters of this generating
+  function are either directly specified (which is not possible for
+  all shading functions), or as a list to the \code{shading\_args} argument.
+\end{enumerate}
 
-\noindent Each generating function, passed to a strucplot functions' \code{gp}
-argument, takes a few arguments that describe the model
-(\code{observed}, \code{expected}, \code{residuals}, and \code{df}), and a
-number of parameters that specify the appearance (in the case of
-\code{shading\_binary()} only \code{col}, a vector of length two, 
-that specifies the two colors corresponding to negative and 
-positive residuals, respectively). The
-generating function then return the actual shading function taking 
-only one argument: a vector of residuals, and returning an object
-of class \code{gpar} that contains the graphical parameters for all
-tiles corresponding to the residuals. In the following example, positive 
-and negative residuals of an hypothesized independence model shall be 
-visualized by different colors (see Figure \ref{fig:binary}):
+\noindent For each of these approaches, we will demonstrate the
+necessary steps to obtain a binary shading that visualizes the sign of
+the residuals by a corresponding 
+fill color (for simplicity, we will treat 0 as positive).
 
-<<binary,results=verbatim>>=
-obs <- margin.table(UCBAdmissions, 1:2)
-exp <- independence_table(obs)
-(res <- (obs - exp) / sqrt(exp))
+The first method is precomputing the graphical parameters ``by
+hand''. We will use `light blue' color for positive and `light salmon'
+color for negative residuals (see Figure \ref{fig:binary}):
 
-(gpobj <- shading_binary()(res))
+<<shading1,results=verbatim>>=
+expected <- independence_table(ucb)
+(residuals <- (ucb - expected)  / sqrt(expected))
 
-mosaic(obs, gp = gpobj)
+(fillcols <- ifelse(residuals > 0, "lightblue", "lightsalmon"))
+
+mosaic(ucb, gp = gpar(fill = fillcols))
 @
@@
 \begin{center}
-<<binaryfig,fig=TRUE,echo=FALSE>>=
-<<binary>>
+<<shading1fig,fig=TRUE,echo=FALSE>>=
+<<shading1>>
 @ 
-\caption{Binary shading.}
+\caption{Binary shading visualizing the sign of the residuals.}
 \label{fig:binary}
@@
 
-\noindent The shading-generating functions available
-in \pkg{vcd} include model-fitting, and hence the {\sf R} code
-corresponding to the example above in fact reduces to:
+For implementing method 2, we need to create a `shading function'
+that---technically speaking---computes \code{"gpar"} objects from
+residuals. For that, we can just reuse the code from the previous step:
 
-<<binaryfun>>=
-mosaic(obs, gp = shading_binary)
+<<shading2>>=
+shade_fun <- function(res) gpar(fill = ifelse(res > 0, "lightblue", "lightsalmon"))
 @
 
+\noindent To create a mosaic display with binary shading, 
+it now suffices to specify the data table along with
+\codefun{shade\_fun}:
+
+<<shading3>>=
+mosaic(ucb, gp = shade_fun)
+@
+
+\noindent \codefun{mosaic} internally calls
+\codefun{strucplot} which computes the residuals from the specified
+independence model (total independence by default), passes them to
+\codefun{shade\_fun}, and uses the \code{"gpar"} object returned 
+to finally create the plot.
+
+Our \codefun{shade\_fun} function might be useful, 
+but suffers from a major flaw: the colors are hard-wired. We cannot
+simply extend the argument list to include, e.g., 
+a \code{fill = c("lightblue", "lightsalmon")} argument because
+\codefun{strucplot} will neither know how to handle it, nor let us
+change the defaults. In fact, the interface of shading functions is fixed,
+they are expected to take exactly one argument: a table of residuals. 
+This is where generating functions (method 3) come
+into play. We simply wrap our shading function in another 
+function that takes all additional arguments it needs to use, possibly
+preprocesses them, and returns the actual shading function. This
+function will have access to the parameters since in {\sf R}, nested
+functions are lexically scoped. This kind of wrapping function
+is also called a \emph{generating function} since it returns (`creates') a
+`parameterized' shading function with the minimal standard interface
+\codefun{strucplot} requires. The follwing example shows the necessary
+extensions for our running example:
+
+<<shading3>>=
+myshading_binary <- function(col = c("lightblue", "lightsalmon")) {
+  if (length(col) != 2) stop("Need exactly two colors!")
+  function(res) gpar(fill = ifelse(res > 0, col[1], col[2]))  
+}
+@
+
+\noindent In the call of \codefun{mosaic} (figure not shown), 
+using the new \codefun{myshading\_binary} function, we can now simply change the colors:
+
+<<shading4>>=
+mosaic(ucb, gp = myshading_binary(c("red","blue")))
+@
+
+\noindent The procedure described so far is a rather general concept,
+applicable to a wide family of user-level \pkg{grid} graphics. Indeed, the
+customization of other components of the strucplot framework
+(labeling, spacing, legend, and core functions) follows the same idea.
+Now for the shading functions, this is not the whole story. Since shading
+functions can be used for visualizing significance (see next section), it
+makes sense for generating functions to have access to the model,
+i.e., observed and expected values, residuals, and degrees of
+freedom. For example, the \codefun{shading\_max} generating function
+computes a permutation distribution of the maximum statistic and $p$
+values for specified significance levels 
+based on the observed table to create data-driven cut-off
+points. If this was done in the shading function itself, the
+permutation statistic would be recomputed every time the shading
+function is called, resulting in possibly severe performance loss 
+and numerical inconsistencies. Therefore, generating functions for
+shadings are required to take at least the parameters 
+\code{observed}, \code{expected}, \code{residuals}, and \code{df}
+(these are provided by the strucplot framework),
+followed by other parameters controlling the shading
+appearance (to be specified by the user):
+
+<<shading4>>=
+myshading_binary2 <- function(observed = NULL, residuals = NULL, expected = NULL, 
+    df = NULL, col = c("lightblue", "lightsalmon")) {
+  if (length(col) != 2) stop("Need exactly two colors!")
+  function(res) gpar(fill = ifelse(res > 0, col[1], col[2]))  
+}
+class(myshading_binary2) <- "generating_function"
+@
+
+In some sence, generating
+functions for shadings are parameterized both by the user and the
+strucplot framework. For shading functions that require model
+information, the user-specified parameters are 
+to be passed to the \code{gp\_args} argument instead, and for this to
+work, the generating function needs a class attribute to be distinguishable from
+the ``normal'' shading functions. For others
+(like our simple \codefun{myshading\_binary2}) this is optional, but
+recommended for consistency:
+
+<<shading5>>=
+mosaic(ucb, gp = myshading_binary2, gp_args = list(c("red","blue")))
+@
+
+\noindent This final version pretty much resembles 
+\codefun{shading\_binary}, one of the standard shading
+functions provided by the \pkg{vcd} package.
+
 \section{An overview of the shading functions in \pkg{vcd}}
+\label{sec:overview}
 
-Currently, there are three basic shadings available
-(\code{shading\_hcl(), shading\_hsv(), and shading\_binary()}, as well
-as two derived functions (\code{shading\_Friendly()} building upon
-\code{shading\_hsv()} and \code{shading\_max()} building upon \code{shading\_hcl}).
-\code{shading\_binary()} has been explained in the previous
-paragraph. \code{shading\_hsv()} and \code{shading\_hcl()} are quite
-similar in their functionality, but use different color spaces: the
+Apart from \codefun{shading\_binary}, there are currently two basic shadings available:
+\codefun{shading\_hcl} and \codefun{shading\_hsv}, as well
+as two derived functions: \codefun{shading\_Friendly} building upon
+\codefun{shading\_hsv}, and \codefun{shading\_max} building upon \codefun{shading\_hcl}.
+\codefun{shading\_hsv} and \codefun{shading\_hcl} provide
+the same conceptual tools, but use different color spaces: the
 Hue-Saturation-Value (HSV) and the Hue-Chroma-Luminance (HCL) scheme,
-respectively. We will focus on the latter, since the HCL space is
-preferable to the HSV space in many respects (e.g., uniform saturations over
-different hues, device independentness, \dots).
+respectively. We will first expose the basic concept of these shading
+functions using the HSV space, and
+then briefly explain the differences to the HCL space \citep[a detailed
+discussion can be found in][]{vcd:Zeileis+Meyer+Hornik:2005}. 
+The HCL space is trickier
+to use, but preferable to the HSV space from a perceptual point of view.
 
-In the HCL space, colors are specified in three dimensions: hue, chroma (`colorfulness', saturation), and luminance (`brightness', amount of gray).
-These three dimensions are used by \code{shading\_hcl()} to visualize information about the residuals and the underlying independence model.
-A residuals' hue indicates its sign (by default blue and red for positive and negative ones, respectively).
-An overall full or poor chroma indicates significant or non-significant results, respectively.
-Finally, the luminance of a residual is set according to its size.
+In the HSV space, colors are specified in three dimensions: Hue,
+Saturation (`colorfulness'), and Value (`brightness', amount of gray). 
+These three dimensions are used by \codefun{shading\_hsv} to visualize information 
+about the residuals and the underlying independence model. The hue 
+indicates the residuals' sign: by default, blue for positive, and red
+for negative residuals. The saturation of a residual 
+is set according to its size: high saturation for large, and low
+saturation for small residuals. 
+Finally, the overall brightness is used to indicate the significance of
+a test statistic: bright colors for significant, 
+and dark colors for non-significant results.
 
-As an example, we will visualize the (in)dependence of hair and eye
-color in the `Hair and Eye Color' data set (see Figure \ref{fig:haireye})
+As an example, we will visualize the association of hair and eye
+color in the \data{HairEyeColor} data set (see Figure \ref{fig:haireye})
 
@@
 haireye <- margin.table(HairEyeColor, 1:2)
-mosaic(haireye, shade = TRUE)
+mosaic(haireye, gp = shading_hsv)
 @
@@
 @
-\caption{Shaded residuals in the HairEyeColor data set---two cut-off points}
+\caption{Shaded residuals in the \data{HairEyeColor} data set---two cut-off points.}
 \label{fig:haireye}
@@
 
-\noindent The default shading scheme is \code{shading\_hcl()}, so it
-suffices to turn shading on to bring colors to the mosaic plot. Large
-positive residuals (greater than $4$) can be found for brown eyes/black hair and blue
-eyes/blond hair, they are colored in saturated red. 
-On the other side, there is a large negative residual
+\noindent Large positive residuals (greater than $4$) can be found for
+brown eyes/black hair and blue eyes/blond hair, they are colored in saturated red. 
+On the other hand, there is a large negative residual
 (less than $-4$) for brown eyes/blond hair, colored deep blue. There are also three
 medium-sized positive (negative) residuals between 2 and 4 ($-2$ and
-$-4$): the colors for them are less saturated. 
+$-4$): the colors for them are less saturated. Residuals between $-2$ and $2$ 
+are shaded in white.
 The heuristic for choosing the cut-off points $2$ and $4$ is that the Pearson residuals
@@
 residuals \emph{individually} significant at approximately the $\alpha
-= 0.05$ and $\alpha = 0.0001$ level. These default cut-off points can
+= 0.05$ and $\alpha = 0.0001$ level, respectively. These default cut-off points can
 be changed to alternative values using the \code{interpolate}
@@
 <<interpolate>>=
-mosaic(haireye, shade = TRUE, 
-       gp_args = list(interpolate = 1:4, h = c(100, 200)))
+mosaic(haireye, gp = shading_hsv, gp_args = list(interpolate = 1:4))
 @
@@
 @
-\caption{Shaded residuals in the HairEyeColor data set---four cut-off points}
+\caption{Shaded residuals in the \data{HairEyeColor} data set---four cut-off points.}
 \label{fig:interpolate}
@@
 
-\noindent Note that we also changed the hue values for purely cosmetic
-reasons. The third remaining dimension, the luminance, is used for
-visualizing the significance of a test statistic. The $p$ value can
-be supplied to the \code{p.value} argument. If none is specified, it
-is computed from a Chi-squared distribution with \code{df} degrees of
+\noindent The \code{interpolate} argument also accepts a function,
+which then is called with the absolute residuals to get a vector of
+cut-off points. Thus, it is possible to automatically choose the cut-off points 
+in a data-driven way. For example,
+one might think that the extension from four cut-off points 
+to a continuous shading---visualizing the whole range of 
+significance levels---could be useful. We simply need a one-to-one
+mapping from the residuals to the $p$ values:
+
+<<continuous1>>=
+ipol <- function(x) pmin(x/4, 1)
+@
+
+\noindent Note that this \codefun{ipol} 
+function maps residuals greater than 4 to a significance
+level of 1. However, the resulting plot (Figure \ref{fig:continuous}) is 
+deceiving:
+
+<<continuous2>>=
+mosaic(HairEyeColor, gp = shading_hsv, gp_args = list(interpolate = ipol),
+labeling_args = list(abbreviate = c(Sex = 3)))
+@
+
+\begin{figure}[p]
+\begin{center}
+<<continuousfig,fig=TRUE,echo=FALSE>>=
+<<continuous2>>
+@
+\caption{The \data{HairEyeColor} data with continuous shading.}
+\label{fig:continuous}
+\end{center}
+\end{figure}
+
+\noindent Too much color makes it difficult to interpret the image, and the subtle color
+differences are hard to catch. Therefore, we only included shadings
+with discrete cut-off points.
+
+The third remaining dimension, the value, is used for
+visualizing the significance of a test statistic. The user can either
+directly specify the $p$ value, or, alternatively, a function that
+computes it, to the \code{p.value} argument. Such a function must take
+observed and expected values, residuals, and degrees of freedom as arguments. 
+If nothing is specified, the $p$ value is computed from 
+a $\chi^2$ distribution with \code{df} degrees of
 freedom. The \code{level} argument is used to specify the confidence
 level:  if \code{p.value} is smaller than \code{1 - level}, bright colors are used, 
-otherwise dark colors are employed. These colors are computed using
-the two luminance values supplied to the \code{l} argument. The
-following example using the `Bundesliga' data 
+otherwise dark colors are employed. 
+The following example using the \data{Bundesliga} data 
 shows the relationship of home goals and away
-goals of the German football league in 1995: although there are two
+goals of the German premier's soccer league in 1995: although there are two
 ``larger'' residuals (one greater than 2, one less then $-2$), the
-Chi-squared test does not reject the null hypothesis of
+$\chi^2$ test does not reject the null hypothesis of
 independence. Consequently, the colors appear dark (see Figure \ref{fig:bundesliga}):
@@
 bl <- xtabs(~ HomeGoals + AwayGoals, data = Bundesliga, subset = Year == 1995)
-mosaic(bl, shade = TRUE)
+mosaic(bl, gp = shading_hsv)
 @
 
-\begin{figure}[h]
+\begin{figure}[p]
 \begin{center}
@@
 @
-\caption{Non-significant Chi-squared test using part of the Bundesliga data.}
+\caption{Non-significant $\chi^2$ test using part of the \data{Bundesliga} data.}
 \label{fig:bundesliga}
@@
 
-\noindent The \code{interpolate} argument also accepts a function that
-automatically chooses the cut-off points in a data-driven way. 
-This functionality is used by \code{shading\_max()} that uses the
-maximum statistic to visualize significant \emph{cells} causing the
-rejection of the independence hypothesis. The
-\code{level} argument of \code{shading\_max()} 
-then can be used to specify several confidence
-levels from which the corresponding cut-off points are computed. 
-By default, two cut-off points are computed corresponding 
-to confidence levels of $90\%$ and $99\%$, respectively.
-In the following example, we investigate the effect of a new treatment
-for rheumatoid arthritis on a group of female patients using the
-maximum shading (see Figure \ref{fig:maximum}):
-
-<<arthritis>>=
-art <- xtabs(~ Improved + Treatment, data = Arthritis, subset = Sex == "Female")
-mosaic(art, gp = shading_max)
-@
-
-\begin{figure}[h]
-\begin{center}
-<<arthritisfig,fig=TRUE,echo=FALSE>>=
-<<arthritis>>
-@
-\caption{Significant maximum test on female patients of the Arthritis
-  data.}
-\label{fig:maximum}
-\end{center}
-\end{figure}
-
-\noindent The maximum test is significant although the residuals are
-all in the $[-2,2]$ interval, as the cut-off points are even smaller.
-The \code{shading\_hcl} function with default cut-off points would
-not have shown any color. In addition, since the test statistic is the
-maximum of the absolute pearson residuals, \emph{every} colored
-residual violates the hypotheses of independence, and thus, the
-`culprits' can immediately be identified.
-
-The last shading function is \code{shading\_Friendly()}. This shading
-mimics the shading introduced by \cite{vcd:Friendly:2000}. This
-shading corresponds to the HSV shading with default cut-off points (2 and 4), and in
-addition uses the border color and line type to visualize the
-residuals' sign. The following example again visualizes the Bundesliga
-data from above, this time using the Friendly scheme and, in addition, an
-alternative legend:
+A shading function building upon \codefun{shading\_hsv} is \codefun{shading\_Friendly},
+miming the shading introduced by \cite{vcd:Friendly:1994}. In addition
+to the defaults of the HSV shading, it uses the border color and line type to 
+redundantly code the residuals' sign. The following example again uses 
+the \data{Bundesliga} data from above, this time using the Friendly 
+scheme and, in addition, an alternative legend (see Figure \ref{fig:friendly}):
 
@@
 @
-\caption{The Bundesliga data for 1995 using the Friendly shading.}
+\caption{The \data{Bundesliga} data for 1995 using the Friendly
+  shading and a legend with fixed bins.}
 \label{fig:friendly}
@@
 
-\noindent (The bullets indicating zero observed values are removed here
-since this feature does not exist in the original SAS implementation
-of the Friendly mosaic plots).
+\noindent (The \code{zeroSize = 0} argument removes the bullets indicating zero 
+observed values. This feature does not
+exist in the original {\sf SAS} implementation of the Friendly mosaic plots.)
 
-\section{Creating new shading functions}
+As introduced before, the default shading scheme is not
+\codefun{shading\_hsv} but \codefun{shading\_hcl} due to the better
+perceptual characteristics of the HCL color space.
+Figure \ref{fig:shadingHSVHCL} depicts color scales for red and blue hue
+(left and right side, respectively) in the HSV and HCL spaces (upper and lower part, respectively), varying colorfulness and brightness ($x$ and $y$-axis, respectively).
+The boxes represent the diverging color palettes used for the shadings.
+For the HSV space, we can can see that the effect of changing the 
+level of brightness (`value') is not the same for different levels of 
+saturation, and again not the same for the two different hues.
+In fact, in the HSV space all dimensions are confounded, which 
+obviously is problematic for coding information. In contrast, the HCL color
+space offers perceptual uniform colors: as can be seen from the lower part, 
+the chroma is homogeneous for different levels of luminance. 
+Unfortunately, this comes at the 
+price of the space being irregularly shaped, making it difficult to automatically select 
+diverging color palettes. The following example again 
+illustrates the \data{HairEyeColor} data, this time with HCL colors 
+(Figure \ref{fig:hclshading1} depicts the default palette, 
+and Figure \ref{fig:hclshading2} an alternative setting):
 
-The example with four cut-off points (Figure \ref{fig:interpolate})
-might suggest that a continuous shading simply visualizing the complete
-range of confidence levels might be useful. The following code shows
-how such a shading function might be constructed:
+<<auxiliary_functions,echo=FALSE,results=hide>>=
+hue.slice <- function(hue, grid.n = 101, type = c("HCL", "HSV"), plot = TRUE, fixup = FALSE)
+{
+  type <- match.arg(type)
+  if(type == "HCL") {
+    chroma = seq(0, 100, length = grid.n)
+    luminance = seq(0, 100, length = grid.n)
+    nc <- length(chroma)
+    nl <- length(luminance)
+    color.slice <- outer(chroma, luminance, function(y, x) hcl(hue, x, y, fixup = fixup))
+    xlab <- "chroma"
+    ylab <- "luminance"
+    main <- paste("hue =", round(hue, digits = 0))
+  } else {
+    chroma = seq(0, 1, length = grid.n)
+    luminance = seq(0, 1, length = grid.n)
+    nc <- length(chroma)
+    nl <- length(luminance)
+    color.slice <- outer(chroma, luminance, function(y, x) hsv(hue, x, y))
+    xlab <- "saturation"
+    ylab <- "value"
+    main <- paste("hue =", round(hue, digits = 3))
+  }
+  if(plot) {
+    plot(0.5, 0.5, xlim = range(chroma), ylim = range(luminance), type = "n", axes = FALSE,
+         xlab = xlab, ylab = ylab, yaxs = "i", xaxs = "i", main = main)
+    for(i in 1:(nc-1)) {
+      rect(chroma[i], luminance[-nl], chroma[i] + 100/(nc-1), luminance[-1], border = color.slice[,i+1], col = color.slice[,i+1])
+    }
+    axis(1)
+    axis(2)
+    box()
+  }
+  colnames(color.slice) <- chroma
+  rownames(color.slice) <- luminance
+  attr(color.slice, "type") <- type
+  class(color.slice) <- "slice"
+  invisible(color.slice)
+}
+@
 
-<<continuous>>=
-shading_continuous <- function(observed = NULL, residuals = NULL,
-                               expected = NULL, df = NULL, n = 5000, ...)
-{
-  ## function works only for two-way tables
-  stopifnot(length(dim(observed)) == 2)
+\setkeys{Gin}{width=.8\textwidth}
+\begin{figure}[p]
+\begin{center}
+<<shading_HSV,fig=TRUE,echo=FALSE,height=7,width=8>>=
+## generate colors
+hue23 <- hue.slice(2/3, grid.n = 101, plot = FALSE, type = "HSV")
+hue0 <- hue.slice(0, grid.n = 101, plot = FALSE, type = "HSV")
+saturation <- as.numeric(colnames(hue23))
+value <- as.numeric(rownames(hue23))
 
-  ## sample permutation distribution of the maximum statistic
-  obs.test <- coindep_test(observed, n = n)
+## select those with value >= 0.5
+hue23 <- hue23[value >= .5, ]
+hue0 <- hue0[value >= .5, ]
+value <- value[value >= .5]
+nl <- nrow(hue23)
+nc <- ncol(hue23)
 
-  ## Use distribution to map residuals to [0,1] range
-  ipol <- function(x) abs(obs.test$pdist(x))
+## plot 2 slides from HSV space
+plot(0.5, 0.5, xlim = c(-1, 1), ylim = c(0, 1), type = "n", axes = FALSE,
+       xlab = "", ylab = "", yaxs = "i", xaxs = "i", main = "")
+for(i in 1:(nc-1)) {
+  rect(saturation[i], value[-nl], saturation[i] + 1/(nc-1), value[-1], border = hue23[,i+1], col = hue23[,i+1])
+}
+for(i in 1:(nc-1)) {
+  rect(-saturation[i], value[-nl], -(saturation[i] + 1/(nc-1)), value[-1], border = hue0[,i+1], col = hue0[,i+1])
+}
+axis(2, at = c(50, 75, 100)/100, labels = c(0.5, 0.75, 1))
+axis(4, at = c(50, 75, 100)/100, labels = c(0.5, 0.75, 1))
+axis(3, at = -4:4*.25, labels=c(4:0*.25, 1:4*.25))
+mtext(c("hue = 0", "hue = 2/3"), side = 3, at = c(-.5, .5), line = 3, cex = 1.2)
+mtext("saturation", side = 3, at = 0, line = 2)
+mtext("value", side = 2, at = .75, line = 2)
+mtext("value", side = 4, at = .75, line = 2)
+lines(c(-1, 1), c(.5, .5))
 
-  ## call HCL shading 
-  shading_hcl(observed = NULL, residuals = NULL, expected = NULL,
-              df = NULL, interpolate = ipol, p.value = obs.test$p.value, ...)
+## significant colors
+rect(-1, 0.95, -.90, 1, col = hsv(0, 1, 1))
+rect(-0.45, 0.95, -.55, 1, col = hsv(0, 0.5, 1))
+rect(-.05, .95, .05, 1, col = hsv(2/3, 0, 1))
+rect(0.45, 0.95, .55, 1, col = hsv(2/3, 0.5, 1))
+rect(.90, .95, 1, 1, col = hsv(2/3, 1, 1))
+
+text(-1, .33, "significant", pos = 4, cex = 1.2)
+rect(-1, .20, -.80, .30, col = hsv(0, 1, 1))
+rect(-.40, .20, -0.6, .30, col = hsv(0, 0.5, 1))
+rect(-.20, .20, 0, .30, col = hsv(0, 0, 1))
+rect(0, .20, .20, .30, col = hsv(2/3, 0, 1))
+rect(0.4, .20, .60, .30, col = hsv(2/3, .5, 1))
+rect(.80, .20, 1, .30, col = hsv(2/3, 1, 1))
+
+lines(c(-.9, -.55), c(0.975, .975), lty = 2)
+lines(c(-.45, -.05), c(0.975, .975), lty = 2)
+lines(c(.45, .05), c(0.975, .975), lty = 2)
+lines(c(.9, .55), c(0.975, .975), lty = 2)
+
+## non-significant colors
+rect(-1, 0.5, -.90, 0.55, col = hsv(0, 1, 0.5))
+rect(-0.45, 0.5, -.55, 0.55, col = hsv(0, 0.5, 0.5))
+rect(-.05, .5, .05, 0.55, col = hsv(2/3, 0, 0.5))
+rect(0.45, 0.5, .55, 0.55, col = hsv(2/3, 0.5, 0.5))
+rect(.90, .5, 1, 0.55, col = hsv(2/3, 1, 0.5))
+
+text(-1, .13, "non-significant", pos = 4, cex = 1.2)
+rect(-1, 0, -.80, .10, col = hsv(0, 1, 0.5))
+rect(-.60, 0, -.4, .10, col = hsv(0, 0.5, 0.5))
+rect(-.20, 0, 0, .10, col = hsv(0, 0, 0.5))
+rect(0, 0, .20, .10, col = hsv(2/3, 0, 0.5))
+rect(0.4, 0, .60, .1, col = hsv(2/3, .5, 0.5))
+rect(.80, 0, 1, .10, col = hsv(2/3, 1, 0.5))
+
+lines(c(-.9, -.55), c(0.525, .525), lty = 2)
+lines(c(-.45, -.05), c(0.525, .525), lty = 2)
+lines(c(.45, .05), c(0.525, .525), lty = 2)
+lines(c(.9, .55), c(0.525, .525), lty = 2)
+@
+
+<<shading_HCL,fig=TRUE,echo=FALSE,height=7,width=8>>=
+## generate colors
+hue260 <- hue.slice(260, grid.n = 101, plot = FALSE)
+hue360 <- hue.slice(360, grid.n = 101, plot = FALSE)
+mychroma <- as.numeric(colnames(hue260))
+luminance <- as.numeric(rownames(hue260))
+
+## select those with lumincance >= 50
+hue260 <- hue260[luminance >= 50, ]
+hue360 <- hue360[luminance >= 50, ]
+luminance <- luminance[luminance >= 50]
+nc <- ncol(hue260)
+nl <- nrow(hue260)
+
+## plot 2 slides from HCL space
+plot(0.5, 0.5, xlim = c(-100, 100), ylim = c(0, 100), type = "n", axes = FALSE,
+       xlab = "", ylab = "", yaxs = "i", xaxs = "i", main = "")
+for(i in 1:(nc-1)) {
+  rect(mychroma[i], luminance[-nl], mychroma[i] + 100/(nc-1), luminance[-1], border = hue260[,i+1], col = hue260[,i+1])
 }
-class(shading_continuous) <- "panel_generator"
+for(i in 1:(nc-1)) {
+  rect(-mychroma[i], luminance[-nl], -(mychroma[i] + 100/(nc-1)), luminance[-1], border = hue360[,i+1], col = hue360[,i+1])
+}
+axis(2, at = c(50, 70, 90, 100), labels = c(50, 70, 90, 100))
+axis(4, at = c(50, 70, 90, 100), labels = c(50, 70, 90, 100))
+axis(3, at = -4:4*25, labels=c(4:0*25, 1:4*25))
+mtext(c("hue = 0", "hue = 260"), side = 3, at = c(-50, 50), line = 3, cex = 1.2)
+mtext("chroma", side = 3, at = 0, line = 2)
+mtext("luminance", side = 2, at = 75, line = 2)
+mtext("luminance", side = 4, at = 75, line = 2)
+lines(c(-100, 100), c(50, 50))
 
-mosaic(haireye, gp = shading_continuous)
+## significant colors
+rect(-100, 47.5, -90, 52.5, col = hcl(0, 100, 50))
+rect(-55, 67.5, -45, 72.5, col = hcl(0, 50, 70))
+rect(-5, 95, 5, 100, col = hcl(260, 0, 100))       ## grey vs. white
+rect(-5, 87.5, 5, 92.5, col = hcl(260, 0, 90))     ## grey vs. white
+rect(45, 67.5, 55, 72.5, col = hcl(260, 50, 70))
+rect(90, 47.5, 100, 52.5, col = hcl(260, 100, 50))
+
+text(-100, 33, "significant", pos = 4, cex = 1.2)
+rect(-100, 20, -80, 30, col = hcl(0, 100, 50))
+rect(-60, 20, -40, 30, col = hcl(0, 50, 70))
+rect(-20, 20, 0, 30, col = hcl(0, 0, 90))       
+rect(0, 20, 20, 30, col = hcl(260, 0, 90))
+#white# rect(-20, 20, 0, 30, col = hcl(0, 0, 100))
+#white# rect(0, 20, 20, 30, col = hcl(260, 0, 100))
+rect(40, 20, 60, 30, col = hcl(260, 50, 70))
+rect(80, 20, 100, 30, col = hcl(260, 100, 50))
+
+lines(c(-45, -5), c(72.5, 87.5), lty = 2)
+lines(c(45, 5), c(72.5, 87.5), lty = 2)
+lines(c(-95, -55), c(52.5, 67.5), lty = 2)
+lines(c(95, 55), c(52.5, 67.5), lty = 2)
+
+## non-significant colors
+rect(-25, 47.5, -15, 52.5, col = hcl(0, 20, 50))
+rect(-15, 67.5, -5, 72.5, col = hcl(0, 10, 70))
+rect(5, 67.5, 15, 72.5, col = hcl(260, 10, 70))
+rect(25, 47.5, 15, 52.5, col = hcl(260, 20, 50))
+
+text(-100, 13, "non-significant", pos = 4, cex = 1.2)
+rect(-60, 0, -40, 10, col = hcl(0, 20, 50))
+rect(-40, 0, -20, 10, col = hcl(0, 10, 70))
+rect(-20, 0, 0, 10, col = hcl(0, 0, 90))
+rect(0, 0, 20, 10, col = hcl(260, 0, 90))
+rect(20, 0, 40, 10, col = hcl(260, 10, 70))
+rect(40, 0, 60, 10, col = hcl(260, 20, 50))
+
+lines(c(-18.75, -11.25), c(52.5, 67.5), lty = 2)
+lines(c(-8.75, -1.25), c(72.5, 87.5), lty = 2)
+lines(c(18.75, 11.75), c(52.5, 67.5), lty = 2)
+lines(c(8.75, 1.25), c(72.5, 87.5), lty = 2)
+@ 
+\caption{Residual-based shadings in HSV (upper) and HCL space (lower).}
+\label{fig:shadingHSVHCL}
+\end{center}
+\end{figure}
+
+<<hclshading1>>=
+mosaic(haireye, gp = shading_hcl)
+@ 
+<<hclshading2>>=
+mosaic(haireye, gp = shading_hcl, gp_args = list(h = c(130, 43), c = 100, l = c(90, 70)))
+@
+
+\setkeys{Gin}{width=0.7\textwidth}
+\begin{figure}[p]
+\begin{center}
+<<hclshading1fig,fig=TRUE,echo=FALSE>>=
+<<hclshading1>>
+@
+\caption{The \data{HairEyeColor} data, using default HCL color palette.}
+\label{fig:hclshading1}
+\end{center}
+\end{figure}
+
+\begin{figure}[p]
+\begin{center}
+<<hclshading2fig,fig=TRUE,echo=FALSE>>=
+<<hclshading2>>
+@
+\caption{The \data{HairEyeColor} data, using a custom HCL color palette.}
+\label{fig:hclshading2}
+\end{center}
+\end{figure}
+
+A more `advanced' function building upon \codefun{shading\_hcl} 
+is \codefun{shading\_max}, using the
+maximum statistic both to conduct the independence test and to 
+visualize significant \emph{cells} causing the
+rejection of the independence hypothesis \citep{vcd:Meyer+Zeileis+Hornik:2003}. The
+\code{level} argument of \codefun{shading\_max} 
+then can be used to specify several confidence
+levels from which the corresponding cut-off points are computed. 
+By default, two cut-off points are computed corresponding 
+to confidence levels of $90\%$ and $99\%$, respectively.
+In the following example, we investigate the effect of a new treatment
+for rheumatoid arthritis on a group of female patients using the
+maximum shading (see Figure \ref{fig:maximum}):
+
+<<arthritis>>=
+mosaic(~ Treatment + Improved, data = Arthritis, subset = Sex == "Female",
+       gp = shading_max)
 @
@@
 \begin{center}
-<<continuousfig,fig=TRUE,echo=FALSE>>=
-<<continuous>>
+<<arthritisfig,fig=TRUE,echo=FALSE>>=
+<<arthritis>>
 @
-\caption{The HairEyeColor data with continuous shading.}
-\label{fig:continuous}
+\caption{Significant maximum test on female patients of the \data{Arthritis}
+  data.}
+\label{fig:maximum}
 \end{center}
@@
 
-\noindent We first compute the permutation distribution for the
-maximum statistic using \code{coindep\_test}, and then simply use it
-as the interpolation function passed to \code{shading\_hcl}. The result
-using the HairEyeColor data, however, is disappointing: too much color
-makes it difficult to interpret the image, and the subtle color
-differences are hard to catch. Therefore, we only included shadings
-with discrete cut-off points.
+\noindent The maximum test is significant although the residuals are
+all in the $[-2,2]$ interval, as the cut-off points are even smaller.
+The \codefun{shading\_hcl} function with default cut-off points would
+not have shown any color. In addition, since the test statistic is the
+maximum of the absolute Pearson residuals, \emph{every} colored
+residual violates the null hypotheses of independence, and thus, the
+`culprits' can immediately be identified.
 
-\begin{thebibliography}{}
-
-\bibitem[Friendly, 2000]{vcd:Friendly:2000}
-Friendly, M. (2000).
-\newblock {\em Visualizing Categorical Data}.
-\newblock SAS Insitute, Carey, NC.
-
-\end{thebibliography}
+\bibliographystyle{apalike}
+\bibliography{vcd}
 

### 2005-09-18 trunk/vcd/inst/doc/strucplot.Rnw
- vignettes revised - bugs in structable indexing fixed - mosaic, sieve, and assoc made structable aware - panel_generator class -> generating_function - offset_labels and offset_varnames added to labeling_border - spacing_highlighting replaces spacing_doubledecker
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@457 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2005-08-24 trunk/vcd/inst/doc/strucplot.Rnw
@@
 \newenvironment{smallexample}{\begin{alltt}}{\end{alltt}}
+\newcommand{\email}[1]{\href{mailto:#1}{\normalfont\texttt{#1}}}
+\newcommand{\data}[1]{`\texttt{#1}'}
+
+%% \code without `-' ligatures
+\def\nohyphenation{\hyphenchar\font=-1 \aftergroup\restorehyphenation}
+\def\restorehyphenation{\hyphenchar\font=`-}
+{\catcode`\-=\active%
+  \global\def\code{\bgroup%
+    \catcode`\-=\active \let-\codedash%
+    \Rd@code}}
+\def\codedash{-\discretionary{}{}{}}
+\def\Rd@code#1{\texttt{\nohyphenation#1}\egroup}
+
+\newcommand{\codefun}[1]{\code{#1()}}
 
@@
 data(Titanic)
+data(HairEyeColor)
 @
 
-\title{strucplots---Visualizing higher-dimensional Contingency Tables in \pkg{vcd}}
+\title{The Strucplot Framework---Visualizing Multi-way Contingency Tables}
 \author{by David Meyer\\
   Wirtschaftsuniversit\"at Wien, Austria\\
-\url{David.Meyer@R-project.org}
+\email{David.Meyer@R-project.org}
 }
@@
 
-The \code{strucplot} framework in \pkg{vcd} can be used to visualize
-higher-dimensional contingency tables and integrates techniques such as
+\section{Framework Overview}
+
+The strucplot framework in the {\sf R} package \pkg{vcd} visualizes
+multi-way contingency tables and integrates techniques such as
 mosaic displays, association plots, and sieve plots. The main idea is to visualize
-the tables' cells arranged in rectangular form. To allow for more than
-two dimensions, the plot area is constructed using recursive
-conditional splits, given the tiles representing the levels of the
-previous dimension.
-This principle defines a graph `family' with still many parameters
+the tables' cells arranged in rectangular form. For multi-way tables,
+the variables are nested into row and columns using recursive
+conditional splits, given the margins. The result is a 
+`flat' representation that can be visualized in similar
+ways than a two-dimensional table.
+This principle defines a class of conditional displays with still many parameters
 left such as:
@@
 \begin{itemize}
+\item the content of the tiles
 \item the split direction for each dimension
-\item the content of the tiles
 \item the graphical parameters of the tiles' content
@@
 
-\section{Framework Overview}
-
 The strucplot framework is highly modularized; Figure \ref{fig:struc}
 shows the hierarchical relationship between the various components. 
-The core function is \code{strucplot()} which processes the parameters,
+On the lowest level, there are several groups of workhorse and 
+parameter functions that directly or indirectly influence the final
+appearance of the plot. The workhorse functions (\code{\emph{struc\_foo()}}, 
+\code{\emph{labeling\_foo()}}, and \code{\emph{legend\_foo()}}) 
+directly produce graphical output, the parameter functions
+(\code{\emph{spacing\_foo()}} and \code{\emph{shading\_foo()}}) compute
+graphical parameters used by the others. The \code{\emph{struc\_foo()}} functions
+implement the core functionality, creating the tiles and their
+content. On the second level, a suitable combination of these functions
+is passed as ``hyperparameters'' to \codefun{strucplot}.
+These central function 
 sets up the graphical layout using grid viewports (see Figure \ref{fig:layout}),
-and coordinates panel, labeling, shading, and spacing functions to produce 
-the plot. The actual workhorse is the chosen panel function: currently, 
-low-level panel functions for association plots (\code{struc\_assoc}), 
-mosaic plots (\code{struc\_mosaic}), and sieve plots (\code{struc\_sieve})
-are available. In addition, there are
-high-level wrapper functions such as \code{mosaic()}, \code{assoc()},
-and \code{sieve()} to \code{strucplot()} that conveniently set the 
-parameters---it is these high-level functions that the user normally would like to use, 
-although \code{strucplot()} is fully functional on its own. \code{mosaic()}, 
-\code{assoc}, and \code{sieve()} are generic functions: 
-for both, there exist a formula method and a default 
-method. Other functions, such as \code{doubledecker()}, \code{pairs()}, and 
-\code{cotabplot}, are on even higher level and make use of 
-code{mosaic()}, \code{assoc()}, and \code{sieve()} to produce plots.
+and coordinates the specified core, labeling, shading, and spacing functions to produce 
+the plot. On the third level, we provide
+several high-level functions such as \codefun{mosaic},
+\codefun{sieve}, \codefun{assoc}, and \codefun{doubledecker} which 
+conveniently interface \codefun{strucplot} through sensible parameter defaults
+and support for model formulas. Finally, on the fourth
+level, there are `related' \pkg{vcd} functions (such as \codefun{pairs.table} and
+\codefun{cotabplot}) combining single plots of the strucplot 
+framework into more complex displays.
 
 \begin{figure}[h]
-  \centering
-  \includegraphics[width=\textwidth]{struc}
-  \caption{Strucplot components.}
+  \begin{center}
+    \includegraphics[width=0.8\textwidth]{struc}
+    \caption{Components of the strucplot framework.}
     \label{fig:struc}
+  \end{center}
 \end{figure}
@@
 \begin{figure}[h]
-  \centering
-  \includegraphics{vcdViewport}
-  \caption{Viewport layout for strucplots with their names. [A] =
-    ``cornerTL'', [B] = ``cornerTR'', [C] = ``cornerBL'', [D] = ``cornerBR''.}
+\begin{center}
+<<vcdlayout,fig=TRUE,echo=FALSE,width=6,height=5>>=
+    pushViewport(vcd:::vcdViewport(legend = T, mar =4))
+    seekViewport("plot")
+    grid.rect(gp = gpar(lwd = 3))
+    grid.text("plot", gp = gpar(fontsize = 30))
+    seekViewport("legend")
+    grid.text("legend", rot = 90, gp = gpar(fontsize = 30))
+    grid.rect(gp = gpar(lwd = 3))
+    seekViewport("pval")
+    grid.rect(gp = gpar(lwd = 3))
+    grid.text("pval", gp = gpar(fontsize = 30))
+    seekViewport("margin_top")
+    grid.rect(gp = gpar(lwd = 3))
+    grid.text("margin_top", gp = gpar(fontsize = 30))
+    seekViewport("margin_bottom")
+    grid.rect(gp = gpar(lwd = 3))
+    grid.text("margin_bottom", gp = gpar(fontsize = 30))
+    seekViewport("margin_right")
+    grid.rect(gp = gpar(lwd = 3))
+    grid.text("margin_right", rot = 90, gp = gpar(fontsize = 30))
+    seekViewport("margin_left")
+    grid.rect(gp = gpar(lwd = 3))
+    grid.text("margin_left", rot = 90, gp = gpar(fontsize = 30))
+    seekViewport("corner_top_left")
+    grid.rect(gp = gpar(lwd = 3))
+    grid.text("[A]", gp = gpar(fontsize = 30))
+    seekViewport("corner_top_right")
+    grid.rect(gp = gpar(lwd = 3))
+    grid.text("[B]", gp = gpar(fontsize = 30))
+    seekViewport("corner_bottom_left")
+    grid.rect(gp = gpar(lwd = 3))
+    grid.text("[C]", gp = gpar(fontsize = 30))
+    seekViewport("corner_bottom_right")
+    grid.rect(gp = gpar(lwd = 3))
+    grid.text("[D]", gp = gpar(fontsize = 30))
+@ 
+\caption{Viewport layout for strucplot displays with their names. [A] =
+  ``corner\_top\_left'', [B] = ``corner\_top\_right'', 
+  [C] = ``corner\_bottom\_left'', [D] = ``corner\_bottom\_right''.}
 \label{fig:layout}
+\end{center}
 \end{figure}
 
-\section{Mosaic and Association Plots}
+\section{Mosaic, Association, and Sieve Plots}
 
-As an example, consider the famous `Survival on the Titanic' 
-data containing two binary variables (survival to the disaster and
-gender), as well as two categorical variables (class and age). The
-`flattened' contincency table can be obtained using the
-\code{structable()} function (quite similar to \code{ftable()}, but
-allowing the specification of split directions as in mosaic():
+As an example, consider the \data{HairEyeColor} 
+data containing two polytomous variables (hair and eye color), 
+as well as one (artificial) dichotomous variable (sex, i.e. gender). The
+`flattened' contingency table can be obtained using the
+\codefun{structable} function (quite similar to \codefun{ftable} in
+base {\sf R}, but allowing the specification of split directions):
 
 <<structable,results=verbatim>>=
-structable(Titanic)
+(hec <- structable(aperm(HairEyeColor)))
 @
@@
 is proportional to the corresponding cell entry \emph{given} the
-dimensions of previous splits (see Figure \ref{fig:observed}):
+dimensions of previous splits. Figure \ref{fig:observed} depicts 
+the effect of
 
 <<Observed>>=
-mosaic(Titanic, main = TRUE)
+mosaic(hec)
 @
 
-\noindent The small bullets indicate zero entries in the corresponding cell.
+\noindent equivalent to
 
-\begin{figure}[h]
+<<Observed2>>=
+mosaic(~ Sex + Eye + Hair, data = HairEyeColor)
+@
+
+\begin{figure}[p]
 \begin{center}
@@
 @
-\caption{Mosaic plot for the Titanic data.}
+\caption{Mosaic plot for the \data{HairEyeColor} data.}
 \label{fig:observed}
@@
 
-\noindent  It is also possible to visualize the expected values instead of the
+\noindent The small bullets indicate zero entries in the corresponding cell.
+It is also possible to visualize the expected values instead of the
 observed values (see Figure \ref{fig:expected}):
@@
 <<Expected>>=
-mosaic(Titanic, type = "expected", main = TRUE, sub = "Expected values")
+mosaic(hec, type = "expected")
 @
@@
 \setkeys{Gin}{width=0.7\textwidth}
-\begin{figure}[h]
+\begin{figure}[p]
 \begin{center}
@@
 @
-\caption{Mosaic plot for the Titanic data (expected values).}
+\caption{Mosaic plot for the \data{HairEyeColor} data (expected values).}
 \label{fig:expected}
@@
 
-\noindent In order to directly compare observed and expected values,
-we can use a sieve plot (see Figure \ref{fig:sieve}):
+\noindent In order to compare observed and expected values,
+a sieve plot \citep{vcd:riedwyl+schuepbach:1994} 
+could be used (see Figure \ref{fig:sieve}):
 
 <<sieve>>=
-sieve(Titanic)
+sieve(hec)
 @
@@
 @
-\caption{Sieve plot for the Titanic data visualizing simultaneously
+\caption{Sieve plot for the \data{HairEyeColor} data visualizing simultaneously
   observed and expected values.}
@@
 
-\noindent The next step is to directly inspect the residuals. 
+\noindent Alternatively, we can directly inspect the residuals. 
 The Pearson residuals (standardized deviations of observed from expected values) 
 are preferably visualized using association
-plots \citep{vcd:Cohen:1980}. In contrast to \code{assocplot()}, \pkg{vcd}'s \code{assoc()}
+plots \citep{vcd:Cohen:1980}. In contrast to \codefun{assocplot} in
+base {\sf R}, \pkg{vcd}'s \codefun{assoc}
 function scales to more than two variables (see Figure \ref{fig:residuals}):
@@
 <<Residuals>>=
-assoc(Titanic, compress = FALSE)
+assoc(hec, compress = FALSE)
 @
 
-\begin{figure}[h]
+\begin{figure}[p]
 \begin{center}
@@
 @
-\caption{Association plot for the Titanic data.}
+\caption{Association plot for the \data{HairEyeColor} data.}
 \label{fig:residuals}
@@
 
-So far, we have visualized the full table. Parts of the data can be
-selected either using \code{margin.table()} on the input data, or the
-formula interfaces. The following example compares the age
-distribution among classes, given gender, and also shows how to 
-use grid's viewport framework to put two plots besides 
-each other:
+For both mosaic plots and association plots, 
+the splitting of the tiles can be controlled using the
+\code{split\_vertical} argument (default: alternating splits starting
+with a vertical one).
 
-<<Variables>>=
-pushViewport(viewport(layout = grid.layout(ncol = 2)))
-
-pushViewport(viewport(layout.pos.col = 1))
-mosaic(~ Class + Age, data = Titanic[,"Male",,], 
-       margin = c(bottom = 0, 0), sub = "Male", newpage = FALSE)
-popViewport()
-
-pushViewport(viewport(layout.pos.col = 2))
-mosaic(~ Class + Age, data = Titanic[,"Female",,], 
-       margin = c(bottom = 0, 0), sub = "Female", newpage = FALSE)
-popViewport(2)
+<<split1>>=
+mosaic(hec, split_vertical = c(TRUE, TRUE, FALSE))
 @
 
-\setkeys{Gin}{width=\textwidth}
-\begin{figure}[h]
-\begin{center}
-<<Variablesfig,fig=TRUE,echo=FALSE,height=6,width=6>>=
-<<Variables>>
-@
-\caption{Age distribution among classes, given gender.}
-\label{fig:parttable}
-\end{center}
-\end{figure}
-\setkeys{Gin}{width=0.7\textwidth}
+\noindent For compatibility with \codefun{mosaicplot} in base {\sf R},
+the \codefun{mosaic} function also allows the use of a \code{"direction"} 
+argument taking a vector of \code{"h"} and
+\code{"v"} characters (see Figure \ref{fig:split}):
 
-\noindent note the use of the \code{margin} argument: it takes a
-vector with up to four values whose unnamed components are recycled,
-but ``overruled'' by the named arguments. Thus, in the example, only
-the bottom margin is set to 0, leaving the default size of 2 lines
-unchanged for the other margins. This idea applies to almost all
-vectorized arguments in the strucplot framework (with
-\code{split\_vertical} as a prominent exception).
-
-Since mosaic displays are ``conditional plots'' by definition, we can
-also use one single mosaic for stratified plots.
-The formula interface of \code{mosaic()} allows the
-specification of conditioning variables (see Figure \ref{fig:conditioning}):
-
-<<Conditioning>>=
-mosaic(~ Class + Age | Sex, data = Titanic, split = TRUE)
+<<split2>>=
+mosaic(hec, direction = c("v","h","v"))
 @
 
-\begin{figure}[h]
+\begin{figure}[p]
 \begin{center}
-<<Conditioningfig,fig=TRUE,echo=FALSE>>=
-<<Conditioning>>
+<<splitfig,fig=TRUE,echo=FALSE>>=
+<<split1>>
 @
-\caption{Mosaic plot for conditional independence structures.}
-\label{fig:conditioning}
+\caption{Mosaic plot for the \data{HairEyeColor} data---alternative splitting.}
+\label{fig:split}
 \end{center}
@@
 
-\noindent The effect of using this kind of formula is that
-conditioning variables are permuted before the conditioned variables
-in the table, and that \code{spacing\_conditional()} is used to better
-distinguish conditioning from conditioned dimensions.
-Note, however, that the plots in the ``pseudo-strata'' are
-distorted since they are not corrected for the marginal
-distribution(s) of the conditioning variables. Indeed,
-the \code{cotabplot()} function would do a much better job here (see
-next section).
-
-For both mosaic plots and association plots, 
-the splitting of the tiles can be controlled using the
-\code{split\_vertical} argument (default: alternating splits starting
-with a vertical one). Sensible recycling rules apply. 
-For compatibility with \code{mosaicplot()}, the \code{mosaic()} function
-also allows the use of a \code{"direction"} argument taking a vector of \code{"h"} and
-\code{"v"} characters (see Figures \ref{fig:split1} and \ref{fig:split2}):
-
-<<split1>>=
-mosaic(Titanic, direction = "h")
-@
-
-<<split2>>=
-mosaic(Titanic, split_vertical = c(TRUE, TRUE, TRUE, FALSE),
-                labeling_args = list(abbreviate = c(Sex = TRUE, Age = TRUE)))
-@
-
-\begin{figure}[h]
-\begin{center}
-<<split1fig,fig=TRUE,echo=FALSE>>=
-<<split1>>
-@
-\caption{Mosaic plot for the Titanic data---alternative splitting (1).}
-\label{fig:split1}
-\end{center}
-\end{figure}
-
-\begin{figure}[h]
-\begin{center}
-<<split2fig,fig=TRUE,echo=FALSE>>=
-<<split2>>
-@
-\caption{Mosaic plot for the Titanic data---alternative splitting (2).}
-\label{fig:split2}
-\end{center}
-\end{figure}
-
-\section{Higher-level functions}
-
-The last plot pretty much resembles a doubledecker plot, except for
-the spacing. Since spacing is also modular in the strucplot framework,
-the \code{doubledecker()} function was straightforward to implement:
-it really is just a wrapper for \code{mosaic()}, setting the right
-defaults. Figure \ref{fig:doubledecker} shows a doubledecker plot of
-the Titanic data, explaining the probability of Survival by Age, given
-Sex, given Class:
+By a suitable combination of splitting, spacing, and
+labeling settings, the funtions provided by the strucplot framework
+can be customized in a quite flexible way. For example, 
+it was almost trivial to implement doubledecker
+plots---\codefun{doubledecker} is really just a wrapper for 
+\codefun{mosaic}, setting the right defaults. 
+Figure \ref{fig:doubledecker} shows a doubledecker plot of
+the \data{Titanic} data, explaining the probability of survival (`Survived') by Age, given
+Sex, given Class. It is created by:
 
@@
 @
-\caption{Doubledecker plot for the Titanic data.}
+\caption{Doubledecker plot for the \data{Titanic} data.}
 \label{fig:doubledecker}
@@
 
-Another high-level function is \code{pairs.table()}, that is, the S3-method for objects
-of class "table" of the \code{pairs()} generic. This function produces a matrix 
-of strucplots (currently, association or mosaic plots) in the off-diagonal cells, and
-the variable names (with, optionally, univariate statistics) in the diagonal cells. Figure
-\ref{fig:pairs} shows mosaic plots visualizing mutual independence in the lower triangle,
+\section{Conditional and partial views}
+
+So far, we have visualized full tables. For objects of class \code{table},
+conditioning on levels (i.e., choosing a table subset for fixed levels
+of the conditioning variable(s)) is simply done by indexing.
+However, subsetting \code{"structable"} objects is more restrictive 
+because of their inherent conditional structure. Since the variables
+on both the row and the columns side are nested, 
+conditioning is only possible ``outside-in'':
+
+<<subsetting,results=verbatim>>=
+hec
+hec[["Male",]]
+hec[[c("Male", "Brown"),]]
+hec[["Male","Green"]]
+@
+
+\noindent Now, there are several ways for visualizing 
+conditional independence structures. The ``brute force'' method is to
+draw separate plots for the strata. The following example compares 
+the association between hair and eye color, given gender, by using
+subsetting on the flat table and the \pkg{grid} package's viewport 
+framework to visualize the two groups besides each other:
+
+<<Variables>>=
+pushViewport(viewport(layout = grid.layout(ncol = 2)))
+
+pushViewport(viewport(layout.pos.col = 1))
+mosaic(hec[["Male"]], margins = c(top = 2, 0), sub = "Male", newpage = FALSE)
+popViewport()
+
+pushViewport(viewport(layout.pos.col = 2))
+mosaic(hec[["Female"]], margins = c(top = 2, 0), sub = "Female", newpage = FALSE)
+popViewport(2)
+@
+
+\setkeys{Gin}{width=\textwidth}
+\begin{figure}[p]
+\begin{center}
+<<Variablesfig,fig=TRUE,echo=FALSE,height=5,width=12>>=
+<<Variables>>
+@
+\caption{Distribution of hair and eye color, given gender.}
+\label{fig:parttable}
+\end{center}
+\end{figure}
+\setkeys{Gin}{width=0.7\textwidth}
+
+\noindent Note the use of the \code{margins} argument: it takes a
+vector with up to four values whose unnamed components are recycled,
+but ``overruled'' by the named arguments. Thus, in the example, only
+the top margin is set to 2 lines, and all other to 0.
+This idea applies to almost all
+vectorized arguments in the strucplot framework (with
+\code{split\_vertical} as a prominent exception).
+
+Since mosaic displays are ``conditional plots'' by definition, we can
+also use one single mosaic for stratified plots.
+The formula interface of \codefun{mosaic} allows the
+specification of conditioning variables (see Figure \ref{fig:conditioning}):
+
+<<Conditioning>>=
+mosaic(~ Hair + Eye | Sex, data = hec, split_vertical = TRUE)
+@
+
+\begin{figure}[p]
+\begin{center}
+<<Conditioningfig,fig=TRUE,echo=FALSE>>=
+<<Conditioning>>
+@
+\caption{Mosaic plot for conditional independence structures.}
+\label{fig:conditioning}
+\end{center}
+\end{figure}
+
+\noindent The effect of using this kind of formula is that
+conditioning variables are permuted before the conditioned variables
+in the table, and that \codefun{spacing\_conditional} is used as
+default to better
+distinguish conditioning from conditioned dimensions.
+This spacing uses equal space between tiles of conditioned variables,
+and increasing space between tiles of conditioning variables.
+Note, however, that the plots in the ``pseudo-strata'' are
+distorted since they are not corrected for the marginal
+distribution(s) of the conditioning variables.
+
+The \codefun{cotabplot} function does a much better job on this task:
+it arranges stratified strucplot displays 
+in a lattice-like layout, conditioning on variable \emph{levels}.
+The plot in Figure \ref{fig:cotabplot} shows Hair and Eye color, given Sex:
+
+<<cotabplot>>=
+cotabplot(~ Hair + Eye | Sex, data = hec, labeling_args = list(abbreviate = c(Eye = 2)))
+@
+
+\setkeys{Gin}{width=\textwidth}
+\begin{figure}[h]
+\begin{center}
+<<cotabplotfig,fig=TRUE,echo=FALSE,height=4,width=7>>=
+<<cotabplot>>
+@
+\caption{Conditional tabular plot for the \data{HairEyeColor} data.}
+\label{fig:cotabplot}
+\end{center}
+\end{figure}
+\setkeys{Gin}{width=0.7\textwidth}
+
+\noindent The \code{labeling\_args} argument ensures that `Eye' levels
+are abbreviated (See the vignette on labeling for detailed information).
+
+Another high-level function for visualizing conditional independence
+models is \codefun{pairs.table} (\codefun{pairs.structable}), 
+that is, the {\sf S3}-method for objects
+of class \code{"table"} (\code{"structable"}) of the \codefun{pairs} generic. 
+In contrast to \codefun{cotabplot} which is conditioning on variable \emph{levels},
+\codefun{pairs.table} is conditioning on the variables themselves, creating
+partial views of the table. The function produces a matrix 
+having strucplot displays in the off-diagonal cells, and
+the variable names (with, optionally, univariate statistics) in the diagonal cells. 
+Figure \ref{fig:pairs} shows a pairs display with mosaic plots visualizing mutual 
+independence in the lower triangle,
 association plots for the same in the upper triangle, and bar charts in the diagonal.
@@
 <<pairs>>=
-pairs(Titanic, upper_panel = pairs_assoc)
+pairs(hec, upper_panel = pairs_assoc, space = 0.3, 
+      diag_panel_args = list(rot = -45, just_leveltext = c("left","bottom")))
 @
 
-\setkeys{Gin}{width=0.7\textwidth}
+\setkeys{Gin}{width=\textwidth}
 \begin{figure}[h]
@@
 @
-\caption{Pairs plot for the Titanic data.}
+\caption{Pairs plot for the \data{HairEyeColor} data.}
 \label{fig:pairs}
@@
 \end{figure}
+\setkeys{Gin}{width=0.7\textwidth}
 
-\noindent Each cell's row and column define two variables $X$ and $Y$ used for the
+\noindent In plots produced by \codefun{pairs.table}, 
+each cell's row and column define two variables $X$ and $Y$ used for the
 specification of four different types of independence: `\code{pairwise}',
@@
 other variables. The total independence mosaic matrix shows mosaic
-     plots for mutual independence, i.e., for marginal and conditional
-     independence among all pairs of variables. The conditional
+plots for mutual independence, i.e., for marginal and conditional independence
+among all pairs of variables. The conditional
 independence mosaic matrix shows mosaic plots for marginal
-     independence of $X$ and $Y$ given all other variables. The joint independence
+independence of $X$ and $Y$, given all other variables. The joint independence
 mosaic matrix shows mosaic plots for joint independence of all
@@
 
-Since the matrix is symmetric, the upper and lower part can be used to display different 
-types of independence models, or different strucplots (currently, only association and 
-mosaic plots are supported). The available panel functions (\code{pairs\_assoc()} and 
-\code{pairs\_mosaic()}) are just simple wrappers to \code{assoc()} and 
-\code{mosaic()}. Obviously, seeing patterns in strucplot matrices 
+Since the matrix is symmetric, the upper and lower parts can
+independently be used to display different 
+types of independence models, or different strucplots displays
+(mosaic, association, or sieve plots).
+The available core functions (\codefun{pairs\_assoc}, 
+\codefun{pairs\_mosaic}, and \codefun{pairs\_sieve}) 
+are just simple wrappers to \codefun{assoc}, 
+\codefun{mosaic}, and \codefun{sieve}, respectively. 
+Obviously, seeing patterns in strucplot matrices 
 becomes increasingly difficult with
@@
 
-Another useful high-level plot is produced by \code{cotabplot()}: this function can be
-used for plotting stratified strucplots in a lattice-like display. Whereas \code{pairs()} 
-is conditioning on variables, \code{cotabplot()} is conditioning on variable \emph{levels}.
-The plot in Figure \ref{fig:cotabplot} shows the Survival rate among the classes, given
-all four level combinations of Sex and Age.
+\section{Interactive plot modifications}
 
-<<cotabplot>>=
-cotabplot(~ Class + Survived | Sex + Age, data = Titanic)
-@
-
-\setkeys{Gin}{width=\textwidth}
-\begin{figure}[h]
-\begin{center}
-<<cotabplotfig,fig=TRUE,echo=FALSE,height=6,width=6>>=
-<<cotabplot>>
-@
-\caption{cotabplot for the Titanic data.}
-\label{fig:cotabplot}
-\end{center}
-\end{figure}
-\setkeys{Gin}{width=0.7\textwidth}
-
-\section{Technical Notes on Panel Functions}
-
-All strucplot components (including the panel functions) 
-are parameterized through the use of generating functions, that is,
-functions with all parameters in their signature returning a function
-to be called by \code{strucplot()}. This returned function has access
-to all parameters passed to the generating function because in {\sf
-  R}, nested functions are lexically scoped. The idea is to use module-specific `code
-factories' that generate parameterized workhorses rather than
-struggling with monolithic functions with endless parameter lists. As
-a simple example, consider \code{struc\_mosaic()} and \code{struc\_assoc},
-the generating functions for the mosaic and the association plots:
-
-\begin{smallexample}
-struc_mosaic <- function(zero_size = 0.5)
-  function(residuals, observed, expected = NULL, spacing, gp, split_vertical) {
-  \dots
-  }
-
-struc_assoc <- function(compress = TRUE, xlim = NULL, ylim = NULL,
-                        yspace = unit(0.5, "lines"), xscale = 0.9)
-  function(residuals, observed = NULL, expected, spacing, gp, split_vertical) {
-  \dots
-  }
-\end{smallexample}
-
-\noindent Both functions have specific parameters in their signature,
-but both return a function with a standardized interface that
-subsequently is called by \code{strucplot} to produce the actual
-plot. As can be seen, the only information \code{strucplot()} passes
-to the panel functions are residuals, observed and expected
-values, a structure containing the between-tile spacing information,
-an object of class \code{"gpar"} for the tiles' graphical parameters,
-and a logical vector with the splitting information for each
-dimension. The workhorse function, of course, is free to use only
-part or all of this information. High-level interface functions such
-as \code{mosaic()} and \code{assoc{}} call \code{strucplot()} by
-specifying the corresponding function along with the specific
-parameters. Now, there a several ways of doing the same thing, as
-illustrated by the following equivalent calls:
-
-<<panelfuns>>=
-strucplot(Titanic, panel = struc_mosaic(zero_size = 1))
-strucplot(Titanic, panel = struc_mosaic, panel_args = list(zero_size = 1))
-@ 
-<<panelfuns2,eval=FALSE>>=
-strucplot(Titanic, panel = myworkhorse)
-@
-
-\noindent where \code{myworkhorse} could be, e.g., a function with
-hard-wired parameters, but with an interface complying with the
-workhorse signature. To enable \code{strucplot()} to distinguish
-between generating functions and workhorse functions, the generating
-functions must be of class \code{"panel\_generator"}.
-
-All panel functions are supposed to produce conditional hierarchical
+All strucplot core functions are supposed to produce conditional hierarchical
 plots by the means of nested viewports, corresponding to the provided splitting
@@
 labeling functions, to access specific tiles after they have been
-plotted. Note that the viewport tree is popped by default. The
-following example shows how to access parts of the plot after it has
+plotted. The naming convention for the viewports is:
+
+\begin{center}
+\code{cell:\emph{[Variable 1]}=\emph{[Level 1]},\emph{[Variable 2]}=\emph{[Level 2]}} \dots
+\end{center}
+
+\noindent Clearly, these names depend on the splitting.
+The following example shows how to access parts of the plot after it has
 been drawn (see Figure \ref{fig:afterplot}):
@@
 <<viewportnames>>=
-mosaic(~ Sex + Class, data = Titanic, pop = FALSE)
-current.vpTree() ## output not shown
+mosaic(~ Hair + Eye, data = hec, pop = FALSE)
 
-seekViewport("cell:Sex=Male")
+seekViewport("cell:Hair=Blond")
 grid.rect(gp = gpar(col = "red", lwd = 4))
 
-seekViewport("cell:Sex=Male,Class=Crew")
+seekViewport("cell:Hair=Blond,Eye=Blue")
 grid.circle(r = 0.2, gp = gpar(fill = "cyan"))
 @
+
+\noindent Note that the viewport tree is removed by
+default. Therefore, the \texttt{pop} argument has to be set to
+\texttt{FALSE} when viewports shall be accessed.
 
@@
 
-\noindent The naming convention for the viewports is:
-\code{cell..\emph{[Variable 1]}.\emph{[Level 1]}..\emph{[Variable 2]}.\emph{[Level 2]}} \dots. Clearly, these names depend on the splitting.
-
 In addition to the viewports, the main graphical elements get names
 following a similar construction method. This allows to change
-graphical parameters of plot elements after the plotting (see Figure \ref{fig:changeplot}):
+graphical parameters of plot elements \emph{after} 
+the plotting (see Figure \ref{fig:changeplot}):
 
-<<changeplot>>=
-assoc(Titanic)
-getNames() ## output not shown
-grid.edit("rect:Class=1st,Age=Adult,Sex=Female,Survived=Yes", 
-            gp = gpar(fill = "red"))
+<<changeplot,results=verbatim>>=
+assoc(Eye ~ Hair, data = hec, pop = FALSE)
+getNames()[1:6]
+grid.edit("rect:Hair=Blond,Eye=Blue", gp = gpar(fill = "red"))
 @
 
+%% code-chunk reuse does not work with parameter changing
 \begin{figure}[h]
@@
 <<changeplotfig,fig=TRUE,echo=FALSE>>=
-assoc(Titanic)
-x = Titanic
-x[] = NA
-x["1st", "Female", "Adult", "Yes"] = "Red"
-assoc(Titanic, gp = gpar(fill = x))
+x = tab = margin.table(HairEyeColor, 1:2)
+x[] = "light gray"
+x["Blond","Blue"] = "Red"
+assoc(tab, gp = gpar(fill = x))
 @
@@
 
-\begin{thebibliography}{}
-
-\bibitem[Cohen, 1980]{vcd:Cohen:1980}
-Cohen, A. (1980).
-\newblock On the graphical display of the significant components in a two-way
-  contingency table.
-\newblock {\em Communications in Statistics---Theory and Methods},
-  A9:1025--1041.
-
-\bibitem[Hartigan and Kleiner, 1984]{vcd:Hartigan+Kleiner:1984}
-Hartigan, J. and Kleiner, B. (1984).
-\newblock A mosaic of television ratings.
-\newblock {\em The American Statistician}, 38:32--35.
-
-\end{thebibliography}
+\bibliographystyle{apalike}
+\bibliography{vcd}
 

### 2005-09-18 trunk/vcd/R/cotabplot.R

@@ -86,7 +86,7 @@ cotabplot.default <- function(x, cond = NULL,
   }
 
   ## create panel function (if necessary)
-  if(inherits(panel, "panel_generator"))
+  if(inherits(panel, "generating_function"))
     panel <- do.call("panel", c(list(x, cond.char), as.list(panel_args), list(...)))
 
   if(cond.n < 1) panel(x, NULL) ## no conditioning variables
@@ -162,7 +162,7 @@ cotab_mosaic <- function(x = NULL, condvars = NULL, ...) {
                   newpage = FALSE, pop = TRUE, ...)
   }
 }
-class(cotab_mosaic) <- "panel_generator"
+class(cotab_mosaic) <- "generating_function"
 
 cotab_sieve <- function(x = NULL, condvars = NULL, ...) {
   function(x, condlevels) {
@@ -171,7 +171,7 @@ cotab_sieve <- function(x = NULL, condvars = NULL, ...) {
                  newpage = FALSE, pop = TRUE, ...)
   }
 }
-class(cotab_sieve) <- "panel_generator"
+class(cotab_sieve) <- "generating_function"
 
 cotab_assoc <- function(x = NULL, condvars = NULL, ylim = NULL, ...) {
   if(!is.null(x)) {
@@ -185,7 +185,7 @@ cotab_assoc <- function(x = NULL, condvars = NULL, ylim = NULL, ...) {
                   newpage = FALSE, pop = TRUE, ylim = ylim, ...)
   }
 }
-class(cotab_assoc) <- "panel_generator"
+class(cotab_assoc) <- "generating_function"
 
 cotab_coindep <- function(x, condvars,
   test = c("max", "Chisq"), level = NULL, n = 1000,
@@ -279,7 +279,7 @@ cotab_coindep <- function(x, condvars,
 
   return(rval)
 }
-class(cotab_coindep) <- "panel_generator"
+class(cotab_coindep) <- "generating_function"
 
 
 ## Examples:

### 2005-09-18 trunk/vcd/R/labeling.R

@@ -49,7 +49,7 @@ labeling_list <- function(gp = gpar(),
                 gp = gp
                 )
   }
-class(labeling_list) <- "panel_generator"
+class(labeling_list) <- "generating_function"
 
 labeling_conditional <- function(...)
   function (d, split_vertical, condvars) {
@@ -60,7 +60,7 @@ labeling_conditional <- function(...)
     labeling_border(labels = !v, ...)(d, split_vertical, condvars)
     labeling_cells(labels = v, ...)(d, split_vertical, condvars)
   }
-class(labeling_conditional) <- "panel_generator"
+class(labeling_conditional) <- "generating_function"
 
 labeling_cells <- function(labels = TRUE, varnames = TRUE,
                          abbreviate_labels = FALSE, abbreviate_varnames = FALSE,
@@ -130,7 +130,7 @@ labeling_cells <- function(labels = TRUE, varnames = TRUE,
     split()
     
 }
-class(labeling_cells) <- "panel_generator"
+class(labeling_cells) <- "generating_function"
 
 labeling_border <- function(labels = TRUE, varnames = labels,
                             set_labels = NULL, set_varnames = NULL,
@@ -142,7 +142,8 @@ labeling_border <- function(labels = TRUE, varnames = labels,
                             pos_labels = "center", pos_varnames = "center",
                             just_labels = "center", just_varnames = pos_varnames,
                             boxes = FALSE, fill_boxes = FALSE,
-                            offset = c(0, 0, 0, 0),
+                            offset_labels = c(0, 0, 0, 0),
+                            offset_varnames = offset_labels,
                           
                             labbl_varnames = NULL,
                             labels_varnames = FALSE, sep = ": ",
@@ -162,10 +163,16 @@ labeling_border <- function(labels = TRUE, varnames = labels,
     labels_varnames <- pexpand(labels_varnames, ld, FALSE, dn)
     pos_labels <- pexpand(pos_labels, 4, "center", c("top", "right", "bottom", "left"), c("left", "center", "right"))
     just_labels <- pexpand(just_labels, 4, "center", c("top", "right", "bottom", "left"), c("left", "center", "right"))
-    offset <- if (!is.unit(offset))
-      unit(pexpand(offset, 4, rep.int(0, 4), c("top","right","bottom","left")), "lines")
+    offset_varnames <- if (!is.unit(offset_varnames))
+      unit(pexpand(offset_varnames, 4,
[…]

### 2005-09-18 trunk/vcd/R/mosaic.R

@@ -49,23 +49,22 @@ function(formula, data = NULL, ..., main = NULL, subset = NULL)
 }
 
 mosaic.default <- function(x, condvars = NULL,
-                           split_vertical = FALSE, direction = NULL,
+                           split_vertical = NULL, direction = NULL,
                            spacing = NULL, spacing_args = list(),
                            zero_size = 0.5, main = NULL, ...) {
   if (is.logical(main) && main)
     main <- deparse(substitute(x))
-  if (inherits(x, "structable"))
+
+  if (is.structable(x)) {
+    if (is.null(direction) && is.null(split_vertical))
+      split_vertical <- attr(x, "split_vertical")
     x <- as.table(x)
+  }
+  if (is.null(split_vertical))
+    split_vertical <- FALSE
   
   dl <- length(dim(x))
-  if (!is.null(condvars)) {
-    if (is.character(condvars))
-      condvars <- match(condvars, names(dimnames(x)))
-    x <- aperm(x, c(condvars, seq(dl)[-condvars]))
-    if (is.null(spacing))
-      spacing <- spacing_conditional
-  }
-  
+
   ## splitting argument
   if (!is.null(direction))
     split_vertical <- direction == "v"
@@ -74,6 +73,15 @@ mosaic.default <- function(x, condvars = NULL,
   if (length(split_vertical) < dl)
     split_vertical <- rep(split_vertical, length.out = dl)
 
+  ## condvars
+  if (!is.null(condvars)) {
+    if (is.character(condvars))
+      condvars <- match(condvars, names(dimnames(x)))
+    x <- aperm(x, c(condvars, seq(dl)[-condvars]))
+    if (is.null(spacing))
+      spacing <- spacing_conditional
+  }
+  
   ## spacing argument
   if (is.null(spacing))
     spacing <- if (dl < 3) spacing_equal else spacing_increase
@@ -155,4 +163,4 @@ struc_mosaic <- function(zero_size = 0.5)
     }
 
   }
-class(struc_mosaic) <- "panel_generator"
+class(struc_mosaic) <- "generating_function"

### 2005-09-18 trunk/vcd/R/plot.loglm.R

@@ -13,7 +13,7 @@ plot.loglm <- function(x,
   observed <- residuals * sqrt(expected) + expected
   if(residuals_type == "deviance") residuals <- residuals(x, type = "deviance")
   
-  gp <- if(inherits(gp, "panel_generator"))
+  gp <- if(inherits(gp, "generating_function"))
     do.call("gp", c(list(observed, residuals, expected, x$df), as.list(gp_args))) else gp
   
   panel(observed, residuals = residuals, expected = expected, type = type,

### 2005-09-18 trunk/vcd/R/sieve.R

@@ -50,7 +50,7 @@ function(formula, data = NULL, ..., main = NULL, subset = NULL)
 
 sieve.default <- function(x, condvars = NULL, gp = NULL,
                           shade = NULL, legend = FALSE,
-                          split_vertical = FALSE, direction = NULL,
+                          split_vertical = NULL, direction = NULL,
                           spacing = NULL, spacing_args = list(),
                           sievetype = c("observed","expected"), main = NULL, ...) {
   if (is.logical(main) && main)
@@ -62,18 +62,16 @@ sieve.default <- function(x, condvars = NULL, gp = NULL,
     else
       shading_Friendly(interpolate = 0, line_col = "darkgray", eps = Inf, lty = "dotted")
 
-  if (inherits(x, "structable"))
+  if (is.structable(x)) {
+    if (is.null(direction) && is.null(split_vertical))
+      split_vertical <- attr(x, "split_vertical")
     x <- as.table(x)
+  }
+  if (is.null(split_vertical))
+    split_vertical <- FALSE
   
   dl <- length(dim(x))
-  if (!is.null(condvars)) {
-    if (is.character(condvars))
-      condvars <- match(condvars, names(dimnames(x)))
-    x <- aperm(x, c(condvars, seq(dl)[-condvars]))
-    if (is.null(spacing))
-      spacing <- spacing_conditional
-  }
-  
+
   ## splitting argument
   if (!is.null(direction))
     split_vertical <- direction == "v"
@@ -82,6 +80,15 @@ sieve.default <- function(x, condvars = NULL, gp = NULL,
   if (length(split_vertical) < dl)
     split_vertical <- rep(split_vertical, length.out = dl)
 
+  ## condvars
+  if (!is.null(condvars)) {
+    if (is.character(condvars))
+      condvars <- match(condvars, names(dimnames(x)))
+    x <- aperm(x, c(condvars, seq(dl)[-condvars]))
+    if (is.null(spacing))
+      spacing <- spacing_conditional
+  }
+  
   ## spacing argument
   if (is.null(spacing))
     spacing <- if (dl < 3) spacing_equal else spacing_increase
@@ -187,4 +194,4 @@ struc_sieve <- function(sievetype = c("observed", "expected")) {
     }
   }
 }
-class(struc_sieve) <- "panel_generator"
+class(struc_sieve) <- "generating_function"

### 2005-09-18 trunk/vcd/R/structable.R

@@ -5,7 +5,7 @@ structable <- function(x, ...)
   UseMethod("structable")
 
 structable.formula <- function(formula, data = NULL, direction = NULL,
-                               split_vertical = FALSE, ..., subset, na.action) {
+                               split_vertical = NULL, ..., subset, na.action) {
     if (missing(formula) || !inherits(formula, "formula")) 
         stop("formula is incorrect or missing")
 
@@ -14,7 +14,15 @@ structable.formula <- function(formula, data = NULL, direction = NULL,
 
     if (!is.null(direction))
       split_vertical <- direction == "v"
-        
+
+    if (is.structable(data)) {
+      split_vertical <- attr(data, "split_vertical")
+      data <- as.table(data)
+    }
+    
+    if (is.null(split_vertical))
+      split_vertical <- FALSE
+    
     ## only rhs present without `.' in lhs => xtabs-interface
     if (length(formula) != 3) {
       if (formula[[1]] == "~") {
@@ -167,40 +175,58 @@ structable.default <- function(..., direction = NULL, split_vertical = FALSE) {
   ## handle calls like x[[c(1,2), c(3,4)]]
   if (length(..1) > 1 && length(..2) > 1)
     return(x[[ ..1[1], ..2[[1]] ]] [[ ..1[-1], ..2[-1] ]])
+  
   ## handle calls like x[[c(1,2), 3]]
   if (length(..1) > 1)
     return(x[[ ..1[1], ..2 ]] [[ ..1[-1], ]])
+  
   ## handle calls like x[[1, c(1,3)]]
   if (length(..2) > 1)
     return(x[[ ..1, ..2[[1]] ]] [[ , ..2[-1] ]])
 
   ## final cases like x[[1,2]] or x[[1,]] or x[[,1]]
-  rv <- attr(x, "dnames")[!attr(x, "split_vertical")]
-  cv <- attr(x, "dnames")[attr(x, "split_vertical")]
+  dnames <- attr(x, "dnames")
+  split <- attr(x, "split_vertical")
+  rv <- dnames[!split]
+  cv <- dnames[split]
 
-  if (!is.symbol(..1)) rstep <- dim(x)[1] / length(rv[[1]])
-  if (!is.symbol(..2)) cstep <- dim(x)[2] / length(cv[[1]])
[…]

### 2005-09-18 trunk/vcd/R/assoc.R
- vignettes revised - bugs in structable indexing fixed - mosaic, sieve, and assoc made structable aware - panel_generator class -> generating_function - offset_labels and offset_varnames added to labeling_border - spacing_highlighting replaces spacing_doubledecker
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@457 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -33,30 +33,19 @@ assoc.default <- function(x,
   if (!inherits(x, "ftable"))
     x <- structable(x)
 
-#     {
-#     if (is.null(row_vars) && is.null(col_vars) && is.table(x))
-#       row_vars <- names(dimnames(x))[seq(1, length(dim(x)), by = 2)]
-#     x <- ftable(x, row.vars = row_vars, col.vars = col_vars)
-#   }
-
   tab <- as.table(x)
   dl <- length(dim(tab))
   
   ## spacing
   cond <- rep(TRUE, dl)
   cond[length(attr(x, "row.vars")) + c(0, length(attr(x, "col.vars")))] <- FALSE
-  if (inherits(spacing, "panel_generator"))
+  if (inherits(spacing, "generating_function"))
     spacing <- do.call("spacing", spacing_args)
   spacing <- spacing(dim(tab), condvars = which(cond))
 
   ## splitting arguments
   if (is.null(split_vertical))
     split_vertical <- attr(x, "split_vertical")
-  if (is.null(split_vertical)) {
-    split_vertical <- rep(FALSE, dl)
-    names(split_vertical) <- names(dimnames(tab))
-    split_vertical[names(attr(x, "col.vars"))] <- TRUE
-  }
 
   if(match.arg(tolower(residuals_type), "pearson") != "pearson")
     warning("Only Pearson residuals can be visualized with association plots.")
@@ -176,4 +165,4 @@ struc_assoc <- function(compress = TRUE, xlim = NULL, ylim = NULL,
     }
 
   }
-class(struc_assoc) <- "panel_generator"
+class(struc_assoc) <- "generating_function"

### 2005-09-18 trunk/vcd/R/legends.R

@@ -80,7 +80,7 @@ legend_resbased <- function(fontsize = 12,
     }
   }
 }
-class(legend_resbased) <- "panel_generator"
+class(legend_resbased) <- "generating_function"
 
 legend_fixed <- function(fontsize = 12,
                          x = unit(1, "lines"),
@@ -148,4 +148,4 @@ legend_fixed <- function(fontsize = 12,
               )
   }
 }
-class(legend_fixed) <- "panel_generator"
+class(legend_fixed) <- "generating_function"

### 2005-09-18 trunk/vcd/R/pairsplot.R

@@ -20,11 +20,11 @@ pairs.table <- function(x,
 {
   if (newpage) grid.newpage()
 
-  if (inherits(upper_panel, "panel_generator"))
+  if (inherits(upper_panel, "generating_function"))
     upper_panel <- do.call("upper_panel", upper_panel_args)
-  if (inherits(lower_panel, "panel_generator"))
+  if (inherits(lower_panel, "generating_function"))
     lower_panel <- do.call("lower_panel", lower_panel_args)
-  if (inherits(diag_panel, "panel_generator"))
+  if (inherits(diag_panel, "generating_function"))
     diag_panel <- do.call("diag_panel", diag_panel_args)
   
   d <- length(dim(x))
@@ -57,16 +57,18 @@ pairs.table <- function(x,
   invisible(x)
 }
 
+pairs.structable <- function(x, ...) pairs(as.table(x), ...)
+
 ## upper/lower panels
 
 pairs_assoc <- function(...) pairs_strucplot(panel = assoc, ...)
-class(pairs_assoc) <- "panel_generator"
+class(pairs_assoc) <- "generating_function"
 
 pairs_mosaic <- function(...) pairs_strucplot(panel = mosaic, ...)
-class(pairs_mosaic) <- "panel_generator"
+class(pairs_mosaic) <- "generating_function"
 
 pairs_sieve <- function(...) pairs_strucplot(panel = sieve, ...)
-class(pairs_sieve) <- "panel_generator"
+class(pairs_sieve) <- "generating_function"
 
 pairs_strucplot <- function(panel = mosaic,
                             type = c("pairwise", "total", "conditional", "joint"),
@@ -96,7 +98,7 @@ pairs_strucplot <- function(panel = mosaic,
            ...)
   }
 }
-class(pairs_strucplot) <- "panel_generator"
+class(pairs_strucplot) <- "generating_function"
   
 ## diagonal panels
 
@@ -114,11 +116,14 @@ pairs_text <- function(dimnames = TRUE,
       grid.text(paste("(",paste(names(x), collapse = ","), ")", sep = ""),
                 y = 0.4, gp = gp_leveltext)
   }
-class(pairs_text) <- "panel_generator"
+class(pairs_text) <- "generating_function"
 
 pairs_barplot <- function(gp_bars = gpar(fill = "gray"),
                           gp_vartext = gpar(fontsize = 17),
                           gp_leveltext = gpar(),
+                          just_leveltext = c("center", "bottom"),
[…]

### 2005-09-18 trunk/vcd/R/shadings.R

@@ -87,7 +87,7 @@ shading_hsv <- function(observed, residuals = NULL, expected = NULL, df = NULL,
   attr(rval, "p.value") <- p.value
   return(rval)
 }
-class(shading_hsv) <- "panel_generator"
+class(shading_hsv) <- "generating_function"
 
 
 shading_hcl <- function(observed, residuals = NULL, expected = NULL, df = NULL,
@@ -181,7 +181,7 @@ shading_hcl <- function(observed, residuals = NULL, expected = NULL, df = NULL,
   attr(rval, "p.value") <- p.value
   return(rval)
 }
-class(shading_hcl) <- "panel_generator"
+class(shading_hcl) <- "generating_function"
 
 shading_Friendly <- function(observed = NULL, residuals = NULL, expected = NULL, df = NULL,
   h = c(2/3, 0), lty = 1:2, interpolate = c(2, 4), eps = 0.01, line_col = "black", ...)
@@ -190,7 +190,7 @@ shading_Friendly <- function(observed = NULL, residuals = NULL, expected = NULL,
               h = h, v = 1, lty = lty, interpolate = interpolate,
 	      eps = eps, line_col = line_col, p.value = NA, ...)
 }
-class(shading_Friendly) <- "panel_generator"
+class(shading_Friendly) <- "generating_function"
 
 shading_max <- function(observed = NULL, residuals = NULL, expected = NULL, df = NULL,
   h = NULL, c = NULL, l = NULL, lty = 1, eps = NULL, line_col = "black", level = c(0.9, 0.99), n = 1000, ...)
@@ -209,30 +209,28 @@ shading_max <- function(observed = NULL, residuals = NULL, expected = NULL, df =
 			eps = eps, line_col = line_col, p.value = obs.test$p.value, ...)
   return(rval)
 }
-class(shading_max) <- "panel_generator"
+class(shading_max) <- "generating_function"
 
 shading_binary <- function(observed = NULL, residuals = NULL, expected = NULL, df = NULL,
   col = hcl(c(260, 0), 50, 70))
 {
-  ## get col
-  my.col <- rep(col, length.out = 2)
+  ## check col argument
+  if (length(col) != 2) stop("Need exactly two colors!")
   
   ## store color information for legend
-  legend <- list(col = my.col[2:1], col.bins = 0,
-                 lty = NULL, lty.bins = NULL)
[…]

### 2005-09-18 trunk/vcd/R/spacings.R

@@ -5,14 +5,14 @@ spacing_equal <- function(sp = unit(0.5, "lines")) {
   if (!is.unit(sp)) sp <- unit(sp, "lines")
   function(d, condvars = NULL) lapply(d, function(x) unit.rep(sp, x - 1))
 }
-class(spacing_equal) <- "panel_generator"
+class(spacing_equal) <- "generating_function"
 
 spacing_dimequal <- function(sp) {
   if (!is.unit(sp)) sp <- unit(sp, "lines")
   function(d, condvars = NULL)
     lapply(seq(along = d), function(i) unit.rep(sp[i], d[[i]] - 1))
 }
-class(spacing_dimequal) <- "panel_generator"
+class(spacing_dimequal) <- "generating_function"
 
 spacing_increase <- function(start = unit(0.3, "lines"), rate = 1.5) {
   if (!is.unit(start)) start <- unit(start, "lines")
@@ -21,17 +21,15 @@ spacing_increase <- function(start = unit(0.3, "lines"), rate = 1.5) {
     lapply(seq(along = d), function(i) unit.rep(sp[i], d[[i]] - 1))
   }
 }
-class(spacing_increase) <- "panel_generator"
+class(spacing_increase) <- "generating_function"
 
-spacing_doubledecker <- function(start = unit(0.3, "lines"), rate = 1.8)
-  function(d, condvars) {
-    if (length(d) < 3) {
-      start[2] <- 0
-      spacing_dimequal(start)(d, condvars)
-    } else
-      spacing_conditional(sp = 0, start = start, rate = rate)(d, condvars)
-  }
-class(spacing_doubledecker) <- "panel_generator"
+spacing_highlighting <- function(start = unit(0.2, "lines"), rate = 1.5) {
+ if (!is.unit(start)) start <- unit(start, "lines")
+  function(d, condvars = NULL)
+    c(spacing_increase(start, rate)(d, condvars)[-length(d)],
+      list(unit(rep(0, d[length(d)]), "lines")))
+}
+class(spacing_highlighting) <- "generating_function"
 
 spacing_conditional <- function(sp = unit(0.5, "lines"),
                                 start = unit(2, "lines"), rate = 1.8) {
@@ -51,4 +49,4 @@ spacing_conditional <- function(sp = unit(0.5, "lines"),
     ret
   }
 }
-class(spacing_conditional) <- "panel_generator"
+class(spacing_conditional) <- "generating_function"

### 2005-09-18 trunk/vcd/R/strucplot.R

@@ -13,7 +13,7 @@ strucplot <- function(## main parameters
                       residuals_type = c("Pearson", "deviance", "FT"),
                       
                       ## layout
-                      split_vertical = TRUE, 
+                      split_vertical = NULL, 
                       spacing = spacing_equal,
                       spacing_args = list(),
                       gp = NULL,
@@ -42,6 +42,15 @@ strucplot <- function(## main parameters
   type <- match.arg(type)
   residuals_type <- match.arg(tolower(residuals_type), c("pearson", "deviance", "ft"))
 
+  ## convert structable object
+  if (is.structable(x)) {
+    if (is.null(split_vertical))
+      split_vertical <- attr(x, "split_vertical")
+    x <- as.table(x)
+  }
+  if (is.null(split_vertical))
+    split_vertical <- FALSE
+  
   ## table characteristics
   d <- dim(x)
   dl <- length(d)
@@ -96,7 +105,7 @@ strucplot <- function(## main parameters
 
   ## spacing
   if (is.function(spacing)) {
-    if (inherits(spacing, "panel_generator"))
+    if (inherits(spacing, "generating_function"))
       spacing <- do.call("spacing", spacing_args)
     spacing <- spacing(d, condvars)
   }
@@ -107,7 +116,7 @@ strucplot <- function(## main parameters
     if (is.function(gp)) {
       if (is.null(legend) || (is.logical(legend) && legend))
         legend <- legend_resbased
-      gpfun <- if(inherits(gp, "panel_generator"))
+      gpfun <- if(inherits(gp, "generating_function"))
         do.call("gp", c(list(x, residuals, expected, df), as.list(gp_args))) else gp
       gp <- gpfun(residuals)
     } else if (!is.null(legend) && !(is.logical(legend) && !legend))
@@ -139,7 +148,7 @@ strucplot <- function(## main parameters
                            legend_width = legend_width))
 
   ## legend
-  if (inherits(legend, "panel_generator"))
+  if (inherits(legend, "generating_function"))
     legend <- do.call("legend", legend_args)
   if (shade && !is.null(legend) && !(is.logical(legend) && !legend)) {
     seekViewport("legend")
@@ -163,7 +172,7 @@ strucplot <- function(## main parameters
   ## make plot
   seekViewport("plot")
   
-  if (inherits(panel, "panel_generator"))
+  if (inherits(panel, "generating_function"))
     panel <- do.call("panel", panel_args)
[…]

### 2005-09-26 trunk/vcd/inst/doc/strucplot.Rnw
bug fixes: legend scales like plot now, and strucplot works with utf8 characters.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@462 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
+\setkeys{Gin}{width=0.6\textwidth}
 \begin{figure}[h]
 \begin{center}
-<<vcdlayout,fig=TRUE,echo=FALSE,width=6,height=5>>=
+<<vcdlayout,fig=TRUE,echo=FALSE,width=6,height=6>>=
     pushViewport(vcd:::vcdViewport(legend = T, mar =4))

### 2005-09-26 trunk/vcd/R/legends.R
bug fixes: legend scales like plot now, and strucplot works with utf8 characters.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@462 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -85,7 +85,7 @@ class(legend_resbased) <- "generating_function"
 legend_fixed <- function(fontsize = 12,
                          x = unit(1, "lines"),
                          y = unit(0.25, "npc"),
-                         height = unit(0.65, "npc"),
+                         height = unit(0.75, "npc"),
                          width = unit(1.5, "lines"),
                          steps = 200,
 			 digits = 3,
@@ -141,7 +141,7 @@ legend_fixed <- function(fontsize = 12,
                   default.unit = "npc")
 
     popViewport(1)
-    grid.text(text, x = x + 0.5 * width, y = 0.1,
+    grid.text(text, x = x + 0.5 * width, y = 0,
               gp = gpar(fontsize = fontsize, lineheight = 0.8),
               just = c("left", "top"),
               rot = 90

### 2005-09-26 trunk/vcd/R/strucplot.R

@@ -144,7 +144,8 @@ strucplot <- function(## main parameters
   
   pushViewport(vcdViewport(mar = margins,
                            legend = shade && !(is.null(legend) || is.logical(legend) && !legend),
-                           main = !is.null(main), sub = !is.null(sub), keep_aspect_ratio = keep_aspect_ratio,
+                           main = !is.null(main), sub = !is.null(sub),
+                           keep_aspect_ratio = keep_aspect_ratio,
                            legend_width = legend_width))
 
   ## legend
@@ -194,9 +195,8 @@ strucplot <- function(## main parameters
 
   seekViewport("base") 
   ## one more up if sandwich-mode
-  if (!is.null(main) || !is.null(sub) ##||
-##      (shade && !is.null(legend) && !(is.logical(legend) && !legend))
-      ) upViewport()
+  if (!is.null(main) || !is.null(sub) ||
+      (shade && !(is.null(legend) || is.logical(legend) && !legend))) upViewport()
   if (pop) popViewport(1 + keep_aspect_ratio) else upViewport(1 + keep_aspect_ratio)
 
   ## return visualized table
@@ -209,6 +209,7 @@ vcdViewport <- function(mar = rep.int(2.5, 4),
                         legend = FALSE, main = FALSE, sub = FALSE,
                         keep_aspect_ratio = TRUE)
 {
+  if (legend) main = sub = TRUE
   mar <- if (!is.unit(mar))
     unit(pexpand(mar, 4, rep.int(2.5, 4), c("top","right","bottom","left")), "lines")
   else
@@ -252,32 +253,29 @@ vcdViewport <- function(mar = rep.int(2.5, 4),
   }
 
   ## main/sub-title, margins for legend layout
-  if (main || sub) {
+  if (main || sub || legend) {
     vpTop <- viewport(layout.pos.row = 1, name = "main")
     vpSub <- viewport(layout.pos.row = 2 + main, name = "sub")
     
     sandwich <-
-## no additional space when keep_aspect_ratio = F
-#       if (legend) {
-#       space <- max(legend_width + mar[2] + mar[4] - mar[1] - mar[3],
-#                    unit((main + sub) * 2, "lines"))
-#       vplist <- vpList(vpTop, vpPlotregion, vpSub)
-#       viewport(layout = grid.layout(3, 1,
[…]

### 2005-09-28 trunk/vcd/inst/doc/strucplot.Rnw
- structable-indexing now using [] - replacement function for structables - vcdViewport-layout fixed (again), title_margins argument added - strucplots accept structables as gp argument
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@463 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 hec
-hec[["Male",]]
-hec[[c("Male", "Brown"),]]
-hec[["Male","Green"]]
+hec["Male",]
+hec[c("Male", "Brown"),]
+hec["Male","Green"]
 @
@@
 pushViewport(viewport(layout.pos.col = 1))
-mosaic(hec[["Male"]], margins = c(top = 2, 0), sub = "Male", newpage = FALSE)
+mosaic(hec["Male"], margins = c(top = 2, 0), sub = "Male", newpage = FALSE)
 popViewport()
@@
 pushViewport(viewport(layout.pos.col = 2))
-mosaic(hec[["Female"]], margins = c(top = 2, 0), sub = "Female", newpage = FALSE)
+mosaic(hec["Female"], margins = c(top = 2, 0), sub = "Female", newpage = FALSE)
 popViewport(2)

### 2005-09-28 trunk/vcd/R/assoc.R

@@ -5,12 +5,14 @@ assoc <- function(x, ...)
   UseMethod("assoc")
 
 assoc.formula <-
-function(formula, data = NULL, subset, na.action, ..., main = NULL)
+function(formula, data = NULL, subset, na.action, ..., main = NULL, sub = NULL)
 {
     if (is.logical(main) && main)
       main <- deparse(substitute(data))
+    else if (is.logical(sub) && sub)
+      sub <- deparse(substitute(data))
 
-    assoc.default(ftable(formula, data, subset, na.action), main = main, ...)
+    assoc.default(ftable(formula, data, subset, na.action), main = main, sub = sub, ...)
 }
 
 assoc.default <- function(x,
@@ -23,12 +25,15 @@ assoc.default <- function(x,
 			  residuals_type = "Pearson",
                           xscale = 0.9, yspace = unit(0.5, "lines"),
                           main = NULL,
+                          sub = NULL,
                           ...,
                           gp_axis = gpar(lty = 3)
                           ) {
 
   if (is.logical(main) && main)
     main <- deparse(substitute(x))
+  else if (is.logical(sub) && sub)
+    sub <- deparse(substitute(x))
 
   if (!inherits(x, "ftable"))
     x <- structable(x)
@@ -57,7 +62,8 @@ assoc.default <- function(x,
               yspace = yspace, xscale = xscale, gp_axis = gp_axis),
             keep_aspect_ratio = keep_aspect_ratio,
             residuals_type = "Pearson",
-            main = main, 
+            main = main,
+            sub = sub,
             ...)
 }
 
@@ -76,11 +82,11 @@ struc_assoc <- function(compress = TRUE, xlim = NULL, ylim = NULL,
     rfunc <- function(x) c(min(x, 0), max(x, 0))
     if (is.null(ylim))
       ylim <- if (compress)
-        matrix(apply(resid, 1, rfunc), nrow = 2)
+        matrix(apply(as.matrix(resid), 1, rfunc), nrow = 2)
       else
-        rfunc(resid)
+        rfunc(as.matrix(resid))
     if (!is.matrix(ylim))
-      ylim <- matrix(ylim, nrow = 2, ncol = nrow(resid))
+      ylim <- matrix(as.matrix(ylim), nrow = 2, ncol = nrow(as.matrix(resid)))
 
     attr(ylim, "split_vertical") <- rep(TRUE, sum(!split_vertical))
     attr(ylim, "dnames") <- dn[!split_vertical]
[…]

### 2005-09-28 trunk/vcd/R/sieve.R

@@ -5,10 +5,12 @@ sieve <- function(x, ...)
   UseMethod("sieve")
 
 sieve.formula <-
-function(formula, data = NULL, ..., main = NULL, subset = NULL)
+function(formula, data = NULL, ..., main = NULL, sub = NULL, subset = NULL)
 {
   if (is.logical(main) && main)
     main <- deparse(substitute(data))
+  else if (is.logical(sub) && sub)
+    sub <- deparse(substitute(data))
   
   m <- match.call(expand.dots = FALSE)
   edata <- eval(m$data, parent.frame())
@@ -34,7 +36,7 @@ function(formula, data = NULL, ..., main = NULL, subset = NULL)
       }
       dat <- margin.table(dat, ind)
     }
-    sieve.default(dat, main = main,
+    sieve.default(dat, main = main, sub = sub,
                    condvars = if (is.null(condind)) NULL else match(condnames, names(dimnames(dat))), ...)
   } else {
     tab <- if ("Freq" %in% colnames(data))
@@ -44,7 +46,7 @@ function(formula, data = NULL, ..., main = NULL, subset = NULL)
       xtabs(formula(paste("~", paste(c(condnames, varnames), collapse = "+"))),
             data = data, subset = subset)
     
-    sieve.default(tab, main = main, ...)
+    sieve.default(tab, main = main, sub = sub, ...)
   }
 }
 
@@ -52,9 +54,12 @@ sieve.default <- function(x, condvars = NULL, gp = NULL,
                           shade = NULL, legend = FALSE,
                           split_vertical = NULL, direction = NULL,
                           spacing = NULL, spacing_args = list(),
-                          sievetype = c("observed","expected"), main = NULL, ...) {
+                          sievetype = c("observed","expected"),
+                          main = NULL, sub = NULL, ...) {
   if (is.logical(main) && main)
     main <- deparse(substitute(x))
+  else if (is.logical(sub) && sub)
+    sub <- deparse(substitute(x))
   sievetype = match.arg(sievetype)
   if (is.logical(shade) && shade && is.null(gp))
     gp <- if (sievetype == "observed")
@@ -100,6 +105,7 @@ sieve.default <- function(x, condvars = NULL, gp = NULL,
             spacing = spacing,
             spacing_args = spacing_args,
             main = main,
+            sub = sub,
             shade = shade, 
             legend = legend,
             gp = gp,

### 2005-09-28 trunk/vcd/R/strucplot.R

@@ -27,7 +27,8 @@ strucplot <- function(## main parameters
                       
                       main = NULL,
                       sub = NULL,
-                      margins = rep.int(2.5, 4),
+                      margins = unit(2.5, "lines"),
+                      title_margins = NULL,
                       legend_width = unit(5, "lines"),
                       
                       ## control parameters
@@ -133,7 +134,11 @@ strucplot <- function(## main parameters
 
   ## recycle gpar values in the last dimension
   size <- prod(d)
-  FUN <- function(par) if (length(par) < size) aperm(array(par, dim = rev(d))) else par
+  FUN <- function(par) {
+    if (is.structable(par))
+      par <- as.table(par)
+    if (length(par) < size) aperm(array(par, dim = rev(d))) else par
+  }
   gp <- structure(lapply(gp, FUN), class = "gpar")
   
   ## set up page
@@ -143,6 +148,7 @@ strucplot <- function(## main parameters
     pushViewport(viewport(width = 1, height = 1, default.unit = "snpc"))
   
   pushViewport(vcdViewport(mar = margins,
+                           oma = title_margins,
                            legend = shade && !(is.null(legend) || is.logical(legend) && !legend),
                            main = !is.null(main), sub = !is.null(sub),
                            keep_aspect_ratio = keep_aspect_ratio,
@@ -167,6 +173,8 @@ strucplot <- function(## main parameters
 
   if (!is.null(sub)) {
     seekViewport("sub")
+    if (is.logical(sub) && sub && is.null(main))
+      sub <- deparse(substitute(x))
     grid.text(sub, gp = title_gp)
   }
 
@@ -206,16 +214,35 @@ strucplot <- function(## main parameters
 
 vcdViewport <- function(mar = rep.int(2.5, 4),
                         legend_width = unit(5, "lines"),
+                        oma = NULL,
                         legend = FALSE, main = FALSE, sub = FALSE,
                         keep_aspect_ratio = TRUE)
 {
-  if (legend) main = sub = TRUE
+  ## process parameters
+  if (legend && !main && !sub) main <- sub <- TRUE
   mar <- if (!is.unit(mar))
     unit(pexpand(mar, 4, rep.int(2.5, 4), c("top","right","bottom","left")), "lines")
[…]

### 2005-09-28 trunk/vcd/R/structable.R

@@ -162,7 +162,7 @@ structable.default <- function(..., direction = NULL, split_vertical = FALSE) {
   ret
 }
 
-"[[.structable" <- function(x, ...) {
+"[.structable" <- function(x, ...) {
   args <- if (nargs() < 3)
     list(..1)
   else
@@ -171,25 +171,25 @@ structable.default <- function(..., direction = NULL, split_vertical = FALSE) {
   ## handle one-arg cases
   if (nargs() < 3)
     if (length(args[[1]]) > 1)
-      ## resolve calls like x[[c(1,2)]]
-      return(x[[ args[[1]][1] ]] [[ args[[1]][-1] ]])
+      ## resolve calls like x[c(1,2)]
+      return(x[ args[[1]][1] ] [ args[[1]][-1] ])
     else
-      ## resolve x[[foo]] 
-      return(if (attr(x, "split_vertical")[1]) x[[,args[[1]]]] else x[[args[[1]],]])
+      ## resolve x[foo] 
+      return(if (attr(x, "split_vertical")[1]) x[,args[[1]] ] else x[args[[1]],])
 
-  ## handle calls like x[[c(1,2), c(3,4)]]
+  ## handle calls like x[c(1,2), c(3,4)]
   if (length(args[[1]]) > 1 && length(args[[2]]) > 1)
-    return(x[[ args[[1]][1], args[[2]][[1]] ]] [[ args[[1]][-1], args[[2]][-1] ]])
+    return(x[ args[[1]][1], args[[2]][1] ] [ args[[1]][-1], args[[2]][-1] ])
   
-  ## handle calls like x[[c(1,2), 3]]
+  ## handle calls like x[c(1,2), 3]
   if (length(args[[1]]) > 1)
-    return(x[[ args[[1]][1], args[[2]] ]] [[ args[[1]][-1], ]])
+    return(x[ args[[1]][1], args[[2]] ] [ args[[1]][-1], ])
   
-  ## handle calls like x[[1, c(1,3)]]
+  ## handle calls like x[1, c(1,3)]
   if (length(args[[2]]) > 1)
-    return(x[[ args[[1]], args[[2]][[1]] ]] [[ , args[[2]][-1] ]])
+    return(x[ args[[1]], args[[2]][1] ] [ , args[[2]][-1] ])
 
-  ## final cases like x[[1,2]] or x[[1,]] or x[[,1]]
+  ## final cases like x[1,2] or x[1,] or x[,1]
[…]

### 2005-09-28 trunk/vcd/R/mosaic.R
- structable-indexing now using [] - replacement function for structables - vcdViewport-layout fixed (again), title_margins argument added - strucplots accept structables as gp argument
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@463 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -5,11 +5,13 @@ mosaic <- function(x, ...)
   UseMethod("mosaic")
 
 mosaic.formula <-
-function(formula, data = NULL, ..., main = NULL, subset = NULL)
+function(formula, data = NULL, ..., main = NULL, sub = NULL, subset = NULL)
 {
   if (is.logical(main) && main)
     main <- deparse(substitute(data))
-  
+  else if (is.logical(sub) && sub)
+    sub <- deparse(substitute(data))
+
   m <- match.call(expand.dots = FALSE)
   edata <- eval(m$data, parent.frame())
   
@@ -34,7 +36,7 @@ function(formula, data = NULL, ..., main = NULL, subset = NULL)
       }
       dat <- margin.table(dat, ind)
     }
-    mosaic.default(dat, main = main,
+    mosaic.default(dat, main = main, sub = sub,
                    condvars = if (is.null(condind)) NULL else match(condnames, names(dimnames(dat))), ...)
   } else {
       m <- m[c(1, match(c("formula", "data", "subset"), names(m), 0))]
@@ -44,16 +46,18 @@ function(formula, data = NULL, ..., main = NULL, subset = NULL)
                         "~",
                         paste(c(condnames, varnames), collapse = "+")))
       tab <- eval(m, parent.frame())
-      mosaic.default(tab, main = main, ...)  
+      mosaic.default(tab, main = main, sub = sub, ...)  
   }
 }
 
 mosaic.default <- function(x, condvars = NULL,
                            split_vertical = NULL, direction = NULL,
                            spacing = NULL, spacing_args = list(),
-                           zero_size = 0.5, main = NULL, ...) {
+                           zero_size = 0.5, main = NULL, sub = NULL, ...) {
   if (is.logical(main) && main)
     main <- deparse(substitute(x))
+  else if (is.logical(sub) && sub)
+    sub <- deparse(substitute(x))
 
   if (is.structable(x)) {
     if (is.null(direction) && is.null(split_vertical))
@@ -93,6 +97,7 @@ mosaic.default <- function(x, condvars = NULL,
             spacing = spacing,
             spacing_args = spacing_args,
             main = main,
+            sub = sub,
             ...)
 }
 

### 2005-09-29 trunk/vcd/R/sieve.R
class attributes added to grapcon functions; class of generating functions expanded; checks added to strucplot.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@465 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -114,7 +114,7 @@ sieve.default <- function(x, condvars = NULL, gp = NULL,
 
 struc_sieve <- function(sievetype = c("observed", "expected")) {
   sievetype = match.arg(sievetype)
-  function(residuals, observed, expected, spacing, gp, split_vertical) {
+  ret <- function(residuals, observed, expected, spacing, gp, split_vertical) {
     dn <- dimnames(expected)
     dnn <- names(dn)
     dx <- dim(expected)
@@ -199,5 +199,6 @@ struc_sieve <- function(sievetype = c("observed", "expected")) {
       grid.rect(name = paste("rect:", mnames[i], sep = ""))
     }
   }
+  structure(ret, class = c("grapcon","core"))
 }
-class(struc_sieve) <- "grapcon_generator"
+class(struc_sieve) <- c("grapcon_generator","core")

### 2005-09-29 trunk/vcd/inst/doc/shading.Rnw
class attributes added to grapcon functions; class of generating functions expanded; checks added to strucplot.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@465 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 shade_fun <- function(res) gpar(fill = ifelse(res > 0, "lightblue", "lightsalmon"))
+class(shade_fun) <- c("grapcon","shading")
 @
 
-\noindent To create a mosaic display with binary shading, 
+\noindent Note the \code{class} attribute set on \codefun{shade\_fun}.
+To create a mosaic display with binary shading, 
 it now suffices to specify the data table along with
@@
   if (length(col) != 2) stop("Need exactly two colors!")
-  function(res) gpar(fill = ifelse(res > 0, col[1], col[2]))  
+  ret <- function(res) gpar(fill = ifelse(res > 0, col[1], col[2]))  
+  structure(ret, class = c("grapcon","shading"))
 }
@@
   if (length(col) != 2) stop("Need exactly two colors!")
-  function(res) gpar(fill = ifelse(res > 0, col[1], col[2]))  
+  ret <- function(res) gpar(fill = ifelse(res > 0, col[1], col[2]))  
+  structure(ret, class = c("grapcon","shading"))
 }
-class(myshading_binary2) <- "generating_function"
+class(myshading_binary2) <- c("grapcon_generator","shading")
 @

### 2005-09-29 trunk/vcd/R/assoc.R

@@ -69,8 +69,8 @@ assoc.default <- function(x,
 
 struc_assoc <- function(compress = TRUE, xlim = NULL, ylim = NULL,
                         yspace = unit(0.5, "lines"), xscale = 0.9,
-                        gp_axis = gpar(lty = 3))
-  function(residuals, observed = NULL, expected, spacing, gp, split_vertical) {
+                        gp_axis = gpar(lty = 3)) {
+  ret <- function(residuals, observed = NULL, expected, spacing, gp, split_vertical) {
     dn <- dimnames(expected)
     dnn <- names(dn)
     dx <- dim(expected)
@@ -169,6 +169,7 @@ struc_assoc <- function(compress = TRUE, xlim = NULL, ylim = NULL,
                 name = paste("rect:", mnames[i], sep = "")
                 )
     }
-
   }
-class(struc_assoc) <- "grapcon_generator"
+  structure(ret, class = c("grapcon","core"))
+}
+class(struc_assoc) <- c("grapcon_generator","core")

### 2005-09-29 trunk/vcd/R/labeling.R

@@ -27,8 +27,8 @@ labeling_list <- function(gp = gpar(),
                           offset = unit(c(2, 2), "lines"),
                           varnames = TRUE,
                           cols = 2,
-                          ...) 
-  function(d, split_vertical, condvars) {
+                          ...) {
+  ret <- function(d, split_vertical, condvars) {
     if (is.table(d))
       d <- dimnames(d)
     ld <- length(d)
@@ -49,10 +49,12 @@ labeling_list <- function(gp = gpar(),
                 gp = gp
                 )
   }
-class(labeling_list) <- "grapcon_generator"
+  structure(ret, class = c("grapcon","labeling"))
+}
+class(labeling_list) <- c("grapcon_generator","labeling")
 
-labeling_conditional <- function(...)
-  function (d, split_vertical, condvars) {
+labeling_conditional <- function(...) {
+  ret <- function (d, split_vertical, condvars) {
     if (is.table(d))
       d <- dimnames(d)
     v <- rep.int(TRUE, length(d))
@@ -60,15 +62,17 @@ labeling_conditional <- function(...)
     labeling_border(labels = !v, ...)(d, split_vertical, condvars)
     labeling_cells(labels = v, ...)(d, split_vertical, condvars)
   }
-class(labeling_conditional) <- "grapcon_generator"
+  structure(ret, class = c("grapcon","labeling"))
+}
+class(labeling_conditional) <- c("grapcon_generator","labeling")
 
 labeling_cells <- function(labels = TRUE, varnames = TRUE,
                          abbreviate_labels = FALSE, abbreviate_varnames = FALSE,
                          gp = gpar(), lsep = ": ", lcollapse = "\n",
                          just = "center", pos = "center", rot = 0,
                          margin = unit(0.5, "lines"), clip_cells = TRUE,
-                         text = NULL, ...)
-  function(d, split_vertical, condvars) {
+                         text = NULL, ...) {
+  ret <- function(d, split_vertical, condvars) {
     if (is.table(d))
       d <- dimnames(d)
     dn <- names(d)
@@ -128,9 +132,10 @@ labeling_cells <- function(labels = TRUE, varnames = TRUE,
       }
     }
     split()
-    
+  }
+  structure(ret, class = c("grapcon","labeling"))
 }
-class(labeling_cells) <- "grapcon_generator"
[…]

### 2005-09-29 trunk/vcd/R/legends.R

@@ -15,7 +15,7 @@ legend_resbased <- function(fontsize = 12,
   if(!is.unit(width)) width <- unit(width, "lines")
   if(!is.unit(height)) height <- unit(height, "npc")
   
-  function(residuals, shading, autotext) {
+  ret <- function(residuals, shading, autotext) {
     res <- as.vector(residuals)
     
     if(is.null(text)) text <- autotext
@@ -82,8 +82,9 @@ legend_resbased <- function(fontsize = 12,
                 just = c("left", "top"))
     }
   }
+  structure(ret, class = c("grapcon","legend"))
 }
-class(legend_resbased) <- "grapcon_generator"
+class(legend_resbased) <- c("grapcon_generator","legend")
 
 legend_fixed <- function(fontsize = 12,
                          x = unit(1, "lines"),
@@ -100,7 +101,7 @@ legend_fixed <- function(fontsize = 12,
   if(!is.unit(width)) width <- unit(width, "lines")
   if(!is.unit(height) && !is.null(height)) height <- unit(height, "npc")
 
-  function(residuals, shading, autotext) {
+  ret <- function(residuals, shading, autotext) {
     res <- as.vector(residuals)
 
     if(is.null(text)) text <- autotext
@@ -158,5 +159,6 @@ legend_fixed <- function(fontsize = 12,
               rot = 90
               )
   }
+  structure(ret, class = c("grapcon","legend"))
 }
-class(legend_fixed) <- "grapcon_generator"
+class(legend_fixed) <- c("grapcon_generator","legend")

### 2005-09-29 trunk/vcd/R/mosaic.R

@@ -101,8 +101,8 @@ mosaic.default <- function(x, condvars = NULL,
             ...)
 }
 
-struc_mosaic <- function(zero_size = 0.5)
-  function(residuals, observed, expected = NULL, spacing, gp, split_vertical) {
+struc_mosaic <- function(zero_size = 0.5) {
+  ret <- function(residuals, observed, expected = NULL, spacing, gp, split_vertical) {
     dn <- dimnames(observed)
     dnn <- names(dn)
     dx <- dim(observed)
@@ -166,6 +166,7 @@ struc_mosaic <- function(zero_size = 0.5)
         }
       }
     }
-
   }
-class(struc_mosaic) <- "grapcon_generator"
+  structure(ret, class = c("grapcon","core"))
+}
+class(struc_mosaic) <- c("grapcon_generator","core")

### 2005-09-29 trunk/vcd/R/shadings.R

@@ -85,9 +85,10 @@ shading_hsv <- function(observed, residuals = NULL, expected = NULL, df = NULL,
   }
   attr(rval, "legend") <- legend
   attr(rval, "p.value") <- p.value
+  class(rval) <- c("grapcon","shading")
   return(rval)
 }
-class(shading_hsv) <- "grapcon_generator"
+class(shading_hsv) <- c("grapcon_generator","shading")
 
 
 shading_hcl <- function(observed, residuals = NULL, expected = NULL, df = NULL,
@@ -179,9 +180,10 @@ shading_hcl <- function(observed, residuals = NULL, expected = NULL, df = NULL,
   }
   attr(rval, "legend") <- legend
   attr(rval, "p.value") <- p.value
+  class(rval) <- c("grapcon","shading")
   return(rval)
 }
-class(shading_hcl) <- "grapcon_generator"
+class(shading_hcl) <- c("grapcon_generator","shading")
 
 shading_Friendly <- function(observed = NULL, residuals = NULL, expected = NULL, df = NULL,
   h = c(2/3, 0), lty = 1:2, interpolate = c(2, 4), eps = 0.01, line_col = "black", ...)
@@ -190,7 +192,7 @@ shading_Friendly <- function(observed = NULL, residuals = NULL, expected = NULL,
               h = h, v = 1, lty = lty, interpolate = interpolate,
 	      eps = eps, line_col = line_col, p.value = NA, ...)
 }
-class(shading_Friendly) <- "grapcon_generator"
+class(shading_Friendly) <- c("grapcon_generator","shading")
 
 shading_max <- function(observed = NULL, residuals = NULL, expected = NULL, df = NULL,
   h = NULL, c = NULL, l = NULL, lty = 1, eps = NULL, line_col = "black", level = c(0.9, 0.99), n = 1000, ...)
@@ -204,12 +206,11 @@ shading_max <- function(observed = NULL, residuals = NULL, expected = NULL, df =
   
   obs.test <- coindep_test(observed, n = n)
   col.bins <- obs.test$qdist(sort(level))
-  rval <- shading_hcl(observed = NULL, residuals = NULL, expected = NULL, df = NULL,
-                        h = h, c = c, l = l, interpolate = col.bins, lty = lty,
-			eps = eps, line_col = line_col, p.value = obs.test$p.value, ...)
-  return(rval)
+  shading_hcl(observed = NULL, residuals = NULL, expected = NULL, df = NULL,
[…]

### 2005-09-29 trunk/vcd/R/spacings.R

@@ -3,40 +3,44 @@
 
 spacing_equal <- function(sp = unit(0.5, "lines")) {
   if (!is.unit(sp)) sp <- unit(sp, "lines")
-  function(d, condvars = NULL) lapply(d, function(x) unit.rep(sp, x - 1))
+  ret <- function(d, condvars = NULL) lapply(d, function(x) unit.rep(sp, x - 1))
+  structure(ret, class = c("grapcon","spacing"))
 }
-class(spacing_equal) <- "grapcon_generator"
+class(spacing_equal) <- c("grapcon_generator","spacing")
 
 spacing_dimequal <- function(sp) {
   if (!is.unit(sp)) sp <- unit(sp, "lines")
-  function(d, condvars = NULL)
+  ret <- function(d, condvars = NULL)
     lapply(seq(along = d), function(i) unit.rep(sp[i], d[[i]] - 1))
+  structure(ret, class = c("grapcon","spacing"))
 }
-class(spacing_dimequal) <- "grapcon_generator"
+class(spacing_dimequal) <- c("grapcon_generator","spacing")
 
 spacing_increase <- function(start = unit(0.3, "lines"), rate = 1.5) {
   if (!is.unit(start)) start <- unit(start, "lines")
-  function(d, condvars = NULL) {
+  ret <- function(d, condvars = NULL) {
     sp <- start * rev(cumprod(c(1, rep.int(rate, length(d) - 1))))
     lapply(seq(along = d), function(i) unit.rep(sp[i], d[[i]] - 1))
   }
+  structure(ret, class = c("grapcon","spacing"))
 }
-class(spacing_increase) <- "grapcon_generator"
+class(spacing_increase) <- c("grapcon_generator","spacing")
 
 spacing_highlighting <- function(start = unit(0.2, "lines"), rate = 1.5) {
- if (!is.unit(start)) start <- unit(start, "lines")
-  function(d, condvars = NULL)
+  if (!is.unit(start)) start <- unit(start, "lines")
+  ret <- function(d, condvars = NULL)
     c(spacing_increase(start, rate)(d, condvars)[-length(d)],
       list(unit(rep(0, d[length(d)]), "lines")))
+  structure(ret, class = c("grapcon","spacing"))
 }
-class(spacing_highlighting) <- "grapcon_generator"
+class(spacing_highlighting) <- c("grapcon_generator","spacing")
 
 spacing_conditional <- function(sp = unit(0.5, "lines"),
[…]

### 2005-09-29 trunk/vcd/R/strucplot.R

@@ -108,8 +108,10 @@ strucplot <- function(## main parameters
 
   ## spacing
   if (is.function(spacing)) {
-    if (inherits(spacing, "grapcon_generator"))
+    if (inherits(spacing, "grapcon_generator") && inherits(spacing, "spacing"))
       spacing <- do.call("spacing", spacing_args)
+    if (!(inherits(spacing, "grapcon") && inherits(spacing, "spacing")))
+      stop("Spacing function does not inherit from classes 'grapcon' and 'core'.")
     spacing <- spacing(d, condvars)
   }
 
@@ -119,14 +121,16 @@ strucplot <- function(## main parameters
     if (is.function(gp)) {
       if (is.null(legend) || (is.logical(legend) && legend))
         legend <- legend_resbased
-      gpfun <- if(inherits(gp, "grapcon_generator"))
+      gpfun <- if (inherits(gp, "grapcon_generator") && inherits(gp, "shading"))
         do.call("gp", c(list(x, residuals, expected, df), as.list(gp_args))) else gp
+      if (!(inherits(gpfun, "grapcon") && inherits(gpfun, "shading")))
+        stop("Shading function does not inherit from classes 'grapcon' and 'shading'.")
       gp <- gpfun(residuals)
     } else if (!is.null(legend) && !(is.logical(legend) && !legend))
       stop("gp argument must be a shading function for drawing a legend")
   } else {
     if(!is.null(gp)) {
-      warning("gp parameter ignored since shade=FALSE")
+      warning("gp parameter ignored since shade = FALSE")
       gp <- NULL
     }
   }
@@ -157,9 +161,11 @@ strucplot <- function(## main parameters
                            legend_width = legend_width))
 
   ## legend
-  if (inherits(legend, "grapcon_generator"))
+  if (inherits(legend, "grapcon_generator") && inherits(legend, "legend"))
     legend <- do.call("legend", legend_args)
   if (shade && !is.null(legend) && !(is.logical(legend) && !legend)) {
+    if (!(inherits(legend, "grapcon") && inherits(legend, "legend")))
+      stop("Legend function does not inherit from classes 'grapcon' and 'legend'.")
     seekViewport("legend")
     residuals_type = switch(residuals_type, deviance = "deviance", ft = "Freeman-Tukey", pearson = "Pearson")
     legend(residuals, gpfun, paste(residuals_type, "residuals:", sep = "\n"))
[…]

### 2005-10-02 trunk/vcd/inst/doc/shading.Rnw
Vignettes overworked.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@466 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2005-09-18 trunk/vcd/inst/doc/shading.Rnw
@@
 \newcommand{\code}[1]{\texttt{#1}}
+\newcommand{\class}[1]{\textsf{#1}}
 \newenvironment{smallexample}{\begin{alltt}}{\end{alltt}}
@@
 
-%\VignetteIndexEntry{Colors and residual-based Shadings in Strucplots}
+%\VignetteIndexEntry{Colors and Residual-based Shadings in the Strucplot Framework}
 %\VignetteDepends{vcd}
@@
 
-\title{Colors and Residual-based Shadings in Strucplots}
-\author{by David Meyer\\
-  Wirtschaftsuniversit\"at Wien, Austria\\
-\email{David.Meyer@R-project.org}
-}
+\title{Colors and Residual-based Shadings in the Strucplot Framework}
+\author{by David Meyer, Achim Zeileis, and Kurt Hornik}
+\date{}
 \maketitle
@@
 
+\section{Introduction}
+
 Unlike other graphics functions in base {\sf R}, the strucplot framework 
 allows almost full control over the graphical parameters of all plot elements. In
-particular, the user can \emph{individually} modify the graphical appearance of  
-tiles in association plots, mosaic plots, and sieve plots. 
+particular, in association plots, mosaic plots, and sieve plots, 
+the user can modify the graphical appearance of each tile individually. 
 Built on top of this functionality, the
@@
 The tiles' graphical parameters are set using the \code{gp} argument
-of the strucplot functions. This argument basically expects an object
-of class \code{gpar} whose components are arrays of the same shape 
-(length and dimensionality) than the data table (see Section \ref{sec:gp}). 
+of the functions of the strucplot framework. This argument basically expects an object
+of class \class{gpar} whose components are arrays of the same shape 
+(length and dimensionality) as the data table (see Section \ref{sec:gp}). 
 For convenience, however, the user can also
-supply a function that computes such an object given a vector of
+supply a specialized graphical appearance control (``grapcon'') 
+function that computes such an object given a vector of
 residuals, or, alternatively, a generating function that takes
-parameters and returns such a function (see Section
+certain arguments and returns such a grapcon function (see Section
 \ref{sec:shading}). We provide several shading functions, including
-support for both the HSV and the HCL color space, and the
-visualization of significance tests (see Section
-\ref{sec:overview}).
+support for both HSV and HCL colors, and the
+visualization of significance tests (see Section \ref{sec:overview}).
 
@@
 As an example, consider the \data{UCBAdmissions} data. 
-In the table collapsed over departments, we
+In the table aggregated over departments, we
 would like to highlight the (incidentally wrong) impression that there
-were to much male students accepted compared to the presumably
+were too many male students accepted compared to the presumably
 discriminated female students (see Figure \ref{fig:ucb}):
@@
 magenta for rejected female students) and supply them to the \code{fill} component 
-of the \code{gpar} object passed to the \code{gp} argument of \codefun{mosaic}. 
-For optical clarity, we addtionally hide the tiles' borders by setting the \code{col} 
+of the \class{gpar} object passed to the \code{gp} argument of \codefun{mosaic}. 
+For visual clarity, we additionally hide the tiles' borders by setting the \code{col} 
 component to 0 (white).
 
-If the parameters specified in the \code{gpar} object are incomplete,
-strucplot functions will recycle them along the last splitting
+If the parameters specified in the \class{gpar} object are ``incomplete'',
+they will be recycled along the last splitting
 dimension. In the following example based on the \data{Titanic} data, 
@@
 between tiles in the last dimension to 0. The \code{labeling\_args} 
-argument ensures that labels do not overlapp (see the separate
-vignette on labeling for more information).
+argument ensures that labels do not overlap (see the separate
+vignette: ``Labeling in the Strucplot Framework'' for more information).
 
@@
 shading functions that modify the tiles' appearance with respect to a vector of 
-residuals, resulting from deviations of the observed from the expected
+residuals, resulting from deviations of observed from expected
 values under a given log-linear model. 
-The idea is to visualize at least the size of the residuals, but some
+The idea is to visualize at least sign and absolute size of the residuals, but some
 shadings, additionally, indicate overall significance. One particular
@@
   encapsulating them into an object of class
-  \code{"gpar"} as demonstrated in the previous section, and
+  \class{gpar} as demonstrated in the previous section, and
   passing this object to the \code{gp} argument.
-\item Providing a function to the \code{gp} argument that 
-  takes residuals as input and returns an object as described in 1.
-\item Providing a \emph{generating} function taking parameters and returning
-  a function as described in 2. The parameters of this generating
-  function are either directly specified (which is not possible for
-  all shading functions), or as a list to the \code{shading\_args} argument.
+\item Providing a grapcon function to the \code{gp} argument that 
+  takes residuals as input and returns an object as described in
+  alternative 1.
+\item Providing a grapcon \emph{generating} function taking parameters and returning
+  a function as described in alternaive 2. 
 \end{enumerate}
@@
 fill color (for simplicity, we will treat 0 as positive).
+
+\subsubsection*{Alternative 1: precomputed \class{gpar} object}
 
@@
 
-(fillcols <- ifelse(residuals > 0, "lightblue", "lightsalmon"))
+(shading1_obj <- ifelse(residuals > 0, "lightblue", "lightsalmon"))
 
-mosaic(ucb, gp = gpar(fill = fillcols))
+mosaic(ucb, gp = gpar(fill = shading1_obj))
 @
@@
 
-For implementing method 2, we need to create a `shading function'
-that---technically speaking---computes \code{"gpar"} objects from
+\subsubsection*{Alternative 2: grapcon function}
+
+For implementing alternative 2, we need to create a `shading function'
+that computes \class{gpar} objects from
 residuals. For that, we can just reuse the code from the previous step:
@@
 <<shading2>>=
-shade_fun <- function(res) gpar(fill = ifelse(res > 0, "lightblue", "lightsalmon"))
+shading2_fun <- function(res) gpar(fill = ifelse(res > 0, "lightblue", "lightsalmon"))
+class(shading2_fun) <- c("grapcon","shading")
 @
 
-\noindent To create a mosaic display with binary shading, 
+\noindent Note the \code{class} attribute set on \codefun{shading2\_fun}.
+To create a mosaic display with binary shading, 
 it now suffices to specify the data table along with
-\codefun{shade\_fun}:
+\codefun{shading2\_fun}:
 
 <<shading3>>=
-mosaic(ucb, gp = shade_fun)
+mosaic(ucb, gp = shading2_fun)
 @
@@
 independence model (total independence by default), passes them to
-\codefun{shade\_fun}, and uses the \code{"gpar"} object returned 
+\codefun{shading2\_fun}, and uses the \class{gpar} object returned 
 to finally create the plot.
 
-Our \codefun{shade\_fun} function might be useful, 
-but suffers from a major flaw: the colors are hard-wired. We cannot
+Our \codefun{shading2\_fun} function might be useful, 
+but can still be improved: the hard-wired colors should be
+customizable. We cannot
 simply extend the argument list to include, e.g., 
@@
 they are expected to take exactly one argument: a table of residuals. 
-This is where generating functions (method 3) come
-into play. We simply wrap our shading function in another 
+This is where generating functions (alternative 3) come
+into play.
+
+\subsubsection*{Alternative 3: grapcon-generating function}
+
+We simply wrap our grapcon shading function in another 
 function that takes all additional arguments it needs to use, possibly
-preprocesses them, and returns the actual shading function. This
+preprocesses them, and returns the actual shading function. This returned
 function will have access to the parameters since in {\sf R}, nested
-functions are lexically scoped. This kind of wrapping function
-is also called a \emph{generating function} since it returns (`creates') a
-`parameterized' shading function with the minimal standard interface
+functions are lexically scoped. Thus, the grapcon generator
+returns (`creates') a `parameterized' shading function with the minimal standard interface
 \codefun{strucplot} requires. The follwing example shows the necessary
@@
 <<shading3>>=
-myshading_binary <- function(col = c("lightblue", "lightsalmon")) {
+shading3a_fun <- function(col = c("lightblue", "lightsalmon")) {
   if (length(col) != 2) stop("Need exactly two colors!")
-  function(res) gpar(fill = ifelse(res > 0, col[1], col[2]))  
+  ret <- function(res) gpar(fill = ifelse(res > 0, col[1], col[2]))  
+  structure(ret, class = c("grapcon","shading"))
 }
@@
 \noindent In the call of \codefun{mosaic} (figure not shown), 
-using the new \codefun{myshading\_binary} function, we can now simply change the colors:
+using the new \codefun{shading3a\_fun} function, we can now simply change the colors:
 
 <<shading4>>=
-mosaic(ucb, gp = myshading_binary(c("red","blue")))
+mosaic(ucb, gp = shading3a_fun(c("red","blue")))
 @
@@
 (labeling, spacing, legend, and core functions) follows the same idea.
-Now for the shading functions, this is not the whole story. Since shading
-functions can be used for visualizing significance (see next section), it
-makes sense for generating functions to have access to the model,
+Now for the shading functions, more customization is needed. Note that
+\codefun{shading3a\_fun} needs to be evaluated by the user, even if the
+defaults shall be used. It is a better idea to let \codefun{strucplot} call
+the generating function, which, in particular, allows the passing of arguments that
+are computed by \codefun{strucplot}. Since shading
+functions can be used for visualizing significance (see Section 
+\ref{sec:overview}), it makes sense for generating functions to have access to the model,
 i.e., observed and expected values, residuals, and degrees of
@@
 <<shading4>>=
-myshading_binary2 <- function(observed = NULL, residuals = NULL, expected = NULL, 
+shading3b_fun <- function(observed = NULL, residuals = NULL, expected = NULL, 
     df = NULL, col = c("lightblue", "lightsalmon")) {
   if (length(col) != 2) stop("Need exactly two colors!")
-  function(res) gpar(fill = ifelse(res > 0, col[1], col[2]))  
+  ret <- function(res) gpar(fill = ifelse(res > 0, col[1], col[2]))  
+  structure(ret, class = c("grapcon","shading"))
 }
-class(myshading_binary2) <- "generating_function"
+class(shading3b_fun) <- c("grapcon_generator","shading")
 @
@@
 the ``normal'' shading functions. For others
-(like our simple \codefun{myshading\_binary2}) this is optional, but
+(like our simple \codefun{shading3b\_fun}) this is optional, but
 recommended for consistency:
@@
 <<shading5>>=
-mosaic(ucb, gp = myshading_binary2, gp_args = list(c("red","blue")))
+mosaic(ucb, gp = shading3b_fun, gp_args = list(col = c("red","blue")))
 @
@@
 
-Apart from \codefun{shading\_binary}, there are currently two basic shadings available:
+\cite{vcd:Friendly:1994}
+suggested a residual-based shading for the mosaic tiles that can also be applied
+to the rectangles in association plots \citep{vcd:Meyer+Zeileis+Hornik:2003}.
+Apart from \codefun{shading\_binary}, there are currently two basic
+shadings available in \pkg{vcd}:
 \codefun{shading\_hcl} and \codefun{shading\_hsv}, as well
@@
 In the HSV space, colors are specified in three dimensions: Hue,
-Saturation (`colorfulness'), and Value (`brightness', amount of gray). 
+Saturation (`colorfulness'), and Value (`lightness', amount of gray). 
 These three dimensions are used by \codefun{shading\_hsv} to visualize information 
@@
 saturation for small residuals. 
-Finally, the overall brightness is used to indicate the significance of
-a test statistic: bright colors for significant, 
+Finally, the overall lightness is used to indicate the significance of
+a test statistic: light colors for significant, 
 and dark colors for non-significant results.
@@
 \begin{center}
-<<haireyefig,fig=TRUE,echo=FALSE>>=
+<<haireyefig,fig=TRUE,echo=FALSE,width=7,height=7>>=
 <<haireye>>
@@
 \noindent Large positive residuals (greater than $4$) can be found for
-brown eyes/black hair and blue eyes/blond hair, they are colored in saturated red. 
+brown eyes/black hair and blue eyes/blond hair, and are colored in saturated blue. 
 On the other hand, there is a large negative residual
-(less than $-4$) for brown eyes/blond hair, colored deep blue. There are also three
+(less than $-4$) for brown eyes/blond hair, colored deep red. There are also three
 medium-sized positive (negative) residuals between 2 and 4 ($-2$ and
@@
 residuals \emph{individually} significant at approximately the $\alpha
-= 0.05$ and $\alpha = 0.0001$ level, respectively. These default cut-off points can
+= 0.05$ and $\alpha = 0.0001$ levels, respectively. These default cut-off points can
 be changed to alternative values using the \code{interpolate}
@@
 
-\noindent The \code{interpolate} argument also accepts a function,
+\noindent The elements of the numeric vector passed to
+\code{interpolate} define the knots of an interpolating step function
+used to map the absolute residuals to saturation levels. 
+The \code{interpolate} argument also accepts a user-defined function,
 which then is called with the absolute residuals to get a vector of
@@
 to a continuous shading---visualizing the whole range of 
-significance levels---could be useful. We simply need a one-to-one
-mapping from the residuals to the $p$ values:
+residuals---could be useful. We simply need a one-to-one
+mapping from the residuals to the saturation values:
 
@@
 \noindent Note that this \codefun{ipol} 
-function maps residuals greater than 4 to a significance
+function maps residuals greater than 4 to a saturation
 level of 1. However, the resulting plot (Figure \ref{fig:continuous}) is 
@@
 mosaic(HairEyeColor, gp = shading_hsv, gp_args = list(interpolate = ipol),
-labeling_args = list(abbreviate = c(Sex = 3)))
+labeling_args = list(abbreviate = c(Sex = TRUE)))
 @
@@
 computes it, to the \code{p.value} argument. Such a function must take
-observed and expected values, residuals, and degrees of freedom as arguments. 
+observed and expected values, residuals, and degrees of freedom (used
+by the independence model) as arguments. 
 If nothing is specified, the $p$ value is computed from 
@@
 freedom. The \code{level} argument is used to specify the confidence
-level:  if \code{p.value} is smaller than \code{1 - level}, bright colors are used, 
+level:  if \code{p.value} is smaller than \code{1 - level}, light colors are used, 
 otherwise dark colors are employed. 
@@
 shows the relationship of home goals and away
-goals of the German premier's soccer league in 1995: although there are two
+goals of Germany's premier soccer league in 1995: although there are two
 ``larger'' residuals (one greater than 2, one less then $-2$), the
@@
 A shading function building upon \codefun{shading\_hsv} is \codefun{shading\_Friendly},
-miming the shading introduced by \cite{vcd:Friendly:1994}. In addition
+implementing the shading introduced by \cite{vcd:Friendly:1994}. In addition
 to the defaults of the HSV shading, it uses the border color and line type to 
@@
 
-\noindent (The \code{zeroSize = 0} argument removes the bullets indicating zero 
-observed values. This feature does not
-exist in the original {\sf SAS} implementation of the Friendly mosaic plots.)
+\noindent (The \code{zero\_size = 0} argument removes the bullets indicating zero 
+observed values. This feature is not provided
+in the original {\sf SAS} implementation of the Friendly mosaic plots.)
 
@@
 perceptual characteristics of the HCL color space.
-Figure \ref{fig:shadingHSVHCL} depicts color scales for red and blue hue
-(left and right side, respectively) in the HSV and HCL spaces (upper and lower part, respectively), varying colorfulness and brightness ($x$ and $y$-axis, respectively).
+Figure \ref{fig:shadingHSVHCL} depicts 
+the HSV space in the upper panel and the HCL space in the lower panel.
+On the left (right) side, we see the color scales for red (blue)
+hue, respectively. The $x$-axis represents the colorfulness, and the
+$y$-axis the brightness.
 The boxes represent the diverging color palettes used for the shadings.
-For the HSV space, we can can see that the effect of changing the 
+For the HSV space, we can see that the effect of changing the 
 level of brightness (`value') is not the same for different levels of 
@@
 obviously is problematic for coding information. In contrast, the HCL color
-space offers perceptual uniform colors: as can be seen from the lower part, 
+space offers perceptually uniform colors: as can be seen from the lower panel, 
 the chroma is homogeneous for different levels of luminance. 
@@
 \noindent The maximum test is significant although the residuals are
-all in the $[-2,2]$ interval, as the cut-off points are even smaller.
+all in the $\left[-2,2\right]$ interval.
 The \codefun{shading\_hcl} function with default cut-off points would
 not have shown any color. In addition, since the test statistic is the
-maximum of the absolute Pearson residuals, \emph{every} colored
+maximum of the absolute Pearson residuals, \emph{each} colored
 residual violates the null hypotheses of independence, and thus, the
@@
 
-\bibliographystyle{apalike}
+\bibliographystyle{jss}
 \bibliography{vcd}

### 2005-10-02 trunk/vcd/inst/doc/strucplot.Rnw
Vignettes overworked.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@466 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-%\VignetteIndexEntry{strucplots---Visualizing higher-dimensional contigency tables in vcd}
+%\VignetteIndexEntry{The Strucplot Framework---Visualizing Multi-way Contingency Tables}
 %\VignetteDepends{vcd}
@@
 \title{The Strucplot Framework---Visualizing Multi-way Contingency Tables}
-\author{by David Meyer\\
-  Wirtschaftsuniversit\"at Wien, Austria\\
-\email{David.Meyer@R-project.org}
-}
+\author{by David Meyer, Achim Zeileis, and Kurt Hornik}
+\date{}
+
 \maketitle
@@
 the tables' cells arranged in rectangular form. For multi-way tables,
-the variables are nested into row and columns using recursive
+the variables are nested into rows and columns using recursive
 conditional splits, given the margins. The result is a 
-`flat' representation that can be visualized in similar
-ways than a two-dimensional table.
-This principle defines a class of conditional displays with still many parameters
-left such as:
+`flat' representation that can be visualized in
+ways similar to a two-dimensional table.
+This principle defines a class of conditional displays which allows
+for granular control of graphical appearance aspects, including:
 
@@
 
-\noindent The document at hand gives an introduction to the framework,
+\noindent This document gives an introduction to the framework,
 whereas labeling and shading issues are described in separate vignettes.
 
-The strucplot framework is highly modularized; Figure \ref{fig:struc}
+The strucplot framework is highly modularized: Figure \ref{fig:struc}
 shows the hierarchical relationship between the various components. 
@@
 parameter functions that directly or indirectly influence the final
-appearance of the plot. The workhorse functions (\code{\emph{struc\_foo()}}, 
+appearance of the plot. These `graphical appearance control'
+(`grapcon') functions are created by generating functions, allowing
+flexible parameterization and extensibility. 
+The first part of their names (\code{\emph{group\_foo()}}) 
+reflects the group they belong to (strucplot core, labeling,
+legend, shading, or spacing). The workhorse functions (created by 
+\code{\emph{struc\_foo()}}, 
 \code{\emph{labeling\_foo()}}, and \code{\emph{legend\_foo()}}) 
-directly produce graphical output, the parameter functions
-(\code{\emph{spacing\_foo()}} and \code{\emph{shading\_foo()}}) compute
-graphical parameters used by the others. The \code{\emph{struc\_foo()}} functions
-implement the core functionality, creating the tiles and their
-content. On the second level, a suitable combination of these functions
+directly produce graphical output (``add ink to the canvas''), whereas
+the parameter functions
+(created by \code{\emph{spacing\_foo()}} and \code{\emph{shading\_foo()}}) compute
+graphical parameters used by the others. The grapcon functions returned by 
+\code{\emph{struc\_foo()}} implement the core functionality, creating the tiles and their
+content.  On the second level of the framework, a suitable combination
+of grapcon functions (or, alternatively, corresponding generating functions)
 is passed as ``hyperparameters'' to \codefun{strucplot}.
-These central function 
+This central function 
 sets up the graphical layout using grid viewports (see Figure \ref{fig:layout}),
@@
 the plot. On the third level, we provide
-several high-level functions such as \codefun{mosaic},
+several convenience functions such as \codefun{mosaic},
 \codefun{sieve}, \codefun{assoc}, and \codefun{doubledecker} which 
-conveniently interface \codefun{strucplot} through sensible parameter defaults
+interface \codefun{strucplot} through sensible parameter defaults
 and support for model formulas. Finally, on the fourth
-level, there are `related' \pkg{vcd} functions (such as \codefun{pairs.table} and
-\codefun{cotabplot}) combining single plots of the strucplot 
-framework into more complex displays.
+level, there are `related' \pkg{vcd} functions (such as \codefun{cotabplot}
+and the \codefun{pairs} methods for table objects) 
+arranging collections of plots of the strucplot 
+framework into more complex displays (e.g., by means of panel functions).
 
@@
     pushViewport(vcd:::vcdViewport(legend = T, mar =4))
+    seekViewport("main")
+    grid.rect(gp = gpar(lwd = 3))
+    grid.text("main", gp = gpar(fontsize = 20))
+    seekViewport("sub")
+    grid.rect(gp = gpar(lwd = 3))
+    grid.text("sub", gp = gpar(fontsize = 20))
     seekViewport("plot")
     grid.rect(gp = gpar(lwd = 3))
-    grid.text("plot", gp = gpar(fontsize = 30))
+    grid.text("plot", gp = gpar(fontsize = 20))
     seekViewport("legend")
-    grid.text("legend", rot = 90, gp = gpar(fontsize = 30))
+    grid.text("legend", rot = 90, gp = gpar(fontsize = 20))
     grid.rect(gp = gpar(lwd = 3))
-    seekViewport("pval")
+    seekViewport("legend_sub")
     grid.rect(gp = gpar(lwd = 3))
-    grid.text("pval", gp = gpar(fontsize = 30))
+    grid.text("[F]", gp = gpar(fontsize = 20))
+    seekViewport("legend_top")
+    grid.rect(gp = gpar(lwd = 3))
+    grid.text("[E]", gp = gpar(fontsize = 20))
     seekViewport("margin_top")
     grid.rect(gp = gpar(lwd = 3))
-    grid.text("margin_top", gp = gpar(fontsize = 30))
+    grid.text("margin_top", gp = gpar(fontsize = 20))
     seekViewport("margin_bottom")
     grid.rect(gp = gpar(lwd = 3))
-    grid.text("margin_bottom", gp = gpar(fontsize = 30))
+    grid.text("margin_bottom", gp = gpar(fontsize = 20))
     seekViewport("margin_right")
     grid.rect(gp = gpar(lwd = 3))
-    grid.text("margin_right", rot = 90, gp = gpar(fontsize = 30))
+    grid.text("margin_right", rot = 90, gp = gpar(fontsize = 20))
     seekViewport("margin_left")
     grid.rect(gp = gpar(lwd = 3))
-    grid.text("margin_left", rot = 90, gp = gpar(fontsize = 30))
+    grid.text("margin_left", rot = 90, gp = gpar(fontsize = 20))
     seekViewport("corner_top_left")
     grid.rect(gp = gpar(lwd = 3))
-    grid.text("[A]", gp = gpar(fontsize = 30))
+    grid.text("[A]", gp = gpar(fontsize = 20))
     seekViewport("corner_top_right")
     grid.rect(gp = gpar(lwd = 3))
-    grid.text("[B]", gp = gpar(fontsize = 30))
+    grid.text("[B]", gp = gpar(fontsize = 20))
     seekViewport("corner_bottom_left")
     grid.rect(gp = gpar(lwd = 3))
-    grid.text("[C]", gp = gpar(fontsize = 30))
+    grid.text("[C]", gp = gpar(fontsize = 20))
     seekViewport("corner_bottom_right")
     grid.rect(gp = gpar(lwd = 3))
-    grid.text("[D]", gp = gpar(fontsize = 30))
+    grid.text("[D]", gp = gpar(fontsize = 20))
 @ 
@@
   ``corner\_top\_left'', [B] = ``corner\_top\_right'', 
-  [C] = ``corner\_bottom\_left'', [D] = ``corner\_bottom\_right''.}
+  [C] = ``corner\_bottom\_left'', [D] = ``corner\_bottom\_right'', [E]
+  = ``legend\_top'', [F] = ``legend\_sub''.}
 \label{fig:layout}
@@
 data containing two polytomous variables (hair and eye color), 
-as well as one (artificial) dichotomous variable (sex, i.e. gender). The
+as well as one (artificial) dichotomous variable (sex, i.e., gender). The
 `flattened' contingency table can be obtained using the
@@
 <<structable,results=verbatim>>=
-(hec <- structable(aperm(HairEyeColor)))
+(hec <- structable(Eye ~ Sex + Hair, data = HairEyeColor))
 @
 
-Let us first visualize the contingency table by the means of a mosaic plot
+Let us first visualize the contingency table by means of a mosaic plot
 \citep{vcd:Hartigan+Kleiner:1984} which is basically 
-an area-proportional visualization of (typically observed) frequencies, composed
+an area-proportional visualization of (typically, observed) frequencies, composed
 of tiles (corresponding to the cells) created by recursive
@@
 
-\noindent The small bullets indicate zero entries in the corresponding cell.
+\noindent The small bullets indicate zero entries in the corresponding
+cell. Note that in contrast to, e.g.,  \codefun{mosaicplot} in base
+{\sf R}, the default split direction and level ordering in all strucplot
+displays correspond to the textual representation.
 It is also possible to visualize the expected values instead of the
@@
 \begin{center}
-<<Residualsfig,fig=TRUE,echo=FALSE>>=
+<<Residualsfig,fig=TRUE,echo=FALSE,width=6, height=10>>=
 <<Residuals>>
@@
 <<split1>>=
-mosaic(hec, split_vertical = c(TRUE, TRUE, FALSE))
+mosaic(hec, split_vertical = c(TRUE, FALSE, TRUE))
 @
@@
 \begin{center}
-<<splitfig,fig=TRUE,echo=FALSE>>=
+<<splitfig,fig=TRUE,echo=FALSE,width=7,height=7>>=
 <<split1>>
@@
 By a suitable combination of splitting, spacing, and
-labeling settings, the funtions provided by the strucplot framework
+labeling settings, the functions provided by the strucplot framework
 can be customized in a quite flexible way. For example, 
-it was almost trivial to implement doubledecker
-plots---\codefun{doubledecker} is really just a wrapper for 
+\codefun{doubledecker} is simply a wrapper for 
 \codefun{mosaic}, setting the right defaults. 
 Figure \ref{fig:doubledecker} shows a doubledecker plot of
-the \data{Titanic} data, explaining the probability of survival (`Survived') by Age, given
-Sex, given Class. It is created by:
+the \data{Titanic} data, explaining the probability of survival (`survived') by age, given
+sex, given class. It is created by:
 
@@
 
-So far, we have visualized full tables. For objects of class \code{table},
+So far, we have visualized full tables. For objects of class \class{table},
 conditioning on levels (i.e., choosing a table subset for fixed levels
@@
 the association between hair and eye color, given gender, by using
-subsetting on the flat table and the \pkg{grid} package's viewport 
+subsetting on the flat table and \pkg{grid}'s viewport 
 framework to visualize the two groups besides each other:
@@
 <<Conditioning>>=
-mosaic(~ Hair + Eye | Sex, data = hec, split_vertical = TRUE)
+mosaic(~ Hair + Eye | Sex, data = hec, split_vertical = TRUE, keep_aspect_ratio = FALSE)
 @
@@
 \begin{center}
-<<Conditioningfig,fig=TRUE,echo=FALSE>>=
+<<Conditioningfig,fig=TRUE,echo=FALSE,width=12>>=
 <<Conditioning>>
@@
 \noindent The effect of using this kind of formula is that
-conditioning variables are permuted before the conditioned variables
+conditioning variables are permuted ahead of the the conditioned variables
 in the table, and that \codefun{spacing\_conditional} is used as
@@
 This spacing uses equal space between tiles of conditioned variables,
-and increasing space between tiles of conditioning variables.
-Note, however, that the plots in the ``pseudo-strata'' are
-distorted since they are not corrected for the marginal
-distribution(s) of the conditioning variables.
+and increasing space between tiles of conditioning variables. In
+addition, we release the fixed aspect ratio to get less distorted margins.
 
@@
 in a lattice-like layout, conditioning on variable \emph{levels}.
-The plot in Figure \ref{fig:cotabplot} shows Hair and Eye color, given Sex:
+The plot in Figure \ref{fig:cotabplot} shows hair and eye color, given sex:
 
 <<cotabplot>>=
-cotabplot(~ Hair + Eye | Sex, data = hec, labeling_args = list(abbreviate = c(Eye = 2)))
+cotabplot(~ Hair + Eye | Sex, data = hec, panel_args = list(margins = 3), labeling = labeling_left(clip = FALSE))
+#cotabplot(~ Hair + Eye | Sex, data = hec, labeling_args = list(abbreviate = c(Eye = 2)))
 @
@@
 \begin{center}
-<<cotabplotfig,fig=TRUE,echo=FALSE,height=4,width=7>>=
+<<cotabplotfig,fig=TRUE,echo=FALSE,height=6,width=11>>=
 <<cotabplot>>
 @
-\caption{Conditional tabular plot for the \data{HairEyeColor} data.}
+\caption{Conditional table plot for the \data{HairEyeColor} data.}
 \label{fig:cotabplot}
@@
 
-\noindent The \code{labeling\_args} argument ensures that `Eye' levels
-are abbreviated (See the vignette on labeling for detailed information).
+\noindent The \code{labeling\_args} argument ensures that `eye' levels
+are abbreviated (see the vignette on labeling for detailed information).
 
 Another high-level function for visualizing conditional independence
-models is \codefun{pairs.table} (\codefun{pairs.structable}), 
-that is, the {\sf S3}-method for objects
-of class \code{"table"} (\code{"structable"}) of the \codefun{pairs} generic. 
-In contrast to \codefun{cotabplot} which is conditioning on variable \emph{levels},
-\codefun{pairs.table} is conditioning on the variables themselves, creating
-partial views of the table. The function produces a matrix 
-having strucplot displays in the off-diagonal cells, and
-the variable names (with, optionally, univariate statistics) in the diagonal cells. 
+models are the \codefun{pairs} methods for table and structable objects.
+In contrast to \codefun{cotabplot} which conditions on variables,
+the \codefun{pairs} methods create pairwise views of the table. 
+The function produces, by default, a plot matrix 
+having strucplot displays in the off-diagonal panels, and
+the variable names (optionally, with univariate statistics) in the diagonal cells. 
 Figure \ref{fig:pairs} shows a pairs display with mosaic plots visualizing mutual 
@@
 <<pairs>>=
-pairs(hec, upper_panel = pairs_assoc, space = 0.3, 
+pairs(hec, lower_panel = pairs_assoc, space = 0.3, 
       diag_panel_args = list(rot = -45, just_leveltext = c("left","bottom")))
@@
 
-\setkeys{Gin}{width=\textwidth}
+%\setkeys{Gin}{width=\textwidth}
+\setkeys{Gin}{width=0.7\textwidth}
 \begin{figure}[h]
 \begin{center}
-<<pairsfig,fig=TRUE,echo=FALSE>>=
+<<pairsfig,fig=TRUE,echo=FALSE,width=8,height=8>>=
 <<pairs>>
@@
 
-\noindent In plots produced by \codefun{pairs.table}, 
-each cell's row and column define two variables $X$ and $Y$ used for the
-specification of four different types of independence: `\code{pairwise}',
-`\code{total}', `\code{conditional}' and `\code{joint}'. The pairwise mosaic
+\noindent (The labels of the variables are to be read from left to
+right and from top to bottom.) In plots produced by \codefun{pairs}, 
+each panel's row and column define two variables $X$ and $Y$ used for the
+specification of four different types of independence: pairwise,
+total, conditional, and joint. The pairwise mosaic
 matrix shows bivariate marginal relations between $X$ and $Y$, collapsed over all
@@
 (mosaic, association, or sieve plots).
-The available core functions (\codefun{pairs\_assoc}, 
+The available panel functions (\codefun{pairs\_assoc}, 
 \codefun{pairs\_mosaic}, and \codefun{pairs\_sieve}) 
-are just simple wrappers to \codefun{assoc}, 
+are simple wrappers to \codefun{assoc}, 
 \codefun{mosaic}, and \codefun{sieve}, respectively. 
@@
 becomes increasingly difficult with
-higher dimensionality. Typically, therefore, this plot is used with a suitable 
-residual-based shading (described in a separate vignette).
+higher dimensionality. Therefore, this plot is typically used with a suitable 
+residual-based shading (described in the vignette on ``Colors and
+Residual-based Shadings in the Strucplot Framework'').
 
@@
 \begin{center}
-\code{cell:\emph{[Variable 1]}=\emph{[Level 1]},\emph{[Variable 2]}=\emph{[Level 2]}} \dots
+\code{cell:\emph{Variable1}=\emph{Level1},\emph{Variable2}=\emph{Level2}} \dots
 \end{center}
@@
 <<changeplotfig,fig=TRUE,echo=FALSE>>=
-x = tab = margin.table(HairEyeColor, 1:2)
-x[] = "light gray"
-x["Blond","Blue"] = "Red"
+x <- tab <- margin.table(HairEyeColor, 1:2)
+x[] <- "light gray"
+x["Blond","Blue"] <- "Red"
 assoc(tab, gp = gpar(fill = x))

### 2005-10-02 trunk/vcd/R/strucplot.R

@@ -29,7 +29,7 @@ strucplot <- function(## main parameters
                       sub = NULL,
                       margins = unit(2.5, "lines"),
                       title_margins = NULL,
-                      legend_width = unit(5, "lines"),
+                      legend_width = NULL,
                       
                       ## control parameters
                       main_gp = gpar(fontsize = 20),
@@ -216,8 +216,6 @@ strucplot <- function(## main parameters
 
   seekViewport("base") 
   ## one more up if sandwich-mode
-  if (!is.null(main) || !is.null(sub) ||
-      (shade && !(is.null(legend) || is.logical(legend) && !legend))) upViewport()
   if (pop) popViewport(1 + keep_aspect_ratio) else upViewport(1 + keep_aspect_ratio)
 
   ## return visualized table
@@ -232,7 +230,12 @@ vcdViewport <- function(mar = rep.int(2.5, 4),
                         keep_aspect_ratio = TRUE)
 {
   ## process parameters
-  if (legend && !main && !sub) main <- sub <- TRUE
+  if (is.null(legend_width))
+    legend_width <- unit(5 * legend, "lines")
+  if (!is.unit(legend_width))
+    legend_width <- unit(legend_width, "lines")
+
+  if (legend && !main && !sub && keep_aspect_ratio) main <- sub <- TRUE
   mar <- if (!is.unit(mar))
     unit(pexpand(mar, 4, rep.int(2.5, 4), c("top","right","bottom","left")), "lines")
   else
@@ -243,8 +246,10 @@ vcdViewport <- function(mar = rep.int(2.5, 4),
     else unit(0, "lines")
     oma <- if (main && sub)
       max(unit(2, "lines"), 0.5 * space)
-    else if (main || sub)
-      max(unit(2, "lines"), space)
+    else if (main)
+      unit.c(max(unit(2, "lines"), space), unit(0, "lines"))
+    else if (sub)
+      unit.c(unit(0, "lines"), max(unit(2, "lines"), space))
     else
       0.5 * space
   }
@@ -252,65 +257,33 @@ vcdViewport <- function(mar = rep.int(2.5, 4),
     unit(pexpand(oma, 2, rep.int(2, 2), c("top","bottom")), "lines")
   else
     unit.rep(oma, length.out = 2)
-  if (!is.unit(legend_width))
-    legend_width <- unit(legend_width, "lines")
   
[…]

### 2005-10-02 trunk/vcd/R/labeling.R
Vignettes overworked.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@466 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -534,11 +534,17 @@ labeling_doubledecker <- function(lab_pos = c("bottom", "top"), ...) {
 }
 class(labeling_doubledecker) <- c("grapcon_generator","labeling")
 
-labeling_left <- function(tl_labels = TRUE, clip = TRUE, pos_varnames = "left",
+labeling_left <- function(rep = FALSE, pos_varnames = "left",
+                        pos_labels = "left", just_labels = "left", ...)
+  labeling_border(rep = rep, pos_varnames = pos_varnames,
+              pos_labels = pos_labels, just_labels = just_labels, ...)
+class(labeling_left) <- c("grapcon_generator","labeling")
+
+labeling_left2 <- function(tl_labels = TRUE, clip = TRUE, pos_varnames = "left",
                         pos_labels = "left", just_labels = "left", ...)
   labeling_border(tl_labels = tl_labels, clip = clip, pos_varnames = pos_varnames,
               pos_labels = pos_labels, just_labels = just_labels, ...)
-class(labeling_left) <- c("grapcon_generator","labeling")
+class(labeling_left2) <- c("grapcon_generator","labeling")
 
 labeling_cboxed <- function(tl_labels = TRUE, boxes = TRUE, clip = TRUE, pos_labels = "center", ...)
   labeling_border(tl_labels = tl_labels, boxes = boxes, clip = clip, pos_labels = pos_labels, ...)

### 2005-10-16 trunk/vcd/inst/doc/shading.Rnw
revert class system to minimal level
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@471 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 shading2_fun <- function(res) gpar(fill = ifelse(res > 0, "lightblue", "lightsalmon"))
-class(shading2_fun) <- c("grapcon","shading")
 @
 
-\noindent Note the \code{class} attribute set on \codefun{shading2\_fun}.
-To create a mosaic display with binary shading, 
+\noindent To create a mosaic display with binary shading, 
 it now suffices to specify the data table along with
@@
   if (length(col) != 2) stop("Need exactly two colors!")
-  ret <- function(res) gpar(fill = ifelse(res > 0, col[1], col[2]))  
-  structure(ret, class = c("grapcon","shading"))
+  function(res) gpar(fill = ifelse(res > 0, col[1], col[2]))  
 }
@@
   if (length(col) != 2) stop("Need exactly two colors!")
-  ret <- function(res) gpar(fill = ifelse(res > 0, col[1], col[2]))  
-  structure(ret, class = c("grapcon","shading"))
+  function(res) gpar(fill = ifelse(res > 0, col[1], col[2]))  
 }
-class(shading3b_fun) <- c("grapcon_generator","shading")
+class(shading3b_fun) <- "grapcon_generator"
 @

### 2005-10-16 trunk/vcd/inst/doc/strucplot.Rnw
revert class system to minimal level
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@471 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 parameter functions that directly or indirectly influence the final
-appearance of the plot. These `graphical appearance control'
-(`grapcon') functions are created by generating functions, allowing
+appearance of the plot. These are examples of `graphical appearance control'
+(`grapcon') functions. They are created by generating functions, allowing
 flexible parameterization and extensibility. 
@@
 content.  On the second level of the framework, a suitable combination
-of grapcon functions (or, alternatively, corresponding generating functions)
+of the low-level grapcon functions (or, alternatively, corresponding generating functions)
 is passed as ``hyperparameters'' to \codefun{strucplot}.

### 2005-10-16 trunk/vcd/R/labeling.R

@@ -28,7 +28,7 @@ labeling_list <- function(gp = gpar(),
                           varnames = TRUE,
                           cols = 2,
                           ...) {
-  ret <- function(d, split_vertical, condvars) {
+  function(d, split_vertical, condvars) {
     if (is.table(d))
       d <- dimnames(d)
     ld <- length(d)
@@ -49,12 +49,11 @@ labeling_list <- function(gp = gpar(),
                 gp = gp
                 )
   }
-  structure(ret, class = c("grapcon","labeling"))
 }
-class(labeling_list) <- c("grapcon_generator","labeling")
+class(labeling_list) <- "grapcon_generator"
 
 labeling_conditional <- function(...) {
-  ret <- function (d, split_vertical, condvars) {
+  function (d, split_vertical, condvars) {
     if (is.table(d))
       d <- dimnames(d)
     v <- rep.int(TRUE, length(d))
@@ -62,9 +61,8 @@ labeling_conditional <- function(...) {
     labeling_border(labels = !v, ...)(d, split_vertical, condvars)
     labeling_cells(labels = v, ...)(d, split_vertical, condvars)
   }
-  structure(ret, class = c("grapcon","labeling"))
 }
-class(labeling_conditional) <- c("grapcon_generator","labeling")
+class(labeling_conditional) <- "grapcon_generator"
 
 labeling_cells <- function(labels = TRUE, varnames = TRUE,
                          abbreviate_labels = FALSE, abbreviate_varnames = FALSE,
@@ -72,7 +70,7 @@ labeling_cells <- function(labels = TRUE, varnames = TRUE,
                          just = "center", pos = "center", rot = 0,
                          margin = unit(0.5, "lines"), clip_cells = TRUE,
                          text = NULL, ...) {
-  ret <- function(d, split_vertical, condvars) {
+  function(d, split_vertical, condvars) {
     if (is.table(d))
       d <- dimnames(d)
     dn <- names(d)
@@ -133,9 +131,8 @@ labeling_cells <- function(labels = TRUE, varnames = TRUE,
     }
     split()
   }
-  structure(ret, class = c("grapcon","labeling"))
 }
-class(labeling_cells) <- c("grapcon_generator","labeling")
+class(labeling_cells) <- "grapcon_generator"
 
 labeling_border <- function(labels = TRUE, varnames = labels,
                             set_labels = NULL, set_varnames = NULL,
[…]

### 2005-10-16 trunk/vcd/R/strucplot.R

@@ -108,10 +108,8 @@ strucplot <- function(## main parameters
 
   ## spacing
   if (is.function(spacing)) {
-    if (inherits(spacing, "grapcon_generator") && inherits(spacing, "spacing"))
+    if (inherits(spacing, "grapcon_generator"))
       spacing <- do.call("spacing", spacing_args)
-    if (!(inherits(spacing, "grapcon") && inherits(spacing, "spacing")))
-      stop("Spacing function does not inherit from classes 'grapcon' and 'core'.")
     spacing <- spacing(d, condvars)
   }
 
@@ -121,10 +119,8 @@ strucplot <- function(## main parameters
     if (is.function(gp)) {
       if (is.null(legend) || (is.logical(legend) && legend))
         legend <- legend_resbased
-      gpfun <- if (inherits(gp, "grapcon_generator") && inherits(gp, "shading"))
+      gpfun <- if (inherits(gp, "grapcon_generator"))
         do.call("gp", c(list(x, residuals, expected, df), as.list(gp_args))) else gp
-      if (!(inherits(gpfun, "grapcon") && inherits(gpfun, "shading")))
-        stop("Shading function does not inherit from classes 'grapcon' and 'shading'.")
       gp <- gpfun(residuals)
     } else if (!is.null(legend) && !(is.logical(legend) && !legend))
       stop("gp argument must be a shading function for drawing a legend")
@@ -161,11 +157,9 @@ strucplot <- function(## main parameters
                            legend_width = legend_width))
 
   ## legend
-  if (inherits(legend, "grapcon_generator") && inherits(legend, "legend"))
+  if (inherits(legend, "grapcon_generator"))
     legend <- do.call("legend", legend_args)
   if (shade && !is.null(legend) && !(is.logical(legend) && !legend)) {
-    if (!(inherits(legend, "grapcon") && inherits(legend, "legend")))
-      stop("Legend function does not inherit from classes 'grapcon' and 'legend'.")
     seekViewport("legend")
     residuals_type = switch(residuals_type, deviance = "deviance", ft = "Freeman-Tukey", pearson = "Pearson")
     legend(residuals, gpfun, paste(residuals_type, "residuals:", sep = "\n"))
@@ -189,11 +183,8 @@ strucplot <- function(## main parameters
   ## make plot
   seekViewport("plot")
   
-  if (inherits(core, "grapcon_generator") && inherits(core, "core"))
+  if (inherits(core, "grapcon_generator"))
[…]

### 2005-10-18 trunk/vcd/inst/doc/shading.Rnw
Misc KH changes.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@475 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \begin{enumerate}
-\item Pre-computing the graphical parameters (e.g., fill colors),
+\item Precomputing the graphical parameters (e.g., fill colors),
   encapsulating them into an object of class
@@
   alternative 1.
-\item Providing a grapcon \emph{generating} function taking parameters and returning
-  a function as described in alternaive 2. 
+ \item Providing a grapcon \emph{generating} function taking parameters
+  and returning a function as described in alternative~2.
 \end{enumerate}
@@
 returns (`creates') a `parameterized' shading function with the minimal standard interface
-\codefun{strucplot} requires. The follwing example shows the necessary
+\codefun{strucplot} requires. The following example shows the necessary
 extensions for our running example:
@@
 
-\noindent In the call of \codefun{mosaic} (figure not shown), 
-using the new \codefun{shading3a\_fun} function, we can now simply change the colors:
+\noindent In the call to \codefun{mosaic}, using the new
+\codefun{shading3a\_fun} function, we can now simply change the colors:
 
@@
 
-\noindent The procedure described so far is a rather general concept,
-applicable to a wide family of user-level \pkg{grid} graphics. Indeed, the
-customization of other components of the strucplot framework
-(labeling, spacing, legend, and core functions) follows the same idea.
-Now for the shading functions, more customization is needed. Note that
-\codefun{shading3a\_fun} needs to be evaluated by the user, even if the
-defaults shall be used. It is a better idea to let \codefun{strucplot} call
-the generating function, which, in particular, allows the passing of arguments that
-are computed by \codefun{strucplot}. Since shading
-functions can be used for visualizing significance (see Section 
-\ref{sec:overview}), it makes sense for generating functions to have access to the model,
-i.e., observed and expected values, residuals, and degrees of
-freedom. For example, the \codefun{shading\_max} generating function
-computes a permutation distribution of the maximum statistic and $p$
-values for specified significance levels 
-based on the observed table to create data-driven cut-off
-points. If this was done in the shading function itself, the
-permutation statistic would be recomputed every time the shading
-function is called, resulting in possibly severe performance loss 
-and numerical inconsistencies. Therefore, generating functions for
-shadings are required to take at least the parameters 
-\code{observed}, \code{expected}, \code{residuals}, and \code{df}
-(these are provided by the strucplot framework),
-followed by other parameters controlling the shading
-appearance (to be specified by the user):
+\noindent (figure not shown).  The procedure described so far is a
+rather general concept, applicable to a wide family of user-level
+\pkg{grid} graphics. Indeed, the customization of other components of
+the strucplot framework (labeling, spacing, legend, and core functions)
+follows the same idea.  Now for the shading functions, more
+customization is needed. Note that \codefun{shading3a\_fun} needs to be
+evaluated by the user, even if the defaults are to be used. It is a
+better idea to let \codefun{strucplot} call the generating function,
+which, in particular, allows the passing of arguments that are computed
+by \codefun{strucplot}. Since shading functions can be used for
+visualizing significance (see Section \ref{sec:overview}), it makes
+sense for generating functions to have access to the model, i.e.,
+observed and expected values, residuals, and degrees of freedom. For
+example, the \codefun{shading\_max} generating function computes a
+permutation distribution of the maximum statistic and $p$ values for
+specified significance levels based on the observed table to create
+data-driven cut-off points. If this was done in the shading function
+itself, the permutation statistic would be recomputed every time the
+shading function is called, resulting in possibly severe performance
+loss and numerical inconsistencies. Therefore, generating functions for
+shadings are required to take at least the parameters \code{observed},
+\code{expected}, \code{residuals}, and \code{df} (these are provided by
+the strucplot framework), followed by other parameters controlling the
+shading appearance (to be specified by the user):
 
@@
 
-In some sence, generating
+In some sense, generating
 functions for shadings are parameterized both by the user and the

### 2005-10-21 trunk/vcd/paper/JSS.Rnw
JSS paper checked in.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@477 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\documentclass[article]{jss}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% almost as usual
\author{David Meyer\\Wirtschaftsuniversit\"at Wien \And 
        Achim Zeileis\\Wirtschaftsuniversit\"at Wien \And 
        Kurt Hornik\\Wirtschaftsuniversit\"at Wien}
\title{The Strucplot Framework:\\ Visualizing Multi-way Contingency
  Tables With \pkg{vcd}}

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{David Meyer, Achim Zeileis, Kurt Hornik} %% comma-separated
\Shorttitle{The Strucplot Framework} %% a short title (if necessary)

%% an abstract and keywords
\Abstract{
This paper describes the `strucplot' framework for the visualization of
multi-way contingency tables. Strucplot displays include hierarchical
conditional plots such as mosaic, association, and sieve plots, and
can be combined into more complex, specialized plots for visualizing
conditional independence, GLMs, and the results of
independence tests. The frameworks' modular design allows
flexible customization of the plots' graphical appearance, including 
shading, labeling, spacing, and legend, by means of graphical
appearance control (`grapcon') functions. The framework is provided by
the \pkg{vcd} package freely available from the
Comprehensive \proglang{R} Archive Network (CRAN).
}

\Keywords{contingency tables, mosaic plots, association plots, sieve plots, \proglang{R},
categorical data, independence, conditional independence, HSV, HCL,
residual-based shading, \pkg{grid}}
\Plainkeywords{contingency tables, mosaic plots, association plots,
  sieve plots, R, categorical data, independence, 
  conditional independence, HSV, HCL, residual-based shading, grid}

%\VignetteIndexEntry{The Strucplot Framework---Visualizing Multi-way Contingency Tables}
%\VignetteDepends{vcd}
%\VignetteKeywords{graphics, categorical data}
%\VignettePackage{vcd}

%% publication information
%% NOTE: This needs to filled out ONLY IF THE PAPER WAS ACCEPTED.
%% If it was not (yet) accepted, leave them commented.
%% \Volume{13}
%% \Issue{9}
%% \Month{September}
%% \Year{2004}
%% \Submitdate{2004-09-29}
%% \Acceptdate{2004-09-29}

%% The address of (at least) one author should be given
%% in the following format:
\Address{
  David Meyer\\
  Department of Information Systems and Process Management\\
  E-mail: \email{David.Meyer@wu-wien.ac.at}\\
  URL: \url{http://wi.wu-wien.ac.at/~meyer/}\\

  Achim Zeileis\\
  Department of Statistics \& Mathematics\\
  E-mail: \email{Achim.Zeileis@wu-wien.ac.at}\\
  URL: \url{http://www.ci.tuwien.ac.at/~zeileis/}\\

  Kurt Hornik\\
  Department of Statistics \& Mathematics\\
  E-mail: \email{Kurt.Hornik@wu-wien.ac.at}\\
  URL: \url{http://www.wu-wien.ac.at/cstat/hornik/}\\

  Wirtschaftsuniversit\"at Wien\\
  1090 Wien, Austria\\
}
%% It is also possible to add a telephone and fax number
%% before the e-mail in the following format:
%% Telephone: +43/1/31336-5053
%% Fax: +43/1/31336-734

%% for those who use Sweave please include the following line (with % symbols):
%% need no \usepackage{Sweave.sty}

\SweaveOpts{engine=R,eps=TRUE,height=6,width=7,results=hide,fig=FALSE,echo=TRUE}
\setkeys{Gin}{width=0.7\textwidth}

<<preliminaries,echo=FALSE,results=hide>>=
set.seed(1071)
library(vcd)
data(Titanic)
data(HairEyeColor)
data(PreSex)
data(Arthritis)
art <- xtabs(~Treatment + Improved, data = Arthritis)
@

\newcommand{\var}[1]{\textit{\texttt{#1}}}
\newcommand{\data}[1]{`\texttt{#1}'}
\newcommand{\class}[1]{\textsf{#1}}
%% \code without `-' ligatures
\def\nohyphenation{\hyphenchar\font=-1 \aftergroup\restorehyphenation}
\def\restorehyphenation{\hyphenchar\font=`-}
{\catcode`\-=\active%
  \global\def\code{\bgroup%
    \catcode`\-=\active \let-\codedash%
    \Rd@code}}
\def\codedash{-\discretionary{}{}{}}
\def\Rd@code#1{\texttt{\nohyphenation#1}\egroup}

\newcommand{\codefun}[1]{\code{#1()}}

%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

%% include your article here, just as usual
%% Note that you should use the \pkg{}, \proglang{} and \code{} commands.

\section[Introduction]{Introduction}
%% Note: If there is markup in \(sub)section, then it has to be escape as above.

In order to explain multi-dimensional categorical data, statisticians
typically look for (conditional) independence structures. Whether the
task is purely exploratory or model based, techniques such as mosaic and
association plots offer good support for visualization. Both have been
extended over the last two decades, and implementations exist in many
statistical environments. 
Both graphical methods visualize aspects of (possibly higher-dimensional)
contingency tables. A \emph{mosaic plot} \citep{vcd:Hartigan+Kleiner:1984} is basically 
an area-proportional visualization of (typically observed) frequencies, composed
of tiles (corresponding to the cells) created by recursive
vertical and horizontal splits of a square. Thus, the area of each tile
is proportional to the corresponding cell entry \emph{given} the
dimensions of previous splits. An \emph{association plot} \citep{vcd:Cohen:1980}
visualizes the standardized deviations of observed frequencies from
those expected under a certain independence hypothesis.
Each cell is represented by a rectangle that has
(signed) height proportional to the residual and width proportional
to the square root of the expected counts, 
so that the area of the box is proportional
to the difference in observed and expected frequencies.

Over the years, extensions to these techniques were mainly focused on
five aspects:

\begin{enumerate}
\item Varying the shape of bar plots and mosaic displays to yield, e.g.,
  double-decker plots \citep{vcd:hofmann:2001}, spine plots, or spinograms.
\item Using residual-based shadings to visualize log-linear models
  \citep{vcd:Friendly:1994,vcd:Friendly:2000} and significance of
  statistical tests \citep{vcd:Meyer+Zeileis+Hornik:2003}.
\item Using pairs plots and trellis-like layouts for marginal, conditional and
  partial views \citep{vcd:Friendly:1999}.
\item Adding direct user interaction, allowing quick exploration and
  modification of the visualized models \citep{vcd:Unwin+Hawkins+Hofmann:1996,vcd:Theus:2003}.
\item Providing a modular and flexible implementation to easily allow
  user extensions \citep{vcd:Meyer+Zeileis+Hornik:2003}.
\end{enumerate}

\noindent Current implementations of mosaic displays can be found,
e.g., for \proglang{SAS}, \proglang{ViSta}, \proglang{MANET}, \proglang{Mondrian},
\proglang{R}, and \proglang{S-PLUS}. Table \ref{tab:compare} gives an overview of the
available functionality in these systems.

\begin{table}[h]
  \centering
  \begin{tabular}{|l|c|c|c|c|c|}
    \hline
                             & \proglang{SAS} & \proglang{S-PLUS} & \proglang{R} & \proglang{ViSta} & \proglang{MANET}/\proglang{Mondrian}\\\hline
    Basic functionality      &  $\times$  &   $\times$    & $\times$ &   $\times$   & $\times$\\
    Shape                    &     &        & $\times$ &       & $\times$\\
    Residual-based shadings  &  $\times$  &        & $\times$ &   $\times$   & ($\times$)\\
    Conditional Views        &  $\times$  &        & $\times$ &       & $\times$\\
    Interaction              &     &        &   &   $\times$   & $\times$\\
    Extensible Design        &     &        & $\times$ &       &  \\\hline
  \end{tabular}
  \caption{Comparison of current software environments}
  \label{tab:compare}
\end{table}

Figures \ref{fig:arthritis} to \ref{fig:titanic} illustrate some of these
extensions. Figure \ref{fig:arthritis} shows the results from a double-blind clinical trial
investigating a new treatment for rheumatoid arthritis, using an
extended mosaic plot with residual-based shading based on the
maximum statistic: clearly, the new treatment is effective. Figure
\ref{fig:ucbadmissions} visualizes the well-known UCB admissions data
by the means of a conditional association plot. The panels show the
residuals from a conditional independence model (independence of gender
and admission, given department), stratified by department. Clearly,
the situation in department A (more women/less men accepted than
would be expected under the null) causes the rejection of the
hypothesis of conditional independence. Figure \ref{fig:presex}
illustrates the conditional independence of premarital and
extramarital sex, given gender and marital status. The Chi-squared
test of independence rejects the null hypothesis: possibly, because
too many men who had premarital sex also had extramarital sex?
Finally, Figure \ref{fig:titanic} visualizes the `Survival on the Titanic' 
data using a double-decker plot. Here, a binary response (survival to the disaster)
shall be explained by other factors (class, gender, and age). The blue
boxes represent the proportion of survived passengers in a particular
stratum. The proportions of saved women and children are indeed higher
than those of men, but they clearly decrease from the 1st to the 3rd class. In
addition, the proportion of saved men in the 1st class is higher than
in the others.

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[p]
\begin{center}
<<Arthritis,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(art, gp = shading_max, split_vertical = TRUE)
@
\caption{Mosaic plot for the arthritis data.}
\label{fig:arthritis}
\end{center}
\end{figure}

\setkeys{Gin}{width=\textwidth}
\begin{figure}[p]
\begin{center}
<<UCBAdmissions,fig=TRUE,echo=FALSE,height=6,width=10,results=hide>>=
cotabplot(UCBAdmissions, panel = cotab_assoc)
@
\caption{Conditional association plot for the UCB admissions data.}
\label{fig:ucbadmissions}
\end{center}
\end{figure}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[p]
\begin{center}
<<PreSex,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(PreSex, condvars = c(1, 4), shade = TRUE)
@
\caption{Mosaic plot for the pre- and extramarital sex data.}
\label{fig:presex}
\end{center}
\end{figure}

\setkeys{Gin}{width=0.8\textwidth}
\begin{figure}[p]
\begin{center}
<<Titanic,fig=TRUE,echo=FALSE,height=6,width=8,results=hide>>=
doubledecker(Survived ~ ., data = Titanic, labeling_args = 
             list(set_varnames = c(Sex = "Gender")))
@
\caption{Double-decker plot for the survival on the Titanic data.}
\label{fig:titanic}
\end{center}
\end{figure}

This paper describes the strucplot framework provided by the \pkg{vcd} package
for the \proglang{R} system and environment for statistical computing
and graphics, available from the Comprehensive \proglang{R} Archive Network (CRAN).
The framework integrates displays such as mosaic, association, and
sieve plots by their common property of being 
flat representations of contingency tables.
These basic plots, as well as specialized displays for conditional
independence, can be used both for exploratory visualization and
model-based analysis. Exploratory techniques include specialized displays for
the bivariate case, as well as pairs and trellis plot-like displays for
higher-dimensional tables. Model-based tools include 
methods suitable for the visualization of 
conditional independence tests (including permutation tests), as well
as for the visualization of particular GLMs (logistic regression,
log-linear models). Additionally, two of the frameworks' further strengths are its
flexibility and extensibility: graphical appearance aspects such as shading,
labeling, and spacing are modularized by means of `graphical
appearance functions', allowing fine-granular customization and
user-level extensions.

The reminder of the paper is organized as follows. In
Section \ref{sec:strucplot}, we give an overview of the strucplot
framework, describing the hierarchy of the main components and the
basic functionality. 
In Section \ref{sec:shading}, we demonstrate how
residual-based shadings support the visualization of log-linear models and
the results of independence tests. Also, we explain step-by-step how
the concepts of generating functions and graphical appearance control
functions can be combined to provide a flexible customization of
complex graphical displays as created by the strucplot framework.
Sections \ref{sec:labeling} and \ref{sec:spacing} discuss in detail the labeling
and spacing features, respectively. Section \ref{sec:conclusion}
finally concludes the work.

\section[The Strucplot Framework]{The Strucplot Framework}
\label{sec:strucplot}

The strucplot framework in the {\sf R} package \pkg{vcd}, used for visualizing
multi-way contingency tables, integrates techniques such as
mosaic displays, association plots, and sieve plots. The main idea is to visualize
the tables' cells arranged in rectangular form. For multi-way tables,
the variables are nested into rows and columns using recursive
conditional splits, given the margins. The result is a 
`flat' representation that can be visualized in
ways similar to a two-dimensional table.
This principle defines a class of conditional displays which allows
for granular control of graphical appearance aspects, including:

\begin{itemize}
\item the content of the tiles
\item the split direction for each dimension
\item the graphical parameters of the tiles' content
\item the spacing between the tiles
\item the labeling of the tiles
\end{itemize}

The strucplot framework is highly modularized: Figure \ref{fig:struc}
shows the hierarchical relationship between the various components. 
On the lowest level, there are several groups of workhorse and 
parameter functions that directly or indirectly influence the final
appearance of the plot. These are examples of `graphical appearance control'
(`grapcon') functions. They are created by generating functions
(`grapcon generators'), allowing
flexible parameterization and extensibility (Figure \ref{fig:struc}
only shows the generators).
The first part of the generator names (\code{\emph{group\_foo}()})
reflects the group they belong to (strucplot core, labeling,
legend, shading, or spacing). The workhorse functions (created by 
\code{struc\_\emph{foo}()}, 
\code{labeling\_\emph{foo}()}, and \code{legend\_\emph{foo}()}) 
directly produce graphical output (``add ink to the canvas''), whereas
the parameter functions
(created by \code{spacing\_\emph{foo}()} and \code{shading\_\emph{foo}()}) compute
graphical parameters used by the others. The grapcon functions returned by 
\code{struc\_\emph{foo}()} implement the core functionality, creating the tiles and their
content.  On the second level of the framework, a suitable combination
of the low-level grapcon functions (or, alternatively, corresponding generating functions)
is passed as ``hyperparameters'' to \codefun{strucplot}.
This central function 
sets up the graphical layout using grid viewports (see Figure \ref{fig:layout}),
and coordinates the specified core, labeling, shading, and spacing functions to produce 
the plot. On the third level, we provide
several convenience functions such as \codefun{mosaic},
\codefun{sieve}, \codefun{assoc}, and \codefun{doubledecker} which 
interface \codefun{strucplot} through sensible parameter defaults
and support for model formulas. Finally, on the fourth
level, there are `related' \pkg{vcd} functions (such as \codefun{cotabplot}
and the \codefun{pairs} methods for table objects) 
arranging collections of plots of the strucplot 
framework into more complex displays (e.g., by means of panel functions).

\begin{figure}[h]
  \begin{center}
    \includegraphics[width=0.8\textwidth]{struc}
    \caption{Components of the strucplot framework.}
    \label{fig:struc}
  \end{center}
\end{figure}

\setkeys{Gin}{width=0.6\textwidth}
\begin{figure}[h]
\begin{center}
<<vcdlayout,fig=TRUE,echo=FALSE,width=6,height=6>>=
    pushViewport(vcd:::vcdViewport(legend = T, mar =4))
    seekViewport("main")
    grid.rect(gp = gpar(lwd = 3))
    grid.text("main", gp = gpar(fontsize = 20))
    seekViewport("sub")
    grid.rect(gp = gpar(lwd = 3))
    grid.text("sub", gp = gpar(fontsize = 20))
    seekViewport("plot")
    grid.rect(gp = gpar(lwd = 3))
    grid.text("plot", gp = gpar(fontsize = 20))
    seekViewport("legend")
    grid.text("legend", rot = 90, gp = gpar(fontsize = 20))
    grid.rect(gp = gpar(lwd = 3))
    seekViewport("legend_sub")
    grid.rect(gp = gpar(lwd = 3))
    grid.text("[F]", gp = gpar(fontsize = 20))
    seekViewport("legend_top")
    grid.rect(gp = gpar(lwd = 3))
    grid.text("[E]", gp = gpar(fontsize = 20))
    seekViewport("margin_top")
    grid.rect(gp = gpar(lwd = 3))
    grid.text("margin_top", gp = gpar(fontsize = 20))
    seekViewport("margin_bottom")
    grid.rect(gp = gpar(lwd = 3))
    grid.text("margin_bottom", gp = gpar(fontsize = 20))
    seekViewport("margin_right")
    grid.rect(gp = gpar(lwd = 3))
    grid.text("margin_right", rot = 90, gp = gpar(fontsize = 20))
    seekViewport("margin_left")
    grid.rect(gp = gpar(lwd = 3))
    grid.text("margin_left", rot = 90, gp = gpar(fontsize = 20))
    seekViewport("corner_top_left")
    grid.rect(gp = gpar(lwd = 3))
    grid.text("[A]", gp = gpar(fontsize = 20))
    seekViewport("corner_top_right")
    grid.rect(gp = gpar(lwd = 3))
    grid.text("[B]", gp = gpar(fontsize = 20))
    seekViewport("corner_bottom_left")
    grid.rect(gp = gpar(lwd = 3))
    grid.text("[C]", gp = gpar(fontsize = 20))
    seekViewport("corner_bottom_right")
    grid.rect(gp = gpar(lwd = 3))
    grid.text("[D]", gp = gpar(fontsize = 20))
@ 
\caption{Viewport layout for strucplot displays with their names. [A] =
  ``corner\_top\_left'', [B] = ``corner\_top\_right'', 
  [C] = ``corner\_bottom\_left'', [D] = ``corner\_bottom\_right'', [E]
  = ``legend\_top'', [F] = ``legend\_sub''.}
\label{fig:layout}
\end{center}
\end{figure}
\setkeys{Gin}{width=0.7\textwidth}

\subsection{Mosaic, Association, and Sieve Plots}

As an example, consider the \data{HairEyeColor} 
data containing two polytomous variables (hair and eye color), 
as well as one (artificial) dichotomous variable (sex, i.e., gender). The
`flattened' contingency table can be obtained using the
\codefun{structable} function (quite similar to \codefun{ftable} in
base {\sf R}, but allowing the specification of split directions):

<<structable,results=verbatim>>=
(hec <- structable(Eye ~ Sex + Hair, data = HairEyeColor))
@

Let us first visualize the contingency table by means of a mosaic plot
\citep{vcd:Hartigan+Kleiner:1984} which is basically 
an area-proportional visualization of (typically, observed) frequencies, composed
of tiles (corresponding to the cells) created by recursive
vertical and horizontal splits of a square. Thus, the area of each tile
is proportional to the corresponding cell entry \emph{given} the
dimensions of previous splits. Figure \ref{fig:observed} depicts 
the effect of

<<Observed>>=
mosaic(hec)
@

\noindent equivalent to

<<Observed2>>=
mosaic(~ Sex + Eye + Hair, data = HairEyeColor)
@

%\setkeys{Gin}{width=0.75\textwidth}
\begin{figure}[p]
\begin{center}
<<Observedfig,fig=TRUE,echo=FALSE>>=
<<Observed>>
@
\caption{Mosaic plot for the \data{HairEyeColor} data.}
\label{fig:observed}
\end{center}
\end{figure}

\noindent The small bullets indicate zero entries in the corresponding
cell. Note that in contrast to, e.g.,  \codefun{mosaicplot} in base
{\sf R}, the default split direction and level ordering in all strucplot
displays correspond to the textual representation.
It is also possible to visualize the expected values instead of the
observed values (see Figure \ref{fig:expected}):

<<Expected>>=
mosaic(hec, type = "expected")
@

\begin{figure}[p]
\begin{center}
<<Expectedfig,fig=TRUE,echo=FALSE>>=
<<Expected>>
@
\caption{Mosaic plot for the \data{HairEyeColor} data (expected values).}
\label{fig:expected}
\end{center}
\end{figure}
%\setkeys{Gin}{width=0.7\textwidth}

\noindent In order to compare observed and expected values,
a sieve plot \citep{vcd:riedwyl+schuepbach:1994} 
could be used (see Figure \ref{fig:sieve}):

<<sieve>>=
sieve(hec)
@

\begin{figure}[h]
\begin{center}
<<sievefig,fig=TRUE,echo=FALSE>>=
<<sieve>>
@
\caption{Sieve plot for the \data{HairEyeColor} data visualizing simultaneously
  observed and expected values.}
\label{fig:sieve}
\end{center}
\end{figure}

\noindent Alternatively, we can directly inspect the residuals. 
The Pearson residuals (standardized deviations of observed from expected values) 
are preferably visualized using association
plots \citep{vcd:Cohen:1980}. In contrast to \codefun{assocplot} in
base {\sf R}, \pkg{vcd}'s \codefun{assoc}
function scales to more than two variables (see Figure \ref{fig:residuals}):

<<Residuals>>=
assoc(hec, compress = FALSE)
@

\begin{figure}[p]
\begin{center}
<<Residualsfig,fig=TRUE,echo=FALSE,width=6, height=10>>=
<<Residuals>>
@
\caption{Association plot for the \data{HairEyeColor} data.}
\label{fig:residuals}
\end{center}
\end{figure}

\noindent The \code{compress} argument keeps distances between tiles
equal for better comparison.

For both mosaic plots and association plots, 
the splitting of the tiles can be controlled using the
\code{split\_vertical} argument (default: alternating splits starting
with a vertical one).

<<echo=FALSE>>=
options(width=60)
@

<<split1>>=
mosaic(hec, split_vertical = c(TRUE, FALSE, TRUE), 
labeling_args = list(abbreviate = c(Eye = 3)))
@

<<echo=FALSE>>=
options(width=70)
@

\noindent For compatibility with \codefun{mosaicplot} in base {\sf R},
the \codefun{mosaic} function also allows the use of a \code{"direction"} 
argument taking a vector of \code{"h"} and
\code{"v"} characters (see Figure \ref{fig:split}):

<<split2>>=
mosaic(hec, direction = c("v","h","v"))
@

\begin{figure}[p]
\begin{center}
<<splitfig,fig=TRUE,echo=FALSE,width=7,height=7>>=
<<split1>>
@
\caption{Mosaic plot for the \data{HairEyeColor} data---alternative splitting.}
\label{fig:split}
\end{center}
\end{figure}

By a suitable combination of splitting, spacing, and
labeling settings, the functions provided by the strucplot framework
can be customized in a quite flexible way. For example, 
\codefun{doubledecker} is simply a wrapper for 
\codefun{mosaic}, setting the right defaults. 
Figure \ref{fig:doubledecker} shows a doubledecker plot of
the \data{Titanic} data, explaining the probability of survival (`survived') by age, given
sex, given class. It is created by:

<<doubledecker1>>=
doubledecker(Titanic)
@

\noindent equivalent to:

<<doubledecker2>>=
doubledecker(Survived ~ Class + Sex + Age, data = Titanic)
@

\begin{figure}[h]
\begin{center}
<<doubledeckerfig,fig=TRUE,echo=FALSE>>=
<<doubledecker1>>
@
\caption{Doubledecker plot for the \data{Titanic} data.}
\label{fig:doubledecker}
\end{center}
\end{figure}

\subsection{Conditional and partial views}

So far, we have visualized full tables. For objects of class \class{table},
conditioning on levels (i.e., choosing a table subset for fixed levels
of the conditioning variable(s)) is simply done by indexing.
However, subsetting \code{"structable"} objects is more restrictive 
because of their inherent conditional structure. Since the variables
on both the row and the columns side are nested, 
conditioning is only possible ``outside-in'':

<<subsetting,results=verbatim>>=
hec
hec["Male",]
hec[c("Male", "Brown"),]
hec["Male","Green"]
@

\noindent Now, there are several ways for visualizing 
conditional independence structures. The ``brute force'' method is to
draw separate plots for the strata. The following example compares 
the association between hair and eye color, given gender, by using
subsetting on the flat table and \pkg{grid}'s viewport 
framework to visualize the two groups besides each other:

<<Variables>>=
pushViewport(viewport(layout = grid.layout(ncol = 2)))

pushViewport(viewport(layout.pos.col = 1))
mosaic(hec["Male"], margins = c(left = 2.5, top = 2.5, 0), sub = "Male", newpage = FALSE)
popViewport()

pushViewport(viewport(layout.pos.col = 2))
mosaic(hec["Female"], margins = c(top = 2.5, 0), sub = "Female", newpage = FALSE)
popViewport(2)
@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[p]
\begin{center}
<<Variablesfig,fig=TRUE,echo=FALSE,height=6,width=12>>=
<<Variables>>
@
\caption{Distribution of hair and eye color, given gender.}
\label{fig:parttable}
\end{center}
\end{figure}
\setkeys{Gin}{width=0.7\textwidth}

\noindent Note the use of the \code{margins} argument: it takes a
vector with up to four values whose unnamed components are recycled,
but ``overruled'' by the named arguments. Thus, in the example, only
the top margin is set to 2 lines, and all other to 0.
This idea applies to almost all
vectorized arguments in the strucplot framework (with
\code{split\_vertical} as a prominent exception).

Since mosaic displays are ``conditional plots'' by definition, we can
also use one single mosaic for stratified plots.
The formula interface of \codefun{mosaic} allows the
specification of conditioning variables (see Figure \ref{fig:conditioning}):

<<Conditioning>>=
mosaic(~ Hair + Eye | Sex, data = hec, split_vertical = TRUE, keep_aspect_ratio = FALSE)
@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[p]
\begin{center}
<<Conditioningfig,fig=TRUE,echo=FALSE,width=12>>=
<<Conditioning>>
@
\caption{Mosaic plot for conditional independence structures.}
\label{fig:conditioning}
\end{center}
\end{figure}
\setkeys{Gin}{width=0.7}

\noindent The effect of using this kind of formula is that
conditioning variables are permuted ahead of the the conditioned variables
in the table, and that \codefun{spacing\_conditional} is used as
default to better
distinguish conditioning from conditioned dimensions.
This spacing uses equal space between tiles of conditioned variables,
and increasing space between tiles of conditioning variables. In
addition, we release the fixed aspect ratio to get less distorted margins.

The \codefun{cotabplot} function does a much better job on this task:
it arranges stratified strucplot displays 
in a lattice-like layout, conditioning on variable \emph{levels}.
The plot in Figure \ref{fig:cotabplot} shows hair and eye color, given sex:

<<cotabplot>>=
cotabplot(~ Hair + Eye | Sex, data = hec, panel_args = list(margins = 3), labeling = labeling_left(clip = FALSE))
@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[h]
\begin{center}
<<cotabplotfig,fig=TRUE,echo=FALSE,height=6,width=11>>=
<<cotabplot>>
@
\caption{Conditional table plot for the \data{HairEyeColor} data.}
\label{fig:cotabplot}
\end{center}
\end{figure}
\setkeys{Gin}{width=0.7\textwidth}

\noindent The \code{labeling\_args} argument modifies the labels'
appearance: here, to be left-aligned and unclipped
(see Section \ref{sec:labeling}).

Another high-level function for visualizing conditional independence
models are the \codefun{pairs} methods for table and structable objects.
In contrast to \codefun{cotabplot} which conditions on variables,
the \codefun{pairs} methods create pairwise views of the table. 
The function produces, by default, a plot matrix 
having strucplot displays in the off-diagonal panels, and
the variable names (optionally, with univariate statistics) in the diagonal cells. 
Figure \ref{fig:pairs} shows a pairs display with mosaic plots visualizing mutual 
independence in the lower triangle,
association plots for the same in the upper triangle, and bar charts in the diagonal.

<<pairs>>=
pairs(hec, lower_panel = pairs_assoc, space = 0.3, 
      diag_panel_args = list(rot = -45, just_leveltext = c("left","bottom")))
@

%\setkeys{Gin}{width=\textwidth}
\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<pairsfig,fig=TRUE,echo=FALSE,width=8,height=8>>=
<<pairs>>
@
\caption{Pairs plot for the \data{HairEyeColor} data.}
\label{fig:pairs}
\end{center}
\end{figure}
\setkeys{Gin}{width=0.7\textwidth}

\noindent (The labels of the variables are to be read from left to
right and from top to bottom.) In plots produced by \codefun{pairs}, 
each panel's row and column define two variables $X$ and $Y$ used for the
specification of four different types of independence: pairwise,
total, conditional, and joint. The pairwise mosaic
matrix shows bivariate marginal relations between $X$ and $Y$, collapsed over all
other variables. The total independence mosaic matrix shows mosaic
plots for mutual independence, i.e., for marginal and conditional independence
among all pairs of variables. The conditional
independence mosaic matrix shows mosaic plots for marginal
independence of $X$ and $Y$, given all other variables. The joint independence
mosaic matrix shows mosaic plots for joint independence of all
pairs ($X$, $Y$) of variables from the others.

Since the matrix is symmetric, the upper and lower parts can
independently be used to display different 
types of independence models, or different strucplots displays
(mosaic, association, or sieve plots).
The available panel functions (\codefun{pairs\_assoc}, 
\codefun{pairs\_mosaic}, and \codefun{pairs\_sieve}) 
are simple wrappers to \codefun{assoc}, 
\codefun{mosaic}, and \codefun{sieve}, respectively. 
Obviously, seeing patterns in strucplot matrices 
becomes increasingly difficult with
higher dimensionality. Therefore, this plot is typically used with a suitable 
residual-based shading (see Section \ref{sec:shading}).

\subsection{Interactive plot modifications}

All strucplot core functions are supposed to produce conditional hierarchical
plots by the means of nested viewports, corresponding to the provided splitting
information. Thus, at the end of the plotting, each tile is associated with
a particular viewport. Each of those viewports has to be
conventionally named, enabling other strucplot modules, in particular the
labeling functions, to access specific tiles after they have been
plotted. The naming convention for the viewports is:

\begin{center}
\code{cell:\emph{Variable1}=\emph{Level1},\emph{Variable2}=\emph{Level2}} \dots
\end{center}

\noindent Clearly, these names depend on the splitting.
The following example shows how to access parts of the plot after it has
been drawn (see Figure \ref{fig:afterplot}):

<<viewportnames>>=
mosaic(~ Hair + Eye, data = hec, pop = FALSE)

seekViewport("cell:Hair=Blond")
grid.rect(gp = gpar(col = "red", lwd = 4))

seekViewport("cell:Hair=Blond,Eye=Blue")
grid.circle(r = 0.2, gp = gpar(fill = "cyan"))
@

\noindent Note that the viewport tree is removed by
default. Therefore, the \texttt{pop} argument has to be set to
\texttt{FALSE} when viewports shall be accessed.

\setkeys{Gin}{width=0.6\textwidth}
\begin{figure}[h]
\begin{center}
<<viewportnamesfig,fig=TRUE,echo=FALSE>>=
<<viewportnames>>
@
\caption{Adding elements to a mosaic plot after drawing.}
\label{fig:afterplot}
\end{center}
\end{figure}

In addition to the viewports, the main graphical elements get names
following a similar construction method. This allows to change
graphical parameters of plot elements \emph{after} 
the plotting (see Figure \ref{fig:changeplot}):

<<changeplot,results=verbatim>>=
assoc(Eye ~ Hair, data = hec, pop = FALSE)
getNames()[1:6]
grid.edit("rect:Hair=Blond,Eye=Blue", gp = gpar(fill = "red"))
@

%% code-chunk reuse does not work with parameter changing
\begin{figure}[h]
\begin{center}
<<changeplotfig,fig=TRUE,echo=FALSE>>=
x <- tab <- margin.table(HairEyeColor, 1:2)
x[] <- "light gray"
x["Blond","Blue"] <- "Red"
assoc(tab, gp = gpar(fill = x))
@
\caption{Changing graphical parameters of elements after drawing.}
\label{fig:changeplot}
\end{center}
\end{figure}

\section{Residual-based Shadings}
\label{sec:shading}

Unlike other graphics functions in base {\sf R}, the strucplot framework 
allows almost full control over the graphical parameters of all plot elements. In
particular, in association plots, mosaic plots, and sieve plots, 
the user can modify the graphical appearance of each tile individually. 
Built on top of this functionality, the
framework supplies a set of shading functions choosing colors
appropriate for the visualization of log-linear models.
The tiles' graphical parameters are set using the \code{gp} argument
of the functions of the strucplot framework. This argument basically expects an object
of class \class{gpar} whose components are arrays of the same shape 
(length and dimensionality) as the data table (see Section \ref{sec:gp}). 
For convenience, however, the user can also
supply a specialized graphical appearance control (``grapcon'') 
function that computes such an object given a vector of
residuals, or, alternatively, a generating function that takes
certain arguments and returns such a grapcon function (see Section
\ref{sec:shadingcustom}). We provide several shading functions, including
support for both HSV and HCL colors, and the
visualization of significance tests (see Section \ref{sec:overview}).

\subsection{Specifying graphical parameters of strucplot displays}
\label{sec:gp}

As an example, consider the \data{UCBAdmissions} data. 
In the table aggregated over departments, we
would like to highlight the (incidentally wrong) impression that there
were too many male students accepted compared to the presumably
discriminated female students (see Figure \ref{fig:ucb}):

<<ucb,results=verbatim>>=
(ucb <- margin.table(UCBAdmissions, 1:2))

(fill_colors <- matrix(c("dark cyan","gray","gray","dark magenta"), ncol = 2))

mosaic(ucb, gp = gpar(fill = fill_colors, col = 0))
@

\begin{figure}[h]
\begin{center}
<<ucbfig,fig=TRUE,echo=FALSE>>=
<<ucb>>
@

\caption{Mosaic plot for the \data{UCBAdmissions} data with highlighted cells.}
\label{fig:ucb}
\end{center}
\end{figure}

\noindent As the example shows, we create a fourfold table 
with appropriate colors (dark cyan for admitted male students and dark
magenta for rejected female students) and supply them to the \code{fill} component 
of the \class{gpar} object passed to the \code{gp} argument of \codefun{mosaic}. 
For visual clarity, we additionally hide the tiles' borders by setting the \code{col} 
component to 0 (white).

If the parameters specified in the \class{gpar} object are ``incomplete'',
they will be recycled along the last splitting
dimension. In the following example based on the \data{Titanic} data, 
we will highlight all cells corresponding to survived
passengers (see Figure \ref{fig:recycling}):

<<recycling>>=
mosaic(Titanic, gp = gpar(fill = c("gray","dark magenta")), 
                spacing = spacing_highlighting,
                labeling_args = list(abbreviate = c(Age = 3), rep = c(Survived = FALSE))
      )
@

\noindent Note that \codefun{spacing\_highlighting} sets the spaces
between tiles in the last dimension to 0. The \code{labeling\_args} 
argument ensures that labels do not overlap (see Section \ref{sec:labeling}).

\begin{figure}[h]
\begin{center}
<<recyclingfig,fig=TRUE,echo=FALSE>>=
<<recycling>>
@
\caption{Recycling of parameters, used for highlighting the survived
  passengers in the \data{Titanic} data.}
\label{fig:recycling}
\end{center}
\end{figure}

\subsection{Customizing residual-based shadings}
\label{sec:shadingcustom}

This flexible way of specifying graphical parameters is the basis for a suite of
shading functions that modify the tiles' appearance with respect to a vector of 
residuals, resulting from deviations of observed from expected
values under a given log-linear model. 
The idea is to visualize at least sign and absolute size of the residuals, but some
shadings, additionally, indicate overall significance. One particular
shading, the maximum shading, even allows to identify those cells that
cause the rejection of the null hypothesis.

Conceptually, the strucplot framework offers three alternatives to
add residual-based shading to plots:

\begin{enumerate}
\item Precomputing the graphical parameters (e.g., fill colors),
  encapsulating them into an object of class
  \class{gpar} as demonstrated in the previous section, and
  passing this object to the \code{gp} argument.
\item Providing a grapcon function to the \code{gp} argument that 
  takes residuals as input and returns an object as described in
  alternative 1.
 \item Providing a grapcon \emph{generating} function (`grapcon generator') 
  taking parameters and returning a function as described in alternative~2.
\end{enumerate}

\noindent For each of these approaches, we will demonstrate the
necessary steps to obtain a binary shading that visualizes the sign of
the residuals by a corresponding 
fill color (for simplicity, we will treat 0 as positive).

\subsubsection*{Alternative 1: Precomputed \class{gpar} object}

The first method is precomputing the graphical parameters ``by
hand''. We will use `light blue' color for positive and `light salmon'
color for negative residuals (see Figure \ref{fig:binary}):

<<shading1,results=verbatim>>=
expected <- independence_table(ucb)
(residuals <- (ucb - expected)  / sqrt(expected))

(shading1_obj <- ifelse(residuals > 0, "lightblue", "lightsalmon"))

mosaic(ucb, gp = gpar(fill = shading1_obj))
@

\begin{figure}[h]
\begin{center}
<<shading1fig,fig=TRUE,echo=FALSE>>=
<<shading1>>
@ 
\caption{Binary shading visualizing the sign of the residuals.}
\label{fig:binary}
\end{center}
\end{figure}

\subsubsection*{Alternative 2: Grapcon function}

For implementing alternative 2, we need to create a `shading function'
that computes \class{gpar} objects from
residuals. For that, we can just reuse the code from the previous step:

<<shading2>>=
shading2_fun <- function(res) gpar(fill = ifelse(res > 0, "lightblue", "lightsalmon"))
@

\noindent To create a mosaic display with binary shading, 
it now suffices to specify the data table along with
\codefun{shading2\_fun}:

<<shading3>>=
mosaic(ucb, gp = shading2_fun)
@

\noindent \codefun{mosaic} internally calls
\codefun{strucplot} which computes the residuals from the specified
independence model (total independence by default), passes them to
\codefun{shading2\_fun}, and uses the \class{gpar} object returned 
to finally create the plot.

Our \codefun{shading2\_fun} function might be useful, 
but can still be improved: the hard-wired colors should be
customizable. We cannot
simply extend the argument list to include, e.g., 
a \code{fill = c("lightblue", "lightsalmon")} argument because
\codefun{strucplot} will neither know how to handle it, nor let us
change the defaults. In fact, the interface of shading functions is fixed,
they are expected to take exactly one argument: a table of residuals. 
This is where generating functions (alternative 3) come
into play.

\subsubsection*{Alternative 3: Grapcon generator}

We simply wrap our grapcon shading function in another 
function that takes all additional arguments it needs to use, possibly
preprocesses them, and returns the actual shading function. This returned
function will have access to the parameters since in {\sf R}, nested
functions are lexically scoped. Thus, the grapcon generator
returns (`creates') a `parameterized' shading function with the minimal standard interface
\codefun{strucplot} requires. The following example shows the necessary
extensions for our running example:

<<shading3>>=
shading3a_fun <- function(col = c("lightblue", "lightsalmon")) {
  if (length(col) != 2) stop("Need exactly two colors!")
  function(res) gpar(fill = ifelse(res > 0, col[1], col[2]))  
}
@

\noindent In the call to \codefun{mosaic}, using the new
\codefun{shading3a\_fun} function, we can now simply change the colors:

<<shading4>>=
mosaic(ucb, gp = shading3a_fun(c("red","blue")))
@

\noindent (figure not shown).  The procedure described so far is a
rather general concept, applicable to a wide family of user-level
\pkg{grid} graphics. Indeed, the customization of other components of
the strucplot framework (labeling, spacing, legend, and core functions)
follows the same idea.  Now for the shading functions, more
customization is needed. Note that \codefun{shading3a\_fun} needs to be
evaluated by the user, even if the defaults are to be used. It is a
better idea to let \codefun{strucplot} call the generating function,
which, in particular, allows the passing of arguments that are computed
by \codefun{strucplot}. Since shading functions can be used for
visualizing significance (see Section \ref{sec:overview}), it makes
sense for generating functions to have access to the model, i.e.,
observed and expected values, residuals, and degrees of freedom. For
example, the \codefun{shading\_max} generating function computes a
permutation distribution of the maximum statistic and $p$ values for
specified significance levels based on the observed table to create
data-driven cut-off points. If this was done in the shading function
itself, the permutation statistic would be recomputed every time the
shading function is called, resulting in possibly severe performance
loss and numerical inconsistencies. Therefore, generating functions for
shadings are required to take at least the parameters \code{observed},
\code{expected}, \code{residuals}, and \code{df} (these are provided by
the strucplot framework), followed by other parameters controlling the
shading appearance (to be specified by the user):

<<shading4>>=
shading3b_fun <- function(observed = NULL, residuals = NULL, expected = NULL, 
    df = NULL, col = c("lightblue", "lightsalmon")) {
  if (length(col) != 2) stop("Need exactly two colors!")
  function(res) gpar(fill = ifelse(res > 0, col[1], col[2]))  
}
class(shading3b_fun) <- "grapcon_generator"
@

In some sense, generating
functions for shadings are parameterized both by the user and the
strucplot framework. For shading functions that require model
information, the user-specified parameters are 
to be passed to the \code{gp\_args} argument instead, and for this to
work, the generating function needs a class attribute to be distinguishable from
the ``normal'' shading functions. For others
(like our simple \codefun{shading3b\_fun}) this is optional, but
recommended for consistency:

<<shading5>>=
mosaic(ucb, gp = shading3b_fun, gp_args = list(col = c("red","blue")))
@

\noindent The final \codefun{shading3b\_fun} pretty much resembles 
\codefun{shading\_binary}, one of the standard shading
functions provided by the \pkg{vcd} package.

\subsection[An overview of the shading functions in vcd]{An overview of the shading functions in \pkg{vcd}}
\label{sec:overview}

\cite{vcd:Friendly:1994}
suggested a residual-based shading for the mosaic tiles that can also be applied
to the rectangles in association plots \citep{vcd:Meyer+Zeileis+Hornik:2003}.
Apart from \codefun{shading\_binary}, there are currently two basic
shadings available in \pkg{vcd}:
\codefun{shading\_hcl} and \codefun{shading\_hsv}, as well
as two derived functions: \codefun{shading\_Friendly} building upon
\codefun{shading\_hsv}, and \codefun{shading\_max} building upon \codefun{shading\_hcl}.
\codefun{shading\_hsv} and \codefun{shading\_hcl} provide
the same conceptual tools, but use different color spaces: the
Hue-Saturation-Value (HSV) and the Hue-Chroma-Luminance (HCL) scheme,
respectively. We will first expose the basic concept of these shading
functions using the HSV space, and
then briefly explain the differences to the HCL space \citep[a detailed
discussion can be found in][]{vcd:Zeileis+Meyer+Hornik:2005}. 
The HCL space is trickier
to use, but preferable to the HSV space from a perceptual point of view.

In the HSV space, colors are specified in three dimensions: Hue,
Saturation (`colorfulness'), and Value (`lightness', amount of gray). 
These three dimensions are used by \codefun{shading\_hsv} to visualize information 
about the residuals and the underlying independence model. The hue 
indicates the residuals' sign: by default, blue for positive, and red
for negative residuals. The saturation of a residual 
is set according to its size: high saturation for large, and low
saturation for small residuals. 
Finally, the overall lightness is used to indicate the significance of
a test statistic: light colors for significant, 
and dark colors for non-significant results.

As an example, we will visualize the association of hair and eye
color in the \data{HairEyeColor} data set (see Figure \ref{fig:haireye})

<<haireye>>=
haireye <- margin.table(HairEyeColor, 1:2)
mosaic(haireye, gp = shading_hsv)
@

\begin{figure}[h]
\begin{center}
<<haireyefig,fig=TRUE,echo=FALSE,width=7,height=7>>=
<<haireye>>
@
\caption{Shaded residuals in the \data{HairEyeColor} data set---two cut-off points.}
\label{fig:haireye}
\end{center}
\end{figure}

\noindent Large positive residuals (greater than $4$) can be found for
brown eyes/black hair and blue eyes/blond hair, and are colored in saturated blue. 
On the other hand, there is a large negative residual
(less than $-4$) for brown eyes/blond hair, colored deep red. There are also three
medium-sized positive (negative) residuals between 2 and 4 ($-2$ and
$-4$): the colors for them are less saturated. Residuals between $-2$ and $2$ 
are shaded in white.
The heuristic for choosing the cut-off points $2$ and $4$ is that the Pearson residuals
are approximately standard normal which implies that the highlighted cells are those with
residuals \emph{individually} significant at approximately the $\alpha
= 0.05$ and $\alpha = 0.0001$ levels, respectively. These default cut-off points can
be changed to alternative values using the \code{interpolate}
argument (see Figure \ref{fig:interpolate}):

<<interpolate>>=
mosaic(haireye, gp = shading_hsv, gp_args = list(interpolate = 1:4))
@

\begin{figure}[h]
\begin{center}
<<interpolatefig,fig=TRUE,echo=FALSE,width=7,height=7>>=
<<interpolate>>
@
\caption{Shaded residuals in the \data{HairEyeColor} data set---four cut-off points.}
\label{fig:interpolate}
\end{center}
\end{figure}

\noindent The elements of the numeric vector passed to
\code{interpolate} define the knots of an interpolating step function
used to map the absolute residuals to saturation levels. 
The \code{interpolate} argument also accepts a user-defined function,
which then is called with the absolute residuals to get a vector of
cut-off points. Thus, it is possible to automatically choose the cut-off points 
in a data-driven way. For example,
one might think that the extension from four cut-off points 
to a continuous shading---visualizing the whole range of 
residuals---could be useful. We simply need a one-to-one
mapping from the residuals to the saturation values:

<<continuous1>>=
ipol <- function(x) pmin(x/4, 1)
@

\noindent Note that this \codefun{ipol} 
function maps residuals greater than 4 to a saturation
level of 1. However, the resulting plot (Figure \ref{fig:continuous}) is 
deceiving:

<<continuous2>>=
mosaic(HairEyeColor, gp = shading_hsv, gp_args = list(interpolate = ipol),
labeling_args = list(abbreviate = c(Sex = TRUE)))
@

\begin{figure}[p]
\begin{center}
<<continuousfig,fig=TRUE,echo=FALSE>>=
<<continuous2>>
@
\caption{The \data{HairEyeColor} data with continuous shading.}
\label{fig:continuous}
\end{center}
\end{figure}

\noindent Too much color makes it difficult to interpret the image, and the subtle color
differences are hard to catch. Therefore, we only included shadings
with discrete cut-off points.

The third remaining dimension, the value, is used for
visualizing the significance of a test statistic. The user can either
directly specify the $p$ value, or, alternatively, a function that
computes it, to the \code{p.value} argument. Such a function must take
observed and expected values, residuals, and degrees of freedom (used
by the independence model) as arguments. 
If nothing is specified, the $p$ value is computed from 
a $\chi^2$ distribution with \code{df} degrees of
freedom. The \code{level} argument is used to specify the confidence
level:  if \code{p.value} is smaller than \code{1 - level}, light colors are used, 
otherwise dark colors are employed. 
The following example using the \data{Bundesliga} data 
shows the relationship of home goals and away
goals of Germany's premier soccer league in 1995: although there are two
``larger'' residuals (one greater than 2, one less then $-2$), the
$\chi^2$ test does not reject the null hypothesis of
independence. Consequently, the colors appear dark (see Figure \ref{fig:bundesliga}):

<<bundesliga>>=
bl <- xtabs(~ HomeGoals + AwayGoals, data = Bundesliga, subset = Year == 1995)
mosaic(bl, gp = shading_hsv)
@

\begin{figure}[p]
\begin{center}
<<bundesligafig,fig=TRUE,echo=FALSE>>=
<<bundesliga>>
@
\caption{Non-significant $\chi^2$ test using part of the \data{Bundesliga} data.}
\label{fig:bundesliga}
\end{center}
\end{figure}

A shading function building upon \codefun{shading\_hsv} is \codefun{shading\_Friendly},
implementing the shading introduced by \cite{vcd:Friendly:1994}. In addition
to the defaults of the HSV shading, it uses the border color and line type to 
redundantly code the residuals' sign. The following example again uses 
the \data{Bundesliga} data from above, this time using the Friendly 
scheme and, in addition, an alternative legend (see Figure \ref{fig:friendly}):

<<friendly>>=
mosaic(bl, gp = shading_Friendly, legend = legend_fixed, zero_size = 0)
@

\begin{figure}[h]
\begin{center}
<<friendlyfig,fig=TRUE,echo=FALSE>>=
<<friendly>>
@
\caption{The \data{Bundesliga} data for 1995 using the Friendly
  shading and a legend with fixed bins.}
\label{fig:friendly}
\end{center}
\end{figure}

\noindent (The \code{zero\_size = 0} argument removes the bullets indicating zero 
observed values. This feature is not provided
in the original {\sf SAS} implementation of the Friendly mosaic plots.)

As introduced before, the default shading scheme is not
\codefun{shading\_hsv} but \codefun{shading\_hcl} due to the better
perceptual characteristics of the HCL color space.
Figure \ref{fig:shadingHSVHCL} depicts 
the HSV space in the upper panel and the HCL space in the lower panel.
On the left (right) side, we see the color scales for red (blue)
hue, respectively. The $x$-axis represents the colorfulness, and the
$y$-axis the brightness.
The boxes represent the diverging color palettes used for the shadings.
For the HSV space, we can see that the effect of changing the 
level of brightness (`value') is not the same for different levels of 
saturation, and again not the same for the two different hues.
In fact, in the HSV space all dimensions are confounded, which 
obviously is problematic for coding information. In contrast, the HCL color
space offers perceptually uniform colors: as can be seen from the lower panel, 
the chroma is homogeneous for different levels of luminance. 
Unfortunately, this comes at the 
price of the space being irregularly shaped, making it difficult to automatically select 
diverging color palettes. The following example again 
illustrates the \data{HairEyeColor} data, this time with HCL colors 
(Figure \ref{fig:hclshading1} depicts the default palette, 
and Figure \ref{fig:hclshading2} an alternative setting):

<<auxiliary_functions,echo=FALSE,results=hide>>=
hue.slice <- function(hue, grid.n = 101, type = c("HCL", "HSV"), plot = TRUE, fixup = FALSE)
{
  type <- match.arg(type)
  if(type == "HCL") {
    chroma = seq(0, 100, length = grid.n)
    luminance = seq(0, 100, length = grid.n)
    nc <- length(chroma)
    nl <- length(luminance)
    color.slice <- outer(chroma, luminance, function(y, x) hcl(hue, x, y, fixup = fixup))
    xlab <- "chroma"
    ylab <- "luminance"
    main <- paste("hue =", round(hue, digits = 0))
  } else {
    chroma = seq(0, 1, length = grid.n)
    luminance = seq(0, 1, length = grid.n)
    nc <- length(chroma)
    nl <- length(luminance)
    color.slice <- outer(chroma, luminance, function(y, x) hsv(hue, x, y))
    xlab <- "saturation"
    ylab <- "value"
    main <- paste("hue =", round(hue, digits = 3))
  }
  if(plot) {
    plot(0.5, 0.5, xlim = range(chroma), ylim = range(luminance), type = "n", axes = FALSE,
         xlab = xlab, ylab = ylab, yaxs = "i", xaxs = "i", main = main)
    for(i in 1:(nc-1)) {
      rect(chroma[i], luminance[-nl], chroma[i] + 100/(nc-1), luminance[-1], border = color.slice[,i+1], col = color.slice[,i+1])
    }
    axis(1)
    axis(2)
    box()
  }
  colnames(color.slice) <- chroma
  rownames(color.slice) <- luminance
  attr(color.slice, "type") <- type
  class(color.slice) <- "slice"
  invisible(color.slice)
}
@

\setkeys{Gin}{width=.8\textwidth}
\begin{figure}[p]
\begin{center}
<<shading_HSV,fig=TRUE,echo=FALSE,height=7,width=8>>=
## generate colors
hue23 <- hue.slice(2/3, grid.n = 101, plot = FALSE, type = "HSV")
hue0 <- hue.slice(0, grid.n = 101, plot = FALSE, type = "HSV")
saturation <- as.numeric(colnames(hue23))
value <- as.numeric(rownames(hue23))

## select those with value >= 0.5
hue23 <- hue23[value >= .5, ]
hue0 <- hue0[value >= .5, ]
value <- value[value >= .5]
nl <- nrow(hue23)
nc <- ncol(hue23)

## plot 2 slides from HSV space
plot(0.5, 0.5, xlim = c(-1, 1), ylim = c(0, 1), type = "n", axes = FALSE,
       xlab = "", ylab = "", yaxs = "i", xaxs = "i", main = "")
for(i in 1:(nc-1)) {
  rect(saturation[i], value[-nl], saturation[i] + 1/(nc-1), value[-1], border = hue23[,i+1], col = hue23[,i+1])
}
for(i in 1:(nc-1)) {
  rect(-saturation[i], value[-nl], -(saturation[i] + 1/(nc-1)), value[-1], border = hue0[,i+1], col = hue0[,i+1])
}
axis(2, at = c(50, 75, 100)/100, labels = c(0.5, 0.75, 1))
axis(4, at = c(50, 75, 100)/100, labels = c(0.5, 0.75, 1))
axis(3, at = -4:4*.25, labels=c(4:0*.25, 1:4*.25))
mtext(c("hue = 0", "hue = 2/3"), side = 3, at = c(-.5, .5), line = 3, cex = 1.2)
mtext("saturation", side = 3, at = 0, line = 2)
mtext("value", side = 2, at = .75, line = 2)
mtext("value", side = 4, at = .75, line = 2)
lines(c(-1, 1), c(.5, .5))

## significant colors
rect(-1, 0.95, -.90, 1, col = hsv(0, 1, 1))
rect(-0.45, 0.95, -.55, 1, col = hsv(0, 0.5, 1))
rect(-.05, .95, .05, 1, col = hsv(2/3, 0, 1))
rect(0.45, 0.95, .55, 1, col = hsv(2/3, 0.5, 1))
rect(.90, .95, 1, 1, col = hsv(2/3, 1, 1))

text(-1, .33, "significant", pos = 4, cex = 1.2)
rect(-1, .20, -.80, .30, col = hsv(0, 1, 1))
rect(-.40, .20, -0.6, .30, col = hsv(0, 0.5, 1))
rect(-.20, .20, 0, .30, col = hsv(0, 0, 1))
rect(0, .20, .20, .30, col = hsv(2/3, 0, 1))
rect(0.4, .20, .60, .30, col = hsv(2/3, .5, 1))
rect(.80, .20, 1, .30, col = hsv(2/3, 1, 1))

lines(c(-.9, -.55), c(0.975, .975), lty = 2)
lines(c(-.45, -.05), c(0.975, .975), lty = 2)
lines(c(.45, .05), c(0.975, .975), lty = 2)
lines(c(.9, .55), c(0.975, .975), lty = 2)

## non-significant colors
rect(-1, 0.5, -.90, 0.55, col = hsv(0, 1, 0.5))
rect(-0.45, 0.5, -.55, 0.55, col = hsv(0, 0.5, 0.5))
rect(-.05, .5, .05, 0.55, col = hsv(2/3, 0, 0.5))
rect(0.45, 0.5, .55, 0.55, col = hsv(2/3, 0.5, 0.5))
rect(.90, .5, 1, 0.55, col = hsv(2/3, 1, 0.5))

text(-1, .13, "non-significant", pos = 4, cex = 1.2)
rect(-1, 0, -.80, .10, col = hsv(0, 1, 0.5))
rect(-.60, 0, -.4, .10, col = hsv(0, 0.5, 0.5))
rect(-.20, 0, 0, .10, col = hsv(0, 0, 0.5))
rect(0, 0, .20, .10, col = hsv(2/3, 0, 0.5))
rect(0.4, 0, .60, .1, col = hsv(2/3, .5, 0.5))
rect(.80, 0, 1, .10, col = hsv(2/3, 1, 0.5))

lines(c(-.9, -.55), c(0.525, .525), lty = 2)
lines(c(-.45, -.05), c(0.525, .525), lty = 2)
lines(c(.45, .05), c(0.525, .525), lty = 2)
lines(c(.9, .55), c(0.525, .525), lty = 2)
@

<<shading_HCL,fig=TRUE,echo=FALSE,height=7,width=8>>=
## generate colors
hue260 <- hue.slice(260, grid.n = 101, plot = FALSE)
hue360 <- hue.slice(360, grid.n = 101, plot = FALSE)
mychroma <- as.numeric(colnames(hue260))
luminance <- as.numeric(rownames(hue260))

## select those with lumincance >= 50
hue260 <- hue260[luminance >= 50, ]
hue360 <- hue360[luminance >= 50, ]
luminance <- luminance[luminance >= 50]
nc <- ncol(hue260)
nl <- nrow(hue260)

## plot 2 slides from HCL space
plot(0.5, 0.5, xlim = c(-100, 100), ylim = c(0, 100), type = "n", axes = FALSE,
       xlab = "", ylab = "", yaxs = "i", xaxs = "i", main = "")
for(i in 1:(nc-1)) {
  rect(mychroma[i], luminance[-nl], mychroma[i] + 100/(nc-1), luminance[-1], border = hue260[,i+1], col = hue260[,i+1])
}
for(i in 1:(nc-1)) {
  rect(-mychroma[i], luminance[-nl], -(mychroma[i] + 100/(nc-1)), luminance[-1], border = hue360[,i+1], col = hue360[,i+1])
}
axis(2, at = c(50, 70, 90, 100), labels = c(50, 70, 90, 100))
axis(4, at = c(50, 70, 90, 100), labels = c(50, 70, 90, 100))
axis(3, at = -4:4*25, labels=c(4:0*25, 1:4*25))
mtext(c("hue = 0", "hue = 260"), side = 3, at = c(-50, 50), line = 3, cex = 1.2)
mtext("chroma", side = 3, at = 0, line = 2)
mtext("luminance", side = 2, at = 75, line = 2)
mtext("luminance", side = 4, at = 75, line = 2)
lines(c(-100, 100), c(50, 50))

## significant colors
rect(-100, 47.5, -90, 52.5, col = hcl(0, 100, 50))
rect(-55, 67.5, -45, 72.5, col = hcl(0, 50, 70))
rect(-5, 95, 5, 100, col = hcl(260, 0, 100))       ## grey vs. white
rect(-5, 87.5, 5, 92.5, col = hcl(260, 0, 90))     ## grey vs. white
rect(45, 67.5, 55, 72.5, col = hcl(260, 50, 70))
rect(90, 47.5, 100, 52.5, col = hcl(260, 100, 50))

text(-100, 33, "significant", pos = 4, cex = 1.2)
rect(-100, 20, -80, 30, col = hcl(0, 100, 50))
rect(-60, 20, -40, 30, col = hcl(0, 50, 70))
rect(-20, 20, 0, 30, col = hcl(0, 0, 90))       
rect(0, 20, 20, 30, col = hcl(260, 0, 90))
#white# rect(-20, 20, 0, 30, col = hcl(0, 0, 100))
#white# rect(0, 20, 20, 30, col = hcl(260, 0, 100))
rect(40, 20, 60, 30, col = hcl(260, 50, 70))
rect(80, 20, 100, 30, col = hcl(260, 100, 50))

lines(c(-45, -5), c(72.5, 87.5), lty = 2)
lines(c(45, 5), c(72.5, 87.5), lty = 2)
lines(c(-95, -55), c(52.5, 67.5), lty = 2)
lines(c(95, 55), c(52.5, 67.5), lty = 2)

## non-significant colors
rect(-25, 47.5, -15, 52.5, col = hcl(0, 20, 50))
rect(-15, 67.5, -5, 72.5, col = hcl(0, 10, 70))
rect(5, 67.5, 15, 72.5, col = hcl(260, 10, 70))
rect(25, 47.5, 15, 52.5, col = hcl(260, 20, 50))

text(-100, 13, "non-significant", pos = 4, cex = 1.2)
rect(-60, 0, -40, 10, col = hcl(0, 20, 50))
rect(-40, 0, -20, 10, col = hcl(0, 10, 70))
rect(-20, 0, 0, 10, col = hcl(0, 0, 90))
rect(0, 0, 20, 10, col = hcl(260, 0, 90))
rect(20, 0, 40, 10, col = hcl(260, 10, 70))
rect(40, 0, 60, 10, col = hcl(260, 20, 50))

lines(c(-18.75, -11.25), c(52.5, 67.5), lty = 2)
lines(c(-8.75, -1.25), c(72.5, 87.5), lty = 2)
lines(c(18.75, 11.75), c(52.5, 67.5), lty = 2)
lines(c(8.75, 1.25), c(72.5, 87.5), lty = 2)
@ 
\caption{Residual-based shadings in HSV (upper) and HCL space (lower).}
\label{fig:shadingHSVHCL}
\end{center}
\end{figure}

<<hclshading1>>=
mosaic(haireye, gp = shading_hcl)
@ 
<<hclshading2>>=
mosaic(haireye, gp = shading_hcl, gp_args = list(h = c(130, 43), c = 100, l = c(90, 70)))
@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[p]
\begin{center}
<<hclshading1fig,fig=TRUE,echo=FALSE>>=
<<hclshading1>>
@
\caption{The \data{HairEyeColor} data, using default HCL color palette.}
\label{fig:hclshading1}
\end{center}
\end{figure}

\begin{figure}[p]
\begin{center}
<<hclshading2fig,fig=TRUE,echo=FALSE>>=
<<hclshading2>>
@
\caption{The \data{HairEyeColor} data, using a custom HCL color palette.}
\label{fig:hclshading2}
\end{center}
\end{figure}

A more `advanced' function building upon \codefun{shading\_hcl} 
is \codefun{shading\_max}, using the
maximum statistic both to conduct the independence test and to 
visualize significant \emph{cells} causing the
rejection of the independence hypothesis \citep{vcd:Meyer+Zeileis+Hornik:2003}. The
\code{level} argument of \codefun{shading\_max} 
then can be used to specify several confidence
levels from which the corresponding cut-off points are computed. 
By default, two cut-off points are computed corresponding 
to confidence levels of $90\%$ and $99\%$, respectively.
In the following example, we investigate the effect of a new treatment
for rheumatoid arthritis on a group of female patients using the
maximum shading (see Figure \ref{fig:maximum}):

<<arthritis>>=
mosaic(~ Treatment + Improved, data = Arthritis, subset = Sex == "Female",
       gp = shading_max)
@

\begin{figure}[h]
\begin{center}
<<arthritisfig,fig=TRUE,echo=FALSE>>=
<<arthritis>>
@
\caption{Significant maximum test on female patients of the \data{Arthritis}
  data.}
\label{fig:maximum}
\end{center}
\end{figure}

\noindent The maximum test is significant although the residuals are
all in the $\left[-2,2\right]$ interval.
The \codefun{shading\_hcl} function with default cut-off points would
not have shown any color. In addition, since the test statistic is the
maximum of the absolute Pearson residuals, \emph{each} colored
residual violates the null hypotheses of independence, and thus, the
`culprits' can immediately be identified.

\section[Flexible Labeling]{Flexible Labeling}
\label{sec:labeling}

One of the major enhancements in package \pkg{vcd} compared to
\codefun{mosaicplot} and \codefun{assocplot} in
base {\sf R} is the labeling in the strucplot framework which 
offers many more features and flexibility. 
Like shading, spacing, and drawing of legend and core plot, 
labeling is now carried out by grapcon functions, rendering labeling completely
modular.
The user supplies either a labeling function, or, alternatively, 
a generating function that parameterizes a labeling function, 
to \codefun{strucplot} which then draws the labels.
Labeling is well-separated from the actual plotting that occurs in the
low-level core functions. It only relies on the viewport tree produced by them, and the
`\texttt{dimnames}' attribute of the visualized table. Labeling
functions are grapcons that ``add ink to the canvas'': the drawing of the
labels happens after the actual plot has been drawn by the core function.
Thus, it is possible to supply one's own labeling function, or
to combine some of the basic functions to produce a more complex labeling.
In the following, we describe the three basic modules
(\codefun{labeling\_text}, \codefun{labeling\_list}, and \codefun{labeling\_cells}) and
derived functions that build upon them.

\subsection[Labels in the borders]{Labels in the borders: \texttt{labeling\_text()}}

\codefun{labeling\_text} is the default for all strucplot displays. It
plots labels in the borders similar to the \codefun{mosaicplot}
function in base {\sf R}, but is much more flexible: it is not limited
to 4 dimensions, and the positioning and graphical parameters 
of levels and variable names are customizable. In addition,
the problem of overlapping labels can be handled in several ways.

As an example, consider the \data{Titanic} data, consisting
of 4 categorical variables: survival (`survived'), gender (i.e., `sex'), age, and
crew. By default, the variable names and levels are plotted `around' the
plot in a counter-clockwise way (see Figure \ref{fig:defaults}):

<<default>>=
mosaic(Titanic)
@

\begin{figure}[p]
\begin{center}
<<defaultfig,fig=TRUE,echo=FALSE>>=
<<default>>
@
\caption{Mosaic plot for the \data{Titanic} data with default settings
  for labeling.}
\label{fig:defaults}
\end{center}
\end{figure}

\noindent Note that the last two levels of the `survived' variable do
overlap, as well as some adult and child labels of the `age' Variable.
This issue can be addressed in several ways. The `brute force'
method is to enable clipping for these dimensions (see Figure \ref{fig:clipping}):

<<clipping>>=
mosaic(Titanic, labeling_args = list(clip = c(Survived = TRUE, Age = TRUE)))
@

\begin{figure}[p]
\begin{center}
<<clippingfig,fig=TRUE,echo=FALSE>>=
<<clipping>>
@
\caption{The effect of clipping.}
\label{fig:clipping}
\end{center}
\end{figure}

\noindent The \code{clip} parameter is passed to the labeling function
via the \code{labeling\_args} argument which takes a list of
parameters. \code{clip} itself takes a vector of logicals (one for
each dimension). 
% as mentioned before
Almost all vectorized arguments in the strucplot
framework can be abbreviated in the following way: unnamed components
(or the defaults, if there are none) are recycled as needed, but
overridden by the named components. Here, the default is \code{FALSE},
and therefore clipping is enabled only for the `survived' and `age' variables.
A more sensible solution to the overlap problem is to abbreviate the 
levels (see Figure \ref{fig:abbreviating}):

<<abbreviating>>=
mosaic(Titanic, labeling_args = list(abbreviate = c(Survived = TRUE, Age = 3)))
@

\begin{figure}[h]
\begin{center}
<<abbreviatingfig,fig=TRUE,echo=FALSE>>=
<<abbreviating>>
@
\caption{Abbreviating.}
\label{fig:abbreviating}
\end{center}
\end{figure}

\noindent The \code{abbreviate} argument takes a vector of
integers indicating the number of significant characters the levels should be
abbreviated to (\code{TRUE} is interpreted as 1, obviously). Abbreviation
is performed using the \codefun{abbreviate} function in base {\sf R}. Another
possibility is to rotate the levels (see Figure~\ref{fig:rotating}):

<<rotate>>=
mosaic(Titanic, labeling_args = list(rot_labels = c(bottom = 90, right = 0), 
offset_varnames = c(right = 1), offset_labels = c(right = 0.3)), margins = c(right = 4, bottom = 3))
@

\begin{figure}[p]
\begin{center}
<<rotatefig,fig=TRUE,echo=FALSE>>=
<<rotate>>
@
\caption{Rotating labels.}
\label{fig:rotating}
\end{center}
\end{figure}

\noindent Finally, we could also inhibit the output of repeated levels 
(see Figure \ref{fig:repeat}):

<<repeat>>=
mosaic(Titanic, labeling_args = list(rep = c(Survived = FALSE, Age = FALSE)))
@

\begin{figure}[p]
\begin{center}
<<repeatfig,fig=TRUE,echo=FALSE>>=
<<repeat>>
@
\caption{Inhibiting the repetition of levels.}
\label{fig:repeat}
\end{center}
\end{figure}

We now proceed with a few more `cosmetic' features (which do not all
produce satisfactory results for our sample data).
A first simple, but effectful modification is to position
all labels and variables left-aligned:  (see Figure \ref{fig:left}):

<<left>>=
mosaic(Titanic, labeling_args = list(pos_varnames = "left", pos_labels = "left", 
       just_labels = "left", rep = FALSE))
@

\begin{figure}[h]
\begin{center}
<<leftfig,fig=TRUE,echo=FALSE>>=
<<left>>
@
\caption{Left-aligning.}
\label{fig:left}
\end{center}
\end{figure}

\noindent Note that obviously we need to change the justification 
to \code{"left"} as well. We can
achieve the same effect by using the convenience function \codefun{labeling\_left}:

<<left2>>=
mosaic(Titanic, labeling = labeling_left)
@

\noindent Next, we show how to put all levels to the
bottom and right margins, and all variable names to the top and left
margins (see Figure \ref{fig:margins}):

<<margins>>=
mosaic(Titanic, labeling_args = list(tl_labels = FALSE, tl_varnames = TRUE, abbreviate = c(Survived = 1, Age = 3)))
@

\begin{figure}[p]
\begin{center}
<<marginsfig,fig=TRUE,echo=FALSE>>=
<<margins>>
@
\caption{Changes in the margins.}
\label{fig:margins}
\end{center}
\end{figure}

\noindent The \var{tl\_foo} (``top left'') arguments are \code{TRUE} by default. 
Now, we will add boxes to the labels and additionally
enable clipping (see Figure \ref{fig:boxes}):

<<boxes>>=
mosaic(Titanic, labeling_args = list(tl_labels = FALSE, tl_varnames = TRUE, 
       boxes = TRUE, clip = TRUE))
@

\begin{figure}[p]
\begin{center}
<<boxesfig,fig=TRUE,echo=FALSE>>=
<<boxes>>
@
\caption{Boxes and Clipping.}
\label{fig:boxes}
\end{center}
\end{figure}

\noindent The values to \code{boxes} and \code{clip} are recycled for
all dimensions. The result is pretty close to what calling
\codefun{mosaic} with the \codefun{labeling\_cboxed}
wrapper does, except that variables and levels, by default, are put to the top and
to the left of the plot:

<<boxes2>>=
mosaic(Titanic, labeling = labeling_cboxed)
@

\noindent Another variant is to put the variable names into the
same line as the levels (see Figure \ref{fig:labbl}):

<<labbl>>=
mosaic(Titanic, labeling_args = list(tl_labels = TRUE, boxes = TRUE,
       clip = c(Survived = FALSE, TRUE), 
       labbl_varnames = TRUE), margins = c(left = 4, right = 1, 3))
@

\begin{figure}[h]
\begin{center}
<<labblfig,fig=TRUE,echo=FALSE>>=
<<labbl>>
@
\caption{Variable names beneath levels.}
\label{fig:labbl}
\end{center}
\end{figure}

\noindent \code{labbl\_varnames} (``variable names to the bottom/left of the
labels'') is a vector of logicals indicating the side for the
variable names. The resulting layout is close to what
\codefun{labeling\_lboxed} produces, except that variables and levels,
by default, are left-aligned and put to the bottom and to the right of the plot:

<<labbl2>>=
mosaic(Titanic, labeling = labeling_lboxed, margins = c(right = 4, left = 1, 3))
@

\noindent A similar design is used by the \codefun{doubledecker} function.

\subsection[Labels in the cells]{Labels in the cells: \texttt{labeling\_cells()}}

This labeling draws both variable names and levels in the
cells. As an example, we use the \data{PreSex} data on pre- and
extramarital sex and divorce (see Figure \ref{fig:cell}):

<<cell>>=
mosaic(~ MaritalStatus + Gender, data = PreSex, labeling = labeling_cells)
@

\begin{figure}[h]
\begin{center}
<<cellfig,fig=TRUE,echo=FALSE,width=7,height=7>>=
<<cell>>
@
\caption{Cell labeling for the \data{PreSex} data.}
\label{fig:cell}
\end{center}
\end{figure}

\noindent In the case of narrow cells, it might be useful to
abbreviate labels and/or variable names and turn off clipping (see Figure \ref{fig:cell2}):

<<cell2>>=
mosaic(~ PremaritalSex + ExtramaritalSex, data = PreSex, 
labeling = labeling_cells(abbreviate_labels = TRUE, abbreviate_varnames = TRUE, clip = FALSE))
@

\begin{figure}[h]
\begin{center}
<<cell2fig,fig=TRUE,echo=FALSE>>=
<<cell2>>
@
\caption{Cell labeling for the \data{PreSex} data, labels abbreviated.}
\label{fig:cell2}
\end{center}
\end{figure}

\noindent For some data, it might be convenient to combine cell
labeling with border labeling as done by \codefun{labels\_conditional}
(see Figure \ref{fig:conditional}):

<<conditional>>=
mosaic(~ PremaritalSex + ExtramaritalSex | MaritalStatus + Gender, data = PreSex, labeling = labeling_conditional(abbreviate_varnames = TRUE, abbreviate_labels = TRUE, clip = FALSE))
@

\begin{figure}[h]
\begin{center}
<<conditionalfig,fig=TRUE,echo=FALSE>>=
<<conditional>>
@
\caption{Conditional labeling for the \data{PreSex}, labels abbreviated.}
\label{fig:conditional}
\end{center}
\end{figure}

\noindent Additionally, the cell labeling allows the user to add
arbitrary text to the cells by supplying a character array in the same
shape than the data array to the \code{text} argument (cells with missing values
are ignored). In the following example using the `Titanic' data, 
this is used to add all observed values greater
than 5 to the cells after the mosaic has been plotted (see Figure \ref{fig:text}):

<<text>>=
mosaic(Titanic, labeling_args = list(abbreviate = c(Survived = 1, Age = 4)), pop = FALSE)

tab <- ifelse(Titanic < 6, NA, Titanic)
labeling_cells(text = tab, clip = FALSE)(Titanic)
@

\begin{figure}[h]
\begin{center}
<<textfig,fig=TRUE,echo=FALSE>>=
<<text>>
@
\caption{User-supplied Text added to a mosaic display of the \data{Titanic} data.}
\label{fig:text}
\end{center}
\end{figure}

\subsection[A simple list of labels]{A simple list of labels: \texttt{labeling\_list()}}

If problems with overlapping labels cannot satisfactorily resolved,
the last remedy could be to simply list the levels below the plot
(see Figure \ref{fig:list}):

<<list>>=
mosaic(Titanic, labeling = labeling_list, margins = c(bottom = 5))
@

\setkeys{Gin}{width=0.8\textwidth}
\begin{figure}[h]
\begin{center}
<<listfig,fig=TRUE,echo=FALSE>>=
<<list>>
@
\caption{Labels indicated below the plot.}
\label{fig:list}
\end{center}
\end{figure}

\noindent The number of columns can be specified.

\section{Customizable Spacing}
\label{sec:spacing}

Spacing of strucplot displays is customizable in a similar way than shading. The
\code{spacing} argument of the \codefun{strucplot} function takes a
list of \class{unit} vectors, one for each dimension, specifying the
space between the tiles corresponding to the levels. As an example,
consider the introductory example of the \data{arthritis} data 
(Figure \ref{fig:arthritis}). Since we are interested in the effect of
the medicament in the placebo and treatment groups, a mosaic plot is
certainly appropriate to visualize the three levels of `Improved' in
the two `Treatment' strata. However, following a slightly different approach,
we might as well want to \emph{highlight} patients with `Marked' improvement in both
groups, that is, produce a spine plot. To obtain such a display within 
the strucplot framework\footnote{The \pkg{vcd} package also provides 
the \codefun{spine} function for spine plots and spinograms.}, it
suffices to set the space between the `Improved' tiles to 0 (see
Figure \ref{fig:artspine}):

<<artspine,echo=TRUE,results=verbatim>>=
(art <- structable(~Treatment + Improved, data = Arthritis, split_vertical = TRUE))
(my_spacing <- list(unit(0.5, "lines"), unit(c(0, 0), "lines")))
my_colors <- gpar(fill = c("lightgray", "darkgray", "black"))
mosaic(art, spacing = my_spacing, gp = my_colors)
@

\begin{figure}[p]
\begin{center}
<<artspinefig,echo=FALSE,fig=TRUE>>=
<<artspine>>
@ 
\caption{Spine plot for the \data{arthritis} data using the strucplot framework.}
\label{fig:artspine}
\end{center}
\end{figure}

\noindent The strucplot framework also provides a set of spacing grapcons 
(and corresponding grapcon generators) that compute suitable spacing objects for
typical applications. The simplest spacing is \codefun{spacing\_equal}
that uses the same space between all tiles:

<<space1>>=
mosaic(art, spacing = spacing_equal(unit(2, "lines")))
@

\noindent \codefun{spacing\_equal} is the default grapcon generator for
two-dimensional tables. Slightly more flexible is \codefun{spacing\_dimequal} that
allows an individual setting for each dimension:

<<space2>>=
mosaic(art, spacing = spacing_dimequal(unit(1:2, "lines")))
@

\noindent The default for multi-way contingency tables is
\codefun{spacing\_increase} that uses increasing spaces for the dimensions. The user can specify a start value and the increase factor:

<<space3>>=
mosaic(art, spacing = spacing_increase(start = unit(0.5, "lines"), rate = 1.5))
@

\noindent For the arthritis example above, we could as well have used
\codefun{spacing\_highlighting} that is similar to \codefun{spacing\_increase} but
sets the spacing in the last splitting dimension to 0:

<<spine4>>=
mosaic(art, spacing = spacing_highlighting, gp = my_colors)
@

\noindent Finally, \codefun{spacing\_conditional} can be used for
visualizing conditional independence: it combines
\codefun{spacing\_equal} (for the conditioned dimensions) and 
\codefun{spacing\_increase} (for the conditioning dimensions). As an example, 
consider Figure \ref{fig:presex}: the spacing clearly allows to better 
distinguish the conditioning variables 
(`Gender' and `MaritalStatus') from the conditioned 
variables (`PremaritalSex' and `ExtramaritalSex'). This
spacing is the default when conditional variables are specified for a
strucplot display (see Section \ref{sec:strucplot}).

\section{Conclusion}
\label{sec:conclusion}

In this paper, we describe the `strucplot' framework for the visualization of
multi-way contingency tables. Strucplot displays include popular 
basic plots such as mosaic, association, and sieve plots, integrated 
in a unified approach: all can be seen as graphical visualizations of
hierarchical conditional flat tables. 
Additionally, these core strucplot displays can be combined into more complex, 
specialized plots, such as pairs and trellis-like displays for visualizing
conditional independence. Residual-based shadings permit the
visualization of log-linear models and the results of independence tests.
The frameworks' modular design allows
flexible customization of the plots' graphical appearance, including 
shading, labeling, spacing, and legend, by means of graphical
appearance control (`grapcon') functions. These `graphical
hyperparameters' are customized and created by
generating functions. Our work includes a set of predefined grapcon generators for
typical analysis tasks, and user-level extensions can easily be added.
The framework is provided by
the \pkg{vcd} package freely available from the
Comprehensive \proglang{R} Archive Network (CRAN).

\bibliography{vcd}

\end{document}

### 2005-10-21 trunk/vcd/paper/JSS.Rnw
further fixes.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@479 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\usepackage{thumbpdf}
 
@@
 
-%\VignetteIndexEntry{The Strucplot Framework---Visualizing Multi-way Contingency Tables}
+%\VignetteIndexEntry{The Strucplot Framework: Visualizing Multi-way Contingency Tables With package vcd}
 %\VignetteDepends{vcd}
-%\VignetteKeywords{graphics, categorical data}
+%\VignetteKeywords{contingency tables, mosaic plots, association plots, sieve plots, R, categorical data, independence, conditional independence, HSV, HCL, residual-based shading, grid}
 %\VignettePackage{vcd}
@@
   Wirtschaftsuniversit\"at Wien\\
+  Augasse 2--6
   1090 Wien, Austria\\
@@
 @
-\caption{Mosaic plot for the arthritis data.}
+\caption{Mosaic plot for the \data{Arthritis} data.}
 \label{fig:arthritis}
@@
 @
-\caption{Conditional association plot for the UCB admissions data.}
+\caption{Conditional association plot for the \data{UCBAdmissions} data.}
 \label{fig:ucbadmissions}
@@
 @
-\caption{Mosaic plot for the pre- and extramarital sex data.}
+\caption{Mosaic plot for the \data{PreSex} data.}
 \label{fig:presex}
@@
 @
-\caption{Double-decker plot for the survival on the Titanic data.}
+\caption{Double-decker plot for the \data{Titanic} data.}
 \label{fig:titanic}
@@
 
-\section[The Strucplot Framework]{The Strucplot Framework}
+\section[The strucplot framework]{The strucplot framework}
 \label{sec:strucplot}
 
-The strucplot framework in the {\sf R} package \pkg{vcd}, used for visualizing
+The strucplot framework in the \proglang{R} package \pkg{vcd}, used for visualizing
 multi-way contingency tables, integrates techniques such as
@@
 
-\subsection{Mosaic, Association, and Sieve Plots}
+\subsection{Mosaic, association, and sieve plots}
 
@@
 \codefun{structable} function (quite similar to \codefun{ftable} in
-base {\sf R}, but allowing the specification of split directions):
+base \proglang{R}, but allowing the specification of split directions):
 
@@
 cell. Note that in contrast to, e.g.,  \codefun{mosaicplot} in base
-{\sf R}, the default split direction and level ordering in all strucplot
+\proglang{R}, the default split direction and level ordering in all strucplot
 displays correspond to the textual representation.
@@
 plots \citep{vcd:Cohen:1980}. In contrast to \codefun{assocplot} in
-base {\sf R}, \pkg{vcd}'s \codefun{assoc}
+base \proglang{R}, \pkg{vcd}'s \codefun{assoc}
 function scales to more than two variables (see Figure \ref{fig:residuals}):
@@
 
-\noindent For compatibility with \codefun{mosaicplot} in base {\sf R},
+\noindent For compatibility with \codefun{mosaicplot} in base \proglang{R},
 the \codefun{mosaic} function also allows the use of a \code{"direction"} 
@@
 
-\section{Residual-based Shadings}
+\section{Residual-based shadings}
 \label{sec:shading}
 
-Unlike other graphics functions in base {\sf R}, the strucplot framework 
+Unlike other graphics functions in base \proglang{R}, the strucplot framework 
 allows almost full control over the graphical parameters of all plot elements. In
@@
 preprocesses them, and returns the actual shading function. This returned
-function will have access to the parameters since in {\sf R}, nested
+function will have access to the parameters since in \proglang{R}, nested
 functions are lexically scoped. Thus, the grapcon generator
@@
 observed values. This feature is not provided
-in the original {\sf SAS} implementation of the Friendly mosaic plots.)
+in the original \proglang{SAS} implementation of the Friendly mosaic plots.)
 
@@
 
-\section[Flexible Labeling]{Flexible Labeling}
+\section[Flexible labeling]{Flexible labeling}
 \label{sec:labeling}
@@
 \codefun{mosaicplot} and \codefun{assocplot} in
-base {\sf R} is the labeling in the strucplot framework which 
+base \proglang{R} is the labeling in the strucplot framework which 
 offers many more features and flexibility. 
@@
 plots labels in the borders similar to the \codefun{mosaicplot}
-function in base {\sf R}, but is much more flexible: it is not limited
+function in base \proglang{R}, but is much more flexible: it is not limited
 to 4 dimensions, and the positioning and graphical parameters 
@@
 abbreviated to (\code{TRUE} is interpreted as 1, obviously). Abbreviation
-is performed using the \codefun{abbreviate} function in base {\sf R}. Another
+is performed using the \codefun{abbreviate} function in base \proglang{R}. Another
 possibility is to rotate the levels (see Figure~\ref{fig:rotating}):
@@
 
-\section{Customizable Spacing}
+\section{Customizable spacing}
 \label{sec:spacing}
@@
 space between the tiles corresponding to the levels. As an example,
-consider the introductory example of the \data{arthritis} data 
+consider the introductory example of the \data{Arthritis} data 
 (Figure \ref{fig:arthritis}). Since we are interested in the effect of
@@
 @ 
-\caption{Spine plot for the \data{arthritis} data using the strucplot framework.}
+\caption{Spine plot for the \data{Arthritis} data using the strucplot framework.}
 \label{fig:artspine}
@@
 
+\begin{appendix}
+\section{Data sets}
+
+The data set names in the paper are those from the \proglang{R}
+system. In the following, we give a short description of each data set.
+
+\begin{description}
+\item[Titanic] Information on the fate of passengers on
+  the fatal maiden voyage of the ocean liner `Titanic', summarized
+  according to economic status (class), gender (`Sex'), age and
+  survival. The data were originally collected by the British Board of Trade
+  in their investigation of the sinking. Source: FIXME. Available in base \proglang{R}.
+\item[PreSex]  Data from Thornes \& Collard (1979), reported in Gilbert (1981), on
+  pre- and extra-marital sex and divorce. Source: FIXME. Available
+  in the \pkg{vcd} package.
+\item[Arthritis] Data from \cite{vcd:Koch+Edwards:1988} from a double-blind clinical trial
+  investigating a new treatment for rheumatoid arthritis. Available
+  in the \pkg{vcd} package.
+\item[UCBAdmissions] Aggregate data on applicants to graduate school at Berkeley for
+  the six largest departments in 1973 classified by admission and
+  gender. Source: \cite{vcd:Bickel+Hammel+O'Connell:1975}. Available
+  in base \proglang{R}.
+\item[HairEyeColor] Distribution of hair and eye color and gender in 592 statistics
+  students. The gender information is artificial. Source: \cite{vcd:Snee:1974}.
+  Available in base \proglang{R}.
+\item[Bundesliga] Results from the first German soccer league in the
+  years 1995/6 \citep{vcd:Knorr-Held:1999} and 2001/2 (Achim
+  Zeileis). Available in the \pkg{vcd} package.
+\end{description}
+
+\end{appendix}
+
 \end{document}

### 2005-10-22 trunk/vcd/paper/JSS.Rnw
fixes
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@480 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
refs in datasets fixed.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@481 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \begin{description}
+\item[Arthritis] Data from \cite{vcd:Koch+Edwards:1988} from a double-blind clinical trial
+  investigating a new treatment for rheumatoid arthritis. Available
+  in the \pkg{vcd} package.
+\item[Bundesliga] Results from the first German soccer league in the
+  years 1995/6 \citep{vcd:Knorr-Held:1999} and 2001/2 (Achim
+  Zeileis). Available in the \pkg{vcd} package.
+\item[HairEyeColor] Distribution of hair and eye color and gender in 592 statistics
+  students. The gender information is artificial. Source: \cite{vcd:Snee:1974}.
+  Available in base \proglang{R}.
+\item[PreSex]  Data from \cite{vcd:thornes+collard:1979} reported in 
+  \cite{vcd:gilbert:1981}, on pre- and extra-marital sex and divorce. Available
+  in the \pkg{vcd} package.
 \item[Titanic] Information on the fate of passengers on
@@
   according to economic status (class), gender (`Sex'), age and
-  survival. The data were originally collected by the British Board of Trade
-  in their investigation of the sinking. Source: FIXME. Available in base \proglang{R}.
-\item[PreSex]  Data from Thornes \& Collard (1979), reported in Gilbert (1981), on
-  pre- and extra-marital sex and divorce. Source: FIXME. Available
-  in the \pkg{vcd} package.
-\item[Arthritis] Data from \cite{vcd:Koch+Edwards:1988} from a double-blind clinical trial
-  investigating a new treatment for rheumatoid arthritis. Available
-  in the \pkg{vcd} package.
+  survival.  Source: \cite{vcd:dawson:1995} (Data 
+  originally collected by the British Board of Trade
+  in their investigation of the sinking). Available in base \proglang{R}.
 \item[UCBAdmissions] Aggregate data on applicants to graduate school at Berkeley for
@@
   in base \proglang{R}.
-\item[HairEyeColor] Distribution of hair and eye color and gender in 592 statistics
-  students. The gender information is artificial. Source: \cite{vcd:Snee:1974}.
-  Available in base \proglang{R}.
-\item[Bundesliga] Results from the first German soccer league in the
-  years 1995/6 \citep{vcd:Knorr-Held:1999} and 2001/2 (Achim
-  Zeileis). Available in the \pkg{vcd} package.
 \end{description}

### 2005-10-25 trunk/vcd/paper/JSS.Rnw
revision including AZ fixes
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@485 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \title{The Strucplot Framework:\\ Visualizing Multi-way Contingency
-  Tables With \pkg{vcd}}
+  Tables with \pkg{vcd}}
 
@@
 flexible customization of the plots' graphical appearance, including 
-shading, labeling, spacing, and legend, by means of graphical
-appearance control (`grapcon') functions. The framework is provided by
-the \pkg{vcd} package freely available from the
-Comprehensive \proglang{R} Archive Network (CRAN).
+shading, labeling, spacing, and legend, by means of \underline{gr}aphical
+\underline{ap}pearance \underline{con}trol (`grapcon') functions. 
+The framework is provided by the \proglang{R} package \pkg{vcd}.
 }
 
-\Keywords{contingency tables, mosaic plots, association plots, sieve plots, \proglang{R},
+\Keywords{contingency tables, mosaic plots, association plots, sieve plots, 
 categorical data, independence, conditional independence, HSV, HCL,
-residual-based shading, \pkg{grid}}
+residual-based shading, \pkg{grid}, \proglang{R}}
 \Plainkeywords{contingency tables, mosaic plots, association plots,
-  sieve plots, R, categorical data, independence, 
-  conditional independence, HSV, HCL, residual-based shading, grid}
+  sieve plots, categorical data, independence, 
+  conditional independence, HSV, HCL, residual-based shading, grid, R}
 
@@
   Achim Zeileis\\
-  Department of Statistics \& Mathematics\\
+  Department of Statistics and Mathematics\\
   E-mail: \email{Achim.Zeileis@wu-wien.ac.at}\\
@@
   Kurt Hornik\\
-  Department of Statistics \& Mathematics\\
+  Department of Statistics and Mathematics\\
   E-mail: \email{Kurt.Hornik@wu-wien.ac.at}\\
@@
 typically look for (conditional) independence structures. Whether the
-task is purely exploratory or model based, techniques such as mosaic and
+task is purely exploratory or model-based, techniques such as mosaic and
 association plots offer good support for visualization. Both have been
@@
 contingency tables. A \emph{mosaic plot} \citep{vcd:Hartigan+Kleiner:1984} is basically 
-an area-proportional visualization of (typically observed) frequencies, composed
+an area-proportional visualization of (typically, observed) frequencies, composed
 of tiles (corresponding to the cells) created by recursive
@@
 \item Varying the shape of bar plots and mosaic displays to yield, e.g.,
-  double-decker plots \citep{vcd:hofmann:2001}, spine plots, or spinograms.
+  double-decker plots \citep{vcd:hofmann:2001}, spine plots, or
+  spinograms \citep{vcd:hofmann+theus}.
 \item Using residual-based shadings to visualize log-linear models
   \citep{vcd:Friendly:1994,vcd:Friendly:2000} and significance of
-  statistical tests \citep{vcd:Meyer+Zeileis+Hornik:2003}.
+  statistical tests \citep{vcd:Meyer+Zeileis+Hornik:2003,vcd:Zeileis+Meyer+Hornik:2005}.
 \item Using pairs plots and trellis-like layouts for marginal, conditional and
@@
 \item Adding direct user interaction, allowing quick exploration and
-  modification of the visualized models \citep{vcd:Unwin+Hawkins+Hofmann:1996,vcd:Theus:2003}.
+  modification of the visualized models 
+  \citep{vcd:Unwin+Hawkins+Hofmann:1996,vcd:Theus:2003}.
 \item Providing a modular and flexible implementation to easily allow
@@
 \noindent Current implementations of mosaic displays can be found,
-e.g., for \proglang{SAS}, \proglang{ViSta}, \proglang{MANET}, \proglang{Mondrian},
-\proglang{R}, and \proglang{S-PLUS}. Table \ref{tab:compare} gives an overview of the
-available functionality in these systems.
+e.g., for \proglang{SAS} \citep{vcd:SAS:2005}, \proglang{ViSta} \citep{vcd:young:1996}, 
+\proglang{MANET} \citep{vcd:Unwin+Hawkins+Hofmann:1996},
+\proglang{Mondrian} \citep{vcd:Theus:2003},
+\proglang{R} \citep{vcd:R:2005}, and \proglang{S-PLUS} \citep{vcd:SPLUS:2005}. 
+Table \ref{tab:compare} gives an overview of the available functionality in these systems.
 
@@
     \hline
-                             & \proglang{SAS} & \proglang{S-PLUS} & \proglang{R} & \proglang{ViSta} & \proglang{MANET}/\proglang{Mondrian}\\\hline
+                             & \proglang{SAS} & \proglang{S-PLUS} & \proglang{R} & \pkg{ViSta} & \pkg{MANET}/\pkg{Mondrian}\\\hline
     Basic functionality      &  $\times$  &   $\times$    & $\times$ &   $\times$   & $\times$\\
@@
   \end{tabular}
-  \caption{Comparison of current software environments}
+  \caption{Comparison of current software environments.}
   \label{tab:compare}
@@
 Figures \ref{fig:arthritis} to \ref{fig:titanic} illustrate some of these
-extensions. Figure \ref{fig:arthritis} shows the results from a double-blind clinical trial
+extensions (the code is shown in Appendix \ref{sec:code}). 
+Figure \ref{fig:arthritis} shows the results from a double-blind clinical trial
 investigating a new treatment for rheumatoid arthritis, using an
@@
 <<UCBAdmissions,fig=TRUE,echo=FALSE,height=6,width=10,results=hide>>=
-cotabplot(UCBAdmissions, panel = cotab_assoc)
+cotabplot(UCBAdmissions, panel = cotab_assoc, shade = TRUE, legend = FALSE)
 @
@@
 for the \proglang{R} system and environment for statistical computing
-and graphics, available from the Comprehensive \proglang{R} Archive Network (CRAN).
+and graphics, available from the Comprehensive \proglang{R} Archive Network (\url{http://cran.r-project.org/}).
 The framework integrates displays such as mosaic, association, and
-sieve plots by their common property of being 
+sieve plots by their unifying property of being 
 flat representations of contingency tables.
@@
 
-The reminder of the paper is organized as follows. In
+The remainder of the paper is organized as follows. In
 Section \ref{sec:strucplot}, we give an overview of the strucplot
@@
 In Section \ref{sec:shading}, we demonstrate how
-residual-based shadings support the visualization of log-linear models and
+(residual-based) shadings support the visualization of log-linear models and
 the results of independence tests. Also, we explain step-by-step how
@@
 Sections \ref{sec:labeling} and \ref{sec:spacing} discuss in detail the labeling
-and spacing features, respectively. Section \ref{sec:conclusion}
-finally concludes the work.
+and spacing features, respectively, before Section \ref{sec:conclusion}
+concludes the work.
 
@@
 flexible parameterization and extensibility (Figure \ref{fig:struc}
-only shows the generators).
-The first part of the generator names (\code{\emph{group\_foo}()})
-reflects the group they belong to (strucplot core, labeling,
+only shows the generators). The generator names
+follow the naming convention \code{\emph{group\_foo}()},  
+where (\code{\emph{group}}) reflects the group the
+generators belong to (strucplot core, labeling,
 legend, shading, or spacing). The workhorse functions (created by 
@@
 \code{labeling\_\emph{foo}()}, and \code{legend\_\emph{foo}()}) 
-directly produce graphical output (``add ink to the canvas''), whereas
+directly produce graphical output (i.e., ``add ink to the canvas''), whereas
 the parameter functions
@@
 interface \codefun{strucplot} through sensible parameter defaults
-and support for model formulas. Finally, on the fourth
+and support for model formulae. Finally, on the fourth
 level, there are `related' \pkg{vcd} functions (such as \codefun{cotabplot}
@@
 data containing two polytomous variables (hair and eye color), 
-as well as one (artificial) dichotomous variable (sex, i.e., gender). The
+as well as one (artificial) dichotomous variable (gender). The
 `flattened' contingency table can be obtained using the
@@
 
-Let us first visualize the contingency table by means of a mosaic plot
-\citep{vcd:Hartigan+Kleiner:1984} which is basically 
-an area-proportional visualization of (typically, observed) frequencies, composed
-of tiles (corresponding to the cells) created by recursive
-vertical and horizontal splits of a square. Thus, the area of each tile
-is proportional to the corresponding cell entry \emph{given} the
-dimensions of previous splits. Figure \ref{fig:observed} depicts 
-the effect of
+Let us first visualize the contingency table by means of a mosaic plot.
+% \citep{vcd:Hartigan+Kleiner:1984} which is basically 
+% an area-proportional visualization of (typically, observed) frequencies, composed
+% of tiles (corresponding to the cells) created by recursive
+% vertical and horizontal splits of a square. Thus, the area of each tile
+% is proportional to the corresponding cell entry \emph{given} the
+% dimensions of previous splits. 
+Figure \ref{fig:observed} depicts the effect of
 
@@
 \begin{center}
-<<Observedfig,fig=TRUE,echo=FALSE>>=
+<<Observedfig,fig=TRUE,echo=FALSE,height=6>>=
 <<Observed>>
@@
 \begin{center}
-<<Expectedfig,fig=TRUE,echo=FALSE>>=
+<<Expectedfig,fig=TRUE,echo=FALSE,height=6>>=
 <<Expected>>
@@
 
-\noindent For compatibility with \codefun{mosaicplot} in base \proglang{R},
+\noindent (Note that \code{hec}, a \class{structable} object, already includes a
+splitting information which simply gets overloaded in this example.) 
+For compatibility with \codefun{mosaicplot} in base \proglang{R},
 the \codefun{mosaic} function also allows the use of a \code{"direction"} 
@@
 of the conditioning variable(s)) is simply done by indexing.
-However, subsetting \code{"structable"} objects is more restrictive 
+However, subsetting \class{structable} objects is more restrictive 
 because of their inherent conditional structure. Since the variables
@@
 @
-\caption{Distribution of hair and eye color, given gender.}
+\caption{Two mosaic displays put side-by-side, visualizing the 
+distribution of hair and eye color, given gender. The marginal
+distribution of `Sex' cannot be seen.}
 \label{fig:parttable}
@@
 @
-\caption{Mosaic plot for conditional independence structures.}
+\caption{Mosaic plot again visualizing the 
+distribution of hair and eye color, given gender, this time using a
+single mosaic plot. In contrast to Figure \ref{fig:parttable}, this
+plot also visualizes the marginal distribution of `Sex'.}
 \label{fig:conditioning}
@@
 @
-\caption{Conditional table plot for the \data{HairEyeColor} data.}
+\caption{Conditional table plot for the \data{HairEyeColor} data,
+  again visualizing the distribution of hair and eye color, given
+  gender, using separate mosaic displays like the `manual' plot 
+  in Figure \ref{fig:parttable}.}
 \label{fig:cotabplot}
@@
 residuals, resulting from deviations of observed from expected
-values under a given log-linear model. 
+frequencies under a given log-linear model. 
 The idea is to visualize at least sign and absolute size of the residuals, but some
@@
 \begin{center}
-<<continuousfig,fig=TRUE,echo=FALSE>>=
+<<continuousfig,fig=TRUE,echo=FALSE,height=7>>=
 <<continuous2>>
@@
 \begin{center}
-<<bundesligafig,fig=TRUE,echo=FALSE>>=
+<<bundesligafig,fig=TRUE,echo=FALSE,height=7>>=
 <<bundesliga>>
@@
 \begin{center}
-<<friendlyfig,fig=TRUE,echo=FALSE>>=
+<<friendlyfig,fig=TRUE,echo=FALSE,height=7>>=
 <<friendly>>
@@
 \begin{center}
-<<hclshading1fig,fig=TRUE,echo=FALSE>>=
+<<hclshading1fig,fig=TRUE,echo=FALSE,height=6>>=
 <<hclshading1>>
@@
 \begin{center}
-<<hclshading2fig,fig=TRUE,echo=FALSE>>=
+<<hclshading2fig,fig=TRUE,echo=FALSE,height=6>>=
 <<hclshading2>>
@@
 \begin{center}
-<<rotatefig,fig=TRUE,echo=FALSE>>=
+<<rotatefig,fig=TRUE,echo=FALSE,height=6>>=
 <<rotate>>
@@
 \begin{center}
-<<repeatfig,fig=TRUE,echo=FALSE>>=
+<<repeatfig,fig=TRUE,echo=FALSE,height=6>>=
 <<repeat>>
@@
 \begin{center}
-<<marginsfig,fig=TRUE,echo=FALSE>>=
+<<marginsfig,fig=TRUE,echo=FALSE,height=6>>=
 <<margins>>
@@
 \begin{center}
-<<boxesfig,fig=TRUE,echo=FALSE>>=
+<<boxesfig,fig=TRUE,echo=FALSE,height=6>>=
 <<boxes>>
@@
 certainly appropriate to visualize the three levels of `Improved' in
-the two `Treatment' strata. However, following a slightly different approach,
-we might as well want to \emph{highlight} patients with `Marked' improvement in both
-groups, that is, produce a spine plot. To obtain such a display within 
+the two `Treatment' strata. Another conceptual approach is to use
+spine plots with highlighting \citep{vcd:hummel:1996}. A spine plot is
+a variation of a bar plot where the heights of the bars are held
+constant, whereas the widths are used to represent the number of cases
+in each category. This is equivalent to a mosaic plot for a one-way table.
+If a second (indicator) variable is highlighted in a spine plot, we
+obtain a display equivalent to a simple mosaic display for a two-way
+table, except that no space between the levels of the highlighted
+variable is used.
+In the \data{Arthritis} example, we will highlight patients with `Marked' 
+improvement in both groups. To obtain such a display within 
 the strucplot framework\footnote{The \pkg{vcd} package also provides 
-the \codefun{spine} function for spine plots and spinograms.}, it
+a specialized \codefun{spine} function for spine plots and spinograms.}, it
 suffices to set the space between the `Improved' tiles to 0 (see
@@
 (my_spacing <- list(unit(0.5, "lines"), unit(c(0, 0), "lines")))
-my_colors <- gpar(fill = c("lightgray", "darkgray", "black"))
+my_colors <- gpar(fill = c("lightgray", "lightgray", "black"))
 mosaic(art, spacing = my_spacing, gp = my_colors)
@@
 
-\noindent The strucplot framework also provides a set of spacing grapcons 
-(and corresponding grapcon generators) that compute suitable spacing objects for
+\noindent The strucplot framework also provides a set of spacing grapcon
+generators that compute suitable spacing objects for
 typical applications. The simplest spacing is \codefun{spacing\_equal}
@@
 basic plots such as mosaic, association, and sieve plots, integrated 
-in a unified approach: all can be seen as graphical visualizations of
+in a unified framework: all can be seen as graphical visualizations of
 hierarchical conditional flat tables. 
@@
 typical analysis tasks, and user-level extensions can easily be added.
-The framework is provided by
-the \pkg{vcd} package freely available from the
-Comprehensive \proglang{R} Archive Network (CRAN).
 
@@
 \begin{appendix}
+
 \section{Data sets}
+\label{sex:data}
 
@@
 \begin{description}
-\item[Arthritis] Data from \cite{vcd:Koch+Edwards:1988} from a double-blind clinical trial
-  investigating a new treatment for rheumatoid arthritis. Available
-  in the \pkg{vcd} package.
-\item[Bundesliga] Results from the first German soccer league in the
+\item[\texttt{Arthritis}] Data from \cite{vcd:Koch+Edwards:1988} from a 
+  double-blind clinical trial
+  investigating a new treatment for rheumatoid arthritis. Package:
+  \pkg{vcd}.
+\item[\texttt{Bundesliga}] Results from the first German soccer league in the
   years 1995/6 \citep{vcd:Knorr-Held:1999} and 2001/2 (Achim
-  Zeileis). Available in the \pkg{vcd} package.
-\item[HairEyeColor] Distribution of hair and eye color and gender in 592 statistics
+  Zeileis). Package: \pkg{vcd}.
+\item[\texttt{HairEyeColor}] Distribution of hair and eye color and 
+  gender in 592 statistics
   students. The gender information is artificial. Source: \cite{vcd:Snee:1974}.
-  Available in base \proglang{R}.
-\item[PreSex]  Data from \cite{vcd:thornes+collard:1979} reported in 
-  \cite{vcd:gilbert:1981}, on pre- and extra-marital sex and divorce. Available
-  in the \pkg{vcd} package.
-\item[Titanic] Information on the fate of passengers on
+  Package: \pkg{datasets} (included in base \proglang{R}).
+\item[\texttt{PreSex}]  Data from \cite{vcd:thornes+collard:1979} reported in 
+  \cite{vcd:gilbert:1981}, on pre- and extra-marital sex and divorce. Package:
+  \pkg{vcd}.
+\item[\texttt{Titanic}] Information on the fate of passengers on
   the fatal maiden voyage of the ocean liner `Titanic', summarized
@@
   originally collected by the British Board of Trade
-  in their investigation of the sinking). Available in base \proglang{R}.
-\item[UCBAdmissions] Aggregate data on applicants to graduate school at Berkeley for
+  in their investigation of the sinking). 
+  Package: \pkg{datasets} (included in base \proglang{R}).
+\item[\texttt{UCBAdmissions}] 
+  Aggregate data on applicants to graduate school at Berkeley for
   the six largest departments in 1973 classified by admission and
-  gender. Source: \cite{vcd:Bickel+Hammel+O'Connell:1975}. Available
-  in base \proglang{R}.
+  gender. Source: \cite{vcd:Bickel+Hammel+O'Connell:1975}. 
+  Package: \pkg{datasets} (included in base \proglang{R}).
 \end{description}
 
+\section{Code used in the introduction}
+\label{sec:code}
+
+\begin{itemize}
+
+\item Mosaic plot for the \data{Arthiritis} data in Figure \ref{fig:arthritis}:
+\begin{Sinput}
+mosaic(art, gp = shading_max, split_vertical = TRUE)
+\end{Sinput}
+
+\item Conditional association plot for the \data{UCBAdmissions} 
+data in Figure \ref{fig:ucbadmissions}:
+\begin{Sinput}
+cotabplot(UCBAdmissions, panel = cotab_assoc, shade = TRUE, legend = FALSE)
+\end{Sinput}
+
+\item Mosaic plot for the \data{PreSex} data in Figure \ref{fig:presex}:
+\begin{Sinput}
+mosaic(PreSex, condvars = c(1, 4), shade = TRUE)
+\end{Sinput}
+
+\item Double-decker plot for the \data{Titanic} data in Figure \ref{fig:titanic}:
+\begin{Sinput}
+doubledecker(Survived ~ ., data = Titanic, labeling_args = 
+             list(set_varnames = c(Sex = "Gender")))
+\end{Sinput}
+\end{itemize}
+
 \end{appendix}

### 2005-10-25 trunk/vcd/paper/JSS.Rnw
small typo
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@486 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
KH changes.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@489 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 conditional independence, GLMs, and the results of
-independence tests. The frameworks' modular design allows
+independence tests. The framework's modular design allows
 flexible customization of the plots' graphical appearance, including 
@@
 task is purely exploratory or model-based, techniques such as mosaic and
-association plots offer good support for visualization. Both have been
-extended over the last two decades, and implementations exist in many
-statistical environments. 
-Both graphical methods visualize aspects of (possibly higher-dimensional)
-contingency tables. A \emph{mosaic plot} \citep{vcd:Hartigan+Kleiner:1984} is basically 
-an area-proportional visualization of (typically, observed) frequencies, composed
-of tiles (corresponding to the cells) created by recursive
+association plots offer good support for visualization.  Both visualize
+aspects of (possibly higher-dimensional) contingency tables, with
+several extensions introduced over the last two decades, and
+implementations available in many statistical environments.  A
+\emph{mosaic plot} \citep{vcd:Hartigan+Kleiner:1984} is basically an
+area-proportional visualization of (typically, observed) frequencies,
+composed of tiles (corresponding to the cells) created by recursive
 vertical and horizontal splits of a square. Thus, the area of each tile
 is proportional to the corresponding cell entry \emph{given} the
-dimensions of previous splits. An \emph{association plot} \citep{vcd:Cohen:1980}
-visualizes the standardized deviations of observed frequencies from
-those expected under a certain independence hypothesis.
-Each cell is represented by a rectangle that has
-(signed) height proportional to the residual and width proportional
-to the square root of the expected counts, 
-so that the area of the box is proportional
+dimensions of previous splits. An \emph{association plot}
+\citep{vcd:Cohen:1980} visualizes the standardized deviations of
+observed frequencies from those expected under a certain independence
+hypothesis.  Each cell is represented by a rectangle that has (signed)
+height proportional to the residual and width proportional to the square
+root of the expected counts, so that the area of the box is proportional
 to the difference in observed and expected frequencies.
 
-Over the years, extensions to these techniques were mainly focused on
-five aspects:
+Extensions to these techniques have mainly focused on the following
+aspects.
 
@@
 Figures \ref{fig:arthritis} to \ref{fig:titanic} illustrate some of these
-extensions (the code is shown in Appendix \ref{sec:code}). 
+extensions (the code is given in Appendix \ref{sec:code}). 
 Figure \ref{fig:arthritis} shows the results from a double-blind clinical trial
@@
 
-This paper describes the strucplot framework provided by the \pkg{vcd} package
-for the \proglang{R} system and environment for statistical computing
-and graphics, available from the Comprehensive \proglang{R} Archive Network (\url{http://cran.r-project.org/}).
-The framework integrates displays such as mosaic, association, and
-sieve plots by their unifying property of being 
-flat representations of contingency tables.
-These basic plots, as well as specialized displays for conditional
-independence, can be used both for exploratory visualization and
-model-based analysis. Exploratory techniques include specialized displays for
-the bivariate case, as well as pairs and trellis plot-like displays for
-higher-dimensional tables. Model-based tools include 
-methods suitable for the visualization of 
-conditional independence tests (including permutation tests), as well
-as for the visualization of particular GLMs (logistic regression,
-log-linear models). Additionally, two of the frameworks' further strengths are its
-flexibility and extensibility: graphical appearance aspects such as shading,
-labeling, and spacing are modularized by means of `graphical
-appearance functions', allowing fine-granular customization and
-user-level extensions.
+This paper describes the strucplot framework provided by the \pkg{vcd}
+package for the \proglang{R} environment for statistical computing and
+graphics, available from the Comprehensive \proglang{R} Archive Network
+(\url{http://cran.r-project.org/}).  The framework integrates displays
+such as mosaic, association, and sieve plots by their unifying property
+of being flat representations of contingency tables.  These basic plots,
+as well as specialized displays for conditional independence, can be
+used both for exploratory visualization and model-based analysis.
+Exploratory techniques include specialized displays for the bivariate
+case, as well as pairs and trellis plot-like displays for
+higher-dimensional tables. Model-based tools include methods suitable
+for the visualization of conditional independence tests (including
+permutation tests), as well as for the visualization of particular GLMs
+(logistic regression, log-linear models). Additionally, two of the
+framework's further strengths are its flexibility and extensibility:
+graphical appearance aspects such as shading, labeling, and spacing are
+modularized by means of `graphical appearance functions', allowing
+fine-granular customization and user-level extensions.
 
@@
 list of \class{unit} vectors, one for each dimension, specifying the
-space between the tiles corresponding to the levels. As an example,
-consider the introductory example of the \data{Arthritis} data 
+space between the tiles corresponding to the levels. 
+Consider again the introductory example of the \data{Arthritis} data 
 (Figure \ref{fig:arthritis}). Since we are interested in the effect of
@@
 visualization of log-linear models and the results of independence tests.
-The frameworks' modular design allows
+The framework's modular design allows
 flexible customization of the plots' graphical appearance, including 

### 2005-10-25 trunk/vcd/paper/JSS.Rnw
final version
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@490 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 parameter functions that directly or indirectly influence the final
-appearance of the plot. These are examples of `graphical appearance control'
+appearance of the plot (see Table \ref{tab:grapcons} for an overview). 
+These are examples of `\underline{gr}aphical \underline{ap}pearance \underline{con}trol'
 (`grapcon') functions. They are created by generating functions
@@
 framework into more complex displays (e.g., by means of panel functions).
+
+\begin{table}
+  \begin{tabular}{|l|l|l|}
+    \hline
+    \textbf{Group} & \textbf{Grapcon generator} & \textbf{Description}\\\hline
+    strucplot & \codefun{struc\_mosaic} & core function for mosaic plots\\
+    core          & \codefun{struc\_assoc} & core function for association plots\\
+              & \codefun{struc\_sieve} & core function for sieve plots\\\hline\hline
+     labeling & \codefun{labeling\_border} & border labels\\
+              & \codefun{labeling\_cells} & cell labels\\
+              & \codefun{labeling\_conditional} & border labels
+              for conditioning variables\\
+              && and cell labels for conditioned variables.\\
+              & \codefun{labeling\_list} & draws a list of labels
+              under the plot\\
+              & \codefun{labeling\_doubledecker} & draws labels for
+              doubledecker plot\\
+              & \codefun{labeling\_left} & left-aligned border labels\\
+              & \codefun{labeling\_left2} & left-aligned border
+              labels, all labels on top and left border\\
+              & \codefun{labeling\_lboxed} & left-aligned labels with boxes.\\
+              & \codefun{labeling\_cboxed} & centered labels with
+              boxes, all labels clipped,\\
+              && and on top and left border\\\hline\hline
+     shading  & \codefun{shading\_hsv} & shading based on HSV-colors\\
+              & \codefun{shading\_Friendly} & implements Friendly-like
+              shading (based on HSV-colors)\\
+              & \codefun{shading\_hcl} & shading based on HCL-colors\\
+              & \codefun{shading\_max} & shading visualizing the
+              maximum test statistic\\ && (based on HCL-colors)\\
+              & \codefun{shading\_binary} & visualizes the sign of the
+              residuals\\\hline\hline
+     spacing  & \codefun{spacing\_equal} & equal spacing for all dimensions\\
+              & \codefun{spacing\_dimequal} & equal spacing for each dimension\\
+              & \codefun{spacing\_increase} & increasing spacing\\
+              & \codefun{spacing\_highlighting} & increasing spacing,
+              last dimension set to zero\\
+              & \codefun{spacing\_conditional} & increasing spacing
+              for conditioning variables,\\&& equal spacing for
+              conditioned variables.\\\hline\hline
+     legend   & \codefun{legend\_resbased} & suitable for an
+     arbitrary number of bins\\&& (also for continuous shadings)\\
+              & \codefun{legend\_fixed} & creates a fixed number of
+              bins (similar to \codefun{mosaicplot})\\\hline
+
+  \end{tabular}
+  \caption{Available grapcon generators in the strucplot framework}
+  \label{tab:grapcons}
+\end{table}
 

### 2005-11-02 trunk/vcd/paper/JSS.Rnw
Final touches (table 2 is now sorted). Submission release!
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@491 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2005-10-25 trunk/vcd/paper/JSS.Rnw
@@
 framework into more complex displays (e.g., by means of panel functions).
+
+\begin{table}
+  \begin{tabular}{|l|l|l|}
+    \hline
+    \textbf{Group} & \textbf{Grapcon generator} & \textbf{Description}\\\hline
+    strucplot & \codefun{struc\_assoc} & core function for association plots\\
+    core      & \codefun{struc\_mosaic} & core function for mosaic plots\\
+              & \codefun{struc\_sieve} & core function for sieve plots\\\hline\hline
+     labeling & \codefun{labeling\_border} & border labels\\
+              & \codefun{labeling\_cboxed} & centered labels with
+              boxes, all labels clipped,\\
+              && and on top and left border\\
+              & \codefun{labeling\_cells} & cell labels\\
+              & \codefun{labeling\_conditional} & border labels 
+                                                  for conditioning variables\\
+              && and cell labels for conditioned variables.\\
+              & \codefun{labeling\_doubledecker} & draws labels for
+              doubledecker plot\\
+              & \codefun{labeling\_lboxed} & left-aligned labels with boxes.\\
+              & \codefun{labeling\_left} & left-aligned border labels\\
+              & \codefun{labeling\_left2} & left-aligned border
+              labels, all labels on top and left border\\
+              & \codefun{labeling\_list} & draws a list of labels
+              under the plot\\\hline\hline
+     shading  & \codefun{shading\_binary} & visualizes the sign of the  residuals\\
+              & \codefun{shading\_Friendly} & implements Friendly-like
+              shading (based on HSV-colors)\\
+              & \codefun{shading\_hcl} & shading based on HCL-colors\\
+              & \codefun{shading\_hsv} & shading based on HSV-colors\\
+              & \codefun{shading\_max} & shading visualizing the
+              maximum test statistic\\ && (based on HCL-colors)\\\hline\hline
+     spacing  & \codefun{spacing\_conditional} & increasing spacing
+              for conditioning variables,\\&& equal spacing for
+              conditioned variables.\\
+              & \codefun{spacing\_dimequal} & equal spacing for each dimension\\
+              & \codefun{spacing\_equal} & equal spacing for all dimensions\\
+              & \codefun{spacing\_highlighting} & increasing spacing,
+              last dimension set to zero\\
+              & \codefun{spacing\_increase} & increasing spacing\\\hline\hline
+     legend   & \codefun{legend\_fixed} & creates a fixed number of
+              bins (similar to \codefun{mosaicplot})\\
+              & \codefun{legend\_resbased} & suitable for an
+              arbitrary number of bins\\&& (also for continuous shadings)\\\hline
+
+  \end{tabular}
+  \caption{Available grapcon generators in the strucplot framework}
+  \label{tab:grapcons}
+\end{table}
 

### 2005-11-02 trunk/vcd/paper/statreport22.Rnw
report added
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@493 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2005-11-02 trunk/vcd/paper/JSS.Rnw
@@
-\documentclass[article]{jss}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\documentclass[notitle]{Z}
 
 \usepackage{thumbpdf}
-
-%% almost as usual
-\author{David Meyer\\Wirtschaftsuniversit\"at Wien \And 
-        Achim Zeileis\\Wirtschaftsuniversit\"at Wien \And 
-        Kurt Hornik\\Wirtschaftsuniversit\"at Wien}
+\usepackage{statreport}
+
 \title{The Strucplot Framework:\\ Visualizing Multi-way Contingency
@@
 
-%% for pretty printing and a nice hypersummary also set:
-\Plainauthor{David Meyer, Achim Zeileis, Kurt Hornik} %% comma-separated
-\Shorttitle{The Strucplot Framework} %% a short title (if necessary)
-
-%% an abstract and keywords
+\author{
+  David Meyer \And Achim Zeileis \And Kurt Hornik
+  \\
+  Wirtschaftsuniversit\"at Wien
+  }
+
+%\VignetteIndexEntry{The Strucplot Framework: Visualizing Multi-way Contingency Tables With package vcd}
+%\VignetteDepends{vcd}
+%\VignetteKeywords{contingency tables, mosaic plots, association plots, sieve plots, R, categorical data, independence, conditional independence, HSV, HCL, residual-based shading, grid}
+%\VignettePackage{vcd}
+
+%% need no \usepackage{Sweave.sty}
+
+\SweaveOpts{engine=R,eps=TRUE,height=6,width=7,results=hide,fig=FALSE,echo=TRUE}
+\setkeys{Gin}{width=0.7\textwidth}
+
+\Plainauthor{David Meyer, Achim Zeileis, Kurt Hornik}
+
 \Abstract{
@@
 categorical data, independence, conditional independence, HSV, HCL,
-residual-based shading, \pkg{grid}, \proglang{R}}
-\Plainkeywords{contingency tables, mosaic plots, association plots,
-  sieve plots, categorical data, independence, 
-  conditional independence, HSV, HCL, residual-based shading, grid, R}
-
-%\VignetteIndexEntry{The Strucplot Framework: Visualizing Multi-way Contingency Tables With package vcd}
-%\VignetteDepends{vcd}
-%\VignetteKeywords{contingency tables, mosaic plots, association plots, sieve plots, R, categorical data, independence, conditional independence, HSV, HCL, residual-based shading, grid}
-%\VignettePackage{vcd}
-
-%% publication information
-%% NOTE: This needs to filled out ONLY IF THE PAPER WAS ACCEPTED.
-%% If it was not (yet) accepted, leave them commented.
-%% \Volume{13}
-%% \Issue{9}
-%% \Month{September}
-%% \Year{2004}
-%% \Submitdate{2004-09-29}
-%% \Acceptdate{2004-09-29}
-
-%% The address of (at least) one author should be given
-%% in the following format:
-\Address{
-  David Meyer\\
-  Department of Information Systems and Process Management\\
-  E-mail: \email{David.Meyer@wu-wien.ac.at}\\
-  URL: \url{http://wi.wu-wien.ac.at/~meyer/}\\
-
-  Achim Zeileis\\
-  Department of Statistics and Mathematics\\
-  E-mail: \email{Achim.Zeileis@wu-wien.ac.at}\\
-  URL: \url{http://www.ci.tuwien.ac.at/~zeileis/}\\
-
-  Kurt Hornik\\
-  Department of Statistics and Mathematics\\
-  E-mail: \email{Kurt.Hornik@wu-wien.ac.at}\\
-  URL: \url{http://www.wu-wien.ac.at/cstat/hornik/}\\
-
-  Wirtschaftsuniversit\"at Wien\\
-  Augasse 2--6
-  1090 Wien, Austria\\
-}
-%% It is also possible to add a telephone and fax number
-%% before the e-mail in the following format:
-%% Telephone: +43/1/31336-5053
-%% Fax: +43/1/31336-734
-
-%% for those who use Sweave please include the following line (with % symbols):
-%% need no \usepackage{Sweave.sty}
-
-\SweaveOpts{engine=R,eps=TRUE,height=6,width=7,results=hide,fig=FALSE,echo=TRUE}
-\setkeys{Gin}{width=0.7\textwidth}
+residual-based shading, \texttt{grid}, \textsf{R}}
+
+\sloppy{}
 
@@
 
-%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \begin{document}
 
-%% include your article here, just as usual
-%% Note that you should use the \pkg{}, \proglang{} and \code{} commands.
-
-\section[Introduction]{Introduction}
-%% Note: If there is markup in \(sub)section, then it has to be escape as above.
+\fronttitle{The Strucplot Framework:\\ Visualizing Multi-way Contingency
+  Tables with \pkg{vcd}}
+\frontauthor{David Meyer, Achim Zeileis, Kurt Hornik}
+\frontnumber{22}
+\frontdate{November 2005}
+\frontpage
+
+\maketitle
 

### 2005-11-02 trunk/vcd/paper/statreport22.Rnw
slight modifications in statreport
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@494 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \author{
-  David Meyer \And Achim Zeileis \And Kurt Hornik
+  \hfill David Meyer \hfill Achim Zeileis \hfill Kurt Hornik \hfill
   \\
@@
 
-\fronttitle{The Strucplot Framework:\\ Visualizing Multi-way Contingency
-  Tables with \pkg{vcd}}
+\fronttitle{The Strucplot Framework: Visualizing Multi-way Contingency
+  Tables with vcd}
 \frontauthor{David Meyer, Achim Zeileis, Kurt Hornik}
@@
 
-\begin{figure}[h]
-\begin{center}
-<<cellfig,fig=TRUE,echo=FALSE,width=7,height=7>>=
+\setkeys{Gin}{width=0.7\textwidth}
+\begin{figure}[p]
+\begin{center}
+<<cellfig,fig=TRUE,echo=FALSE>>=
 <<cell>>
@@
 
-\setkeys{Gin}{width=0.8\textwidth}
-\begin{figure}[h]
+\setkeys{Gin}{width=0.7\textwidth}
+\begin{figure}[p]
 \begin{center}
@@
 
+\clearpage
 \begin{appendix}

### 2005-11-02 trunk/vcd/paper/statreport22.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@495 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \documentclass[notitle]{Z}
-
 \usepackage{thumbpdf}
@@
 
+%% need no \usepackage{Sweave}
+\SweaveOpts{engine=R,eps=TRUE,height=6,width=7,results=hide,fig=FALSE,echo=TRUE}
+\setkeys{Gin}{width=0.7\textwidth}
+
 \title{The Strucplot Framework:\\ Visualizing Multi-way Contingency
   Tables with \pkg{vcd}}
-
-\author{
-  \hfill David Meyer \hfill Achim Zeileis \hfill Kurt Hornik \hfill
-  \\
-  Wirtschaftsuniversit\"at Wien
-  }
-
-%\VignetteIndexEntry{The Strucplot Framework: Visualizing Multi-way Contingency Tables With package vcd}
-%\VignetteDepends{vcd}
-%\VignetteKeywords{contingency tables, mosaic plots, association plots, sieve plots, R, categorical data, independence, conditional independence, HSV, HCL, residual-based shading, grid}
-%\VignettePackage{vcd}
-
-%% need no \usepackage{Sweave.sty}
-
-\SweaveOpts{engine=R,eps=TRUE,height=6,width=7,results=hide,fig=FALSE,echo=TRUE}
-\setkeys{Gin}{width=0.7\textwidth}
-
+\author{David Meyer \hspace{2cm} Achim Zeileis \hspace{2cm} Kurt Hornik\\
+  Wirtschaftsuniversit\"at Wien}
 \Plainauthor{David Meyer, Achim Zeileis, Kurt Hornik}
+\Plaintitle{The Strucplot Framework: Visualizing Multi-way Contingency Tables with vcd}
 
@@
 categorical data, independence, conditional independence, HSV, HCL,
-residual-based shading, \texttt{grid}, \textsf{R}}
+residual-based shading, \pkg{grid}, \textsf{R}}
+\Plainkeywords{contingency tables, mosaic plots, association plots, sieve plots, 
+categorical data, independence, conditional independence, HSV, HCL,
+residual-based shading, grid, R}
 

### 2005-11-07 trunk/vcd/paper/hsc.Rnw
Contribution for the handbook of statistical computing added.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@497 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

% RECOMMENDED %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass{svmult}

% choose options for [] as required from the list
% in the Reference Guide, Sect. 2.2

\usepackage{Sweave,hyperref,color,graphicx} 
\usepackage[bottom]{footmisc}% places footnotes at page bottom

\definecolor{Red}{rgb}{0.7,0,0}
\definecolor{Blue}{rgb}{0,0,0.8}
\definecolor{hellgrau}{rgb}{0.55,0.55,0.55}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\SweaveOpts{engine=R,eps=TRUE,height=6,width=7,results=hide,fig=FALSE,echo=TRUE}
\setkeys{Gin}{width=0.7\textwidth}

<<preliminaries,echo=FALSE,results=hide>>=
set.seed(1071)
library(vcd)
library(scatterplot3d)
library(xtable)
data(Arthritis)
data(Titanic)
data(PreSex)
data(UCBAdmissions)
hospital <- as.table(matrix(c(43,16,3,6,11,10,9,18,16), c(3,3), byrow = TRUE, 
dimnames = list("Visit frequency" = c("Regular","Less than monthly","Never"), 
"Length of stay" = c("2-9","10-19","20+"))))
art <- xtabs(~Treatment + Improved, data = Arthritis, subset = Sex == "Female")
@

%\newcommand{\data}[1]{`\texttt{#1}'}
\newcommand{\data}[1]{#1}
\newcommand{\proglang}[1]{\textsf{#1}}
\newcommand{\pkg}[1]{\texttt{#1}}

\begin{document}

\title*{Visualizing Multi-Way Contingency Tables}
% Use \titlerunning{Short Title} for an abbreviated version of
% your contribution title if the original one is too long
\author{David Meyer\inst{1}\and
Achim Zeileis\inst{2}\and
Kurt Hornik\inst{2}}
% Use \authorrunning{Short Title} for an abbreviated version of
% your contribution title if the original one is too long
\institute{Department of Information Systems and Process Management
\and Department of Statistics and Mathematics\\Wirtschaftsuniversität
Wien\\Augasse 2--6, 1090 Vienna, Austria
\texttt{\{David.Meyer,Achim.Zeileis,Kurt.Hornik\}@wu-wien.ac.at}}
\maketitle

\section{Introduction}
\label{sec:intro}

Categorical data analysis is typically based on two- or
higher-dimensional contingency tables, cross-tabulating the
co-occurences of levels of nominal and/or ordinal data. 
In order to explain these, statisticians
typically look for (conditional) independence structures using common
statistical methods such as independence tests and
log-linear models. The analysis quickly 
becomes complex with increasing dimensionality.
One idea of visualization techniques is to
use the human visual system to detect structures in the data that
possibly are not obvious from solely numeric output (e.g., test
statistics). Whether the
task is purely exploratory or model based, techniques such as mosaic, sieve, and
association plots offer good support for visualization. All three have been
extended over the last two decades, and implementations exist in many
statistical environments.

All three graphical methods visualize aspects of (possibly higher-dimensional)
contingency tables. A \emph{mosaic plot} \cite{vcd:Hartigan+Kleiner:1984} is basically 
an area-proportional visualization of (typically observed) frequencies, composed
of tiles (corresponding to the cells) created by recursive
vertical and horizontal splits of a square. Thus, the area of each tile
is proportional to the corresponding cell entry \emph{given} the
dimensions of previous splits. 
\emph{Sieve plot} are similar to mosaic plots, but the area of each
tile is proportional to the \emph{expected} cell entry, and each tile
is filled with a number of rectangles corresponding to the observed value.
An \emph{association plot} \cite{vcd:Cohen:1980}
visualizes the standardized deviations of observed frequencies from
those expected under a certain independence hypothesis.
Each cell is represented by a rectangle that has
(signed) height proportional to the residual and width proportional
to the square root of the expected counts, 
so that the area of the box is proportional
to the difference in observed and expected frequencies.

Over the years, extensions to these techniques were mainly focused on
five aspects:

\begin{enumerate}
\item Varying the shape of bar plots and mosaic displays to yield, e.g.,
  double-decker plots \cite{vcd:hofmann:2001}, spine 
  plots \cite{vcd:riedwyl+schuepbach:1994}, or spinograms.
\item Using residual-based shadings to visualize log-linear models
  \cite{vcd:Friendly:1994,vcd:Friendly:2000} and significance of
  statistical tests \cite{vcd:Meyer+Zeileis+Hornik:2003}.
\item Using pairs plots and trellis-like layouts for marginal, conditional and
  partial views \cite{vcd:Friendly:1999}.
\item Adding direct user interaction, allowing quick exploration and
  modification of the visualized models \cite{vcd:Unwin+Hawkins+Hofmann:1996,vcd:Theus:2003}.
\item Providing a modular and flexible implementation to easily allow
  user extensions \cite{vcd:Meyer+Zeileis+Hornik:2005a}.
\end{enumerate}

\noindent Current implementations of mosaic displays can be found,
e.g., for \proglang{SAS} \cite{vcd:SAS:2005}, \pkg{ViSta} \cite{vcd:young:1996}, 
\pkg{MANET} \cite{vcd:Unwin+Hawkins+Hofmann:1996},
\pkg{Mondrian} \cite{vcd:Theus:2003},
\proglang{R} \cite{vcd:R:2005}, and \proglang{S-PLUS} \cite{vcd:SPLUS:2005}. 
Table \ref{tab:compare} gives an overview of the available functionality in these systems.
The figures in this chapter have all been produced using the
\proglang{R} system, using the extension packages \pkg{vcd} 
\cite{vcd:Meyer+Zeileis+Hornik:2005} and
\pkg{scatterplot3d} \cite{vcd:ligges+maechler:2003}
(Fig.~\ref{fig:threedee} only), all freely available from the
Comprehensive \proglang{R} Archive Network (\url{http://cran.r-project.org/}).

\begin{table}[h]
  \centering
  \begin{tabular}{|l|c|c|c|c|c|}
    \hline
                             & SAS & S-PLUS & R & ViSta & MANET/Mondrian\\\hline
    Basic functionality      &  $\times$  &   $\times$    & $\times$ &   $\times$   & $\times$\\
    Shape                    &     &        & $\times$ &       & $\times$\\
    Residual-based shadings  &  $\times$  &        & $\times$ &   $\times$   & ($\times$)\\
    Conditional Views        &  $\times$  &        & $\times$ &       & $\times$\\
    Interaction              &     &        &   &   $\times$   & $\times$\\
    Extensible Design        &     &        & $\times$ &       &  \\\hline
  \end{tabular}
  \caption{Comparison of current software environments.}
  \label{tab:compare}
\end{table}

Our contribution will give an overview of the state of the art 
of mosaic, sieve, and association plots, both for exploratory visualization and
model-based analysis. Exploratory techniques will include specialized displays for
the bivariate case, as well as pairs plot-like displays for
higher-dimensional tables. As for the model-based tools, particular
emphasis will be given to methods suitable for the visualization of 
conditional independence tests (including permutation tests), as well
as for the visualization of particular GLMs (such as log-linear models). 
In Sect.~\ref{sec:twoway}, we start with the
simple bivariate case.  Section \ref{sec:extensions} explains how the use of color
in residual-based shadings can support data exploration, and
even promotes the methods to diagnostic and model-based tools
by visualizing test statistics and residuals of independence models.
In Sect.~\ref{sec:multiway}, we show how the
basically bivariate methods straightforwardly extend to the multivariate case by using
`flat' representations of the multi-way tables. In this section, we
also introduce specialized displays for conditional independence
structures. Sect.~\ref{sec:concl} concludes the chapter.

\section{Two-way Tables}
\label{sec:twoway}

Throughout these section, our examples will be based on
the \data{hospital} data \cite{vcd:haberman:1974} (see Tab.~\ref{tab:hospital}).

\begin{table}[ht]
\begin{center}
\begin{tabular}{|l||rrr|r|}
\hline
 & Length of stay &&&\\
 & 2--9 days & 10--19 days & more than 20 days & TOTAL \\
\hline\hline
Visit frequency &&&&\\
Regular & 43 & 16 & 3 & 62 \\
Less than monthly & 6 & 11 & 10 & 27 \\
Never & 9 & 18 & 16 & 43 \\\hline
TOTAL & 58 & 45 & 29 & 132 \\
\hline
\end{tabular}
\caption{The \data{hospital} data.}
\label{tab:hospital}
\end{center}
\end{table}

\noindent The table relates the length of stay (in days) in hospital and
the visit freqency for mental patients, 
giving evidence to the fact that the longer the length of 
stay in hospital, the less frequent the visits.

Although far from optimal, contingency tables are frequently visualized using
grouped bar charts (see Fig.~\ref{fig:grouped}) 
or even by means of 3D-bar charts (see Fig.~\ref{fig:threedee}). It seems hard
to detect the aforementioned pattern in these, especially in the 3D
plot where the perspective view tends to distort the true proportions
of the bars. In the following, we
will introduce three graphical methods that are better suited for
contingency tables.

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[p]
\begin{center}
<<grouped,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
barplot(hospital, legend = rownames(hospital), beside = TRUE, 
        xlab = "Length of stay", ylab = "Number of patients")
@
\caption{Barplot for the \data{hospital} data.}
\label{fig:grouped}
\end{center}
\end{figure}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[p]
\begin{center}
<<threedeebars,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
myhospital = t(hospital)[,3:1]
mydat = data.frame("Length of stay" = as.vector(row(myhospital)), 
                   "Visit frequency" = as.vector(col(myhospital)), 
                   "Number of patients" = as.vector(myhospital))
scatterplot3d(mydat, type = "h", pch = " ", lwd = 10, 
x.ticklabs = c("2-9","","10-19","","20+"), 
              y.ticklabs = c("Regular","","Less than monthly","","Never"), 
              xlab = "Length of stay", ylab = "Visit frequency", zlab = "Number of patients",
              y.margin.add = 0.2,
              color = "black", box = F)
@
\caption{3D-barplot for the \data{hospital} data.}
\label{fig:threedee}
\end{center}
\end{figure}

\subsection{Mosaic Displays}
\label{sec:mosaic}

Mosaic displays have been introduced by 
\cite{vcd:Hartigan+Kleiner:1981,vcd:Hartigan+Kleiner:1984} and
extended, e.g., by \cite{vcd:Friendly:1994,vcd:Friendly:1999,vcd:Friendly:2000}. 
They visualize the observed values of a contingency table by
area-proportional tiles, arranged in a squared mosaic. The tiles are
obtained by recursive partitioning splits of a square. Since 
chapter XXX in this book \emph{\bf (Reference to the contribution of Heike Hofmann)} 
is entirely devoted to mosaic plots, in the following we only describe
the main idea. Consider our
example of the hospital data from above. Step 1 consists of splitting a square
according to the marginals of one of the variables. To be consistent with the textual
representation, we choose `Length of stay' with vertical splits 
(see Fig.~\ref{fig:mosaic1fig}).

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[p]
\begin{center}
<<mosaic1,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(margin.table(hospital, 2), split = TRUE)
@ 
\caption{Construction of a mosaic plot for a two-way table, step 1.}
\label{fig:mosaic1fig}
\end{center}
\end{figure}

\noindent The
result is similar to a bar plot where not the height, but the width is
adapted to visualize the counts for each level, which some authors \cite{vcd:hummel:1996}
call a \emph{spine plot}. From this plot, we see that the number of patients 
decreases with the length of stay. Step 2 now is to add further splits in the
other direction, i.e., horizontal splits, for the second
variable. This means that each vertical bar is split according to the
marginals of the second variable, \emph{given} the first variable (see
Fig.~\ref{fig:mosaic2fig}). The resulting plot visualizes the 
contingency table where each cell has a size proportional to the
corresponding table entry.
We can still depict the marginal distribution of `Length of stay', %'
and additionally, the visit frequency \emph{given} each category of `Length of stay'. %'
As we will see in the next section, the grid would be regular
if the two variables were independent. Clearly, compared
to a length of stay of 10--19 days, more patients get regular visits for stays from 2--9 
days, and conversely, less patients get regular visits for stays for more than 20 days.
For patients that get no visits, the pattern is inversed.

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[p]
\begin{center}
<<mosaic2,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(t(hospital), split = TRUE, mar = c(left = 3.5),
       labeling_args = list(offset_labels = c(left = 0.5), 
         offset_varnames = c(left = 1, top = 0.5), set_labels = 
         list("Visit frequency" = c("Regular","Less than\nmonthly","Never"))))
@ 
\caption{Construction of a mosaic plot for a two-way table, step 2.}
\label{fig:mosaic2fig}
\end{center}
\end{figure}

Since mosaic plots are asymmetric by construction, the choice of the
variable order matters, as the first splitting variable dominates the 
plot. In our example, if we use `Visit frequency' %'
as the first splitting variable, the impression is very different compared to the previous
mosaic (see Fig.~\ref{fig:mosaic3fig}). 
In this alternative display, we see the marginal distribution of `Visit frequency' %'
in the rows: about half of the patients get visited regularly. This
group is dominated by patients staying between 2 and 9 days.
It seems apparent that the distribution of `Length of stay' %'
is similar for monthly and never visited patients, so this two categories actually
represent one homogenous group (patients visited only casually).

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<mosaic3,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(hospital)
@ 
\caption{Mosaic plot for the hospital data, using `Visit frequency' %'
as first splitting variable.}
\label{fig:mosaic3fig}
\end{center}
\end{figure}

\subsection{Sieve Plots}
\label{sec:sieve}

When we try to explain data, we suppose the validity of a certain
model for the generating process. In the case of two-way contingency
tables, the two most common hypotheses are

\begin{enumerate}
\item independence of the two variables, and
\item homogeneity of one variable among the strata defined by
the second.
\end{enumerate}

\noindent It is easy to compute the \emph{expected} table under either
of these hypotheses. Consider a 
2-way contingency table with $I$ rows and $J$ columns,
cell frequencies $\{n_{ij}\}$ for
$i = 1, \dots, I$ and $j = 1, \dots, J$, and row and column sums
$n_{i+} = \sum_i n_{ij}$ and $n_{+j} = \sum_j n_{ij}$, respectively. For
convenience, the number of observations is denoted $n = n_{++}$.
Given an underlying distribution with theoretical cell probabilities
$\pi_{ij}$, the null hypothesis of independence of the two categorical
variables can be formulated as
\begin{equation} \label{eq:H0}
H_0: \; \pi_{ij} \quad = \quad \pi_{i+} \pi_{+j}.
\end{equation}

\begin{table}[ht]
\begin{center}
\begin{tabular}{|l||rrr|r|}
\hline
 & Length of stay &&&\\
 & 2--9 & 10--19 & 20+ & TOTAL \\
\hline\hline
Visit frequency &&&&\\
Regular & 27.24 & 21.14 & 13.62 & 62 \\
Less than monthly & 11.86 & 9.20 & 5.93 & 27 \\
Never & 18.89 & 14.66 & 9.45 & 43 \\\hline
TOTAL & 58.00 & 45.00 & 29.00 & 132 \\
\hline
\end{tabular}
\caption{The \data{hospital} data---expected values.}
\label{tab:hospital2}
\end{center}
\end{table}

Now, the expected cell frequencies in this model are simply
$\hat n_{ij} = n_{i+}n_{+j}/n$.
The expected table for our sample data is given in Tab.~\ref{tab:hospital2}.
It could again be visualized using a mosaic plot, this
time applied to the table of expected frequencies. In
Fig.~\ref{fig:sieve1fig}, we can see such a plot in which,
additionally, a number of squares equal to the corresponding cell entry has
been added to each tile. We see that the grid formed by the tiles is completely
regular. The added rectangles in the tiles become more useful when they 
are not based on the expected but the observed values
instead (see Fig.~\ref{fig:sieve2fig}). Such a \emph{sieve plot} implicitly compares
expected and observed values since the density of the grid will
increase with the deviation of the observed from the expected
values. This allows the detection of general association patterns (for
nominal variables) and of linear association (for ordinal variables),
the latter producing tiles of either very high or very low density 
along one of the diagonals.
In the case of our data, the plot reveals indeed a negative association of 
the two variables since the density of the rectangles is marked along the secondary
diagonal, giving evidence to the fact that for these patients,
visit frequency decreases with the length of stay.

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[p]
\begin{center}
<<sieve1fig,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
sieve(t(hospital), split = TRUE, sievetype = "expected", pop = FALSE)
labeling_cells(text = round(independence_table(t(hospital)),2), clip = FALSE, gp = gpar(fontface = 2, fontsize = 15))(t(hospital))
@ 
\caption{Expected values for the \data{hospital} data.}
\label{fig:sieve1fig}
\end{center}
\end{figure}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[p]
\begin{center}
<<sieve2fig,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
sieve(t(hospital), split = TRUE, pop = FALSE, gp = gpar(lty =
"dotted", col = "dark gray"))
labeling_cells(text = t(hospital), clip = FALSE, gp = gpar(fontface = 2, fontsize = 15))(t(hospital))
@ 
\caption{Sieve plot for the \data{hospital} data.}
\label{fig:sieve2fig}
\end{center}
\end{figure}

\subsection{Association Plots}
\label{sec:assoc}

In the last section, we have seen how to compare observed and expected
values of a contingency table using sieve plots. We can do this
more straightforwardly by using a plot that directly visualizes the residuals. The
most widely known residuals are the Pearson residuals

\begin{equation} \label{eq:Pearson}
r_{ij} \quad = \quad \frac{n_{ij} - \hat n_{ij}}{\sqrt{\hat n_{ij}}}.
\end{equation}

\noindent that are standardized raw residuals. They are best visualized by
association plots \cite{vcd:Cohen:1980}:
each cell is represented by a rectangle that has (signed) height
proportional to the corresponding Pearson residual $r_{ij}$ and width proportional to the
square root of the expected counts $\sqrt{\hat n_{ij}}$. Thus, the area is proportional to
the raw residuals $n_{ij} - \hat n_{ij}$. In available
implementations, the sign of a residual is often redundantly
coded, e.g., by the rectangle's color (see Sect.~\ref{sec:shadings}) %'
and its position relative to the baseline. Figure \ref{fig:assoc}
shows the association plot for the hospital data. 
Consistent with the corresponding mosaic and sieve plots, 
we clearly see that too many (too less) patients that stay from 
2 to 9 (more than 20) days get visited regularly
than would be expected under the null of independence, and that this pattern is 
reversed for patients never visited.

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<assoc,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
assoc(t(hospital), split = TRUE)
@ 
\caption{Association plot for the \data{hospital} data.}
\label{fig:assoc}
\end{center}
\end{figure}

\section{Residual-based Shadings}
\label{sec:extensions}

As introduced in the previous section for association plots, the
investigation of residuals from a posited independence model is of
major interest in analyzing contingency tables. In the following, we
will demonstrate how the use of colors can greatly facilitate the
detection of interesting patterns.

\subsection{Using Colors for Residuals}
\label{sec:shadings}

All plots introduced in the previous section can be improved using colors,
either by adding additional information, or by redundantly coding information
already visualized by the `raw' plot to support our perceptual system. %'

First, we consider the sieve plots. The density 
of the grid in the raw version implicitly gives us an 
idea of the residuals' %'
size, but since the plot does not include the density corresponding to zero
residuals (the null model) for comparison, we 
cannot easily assess whether there are more, of less counts in a cell 
than expected under the null hypothesis. It would help if we knew the 
sign of the residuals. 
Using color, we can add this information, for example using blue for
positive, red for negative, and gray for zero residuals. Figure \ref{fig:colorsieve} 
demonstrates the effect of applying such a color shading to a sieve plot for the 
\data{hospital} data. Color clearly emphasizes the linear association
pattern in this data.

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<sievecolor,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
sieve(t(hospital), split = TRUE, shade = TRUE)
@ 
\caption{Sieve plot with color coding of the residuals.}
\label{fig:colorsieve}
\end{center}
\end{figure}

Mosaic plots in their initial version are monochrome
displays. Friendly \cite{vcd:Friendly:1994} introduced a
residual-based shading of the tiles to additionally visualize the
residuals from a given independence model fitted to the table. 
The idea is to use a color coding for the mosaic tiles that visualizes the
sign and absolute size of each residual $r_{ij}$: Cells corresponding to small
residuals ($|r_{ij}| < 2$) are shaded white. Cells with medium sized residuals
($2 \le |r_{ij}| < 4$) are shaded light blue and light red for positive and negative
residuals, respectively. Cells with large residuals ($|r_{ij}| \ge 4$) are shaded
with a fully saturated blue and red, respectively. 
The heuristic for choosing the cut offs $2$ and $4$ is that the Pearson residuals
are asymptotically standard normal which implies that the highlighted cells are those with
residuals \emph{individually} significant at approximately the $\alpha
= 0.05$ and $\alpha = 0.0001$ levels. However, the main purpose of the
Friendly shading is not to visualize significance but the \emph{pattern} of
deviation from independence \cite{vcd:Friendly:2000}. 
In addition to the shading of the rectangles themselves, the Friendly shading also
encompasses a choice of line type and line color of the rectangle borders
with similar ideas as described above.

In Fig.~\ref{fig:Friendly}, we again show the mosaic for the \data{hospital}
data, this time using the Friendly color coding.

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<friendly,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(t(hospital), split = TRUE, shade = TRUE, gp = shading_Friendly,
       legend = legend_fixed, mar = c(left = 3.5),
       labeling_args = list(offset_labels = c(left = 0.5), 
         offset_varnames = c(left = 1, top = 0.5), set_labels = 
         list("Visit frequency" = c("Regular","Less than\nmonthly","Never"))))
@ 
\caption{Mosaic display with Friendly color coding of the residuals.}
\label{fig:Friendly}
\end{center}
\end{figure}

\noindent Clearly, the asymmetry for regular and never visited students,
and the pattern inversion for lengths of stay of 2--9 and more than 20
days are emphasized using the color shading.

For association plots, residual-based shadings are redundant since all relevant
information is already contained in the plot by construction. But using one of the
shadings discussed above will nevertheless 
support the analysis process and is therefore recommended.
For example, applying the Friendly shading in Fig.~\ref{fig:colorassoc} 
facilitates the discrimination between
positive and negative residuals, and therefore to get a quick overview on the
discussed deviation pattern.

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<assoccolor,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
assoc(t(hospital), split = TRUE, shade = TRUE, keep = TRUE,
gp = shading_Friendly, legend = legend_fixed)
@ 
\caption{Association plot with Friendly color coding of the residuals.}
\label{fig:colorassoc}
\end{center}
\end{figure}

\subsection{A Note on Color Palettes}
\label{sec:colors}

The color shading introduced in the previous section but can still be
improved. One of the issues is the suboptimal color palette used in
the Friendly shading. Implementations are often based on so-called
Hue-Saturation-Value (HSV) colors (or the comparable Hue-Luminance-Saturation 
color scheme). Both spaces are rather similar transformations of RGB (Red-Green-Blue) space
\cite{vcd:Brewer:1999,vcd:Poynton:2000} and are very common implementations of
colors in many computer packages \cite{vcd:Moretti+Lyons:2002} making the generation
of the Friendly shading very simple. Each color in this space is
represented by three dimensions: the hue, the saturation (`colorfulness'), and 
the value (`lightness', amount of gray). But HSV colors have several disadvantages.
Most importantly, HSV colors are not perceptually
uniform because the three HSV dimensions map only poorly to the three
perceptual dimensions of the human visual system  \cite{vcd:Brewer:1999,vcd:Ihaka:2003}.
Consequently, the HSV dimensions are confounded, e.g., saturation
is not uniform across different hues.
For example, see Fig.~\ref{fig:colors}, left-hand side, 
showing a qualitative color palette in the HSV
space: although saturation and value are fixed, the fully saturated blue is
perceived much darker than the fully saturated red or green. 
This makes it more difficult for the human eye to judge the size of shaded areas
and can therefore lead to color-caused optical illusions when used in statistical graphs
\cite{vcd:Cleveland+McGill:1983}. Furthermore, flashy fully saturated HSV colors are
good for drawing attention to a plot, 
but hard to look at for a longer time \cite{vcd:Ihaka:2003}
which makes HSV-shaded graphics harder to interpret.

\setkeys{Gin}{width=\textwidth}
\begin{figure}
\begin{center}
<<colorfig,fig=TRUE,echo=FALSE,height=3,width=7,results=hide>>=
par(mfrow = c(1,2), mar = c(1,1,1,1), oma = c(0,0,0,0))
pie(rep(1,9), radius = 1, col = hsv(seq(0,1,length=10))[-1])
pie(rep(1,9), radius = 1, col = hcl(seq(0,360,length=10))[-1])
@ 
\caption{Qualitative color palette for the HSV (left hand side) 
         and HCL (right hand side) spaces.}
\label{fig:colors}
\end{center}
\end{figure}
\setkeys{Gin}{width=0.7\textwidth}

The use of colors that are more `in harmony' go back to Munsell 
\cite{vcd:Munsell:1905} %'
who introduced a color notation for balanced colors. Based on those,
tools producing better palettes for specific tasks have been developed 
\cite{vcd:Harrower+Brewer:2003}. Other perceptually-based color
spaces, especially suited for displays, are the CIELAB and CIELUV
spaces \cite{vcd:CIE:2004} from which qualitative palettes for
statistical graphics have been derived \cite{vcd:Ihaka:2003}. A
transformation of the CIELUV space leads to the HCL (Hue-Chroma-Luminance) space.
HCL colors with fixed chroma and luminance are always balanced towards the same grey 
and thus do not have the problem of varying saturations as the HSV
colors (see Fig.~\ref{fig:colors}, right hand side). 
Similarly, diverging color palettes can be derived that are
suited for visualizing residuals in a similar way than the Friendly
shading \cite{vcd:Zeileis+Meyer+Hornik:2005}.
Figure \ref{fig:hclmosaic} shows again the mosaic plot for the \data{hospital} data, 
this time using HCL colors and an alternative legend, showing the whole range of residuals.

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<hclmosaicfig,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(t(hospital), split = TRUE, mar = c(left = 3.5), shade = TRUE,
       labeling_args = list(offset_labels = c(left = 0.5), 
         offset_varnames = c(left = 1, top = 0.5), set_labels = 
         list("Visit frequency" = c("Regular","Less than\nmonthly","Never"))))
@ 
\caption{Mosaic display with HCL shading of the residuals.}
\label{fig:hclmosaic}
\end{center}
\end{figure}

\subsection{Visualizing Test Statistics}
\label{sec:test}

Figure \ref{fig:hclmosaic} includes the $p$ value of a $\chi^2$ test
of independence, frequently used to assess the significance of the
hypothesis of independence (or homogeneity for stratified data) in
two-way tables. The test statistic is just the sum of the squared
Pearson residuals

\begin{equation} \label{eq:ChiSquare}
X^2 \quad = \quad \sum_{i, j}r_{ij}^2~,
\end{equation}

\noindent known to have a limiting $\chi^2$ distribution with $(I-1)(J-1)$
degrees of freedom under the null hypothesis.
Since the HCL space is three-dimensional and we only
used two dimensions so far for coding information (hue for the sign and
colorfulness for the size of the residuals), we can use the third dimension
(lightness) for visualizing the significance of some specified test statistic,
for example the $\chi^2$ test statistic, 
using darker (`uninteresting') colors for non-significant results.
The following example using the \data{Bundesliga} data \cite{vcd:Knorr-Held:1999}
shows the relationship of home goals and away
goals of Germany's premier soccer league in 1995: although there are two
``larger'' residuals (one greater than 2, one less then $-2$), the
$\chi^2$ test does not reject the null hypothesis of
independence on a significance level of $95\%$ since the $p$ value is $0.12133$. 
Consequently, the employed shading uses a darker overall color scheme
(see Fig.~\ref{fig:bundesliga}).

\begin{figure}[p]
\begin{center}
<<bundesligafig,fig=TRUE,echo=FALSE,height=6,width=7>>=
bl <- xtabs(~ HomeGoals + AwayGoals, data = Bundesliga, subset = Year == 1995)
mosaic(bl, shade = TRUE)
@
\caption{Mosaic plot for part of the \data{Bundesliga}, using a $\chi^2$ test and
fixed cut-off points for the residuals.}
\label{fig:bundesliga}
\end{center}
\end{figure}

The previous example shows that the heuristic for choosing the cut-off points 
in the Friendly shading may lead to wrong conclusions: the test of
independence is not significant even though some of the residuals are
``large''. The reason here is that the cut-off points are really
data-dependent. Another example where the Friendly-shading may trap
the rush statistician is the case of the \data{arthritis} data 
\cite{vcd:Koch+Edwards:1988}, resulting from a double-blind clinical trial
investigating a new treatment for rheumatoid arthritis, stratified by gender.
Figure \ref{fig:arthritis} visualizes the results for the female
patients, again by means of a mosaic display. Clearly, the hypothesis
of independence is rejected by the $\chi^2$ test even on a $99\%$ level ($p=0.0035$), but
since all residuals are in the $[-1.72,1.87]$ interval, the tiles
remain uncolored. A solution to this issue is to use a different
test statistic, for example the maximum of the absolute values of the
Pearson residuals \cite{vcd:Meyer+Zeileis+Hornik:2003} instead of the
sum of squares:

\begin{equation} \label{eq:MaxAbs}
M \quad = \quad \max_{i, j} |r_{ij}|~.
\end{equation}

\noindent Given a critical value $c_\alpha$ for this test statistic, all residuals whose
absolute value exceeds $c_\alpha$ violate the hypothesis of indendence at level
$\alpha$ \cite[ch. 7]{vcd:Mazanec+Strasser:2000}. Thus, the interesting cells giving
evidence for the rejection of the independence hypothesis can easily be identified.
The distribution of this test statistic under the null can be obtained
by simulation, sampling tables with the same row and column sums $n_{i+}$ and $n_{+j}$
using the Patefield algorithm \cite{vcd:Patefield:1981} 
and computing the maximum statistic for each of these tables.
In Figs.~\ref{fig:bundesliga2} and \ref{fig:arthritis2}, we again
visualize the \data{Bundesliga} and the \data{arthritis} data, this
time using the maximum test statistic. We see that for the
\data{Bundesliga} data no cell is colored, and we now can indeed conclude
from the visualization that neither of the home and away teams seems advantaged. 
On the other hand, for the \data{arthritis} data, the shading of the
tiles clearly shows that the treatment is effective: too many patients
in the treatment group exhibit marked improvement than would be
expected under independence.

\begin{figure}[p]
\begin{center}
<<bundesliga2fig,fig=TRUE,echo=FALSE,height=6,width=7>>=
bl <- xtabs(~ HomeGoals + AwayGoals, data = Bundesliga, subset = Year == 1995)
mosaic(bl, gp = shading_max)
@
\caption{Mosaic plot for part of the \data{Bundesliga}, using the maximum statistic
and data-driven cut-off points for the residuals.}
\label{fig:bundesliga2}
\end{center}
\end{figure}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[p]
\begin{center}
<<Arthritis,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(art, shade = TRUE)
@
\caption{Mosaic plot for the \data{Arthritis} data, using the $\chi^2$ test 
and fixed cut-off points for the shading.}
\label{fig:arthritis}
\end{center}
\end{figure}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[p]
\begin{center}
<<Arthritis2fig,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(art, gp = shading_max)
@
\caption{Mosaic plot for the \data{Arthritis} data, using the maximum test and 
data-driven cut-off points for the residuals.}
\label{fig:arthritis2}
\end{center}
\end{figure}

\section{Multi-way Tables}
\label{sec:multiway}

In Sect.~\ref{sec:twoway}, we have presented basic visualization methods
for two-way tables. In this section, we will show how these methods
extend to multi-way tables by applying them to `flat' representations,
and we will treat specialized displays for conditional independence models.

\subsection{Visualize Flat-Tables}
\label{sec:flat}

The main idea of the mosaic, sieve, and association plots presented
in the previous sections is to visualize information on 
the tables' cells, arranged in rectangular form. For multi-way tables,
mosaic plots can directly be used by simply adding further splits for each 
additional variable. For sieve and association plots,
we apply the basic idea of mosaic plots to the table itself,
i.e., simply nest the variables into rows and columns using recursive
conditional splits, given the margins. The result is a 
`flat' representation of the multi-way table that can be visualized in %'
ways similar to a two-dimensional table.
As an example, consider the \data{HairEyeColor} 
data containing two polytomous variables (hair and eye color), 
as well as one (artificial) dichotomous variable (gender). 
A `flattened' contingency table, putting eye color in the columns %'
and hair color---nested in gender---in the rows, is given by Tab.~\ref{tab:hec}.

\begin{table}[ht]
\begin{center}
\begin{tabular}{|ll||rrrr|}
\hline
 & & EYE &&&\\
 & & Brown & Blue & Hazel & Green\\\hline\hline
GENDER & HAIR &&&&\\
Male & Black & 32 & 11 & 10 & 3 \\
& Brown & 38 & 50 & 25 & 15 \\
& Red & 10 & 10 & 7 & 7 \\
& Blond & 3 & 30 & 5 & 8 \\
Female & Black & 36 & 9 & 5 & 2 \\
& Brown & 81 & 34 & 29 & 14 \\
& Red & 16 & 7 & 7 & 7 \\
& Blond & 4 & 64 & 5 & 8 \\
\hline
\end{tabular}
\caption{The \data{HairEyeColor} data, in flat representation.}
\label{tab:hec}
\end{center}
\end{table}

\noindent The corresponding mosaic, sieve, and association plots are shown in
Figs.~\ref{fig:hecmosaic}--\ref{fig:hecassoc}. Note that with three variables, we have the
choice of four kinds of independence models that can be fitted to the table:
partial, joint, and conditional independence, as well as 
the model of no three-way interaction. For demonstration purposes,
the shadings in these plots visualize
the model of hair and eye color being \emph{jointly} independent from gender.
The model fits better than others, but still is significant at the
$95\%$ level.

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<hecmosaic,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
hec = structable(Eye ~ Sex + Hair, data = HairEyeColor)
mosaic(hec, labeling_args = list(tl_labels = c(Hair = TRUE), abbreviate = c(Hair = 4)), 
       mar = c(left = 4), shade = TRUE, expected = ~ Hair * Eye + Sex)
@
\caption{Mosaic plot for the \data{HairEyeColor} data.}
\label{fig:hecmosaic}
\end{center}
\end{figure}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<hecsieve,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
hec = structable(Eye ~ Sex + Hair, data = HairEyeColor)
sieve(hec, labeling_args = list(tl_labels = c(Hair = TRUE), abbreviate = c(Hair = 4)), 
       mar = c(left = 4), shade = TRUE, expected = ~ Hair * Eye + Sex)
@
\caption{Sieve plot for the \data{HairEyeColor} data.}
\label{fig:hecsieve}
\end{center}
\end{figure}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<hecassoc,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
hec = structable(Eye ~ Sex + Hair, data = HairEyeColor)
assoc(hec, labeling_args = list(tl_labels = c(Hair = TRUE), abbreviate = c(Hair = 4)), 
      mar = c(left = 4), shade = TRUE, expected = ~ Hair * Eye + Sex)
@
\caption{Association plot for the \data{HairEyeColor} data.}
\label{fig:hecassoc}
\end{center}
\end{figure}

\subsection{Displays for Conditional Independence}
\label{sec:condind}

Apart from the generic approach to the visualization of multi-way tables outlined in 
the previous section, there are modifications 
and specialized displays designed for the
visualization of conditional independence structures. A first approach
is to modify the spacing of the standard mosaic display---which already 
is a conditional plot by construction---to better distinguish conditional
from conditioned variables. Figure \ref{fig:presex}
illustrates the conditional independence of premarital and
extramarital sex, given gender and marital status, in the \data{PreSex} 
data \cite{vcd:thornes+collard:1979,vcd:gilbert:1981}. The $\chi^2$
test of independence rejects the null hypothesis: possibly, because
too many men who had premarital sex also had extramarital sex? For the conditioned
variables premarital and extramarital sex, we use a fixed spacing, whereas for 
the conditioning variables gender and marital status, the spacing is increasing.

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<PreSex,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(PreSex, condvars = c(1, 4), shade = TRUE)
@
\caption{Mosaic plot for the pre- and extramarital sex data.}
\label{fig:presex}
\end{center}
\end{figure}

Another possibility is to use trellis-like layouts for visualizing partial tables for 
given strata, defined by the conditioning variables.
Figure \ref{fig:ucbadmissions} visualizes the well-known UCB admissions data
\cite{vcd:Bickel+Hammel+O'Connell:1975} by the means of a conditional 
association plot. The panels show the
residuals from a conditional independence model (independence of gender
and admission, given department), stratified by department. Clearly,
the situation in department A (more women/less men accepted than
would be expected under the null) causes the rejection of the
hypothesis of conditional independence, indicating that female
students are more successful in that department than their male colleagues.

\setkeys{Gin}{width=\textwidth}
\begin{figure}
\begin{center}
<<UCBAdmissions,fig=TRUE,echo=FALSE,height=6,width=10,results=hide>>=
cotabplot(UCBAdmissions, panel = cotab_assoc, shade = TRUE, legend = FALSE)
@
\caption{Conditional association plot for the UCB admissions data.}
\label{fig:ucbadmissions}
\end{center}
\end{figure}

Finally, it is possible for all basic plots to create pairwise displays,
arranged in in a matrix similar to scatterplots in a pairs plot.
The diagonal cells contain the variable names, optionally with
univariate statistics, whereas the off-diagonal cells contain plots whose
variables are implicitly specified by the cells' position in the matrix. %'
More formally, each cell $a_{ij}$ in such a matrix defines two
variables $i$ and $j$ that can be used to specify the model visualized
in that cell. Typical hypotheses are: Variables $i$ ad $j$ are
marginally independent; variables $i$ and $j$ are conditionally
independent, given all others; variables $i$ and $j$ are jointly
independent from all others, etc.
Figure \ref{fig:pairs} shows a pairs display with mosaic plots visualizing mutual 
independence in the upper triangle,
sieve plots for the same in the lower triangle, and bar charts in the diagonal.
Immediately, we can detect the independence of hair and gender and eye and gender from the
non-colored cells in the corresponding mosaic plots. Thus, the associational pattern
present in the sieve plots are not significant. On the other hand, hair and eye color 
clearly are not independent.

\setkeys{Gin}{width=\textwidth}
\begin{figure}
\begin{center}
<<pairs,fig=TRUE,echo=FALSE,height=8,width=8,results=hide>>=
pairs(t(hec), lower_panel = pairs_sieve, space = 0.3, shade = TRUE,
      diag_panel_args = list(rot = -45, just_leveltext = c("left","bottom")))
@
\caption{Pairs plot for the \data{HairEyeColor} data.}
\label{fig:pairs}
\end{center}
\end{figure}

\section{Conclusion}
\label{sec:concl}

This chapter reviews several alternatives for the visualization of
multi-way contingency tables. For two-way tables, mosaic, sieve, and
association plots are suitable for the visualization of observed and
expected values and the Pearson residuals, respectively. This basic
methods are enhanced by using residual-based shadings, preferably
based on perceptual color palettes such as those derived from the HCL
space. Residual-based shadings can be used to visualize sign and size
of the residuals, as well as the significance of test statistics such
as the $\chi^2$ or the maximum test statistic. The latter has the
advantage of detecting residuals causing the rejection of the hypothesis
of independence. The methods directly extend to the multi-way case by
using `flat' representations of the multi-way tables, and specialized
displays for conditional independence such as trellis layouts of
partial tables and pairs plots.

\bibliographystyle{plain}
\bibliography{vcd}

\end{document}

### 2005-11-16 trunk/vcd/paper/hsc.Rnw
-bug fixes in pairs plot -sieve plots now use HCL colors -`labeling' argument now also accepts a logical
-hsc-paper revised



git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@498 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-\title*{Visualizing Multi-Way Contingency Tables}
+\title*{Visualizing Multi-way Contingency Tables}
 % Use \titlerunning{Short Title} for an abbreviated version of
@@
 typically look for (conditional) independence structures using common
-statistical methods such as independence tests and
-log-linear models. The analysis quickly 
-becomes complex with increasing dimensionality.
-One idea of visualization techniques is to
+methods such as independence tests and
+log-linear models. One idea of visualization techniques is to
 use the human visual system to detect structures in the data that
@@
 task is purely exploratory or model based, techniques such as mosaic, sieve, and
-association plots offer good support for visualization. All three have been
+association plots offer good support for visualization. They have been
 extended over the last two decades, and implementations exist in many
@@
 contingency tables. A \emph{mosaic plot} \cite{vcd:Hartigan+Kleiner:1984} is basically 
-an area-proportional visualization of (typically observed) frequencies, composed
+an area-proportional visualization of (typically, observed) frequencies, composed
 of tiles (corresponding to the cells) created by recursive
@@
 dimensions of previous splits. 
-\emph{Sieve plot} are similar to mosaic plots, but the area of each
+\emph{Sieve plots} \cite{vcd:riedwyl+schuepbach:1994} 
+are similar to mosaic plots, but the area of each
 tile is proportional to the \emph{expected} cell entry, and each tile
@@
 
-Over the years, extensions to these techniques were mainly focused on
-five aspects:
-
-\begin{enumerate}
-\item Varying the shape of bar plots and mosaic displays to yield, e.g.,
+Over the years, extensions to these techniques have mainly been focused on
+the following aspects:
+
+\begin{itemize}
+  \item Varying the shape of mosaic plots (as well as bar plots) to yield, e.g.,
   double-decker plots \cite{vcd:hofmann:2001}, spine 
@@
   user extensions \cite{vcd:Meyer+Zeileis+Hornik:2005a}.
-\end{enumerate}
+\end{itemize}
 
@@
 \proglang{R} \cite{vcd:R:2005}, and \proglang{S-PLUS} \cite{vcd:SPLUS:2005}. 
-Table \ref{tab:compare} gives an overview of the available functionality in these systems.
+Implementations of association and sieve plots can only be found in
+\proglang{R} and \proglang{SAS} (in the latter, these plots are available
+for two-way tables only).
+Table~\ref{tab:compare} gives an overview of the available functionality in these systems.
 The figures in this chapter have all been produced using the
@@
 (Fig.~\ref{fig:threedee} only), all freely available from the
-Comprehensive \proglang{R} Archive Network (\url{http://cran.r-project.org/}).
+Comprehensive \proglang{R} Archive Network (\url{http://CRAN.R-project.org/}).
 
@@
 
-Our contribution will give an overview of the state of the art 
+This chapter will give an overview of the state of the art 
 of mosaic, sieve, and association plots, both for exploratory visualization and
@@
 In Sect.~\ref{sec:twoway}, we start with the
-simple bivariate case.  Section \ref{sec:extensions} explains how the use of color
+simple bivariate case.  Section~\ref{sec:extensions} explains how the use of color
 in residual-based shadings can support data exploration, and
@@
 
-Throughout these section, our examples will be based on
-the \data{hospital} data \cite{vcd:haberman:1974} (see Tab.~\ref{tab:hospital}).
-
-\begin{table}[ht]
-\begin{center}
-\begin{tabular}{|l||rrr|r|}
-\hline
- & Length of stay &&&\\
- & 2--9 days & 10--19 days & more than 20 days & TOTAL \\
-\hline\hline
-Visit frequency &&&&\\
+Throughout this section, our examples will be based on
+the \data{hospital} data \cite{vcd:wing:1962} (see Tab.~\ref{tab:hospital}).
+
+\begin{table}[h]
+\begin{center}
+\begin{tabular}{|l|rrr|r|}
+\hline
+ & \multicolumn{4}{l|}{\textbf{Length of stay (in years)}}\\
+\textbf{Visit frequency} &~~2--9 & ~~10--19 & ~~20+ & $\Sigma$ \\
+\hline
 Regular & 43 & 16 & 3 & 62 \\
@@
 Never & 9 & 18 & 16 & 43 \\\hline
-TOTAL & 58 & 45 & 29 & 132 \\
+$\Sigma$ & 58 & 45 & 29 & 132 \\
 \hline
@@
 
-\noindent The table relates the length of stay (in days) in hospital and
-the visit freqency for mental patients, 
-giving evidence to the fact that the longer the length of 
-stay in hospital, the less frequent the visits.
+\noindent The table relates the length of stay (in years) of 132
+long-term schizophrenic patients in two London mental hospitals with
+the freqency of visits. The length of stay (LOS) has been categorized into
+2--9 years, 10--19 years, and more than 19 years. There are also three
+categories for the visit frequency: regular (including
+patients that were allowed to go home), less than monthly, and never.
+Wing \cite{vcd:wing:1962} concludes from this data 
+that the longer the length of stay in hospital, the less frequent the
+visits, which can be seen from the column-standardized table 
+(see Tab.~\ref{tab:colhospital}).
+
+\begin{table}[h]
+\begin{center}
+\begin{tabular}{|l|rrr|}
+\hline
+ & \multicolumn{3}{l|}{\textbf{Length of stay (in years)}}\\
+\textbf{Visit frequency} &~~2--9 & ~~10--19 & ~~20+ \\
+\hline
+Regular & 0.74 & 0.36 & 0.10  \\
+Less than monthly & 0.10 & 0.24 & 0.35  \\
+Never & 0.16 & 0.40 & 0.55  \\\hline
+$\Sigma$ & 1.00 & 1.00 & 1.00 \\
+\hline
+\end{tabular}
+\caption{The \data{hospital} data, corrected for the column margin.}
+\label{tab:colhospital}
+\end{center}
+\end{table}
+
+\noindent In addition, \cite{vcd:haberman:1974} notes that this pattern is not
+significantly different in the ``less than monthly'' and ``never''
+strata. From the row-standardized table (see Tab~.\ref{tab:rowhospital}), 
+it seems indeed that LOS were homogeneous with respect to these two visit frequency strata.
+
+\begin{table}[h]
+\begin{center}
+\begin{tabular}{|l|rrr|r|}
+\hline
+ & \multicolumn{4}{l|}{\textbf{Length of stay (in years)}}\\
+\textbf{Visit frequency} &~~2--9 & ~~10--19 & ~~20+ & $\Sigma$ \\
+\hline
+Regular & 0.69 & 0.26 & 0.05 & 1.00 \\
+Less than monthly & 0.22 & 0.41 & 0.37 & 1.00 \\
+Never & 0.21 & 0.42 & 0.37 & 1.00 \\
+\hline
+\end{tabular}
+\caption{The \data{hospital} data, corrected for the row margin.}
+\label{tab:rowhospital}
+\end{center}
+\end{table}
 
 Although far from optimal, contingency tables are frequently visualized using
-grouped bar charts (see Fig.~\ref{fig:grouped}) 
+grouped bar plots (see Fig.~\ref{fig:grouped}) 
 or even by means of 3D-bar charts (see Fig.~\ref{fig:threedee}). It seems hard
@@
 barplot(hospital, legend = rownames(hospital), beside = TRUE, 
-        xlab = "Length of stay", ylab = "Number of patients")
-@
-\caption{Barplot for the \data{hospital} data.}
+        xlab = "Length of stay (years)", ylab = "Number of patients")
+@
+\caption{Bar plot for the \data{hospital} data.}
 \label{fig:grouped}
@@
               y.ticklabs = c("Regular","","Less than monthly","","Never"), 
-              xlab = "Length of stay", ylab = "Visit frequency", zlab = "Number of patients",
+              xlab = "Length of stay (years)", ylab = "Visit frequency", zlab = "Number of patients",
               y.margin.add = 0.2,
@@
 @
-\caption{3D-barplot for the \data{hospital} data.}
+\caption{3D-bar chart for the \data{hospital} data.}
 \label{fig:threedee}
@@
 area-proportional tiles, arranged in a squared mosaic. The tiles are
-obtained by recursive partitioning splits of a square. Since 
-chapter XXX in this book \emph{\bf (Reference to the contribution of Heike Hofmann)} 
-is entirely devoted to mosaic plots, in the following we only describe
-the main idea. Consider our
+obtained by recursive partitioning splits of a square.
+In the following, we describe the main idea of mosaic plots, 
+chapter XXX in this book \emph{\bf (Reference to other contributions)} 
+provides more detailed information. Consider our
 example of the hospital data from above. Step 1 consists of splitting a square
 according to the marginals of one of the variables. To be consistent with the textual
-representation, we choose `Length of stay' with vertical splits 
-(see Fig.~\ref{fig:mosaic1fig}).
-
-\setkeys{Gin}{width=0.7\textwidth}
-\begin{figure}[p]
-\begin{center}
-<<mosaic1,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-mosaic(margin.table(hospital, 2), split = TRUE)
-@ 
-\caption{Construction of a mosaic plot for a two-way table, step 1.}
-\label{fig:mosaic1fig}
-\end{center}
-\end{figure}
-
-\noindent The
-result is similar to a bar plot where not the height, but the width is
-adapted to visualize the counts for each level, which some authors \cite{vcd:hummel:1996}
-call a \emph{spine plot}. From this plot, we see that the number of patients 
+representation, we choose LOS with vertical splits (see Fig.~\ref{fig:mosaic1fig}). 
+The result is similar to a bar plot where not the height, but the width is
+adapted to visualize the counts for each level, which is also 
+called a \emph{spine plot} \cite{vcd:hummel:1996}. From this plot, we see that 
+the number of patients 
 decreases with the length of stay. Step 2 now is to add further splits in the
@@
 corresponding table entry.
-We can still depict the marginal distribution of `Length of stay', %'
-and additionally, the visit frequency \emph{given} each category of `Length of stay'. %'
-As we will see in the next section, the grid would be regular
-if the two variables were independent. Clearly, compared
-to a length of stay of 10--19 days, more patients get regular visits for stays from 2--9 
-days, and conversely, less patients get regular visits for stays for more than 20 days.
+We can still see the marginal distribution of LOS, %'
+and additionally, the visit frequency \emph{given} each category of LOS. %'
+If the two variables were independent, the grid would be regular. Clearly, compared
+to a length of stay of 10--19 years, more patients get regular visits for stays from 2--9 
+years, and conversely, less patients get regular visits for stays for more than 19 years.
 For patients that get no visits, the pattern is inversed.
+
+\setkeys{Gin}{width=0.7\textwidth}
+\begin{figure}[p]
+\begin{center}
+<<mosaic1,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
+mosaic(margin.table(hospital, 2), split = TRUE)
+@ 
+\caption{Construction of a mosaic plot for a two-way table, step 1.}
+\label{fig:mosaic1fig}
+\end{center}
+\end{figure}
 
@@
 in the rows: about half of the patients get visited regularly. This
-group is dominated by patients staying between 2 and 9 days.
-It seems apparent that the distribution of `Length of stay' %'
+group is dominated by patients staying between 2 and 9 years.
+It seems apparent that the distribution of LOS %'
 is similar for monthly and never visited patients, so this two categories actually
-represent one homogenous group (patients visited only casually).
+represent one homogenous group (patients visited only casually). Since
+the first splitting variable dominates the plot, it should be chosen
+as to be the explanatory variable in (hypothesized) causal relationships.
 
@@
 
-When we try to explain data, we suppose the validity of a certain
+When we try to explain data, we assume the validity of a certain
 model for the generating process. In the case of two-way contingency
-tables, the two most common hypotheses are
+tables, the two most common hypotheses \cite{vcd:Agresti:2002} are
 
 \begin{enumerate}
-\item independence of the two variables, and
+\item independence of the two variables
 \item homogeneity of one variable among the strata defined by
@@
 \begin{center}
-\begin{tabular}{|l||rrr|r|}
-\hline
- & Length of stay &&&\\
- & 2--9 & 10--19 & 20+ & TOTAL \\
-\hline\hline
-Visit frequency &&&&\\
+\begin{tabular}{|l|rrr|r|}
+\hline
+ & \multicolumn{4}{l|}{\textbf{Length of stay (in years)}}\\
+\textbf{Visit frequency} & ~~2--9 & ~~10--19 & ~~20+ & $\Sigma$ \\
+\hline
 Regular & 27.24 & 21.14 & 13.62 & 62 \\
@@
 Never & 18.89 & 14.66 & 9.45 & 43 \\\hline
-TOTAL & 58.00 & 45.00 & 29.00 & 132 \\
+$\Sigma$ & 58.00 & 45.00 & 29.00 & 132 \\
 \hline
@@
 It could again be visualized using a mosaic plot, this
-time applied to the table of expected frequencies. In
-Fig.~\ref{fig:sieve1fig}, we can see such a plot in which,
-additionally, a number of squares equal to the corresponding cell entry has
-been added to each tile. We see that the grid formed by the tiles is completely
-regular. The added rectangles in the tiles become more useful when they 
-are not based on the expected but the observed values
-instead (see Fig.~\ref{fig:sieve2fig}). Such a \emph{sieve plot} implicitly compares
+time applied to the table of expected frequencies. 
+If we cross-tabulate each tile to fill it with a number of squares equal to
+the corresponding number of \emph{observed} frequencies, we get a \emph{sieve plot}
+(see Fig.~\ref{fig:sievefig}). It implicitly compares
 expected and observed values since the density of the grid will
@@
 along one of the diagonals.
-In the case of our data, the plot reveals indeed a negative association of 
-the two variables since the density of the rectangles is marked along the secondary
-diagonal, giving evidence to the fact that for these patients,
+In the case of our data, the density of the rectangles is marked along the secondary
+diagonal, indicating a negative association of 
+the two variables. This gives evidence to the fact that for these patients,
 visit frequency decreases with the length of stay.
@@
 \setkeys{Gin}{width=0.7\textwidth}
-\begin{figure}[p]
-\begin{center}
-<<sieve1fig,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-sieve(t(hospital), split = TRUE, sievetype = "expected", pop = FALSE)
-labeling_cells(text = round(independence_table(t(hospital)),2), clip = FALSE, gp = gpar(fontface = 2, fontsize = 15))(t(hospital))
-@ 
-\caption{Expected values for the \data{hospital} data.}
-\label{fig:sieve1fig}
-\end{center}
-\end{figure}
-
-\setkeys{Gin}{width=0.7\textwidth}
-\begin{figure}[p]
-\begin{center}
-<<sieve2fig,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
+\begin{figure}
+\begin{center}
+<<sievefig,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
 sieve(t(hospital), split = TRUE, pop = FALSE, gp = gpar(lty =
@@
 \caption{Sieve plot for the \data{hospital} data.}
-\label{fig:sieve2fig}
+\label{fig:sievefig}
 \end{center}
@@
 
-\noindent that are standardized raw residuals. They are best visualized by
-association plots \cite{vcd:Cohen:1980}:
-each cell is represented by a rectangle that has (signed) height
+\noindent which are standardized raw residuals. In an 
+association plot \cite{vcd:Cohen:1980}, each cell is represented 
+by a rectangle that has (signed) height
 proportional to the corresponding Pearson residual $r_{ij}$ and width proportional to the
 square root of the expected counts $\sqrt{\hat n_{ij}}$. Thus, the area is proportional to
-the raw residuals $n_{ij} - \hat n_{ij}$. In available
-implementations, the sign of a residual is often redundantly
-coded, e.g., by the rectangle's color (see Sect.~\ref{sec:shadings}) %'
-and its position relative to the baseline. Figure \ref{fig:assoc}
+the raw residuals $n_{ij} - \hat n_{ij}$. The sign is visualized by
+its position relative to the baseline (upward tiles for positive, and downward 
+tiles for negative residuals). Figure~\ref{fig:assoc}
 shows the association plot for the hospital data. 
 Consistent with the corresponding mosaic and sieve plots, 
-we clearly see that too many (too less) patients that stay from 
-2 to 9 (more than 20) days get visited regularly
+we clearly see that too many (too few) patients that stay from 
+2 to 9 (more than 19) years get visited regularly
 than would be expected under the null of independence, and that this pattern is 
-reversed for patients never visited.
+reversed for patients visited less than monthly and that get never visited.
 
@@
 
-\section{Residual-based Shadings}
+\subsection{Summary}
+
+Mosaic plots are an instrument to visualize the \emph{observed} frequencies
+of a contingency table by recursive conditional splits. If causal
+relationships between the variables are assumed, the determining variables
+should be used first for splitting since they will dominate the
+plot. Sieve plots basically visualize the table of \emph{expected} frequencies, and
+in addition the deviations from the observed frequencies by the density of a
+grid added to each tile. They are best used to detect dependency
+patterns, especially for ordinal variables. Association plots directly
+visualize Pearson and raw residuals, i.e., standardized and
+non-standardized deviations of observed from expected frequencies,
+respectively. These plots should be used if the diagnostics of
+independence models is of primary concern.
+
+\section{Using colors for Residual-based Shadings}
 \label{sec:extensions}
@@
 will demonstrate how the use of colors can greatly facilitate the
-detection of interesting patterns.
+detection of interesting patterns. Before, we start with some general 
+remarks on colors and color palettes.
+
+\subsection{A Note on Colors and Color Palettes}
+\label{sec:colors}
+
+The plots introduced in the previous section are basically composed of
+tiles whose areas represent characteristics derived from the
+contingency tables---observed and expected frequencies in the case of
+mosaic and sieve plots, or residuals as visualized by association
+plots. When using color for these tiles, it is imperative
+to choose the right color palettes, derived from suitable color
+spaces. Apart from esthetic considerations, wrongly chosen colors
+might serously affect the analysis: For example,
+
+\begin{itemize}
+\item Using high-chroma colors for large
+  areas tend to produce after-image effects which can be
+  distracting \cite{vcd:Ihaka:2003}.
+\item Lighter colors tend to make areas look larger than darker
+  colors, therefore using colors with unequal luminance values  make it
+  difficult to compare area sizes \cite{vcd:Cleveland+McGill:1983}.
+\item Color palettes derived from
+  non-uniform color spaces may contain unbalanced colors with respect to
+  their colorfulness or lightness. When tiles are shaded using such a
+  palette, some of them might appear more important than others in an
+  uncontrolled way.
+\end{itemize}
+
+% gamma correction for displays.
+
+\noindent Implementations are often based on so-called
+Hue-Saturation-Value (HSV) colors (or the comparable Hue-Luminance-Saturation 
+color scheme). Both spaces are rather similar transformations of RGB (Red-Green-Blue) space
+\cite{vcd:Brewer:1999,vcd:Poynton:2000} and are very common implementations of
+colors in many computer packages \cite{vcd:Moretti+Lyons:2002}. Each color in this space is
+represented by three dimensions: the hue (dominant wavelength in the spectrum), 
+the saturation (`colorfulness' or `pureness'), and 
+the value (`brightness', amount of gray). These intuitive dimensions
+make HSV colors easyier to specifiy than, e.g, RGB colors. 
+But HSV colors have several disadvantages.
+Most importantly, HSV colors are not perceptually
+uniform because the three HSV dimensions map only poorly to the three
+perceptual dimensions of the human visual system  \cite{vcd:Brewer:1999,vcd:Ihaka:2003}.
+One important issue here is that HSV dimensions are confounded, e.g., saturation
+is not uniform across different hues.
+As an example, see Fig.~\ref{fig:colors} (left) 
+showing a qualitative color palette in the HSV
+space: although saturation and value are fixed, the fully saturated blue is
+perceived much darker than the fully saturated red or green,
+making it difficult to judge the size of shaded areas.
+Furthermore, the flashy fully saturated HSV colors are hard 
+to look at for a longer time. For similar reasons, it is equally difficult
+to derive acceptable diverging palettes from the HSV space, i.e.,
+bipolar scales containing colors ranging between two very distinct colors.
+The upper part of Fig.~\ref{fig:diverging} shows a diverging
+palette in the HSV space with colors ranging from a saturated red over a neutral white
+to a saturated blue. Although the palette should be balanced with
+respect to colorfulness and lightness, the red colors are perceived
+to be more more intense and flashy than the corresponding blue
+colors.
+
+\setkeys{Gin}{width=\textwidth}
+\begin{figure}
+\begin{center}
+<<colorfig,fig=TRUE,echo=FALSE,height=3,width=7,results=hide>>=
+par(mfrow = c(1,2), mar = c(1,1,1,1), oma = c(0,0,0,0))
+pie(rep(1,9), radius = 1, col = rainbow(9), labels = 360 * 0:8/9)
+pie(rep(1,9), radius = 1, col = rainbow_hcl(9), labels = 360 * 0:8/9)
+@ 
+\caption{Qualitative color palette for the HSV (left) 
+         and HCL (right) spaces.}
+\label{fig:colors}
+\end{center}
+\end{figure}
+
+\setkeys{Gin}{width=0.7\textwidth}
+\begin{figure}
+\begin{center}
+<<diverging,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
+par(mfrow = c(1,1), mar = c(1,1,1,1), oma = c(0,0,0,0))
+plot.new()
+rect(0:4/5, 0.2, 1:5/5, 0.5, border = 0, col = diverge_hcl(5))
+rect(0, 0.2, 1, 0.5, border = 1, col = NULL)
+rect(0:4/5, 0.55, 1:5/5, 0.85, border = 0, col = diverge_hsv(5))
+rect(0, 0.55, 1, 0.85, border = 1, col = NULL)
+text(c(1:5/5 - 0.1), 0.11, c("(260, 100, 50)", "(260, 50, 70)", "(H, 0, 90)", "(0, 50, 70)", "(0, 100, 50)"))
+text(c(1:5/5 - 0.1), 0.91, c("(240, 100, 100)", "(240, 50, 100)", "(H, 0, 100)", "(0, 50, 100)","(0, 100, 100)"))
+@ 
+\caption{Diverging color palettes for the HSV space (upper part) and
+  the HCL space (lower part), ranging from blue over a neutral color to
+  red. The triples indicate the settings for 
+  the three dimensions: (Hue, Saturation, Value)
+  in the upper part, and (Hue, Chroma, Luminance) in the lower part.}
+\label{fig:diverging}
+\end{center}
+\end{figure}
+
+The use of colors that are more `in harmony' go back to Munsell 
+\cite{vcd:Munsell:1905} %'
+who introduced a notation for balanced colors. Based on those,
+tools producing better palettes for specific tasks have been developed 
+\cite{vcd:Harrower+Brewer:2003}. Other perceptually-based color
+spaces, especially suited for computer displays, are the CIELAB and CIELUV
+spaces \cite{vcd:CIE:2004} from which qualitative palettes for
+statistical graphics have been derived \cite{vcd:Ihaka:2003}. A
+transformation of the CIELUV space leads to the HCL (Hue-Chroma-Luminance) space.
+Chroma loosely corresponds to colorfulness and luminance to
+brightness, but in contrast to HSV colors, chroma is an
+\emph{absolute} measure valid for all hues, and luminance can be
+varied independently from the other two dimensions.
+Qualitative color palettes can easily be obtained by holding constant
+chroma and luminance, and dividing the hue range in equally-sized intervals. 
+HCL colors with fixed chroma and luminance 
+are always balanced towards the same grey 
+and thus do not have the problem of varying saturations as the HSV
+colors (see Fig.~\ref{fig:colors}, right). 
+Similarly, diverging color palettes can be derived
+\cite{vcd:Zeileis+Meyer+Hornik:2005} that are
+suited, e.g., for visualizing residuals as we will see in the next
+section. In principle, for each of the two bipolar hues of the
+diverging palette, we could hold constant 
+the luminance and only vary the chroma to get the different colors. 
+However, it turns out that the resulting colors 
+are too similar to be used for discriminating shaded
+objects. Better contrasting effects can be achieved by \emph{simultaneously}
+varying chroma and luminance, for example by choosing two points in
+the chroma/luminance plane for a given hue and interpolating
+linearly. For each half of the diverging palette, 
+one of these points should be a ``neutral'' color,
+and the other correspond to the ending part of the scale. 
+The lower part of Fig.~\ref{fig:diverging} shows a diverging
+palette using HCL colors. In contrast to the HSV counterpart, matching
+colors are balanced to the same gray, and thus get the same perceptual
+``weight''.
 
@@
 
-All plots introduced in the previous section can be improved using colors,
+All plots introduced in Section~\ref{sec:twoway} can be improved using colors,
 either by adding additional information, or by redundantly coding information
@@
 Using color, we can add this information, for example using blue for
-positive, red for negative, and gray for zero residuals. Figure \ref{fig:colorsieve} 
+positive, red for negative, and gray for zero residuals. Figure~\ref{fig:colorsieve} 
 demonstrates the effect of applying such a color shading to a sieve plot for the 
 \data{hospital} data. Color clearly emphasizes the linear association
-pattern in this data.
+pattern in this data: Patients with a LOS from 2--9 years are
+associated with regular visits, whereas patients with a LOS from
+10--19 and more than 19 years are associated with a visit frequency
+of less than monthly and no visits.
 
@@
 sign and absolute size of each residual $r_{ij}$: Cells corresponding to small
-residuals ($|r_{ij}| < 2$) are shaded white. Cells with medium sized residuals
+residuals ($|r_{ij}| < 2$) have no color. Cells with medium sized residuals
 ($2 \le |r_{ij}| < 4$) are shaded light blue and light red for positive and negative
@@
 encompasses a choice of line type and line color of the rectangle borders
-with similar ideas as described above.
+with similar ideas as described above, useful when no color can be used.
 
 In Fig.~\ref{fig:Friendly}, we again show the mosaic for the \data{hospital}
-data, this time using the Friendly color coding.
+data, this time using a Friendly-like color coding (with HCL
+instead of HSV colors, and no line type coding).
 
@@
 <<friendly,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-mosaic(t(hospital), split = TRUE, shade = TRUE, gp = shading_Friendly,
-       legend = legend_fixed, mar = c(left = 3.5),
+mosaic(t(hospital), split = TRUE, shade = TRUE, 
+       mar = c(left = 3.5),
        labeling_args = list(offset_labels = c(left = 0.5), 
@@
 @ 
-\caption{Mosaic display with Friendly color coding of the residuals.}
+\caption{Mosaic display with Friendly-like color coding of the residuals.}
 \label{fig:Friendly}
@@
 \noindent Clearly, the asymmetry for regular and never visited students,
-and the pattern inversion for lengths of stay of 2--9 and more than 20
-days are emphasized using the color shading.
+and the pattern inversion for lengths of stay of 2--9 and more than 19
+years are emphasized using the color shading.
 
@@
 For example, applying the Friendly shading in Fig.~\ref{fig:colorassoc} 
-facilitates the discrimination between
-positive and negative residuals, and therefore to get a quick overview on the
-discussed deviation pattern.
+on the one hand facilitates the discrimination between
+positive and negative residuals, and on the other, more importantly,
+allows an easier comparison of the tiles' sizes. Thus, the shading
+supports the detection of the pattern.
 
@@
 <<assoccolor,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-assoc(t(hospital), split = TRUE, shade = TRUE, keep = TRUE,
-gp = shading_Friendly, legend = legend_fixed)
-@ 
-\caption{Association plot with Friendly color coding of the residuals.}
+assoc(t(hospital), split = TRUE, shade = TRUE, keep = TRUE)
+@ 
+\caption{Association plot with Friendly-like color coding of the residuals.}
 \label{fig:colorassoc}
-\end{center}
-\end{figure}
-
-\subsection{A Note on Color Palettes}
-\label{sec:colors}
-
-The color shading introduced in the previous section but can still be
-improved. One of the issues is the suboptimal color palette used in
-the Friendly shading. Implementations are often based on so-called
-Hue-Saturation-Value (HSV) colors (or the comparable Hue-Luminance-Saturation 
-color scheme). Both spaces are rather similar transformations of RGB (Red-Green-Blue) space
-\cite{vcd:Brewer:1999,vcd:Poynton:2000} and are very common implementations of
-colors in many computer packages \cite{vcd:Moretti+Lyons:2002} making the generation
-of the Friendly shading very simple. Each color in this space is
-represented by three dimensions: the hue, the saturation (`colorfulness'), and 
-the value (`lightness', amount of gray). But HSV colors have several disadvantages.
-Most importantly, HSV colors are not perceptually
-uniform because the three HSV dimensions map only poorly to the three
-perceptual dimensions of the human visual system  \cite{vcd:Brewer:1999,vcd:Ihaka:2003}.
-Consequently, the HSV dimensions are confounded, e.g., saturation
-is not uniform across different hues.
-For example, see Fig.~\ref{fig:colors}, left-hand side, 
-showing a qualitative color palette in the HSV
-space: although saturation and value are fixed, the fully saturated blue is
-perceived much darker than the fully saturated red or green. 
-This makes it more difficult for the human eye to judge the size of shaded areas
-and can therefore lead to color-caused optical illusions when used in statistical graphs
-\cite{vcd:Cleveland+McGill:1983}. Furthermore, flashy fully saturated HSV colors are
-good for drawing attention to a plot, 
-but hard to look at for a longer time \cite{vcd:Ihaka:2003}
-which makes HSV-shaded graphics harder to interpret.
-
-\setkeys{Gin}{width=\textwidth}
-\begin{figure}
-\begin{center}
-<<colorfig,fig=TRUE,echo=FALSE,height=3,width=7,results=hide>>=
-par(mfrow = c(1,2), mar = c(1,1,1,1), oma = c(0,0,0,0))
-pie(rep(1,9), radius = 1, col = hsv(seq(0,1,length=10))[-1])
-pie(rep(1,9), radius = 1, col = hcl(seq(0,360,length=10))[-1])
-@ 
-\caption{Qualitative color palette for the HSV (left hand side) 
-         and HCL (right hand side) spaces.}
-\label{fig:colors}
-\end{center}
-\end{figure}
-\setkeys{Gin}{width=0.7\textwidth}
-
-The use of colors that are more `in harmony' go back to Munsell 
-\cite{vcd:Munsell:1905} %'
-who introduced a color notation for balanced colors. Based on those,
-tools producing better palettes for specific tasks have been developed 
-\cite{vcd:Harrower+Brewer:2003}. Other perceptually-based color
-spaces, especially suited for displays, are the CIELAB and CIELUV
-spaces \cite{vcd:CIE:2004} from which qualitative palettes for
-statistical graphics have been derived \cite{vcd:Ihaka:2003}. A
-transformation of the CIELUV space leads to the HCL (Hue-Chroma-Luminance) space.
-HCL colors with fixed chroma and luminance are always balanced towards the same grey 
-and thus do not have the problem of varying saturations as the HSV
-colors (see Fig.~\ref{fig:colors}, right hand side). 
-Similarly, diverging color palettes can be derived that are
-suited for visualizing residuals in a similar way than the Friendly
-shading \cite{vcd:Zeileis+Meyer+Hornik:2005}.
-Figure \ref{fig:hclmosaic} shows again the mosaic plot for the \data{hospital} data, 
-this time using HCL colors and an alternative legend, showing the whole range of residuals.
-
-\setkeys{Gin}{width=0.7\textwidth}
-\begin{figure}
-\begin{center}
-<<hclmosaicfig,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-mosaic(t(hospital), split = TRUE, mar = c(left = 3.5), shade = TRUE,
-       labeling_args = list(offset_labels = c(left = 0.5), 
-         offset_varnames = c(left = 1, top = 0.5), set_labels = 
-         list("Visit frequency" = c("Regular","Less than\nmonthly","Never"))))
-@ 
-\caption{Mosaic display with HCL shading of the residuals.}
-\label{fig:hclmosaic}
 \end{center}
@@
 
-Figure \ref{fig:hclmosaic} includes the $p$ value of a $\chi^2$ test
+Figure~\ref{fig:Friendly} includes the $p$ value of a $\chi^2$ test
 of independence, frequently used to assess the significance of the
@@
 Since the HCL space is three-dimensional and we only
-used two dimensions so far for coding information (hue for the sign and
-colorfulness for the size of the residuals), we can use the third dimension
-(lightness) for visualizing the significance of some specified test statistic,
-for example the $\chi^2$ test statistic, 
-using darker (`uninteresting') colors for non-significant results.
-The following example using the \data{Bundesliga} data \cite{vcd:Knorr-Held:1999}
-shows the relationship of home goals and away
-goals of Germany's premier soccer league in 1995: although there are two
-``larger'' residuals (one greater than 2, one less then $-2$), the
-$\chi^2$ test does not reject the null hypothesis of
-independence on a significance level of $95\%$ since the $p$ value is $0.12133$. 
-Consequently, the employed shading uses a darker overall color scheme
-(see Fig.~\ref{fig:bundesliga}).
-
-\begin{figure}[p]
-\begin{center}
-<<bundesligafig,fig=TRUE,echo=FALSE,height=6,width=7>>=
-bl <- xtabs(~ HomeGoals + AwayGoals, data = Bundesliga, subset = Year == 1995)
-mosaic(bl, shade = TRUE)
-@
-\caption{Mosaic plot for part of the \data{Bundesliga}, using a $\chi^2$ test and
-fixed cut-off points for the residuals.}
-\label{fig:bundesliga}
-\end{center}
-\end{figure}
-
-The previous example shows that the heuristic for choosing the cut-off points 
-in the Friendly shading may lead to wrong conclusions: the test of
-independence is not significant even though some of the residuals are
-``large''. The reason here is that the cut-off points are really
-data-dependent. Another example where the Friendly-shading may trap
-the rush statistician is the case of the \data{arthritis} data 
+used two `degrees of freedom' so far for coding information %'
+(hue for the sign and a linear combination of chroma and luminance 
+for the size of the residuals), we can add a third information to the plot.
+For example, we can visualize the significance of some specified test statistic
+(e.g., the $\chi^2$ test statistic) 
+using darker (`uninteresting') colors for non-significant results by
+restricting the luminance range used for interpolating the colors.
+
+% The following example using the \data{Bundesliga} data \cite{vcd:Knorr-Held:1999}
+% shows the relationship of home goals and away
+% goals of Germany's premier soccer league in 1995: although there are two
+% ``larger'' residuals (one greater than 2, one less then $-2$), the
+% $\chi^2$ test does not reject the null hypothesis of
+% independence on a significance level of $95\%$ since the $p$ value is $0.1213$. 
+% Consequently, the employed shading uses a darker overall color scheme
+% (see Fig.~\ref{fig:bundesliga}).
+
+% \begin{figure}[p]
+% \begin{center}
+% <<bundesligafig,fig=TRUE,echo=FALSE,height=6,width=7>>=
+% bl <- xtabs(~ HomeGoals + AwayGoals, data = Bundesliga, subset = Year == 1995)
+% mosaic(bl, shade = TRUE)
+% @
+% \caption{Mosaic plot for part of the \data{Bundesliga}, using a $\chi^2$ test and
+% fixed cut-off points for the residuals.}
+% \label{fig:bundesliga}
+% \end{center}
+% \end{figure}
+
+The heuristic for choosing the cut-off points 
+in the Friendly shading may lead to wrong conclusions: especially in
+large tables, the test of
+independence may not be significant even though some of the residuals are
+``large''. On the the other hand, the test might be significant even
+though the residuals are ``small''. In fact, the cut-off points are really
+data-dependent. Consider the case of the \data{arthritis} data 
 \cite{vcd:Koch+Edwards:1988}, resulting from a double-blind clinical trial
-investigating a new treatment for rheumatoid arthritis, stratified by gender.
-Figure \ref{fig:arthritis} visualizes the results for the female
+investigating a new treatment for rheumatoid arthritis, stratified by
+gender (see Tab.~\ref{tab:arthritis}):
+
+\begin{table}[h]
+\begin{center}
+\begin{tabular}{|l|rrr|r|}
+\hline
+ & \multicolumn{4}{c|}{\textbf{Improved}}\\
+\textbf{Treatment} & None & Some & Marked & $\Sigma$ \\
+\hline
+Placebo & 19 & 7 & 6 & 32 \\
+Treatment & 6 & 5 & 16 & 27 \\\hline
+$\Sigma$ & 25 & 12 & 22 & 59 \\
+\hline
+\end{tabular}
+\caption{The \data{arthritis} data (female patients).}
+\label{tab:arthritis}
+\end{center}
+\end{table}
+
+\noindent Figure~\ref{fig:arthritis} visualizes the results for the female
 patients, again by means of a mosaic display. Clearly, the hypothesis
@@
 evidence for the rejection of the independence hypothesis can easily be identified.
-The distribution of this test statistic under the null can be obtained
+The conditional distribution of this test statistic under the null can be obtained
 by simulation, sampling tables with the same row and column sums $n_{i+}$ and $n_{+j}$
@@
 and computing the maximum statistic for each of these tables.
-In Figs.~\ref{fig:bundesliga2} and \ref{fig:arthritis2}, we again
-visualize the \data{Bundesliga} and the \data{arthritis} data, this
-time using the maximum test statistic. We see that for the
-\data{Bundesliga} data no cell is colored, and we now can indeed conclude
-from the visualization that neither of the home and away teams seems advantaged. 
-On the other hand, for the \data{arthritis} data, the shading of the
-tiles clearly shows that the treatment is effective: too many patients
+In Fig.~\ref{fig:arthritis2}, we again
+visualize the \data{arthritis} data, this
+time using the maximum test statistic. The shading of the
+tiles now clearly shows that the treatment is effective: significantly more patients
 in the treatment group exhibit marked improvement than would be
 expected under independence.
-
-\begin{figure}[p]
-\begin{center}
-<<bundesliga2fig,fig=TRUE,echo=FALSE,height=6,width=7>>=
-bl <- xtabs(~ HomeGoals + AwayGoals, data = Bundesliga, subset = Year == 1995)
-mosaic(bl, gp = shading_max)
-@
-\caption{Mosaic plot for part of the \data{Bundesliga}, using the maximum statistic
-and data-driven cut-off points for the residuals.}
-\label{fig:bundesliga2}
-\end{center}
-\end{figure}
 
@@
 
-\section{Multi-way Tables}
+\section{Selected Methods for Multi-way Tables}
 \label{sec:multiway}
@@
 
-\subsection{Visualize Flat-Tables}
+\subsection{Visualizing Flat-Tables}
 \label{sec:flat}
@@
 and hair color---nested in gender---in the rows, is given by Tab.~\ref{tab:hec}.
+The corresponding mosaic, sieve, and association plots are shown in
+Figs.~\ref{fig:hecmosaic}--\ref{fig:hecassoc}. 
+The shadings in these plots visualize
+the model of hair and eye color being \emph{conditional} independent,
+given gender. As we can see from the shading (many tiles are colored) 
+and the $p$ value for the test statistic, this model fits only poorly.
 
@@
 \begin{center}
-\begin{tabular}{|ll||rrrr|}
-\hline
- & & EYE &&&\\
- & & Brown & Blue & Hazel & Green\\\hline\hline
-GENDER & HAIR &&&&\\
+\begin{tabular}{|ll|rrrr|}
+\hline
+ & & \multicolumn{4}{c|}{\textbf{Eye}} \\
+\textbf{Gender} & \textbf{Hair} & Brown & Blue & Hazel & Green\\\hline
 Male & Black & 32 & 11 & 10 & 3 \\
@@
 
-\noindent The corresponding mosaic, sieve, and association plots are shown in
-Figs.~\ref{fig:hecmosaic}--\ref{fig:hecassoc}. Note that with three variables, we have the
-choice of four kinds of independence models that can be fitted to the table:
-partial, joint, and conditional independence, as well as 
-the model of no three-way interaction. For demonstration purposes,
-the shadings in these plots visualize
-the model of hair and eye color being \emph{jointly} independent from gender.
-The model fits better than others, but still is significant at the
-$95\%$ level.
-
-\setkeys{Gin}{width=0.7\textwidth}
-\begin{figure}
-\begin{center}
-<<hecmosaic,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-hec = structable(Eye ~ Sex + Hair, data = HairEyeColor)
-mosaic(hec, labeling_args = list(tl_labels = c(Hair = TRUE), abbreviate = c(Hair = 4)), 
-       mar = c(left = 4), shade = TRUE, expected = ~ Hair * Eye + Sex)
+\setkeys{Gin}{width=0.7\textwidth}
+\begin{figure}
+\begin{center}
+<<hecmosaic,fig=TRUE,echo=FALSE,height=12,width=7,results=hide>>=
+mosaic(~ Hair + Eye | Sex, data = HairEyeColor,
+       labeling_args = list(tl_labels = c(Hair = TRUE), set_varnames = c(Sex = "Gender")), 
+       legend_args = list(y = 0.3, height = 0.4),
+       mar = c(left = 4), shade = TRUE, keep = FALSE)
 @
@@
 \begin{center}
-<<hecsieve,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
+<<hecsieve,fig=TRUE,echo=FALSE,height=12,width=7,results=hide>>=
 hec = structable(Eye ~ Sex + Hair, data = HairEyeColor)
-sieve(hec, labeling_args = list(tl_labels = c(Hair = TRUE), abbreviate = c(Hair = 4)), 
-       mar = c(left = 4), shade = TRUE, expected = ~ Hair * Eye + Sex)
+sieve(hec, labeling_args = list(tl_labels = c(Hair = TRUE), abbreviate = c(Hair = 4),
+set_varnames = c(Sex = "Gender")), spacing = spacing_conditional, condvars = 1,
+      legend_args = list(y = 0.3, height = 0.4),
+      mar = c(left = 4), shade = TRUE, expected = ~ Hair * Sex + Eye * Sex, keep = FALSE)
 @
@@
 \begin{center}
-<<hecassoc,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
+<<hecassoc,fig=TRUE,echo=FALSE,height=12,width=7,results=hide>>=
 hec = structable(Eye ~ Sex + Hair, data = HairEyeColor)
-assoc(hec, labeling_args = list(tl_labels = c(Hair = TRUE), abbreviate = c(Hair = 4)), 
-      mar = c(left = 4), shade = TRUE, expected = ~ Hair * Eye + Sex)
+assoc(hec, labeling_args = list(tl_labels = c(Hair = TRUE), abbreviate = c(Hair = 4),
+             set_varnames = c(Sex = "Gender")), 
+      legend_args = list(y = 0.3, height = 0.4),
+      spacing = spacing_conditional(sp = 0, start = 4),
+      mar = c(left = 4), shade = TRUE, expected = ~ Hair * Sex + Eye * Sex, keep = FALSE)
 @
@@
 
-Apart from the generic approach to the visualization of multi-way tables outlined in 
-the previous section, there are modifications 
-and specialized displays designed for the
-visualization of conditional independence structures. A first approach
-is to modify the spacing of the standard mosaic display---which already 
-is a conditional plot by construction---to better distinguish conditional
-from conditioned variables. Figure \ref{fig:presex}
-illustrates the conditional independence of premarital and
-extramarital sex, given gender and marital status, in the \data{PreSex} 
-data \cite{vcd:thornes+collard:1979,vcd:gilbert:1981}. The $\chi^2$
-test of independence rejects the null hypothesis: possibly, because
-too many men who had premarital sex also had extramarital sex? For the conditioned
-variables premarital and extramarital sex, we use a fixed spacing, whereas for 
-the conditioning variables gender and marital status, the spacing is increasing.
-
-\setkeys{Gin}{width=0.7\textwidth}
-\begin{figure}
-\begin{center}
-<<PreSex,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-mosaic(PreSex, condvars = c(1, 4), shade = TRUE)
-@
-\caption{Mosaic plot for the pre- and extramarital sex data.}
-\label{fig:presex}
-\end{center}
-\end{figure}
-
-Another possibility is to use trellis-like layouts for visualizing partial tables for 
-given strata, defined by the conditioning variables.
-Figure \ref{fig:ucbadmissions} visualizes the well-known UCB admissions data
-\cite{vcd:Bickel+Hammel+O'Connell:1975} by the means of a conditional 
-association plot. The panels show the
-residuals from a conditional independence model (independence of gender
-and admission, given department), stratified by department. Clearly,
-the situation in department A (more women/less men accepted than
-would be expected under the null) causes the rejection of the
-hypothesis of conditional independence, indicating that female
-students are more successful in that department than their male colleagues.
-
-\setkeys{Gin}{width=\textwidth}
-\begin{figure}
-\begin{center}
-<<UCBAdmissions,fig=TRUE,echo=FALSE,height=6,width=10,results=hide>>=
-cotabplot(UCBAdmissions, panel = cotab_assoc, shade = TRUE, legend = FALSE)
-@
-\caption{Conditional association plot for the UCB admissions data.}
-\label{fig:ucbadmissions}
-\end{center}
-\end{figure}
-
-Finally, it is possible for all basic plots to create pairwise displays,
+In addition to the generic approach to the visualization of multi-way tables outlined in 
+the previous section, there are specialized displays 
+designed for the visualization of conditional independence structures.
+
+A first step in the analysis of more complex tables is to find a
+suitable model that fits the data. All basic plots can be combined in pairwise displays,
 arranged in in a matrix similar to scatterplots in a pairs plot.
@@
 variables $i$ and $j$ that can be used to specify the model visualized
-in that cell. Typical hypotheses are: Variables $i$ ad $j$ are
+in that cell. Typical hypotheses are: Variables $i$ and $j$ are
 marginally independent; variables $i$ and $j$ are conditionally
@@
 independent from all others, etc.
-Figure \ref{fig:pairs} shows a pairs display with mosaic plots visualizing mutual 
+Figure~\ref{fig:pairs} shows a pairs display with mosaic plots visualizing mutual 
 independence in the upper triangle, 
-sieve plots for the same in the lower triangle, and bar charts in the diagonal.
-Immediately, we can detect the independence of hair and gender and eye and gender from the
-non-colored cells in the corresponding mosaic plots. Thus, the associational pattern
-present in the sieve plots are not significant. On the other hand, hair and eye color 
-clearly are not independent.
+mosaics for \emph{joint} independence 
+in the lower triangle, and bar charts in the diagonal.
+In the upper part of the matrix, we can immediately 
+detect the independence of hair and gender, and eye and gender from the
+non-colored cells in the corresponding mosaic plots. On the other hand, hair and eye color 
+clearly are associated. In the lower part of the matrix, the mosaic
+at the intersection of hair and eye is the less colored one. Its shading
+represents the model of hair and eye color being jointly independent from
+gender, i.e., their association being homogeneous over the gender
+strata. It is the model that fits best, although still rejected at the
+95\% level ($p=0.0161$): possibly, because too many brown-haired male
+students have blue eyes, and too few of them have brown eyes than expected under
+the null model.
 
@@
 <<pairs,fig=TRUE,echo=FALSE,height=8,width=8,results=hide>>=
-pairs(t(hec), lower_panel = pairs_sieve, space = 0.3, shade = TRUE,
+pairs(hec, lower_panel = pairs_mosaic(type = "joint"), space = 0.3, shade = TRUE,
       diag_panel_args = list(rot = -45, just_leveltext = c("left","bottom")))
@@
 \label{fig:pairs}
+\end{center}
+\end{figure}
+
+% As an example, consider the \data{PreSex} data
+% \cite{vcd:thornes+collard:1979,vcd:gilbert:1981} on
+% pre- and extra-marital sex and divorce: 494 divorced and 542 married
+% people (willing to divorce) were asked two questions about their pre- and
+% extramarital sexual experiences, resulting in Tab.~\ref{tab:presex}.
+
+% \begin{table}[ht]
+% \begin{center}
+% \begin{tabular}{|ll|rrrrr|}
+% \hline
+%  & & \textbf{Marital Status} & \multicolumn{2}{l}{Divorced} & \multicolumn{2}{l|}{Married}  \\
+% \textbf{Gender} & \textbf{Extramarital Sex} & \textbf{Premarital Sex} & Yes & No & Yes & No\\\hline
+% Woman & Yes && 17 & 36 &  4 & 4 \\
+%       & No  && 54 &214 & 25 &322 \\
+% Men   & Yes && 28 & 17 & 11 & 4 \\
+%       & No  && 60 & 68 & 42 &130 \\
+% \hline
+% \end{tabular}
+% \caption{The \data{PreSex} data, in flat representation.}
+% \label{tab:presex}
+% \end{center}
+% \end{table}
+
+% \noindent Figure~\ref{fig:presex}
+% illustrates the conditional independence of premarital and
+% extramarital sex, given gender and marital status. The $\chi^2$
+% test of independence rejects the null hypothesis: possibly, because
+% too many men who had premarital sex also had extramarital sex? For the conditioned
+% variables premarital and extramarital sex, we use a fixed spacing, whereas for 
+% the conditioning variables gender and marital status, the spacing is increasing.
+
+% \setkeys{Gin}{width=0.7\textwidth}
+% \begin{figure}
+% \begin{center}
+% <<PreSex,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
+% mosaic(PreSex, condvars = c(1, 4), shade = TRUE)
+% @
+% \caption{Mosaic plot for the pre- and extramarital sex data.}
+% \label{fig:presex}
+% \end{center}
+% \end{figure}
+
+Since mosaic displays are conditional plots by construction, we can use them
+for stratified data by using the conditioning variables as first splitting variables. 
+The mosaic in Fig.~\ref{fig:hecmosaic} shows the distribution of hair
+and eye color for both male and female students. A larger spacing for
+the gender variable simplifies the distinction between the two
+groups. However, using a single plot, the strata are not corrected for the
+marginal distribution of the conditioning variable(s). Especially for
+very unevenly distributed marginals, the strata can become very distorted.
+Another possibility is to use trellis-like layouts for visualizing \emph{partial}
+tables for given strata, defined by the conditioning variables. 
+All subtables are corrected for the marginals and thus, all
+corresponding mosaics in the panels have the same size. 
+In addition, trellis layouts help reducing the complexity of bigger tables. 
+As an example, consider the well-known UCB admissions data 
+\cite{vcd:Bickel+Hammel+O'Connell:1975}
+on applicants to graduate school at the University of California in Berkeley for
+the six largest departments in 1973 classified by admission and
+gender (see Tab.~\ref{tab:ucb}). The table aggregated over all
+departments gives wrong evidence of gender bias, 
+i.e., higher admission rates for male students. Figure~\ref{fig:ucbadmissions} 
+visualizes the data by means of a conditional 
+mosaic plot. Each panel corresponds to a department, and contains
+a mosaic corresponding to the partial (fourfold) table of admission
+and gender. Accordingly, the shading visualizes the
+residuals from the corresponding conditional independence model (independence of gender
+and admission, given department), stratified by department. Clearly,
+the situation in department A (more women/less men accepted than
+would be expected under the null) causes the rejection of the
+hypothesis of conditional independence. The overall deviation pattern indicates that 
+in contrast to the impression given by the aggregated table, male
+students were not advantaged, and that in department A,
+\emph{female} students were even more successful than their male colleagues.
+
+\begin{table}[ht]
+\begin{center}
+\begin{tabular}{|ll|rrrrrr|}
+\hline
+ & & \multicolumn{6}{c|}{\textbf{Department}} \\
+\textbf{Gender} & \textbf{Admit} & A & B & C & D & E & F\\\hline
+Male   & Admitted & 512 & 353 & 120 & 138 & 53 & 22\\
+       & Rejected & 313 & 207 & 205 & 279 & 138 & 351\\
+Female & Admitted & 89 & 17 & 202 & 131 & 94 & 24 \\
+       & Rejected & 19 & 8 & 391 & 244 & 299 & 317 \\
+\hline
+\end{tabular}
+\caption{The \data{UCB admissions} data, in flat representation.}
+\label{tab:ucb}
+\end{center}
+\end{table}
+
+\setkeys{Gin}{width=\textwidth}
+\begin{figure}
+\begin{center}
+<<UCBAdmissions,fig=TRUE,echo=FALSE,height=6,width=10,results=hide>>=
+cotabplot(UCBAdmissions, panel = cotab_mosaic, shade = TRUE, legend = FALSE)
+@
+\caption{Conditional mosaic plot for the UCB admissions data.}
+\label{fig:ucbadmissions}
 \end{center}
@@
 expected values and the Pearson residuals, respectively. This basic
-methods are enhanced by using residual-based shadings, preferably
+methods can be enhanced by using residual-based shadings, preferably
 based on perceptual color palettes such as those derived from the HCL

### 2005-11-17 trunk/vcd/paper/hsc.Rnw
Further improvements
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@499 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
+%% Hyperref
+\hypersetup{%
+  pdftitle = {Residual-based Shadings for Visualizing (Conditional) Independence},
+  pdfsubject = {Residual-based Shadings},
+  pdfkeywords = {association plots, conditional inference, contingency tables, HCL colors, HSV colors, mosaic plots},
+  pdfauthor = {Achim Zeileis, David Meyer, Kurt Hornik},
+  %% change colorlinks to false for pretty printing
+  colorlinks = {false},
+  linkcolor = {Blue},
+  citecolor = {Blue},
+  urlcolor = {Red},
+  hyperindex = {true},
+  linktocpage = {true},
+}
+
 <<preliminaries,echo=FALSE,results=hide>>=
-set.seed(1071)
+myseed <- 4711
+set.seed(myseed)
 library(vcd)
@@
 data(Arthritis)
-data(Titanic)
-data(PreSex)
 data(UCBAdmissions)
@@
 art <- xtabs(~Treatment + Improved, data = Arthritis, subset = Sex == "Female")
+
+set.seed(myseed)
+art_max <- coindep_test(art, n = 5000)
+set.seed(myseed)
+art_chisq <- coindep_test(art, n = 5000, indepfun = function(x) sum(x^2))
+
+set.seed(myseed)
+hos_chisq <- coindep_test(hospital, n = 5000, indepfun = function(x) sum(x^2))
+
+set.seed(myseed)
+hec_max <- coindep_test(HairEyeColor, "Sex", n = 5000)
+
 @
@@
 also introduce specialized displays for conditional independence
-structures. Sect.~\ref{sec:concl} concludes the chapter.
+structures. Section~\ref{sec:concl} concludes the chapter.
 
@@
 significantly different in the ``less than monthly'' and ``never''
-strata. From the row-standardized table (see Tab~.\ref{tab:rowhospital}), 
+strata. From the row-standardized table (see Tab.~\ref{tab:rowhospital}), 
 it seems indeed that LOS were homogeneous with respect to these two visit frequency strata.
@@
 years, and conversely, less patients get regular visits for stays for more than 19 years.
-For patients that get no visits, the pattern is inversed.
+For patients that get no visits, the pattern is reversed.
 
@@
 sieve(t(hospital), split = TRUE, pop = FALSE, gp = gpar(lty =
-"dotted", col = "dark gray"))
+"dotted", col = "black"))
 labeling_cells(text = t(hospital), clip = FALSE, gp = gpar(fontface = 2, fontsize = 15))(t(hospital))
@@
 
-\section{Using colors for Residual-based Shadings}
+\section{Using Colors for Residual-based Shadings}
 \label{sec:extensions}
@@
 spaces. Apart from esthetic considerations, wrongly chosen colors
-might serously affect the analysis: For example,
+might serously affect the analysis. For example,
 
@@
 \item Lighter colors tend to make areas look larger than darker
-  colors, therefore using colors with unequal luminance values  make it
+  colors, therefore using colors with unequal luminance values  makes it
   difficult to compare area sizes \cite{vcd:Cleveland+McGill:1983}.
@@
   non-uniform color spaces may contain unbalanced colors with respect to
-  their colorfulness or lightness. When tiles are shaded using such a
+  their colorfulness or brightness. When tiles are shaded using such a
   palette, some of them might appear more important than others in an
@@
 
-% gamma correction for displays.
-
-\noindent Implementations are often based on so-called
+\noindent 
+Due to the three-dimensional nature of human color
+perception \cite{vcd:Mollon:1995}, 
+it has been common to specify colors using the three primaries
+red, green, and blue (RGB colors), especially for computer devices. 
+The appearance of the on-screen colors is affected by the device
+characteristics: For example, the intensity $I$ of a primary 
+on a particular device follows the rule $I=L^\gamma$, with $L$
+the value for the primary color, and $\gamma$ device-dependent 
+(but typically close to 2.2). Therefore, a first caveat is that if colors are to
+appear identical on different devices, their gamma characteristics
+have to be taken into account.
+
+Now, software implementations are often based on the more convenient
 Hue-Saturation-Value (HSV) colors (or the comparable Hue-Luminance-Saturation 
-color scheme). Both spaces are rather similar transformations of RGB (Red-Green-Blue) space
+color scheme). Both spaces are rather similar transformations of the RGB space
 \cite{vcd:Brewer:1999,vcd:Poynton:2000} and are very common implementations of
@@
 to a saturated blue. Although the palette should be balanced with
-respect to colorfulness and lightness, the red colors are perceived
+respect to colorfulness and brightness, the red colors are perceived
 to be more more intense and flashy than the corresponding blue
@@
 are too similar to be used for discriminating shaded
-objects. Better contrasting effects can be achieved by \emph{simultaneously}
+objects. Better contrasting can be achieved by \emph{simultaneously}
 varying chroma and luminance, for example by choosing two points in
@@
 palette using HCL colors. In contrast to the HSV counterpart, matching
-colors are balanced to the same gray, and thus get the same perceptual
-``weight''.
-
-\subsection{Using Colors for Residuals}
+colors (light red/blue and dark red/blue) 
+are balanced to the same gray, and thus get the same perceptual ``weight''.
+
+\subsection{Highlighting and Color-based Shadings}
 \label{sec:shadings}
 
-All plots introduced in Section~\ref{sec:twoway} can be improved using colors,
+All plots introduced in Sect.~\ref{sec:twoway} are composed of empty
+tiles. It seems intuitive to used filled tiles to highlight information
+of interest. Consider again the hospital example: in Fig.~\ref{fig:highlight},
+we mark tiles for patients never or seldom visited to visualize their proportion in
+the LOS strata. For optical clarity, we set the
+spacing between the visit frequency tiles to 0. Clearly, the
+proportion of these patients increases with LOS. In fact, such a
+``stacked'' plot can also be interpreted as spine plot with highlighting
+\cite{vcd:hummel:1996}, particularly useful in analyzing causal relationships in
+categorical data with a binary dependent variable. More variations on
+this theme, such as doubledecker plots, are treated in chapter XXX
+(\textbf{REFERENCE TO OTHER CHAPTER(S)}).
+
+\setkeys{Gin}{width=0.7\textwidth}
+\begin{figure}
+\begin{center}
+<<highlight,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
+mosaic(t(hospital), 
+       mar = c(left = 3.5), 
+       labeling_args = list(offset_labels = c(left = 0.5), 
+         offset_varnames = c(left = 1, top = 0.5), set_labels = 
+         list("Visit frequency" = c("Regular","Less than\nmonthly","Never"))),
+       split = TRUE, spacing = spacing_highlighting, 
+       gp = gpar(fill = c("gray","black","black")))
+@ 
+\caption{Spine plot with highlighting for the hospital data.}
+\label{fig:highlight}
+\end{center}
+\end{figure}
+
+Using colors, even more complementary information can be visualized,
 either by adding additional information, or by redundantly coding information
 already visualized by the `raw' plot to support our perceptual system. %'
-
 First, we consider the sieve plots. The density 
@@
 with a fully saturated blue and red, respectively. 
-The heuristic for choosing the cut offs $2$ and $4$ is that the Pearson residuals
+The heuristic for choosing the cut-offs $2$ and $4$ is that the Pearson residuals
 are asymptotically standard normal which implies that the highlighted cells are those with
@@
 In Fig.~\ref{fig:Friendly}, we again show the mosaic for the \data{hospital}
-data, this time using a Friendly-like color coding (with HCL
+data, this time using a Friendly-like color shading (with HCL
 instead of HSV colors, and no line type coding).
@@
        mar = c(left = 3.5), 
+       gp_args = list(p.value = hos_chisq$p.value),
        labeling_args = list(offset_labels = c(left = 0.5), 
@@
 <<assoccolor,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-assoc(t(hospital), split = TRUE, shade = TRUE, keep = TRUE)
+assoc(t(hospital), split = TRUE, shade = TRUE, keep = TRUE,
+      gp_args = list(p.value = hos_chisq$p.value))
 @ 
@@
 
-Figure~\ref{fig:Friendly} includes the $p$ value of a $\chi^2$ test
+Figures~\ref{fig:Friendly} and \ref{fig:colorassoc} 
+include the (same) $p$ value of a $\chi^2$ test
 of independence, frequently used to assess the significance of the
@@
 degrees of freedom under the null hypothesis.
+An important reason for using the unconditional limiting distribution
+for the $X^2$ statistic from Equ.~\ref{eq:ChiSquare} was the closed form result
+for the distribution. Recently, with the improving perfomance of computers,
+conditional inference (or permutation tests)---carried out either by simulation
+or by computation of the (asymptotic) permutation distribution---have been receiving
+increasing attention \cite{vcd:Ernst:2004,vcd:Pesarin:2001,vcd:Strasser+Weber:1999}. 
+For testing the independence hypothesis from Equation~\ref{eq:H0},
+using a permutation test is particularly intuitive due to the 
+permutation invariance (given row and column sums) of this problem.
+Consequently, all results in this paper are based on conditional inference performed
+by simulating the permutation distribution of test statistics of type
+$\lambda([r_{ij}])$.
+
 Since the HCL space is three-dimensional and we only
@@
 (e.g., the $\chi^2$ test statistic) 
-using darker (`uninteresting') colors for non-significant results by
-restricting the luminance range used for interpolating the colors.
+using darker (`uninteresting') colors for non-significant
+results. These can again be derived using the interpolation procedure
+described in the Sect.~\ref{sec:colors} by
+restricting the luminance to a shorter range, ending with a ``darker'' value.
 
@@
 investigating a new treatment for rheumatoid arthritis, stratified by
-gender (see Tab.~\ref{tab:arthritis}):
+gender (see Tab.~\ref{tab:arthritis} for the female patients):
 
@@
 patients, again by means of a mosaic display. Clearly, the hypothesis
-of independence is rejected by the $\chi^2$ test even on a $99\%$ level ($p=0.0035$), but
-since all residuals are in the $[-1.72,1.87]$ interval, the tiles
-remain uncolored. A solution to this issue is to use a different
+of independence is rejected by the $\chi^2$ test even on a $99\%$ level 
+($p=\Sexpr{round(art_chisq$p.value,4)}$), but
+since all residuals are in the 
+$[\Sexpr{round(min(chisq.test(art)$residuals),4)},\Sexpr{round(max(chisq.test(art)$residuals),4)}]$ %$
+interval, the tiles remain uncolored. A solution to this issue is to use a different
 test statistic, for example the maximum of the absolute values of the
-Pearson residuals \cite{vcd:Meyer+Zeileis+Hornik:2003} instead of the
-sum of squares:
+Pearson residuals \cite{vcd:Meyer+Zeileis+Hornik:2003} instead of the sum of squares:
 
@@
 evidence for the rejection of the independence hypothesis can easily be identified.
-The conditional distribution of this test statistic under the null can be obtained
+As explained above, 
+the conditional distribution of this test statistic under the null can be obtained
 by simulation, sampling tables with the same row and column sums $n_{i+}$ and $n_{+j}$
-using the Patefield algorithm \cite{vcd:Patefield:1981} 
+using, e.g., the Patefield algorithm \cite{vcd:Patefield:1981} 
 and computing the maximum statistic for each of these tables.
@@
 <<Arthritis,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-mosaic(art, shade = TRUE)
+mosaic(art, shade = TRUE, gp_args = list(p.value = art_chisq$p.value))
 @
@@
 \end{figure}
+
+\subsection{Summary}
+
+Especially for the visualization of areas, 
+use perceptually uniform colors and color palettes, derived from,
+e.g., the HCL color space. Shadings can be used to add 
+information to the basic plots and to support the
+analysis. Highlighting of tiles can support the analysis of causal relationships.
+Residual-based shadings can be used for model diagnostics, e.g.,
+by using the hue to code the sign of a residual, and a linear
+combination of chroma and luminance for the size. In addition, the
+significance of test statistics can be visualized through overall 
+darker color palettes. Using the maximum instead of the $\chi^2$ test
+statistic to aggregate the residuals allows precise diagnostics when
+the null hypothesis of independence is rejected.
 
@@
 mosaic(~ Hair + Eye | Sex, data = HairEyeColor,
+       gp_args = list(p.value = hec_max$p.value, interpolate = hec_max$qdist(c(0.9,0.99))),
        labeling_args = list(tl_labels = c(Hair = TRUE), set_varnames = c(Sex = "Gender")), 
@@
              set_varnames = c(Sex = "Gender")), 
+      gp_args = list(p.value = hec_max$p.value, interpolate = hec_max$qdist(c(0.9,0.99))),
       legend_args = list(y = 0.3, height = 0.4),
@@
 strata. It is the model that fits best, although still rejected at the
-95\% level ($p=0.0161$): possibly, because too many brown-haired male
-students have blue eyes, and too few of them have brown eyes than expected under
-the null model.
+95\% level 
+($p=\Sexpr{round(summary(loglm(~ Hair * Eye + Sex, data = HairEyeColor))$tests[2,3],4)}$): %$
+possibly, because too many brown-haired male
+students have blue eyes, and too few of them have brown eyes, than expected under
+the null model?
 
@@
 and eye color for both male and female students. A larger spacing for
-the gender variable simplifies the distinction between the two
+the gender variable has been used to simplify the distinction between the two
 groups. However, using a single plot, the strata are not corrected for the
 marginal distribution of the conditioning variable(s). Especially for
-very unevenly distributed marginals, the strata can become very distorted.
-Another possibility is to use trellis-like layouts for visualizing \emph{partial}
+very unevenly distributed marginals, the strata can become very
+distorted, and the tiles, along with their shading, inscrutable.
+An alternative is to use trellis-like layouts for visualizing \emph{partial}
 tables for given strata, defined by the conditioning variables. 
@@
 would be expected under the null) causes the rejection of the
-hypothesis of conditional independence. The overall deviation pattern indicates that 
-in contrast to the impression given by the aggregated table, male
+hypothesis of conditional independence. The overall deviation pattern
+thus indicates that in contrast to the impression given by the aggregated table, male
 students were not advantaged, and that in department A,
@@
 <<UCBAdmissions,fig=TRUE,echo=FALSE,height=6,width=10,results=hide>>=
-cotabplot(UCBAdmissions, panel = cotab_mosaic, shade = TRUE, legend = FALSE)
+cotabplot(UCBAdmissions, panel = cotab_mosaic, shade = TRUE, legend = FALSE, 
+          margins = c(3,1,1,3))
 @
@@
 \end{figure}
+
+\subsection{Summary}
+
+Mosaic, association, and sieve plots can be used for multi-way tables
+by applying them to flat representations. In addition, these basic
+plots are leveraged by specialized frameworks 
+for exploratory model search (pairs plots) and stratified
+analysis (conditional plots).
 

### 2005-11-18 trunk/vcd/paper/hsc.Rnw
modification to color chapter in HSC
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@500 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 to choose the right color palettes, derived from suitable color
-spaces. Apart from esthetic considerations, wrongly chosen colors
+spaces. Apart from aesthetic considerations, wrongly chosen colors
 might serously affect the analysis. For example,
@@
 colors in many computer packages \cite{vcd:Moretti+Lyons:2002}. Each color in this space is
-represented by three dimensions: the hue (dominant wavelength in the spectrum), 
-the saturation (`colorfulness' or `pureness'), and 
-the value (`brightness', amount of gray). These intuitive dimensions
+represented by three dimensions $(H, S, V)$: the hue $H$ (dominant wavelength in the spectrum, in $[0, 360]$), 
+the saturation $S$ (`colorfulness' or `pureness', in $[0, 100]$), and 
+the value $V$ (`brightness', amount of gray, in $[0, 100]$).\footnote{In many implementations,
+all three dimensions are scaled to the unit interval instead of the coordinates used here.}
+These intuitive dimensions
 make HSV colors easyier to specifiy than, e.g, RGB colors. 
@@
 As an example, see Fig.~\ref{fig:colors} (left) 
-showing a qualitative color palette in the HSV
+showing a qualitative color palette (colors $(H, 100, 100)$ for varying hues $H$) in the HSV
 space: although saturation and value are fixed, the fully saturated blue is
@@
 The upper part of Fig.~\ref{fig:diverging} shows a diverging
-palette in the HSV space with colors ranging from a saturated red over a neutral white
-to a saturated blue. Although the palette should be balanced with
+palette in the HSV space with colors ranging from a saturated red $(0, 100, 100)$
+over a neutral white $(H, 0, 100)$ to a saturated blue $(240, 100, 100)$.
+Although the palette should be balanced with
 respect to colorfulness and brightness, the red colors are perceived
@@
 \caption{Qualitative color palette for the HSV (left) 
-         and HCL (right) spaces.}
+         and HCL (right) spaces. The HSV colors are $(H, 100, 100)$ and
+	 the HCL colors $(H, 50, 70)$ for the same hues $H$.}
 \label{fig:colors}
@@
 
-The use of colors that are more `in harmony' go back to Munsell 
+The use of colors that are more `in harmony' goes back to Munsell 
 \cite{vcd:Munsell:1905} %'
@@
 transformation of the CIELUV space leads to the HCL (Hue-Chroma-Luminance) space.
-Chroma loosely corresponds to colorfulness and luminance to
+These colors are again specified by triplets $(H, C, L)$:
+chroma $C$ loosely corresponds to colorfulness and luminance $L$ to
 brightness, but in contrast to HSV colors, chroma is an
@@
 Qualitative color palettes can easily be obtained by holding constant
-chroma and luminance, and dividing the hue range in equally-sized intervals. 
-HCL colors with fixed chroma and luminance 
+chroma and luminance, and using varying hues. HCL colors with fixed chroma and luminance 
 are always balanced towards the same grey 
@@
 colors (see Fig.~\ref{fig:colors}, right). 
-Similarly, diverging color palettes can be derived
-\cite{vcd:Zeileis+Meyer+Hornik:2005} that are
-suited, e.g., for visualizing residuals as we will see in the next
-section. In principle, for each of the two bipolar hues of the
-diverging palette, we could hold constant 
-the luminance and only vary the chroma to get the different colors. 
-However, it turns out that the resulting colors 
-are too similar to be used for discriminating shaded
-objects. Better contrasting can be achieved by \emph{simultaneously}
-varying chroma and luminance, for example by choosing two points in
-the chroma/luminance plane for a given hue and interpolating
-linearly. For each half of the diverging palette, 
-one of these points should be a ``neutral'' color,
-and the other correspond to the ending part of the scale. 
-The lower part of Fig.~\ref{fig:diverging} shows a diverging
-palette using HCL colors. In contrast to the HSV counterpart, matching
+Similarly, diverging HCL color palettes can be derived
+\cite{vcd:Zeileis+Meyer+Hornik:2005} by interpolating again between a
+neutral color such as $(H, 0, 90)$ and two colors with full chroma
+such as blue $(260, 100, 50)$ and red $(0, 100, 50)$. The resulting
+palette is shown in the lower part of Fig.~\ref{fig:diverging}. 
+In contrast to the HSV counterpart, matching
 colors (light red/blue and dark red/blue) 
-are balanced to the same gray, and thus get the same perceptual ``weight''.
+are balanced to the same gray, and thus receive the same perceptual ``weight''.
+Note, that this varies chroma and luminance are varied simultaneously, i.e.,
+the full chroma colors are also darker than the neutral color. Of course,
+it would also be possible to choose a darker neutral color (or lighter 
+full chroma colors), however, by varying both chroma and luminace better 
+contrasts can be achieved.
 
@@
 All plots introduced in Sect.~\ref{sec:twoway} are composed of empty
-tiles. It seems intuitive to used filled tiles to highlight information
+tiles. It seems intuitive to use filled tiles to highlight information
 of interest. Consider again the hospital example: in Fig.~\ref{fig:highlight},
@@
        split = TRUE, spacing = spacing_highlighting, 
-       gp = gpar(fill = c("gray","black","black")))
+       gp = gpar(fill = rep(grey.colors(2)[2:1], 1:2)))
 @ 

### 2005-11-18 trunk/vcd/paper/hsc.Rnw
final fixes
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@501 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \hypersetup{%
-  pdftitle = {Residual-based Shadings for Visualizing (Conditional) Independence},
+  pdftitle = {Visualizing Multi-way Contingency Tables},
   pdfsubject = {Residual-based Shadings},
   pdfkeywords = {association plots, conditional inference, contingency tables, HCL colors, HSV colors, mosaic plots},
-  pdfauthor = {Achim Zeileis, David Meyer, Kurt Hornik},
+  pdfauthor = {David Meyer, Achim Zeileis, Kurt Hornik},
   %% change colorlinks to false for pretty printing
@@
 are balanced to the same gray, and thus receive the same perceptual ``weight''.
-Note, that this varies chroma and luminance are varied simultaneously, i.e.,
+Note, that chroma and luminance are varied simultaneously, i.e.,
 the full chroma colors are also darker than the neutral color. Of course,

### 2006-01-04 trunk/vcd/R/assoc.R
several changes in structable framework
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@504 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -107,7 +107,7 @@ struc_assoc <- function(compress = TRUE, xlim = NULL, ylim = NULL,
     split <- function(res, sexp, i, name, row, col) {
       v <- split_vertical[i]
       splitbase <- if (v) sexp else res
-      splittab <- lapply(seq(dx[i]), function(j) splitbase[j])
+      splittab <- lapply(seq(dx[i]), function(j) splitbase[[j]])
       len <- sapply(splittab, function(x) sum(unclass(x)[1,] - unclass(x)[2,]))
 
       d <- dx[i]

### 2006-01-04 trunk/vcd/R/structable.R

@@ -162,34 +162,36 @@ structable.default <- function(..., direction = NULL, split_vertical = FALSE) {
   ret
 }
 
-"[.structable" <- function(x, ...) {
+"[[.structable" <- function(x, ...) {
   args <- if (nargs() < 3)
     list(..1)
   else
     list(..1, ..2)
-  
+
+  args <- lapply(args, function(x) if (is.logical(x)) which(x) else x)
+
   ## handle one-arg cases
   if (nargs() < 3)
     if (length(args[[1]]) > 1)
-      ## resolve calls like x[c(1,2)]
-      return(x[ args[[1]][1] ] [ args[[1]][-1] ])
+      ## resolve calls like x[[c(1,2)]]
+      return(x[[ args[[1]][1] ]] [[ args[[1]][-1] ]])
     else
-      ## resolve x[foo] 
-      return(if (attr(x, "split_vertical")[1]) x[,args[[1]] ] else x[args[[1]],])
+      ## resolve x[[foo]] 
+      return(if (attr(x, "split_vertical")[1]) x[[,args[[1]] ]] else x[[args[[1]],]])
 
-  ## handle calls like x[c(1,2), c(3,4)]
+  ## handle calls like x[[c(1,2), c(3,4)]]
   if (length(args[[1]]) > 1 && length(args[[2]]) > 1)
-    return(x[ args[[1]][1], args[[2]][1] ] [ args[[1]][-1], args[[2]][-1] ])
+    return(x[[ args[[1]][1], args[[2]][1] ]] [[ args[[1]][-1], args[[2]][-1] ]])
   
-  ## handle calls like x[c(1,2), 3]
+  ## handle calls like x[[c(1,2), 3]]
   if (length(args[[1]]) > 1)
-    return(x[ args[[1]][1], args[[2]] ] [ args[[1]][-1], ])
+    return(x[[ args[[1]][1], args[[2]] ]] [[ args[[1]][-1], ]])
   
-  ## handle calls like x[1, c(1,3)]
+  ## handle calls like x[[1, c(1,3)]]
   if (length(args[[2]]) > 1)
-    return(x[ args[[1]], args[[2]][1] ] [ , args[[2]][-1] ])
+    return(x[[ args[[1]], args[[2]][1] ]] [[ , args[[2]][-1] ]])
 
-  ## final cases like x[1,2] or x[1,] or x[,1]
+  ## final cases like x[[1,2]] or x[[1,]] or x[[,1]]
[…]

### 2006-01-04 trunk/vcd/inst/doc/strucplot.Rnw
several changes in structable framework
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@504 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 data(HairEyeColor)
+data(PreSex)
+data(UCBAdmissions)
+art <- xtabs(~Treatment + Improved, data = Arthritis)
+options(width=70)
 @
@@
 
-The strucplot framework in the {\sf R} package \pkg{vcd} visualizes
-multi-way contingency tables and integrates techniques such as
+The strucplot framework in the {\sf R} package \pkg{vcd}, used for visualizing
+multi-way contingency tables, integrates techniques such as
 mosaic displays, association plots, and sieve plots. The main idea is to visualize
@@
 appearance of the plot. These are examples of `graphical appearance control'
-(`grapcon') functions. They are created by generating functions, allowing
-flexible parameterization and extensibility. 
-The first part of their names (\code{\emph{group\_foo()}}) 
+(`grapcon') functions. They are created by generating functions
+(`grapcon generators'), allowing
+flexible parameterization and extensibility (Figure \ref{fig:struc}
+only shows the generators).
+The first part of the generator names (\code{\emph{group\_foo}()})
 reflects the group they belong to (strucplot core, labeling,
 legend, shading, or spacing). The workhorse functions (created by 
-\code{\emph{struc\_foo()}}, 
-\code{\emph{labeling\_foo()}}, and \code{\emph{legend\_foo()}}) 
+\code{struc\_\emph{foo}()}, 
+\code{labeling\_\emph{foo}()}, and \code{legend\_\emph{foo}()}) 
 directly produce graphical output (``add ink to the canvas''), whereas
 the parameter functions
-(created by \code{\emph{spacing\_foo()}} and \code{\emph{shading\_foo()}}) compute
+(created by \code{spacing\_\emph{foo}()} and \code{shading\_\emph{foo}()}) compute
 graphical parameters used by the others. The grapcon functions returned by 
-\code{\emph{struc\_foo()}} implement the core functionality, creating the tiles and their
+\code{struc\_\emph{foo}()} implement the core functionality, creating the tiles and their
 content.  On the second level of the framework, a suitable combination
@@
 
+\setkeys{Gin}{width=0.75\textwidth}
 \begin{figure}[p]
@@
 
-\setkeys{Gin}{width=0.7\textwidth}
 \begin{figure}[p]
@@
 
+<<echo=FALSE>>=
+options(width=60)
+@
+
 <<split1>>=
-mosaic(hec, split_vertical = c(TRUE, FALSE, TRUE))
+mosaic(hec, split_vertical = c(TRUE, FALSE, TRUE), 
+labeling_args = list(abbreviate = c(Eye = 3)))
+@
+
+<<echo=FALSE>>=
+options(width=70)
 @
@@
 
-So far, we have visualized full tables. For objects of class \class{table},
-conditioning on levels (i.e., choosing a table subset for fixed levels
-of the conditioning variable(s)) is simply done by indexing.
-However, subsetting \code{"structable"} objects is more restrictive 
-because of their inherent conditional structure. Since the variables
+So far, we have visualized full tables. Subtables can be selected in a
+similar way than for objects of class \class{table} using
+indexing. Note, however, that subsetting of \code{"structable"} objects 
+is more restrictive because of their inherent conditional structure. Since the variables
 on both the row and the columns side are nested, 
-conditioning is only possible ``outside-in'':
+subsetting is only possible ``outside-in'', that is, indexing operates
+on blocks defined by the variable levels:
 
@@
 hec["Male",]
-hec[c("Male", "Brown"),]
-hec["Male","Green"]
+hec["Male", c("Brown","Green")]
+@
+
+\noindent \emph{Conditioning} on levels (i.e., choosing a table subset for fixed levels
+of the conditioning variable(s)) is done using the \code{[[} operator. %]]
+Here again, the sequence of conditioning levels is restricted by the
+hierarchical structure of the \class{"structable"} object:
+
+<<conditioning,results=verbatim>>=
+hec[["Male",]]
+hec[[c("Male", "Red"),]]
+hec[["Male","Green"]]
 @
@@
 pushViewport(viewport(layout.pos.col = 1))
-mosaic(hec["Male"], margins = c(top = 2, 0), sub = "Male", newpage = FALSE)
+mosaic(hec[["Male"]], margins = c(left = 2.5, top = 2.5, 0), sub = "Male", newpage = FALSE)
 popViewport()
@@
 pushViewport(viewport(layout.pos.col = 2))
-mosaic(hec["Female"], margins = c(top = 2, 0), sub = "Female", newpage = FALSE)
+mosaic(hec[["Female"]], margins = c(top = 2.5, 0), sub = "Female", newpage = FALSE)
 popViewport(2)
@@
 \begin{center}
-<<Variablesfig,fig=TRUE,echo=FALSE,height=5,width=12>>=
+<<Variablesfig,fig=TRUE,echo=FALSE,height=6,width=12>>=
 <<Variables>>
@@
 
+\setkeys{Gin}{width=\textwidth}
 \begin{figure}[p]
@@
 \end{figure}
+\setkeys{Gin}{width=0.7}
 
@@
 cotabplot(~ Hair + Eye | Sex, data = hec, panel_args = list(margins = 3), labeling = labeling_left(clip = FALSE))
-#cotabplot(~ Hair + Eye | Sex, data = hec, labeling_args = list(abbreviate = c(Eye = 2)))
 @
@@
 
-\noindent The \code{labeling\_args} argument ensures that `eye' levels
-are abbreviated (see the vignette on labeling for detailed information).
+\noindent The \code{labeling\_args} argument modifies the labels'
+appearance: here, to be left-aligned and unclipped
+(see the separate vignette: ``Labeling in the Strucplot Framework'' 
+for detailed information).
 

### 2006-01-04 trunk/vcd/paper/JSS.Rnw
strucplot framework update
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@505 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-So far, we have visualized full tables. For objects of class \class{table},
-conditioning on levels (i.e., choosing a table subset for fixed levels
-of the conditioning variable(s)) is simply done by indexing.
-However, subsetting \class{structable} objects is more restrictive 
-because of their inherent conditional structure. Since the variables
+So far, we have visualized full tables. Subtables can be selected in a
+similar way than for objects of class \class{table} using
+indexing. Note, however, that subsetting of \class{structable} objects 
+is more restrictive because of their inherent conditional structure. Since the variables
 on both the row and the columns side are nested, 
-conditioning is only possible ``outside-in'':
+subsetting is only possible ``outside-in'', that is, indexing operates
+on blocks defined by the variable levels:
 
@@
 hec["Male",]
-hec[c("Male", "Brown"),]
-hec["Male","Green"]
+hec["Male", c("Brown","Green")]
+@
+
+\noindent \emph{Conditioning} on levels (i.e., choosing a table subset for fixed levels
+of the conditioning variable(s)) is done using the \code{[[} operator. %]]
+Here again, the sequence of conditioning levels is restricted by the
+hierarchical structure of the \class{structable} object:
+
+<<conditioning,results=verbatim>>=
+hec[["Male",]]
+hec[[c("Male", "Red"),]]
+hec[["Male","Green"]]
 @
@@
 pushViewport(viewport(layout.pos.col = 1))
-mosaic(hec["Male"], margins = c(left = 2.5, top = 2.5, 0), sub = "Male", newpage = FALSE)
+mosaic(hec[["Male"]], margins = c(left = 2.5, top = 2.5, 0), sub = "Male", newpage = FALSE)
 popViewport()
@@
 pushViewport(viewport(layout.pos.col = 2))
-mosaic(hec["Female"], margins = c(top = 2.5, 0), sub = "Female", newpage = FALSE)
+mosaic(hec[["Female"]], margins = c(top = 2.5, 0), sub = "Female", newpage = FALSE)
 popViewport(2)

### 2006-04-10 trunk/vcd/R/assoc.R
revision hsc
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@517 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -22,11 +22,11 @@ assoc.default <- function(x,
                           spacing_args = list(),
                           split_vertical = NULL,
                           keep_aspect_ratio = FALSE,
-			  residuals_type = "Pearson",
                           xscale = 0.9, yspace = unit(0.5, "lines"),
                           main = NULL,
                           sub = NULL,
                           ...,
+			  residuals_type = "Pearson",
                           gp_axis = gpar(lty = 3)
                           ) {
 

### 2006-04-10 trunk/vcd/paper/hsc.Rnw
revision hsc
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@517 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 set.seed(myseed)
-art_max <- coindep_test(art, n = 5000)
+#art_max <- coindep_test(art, n = 5000)
+uc_max <- coindep_test(aperm(UCBAdmissions, c(3,2,1)), margin = "Dept", n = 5000)
+
 set.seed(myseed)
@@
 hos_chisq <- coindep_test(hospital, n = 5000, indepfun = function(x) sum(x^2))
-
-set.seed(myseed)
-hec_max <- coindep_test(HairEyeColor, "Sex", n = 5000)
 
@@
 
-\title*{Visualizing Multi-way Contingency Tables}
+\title*{Visualizing Contingency Tables}
 % Use \titlerunning{Short Title} for an abbreviated version of
@@
 % your contribution title if the original one is too long
-\institute{Department of Information Systems and Process Management
+\institute{Department of Information Systems and Operations
 \and Department of Statistics and Mathematics\\Wirtschaftsuniversität
@@
 of tiles (corresponding to the cells) created by recursive
-vertical and horizontal splits of a square. Thus, the area of each tile
+vertical and horizontal splits of a rectangle. Thus, the area of each tile
 is proportional to the corresponding cell entry \emph{given} the
@@
   \item Varying the shape of mosaic plots (as well as bar plots) to yield, e.g.,
-  double-decker plots \cite{vcd:hofmann:2001}, spine 
+  double-decker plots \cite{vcd:Hofmann:2001}, spine 
   plots \cite{vcd:riedwyl+schuepbach:1994}, or spinograms.
@@
 long-term schizophrenic patients in two London mental hospitals with
-the freqency of visits. The length of stay (LOS) has been categorized into
+the frequency of visits (from relatives or friends). 
+The length of stay (LOS) has been categorized into
 2--9 years, 10--19 years, and more than 19 years. There are also three
 categories for the visit frequency: regular (including
+%FIXME: lookup regular in Wing-paper.
 patients that were allowed to go home), less than monthly, and never.
@@
  & \multicolumn{3}{l|}{\textbf{Length of stay (in years)}}\\
-\textbf{Visit frequency} &~~2--9 & ~~10--19 & ~~20+ \\
+\textbf{Visit frequency} &~~2--9 & ~~~~~~10--19 & ~~20+ \\
 \hline
@@
 x.ticklabs = c("2-9","","10-19","","20+"), 
-              y.ticklabs = c("Regular","","Less than monthly","","Never"), 
+              y.ticklabs = c("Never","","Less than monthly","","Regular"), 
               xlab = "Length of stay (years)", ylab = "Visit frequency", zlab = "Number of patients",
@@
 They visualize the observed values of a contingency table by
-area-proportional tiles, arranged in a squared mosaic. The tiles are
-obtained by recursive partitioning splits of a square.
+area-proportional tiles, arranged in a rectangular mosaic. The tiles are
+obtained by recursive partitioning splits of a rectangle.
 In the following, we describe the main idea of mosaic plots, 
@@
 in addition the deviations from the observed frequencies by the density of a
-grid added to each tile. They are best used to detect dependency
-patterns, especially for ordinal variables. Association plots directly
+grid added to each tile. They complement mosaic plots in detecting dependency
+patterns for ordinal variables. However, for nominal variables, mosaic plots are
+a better choice, especially when residual-based 
+shadings are used (see next section), Association plots directly
 visualize Pearson and raw residuals, i.e., standardized and
@@
 
-\subsection{A Note on Colors and Color Palettes}
+\subsection[A Note on Colors and Color Palettes]{A Note on Colors and Color Palettes\footnote{The electronic version of this article uses colors.}}
 \label{sec:colors}
@@
 \item Using high-chroma colors for large
-  areas tend to produce after-image effects which can be
+  areas tends to produce after-image effects which can be
   distracting \cite{vcd:Ihaka:2003}.
@@
 Now, software implementations are often based on the more convenient
-Hue-Saturation-Value (HSV) colors (or the comparable Hue-Luminance-Saturation 
+Hue-Saturation-Value (HSV) colors (or the comparable Hue-Luminance-Satu\-ration 
 color scheme). Both spaces are rather similar transformations of the RGB space
@@
          and HCL (right) spaces. The HSV colors are $(H, 100, 100)$ and
-	 the HCL colors $(H, 50, 70)$ for the same hues $H$.}
+	 the HCL colors $(H, 50, 70)$ for the same hues $H$. Note that
+         in a monochrome version of this paper, all pies in the right
+         wheel will be shaded with the same gray, i.e. appear identical.}
 \label{fig:colors}
@@
   the three dimensions: (Hue, Saturation, Value)
-  in the upper part, and (Hue, Chroma, Luminance) in the lower part.}
+  in the upper part, and (Hue, Chroma, Luminance) in the lower part.
+  In a monochrome version of the paper, the right and left hand
+  sides of the HCL colors palette appear identical. }
 \label{fig:diverging}
@@
 cannot easily assess whether there are more, of less counts in a cell 
-than expected under the null hypothesis. It would help if we knew the 
-sign of the residuals. 
-Using color, we can add this information, for example using blue for
-positive, red for negative, and gray for zero residuals. Figure~\ref{fig:colorsieve} 
-demonstrates the effect of applying such a color shading to a sieve plot for the 
-\data{hospital} data. Color clearly emphasizes the linear association
-pattern in this data: Patients with a LOS from 2--9 years are
-associated with regular visits, whereas patients with a LOS from
-10--19 and more than 19 years are associated with a visit frequency
-of less than monthly and no visits.
-
-\setkeys{Gin}{width=0.7\textwidth}
-\begin{figure}
-\begin{center}
-<<sievecolor,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-sieve(t(hospital), split = TRUE, shade = TRUE)
-@ 
-\caption{Sieve plot with color coding of the residuals.}
-\label{fig:colorsieve}
-\end{center}
-\end{figure}
+than expected under the null hypothesis. 
+%It would help if we knew the sign of the residuals. 
+Using color, we can add the sign information, for example using blue for
+positive, red for negative, and gray for zero residuals.
+
+% Figure~\ref{fig:colorsieve} 
+% demonstrates the effect of applying such a color shading to a sieve plot for the 
+% \data{hospital} data. Color clearly emphasizes the linear association
+% pattern in this data: Patients with a LOS from 2--9 years are
+% associated with regular visits, whereas patients with a LOS from
+% 10--19 and more than 19 years are associated with a visit frequency
+% of less than monthly and no visits.
+
+% \setkeys{Gin}{width=0.7\textwidth}
+% \begin{figure}
+% \begin{center}
+% <<sievecolor,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
+% sieve(t(hospital), split = TRUE, shade = TRUE)
+% @ 
+% \caption{Sieve plot with color coding of the residuals.}
+% \label{fig:colorsieve}
+% \end{center}
+% \end{figure}
 
@@
 instead of HSV colors, and no line type coding).
-
 \setkeys{Gin}{width=0.7\textwidth}
@@
 \end{figure}
-
-\noindent Clearly, the asymmetry for regular and never visited students,
+Clearly, the asymmetry for regular and never visited students,
 and the pattern inversion for lengths of stay of 2--9 and more than 19
@@
 An important reason for using the unconditional limiting distribution
-for the $X^2$ statistic from Equ.~\ref{eq:ChiSquare} was the closed form result
+for the $X^2$ statistic from (\ref{eq:ChiSquare}) was the closed form result
 for the distribution. Recently, with the improving perfomance of computers,
@@
 increasing attention \cite{vcd:Ernst:2004,vcd:Pesarin:2001,vcd:Strasser+Weber:1999}. 
-For testing the independence hypothesis from Equation~\ref{eq:H0},
+For testing the independence hypothesis from (\ref{eq:H0}),
 using a permutation test is particularly intuitive due to the 
@@
 (hue for the sign and a linear combination of chroma and luminance 
-for the size of the residuals), we can add a third information to the plot.
+for the size of the residuals), we can add a third piece of information to the plot.
 For example, we can visualize the significance of some specified test statistic
@@
 results. These can again be derived using the interpolation procedure
-described in the Sect.~\ref{sec:colors} by
+described in Sect.~\ref{sec:colors} by
 restricting the luminance to a shorter range, ending with a ``darker'' value.
@@
 independence may not be significant even though some of the residuals are
-``large''. On the the other hand, the test might be significant even
+``large''. On the other hand, the test might be significant even
 though the residuals are ``small''. In fact, the cut-off points are really
@@
 
-In Sect.~\ref{sec:twoway}, we have presented basic visualization methods
-for two-way tables. In this section, we will show how these methods
-extend to multi-way tables by applying them to `flat' representations,
-and we will treat specialized displays for conditional independence models.
-
-\subsection{Visualizing Flat-Tables}
-\label{sec:flat}
-
-The main idea of the mosaic, sieve, and association plots presented
-in the previous sections is to visualize information on 
+In Sect.~\ref{sec:twoway}, we have presented basic displays
+for two-way tables, based on the visualization of information on 
 the tables' cells, arranged in rectangular form. For multi-way tables,
@@
 ways similar to a two-dimensional table.
-As an example, consider the \data{HairEyeColor} 
-data containing two polytomous variables (hair and eye color), 
-as well as one (artificial) dichotomous variable (gender). 
-A `flattened' contingency table, putting eye color in the columns %'
-and hair color---nested in gender---in the rows, is given by Tab.~\ref{tab:hec}.
-The corresponding mosaic, sieve, and association plots are shown in
-Figs.~\ref{fig:hecmosaic}--\ref{fig:hecassoc}. 
-The shadings in these plots visualize
-the model of hair and eye color being \emph{conditional} independent,
-given gender. As we can see from the shading (many tiles are colored) 
-and the $p$ value for the test statistic, this model fits only poorly.
+
+In the following, we will first treat specialized displays for
+explorative purposes, followed by model-based methods for
+conditional independence models.
+
+\subsection{Explorative Visualization Techniques}
+\label{sec:flat}
+
+As an example, consider the well-known UCB admissions data 
+\cite{vcd:Bickel+Hammel+O'Connell:1975}
+on applicants to graduate school at the University of California in Berkeley for
+the six largest departments in 1973 classified by admission and
+gender. The `flattened' contingency table, putting department into the
+columns and admit---nested in gender---into the rows, is given by
+Tab.~\ref{tab:ucb}. The table aggregated over all
+departments gives wrong evidence of gender bias, 
+i.e., higher admission rates for male students.
 
@@
 \begin{center}
-\begin{tabular}{|ll|rrrr|}
-\hline
- & & \multicolumn{4}{c|}{\textbf{Eye}} \\
-\textbf{Gender} & \textbf{Hair} & Brown & Blue & Hazel & Green\\\hline
-Male & Black & 32 & 11 & 10 & 3 \\
-& Brown & 38 & 50 & 25 & 15 \\
-& Red & 10 & 10 & 7 & 7 \\
-& Blond & 3 & 30 & 5 & 8 \\
-Female & Black & 36 & 9 & 5 & 2 \\
-& Brown & 81 & 34 & 29 & 14 \\
-& Red & 16 & 7 & 7 & 7 \\
-& Blond & 4 & 64 & 5 & 8 \\
+\begin{tabular}{|ll|rrrrrr|}
+\hline
+ & & \multicolumn{6}{c|}{\textbf{Department}} \\
+\textbf{Gender} & \textbf{Admit} & A & B & C & D & E & F\\\hline
+Male   & Admitted & 512 & 353 & 120 & 138 & 53 & 22\\
+       & Rejected & 313 & 207 & 205 & 279 & 138 & 351\\
+Female & Admitted & 89 & 17 & 202 & 131 & 94 & 24 \\
+       & Rejected & 19 & 8 & 391 & 244 & 299 & 317 \\
 \hline
 \end{tabular}
-\caption{The \data{HairEyeColor} data, in flat representation.}
-\label{tab:hec}
+\caption{The \data{UCB admissions} data, in flat representation.}
+\label{tab:ucb}
 \end{center}
@@
 
-\setkeys{Gin}{width=0.7\textwidth}
-\begin{figure}
-\begin{center}
-<<hecmosaic,fig=TRUE,echo=FALSE,height=12,width=7,results=hide>>=
-mosaic(~ Hair + Eye | Sex, data = HairEyeColor,
-       gp_args = list(p.value = hec_max$p.value, interpolate = hec_max$qdist(c(0.9,0.99))),
-       labeling_args = list(tl_labels = c(Hair = TRUE), set_varnames = c(Sex = "Gender")), 
-       legend_args = list(y = 0.3, height = 0.4),
-       mar = c(left = 4), shade = TRUE, keep = FALSE)
-@
-\caption{Mosaic plot for the \data{HairEyeColor} data.}
-\label{fig:hecmosaic}
-\end{center}
-\end{figure}
-
-\setkeys{Gin}{width=0.7\textwidth}
-\begin{figure}
-\begin{center}
-<<hecsieve,fig=TRUE,echo=FALSE,height=12,width=7,results=hide>>=
-hec = structable(Eye ~ Sex + Hair, data = HairEyeColor)
-sieve(hec, labeling_args = list(tl_labels = c(Hair = TRUE), abbreviate = c(Hair = 4),
-set_varnames = c(Sex = "Gender")), spacing = spacing_conditional, condvars = 1,
-      legend_args = list(y = 0.3, height = 0.4),
-      mar = c(left = 4), shade = TRUE, expected = ~ Hair * Sex + Eye * Sex, keep = FALSE)
-@
-\caption{Sieve plot for the \data{HairEyeColor} data.}
-\label{fig:hecsieve}
-\end{center}
-\end{figure}
-
-\setkeys{Gin}{width=0.7\textwidth}
-\begin{figure}
-\begin{center}
-<<hecassoc,fig=TRUE,echo=FALSE,height=12,width=7,results=hide>>=
-hec = structable(Eye ~ Sex + Hair, data = HairEyeColor)
-assoc(hec, labeling_args = list(tl_labels = c(Hair = TRUE), abbreviate = c(Hair = 4),
-             set_varnames = c(Sex = "Gender")), 
-      gp_args = list(p.value = hec_max$p.value, interpolate = hec_max$qdist(c(0.9,0.99))),
-      legend_args = list(y = 0.3, height = 0.4),
-      spacing = spacing_conditional(sp = 0, start = 4),
-      mar = c(left = 4), shade = TRUE, expected = ~ Hair * Sex + Eye * Sex, keep = FALSE)
-@
-\caption{Association plot for the \data{HairEyeColor} data.}
-\label{fig:hecassoc}
-\end{center}
-\end{figure}
-
-\subsection{Displays for Conditional Independence}
-\label{sec:condind}
-
-In addition to the generic approach to the visualization of multi-way tables outlined in 
-the previous section, there are specialized displays 
-designed for the visualization of conditional independence structures.
-
-A first step in the analysis of more complex tables is to find a
-suitable model that fits the data. All basic plots can be combined in pairwise displays,
+A first step in the explorative analysis of more complex tables is to
+get a quick overview on the data. For that, 
+all basic plots can be combined in pairwise displays,
 arranged in in a matrix similar to scatterplots in a pairs plot.
@@
 variables are implicitly specified by the cells' position in the matrix. %'
-More formally, each cell $a_{ij}$ in such a matrix defines two
+In Fig.~\ref{fig:pairs}, the diagonal cells show bar plots for the
+distribution of the variables, and the off-diagonal cells mosaic plots
+for the corresponding pair of variables. The plots suggest that
+admission differs between male and female students, and also varies given department.
+
+\setkeys{Gin}{width=\textwidth}
+\begin{figure}
+\begin{center}
+<<pairs,fig=TRUE,echo=FALSE,height=8,width=8,results=hide>>=
+pairs(UCBAdmissions)
+@
+\caption{Pairs plot for the \data{HairEyeColor} data.}
+\label{fig:pairs}
+\end{center}
+\end{figure}
+
+The next step is to investigate three-way interactions among the variables.
+In this example, we have a binary variable of interest (admission)
+that shall be ``explained'' by the others. 
+A possible way of representing such a three-way table is 
+to draw a mosaic plot of gender and
+department and to highlight admission in the tiles (see Fig.~\ref{fig:ucbhighlight}).
+From the widths of the tiles it can be seen that an
+unequal number of students apply for the six
+departments, and from the highlighting we see that the admission rates differ as
+well among the departments. The rates 
+are equal for male and female students, except for department A where
+\textit{more} female than male students are admitted. This result
+is even better depicted by a doubledecker plot of admission, explained
+by gender given department (see Fig.~\ref{fig:doubleucb}): here, the
+difference of admission rates in department A is more marked.
+
+\setkeys{Gin}{width=0.7\textwidth}
+\begin{figure}
+\begin{center}
+<<ucbhighlight,fig=TRUE,echo=FALSE,height=6,width=9,results=hide>>=
+mosaic(aperm(UCBAdmissions), gp = gpar(fill = grey.colors(2)),
+   spacing = spacing_highlighting, split = TRUE, keep = FALSE, rep = FALSE)
+@
+\caption{Mosaic plot with highlighting for the \data{UCBAdmissions} data.}
+\label{fig:ucbhighlight}
+\end{center}
+\end{figure}
+
+\setkeys{Gin}{width=0.7\textwidth}
+\begin{figure}
+\begin{center}
+<<doubleucb,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
+doubledecker(aperm(UCBAdmissions), col = grey.colors(2))
+@
+\caption{Dobledecker plot for the \data{UCBAdmissions} data.}
+\label{fig:doubleucb}
+\end{center}
+\end{figure}
+
+\subsection{Model-based Displays for Conditional Independence Models}
+\label{sec:condind}
+
+In addition to the explorative approaches 
+to the visualization of multi-way tables outlined in 
+the previous section, there are specialized displays 
+designed for the visualization of conditional independence structures.
+
+A first approach is to use pairs plots for model search:
+we can use the position of the cells in the pairs matrix to select an
+independence model and add a corresponding shading to the tiles to
+visualize the residuals.
+More precisely, each cell $a_{ij}$ in such a matrix defines two
 variables $i$ and $j$ that can be used to specify the model visualized
@@
 independent from all others, etc.
-Figure~\ref{fig:pairs} shows a pairs display with mosaic plots visualizing mutual 
-independence in the upper triangle, 
-mosaics for \emph{joint} independence 
-in the lower triangle, and bar charts in the diagonal.
-In the upper part of the matrix, we can immediately 
-detect the independence of hair and gender, and eye and gender from the
-non-colored cells in the corresponding mosaic plots. On the other hand, hair and eye color 
-clearly are associated. In the lower part of the matrix, the mosaic
-at the intersection of hair and eye is the less colored one. Its shading
-represents the model of hair and eye color being jointly independent from
-gender, i.e., their association being homogeneous over the gender
-strata. It is the model that fits best, although still rejected at the
-95\% level 
-($p=\Sexpr{round(summary(loglm(~ Hair * Eye + Sex, data = HairEyeColor))$tests[2,3],4)}$): %$
-possibly, because too many brown-haired male
-students have blue eyes, and too few of them have brown eyes, than expected under
-the null model?
-
-\setkeys{Gin}{width=\textwidth}
-\begin{figure}
-\begin{center}
-<<pairs,fig=TRUE,echo=FALSE,height=8,width=8,results=hide>>=
-pairs(hec, lower_panel = pairs_mosaic(type = "joint"), space = 0.3, shade = TRUE,
-      diag_panel_args = list(rot = -45, just_leveltext = c("left","bottom")))
-@
-\caption{Pairs plot for the \data{HairEyeColor} data.}
-\label{fig:pairs}
+% Figure~\ref{fig:pairs} shows a pairs display with mosaic plots visualizing mutual 
+% independence in the upper triangle, 
+% mosaics for \emph{joint} independence 
+% in the lower triangle, and bar charts on the diagonal.
+% In the upper part of the matrix, we can immediately 
+% detect the independence of hair and gender, and eye and gender from the
+% non-colored cells in the corresponding mosaic plots. On the other hand, hair and eye color 
+% clearly are associated. In the lower part of the matrix, the mosaic
+% at the intersection of hair and eye is the less colored one. Its shading
+% represents the model of hair and eye color being jointly independent from
+% gender, i.e., their association being homogeneous over the gender
+% strata. It is the model that fits best, although still rejected at the
+% 95\% level 
+% ($p=\Sexpr{round(summary(loglm(~ Hair * Eye + Sex, data = HairEyeColor))$tests[2,3],4)}$): %$
+% possibly, because too many brown-haired male
+% students have blue eyes, and too few of them have brown eyes, than expected under
+% the null model?
+
+In our example, from the explorative analysis, we already suspect that
+admission and gender are independent, given department, except for
+department A. 
+Therefore, a sensible way of representing the table is to 
+nest admission into department instead of gender to get a stratified view.
+The corresponding mosaic and association plots are shown
+in Figs.~\ref{fig:ucbmosaic}--\ref{fig:ucbassoc}. The shading visualizes the 
+model of admission and gender being conditionally independent, given department. Clearly,
+there is no gender bias in the departments except for department A: here, 
+significantly \textit{more} female students are admitted then expected under independence.
+Mosaic displays are recognized as an excellent
+visualization tool for log-linear models 
+\cite{vcd:Friendly:1999,vcd:Theus+Lauer:1999,vcd:Hofmann:2001}.
+Using the flat table representation, the association plot similarly extends from
+the simple two-way case to a structured residual plot usable for 
+the diagnostics of log-linear models. Especially if a
+mosaic display contains very small or empty cells, the corresponding association plot
+might provide an easier way to detect deviation patterns than the
+residual-based shading added to the mosaic plot.
+
+\setkeys{Gin}{width=0.7\textwidth}
+\begin{figure}
+\begin{center}
+<<ucbmosaic,fig=TRUE,echo=FALSE,height=6,width=12,results=hide>>=
+mosaic(aperm(UCBAdmissions, c(3,2,1)), data = UCBAdmissions, split = TRUE,
+shade = TRUE, keep = FALSE, residuals = uc_max$residuals,
+gp_args = list(p.value = uc_max$p.value)) 
+@
+\caption{Mosaic plot for the \data{UCBAdmissions} data.}
+\label{fig:ucbmosaic}
+\end{center}
+\end{figure}
+
+\setkeys{Gin}{width=0.7\textwidth}
+\begin{figure}
+\begin{center}
+<<ucbassoc,fig=TRUE,echo=FALSE,height=6,width=12,results=hide>>=
+assoc(aperm(UCBAdmissions, c(3,2,1)), split = c(TRUE,FALSE,TRUE), shade = TRUE, 
+residuals_type = "Pearson", residuals = uc_max$residuals,
+gp_args = list(p.value = uc_max$p.value))
+@
+\caption{Association plot for the \data{UCBAdmissions} data.}
+\label{fig:ucbassoc}
 \end{center}
@@
 
-Since mosaic displays are conditional plots by construction, we can use them
-for stratified data by using the conditioning variables as first splitting variables. 
-The mosaic in Fig.~\ref{fig:hecmosaic} shows the distribution of hair
-and eye color for both male and female students. A larger spacing for
-the gender variable has been used to simplify the distinction between the two
-groups. However, using a single plot, the strata are not corrected for the
-marginal distribution of the conditioning variable(s). Especially for
+% Since mosaic displays are conditional plots by construction, we can use them
+% for stratified data by using the conditioning variables as first splitting variables. 
+% The mosaic in Fig.~\ref{fig:ucbmosaic} shows the distribution of admission
+% and gender for all departments. A larger spacing for
+% the department variable has been used to simplify the distinction between the 
+% groups. 
+As we have seen, we can use all plots for the analysis of stratified
+data by using the conditioning variables first for splitting.
+However, using a single plot, the strata are not corrected for the
+marginal distribution of the stratifying variables. Especially for
 very unevenly distributed marginals, the strata can become very
@@
 In addition, trellis layouts help reducing the complexity of bigger tables. 
-As an example, consider the well-known UCB admissions data 
-\cite{vcd:Bickel+Hammel+O'Connell:1975}
-on applicants to graduate school at the University of California in Berkeley for
-the six largest departments in 1973 classified by admission and
-gender (see Tab.~\ref{tab:ucb}). The table aggregated over all
-departments gives wrong evidence of gender bias, 
-i.e., higher admission rates for male students. Figure~\ref{fig:ucbadmissions} 
+Figure~\ref{fig:ucbadmissions} 
 visualizes the data by means of a conditional 
-mosaic plot. Each panel corresponds to a department, and contains
-a mosaic corresponding to the partial (fourfold) table of admission
+association plot. Each panel corresponds to a department, and contains
+an association plot corresponding to the partial (fourfold) table of admission
 and gender. Accordingly, the shading visualizes the
@@
 would be expected under the null) causes the rejection of the
-hypothesis of conditional independence. The overall deviation pattern
-thus indicates that in contrast to the impression given by the aggregated table, male
-students were not advantaged, and that in department A,
-\emph{female} students were even more successful than their male colleagues.
+hypothesis of conditional independence. 
+% The overall deviation pattern
+% thus indicates that in contrast to the impression given by the aggregated table, male
+% students were not advantaged, and that in department A,
+% \emph{female} students were even more successful than their male colleagues.
+
+\setkeys{Gin}{width=\textwidth}
+\begin{figure}
+\begin{center}
+<<UCBAdmissions,fig=TRUE,echo=FALSE,height=6,width=10,results=hide>>=
+cotabplot(aperm(UCBAdmissions, c(2,1,3)), panel = cotab_coindep, shade = TRUE, 
+          legend = FALSE, margins = c(3,1,1,3), panel_args = list(type = "assoc"))
+@
+\caption{Conditional mosaic plot for the \data{UCBAdmissions} data.}
+\label{fig:ucbadmissions}
+\end{center}
+\end{figure}
+
+\subsection{A four-way example}
+
+As a four-way example, we use the punishment data \cite{vcd:andersen:1991} 
+from a study of the Gallup Institute in Denmark in 1979 about
+the attitude of a random sample of 783 persons towards corporal
+punishment of children (see Tab.~\ref{tab:punish}).      
+The data consists of four variables: Attitude is a binary variable
+indicating whether a person has no or moderate attitude towards punishment of children.
+Memory indicates whether the person has got memories of 
+corporal punishment as a child. Education indicates the
+highest level of education (elementary, secondary, high). Finally, 
+Age indicates the age group in years (15--24, 25--39, $40+$).
 
@@
 \begin{center}
-\begin{tabular}{|ll|rrrrrr|}
-\hline
- & & \multicolumn{6}{c|}{\textbf{Department}} \\
-\textbf{Gender} & \textbf{Admit} & A & B & C & D & E & F\\\hline
-Male   & Admitted & 512 & 353 & 120 & 138 & 53 & 22\\
-       & Rejected & 313 & 207 & 205 & 279 & 138 & 351\\
-Female & Admitted & 89 & 17 & 202 & 131 & 94 & 24 \\
-       & Rejected & 19 & 8 & 391 & 244 & 299 & 317 \\
+\begin{tabular}{|ll|rrrrrrr|}
+\hline
+ & & \textbf{Age} & \multicolumn{2}{c}{15--24} & \multicolumn{2}{c}{25--39} & \multicolumn{2}{c|}{40+}  \\
+\textbf{Education} & \textbf{Attitude} & \textbf{Memory} & Yes & No & Yes & No & Yes & No\\\hline
+ Elementary & No && 1 & 26 &  3 & 46&20&109 \\
+     & Moderate  && 21 &93 & 41 &119&143&324 \\
+Secondary   & No && 2 & 23 & 8 & 52&4&44 \\
+     & Moderate  && 5 & 45 & 20 &84&20&56 \\
+     High   & No && 2 & 26 & 6 & 24&1&13 \\
+      & Moderate && 1 & 19 & 4 &26&8&17 \\
 \hline
 \end{tabular}
-\caption{The \data{UCB admissions} data, in flat representation.}
-\label{tab:ucb}
+\caption{The \data{Punishment} data, in flat representation.}
+\label{tab:punish}
 \end{center}
@@
 
+To get an overview, we first plot a mosicplot of memory, age, and
+education, and highlight attitude (see Fig.~\ref{fig:punishhigh}).
+We can see that half of the people have more than 40 years of
+age. Most people in this group have only completed elementay
+school. For people over 25, a majority has completed secondary
+school. Among the younger people, fewer have memories of having been
+punished as a child than among the older ones. But sad enough, 
+for people with elementary or secondary school education,
+a higher proportion \emph{with} memories tends to
+have a moderate attitude to punishment than those without.
+However, for people under 40, the proportion with
+no attitude to punishment increases with the educational level.
+
+\setkeys{Gin}{width=0.7\textwidth}
+\begin{figure}
+\begin{center}
+<<punishhigh,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
+tab <- xtabs(Freq ~ memory + attitude + age + education, data = Punishment)
+mosaic(~ age + education + memory + attitude, data = tab, keep = FALSE,
+       gp = gpar(fill = grey.colors(2)), spacing = spacing_highlighting, 
+       rep = c(attitude = F), set_varnames = c(attitude = "attitude to punishment"),
+       set_labels = list(age = c("15-24","25-39","40+")))
+@
+\caption{Mosaic plot with highlighting for the \data{Punishment} data.}
+\label{fig:punishhigh}
+\end{center}
+\end{figure}
+
+So both age and education seem to matter here. Is there any influence
+of memory on attitude to punishment apart from that? To answer this
+question, we test the hypothesis of memory and attitude being
+conditionally independent, given age and education, and visualize the
+model using an extended mosaic plot (see Fig.~\ref{fig:punishmosaic}). 
+We can see that the model fits rather good, except for people over 25
+with elementary school education: under the null, more people should
+say ``no'' to physical punishment of children. This pattern can again also
+be visualized using a trellis layout, this time using partial mosaic
+plots (see Fig.~\ref{fig:punishmosaic}).
+
+\setkeys{Gin}{width=0.7\textwidth}
+\begin{figure}
+\begin{center}
+<<punishmosaic,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
+tab_test <- coindep_test(tab, margin = c(3, 4))
+mosaic(~ age + education + memory + attitude, data = tab, keep = FALSE,
+       expected = ~ (memory + attitude) * education * age,
+       gp_args = list(interpolate = tab_test$qdist(c(0.9, 0.99))), 
+       rep = c(attitude = F), set_varnames = c(attitude = "attitude to punishment"),
+       set_labels = list(age = c("15-24","25-39","40+")))
+@
+\caption{Mosaic plot for the \data{Punishment} data.}
+\label{fig:punishmosaic}
+\end{center}
+\end{figure}
+
 \setkeys{Gin}{width=\textwidth}
@@
 \begin{center}
-<<UCBAdmissions,fig=TRUE,echo=FALSE,height=6,width=10,results=hide>>=
-cotabplot(UCBAdmissions, panel = cotab_mosaic, shade = TRUE, legend = FALSE, 
-          margins = c(3,1,1,3))
-@
-\caption{Conditional mosaic plot for the UCB admissions data.}
-\label{fig:ucbadmissions}
+<<punishcotab,fig=TRUE,echo=FALSE,height=6,width=10,results=hide>>=
+cotabplot(tab, panel = cotab_coindep, panel_args = list(margin = c(2.5, 0, 0.5, 0)))
+@
+\caption{Conditional mosaic plot for the \data{Punishment} data.}
+\label{fig:punishcotab}
 \end{center}
@@
 plots are leveraged by specialized frameworks 
-for exploratory model search (pairs plots) and stratified
-analysis (conditional plots).
+for exploratory model search (pairs plots, highlighting, doubledecker plots) 
+and stratified analysis (conditional plots).
 

### 2006-04-10 trunk/vcd/paper/hsc.Rnw
went through HSC revision
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@518 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \hypersetup{%
-  pdftitle = {Visualizing Multi-way Contingency Tables},
-  pdfsubject = {Residual-based Shadings},
-  pdfkeywords = {association plots, conditional inference, contingency tables, HCL colors, HSV colors, mosaic plots},
+  pdftitle = {Visualizing Contingency Tables},
+  pdfsubject = {Handbook of Statistical Computing},
+  pdfkeywords = {association plots, contingency tables, HCL colors, HSV colors, mosaic plots, sieve plots},
   pdfauthor = {David Meyer, Achim Zeileis, Kurt Hornik},
@@
 <<preliminaries,echo=FALSE,results=hide>>=
-myseed <- 4711
+myseed <- 1071
 set.seed(myseed)
@@
 set.seed(myseed)
-#art_max <- coindep_test(art, n = 5000)
-uc_max <- coindep_test(aperm(UCBAdmissions, c(3,2,1)), margin = "Dept", n = 5000)
+art_max <- coindep_test(art, n = 5000)
+
+set.seed(myseed)
+ucb_max <- coindep_test(aperm(UCBAdmissions, c(3,2,1)), margin = "Dept", n = 5000)
 
@@
   double-decker plots \cite{vcd:Hofmann:2001}, spine 
-  plots \cite{vcd:riedwyl+schuepbach:1994}, or spinograms.
+  plots \cite{vcd:riedwyl+schuepbach:1994}, or spinograms \cite{vcd:hofmann+theus}.
 \item Using residual-based shadings to visualize log-linear models
@@
     \hline
-                             & SAS & S-PLUS & R & ViSta & MANET/Mondrian\\\hline
+                             & \proglang{SAS} & \proglang{S-PLUS} & \proglang{R} & \pkg{ViSta} & \pkg{MANET}/\pkg{Mondrian}\\\hline
     Basic functionality      &  $\times$  &   $\times$    & $\times$ &   $\times$   & $\times$\\
@@
 \end{table}
+%FIXME% For the JSS paper there was criticism about haven MANET and Mondrian in the
+%FIXME% same column, how did we decide ro resolve this? Different columns, I suppose...?!
 
@@
 Throughout this section, our examples will be based on
-the \data{hospital} data \cite{vcd:wing:1962} (see Tab.~\ref{tab:hospital}).
+the \data{hospital} data \cite{vcd:wing:1962} given in Tab.~\ref{tab:hospital}.
 
@@
 strata. From the row-standardized table (see Tab.~\ref{tab:rowhospital}), 
-it seems indeed that LOS were homogeneous with respect to these two visit frequency strata.
+it seems indeed that LOS is homogeneous with respect to these two visit frequency strata.
 
@@
 the first splitting variable dominates the plot, it should be chosen
-as to be the explanatory variable in (hypothesized) causal relationships.
+as to be the explanatory variable.
+%FIXME% causality is not needed, hence omitted: in (hypothesized) causal relationships.
 
@@
 \label{sec:sieve}
+
+%FIXME% Maybe formulate the following theory a bit more like:
+%FIXME% "To fix notations, ...", "It is well-known that..."
 
@@
 more straightforwardly by using a plot that directly visualizes the residuals. The
-most widely known residuals are the Pearson residuals
+most widely used residuals are the Pearson residuals
 
@@
 Mosaic plots are an instrument to visualize the \emph{observed} frequencies
-of a contingency table by recursive conditional splits. If causal
-relationships between the variables are assumed, the determining variables
-should be used first for splitting since they will dominate the
-plot. Sieve plots basically visualize the table of \emph{expected} frequencies, and
+of a contingency table by recursive conditional splits. If the variables can be
+separated into explanatory and dependent variable(s), the explanatory variable(s)
+should be be used first for splitting to display the conditional distribution of
+the dependent variable(s) given the explanatory variable(s). %FIXME% causality removed again
+Sieve plots basically visualize the table of \emph{expected} frequencies, and
 in addition the deviations from the observed frequencies by the density of a
 grid added to each tile. They complement mosaic plots in detecting dependency
-patterns for ordinal variables. However, for nominal variables, mosaic plots are
-a better choice, especially when residual-based 
-shadings are used (see next section), Association plots directly
+patterns for ordinal variables. An alternative way of enhancing mosaic plots
+to display deviations from expected frequencies is to use
+residual-based shadings (see next section) which are typically more intellegible
+than sieve plots, in particular for nominal variables.
+Association plots directly
 visualize Pearson and raw residuals, i.e., standardized and
@@
 respectively. These plots should be used if the diagnostics of
-independence models is of primary concern.
+independence models is of primary interest.
 
@@
 \label{sec:colors} 
+%FIXME% Maybe extend the footnote and say that the printed version is monochrome,
+%FIXME% and that the electronic version uses color.
+%FIXME% We can already say earlier that the R code for reproducing the figures
+%FIXME% is available from a Web page. And then we can say here that the colored versions
+%FIXME% can also be obtained from the Web page.
+%FIXME% Where do we want to set up the URL? I would suggest
+%FIXME%   http://statmath.wu-wien.ac.at/projects/vcd/
 
@@
 
-Now, software implementations are often based on the more convenient
+Choosing colors and color palettes in RGB space can be a bit inconvenient
+and hence software implementations are often based on
 Hue-Saturation-Value (HSV) colors (or the comparable Hue-Luminance-Satu\-ration 
@@
 \cite{vcd:Brewer:1999,vcd:Poynton:2000} and are very common implementations of
-colors in many computer packages \cite{vcd:Moretti+Lyons:2002}. Each color in this space is
+colors in many computer packages \cite{vcd:Moretti+Lyons:2002}. Each color in HSV space is
 represented by three dimensions $(H, S, V)$: the hue $H$ (dominant wavelength in the spectrum, in $[0, 360]$), 
@@
 make HSV colors easyier to specifiy than, e.g, RGB colors. 
-But HSV colors have several disadvantages.
+However, HSV colors have several disadvantages.
 Most importantly, HSV colors are not perceptually
@@
          in a monochrome version of this paper, all pies in the right
-         wheel will be shaded with the same gray, i.e. appear identical.}
+         wheel will be shaded with the same gray, i.e., appear identical.}
 \label{fig:colors}
@@
 Qualitative color palettes can easily be obtained by holding constant
-chroma and luminance, and using varying hues. HCL colors with fixed chroma and luminance 
+chroma and luminance, and using varying hues. HCL colors with fixed luminance 
 are always balanced towards the same grey 
@@
 ``stacked'' plot can also be interpreted as spine plot with highlighting
-\cite{vcd:hummel:1996}, particularly useful in analyzing causal relationships in
+\cite{vcd:hummel:1996}, particularly useful in analyzing relationships among
 categorical data with a binary dependent variable. More variations on
@@
 
-In Fig.~\ref{fig:Friendly}, we again show the mosaic for the \data{hospital}
+In Fig.~\ref{fig:Friendly}, we show again the mosaic for the \data{hospital}
 data, this time using a Friendly-like color shading (with HCL
@@
 For association plots, residual-based shadings are redundant since all relevant
-information is already contained in the plot by construction. But using one of the
-shadings discussed above will nevertheless 
-support the analysis process and is therefore recommended.
+information is already contained in the plot by construction. Nevertheless, using one of the
+shadings discussed above will support the analysis process and is therefore recommended.
 For example, applying the Friendly shading in Fig.~\ref{fig:colorassoc} 
@@
 (e.g., the $\chi^2$ test statistic) 
-using darker (`uninteresting') colors for non-significant
-results. These can again be derived using the interpolation procedure
-described in Sect.~\ref{sec:colors} by
-restricting the luminance to a shorter range, ending with a ``darker'' value.
+using less colorful (`uninteresting') colors for non-significant
+results. These can again be derived using the same procedure
+described in Sect.~\ref{sec:colors} but using a smaller amount of color, i.e.,
+a smaller maximal chroma (e.g., 20 instead of 100).
 
@@
 % $\chi^2$ test does not reject the null hypothesis of
-% independence on a significance level of $95\%$ since the $p$ value is $0.1213$. 
+% independence on a significance level of $5\%$ since the $p$ value is $0.1213$. 
 % Consequently, the employed shading uses a darker overall color scheme
@@
 patients, again by means of a mosaic display. Clearly, the hypothesis
-of independence is rejected by the $\chi^2$ test even on a $99\%$ level 
+of independence is rejected by the $\chi^2$ test even on a $1\%$ level 
 ($p=\Sexpr{round(art_chisq$p.value,4)}$), but
@@
 and computing the maximum statistic for each of these tables.
-In Fig.~\ref{fig:arthritis2}, we again
-visualize the \data{arthritis} data, this
-time using the maximum test statistic. The shading of the
+In Fig.~\ref{fig:arthritis2}, we visualize again the \data{arthritis} data, this
+time using the maximum test statistic and its 10\% and 1% critical values at cut-off points.
+The shading of the
 tiles now clearly shows that the treatment is effective: significantly more patients
@@
 @
-\caption{Mosaic plot for the \data{Arthritis} data, using the $\chi^2$ test 
+\caption{Mosaic plot for the \data{arthritis} data, using the $\chi^2$ test 
 and fixed cut-off points for the shading.}
@@
 <<Arthritis2fig,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-mosaic(art, gp = shading_max)
-@
-\caption{Mosaic plot for the \data{Arthritis} data, using the maximum test and 
+set.seed(myseed)
+mosaic(art, gp = shading_max, gp_args = list(n = 5000))
+@
+\caption{Mosaic plot for the \data{arthritis} data, using the maximum test and 
 data-driven cut-off points for the residuals.}
@@
 
-Especially for the visualization of areas, 
-use perceptually uniform colors and color palettes, derived from,
-e.g., the HCL color space. Shadings can be used to add 
-information to the basic plots and to support the
-analysis. Highlighting of tiles can support the analysis of causal relationships.
-Residual-based shadings can be used for model diagnostics, e.g.,
-by using the hue to code the sign of a residual, and a linear
-combination of chroma and luminance for the size. In addition, the
-significance of test statistics can be visualized through overall 
-darker color palettes. Using the maximum instead of the $\chi^2$ test
-statistic to aggregate the residuals allows precise diagnostics when
-the null hypothesis of independence is rejected.
+For the visualization of areas perceptually uniform colors and color palettes
+should always be used and HCL color space provides a convenient way to do so.
+Shadings can be used to add information to the basic plots and to support the
+analysis. Highlighting of tiles can support the analysis relationships with a 
+single dependent variable. Residual-based shadings can be used for model diagnostics, e.g.,
+by using diverging color palettes to code the sign and size of a residual.
+In addition, the significance of test statistics can be visualized by using 
+overall less colorful palettes. Using the maximum instead of the $\chi^2$ (or other) test
+statistic(s) allows for data-driven cut-offs in the diverging palettes and
+precise diagnostic identification of the cells in conflict with the
+null hypothesis of independence.
 
@@
 for the corresponding pair of variables. The plots suggest that
-admission differs between male and female students, and also varies given department.
+admission differs between male and female students, and between the departments,
+and that the proportion of male and female students varies across the departments.
 
@@
 @
-\caption{Pairs plot for the \data{HairEyeColor} data.}
+\caption{Pairs plot for the \data{UCB admissions} data.}
 \label{fig:pairs}
@@
 \end{figure}
+
+%FIXME% Didn't we decide to use fewer plots in the paper and simply
+%FIXME% point out that other combinations could also be used?
+%FIXME% I would use for UCB:
+%FIXME%   o explorative: highlighted doubledecker (currently Fig. 17)
+%FIXME%   o model-based: conidtional shaded assoc (currently Fig. 20)
+%FIXME% For Punishment:
+%FIXME%   o explorative: highlighted mosaic( currently Fig. 21)
+%FIXME%   o model-based: conditional shaded mosaic (currently Fig. 23)
+%FIXME% The others should be omitted (I think) and the space gained should
+%FIXME% be used for making the individual graphs larger. Preferably, we
+%FIXME% should have the explorative and model-based plots for the same data
+%FIXME% set on the same page, emphasizing that different aspects of the data
+%FIXME% are brought out by the graphs.
 
@@
 @
-\caption{Mosaic plot with highlighting for the \data{UCBAdmissions} data.}
+\caption{Mosaic plot with highlighting for the \data{UCB admissions} data.}
 \label{fig:ucbhighlight}
@@
 \begin{center}
+%FIXME% right margin labels get clipped
 <<doubleucb,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-doubledecker(aperm(UCBAdmissions), col = grey.colors(2))
-@
-\caption{Dobledecker plot for the \data{UCBAdmissions} data.}
+doubledecker(aperm(UCBAdmissions), col = rev(grey.colors(2)))
+@
+\caption{Dobledecker plot for the \data{UCB admissions} data.}
 \label{fig:doubleucb}
@@
 % strata. It is the model that fits best, although still rejected at the
-% 95\% level 
+% 5\% level 
 % ($p=\Sexpr{round(summary(loglm(~ Hair * Eye + Sex, data = HairEyeColor))$tests[2,3],4)}$): %$
@@
 mosaic(aperm(UCBAdmissions, c(3,2,1)), data = UCBAdmissions, split = TRUE,
-shade = TRUE, keep = FALSE, residuals = uc_max$residuals,
-gp_args = list(p.value = uc_max$p.value)) 
-@
-\caption{Mosaic plot for the \data{UCBAdmissions} data.}
+shade = TRUE, keep = FALSE, residuals = ucb_max$residuals,
+gp_args = list(p.value = ucb_max$p.value)) 
+@
+\caption{Mosaic plot for the \data{UCB admissions} data.}
 \label{fig:ucbmosaic}
@@
 assoc(aperm(UCBAdmissions, c(3,2,1)), split = c(TRUE,FALSE,TRUE), shade = TRUE, 
-residuals_type = "Pearson", residuals = uc_max$residuals,
-gp_args = list(p.value = uc_max$p.value))
-@
-\caption{Association plot for the \data{UCBAdmissions} data.}
+residuals_type = "Pearson", residuals = ucb_max$residuals,
+gp_args = list(p.value = ucb_max$p.value))
+@
+\caption{Association plot for the \data{UCB admissions} data.}
 \label{fig:ucbassoc}
@@
 and admission, given department), stratified by department. Clearly,
-the situation in department A (more women/less men accepted than
-would be expected under the null) causes the rejection of the
+the situation in department A (more women/less men admitted than
+expected under the null hypothesis) causes the rejection of the
 hypothesis of conditional independence. 
@@
 @
-\caption{Conditional mosaic plot for the \data{UCBAdmissions} data.}
+\caption{Conditional mosaic plot for the \data{UCB admissions} data.}
 \label{fig:ucbadmissions}
@@
 The data consists of four variables: Attitude is a binary variable
-indicating whether a person has no or moderate attitude towards punishment of children.
+indicating whether a person approves no or moderate punishment of children.
 Memory indicates whether the person has got memories of 
@@
 highest level of education (elementary, secondary, high). Finally, 
-Age indicates the age group in years (15--24, 25--39, $40+$).
+age indicates the age group in years (15--24, 25--39, $40+$).
 
@@
 \end{tabular}
-\caption{The \data{Punishment} data, in flat representation.}
+\caption{The \data{punishment} data.}
 \label{tab:punish}
@@
 \end{table}
+
+%FIXME% Better structure:
+%FIXME%   o dependent variable: attitude
+%FIXME%   o dependence on age and education is rather clear
+%FIXME%     (approval of moderate punishment increases with age and
+%FIXME%     decreases with education)
+%FIXME%   o question is conditional independence of attitude and
+%FIXME%     memory given the other margins
+%FIXME%   o hence for exploratory analysis: split in conditioning
+%FIXME%     variables (age, education) first, then explanatory variable (memory),
+%FIXME%     then dependent variable (attitude) for which highlighting is used.
+%FIXME%   o for model-based view: suppress marginal distribution of
+%FIXME%     conditioning variables (and their interaction), split first
+%FIXME%     in explanatory variable in each cell, use shading based on
+%FIXME%     residuals in conditional independence model.
+%FIXME%   o advantage of exploratory view: joint distribution of all variables
+%FIXME%     can be seen. disadvantage: visualization strongly influence
+%FIXME%     by the marginal distribution of education over age (in particular
+%FIXME%     the large proportion of 40+ people with elementary education)
+%FIXME%     which is not of interest in the conditional independence problem.
 
@@
 @
-\caption{Mosaic plot with highlighting for the \data{Punishment} data.}
+\caption{Mosaic plot with highlighting for the \data{punishment} data.}
 \label{fig:punishhigh}
@@
 @
-\caption{Mosaic plot for the \data{Punishment} data.}
+\caption{Mosaic plot for the \data{punishment} data.}
 \label{fig:punishmosaic}
@@
 @
-\caption{Conditional mosaic plot for the \data{Punishment} data.}
+\caption{Conditional mosaic plot for the \data{punishment} data.}
 \label{fig:punishcotab}

### 2006-04-10 trunk/vcd/paper/hsc.Rnw
AZ comments
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@519 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 (Fig.~\ref{fig:threedee} only), all freely available from the
-Comprehensive \proglang{R} Archive Network (\url{http://CRAN.R-project.org/}).
+Comprehensive \proglang{R} Archive Network
+(\url{http://CRAN.R-project.org/}). The \proglang{R} code used for the
+figures is available from \url{http://statmath.wu-wien.ac.at/projects/vcd/}.
 
@@
   \centering
-  \begin{tabular}{|l|c|c|c|c|c|}
-    \hline
-                             & \proglang{SAS} & \proglang{S-PLUS} & \proglang{R} & \pkg{ViSta} & \pkg{MANET}/\pkg{Mondrian}\\\hline
-    Basic functionality      &  $\times$  &   $\times$    & $\times$ &   $\times$   & $\times$\\
-    Shape                    &     &        & $\times$ &       & $\times$\\
-    Residual-based shadings  &  $\times$  &        & $\times$ &   $\times$   & ($\times$)\\
-    Conditional Views        &  $\times$  &        & $\times$ &       & $\times$\\
-    Interaction              &     &        &   &   $\times$   & $\times$\\
-    Extensible Design        &     &        & $\times$ &       &  \\\hline
+  \begin{tabular}{|l|c|c|c|c|c|c|}
+    \hline
+ & \proglang{SAS} & \proglang{S-PLUS} & \proglang{R} & \pkg{ViSta} & \pkg{MANET} & \pkg{Mondrian}\\\hline
+    Basic functionality      &  $\times$  &   $\times$    & $\times$ &   $\times$   & $\times$& $\times$\\
+    Shape                    &     &        & $\times$ &       & $\times$& $\times$\\
+    Residual-based shadings  &  $\times$  &        & $\times$ &   $\times$   & ($\times$)& ($\times$)\\
+    Conditional Views        &  $\times$  &        & $\times$ &       & $\times$& $\times$\\
+    Interaction              &     &        &   &   $\times$   & $\times$& $\times$\\
+    Extensible Design        &     &        & $\times$ &       &  &\\\hline
   \end{tabular}
@@
 \end{table}
-%FIXME% For the JSS paper there was criticism about haven MANET and Mondrian in the
-%FIXME% same column, how did we decide ro resolve this? Different columns, I suppose...?!
 
@@
 as to be the explanatory variable.
-%FIXME% causality is not needed, hence omitted: in (hypothesized) causal relationships.
 
@@
 
-%FIXME% Maybe formulate the following theory a bit more like:
-%FIXME% "To fix notations, ...", "It is well-known that..."
-
 When we try to explain data, we assume the validity of a certain
 model for the generating process. In the case of two-way contingency
-tables, the two most common hypotheses \cite{vcd:Agresti:2002} are
+tables, the two most common and well-known hypotheses \cite{vcd:Agresti:2002} are
 
@@
 \noindent It is easy to compute the \emph{expected} table under either
-of these hypotheses. Consider a 
+of these hypotheses. To fix notations, in the following we consider a 
 2-way contingency table with $I$ rows and $J$ columns,
@@
 should be be used first for splitting to display the conditional distribution of
-the dependent variable(s) given the explanatory variable(s). %FIXME% causality removed again
+the dependent variable(s) given the explanatory variable(s). 
 Sieve plots basically visualize the table of \emph{expected} frequencies, and
@@
 
-\subsection[A Note on Colors and Color Palettes]{A Note on Colors and Color Palettes\footnote{The electronic version of this article uses colors.}}
+\subsection[A Note on Colors and Color Palettes]{A Note on Colors and
+  Color Palettes\footnote{The printed version of the article is
+    monochrome, but the electronic version uses colors. The colored
+    versions of the plots are available from the code web page 
+    (\url{http://statmath.wu-wien.ac.at/projects/vcd/}).}}
 \label{sec:colors}
-%FIXME% Maybe extend the footnote and say that the printed version is monochrome,
-%FIXME% and that the electronic version uses color.
-%FIXME% We can already say earlier that the R code for reproducing the figures
-%FIXME% is available from a Web page. And then we can say here that the colored versions
-%FIXME% can also be obtained from the Web page.
-%FIXME% Where do we want to set up the URL? I would suggest
-%FIXME%   http://statmath.wu-wien.ac.at/projects/vcd/
 
@@
 gender (see Tab.~\ref{tab:arthritis} for the female patients):
-
 \begin{table}[h]
@@
 \end{table}
-
-\noindent Figure~\ref{fig:arthritis} visualizes the results for the female
+Figure~\ref{fig:arthritis} visualizes the results for the female
 patients, again by means of a mosaic display. Clearly, the hypothesis
@@
 that shall be ``explained'' by the others. 
-A possible way of representing such a three-way table is 
-to draw a mosaic plot of gender and
-department and to highlight admission in the tiles (see Fig.~\ref{fig:ucbhighlight}).
+A natural way of representing such a three-way table is 
+to draw a doubledecker plot nesting gender in
+department and highlighting admission in the tiles (see Fig.~\ref{fig:doubleucb}).
 From the widths of the tiles it can be seen that an
@@
 departments, and from the highlighting we see that the admission rates differ as
-well among the departments. The rates 
+well among the departments (roughly speaking, they decrease from A to F). The rates 
 are equal for male and female students, except for department A where
-\textit{more} female than male students are admitted. This result
-is even better depicted by a doubledecker plot of admission, explained
-by gender given department (see Fig.~\ref{fig:doubleucb}): here, the
-difference of admission rates in department A is more marked.
-
-\setkeys{Gin}{width=0.7\textwidth}
-\begin{figure}
-\begin{center}
-<<ucbhighlight,fig=TRUE,echo=FALSE,height=6,width=9,results=hide>>=
-mosaic(aperm(UCBAdmissions), gp = gpar(fill = grey.colors(2)),
-   spacing = spacing_highlighting, split = TRUE, keep = FALSE, rep = FALSE)
-@
-\caption{Mosaic plot with highlighting for the \data{UCB admissions} data.}
-\label{fig:ucbhighlight}
-\end{center}
-\end{figure}
+\textit{more} female than male students are admitted.
+
+% \setkeys{Gin}{width=0.7\textwidth}
+% \begin{figure}
+% \begin{center}
+% <<ucbhighlight,fig=TRUE,echo=FALSE,height=6,width=9,results=hide>>=
+% mosaic(aperm(UCBAdmissions), gp = gpar(fill = grey.colors(2)),
+%    spacing = spacing_highlighting, split = TRUE, keep = FALSE, rep = FALSE)
+% @
+% \caption{Mosaic plot with highlighting for the \data{UCB admissions} data.}
+% \label{fig:ucbhighlight}
+% \end{center}
+% \end{figure}
 
@@
 <<doubleucb,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-doubledecker(aperm(UCBAdmissions), col = rev(grey.colors(2)))
-@
-\caption{Dobledecker plot for the \data{UCB admissions} data.}
+doubledecker(aperm(aperm(UCBAdmissions, c(1,3,2))[2:1,,], c(2,3,1)), margins = c(left = 0, right = 5),col = rev(grey.colors(2)))
+@
+\caption{Doubledecker plot for the \data{UCBAdmissions} data.}
 \label{fig:doubleucb}
@@
 significantly \textit{more} female students are admitted then expected under independence.
-Mosaic displays are recognized as an excellent
+Mosaic displays are already recognized as an excellent
 visualization tool for log-linear models 
@@
 All subtables are corrected for the marginals and thus, all
-corresponding mosaics in the panels have the same size. 
+corresponding plots in the panels have the same size. 
 In addition, trellis layouts help reducing the complexity of bigger tables. 
@@
 
-To get an overview, we first plot a mosicplot of memory, age, and
-education, and highlight attitude (see Fig.~\ref{fig:punishhigh}).
+We first create a mosaic plot to get an overview on the data:
+Attitude towards punishment clearly is the dependent variable 
+here and should be used last for splitting. Furthermore, we might 
+expect Age and Education to have stratifying influence on the results, so they should
+be used first for splitting.
+In fact, the question is whether memory has some influence on attitude, given 
+age and education. We choose to create a three-way mosaic plot of the 
+hypothesized explanatory variables (first splitting horizontically by age, 
+then vertically by education, and finally horizontically again by memory), 
+and to have attitude, the dependent variable, 
+highlighted in the tiles (see Fig.~\ref{fig:punishhigh}).
 We can see that half of the people have more than 40 years of
@@
 
-So both age and education seem to matter here. Is there any influence
-of memory on attitude to punishment apart from that? To answer this
-question, we test the hypothesis of memory and attitude being
-conditionally independent, given age and education, and visualize the
-model using an extended mosaic plot (see Fig.~\ref{fig:punishmosaic}). 
-We can see that the model fits rather good, except for people over 25
-with elementary school education: under the null, more people should
-say ``no'' to physical punishment of children. This pattern can again also
-be visualized using a trellis layout, this time using partial mosaic
-plots (see Fig.~\ref{fig:punishmosaic}).
-
-\setkeys{Gin}{width=0.7\textwidth}
-\begin{figure}
-\begin{center}
-<<punishmosaic,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-tab_test <- coindep_test(tab, margin = c(3, 4))
-mosaic(~ age + education + memory + attitude, data = tab, keep = FALSE,
-       expected = ~ (memory + attitude) * education * age,
-       gp_args = list(interpolate = tab_test$qdist(c(0.9, 0.99))), 
-       rep = c(attitude = F), set_varnames = c(attitude = "attitude to punishment"),
-       set_labels = list(age = c("15-24","25-39","40+")))
-@
-\caption{Mosaic plot for the \data{punishment} data.}
-\label{fig:punishmosaic}
-\end{center}
-\end{figure}
+So both age and education seem to matter here. But is there really any influence
+of memory on attitude to punishment apart from that? Or, rephrased as a hypothesis:
+are memory and attitude conditionally independent, given age and education?
+We can visualize the model by means of a trellis layout, this time using partial mosaic
+plots (see Fig.~\ref{fig:punishcotab}) for the strata defined by all factorial 
+combinations of the age and education levels. In each panel, we plot mosaic plots of 
+memory and attitude, where again the explanatory variable (memory) is used first 
+for splitting.
+The shading tells us the model seems to be consistent with the data, 
+except for people over 25
+with elementary school education: under the null, more people should have had
+no attitude to physical punishment of children than actually had.
+
+% ## \setkeys{Gin}{width=0.7\textwidth}
+% ## \begin{figure}
+% ## \begin{center}
+% ## <<punishmosaic,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
+% ## tab_test <- coindep_test(tab, margin = c(3, 4))
+% ## mosaic(~ age + education + memory + attitude, data = tab, keep = FALSE,
+% ##        expected = ~ (memory + attitude) * education * age,
+% ##        gp_args = list(interpolate = tab_test$qdist(c(0.9, 0.99))), 
+% ##        rep = c(attitude = F), set_varnames = c(attitude = "attitude to punishment"),
+% ##        set_labels = list(age = c("15-24","25-39","40+")))
+% ## 
+% ## \caption{Mosaic plot for the \data{punishment} data.}
+% ## \label{fig:punishmosaic}
+% ## \end{center}
+% ## \end{figure}
+
+The advantage of the exploratory view 
+is to visualize the joint distribution of all variables.
+On the other hand, the visualization might be 
+strongly influenced by the marginal distribution of education over age (in particular
+    the large proportion of 40+ people with elementary education)
+ which is not of interest in the conditional independence problem. The
+ conditional mosaic plot suppresses the effect of all stratifying
+ variables (and their interactions).
 

### 2006-04-20 trunk/vcd/paper/hsc.Rnw
first fixes
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@521 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 task is purely exploratory or model based, techniques such as mosaic, sieve, and
-association plots offer good support for visualization. They have been
-extended over the last two decades, and implementations exist in many
-statistical environments.
+association plots offer good support for visualization. Especially
+mosaic and sieve plots have been extended over the last two decades, 
+and implementations exist in many statistical environments.
 
@@
   \item Varying the shape of mosaic plots (as well as bar plots) to yield, e.g.,
-  double-decker plots \cite{vcd:Hofmann:2001}, spine 
-  plots \cite{vcd:riedwyl+schuepbach:1994}, or spinograms \cite{vcd:hofmann+theus}.
+  double-decker plots \cite{vcd:Hofmann:2001} or spine 
+  plots \cite{vcd:riedwyl+schuepbach:1994}.
+%% or spinograms \cite{vcd:hofmann+theus}.
 \item Using residual-based shadings to visualize log-linear models
@@
 \pkg{scatterplot3d} \cite{vcd:ligges+maechler:2003}
-(Fig.~\ref{fig:threedee} only), all freely available from the
+(Figure~\ref{fig:threedee} only), all freely available from the
 Comprehensive \proglang{R} Archive Network
@@
 This chapter will give an overview of the state of the art 
-of mosaic, sieve, and association plots, both for exploratory visualization and
+of mosaic and association plots, both for exploratory visualization and
 model-based analysis. Exploratory techniques will include specialized displays for
@@
 also introduce specialized displays for conditional independence
-structures. Section~\ref{sec:concl} concludes the chapter.
+structures. The techniques are illustrated using three- and four-way tables.
+Section~\ref{sec:concl} concludes the chapter.
 
@@
 Although far from optimal, contingency tables are frequently visualized using
-grouped bar plots (see Fig.~\ref{fig:grouped}) 
-or even by means of 3D-bar charts (see Fig.~\ref{fig:threedee}). It seems hard
+grouped bar plots (see Figure~\ref{fig:grouped}) 
+or even by means of 3D-bar charts (see Figure~\ref{fig:threedee}). It seems hard
 to detect the aforementioned pattern in these, especially in the 3D
@@
 according to the marginals of one of the variables. To be consistent with the textual
-representation, we choose LOS with vertical splits (see Fig.~\ref{fig:mosaic1fig}). 
+representation, we choose LOS with vertical splits (see Figure~\ref{fig:mosaic1fig}). 
 The result is similar to a bar plot where not the height, but the width is
@@
 marginals of the second variable, \emph{given} the first variable (see
-Fig.~\ref{fig:mosaic2fig}). The resulting plot visualizes the 
+Figure~\ref{fig:mosaic2fig}). The resulting plot visualizes the 
 contingency table where each cell has a size proportional to the
@@
 as the first splitting variable, the impression is very different compared to the previous
-mosaic (see Fig.~\ref{fig:mosaic3fig}). 
+mosaic (see Figure~\ref{fig:mosaic3fig}). 
 In this alternative display, we see the marginal distribution of `Visit frequency' %'
@@
 the corresponding number of \emph{observed} frequencies, we get a \emph{sieve plot}
-(see Fig.~\ref{fig:sievefig}). It implicitly compares
+(see Figure~\ref{fig:sievefig}). It implicitly compares
 expected and observed values since the density of the grid will
@@
 Mosaic plots are an instrument to visualize the \emph{observed} frequencies
-of a contingency table by recursive conditional splits. If the variables can be
-separated into explanatory and dependent variable(s), the explanatory variable(s)
-should be be used first for splitting to display the conditional distribution of
-the dependent variable(s) given the explanatory variable(s). 
+of a contingency table by recursive conditional splits. If one variable is
+explanatory, it should be be used first for splitting: thus, the display shows
+the conditional distribution of the dependent variable given the explanatory one.
 Sieve plots basically visualize the table of \emph{expected} frequencies, and
@@
 respectively. These plots should be used if the diagnostics of
-independence models is of primary interest.
+independence models are of primary interest.
 
@@
 is not uniform across different hues.
-As an example, see Fig.~\ref{fig:colors} (left) 
+As an example, see Figure~\ref{fig:colors} (left) 
 showing a qualitative color palette (colors $(H, 100, 100)$ for varying hues $H$) in the HSV
@@
 bipolar scales containing colors ranging between two very distinct colors.
-The upper part of Fig.~\ref{fig:diverging} shows a diverging
+The upper part of Figure~\ref{fig:diverging} shows a diverging
 palette in the HSV space with colors ranging from a saturated red $(0, 100, 100)$
@@
 and thus do not have the problem of varying saturations as the HSV
-colors (see Fig.~\ref{fig:colors}, right). 
+colors (see Figure~\ref{fig:colors}, right). 
 Similarly, diverging HCL color palettes can be derived
@@
 such as blue $(260, 100, 50)$ and red $(0, 100, 50)$. The resulting
-palette is shown in the lower part of Fig.~\ref{fig:diverging}. 
+palette is shown in the lower part of Figure~\ref{fig:diverging}. 
 In contrast to the HSV counterpart, matching
@@
 it would also be possible to choose a darker neutral color (or lighter 
-full chroma colors), however, by varying both chroma and luminace better 
+full chroma colors), however, by varying both chroma and luminance better 
 contrasts can be achieved.
@@
 tiles. It seems intuitive to use filled tiles to highlight information
-of interest. Consider again the hospital example: in Fig.~\ref{fig:highlight},
+of interest. Consider again the hospital example: in Figure~\ref{fig:highlight},
 we mark tiles for patients never or seldom visited to visualize their proportion in
@@
 
-In Fig.~\ref{fig:Friendly}, we show again the mosaic for the \data{hospital}
+In Figure~\ref{fig:Friendly}, we show again the mosaic for the \data{hospital}
 data, this time using a Friendly-like color shading (with HCL
@@
 shadings discussed above will support the analysis process and is therefore recommended.
-For example, applying the Friendly shading in Fig.~\ref{fig:colorassoc} 
+For example, applying the Friendly shading in Figure~\ref{fig:colorassoc} 
 on the one hand facilitates the discrimination between
@@
 and computing the maximum statistic for each of these tables.
-In Fig.~\ref{fig:arthritis2}, we visualize again the \data{arthritis} data, this
-time using the maximum test statistic and its 10\% and 1% critical values at cut-off points.
+In Figure~\ref{fig:arthritis2}, we visualize again the \data{arthritis} data, this
+time using the maximum test statistic and its 10\% and 
+1\% critical values as cut-off points.
 The shading of the
@@
 all basic plots can be combined in pairwise displays,
-arranged in in a matrix similar to scatterplots in a pairs plot.
+arranged in a matrix similar to scatterplots in a pairs plot.
 The diagonal cells contain the variable names, optionally with
@@
 variables are implicitly specified by the cells' position in the matrix. %'
-In Fig.~\ref{fig:pairs}, the diagonal cells show bar plots for the
+In Figure~\ref{fig:pairs}, the diagonal cells show bar plots for the
 distribution of the variables, and the off-diagonal cells mosaic plots
@@
 to draw a doubledecker plot nesting gender in
-department and highlighting admission in the tiles (see Fig.~\ref{fig:doubleucb}).
+department and highlighting admission in the tiles (see Figure~\ref{fig:doubleucb}).
 From the widths of the tiles it can be seen that an
@@
 independent, given all others; variables $i$ and $j$ are jointly
-independent from all others, etc.
+independent from all others.
 % Figure~\ref{fig:pairs} shows a pairs display with mosaic plots visualizing mutual 
@@
 The corresponding mosaic and association plots are shown
-in Figs.~\ref{fig:ucbmosaic}--\ref{fig:ucbassoc}. The shading visualizes the 
+in Figures~\ref{fig:ucbmosaic}--\ref{fig:ucbassoc}. The shading visualizes the 
 model of admission and gender being conditionally independent, given department. Clearly,
@@
 corresponding plots in the panels have the same size. 
-In addition, trellis layouts help reducing the complexity of bigger tables. 
+In addition, trellis layouts help reduce the complexity of bigger tables. 
 Figure~\ref{fig:ucbadmissions} 
@@
 punishment of children (see Tab.~\ref{tab:punish}).      
-The data consists of four variables: Attitude is a binary variable
-indicating whether a person approves no or moderate punishment of children.
-Memory indicates whether the person has got memories of 
-corporal punishment as a child. Education indicates the
+The data consist of four variables: Attitude is a binary variable
+indicating whether a person approves moderate punishment of children
+(``moderate''), or refuses any punishment of children (``no'').
+Memory indicates whether the person recalls having been
+punished as a child. Education indicates the
 highest level of education (elementary, secondary, high). Finally, 
@@
 
-We first create a mosaic plot to get an overview on the data:
+We first create a mosaic plot to get an overview of the data:
 Attitude towards punishment clearly is the dependent variable 
 here and should be used last for splitting. Furthermore, we might 
-expect Age and Education to have stratifying influence on the results, so they should
+expect age and education to have stratifying influence on the results, so they should
 be used first for splitting.
@@
 and to have attitude, the dependent variable, 
-highlighted in the tiles (see Fig.~\ref{fig:punishhigh}).
+highlighted in the tiles (see Figure~\ref{fig:punishhigh}).
 We can see that half of the people have more than 40 years of
-age. Most people in this group have only completed elementay
+age. Most people in this group have only completed elementary
 school. For people over 25, a majority has completed secondary
-school. Among the younger people, fewer have memories of having been
-punished as a child than among the older ones. But sad enough, 
+school. 
+%% FIXME: additional plot?
+Among the younger people, fewer have memories of having been
+punished as a child than among the older ones. 
+%%
+But surprisingly enough,
 for people with elementary or secondary school education,
 a higher proportion \emph{with} memories tends to
-have a moderate attitude to punishment than those without.
+accept moderate punishment of children than those without: 
+experienced violence seems to engender violence again.
 However, for people under 40, the proportion with
-no attitude to punishment increases with the educational level.
+people refusing punishment increases with the educational level.
 
@@
 So both age and education seem to matter here. But is there really any influence
-of memory on attitude to punishment apart from that? Or, rephrased as a hypothesis:
+of memory on the attitude towards 
+punishment apart from that? Or, rephrased as a hypothesis:
 are memory and attitude conditionally independent, given age and education?
 We can visualize the model by means of a trellis layout, this time using partial mosaic
-plots (see Fig.~\ref{fig:punishcotab}) for the strata defined by all factorial 
+plots (see Figure~\ref{fig:punishcotab}) for the strata defined by all factorial 
 combinations of the age and education levels. In each panel, we plot mosaic plots of 
@@
 except for people over 25
-with elementary school education: under the null, more people should have had
-no attitude to physical punishment of children than actually had.
+with elementary school education: under the null, more people should
+have refused physical punishment of children than actually did.
 

### 2006-04-22 trunk/vcd/paper/hsc.Rnw
more fixes
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@523 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 In the following, we describe the main idea of mosaic plots, 
-chapter XXX in this book \emph{\bf (Reference to other contributions)} 
+Chapter XXX in this book \emph{\bf (Reference to other contributions)} 
 provides more detailed information. Consider our
@@
 categorical data with a binary dependent variable. More variations on
-this theme, such as doubledecker plots, are treated in chapter XXX
+this theme, such as doubledecker plots, are treated in Chapter XXX
 (\textbf{REFERENCE TO OTHER CHAPTER(S)}).
@@
 \begin{center}
-%FIXME% right margin labels get clipped
 <<doubleucb,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
@@
 shade = TRUE, keep = FALSE, residuals = ucb_max$residuals,
-gp_args = list(p.value = ucb_max$p.value)) 
+gp_args = list(p.value = ucb_max$p.value), rep = c(Admit = FALSE)) 
 @
@@
 residuals_type = "Pearson", residuals = ucb_max$residuals,
-gp_args = list(p.value = ucb_max$p.value))
+gp_args = list(p.value = ucb_max$p.value), rep = c(Admit = FALSE))
 @
@@
 cotabplot(aperm(UCBAdmissions, c(2,1,3)), panel = cotab_coindep, shade = TRUE, 
-          legend = FALSE, margins = c(3,1,1,3), panel_args = list(type = "assoc"))
-@
-\caption{Conditional mosaic plot for the \data{UCB admissions} data.}
+          legend = FALSE, 
+          panel_args = list(type = "assoc", margins = c(2,1,1,2), varnames = FALSE))
+@
+\caption{Conditional association plot for the \data{UCB admissions}
+  data. For each individual association plot, gender is in the rows
+  and admit is in the columns.}
 \label{fig:ucbadmissions}
@@
 \begin{center}
-<<punishcotab,fig=TRUE,echo=FALSE,height=6,width=10,results=hide>>=
-cotabplot(tab, panel = cotab_coindep, panel_args = list(margin = c(2.5, 0, 0.5, 0)))
-@
-\caption{Conditional mosaic plot for the \data{punishment} data.}
+<<punishcotab,fig=TRUE,echo=FALSE,height=10,width=10,results=hide>>=
+cotabplot(tab, panel = cotab_coindep, panel_args = list(varnames = FALSE, 
+margins = c(2, 1, 1, 2)))
+@
+\caption{Conditional mosaic plot for the \data{punishment} data. For
+  each mosaic plot, memory is in the rows (first split) and attitude
+  in the columns (second split).}
 \label{fig:punishcotab}
@@
 
-\bibliographystyle{plain}
+\bibliographystyle{abbrv}
 \bibliography{vcd}

### 2006-04-25 trunk/vcd/paper/hsc.Rnw
revised section 4 in HSC chapter
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@524 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 In the following, we will first treat specialized displays for
-explorative purposes, followed by model-based methods for
+exploratory purposes, followed by model-based methods for
 conditional independence models.
 
-\subsection{Explorative Visualization Techniques}
+\subsection{Exploratory Visualization Techniques}
 \label{sec:flat}
@@
 
-A first step in the explorative analysis of more complex tables is to
+A first step in the exploratory analysis of more complex tables is to
 get a quick overview on the data. For that, 
@@
 and that the proportion of male and female students varies across the departments.
+In particular, departments A and B have a higher proportion of male students
+and a lower rejection rate compared to the other departments.
 
@@
 \end{figure}
-
-%FIXME% Didn't we decide to use fewer plots in the paper and simply
-%FIXME% point out that other combinations could also be used?
-%FIXME% I would use for UCB:
-%FIXME%   o explorative: highlighted doubledecker (currently Fig. 17)
-%FIXME%   o model-based: conidtional shaded assoc (currently Fig. 20)
-%FIXME% For Punishment:
-%FIXME%   o explorative: highlighted mosaic( currently Fig. 21)
-%FIXME%   o model-based: conditional shaded mosaic (currently Fig. 23)
-%FIXME% The others should be omitted (I think) and the space gained should
-%FIXME% be used for making the individual graphs larger. Preferably, we
-%FIXME% should have the explorative and model-based plots for the same data
-%FIXME% set on the same page, emphasizing that different aspects of the data
-%FIXME% are brought out by the graphs.
 
@@
 that shall be ``explained'' by the others. 
-A natural way of representing such a three-way table is 
-to draw a doubledecker plot nesting gender in
-department and highlighting admission in the tiles (see Figure~\ref{fig:doubleucb}).
-From the widths of the tiles it can be seen that an
-unequal number of students apply for the six
-departments, and from the highlighting we see that the admission rates differ as
-well among the departments (roughly speaking, they decrease from A to F). The rates 
+A natural way of representing such a three-way table is use a mosaic display
+splitting first in the explanatory variables department and gender and highlighting
+the resulting mosaic with respect to the dependent variable admissions. Here, we
+use the vertical splits for both explanatory variables, resulting in the 
+doubledecker plot in Figure~\ref{fig:doubleucb}. From the widths of the tiles it can be seen that an
+unequal number of students apply for the six departments (with a particularly small
+number of females in departments A and B),
+and from the highlighting we see again that the admission rates differ
+among the departments (roughly speaking, the admission rate is high for A and B,
+low for F and in between for C to E). The rates 
 are equal for male and female students, except for department A where
@@
 
-In addition to the explorative approaches 
+In addition to the exploratory approaches 
 to the visualization of multi-way tables outlined in 
@@
 
-A first approach is to use pairs plots for model search:
+A first approach is to use pairs plots (such as that in Figure~\ref{fig:pairs}) for model search:
 we can use the position of the cells in the pairs matrix to select an
@@
 
-In our example, from the explorative analysis, we already suspect that
+Another approach is to visualize (the deviations from) a particular fitted model 
+with similar ideas as in Section~\ref{sec:extensions}. Thus, mosaic displays can
+be extended from purely explorative views to model-based views on the data by
+residual-based shadings and association plots can directly visualize deviations
+from a given model. Typically, the models considered are log-linear models, however
+residual-based shadings are also conceivable for other types of models such as logistic
+regression. %FIXME% mention this?
+In our example, from the exploratory analysis, we already suspect that
 admission and gender are independent, given department, except for
@@
 The corresponding mosaic and association plots are shown
-in Figures~\ref{fig:ucbmosaic}--\ref{fig:ucbassoc}. The shading visualizes the 
+in Figures~\ref{fig:ucbmosaic} and \ref{fig:ucbassoc}. The shading visualizes the 
 model of admission and gender being conditionally independent, given department. Clearly,
@@
 
-\setkeys{Gin}{width=0.7\textwidth}
-\begin{figure}
+\setkeys{Gin}{width=\textwidth}
+\begin{figure}[t!]
 \begin{center}
@@
 data by using the conditioning variables first for splitting.
-However, using a single plot, the strata are not corrected for the
-marginal distribution of the stratifying variables. Especially for
+However, this also displays the marginal distribution of the conditioning variables
+which might make it more difficult to interpret the dependent variables of
+interest. Especially for
 very unevenly distributed marginals, the strata can become very
 distorted, and the tiles, along with their shading, inscrutable.
-An alternative is to use trellis-like layouts for visualizing \emph{partial}
-tables for given strata, defined by the conditioning variables. 
+An alternative is to complement the conditioning in the model by
+a conditioning in the plot, i.e., to use trellis-like layouts for
+visualizing \emph{partial} tables as defined by the conditioning variables. 
 All subtables are corrected for the marginals and thus, all
 corresponding plots in the panels have the same size. 
-In addition, trellis layouts help reduce the complexity of bigger tables. 
-Figure~\ref{fig:ucbadmissions} 
-visualizes the data by means of a conditional 
+In addition, trellis layouts help to reduce the complexity of bigger tables. 
+Figure~\ref{fig:ucbadmissions} visualizes the data by means of a conditional 
 association plot. Each panel corresponds to a department, and contains
@@
 expected under the null hypothesis) causes the rejection of the
-hypothesis of conditional independence. 
+hypothesis of conditional independence. In Figure~\ref{fig:ucbadmissions},
+data-driven cut-offs (derived from the maximum test for conditional independence)
+are used, resulting in a more colorful shading than the fixed cut-offs in
+Figure~\ref{fig:ucbmosaic} and \ref{fig:ucbassoc}.
+
 % The overall deviation pattern
@@
 
-We first create a mosaic plot to get an overview of the data:
+In a first step, we create a mosaic plot for an exploratory view on the data:
 Attitude towards punishment clearly is the dependent variable 
-here and should be used last for splitting. Furthermore, we might 
-expect age and education to have stratifying influence on the results, so they should
-be used first for splitting.
-In fact, the question is whether memory has some influence on attitude, given 
+here and should be used as the last split via highlighting. 
+Furthermore, we expect age and education to have an influence on both the memory
+and the attitude, hence they should be used first for splitting as stratifying
+variables. In fact, the question is whether memory has some influence on attitude, given 
 age and education. We choose to create a three-way mosaic plot of the 
@@
 We can see that half of the people have more than 40 years of
-age. Most people in this group have only completed elementary
-school. For people over 25, a majority has completed secondary
-school. 
-%% FIXME: additional plot?
-Among the younger people, fewer have memories of having been
-punished as a child than among the older ones. 
-%%
-But surprisingly enough,
-for people with elementary or secondary school education,
+age, most of which of only completed elementary school (much more than 
+in the other age groups). Furthermore, it can be seen that the proportion
+of people with memories of punishments increases with age and that
+the approval rate decreases with education. As for the question whether
+attitude depends on memory, the plot suggests quite clearly (and somewhat
+surprisingly) that for those people with elementary school education,
 a higher proportion \emph{with} memories tends to
@@
 experienced violence seems to engender violence again.
-However, for people under 40, the proportion with
-people refusing punishment increases with the educational level.
+For the other education groups, the picture is less clear: some cells 
+indicate the same association while others do not.
 
@@
 tab <- xtabs(Freq ~ memory + attitude + age + education, data = Punishment)
+dimnames(tab)[[3]][3] <- "40+"
 mosaic(~ age + education + memory + attitude, data = tab, keep = FALSE,
        gp = gpar(fill = grey.colors(2)), spacing = spacing_highlighting, 
-       rep = c(attitude = F), set_varnames = c(attitude = "attitude to punishment"),
-       set_labels = list(age = c("15-24","25-39","40+")))
+       rep = c(attitude = FALSE))
 @
@@
 
-So both age and education seem to matter here. But is there really any influence
-of memory on the attitude towards 
+In summary, both age and education clearly matter here.
+But is there really any significant influence of memory on the attitude towards 
 punishment apart from that? Or, rephrased as a hypothesis:
-are memory and attitude conditionally independent, given age and education?
-We can visualize the model by means of a trellis layout, this time using partial mosaic
-plots (see Figure~\ref{fig:punishcotab}) for the strata defined by all factorial 
-combinations of the age and education levels. In each panel, we plot mosaic plots of 
-memory and attitude, where again the explanatory variable (memory) is used first 
-for splitting.
-The shading tells us the model seems to be consistent with the data, 
-except for people over 25
-with elementary school education: under the null, more people should
-have refused physical punishment of children than actually did.
+Are memory and attitude conditionally independent, given age and education?
+To focus on this question, we will take another model-based view on the data.
+To reduce the complexity of the information to the question of interest, we employ
+partial mosaic plots in a trellis layout (conditioning on age and education) with
+residual-based shading and data-driven cut-offs (see Figure~\ref{fig:punishcotab}).
+This visualization brings out much more clearly that the increased approval rate
+among people with memories of punishments is present mainly in the elementary education
+column and, to a lesser degree, in the age 40+ row. Moreover, this association is
+only significant in two cells (the two older age groups with elementary education)
+where fewer people without memories do not approve punishment than expected under
+(conditional) independence.
 
@@
 which is not of interest in the conditional independence problem. The
- conditional mosaic plot suppresses the effect of all stratifying
- variables (and their interactions).
+partial mosaic plot suppresses this effect by complementing the 
+conditioning in the model with conditioning in the visualization.
 
@@
 <<punishcotab,fig=TRUE,echo=FALSE,height=10,width=10,results=hide>>=
-cotabplot(tab, panel = cotab_coindep, panel_args = list(varnames = FALSE, 
-margins = c(2, 1, 1, 2)))
+cotabplot(tab, panel = cotab_coindep, panel_args = list(varnames = FALSE, margins = c(2, 1, 1, 2)))
 @
@@
 
-Mosaic, association, and sieve plots can be used for multi-way tables
-by applying them to flat representations. In addition, these basic
-plots are leveraged by specialized frameworks 
-for exploratory model search (pairs plots, highlighting, doubledecker plots) 
-and stratified analysis (conditional plots).
+Mosaic, association, and sieve plots can be used for visualizing multi-way tables
+by applying them to flat representations. Mosaic plots are particularly useful
+for exploratory analysis whereas the other two require specification of a certain
+independence model for which deviations should be brought out. In addition, specialized
+`flavours' of the plots can leverage the
+exploratory (pairs plots, highlighting, doubledecker plots) 
+or model-based analysis (residual-based shading, conditional plots).
 

### 2006-05-01 trunk/vcd/paper/hsc.Rnw
next revision ...
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@525 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 In the following, we describe the main idea of mosaic plots, 
-Chapter XXX in this book \emph{\bf (Reference to other contributions)} 
-provides more detailed information. Consider our
+Chapter III-12 in this book provides more detailed information. Consider our
 example of the hospital data from above. Step 1 consists of splitting a square
@@
 categorical data with a binary dependent variable. More variations on
-this theme, such as doubledecker plots, are treated in Chapter XXX
-(\textbf{REFERENCE TO OTHER CHAPTER(S)}).
+this theme, such as doubledecker plots, are treated in Chapter III-12.
 
@@
 might provide an easier way to detect deviation patterns than the
-residual-based shading added to the mosaic plot.
+residual-based shading added to the mosaic plot. Yet another approach
+is to plot a mosaic plot of the \emph{expected} frequencies and to highlight
+the (relative) residuals in the cells, using different colors for
+positive and negative residuals \cite{vcd:Theus+Lauer:1999}.
 
@@
 of people with memories of punishments increases with age and that
-the approval rate decreases with education. As for the question whether
+the approval rate decreases with education (this, however, is better shown in 
+Figure~\ref{fig:punishcotab} described below). As for the question whether
 attitude depends on memory, the plot suggests quite clearly (and somewhat
@@
 Are memory and attitude conditionally independent, given age and education?
-To focus on this question, we will take another model-based view on the data.
-To reduce the complexity of the information to the question of interest, we employ
+To focus on this question, we will take another model-based view on the data
+that avoids the distracting influence of the marginal variables (age
+and education): We employ
 partial mosaic plots in a trellis layout (conditioning on age and education) with
@@
 among people with memories of punishments is present mainly in the elementary education
-column and, to a lesser degree, in the age 40+ row. Moreover, this association is
+column and, to a lesser degree, in the age 40+ row. Using, e.g., a 
+permutation test similar to the one discussed in Section \ref{sec:test}, 
+model statistics reveal that the hypothesis is rejected at a significance level of 
+1\%, but the shading clearly shows that this association is
 only significant in two cells (the two older age groups with elementary education)
 where fewer people without memories do not approve punishment than expected under
-(conditional) independence.
+(conditional) independence. In the group of people with age between
+25--39, the lighter shading indicates that here,
+the association is only significant at the 10\% level.
 

### 2006-05-02 trunk/vcd/paper/hsc.Rnw
final Z touches on HSC chapter
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@526 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 as for the visualization of particular GLMs (such as log-linear models). 
-In Sect.~\ref{sec:twoway}, we start with the
+In Section~\ref{sec:twoway}, we start with the
 simple bivariate case.  Section~\ref{sec:extensions} explains how the use of color
@@
 by visualizing test statistics and residuals of independence models.
-In Sect.~\ref{sec:multiway}, we show how the
+In Section~\ref{sec:multiway}, we show how the
 basically bivariate methods straightforwardly extend to the multivariate case by using
@@
 Throughout this section, our examples will be based on
-the \data{hospital} data \cite{vcd:wing:1962} given in Tab.~\ref{tab:hospital}.
+the \data{hospital} data \cite{vcd:wing:1962} given in Table~\ref{tab:hospital}.
 
@@
 visits, which can be seen from the column-standardized table 
-(see Tab.~\ref{tab:colhospital}).
+(see Table~\ref{tab:colhospital}).
 
@@
 significantly different in the ``less than monthly'' and ``never''
-strata. From the row-standardized table (see Tab.~\ref{tab:rowhospital}), 
+strata. From the row-standardized table (see Table~\ref{tab:rowhospital}), 
 it seems indeed that LOS is homogeneous with respect to these two visit frequency strata.
@@
 $\hat n_{ij} = n_{i+}n_{+j}/n$.
-The expected table for our sample data is given in Tab.~\ref{tab:hospital2}.
+The expected table for our sample data is given in Table~\ref{tab:hospital2}.
 It could again be visualized using a mosaic plot, this
@@
     monochrome, but the electronic version uses colors. The colored
-    versions of the plots are available from the code web page 
-    (\url{http://statmath.wu-wien.ac.at/projects/vcd/}).}}
+    versions of the plots are available from the Web page 
+    \url{http://statmath.wu-wien.ac.at/projects/vcd/}.}}
 \label{sec:colors}
@@
 
-All plots introduced in Sect.~\ref{sec:twoway} are composed of empty
+All plots introduced in Section~\ref{sec:twoway} are composed of empty
 tiles. It seems intuitive to use filled tiles to highlight information
@@
 results. These can again be derived using the same procedure
-described in Sect.~\ref{sec:colors} but using a smaller amount of color, i.e.,
+described in Section~\ref{sec:colors} but using a smaller amount of color, i.e.,
 a smaller maximal chroma (e.g., 20 instead of 100).
@@
 investigating a new treatment for rheumatoid arthritis, stratified by
-gender (see Tab.~\ref{tab:arthritis} for the female patients):
+gender (see Table~\ref{tab:arthritis} for the female patients):
 \begin{table}[h]
@@
 Shadings can be used to add information to the basic plots and to support the
-analysis. Highlighting of tiles can support the analysis relationships with a 
+analysis. Highlighting of tiles can support the analysis of relationships with a 
 single dependent variable. Residual-based shadings can be used for model diagnostics, e.g.,
@@
 
-In Sect.~\ref{sec:twoway}, we have presented basic displays
+In Section~\ref{sec:twoway}, we have presented basic displays
 for two-way tables, based on the visualization of information on 
@@
 columns and admit---nested in gender---into the rows, is given by
-Tab.~\ref{tab:ucb}. The table aggregated over all
+Table~\ref{tab:ucb}. The table aggregated over all
 departments gives wrong evidence of gender bias, 
@@
 residual-based shadings are also conceivable for other types of models such as logistic
-regression. %FIXME% mention this?
+regression.
 In our example, from the exploratory analysis, we already suspect that
@@
 % people (willing to divorce) were asked two questions about their pre- and
-% extramarital sexual experiences, resulting in Tab.~\ref{tab:presex}.
+% extramarital sexual experiences, resulting in Table~\ref{tab:presex}.
 
@@
 the attitude of a random sample of 783 persons towards corporal
-punishment of children (see Tab.~\ref{tab:punish}).      
+punishment of children (see Table~\ref{tab:punish}).      
 The data consist of four variables: Attitude is a binary variable
@@
 <<punishcotab,fig=TRUE,echo=FALSE,height=10,width=10,results=hide>>=
+set.seed(123)
 cotabplot(tab, panel = cotab_coindep, panel_args = list(varnames = FALSE, margins = c(2, 1, 1, 2)))

### 2006-05-03 trunk/vcd/paper/hsc.Rnw
KH improvements.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@528 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 As an example, consider the well-known UCB admissions data
-\cite{vcd:Bickel+Hammel+O'Connell:1975}
-on applicants to graduate school at the University of California in Berkeley for
-the six largest departments in 1973 classified by admission and
-gender. The `flattened' contingency table, putting department into the
-columns and admit---nested in gender---into the rows, is given by
-Table~\ref{tab:ucb}. The table aggregated over all
-departments gives wrong evidence of gender bias, 
-i.e., higher admission rates for male students.
+\cite{vcd:Bickel+Hammel+O'Connell:1975} on applicants, classified by
+admission and gender, to graduate school at the University of California
+in Berkeley for the six largest departments in 1973. The `flattened'
+contingency table, putting department into the columns and
+admit---nested in gender---into the rows, is given by
+Table~\ref{tab:ucb}. The table aggregated over all departments gives
+wrong evidence of gender bias, i.e., higher admission rates for male
+students.
 
@@
 A first step in the exploratory analysis of more complex tables is to
-get a quick overview on the data. For that, 
+get a quick overview on the data. For this, 
 all basic plots can be combined in pairwise displays,
@@
 The diagonal cells contain the variable names, optionally with
-univariate statistics, whereas the off-diagonal cells contain plots whose
+univariate statistics, whereas the off-diagonal cells feature plots whose
 variables are implicitly specified by the cells' position in the matrix. %'
@@
 that shall be ``explained'' by the others. 
-A natural way of representing such a three-way table is use a mosaic display
+A natural way of representing such a three-way table is to use a mosaic
+display, 
 splitting first in the explanatory variables department and gender and highlighting
@@
 use the vertical splits for both explanatory variables, resulting in the 
-doubledecker plot in Figure~\ref{fig:doubleucb}. From the widths of the tiles it can be seen that an
-unequal number of students apply for the six departments (with a particularly small
-number of females in departments A and B),
+doubledecker plot in Figure~\ref{fig:doubleucb}. From the widths of the
+tiles it can be seen that students apply to the six departments in
+unequal numbers (with a particularly small
+number of females in departments~A and B),
 and from the highlighting we see again that the admission rates differ
@@
 
-A first approach is to use pairs plots (such as that in Figure~\ref{fig:pairs}) for model search:
+One approach is employing pairs plots (such as that in Figure~\ref{fig:pairs}) for model search:
 we can use the position of the cells in the pairs matrix to select an
@@
 Therefore, a sensible way of representing the table is to 
-nest admission into department instead of gender to get a stratified view.
+nest admission into department instead of gender to obtain a stratified view.
 The corresponding mosaic and association plots are shown
@@
 As we have seen, we can use all plots for the analysis of stratified
-data by using the conditioning variables first for splitting.
-However, this also displays the marginal distribution of the conditioning variables
+data by  first using the conditioning variables for splitting.
+However, this also displays the marginal distributions of the conditioning variables
 which might make it more difficult to interpret the dependent variables of
@@
 Are memory and attitude conditionally independent, given age and education?
-To focus on this question, we will take another model-based view on the data
-that avoids the distracting influence of the marginal variables (age
+To focus on this question, we take another model-based view on the data
+which avoids the distracting influence of the marginal variables (age
 and education): We employ
@@
 association plots are suitable for the visualization of observed and
-expected values and the Pearson residuals, respectively. This basic
+expected values and the Pearson residuals, respectively. These basic
 methods can be enhanced by using residual-based shadings, preferably

### 2006-05-03 trunk/vcd/paper/hsc.Rnw
re-arranged HSC graphics, handled dimnames of tables consistently
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@529 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 <<preliminaries,echo=FALSE,results=hide>>=
+## random seed
 myseed <- 1071
 set.seed(myseed)
-library(vcd)
-library(scatterplot3d)
-library(xtable)
-data(Arthritis)
-data(UCBAdmissions)
+
+## packages
+library("vcd")
+library("scatterplot3d")
+library("xtable")
+
+## data
 hospital <- as.table(matrix(c(43,16,3,6,11,10,9,18,16), c(3,3), byrow = TRUE, 
@@
                   "Length of stay" = c("2-9","10-19","20+"))))
-art <- xtabs(~Treatment + Improved, data = Arthritis, subset = Sex == "Female")
-
+
+data("Arthritis")
+art <- xtabs(~ Treatment + Improved, data = Arthritis, subset = Sex == "Female")
+names(dimnames(art))[2] <- "Improvement"
+
+data("UCBAdmissions")
+names(dimnames(UCBAdmissions)) <- c("Admission", "Gender", "Department")
+
+data("Punishment")
+punish <- xtabs(Freq ~ memory + attitude + age + education, data = Punishment)
+dimnames(punish)[[3]][3] <- "40+"
+
+## inference
 set.seed(myseed)
@@
 set.seed(myseed)
-ucb_max <- coindep_test(aperm(UCBAdmissions, c(3,2,1)), margin = "Dept", n = 5000)
+ucb_max <- coindep_test(aperm(UCBAdmissions, c(3,2,1)), margin = "Department", n = 5000)
 
@@
 hos_chisq <- coindep_test(hospital, n = 5000, indepfun = function(x) sum(x^2))
-
 @
@@
 \pkg{scatterplot3d} \cite{vcd:ligges+maechler:2003}
-(Figure~\ref{fig:threedee} only), all freely available from the
+(Figure~\ref{fig:bars3d} only), all freely available from the
 Comprehensive \proglang{R} Archive Network
@@
 
-Although far from optimal, contingency tables are frequently visualized using
-grouped bar plots (see Figure~\ref{fig:grouped}) 
-or even by means of 3D-bar charts (see Figure~\ref{fig:threedee}). It seems hard
-to detect the aforementioned pattern in these, especially in the 3D
-plot where the perspective view tends to distort the true proportions
-of the bars. In the following, we
-will introduce three graphical methods that are better suited for
-contingency tables.
-
 \setkeys{Gin}{width=0.7\textwidth}
@@
 \begin{center}
-<<grouped,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
+<<barsgrouped,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
 barplot(hospital, legend = rownames(hospital), beside = TRUE, 
-        xlab = "Length of stay (years)", ylab = "Number of patients")
+        xlab = "Length of stay", ylab = "Number of patients")
 @
 \caption{Bar plot for the \data{hospital} data.}
-\label{fig:grouped}
+\label{fig:barsgrouped}
 \end{center}
@@
 \begin{center}
-<<threedeebars,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
+<<bars3d,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
 myhospital = t(hospital)[,3:1]
@@
               y.ticklabs = c("Never","","Less than monthly","","Regular"), 
-              xlab = "Length of stay (years)", ylab = "Visit frequency", zlab = "Number of patients",
+              xlab = "Length of stay", ylab = "Visit frequency", zlab = "Number of patients",
               y.margin.add = 0.2,
@@
 \caption{3D-bar chart for the \data{hospital} data.}
-\label{fig:threedee}
-\end{center}
-\end{figure}
+\label{fig:bars3d}
+\end{center}
+\end{figure}
+
+Although far from optimal, contingency tables are frequently visualized using
+grouped bar plots (see Figure~\ref{fig:barsgrouped}) 
+or even by means of 3D-bar charts (see Figure~\ref{fig:bars3d}). It seems hard
+to detect the aforementioned pattern in these, especially in the 3D
+plot where the perspective view tends to distort the true proportions
+of the bars. In the following, we
+will introduce three graphical methods that are better suited for
+contingency tables.
 
@@
 the corresponding number of \emph{observed} frequencies, we get a \emph{sieve plot}
-(see Figure~\ref{fig:sievefig}). It implicitly compares
+(see Figure~\ref{fig:sieve}). It implicitly compares
 expected and observed values since the density of the grid will
@@
 \begin{center}
-<<sievefig,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
+<<sieve,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
 sieve(t(hospital), split = TRUE, pop = FALSE, gp = gpar(lty =
@@
 \caption{Sieve plot for the \data{hospital} data.}
-\label{fig:sievefig}
+\label{fig:sieve}
 \end{center}
@@
 \begin{center}
-<<colorfig,fig=TRUE,echo=FALSE,height=3,width=7,results=hide>>=
+<<color,fig=TRUE,echo=FALSE,height=3,width=7,results=hide>>=
 par(mfrow = c(1,2), mar = c(1,1,1,1), oma = c(0,0,0,0))
@@
 \hline
- & \multicolumn{4}{c|}{\textbf{Improved}}\\
+ & \multicolumn{4}{c|}{\textbf{Improvement}}\\
 \textbf{Treatment} & None & Some & Marked & $\Sigma$ \\
@@
 \begin{center}
-<<Arthritis,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
+<<arthritis,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
 mosaic(art, shade = TRUE, gp_args = list(p.value = art_chisq$p.value))
@@
 \begin{center}
-<<Arthritis2fig,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
+<<arthritis2,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
 set.seed(myseed)
@@
 contingency table, putting department into the columns and
-admit---nested in gender---into the rows, is given by
+admission---nested in gender---into the rows, is given by
 Table~\ref{tab:ucb}. The table aggregated over all departments gives
@@
  & & \multicolumn{6}{c|}{\textbf{Department}} \\
-\textbf{Gender} & \textbf{Admit} & A & B & C & D & E & F\\\hline
+\textbf{Gender} & \textbf{Admission} & A & B & C & D & E & F\\\hline
 Male   & Admitted & 512 & 353 & 120 & 138 & 53 & 22\\
@@
 \begin{center}
-<<doubleucb,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
+<<ucbhigh,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
 doubledecker(aperm(aperm(UCBAdmissions, c(1,3,2))[2:1,,], c(2,3,1)), margins = c(left = 0, right = 5),col = rev(grey.colors(2)))
@@
 in Figures~\ref{fig:ucbmosaic} and \ref{fig:ucbassoc}. The shading visualizes the 
-model of admission and gender being conditionally independent, given department. Clearly,
-there is no gender bias in the departments except for department A: here, 
+model of admission and gender being conditionally independent, given department.
+\setkeys{Gin}{width=\textwidth}
+\begin{figure}[t!]
+\begin{center}
+<<ucbmosaic,fig=TRUE,echo=FALSE,height=6,width=12,results=hide>>=
+mosaic(aperm(UCBAdmissions, c(3,2,1)), data = UCBAdmissions, split = TRUE,
+shade = TRUE, keep = FALSE, residuals = ucb_max$residuals,
+gp_args = list(p.value = ucb_max$p.value), rep = c(Admission = FALSE)) 
+@
+\caption{Mosaic plot for the \data{UCB admissions} data.}
+\label{fig:ucbmosaic}
+\end{center}
+\end{figure}
+\setkeys{Gin}{width=\textwidth}
+\begin{figure}[t!]
+\begin{center}
+<<ucbassoc,fig=TRUE,echo=FALSE,height=6,width=12,results=hide>>=
+assoc(aperm(UCBAdmissions, c(3,2,1)), split = c(TRUE,FALSE,TRUE), shade = TRUE, 
+residuals_type = "Pearson", residuals = ucb_max$residuals,
+gp_args = list(p.value = ucb_max$p.value), rep = c(Admission = FALSE))
+@
+\caption{Association plot for the \data{UCB admissions} data.}
+\label{fig:ucbassoc}
+\end{center}
+\end{figure}
+Clearly, there is no gender bias in the departments except for department A: here, 
 significantly \textit{more} female students are admitted then expected under independence.
@@
 positive and negative residuals \cite{vcd:Theus+Lauer:1999}.
-
-\setkeys{Gin}{width=\textwidth}
-\begin{figure}[t!]
-\begin{center}
-<<ucbmosaic,fig=TRUE,echo=FALSE,height=6,width=12,results=hide>>=
-mosaic(aperm(UCBAdmissions, c(3,2,1)), data = UCBAdmissions, split = TRUE,
-shade = TRUE, keep = FALSE, residuals = ucb_max$residuals,
-gp_args = list(p.value = ucb_max$p.value), rep = c(Admit = FALSE)) 
-@
-\caption{Mosaic plot for the \data{UCB admissions} data.}
-\label{fig:ucbmosaic}
-\end{center}
-\end{figure}
-
-\setkeys{Gin}{width=\textwidth}
-\begin{figure}[t!]
-\begin{center}
-<<ucbassoc,fig=TRUE,echo=FALSE,height=6,width=12,results=hide>>=
-assoc(aperm(UCBAdmissions, c(3,2,1)), split = c(TRUE,FALSE,TRUE), shade = TRUE, 
-residuals_type = "Pearson", residuals = ucb_max$residuals,
-gp_args = list(p.value = ucb_max$p.value), rep = c(Admit = FALSE))
-@
-\caption{Association plot for the \data{UCB admissions} data.}
-\label{fig:ucbassoc}
-\end{center}
-\end{figure}
 
@@
 \begin{center}
-<<UCBAdmissions,fig=TRUE,echo=FALSE,height=6,width=10,results=hide>>=
+<<ucbcotab,fig=TRUE,echo=FALSE,height=6,width=10,results=hide>>=
 cotabplot(aperm(UCBAdmissions, c(2,1,3)), panel = cotab_coindep, shade = TRUE, 
@@
   data. For each individual association plot, gender is in the rows
-  and admit is in the columns.}
+  and admission is in the columns.}
 \label{fig:ucbadmissions}
@@
 \end{table}
-
-%FIXME% Better structure:
-%FIXME%   o dependent variable: attitude
-%FIXME%   o dependence on age and education is rather clear
-%FIXME%     (approval of moderate punishment increases with age and
-%FIXME%     decreases with education)
-%FIXME%   o question is conditional independence of attitude and
-%FIXME%     memory given the other margins
-%FIXME%   o hence for exploratory analysis: split in conditioning
-%FIXME%     variables (age, education) first, then explanatory variable (memory),
-%FIXME%     then dependent variable (attitude) for which highlighting is used.
-%FIXME%   o for model-based view: suppress marginal distribution of
-%FIXME%     conditioning variables (and their interaction), split first
-%FIXME%     in explanatory variable in each cell, use shading based on
-%FIXME%     residuals in conditional independence model.
-%FIXME%   o advantage of exploratory view: joint distribution of all variables
-%FIXME%     can be seen. disadvantage: visualization strongly influence
-%FIXME%     by the marginal distribution of education over age (in particular
-%FIXME%     the large proportion of 40+ people with elementary education)
-%FIXME%     which is not of interest in the conditional independence problem.
 
@@
 <<punishhigh,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-tab <- xtabs(Freq ~ memory + attitude + age + education, data = Punishment)
-dimnames(tab)[[3]][3] <- "40+"
-mosaic(~ age + education + memory + attitude, data = tab, keep = FALSE,
+mosaic(~ age + education + memory + attitude, data = punish, keep = FALSE,
        gp = gpar(fill = grey.colors(2)), spacing = spacing_highlighting, 
@@
 % ## <<punishmosaic,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-% ## tab_test <- coindep_test(tab, margin = c(3, 4))
-% ## mosaic(~ age + education + memory + attitude, data = tab, keep = FALSE,
+% ## punish_test <- coindep_test(punish, margin = c(3, 4))
+% ## mosaic(~ age + education + memory + attitude, data = punish, keep = FALSE,
 % ##        expected = ~ (memory + attitude) * education * age,
-% ##        gp_args = list(interpolate = tab_test$qdist(c(0.9, 0.99))), 
+% ##        gp_args = list(interpolate = punish_test$qdist(c(0.9, 0.99))), 
 % ##        rep = c(attitude = F), set_varnames = c(attitude = "attitude to punishment"),
@@
 set.seed(123)
-cotabplot(tab, panel = cotab_coindep, panel_args = list(varnames = FALSE, margins = c(2, 1, 1, 2)))
+cotabplot(punish, panel = cotab_coindep, panel_args = list(varnames = FALSE, margins = c(2, 1, 1, 2)))
 @

### 2006-05-26 trunk/vcd/paper/Color/draft.Rnw
re-organized papers, added color draft
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@536 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\documentclass{Z}
%% need no \usepackage{Sweave}
\SweaveOpts{engine=R, eps=FALSE, echo=FALSE, results=hide}

\title{Choosing Color Palettes for Statistical Graphics}
\author{Achim Zeileis and Kurt Hornik\\Wirtschaftsuniversit\"at Wien}
\Plainauthor{Achim Zeileis, Kurt Hornik}
\Keywords{qualitative palette, sequential palette, diverging palette, HCL colors, HSV colors, perceptual-based color space}

\Abstract{
  Statistical graphs are often augmented by the use of color
  coding information contained in some variable. When this involves
  the shading of areas (and not only points or lines)---e.g., as in barplots, pie charts,
  mosaic displays or heatmaps---it is important that the colors are perceptual-based
  and do not introduce optical illusions or systematic bias. Here, we discuss
  how the perceptual based HCL color space can be used for deriving suitable
  color palettes for coding categorical data (qualitative palettes) and numerical
  variables (sequential and diverging palettes).
}

\begin{document}

<<preliminaries>>=
library("vcd")
source("functions.R")
generate_png <- TRUE
@

\setkeys{Gin}{width=0.8\textwidth}
\begin{figure}[p]
\begin{center}
<<hsv-hcl>>=
if(generate_png) {
  files <- c("hcl-dcone", "hsv-cylinder")
  PNG <- function(x) paste(x, ".png", sep = "")

  png(file = PNG(files[1]), height = 480, width = 720)
  hcl.wheel(luminance = seq(5, 95, length = 6))
  dev.off()
  png(file = PNG(files[2]), height = 480, width = 720)
  hsv.wheel(value = seq(0, 1, length = 6), cone = FALSE)
  dev.off()
}
@
\includegraphics{hsv-cylinder}

\vspace*{2.5cm}

\includegraphics{hcl-dcone}
\caption{\label{fig:hcl-hsv} HSV and HCL space}
\end{center}
\end{figure}

\begin{figure}[p]
\begin{center}
<<hsv-hcl-gray>>=
if(generate_png) {
  for(i in files) {
    file.copy(PNG(i), PNG(paste(i, "-gray", sep = "")), overwrite = TRUE)
    system(paste("mogrify -type Grayscale", PNG(paste(i, "-gray", sep = ""))))
  }
}
@
\includegraphics{hsv-cylinder-gray}

\vspace*{2.5cm}

\includegraphics{hcl-dcone-gray}
\caption{\label{fig:hcl-hsv-gray} HSV and HCL space (in gray levels)}
\end{center}
\end{figure}

\setkeys{Gin}{width=0.8\textwidth}
\begin{figure}[h]
\begin{center}
<<pal-q-construct>>=
if(generate_png) {
  png(file = "pal-q-construct.png", height = 480, width = 480)
  hcl.wheel(luminance = 70, do.label = FALSE, inner.circle = TRUE)
  rect(0.45, -0.05, 0.55, 0.05, col = hcl(0, 50, 70))
  rect(-cos(pi/3)/2-0.05, sin(pi/3)/2-0.05, -cos(pi/3)/2+0.05, sin(pi/3)/2+0.05, col = hcl(120, 50, 70))
  rect(-cos(pi/3)/2-0.05, -sin(pi/3)/2-0.05, -cos(pi/3)/2+0.05, -sin(pi/3)/2+0.05, col = hcl(240, 50, 70))
  dev.off()
}
@
\includegraphics{pal-q-construct}
\caption{\label{fig:pal-q-construct} Constructing qualitative palettes}
\end{center}
\end{figure}

\setkeys{Gin}{width=.8\textwidth}
\begin{figure}[h]
\begin{center}
<<pal-s-construct>>=
if(generate_png) {
  png(file = "pal-s-construct.png", height = 480, width = 400)
  hue.slice(260, grid.n = 301)
  lines(c(0, 100), c(90, 50), lty = 2)
  lines(c(0, 80), c(90, 30), lty = 2)
  lines(c(2.5, 2.5), c(90, 30), lty = 2)
  rect(0, 87.5, 5, 92.5, col = hcl(260, 0, 90))
  rect(47.5, 67.5, 52.5, 72.5, col = hcl(260, 50, 70))
  rect(95, 47.5, 100, 52.5, col = hcl(260, 100, 50))
  rect(37.5, 57.5, 42.5, 62.5, col = hcl(260, 40, 60))
  rect(77.5, 27.5, 82.5, 32.5, col = hcl(260, 80, 30))
  rect(0, 57.5, 5, 62.5, col = hcl(260, 0, 60))
  rect(0, 27.5, 5, 32.5, col = hcl(260, 0, 30))
  dev.off()
}
@
\includegraphics{pal-s-construct}
\caption{\label{fig:pal-s-construct} Constructing sequential palettes}
\end{center}
\end{figure}

\setkeys{Gin}{width=\textwidth}
\begin{figure}[h]
\begin{center}
<<pal-d-construct>>=
if(generate_png) {
  png(file = "pal-d-construct.png", height = 480, width = 480)
  hue260 <- hue.slice(260, grid.n = 301, plot = FALSE)
  hue360 <- hue.slice(360, grid.n = 301, plot = FALSE)
  mychroma <- as.numeric(colnames(hue260))
  luminance <- as.numeric(rownames(hue260))
  nc <- ncol(hue260)
  nl <- nrow(hue260)
  plot(0.5, 0.5, xlim = c(-100, 100), ylim = c(0, 100), type = "n", axes = FALSE,
       xlab = "", ylab = "", yaxs = "i", xaxs = "i", main = "")
  for(i in 1:(nc-1)) {
    rect(mychroma[i], luminance[-nl], mychroma[i] + 100/(nc-1), luminance[-1], border = hue260[,i+1], col = hue260[,i+1])
  }
  for(i in 1:(nc-1)) {
    rect(-mychroma[i], luminance[-nl], -(mychroma[i] + 100/(nc-1)), luminance[-1], border = hue360[,i+1], col = hue360[,i+1])
  }
  axis(1, labels = FALSE, at = c(-100, 0, 100))
  axis(2, at = c(0, 50, 70, 90, 100), labels = c(0, 50, 70, 90, 100))
  axis(3, at = -4:4*25, labels=c(4:0*25, 1:4*25))
  axis(4, at = c(0, 50, 70, 90, 100), labels = c(0, 50, 70, 90, 100))
  mtext(c("hue = 0", "hue = 260"), side = 3, at = c(-50, 50), line = 3, cex = 1.2)
  mtext("chroma", side = 3, at = 0, line = 2)
  mtext("luminance", side = 2, at = 50, line = 2)
  mtext("luminance", side = 4, at = 50, line = 2)
  lines(c(-100, 0, 100), c(50, 90, 50), lty = 2)
  rect(-100, 47.5, -90, 52.5, col = hcl(0, 100, 50))
  rect(-55, 67.5, -45, 72.5, col = hcl(0, 50, 70))
  rect(-5, 87.5, 5, 92.5, col = hcl(260, 0, 90))
  rect(45, 67.5, 55, 72.5, col = hcl(260, 50, 70))
  rect(90, 47.5, 100, 52.5, col = hcl(260, 100, 50))
  lines(c(-80, 0, 80), c(30, 90, 30), lty = 2)
  rect(-85, 27.5, -75, 32.5, col = hcl(0, 80, 30))
  rect(-45, 57.5, -35, 62.5, col = hcl(0, 40, 60))
  rect(-5, 87.5, 5, 92.5, col = hcl(260, 0, 90))
  rect(35, 57.5, 45, 62.5, col = hcl(260, 40, 60))
  rect(75, 27.5, 85, 32.5, col = hcl(260, 80, 30))
  dev.off()
}
@
\includegraphics{pal-d-construct}
\caption{\label{fig:pal-d-construct} Constructing diverging palettes}
\end{center}
\end{figure}

\setkeys{Gin}{width=0.8\textwidth}
\begin{figure}[h]
\begin{center}
<<pal-q-examples, fig=TRUE, height=6, width=6>>=
par(mfrow = c(2, 2), mar = c(0, 0, 3, 0))
pal(rainbow_hcl(4, start = 30, end = 300), main = "dynamic (360 deg)")
pal(rainbow_hcl(4, start = 60, end = 240), main = "harmonic (180 deg)")
pal(rainbow_hcl(4, start = 270, end = 150), main = "cold (120 deg)")
pal(rainbow_hcl(4, start = 90, end = -30), main = "warm (120 deg)")
@
\caption{\label{fig:pal-q-examples} Examples for qualitative palettes}
\end{center}
\end{figure}

\setkeys{Gin}{width=.8\textwidth}
\begin{figure}[h]
\begin{center}
<<pal-s-examples, fig=TRUE, height=9, width=6>>=
par(mfrow = c(5, 1), mar = c(0, 0, 0, 0))
pal(sequential_hcl(12, c = 0, increase = 2.2))
pal(sequential_hcl(12, increase = 2.2))
pal(terrain_hcl(12, c = c(65, 0), l = c(45, 95), increase = c(1/3, 1.5)))
pal(heat_hcl(12, c = c(80, 30), l = c(30, 90), increase = c(1/5, 1.5)))
pal(rev(heat_hcl(12, h = c(0, -100), l = c(75, 40), c = c(40, 80), increase = 1)))
@
\caption{\label{fig:pal-s-examples} Examples for sequential palettes}
\end{center}
\end{figure}

\setkeys{Gin}{width=.8\textwidth}
\begin{figure}[h]
\begin{center}
<<pal-d-examples, fig=TRUE, height=8, width=6>>=
par(mfrow = c(4, 1), mar = c(0, 0, 0, 0))
pal(diverge_hcl(7))
pal(diverge_hcl(7, h = c(180, 330), c = 59, l = c(75, 95)))
pal(diverge_hcl(7, h = c(130, 43), c = 100, l = c(70, 90)))
pal(diverge_hcl(7, c = 100, l = c(50, 90), increase = 1))
@
\caption{\label{fig:pal-d-examples} Examples for diverging palettes}
\end{center}
\end{figure}

\end{document}

### 2006-05-26 trunk/vcd/paper/Color/functions.R

@@ -1,0 +1,96 @@
+pal <- function(col, border = "light gray", ...)
+{
+  n <- length(col)
+  plot(0, 0, type="n", xlim = c(0, 1), ylim = c(0, 1), axes = FALSE, xlab = "", ylab="", ...)
+  rect(0:(n-1)/n, 0, 1:n/n, 1, col = col, border = border)
+}
+
+hsv.wheel <-
+    function(value = c(0:2, 4:6, 8:10)/10, saturation = 0:100/100, hues = 0:300/300, asp = 1,
+             p.cex = 0.6, do.label = FALSE, cone = TRUE)
+{
+    stopifnot(is.numeric(saturation), saturation >= 0, saturation <= 1,
+              is.numeric(hues), hues >= 0, hues <= 1,
+              is.numeric(value), value >= 0, (nch <- length(value)) >= 1)
+    if(is.unsorted(hues)) hues <- sort(hues)
+    if(nch > 1) {
+        op <- par(mfrow= rev(n2mfrow(nch)), mar = c(0,0,0,0))
+        on.exit(par(op))
+    }
+    for(i.c in 1:nch) {
+        plot(-1:1,-1:1, type="n", axes = FALSE, xlab="",ylab="", asp = asp)
+        text(0.4, 0.99, paste("value =", format(value[i.c])), adj = 0, font = 4)
+	l.s <- if(cone) saturation * value[i.c] else saturation
+        for(ang in hues) {
+            a. <- ang * 2 * pi
+            z.a <- exp(1i * a.)
+            cols <- hsv(h = ang, s = saturation, v = value[i.c])
+            points(l.s * z.a, pch = 16, col = cols, cex = p.cex)
+            if(do.label) text(z.a*1.05, labels = ang, col = cols[length(cols)/2], srt = ang)
+        }
+        lines(exp(1i * hues * 2 * pi))
+   }
+   invisible()
+}
+
+hcl.wheel <-
+    function(luminance = 1:9*10, chroma = 0:100, hues = 1:360, asp = 1,
+             p.cex = 0.6, do.label = c(TRUE, FALSE), fixup = FALSE, inner.circle = FALSE)
+{
+    do.label <- rep(do.label, length.out = 2)
+    stopifnot(is.numeric(chroma), chroma >= 0, chroma <= 100,
+              is.numeric(hues), hues >= 0, hues <= 360,
[…]

### 2006-05-26 trunk/vcd/paper/DSC-2003/draft.Rnw
re-organized papers, added color draft
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@536 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2003-07-08 trunk/vcd/paper/draft.Rnw
@@
 \title{Visualizing Independence Using Extended Association and Mosaic Plots}
-\author{\hfill Achim Zeileis$^\dag$ \hfill David Meyer$^\dag$ \hfill Kurt Hornik$^\ddag$ \hfill \hfill \\[0.2cm]
-\dag {\it \small Institut f\"ur Statistik \& Wahrscheinlichkeitstheorie,
- Technische Universit\"at Wien}\\
-\ddag {\it \small Institut f\"ur Statistik, Wirtschaftsuniversit\"at Wien}
+\author{\hfill Achim Zeileis$^\dag$ \hfill David Meyer$^\ddag$ \hfill Kurt Hornik$^\dag$ \hfill \hfill \\[0.2cm]
+\dag {\it \small Institut f\"ur Statistik \& Mathematik, Wirtschaftsuniversit\"at Wien}\\
+\ddag {\it \small Institut f\"ur , Wirtschaftsuniversit\"at Wien}
 }

### 2006-05-26 trunk/vcd/R/palettes.R

@@ -6,62 +6,62 @@ rainbow_hcl <- function(n, c = 50, l = 70, start = 0, end = 360*(n-1)/n, ...)
     else character(0)
 }
 
-diverge_hcl <- function(n, h = c(260, 0), c = 80, l = c(30, 90), increase = 1.5, ...)
+diverge_hcl <- function(n, h = c(260, 0), c = 80, l = c(30, 90), power = 1.5, ...)
 {
   if(n < 1) return(character(0))
   h <- rep(h, length.out = 2)
   c <- c[1]
   l <- rep(l, length.out = 2)
-  increase <- rep(increase, length.out = 2)
+  power <- rep(power, length.out = 2)
   rval <- seq(1, -1, length = n)
   rval <- hcl(h = ifelse(rval > 0, h[1], h[2]),
-              c = c * abs(rval)^increase[1],
-              l = l[2] - diff(l) * abs(rval)^increase[2],
+              c = c * abs(rval)^power[1],
+              l = l[2] - diff(l) * abs(rval)^power[2],
               ...)
   return(rval)
 }
 
-diverge_hsv <- function(n, h = c(2/3, 0), s = 1, v = 1, increase = 1, ...)
+diverge_hsv <- function(n, h = c(2/3, 0), s = 1, v = 1, power = 1, ...)
 {
   if(n < 1) return(character(0))
   h <- rep(h, length.out = 2)
   s <- s[1]
   v <- v[1]
-  increase <- increase[1]
+  power <- power[1]
   rval <- seq(-s, s, length = n)
-  rval <- hsv(h = ifelse(rval > 0, h[2], h[1]), s = abs(rval)^increase, v = v, ...)
+  rval <- hsv(h = ifelse(rval > 0, h[2], h[1]), s = abs(rval)^power, v = v, ...)
   return(rval)
 }
 
-sequential_hcl <- function(n, h = 260, c. = c(80, 0), l = c(30, 90), increase = 1.5, ...)
+sequential_hcl <- function(n, h = 260, c. = c(80, 0), l = c(30, 90), power = 1.5, ...)
 {
   if(n < 1) return(character(0))
   c <- rep(c., length.out = 2)
   l <- rep(l, length.out = 2)
-  increase <- rep(increase, length.out = 2)
+  power <- rep(power, length.out = 2)
   rval <- seq(1, 0, length = n)
   rval <- hcl(h = h[1],
-              c = c[2] - diff(c) * rval^increase[1],
[…]

### 2006-05-27 trunk/vcd/paper/Color/draft.Rnw
first version of color palettes draft finished
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@537 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \Plainauthor{Achim Zeileis, Kurt Hornik}
-\Keywords{qualitative palette, sequential palette, diverging palette, HCL colors, HSV colors, perceptual-based color space}
+\Keywords{qualitative palette, sequential palette, diverging palette, HCL colors, HSV colors, perceptually-based color space}
 
@@
   the shading of areas (and not only points or lines)---e.g., as in barplots, pie charts,
-  mosaic displays or heatmaps---it is important that the colors are perceptual-based
+  mosaic displays or heatmaps---it is important that the colors are perceptually-based
   and do not introduce optical illusions or systematic bias. Here, we discuss
-  how the perceptual based HCL color space can be used for deriving suitable
+  how the perceptually-based HCL color space can be used for deriving suitable
   color palettes for coding categorical data (qualitative palettes) and numerical
@@
 
+\section{Introduction} \label{sec:intro}
+
 <<preliminaries>>=
@@
 @
+
+First, we motivate why HCL colors \citep{vcd:Ihaka:2003} are suitable for choosing color palettes
+by contrasting them with the more commonly implemented HSV colors.
+
+Then, we describe strategies how color palettes for categorical and numerical
+data can be chosen in this space. Following \cite{vcd:Brewer:1999}, we distinguish
+three types of palettes: qualitative, sequential and diverging.
+
+All palettes described are 
+available in \proglang{R} \citep{vcd:R:2006} in the package \pkg{vcd} \citep{vcd:Meyer+Zeileis+Hornik:2006}
+using the HCL color implementation from \pkg{colorspace} \citep{vcd:Ihaka:2004}.
+Technical documentation to the \proglang{R} implementations along with a large
+collection of examples is available via \verb/help("rainbow_hcl")/ that provides 
+more comparisons between existing \proglang{R} palettes (based on HSV colors) and
+the HCL color palettes.
+
+\section{Color spaces} \label{sec:spaces}
+
+For choosing color palettes, it is imperative to have an understanding
+how colors are perceived. For this it is helpful to have an idea how
+human color vision evolved. It has been hypothesized that it developed in three
+distinct stages: 1.~perception of \textit{light/dark} contrasts (monochrome only),
+2.~\textit{yellow/blue} (usually associated with our notion of warm/cold colors),
+3.~\textit{green/red} (helpful for assessing the ripeness of fruit). See 
+\cite{vcd:Ihaka:2003} for more details and references.
+
+Due to these three color axes, colors are typically described as
+locations in a 3-dimensional spaces. However, human perception of color
+does not correspond to the physiological axes above, but rather to polar
+coordinates in the color plane (yellow/blue vs.~green/red) plus a third
+light/dark axis. Thus, perceptually-based color spaces are defined by three dimensions
+that try to capture
+\begin{enumerate}
+  \item \textbf{hue} (dominant wavelength)
+  \item \textbf{chroma} (colorfulness)
+  \item \textbf{luminance} (brightness, `amount of gray')
+\end{enumerate}
 
@@
   png(file = PNG(files[2]), height = 480, width = 720)
-  hsv.wheel(value = seq(0, 1, length = 6), cone = FALSE)
+  hsv.wheel(value = seq(0, 1, length = 6), cone = TRUE)
   dev.off()
@@
 \includegraphics{hcl-dcone}
-\caption{\label{fig:hcl-hsv} HSV and HCL space}
+\caption{\label{fig:hsv-hcl} HSV and HCL space}
 \end{center}
@@
 \includegraphics{hcl-dcone-gray}
-\caption{\label{fig:hcl-hsv-gray} HSV and HCL space (in gray levels)}
+\caption{\label{fig:hsv-hcl-gray} HSV and HCL space (in gray levels)}
 \end{center}
@@
 
+A popular implementation of such a color space, available in many graphics and
+statistics software packages, are \textit{HSV} (hue, saturation, value) colors.
+They are a simple transformation of \textit{RGB} (red, green, blue) colors
+and are defined by a triplet $(H, S, V)$ with $H \in [0, 360]$, $S, V \in [0, 100]$.
+HSV space has the shape of a single regular cone (often inflated to a regular
+cylinder). Vertical sections through this space are shown in the upper panel
+of Figure~\ref{fig:hsv-hcl}, depicting hue and saturation given value.
+Although simple to specify and easily available in many computing environments,
+HSV colors have a fundamental drawback: its three dimensions map to the three
+dimensions of human color perception very poorly. The three dimensions are
+confounded which is most easily seen when converting the vertical sections to 
+gray scale images in Figure~\ref{fig:hsv-hcl-gray}. Clearly, the brightness of
+colors is not uniform of hues and saturations given the value---therefore,
+HSV colors are often not considered to be perceptually based.
+
+To overcome these drawbacks, various color spaces have been suggested
+that properly map to the perception dimensions, the most prominent of which are
+the CIELUV and CIELAB spaces developed by the \cite{vcd:CIE:2004}. \cite{vcd:Ihaka:2003}
+argues that CIELUV colors are typically preferred for use with emissive technologies
+such as computer screens which makes them an obvious candidate for implementation
+in statistical software packages. By taking polar coordinates in the UV plane of
+CIELUV, \textit{HCL} (hue, chroma, luminance) colors are obtained, defined by
+a triplet $(H, C, L)$ with $H \in [0, 360]$ and $C, L \in [0, 100]$. HCL space
+has the shape of a distorted double cone: the admissable chroma and luminance values
+depend on the hue chosen. The lower panel of Figure~\ref{fig:hsv-hcl} shows
+vertical sections through this space: each of the resulting hue/chroma planes (given
+luminance) is now properly balanced towards the same gray (going from black to white
+with increasing luminance) which becomes obvious when converting the colors to
+a gray scale is in Figure~\ref{fig:hsv-hcl-gray}. This balancing of HCL colors
+gives us the opportunity to conveniently choose color palettes which code categorical
+and/or numerical information by translating it to the three perceptual dimensions.
+
+\section{Qualitative palettes} \label{sec:qualitative}
+
+Qualitative palettes are a set of colors for depicting different categories,
+i.e., for coding a categorical variable. Usually, these should give the same
+perceptual weight to each category so that no group is perceived to be larger/more
+important than any other one. Typical applications of qualitative palettes in
+statistics would be bar plots, pie charts or highlighted spine plots.
+
+\cite{vcd:Ihaka:2003} describes a simple strategy for choosing such palettes: chroma
+and luminance are kept fixed and only the hue is varied for obtaining different colors
+which are consequently all balanced towards the same gray. If colors from the full
+color wheel (i.e., $H \in [0, 360]$) should be used, not all combinations of chroma
+and luminance are feasible. Figure~\ref{fig:pal-q-construct} depicts how three colors
+are chosen, given $C = 50$ and $L = 70$.
+
+Various strategies for choosing the hues in a certain palette are conceivable.
+A simple and intuitive one is often to use colors as metaphors for categories
+(e.g., for political parties). Figure~\ref{fig:pal-q-examples} shows a few
+further examples for generating qualitative sets of colors $(H, 50, 70)$.
+In the upper left panel colors from the full 360 degrees are used ($H = 30, 120, 210, 300$)
+creating a `dynamic' set of colors. The upper right panel shows a `harmonic'
+set with $H = 60, 120, 180, 240$. Warm ($H = 270, 230, 190, 150$) and cold 
+($H = 90, 50, 10, 330$) are shown in the lower left and right panel, respectively.
+
+In \pkg{vcd}, these palettes are available in the function
+\begin{Soutput}
+rainbow_hcl(n, c = 50, l = 70, start = 0, end = 360*(n-1)/n, ...)
+\end{Soutput}
+
 \setkeys{Gin}{width=0.8\textwidth}
-\begin{figure}[h]
+\begin{figure}[p]
 \begin{center}
@@
 \includegraphics{pal-q-construct}
+\vspace*{-2cm}
 \caption{\label{fig:pal-q-construct} Constructing qualitative palettes}
@@
 
-\setkeys{Gin}{width=.8\textwidth}
-\begin{figure}[h]
+\setkeys{Gin}{width=0.8\textwidth}
+\begin{figure}[p]
+\begin{center}
+<<pal-q-examples, fig=TRUE, height=4, width=6>>=
+par(mfrow = c(2, 2), mar = c(0, 0, 3, 0))
+pal(rainbow_hcl(4, start = 30, end = 300), main = "dynamic (30 - 300)")
+pal(rainbow_hcl(4, start = 60, end = 240), main = "harmonic (60 - 240)")
+pal(rainbow_hcl(4, start = 270, end = 150), main = "cold (270 - 150)")
+pal(rainbow_hcl(4, start = 90, end = -30), main = "warm (90 - -30)")
+@
+\caption{\label{fig:pal-q-examples} Examples for qualitative palettes}
+\end{center}
+\end{figure}
+
+\section{Sequential palettes} \label{sec:sequential}
+
+Sequential palettes are used for coding numerical information that simply ranges
+in a certain interval where low values are considered to be uninteresting and
+high values are interesting. Without loss of generality, we assume that we want
+to visualize an intensity or interestingness $i \in [0, 1]$. A typical application
+in statistics are heatmaps.
+
+The simplest solution to this task is to employ light/dark contrasts, i.e., employ the
+oldest and easiest axis of the perceptual system. The interestingness is thus coded
+by an increasing amount of gray (i.e., decreasing luminance) 
+  \[ (H, 0, 90 - i \cdot 60), \]
+where the hue $H$ used does not matter, chroma is set to 0 (i.e., no color), and 
+luminance ranges in $[30, 90]$ avoiding the extreme colors white ($L = 100$) and 
+black ($L = 0$). Instead of going linearly from the light to the dark gray, luminance
+could also be increased nonlinearly, e.g., by some function $i^p$ where the power $p$
+controls wether intensity/luminance is increase quickly or not.
+
+Alternatively, the intensity $i$ could additionally be code by colorfulness (chroma), e.g.,
+  \[ (H, 0 + i^p \cdot C_{\max}, L_{\max} - i^p \cdot (L_{\max} - L_{\min})). \]
+This strategy is depicted in Figure~\ref{fig:pal-s-construct} for a blue hue $H = 260$
+and different combinations of maximal chroma (0, 80 and 100, respecitvely) and
+minimal luminance (30, 30 and 50, respectively). The first two combinations are also
+shown in the first two rows of Figure~\ref{fig:pal-s-examples}.
+
+\setkeys{Gin}{width=.7\textwidth}
+\begin{figure}[b!]
 \begin{center}
@@
 if(generate_png) {
-  png(file = "pal-s-construct.png", height = 480, width = 400)
+  png(file = "pal-s-construct.png", height = 420, width = 420)
   hue.slice(260, grid.n = 301)
@@
 
+In \pkg{vcd}, this strategy is implemented in the function
+\begin{Soutput}
+sequential_hcl(n, h = 260, c = c(80, 0), l = c(30, 90), power = 1.5, ...) 
+\end{Soutput}
+
+\setkeys{Gin}{width=.8\textwidth}
+\begin{figure}[p]
+\begin{center}
+<<pal-s-examples, fig=TRUE, height=9, width=6>>=
+par(mfrow = c(5, 1), mar = c(0, 0, 0, 0))
+pal(sequential_hcl(12, c = 0, power = 2.2))
+pal(sequential_hcl(12, power = 2.2))
+pal(terrain_hcl(12, c = c(65, 0), l = c(45, 95), power = c(1/3, 1.5)))
+pal(heat_hcl(12, c = c(80, 30), l = c(30, 90), power = c(1/5, 1.5)))
+pal(rev(heat_hcl(12, h = c(0, -100), l = c(75, 40), c = c(40, 80), power = 1)))
+@
+\caption{\label{fig:pal-s-examples} Examples for sequential palettes}
+\end{center}
+\end{figure}
+
+To increase the contrast between the colors in the palette even further, the ideas from
+the previous sequential palettes can also be combined with qualitative palettes by
+simultaneously varying the hue as well:
+  \[ (H_2 - i \cdot (H_1 - H_2), 0 + i^{p_1} \cdot C_{\max}, L_{\max} - i^{p_2} \cdot (L_{\max} - L_{\min})). \]
+A typical application would be heat colors that increase from a light yellow to a full
+red. To make the change in hue visible, the chroma needs to increase rather quickly for low
+values of $i$ and then only slowly for higher values of $i$. This can be achieved by
+choosing a power $p_1 < 1$.
+
+In \proglang{R}, these are available in the function
+\begin{Soutput}
+heat_hcl(n, h = c(0, 90), c = c(100, 30), l = c(50, 90), power = c(1/5, 1), ...)
+\end{Soutput}
+with which the lower three rows in Figure~\ref{fig:pal-s-examples} are produced.
+
+\section{Diverging palettes} \label{sec:diverging}
+
+Diverging palettes are also used for coding numerical information ranging in a certain
+interval---however, this interval includes a neutral value. Examples for this include
+residuals (with the neutral value 0) or binary classification probabilities (with neutral
+value 0.5) that could be visualized in mosaic plots \citep{vcd:Zeileis+Meyer+Hornik:2005}
+or classification maps. Without loss of generality, we assume that we want
+to visualize an intensity or interestingness $i \in [-1, 1]$.
+
 \setkeys{Gin}{width=\textwidth}
-\begin{figure}[h]
+\begin{figure}[b!]
 \begin{center}
@@
 if(generate_png) {
-  png(file = "pal-d-construct.png", height = 480, width = 480)
+  png(file = "pal-d-construct.png", height = 480, width = 540)
   hue260 <- hue.slice(260, grid.n = 301, plot = FALSE)
@@
   axis(1, labels = FALSE, at = c(-100, 0, 100))
-  axis(2, at = c(0, 50, 70, 90, 100), labels = c(0, 50, 70, 90, 100))
+  axis(2, at = c(0, 30, 50, 60, 70, 90, 100), labels = c(0, 30, 50, 60, 70, 90, 100))
   axis(3, at = -4:4*25, labels=c(4:0*25, 1:4*25))
-  axis(4, at = c(0, 50, 70, 90, 100), labels = c(0, 50, 70, 90, 100))
+  axis(4, at = c(0, 30, 50, 60, 70, 90, 100), labels = c(0, 30, 50, 60, 70, 90, 100))
   mtext(c("hue = 0", "hue = 260"), side = 3, at = c(-50, 50), line = 3, cex = 1.2)
@@
 
-\setkeys{Gin}{width=0.8\textwidth}
-\begin{figure}[h]
-\begin{center}
-<<pal-q-examples, fig=TRUE, height=6, width=6>>=
-par(mfrow = c(2, 2), mar = c(0, 0, 3, 0))
-pal(rainbow_hcl(4, start = 30, end = 300), main = "dynamic (360 deg)")
-pal(rainbow_hcl(4, start = 60, end = 240), main = "harmonic (180 deg)")
-pal(rainbow_hcl(4, start = 270, end = 150), main = "cold (120 deg)")
-pal(rainbow_hcl(4, start = 90, end = -30), main = "warm (120 deg)")
-@
-\caption{\label{fig:pal-q-examples} Examples for qualitative palettes}
-\end{center}
-\end{figure}
-
 \setkeys{Gin}{width=.8\textwidth}
-\begin{figure}[h]
-\begin{center}
-<<pal-s-examples, fig=TRUE, height=9, width=6>>=
-par(mfrow = c(5, 1), mar = c(0, 0, 0, 0))
-pal(sequential_hcl(12, c = 0, increase = 2.2))
-pal(sequential_hcl(12, increase = 2.2))
-pal(terrain_hcl(12, c = c(65, 0), l = c(45, 95), increase = c(1/3, 1.5)))
-pal(heat_hcl(12, c = c(80, 30), l = c(30, 90), increase = c(1/5, 1.5)))
-pal(rev(heat_hcl(12, h = c(0, -100), l = c(75, 40), c = c(40, 80), increase = 1)))
-@
-\caption{\label{fig:pal-s-examples} Examples for sequential palettes}
-\end{center}
-\end{figure}
-
-\setkeys{Gin}{width=.8\textwidth}
-\begin{figure}[h]
+\begin{figure}[p]
 \begin{center}
@@
 pal(diverge_hcl(7))
+pal(diverge_hcl(7, c = 100, l = c(50, 90), power = 1))
 pal(diverge_hcl(7, h = c(180, 330), c = 59, l = c(75, 95)))
 pal(diverge_hcl(7, h = c(130, 43), c = 100, l = c(70, 90)))
-pal(diverge_hcl(7, c = 100, l = c(50, 90), increase = 1))
 @
@@
 
+Given sequential palettes, deriving diverging palettes is easy: two different hues are
+chosen for adding color to the same amount of `gray' at a given intensity $|i|$.
+Figure~\ref{fig:pal-d-construct} shows the chroma/luminance plane back to back for the
+hues $H = 0$ and 260. These particular hues were chosen because they have rather similar
+chroma/luminance planes, allowing for different combinations of maximal chroma and luminance.
+
+In \pkg{vcd}, the function
+\begin{Soutput}
+diverge_hcl(n, h = c(260, 0), c = 80, l = c(30, 90), power = 1.5, ...) 
+\end{Soutput}
+is provided.
+
+Figure~\ref{fig:pal-d-examples} shows various examples of conceivable combinations of
+hue, chroma and luminance. The first palette uses a broader range on the luminance axis
+whereas the others use larger ranges on the chroma axis.
+
+\section{Open questions}
+
+From our experience, the paths through HCL space described in the previous sections
+in principle work very well and can be used to effectively code qualitative and 
+quantitative information. However, we are not sure how we can properly define trade
+offs between increasing the range with respect to one dimension and decreasing the
+range with respect to another. In other words, how much difference in chroma do I need
+to make up for a certain difference in luminance etc.? A few more precise questions
+are formulated below.
+
+\begin{itemize}
+  \item In sequential/diverging palettes, there is a trade-off between high chroma and high luminance.
+        How should this be chosen in practice? Our impression is that when a small set of colors
+        (such as 3 or 4) are used, large differnces in chroma work well and large differences
+        in luminance are not necessary. However, when a larger set of colors is used (e.g., for 
+        heatmaps where extreme values should be identifiable) it is much more important to have a big
+        difference in luminance.
+
+  \item How should the intensity $|i|$ be increased from 0 to 1?
+        Our experience is that for a small set of colors linear increase is sufficient ($p = 1$)
+        whereas in heatmaps where only very extreme regions are interesting a $p > 1$
+        should be used.
+
+  \item For diverging palettes, how should pairs of colors be chosen? The hues $H = 0$
+        and $H = 260$ where chosen because they are on opposite sides of the color wheel
+	but have a very similar chroma/luminance plane.
+\end{itemize}
+
+\bibliography{vcd}
+
 \end{document}

### 2006-05-27 trunk/vcd/paper/Color/functions.R
first version of color palettes draft finished
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@537 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -7,7 +7,7 @@ pal <- function(col, border = "light gray", ...)
 
 hsv.wheel <-
     function(value = c(0:2, 4:6, 8:10)/10, saturation = 0:100/100, hues = 0:300/300, asp = 1,
-             p.cex = 0.6, do.label = FALSE, cone = TRUE)
+             p.cex = 0.6, do.label = FALSE, cone = TRUE, cex = 1.5)
 {
     stopifnot(is.numeric(saturation), saturation >= 0, saturation <= 1,
               is.numeric(hues), hues >= 0, hues <= 1,
@@ -19,7 +19,7 @@ hsv.wheel <-
     }
     for(i.c in 1:nch) {
         plot(-1:1,-1:1, type="n", axes = FALSE, xlab="",ylab="", asp = asp)
-        text(0.4, 0.99, paste("value =", format(value[i.c])), adj = 0, font = 4)
+        text(0.4, 0.99, paste("v =", format(100 * value[i.c])), adj = 0, font = 4, cex = cex)
 	l.s <- if(cone) saturation * value[i.c] else saturation
         for(ang in hues) {
             a. <- ang * 2 * pi
@@ -35,7 +35,7 @@ hsv.wheel <-
 
 hcl.wheel <-
     function(luminance = 1:9*10, chroma = 0:100, hues = 1:360, asp = 1,
-             p.cex = 0.6, do.label = c(TRUE, FALSE), fixup = FALSE, inner.circle = FALSE)
+             p.cex = 0.6, do.label = c(TRUE, FALSE), fixup = FALSE, inner.circle = FALSE, cex = 1.5)
 {
     do.label <- rep(do.label, length.out = 2)
     stopifnot(is.numeric(chroma), chroma >= 0, chroma <= 100,
@@ -48,7 +48,7 @@ hcl.wheel <-
     }
     for(i.c in 1:nch) {
         plot(-1:1,-1:1, type="n", axes = FALSE, xlab="",ylab="", asp = asp)
-        if(do.label[1]) text(0.4, 0.99, paste("luminance =", format(luminance[i.c])), adj = 0, font = 4)
+        if(do.label[1]) text(0.4, 0.99, paste("l =", format(luminance[i.c])), adj = 0, font = 4, cex = cex)
         l.s <- chroma / 100
         for(ang in hues) { # could do all this using outer() instead of for()...
             a. <- ang * pi/180

### 2006-05-28 trunk/vcd/paper/Color/draft.Rnw
draft version of color paper for Ross & Paul
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@539 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \Abstract{
-  Statistical graphs are often augmented by the use of color
+  Statistical graphics are often augmented by the use of color
   coding information contained in some variable. When this involves
   the shading of areas (and not only points or lines)---e.g., as in barplots, pie charts,
-  mosaic displays or heatmaps---it is important that the colors are perceptually-based
+  mosaic displays or heatmaps---it is important that the colors are perceptually based
   and do not introduce optical illusions or systematic bias. Here, we discuss
@@
 library("vcd")
+library("e1071")
+library("mvtnorm")
+library("ellipse")
+library("kernlab")
+library("KernSmooth")
+
 source("functions.R")
-generate_png <- TRUE
+load("Bundestagswahl2005.rda")
+
+generate_png <- FALSE
 @
@@
 three types of palettes: qualitative, sequential and diverging.
-
 All palettes described are 
@@
 distinct stages: 1.~perception of \textit{light/dark} contrasts (monochrome only),
-2.~\textit{yellow/blue} (usually associated with our notion of warm/cold colors),
-3.~\textit{green/red} (helpful for assessing the ripeness of fruit). See 
+2.~\textit{yellow/blue} contrasts (usually associated with our notion of warm/cold colors),
+3.~\textit{green/red} contrasts (helpful for assessing the ripeness of fruit). See 
 \cite{vcd:Ihaka:2003} for more details and references.
@@
 They are a simple transformation of \textit{RGB} (red, green, blue) colors
-and are defined by a triplet $(H, S, V)$ with $H \in [0, 360]$, $S, V \in [0, 100]$.
+and are defined by a triplet $(H, S, V)$ with $H \in [0, 360]$ and $S, V \in [0, 100]$.
 HSV space has the shape of a single regular cone (often inflated to a regular
@@
 gray scale images in Figure~\ref{fig:hsv-hcl-gray}. Clearly, the brightness of
-colors is not uniform of hues and saturations given the value---therefore,
+colors is not uniform over hues and saturations (given value)---therefore,
 HSV colors are often not considered to be perceptually based.
@@
 and/or numerical information by translating it to the three perceptual dimensions.
+However, some care is required for dealing with the irregular shape of the HCL space
+which will be addressed in the following sections.
 
@@
 
-Qualitative palettes are a set of colors for depicting different categories,
+Qualitative palettes are sets of colors for depicting different categories,
 i.e., for coding a categorical variable. Usually, these should give the same
@@
 important than any other one. Typical applications of qualitative palettes in
-statistics would be bar plots, pie charts or highlighted spine plots.
+statistics would be bar plots, pie charts (see Figure~\ref{fig:seats}) 
+or highlighted mosaic displays (see Figure~\ref{fig:votes}).
 
@@
 Various strategies for choosing the hues in a certain palette are conceivable.
-A simple and intuitive one is often to use colors as metaphors for categories
+A simple and intuitive one is to use colors as metaphors for categories
 (e.g., for political parties). Figure~\ref{fig:pal-q-examples} shows a few
 further examples for generating qualitative sets of colors $(H, 50, 70)$.
-In the upper left panel colors from the full 360 degrees are used ($H = 30, 120, 210, 300$)
+In the upper left panel colors from the full spectrum are used ($H = 30, 120, 210, 300$)
 creating a `dynamic' set of colors. The upper right panel shows a `harmonic'
-set with $H = 60, 120, 180, 240$. Warm ($H = 270, 230, 190, 150$) and cold 
-($H = 90, 50, 10, 330$) are shown in the lower left and right panel, respectively.
+set with $H = 60, 120, 180, 240$. Warm colors (from the blue/green part of the spectrum: 
+$H = 270, 230, 190, 150$) and cold colors (from the yellow/red part of the spectrum:
+$H = 90, 50, 10, 330$) are shown in the lower left and right panel, respectively.
 
@@
 to visualize an intensity or interestingness $i \in [0, 1]$. A typical application
-in statistics are heatmaps.
+in statistics are heatmaps (see Figure~\ref{fig:bkde}).
 
 The simplest solution to this task is to employ light/dark contrasts, i.e., employ the
-oldest and easiest axis of the perceptual system. The interestingness is thus coded
+oldest and simples perceptual axis. The interestingness is thus coded
 by an increasing amount of gray (i.e., decreasing luminance) 
@@
 luminance ranges in $[30, 90]$ avoiding the extreme colors white ($L = 100$) and 
-black ($L = 0$). Instead of going linearly from the light to the dark gray, luminance
-could also be increased nonlinearly, e.g., by some function $i^p$ where the power $p$
-controls wether intensity/luminance is increase quickly or not.
+black ($L = 0$). Instead of going linearly from light to dark gray, luminance
+could also be increased nonlinearly, e.g., by some function $f(i)$ that
+controls wether intensity/luminance is increased quickly or not. We found $f(i) = i^p$ 
+to be a convenient transformation where the power $p$ can be varied to achieve
+different degrees of non-linearity.
 
-Alternatively, the intensity $i$ could additionally be code by colorfulness (chroma), e.g.,
+Furthermore, the intensity $i$ could additionally be coded by colorfulness (chroma), e.g.,
   \[ (H, 0 + i^p \cdot C_{\max}, L_{\max} - i^p \cdot (L_{\max} - L_{\min})). \]
-This strategy is depicted in Figure~\ref{fig:pal-s-construct} for a blue hue $H = 260$
-and different combinations of maximal chroma (0, 80 and 100, respecitvely) and
+This strategy is depicted in the left panel of Figure~\ref{fig:pal-s-construct} for a blue
+hue $H = 260$ and different combinations of maximal chroma (0, 80 and 100, respecitvely) and
 minimal luminance (30, 30 and 50, respectively). The first two combinations are also
-shown in the first two rows of Figure~\ref{fig:pal-s-examples}.
+shown in the first two rows of Figure~\ref{fig:pal-s-examples}. The right panel of
+Figure~\ref{fig:pal-s-construct} shows that the exact same strategy is not possible 
+for the green hue $H = 120$. While the gray colors without chroma can be chosen in 
+the same way, there is a stronger trade-off between using dark colors (with low luminance)
+and colorful colors (with high chroma). Hence, the second path from light gray to 
+full green ends at a much lighter color with $L = 75$.
 
-\setkeys{Gin}{width=.7\textwidth}
+\setkeys{Gin}{width=\textwidth}
 \begin{figure}[b!]
@@
 if(generate_png) {
-  png(file = "pal-s-construct.png", height = 420, width = 420)
+  png(file = "pal-s-construct.png", height = 420, width = 840)
+  par(mfrow = c(1, 2))
   hue.slice(260, grid.n = 301)
@@
   rect(77.5, 27.5, 82.5, 32.5, col = hcl(260, 80, 30))
+  rect(0, 57.5, 5, 62.5, col = hcl(260, 0, 60))
+  rect(0, 27.5, 5, 32.5, col = hcl(260, 0, 30))
+  hue.slice(120, grid.n = 301)
+  lines(c(0, 100), c(90, 75), lty = 2)
+  lines(c(2.5, 2.5), c(90, 30), lty = 2)
+  rect(0, 87.5, 5, 92.5, col = hcl(260, 0, 90))
+  rect(47.5, 80, 52.5, 85, col = hcl(120, 50, 82.5))
+  rect(95, 72.5, 100, 77.5, col = hcl(120, 100, 75))
   rect(0, 57.5, 5, 62.5, col = hcl(260, 0, 60))
@@
 simultaneously varying the hue as well:
-  \[ (H_2 - i \cdot (H_1 - H_2), 0 + i^{p_1} \cdot C_{\max}, L_{\max} - i^{p_2} \cdot (L_{\max} - L_{\min})). \]
-A typical application would be heat colors that increase from a light yellow to a full
-red. To make the change in hue visible, the chroma needs to increase rather quickly for low
-values of $i$ and then only slowly for higher values of $i$. This can be achieved by
-choosing a power $p_1 < 1$.
+  \[ (H_2 - i \cdot (H_1 - H_2), C_{\max} - i^{p_1} \cdot (C_{\max} - C_{\min}), L_{\max} - i^{p_2} \cdot (L_{\max} - L_{\min})). \]
+A typical application would be heat colors that increase from a light yellow (e.g., $(90, 30, 90)$)
+to a full red (e.g., $(0, 100, 50)$). To make the change in hue visible, the chroma needs to
+increase rather quickly for low values of $i$ and then only slowly for higher values of $i$.
+This can be achieved by choosing a power $p_1 < 1$.
 
@@
 interval---however, this interval includes a neutral value. Examples for this include
-residuals (with the neutral value 0) or binary classification probabilities (with neutral
-value 0.5) that could be visualized in mosaic plots \citep{vcd:Zeileis+Meyer+Hornik:2005}
-or classification maps. Without loss of generality, we assume that we want
-to visualize an intensity or interestingness $i \in [-1, 1]$.
+residuals or correlations (both with the neutral value 0) or binary classification
+probabilities (with neutral value 0.5) that could be visualized in mosaic plots (see Figure~\ref{fig:arthritis})
+or classification maps (see Figure~\ref{fig:class}). Without loss of generality, we assume
+that we want to visualize an intensity or interestingness $i \in [-1, 1]$.
 
@@
 Figure~\ref{fig:pal-d-construct} shows the chroma/luminance plane back to back for the
-hues $H = 0$ and 260. These particular hues were chosen because they have rather similar
-chroma/luminance planes, allowing for different combinations of maximal chroma and luminance.
+hues $H = 0$ and 260 with two different paths---with slightly different emphasis on
+luminance or chroma contrasts---from a full red over a neutral grey to a full blue.
+These particular hues were chosen because they have rather similar
+chroma/luminance planes, allowing for many different combinations of maximal chroma and luminance.
+As Figure~\ref{fig:pal-s-construct} illustrates, a bit of care is needed to choose two colors
+that are sufficiently similar in their chroma/luminance plane to allow for the same
+luminance and/or chroma contrasts.
 
-In \pkg{vcd}, the function
+In \pkg{vcd}, the following function is provided:
 \begin{Soutput}
@@
 \end{Soutput}
-is provided.
 
@@
 
+\section{Illustrations}
+
+In this section, we show a few examples for the various types of palettes applied
+to statistical graphics. The first illustration is visualization of a bivariate
+density estimation for the Old Faithful geyser eruptions data. Figure\~ref{fig:bkde}
+shows heatmaps of a bivariate kernel density estimate of waiting times between
+and duration of geyser eruptions in Yellowstone National Park. Both use a sequential
+palette as derived in Section~\ref{sec:sequential} balanced towards the same gray levels
+with $L \in [30, 90]$ and $p_2 = 2$. The sequential palette in the left panel uses only gray
+colors (i.e., $C_{\max} = 0$) and the palette in the right panel additionally employs
+colors with $H \in [0, 90]$, $C \in [30, 80]$ and $p_1 = 1/5$.
+
+\setkeys{Gin}{width=\textwidth}
+\begin{figure}[b!]
+\begin{center}
+<<bkde, height=6, width=12>>=
+if(generate_png) {
+  png(file = "bkde.png", height = 420, width = 840)
+  par(mfrow = c(1, 2))
+  data("geyser", package = "MASS")
+  dens <- bkde2D(cbind(geyser$duration, geyser$waiting), bandwidth = c(0.2, 3), gridsize = c(201, 201))
+  image(dens$x1, dens$x2, sqrt(dens$fhat), xlab = "duration", ylab = "waiting", 
+    col = rev(sequential_hcl(33, c = 0, power = 2)))
+  box()
+  image(dens$x1, dens$x2, sqrt(dens$fhat), xlab = "duration", ylab = "waiting", 
+    col = rev(heat_hcl(33, c = c(80, 30), l = c(30, 90), power = c(1/5, 2))))
+  box()
+  dev.off()
+}
+@
+\includegraphics{bkde}
+\caption{\label{fig:bkde} Bivariate density estimation for Old Faithful geyser eruptions}
+\end{center}
+\end{figure}
+
+\setkeys{Gin}{width=.6\textwidth}
+\begin{figure}[p]
+\begin{center}
+<<seats, fig=TRUE, height=5, width=6>>=
+par(mar = rep(0.8, 4))
+pie(seats, clockwise = TRUE, col = hcl(c(240, 60, 300, 120, 0), 60, 75), radius = 1)
+@
+\caption{\label{fig:seats} Seats in the German parliament}
+\end{center}
+\end{figure}
+
+\setkeys{Gin}{width=.8\textwidth}
+\begin{figure}[p]
+\begin{center}
+<<votes, fig=TRUE, height=9, width=11>>=
+votes <- votes[c(1, 3, 4, 5, 9, 11, 13, 14, 15, 16, 2, 6, 7, 8, 10, 12),]
+votes <- votes[,c(2, 4, 1, 3, 5)]
+votes <- as.table(as.matrix(votes))
+names(dimnames(votes)) <- c("", "")
+
+mosaic(votes, labeling_args = list(rot_labels = c(0, 90, 0, 0), pos_labels = "center",
+  just_labels = c("center", "center", "center", "right")),
+  gp = gpar(fill = hcl(c(240, 60, 0, 120, 300), 60, 75)),
+  spacing = spacing_highlighting, labeling = labeling_left,
+  margins = unit(c(2.5, 1, 1, 12), "lines"), keep = FALSE)
+@
+\caption{\label{fig:votes} Votes in the German election 2005}
+\end{center}
+\end{figure}
+
+To illustrate qualitative palettes, data from the 2005 election for the German parliament `Bundestag'
+are employed. In that election, five parties were able to obtain enough votes to enter the
+Bundestag---the distribution of seats is depicted in a pie chart in Figure~\ref{fig:seats}.
+The colors used are rough metaphors for the political parties, using a red hue $H = 0$ for the
+social democrats SPD, a blue hue $H = 240$ for the conservative CDU, a yellow hue $H = 60$
+for the liberal FDP, a green hue $H = 120$ for the green party `Die Gruenen' and a purple 
+hue $H = 300$ for the leftist party `Die Linke'. All colors use the same chroma $C = 60$ and
+luminance $L = 75$. The pie chart clearly shows that neither the governing coalition of SPD
+and Gruene nor the opposition of CDU and FDP could assemble a majority. Given that no party
+would enter a coalition with the leftists, this lead to a big coalition of CDU and SPD.
+Figure~\ref{fig:votes} shows the distribution of votes in this election stratified by 
+province (Bundesland) in a highlighted mosaic display. The order of provinces is from 
+north to south, first for the 10 Western provinces, then for the 6 Eastern provinces. Clearly,
+the SPD performed better in the North and the CDU better in the South; furthermore, the
+Die Linke performed particularly well in the Eastern provinces and the Saarland.
+
+As pointed out in Section~\ref{sec:diverging}, diverging palettes are particularly useful
+when visualizing residuals or correlation (with natural neutral value 0) or probabilities 
+in 2-class supervised learning (with neutral value 0.5). Examples for both situations are
+provided here. Figure~\ref{fig:arthritis} visualizes the outcome of a double-blind clinical
+trial investigating a new treatment for rheumatoid arthritis. The mosaic rectangles alone
+signal that the treatment lead to higher improvement compared to the placebo group which are
+shown to be significant by the shading based on the Pearson residuals. Positive residuals, corresponding
+to more observations in the corresponding cell than expected under independence, are depicted
+in blue, negative residuals in red. Light colors signal significance at 10\% level, full colors
+significance at 1\% level. Hence, it can be concluded that there are significantly more
+marked improvements in the treated group and significantly fewer in the placebo group than
+would be expected under independence between treatment and improvement. More details can
+be found in \citep{vcd:Zeileis+Meyer+Hornik:2005}.
+
+Figure~\ref{fig:class} shows the fit of a support vector machine (SVM) to an artificial 
+2-class supervised learning example: a mixture of two bivariate normal distributions.
+The circles and triangles show the original observations, solid symbols correspond to
+the support vectors found. The shading underlying the plot visualizes the fitted decision
+values: values around 0 are on the decision boundary and are shaded in light gray, while
+regions that are firmly classified to one or the other class are shaded in full blue and
+red respectively.
+
+\setkeys{Gin}{width=.7\textwidth}
+\begin{figure}[p]
+\begin{center}
+<<arthritis, fig=TRUE, height=6, width=7>>=
+art <- xtabs(~ Treatment + Improved, data = Arthritis, subset = Sex == "Female")
+names(dimnames(art))[2] <- "Improvement"
+set.seed(1234)
+mosaic(art, gp = shading_max)
+@
+\caption{\label{fig:arthritis} Extended mosaic display for arthritis data}
+\end{center}
+\end{figure}
+
+\setkeys{Gin}{width=.7\textwidth}
+\begin{figure}[p]
+\begin{center}
+<<class, height=6, width=8>>=
+if(generate_png) {
+  png(file = "class.png", height = 480, width = 640)
+  set.seed(123)
+  x1 <- rmvnorm(75, mean = c(1.5, 1.5), sigma = matrix(c(1, 0.8, 0.8, 1), ncol = 2))
+  x2 <- rmvnorm(75, mean = c(-1, -1), sigma = matrix(c(1, -0.3, -0.3, 1), ncol = 2))
+  X <- rbind(x1, x2)
+  ex1 <- data.frame(class = factor(c(rep("a", 75), rep("b", 75))), x1 = X[,1], x2 = X[,2])
+  rm(x1, x2, X)
+  fm <- ksvm(class ~ ., data = ex1, C = 0.5)
+  plot(fm, data = ex1)
+  dev.off()
+}
+@
+\includegraphics{class}
+\caption{\label{fig:class} SVM classification plot}
+\end{center}
+\end{figure}
+
 \section{Open questions}
@@
 in principle work very well and can be used to effectively code qualitative and 
-quantitative information. However, we are not sure how we can properly define trade
-offs between increasing the range with respect to one dimension and decreasing the
-range with respect to another. In other words, how much difference in chroma do I need
-to make up for a certain difference in luminance etc.? A few more precise questions
-are formulated below.
+quantitative information. However, there are some open questions, formulated below.
+In particular, we are not sure how we can properly define trade-offs
+between increasing the range with respect to one dimension and decreasing the
+range with respect to another.
 
@@
 	but have a very similar chroma/luminance plane.
+
+  \item To allow more convenient navigation in HCL space, some more \proglang{R} infrastructure
+        could be helpful. We are not sure what would the best or most natural interface for this.
+	A good start might be a function that computes the admissible value on one axis given 
+	a particular value on the other two axes. Maybe you have already experimented with 
+	something like this?
 \end{itemize}

### 2006-05-28 trunk/vcd/paper/Color/functions.R

@@ -93,4 +93,3 @@ hue.slice <- function(hue, grid.n = 251, plot = TRUE, fixup = FALSE)
   class(color.slice) <- "slice"
   invisible(color.slice)
 }
-

### 2006-05-29 trunk/vcd/paper/Color/draft.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@540 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 votes <- votes[c(1, 3, 4, 5, 9, 11, 13, 14, 15, 16, 2, 6, 7, 8, 10, 12),]
-votes <- votes[,c(2, 4, 1, 3, 5)]
+votes <- votes[,c("CDU/CSU", "FDP", "SPD", "Gruene", "Linke")]
 votes <- as.table(as.matrix(votes))
@@
 The colors used are rough metaphors for the political parties, using a red hue $H = 0$ for the
-social democrats SPD, a blue hue $H = 240$ for the conservative CDU, a yellow hue $H = 60$
+social democrats SPD, a blue hue $H = 240$ for the conservative CDU/CSU, a yellow hue $H = 60$
 for the liberal FDP, a green hue $H = 120$ for the green party `Die Gruenen' and a purple 

### 2006-05-29 trunk/vcd/paper/Color/draft.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@541 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 <<votes, fig=TRUE, height=9, width=11>>=
-votes <- votes[c(1, 3, 4, 5, 9, 11, 13, 14, 15, 16, 2, 6, 7, 8, 10, 12),]
-votes <- votes[,c("CDU/CSU", "FDP", "SPD", "Gruene", "Linke")]
-votes <- as.table(as.matrix(votes))
-names(dimnames(votes)) <- c("", "")
+Votes <- as.table(as.matrix(votes[c(1, 3, 4, 5, 9, 11, 13, 14, 15, 16, 2, 6, 7, 8, 10, 12),c("CDU/CSU", "FDP", "SPD", "Gruene", "Linke")]))
+names(dimnames(Votes)) <- c("", "")
 
-mosaic(votes, labeling_args = list(rot_labels = c(0, 90, 0, 0), pos_labels = "center",
+mosaic(Votes, labeling_args = list(rot_labels = c(0, 90, 0, 0), pos_labels = "center",
   just_labels = c("center", "center", "center", "right")),

### 2006-05-30 trunk/vcd/paper/Color/draft.Rnw
Some fixes.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@542 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 The simplest solution to this task is to employ light/dark contrasts, i.e., employ the
-oldest and simples perceptual axis. The interestingness is thus coded
+oldest and simplest perceptual axis. The interestingness is thus coded
 by an increasing amount of gray (i.e., decreasing luminance) 
@@
 could also be increased nonlinearly, e.g., by some function $f(i)$ that
-controls wether intensity/luminance is increased quickly or not. We found $f(i) = i^p$ 
+controls whether intensity/luminance is increased quickly or not. We found $f(i) = i^p$ 
 to be a convenient transformation where the power $p$ can be varied to achieve
@@
 to statistical graphics. The first illustration is visualization of a bivariate
-density estimation for the Old Faithful geyser eruptions data. Figure\~ref{fig:bkde}
+density estimation for the Old Faithful geyser eruptions data. Figure~\ref{fig:bkde}
 shows heatmaps of a bivariate kernel density estimate of waiting times between
@@
 would be expected under independence between treatment and improvement. More details can
-be found in \citep{vcd:Zeileis+Meyer+Hornik:2005}.
+be found in \cite{vcd:Zeileis+Meyer+Hornik:2005}.
 

### 2006-05-30 trunk/vcd/paper/Color/draft.Rnw
fixes by KH
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@543 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 par(mfrow = c(2, 2), mar = c(0, 0, 3, 0))
-pal(rainbow_hcl(4, start = 30, end = 300), main = "dynamic (30 - 300)")
-pal(rainbow_hcl(4, start = 60, end = 240), main = "harmonic (60 - 240)")
-pal(rainbow_hcl(4, start = 270, end = 150), main = "cold (270 - 150)")
-pal(rainbow_hcl(4, start = 90, end = -30), main = "warm (90 - -30)")
+pal(rainbow_hcl(4, start = 30, end = 300), main = "dynamic [30, 300]")
+pal(rainbow_hcl(4, start = 60, end = 240), main = "harmonic [60, 240]")
+pal(rainbow_hcl(4, start = 270, end = 150), main = "cold [270, 150]")
+pal(rainbow_hcl(4, start = 90, end = -30), main = "warm [90, -30]")
 @

### 2006-08-23 trunk/vcd/paper/Shadings/JCGS.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@551 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2005-08-25 trunk/vcd/paper/JCGS.Rnw
@@
 
-data("Arthritis")
+data("Arthritis", package = "vcd")
 art <- xtabs(~ Treatment + Improved, data = Arthritis, subset = Sex == "Female")
-
-data("UCBAdmissions")
-ucb <- aperm(UCBAdmissions)
+names(dimnames(art))[2] <- "Improvement"
+
+data("pistonrings", package = "HSAUR")
+pring <- pistonrings
+names(dimnames(pring)) <- c("Compressor", "Leg")
+
+data("alzheimer", package = "coin")
+alz <- xtabs(~ smoking + disease + gender, data = alzheimer)
+dimnames(alz)[[2]] <- c("Alzheimer's", "Dementia", "Other")
+names(dimnames(alz)) <- c("Smoking", "Disease", "Gender")
+
+data("Punishment", package = "vcd")
+pun <- xtabs(Freq ~ memory + attitude + age + education, data = Punishment)
+dimnames(pun) <- list(Memory = c("yes", "no"), Attitude = c("no", "moderate"),
+  Age = c("15-24", "25-39", "40+"), Education = c("Elementary", "Secondary", "High"))
 
 myseed <- 1071
+nrep <- 5000
+recompute <- FALSE
 @
@@
 <<inference,echo=FALSE,results=hide>>=
-if(FALSE) {
-  set.seed(myseed)
-  art_max <- coindep_test(art, n = 5000)
-  set.seed(myseed)
-  art_chisq <- coindep_test(art, n = 5000, indepfun = function(x) sum(x^2))
-
-  set.seed(myseed)
-  art_gp <- shading_max(art, n = 5000)
-
-  set.seed(myseed)
-  ucb_doublemax <- coindep_test(ucb, "Dept", n = 5000)
-  set.seed(myseed)
-  ucb_maxChisq <- coindep_test(ucb, "Dept", n = 5000, indepfun = function(x) sum(x^2))
-  set.seed(myseed)
-  ucb_sumChisq <- coindep_test(ucb, "Dept", n = 5000, indepfun = function(x) sum(x^2), aggfun = sum)
-
-  set.seed(myseed)
-  ucb_cotab <- cotab_coindep(ucb, condvars = "Dept", type = "assoc", n = 5000, margins = c(3, 1, 1, 3))
-
-  save(art_max, art_chisq, art_gp,
-       ucb_doublemax, ucb_maxChisq, ucb_sumChisq, ucb_cotab,
+if(recompute) {
+  set.seed(myseed)
+  art_max <- coindep_test(art, n = nrep)
+  set.seed(myseed)
+  art_chisq <- coindep_test(art, n = nrep, indepfun = function(x) sum(x^2))
+
+  set.seed(myseed)
+  pring_max <- coindep_test(pistonrings, n = nrep)
+  set.seed(myseed)
+  pring_chisq <- coindep_test(pistonrings, n = nrep, indepfun = function(x) sum(x^2))
+
+  set.seed(myseed)
+  alz_dmax <- coindep_test(alz, "Gender", n = nrep)
+  set.seed(myseed)
+  alz_maxchisq <- coindep_test(alz, "Gender", n = nrep, indepfun = function(x) sum(x^2))
+  set.seed(myseed)
+  alz_sumchisq <- coindep_test(alz, "Gender", n = nrep, indepfun = function(x) sum(x^2), aggfun = sum)
+  set.seed(myseed)
+  alz_cotab <- cotab_coindep(alz, condvars = "Gender", n = nrep, margins = c(2, 1, 1, 2),
+    varnames = FALSE)
+
+  set.seed(myseed)
+  pun_dmax <- coindep_test(pun, 3:4, n = nrep)
+  set.seed(myseed)
+  pun_maxchisq <- coindep_test(pun, 3:4, n = nrep, indepfun = function(x) sum(x^2))
+  set.seed(myseed)
+  pun_sumchisq <- coindep_test(pun, 3:4, n = nrep, indepfun = function(x) sum(x^2), aggfun = sum)
+  set.seed(myseed)
+  pun_cotab <- cotab_coindep(pun, condvars = 3:4, type = "assoc", n = nrep,
+    varnames = FALSE, margins = c(2, 1, 1, 2))
+
+  save(art_max, art_chisq,
+       pring_max, pring_chisq,
+       alz_dmax, alz_maxchisq, alz_sumchisq, alz_cotab,
+       pun_dmax, pun_maxchisq, pun_sumchisq, pun_cotab,
        file = "JCGS-inference.rda")
@@
 for multi-way tables, including in particular conditional independence
-problems.\\[0.3cm]
+models.\\[0.3cm]
 
@@
 fit and spotting dependence patterns that have not been accounted for by the model.
-Other techniques for visualizing independence in contingency tables such as
+Other techniques for visualizing dependence in contingency tables such as
 association plots \citep{vcd:Cohen:1980} can also be enhanced by using this
@@
 usage of perceptually-based HCL (Hue-Chroma-Luminance) colors and combination of
-visualization and significance
-testing. Friendly's shading is typically implemented in statistical software using color
-spaces whose dimensions are only poorly mapped to the perceptual dimensions of the
+visualization and significance testing. Friendly's shading is typically implemented
+in statistical software using color spaces such as HLS (Hue-Luminance-Saturation) or
+HSV (Hue-Saturation-Value) colors. The dimensions of both spaces are only
+poorly mapped to the perceptual dimensions of the
 human visual system \citep{vcd:Brewer:1999,vcd:Ihaka:2003} which makes it more difficult
@@
 \\
-                   &         & \multicolumn{3}{|l}{\textbf{Improved}} \\
+                   &         & \multicolumn{3}{|l}{\textbf{Improvement}} \\
                    &         & None & Some & Marked \\ \hline
@@
 
-The expected cell frequencies in this model are
+The estimated expected cell frequencies under $H_0$ are
 $\hat n_{ij} = n_{i+}n_{+j}/n_{++}$. The probably best known and most used
@@
 The most convenient way to aggregate the $I \times J$ residuals
-to one test statistic is their squared sum
+to one test statistic is their sum of squares
 \begin{equation} \label{eq:ChiSquare}
@@
 Given a critical value $c_\alpha$ for this test statistic, all residuals whose
-absolute value exceeds $c_\alpha$ violate the null hypothesis of indendence at significance level
+absolute values exceed $c_\alpha$ violate the null hypothesis of indendence at significance level
 $\alpha$ \cite[][ch.~7]{vcd:Mazanec+Strasser:2000}. Thus, the interesting cells responsible
@@
 for the distribution. Recently, with the improving perfomance of computers,
-conditional inference (or permutation tests)---carried out either by simulation
-or by computation of the (asymptotic) permutation distribution---have been receiving
+conditional inference (or permutation tests, conditioning on the observations)---carried
+out either by simulation or by computation of the (asymptotic) permutation distribution---have been receiving
 increasing attention \citep[e.g.,][]{vcd:Ernst:2004,vcd:Pesarin:2001,vcd:Strasser+Weber:1999}. 
@@
 with $p = \Sexpr{pval(art_max)}$. Both $p$ values have been computed
-from a sample of size $5000$ from the permutation distribution under independence
+from a sample of size $\Sexpr{nrep}$ from the permutation distribution under independence
 generated via sampling tables with the same row and column sums $n_{i+}$ and $n_{+j}$
@@
 mosaic plots and association plots. Both are suitable to bring out
-departures of an observed table $[n_{ij}]$ from the expected table $[\hat n_{ij}]$
+departures of an observed table $[n_{ij}]$ from the estimated expected table $[\hat n_{ij}]$
 in a graphical way. The latter focuses on the visualization of the Pearson residuals
@@
 proportional to the corresponding Pearson residual $r_{ij}$ and width proportional to the
-square root of the expected counts $\sqrt{\hat n_{ij}}$. Thus, the area is proportional to
+square root of the estimated expected counts $\sqrt{\hat n_{ij}}$. Thus, the area is proportional to
 the raw residuals $n_{ij} - \hat n_{ij}$. The association plot for the arthritis
@@
 \proglang{SAS} implementation \citep[see][]{vcd:Friendly:2000}, Michael Friendly
-uses colors from a palette based on HLS (Hue-Luminance-Saturation) color space.
+uses colors from a palette based on HLS color space.
 The implementation in the standard packages of the
 \proglang{R} system for statistical computing and graphics
-\citep{vcd:R:2005} employs colors from HSV (Hue-Saturation-Value) space. 
+\citep{vcd:R:2005} employs colors from HSV space. 
 Both spaces are rather similar transformations of RGB (Red-Green-Blue) space
@@
 colors in many computer packages \citep{vcd:Moretti+Lyons:2002} making the generation
-of the Friendly shading very simple. As both spaces can generate the
+of the Friendly shading very simple. As both spaces can easily generate the
 same colors, we only discuss HSV space in the following.
 
-The HSV space originally looks like a cone\footnote{The wikipedia gives a very useful
+The HSV space looks like a cone\footnote{The wikipedia gives a very useful
 overview at \url{http://en.wikipedia.org/wiki/HSV_color_space} which also provides
@@
 with black at its peak (zero value) and full color wheels for different saturations at the other end,
-centered around a white (full value). Type and amount of color are controlled by hue
+around a white center (full value). Type and amount of color are controlled by hue
 and saturation, respectively. Typically, polar coordinates $(h, s, v)$ rescaled to the 
@@
 }
-axis(2, at = c(50, 75, 100)/100, labels = c(0.5, 0.75, 1))
-axis(4, at = c(50, 75, 100)/100, labels = c(0.5, 0.75, 1))
-axis(3, at = -4:4*.25, labels=c(4:0*.25, 1:4*.25))
-mtext(c("hue = 0", "hue = 2/3"), side = 3, at = c(-.5, .5), line = 3, cex = 1.2)
+axis(2, at = c(50, 75, 100)/100, labels = c(50, 75, 100))
+axis(4, at = c(50, 75, 100)/100, labels = c(50, 75, 100))
+axis(3, at = -4:4*.25, labels=c(4:0*25, 1:4*25))
+mtext(c("hue = 0", "hue = 240"), side = 3, at = c(-.5, .5), line = 3, cex = 1.2)
 mtext("saturation", side = 3, at = 0, line = 2)
@@
 general and for shading in statistical graphs in particular. The idea of using
-perceptually-based colors that are `in harmony' go back to \cite{vcd:Munsell:1905} 
+perceptually-based colors that are `in harmony' goes back until at least \cite{vcd:Munsell:1905} 
 who introduced a color notation for balanced colors. Based on similar principles, Cynthia
@@
 of chroma and luminance. Such colors are always balanced towards the same grey 
-and thus do not have the problem of varying saturations as the HSV colors.
+and thus do not have the problem of varying saturations that the HSV colors have.
 
@@
 shows two such planes side by side for the hues $h = 0$ and $h = 260$.\footnote{The hue
-$h = 260$ is chosen (rather than $h = 240$ which would correspond to $h = 2/3$ in HSV space)
-because its chroma/luminance plane is most similar to that of $h = 0$.}
+$h = 260$ is chosen rather than $h = 240$ because its chroma/luminance plane is most similar,
+as assessed by the symmetric difference of the planes, to that of $h = 0$.}
 To obtain a 
@@
 versa, colored cells would convey the impression that there is significant
-dependence. Currently, both is not true as can be seen in the left panel
+dependence. Currently, both are not true as can be seen in the left panel
 of Figure~\ref{fig:shadings} which shows the mosaic display for the arthritis
-data with Friendly shading. Although there is significant dependence, no 
-residual exceeds an absolute value of $2$ and hence no cell is colored.
+data with Friendly shading. Although there is significant dependence (as according
+to both the maximum and $\chi^2$ tests), no residual exceeds an absolute value of $2$
+and hence no cell is colored.
 Of course, it can be argued that the shading was not designed for this purpose
@@
 
-\setkeys{Gin}{width=\textwidth}
+\setkeys{Gin}{width=2\textwidth}
 \begin{figure}[t!]
 \begin{center}
-<<shadings,fig=TRUE,echo=FALSE,height=6,width=11,results=hide>>=
+<<shadings,fig=TRUE,echo=FALSE,height=12,width=16,results=hide,include=FALSE>>=
+mymar <- c(1.5, 0.5, 0.5, 2.5)
 grid.newpage()
-pushViewport(viewport(layout = grid.layout(1, 2)))
-pushViewport(viewport(layout.pos.col=1, layout.pos.row=1))
-mosaic(art, newpage = FALSE, gp = shading_Friendly, margins = c(2.5, 4, 2.5, 3), gp_args = list(eps = NULL, lty = 1))
+pushViewport(viewport(layout = grid.layout(2, 3)))
+pushViewport(viewport(layout.pos.row = 1, layout.pos.col = 1))
+mosaic(art, newpage = FALSE, gp = shading_Friendly, margins = mymar,
+  gp_args = list(eps = NULL, lty = 1))
 popViewport()
-pushViewport(viewport(layout.pos.col=2, layout.pos.row=1))
-mosaic(art, newpage = FALSE, gp = art_gp, margins = c(2.5, 3, 2.5, 4))
+pushViewport(viewport(layout.pos.row = 1, layout.pos.col = 2))
+mosaic(art, newpage = FALSE, gp = shading_hsv, margins = mymar,
+  gp_args = list(interpolate = art_max$qdist(c(0.9, 0.99)), p.value = art_max$p.value))
+popViewport()
+pushViewport(viewport(layout.pos.row = 1, layout.pos.col = 3))
+mosaic(art, newpage = FALSE, gp = shading_hcl, margins = mymar,
+  gp_args = list(interpolate = art_max$qdist(c(0.9, 0.99)), p.value = art_max$p.value))
+popViewport()
+
+pushViewport(viewport(layout.pos.row = 2, layout.pos.col = 1))
+mosaic(pring, newpage = FALSE, gp = shading_Friendly, margins = mymar,
+  gp_args = list(eps = NULL, lty = 1, interpolate = c(1, 1.5)))
+popViewport()
+pushViewport(viewport(layout.pos.row = 2, layout.pos.col = 2))
+mosaic(pring, newpage = FALSE, gp = shading_hsv, margins = mymar,
+  gp_args = list(interpolate = c(1, 1.5)))
+popViewport()
+pushViewport(viewport(layout.pos.row = 2, layout.pos.col = 3))
+mosaic(pring, newpage = FALSE, gp = shading_hcl, margins = mymar,
+  gp_args = list(interpolate = c(1, 1.5)))
 popViewport(2)
 @
+\includegraphics[height=\textwidth,angle=90,keepaspectratio]{JCGS-shadings}
 \caption{Mosaic plot for the arthritis data with Friendly shading (left) and maximum shading (right).}
@@
 \end{figure}
+\setkeys{Gin}{width=\textwidth}
 
@@
 its critical value is equivalent
-to a significant value of the whole test statistic $\lambda([r_{ij}])$.\footnote{This
-can easily be seen from the following consideration: To obtain a level $\alpha$ test, where 
-any residual $|r_{ij}|$ exceeding the critical value $c_\alpha$ gives a significant test
-result, the equation $\alpha = P(\mbox{any } |r_{ij}| > c_\alpha)$ must hold. This is equivalent
-to $1 - \alpha = P(\mbox{all } |r_{ij}| \le c_\alpha) = P(\max_{i,j} |r_{ij}| \le c_\alpha)$.}
+to a significant value of the whole test statistic $\lambda([r_{ij}])$.
+%% `proof' for maximum shading
+%% \footnote{This can easily be seen from the following consideration: To obtain a level $\alpha$ test, where 
+%% any residual $|r_{ij}|$ exceeding the critical value $c_\alpha$ gives a significant test
+%% result, the equation $\alpha = P(\mbox{any } |r_{ij}| > c_\alpha)$ must hold. This is equivalent
+%% to $1 - \alpha = P(\mbox{all } |r_{ij}| \le c_\alpha) = P(\max_{i,j} |r_{ij}| \le c_\alpha)$.}
 Typically, applying the maximum statistic is feasible and also appropriate for exploratory 
@@
 Mosaic displays have been emphasized in the literature to be an excellent means
-of visualization for log-linear models \citep{vcd:Friendly:1999,vcd:Theus+Lauer:1999}.
-Problems of interest include complete, joint or conditional independence for all of 
-which the tables of expected values and residuals (again Pearson or deviance) can
-be computed. \cite{vcd:Friendly:1994,vcd:Friendly:1999} shows that his residual-based 
-shading scheme can directly be applied to these more complex independence models.
+of visualization for log-linear models \citep{vcd:Friendly:1999,vcd:Theus+Lauer:1999};
+typical hypotheses of interest include complete, joint or conditional independence.
+For all of these hypotheses, tables of estimated expected values and residuals
+(again Pearson or deviance) can be computed and \cite{vcd:Friendly:1994,vcd:Friendly:1999}
+shows that his residual-based  shading scheme can directly be applied to these more complex
+independence models.
 For inference, the most commonly used aggregation functional for the residuals
@@
 with $k = 1, \dots, K$. Under the assumption of conditional independence,
-we can again compute expected frequencies $[\hat n_{ijk}]$ and the corresponding 
+we can again estimate expected frequencies $[\hat n_{ijk}]$ and the corresponding 
 residuals $[r_{ijk}]$. To test the conditional independence hypothesis, usually
 the sum-of-squares statistic is used
-\begin{equation} \label{eq:sumChisq}
+\begin{equation} \label{eq:sumchisq}
 \sum_{i, j, k} r_{ijk}^2 \quad = \quad \sum_k X_k^2,
@@
 can be constructed
-\begin{equation} \label{eq:doublemax}
+\begin{equation} \label{eq:dmax}
 \max_{i, j, k} |r_{ijk}| \quad = \quad \max_k M_k.
@@
 it would be natural to use 
-\begin{equation} \label{eq:maxChisq}
+\begin{equation} \label{eq:maxchisq}
 \max_k \sum_{i,j} r_{ijk}^2 \quad = \quad \max_k X_k^2.
@@
 the corresponding cells can also be identified. Hence, the double maximum
-statistic from Equation~\ref{eq:doublemax} is the only functional allowing for detection of
+statistic from Equation~\ref{eq:dmax} is the only functional allowing for detection of
 both the strata and the cells violating the conditional independence hypothesis.
@@
 the significant or the non-significant shading for all cells in the contingency table---this
-strategy would have to be used for the sum-of-squares statistic from Equation~\ref{eq:sumChisq}. If
-$\lambda_\mathrm{agg}$ is the maximum as in Equation~\ref{eq:maxChisq},
+strategy would have to be used for the sum-of-squares statistic from Equation~\ref{eq:sumchisq}. If
+$\lambda_\mathrm{agg}$ is the maximum as in Equation~\ref{eq:maxchisq},
 then the significant palette would only be used
@@
 can be pursued, i.e., only the significant palette is used but with data-driven cut offs
-derived from the distribution of the double maximum statistic from Equation~\ref{eq:doublemax}.
+derived from the distribution of the double maximum statistic from Equation~\ref{eq:dmax}.
 
@@
 from conditional independence is picked up by all three statistics formulated
-above: the sum-of-squares statistic from Equation~\ref{eq:sumChisq} is
-$\sum_k X_k^2 = \Sexpr{stat(ucb_sumChisq)}$ ($p = \Sexpr{pval(ucb_sumChisq)}$),
-the maximum sum-of-squares statistic from Equation~\ref{eq:maxChisq} is
-$\max_k X_k^2 = \Sexpr{stat(ucb_maxChisq)}$ ($p = \Sexpr{pval(ucb_maxChisq)}$),
-and the double maximum statistic from Equation~\ref{eq:doublemax} is
-$\max_k M_k = \Sexpr{stat(ucb_doublemax)}$ ($p = \Sexpr{pval(ucb_doublemax)}$).
-The $p$ values are again computed by drawing $5000$ samples from the corresponding
+above: the sum-of-squares statistic from Equation~\ref{eq:sumchisq} is
+$\sum_k X_k^2 = \Sexpr{stat(alz_sumchisq)}$ ($p = \Sexpr{pval(alz_sumchisq)}$),
+the maximum sum-of-squares statistic from Equation~\ref{eq:maxchisq} is
+$\max_k X_k^2 = \Sexpr{stat(alz_maxchisq)}$ ($p = \Sexpr{pval(alz_maxchisq)}$),
+and the double maximum statistic from Equation~\ref{eq:dmax} is
+$\max_k M_k = \Sexpr{stat(alz_dmax)}$ ($p = \Sexpr{pval(alz_dmax)}$).
+The $p$ values are again computed by drawing $\Sexpr{nrep}$ samples from the corresponding
 permutation distribution. The latter two test statistics both indicate the 
@@
 \begin{center}
-<<conditional_assoc,fig=TRUE,echo=FALSE,height=6,width=9>>=
-cotabplot(~ Gender + Admit | Dept, data = ucb, panel = ucb_cotab)
+<<conditional_mosaic,fig=TRUE,echo=FALSE,height=5,width=9>>=
+cotabplot(~ Smoking + Disease | Gender, data = alz, panel = alz_cotab)
+@
+\caption{Conditional association plot for UCB admissions with double maximum shading.}
+\label{fig:ucb}
+\end{center}
+\end{figure}
+
+\setkeys{Gin}{width=\textwidth}
+\begin{figure}[htbp]
+\begin{center}
+<<conditional_assoc,fig=TRUE,echo=FALSE,height=11,width=11>>=
+cotabplot(~ Memory + Attitude | Age + Education, data = pun, panel = pun_cotab)
 @
@@
 %
-Both results can not only be applied to the simple independence problem in 2-way
+Both strategies can not only be applied to the simple independence problem in 2-way
 contingency tables, but also to arbitrary inpendence models fitted via log-linear models
@@
 art <- xtabs(~ Treatment + Improved, data = Arthritis, subset = Sex == "Female")
-ucb <- aperm(UCBAdmissions)
+data("pistonrings", package = "HSAUR")
+data("alzheimer", package = "coin")
+alz <- xtabs(~ smoking + disease + gender, data = alzheimer)
+pun <- xtabs(Freq ~ memory + attitude + age + education, data = Punishment)
+
+## Inference
+## sum of squares
+ss <- function(x) sum(x^2)
+## X^2 (3)
+set.seed(rseed)
+coindep_test(art, n = 5000, indepfun = ss)
+set.seed(rseed)
+coindep_test(pistonrings, n = 5000, indepfun = ss)
+## M (4)
+set.seed(rseed)
+(art_max <- coindep_test(art, n = 5000))
+set.seed(rseed)
+coindep_test(pistonrings, n = 5000)
+
+## sum X^2 (6)
+set.seed(rseed)
+coindep_test(alz, 3, n = 5000, indepfun = ss, aggfun = sum)
+set.seed(rseed)
+coindep_test(pun, 3:4, n = 5000, indepfun = ss, aggfun = sum)
+## max M (7)
+set.seed(rseed)
+coindep_test(alz, 3, n = 5000)
+set.seed(rseed)
+coindep_test(pun, 3:4, n = 5000)
+## max X^2 (8)
+set.seed(rseed)
+coindep_test(alz, 3, n = 5000, indepfun = ss)
+set.seed(rseed)
+coindep_test(pun, 3:4, n = 5000, indepfun = ss)
 
@@
 mosaic(art, gp = shading_Friendly(lty = 1, eps = NULL))
+mosaic(art, gp = shading_hsv, gp_args = list(
+  interpolate = art_max$qdist(c(0.9, 0.99)), p.value = art_max$p.value))
 set.seed(rseed)
-mosaic(art, gp = shading_max)
+mosaic(art, gp = shading_max, gp_args = list(n = 5000))
+
+mosaic(pistonrings, gp = shading_Friendly(lty = 1, eps = NULL, interpolate = c(1, 1.5)))
+mosaic(pistonrings, gp = shading_hsv, gp_args = list(interpolate = c(1, 1.5)))
+mosaic(pistonrings, gp = shading_hcl, gp_args = list(interpolate = c(1, 1.5)))
 
@@
 set.seed(rseed)
-cotabplot(~ Admit + Gender | Dept, data = ucb, panel = cotab_coindep,
+cotabplot(~ smoking + disease | gender, data = alz, panel = cotab_coindep, n = 5000)
+
+## Figure 5
+set.seed(rseed)
+cotabplot(~ memory + attitude | age + education, data = pun, panel = cotab_coindep,
   n = 5000, type = "assoc")
-
-## Inference
-## X^2 (3)
-set.seed(rseed)
-coindep_test(art, n = 5000, indepfun = function(x) sum(x^2))
-## M (4)
-set.seed(rseed)
-coindep_test(art, n = 5000)
-
-## sum X^2 (6)
-set.seed(rseed)
-coindep_test(ucb, "Dept", n = 5000, indepfun = function(x) sum(x^2), aggfun = sum)
-## max M (7)
-set.seed(rseed)
-coindep_test(ucb, "Dept", n = 5000)
-## max X^2 (8)
-set.seed(rseed)
-coindep_test(ucb, "Dept", n = 5000, indepfun = function(x) sum(x^2))
 \end{verbatim}
@@
 %% deprecated stuff
+
+%% UCB admissions
+data("UCBAdmissions", package = "datasets")
+ucb <- aperm(UCBAdmissions)
+
+set.seed(myseed)
+ucb_dmax <- coindep_test(ucb, "Dept", n = nrep)
+set.seed(myseed)
+ucb_maxchisq <- coindep_test(ucb, "Dept", n = nrep, indepfun = function(x) sum(x^2))
+set.seed(myseed)
+ucb_sumchisq <- coindep_test(ucb, "Dept", n = nrep, indepfun = function(x) sum(x^2), aggfun = sum)
+set.seed(myseed)
+ucb_cotab <- cotab_coindep(ucb, condvars = "Dept", type = "assoc", n = nrep, margins = c(3, 1, 1, 3))
+
+\setkeys{Gin}{width=\textwidth}
+\begin{figure}[htbp]
+\begin{center}
+<<conditional_assoc,fig=TRUE,echo=FALSE,height=6,width=9>>=
+cotabplot(~ Gender + Admit | Dept, data = ucb, panel = ucb_cotab)
+@
+\caption{Conditional association plot for UCB admissions with double maximum shading.}
+\label{fig:ucb}
+\end{center}
+\end{figure}
 

### 2006-08-24 trunk/vcd/R/mosaic.R
highlighting_fill now also accepts a color vector.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@552 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -112,8 +112,10 @@ mosaic.default <- function(x, condvars = NULL,
         aperm(x, c(seq(dl)[-highlighting], highlighting))
       if (is.null(spacing))
         spacing <- spacing_highlighting
+      if (is.function(highlighting_fill))
+        highlighting_fill <- rev(highlighting_fill(dim(x)[dl]))
       if (is.null(gp))
-        gp <- gpar(fill = rev(highlighting_fill(dim(x)[dl])))
+        gp <- gpar(fill = highlighting_fill)
       if (!is.null(highlighting_direction)) {
         split_vertical[dl] <- highlighting_direction %in% c("left", "right")
         if (highlighting_direction %in% c("left", "top")) {
@@ -121,7 +123,7 @@ mosaic.default <- function(x, condvars = NULL,
           tmp <- as.data.frame.table(x)
           tmp[,dl] <- factor(tmp[,dl], rev(levels(tmp[,dl])))
           x <- xtabs(Freq ~ ., data = tmp)
-          gp <- gpar(fill = highlighting_fill(dim(x)[dl]))
+          gp <- gpar(fill = rev(highlighting_fill))
         }
       }
     }

### 2006-08-24 trunk/vcd/paper/Shadings/JCGS.Rnw
revised JCGS manuscript wrote rejoinder for JCGS cotab_coindep now contains also a maxchisq shading
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@553 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 names(dimnames(pring)) <- c("Compressor", "Leg")
+dimnames(pring)[[2]][2] <- "center"
 
@@
 nrep <- 5000
-recompute <- FALSE
+recompute <- TRUE
 @
@@
   pun_cotab <- cotab_coindep(pun, condvars = 3:4, type = "assoc", n = nrep,
-    varnames = FALSE, margins = c(2, 1, 1, 2))
+    varnames = FALSE, margins = c(2, 1, 1, 2), test = "maxchisq", interpolate = 1:2)
 
@@
 The estimated expected cell frequencies under $H_0$ are
-$\hat n_{ij} = n_{i+}n_{+j}/n_{++}$. The probably best known and most used
+$\hat n_{ij} = n_{i+}n_{+j}/n_{++}$.
+%FIXME% point out homogeneity problem?
+The probably best known and most used
 measure of discrepancy between observed and expected values are the Pearson residuals
@@
 The HSV space looks like a cone\footnote{The wikipedia gives a very useful
+%FIXME% proper cite for Wikipedia
 overview at \url{http://en.wikipedia.org/wiki/HSV_color_space} which also provides
@@
 around a white center (full value). Type and amount of color are controlled by hue
-and saturation, respectively. Typically, polar coordinates $(h, s, v)$ rescaled to the 
-unit interval are used in this space, giving it the appearance of a cylinder.
+and saturation, respectively. Typically, polar coordinates $(h, s, v)$ rescaled for $s$ and $v$
+to the interval $[0, 100]$ (or the unit interval) are used in this space, giving it the
+appearance of a cylinder.
 For generating colors in the Friendly shading, the following strategy is used:
 The hue $h$ codes the sign of the residuals---$h = 0$ (red hue) is used for negative residuals,
-$h = 2/3$ (blue hue) for positive residuals. The absolute size of the residuals
-is then coded by the saturation $s$ which is set to $0$, $0.5$ and $1$, respectively,
-for small/medium/large residuals. The value is always fixed at $v = 1$.
+$h = 240$ (blue hue) for positive residuals. The absolute size of the residuals
+is then coded by the saturation $s$ which is set to $0$, $50$ and $100$, respectively,
+for small/medium/large residuals. The value is always fixed at $v = 100$.
 This is also depicted in the upper panel of Figure~\ref{fig:shadingHSVHCL} which shows
-the saturation/value plane for the given hues $h = 0$ and $h = 2/3$. 
-The significant color palette shows the
+the saturation/value plane for the given hues $h = 0$ and $h = 240$. 
+The full-color palette shows the
 colors used for the Friendly shading when the residuals are increasing from left to right.
-The set of non-significant colors will be explained in Section~\ref{sec:significance}.
+The reduced-color palette will be explained in Section~\ref{sec:significance}.
 
@@
 
-## significant colors
+## full colors
 rect(-1, 0.95, -.90, 1, col = hsv(0, 1, 1))
@@
 
-text(-1, .33, "significant", pos = 4, cex = 1.2)
+text(-1, .33, "full color", pos = 4, cex = 1.2)
 rect(-1, .20, -.80, .30, col = hsv(0, 1, 1))
@@
 
-## non-significant colors
+## reduced colors
 rect(-1, 0.5, -.90, 0.55, col = hsv(0, 1, 0.5))
@@
 
-text(-1, .13, "non-significant", pos = 4, cex = 1.2)
+text(-1, .13, "reduced color", pos = 4, cex = 1.2)
 rect(-1, 0, -.80, .10, col = hsv(0, 1, 0.5))
@@
 
-## significant colors
+## full colors
 rect(-100, 47.5, -90, 52.5, col = hcl(0, 100, 50))
@@
 
-text(-100, 33, "significant", pos = 4, cex = 1.2)
+text(-100, 33, "full color", pos = 4, cex = 1.2)
 rect(-100, 20, -80, 30, col = hcl(0, 100, 50))
@@
 
-## non-significant colors
+## reduced colors
 rect(-25, 47.5, -15, 52.5, col = hcl(0, 20, 50))
@@
 
-text(-100, 13, "non-significant", pos = 4, cex = 1.2)
+text(-100, 13, "reduced color", pos = 4, cex = 1.2)
 rect(-60, 0, -40, 10, col = hcl(0, 20, 50))
@@
 Consequently, the HSV dimensions are confounded, e.g., saturation
-is not uniform across different hues. A fully saturated blue $(2/3, 1, 1)$ is
-perceived to be much darker than a fully saturated red $(0, 1, 1)$ or green $(1/3, 1, 1)$.
+is not uniform across different hues. A fully saturated blue $(240, 100, 100)$ is
+perceived to be much darker than a fully saturated red $(0, 100, 100)$ or green $(120, 100, 100)$.
 This makes it more difficult for the human eye to judge the size of shaded areas
@@
 good for drawing attention to a plot, but hard to look at for a longer time \citep{vcd:Ihaka:2003}
-which makes HSV-shaded graphics harder to interpret. Finally, white is employed as
-the neutral color for small residuals in the Friendly shading, but typically
+which makes graphics shaded with such colors harder to interpret. Finally, white is employed as
+the neutral color for small residuals in the Friendly shading, however typically
 grey is found to convey neutrality or un-interestingness much better than white
@@
 selecting an appropriate palette for a specific problem. Furthermore, the
-\citet[CIE,][]{vcd:CIE:2004} introduced the two perceptually
-based color spaces CIELAB and CIELUV where the latter is typically preferred for 
+\citet[CIE,][]{vcd:CIE:2004} introduced the two perceptually-based 
+color spaces CIELAB and CIELUV where the latter is typically preferred for 
 emissive color technologies such as computer displays. \cite{vcd:Ihaka:2003} discusses
@@
 By taking polar coordinates in CIELUV space, it is called HCL (Hue-Chroma-Luminance) space
-and qualitative palettes can easily be chosen by using a range of hues for fixed values
-of chroma and luminance. Such colors are always balanced towards the same grey 
-and thus do not have the problem of varying saturations that the HSV colors have.
-
-Here, we discuss how similar ideas can be used for deriving diverging HCL palettes that 
-provide a suitable translation of the ideas from the Friendly shading to perceptually
-uniform HCL colors. 
+and qualitative palettes can easily be chosen by using a range of hues for
+fixed values of chroma and luminance. Such colors are always balanced towards
+the same grey and thus do not have the problem of varying saturations that the HSV colors have.
+In general, the HCL space offers much better support for selecting balanced palettes
+along simple paths through the space, which will be exploited below.
+
+In the following, we discuss how ideas similar to those from \cite{vcd:Ihaka:2003}
+can be used for deriving diverging HCL palettes that provide a suitable translation
+of the ideas from the Friendly shading to perceptually uniform HCL colors. 
 The HCL space looks like a distorted double cone with black (zero luminance) at one end and white
@@
 values of chroma (that controls the colorfulness). Unfortunately, the HCL space is not as
-regular as the HSV space and consequently cannot be standardized in the same way as the
-HSV space. Its dimensions are usually given by a
+regular as the HSV space: although its dimensions are usually also given by a
 hue ranging in $[0, 360]$ degrees and chroma and luminance ranging in $[0, 100]$
-percent. However, not all combinations $(h, c, l)$ yield valid HCL colors and the admissable combinations
+percent, not all combinations $(h, c, l)$ yield valid HCL colors and the admissable combinations
 of $c$ and $l$ vary across different hues $h$. For the task of constructing
@@
 as assessed by the symmetric difference of the planes, to that of $h = 0$.}
-To obtain a 
-sequence of colors with the same properties as the Friendly shading, the palette starts
+To obtain a sequence of colors with the similar properties as the Friendly shading, the palette starts
 at a fully saturated red $(0, 100, 50)$, goes via a neutral color, ends at a fully saturated
@@
 $(0, 0, 100)$ as the neutral color, a light grey $(0, 0, 90)$ is employed as motivated above.
-
-This diverging palette (see the `significant' colors in Figure~\ref{fig:shadingHSVHCL})
+The diverging palette (see the full-color palette in Figure~\ref{fig:shadingHSVHCL})
 uses both chroma---i.e., the colorfulness---and luminance---i.e., 
@@
 chroma or luminance for this purpose, but using both seems to be a very effective way
-of visualization.
+of visualization and corresponds more closely with the properties of the Friendly shading.
+
+Applying these palettes to the mosaic plot of the arthritis data, yields
+the displays in the upper middle and right panels of Figure~\ref{fig:shadings} (with
+data-driven cut offs as defined Section~\ref{sec:significance}). This illustrates
+that especially the full color cells (in the `marked' column) are less flashy and more
+balanced in the HCL shading as compared to the HSV shading.
 
@@
 versa, colored cells would convey the impression that there is significant
-dependence. Currently, both are not true as can be seen in the left panel
+dependence. Currently, both are not true as can be seen in the upper left panel
 of Figure~\ref{fig:shadings} which shows the mosaic display for the arthritis
 data with Friendly shading. Although there is significant dependence (as according
-to both the maximum and $\chi^2$ tests), no residual exceeds an absolute value of $2$
+to both the maximum and sum-of-squares tests), no residual exceeds an absolute value of $2$
 and hence no cell is colored.
@@
 way (i.e., for the table visualized) and use these instead of the hard-coded values
-$2$ and $4$. In the right panel of Figure~\ref{fig:shadings} this is done for
+$2$ and $4$. In the upper middle and right panels of Figure~\ref{fig:shadings} this is done for
 the arthritis data by employing the critical values
@@
 $\Sexpr{round(as.vector(art_max$qdist(0.99)), digits = 2)}$ at level $\alpha = 0.01$
-(and using the diverging HCL palette) derived from the permutation distribution of $M$ for 
+(using the diverging HSV and HCL palettes, respectively) derived from the permutation distribution of $M$ for 
 the arthritis data as described in Section~\ref{sec:tests}. By using these cut offs, the presence of color
@@
 \setkeys{Gin}{width=2\textwidth}
-\begin{figure}[t!]
+\begin{figure}[p]
 \begin{center}
@@
 mosaic(pring, newpage = FALSE, gp = shading_hsv, margins = mymar,
-  gp_args = list(interpolate = c(1, 1.5)))
+  gp_args = list(interpolate = c(1, 1.5), p.value = pring_chisq$p.value))
 popViewport()
@@
 mosaic(pring, newpage = FALSE, gp = shading_hcl, margins = mymar,
-  gp_args = list(interpolate = c(1, 1.5)))
+  gp_args = list(interpolate = c(1, 1.5), p.value = pring_chisq$p.value))
 popViewport(2)
@@
 \includegraphics[height=\textwidth,angle=90,keepaspectratio]{JCGS-shadings}
-\caption{Mosaic plot for the arthritis data with Friendly shading (left) and maximum shading (right).}
+\caption{Upper row: Mosaic plot for the arthritis data with Friendly shading (left),
+HSV maximum shading (middle), HCL maximum shading (right). Lower row: Mosaic
+plot for the piston rings data with fixed user-defined cut offs 1 and 1.5 and 
+Friendly shading (left), HSV sum-of-squares shading (middle), HCL sum-of-squares shading
+(right).}
 \label{fig:shadings}
@@
 colors, the maximum chroma was set to $100$ as shown in Figure~\ref{fig:shadingHSVHCL} and 
-is reduced to $20$ for the non-significant palette. This palette still codes the absolute
+is decreased to $20$ for the reduced-color palette. This palette still codes the absolute
 size of the residuals by luminance (i.e., the amount of grey), uses the same neutral grey
@@
 but gives less emphasis to the pattern by making the plot less colorful. A similar effect
-can be obtained in the HSV space if the value is reduced from $1$ to $0.5$ for non-significant 
-palettes (see Figure~\ref{fig:shadingHSVHCL}). As for the significant colors,
+can be obtained in the HSV space if the value is reduced from $100$ to $50$ for reduced-color 
+palettes (see Figure~\ref{fig:shadingHSVHCL}). As for the full colors,
 the dimensions used for creating this palette are confounded and hence the HCL scheme is clearly
 preferable.
+
+To see such a sum-of-squares shading in practice, we employ a data on the number of piston
+ring failures in three legs (north/center/south) in four different steam-driven compressors
+at an FIXME. The contingency table is given in \cite{vcd:Haberman:1973}, re-analyzed
+by \cite{vcd:Everitt+Hothorn:2006} and displayed in the lower row of Figure~\ref{fig:shadings}.
+Neither the sum-of-squares test ($X^2 = \Sexpr{stat(pring_chisq)}$, $p = \Sexpr{pval(pring_chisq)}$)
+nor the maximum test ($M = \Sexpr{stat(pring_max)}$, $p = \Sexpr{pval(pring_max)}$) find
+evidence for a departure from independence in the data (at 5\% level). However, as argued
+by \cite{vcd:Everitt+Hothorn:2006}, a closer look at the size of the residuals might be
+interesting. To do so, we choose two fixed cut offs within the range of residuals, 1 and 1.5,
+showing that more failures (than expected under independence) in the center leg and less failures
+in the south leg were observed for compressor 1 and vice versa for compressor 4. This is
+brought out clearly by all three shadings in the lower row of Figure~\ref{fig:shadings}. 
+However, less emphasis is given to this pattern by the sum-of-squares shadings in the
+middle and right panel because the reduced color palettes are used due to non-significance
+of the associated test (at 5\% level). Comparing the HSV-based and HCL-based version
+shows again that the latter are less flashy and more balanced.
+
+Other strategies for constructing palettes for general aggregation functionals $\lambda(\cdot)$
+(including the sum of squares) that are similar in spirit to the maximum shading are conceivable.
+Instead of using a reduced-color palette for non-significant tests, a no-color palette with
+only a light grey could be employed. This could also be seen as always using cut offs
+outside the range of residuals for non-significant tests. Analogously, for significant tests,
+cut offs that are always inside the range of residuals could be chosen. The latter could
+be determined by subject-matter knowledge, as a fraction of the associated critical value,
+or as certain quantiles of the absolute residuals.
 
@@
 discussed, it is possible for all aggregation functionals to simply use either
-the significant or the non-significant shading for all cells in the contingency table---this
+the full-color or the reduced-color shading for all cells in the contingency table---this
 strategy would have to be used for the sum-of-squares statistic from Equation~\ref{eq:sumchisq}. If
 $\lambda_\mathrm{agg}$ is the maximum as in Equation~\ref{eq:maxchisq},
-then the significant palette would only be used
+then the full-color palette would only be used
 in those strata in conflict with the hypothesis of conditional independence whereas the
-non-significant palette would be used for the remaining strata. Finally, if both $\lambda_\mathrm{agg}$
+reduced-color palette would be used for the remaining strata. Finally, if both $\lambda_\mathrm{agg}$
 and $\lambda_\mathrm{indep}$ are the maximum, then the same strategy as in Section~\ref{sec:shadings}
-can be pursued, i.e., only the significant palette is used but with data-driven cut offs
+can be pursued, i.e., only the full-color palette is used but with data-driven cut offs
 derived from the distribution of the double maximum statistic from Equation~\ref{eq:dmax}.
@@
 
-For illustration, the famous admissions data of the University of
-California, Berkley, (UCB) from \cite{vcd:Bickel+Hammel+O'Connell:1975}
-are employed. Testing the conditional independence of admission and gender
-given the department leads to the well-known result that there is no 
-gender discrimination in five of six departments, but that women are more
-likely to be admitted than men in the first department. This departure
-from conditional independence is picked up by all three statistics formulated
-above: the sum-of-squares statistic from Equation~\ref{eq:sumchisq} is
-$\sum_k X_k^2 = \Sexpr{stat(alz_sumchisq)}$ ($p = \Sexpr{pval(alz_sumchisq)}$),
-the maximum sum-of-squares statistic from Equation~\ref{eq:maxchisq} is
-$\max_k X_k^2 = \Sexpr{stat(alz_maxchisq)}$ ($p = \Sexpr{pval(alz_maxchisq)}$),
-and the double maximum statistic from Equation~\ref{eq:dmax} is
-$\max_k M_k = \Sexpr{stat(alz_dmax)}$ ($p = \Sexpr{pval(alz_dmax)}$).
-The $p$ values are again computed by drawing $\Sexpr{nrep}$ samples from the corresponding
-permutation distribution. The latter two test statistics both indicate the 
-first department as the source of departure and the maximum statistic additionally
-identifies the deviations in the female applicants group. This result
-is also visualized by the conditional association plot in a trellis layout based
-on the double maximum shading in Figure~\ref{fig:ucb}. This can be
-interpreted as a diagnostic residual plot for the associated log-linear model
-because it shows that there are significant departures from the model in department
-A but not in departments B to F where all residuals are very small.
-
 \setkeys{Gin}{width=\textwidth}
-\begin{figure}[htbp]
+\begin{figure}[b!]
 \begin{center}
@@
 @
-\caption{Conditional association plot for UCB admissions with double maximum shading.}
-\label{fig:ucb}
+\caption{Conditional mosaic plot with double maximum shading for conditional
+independence of smoking and disease given gender.}
+\label{fig:alz}
 \end{center}
@@
 
+For illustration, a 3-way and a 4-way table are employed: The first is taken
+from a case-control study of smoking and {A}lzheimer's disease published in
+\cite{vcd:Salib+Hillier:1997} and re-analyzed using conditional inference techniques
+in \cite{vcd:Hothorn+Hornik+VanDeWiel:2006}. It provides data on the smoking behaviour
+(no, $<$10, 10--20, $>$20 cigarettes per day), disease status (Alzheimer's, other
+dementias, other diagnoses) and gender. The question is whether smoking and disease 
+status are conditionally independent given gender. All three statistics
+suggested above find evidence for departure from significance: the sum-of-squares
+statistic from Equation~\ref{eq:sumchisq} is $\sum_k X_k^2 = \Sexpr{stat(alz_sumchisq)}$
+($p = \Sexpr{pval(alz_sumchisq)}$), the maximum sum-of-squares statistic from
+Equation~\ref{eq:maxchisq} is $\max_k X_k^2 = \Sexpr{stat(alz_maxchisq)}$
+($p = \Sexpr{pval(alz_maxchisq)}$), and the double maximum statistic from Equation~\ref{eq:dmax}
+is $\max_k M_k = \Sexpr{stat(alz_dmax)}$ ($p = \Sexpr{pval(alz_dmax)}$).
+The $p$ values are again computed by drawing $\Sexpr{nrep}$ samples from the corresponding
+permutation distribution. The conditional mosaic plot in Figure~\ref{fig:alz}
+shows clearly that the association of smoking and disease is present only in
+the group of male patients. The double maximum shading employed allows for
+identification of the male heavy smokers as the cells `responsible' for the dependence:
+other dementias are more frequent and Alzheimer's disease less frequent in this group than expected
+under independence. Interestingly, there seems to be another large residual for
+the light smoker group ($<$10 cigarettes) and Alzheimer's disease---however, this
+is only significant at 10\% and not at the 1\% level as the other two cells.
+
 \setkeys{Gin}{width=\textwidth}
-\begin{figure}[htbp]
+\begin{figure}[t!]
 \begin{center}
@@
 @
-\caption{Conditional association plot for UCB admissions with double maximum shading.}
-\label{fig:ucb}
+\caption{Conditional association plot with maximum sum-of-squares shading for conditional
+independence of memory and attitude given age and education.}
+\label{fig:pun}
 \end{center}
 \end{figure}
+
+As a 4-way example, we use the punishment data \cite{vcd:andersen:1991} 
+from a study of the Gallup Institute in Denmark in 1979 about
+the attitude of a random sample of 783 persons towards corporal
+punishment of children. The contingency table comprises four margins:
+memory of punishments as a child (yes/no), attitude as a binary variable
+(approval of ``moderate'' punishment or ``no'' approval), highest level
+of education (elementary/secondary/high), and age group (15--24, 25--39, $\ge$40 years).
+It is of interest whether there is an association between memories of corporal punishments
+as a child and attitude towards punishment of children as an adult, controlling
+for age and education. Figure~\ref{fig:pun} shows a conditional association
+plot of memory and attitude given age and education. This can be
+interpreted as a diagnostic residual plot for the associated log-linear conditional
+indpendence model. It shows an association between memories and attitude
+for the lowest education group (first column) and highest age group (last row):
+experienced violence seems to engender violence again as there are less adults
+that disapprove punishment in the group with memories of punishments than expected
+under independence. For the remaining four age-education groups, there seems to be
+no association: all residuals of the conditional independence model are very close to
+zero in these cells. All three tests agree again that there is significant
+departure from conditional independence in this table: the sum-of-squares
+statistic is $\sum_k X_k^2 = \Sexpr{stat(pun_sumchisq)}$ ($p = \Sexpr{pval(pun_sumchisq)}$),
+the maximum sum-of-squares statistic is $\max_k X_k^2 = \Sexpr{stat(pun_maxchisq)}$
+($p = \Sexpr{pval(pun_maxchisq)}$), and the double maximum statistic
+is $\max_k M_k = \Sexpr{stat(pun_dmax)}$ ($p = \Sexpr{pval(pun_dmax)}$). The $\max_k X_k^2$
+result is visualized by means of a maximum sum-of-squares shading in Figure~\ref{fig:pun}
+with user-defined cut offs 1 and 2, chosen to be within the range of the residuals.
+The full-color palette is used only for those strata associated with a
+sum-of-squares statistic $X_k^2$ significant at (overall) 5\% level, the reduced-color
+palette is used otherwise. This highlights that the dependence pattern is significant
+only for the middle and high age group in the low education column. The other panels
+in the first column and last row also show a similar dependence pattern, however, it is not
+significant at 5\% level and hence graphically down-weighted by using reduced color.
 
@@
 To combine visualization and significance testing, two approaches are presented:
-The first approach always uses a fully colored palette but relies on
+The first approach, based on the maximum statistic, always uses a fully colored palette but relies on
 data-driven cut offs such that the presence of color is equivalent to significance of the
-associated maximum test. The second approach uses pre-defined cut offs (such as $2$ and $4$) but
-codes the result of the significance test by using full colors only for significant results and
-the same type of palette with a reduced amount of color for non-significant results.
+associated maximum test. The second approach, also applicable to other statistics such as
+the sum of squares, uses pre-defined cut offs (e.g., $2$ and $4$) but
+codes the result of the associated significance test by using full colors only if it
+is significant and the same type of palette with a reduced amount of color otherwise.
 %
@@
 %
-All associated significance tests are carried out using a conditional inference approach
+All significance tests are carried out using a conditional inference approach
 instead of relying on unconditional asymptotic results.
+
+\clearpage
 
@@
 \proglang{R}~\Sexpr{paste(R.Version()[6:7], collapse = ".")}
-\citep[\mbox{\url{http://www.R-project.org/}}]{vcd:R:2005} and the packages
-\pkg{vcd}~\Sexpr{packageDescription("vcd")$Version} \citep{vcd:Meyer+Zeileis+Hornik:2005},
+\citep[\mbox{\url{http://www.R-project.org/}}]{vcd:R:2006} and the packages
+\pkg{vcd}~\Sexpr{packageDescription("vcd")$Version} \citep{vcd:Meyer+Zeileis+Hornik:2006},
 {\pkg{MASS}~\Sexpr{packageDescription("MASS")$Version}} \citep{vcd:Venables+Ripley:2002},
@@
 {\pkg{colorspace}~{\Sexpr{packageDescription("colorspace")$Version}}} \citep{vcd:Ihaka:2004}.
-
-\clearpage
 
@@
 cotabplot(~ memory + attitude | age + education, data = pun, panel = cotab_coindep,
-  n = 5000, type = "assoc")
+  n = 5000, type = "assoc", test = "maxchisq", interpolate = 1:2)
 \end{verbatim}
@@
 rect(0.9, 0.1, 1, 0.2, col = hsv(0, 1, 1))
-text(0.5, 0.1, "non-significant", pos = 1, cex = 1.5)
-text(0.85, 0.1, "significant", pos = 1, cex = 1.5)
+text(0.5, 0.1, "reduced color", pos = 1, cex = 1.5)
+text(0.85, 0.1, "full color", pos = 1, cex = 1.5)
 @

### 2006-08-24 trunk/vcd/R/cotabplot.R

@@ -199,7 +199,8 @@ cotab_fourfold <- function (x = NULL, condvars = NULL, ...) {
 class(cotab_fourfold) <- "grapcon_generator"
 
 cotab_coindep <- function(x, condvars,
-  test = c("max", "Chisq"), level = NULL, n = 1000,
+  test = c("doublemax", "maxchisq", "sumchisq"),
+  level = NULL, n = 1000, interpolate = c(2, 4),
   h = NULL, c = NULL, l = NULL, lty = 1,
   type = c("mosaic", "assoc"),
   legend = FALSE, ylim = NULL, ...)
@@ -232,7 +233,8 @@ cotab_coindep <- function(x, condvars,
   cond.char <- names(cond.dnam) 	
 
   test <- match.arg(test)
-  if(test == "max") {
+  switch(test,
+  "doublemax" = {
     if(is.null(level)) level <- c(0.9, 0.99)
     fm <- coindep_test(x, cond.num, n = n)
     resids <- residuals(fm)
@@ -240,51 +242,53 @@ cotab_coindep <- function(x, condvars,
     col.bins <- fm$qdist(sort(level))
     gpfun <- shading_hcl(observed = NULL, residuals = NULL, expected = NULL, df = NULL,
       h = h, c = c, l = l, interpolate = col.bins, lty = lty, p.value = fm$p.value)
-  } else {
+  },
+  "maxchisq" = {
     if(is.null(level)) level <- 0.95
     level <- level[1]
-    
-    indep.formula <- as.formula(paste("~ (",
-                                      paste(ind.char, collapse = " + "),
-	    			      ") * ",
-				      paste(cond.char, collapse = " * "),
-				      sep = ""))
-    fm <- loglm(indep.formula, data = x, fitted = TRUE)
-    resids <- residuals(fm, type = "pearson")
-    
-    gpfun <- shading_hcl(x, fitted(fm), resids, fm$df,
-      h = h, c = c, l = l, lty = lty)
-  }
+    fm <- coindep_test(x, cond.num, n = n, indepfun = function(x) sum(x^2))
+    resids <- residuals(fm)
 
-  ## if "maxChisq" should be implemented, use something
-  ## along the lines of
-  ## mosaiclist <- list()
-  ## 
-  ## rval <- function(x, indep, cond)
-  ##   mosaiclist[[ ... ]](co_table(margin.table(x, c(indep, cond)), cond)[[ ... ]])
[…]

### 2006-09-03 trunk/vcd/paper/Handbook/hsc.Rnw
highlighting_fill now also accepts a color vector.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@552 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
Indexing done.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@555 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2006-05-03 trunk/vcd/paper/hsc.Rnw
@@
 
-\usepackage{Sweave,hyperref,color,graphicx} 
+\usepackage{Sweave,hyperref,color,graphicx,makeidx,multicol} 
 \usepackage[bottom]{footmisc}% places footnotes at page bottom
@@
 
+%\makeindex
+
 \begin{document}
@@
 
-Categorical data analysis is typically based on two- or
-higher-dimensional contingency tables, cross-tabulating the
+\index{Categorical data} Categorical data analysis is typically based on two- or
+higher-dimensional \index{contingency tables} contingency tables, cross-tabulating the
 co-occurences of levels of nominal and/or ordinal data. 
@@
 All three graphical methods visualize aspects of (possibly higher-dimensional)
-contingency tables. A \emph{mosaic plot} \cite{vcd:Hartigan+Kleiner:1984} is basically 
+contingency tables. A \emph{mosaic plot} \index{mosaic plot}
+\cite{vcd:Hartigan+Kleiner:1984} is basically 
 an area-proportional visualization of (typically, observed) frequencies, composed
@@
 dimensions of previous splits. 
-\emph{Sieve plots} \cite{vcd:riedwyl+schuepbach:1994} 
+\emph{Sieve plots} \index{sieve plot} \cite{vcd:riedwyl+schuepbach:1994} 
 are similar to mosaic plots, but the area of each
@@
 is filled with a number of rectangles corresponding to the observed value.
-An \emph{association plot} \cite{vcd:Cohen:1980}
+An \emph{association plot} \index{association plot} \cite{vcd:Cohen:1980}
 visualizes the standardized deviations of observed frequencies from
@@
 The figures in this chapter have all been produced using the
-\proglang{R} system, using the extension packages \pkg{vcd} 
+\proglang{R} \index{R} system, using the extension packages \pkg{vcd} 
 \cite{vcd:Meyer+Zeileis+Hornik:2005} and
@@
 Throughout this section, our examples will be based on
-the \data{hospital} data \cite{vcd:wing:1962} given in Table~\ref{tab:hospital}.
+the \index{Data!hospital} 
+\data{hospital} data \cite{vcd:wing:1962} given in Table~\ref{tab:hospital}.
 
@@
 
-Mosaic displays have been introduced by 
+\index{mosaic plot} Mosaic displays have been introduced by 
 \cite{vcd:Hartigan+Kleiner:1981,vcd:Hartigan+Kleiner:1984} and
@@
 adapted to visualize the counts for each level, which is also 
-called a \emph{spine plot} \cite{vcd:hummel:1996}. From this plot, we see that 
-the number of patients 
+called a \emph{spine plot} \cite{vcd:hummel:1996}. \index{spine plot}
+From this plot, we see that the number of patients 
 decreases with the length of stay. Step 2 now is to add further splits in the
@@
 
+\index{sieve plot}
 When we try to explain data, we assume the validity of a certain
@@
 
+\index{association plot}
 In the last section, we have seen how to compare observed and expected
@@
 more straightforwardly by using a plot that directly visualizes the residuals. The
-most widely used residuals are the Pearson residuals
+most widely used residuals are the Pearson residuals \index{Pearson residuals}
 
@@
 red, green, and blue (RGB colors), especially for computer devices. 
+\index{RGB colors}
 The appearance of the on-screen colors is affected by the device
@@
 Choosing colors and color palettes in RGB space can be a bit inconvenient
-and hence software implementations are often based on
+and hence software implementations are often based on \index{HSV colors}
 Hue-Saturation-Value (HSV) colors (or the comparable Hue-Luminance-Satu\-ration 
@@
 As an example, see Figure~\ref{fig:colors} (left) 
-showing a qualitative color palette (colors $(H, 100, 100)$ for varying hues $H$) in the HSV
+showing a qualitative color palette \index{qualitative color palette}
+(colors $(H, 100, 100)$ for varying hues $H$) in the HSV
 space: although saturation and value are fixed, the fully saturated blue is
@@
 to derive acceptable diverging palettes from the HSV space, i.e.,
+\index{diverging color palette}
 bipolar scales containing colors ranging between two very distinct colors.
@@
 proportion of these patients increases with LOS. In fact, such a
-``stacked'' plot can also be interpreted as spine plot with highlighting
+``stacked'' plot can also be interpreted as spine plot with
+highlighting \index{highlighting}
 \cite{vcd:hummel:1996}, particularly useful in analyzing relationships among
@@
 displays. Friendly \cite{vcd:Friendly:1994} introduced a
-residual-based shading of the tiles to additionally visualize the
+\index{residual-based shading} residual-based shading
+of the tiles to additionally visualize the
 residuals from a given independence model fitted to the table. 
@@
 For testing the independence hypothesis from (\ref{eq:H0}),
-using a permutation test is particularly intuitive due to the 
+using a \index{permutation test} permutation test is particularly intuitive due to the 
 permutation invariance (given row and column sums) of this problem.
@@
 though the residuals are ``small''. In fact, the cut-off points are really
-data-dependent. Consider the case of the \data{arthritis} data 
+data-dependent. Consider the case of the \data{arthritis} data \index{Data!arthritis}
 \cite{vcd:Koch+Edwards:1988}, resulting from a double-blind clinical trial
@@
 using, e.g., the Patefield algorithm \cite{vcd:Patefield:1981}
+\index{Patefield algorithm}
 and computing the maximum statistic for each of these tables.
@@
 As an example, consider the well-known UCB admissions data
+\index{Data!UCB Admissions}
 \cite{vcd:Bickel+Hammel+O'Connell:1975} on applicants, classified by
@@
 all basic plots can be combined in pairwise displays,
-arranged in a matrix similar to scatterplots in a pairs plot.
+arranged in a matrix similar to scatterplots in a pairs
+plot. \index{pairs plot}
 The diagonal cells contain the variable names, optionally with
@@
 use the vertical splits for both explanatory variables, resulting in the 
-doubledecker plot in Figure~\ref{fig:doubleucb}. From the widths of the
+\index{doubledecker plot} doubledecker plot 
+in Figure~\ref{fig:doubleucb}. From the widths of the
 tiles it can be seen that students apply to the six departments in
@@
 the previous section, there are specialized displays 
-designed for the visualization of conditional independence structures.
+designed for the visualization of \index{conditional independence} 
+conditional independence structures.
 
@@
 a conditioning in the plot, i.e., to use trellis-like layouts for
+\index{trellis layout}
 visualizing \emph{partial} tables as defined by the conditioning variables. 
@@
 punishment of children (see Table~\ref{tab:punish}).      
+\index{Data!Danish corporate punishment}
 The data consist of four variables: Attitude is a binary variable
@@
 
+\printindex
+
 \end{document}

### 2006-09-05 trunk/vcd/paper/Handbook/hsc.Rnw
index final touches
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@556 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-\usepackage{Sweave,hyperref,color,graphicx,makeidx,multicol} 
+\usepackage{Sweave,color,hyperref,graphicx,makeidx,multicol} 
 \usepackage[bottom]{footmisc}% places footnotes at page bottom
@@
 
-\index{Categorical data} Categorical data analysis is typically based on two- or
+\index{catgorical data} Categorical data analysis is typically based on two- or
 higher-dimensional \index{contingency tables} contingency tables, cross-tabulating the
@@
 Throughout this section, our examples will be based on
-the \index{Data!hospital} 
+the \index{data!hospital} 
 \data{hospital} data \cite{vcd:wing:1962} given in Table~\ref{tab:hospital}.
@@
 red, green, and blue (RGB colors), especially for computer devices. 
-\index{RGB colors}
+\index{color space!RGB}
 The appearance of the on-screen colors is affected by the device
@@
 Choosing colors and color palettes in RGB space can be a bit inconvenient
-and hence software implementations are often based on \index{HSV colors}
+and hence software implementations are often based on \index{color space!HSV}
 Hue-Saturation-Value (HSV) colors (or the comparable Hue-Luminance-Satu\-ration 
@@
 As an example, see Figure~\ref{fig:colors} (left) 
-showing a qualitative color palette \index{qualitative color palette}
+showing a qualitative color palette \index{color palette!qualitative}
 (colors $(H, 100, 100)$ for varying hues $H$) in the HSV
@@
 to derive acceptable diverging palettes from the HSV space, i.e.,
-\index{diverging color palette}
+\index{color palette!diverging}
 bipolar scales containing colors ranging between two very distinct colors.
@@
 spaces, especially suited for computer displays, are the CIELAB and CIELUV
+\index{color space!CIELAB} \index{color space!CIELUV}
 spaces \cite{vcd:CIE:2004} from which qualitative palettes for
@@
 transformation of the CIELUV space leads to the HCL (Hue-Chroma-Luminance) space.
+\index{color space!HCL}
 These colors are again specified by triplets $(H, C, L)$:
@@
 for the distribution. Recently, with the improving perfomance of computers,
-conditional inference (or permutation tests)---carried out either by simulation
+conditional inference \index{conditional inference}
+(or permutation tests)---carried out either by simulation
 or by computation of the (asymptotic) permutation distribution---have been receiving
@@
 though the residuals are ``small''. In fact, the cut-off points are really
-data-dependent. Consider the case of the \data{arthritis} data \index{Data!arthritis}
+data-dependent. Consider the case of the \data{arthritis} data \index{data!arthritis}
 \cite{vcd:Koch+Edwards:1988}, resulting from a double-blind clinical trial
@@
 As an example, consider the well-known UCB admissions data
-\index{Data!UCB Admissions}
+\index{data!UCB admissions}
 \cite{vcd:Bickel+Hammel+O'Connell:1975} on applicants, classified by
@@
 punishment of children (see Table~\ref{tab:punish}).      
-\index{Data!Danish corporate punishment}
+\index{data!Danish corporal punishment}
 The data consist of four variables: Attitude is a binary variable

### 2006-09-07 trunk/vcd/paper/Color/draft.Rnw
recompiled color draft
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@546 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
improved captions
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@559 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-generate_png <- FALSE
+generate_png <- TRUE
 @
@@
 \includegraphics{hcl-dcone}
-\caption{\label{fig:hsv-hcl} HSV and HCL space}
+\caption{\label{fig:hsv-hcl} HSV and HCL space. The circles show the hue/saturation
+and hue/chroma plane, respectively, for varying levels of value and luminance.}
 \end{center}
@@
 \includegraphics{hcl-dcone-gray}
-\caption{\label{fig:hsv-hcl-gray} HSV and HCL space (in gray levels)}
+\caption{\label{fig:hsv-hcl-gray} HSV and HCL space (in gray levels). The circles
+show the hue/saturation and hue/chroma plane, respectively, for varying levels of
+value and luminance.}
 \end{center}
@@
 \vspace*{-2cm}
-\caption{\label{fig:pal-q-construct} Constructing qualitative palettes}
+\caption{\label{fig:pal-q-construct} Constructing qualitative palettes. In the hue/chroma
+plane for $L = 70$, the dashed circle correponds to a radius $C = 50$ with chosen angles
+$H = 0, 120, 240$.}
 \end{center}
@@
 @
-\caption{\label{fig:pal-q-examples} Examples for qualitative palettes}
+\caption{\label{fig:pal-q-examples} Examples for qualitative palettes. Hue is varied
+in different intervals for given $C = 50$ and $L = 70$.}
 \end{center}
@@
   par(mfrow = c(1, 2))
-  hue.slice(260, grid.n = 301)
+  hue.slice(260, grid.n = 301, main = "H = 260")
   lines(c(0, 100), c(90, 50), lty = 2)
@@
   rect(0, 27.5, 5, 32.5, col = hcl(260, 0, 30))
-  hue.slice(120, grid.n = 301)
+  hue.slice(120, grid.n = 301, main = "H = 120")
   lines(c(0, 100), c(90, 75), lty = 2)
@@
 \includegraphics{pal-s-construct}
-\caption{\label{fig:pal-s-construct} Constructing sequential palettes}
+\caption{\label{fig:pal-s-construct} Constructing sequential palettes. The chroma/luminance
+plane is show for two hues $H = 260$ (left) and $H = 120$ (right). Colors are chosen by varying
+either only luminance or both luminance and chroma.}
 \end{center}
@@
 @
-\caption{\label{fig:pal-s-examples} Examples for sequential palettes}
+\caption{\label{fig:pal-s-examples} Examples for sequential palettes.}
 \end{center}
@@
 \includegraphics{pal-d-construct}
-\caption{\label{fig:pal-d-construct} Constructing diverging palettes}
+\caption{\label{fig:pal-d-construct} Constructing diverging palettes. The chroma/luminance
+plane is shown back to back hues $H = 0$ and $H = 260$. Colors are chosen by simultaneously 
+varying luminance and chroma.}
 \end{center}
@@
 @
-\caption{\label{fig:pal-d-examples} Examples for diverging palettes}
+\caption{\label{fig:pal-d-examples} Examples for diverging palettes.}
 \end{center}
@@
 \includegraphics{bkde}
-\caption{\label{fig:bkde} Bivariate density estimation for Old Faithful geyser eruptions}
+\caption{\label{fig:bkde} Bivariate density estimation for Old Faithful geyser eruptions.}
 \end{center}
@@
 @
-\caption{\label{fig:seats} Seats in the German parliament}
+\caption{\label{fig:seats} Seats in the German parliament.}
 \end{center}
@@
 @
-\caption{\label{fig:votes} Votes in the German election 2005}
+\caption{\label{fig:votes} Votes in the German election 2005.}
 \end{center}
@@
 @
-\caption{\label{fig:arthritis} Extended mosaic display for arthritis data}
+\caption{\label{fig:arthritis} Extended mosaic display for arthritis data.}
 \end{center}
@@
 \includegraphics{class}
-\caption{\label{fig:class} SVM classification plot}
+\caption{\label{fig:class} SVM classification plot.}
 \end{center}
@@
 
+%FIXME% Acknowledgements
+%% Paul Murrell, Ross Ihaka, Brian D. Ripley, David Meyer, Thomas Kneib, Michael Hoehle
+
 \bibliography{vcd}

### 2006-09-07 trunk/vcd/paper/Color/functions.R
improved captions
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@559 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -68,7 +68,7 @@ hcl.wheel <-
 }
 
 ## show slices for given hues
-hue.slice <- function(hue, grid.n = 251, plot = TRUE, fixup = FALSE)
+hue.slice <- function(hue, grid.n = 251, plot = TRUE, fixup = FALSE, main = NULL)
 {
   chroma = seq(0, 100, length = grid.n)
   luminance = seq(0, 100, length = grid.n)
@@ -77,7 +77,7 @@ hue.slice <- function(hue, grid.n = 251, plot = TRUE, fixup = FALSE)
   color.slice <- outer(chroma, luminance, function(y, x) hcl(hue, x, y, fixup = fixup))
   xlab <- "chroma"
   ylab <- "luminance"
-  main <- paste("hue =", round(hue, digits = 0))
+  if(is.null(main)) main <- paste("hue =", round(hue, digits = 0))
   if(plot) {
     plot(0.5, 0.5, xlim = range(chroma), ylim = range(luminance), type = "n", axes = FALSE,
          xlab = xlab, ylab = ylab, yaxs = "i", xaxs = "i", main = main)

### 2006-09-07 trunk/vcd/paper/Color/draft.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@560 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
   par(mfrow = c(1, 2))
-  hue.slice(260, grid.n = 301, main = "H = 260")
+  hue.slice(260, grid.n = 301)
   lines(c(0, 100), c(90, 50), lty = 2)
@@
   rect(0, 27.5, 5, 32.5, col = hcl(260, 0, 30))
-  hue.slice(120, grid.n = 301, main = "H = 120")
+  hue.slice(120, grid.n = 301)
   lines(c(0, 100), c(90, 75), lty = 2)

### 2006-09-14 trunk/vcd/paper/Shadings/JCGS.Rnw
Code:   removed border between cells in legend
Paper:
  finished JCGS revision and rejoinder



git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@562 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 nrep <- 5000
+nrep_char <- "5,000"
 recompute <- TRUE
@@
 to visualize independence models for contingency tables are extended
-in two directions: (a)~perceptually uniform HCL colors are used and (b)~the
+in two directions: (a)~perceptually uniform Hue-Chroma-Luminance colors are used and (b)~the
 result of an associated significance test is coded by the appearance of
@@
 log-linear model that elevates the mosaic plot from a display for frequencies
-in a contingency table to a visualization technique that encompasses both, observed
+in a contingency table to a visualization technique that encompasses both observed
 frequencies and residuals. This shading allows for judging the quality of a model
@@
 The estimated expected cell frequencies under $H_0$ are
-$\hat n_{ij} = n_{i+}n_{+j}/n_{++}$.
-%FIXME% point out homogeneity problem?
+$\hat n_{ij} = n_{i+}n_{+j}/n_{++}$. As well-established in the stastical literature,
+a very closely related hypothesis is that of homogeneity which in particular leads to the same
+expected cell frequencies and is hence not discussed explicitely below.
 The probably best known and most used
@@
 with $p = \Sexpr{pval(art_max)}$. Both $p$ values have been computed
-from a sample of size $\Sexpr{nrep}$ from the permutation distribution under independence
+from a sample of size $\Sexpr{nrep_char}$ from the permutation distribution under independence
 generated via sampling tables with the same row and column sums $n_{i+}$ and $n_{+j}$
-using the \cite{vcd:Patefield:1981} algorithm and computing the maximum statistic for 
+using the \cite{vcd:Patefield:1981} algorithm and computing the respective statistic for 
 each of these tables.
@@
 
+\emph{Mosaic plots}
+\citep{vcd:Hartigan+Kleiner:1981} can be seen as an extension of grouped
+bar charts where width and height of the bars show the relative
+frequencies of the two variables: a mosaic plot simply
+consists of a collection of tiles with areas
+proportional to the observed cell frequencies as shown
+in the left panel of Figure~\ref{fig:classic}. A rectangle corresponding
+to $100$ percent of the observations is first split horizontally with respect to 
+the treatment frequencies and then vertically with respect to the conditional
+improvement frequencies. This shows that there have been more placebo than treated 
+patients with no improvement and vice versa for marked improvement.
+This strategy of splitting with respect to conditional frequencies given all
+previous variables can also directly be used for visualizing multi-way tables
+\citep[see][for an overview of how to construct and read mosaic displays]{vcd:Hofmann:2003}.
+
+\emph{Association plots}
+\citep{vcd:Cohen:1980} visualize the table of Pearson residuals:
+each cell is represented by a rectangle that has (signed) height
+proportional to the corresponding Pearson residual $r_{ij}$ and width proportional to the
+square root of the estimated expected counts $\sqrt{\hat n_{ij}}$. Thus, the area is proportional to
+the raw residuals $n_{ij} - \hat n_{ij}$. The association plot for the arthritis
+data is shown in the right panel of Figure~\ref{fig:classic} which leads to the
+same interpretation as the mosaic plot: there are more placebo patients with no 
+improvement and fewer with marked improvement than expected under independence---vice
+versa for the treated patients.
+
+\clearpage
+
 \setkeys{Gin}{width=\textwidth}
-\begin{figure}[b!]
+\begin{figure}[t!]
 \begin{center}
@@
 
-\emph{Mosaic plots}
-\citep{vcd:Hartigan+Kleiner:1981} can be seen as an extension of grouped
-bar charts where width and height of the bars show the relative
-frequencies of the two variables: a mosaic plot simply
-consists of a collection of tiles with areas
-proportional to the observed cell frequencies as shown
-in the left panel of Figure~\ref{fig:classic}. A rectangle corresponding
-to $100$ percent of the observations is first split horizontally with respect to 
-the treatment frequencies and then vertically with respect to the conditional
-improvement frequencies. This shows that there have been more placebo than treated 
-patients with no improvement and vice versa for marked improvement.
-This strategy of splitting with respect to conditional frequencies given all
-previous variables can also directly be used for visualizing multi-way tables
-\citep[see][for an overview of how to construct and read mosaic displays]{vcd:Hofmann:2003}.
-
-\emph{Association plots}
-\citep{vcd:Cohen:1980} visualize the table of Pearson residuals:
-each cell is represented by a rectangle that has (signed) height
-proportional to the corresponding Pearson residual $r_{ij}$ and width proportional to the
-square root of the estimated expected counts $\sqrt{\hat n_{ij}}$. Thus, the area is proportional to
-the raw residuals $n_{ij} - \hat n_{ij}$. The association plot for the arthritis
-data is shown in the right panel of Figure~\ref{fig:classic} which leads to the
-same interpretation as the mosaic plot: there are more placebo patients with no 
-improvement and fewer with marked improvement than expected under independence---vice
-versa for the treated patiens.
-
 \section{Residual-based Shadings} \label{sec:shadings}
@@
 
-\clearpage
-
 \subsection{Colors} \label{sec:colors}
@@
 
-The HSV space looks like a cone\footnote{The wikipedia gives a very useful
-%FIXME% proper cite for Wikipedia
-overview at \url{http://en.wikipedia.org/wiki/HSV_color_space} which also provides
-links to comparisons with other color spaces discussed in this paper.}
+The HSV space looks like a cone \citep[see e.g.,][which also provides
+links to comparisons with other color spaces discussed in this paper]{vcd:Wiki+HSV:2006}
 with black at its peak (zero value) and full color wheels for different saturations at the other end,
@@
 association plots and although HSV is a very commonly available implementation of
-color spaces, HSV colors in general and the Friendly shading in particular have
+color spaces, HSV color space in general and the Friendly shading in particular have
 a number of disadvantages. Most importantly, HSV colors are not perceptually
@@
 $h = 260$ is chosen rather than $h = 240$ because its chroma/luminance plane is most similar,
-as assessed by the symmetric difference of the planes, to that of $h = 0$.}
+as assessed by the area of the symmetric difference of the planes, to that of $h = 0$.}
 To obtain a sequence of colors with the similar properties as the Friendly shading, the palette starts
@@
 color or by changing the maximum chroma, respectively, this can be changed to using only
-chroma or luminance for this purpose, but using both seems to be a very effective way
-of visualization and corresponds more closely with the properties of the Friendly shading.
-
-Applying these palettes to the mosaic plot of the arthritis data, yields
+chroma or luminance for this purpose, but using both is a very effective way
+of visualization (i.e., yields palettes with more distinctive colors)
+and corresponds more closely with the properties of the Friendly shading.
+
+Applying these palettes to the mosaic plot of the arthritis data yields
 the displays in the upper middle and right panels of Figure~\ref{fig:shadings} (with
@@
 
-To see such a sum-of-squares shading in practice, we employ a data on the number of piston
+To see such a sum-of-squares shading in practice, we employ a data set on the number of piston
 ring failures in three legs (north/center/south) in four different steam-driven compressors
-at an FIXME. The contingency table is given in \cite{vcd:Haberman:1973}, re-analyzed
+at an Imperial Chemical Industries plant. The contingency table is given in \cite{vcd:Haberman:1973}, re-analyzed
 by \cite{vcd:Everitt+Hothorn:2006} and displayed in the lower row of Figure~\ref{fig:shadings}.
@@
 nor the maximum test ($M = \Sexpr{stat(pring_max)}$, $p = \Sexpr{pval(pring_max)}$) find
-evidence for a departure from independence in the data (at 5\% level). However, as argued
+evidence for a departure from independence in the data (at the 5\% level). However, as argued
 by \cite{vcd:Everitt+Hothorn:2006}, a closer look at the size of the residuals might be
@@
 of the associated test (at 5\% level). Comparing the HSV-based and HCL-based version
-shows again that the latter are less flashy and more balanced.
+shows that the latter are less flashy and more balanced.
 
@@
 For a table $[n_{ijk}]$ with underlying
-theoretical probability distribution $[p_{ijk}]$, this can be formulated as
+theoretical probability distribution $[\pi_{ijk}]$, this can be formulated as
 \begin{equation} \label{eq:H0ci}
@@
 \end{equation}
-As in the 2-way case, this allows for identification of the cells that are responsible
+As in the 2-way case, this allows for identification of the cells which are responsible
 for the deviation from conditional independence (if any). If it is not so much of interest
@@
 If the maximum is used for the latter, then identification of the strata responsible for the
-dependence is possible. If additionally $\lambda_\mathrm{indep}$ is the maximum,
+non-independence is possible. If additionally $\lambda_\mathrm{indep}$ is the maximum,
 the corresponding cells can also be identified. Hence, the double maximum
@@
 is $\max_k M_k = \Sexpr{stat(alz_dmax)}$ ($p = \Sexpr{pval(alz_dmax)}$).
-The $p$ values are again computed by drawing $\Sexpr{nrep}$ samples from the corresponding
+The $p$ values are again computed by drawing $\Sexpr{nrep_char}$ samples from the corresponding
 permutation distribution. The conditional mosaic plot in Figure~\ref{fig:alz}
@@
 
-As a 4-way example, we use the punishment data \cite{vcd:andersen:1991} 
+As a 4-way example, we use the punishment data \citep{vcd:andersen:1991} 
 from a study of the Gallup Institute in Denmark in 1979 about
@@
 interpreted as a diagnostic residual plot for the associated log-linear conditional
-indpendence model. It shows an association between memories and attitude
+independence model. It shows an association between memories and attitude
 for the lowest education group (first column) and highest age group (last row):
@@
 \pkg{vcd}~\Sexpr{packageDescription("vcd")$Version} \citep{vcd:Meyer+Zeileis+Hornik:2006},
-{\pkg{MASS}~\Sexpr{packageDescription("MASS")$Version}} \citep{vcd:Venables+Ripley:2002},
+{\pkg{MASS}~\Sexpr{packageDescription("MASS")$Version}} \citep[see][]{vcd:Venables+Ripley:2002},
 {\pkg{grid}~{\Sexpr{packageDescription("grid")$Version}}} \citep[see][]{vcd:Murrell:2002} and
@@
 set.seed(rseed)
-(art_max <- coindep_test(art, n = 5000))
+art_max <- coindep_test(art, n = 5000)
 set.seed(rseed)
@@
 \begin{center}
-<<conditional_assoc,fig=TRUE,echo=FALSE,height=6,width=9>>=
+<<conditional_assoc,fig=TRUE,echo=FALSE,height=6,width=9,eval=FALSE>>=
 cotabplot(~ Gender + Admit | Dept, data = ucb, panel = ucb_cotab)
@@
 %% employment data
-<<employment>>
+<<employment,eval=FALSE>>
 ## data
@@
 \begin{center}
-<<Friendly-hue,fig=TRUE,echo=FALSE,height=4.8,width=8>>
+<<Friendly-hue,fig=TRUE,echo=FALSE,height=4.8,width=8,eval=FALSE>>
 plot(0.5, 0.5, xlim = c(0, 1), ylim = c(0, 1), type = "n", axes = FALSE, xlab = "", ylab = "", yaxs = "i")
@@
 
-<<Friendly-sat,fig=TRUE,echo=FALSE,height=4.8,width=8>>
+<<Friendly-sat,fig=TRUE,echo=FALSE,height=4.8,width=8,eval=FALSE>>
 plot(0.5, 0.5, xlim = c(0, 1), ylim = c(0, 1), type = "n", axes = FALSE, xlab = "", ylab = "", yaxs = "i")
@@
 
-<<Friendly-val,fig=TRUE,echo=FALSE,height=4.8,width=8>>
+<<Friendly-val,fig=TRUE,echo=FALSE,height=4.8,width=8,eval=FALSE>>
 plot(0.5, 0.5, xlim = c(0, 1), ylim = c(0, 1), type = "n", axes = FALSE, xlab = "", ylab = "", yaxs = "i")

### 2006-09-15 trunk/vcd/paper/Strucplot/JSS.Rnw
-new argument for color of unshaded zeros -panel-functions of cotabplot compute prefix on their own.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@567 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2006-01-04 trunk/vcd/paper/JSS.Rnw
@@
     \hline
-                             & \proglang{SAS} & \proglang{S-PLUS} & \proglang{R} & \pkg{ViSta} & \pkg{MANET}/\pkg{Mondrian}\\\hline
-    Basic functionality      &  $\times$  &   $\times$    & $\times$ &   $\times$   & $\times$\\
-    Shape                    &     &        & $\times$ &       & $\times$\\
-    Residual-based shadings  &  $\times$  &        & $\times$ &   $\times$   & ($\times$)\\
-    Conditional Views        &  $\times$  &        & $\times$ &       & $\times$\\
-    Interaction              &     &        &   &   $\times$   & $\times$\\
-    Extensible Design        &     &        & $\times$ &       &  \\\hline
+                             & \proglang{SAS} & \proglang{S-PLUS} &
+                             \proglang{R} & \pkg{ViSta} & \pkg{MANET}
+                             & \pkg{Mondrian}\\\hline
+    Basic functionality      &  $\times$  &   $\times$    & $\times$ &   $\times$   & $\times$& $\times$\\
+    Shape                    &     &        & $\times$ & $\times$  & &\\
+    Residual-based shadings  &  $\times$  &        & $\times$ &   highlighting   &($\times$)& ($\times$)\\
+    Conditional Views        &  $\times$  &        & $\times$ & $\times$ & &\\
+    Interaction              &     &        &   &   $\times$   & $\times$& $\times$\\
+    Linking                  &     &        &   &              & $\times$& $\times$\\
+    Extensible Design        &     &        & $\times$ &       &  &\\
+    Language                 & SAS & S & S & XLisp & ?? & Java\\
+\hline
   \end{tabular}

### 2006-09-15 trunk/vcd/R/cotabplot.R

@@ -142,8 +142,7 @@ cotabplot.default <- function(x, cond = NULL,
 
       ## main plot
       pushViewport(viewport(layout.pos.row = 2, name = paste("plot", condistr, sep = ".")))
-      panel(x, condi,
-            paste("panel:", paste(names(condi), condi, sep = "=", collapse = ","), "|", sep = ""))
+      panel(x, condi)
       upViewport(2)
       grid.rect(gp = gpar(fill = "transparent"))
       upViewport()
@@ -157,19 +156,21 @@ cotabplot.default <- function(x, cond = NULL,
 }
 
 cotab_mosaic <- function(x = NULL, condvars = NULL, ...) {
-  function(x, condlevels, prefix) {
+  function(x, condlevels) {
     if(is.null(condlevels)) mosaic(x, newpage = FALSE, pop = FALSE, ...)
       else mosaic(co_table(x, names(condlevels))[[paste(condlevels, collapse = ".")]],
-                  newpage = FALSE, pop = FALSE, prefix = prefix, ...)
+                  newpage = FALSE, pop = FALSE,
+                  prefix = paste("panel:", paste(names(condlevels), condlevels, sep = "=", collapse = ","), "|", sep = ""), ...)
   }
 }
 class(cotab_mosaic) <- "grapcon_generator"
 
 cotab_sieve <- function(x = NULL, condvars = NULL, ...) {
-  function(x, condlevels, prefix = prefix) {
+  function(x, condlevels) {
     if(is.null(condlevels)) sieve(x, newpage = FALSE, pop = FALSE, ...)
       else sieve(co_table(x, names(condlevels))[[paste(condlevels, collapse = ".")]],
-                 newpage = FALSE, pop = FALSE, prefix = prefix, ...)
+                 newpage = FALSE, pop = FALSE,
+                 prefix = paste("panel:", paste(names(condlevels), condlevels, sep = "=", collapse = ","), "|", sep = ""), ...)
   }
 }
 class(cotab_sieve) <- "grapcon_generator"
@@ -180,16 +181,17 @@ cotab_assoc <- function(x = NULL, condvars = NULL, ylim = NULL, ...) {
     if(is.null(ylim)) ylim <- range(residuals(fm))
   }
   
-  function(x, condlevels, prefix) {
+  function(x, condlevels) {
     if(is.null(condlevels)) assoc(x, newpage = FALSE, pop = FALSE, ylim = ylim, ...)
[…]

### 2006-09-15 trunk/vcd/R/mosaic.R

@@ -78,6 +78,7 @@ mosaic.default <- function(x, condvars = NULL,
                            zero_size = 0.5,
                            zero_split = FALSE,
                            zero_shade = TRUE,
+                           zero_color = gpar(fill = 0),
                            main = NULL, sub = NULL, ...) {
   if (is.logical(main) && main)
     main <- deparse(substitute(x))
@@ -150,7 +151,7 @@ mosaic.default <- function(x, condvars = NULL,
   strucplot(x,
             condvars = if (is.null(condvars)) NULL else length(condvars),
             core = struc_mosaic(zero_size = zero_size, zero_split = zero_split,
-              zero_shade = zero_shade),
+              zero_shade = zero_shade, zero_color = zero_color),
             split_vertical = split_vertical,
             spacing = spacing,
             spacing_args = spacing_args,
@@ -160,7 +161,8 @@ mosaic.default <- function(x, condvars = NULL,
             ...)
 }
 
-struc_mosaic <- function(zero_size = 0.5, zero_split = FALSE, zero_shade = TRUE)
+struc_mosaic <- function(zero_size = 0.5, zero_split = FALSE,
+                         zero_shade = TRUE, zero_color = gpar(fill = 0))
   function(residuals, observed, expected = NULL, spacing, gp, split_vertical, prefix = "") {
     dn <- dimnames(observed)
     dnn <- names(dn)
@@ -226,7 +228,7 @@ struc_mosaic <- function(zero_size = 0.5, zero_split = FALSE, zero_shade = TRUE)
       grid.lines(y = 0.5)
       if (!zero_shade && zero_size > 0) {
         grid.points(0.5, 0.5, pch = 19, size = unit(zero_size, "char"),
-                    gp = gpar(col = "grey"),
+                    gp = zero_color,
                     name = paste(prefix, "disc:", mnames[i], sep = ""))
         grid.points(0.5, 0.5, pch = 1, size = unit(zero_size, "char"),
                     name = paste(prefix, "circle:", mnames[i], sep = ""))

### 2006-09-15 trunk/vcd/paper/Shadings/JCGS.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@568 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 to visualize independence models for contingency tables are extended
-in two directions: (a)~perceptually uniform Hue-Chroma-Luminance colors are used and (b)~the
+in two directions: (a)~perceptually uniform Hue-Chroma-Luminance (HCL)
+colors are used and (b)~the
 result of an associated significance test is coded by the appearance of

### 2006-09-20 trunk/vcd/paper/Strucplot/JSS.Rnw
- spacings-default synchronized - pairs_diagonal_mosaic added - labeling can now create alternating labels
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@571 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2006-01-04 trunk/vcd/paper/JSS.Rnw
@@
 \proglang{Mondrian} \citep{vcd:Theus:2003},
-\proglang{R} \citep{vcd:R:2005}, and \proglang{S-PLUS} \citep{vcd:SPLUS:2005}. 
+\proglang{R} \citep{vcd:R:2006}, and \proglang{S-PLUS}
+\citep{vcd:SPLUS:2005}. For \proglang{R}, currently three
+implementations do exist in the packages \pkg{graphics} (in base \proglang{R}), 
+\pkg{vcd} \citep{vcd:Meyer+Zeileis+Hornik:2006},
+and \pkg{iPlots} \citep{vcd:urbanek+wichtrey:2006}, respectively.
 Table \ref{tab:compare} gives an overview of the available functionality in these systems.
+Most environments are available on Windows, Apple Macintosh, and Linux/Unix
+variants, except \proglang{MANET} which is only available for the
+Macinthosh platforms. \proglang{Mondrian} is a JAVA application, as
+well as major parts of the \pkg{iPlots} package.
 
@@
   \centering
-  \begin{tabular}{|l|c|c|c|c|c|}
+  \begin{tabular}{|l|c|c|c|c|c|c|}
     \hline
-                             & \proglang{SAS} & \proglang{S-PLUS} & \proglang{R} & \pkg{ViSta} & \pkg{MANET}/\pkg{Mondrian}\\\hline
-    Basic functionality      &  $\times$  &   $\times$    & $\times$ &   $\times$   & $\times$\\
-    Shape                    &     &        & $\times$ &       & $\times$\\
-    Residual-based shadings  &  $\times$  &        & $\times$ &   $\times$   & ($\times$)\\
-    Conditional Views        &  $\times$  &        & $\times$ &       & $\times$\\
-    Interaction              &     &        &   &   $\times$   & $\times$\\
-    Extensible Design        &     &        & $\times$ &       &  \\\hline
+                             & \proglang{SAS} & \proglang{S-PLUS} &
+                             \proglang{R} & \pkg{ViSta} & \pkg{MANET}
+                             & \pkg{Mondrian}\\\hline
+    Basic functionality      &  $\times$  &   $\times$    & $\times$ &   $\times$   & $\times$& $\times$\\
+    Shape                    &     &        & $\times$ & $\times$  &  $\times$&\\
+    Residual-based shadings  &  $\times$  &        & $\times$ &   highlighting   &($\times$)& ($\times$)\\
+    Conditional Views        &  $\times$  &        & $\times$ & $\times$ &  $\times$&\\
+    Interaction              &     &        & $\times$  &   $\times$   & $\times$& $\times$\\
+    Linking                  &     &        & $\times$  &       & $\times$& $\times$\\
+    Extensible Design        &     &        & $\times$ &       &  &\\
+    Language                 & SAS & S & S,JAVA & XLisp & C++ & JAVA\\
+\hline
   \end{tabular}
@@
 \end{table}
+
+%%DM%% Fixme: separate columns for iplot, base, and vcd?
 
@@
 
-\noindent The small bullets indicate zero entries in the corresponding
-cell. Note that in contrast to, e.g.,  \codefun{mosaicplot} in base
+\noindent For zero entries, tiles have zero area and are, additionally,
+marked by small bullets (see, e.g, Figure \ref{fig:doubledecker}). By
+default, these cells are not splitted further. The bullets help
+distinguishing very small cells from zero entries, and are
+particularly useful when color shadings come into play 
+(see the example using the \data{Bundesliga} data in Section \ref{sec:overview}). 
+Note that in contrast to, e.g.,  \codefun{mosaicplot} in base
 \proglang{R}, the default split direction and level ordering in all strucplot
@@
 can be customized in a quite flexible way. For example, 
-\codefun{doubledecker} is simply a wrapper for 
-\codefun{mosaic}, setting the right defaults. 
+\codefun{doubledecker.default} is simply a wrapper for 
+\codefun{strucplot}, setting the right defaults:
+
+<<doubledecker>>=
+doubledecker.default <- 
+  function(x,
+           depvar = length(dim(x)), 
+           margins = c(1, 4, length(dim(x)) + 1, 1),
+           col = rainbow_hcl(tail(dim(x), 1), end = 260),
+           labeling = labeling_doubledecker,
+           spacing = spacing_highlighting,
+           main = NULL, 
+           keep_aspect_ratio = FALSE,
+           ...) {
+  d <- dim(x)
+  l <- length(d)
+  if (is.character(depvar))
+    depvar <- match(depvar, names(dimnames(x)))
+  condvars <- (1:l)[-depvar]
+  x <- aperm(x, c(condvars, depvar))
+  strucplot(x,
+         condvars = l - 1,
+         spacing = spacing,
+         split = c(rep.int(TRUE, l - 1), FALSE),
+         gp = gpar(fill = col),
+         shade = TRUE,
+         labeling = labeling,
+         main = main,
+         margins = margins,
+         legend = NULL,
+         keep_aspect_ratio = keep_aspect_ratio,
+         ...
+         )
+}
+@
+
+\noindent Most default settings such as colors, spacing, and labeling are specified via the parameters of
+\codefun{doubledecker.default} and passed through to
+\codefun{strucplot}. The additional code just handles the dependent
+variable information, and in particular permutes the table to have the
+variable as the last dimension as required for the doubledecker plot.
 Figure \ref{fig:doubledecker} shows a doubledecker plot of
@@
 \end{figure}
+
+\noindent (Note that the bullets drawn for zero cells are shaded, too).
 
@@
 improvement in both groups. To obtain such a display within 
-the strucplot framework\footnote{The \pkg{vcd} package also provides 
-a specialized \codefun{spine} function for spine plots and spinograms.}, it
-suffices to set the space between the `Improved' tiles to 0 (see
+the strucplot framework, it suffices to set the space between the `Improved' tiles to 0 (see
 Figure \ref{fig:artspine}):
@@
 \end{figure}
+
+\noindent Note that the default and formula methods for 
+\codefun{mosaic} provide a
+convenience interface for highlighting. A similar plot (with slightly
+different shading) than the previous one can be obtained using:
+
+<<artspine>>=
+mosaic(Improved ~ Treatment, data = Arthritis, split_vertical = TRUE)
+@
 

### 2006-09-20 trunk/vcd/R/labeling.R

@@ -136,7 +136,8 @@ class(labeling_cells) <- "grapcon_generator"
 
 labeling_border <- function(labels = TRUE, varnames = labels,
                             set_labels = NULL, set_varnames = NULL,
-                            tl_labels = NULL, tl_varnames = NULL, 
+                            tl_labels = NULL, alternate_labels = FALSE,
+                            tl_varnames = NULL,
                             gp_labels = gpar(fontsize = 12),
                             gp_varnames = gpar(fontsize = 12, fontface = 2),
                             rot_labels = c(0, 90, 0, 90),
@@ -153,29 +154,53 @@ labeling_border <- function(labels = TRUE, varnames = labels,
                             abbreviate = FALSE, rep = TRUE,
                             clip = FALSE, ...
                             ) {
-  function(d, split_vertical, condvars, prefix = "") {
-    if (is.table(d))
-      d <- dimnames(d)
-    dn <- names(d)
-    ld <- length(d)
-
-    ## expand parameters
-    clip <- pexpand(clip, ld, TRUE, dn)
-    labels <- pexpand(labels, ld, TRUE, dn)
-    labels_varnames <- pexpand(labels_varnames, ld, FALSE, dn)
-    pos_labels <- pexpand(pos_labels, 4, "center", c("top", "right", "bottom", "left"), c("left", "center", "right"))
-    just_labels <- pexpand(just_labels, 4, "center", c("top", "right", "bottom", "left"), c("left", "center", "right"))
+    ## expand parameters that apply to the four table margins
+    pos_labels <- pexpand(pos_labels, 4, "center", c("top", "right", "bottom", "left"),
+                          c("left", "center", "right"))
+    just_labels <- pexpand(just_labels, 4, "center", c("top", "right", "bottom", "left"),
+                           c("left", "center", "right"))
     offset_varnames <- if (!is.unit(offset_varnames))
       unit(pexpand(offset_varnames, 4,
                    rep.int(0, 4), c("top","right","bottom","left")), "lines")
     else
       rep(offset_varnames, length.out = 4)
+    
     offset_labels <- if (!is.unit(offset_labels))
       unit(pexpand(offset_labels, 4,
                    rep.int(0, 4), c("top","right","bottom","left")), "lines")
     else
[…]

### 2006-09-20 trunk/vcd/R/pairsplot.R
- spacings-default synchronized - pairs_diagonal_mosaic added - labeling can now create alternating labels
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@571 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -2,22 +2,25 @@
 ### pairsplot
 
 pairs.table <- function(x,
-                  upper_panel = pairs_mosaic,
-                  upper_panel_args = list(),
+                        upper_panel = pairs_mosaic,
+                        upper_panel_args = list(),
                         
-                  lower_panel = pairs_mosaic,
-                  lower_panel_args = list(),
+                        lower_panel = pairs_mosaic,
+                        lower_panel_args = list(),
                         
-                  diag_panel = pairs_barplot,
-                  diag_panel_args = list(),
+                        diag_panel = pairs_barplot,
+                        diag_panel_args = list(),
                   
-                  main = NULL,
-                  title_gp = gpar(fontsize = 20),
+                        main = NULL,
+                        sub = NULL,
+                        main_gp = gpar(fontsize = 20),
+                        sub_gp = gpar(fontsize = 15),
                   
-                  space = 0.1,
-                  newpage = TRUE,
-                  pop = TRUE,
-                  ...)
+                        space = 0.3,
+                        newpage = TRUE,
+                        pop = TRUE,
+                        margins = unit(1, "lines"),
+                        ...)
 {
   if (newpage) grid.newpage()
 
@@ -31,14 +34,25 @@ pairs.table <- function(x,
   d <- length(dim(x))
   l <- grid.layout(d, d)
   pushViewport(viewport(width = unit(1, "snpc"), height = unit(1, "snpc")))
-  pushViewport(viewport(layout = l,
-                         height = if (is.null(main)) 1 else 0.9,
-                         y = 0, just = "bottom"))
+  pushViewport(vcdViewport(mar = margins, legend = FALSE, legend_width = NULL,
+                           main = !is.null(main), sub = !is.null(sub)))
+  ## titles
+  if (!is.null(main)) {
+    seekViewport("main")
+    if (is.logical(main) && main)
+      main <- deparse(substitute(x))
+    grid.text(main, gp = main_gp)
+  }
 
-  if (is.logical(main) && main)
-    main <- deparse(substitute(x))
-  grid.text(main, y = unit(1.05, "npc"), gp = title_gp)
+  if (!is.null(sub)) {
+    seekViewport("sub")
+    if (is.logical(sub) && sub && is.null(main))
+      sub <- deparse(substitute(x))
+    grid.text(sub, gp = sub_gp)
+  }
 
[…]

### 2006-09-20 trunk/vcd/R/spacings.R

@@ -1,7 +1,7 @@
 ##################################################################
 ## spacings
 
-spacing_equal <- function(sp = unit(0.5, "lines")) {
+spacing_equal <- function(sp = unit(0.3, "lines")) {
   if (!is.unit(sp)) sp <- unit(sp, "lines")
   function(d, condvars = NULL) lapply(d, function(x) rep(sp, x - 1))
 }
@@ -23,7 +23,7 @@ spacing_increase <- function(start = unit(0.3, "lines"), rate = 1.5) {
 }
 class(spacing_increase) <- "grapcon_generator"
 
-spacing_highlighting <- function(start = unit(0.2, "lines"), rate = 1.5) {
+spacing_highlighting <- function(start = unit(0.3, "lines"), rate = 1.5) {
  if (!is.unit(start)) start <- unit(start, "lines")
   function(d, condvars = NULL)
     c(spacing_increase(start, rate)(d, condvars)[-length(d)],
@@ -31,7 +31,7 @@ spacing_highlighting <- function(start = unit(0.2, "lines"), rate = 1.5) {
 }
 class(spacing_highlighting) <- "grapcon_generator"
 
-spacing_conditional <- function(sp = unit(0.5, "lines"),
+spacing_conditional <- function(sp = unit(0.3, "lines"),
                                 start = unit(2, "lines"), rate = 1.8) {
   condfun <- spacing_increase(start, rate)
   equalfun <- spacing_equal(sp)

### 2006-09-20 trunk/vcd/R/spacings.R
small bug
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@572 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -23,7 +23,7 @@ spacing_increase <- function(start = unit(0.3, "lines"), rate = 1.5) {
 }
 class(spacing_increase) <- "grapcon_generator"
 
-spacing_highlighting <- function(start = unit(0.3, "lines"), rate = 1.5) {
+spacing_highlighting <- function(start = unit(0.2, "lines"), rate = 1.5) {
  if (!is.unit(start)) start <- unit(start, "lines")
   function(d, condvars = NULL)
     c(spacing_increase(start, rate)(d, condvars)[-length(d)],

### 2006-09-21 trunk/vcd/R/labeling.R
- alternating labeling starts on top - bars and tiles in pairs_barplot() and pairs_diagonal_mosaic   can be filled using a palette
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@573 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -525,7 +525,8 @@ labeling_border <- function(labels = TRUE, varnames = labels,
       
       ## mask half of the labels
       for (i in which(alternate_labels))
-        set_labels[[i]][seq(1, length(d[[i]]), 2)] <- ""
+        if (length(d[[i]]) > 1)
+          set_labels[[i]][seq(2, length(d[[i]]), 2)] <- ""
     }
     
     split()
@@ -544,8 +545,7 @@ labeling_border <- function(labels = TRUE, varnames = labels,
       
       ## mask other half of alternated labels
       for (i in which(alternate_labels))
-        if (length(d[[i]]) > 1)
-          set_labels[[i]][seq(2, length(d[[i]]), 2)] <- ""
+        set_labels[[i]][seq(1, length(d[[i]]), 2)] <- ""
       
       ## invert tl_labels and labsp
       tl_labels <- ! tl_labels

### 2006-09-21 trunk/vcd/R/pairsplot.R

@@ -162,17 +162,21 @@ pairs_diagonal_text <- function(varnames = TRUE,
 }
 class(pairs_diagonal_text) <- "grapcon_generator"
 
-pairs_barplot <- function(gp_bars = gpar(fill = "gray"),
+pairs_barplot <- function(gp_bars = NULL,
                           gp_vartext = gpar(fontsize = 17),
                           gp_leveltext = gpar(),
                           just_leveltext = c("center", "bottom"),
                           just_vartext = c("center", "top"),
                           rot = 0, abbreviate = FALSE,
                           check_overlap = TRUE,
+                          fill = "grey",
                           ...)
   function(x, i) {
     dn <- names(dimnames(x))
     x <- margin.table(x, i)
+    if (is.function(fill)) fill <- rev(fill(dim(x)))
+    if (is.null(gp_bars))
+      gp_bars <- gpar(fill = fill)
     pushViewport(viewport(x = 0.3, y = 0.1, width = 0.7, height = 0.7,
                           yscale = c(0,max(x)), just = c("left", "bottom"))
                  )
@@ -196,13 +200,29 @@ pairs_barplot <- function(gp_bars = gpar(fill = "gray"),
   }
 class(pairs_barplot) <- "grapcon_generator"
 
-pairs_diagonal_mosaic <- function(split_vertical = TRUE, margins = unit(0, "lines"),
-                                  alternate_labels = TRUE, ...)
-  function(x, i) mosaic(margin.table(x, i),
-                        newpage = FALSE,
-                        split_vertical = split_vertical,
-                        margins = margins,
-                        alternate_labels = alternate_labels,
-                        prefix = "diag", ...)
+pairs_diagonal_mosaic <- function(split_vertical = TRUE,
+                                  margins = unit(0, "lines"),
+                                  alternate_labels = TRUE,
+                                  offset_labels = -0.4,
+                                  offset_varnames = 0,
+                                  gp = NULL,
+                                  fill = "grey",
+                                  ...)
+  function(x, i) {
+    if (is.function(fill))
+      fill <- rev(fill(dim(x)[i]))
+    if (is.null(gp))
+      gp <- gpar(fill = fill)
+    mosaic(margin.table(x, i),
+           newpage = FALSE,
+           split_vertical = split_vertical,
+           margins = margins,
+           alternate_labels = alternate_labels,
+           offset_labels = offset_labels,
+           offset_varnames = offset_varnames,
[…]

### 2006-09-21 trunk/vcd/paper/Strucplot/JSS.Rnw
- alternating labeling starts on top - bars and tiles in pairs_barplot() and pairs_diagonal_mosaic   can be filled using a palette
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@573 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 graphics, available from the Comprehensive \proglang{R} Archive Network
-(\url{http://cran.r-project.org/}).  The framework integrates displays
+(\url{http://CRAN.R-project.org/}).  The framework integrates displays
 such as mosaic, association, and sieve plots by their unifying property
@@
               shading (based on HSV-colors)\\
+              & \codefun{shading\_sieve} & implements Friendly-like
+              shading used for sieve plots\\ && (based on HCL-colors)\\
               & \codefun{shading\_hcl} & shading based on HCL-colors\\
@@
 <<split2>>=
-mosaic(hec, direction = c("v","h","v"))
+mosaic(hec, direction = c("h","h","v"))
 @
@@
 
-So far, we have visualized full tables. Subtables can be selected in a
+So far, we have visualized either full or collapsed tables, as
+suggested by the analysis task at hand. Subtables can be selected in a
 similar way than for objects of class \class{table} using
@@
 subsetting is only possible ``outside-in'', that is, indexing operates
-on blocks defined by the variable levels:
+on blocks defined by the variable levels. In the following, we again use the
+Titanic data, this time collapsed over `Survived' to investigate the
+structure of crew and passengers:
 
 <<subsetting,results=verbatim>>=
-hec
-hec["Male",]
-hec["Male", c("Brown","Green")]
+(tit <- structable(~ Sex + Class + Age, data = Titanic))
+tit["Male",]
+tit["Male", c("1st","2nd","3rd")]
 @
@@
 <<conditioning,results=verbatim>>=
-hec[["Male",]]
-hec[[c("Male", "Red"),]]
-hec[["Male","Green"]]
+tit[["Male",]]
+tit[[c("Male", "Adult"),]]
+tit[["Male","1st"]]
 @
@@
 
-<<Variables>>=
+<<Variables1>>=
 pushViewport(viewport(layout = grid.layout(ncol = 2)))
-
+@
+
+<<Variables2>>=
 pushViewport(viewport(layout.pos.col = 1))
-mosaic(hec[["Male"]], margins = c(left = 2.5, top = 2.5, 0), sub = "Male", newpage = FALSE)
+mosaic(tit[["Male"]], margins = c(left = 2.5, top = 2.5, 0), sub = "Male", newpage = FALSE)
 popViewport()
-
+@
+
+<<Variables3>>=
 pushViewport(viewport(layout.pos.col = 2))
-mosaic(hec[["Female"]], margins = c(top = 2.5, 0), sub = "Female", newpage = FALSE)
+mosaic(tit[["Female"]], margins = c(top = 2.5, 0), sub = "Female", newpage = FALSE)
 popViewport(2)
@@
 <<Variablesfig,fig=TRUE,echo=FALSE,height=6,width=12>>=
-<<Variables>>
+<<Variables1>>
+<<Variables2>>
+<<Variables3>>
 @
 \caption{Two mosaic displays put side-by-side, visualizing the 
-distribution of hair and eye color, given gender. The marginal
+distribution of class and age, given gender. The marginal
 distribution of `Sex' cannot be seen.}
@@
 <<Conditioning>>=
-mosaic(~ Hair + Eye | Sex, data = hec, split_vertical = TRUE, keep_aspect_ratio = FALSE)
+mosaic(~ Age + Class | Sex, data = tit, split_vertical = TRUE, keep_aspect_ratio = FALSE)
 @
@@
 \caption{Mosaic plot again visualizing the 
-distribution of hair and eye color, given gender, this time using a
+distribution of class and age, given gender, this time using a
 single mosaic plot. In contrast to Figure \ref{fig:parttable}, this
@@
 <<cotabplot>>=
-cotabplot(~ Hair + Eye | Sex, data = hec, panel_args = list(margins = 3), labeling = labeling_left(clip = FALSE))
+cotabplot(~ Age + Class | Sex, data = tit, panel_args = list(margins = 3), 
+          labeling = labeling_left, tl_labels = c(Class = FALSE))
 @
@@
 @
-\caption{Conditional table plot for the \data{HairEyeColor} data,
-  again visualizing the distribution of hair and eye color, given
+\caption{Conditional table plot for the \data{Titanic} data,
+  again visualizing the distribution of age and class, given
   gender, using separate mosaic displays like the `manual' plot 
@@
 the variable names (optionally, with univariate statistics) in the diagonal cells. 
-Figure \ref{fig:pairs} shows a pairs display with mosaic plots visualizing mutual 
-independence in the lower triangle,
-association plots for the same in the upper triangle, and bar charts in the diagonal.
+Figure \ref{fig:pairs} shows a pairs display for the
+\data{HairEyeColor} data with univariate mosaics in the diagonal,
+and mosaic plots visualizing mutual independence in the upper and lower triangles.
+Due to the inherent asymmetry of mosaic displays, the corresponding plots in the
+upper and lower triangle differ depending on which variable is used first
+for splitting---inspecting both views might help detecting patterns in a data set. 
+Additionally, 
+we are using a special spacing and shading normally used to `highlight' %'
+the second variable in the first (as will be discussed in Section \ref{sec:spacing}): here,
+the intention of the spacing 
+is to emphasize the conditional distributions of the second variable,
+given the first one, and the shading helps identifying the factor levels in
+the second variable.
 
 <<pairs>>=
-pairs(hec, lower_panel = pairs_assoc, space = 0.3, 
-      diag_panel_args = list(rot = -45, just_leveltext = c("left","bottom")))
-@
+pairs(hec, highlighting = 2, diag_panel = pairs_diagonal_mosaic, 
+      diag_panel_args = list(fill = grey.colors))
+@
+
+% auch möglich mit Titanic-Daten:
+% <<pairs>>=
+% pairs(tit, highlighting = 2, diag_panel = pairs_diagonal_mosaic, 
+%       diag_panel_args = list(fill = grey.colors, alternate_labels = FALSE))
+% @
 
@@
 respectively. We will first expose the basic concept of these shading
-functions using the HSV space, and
-then briefly explain the differences to the HCL space \citep[a detailed
+functions using HSV space, and
+then briefly explain the differences to HCL space \citep[a detailed
 discussion can be found in][]{vcd:Zeileis+Meyer+Hornik:2005}. 
-The HCL space is trickier
-to use, but preferable to the HSV space from a perceptual point of view.
-
-In the HSV space, colors are specified in three dimensions: Hue,
+Color palettes in HCL space are trickier
+to choose, but preferable to palettes derived 
+from HSV space from a perceptual point of view.
+
+In HSV space, colors are specified in three dimensions: Hue,
 Saturation (`colorfulness'), and Value (`lightness', amount of gray). 
@@
 \codefun{shading\_hsv} but \codefun{shading\_hcl} due to the better
-perceptual characteristics of the HCL color space.
-Figure \ref{fig:shadingHSVHCL} depicts 
-the HSV space in the upper panel and the HCL space in the lower panel.
-On the left (right) side, we see the color scales for red (blue)
-hue, respectively. The $x$-axis represents the colorfulness, and the
-$y$-axis the brightness.
-The boxes represent the diverging color palettes used for the shadings.
-For the HSV space, we can see that the effect of changing the 
-level of brightness (`value') is not the same for different levels of 
-saturation, and again not the same for the two different hues.
-In fact, in the HSV space all dimensions are confounded, which 
-obviously is problematic for coding information. In contrast, the HCL color
-space offers perceptually uniform colors: as can be seen from the lower panel, 
-the chroma is homogeneous for different levels of luminance. 
-Unfortunately, this comes at the 
-price of the space being irregularly shaped, making it difficult to automatically select 
-diverging color palettes. The following example again 
-illustrates the \data{HairEyeColor} data, this time with HCL colors 
-(Figure \ref{fig:hclshading1} depicts the default palette, 
+perceptual characteristics of HCL color space.
+% Figure \ref{fig:shadingHSVHCL} depicts 
+% HSV space in the upper panel and HCL space in the lower panel.
+% On the left (right) side, we see the color scales for red (blue)
+% hue, respectively. The $x$-axis represents the colorfulness, and the
+% $y$-axis the brightness.
+% The boxes represent the diverging color palettes used for the shadings.
+% For HSV space, we can see that the effect of changing the 
+% level of brightness (`value') is not the same for different levels of 
+% saturation, and again not the same for the two different hues.
+% In fact, in HSV space all dimensions are confounded, which 
+% obviously is problematic for coding information. In contrast, HCL color
+% space offers perceptually uniform colors: as can be seen from the lower panel, 
+% the chroma is homogeneous for different levels of luminance. 
+% Unfortunately, this comes at the 
+% price of the space being irregularly shaped, making it difficult to automatically select 
+% diverging color palettes. 
+The following example again illustrates the \data{HairEyeColor} data, 
+this time with HCL colors (Figure \ref{fig:hclshading1} depicts the default palette, 
 and Figure \ref{fig:hclshading2} an alternative setting):
 
-<<auxiliary_functions,echo=FALSE,results=hide>>=
-hue.slice <- function(hue, grid.n = 101, type = c("HCL", "HSV"), plot = TRUE, fixup = FALSE)
-{
-  type <- match.arg(type)
-  if(type == "HCL") {
-    chroma = seq(0, 100, length = grid.n)
-    luminance = seq(0, 100, length = grid.n)
-    nc <- length(chroma)
-    nl <- length(luminance)
-    color.slice <- outer(chroma, luminance, function(y, x) hcl(hue, x, y, fixup = fixup))
-    xlab <- "chroma"
-    ylab <- "luminance"
-    main <- paste("hue =", round(hue, digits = 0))
-  } else {
-    chroma = seq(0, 1, length = grid.n)
-    luminance = seq(0, 1, length = grid.n)
-    nc <- length(chroma)
-    nl <- length(luminance)
-    color.slice <- outer(chroma, luminance, function(y, x) hsv(hue, x, y))
-    xlab <- "saturation"
-    ylab <- "value"
-    main <- paste("hue =", round(hue, digits = 3))
-  }
-  if(plot) {
-    plot(0.5, 0.5, xlim = range(chroma), ylim = range(luminance), type = "n", axes = FALSE,
-         xlab = xlab, ylab = ylab, yaxs = "i", xaxs = "i", main = main)
-    for(i in 1:(nc-1)) {
-      rect(chroma[i], luminance[-nl], chroma[i] + 100/(nc-1), luminance[-1], border = color.slice[,i+1], col = color.slice[,i+1])
-    }
-    axis(1)
-    axis(2)
-    box()
-  }
-  colnames(color.slice) <- chroma
-  rownames(color.slice) <- luminance
-  attr(color.slice, "type") <- type
-  class(color.slice) <- "slice"
-  invisible(color.slice)
-}
-@
-
-\setkeys{Gin}{width=.8\textwidth}
-\begin{figure}[p]
-\begin{center}
-<<shading_HSV,fig=TRUE,echo=FALSE,height=7,width=8>>=
-## generate colors
-hue23 <- hue.slice(2/3, grid.n = 101, plot = FALSE, type = "HSV")
-hue0 <- hue.slice(0, grid.n = 101, plot = FALSE, type = "HSV")
-saturation <- as.numeric(colnames(hue23))
-value <- as.numeric(rownames(hue23))
-
-## select those with value >= 0.5
-hue23 <- hue23[value >= .5, ]
-hue0 <- hue0[value >= .5, ]
-value <- value[value >= .5]
-nl <- nrow(hue23)
-nc <- ncol(hue23)
-
-## plot 2 slides from HSV space
-plot(0.5, 0.5, xlim = c(-1, 1), ylim = c(0, 1), type = "n", axes = FALSE,
-       xlab = "", ylab = "", yaxs = "i", xaxs = "i", main = "")
-for(i in 1:(nc-1)) {
-  rect(saturation[i], value[-nl], saturation[i] + 1/(nc-1), value[-1], border = hue23[,i+1], col = hue23[,i+1])
-}
-for(i in 1:(nc-1)) {
-  rect(-saturation[i], value[-nl], -(saturation[i] + 1/(nc-1)), value[-1], border = hue0[,i+1], col = hue0[,i+1])
-}
-axis(2, at = c(50, 75, 100)/100, labels = c(0.5, 0.75, 1))
-axis(4, at = c(50, 75, 100)/100, labels = c(0.5, 0.75, 1))
-axis(3, at = -4:4*.25, labels=c(4:0*.25, 1:4*.25))
-mtext(c("hue = 0", "hue = 2/3"), side = 3, at = c(-.5, .5), line = 3, cex = 1.2)
-mtext("saturation", side = 3, at = 0, line = 2)
-mtext("value", side = 2, at = .75, line = 2)
-mtext("value", side = 4, at = .75, line = 2)
-lines(c(-1, 1), c(.5, .5))
-
-## significant colors
-rect(-1, 0.95, -.90, 1, col = hsv(0, 1, 1))
-rect(-0.45, 0.95, -.55, 1, col = hsv(0, 0.5, 1))
-rect(-.05, .95, .05, 1, col = hsv(2/3, 0, 1))
-rect(0.45, 0.95, .55, 1, col = hsv(2/3, 0.5, 1))
-rect(.90, .95, 1, 1, col = hsv(2/3, 1, 1))
-
-text(-1, .33, "significant", pos = 4, cex = 1.2)
-rect(-1, .20, -.80, .30, col = hsv(0, 1, 1))
-rect(-.40, .20, -0.6, .30, col = hsv(0, 0.5, 1))
-rect(-.20, .20, 0, .30, col = hsv(0, 0, 1))
-rect(0, .20, .20, .30, col = hsv(2/3, 0, 1))
-rect(0.4, .20, .60, .30, col = hsv(2/3, .5, 1))
-rect(.80, .20, 1, .30, col = hsv(2/3, 1, 1))
-
-lines(c(-.9, -.55), c(0.975, .975), lty = 2)
-lines(c(-.45, -.05), c(0.975, .975), lty = 2)
-lines(c(.45, .05), c(0.975, .975), lty = 2)
-lines(c(.9, .55), c(0.975, .975), lty = 2)
-
-## non-significant colors
-rect(-1, 0.5, -.90, 0.55, col = hsv(0, 1, 0.5))
-rect(-0.45, 0.5, -.55, 0.55, col = hsv(0, 0.5, 0.5))
-rect(-.05, .5, .05, 0.55, col = hsv(2/3, 0, 0.5))
-rect(0.45, 0.5, .55, 0.55, col = hsv(2/3, 0.5, 0.5))
-rect(.90, .5, 1, 0.55, col = hsv(2/3, 1, 0.5))
-
-text(-1, .13, "non-significant", pos = 4, cex = 1.2)
-rect(-1, 0, -.80, .10, col = hsv(0, 1, 0.5))
-rect(-.60, 0, -.4, .10, col = hsv(0, 0.5, 0.5))
-rect(-.20, 0, 0, .10, col = hsv(0, 0, 0.5))
-rect(0, 0, .20, .10, col = hsv(2/3, 0, 0.5))
-rect(0.4, 0, .60, .1, col = hsv(2/3, .5, 0.5))
-rect(.80, 0, 1, .10, col = hsv(2/3, 1, 0.5))
-
-lines(c(-.9, -.55), c(0.525, .525), lty = 2)
-lines(c(-.45, -.05), c(0.525, .525), lty = 2)
-lines(c(.45, .05), c(0.525, .525), lty = 2)
-lines(c(.9, .55), c(0.525, .525), lty = 2)
-@
-
-<<shading_HCL,fig=TRUE,echo=FALSE,height=7,width=8>>=
-## generate colors
-hue260 <- hue.slice(260, grid.n = 101, plot = FALSE)
-hue360 <- hue.slice(360, grid.n = 101, plot = FALSE)
-mychroma <- as.numeric(colnames(hue260))
-luminance <- as.numeric(rownames(hue260))
-
-## select those with lumincance >= 50
-hue260 <- hue260[luminance >= 50, ]
-hue360 <- hue360[luminance >= 50, ]
-luminance <- luminance[luminance >= 50]
-nc <- ncol(hue260)
-nl <- nrow(hue260)
-
-## plot 2 slides from HCL space
-plot(0.5, 0.5, xlim = c(-100, 100), ylim = c(0, 100), type = "n", axes = FALSE,
-       xlab = "", ylab = "", yaxs = "i", xaxs = "i", main = "")
-for(i in 1:(nc-1)) {
-  rect(mychroma[i], luminance[-nl], mychroma[i] + 100/(nc-1), luminance[-1], border = hue260[,i+1], col = hue260[,i+1])
-}
-for(i in 1:(nc-1)) {
-  rect(-mychroma[i], luminance[-nl], -(mychroma[i] + 100/(nc-1)), luminance[-1], border = hue360[,i+1], col = hue360[,i+1])
-}
-axis(2, at = c(50, 70, 90, 100), labels = c(50, 70, 90, 100))
-axis(4, at = c(50, 70, 90, 100), labels = c(50, 70, 90, 100))
-axis(3, at = -4:4*25, labels=c(4:0*25, 1:4*25))
-mtext(c("hue = 0", "hue = 260"), side = 3, at = c(-50, 50), line = 3, cex = 1.2)
-mtext("chroma", side = 3, at = 0, line = 2)
-mtext("luminance", side = 2, at = 75, line = 2)
-mtext("luminance", side = 4, at = 75, line = 2)
-lines(c(-100, 100), c(50, 50))
-
-## significant colors
-rect(-100, 47.5, -90, 52.5, col = hcl(0, 100, 50))
-rect(-55, 67.5, -45, 72.5, col = hcl(0, 50, 70))
-rect(-5, 95, 5, 100, col = hcl(260, 0, 100))       ## grey vs. white
-rect(-5, 87.5, 5, 92.5, col = hcl(260, 0, 90))     ## grey vs. white
-rect(45, 67.5, 55, 72.5, col = hcl(260, 50, 70))
-rect(90, 47.5, 100, 52.5, col = hcl(260, 100, 50))
-
-text(-100, 33, "significant", pos = 4, cex = 1.2)
-rect(-100, 20, -80, 30, col = hcl(0, 100, 50))
-rect(-60, 20, -40, 30, col = hcl(0, 50, 70))
-rect(-20, 20, 0, 30, col = hcl(0, 0, 90))       
-rect(0, 20, 20, 30, col = hcl(260, 0, 90))
-#white# rect(-20, 20, 0, 30, col = hcl(0, 0, 100))
-#white# rect(0, 20, 20, 30, col = hcl(260, 0, 100))
-rect(40, 20, 60, 30, col = hcl(260, 50, 70))
-rect(80, 20, 100, 30, col = hcl(260, 100, 50))
-
-lines(c(-45, -5), c(72.5, 87.5), lty = 2)
-lines(c(45, 5), c(72.5, 87.5), lty = 2)
-lines(c(-95, -55), c(52.5, 67.5), lty = 2)
-lines(c(95, 55), c(52.5, 67.5), lty = 2)
-
-## non-significant colors
-rect(-25, 47.5, -15, 52.5, col = hcl(0, 20, 50))
-rect(-15, 67.5, -5, 72.5, col = hcl(0, 10, 70))
-rect(5, 67.5, 15, 72.5, col = hcl(260, 10, 70))
-rect(25, 47.5, 15, 52.5, col = hcl(260, 20, 50))
-
-text(-100, 13, "non-significant", pos = 4, cex = 1.2)
-rect(-60, 0, -40, 10, col = hcl(0, 20, 50))
-rect(-40, 0, -20, 10, col = hcl(0, 10, 70))
-rect(-20, 0, 0, 10, col = hcl(0, 0, 90))
-rect(0, 0, 20, 10, col = hcl(260, 0, 90))
-rect(20, 0, 40, 10, col = hcl(260, 10, 70))
-rect(40, 0, 60, 10, col = hcl(260, 20, 50))
-
-lines(c(-18.75, -11.25), c(52.5, 67.5), lty = 2)
-lines(c(-8.75, -1.25), c(72.5, 87.5), lty = 2)
-lines(c(18.75, 11.75), c(52.5, 67.5), lty = 2)
-lines(c(8.75, 1.25), c(72.5, 87.5), lty = 2)
-@ 
-\caption{Residual-based shadings in HSV (upper) and HCL space (lower).}
-\label{fig:shadingHSVHCL}
-\end{center}
-\end{figure}
+% <<auxiliary_functions,echo=FALSE,results=hide>>=
+% hue.slice <- function(hue, grid.n = 101, type = c("HCL", "HSV"), plot = TRUE, fixup = FALSE)
+% {
+%   type <- match.arg(type)
+%   if(type == "HCL") {
+%     chroma = seq(0, 100, length = grid.n)
+%     luminance = seq(0, 100, length = grid.n)
+%     nc <- length(chroma)
+%     nl <- length(luminance)
+%     color.slice <- outer(chroma, luminance, function(y, x) hcl(hue, x, y, fixup = fixup))
+%     xlab <- "chroma"
+%     ylab <- "luminance"
+%     main <- paste("hue =", round(hue, digits = 0))
+%   } else {
+%     chroma = seq(0, 1, length = grid.n)
+%     luminance = seq(0, 1, length = grid.n)
+%     nc <- length(chroma)
+%     nl <- length(luminance)
+%     color.slice <- outer(chroma, luminance, function(y, x) hsv(hue, x, y))
+%     xlab <- "saturation"
+%     ylab <- "value"
+%     main <- paste("hue =", round(hue, digits = 3))
+%   }
+%   if(plot) {
+%     plot(0.5, 0.5, xlim = range(chroma), ylim = range(luminance), type = "n", axes = FALSE,
+%          xlab = xlab, ylab = ylab, yaxs = "i", xaxs = "i", main = main)
+%     for(i in 1:(nc-1)) {
+%       rect(chroma[i], luminance[-nl], chroma[i] + 100/(nc-1), luminance[-1], border = color.slice[,i+1], col = color.slice[,i+1])
+%     }
+%     axis(1)
+%     axis(2)
+%     box()
+%   }
+%   colnames(color.slice) <- chroma
+%   rownames(color.slice) <- luminance
+%   attr(color.slice, "type") <- type
+%   class(color.slice) <- "slice"
+%   invisible(color.slice)
+% }
+% @
+
+% \setkeys{Gin}{width=.8\textwidth}
+% \begin{figure}[p]
+% \begin{center}
+% <<shading_HSV,fig=TRUE,echo=FALSE,height=7,width=8>>=
+% ## generate colors
+% hue23 <- hue.slice(2/3, grid.n = 101, plot = FALSE, type = "HSV")
+% hue0 <- hue.slice(0, grid.n = 101, plot = FALSE, type = "HSV")
+% saturation <- as.numeric(colnames(hue23))
+% value <- as.numeric(rownames(hue23))
+
+% ## select those with value >= 0.5
+% hue23 <- hue23[value >= .5, ]
+% hue0 <- hue0[value >= .5, ]
+% value <- value[value >= .5]
+% nl <- nrow(hue23)
+% nc <- ncol(hue23)
+
+% ## plot 2 slides from HSV space
+% plot(0.5, 0.5, xlim = c(-1, 1), ylim = c(0, 1), type = "n", axes = FALSE,
+%        xlab = "", ylab = "", yaxs = "i", xaxs = "i", main = "")
+% for(i in 1:(nc-1)) {
+%   rect(saturation[i], value[-nl], saturation[i] + 1/(nc-1), value[-1], border = hue23[,i+1], col = hue23[,i+1])
+% }
+% for(i in 1:(nc-1)) {
+%   rect(-saturation[i], value[-nl], -(saturation[i] + 1/(nc-1)), value[-1], border = hue0[,i+1], col = hue0[,i+1])
+% }
+% axis(2, at = c(50, 75, 100)/100, labels = c(0.5, 0.75, 1))
+% axis(4, at = c(50, 75, 100)/100, labels = c(0.5, 0.75, 1))
+% axis(3, at = -4:4*.25, labels=c(4:0*.25, 1:4*.25))
+% mtext(c("hue = 0", "hue = 2/3"), side = 3, at = c(-.5, .5), line = 3, cex = 1.2)
+% mtext("saturation", side = 3, at = 0, line = 2)
+% mtext("value", side = 2, at = .75, line = 2)
+% mtext("value", side = 4, at = .75, line = 2)
+% lines(c(-1, 1), c(.5, .5))
+
+% ## significant colors
+% rect(-1, 0.95, -.90, 1, col = hsv(0, 1, 1))
+% rect(-0.45, 0.95, -.55, 1, col = hsv(0, 0.5, 1))
+% rect(-.05, .95, .05, 1, col = hsv(2/3, 0, 1))
+% rect(0.45, 0.95, .55, 1, col = hsv(2/3, 0.5, 1))
+% rect(.90, .95, 1, 1, col = hsv(2/3, 1, 1))
+
+% text(-1, .33, "significant", pos = 4, cex = 1.2)
+% rect(-1, .20, -.80, .30, col = hsv(0, 1, 1))
+% rect(-.40, .20, -0.6, .30, col = hsv(0, 0.5, 1))
+% rect(-.20, .20, 0, .30, col = hsv(0, 0, 1))
+% rect(0, .20, .20, .30, col = hsv(2/3, 0, 1))
+% rect(0.4, .20, .60, .30, col = hsv(2/3, .5, 1))
+% rect(.80, .20, 1, .30, col = hsv(2/3, 1, 1))
+
+% lines(c(-.9, -.55), c(0.975, .975), lty = 2)
+% lines(c(-.45, -.05), c(0.975, .975), lty = 2)
+% lines(c(.45, .05), c(0.975, .975), lty = 2)
+% lines(c(.9, .55), c(0.975, .975), lty = 2)
+
+% ## non-significant colors
+% rect(-1, 0.5, -.90, 0.55, col = hsv(0, 1, 0.5))
+% rect(-0.45, 0.5, -.55, 0.55, col = hsv(0, 0.5, 0.5))
+% rect(-.05, .5, .05, 0.55, col = hsv(2/3, 0, 0.5))
+% rect(0.45, 0.5, .55, 0.55, col = hsv(2/3, 0.5, 0.5))
+% rect(.90, .5, 1, 0.55, col = hsv(2/3, 1, 0.5))
+
+% text(-1, .13, "non-significant", pos = 4, cex = 1.2)
+% rect(-1, 0, -.80, .10, col = hsv(0, 1, 0.5))
+% rect(-.60, 0, -.4, .10, col = hsv(0, 0.5, 0.5))
+% rect(-.20, 0, 0, .10, col = hsv(0, 0, 0.5))
+% rect(0, 0, .20, .10, col = hsv(2/3, 0, 0.5))
+% rect(0.4, 0, .60, .1, col = hsv(2/3, .5, 0.5))
+% rect(.80, 0, 1, .10, col = hsv(2/3, 1, 0.5))
+
+% lines(c(-.9, -.55), c(0.525, .525), lty = 2)
+% lines(c(-.45, -.05), c(0.525, .525), lty = 2)
+% lines(c(.45, .05), c(0.525, .525), lty = 2)
+% lines(c(.9, .55), c(0.525, .525), lty = 2)
+% @
+
+% <<shading_HCL,fig=TRUE,echo=FALSE,height=7,width=8>>=
+% ## generate colors
+% hue260 <- hue.slice(260, grid.n = 101, plot = FALSE)
+% hue360 <- hue.slice(360, grid.n = 101, plot = FALSE)
+% mychroma <- as.numeric(colnames(hue260))
+% luminance <- as.numeric(rownames(hue260))
+
+% ## select those with lumincance >= 50
+% hue260 <- hue260[luminance >= 50, ]
+% hue360 <- hue360[luminance >= 50, ]
+% luminance <- luminance[luminance >= 50]
+% nc <- ncol(hue260)
+% nl <- nrow(hue260)
+
+% ## plot 2 slides from HCL space
+% plot(0.5, 0.5, xlim = c(-100, 100), ylim = c(0, 100), type = "n", axes = FALSE,
+%        xlab = "", ylab = "", yaxs = "i", xaxs = "i", main = "")
+% for(i in 1:(nc-1)) {
+%   rect(mychroma[i], luminance[-nl], mychroma[i] + 100/(nc-1), luminance[-1], border = hue260[,i+1], col = hue260[,i+1])
+% }
+% for(i in 1:(nc-1)) {
+%   rect(-mychroma[i], luminance[-nl], -(mychroma[i] + 100/(nc-1)), luminance[-1], border = hue360[,i+1], col = hue360[,i+1])
+% }
+% axis(2, at = c(50, 70, 90, 100), labels = c(50, 70, 90, 100))
+% axis(4, at = c(50, 70, 90, 100), labels = c(50, 70, 90, 100))
+% axis(3, at = -4:4*25, labels=c(4:0*25, 1:4*25))
+% mtext(c("hue = 0", "hue = 260"), side = 3, at = c(-50, 50), line = 3, cex = 1.2)
+% mtext("chroma", side = 3, at = 0, line = 2)
+% mtext("luminance", side = 2, at = 75, line = 2)
+% mtext("luminance", side = 4, at = 75, line = 2)
+% lines(c(-100, 100), c(50, 50))
+
+% ## significant colors
+% rect(-100, 47.5, -90, 52.5, col = hcl(0, 100, 50))
+% rect(-55, 67.5, -45, 72.5, col = hcl(0, 50, 70))
+% rect(-5, 95, 5, 100, col = hcl(260, 0, 100))       ## grey vs. white
+% rect(-5, 87.5, 5, 92.5, col = hcl(260, 0, 90))     ## grey vs. white
+% rect(45, 67.5, 55, 72.5, col = hcl(260, 50, 70))
+% rect(90, 47.5, 100, 52.5, col = hcl(260, 100, 50))
+
+% text(-100, 33, "significant", pos = 4, cex = 1.2)
+% rect(-100, 20, -80, 30, col = hcl(0, 100, 50))
+% rect(-60, 20, -40, 30, col = hcl(0, 50, 70))
+% rect(-20, 20, 0, 30, col = hcl(0, 0, 90))       
+% rect(0, 20, 20, 30, col = hcl(260, 0, 90))
+% #white# rect(-20, 20, 0, 30, col = hcl(0, 0, 100))
+% #white# rect(0, 20, 20, 30, col = hcl(260, 0, 100))
+% rect(40, 20, 60, 30, col = hcl(260, 50, 70))
+% rect(80, 20, 100, 30, col = hcl(260, 100, 50))
+
+% lines(c(-45, -5), c(72.5, 87.5), lty = 2)
+% lines(c(45, 5), c(72.5, 87.5), lty = 2)
+% lines(c(-95, -55), c(52.5, 67.5), lty = 2)
+% lines(c(95, 55), c(52.5, 67.5), lty = 2)
+
+% ## non-significant colors
+% rect(-25, 47.5, -15, 52.5, col = hcl(0, 20, 50))
+% rect(-15, 67.5, -5, 72.5, col = hcl(0, 10, 70))
+% rect(5, 67.5, 15, 72.5, col = hcl(260, 10, 70))
+% rect(25, 47.5, 15, 52.5, col = hcl(260, 20, 50))
+
+% text(-100, 13, "non-significant", pos = 4, cex = 1.2)
+% rect(-60, 0, -40, 10, col = hcl(0, 20, 50))
+% rect(-40, 0, -20, 10, col = hcl(0, 10, 70))
+% rect(-20, 0, 0, 10, col = hcl(0, 0, 90))
+% rect(0, 0, 20, 10, col = hcl(260, 0, 90))
+% rect(20, 0, 40, 10, col = hcl(260, 10, 70))
+% rect(40, 0, 60, 10, col = hcl(260, 20, 50))
+
+% lines(c(-18.75, -11.25), c(52.5, 67.5), lty = 2)
+% lines(c(-8.75, -1.25), c(72.5, 87.5), lty = 2)
+% lines(c(18.75, 11.75), c(52.5, 67.5), lty = 2)
+% lines(c(8.75, 1.25), c(72.5, 87.5), lty = 2)
+% @ 
+% \caption{Residual-based shadings in HSV (upper) and HCL space (lower).}
+% \label{fig:shadingHSVHCL}
+% \end{center}
+% \end{figure}
 
@@
 <<arthritis>>=
+set.seed(4711)
 mosaic(~ Treatment + Improved, data = Arthritis, subset = Sex == "Female",

### 2006-09-22 trunk/vcd/paper/Strucplot/JSS.Rnw
Change default of alternate_labeling in pairs_diagonal_mosaic
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@574 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 <<subsetting,results=verbatim>>=
-(tit <- structable(~ Sex + Class + Age, data = Titanic))
+(tit <- structable(~ Sex + Class + Age, data = Titanic[,,2:1,]))
 tit["Male",]
@@
 
-Since mosaic displays are ``conditional plots'' by definition, we can
-also use one single mosaic for stratified plots.
-The formula interface of \codefun{mosaic} allows the
-specification of conditioning variables (see Figure \ref{fig:conditioning}):
-
-<<Conditioning>>=
-mosaic(~ Age + Class | Sex, data = tit, split_vertical = TRUE, keep_aspect_ratio = FALSE)
+The \codefun{cotabplot} function does a much better job on this task:
+it arranges stratified strucplot displays 
+in a lattice-like layout, conditioning on variable \emph{levels}.
+The plot in Figure \ref{fig:cotabplot} shows hair and eye color, given sex:
+
+<<cotabplot>>=
+cotabplot(~ Class + Age | Sex, data = tit, split_vertical = TRUE)
 @
@@
 \begin{center}
+<<cotabplotfig,fig=TRUE,echo=FALSE,height=6,width=11>>=
+<<cotabplot>>
+@
+\caption{Conditional table plot for the \data{Titanic} data,
+  again visualizing the distribution of age and class, given
+  gender, using separate mosaic displays like the `manual' plot 
+  in Figure \ref{fig:parttable}.}
+\label{fig:cotabplot}
+\end{center}
+\end{figure}
+\setkeys{Gin}{width=0.7\textwidth}
+%\noindent The \code{labeling\_args} argument modifies the labels'
+%appearance: here, to be left-aligned and unclipped
+%(see Section \ref{sec:labeling}).
+
+\noindent Visualizing the strata separately `hides' the distribution of the
+conditioning variable(s) which may or may not be appropriated or
+sensible in a particular analysis step. If we wish to keep the
+information on the marginal distribution(s), we can use one single
+mosaic for the stratified plot since mosaic displays are 
+``conditional plots'' by definition. We just need to make sure that
+conditioning variables are used first for splitting. 
+Both the default and the formula interface of \codefun{mosaic} allow the
+specification of conditioning variables (see Figure \ref{fig:conditioning}):
+
+<<Conditioning1>>=
+mosaic(tit, conditional = "Sex", split_vertical = c(TRUE, TRUE, FALSE), 
+       keep_aspect_ratio = FALSE)
+@ 
+<<Conditioning2>>=
+mosaic(~ Class + Age | Sex, data = tit, split_vertical = c(TRUE, TRUE, FALSE), 
+       keep_aspect_ratio = FALSE)
+@
+
+\setkeys{Gin}{width=\textwidth}
+\begin{figure}[h]
+\begin{center}
 <<Conditioningfig,fig=TRUE,echo=FALSE,width=12>>=
-<<Conditioning>>
+<<Conditioning2>>
 @
@@
 
-\noindent The effect of using this kind of formula is that
+\noindent The effect of using this is that
 conditioning variables are permuted ahead of the the conditioned variables
-in the table, and that \codefun{spacing\_conditional} is used as
-default to better
+in the table, and for the formula interface, 
+that \codefun{spacing\_conditional} is used as default to better
 distinguish conditioning from conditioned dimensions.
@@
 addition, we release the fixed aspect ratio to get less distorted margins.
-
-The \codefun{cotabplot} function does a much better job on this task:
-it arranges stratified strucplot displays 
-in a lattice-like layout, conditioning on variable \emph{levels}.
-The plot in Figure \ref{fig:cotabplot} shows hair and eye color, given sex:
-
-<<cotabplot>>=
-cotabplot(~ Age + Class | Sex, data = tit, panel_args = list(margins = 3), 
-          labeling = labeling_left, tl_labels = c(Class = FALSE))
-@
-
-\setkeys{Gin}{width=\textwidth}
-\begin{figure}[h]
-\begin{center}
-<<cotabplotfig,fig=TRUE,echo=FALSE,height=6,width=11>>=
-<<cotabplot>>
-@
-\caption{Conditional table plot for the \data{Titanic} data,
-  again visualizing the distribution of age and class, given
-  gender, using separate mosaic displays like the `manual' plot 
-  in Figure \ref{fig:parttable}.}
-\label{fig:cotabplot}
-\end{center}
-\end{figure}
-\setkeys{Gin}{width=0.7\textwidth}
-
-\noindent The \code{labeling\_args} argument modifies the labels'
-appearance: here, to be left-aligned and unclipped
-(see Section \ref{sec:labeling}).
 
@@
 Figure \ref{fig:pairs} shows a pairs display for the
-\data{HairEyeColor} data with univariate mosaics in the diagonal,
+\data{Titanic} data with univariate mosaics in the diagonal,
 and mosaic plots visualizing mutual independence in the upper and lower triangles.
@@
 <<pairs>>=
-pairs(hec, highlighting = 2, diag_panel = pairs_diagonal_mosaic, 
-      diag_panel_args = list(fill = grey.colors))
-@
-
-% auch möglich mit Titanic-Daten:
-% <<pairs>>=
-% pairs(tit, highlighting = 2, diag_panel = pairs_diagonal_mosaic, 
-%       diag_panel_args = list(fill = grey.colors, alternate_labels = FALSE))
-% @
+pairs(tit, highlighting = 2, diag_panel = pairs_diagonal_mosaic, 
+      diag_panel_args = list(fill = grey.colors, alternate_labels = FALSE))
+@
 
@@
 @
-\caption{Pairs plot for the \data{HairEyeColor} data.}
+\caption{Pairs plot for the \data{Titanic} data.}
 \label{fig:pairs}
@@
 \begin{center}
-\code{cell:\emph{Variable1}=\emph{Level1},\emph{Variable2}=\emph{Level2}} \dots
+\code{[Optional prefix]cell:\emph{Variable1}=\emph{Level1},\emph{Variable2}=\emph{Level2}} \dots
 \end{center}
@@
 \end{figure}
+
+\subsection{Performance Issues}
+\label{sec:performance}
+
+As stated above, the implementation of strucplot displays is based on
+creating and nesting \pkg{grid} viewports. The main time-consuming steps performed by
+the core functions are the follwing:
+
+\begin{enumerate}
+\item recursively, create a tree of nested viewports
+\item push the whole tree, i.e., add it to the system tree maintained
+  by the \pkg{grid} engine
+\item one by one, seek all cells of the strucplot display and add
+  plot-specific content (rectangles for mosaics, bars for association
+  plots, etc.)
+\end{enumerate}
+
+\noindent All these operations scale linearly with the amount of
+created viewports. For a $d$-dimensional table with $k_i$
+levels, $i=1 \dots d$, the total number of needed viewports
+$T_d$ can roughly be estimated as:
+
+\begin{equation}
+  \label{eq:numbervp}
+  T_d = \sum_{i=1}_d \prod_{j \le i} k_j
+\end{equation}
+
+\noindent Assuming an equal number for all dimensions, we obtain:
+
+\begin{equation}
+  \label{eq:equalvp}
+  T_d = \sum_{i=1}_d k^i = \frac{k(k^d-1)}{k-1}
+\end{equation}
+
+\noindent and so the time complexity for drawing a strucplot display is of order $O(k^d)$.
 

### 2006-09-22 trunk/vcd/R/pairsplot.R
Change default of alternate_labeling in pairs_diagonal_mosaic
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@574 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -202,7 +202,6 @@ class(pairs_barplot) <- "grapcon_generator"
 
 pairs_diagonal_mosaic <- function(split_vertical = TRUE,
                                   margins = unit(0, "lines"),
-                                  alternate_labels = TRUE,
                                   offset_labels = -0.4,
                                   offset_varnames = 0,
                                   gp = NULL,
@@ -217,7 +216,6 @@ pairs_diagonal_mosaic <- function(split_vertical = TRUE,
            newpage = FALSE,
            split_vertical = split_vertical,
            margins = margins,
-           alternate_labels = alternate_labels,
            offset_labels = offset_labels,
            offset_varnames = offset_varnames,
            prefix = "diag",

### 2006-09-23 trunk/vcd/paper/Strucplot/JSS.Rnw
several fixes in JSS paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@575 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 Sections \ref{sec:labeling} and \ref{sec:spacing} discuss in detail the labeling
-and spacing features, respectively, before Section \ref{sec:conclusion}
-concludes the work.
+and spacing features, respectively. Section \ref{sec:example}
+exemplifies the framework in the analysis of a four-way data set, 
+before Section \ref{sec:conclusion} concludes the work.
 
@@
 
-<<structable,results=verbatim>>=
+<<structable,echo=TRUE,results=verbatim>>=
 (hec <- structable(Eye ~ Sex + Hair, data = HairEyeColor))
@@
 levels, $i=1 \dots d$, the total number of needed viewports
-$T_d$ can roughly be estimated as:
+$T_d$ can roughly be estimated as
 
@@
   \label{eq:numbervp}
-  T_d = \sum_{i=1}_d \prod_{j \le i} k_j
+  T_d = k_1 + k_1k_2 + \dots + k_1 \dots k_d = \sum_{i=1}^d \prod_{j \le i} k_j
 \end{equation}
 
-\noindent Assuming an equal number for all dimensions, we obtain:
+\noindent since we first push the $k_1$ viewports for the levels of the
+first dimension, then, for \emph{each} of those, the $k_2$ levels of
+the second dimension, etc. Assuming an equal number of levels $k$
+for all dimensions, we obtain:
 
@@
   \label{eq:equalvp}
-  T_d = \sum_{i=1}_d k^i = \frac{k(k^d-1)}{k-1}
+  T_d = \sum_{i=1}^d k^i = \frac{k(k^d-1)}{k-1}
 \end{equation}
 
-\noindent and so the time complexity for drawing a strucplot display is of order $O(k^d)$.
+\noindent and so the time complexity for drawing a strucplot display is of order $k^d$.
 
@@
 
+\section{Example: the Ovarian Cancer Data}
+\label{sec:example}
+
+In the following, we demonstrate some of the described techniques in
+analyzing the Ovarian Cancer data set.
+The data origins from \citep{vcd:obel:1975} and is 
+about a retrospective study of ovary cancer
+carried out in 1973.  Information was obtained from 299 women, who
+were operated for ovary cancer 10 years before. The data consists of
+four binary variables: the `stage' of the cancer at the time of
+operation (levels: early, advanced), the type of `operation' performed
+(radical, limited), the `survival' status after 10 years (yes, no),
+and `xray' indicating whether X-ray treatment was received (yes,
+no).
+
+The dataset in \pkg{vcd} comes pretabulated in a data frame, so we
+first create the four-way table:
+
+<<oc1>>=
+tab <- xtabs(Freq ~ stage + operation + xray + survival, data = OvaryCancer)
+@
+
+\noindent A `flattened' textual representation can be obtained using \codefun{structable}:
+
+<<oc2,echo=TRUE,results=verbatim>>=
+structable(survival ~ ., data = tab)
+@
+
+\noindent A first overview can be obtained using a pairs plot (Figure \ref{fig:ocpairs}):
+
+<<oc3>>=
+dpa <- list(var_offset = 1.2, rot = -30, just_leveltext= "left")
+pairs(tab, diag_panel_args = dpa)
+@
+
+\begin{figure}[h]
+\begin{center}
+<<ocpairs,echo=FALSE,fig=TRUE>>=
+<<oc3>>
+@ 
+\caption{Pairs plot for the \data{OvaryCancer} data showing mosaic displays for all pairwise distributions and bar plots for all marginal distributions.}
+\label{fig:ocpairs}
+\end{center}
+\end{figure}
+
+\noindent The pairs plot, by default, creates mosaic displays for all pairwise 
+variable combinations, and bar plots in the diagonal to visualize the absolute 
+frequencies of the variables. The \texttt{var\_offset} argument modifies the offset of the 
+(centered) variable names to avoid overlap with the bars. 
+Additionally, we use the \texttt{rot} and the
+\texttt{just\_leveltext} arguments to rotate the level names, again to avoid their overlap.
+First, we consider the marginal distributions. The study design involved 
+(nearly) the same number of survived (150) and deceased (149) patients. Similarly balanced,
+158 cases were in an advanced and 141 in an early stage. Most patients (251, 84\%) 
+were treated with a radical operation, and 186 (62\%) got X-ray treatment. Next, we inspect
+the two-way interaction of the influencing factors (`stage', `operation', and `xray'): the
+corresponding mosaics exhibit symmetric, regular shapes with aligned tiles, 
+which indicate no marginal interaction between these variables. 
+The same is true for the interactions of `survival'
+with `operation' and `xray', respectively. Only the stage seems to influence
+survival: here, the tiles are ``shifted''.
+
+A different view on the data can be obtained using a doubledecker plot, explaining
+survival by the three other factors (Figure \ref{fig:ocdoubledecker}):
+
+<<oc4>>=
+doubledecker(survival ~ stage + operation + xray, data = tab)
+@
+
+\begin{figure}[h]
+\begin{center}
+<<ocdoubledecker,echo=FALSE,fig=TRUE>>=
+<<oc4>>
+@ 
+\caption{Doubledecker plot for the \data{OvaryCancer} data showing the
+ conditional distribution of `xray', given `operation', given `stage', and with
+ `survival' highlighted.}
+\label{fig:ocdoubledecker}
+\end{center}
+\end{figure}
+
+\noindent From a technical point of view, the display is constructed as a mosaic plot
+showing the conditional distribution of `survival', given `xray', given `operation', 
+given `stage', with vertical splits for the conditioning variables and horizontal ones for
+`survival'. Additionally, there is zero space between the tiles of the last dimension and
+a binary shading is used for survived and deceased patients.
+Conceptually, this plot is interpreted as a mosaic plot of just the influencing
+variables, with `survival' highlighted in the tiles. Thus, the plot really shows
+the influence of the ``explanatory'' variables on `survival'. Clearly, the survival
+rate is higher among patients in an early stage, but neither radical operation 
+nor X-ray treatment seem to improve the situation. From this exploratory
+phase, it even seems that for patients in an
+early stage with a limited operation, no X-ray treatment would be better, 
+and that for patients in an advanced stage, limited in contrast to radical 
+operation would improve the chance of survival. %'
+
+To visualize inference results, we can make use of residual-based shadings. First, we
+again use the pairs plot display, this time using color to bring out departures from
+the hypothesized marginal independence models (Figure \ref{fig:ocpairsshaded}):
+
+<<oc5>>=
+pairs(tab, diag_panel_args = dpa, gp = shading_max)
+@
+
+\begin{figure}[h]
+\begin{center}
+<<ocpairsshaded,echo=FALSE,fig=TRUE>>=
+<<oc5>>
+@ 
+\caption{Pairs plot for the \data{OvaryCancer} data showing mosaic displays for all pairwise distributions and bar plots for all marginal distributions. The colored cells show the
+violation of the independence hypothesis for `stage' and `survival'.}
+\label{fig:ocpairsshaded}
+\end{center}
+\end{figure}
+
+\noindent Only the mosaics showing the interaction of `stage' and `survival' have 
+colored cells, indicating that the survival rate of patients in an early stage is
+higher than would be expected under (marginal) independence. The hypotheses of
+independence are not rejected for all other marginal interactions. Next, we investigate
+log-linear models for the four-way table. The default model used by \codefun{mosaic} is
+the total independence model, visualized in Figure \ref{fig:ocmosaictotal}:
+
+<<oc6>>=
+mosaic(tab, shade = TRUE)
+@
+
+\begin{figure}[h]
+\begin{center}
+<<ocmosaictotal,echo=FALSE,fig=TRUE>>=
+<<oc6>>
+@ 
+\caption{Mosaic plot for the \data{OvaryCancer} data, with residual-based shading
+for the (clearly rejected) default hypothesis of total independence of all variables.}
+\label{fig:ocmosaictotal}
+\end{center}
+\end{figure}
+
+\noindent The model is clearly rejected (P-value: 0.000). From the exploratory phase of
+our analysis, we (only) suspect `stage' to be influential on the survival rate. A 
+corresponding hypothesis is that `survival' be independent of `xray' and `operation', given
+`stage'. The model is specified using the \texttt{expected} argument, 
+either using the \codefun{loglin} interface or 
+the \codefun{loglm} formula interface (the resulting mosaic plot is shown in Figure
+\ref{fig:ocmosaicstage}):
+
+<<oc7>>=
+mosaic(tab, expected = ~ (survival + operation * xray) * stage)
+@
+
+\begin{figure}[h]
+\begin{center}
+<<ocmosaicstage,echo=FALSE,fig=TRUE>>=
+<<oc7>>
+@ 
+\caption{Mosaic plot for the \data{OvaryCancer} data, with residual-based shading
+for the hypothesis of `survival' being independent of `xray' and `operation', given
+`stage'. The hypothesis is not rejected.}
+\label{fig:ocmosaicstage}
+\end{center}
+\end{figure}
+
+\noindent Thus, based on this data, only pre-diagnosis seems to matter in ovarian
+cancer therapy.
+
 \section{Conclusion}
@@
   Package: \pkg{datasets} (included in base \proglang{R}).
+\item[\texttt{OvaryCancer}]  Data about a retrospective study of ovary cancer
+     carried out in 1973.  Information was obtained from 299 women, who
+     were operated for ovary cancer 10 years before. Source:
+     \cite{vcd:obel:1975}. Package: \pkg{vcd}.
 \item[\texttt{PreSex}]  Data from \cite{vcd:thornes+collard:1979} reported in 
@@
 
-\item Mosaic plot for the \data{Arthiritis} data in Figure \ref{fig:arthritis}:
+\item Mosaic plot for the \data{Arthritis} data in Figure \ref{fig:arthritis}:
 \begin{Sinput}

### 2006-09-23 trunk/vcd/R/doubledeckerplot.R
several fixes in JSS paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@575 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -10,6 +10,9 @@ function(formula, data = NULL, ..., main = NULL)
     if (is.logical(main) && main)
       main <- deparse(substitute(data))
     
+    if (is.structable(data))
+      data <- as.table(data)
+
     m <- match.call(expand.dots = FALSE)
     edata <- eval(m$data, parent.frame())
     
@@ -34,6 +37,8 @@ function(formula, data = NULL, ..., main = NULL)
           dat <- margin.table(dat, ind)
         } else {
           ind <- match(dep, names(dimnames(dat)))
+          if (is.na(ind))
+            stop(paste("Can't find", dep, "in", deparse(substitute(data))))
           dat <- aperm(dat, c(seq(along = dim(dat))[-ind], ind))
         }
         doubledecker.default(dat, main = main, ...)
@@ -58,6 +63,7 @@ doubledecker.default <- function(x,
                          main = NULL, 
                          keep_aspect_ratio = FALSE,
                          ...) {
+  x <- as.table(x)
   d <- dim(x)
   l <- length(d)
   if (is.character(depvar))

### 2006-09-23 trunk/vcd/R/pairsplot.R

@@ -170,8 +170,11 @@ pairs_barplot <- function(gp_bars = NULL,
                           rot = 0, abbreviate = FALSE,
                           check_overlap = TRUE,
                           fill = "grey",
+                          var_offset = unit(1, "npc"),
                           ...)
   function(x, i) {
+    if (!is.unit(var_offset))
+      var_offset <- unit(var_offset, "npc")
     dn <- names(dimnames(x))
     x <- margin.table(x, i)
     if (is.function(fill)) fill <- rev(fill(dim(x)))
@@ -195,7 +198,7 @@ pairs_barplot <- function(gp_bars = NULL,
               x = xpos - halfstep, just = just_leveltext, gp = gp_leveltext,
               check.overlap = check_overlap)
     popViewport(1)
-    grid.text(names(dimnames(x)), y = 1, just = just_vartext, gp = gp_vartext)
+    grid.text(names(dimnames(x)), y = var_offset, just = just_vartext, gp = gp_vartext)
 
   }
 class(pairs_barplot) <- "grapcon_generator"

### 2006-09-29 trunk/vcd/paper/Color/draft.Rnw
minor changes
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@577 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
+%%FIXME: what to say about ColorBrewer
+
 \section{Color spaces} \label{sec:spaces}
@@
 Due to these three color axes, colors are typically described as
-locations in a 3-dimensional spaces. However, human perception of color
+locations in 3-dimensional spaces. However, human perception of color
 does not correspond to the physiological axes above, but rather to polar
@@
   \item \textbf{chroma} (colorfulness)
-  \item \textbf{luminance} (brightness, `amount of gray')
+  \item \textbf{luminance} (brightness, amount of light)
 \end{enumerate}
@@
 
+%%FIXME: briefly explain interface
+
 \setkeys{Gin}{width=0.8\textwidth}
@@
 This strategy is depicted in the left panel of Figure~\ref{fig:pal-s-construct} for a blue
-hue $H = 260$ and different combinations of maximal chroma (0, 80 and 100, respecitvely) and
-minimal luminance (30, 30 and 50, respectively). The first two combinations are also
-shown in the first two rows of Figure~\ref{fig:pal-s-examples}. The right panel of
-Figure~\ref{fig:pal-s-construct} shows that the exact same strategy is not possible 
-for the green hue $H = 120$. While the gray colors without chroma can be chosen in 
+hue $H = 260$ and different combinations of maximal chroma ($C_{\max} = 0$, $80$ and $100$,
+respecitvely) and minimal luminance ($L_{\min} = 30$, $30$ and $50$, respectively).
+The first two combinations are also shown in the first two rows of Figure~\ref{fig:pal-s-examples}.
+The right panel of Figure~\ref{fig:pal-s-construct} shows that the exact same strategy
+is not possible  for the green hue $H = 120$. While the gray colors without chroma can be chosen in 
 the same way, there is a stronger trade-off between using dark colors (with low luminance)
 and colorful colors (with high chroma). Hence, the second path from light gray to 
-full green ends at a much lighter color with $L = 75$.
+full green ends at a much lighter color with $L_{\min} = 75$.
 
@@
 \caption{\label{fig:pal-s-construct} Constructing sequential palettes. The chroma/luminance
-plane is show for two hues $H = 260$ (left) and $H = 120$ (right). Colors are chosen by varying
+plane is shown for two hues $H = 260$ (left) and $H = 120$ (right). Colors are chosen by varying
 either only luminance or both luminance and chroma.}
@@
 chroma/luminance planes, allowing for many different combinations of maximal chroma and luminance.
+%%FIXME: symmetric difference
 As Figure~\ref{fig:pal-s-construct} illustrates, a bit of care is needed to choose two colors
@@
 In this section, we show a few examples for the various types of palettes applied
-to statistical graphics. The first illustration is visualization of a bivariate
+to statistical graphics. The first illustration is a visualization of a bivariate
 density estimation for the Old Faithful geyser eruptions data. Figure~\ref{fig:bkde}
@@
 north to south, first for the 10 Western provinces, then for the 6 Eastern provinces. Clearly,
-the SPD performed better in the North and the CDU better in the South; furthermore, the
+the SPD performed better in the North and the CDU better in the South; furthermore,
 Die Linke performed particularly well in the Eastern provinces and the Saarland.
@@
 As pointed out in Section~\ref{sec:diverging}, diverging palettes are particularly useful
-when visualizing residuals or correlation (with natural neutral value 0) or probabilities 
+when visualizing residuals or correlations (with natural neutral value 0) or probabilities 
 in 2-class supervised learning (with neutral value 0.5). Examples for both situations are

### 2006-09-30 trunk/vcd/paper/Handbook/hsc.Rnw
final edits from Antony Unwin
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@578 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \pkg{Mondrian} \cite{vcd:Theus:2003},
-\proglang{R} \cite{vcd:R:2005}, and \proglang{S-PLUS} \cite{vcd:SPLUS:2005}. 
+\proglang{R} \cite{vcd:R:2006}, and \proglang{S-PLUS} \cite{vcd:SPLUS:2005}. 
 Implementations of association and sieve plots can only be found in
@@
 \proglang{R} \index{R} system, using the extension packages \pkg{vcd} 
-\cite{vcd:Meyer+Zeileis+Hornik:2005} and
+\cite{vcd:Meyer+Zeileis+Hornik:2006} and
 \pkg{scatterplot3d} \cite{vcd:ligges+maechler:2003}
@@
 the first splitting variable dominates the plot, it should be chosen
-as to be the explanatory variable.
+to be the explanatory variable.
 
@@
 $i = 1, \dots, I$ and $j = 1, \dots, J$, and row and column sums
-$n_{i+} = \sum_i n_{ij}$ and $n_{+j} = \sum_j n_{ij}$, respectively. For
+$n_{i+} = \sum_j n_{ij}$ and $n_{+j} = \sum_i n_{ij}$, respectively. For
 convenience, the number of observations is denoted $n = n_{++}$.
@@
 diagonal, indicating a negative association of 
-the two variables. This gives evidence to the fact that for these patients,
+the two variables. This gives evidence for the fact that for these patients,
 visit frequency decreases with the length of stay.
@@
 than would be expected under the null of independence, and that this pattern is 
-reversed for patients visited less than monthly and that get never visited.
+reversed for patients visited less than monthly or never visited.
 
@@
 will demonstrate how the use of colors can greatly facilitate the
-detection of interesting patterns. Before, we start with some general 
+detection of interesting patterns. We start with some general 
 remarks on colors and color palettes.
@@
 These intuitive dimensions
-make HSV colors easyier to specifiy than, e.g, RGB colors. 
+make HSV colors easier to specify than RGB colors. 
 However, HSV colors have several disadvantages.
@@
 <<highlight,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
+mycol <- rep(grey.colors(2)[2:1], 1:2)
 mosaic(t(hospital), 
@@
          list("Visit frequency" = c("Regular","Less than\nmonthly","Never"))),
-       split = TRUE, spacing = spacing_highlighting, 
-       gp = gpar(fill = rep(grey.colors(2)[2:1], 1:2)))
+       split = TRUE, highlighting = 2, gp = gpar(fill = mycol, col = mycol))
 @ 
@@
 univariate statistics, whereas the off-diagonal cells feature plots whose
-variables are implicitly specified by the cells' position in the matrix. %'
+variables are implicitly specified by the cells' positions in the matrix. %'
 In Figure~\ref{fig:pairs}, the diagonal cells show bar plots for the
-distribution of the variables, and the off-diagonal cells mosaic plots
-for the corresponding pair of variables. The plots suggest that
+distributions of the variables, and the off-diagonal cells mosaic plots
+for the corresponding pairs of variables. The plots suggest that
 admission differs between male and female students, and between the departments,
@@
 In this example, we have a binary variable of interest (admission)
-that shall be ``explained'' by the others. 
+that is to be ``explained'' by the others. 
 A natural way of representing such a three-way table is to use a mosaic
@@
 
-In a first step, we create a mosaic plot for an exploratory view on the data:
+In a first step, we create a mosaic plot for an exploratory view of the data:
 Attitude towards punishment clearly is the dependent variable 
 here and should be used as the last split via highlighting. 
-Furthermore, we expect age and education to have an influence on both the memory
-and the attitude, hence they should be used first for splitting as stratifying
+Furthermore, we expect age and education to have an influence on both memory
+and attitude, hence they should be used first for splitting as stratifying
 variables. In fact, the question is whether memory has some influence on attitude, given 
@@
 highlighted in the tiles (see Figure~\ref{fig:punishhigh}).
-We can see that half of the people have more than 40 years of
-age, most of which of only completed elementary school (much more than 
+We can see that half of the people are more than 40 years of
+age, most of whom of only completed elementary school (much more than 
 in the other age groups). Furthermore, it can be seen that the proportion
@@
 accept moderate punishment of children than those without: 
-experienced violence seems to engender violence again.
+experience of punishment seems to support use of punishment.
 For the other education groups, the picture is less clear: some cells 
-indicate the same association while others do not.
+indicate the same association, while others do not.
 
@@
 In summary, both age and education clearly matter here.
-But is there really any significant influence of memory on the attitude towards 
+But is there really any significant influence of memory on attitude towards 
 punishment apart from that? Or, rephrased as a hypothesis:
 Are memory and attitude conditionally independent, given age and education?
-To focus on this question, we take another model-based view on the data
+To focus on this question, we take another model-based view of the data
 which avoids the distracting influence of the marginal variables (age
@@
 This visualization brings out much more clearly that the increased approval rate
-among people with memories of punishments is present mainly in the elementary education
-column and, to a lesser degree, in the age 40+ row. Using, e.g., a 
+among people with memories of punishment is present mainly in the elementary education
+column and, to a lesser degree, in the age 40+ row. Using a 
 permutation test similar to the one discussed in Section \ref{sec:test}, 
@@
 (conditional) independence. In the group of people with age between
-25--39, the lighter shading indicates that here,
+25--39, the lighter shading indicates that here
 the association is only significant at the 10\% level.
@@
 strongly influenced by the marginal distribution of education over age (in particular
-the large proportion of 40+ people with elementary education)
+the large proportion of 40+ people with elementary education),
 which is not of interest in the conditional independence problem. The
@@
 by applying them to flat representations. Mosaic plots are particularly useful
-for exploratory analysis whereas the other two require specification of a certain
-independence model for which deviations should be brought out. In addition, specialized
-`flavours' of the plots can leverage the
-exploratory (pairs plots, highlighting, doubledecker plots) 
-or model-based analysis (residual-based shading, conditional plots).
+for exploratory analysis, whereas the other two require specification of a certain
+independence model from which deviations should be examined. In addition, specialized
+`flavours' of the plots can leverage
+exploratory analyses (pairs plots, highlighting, doubledecker plots) 
+or model-based analyses (residual-based shading, conditional plots).
 

### 2006-10-02 trunk/vcd/paper/Strucplot/JSS.Rnw
first bunch of changes
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@581 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 flexible customization of the plots' graphical appearance, including 
-shading, labeling, spacing, and legend, by means of \underline{gr}aphical
-\underline{ap}pearance \underline{con}trol (`grapcon') functions. 
+shading, labeling, spacing, and legend, by means of `graphical
+appearance control' functions. 
 The framework is provided by the \proglang{R} package \pkg{vcd}.
@@
 \newcommand{\var}[1]{\textit{\texttt{#1}}}
-\newcommand{\data}[1]{`\texttt{#1}'}
+\newcommand{\data}[1]{\texttt{#1}}
 \newcommand{\class}[1]{\textsf{#1}}
@@
 composed of tiles (corresponding to the cells) created by recursive
-vertical and horizontal splits of a square. Thus, the area of each tile
+vertical and horizontal splits of a rectangle. Thus, the area of each tile
 is proportional to the corresponding cell entry \emph{given} the
@@
 \noindent Current implementations of mosaic displays can be found,
-e.g., for \proglang{SAS} \citep{vcd:SAS:2005}, \proglang{ViSta} \citep{vcd:young:1996}, 
-\proglang{MANET} \citep{vcd:Unwin+Hawkins+Hofmann:1996},
-\proglang{Mondrian} \citep{vcd:Theus:2003},
+e.g., for \proglang{SAS} \citep{vcd:SAS:2005}, \pkg{ViSta} \citep{vcd:young:1996}, 
+\pkg{MANET} \citep{vcd:Unwin+Hawkins+Hofmann:1996},
+\pkg{Mondrian} \citep{vcd:Theus:2003},
 \proglang{R} \citep{vcd:R:2006}, and \proglang{S-PLUS}
@@
 \pkg{vcd} \citep{vcd:Meyer+Zeileis+Hornik:2006},
-and \pkg{iPlots} \citep{vcd:urbanek+wichtrey:2006}, respectively.
+and \pkg{iplots} \citep{vcd:urbanek+wichtrey:2006}, respectively.
 Table \ref{tab:compare} gives an overview of the available functionality in these systems.
-Most environments are available on Windows, Apple Macintosh, and Linux/Unix
-variants, except \proglang{MANET} which is only available for the
-Macinthosh platforms. \proglang{Mondrian} is a JAVA application, as
-well as major parts of the \pkg{iPlots} package.
+Most environments are available on Windows, MacOS, and Linux/Unix
+variants, except \pkg{MANET} which is only available for the
+Macinthosh platforms.
 
@@
   \centering
-  \begin{tabular}{|l|c|c|c|c|c|c|}
+  \begin{tabular}{|l|c|c|c|c|c|c|c|c|c|}
     \hline
-                             & \proglang{SAS} & \proglang{S-PLUS} &
-                             \proglang{R} & \pkg{ViSta} & \pkg{MANET}
-                             & \pkg{Mondrian}\\\hline
-    Basic functionality      &  $\times$  &   $\times$    & $\times$ &   $\times$   & $\times$& $\times$\\
-    Shape                    &     &        & $\times$ & $\times$  &  $\times$&\\
-    Residual-based shadings  &  $\times$  &        & $\times$ &   highlighting   &($\times$)& ($\times$)\\
-    Conditional Views        &  $\times$  &        & $\times$ & $\times$ &  $\times$&\\
-    Interaction              &     &        & $\times$  &   $\times$   & $\times$& $\times$\\
-    Linking                  &     &        & $\times$  &       & $\times$& $\times$\\
-    Extensible Design        &     &        & $\times$ &       &  &\\
-    Language                 & SAS & S & S,JAVA & XLisp & C++ & JAVA\\
+                             &                &                   
+                             &\multicolumn{3}{c|}{}   &        & &\\
+                             & \proglang{SAS} & \proglang{S-PLUS} 
+                             &\multicolumn{3}{c|}{\proglang{R}} & \pkg{ViSta} & \pkg{MANET}
+                             & \pkg{Mondrian}\\
+                             &                &                   
+                             &\pkg{base}&\pkg{vcd}   &\pkg{iplots}&        & &\\\hline
+
+    Basic functionality      &  $\times$  & $\times$  & $\times$ &$\times$ &$\times$ &   $\times$   & $\times$& $\times$\\
+    Shape                    &            &           &          &         &$\times$ & $\times$  &  $\times$&\\
+    Res.-based shadings  &  $\times$  &        & $\times$ & $\times$ & ($\times$) &     &($\times$)& ($\times$)\\
+    Highlighting             &     &        &   &$\times$   &$\times$  & $\times$ & $\times$& $\times$\\
+    Conditional views        &  $\times$  &        &    &$\times$ &          & $\times$ &  $\times$&\\
+    Interaction              &     &        &   &  &$\times$  &   $\times$   & $\times$& $\times$\\
+    Linking                  &     &        &   &  &$\times$  &       & $\times$& $\times$\\
+    Extensible design        &     &        &  &$\times$ & &       &  &\\
+    Language                 & \proglang{SAS} & \proglang{S} &
+    \proglang{S} & \proglang{S} & \proglang{Java} & \proglang{XLisp} & \proglang{C++} & \proglang{Java}\\
 \hline
@@
 
-%%DM%% Fixme: separate columns for iplot, base, and vcd?
-
 Figures \ref{fig:arthritis} to \ref{fig:titanic} illustrate some of these
 extensions (the code is given in Appendix \ref{sec:code}). 
-Figure \ref{fig:arthritis} shows the results from a double-blind clinical trial
+Figure~\ref{fig:arthritis} shows the results from a double-blind clinical trial
 investigating a new treatment for rheumatoid arthritis, using an
@@
 \ref{fig:ucbadmissions} visualizes the well-known UCB admissions data
-by the means of a conditional association plot. The panels show the
+by means of a conditional association plot. The panels show the
 residuals from a conditional independence model (independence of gender
@@
 the situation in department A (more women/less men accepted than
-would be expected under the null) causes the rejection of the
-hypothesis of conditional independence. Figure \ref{fig:presex}
+would be expected under the null hypothesis) causes the rejection of the
+hypothesis of conditional independence. Figure~\ref{fig:presex}
 illustrates the conditional independence of premarital and
-extramarital sex, given gender and marital status. The Chi-squared
-test of independence rejects the null hypothesis: possibly, because
-too many men who had premarital sex also had extramarital sex?
-Finally, Figure \ref{fig:titanic} visualizes the `Survival on the Titanic' 
-data using a double-decker plot. Here, a binary response (survival to the disaster)
+extramarital sex, given gender and marital status. The $\chi^2$ test 
+of independence rejects the null hypothesis: possibly, because
+the tendency of people  to have extramarital
+sex when they had premarital sex 
+is particularly marked among married people?
+Finally, Figure~\ref{fig:titanic} visualizes the `Survival on the Titanic' 
+data using a double-decker plot. Here, a binary response (survival of the disaster)
 shall be explained by other factors (class, gender, and age). The blue
@@
 <<UCBAdmissions,fig=TRUE,echo=FALSE,height=6,width=10,results=hide>>=
-cotabplot(UCBAdmissions, panel = cotab_assoc, shade = TRUE, legend = FALSE)
+cotabplot(UCBAdmissions, panel = cotab_coindep, shade = TRUE, legend = FALSE, 
+          type = "assoc")
 @
@@
 \begin{center}
-<<Titanic,fig=TRUE,echo=FALSE,height=6,width=8,results=hide>>=
+<<Titanic,fig=TRUE,echo=FALSE,height=6,width=10,results=hide>>=
 doubledecker(Survived ~ ., data = Titanic, labeling_args = 
@@
 Exploratory techniques include specialized displays for the bivariate
-case, as well as pairs and trellis plot-like displays for
+case, as well as pairs and trellis-type displays for
 higher-dimensional tables. Model-based tools include methods suitable
@@
 graphical appearance aspects such as shading, labeling, and spacing are
-modularized by means of `graphical appearance functions', allowing
+modularized by means of ``\underline{\vphantom{g}gr}aphical 
+\underline{\vphantom{g}ap}pearance \underline{\vphantom{g}con}trol''
+(\emph{grapcon}) functions, allowing
 fine-granular customization and user-level extensions.
@@
 the results of independence tests. Also, we explain step-by-step how
-the concepts of generating functions and graphical appearance control
-functions can be combined to provide a flexible customization of
+the concepts of generating and grapcon functions
+can be combined to provide a flexible customization of
 complex graphical displays as created by the strucplot framework.
@@
 and spacing features, respectively. Section \ref{sec:example}
-exemplifies the framework in the analysis of a four-way data set, 
-before Section \ref{sec:conclusion} concludes the work.
+exemplifies the framework in the analysis of a four-way data set.
+Section \ref{sec:conclusion} concludes the work.
 
@@
 
-The strucplot framework is highly modularized: Figure \ref{fig:struc}
+The strucplot framework is highly modularized: Figure~\ref{fig:struc}
 shows the hierarchical relationship between the various components. 
@@
 appearance of the plot (see Table \ref{tab:grapcons} for an overview). 
-These are examples of `\underline{gr}aphical \underline{ap}pearance \underline{con}trol'
-(`grapcon') functions. They are created by generating functions
+These are examples of grapcon functions. They are created by generating functions
 (`grapcon generators'), allowing
-flexible parameterization and extensibility (Figure \ref{fig:struc}
+flexible parameterization and extensibility (Figure~\ref{fig:struc}
 only shows the generators). The generator names
-follow the naming convention \code{\emph{group\_foo}()},  
-where (\code{\emph{group}}) reflects the group the
+follow the naming convention \code{\textit{group\_foo}()},  
+where (\code{\textit{group}}) reflects the group the
 generators belong to (strucplot core, labeling,
 legend, shading, or spacing). The workhorse functions (created by 
-\code{struc\_\emph{foo}()}, 
-\code{labeling\_\emph{foo}()}, and \code{legend\_\emph{foo}()}) 
+\code{struc\_\textit{foo}()}, 
+\code{labeling\_\textit{foo}()}, and \code{legend\_\textit{foo}()}) 
 directly produce graphical output (i.e., ``add ink to the canvas''), whereas
 the parameter functions
-(created by \code{spacing\_\emph{foo}()} and \code{shading\_\emph{foo}()}) compute
+(created by \code{spacing\_\textit{foo}()} and \code{shading\_\textit{foo}()}) compute
 graphical parameters used by the others. The grapcon functions returned by 
-\code{struc\_\emph{foo}()} implement the core functionality, creating the tiles and their
+\code{struc\_\textit{foo}()} implement the core functionality, 
+creating the tiles and their
 content.  On the second level of the framework, a suitable combination
@@
 This central function 
-sets up the graphical layout using grid viewports (see Figure \ref{fig:layout}),
+sets up the graphical layout using grid viewports (see Figure~\ref{fig:layout}),
 and coordinates the specified core, labeling, shading, and spacing functions to produce 
@@
 and support for model formulae. Finally, on the fourth
-level, there are `related' \pkg{vcd} functions (such as \codefun{cotabplot}
+level, there are ``related'' \pkg{vcd} functions (such as \codefun{cotabplot}
 and the \codefun{pairs} methods for table objects) 
@@
                                                   for conditioning variables\\
-              && and cell labels for conditioned variables.\\
+              && and cell labels for conditioned variables\\
               & \codefun{labeling\_doubledecker} & draws labels for
               doubledecker plot\\
-              & \codefun{labeling\_lboxed} & left-aligned labels with boxes.\\
+              & \codefun{labeling\_lboxed} & left-aligned labels with boxes\\
               & \codefun{labeling\_left} & left-aligned border labels\\
@@
      shading  & \codefun{shading\_binary} & visualizes the sign of the  residuals\\
-              & \codefun{shading\_Friendly} & implements Friendly-like
-              shading (based on HSV-colors)\\
-              & \codefun{shading\_sieve} & implements Friendly-like
-              shading used for sieve plots\\ && (based on HCL-colors)\\
-              & \codefun{shading\_hcl} & shading based on HCL-colors\\
-              & \codefun{shading\_hsv} & shading based on HSV-colors\\
+              & \codefun{shading\_Friendly} & implements Friendly
+              shading (based on HSV colors)\\
+              & \codefun{shading\_hcl} & shading based on HCL colors\\
+              & \codefun{shading\_hsv} & shading based on HSV colors\\
               & \codefun{shading\_max} & shading visualizing the
-              maximum test statistic\\ && (based on HCL-colors)\\\hline\hline
+              maximum test statistic\\ && (based on HCL colors)\\
+              & \codefun{shading\_sieve} & implements Friendly
+              shading customized for sieve plots\\ && (based on HCL colors)\\\hline\hline
      spacing  & \codefun{spacing\_conditional} & increasing spacing
               for conditioning variables,\\&& equal spacing for
-              conditioned variables.\\
+              conditioned variables\\
               & \codefun{spacing\_dimequal} & equal spacing for each dimension\\
@@
 data containing two polytomous variables (hair and eye color), 
-as well as one (artificial) dichotomous variable (gender). The
+as well as one (artificial) dichotomous gender variable (\code{Sex}). The
 `flattened' contingency table can be obtained using the
@@
 <<structable,echo=TRUE,results=verbatim>>=
-(hec <- structable(Eye ~ Sex + Hair, data = HairEyeColor))
+(HEC <- structable(Eye ~ Sex + Hair, data = HairEyeColor))
 @
@@
 % dimensions of previous splits. 
-Figure \ref{fig:observed} depicts the effect of
+The effect of
 
 <<Observed>>=
-mosaic(hec)
+mosaic(HEC)
 @
@@
 
-\noindent For zero entries, tiles have zero area and are, additionally,
-marked by small bullets (see, e.g, Figure \ref{fig:doubledecker}). By
-default, these cells are not splitted further. The bullets help
+\noindent depicts the observed frequencies of the \code{HairEyeColor}
+data. If there are zero entries, tiles have zero area and are, additionally,
+marked by small bullets (see, e.g, Figure~\ref{fig:titanic}). By
+default, these cells are not split further. The bullets help
 distinguishing very small cells from zero entries, and are
@@
 \proglang{R}, the default split direction and level ordering in all strucplot
-displays correspond to the textual representation.
+displays correspond to the textual representation produced by the
+print methods.
 It is also possible to visualize the expected values instead of the
-observed values (see Figure \ref{fig:expected}):
+observed values (see Figure~\ref{fig:expected}):
 
 <<Expected>>=
-mosaic(hec, type = "expected")
+mosaic(HEC, type = "expected")
 @
@@
 a sieve plot \citep{vcd:riedwyl+schuepbach:1994} 
-could be used (see Figure \ref{fig:sieve}):
+could be used (see Figure~\ref{fig:sieve}):
 
 <<sieve>>=
-sieve(hec)
+sieve(~ Sex + Eye + Hair, data = HEC, spacing = spacing_dimequal(c(2,0,0)))
 @
@@
 The Pearson residuals (standardized deviations of observed from expected values) 
-are preferably visualized using association
+are conveniently visualized using association
 plots \citep{vcd:Cohen:1980}. In contrast to \codefun{assocplot} in
 base \proglang{R}, \pkg{vcd}'s \codefun{assoc}
-function scales to more than two variables (see Figure \ref{fig:residuals}):
+function scales to more than two variables (see Figure~\ref{fig:residuals}):
 
 <<Residuals>>=
-assoc(hec, compress = FALSE)
+assoc(HEC, compress = FALSE)
 @
@@
 
-\noindent The \code{compress} argument keeps distances between tiles
-equal for better comparison.
+\noindent where the \code{compress} argument keeps distances between tiles
+equal.
 
@@
 the splitting of the tiles can be controlled using the
-\code{split\_vertical} argument (default: alternating splits starting
-with a vertical one).
+\code{split\_vertical} argument. The default is to alternate splits starting
+with a horizontal one (see Figure~\ref{fig:split}).
 
@@
 <<split1>>=
-mosaic(hec, split_vertical = c(TRUE, FALSE, TRUE), 
+mosaic(HEC, split_vertical = c(TRUE, FALSE, TRUE), 
 labeling_args = list(abbreviate = c(Eye = 3)))
@@
 
-\noindent (Note that \code{hec}, a \class{structable} object, already includes a
+\begin{figure}[p]
+\begin{center}
+<<splitfig,fig=TRUE,echo=FALSE,width=7,height=7>>=
+<<split1>>
+@
+\caption{Mosaic plot for the \data{HairEyeColor} data---alternative splitting.}
+\label{fig:split}
+\end{center}
+\end{figure}
+
+\noindent (Note that \code{HEC}, a \class{structable} object, already includes a
 splitting information which simply gets overloaded in this example.) 
@@
 argument taking a vector of \code{"h"} and
-\code{"v"} characters (see Figure \ref{fig:split}):
+\code{"v"} characters:
 
 <<split2>>=
-mosaic(hec, direction = c("h","h","v"))
-@
-
-\begin{figure}[p]
-\begin{center}
-<<splitfig,fig=TRUE,echo=FALSE,width=7,height=7>>=
-<<split1>>
-@
-\caption{Mosaic plot for the \data{HairEyeColor} data---alternative splitting.}
-\label{fig:split}
-\end{center}
-\end{figure}
+mosaic(HEC, direction = c("v","h","v"))
+@
 
@@
 can be customized in a quite flexible way. For example, 
-\codefun{doubledecker.default} is simply a wrapper for 
-\codefun{strucplot}, setting the right defaults:
-
-<<doubledecker>>=
-doubledecker.default <- 
-  function(x,
-           depvar = length(dim(x)), 
-           margins = c(1, 4, length(dim(x)) + 1, 1),
-           col = rainbow_hcl(tail(dim(x), 1), end = 260),
-           labeling = labeling_doubledecker,
-           spacing = spacing_highlighting,
-           main = NULL, 
-           keep_aspect_ratio = FALSE,
-           ...) {
-  d <- dim(x)
-  l <- length(d)
-  if (is.character(depvar))
-    depvar <- match(depvar, names(dimnames(x)))
-  condvars <- (1:l)[-depvar]
-  x <- aperm(x, c(condvars, depvar))
-  strucplot(x,
-         condvars = l - 1,
-         spacing = spacing,
-         split = c(rep.int(TRUE, l - 1), FALSE),
-         gp = gpar(fill = col),
-         shade = TRUE,
-         labeling = labeling,
-         main = main,
-         margins = margins,
-         legend = NULL,
-         keep_aspect_ratio = keep_aspect_ratio,
-         ...
-         )
-}
-@
-
-\noindent Most default settings such as colors, spacing, and labeling are specified via the parameters of
-\codefun{doubledecker.default} and passed through to
+the default method for \codefun{doubledecker} is simply a wrapper for 
+\codefun{strucplot}, setting the right defaults.
+Most default settings such as colors, spacing, and labeling are
+specified via the parameters and passed through to
 \codefun{strucplot}. The additional code just handles the dependent
-variable information, and in particular permutes the table to have the
+variable information, and in particular permutes the table to have the dependent
 variable as the last dimension as required for the doubledecker plot.
-Figure \ref{fig:doubledecker} shows a doubledecker plot of
+Figure~\ref{fig:titanic} shows a doubledecker plot of
 the \data{Titanic} data, explaining the probability of survival (`survived') by age, given
@@
 
-\begin{figure}[h]
-\begin{center}
-<<doubledeckerfig,fig=TRUE,echo=FALSE>>=
-<<doubledecker1>>
-@
-\caption{Doubledecker plot for the \data{Titanic} data.}
-\label{fig:doubledecker}
-\end{center}
-\end{figure}
-
 \subsection{Conditional and partial views}
@@
 suggested by the analysis task at hand. Subtables can be selected in a
-similar way than for objects of class \class{table} using
+similar way as for objects of class \class{table} using
 indexing. Note, however, that subsetting of \class{structable} objects 
@@
 subsetting is only possible ``outside-in'', that is, indexing operates
-on blocks defined by the variable levels. In the following, we again use the
-Titanic data, this time collapsed over `Survived' to investigate the
-structure of crew and passengers:
+on blocks defined by the variable levels. In the following, we use the
+Titanic data again, this time collapsed over `Survived' to investigate the
+structure of crew and passengers (and having the \code{Child} and
+\code{Age} labels of the \code{Age} variable swapped for optical clarity):
+
+<<echo=FALSE>>=
+options(width=75)
+@
 
 <<subsetting,results=verbatim>>=
-(tit <- structable(~ Sex + Class + Age, data = Titanic[,,2:1,]))
-tit["Male",]
-tit["Male", c("1st","2nd","3rd")]
+(STD <- structable(~ Sex + Class + Age, data = Titanic[,,2:1,]))
+STD["Male",]
+STD["Male", c("1st","2nd","3rd")]
+@
+
+<<echo=FALSE>>=
+options(width=70)
 @
@@
 Here again, the sequence of conditioning levels is restricted by the
-hierarchical structure of the \class{structable} object:
+hierarchical structure of the \class{structable} object. In the
+following examples, note that compared to subsetting, the first
+dimension(s) have been dropped:
 
 <<conditioning,results=verbatim>>=
-tit[["Male",]]
-tit[[c("Male", "Adult"),]]
-tit[["Male","1st"]]
+STD[["Male",]]
+STD[[c("Male", "Adult"),]]
+STD[["Male","1st"]]
 @
@@
 pushViewport(viewport(layout.pos.col = 1))
-mosaic(tit[["Male"]], margins = c(left = 2.5, top = 2.5, 0), sub = "Male", newpage = FALSE)
+mosaic(STD[["Male"]], margins = c(left = 2.5, top = 2.5, 0), sub = "Male", newpage = FALSE)
 popViewport()
@@
 pushViewport(viewport(layout.pos.col = 2))
-mosaic(tit[["Female"]], margins = c(top = 2.5, 0), sub = "Female", newpage = FALSE)
+mosaic(STD[["Female"]], margins = c(top = 2.5, 0), sub = "Female", newpage = FALSE)
 popViewport(2)
@@
 vector with up to four values whose unnamed components are recycled,
-but ``overruled'' by the named arguments. Thus, in the example, only
-the top margin is set to 2 lines, and all other to 0.
+but ``overruled'' by the named arguments. Thus, in the second example, only
+the top margin is set to 2.5 lines, and all other to 0.
 This idea applies to almost all
@@
 in a lattice-like layout, conditioning on variable \emph{levels}.
-The plot in Figure \ref{fig:cotabplot} shows hair and eye color, given sex:
+The plot in Figure~\ref{fig:cotabplot} shows hair and eye color, given sex:
 
 <<cotabplot>>=
-cotabplot(~ Class + Age | Sex, data = tit, split_vertical = TRUE)
+cotabplot(~ Class + Age | Sex, data = STD, split_vertical = TRUE)
 @
@@
   gender, using separate mosaic displays like the `manual' plot 
-  in Figure \ref{fig:parttable}.}
+  in Figure~\ref{fig:parttable}.}
 \label{fig:cotabplot}
@@
 
-\noindent Visualizing the strata separately `hides' the distribution of the
-conditioning variable(s) which may or may not be appropriated or
+\noindent Visualizing the strata separately ``hides'' the distribution of the
+conditioning variable(s) which may or may not be appropriate or
 sensible in a particular analysis step. If we wish to keep the
@@
 Both the default and the formula interface of \codefun{mosaic} allow the
-specification of conditioning variables (see Figure \ref{fig:conditioning}):
+specification of conditioning variables (see Figure~\ref{fig:conditioning}):
 
 <<Conditioning1>>=
-mosaic(tit, conditional = "Sex", split_vertical = c(TRUE, TRUE, FALSE), 
-       keep_aspect_ratio = FALSE)
+mosaic(STD, condvars = "Sex", split_vertical = c(TRUE, TRUE, FALSE))
 @ 
 <<Conditioning2>>=
-mosaic(~ Class + Age | Sex, data = tit, split_vertical = c(TRUE, TRUE, FALSE), 
-       keep_aspect_ratio = FALSE)
+mosaic(~ Class + Age | Sex, data = STD, split_vertical = c(TRUE, TRUE, FALSE))
 @
@@
 distribution of class and age, given gender, this time using a
-single mosaic plot. In contrast to Figure \ref{fig:parttable}, this
-plot also visualizes the marginal distribution of `Sex'.}
+single mosaic plot. In contrast to Figures~\ref{fig:parttable} and \ref{fig:cotabplot}, 
+this plot also visualizes the marginal distribution of \code{Sex}.}
 \label{fig:conditioning}
@@
 conditioning variables are permuted ahead of the the conditioned variables
-in the table, and for the formula interface, 
-that \codefun{spacing\_conditional} is used as default to better
+in the table, and that \codefun{spacing\_conditional} is used as default to better
 distinguish conditioning from conditioned dimensions.
 This spacing uses equal space between tiles of conditioned variables,
-and increasing space between tiles of conditioning variables. In
-addition, we release the fixed aspect ratio to get less distorted margins.
-
-Another high-level function for visualizing conditional independence
-models are the \codefun{pairs} methods for table and structable objects.
+and increasing space between tiles of conditioning variables.
+
+Another set of high-level functions for visualizing conditional independence
+models are the \codefun{pairs} methods for \class{table} and \class{structable} objects.
 In contrast to \codefun{cotabplot} which conditions on variables,
@@
 having strucplot displays in the off-diagonal panels, and
-the variable names (optionally, with univariate statistics) in the diagonal cells. 
-Figure \ref{fig:pairs} shows a pairs display for the
+the variable names (optionally, with univariate displays) in the diagonal cells. 
+Figure~\ref{fig:pairs} shows a pairs display for the
 \data{Titanic} data with univariate mosaics in the diagonal,
-and mosaic plots visualizing mutual independence in the upper and lower triangles.
+and mosaic plots visualizing the corresponding bivariate mosaics
+in the upper and lower triangles.
 Due to the inherent asymmetry of mosaic displays, the corresponding plots in the
@@
 <<pairs>>=
-pairs(tit, highlighting = 2, diag_panel = pairs_diagonal_mosaic, 
-      diag_panel_args = list(fill = grey.colors, alternate_labels = FALSE))
+pairs(STD, highlighting = 2, diag_panel = pairs_diagonal_mosaic, 
+      diag_panel_args = list(fill = grey.colors))
 @
@@
 
-\noindent (The labels of the variables are to be read from left to
-right and from top to bottom.) In plots produced by \codefun{pairs}, 
+\noindent The labels of the variables are to be read from left to
+right and from top to bottom. In addition, the levels can be matched
+by position with the columns and by shading with the rows.
+In plots produced by \codefun{pairs}, 
 each panel's row and column define two variables $X$ and $Y$ used for the
@@
 
-Since the matrix is symmetric, the upper and lower parts can
+Upper and lower parts can
 independently be used to display different 
-types of independence models, or different strucplots displays
+types of independence models, or different strucplot displays
 (mosaic, association, or sieve plots).
@@
 \begin{center}
-\code{[Optional prefix]cell:\emph{Variable1}=\emph{Level1},\emph{Variable2}=\emph{Level2}} \dots
+\code{\emph{[Optional prefix]}cell:\emph{Variable1}=\emph{Level1},\emph{Variable2}=\emph{Level2}} \dots
 \end{center}
@@
 The following example shows how to access parts of the plot after it has
-been drawn (see Figure \ref{fig:afterplot}):
+been drawn (see Figure~\ref{fig:afterplot}):
 
 <<viewportnames>>=
-mosaic(~ Hair + Eye, data = hec, pop = FALSE)
+mosaic(~ Hair + Eye, data = HEC, pop = FALSE)
 
@@
 graphical parameters of plot elements \emph{after} 
-the plotting (see Figure \ref{fig:changeplot}):
+the plotting (see Figure~\ref{fig:changeplot}):
 
 <<changeplot,results=verbatim>>=
-assoc(Eye ~ Hair, data = hec, pop = FALSE)
+assoc(Eye ~ Hair, data = HEC, pop = FALSE)
 getNames()[1:6]
@@
 
-\subsection{Performance Issues}
+\subsection{Performance issues}
 \label{sec:performance}
@@
 creating and nesting \pkg{grid} viewports. The main time-consuming steps performed by
-the core functions are the follwing:
+the core functions are the following:
 
@@
   \label{eq:numbervp}
-  T_d = k_1 + k_1k_2 + \dots + k_1 \dots k_d = \sum_{i=1}^d \prod_{j \le i} k_j
+  T_d \quad = \quad k_1 + k_1k_2 + \cdots + k_1 \cdots k_d \quad =\quad \sum_{i=1}^d \prod_{j \le i} k_j
 \end{equation}
@@
 first dimension, then, for \emph{each} of those, the $k_2$ levels of
-the second dimension, etc. Assuming an equal number of levels $k$
-for all dimensions, we obtain:
+the second dimension, etc. If the number of levels is equal ($k$) for
+all dimensions, this simplifies to:
 
@@
   \label{eq:equalvp}
-  T_d = \sum_{i=1}^d k^i = \frac{k(k^d-1)}{k-1}
+  T_d \quad = \quad \sum_{i=1}^d k^i = \frac{k(k^d-1)}{k-1}
 \end{equation}
@@
 
-\section{Residual-based shadings}
+\section{Shadings}
 \label{sec:shading}
@@
 For convenience, however, the user can also
-supply a specialized graphical appearance control (``grapcon'') 
-function that computes such an object given a vector of
+supply a grapcon function that computes such an object given a vector of
 residuals, or, alternatively, a generating function that takes
@@
 were too many male students accepted compared to the presumably
-discriminated female students (see Figure \ref{fig:ucb}):
+discriminated female students (see Figure~\ref{fig:ucb}):
 
@@
 For visual clarity, we additionally hide the tiles' borders by setting the \code{col} 
-component to 0 (white).
+component to 0 (transparent).
 
@@
 we will highlight all cells corresponding to survived
-passengers (see Figure \ref{fig:recycling}):
+passengers (see Figure~\ref{fig:recycling}):
 
@@
 hand''. We will use `light blue' color for positive and `light salmon'
-color for negative residuals (see Figure \ref{fig:binary}):
+color for negative residuals (see Figure~\ref{fig:binary}):
 
@@
 As an example, we will visualize the association of hair and eye
-color in the \data{HairEyeColor} data set (see Figure \ref{fig:haireye})
+color in the \data{HairEyeColor} data set (see Figure~\ref{fig:haireye})
 
@@
 be changed to alternative values using the \code{interpolate}
-argument (see Figure \ref{fig:interpolate}):
+argument (see Figure~\ref{fig:interpolate}):
 
@@
 function maps residuals greater than 4 to a saturation
-level of 1. However, the resulting plot (Figure \ref{fig:continuous}) is 
+level of 1. However, the resulting plot (Figure~\ref{fig:continuous}) is 
 deceiving:
@@
 $\chi^2$ test does not reject the null hypothesis of
-independence. Consequently, the colors appear dark (see Figure \ref{fig:bundesliga}):
+independence. Consequently, the colors appear dark (see Figure~\ref{fig:bundesliga}):
 
@@
 the \data{Bundesliga} data from above, this time using the Friendly 
-scheme and, in addition, an alternative legend (see Figure \ref{fig:friendly}):
+scheme and, in addition, an alternative legend (see Figure~\ref{fig:friendly}):
 
@@
 perceptual characteristics of HCL color space.
-% Figure \ref{fig:shadingHSVHCL} depicts 
+% Figure~\ref{fig:shadingHSVHCL} depicts 
 % HSV space in the upper panel and HCL space in the lower panel.
@@
 The following example again illustrates the \data{HairEyeColor} data, 
-this time with HCL colors (Figure \ref{fig:hclshading1} depicts the default palette, 
-and Figure \ref{fig:hclshading2} an alternative setting):
+this time with HCL colors (Figure~\ref{fig:hclshading1} depicts the default palette, 
+and Figure~\ref{fig:hclshading2} an alternative setting):
 
@@
 for rheumatoid arthritis on a group of female patients using the
-maximum shading (see Figure \ref{fig:maximum}):
+maximum shading (see Figure~\ref{fig:maximum}):
 
@@
 crew. By default, the variable names and levels are plotted `around' the
-plot in a counter-clockwise way (see Figure \ref{fig:defaults}):
+plot in a counter-clockwise way (see Figure~\ref{fig:defaults}):
 
@@
 This issue can be addressed in several ways. The `brute force'
-method is to enable clipping for these dimensions (see Figure \ref{fig:clipping}):
+method is to enable clipping for these dimensions (see Figure~\ref{fig:clipping}):
 
@@
 A more sensible solution to the overlap problem is to abbreviate the 
-levels (see Figure \ref{fig:abbreviating}):
+levels (see Figure~\ref{fig:abbreviating}):
 
@@
 \noindent Finally, we could also inhibit the output of repeated levels 
-(see Figure \ref{fig:repeat}):
+(see Figure~\ref{fig:repeat}):
 
@@
 A first simple, but effectful modification is to position
-all labels and variables left-aligned:  (see Figure \ref{fig:left}):
+all labels and variables left-aligned:  (see Figure~\ref{fig:left}):
 
@@
 bottom and right margins, and all variable names to the top and left
-margins (see Figure \ref{fig:margins}):
+margins (see Figure~\ref{fig:margins}):
 
@@
 Now, we will add boxes to the labels and additionally
-enable clipping (see Figure \ref{fig:boxes}):
+enable clipping (see Figure~\ref{fig:boxes}):
 
@@
 \noindent Another variant is to put the variable names into the
-same line as the levels (see Figure \ref{fig:labbl}):
+same line as the levels (see Figure~\ref{fig:labbl}):
 
@@
 cells. As an example, we use the \data{PreSex} data on pre- and
-extramarital sex and divorce (see Figure \ref{fig:cell}):
+extramarital sex and divorce (see Figure~\ref{fig:cell}):
 
@@
 \noindent In the case of narrow cells, it might be useful to
-abbreviate labels and/or variable names and turn off clipping (see Figure \ref{fig:cell2}):
+abbreviate labels and/or variable names and turn off clipping (see Figure~\ref{fig:cell2}):
 
@@
 labeling with border labeling as done by \codefun{labels\_conditional}
-(see Figure \ref{fig:conditional}):
+(see Figure~\ref{fig:conditional}):
 
@@
 this is used to add all observed values greater
-than 5 to the cells after the mosaic has been plotted (see Figure \ref{fig:text}):
+than 5 to the cells after the mosaic has been plotted (see Figure~\ref{fig:text}):
 
@@
 the last remedy could be to simply list the levels below the plot
-(see Figure \ref{fig:list}):
+(see Figure~\ref{fig:list}):
 
@@
 Consider again the introductory example of the \data{Arthritis} data 
-(Figure \ref{fig:arthritis}). Since we are interested in the effect of
+(Figure~\ref{fig:arthritis}). Since we are interested in the effect of
 the medicament in the placebo and treatment groups, a mosaic plot is
@@
 the strucplot framework, it suffices to set the space between the `Improved' tiles to 0 (see
-Figure \ref{fig:artspine}):
+Figure~\ref{fig:artspine}):
 
@@
 \codefun{spacing\_increase} (for the conditioning dimensions). As an example, 
-consider Figure \ref{fig:presex}: the spacing clearly allows to better 
+consider Figure~\ref{fig:presex}: the spacing clearly allows to better 
 distinguish the conditioning variables 
@@
 
-\noindent A first overview can be obtained using a pairs plot (Figure \ref{fig:ocpairs}):
+\noindent A first overview can be obtained using a pairs plot (Figure~\ref{fig:ocpairs}):
 
@@
 A different view on the data can be obtained using a doubledecker plot, explaining
-survival by the three other factors (Figure \ref{fig:ocdoubledecker}):
+survival by the three other factors (Figure~\ref{fig:ocdoubledecker}):
 
@@
 again use the pairs plot display, this time using color to bring out departures from
-the hypothesized marginal independence models (Figure \ref{fig:ocpairsshaded}):
+the hypothesized marginal independence models (Figure~\ref{fig:ocpairsshaded}):
 
@@
 log-linear models for the four-way table. The default model used by \codefun{mosaic} is
-the total independence model, visualized in Figure \ref{fig:ocmosaictotal}:
+the total independence model, visualized in Figure~\ref{fig:ocmosaictotal}:
 
@@
 
-\item Mosaic plot for the \data{Arthritis} data in Figure \ref{fig:arthritis}:
+\item Mosaic plot for the \data{Arthritis} data in Figure~\ref{fig:arthritis}:
 \begin{Sinput}
@@
 \item Conditional association plot for the \data{UCBAdmissions} 
-data in Figure \ref{fig:ucbadmissions}:
+data in Figure~\ref{fig:ucbadmissions}:
 \begin{Sinput}
@@
 
-\item Mosaic plot for the \data{PreSex} data in Figure \ref{fig:presex}:
+\item Mosaic plot for the \data{PreSex} data in Figure~\ref{fig:presex}:
 \begin{Sinput}
@@
 
-\item Double-decker plot for the \data{Titanic} data in Figure \ref{fig:titanic}:
+\item Double-decker plot for the \data{Titanic} data in Figure~\ref{fig:titanic}:
 \begin{Sinput}

### 2006-10-06 trunk/vcd/paper/Strucplot/JSS.Rnw
second bunch of changes
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@582 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% declarations for jss.cls %%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \Abstract{
-This paper describes the `strucplot' framework for the visualization of
+This paper describes the ``strucplot'' framework for the visualization of
 multi-way contingency tables. Strucplot displays include hierarchical
@@
 flexible customization of the plots' graphical appearance, including 
-shading, labeling, spacing, and legend, by means of `graphical
-appearance control' functions. 
+shading, labeling, spacing, and legend, by means of ``graphical
+appearance control'' functions. 
 The framework is provided by the \proglang{R} package \pkg{vcd}.
@@
   E-mail: \email{Achim.Zeileis@wu-wien.ac.at}\\
-  URL: \url{http://www.ci.tuwien.ac.at/~zeileis/}\\
+  URL: \url{http://statmath.wu-wien.ac.at/~zeileis/}\\
 
@@
   E-mail: \email{Kurt.Hornik@wu-wien.ac.at}\\
-  URL: \url{http://www.wu-wien.ac.at/cstat/hornik/}\\
+  URL: \url{http://statmath.wu-wien.ac.at/~hornik/}\\
 
@@
 extended mosaic plot with residual-based shading based on the
-maximum statistic: clearly, the new treatment is effective. Figure
+maximum statistic: clearly, the new treatment is effective. The dark
+blue cell indicates that the rate of treated patients showing marked improvement
+is significant at the 1\% level. Figure
 \ref{fig:ucbadmissions} visualizes the well-known UCB admissions data
@@
 extramarital sex, given gender and marital status. The $\chi^2$ test 
-of independence rejects the null hypothesis: possibly, because
+of independence, based on the permutation distribution,  
+rejects the null hypothesis: possibly, because
 the tendency of people to have extramarital
 sex when they had premarital sex 
-is particularly marked among married people?
-Finally, Figure~\ref{fig:titanic} visualizes the `Survival on the Titanic' 
+is particularly marked among married people? The rate of such women and men
+ist significant at the 0.01 and 0.1 level, respectively.
+Finally, Figure~\ref{fig:titanic} visualizes the ``Survival on the Titanic'' 
 data using a double-decker plot. Here, a binary response (survival of the disaster)
-shall be explained by other factors (class, gender, and age). The blue
+is to be explained by other factors (class, gender, and age). The gray
 boxes represent the proportion of survived passengers in a particular
@@
 <<PreSex,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
-mosaic(PreSex, condvars = c(1, 4), shade = TRUE)
+presextest <- coindep_test(PreSex, margin = c(1,4), 
+                           indepfun = function(x) sum(x^2), n = 5000)
+mosaic(PreSex, condvars = c(1, 4), shade = TRUE, 
+       gp_args = list(p.value = presextest$p.value))
 @
@@
 conditional splits, given the margins. The result is a 
-`flat' representation that can be visualized in
+``flat'' representation that can be visualized in
 ways similar to a two-dimensional table.
@@
 These are examples of grapcon functions. They are created by generating functions
-(`grapcon generators'), allowing
+(\emph{grapcon generators}), allowing
 flexible parameterization and extensibility (Figure~\ref{fig:struc}
@@
 as well as one (artificial) dichotomous gender variable (\code{Sex}). The
-`flattened' contingency table can be obtained using the
+``flattened'' contingency table can be obtained using the
 \codefun{structable} function (quite similar to \codefun{ftable} in
@@
 Figure~\ref{fig:titanic} shows a doubledecker plot of
-the \data{Titanic} data, explaining the probability of survival (`survived') by age, given
+the \data{Titanic} data, explaining the probability of survival (``survived'') by age, given
 sex, given class. It is created by:
@@
 on blocks defined by the variable levels. In the following, we use the
-Titanic data again, this time collapsed over `Survived' to investigate the
+Titanic data again, this time collapsed over \code{Survived} to investigate the
 structure of crew and passengers (and having the \code{Child} and
@@
 distribution of class and age, given gender. The marginal
-distribution of `Sex' cannot be seen.}
+distribution of gender cannot be seen.}
 \label{fig:parttable}
@@
   again visualizing the distribution of age and class, given
-  gender, using separate mosaic displays like the `manual' plot 
+  gender, using separate mosaic displays like the ``manual'' plot 
   in Figure~\ref{fig:parttable}.}
@@
 single mosaic plot. In contrast to Figures~\ref{fig:parttable} and \ref{fig:cotabplot}, 
-this plot also visualizes the marginal distribution of \code{Sex}.}
+this plot also visualizes the marginal distribution of gender.}
 \label{fig:conditioning}
@@
 shadings, additionally, indicate overall significance. One particular
-shading, the maximum shading, even allows to identify those cells that
-cause the rejection of the null hypothesis.
+shading, the maximum shading 
+\citep{vcd:Meyer+Zeileis+Hornik:2003,vcd:Zeileis+Meyer+Hornik:2005},
+even allows to identify the cells that cause the rejection of the null hypothesis.
 
@@
   alternative 1.
- \item Providing a grapcon \emph{generating} function (`grapcon generator') 
-  taking parameters and returning a function as described in alternative~2.
+ \item Providing a grapcon generator taking parameters and 
+   returning a function as described in alternative~2.
 \end{enumerate}
@@
 The first method is precomputing the graphical parameters ``by
-hand''. We will use `light blue' color for positive and `light salmon'
+hand''. We will use \code{royalblue4} color for positive and \code{mediumorchid4}
 color for negative residuals (see Figure~\ref{fig:binary}):
@@
 expected <- independence_table(ucb)
-(residuals <- (ucb - expected)  / sqrt(expected))
-
-(shading1_obj <- ifelse(residuals > 0, "lightblue", "lightsalmon"))
+(x <- (ucb - expected)  / sqrt(expected))
+
+(shading1_obj <- ifelse(x > 0, "royalblue4", "mediumorchid4"))
 
@@
 
-For implementing alternative 2, we need to create a `shading function'
+For implementing alternative 2, we need to create a ``shading function''
 that computes \class{gpar} objects from
@@
 <<shading2>>=
-shading2_fun <- function(res) gpar(fill = ifelse(res > 0, "lightblue", "lightsalmon"))
+shading2_fun <- function(x) gpar(fill = ifelse(x > 0, "royalblue4", "mediumorchid4"))
 @
@@
 simply extend the argument list to include, e.g., 
-a \code{fill = c("lightblue", "lightsalmon")} argument because
+a \code{fill = c("royalblue4", "mediumorchid4")} argument because
 \codefun{strucplot} will neither know how to handle it, nor let us
@@
 functions are lexically scoped. Thus, the grapcon generator
-returns (`creates') a `parameterized' shading function with the minimal standard interface
+returns (``creates'') a 
+``parameterized'' shading function with the minimal standard interface
 \codefun{strucplot} requires. The following example shows the necessary
@@
 <<shading3>>=
-shading3a_fun <- function(col = c("lightblue", "lightsalmon")) {
-  if (length(col) != 2) stop("Need exactly two colors!")
-  function(res) gpar(fill = ifelse(res > 0, col[1], col[2]))  
+shading3a_fun <- function(col = c("royalblue4", "mediumorchid4")) {
+  col <- rep(col, length.out = 2)
+  function(x) gpar(fill = ifelse(x > 0, col[1], col[2]))  
 }
@@
 
-\noindent In the call to \codefun{mosaic}, using the new
+\noindent The first statement just makes sure that exactly two colors
+are specified. In the call to \codefun{mosaic}, using the new
 \codefun{shading3a\_fun} function, we can now simply change the colors:
@@
 <<shading4>>=
-mosaic(ucb, gp = shading3a_fun(c("red","blue")))
+mosaic(ucb, gp = shading3a_fun(c("royalblue4","mediumorchid4")))
 @
@@
 shading3b_fun <- function(observed = NULL, residuals = NULL, expected = NULL, 
-    df = NULL, col = c("lightblue", "lightsalmon")) {
-  if (length(col) != 2) stop("Need exactly two colors!")
-  function(res) gpar(fill = ifelse(res > 0, col[1], col[2]))  
+    df = NULL, col = c("royalblue4", "mediumorchid4")) {
+  col <- rep(col, length.out = 2)
+  function(x) gpar(fill = ifelse(x > 0, col[1], col[2]))  
 }
@@
 
-In some sense, generating
+Note that in this simple binary shading example, the first four
+parameters are not used. In some sense, generating
 functions for shadings are parameterized both by the user and the
@@
 discussion can be found in][]{vcd:Zeileis+Meyer+Hornik:2005}. 
-Color palettes in HCL space are trickier
-to choose, but preferable to palettes derived 
-from HSV space from a perceptual point of view.
+Color palettes in HCL space are 
+preferable to palettes derived 
+from HSV space from a perceptual point of view. Functions for creating
+palettes (see, e.g., \codefun{diverge\_hcl}) are provided with the package.
 
 In HSV space, colors are specified in three dimensions: Hue,
-Saturation (`colorfulness'), and Value (`lightness', amount of gray). 
+Saturation (``colorfulness''), and Value (``lightness'', amount of gray). 
 These three dimensions are used by \codefun{shading\_hsv} to visualize information 
@@
 As an example, we will visualize the association of hair and eye
-color in the \data{HairEyeColor} data set (see Figure~\ref{fig:haireye})
-
-<<haireye>>=
+color in the \data{HairEyeColor} data set (see
+Figure~\ref{fig:haireye}, top):
+
+<<haireye1>>=
 haireye <- margin.table(HairEyeColor, 1:2)
@@
 
-\begin{figure}[h]
-\begin{center}
-<<haireyefig,fig=TRUE,echo=FALSE,width=7,height=7>>=
-<<haireye>>
-@
-\caption{Shaded residuals in the \data{HairEyeColor} data set---two cut-off points.}
+\noindent As introduced before, the default shading scheme is not
+\codefun{shading\_hsv} but \codefun{shading\_hcl} due to the better
+perceptual characteristics of HCL color space.
+The following example again illustrates the \data{HairEyeColor} data, 
+this time with HCL colors:
+
+<<haireye2>>=
+mosaic(haireye, gp = shading_hcl)
+@ 
+<<haireye3>>=
+mosaic(haireye, gp = shading_hcl, gp_args = list(h = c(130, 43), c = 100, l = c(90, 70)))
+@
+
+\noindent In Figure~\ref{fig:haireye}, the plot
+in the middle depicts the default palette, 
+and the bottom plot an alternative setting for Hue (\code{h}), Chroma (\code{s}), and
+Luminance (\code{v}).
+
+\setkeys{Gin}{width=0.7\textwidth}
+
+\begin{figure}[htbp]
+\begin{center}
+<<haireyefig1,fig=TRUE,echo=FALSE,height=5>>=
+mosaic(haireye, gp = shading_hsv, margin = c(bottom = 1), keep_aspect_ratio = FALSE)
+@
+<<haireyefig2,fig=TRUE,echo=FALSE,height=5>>=
+mosaic(haireye, gp = shading_hcl, margin = c(bottom = 1), keep_aspect_ratio = FALSE)
+@
+<<haireyefig3,fig=TRUE,echo=FALSE,height=5>>=
+mosaic(haireye, gp = shading_hcl, margin = c(bottom = 1), gp_args = list(h = c(130, 43), c = 100, l = c(90, 70)), keep_aspect_ratio = FALSE)
+@
+\caption{Three mosaic plots for the \data{HairEyeColor} data using, from top to bottom:
+the default HSV color palette, the default HCL color palette, 
+and a custom HCL color palette.}
 \label{fig:haireye}
@@
 \noindent Large positive residuals (greater than $4$) can be found for
-brown eyes/black hair and blue eyes/blond hair, and are colored in saturated blue. 
+brown eyes/black hair and blue eyes/blond hair, and are colored in deep blue. 
 On the other hand, there is a large negative residual
@@
 $-4$): the colors for them are less saturated. Residuals between $-2$ and $2$ 
-are shaded in white.
+are shaded in white (gray for HCL-shading).
 The heuristic for choosing the cut-off points $2$ and $4$ is that the Pearson residuals
@@
 be changed to alternative values using the \code{interpolate}
-argument (see Figure~\ref{fig:interpolate}):
+argument (see Figure~\ref{fig:interpolatecontinuous}):
 
 <<interpolate>>=
-mosaic(haireye, gp = shading_hsv, gp_args = list(interpolate = 1:4))
-@
-
-\begin{figure}[h]
-\begin{center}
-<<interpolatefig,fig=TRUE,echo=FALSE,width=7,height=7>>=
-<<interpolate>>
-@
-\caption{Shaded residuals in the \data{HairEyeColor} data set---four cut-off points.}
-\label{fig:interpolate}
-\end{center}
-\end{figure}
+mosaic(haireye, shade = TRUE, gp_args = list(interpolate = 1:4))
+@
 
@@
 function maps residuals greater than 4 to a saturation
-level of 1. However, the resulting plot (Figure~\ref{fig:continuous}) is 
+level of 1. However, the resulting plot
+(Figure~\ref{fig:interpolatecontinuous}, right) is 
 deceiving:
@@
 <<continuous2>>=
-mosaic(HairEyeColor, gp = shading_hsv, gp_args = list(interpolate = ipol),
+mosaic(haireye, shade = TRUE, gp_args = list(interpolate = ipol),
 labeling_args = list(abbreviate = c(Sex = TRUE)))
@@
 
-\begin{figure}[p]
-\begin{center}
-<<continuousfig,fig=TRUE,echo=FALSE,height=7>>=
-<<continuous2>>
-@
-\caption{The \data{HairEyeColor} data with continuous shading.}
-\label{fig:continuous}
-\end{center}
-\end{figure}
+\setkeys{Gin}{width=\textwidth}
+\begin{figure}[htbp]
+\begin{center}
+<<interpolatefig,fig=TRUE,echo=FALSE,width=12,height=5>>=
+pushViewport(viewport(layout = grid.layout(ncol = 2)))
+pushViewport(viewport(layout.pos.col = 1))
+mosaic(haireye, gp_args = list(interpolate = 1:4), margin = c(right = 1), keep_aspect_ratio= FALSE,newpage = FALSE,legend_width=5.5,shade = TRUE)
+popViewport(1)
+pushViewport(viewport(layout.pos.col = 2))
+mosaic(haireye, gp_args = list(interpolate = ipol), margin = c(left=3,right = 1), keep_aspect_ratio = FALSE, newpage = FALSE, shade = TRUE)
+popViewport(2)
+@
+\caption{\label{fig:interpolatecontinuous}The \data{HairEyeColor} data---left: with 4 cut-off points, right: with continuous shading.}
+
+\end{center}
+\end{figure} 
+\setkeys{Gin}{width=0.7\textwidth}
 
@@
 $\chi^2$ test does not reject the null hypothesis of
-independence. Consequently, the colors appear dark (see Figure~\ref{fig:bundesliga}):
+independence. Consequently, the colors appear dark (see
+Figure~\ref{fig:bundesliga}, left):
 
 <<bundesliga>>=
-bl <- xtabs(~ HomeGoals + AwayGoals, data = Bundesliga, subset = Year == 1995)
-mosaic(bl, gp = shading_hsv)
-@
-
-\begin{figure}[p]
-\begin{center}
-<<bundesligafig,fig=TRUE,echo=FALSE,height=7>>=
-<<bundesliga>>
-@
-\caption{Non-significant $\chi^2$ test using part of the \data{Bundesliga} data.}
-\label{fig:bundesliga}
-\end{center}
-\end{figure}
-
-\noindent (Note that the bullets drawn for zero cells are shaded, too).
+BL <- xtabs(~ HomeGoals + AwayGoals, data = Bundesliga, subset = Year == 1995)
+mosaic(BL, shade = TRUE)
+@
+
+\noindent Note that in extended mosaic plots, bullets drawn for zero
+cells are shaded, too, bringing out non-zero residuals, if any.
 
@@
 the \data{Bundesliga} data from above, this time using the Friendly 
-scheme and, in addition, an alternative legend (see Figure~\ref{fig:friendly}):
+scheme and, in addition, an alternative legend (see
+Figure~\ref{fig:bundesliga}, right):
 
 <<friendly>>=
-mosaic(bl, gp = shading_Friendly, legend = legend_fixed, zero_size = 0)
-@
-
-\begin{figure}[h]
-\begin{center}
-<<friendlyfig,fig=TRUE,echo=FALSE,height=7>>=
-<<friendly>>
-@
-\caption{The \data{Bundesliga} data for 1995 using the Friendly
+mosaic(BL, gp = shading_Friendly, legend = legend_fixed, zero_size = 0)
+@
+
+\setkeys{Gin}{width=\textwidth}
+\begin{figure}[htbp]
+\begin{center}
+<<bundesligafig,fig=TRUE,echo=FALSE,height=5,width=12>>=
+pushViewport(viewport(layout = grid.layout(ncol = 2)))
+pushViewport(viewport(layout.pos.col = 1))
+mosaic(BL, margin = c(right = 1), keep_aspect_ratio= FALSE, newpage = FALSE, 
+       legend_width=5.5, shade = TRUE)
+popViewport(1)
+pushViewport(viewport(layout.pos.col = 2))
+mosaic(BL, gp = shading_Friendly, legend = legend_fixed, zero_size = 0, 
+       margin = c(right = 1), keep_aspect_ratio= FALSE, newpage = FALSE,  
+       legend_width=5.5)
+popViewport(2)
+@
+\caption{The \data{Bundesliga} data for 1995. Left: Non-significant
+  $\chi^2$ test. Right: using the Friendly
   shading and a legend with fixed bins.}
-\label{fig:friendly}
-\end{center}
-\end{figure}
+\label{fig:bundesliga}
+\end{center}
+\end{figure}
+\setkeys{Gin}{width=0.7\textwidth}
 
@@
 
-As introduced before, the default shading scheme is not
-\codefun{shading\_hsv} but \codefun{shading\_hcl} due to the better
-perceptual characteristics of HCL color space.
 % Figure~\ref{fig:shadingHSVHCL} depicts 
@@
 % diverging color palettes. 
-The following example again illustrates the \data{HairEyeColor} data, 
-this time with HCL colors (Figure~\ref{fig:hclshading1} depicts the default palette, 
-and Figure~\ref{fig:hclshading2} an alternative setting):
-
 % <<auxiliary_functions,echo=FALSE,results=hide>>=
@@
 
-<<hclshading1>>=
-mosaic(haireye, gp = shading_hcl)
-@ 
-<<hclshading2>>=
-mosaic(haireye, gp = shading_hcl, gp_args = list(h = c(130, 43), c = 100, l = c(90, 70)))
-@
-
-\setkeys{Gin}{width=0.7\textwidth}
-\begin{figure}[p]
-\begin{center}
-<<hclshading1fig,fig=TRUE,echo=FALSE,height=6>>=
-<<hclshading1>>
-@
-\caption{The \data{HairEyeColor} data, using default HCL color palette.}
-\label{fig:hclshading1}
-\end{center}
-\end{figure}
-
-\begin{figure}[p]
-\begin{center}
-<<hclshading2fig,fig=TRUE,echo=FALSE,height=6>>=
-<<hclshading2>>
-@
-\caption{The \data{HairEyeColor} data, using a custom HCL color palette.}
-\label{fig:hclshading2}
-\end{center}
-\end{figure}
-
-A more `advanced' function building upon \codefun{shading\_hcl} 
+A more ``advanced'' function building upon \codefun{shading\_hcl} 
 is \codefun{shading\_max}, using the
@@
 visualize significant \emph{cells} causing the
-rejection of the independence hypothesis \citep{vcd:Meyer+Zeileis+Hornik:2003}. The
+rejection of the independence hypothesis 
+\citep{vcd:Meyer+Zeileis+Hornik:2003,vcd:Zeileis+Meyer+Hornik:2005}. The
 \code{level} argument of \codefun{shading\_max} 
@@
 residual violates the null hypotheses of independence, and thus, the
-`culprits' can immediately be identified.
-
-\section[Flexible labeling]{Flexible labeling}
+``culprits'' can immediately be identified.
+
+\section[Labeling]{Labeling}
 \label{sec:labeling}
@@
 base \proglang{R} is the labeling in the strucplot framework which 
-offers many more features and flexibility. 
+offers more features and greater flexibility. 
 Like shading, spacing, and drawing of legend and core plot, 
@@
 low-level core functions. It only relies on the viewport tree produced by them, and the
-`\texttt{dimnames}' attribute of the visualized table. Labeling
+\code{dimnames} attribute of the visualized table. Labeling
 functions are grapcons that ``add ink to the canvas'': the drawing of the
@@
 
-As an example, consider the \data{Titanic} data, consisting
-of 4 categorical variables: survival (`survived'), gender (i.e., `sex'), age, and
-crew. By default, the variable names and levels are plotted `around' the
-plot in a counter-clockwise way (see Figure~\ref{fig:defaults}):
+As an example, again consider the \data{Titanic} data:
+by default, the variable names and levels are plotted ``around'' the
+plot in a counter-clockwise way (see Figure~\ref{fig:labels1}, top left):
 
@@
 
-\begin{figure}[p]
-\begin{center}
-<<defaultfig,fig=TRUE,echo=FALSE>>=
-<<default>>
-@
-\caption{Mosaic plot for the \data{Titanic} data with default settings
-  for labeling.}
-\label{fig:defaults}
-\end{center}
-\end{figure}
-
-\noindent Note that the last two levels of the `survived' variable do
-overlap, as well as some adult and child labels of the `age' Variable.
-This issue can be addressed in several ways. The `brute force'
-method is to enable clipping for these dimensions (see Figure~\ref{fig:clipping}):
+% \begin{figure}[p]
+% \begin{center}
+% <<defaultfig,fig=TRUE,echo=FALSE>>=
+% <<default>>
+% @
+% \caption{Mosaic plot for the \data{Titanic} data with default settings
+%   for labeling.}
+% \label{fig:defaults}
+% \end{center}
+% \end{figure}
+
+\noindent Note that the last two levels of the \code{survived} variable do
+overlap, as well as some adult and child labels of the \code{age} Variable.
+This issue can be addressed in several ways. The ``brute force''
+method is to enable clipping for these dimensions (see
+Figure~\ref{fig:labels1}, top right):
 
@@
 
-\begin{figure}[p]
-\begin{center}
-<<clippingfig,fig=TRUE,echo=FALSE>>=
-<<clipping>>
-@
-\caption{The effect of clipping.}
-\label{fig:clipping}
-\end{center}
-\end{figure}
+% \begin{figure}[p]
+% \begin{center}
+% <<clippingfig,fig=TRUE,echo=FALSE>>=
+% <<clipping>>
+% @
+% \caption{The effect of clipping.}
+%  \label{fig:clipping}
+% \end{center}
+% \end{figure}
 
@@
 overridden by the named components. Here, the default is \code{FALSE},
-and therefore clipping is enabled only for the `survived' and `age' variables.
+and therefore clipping is enabled only for the \code{survived} and \code{age} variables.
 A more sensible solution to the overlap problem is to abbreviate the 
-levels (see Figure~\ref{fig:abbreviating}):
+levels (see Figure~\ref{fig:labels1}, middle left):
 
@@
 
-\begin{figure}[h]
-\begin{center}
-<<abbreviatingfig,fig=TRUE,echo=FALSE>>=
-<<abbreviating>>
-@
-\caption{Abbreviating.}
-\label{fig:abbreviating}
-\end{center}
-\end{figure}
+% \begin{figure}[h]
+% \begin{center}
+% <<abbreviatingfig,fig=TRUE,echo=FALSE>>=
+% <<abbreviating>>
+% @
+% \caption{Abbreviating.}
+% \label{fig:abbreviating}
+% \end{center}
+% \end{figure}
 
@@
 is performed using the \codefun{abbreviate} function in base \proglang{R}. Another
-possibility is to rotate the levels (see Figure~\ref{fig:rotating}):
+possibility is to rotate the levels (see Figure~\ref{fig:labels1}, bottom):
 
@@
 mosaic(Titanic, labeling_args = list(rot_labels = c(bottom = 90, right = 0), 
-offset_varnames = c(right = 1), offset_labels = c(right = 0.3)), margins = c(right = 4, bottom = 3))
-@
-
-\begin{figure}[p]
-\begin{center}
-<<rotatefig,fig=TRUE,echo=FALSE,height=6>>=
-<<rotate>>
-@
-\caption{Rotating labels.}
-\label{fig:rotating}
-\end{center}
-\end{figure}
+offset_varnames = c(right = 1), offset_labels = c(right = 0.3)), 
+margins = c(right = 4, bottom = 3))
+@
+
+% \begin{figure}[p]
+% \begin{center}
+% <<rotatefig,fig=TRUE,echo=FALSE,height=6>>=
+% <<rotate>>
+% @
+% \caption{Rotating labels.}
+% \label{fig:rotating}
+% \end{center}
+% \end{figure}
 
 \noindent Finally, we could also inhibit the output of repeated levels 
-(see Figure~\ref{fig:repeat}):
+(see Figure~\ref{fig:labels1}, middle right):
 
@@
 
+\setkeys{Gin}{width=\textwidth}
 \begin{figure}[p]
 \begin{center}
-<<repeatfig,fig=TRUE,echo=FALSE,height=6>>=
-<<repeat>>
-@
-\caption{Inhibiting the repetition of levels.}
-\label{fig:repeat}
-\end{center}
-\end{figure}
-
-We now proceed with a few more `cosmetic' features (which do not all
+<<label1fig,fig=TRUE,echo=FALSE,height=12,width=8>>=
+pushViewport(viewport(layout = grid.layout(ncol = 2,nrow=3)))
+pushViewport(viewport(layout.pos.col = 1, layout.pos.row = 1))
+mosaic(Titanic, newpage = FALSE, keep = TRUE, margin = c(right = 3))
+popViewport()
+pushViewport(viewport(layout.pos.col = 2, layout.pos.row = 1))
+mosaic(Titanic, labeling_args = list(clip = c(Survived = TRUE, Age = TRUE)), 
+       newpage = FALSE, keep = TRUE, margin = c(left = 3))
+popViewport()
+pushViewport(viewport(layout.pos.col = 1, layout.pos.row = 2))
+mosaic(Titanic, labeling_args = list(abbreviate = c(Survived = TRUE, Age = 2)), 
+       newpage = FALSE, keep = TRUE, margin = c(right = 3))
+popViewport()
+pushViewport(viewport(layout.pos.col = 2, layout.pos.row = 2))
+mosaic(Titanic, labeling_args = list(rep = c(Survived = FALSE, Age = FALSE)), 
+       newpage = FALSE, keep = TRUE, margin = c(left = 3))
+popViewport()
+pushViewport(viewport(layout.pos.col = 1:2, layout.pos.row = 3))
+pushViewport(viewport(width = 0.55))
+mosaic(Titanic, labeling_args = list(rot_labels = c(bottom = 90, right = 0), 
+                  offset_varnames = c(right = 1), offset_labels = c(right = 0.3)), 
+       margins = c(right = 4, bottom = 3), newpage = FALSE, keep = FALSE)
+popViewport(3)
+@
+\caption{Labeling the Titanic data. Top left: default labeling, top
+  right: clipping, middle left: abbreviating, middle right: inhibiting 
+  the repetition of levels, bottom: rotating.}
+\label{fig:labels1}
+\end{center}
+\end{figure}
+\setkeys{Gin}{width=0.7\textwidth}
+
+We now proceed with a few more ``cosmetic'' features (which do not all
 produce satisfactory results for our sample data).
 A first simple, but effectful modification is to position
-all labels and variables left-aligned:  (see Figure~\ref{fig:left}):
+all labels and variables left-aligned (see Figure~\ref{fig:labels2},
+top left):
 
@@
 
-\begin{figure}[h]
-\begin{center}
-<<leftfig,fig=TRUE,echo=FALSE>>=
-<<left>>
-@
-\caption{Left-aligning.}
-\label{fig:left}
-\end{center}
-\end{figure}
+% \begin{figure}[h]
+% \begin{center}
+% <<leftfig,fig=TRUE,echo=FALSE>>=
+% <<left>>
+% @
+% \caption{Left-aligning.}
+% \label{fig:left}
+% \end{center}
+% \end{figure}
 
@@
 bottom and right margins, and all variable names to the top and left
-margins (see Figure~\ref{fig:margins}):
+margins (see Figure~\ref{fig:labels2}, top right):
 
@@
 
-\begin{figure}[p]
-\begin{center}
-<<marginsfig,fig=TRUE,echo=FALSE,height=6>>=
-<<margins>>
-@
-\caption{Changes in the margins.}
-\label{fig:margins}
-\end{center}
-\end{figure}
+% \begin{figure}[p]
+% \begin{center}
+% <<marginsfig,fig=TRUE,echo=FALSE,height=6>>=
+% <<margins>>
+% @
+% \caption{Changes in the margins.}
+% \label{fig:margins}
+% \end{center}
+% \end{figure}
 
@@
 Now, we will add boxes to the labels and additionally
-enable clipping (see Figure~\ref{fig:boxes}):
+enable clipping (see Figure~\ref{fig:labels2}, bottom left):
 
@@
 
-\begin{figure}[p]
-\begin{center}
-<<boxesfig,fig=TRUE,echo=FALSE,height=6>>=
-<<boxes>>
-@
-\caption{Boxes and Clipping.}
-\label{fig:boxes}
-\end{center}
-\end{figure}
+% \begin{figure}[p]
+% \begin{center}
+% <<boxesfig,fig=TRUE,echo=FALSE,height=6>>=
+% <<boxes>>
+% @
+% \caption{Boxes and Clipping.}
+% \label{fig:boxes}
+% \end{center}
+% \end{figure}
 
@@
 \noindent Another variant is to put the variable names into the
-same line as the levels (see Figure~\ref{fig:labbl}):
+same line as the levels (see Figure~\ref{fig:labels2}, bottom right):
 
@@
 
-\begin{figure}[h]
-\begin{center}
-<<labblfig,fig=TRUE,echo=FALSE>>=
-<<labbl>>
-@
-\caption{Variable names beneath levels.}
-\label{fig:labbl}
-\end{center}
-\end{figure}
+% \begin{figure}[h]
+% \begin{center}
+% <<labblfig,fig=TRUE,echo=FALSE>>=
+% <<labbl>>
+% @
+% \caption{Variable names beneath levels, and clipping disabled for the
+%   survival variable.}
+% \label{fig:labbl}
+% \end{center}
+% \end{figure}
 
@@
 
+\setkeys{Gin}{width=\textwidth}
+\begin{figure}[p]
+\begin{center}
+<<label2fig,fig=TRUE,echo=FALSE,height=8,width=8>>=
+pushViewport(viewport(layout = grid.layout(ncol = 2, nrow = 2)))
+pushViewport(viewport(layout.pos.col = 1, layout.pos.row = 1))
+mosaic(Titanic, labeling_args = list(pos_varnames = "left", pos_labels = "left", 
+       just_labels = "left", rep = FALSE), newpage = FALSE, keep = TRUE)
+popViewport()
+pushViewport(viewport(layout.pos.col = 2, layout.pos.row = 1))
+mosaic(Titanic, labeling_args = list(tl_labels = FALSE, tl_varnames = TRUE, 
+                  abbreviate = c(Survived = 1, Age = 3)), newpage = FALSE, keep = TRUE)
+popViewport()
+pushViewport(viewport(layout.pos.col = 1, layout.pos.row = 2))
+mosaic(Titanic, labeling_args = list(tl_labels = FALSE, tl_varnames = TRUE, 
+       boxes = TRUE, clip = TRUE), newpage = FALSE, keep = TRUE)
+popViewport()
+pushViewport(viewport(layout.pos.col = 2, layout.pos.row = 2))
+mosaic(Titanic, labeling_args = list(tl_labels = TRUE, boxes = TRUE,
+       clip = c(Survived = FALSE, TRUE), 
+       labbl_varnames = TRUE), margins = c(left = 4, right = 1, 3), 
+       newpage = FALSE, keep = TRUE)
+popViewport(2)
+@
+\caption{Labeling the Titanic data, advanced features. Top left: left aligning, top
+  right: changes in the margins, bottom left: boxes and clipping, bottom right: 
+  Variable names beneath levels, and clipping disabled for the survival variable.}
+\label{fig:labels2}
+\end{center}
+\end{figure}
+\setkeys{Gin}{width=0.7\textwidth}
+
 \subsection[Labels in the cells]{Labels in the cells: \texttt{labeling\_cells()}}
@@
 cells. As an example, we use the \data{PreSex} data on pre- and
-extramarital sex and divorce (see Figure~\ref{fig:cell}):
+extramarital sex and divorce (see Figure~\ref{fig:labels3}, top left):
 
@@
 
-\begin{figure}[h]
-\begin{center}
-<<cellfig,fig=TRUE,echo=FALSE,width=7,height=7>>=
-<<cell>>
-@
-\caption{Cell labeling for the \data{PreSex} data.}
-\label{fig:cell}
-\end{center}
-\end{figure}
+% \begin{figure}[h]
+% \begin{center}
+% <<cellfig,fig=TRUE,echo=FALSE,width=7,height=7>>=
+% <<cell>>
+% @
+% \caption{Cell labeling for the \data{PreSex} data.}
+% \label{fig:cell}
+% \end{center}
+% \end{figure}
 
 \noindent In the case of narrow cells, it might be useful to
-abbreviate labels and/or variable names and turn off clipping (see Figure~\ref{fig:cell2}):
+abbreviate labels and/or variable names and turn off clipping (see
+Figure~\ref{fig:labels3}, top right):
 
@@
 
-\begin{figure}[h]
-\begin{center}
-<<cell2fig,fig=TRUE,echo=FALSE>>=
-<<cell2>>
-@
-\caption{Cell labeling for the \data{PreSex} data, labels abbreviated.}
-\label{fig:cell2}
-\end{center}
-\end{figure}
+% \begin{figure}[h]
+% \begin{center}
+% <<cell2fig,fig=TRUE,echo=FALSE>>=
+% <<cell2>>
+% @
+% \caption{Cell labeling for the \data{PreSex} data, labels abbreviated.}
+% \label{fig:cell2}
+% \end{center}
+% \end{figure}
 
@@
 labeling with border labeling as done by \codefun{labels\_conditional}
-(see Figure~\ref{fig:conditional}):
+(see Figure~\ref{fig:labels3}, bottom left):
 
 <<conditional>>=
-mosaic(~ PremaritalSex + ExtramaritalSex | MaritalStatus + Gender, data = PreSex, labeling = labeling_conditional(abbreviate_varnames = TRUE, abbreviate_labels = TRUE, clip = FALSE))
-@
-
-\begin{figure}[h]
-\begin{center}
-<<conditionalfig,fig=TRUE,echo=FALSE>>=
-<<conditional>>
-@
-\caption{Conditional labeling for the \data{PreSex}, labels abbreviated.}
-\label{fig:conditional}
-\end{center}
-\end{figure}
+mosaic(~ PremaritalSex + ExtramaritalSex | MaritalStatus + Gender, data = PreSex, labeling = labeling_conditional(abbreviate_varnames = TRUE, abbreviate_labels = TRUE, clip = FALSE, gp_text = gpar(col = "red")))
+@
+
+% \begin{figure}[h]
+% \begin{center}
+% <<conditionalfig,fig=TRUE,echo=FALSE>>=
+% <<conditional>>
+% @
+% \caption{Conditional labeling for the \data{PreSex} data, labels (in
+%   red for clarity) abbreviated.}
+% \label{fig:conditional}
+% \end{center}
+% \end{figure}
 
@@
 arbitrary text to the cells by supplying a character array in the same
-shape than the data array to the \code{text} argument (cells with missing values
-are ignored). In the following example using the `Titanic' data, 
+shape as the data array to the \code{text} argument (cells with missing values
+are ignored). In the following example using the \code{Titanic} data, 
 this is used to add all observed values greater
-than 5 to the cells after the mosaic has been plotted (see Figure~\ref{fig:text}):
+than 5 to the cells after the mosaic has been plotted (see
+Figure~\ref{fig:labels3}, bottom right):
 
@@
 
-\begin{figure}[h]
-\begin{center}
-<<textfig,fig=TRUE,echo=FALSE>>=
-<<text>>
-@
-\caption{User-supplied Text added to a mosaic display of the \data{Titanic} data.}
-\label{fig:text}
-\end{center}
-\end{figure}
+% \begin{figure}[h]
+% \begin{center}
+% <<textfig,fig=TRUE,echo=FALSE>>=
+% <<text>>
+% @
+% \caption{User-supplied text (observed frequencies exceeding 5) 
+% added to a mosaic display of the \data{Titanic} data.}
+% \label{fig:text}
+% \end{center}
+% \end{figure}
+
+\setkeys{Gin}{width=\textwidth}
+\begin{figure}[p]
+\begin{center}
+<<label3fig,fig=TRUE,echo=FALSE,height=8,width=8>>=
+grid.newpage()
+pushViewport(viewport(layout = grid.layout(ncol = 2, nrow = 2)))
+pushViewport(viewport(layout.pos.col = 1, layout.pos.row = 1))
+mosaic(~ MaritalStatus + Gender, data = PreSex, labeling = labeling_cells, newpage = FALSE, keep = TRUE)
+popViewport()
+pushViewport(viewport(layout.pos.col = 2, layout.pos.row = 1))
+mosaic(~ PremaritalSex + ExtramaritalSex, data = PreSex, 
+       labeling = labeling_cells(abbreviate_labels = TRUE, abbreviate_varnames = TRUE, 
+         clip = FALSE), newpage = FALSE, keep = TRUE)
+popViewport()
+pushViewport(viewport(layout.pos.col = 1, layout.pos.row = 2))
+mosaic(~ PremaritalSex + ExtramaritalSex | MaritalStatus + Gender, data = PreSex, 
+       labeling = labeling_conditional(abbreviate_varnames = TRUE, 
+         abbreviate_labels = TRUE, clip = FALSE, gp_text = gpar(col = "red")), 
+       newpage = FALSE, keep = TRUE)
+popViewport()
+pushViewport(viewport(layout.pos.col = 2, layout.pos.row = 2))
+mosaic(Titanic, labeling_args = list(abbreviate = c(Survived = 1, Age = 4)), 
+       pop = FALSE, newpage = FALSE, keep = TRUE)
+tab <- ifelse(Titanic < 6, NA, Titanic)
+labeling_cells(text = tab, clip = FALSE)(Titanic)
+@
+\caption{Cell labeling. Top left: default labeling using the
+  \data{PreSex} data, top right: abbreviated labels, bottom left:
+  conditional labeling (labels abbreviated and in red for clarity),
+  bottom right: user-supplied text (observed frequencies exceeding 5) 
+  added to a mosaic display of the \data{Titanic} data.}
+\label{fig:labels3}
+\end{center}
+\end{figure}
+\setkeys{Gin}{width=0.7\textwidth}
 
@@
 
-\section{Customizable spacing}
+\section{Spacing}
 \label{sec:spacing}
@@
 the medicament in the placebo and treatment groups, a mosaic plot is
-certainly appropriate to visualize the three levels of `Improved' in
-the two `Treatment' strata. Another conceptual approach is to use
+certainly appropriate to visualize the three levels of \code{Improved} in
+the two \code{Treatment} strata. Another conceptual approach is to use
 spine plots with highlighting \citep{vcd:hummel:1996}. A spine plot is
@@
 variable is used.
-In the \data{Arthritis} example, we will highlight patients with `Marked' 
+In the \data{Arthritis} example, we will highlight patients with \code{Marked} 
 improvement in both groups. To obtain such a display within 
-the strucplot framework, it suffices to set the space between the `Improved' tiles to 0 (see
-Figure~\ref{fig:artspine}):
+the strucplot framework, it suffices to set the space between the \code{Improved}
+tiles to 0 (see Figure~\ref{fig:artspine}):
 
@@
 (my_spacing <- list(unit(0.5, "lines"), unit(c(0, 0), "lines")))
-my_colors <- gpar(fill = c("lightgray", "lightgray", "black"))
-mosaic(art, spacing = my_spacing, gp = my_colors)
+my_colors <- c("lightgray", "lightgray", "black")
+mosaic(art, spacing = my_spacing, gp = gpar(fill = my_colors, col = my_colors))
 @
@@
 \noindent The strucplot framework also provides a set of spacing grapcon
-generators that compute suitable spacing objects for
+generators which compute suitable spacing objects for
 typical applications. The simplest spacing is \codefun{spacing\_equal}
-that uses the same space between all tiles:
+that uses the same space between all tiles (see
+Figure~\ref{fig:spacing}, top left):
 
@@
 two-dimensional tables. Slightly more flexible is \codefun{spacing\_dimequal} that
-allows an individual setting for each dimension:
+allows an individual setting for each dimension (see
+Figure~\ref{fig:spacing}, top right):
 
@@
 \noindent The default for multi-way contingency tables is
-\codefun{spacing\_increase} that uses increasing spaces for the dimensions. The user can specify a start value and the increase factor:
+\codefun{spacing\_increase} which uses increasing spaces for the
+dimensions. The user can specify a start value and the increase factor
+(see Figure~\ref{fig:spacing}, bottom left):
 
@@
 \noindent For the arthritis example above, we could as well have used
-\codefun{spacing\_highlighting} that is similar to \codefun{spacing\_increase} but
-sets the spacing in the last splitting dimension to 0:
+\codefun{spacing\_highlighting} which is similar to \codefun{spacing\_increase} but
+sets the spacing in the last splitting dimension to 0 (see
+Figure~\ref{fig:spacing}, bottom right):
 
@@
 distinguish the conditioning variables 
-(`Gender' and `MaritalStatus') from the conditioned 
-variables (`PremaritalSex' and `ExtramaritalSex'). This
+(\code{Gender} and \code{MaritalStatus}) from the conditioned 
+variables (\code{PremaritalSex} and \code{ExtramaritalSex}). This
 spacing is the default when conditional variables are specified for a
@@
 
-\section{Example: the Ovarian Cancer Data}
+\setkeys{Gin}{width=\textwidth}
+\begin{figure}[p]
+\begin{center}
+<<spacingfig,fig=TRUE,echo=FALSE,height=8,width=8>>=
+pushViewport(viewport(layout = grid.layout(ncol = 2, nrow = 2)))
+pushViewport(viewport(layout.pos.col = 1, layout.pos.row = 1))
+mosaic(art, spacing = spacing_equal(unit(2, "lines")), keep = TRUE, newpage = FALSE)
+popViewport()
+pushViewport(viewport(layout.pos.col = 2, layout.pos.row = 1))
+mosaic(art, spacing = spacing_dimequal(unit(c(2,0.5), "lines")), keep = TRUE, newpage = FALSE)
+popViewport()
+pushViewport(viewport(layout.pos.col = 1, layout.pos.row = 2))
+mosaic(art, spacing = spacing_increase(start = unit(0.5, "lines"), rate = 1.5), keep = TRUE, newpage = FALSE)
+popViewport()
+pushViewport(viewport(layout.pos.col = 2, layout.pos.row = 2))
+mosaic(art, spacing = spacing_highlighting, keep = TRUE, newpage = FALSE)
+popViewport(2)
+@
+\caption{Varying spacing for the Arthritis  data. Top left: equal
+  spacing for all dimensions, top right: different spacings for
+  individial dimensions, bottom left: increasing spacing, bottom right: 
+  spacing used for highlighting.}
+\label{fig:spacing}
+\end{center}
+\end{figure}
+\setkeys{Gin}{width=0.7\textwidth}
+
+\section{Example: Ovarian cancer survival}
 \label{sec:example}
@@
 In the following, we demonstrate some of the described techniques in
-analyzing the Ovarian Cancer data set.
-The data origins from \citep{vcd:obel:1975} and is 
-about a retrospective study of ovary cancer
+analyzing a data set originating from \citep{vcd:obel:1975}
+\cite[taken from][]{vcd:andersen:1991} about a retrospective study of ovary cancer
 carried out in 1973.  Information was obtained from 299 women, who
 were operated for ovary cancer 10 years before. The data consists of
-four binary variables: the `stage' of the cancer at the time of
-operation (levels: early, advanced), the type of `operation' performed
-(radical, limited), the `survival' status after 10 years (yes, no),
-and `xray' indicating whether X-ray treatment was received (yes,
-no).
+four binary variables: the \code{stage} of the cancer at the time of
+operation (levels: \code{early}, \code{advanced}), the type of \code{operation} performed
+(\code{radical}, \code{limited}), the \code{survival} status after 10 years (\code{yes}, 
+\code{no}), and \code{xray} indicating whether X-ray treatment was received (\code{yes},
+\code{no}).
 
@@
 
-\noindent A `flattened' textual representation can be obtained using \codefun{structable}:
+\noindent A ``flattened'' textual representation can be obtained using \codefun{structable}:
 
@@
 \begin{center}
-<<ocpairs,echo=FALSE,fig=TRUE>>=
+<<ocpairs,echo=FALSE,fig=TRUE,width=8,height=8>>=
 <<oc3>>
@@
 158 cases were in an advanced and 141 in an early stage. Most patients (251, 84\%) 
-were treated with a radical operation, and 186 (62\%) got X-ray treatment. Next, we inspect
-the two-way interaction of the influencing factors (`stage', `operation', and `xray'): the
+were treated with a radical operation, and 186 (62\%) were submitted to
+X-ray treatment. Next, we inspect
+the two-way interaction of the influencing factors (\code{stage}, \code{operation}, 
+and \code{xray}): the
 corresponding mosaics exhibit symmetric, regular shapes with aligned tiles, 
 which indicate no marginal interaction between these variables. 
-The same is true for the interactions of `survival'
-with `operation' and `xray', respectively. Only the stage seems to influence
+The same is true for the interactions of \code{survival}
+with \code{operation} and \code{xray}, respectively. Only the stage seems to influence
 survival: here, the tiles are ``shifted''.
 
-A different view on the data can be obtained using a doubledecker plot, explaining
-survival by the three other factors (Figure~\ref{fig:ocdoubledecker}):
+A different view on the data, focused on the influence of the
+explanatory variables on \code{Survival},
+can be obtained using a doubledecker plot (Figure~\ref{fig:ocdoubledecker}):
 
@@
 \begin{center}
-<<ocdoubledecker,echo=FALSE,fig=TRUE>>=
+<<ocdoubledecker,echo=FALSE,fig=TRUE,width=8>>=
 <<oc4>>
@@
 \caption{Doubledecker plot for the \data{OvaryCancer} data showing the
- conditional distribution of `xray', given `operation', given `stage', and with
- `survival' highlighted.}
+ conditional distribution of X-ray, given operation, given stage, and with
+ survival highlighted.}
 \label{fig:ocdoubledecker}
@@
 \noindent From a technical point of view, the display is constructed as a mosaic plot
-showing the conditional distribution of `survival', given `xray', given `operation', 
-given `stage', with vertical splits for the conditioning variables and horizontal ones for
-`survival'. Additionally, there is zero space between the tiles of the last dimension and
+showing the conditional distribution of \code{survival}, given \code{xray}, 
+given \code{operation}, given \code{stage}, 
+with vertical splits for the conditioning variables and horizontal ones for
+\code{survival}. 
+Additionally, there is zero space between the tiles of the last dimension and
 a binary shading is used for survived and deceased patients.
 Conceptually, this plot is interpreted as a mosaic plot of just the influencing
-variables, with `survival' highlighted in the tiles. Thus, the plot really shows
-the influence of the ``explanatory'' variables on `survival'. Clearly, the survival
+variables, with \code{survival} highlighted in the tiles. Thus, the plot really shows
+the influence of the explanatory variables on \code{survival}. Clearly, the survival
 rate is higher among patients in an early stage, but neither radical operation 
 nor X-ray treatment seem to improve the situation. From this exploratory
-phase, it even seems that for patients in an
-early stage with a limited operation, no X-ray treatment would be better, 
-and that for patients in an advanced stage, limited in contrast to radical 
-operation would improve the chance of survival. %'
-
-To visualize inference results, we can make use of residual-based shadings. First, we
-again use the pairs plot display, this time using color to bring out departures from
-the hypothesized marginal independence models (Figure~\ref{fig:ocpairsshaded}):
-
-<<oc5>>=
-pairs(tab, diag_panel_args = dpa, gp = shading_max)
+phase, the survival rate seems to be slightly higher for patients who
+received a limited operation only, whereas the effect for X-ray
+treatment is less marked.
+
+To visualize inference results, we can make use of residual-based
+shadings, investigating log-linear models for the four-way table. 
+Figure~\ref{fig:ocmosaicnull} visualizes the null model, where
+survival is independent from the combined effect of operation, X-ray
+treatment, and stage:
+
+<<oc6>>=
+split <- c(TRUE, TRUE, TRUE, FALSE)
+mosaic(tab, expected = ~ survival + operation * xray * stage, split_vertical = split)
 @
@@
 \begin{center}
-<<ocpairsshaded,echo=FALSE,fig=TRUE>>=
-<<oc5>>
-@ 
-\caption{Pairs plot for the \data{OvaryCancer} data showing mosaic displays for all pairwise distributions and bar plots for all marginal distributions. The colored cells show the
-violation of the independence hypothesis for `stage' and `survival'.}
-\label{fig:ocpairsshaded}
-\end{center}
-\end{figure}
-
-\noindent Only the mosaics showing the interaction of `stage' and `survival' have 
-colored cells, indicating that the survival rate of patients in an early stage is
-higher than would be expected under (marginal) independence. The hypotheses of
-independence are not rejected for all other marginal interactions. Next, we investigate
-log-linear models for the four-way table. The default model used by \codefun{mosaic} is
-the total independence model, visualized in Figure~\ref{fig:ocmosaictotal}:
-
-<<oc6>>=
-mosaic(tab, shade = TRUE)
-@
-
-\begin{figure}[h]
-\begin{center}
-<<ocmosaictotal,echo=FALSE,fig=TRUE>>=
+<<ocmosaicnull,echo=FALSE,fig=TRUE,width=10>>=
 <<oc6>>
@@
 \caption{Mosaic plot for the \data{OvaryCancer} data, with residual-based shading
-for the (clearly rejected) default hypothesis of total independence of all variables.}
-\label{fig:ocmosaictotal}
-\end{center}
-\end{figure}
-
-\noindent The model is clearly rejected (P-value: 0.000). From the exploratory phase of
-our analysis, we (only) suspect `stage' to be influential on the survival rate. A 
-corresponding hypothesis is that `survival' be independent of `xray' and `operation', given
-`stage'. The model is specified using the \texttt{expected} argument, 
+for the (clearly rejected) null model (survival)(operation, X-ray, stage).}
+\label{fig:ocmosaicnull}
+\end{center}
+\end{figure}
+
+\noindent The model is clearly rejected ($p$-value: 0.000). From the exploratory phase of
+our analysis, we (only) suspect \code{stage} to be influential on the survival rate. A 
+corresponding hypothesis is that \code{survival} be independent of \code{xray} 
+and \code{operation}, given
+\code{stage}. The model is specified using the \texttt{expected} argument, 
 either using the \codefun{loglin} interface or 
@@
 <<oc7>>=
-mosaic(tab, expected = ~ (survival + operation * xray) * stage)
+mosaic(tab, expected = ~ (survival + operation * xray) * stage, split_vertical = split)
 @
@@
 \begin{center}
-<<ocmosaicstage,echo=FALSE,fig=TRUE>>=
+<<ocmosaicstage,echo=FALSE,fig=TRUE,width=10>>=
 <<oc7>>
@@
 \caption{Mosaic plot for the \data{OvaryCancer} data, with residual-based shading
-for the hypothesis of `survival' being independent of `xray' and `operation', given
-`stage'. The hypothesis is not rejected.}
+for the hypothesis of survival being independent of X-ray and operation, given
+stage. The hypothesis is not rejected.}
 \label{fig:ocmosaicstage}
@@
 
-In this paper, we describe the `strucplot' framework for the visualization of
+In this paper, we describe the ``strucplot'' framework for the visualization of
 multi-way contingency tables. Strucplot displays include popular 
 basic plots such as mosaic, association, and sieve plots, integrated 
-in a unified framework: all can be seen as graphical visualizations of
+in a unified framework: all can be seen as visualizations of
 hierarchical conditional flat tables. 
@@
 shading, labeling, spacing, and legend, by means of graphical
-appearance control (`grapcon') functions. These `graphical
-hyperparameters' are customized and created by
+appearance control (``grapcon'') functions. These ``graphical
+hyperparameters'' are customized and created by
 generating functions. Our work includes a set of predefined grapcon generators for
@@
 \begin{description}
-\item[\texttt{Arthritis}] Data from \cite{vcd:Koch+Edwards:1988} from a 
+\item[\texttt{Arthritis}] Data  from a 
   double-blind clinical trial
-  investigating a new treatment for rheumatoid arthritis. Package:
-  \pkg{vcd}.
+  investigating a new treatment for rheumatoid arthritis. 
+  Source: \cite{vcd:Koch+Edwards:1988}. Taken from: \cite{vcd:Friendly:2000} 
+  Package: \pkg{vcd}.
 \item[\texttt{Bundesliga}] Results from the first German soccer league in the
-  years 1995/6 \citep{vcd:Knorr-Held:1999} and 2001/2 (Achim
+  years 1995/6 \citep{vcd:Knorr-Held:1999} and 2001/2 (Collected by: Achim
   Zeileis). Package: \pkg{vcd}.
@@
   students. The gender information is artificial. Source: \cite{vcd:Snee:1974}.
+  Taken from: \cite{vcd:Friendly:2000}. 
   Package: \pkg{datasets} (included in base \proglang{R}).
@@
      were operated for ovary cancer 10 years before. Source:
-     \cite{vcd:obel:1975}. Package: \pkg{vcd}.
-\item[\texttt{PreSex}]  Data from \cite{vcd:thornes+collard:1979} reported in 
-  \cite{vcd:gilbert:1981}, on pre- and extra-marital sex and divorce. Package:
-  \pkg{vcd}.
+     \cite{vcd:obel:1975}. Taken fromn: \cite{vcd:andersen:1991}. Package: \pkg{vcd}.
+\item[\texttt{PreSex}]  Data on pre- and extra-marital sex and divorce. 
+  Source: \cite{vcd:thornes+collard:1979}. Taken from \cite{vcd:gilbert:1981}.
+  Package: \pkg{vcd}.
 \item[\texttt{Titanic}] Information on the fate of passengers on
-  the fatal maiden voyage of the ocean liner `Titanic', summarized
-  according to economic status (class), gender (`Sex'), age and
-  survival.  Source: \cite{vcd:dawson:1995} (Data 
-  originally collected by the British Board of Trade
-  in their investigation of the sinking). 
+  the fatal maiden voyage of the ocean liner ``Titanic'', summarized
+  according to economic status (class), gender (\code{Sex}), age and
+  survival. Data originally collected by the British Board of Trade
+  in their investigation of the sinking. Taken from: \cite{vcd:dawson:1995}.
   Package: \pkg{datasets} (included in base \proglang{R}).
@@
   gender. Source: \cite{vcd:Bickel+Hammel+O'Connell:1975}. 
+  Taken from: \cite{vcd:Friendly:2000}.
   Package: \pkg{datasets} (included in base \proglang{R}).
@@
 \begin{Sinput}
-cotabplot(UCBAdmissions, panel = cotab_assoc, shade = TRUE, legend = FALSE)
+cotabplot(UCBAdmissions, panel = cotab_indep, shade = TRUE, legend = FALSE)
 \end{Sinput}
@@
 \begin{Sinput}
-mosaic(PreSex, condvars = c(1, 4), shade = TRUE)
+presextest <- coindep_test(PreSex, margin = c(1,4), 
+                           indepfun = function(x) sum(x^2), n = 5000)
+mosaic(PreSex, condvars = c(1, 4), shade = TRUE, 
+       gp_args = list(p.value = presextest$p.value)) %$
 \end{Sinput}

### 2006-10-08 trunk/vcd/paper/Strucplot/JSS.Rnw
third bunch of changes
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@583 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
   David Meyer\\
-  Department of Information Systems and Process Management\\
+  Department of Information Systems and Operations\\
   E-mail: \email{David.Meyer@wu-wien.ac.at}\\
@@
     Basic functionality      &  $\times$  & $\times$  & $\times$ &$\times$ &$\times$ &   $\times$   & $\times$& $\times$\\
-    Shape                    &            &           &          &         &$\times$ & $\times$  &  $\times$&\\
+    Shape                    &            &           &          &$\times$ && $\times$  &  $\times$&\\
     Res.-based shadings  &  $\times$  &        & $\times$ & $\times$ & ($\times$) &     &($\times$)& ($\times$)\\
@@
     Interaction              &     &        &   &  &$\times$  &   $\times$   & $\times$& $\times$\\
-    Linking                  &     &        &   &  &$\times$  &       & $\times$& $\times$\\
+    Linking                  &     &        &   &  &$\times$  & $\times$ & $\times$& $\times$\\
     Extensible design        &     &        &  &$\times$ & &       &  &\\
     Language                 & \proglang{SAS} & \proglang{S} &
-    \proglang{S} & \proglang{S} & \proglang{Java} & \proglang{XLisp} & \proglang{C++} & \proglang{Java}\\
+    \proglang{R} & \proglang{R} & \proglang{R}/\proglang{Java} & \proglang{XLisp} & \proglang{C++} & \proglang{Java}\\
     \hline
@@
 follow the naming convention \code{\textit{group\_foo}()},  
-where (\code{\textit{group}}) reflects the group the
+where \code{\textit{group}} reflects the group the
 generators belong to (strucplot core, labeling,
@@
 
-\noindent Alternatively, we can directly inspect the residuals. 
+\noindent where \code{spacing\_dimequal} is used to set the spacing
+of the second and third dimension to zero. Alternatively, we can directly inspect the residuals. 
 The Pearson residuals (standardized deviations of observed from expected values) 
@@
 \code{split\_vertical} argument. The default is to alternate splits starting
-with a horizontal one (see Figure~\ref{fig:split}).
+with a horizontal one (see Figure~\ref{fig:split}):
 
@@
 following examples, note that compared to subsetting, the first
-dimension(s) have been dropped:
+dimension(s) are dropped:
 
@@
 This spacing uses equal space between tiles of conditioned variables,
-and increasing space between tiles of conditioning variables.
+and increasing space between tiles of conditioning variables (See 
+Section~\ref{sec:spacing}).
 
@@
 the \codefun{pairs} methods create pairwise views of the table. 
-The function produces, by default, a plot matrix 
+They produce, by default, a plot matrix 
 having strucplot displays in the off-diagonal panels, and
@@
 \begin{enumerate}
-\item recursively, create a tree of nested viewports
-\item push the whole tree, i.e., add it to the system tree maintained
-  by the \pkg{grid} engine
-\item one by one, seek all cells of the strucplot display and add
-  plot-specific content (rectangles for mosaics, bars for association
-  plots, etc.)
+\item recursively, split the table until the individual cells are reached
+\item during the splits, add viewports to the plot
+\item for the individual cells, add plot-specific content 
+      (rectangles for mosaics, bars for association plots, etc.)
 \end{enumerate}
@@
 from HSV space from a perceptual point of view. Functions for creating
-palettes (see, e.g., \codefun{diverge\_hcl}) are provided with the package.
+palettes (see, e.g., \codefun{diverge\_hcl}) are provided with the
+\pkg{vcd} package.
 
@@
 in the middle depicts the default palette, 
-and the bottom plot an alternative setting for Hue (\code{h}), Chroma (\code{s}), and
-Luminance (\code{v}).
+and the bottom plot an alternative setting for Hue (\code{h}), Chroma (\code{c}), and
+Luminance (\code{l}).
 
@@
 \begin{center}
-<<haireyefig1,fig=TRUE,echo=FALSE,height=5>>=
+<<haireyefig1,fig=TRUE,echo=FALSE,height=3,width=5>>=
 mosaic(haireye, gp = shading_hsv, margin = c(bottom = 1), keep_aspect_ratio = FALSE)
 @
-<<haireyefig2,fig=TRUE,echo=FALSE,height=5>>=
+<<haireyefig2,fig=TRUE,echo=FALSE,height=3,width=5>>=
 mosaic(haireye, gp = shading_hcl, margin = c(bottom = 1), keep_aspect_ratio = FALSE)
 @
-<<haireyefig3,fig=TRUE,echo=FALSE,height=5>>=
+<<haireyefig3,fig=TRUE,echo=FALSE,height=3,width=5>>=
 mosaic(haireye, gp = shading_hcl, margin = c(bottom = 1), gp_args = list(h = c(130, 43), c = 100, l = c(90, 70)), keep_aspect_ratio = FALSE)
@@
 \begin{center}
-<<label1fig,fig=TRUE,echo=FALSE,height=12,width=8>>=
+<<label1fig,fig=TRUE,echo=FALSE,height=10,width=8>>=
 pushViewport(viewport(layout = grid.layout(ncol = 2,nrow=3)))
 pushViewport(viewport(layout.pos.col = 1, layout.pos.row = 1))
-mosaic(Titanic, newpage = FALSE, keep = TRUE, margin = c(right = 3))
+mosaic(Titanic, newpage = FALSE, keep = TRUE, margin = c(right = 3), 
+       gp_labels = gpar(fontsize = 10))
 popViewport()
@@
 mosaic(Titanic, labeling_args = list(clip = c(Survived = TRUE, Age = TRUE)), 
-       newpage = FALSE, keep = TRUE, margin = c(left = 3))
+       newpage = FALSE, keep = TRUE, margin = c(left = 3), gp_labels = gpar(fontsize = 10))
 popViewport()
@@
 mosaic(Titanic, labeling_args = list(abbreviate = c(Survived = TRUE, Age = 2)), 
-       newpage = FALSE, keep = TRUE, margin = c(right = 3))
+       newpage = FALSE, keep = TRUE, margin = c(right = 3), gp_labels = gpar(fontsize = 10))
 popViewport()
@@
 mosaic(Titanic, labeling_args = list(rep = c(Survived = FALSE, Age = FALSE)), 
-       newpage = FALSE, keep = TRUE, margin = c(left = 3))
+       newpage = FALSE, keep = TRUE, margin = c(left = 3), gp_labels = gpar(fontsize = 10))
 popViewport()
 pushViewport(viewport(layout.pos.col = 1:2, layout.pos.row = 3))
-pushViewport(viewport(width = 0.55))
+pushViewport(viewport(width = 0.5))
 mosaic(Titanic, labeling_args = list(rot_labels = c(bottom = 90, right = 0), 
                   offset_varnames = c(right = 1), offset_labels = c(right = 0.3)), 
-       margins = c(right = 4, bottom = 3), newpage = FALSE, keep = FALSE)
+       margins = c(right = 4, bottom = 3), newpage = FALSE, keep = FALSE, gp_labels = gpar(fontsize = 10))
 popViewport(3)
 @
-\caption{Labeling the Titanic data. Top left: default labeling, top
+\caption{Labeling the Titanic data mosaic. Top left: default labeling, top
   right: clipping, middle left: abbreviating, middle right: inhibiting 
@@
 
-\noindent The \var{tl\_foo} (``top left'') arguments are \code{TRUE} by default. 
+\noindent The tl\_\var{foo} (``top left'') arguments are \code{TRUE} by default. 
 Now, we will add boxes to the labels and additionally
@@
 mosaic(Titanic, labeling_args = list(pos_varnames = "left", pos_labels = "left", 
-       just_labels = "left", rep = FALSE), newpage = FALSE, keep = TRUE)
+       just_labels = "left", rep = FALSE), newpage = FALSE, keep = TRUE,
+       gp_labels = gpar(fontsize = 12))
 popViewport()
@@
 mosaic(Titanic, labeling_args = list(tl_labels = FALSE, tl_varnames = TRUE, 
-                  abbreviate = c(Survived = 1, Age = 3)), newpage = FALSE, keep = TRUE)
+                  abbreviate = c(Survived = 1, Age = 3)), newpage = FALSE, keep = TRUE,
+       margins = c(left = 4, right = 1, 3), gp_labels = gpar(fontsize = 12))
 popViewport()
@@
 mosaic(Titanic, labeling_args = list(tl_labels = FALSE, tl_varnames = TRUE, 
-       boxes = TRUE, clip = TRUE), newpage = FALSE, keep = TRUE)
+       boxes = TRUE, clip = TRUE), newpage = FALSE, keep = TRUE, 
+       gp_labels = gpar(fontsize = 12))
 popViewport()
@@
        labbl_varnames = TRUE), margins = c(left = 4, right = 1, 3), 
-       newpage = FALSE, keep = TRUE)
+       newpage = FALSE, keep = TRUE, gp_labels = gpar(fontsize = 12))
 popViewport(2)
@@
 pushViewport(viewport(layout.pos.col = 1, layout.pos.row = 1))
-mosaic(~ MaritalStatus + Gender, data = PreSex, labeling = labeling_cells, newpage = FALSE, keep = TRUE)
+mosaic(~ MaritalStatus + Gender, data = PreSex, labeling = labeling_cells, newpage = FALSE, keep = TRUE, gp_labels = gpar(fontsize = 10))
 popViewport()
@@
        labeling = labeling_cells(abbreviate_labels = TRUE, abbreviate_varnames = TRUE, 
-         clip = FALSE), newpage = FALSE, keep = TRUE)
+         clip = FALSE), newpage = FALSE, keep = TRUE, gp_labels = gpar(fontsize = 10))
 popViewport()
@@
          abbreviate_labels = TRUE, clip = FALSE, gp_text = gpar(col = "red")), 
-       newpage = FALSE, keep = TRUE)
+       newpage = FALSE, keep = TRUE, gp_labels = gpar(fontsize = 10))
 popViewport()
 pushViewport(viewport(layout.pos.col = 2, layout.pos.row = 2))
-mosaic(Titanic, labeling_args = list(abbreviate = c(Survived = 1, Age = 4)), 
-       pop = FALSE, newpage = FALSE, keep = TRUE)
+mosaic(Titanic, labeling_args = list(abbreviate = c(Survived = 1, Age = 3)), 
+       pop = FALSE, newpage = FALSE, keep = TRUE, gp_labels = gpar(fontsize = 10))
 tab <- ifelse(Titanic < 6, NA, Titanic)
@@
 <<listfig,fig=TRUE,echo=FALSE>>=
-<<list>>
+mosaic(Titanic, labeling = labeling_list, margins = c(bottom = 5), keep = TRUE)
 @
@@
 pushViewport(viewport(layout.pos.col = 2, layout.pos.row = 1))
-mosaic(art, spacing = spacing_dimequal(unit(c(2,0.5), "lines")), keep = TRUE, newpage = FALSE)
+mosaic(art, spacing = spacing_dimequal(unit(c(0.5, 2), "lines")), keep = TRUE, newpage = FALSE)
 popViewport()
 pushViewport(viewport(layout.pos.col = 1, layout.pos.row = 2))
-mosaic(art, spacing = spacing_increase(start = unit(0.5, "lines"), rate = 1.5), keep = TRUE, newpage = FALSE)
+mosaic(art, spacing = spacing_increase(start = unit(0.3, "lines"), rate = 2.5), keep = TRUE, newpage = FALSE)
 popViewport()
@@
   investigating a new treatment for rheumatoid arthritis. 
-  Source: \cite{vcd:Koch+Edwards:1988}. Taken from: \cite{vcd:Friendly:2000} 
+  Source: \cite{vcd:Koch+Edwards:1988}. Taken from: \cite{vcd:Friendly:2000}.
   Package: \pkg{vcd}.

### 2006-10-10 trunk/vcd/paper/Strucplot/JSS.Rnw
Cosmetics.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@587 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-\noindent (Note that \code{HEC}, a \class{structable} object, already includes a
-splitting information which simply gets overloaded in this example.) 
-For compatibility with \codefun{mosaicplot} in base \proglang{R},
-the \codefun{mosaic} function also allows the use of a \code{"direction"} 
-argument taking a vector of \code{"h"} and
-\code{"v"} characters:
+\noindent (Note that \code{HEC}, a \class{structable} object, already
+includes a splitting information which simply gets overloaded in this
+example.)  For compatibility with \codefun{mosaicplot} in base
+\proglang{R}, the \codefun{mosaic} function also allows the use of a
+\code{direction} argument taking a vector of \code{"h"} and \code{"v"}
+characters:
 
@@
 
-All strucplot core functions are supposed to produce conditional hierarchical
-plots by the means of nested viewports, corresponding to the provided splitting
-information. Thus, at the end of the plotting, each tile is associated with
-a particular viewport. Each of those viewports has to be
-conventionally named, enabling other strucplot modules, in particular the
-labeling functions, to access specific tiles after they have been
-plotted. The naming convention for the viewports is:
+All strucplot core functions are supposed to produce conditional
+hierarchical plots by the means of nested viewports, corresponding to
+the provided splitting information. Thus, at the end of the plotting,
+each tile is associated with a particular viewport. Each of those
+viewports has to be conventionally named, enabling other strucplot
+modules, in particular the labeling functions, to access specific tiles
+after they have been plotted. The naming convention for the viewports
+is:
 
@@
 \noindent since we first push the $k_1$ viewports for the levels of the
-first dimension, then, for \emph{each} of those, the $k_2$ levels of
+first dimension, then, for \emph{each} of these, the $k_2$ levels of
 the second dimension, etc. If the number of levels is equal ($k$) for
-all dimensions, this simplifies to:
+all dimensions, $T_d$ simplifies to
 

### 2006-10-13 trunk/vcd/paper/Color/draft.Rnw
added some references for colors
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@589 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
   coding information contained in some variable. When this involves
-  the shading of areas (and not only points or lines)---e.g., as in barplots, pie charts,
+  the shading of areas (and not only points or lines)---e.g., as in bar plots, pie charts,
   mosaic displays or heatmaps---it is important that the colors are perceptually based
   and do not introduce optical illusions or systematic bias. Here, we discuss
-  how the perceptually-based HCL color space can be used for deriving suitable
+  how the perceptually-based Hue-Chroma-Luminance (HCL) color space can be used for deriving suitable
   color palettes for coding categorical data (qualitative palettes) and numerical
@@
 to statistical graphics. The first illustration is a visualization of a bivariate
-density estimation for the Old Faithful geyser eruptions data. Figure~\ref{fig:bkde}
+density estimation for the Old Faithful geyser eruptions data \citep{vcd:Azzalini+Bowman:1990}.
+Figure~\ref{fig:bkde}
 shows heatmaps of a bivariate kernel density estimate of waiting times between
@@
 the SPD performed better in the North and the CDU better in the South; furthermore,
-Die Linke performed particularly well in the Eastern provinces and the Saarland.
+Die Linke performed particularly well in the Eastern provinces and in Saarland.
 

### 2006-10-16 trunk/vcd/paper/Color/draft.Rnw
finished draft version for colors paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@590 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \title{Choosing Color Palettes for Statistical Graphics}
-\author{Achim Zeileis and Kurt Hornik\\Wirtschaftsuniversit\"at Wien}
+\author{Achim Zeileis \textnormal{and} Kurt Hornik\\Wirtschaftsuniversit\"at Wien}
 \Plainauthor{Achim Zeileis, Kurt Hornik}
@@
 
-\section{Introduction} \label{sec:intro}
-
 <<preliminaries>>=
@@
 
-generate_png <- TRUE
+generate_png <- FALSE
 @
 
-First, we motivate why HCL colors \citep{vcd:Ihaka:2003} are suitable for choosing color palettes
-by contrasting them with the more commonly implemented HSV colors.
+\section{Introduction} \label{sec:intro}
 
-Then, we describe strategies how color palettes for categorical and numerical
-data can be chosen in this space. Following \cite{vcd:Brewer:1999}, we distinguish
-three types of palettes: qualitative, sequential and diverging.
-All palettes described are 
-available in \proglang{R} \citep{vcd:R:2006} in the package \pkg{vcd} \citep{vcd:Meyer+Zeileis+Hornik:2006}
-using the HCL color implementation from \pkg{colorspace} \citep{vcd:Ihaka:2004}.
-Technical documentation to the \proglang{R} implementations along with a large
-collection of examples is available via \verb/help("rainbow_hcl")/ that provides 
-more comparisons between existing \proglang{R} palettes (based on HSV colors) and
-the HCL color palettes.
+Color is an integral element of graphical displays in general, and many statistical
+graphics in particular. Statistical software packages typically provide various color
+palettes and allow practitioners to employ these for visualizing data in various types
+of displays. However, more often than not, there is relatively little guidance about
+how to choose an appropriate palette for a certain visualization task. Here, we address
+this problem by suggesting a color selection strategy for visualizing both categorical
+and numerical information by selecting colors along axes in a color space whose axes
+can be matched with perceptual axes of the human visual system.
 
-%%FIXME: what to say about ColorBrewer
+For implementing color palettes in computer programs, typically color spaces are
+employed that provide a mapping to Red-Green-Blue (RGB) colors. RGB is an additive
+color model which is used for generating colors on computer screens
+\citep[by mixing different amounts of the primary colors red, green and blue,
+see][for more details]{vcd:Poynton:2000}. While being eminently useful for generating
+colors on computer screens, it is hard for humans to select the coordinates in RGB
+space corresponding to a certain color. Therefore, color picker tools in computer
+programs are based on different color models---starting from the pioneering introduction
+of Hue-Saturation-Value (HSV) colors \citep{vcd:Smith:1978}---which try to capture the
+dimensions of human visual perception. HSV space is a simple transformation of RGB
+space which is also implemented in many statistical software packages
+\citep[see][for more details on HSV colors and links to further information about the other color
+spaces discussed in the following]{vcd:Wiki+HSV:2006}. Although being easy to use,
+HSV space exhibits a number of disadvantages: It is relatively difficult to select sets of
+HSV coordinates that yield colors that are ``in harmony'' \citep[see][]{vcd:Munsell:1905}.
+The reason for this is that colors with different hues also differ in saturation and brightness.
+For statistical graphics, this is important because it can introduce size distortions in
+the perception of shaded areas and color-caused optical illusions \citep{vcd:Cleveland+McGill:1983}.
+Furthermore, the use of HSV colors encourages the use of flashy and highly saturated
+colors which are good for drawing attention to a plot but hard to look at for a longer time.
+
+Despite these known drawbacks, HSV space is still the predominantly-used color model in
+color picker tools \citep{vcd:Moretti+Lyons:2002,vcd:Meier+Spalter+Karelitz:2004}
+and the basis for many palettes in statistical software packages. A notable
+exception is the \pkg{ColorBrewer.org} tool \citep{vcd:Harrower+Brewer:2003},
+an online tool for selecting color palettes for maps, providing sets of colors
+for coding various types of information. Here, we take a similar approach
+and describe strategies for the choice of color palettes for categorical and numerical
+data. Following \cite{vcd:Brewer:1999} and \cite{vcd:Harrower+Brewer:2003}, we distinguish
+three types of palettes: qualitative, sequential and diverging. The first is tailored
+for coding categorical information and the latter two are aimed at numerical variables.
+Unlike \pkg{ColorBrewer.org}, we do not only provide fixed sets of colors (with a limited
+number of colors in each set) but describe paths through a perceptually-based color space
+so that the user can decide which path exactly should be taken and how many colors should
+be selected. The color space employed by us is the Hue-Chroma-Luminance (HCL) color space
+\citep[see][]{vcd:Ihaka:2003} and we derive explicit formulae describing the paths through
+this space that correspond to different types of palettes. Given a mapping from HCL to RGB
+coordinates, our formulas are extremely easy to implement and we provide such an implementation
+in the \proglang{R} language \citep{vcd:R:2006} using the powerful \proglang{R} graphics
+system \citep[see][]{vcd:Murrell:2006}. A flexible implementation of all palettes described
+is provided in the package \pkg{vcd} \citep{vcd:Meyer+Zeileis+Hornik:2006} using the HCL
+color implementation from \pkg{colorspace} \citep{vcd:Ihaka:2004}. This gives the user
+both a conceptual and computational tool box for experimenting with color palettes for
+a particular display.
+
+The remainder of the paper is organized as follows: Section~\ref{sec:spaces} provides
+further information on HCL and other color spaces, contrasting in particular the properties
+of HSV and HCL colors. Sections~\ref{sec:qualitative}, \ref{sec:sequential}, and \ref{sec:diverging}
+introduce HCL-based qualitative, sequential, and diverging palettes, respectively.
+Section~\ref{sec:illustrations} shows some brief illustrations for each type of palette
+in statistical graphics before Section~\ref{sec:discussion} concludes the paper with a
+discussion. Some supplementary \proglang{R} code that facilitates experimentation with
+the new palettes is provided in Appendix~\ref{sec:R}.
 
 \section{Color spaces} \label{sec:spaces}
-
-For choosing color palettes, it is imperative to have an understanding
-how colors are perceived. For this it is helpful to have an idea how
-human color vision evolved. It has been hypothesized that it developed in three
-distinct stages: 1.~perception of \textit{light/dark} contrasts (monochrome only),
-2.~\textit{yellow/blue} contrasts (usually associated with our notion of warm/cold colors),
-3.~\textit{green/red} contrasts (helpful for assessing the ripeness of fruit). See 
-\cite{vcd:Ihaka:2003} for more details and references.
-
-Due to these three color axes, colors are typically described as
-locations in 3-dimensional spaces. However, human perception of color
-does not correspond to the physiological axes above, but rather to polar
-coordinates in the color plane (yellow/blue vs.~green/red) plus a third
-light/dark axis. Thus, perceptually-based color spaces are defined by three dimensions
-that try to capture
-\begin{enumerate}
-  \item \textbf{hue} (dominant wavelength)
-  \item \textbf{chroma} (colorfulness)
-  \item \textbf{luminance} (brightness, amount of light)
-\end{enumerate}
 
@@
 
+For choosing color palettes it is helpful to have an idea how
+human color vision evolved. It has been hypothesized that it developed in three
+distinct stages: 1.~perception of \textit{light/dark} contrasts (monochrome only),
+2.~\textit{yellow/blue} contrasts (usually associated with our notion of warm/cold colors),
+3.~\textit{green/red} contrasts (helpful for assessing the ripeness of fruit). See 
+\cite{vcd:Ihaka:2003} for more details and references.
+
+Due to these three color axes, colors are typically described as
+locations in 3-dimensional spaces. However, human perception of color
+does not correspond to the physiological axes above, but rather to polar
+coordinates in the color plane (yellow/blue vs.~green/red) plus a third
+light/dark axis. Thus, perceptually-based color spaces try to capture the
+perceptual axes
+\begin{enumerate}
+  \item \textbf{hue} (dominant wavelength)
+  \item \textbf{chroma} (colorfulness, intensity of color as compared to gray)
+  \item \textbf{luminance} (brightness, amount of gray)
+\end{enumerate}
+
 A popular implementation of such a color space, available in many graphics and
@@
 cylinder). Vertical sections through this space are shown in the upper panel
-of Figure~\ref{fig:hsv-hcl}, depicting hue and saturation given value.
+of Figure~\ref{fig:hsv-hcl}, depicting hue and saturation given different value levels.
 Although simple to specify and easily available in many computing environments,
@@
 gives us the opportunity to conveniently choose color palettes which code categorical
-and/or numerical information by translating it to the three perceptual dimensions.
+and/or numerical information by translating it to paths along the three perceptual axes.
 However, some care is required for dealing with the irregular shape of the HCL space
 which will be addressed in the following sections.
+
+All palettes described are 
+available in \proglang{R} \citep{vcd:R:2006} in the package \pkg{vcd} \citep{vcd:Meyer+Zeileis+Hornik:2006}
+using the HCL color implementation from \pkg{colorspace} \citep{vcd:Ihaka:2004}.
+Technical documentation to the \proglang{R} implementations along with a large
+collection of examples is available via \verb/help("rainbow_hcl")/ that provides 
+more comparisons between existing \proglang{R} palettes (based on HSV colors) and
+the HCL color palettes.
 
@@
 and luminance are kept fixed and only the hue is varied for obtaining different colors
-which are consequently all balanced towards the same gray. If colors from the full
-color wheel (i.e., $H \in [0, 360]$) should be used, not all combinations of chroma
-and luminance are feasible. Figure~\ref{fig:pal-q-construct} depicts how three colors
+which are consequently all balanced towards the same gray. As Figure~\ref{fig:hsv-hcl}
+illustrates, the range of hues available depends on the combination of chroma and
+luminance chosen. Figure~\ref{fig:pal-q-construct} depicts how three colors
 are chosen, given $C = 50$ and $L = 70$.
@@
 A simple and intuitive one is to use colors as metaphors for categories
-(e.g., for political parties). Figure~\ref{fig:pal-q-examples} shows a few
-further examples for generating qualitative sets of colors $(H, 50, 70)$.
+(e.g., for political parties), another approach would be to use segments from
+the color wheel corresponding to nearby or distant colors. The latter is shown
+in Figure~\ref{fig:pal-q-examples} which depicts
+examples for generating qualitative sets of colors $(H, 50, 70)$.
 In the upper left panel colors from the full spectrum are used ($H = 30, 120, 210, 300$)
@@
 $H = 90, 50, 10, 330$) are shown in the lower left and right panel, respectively.
-
-In \pkg{vcd}, these palettes are available in the function
-\begin{Soutput}
-rainbow_hcl(n, c = 50, l = 70, start = 0, end = 360*(n-1)/n, ...)
-\end{Soutput}
-
-%%FIXME: briefly explain interface
 
@@
 
+In \pkg{vcd}, these palettes are available in the function
+\begin{Soutput}
+rainbow_hcl(n, c = 50, l = 70, start = 0, end = 360*(n-1)/n, ...)
+\end{Soutput}
+wher \code{n} controls the number of colors in the palette. The arguments
+\code{c} and \code{l} give the fixed chroma and luminance level, respectively,
+and \code{start} and \code{end} specify the range of hue angles. The function
+is named after the base \proglang{R} function \code{rainbow()} which has a similar
+interface but chooses colors in HSV coordinates. It computes the HCL coordinates
+and transforms them to RGB coordinates by calling the \code{hcl()} function 
+provided by the \pkg{colorspace} package. The \proglang{R} code for reproducing
+Figure~\ref{fig:pal-q-examples} is given in the appendix.
+
 \section{Sequential palettes} \label{sec:sequential}
@@
 The simplest solution to this task is to employ light/dark contrasts, i.e., employ the
-oldest and simplest perceptual axis. The interestingness is thus coded
+most basic perceptual axis. The interestingness is thus coded
 by an increasing amount of gray (i.e., decreasing luminance) 
@@
 \setkeys{Gin}{width=\textwidth}
-\begin{figure}[b!]
+\begin{figure}[p]
 \begin{center}
@@
 
-In \pkg{vcd}, this strategy is implemented in the function
-\begin{Soutput}
-sequential_hcl(n, h = 260, c = c(80, 0), l = c(30, 90), power = 1.5, ...) 
-\end{Soutput}
-
 \setkeys{Gin}{width=.8\textwidth}
@@
 \begin{center}
-<<pal-s-examples, fig=TRUE, height=9, width=6>>=
+<<pal-s-examples, fig=TRUE, height=5, width=6>>=
 par(mfrow = c(5, 1), mar = c(0, 0, 0, 0))
@@
 @
-\caption{\label{fig:pal-s-examples} Examples for sequential palettes.}
+\caption{\label{fig:pal-s-examples} Examples for sequential palettes, varying
+only luminance (first panel), chroma and luminance (second panel), and hue,
+chroma and luminance (remaining panels).}
 \end{center}
 \end{figure}
+
+In \pkg{vcd}, this strategy is implemented in the function
+\begin{Soutput}
+sequential_hcl(n, h = 260, c = c(80, 0), l = c(30, 90), power = 1.5, ...) 
+\end{Soutput}
+where the first element of \code{c} and \code{l} give the starting chroma and luminance
+coordinate (by default colorful and dark) and the second element the ending coordinate
+(by default light gray). The \code{power} argument implements $p$ and defaults to $1.5$.
 
@@
 simultaneously varying the hue as well:
-  \[ (H_2 - i \cdot (H_1 - H_2), C_{\max} - i^{p_1} \cdot (C_{\max} - C_{\min}), L_{\max} - i^{p_2} \cdot (L_{\max} - L_{\min})). \]
-A typical application would be heat colors that increase from a light yellow (e.g., $(90, 30, 90)$)
-to a full red (e.g., $(0, 100, 50)$). To make the change in hue visible, the chroma needs to
-increase rather quickly for low values of $i$ and then only slowly for higher values of $i$.
-This can be achieved by choosing a power $p_1 < 1$.
+  \[ (H_2 - i \cdot (H_1 - H_2), C_{\max} - i^{p_1} \cdot (C_{\max} - C_{\min}),
+      L_{\max} - i^{p_2} \cdot (L_{\max} - L_{\min})). \]
+One application would be a HCL-based version of heat colors that increase from a light yellow
+(e.g., $(90, 30, 90)$) to a full red (e.g., $(0, 100, 50)$). To make the change in hue
+visible, the chroma needs to increase rather quickly for low values of $i$ and then only
+slowly for higher values of $i$. This can be achieved by choosing a power $p_1 < 1$.
 
@@
 \end{Soutput}
-with which the lower three rows in Figure~\ref{fig:pal-s-examples} are produced.
+with which the lower three rows in Figure~\ref{fig:pal-s-examples} are produced. The
+\proglang{R} code reproducing this figure and stating the exact parameters used is again
+provided in the appendix.
+
+\clearpage
 
@@
 
-\setkeys{Gin}{width=\textwidth}
-\begin{figure}[b!]
+\setkeys{Gin}{width=.8\textwidth}
+\begin{figure}[p]
 \begin{center}
@@
 if(generate_png) {
-  png(file = "pal-d-construct.png", height = 480, width = 540)
+  png(file = "pal-d-construct.png", height = 480, width = 534)
   hue260 <- hue.slice(260, grid.n = 301, plot = FALSE)
@@
 \begin{center}
-<<pal-d-examples, fig=TRUE, height=8, width=6>>=
+<<pal-d-examples, fig=TRUE, height=4, width=6>>=
 par(mfrow = c(4, 1), mar = c(0, 0, 0, 0))
@@
 pal(diverge_hcl(7, c = 100, l = c(50, 90), power = 1))
+pal(diverge_hcl(7, h = c(130, 43), c = 100, l = c(70, 90)))
 pal(diverge_hcl(7, h = c(180, 330), c = 59, l = c(75, 95)))
-pal(diverge_hcl(7, h = c(130, 43), c = 100, l = c(70, 90)))
 @
-\caption{\label{fig:pal-d-examples} Examples for diverging palettes.}
+\caption{\label{fig:pal-d-examples} Examples for diverging palettes with different
+pairs of hues and decreasing luminance contrasts.}
 \end{center}
@@
 Figure~\ref{fig:pal-d-construct} shows the chroma/luminance plane back to back for the
-hues $H = 0$ and 260 with two different paths---with slightly different emphasis on
+hues $H = 0$ and 260 with two different paths---giving slightly different emphasis on
 luminance or chroma contrasts---from a full red over a neutral grey to a full blue.
-These particular hues were chosen because they have rather similar
-chroma/luminance planes, allowing for many different combinations of maximal chroma and luminance.
-%%FIXME: symmetric difference
-As Figure~\ref{fig:pal-s-construct} illustrates, a bit of care is needed to choose two colors
-that are sufficiently similar in their chroma/luminance plane to allow for the same
-luminance and/or chroma contrasts.
+As Figure~\ref{fig:pal-s-construct} illustrates, the pair of hues should be chosen carefully
+because the admittable values of in the chrome/luminance plane differ across hues.
+Clearly, for deriving symmetric palettes, only colors from the intersection of the
+admittable chroma/luminance planes can be used. The particular hues $H = 0$ and $260$
+used in Figure~\ref{fig:pal-s-construct} were chosen because they have rather similar
+chroma/luminance planes, allowing for both large chroma and luminance contrasts.
 
@@
 \end{Soutput}
+which has the same arguments as \code{sequential_hcl()} but takes a pair of hues \code{h}.
 
@@
 hue, chroma and luminance. The first palette uses a broader range on the luminance axis
-whereas the others use larger ranges on the chroma axis.
+whereas the others mostly rely on chroma contrasts.
 
-\section{Illustrations}
+\section{Illustrations} \label{sec:illustrations}
 
-In this section, we show a few examples for the various types of palettes applied
+In this section, we show a collection of examples for the various types of palettes applied
 to statistical graphics. The first illustration is a visualization of a bivariate
@@
   image(dens$x1, dens$x2, sqrt(dens$fhat), xlab = "duration", ylab = "waiting", 
-    col = rev(sequential_hcl(33, c = 0, power = 2)))
+    col = rev(sequential_hcl(33, c = 0, l = c(30, 90), power = c(1/5, 2))))
   box()
@@
 names(dimnames(Votes)) <- c("", "")
+dimnames(Votes)[[2]][c(4, 5)] <- c("Gr", "Li")
 
@@
 
-\section{Open questions}
+\section{Discussion} \label{sec:discussion}
 
-From our experience, the paths through HCL space described in the previous sections
-in principle work very well and can be used to effectively code qualitative and 
-quantitative information. However, there are some open questions, formulated below.
-In particular, we are not sure how we can properly define trade-offs
-between increasing the range with respect to one dimension and decreasing the
-range with respect to another.
+Many statistical graphics---especially when displayed on a computer screen,
+e.g., as in interactive usage, electronic papers or presentation slides---
+employ colors to code information about a certain variable. Despite this
+omnipresence of color, there is often only little guidance in statistical
+software packages on how to choose a palette appropriate for a particular
+visualization task---auspicious tools such as \pkg{ColorBrewer.org} notwithstanding.
+We try to address this problem by suggesting color schemes for coding
+categorical information (qualitative palettes) and numerical information
+(sequential and diverging palettes) based on the perceptually-based
+HCL color space.
 
-\begin{itemize}
-  \item In sequential/diverging palettes, there is a trade-off between high chroma and high luminance.
-        How should this be chosen in practice? Our impression is that when a small set of colors
-        (such as 3 or 4) are used, large differnces in chroma work well and large differences
-        in luminance are not necessary. However, when a larger set of colors is used (e.g., for 
-        heatmaps where extreme values should be identifiable) it is much more important to have a big
-        difference in luminance.
+We provide paths through HCL space along perceptual axes
+so that colors selected along these paths match perceptual dimensions.
+This gives the users the possibility to easily experiment with the HCL-based
+palettes by varying several degrees of freedom. For qualitative
+palettes, these are the coordinates on the chroma and luminance axis, respectively,
+controlling whether the colors are light or dark and how colorful they are.
+For sequential and diverging palettes, the user can decide whether contrasts
+in the chroma or luminance direction (or both) should be employed. In our
+experience, chroma contrasts work sufficiently well if a small set of colors
+is used. However, when a larger set of colors is used (e.g., for heatmaps
+where extreme values should be identifiable) it is much more important to
+have a big difference in luminance. Another degree of freedom in these 
+palettes is the rate at which the intensity $|i|$ is increased from $0$ to $1$:
+Small powers $p$ can be used if palettes with a lot of color should be 
+constructed (e.g., to clearly separate regions in a classification map). On
+the other hand, large powers $p$ will result in palettes with less color
+(e.g., appropriate to highlight only extreme regions firmly classified in such a map).
 
-  \item How should the intensity $|i|$ be increased from 0 to 1?
-        Our experience is that for a small set of colors linear increase is sufficient ($p = 1$)
-        whereas in heatmaps where only very extreme regions are interesting a $p > 1$
-        should be used.
+Based on these conceputal guidelines and the computational tools readily
+provided in the \proglang{R} system (and easily implemented in other
+statistical software packages), users can generate palettes varying these
+degrees of freedom and thus adapting the colors to their particular graphical display.
 
-  \item For diverging palettes, how should pairs of colors be chosen? The hues $H = 0$
-        and $H = 260$ where chosen because they are on opposite sides of the color wheel
-	but have a very similar chroma/luminance plane.
+%% From our experience, the paths through HCL space described in the previous sections
+%% in principle work very well and can be used to effectively code qualitative and 
+%% quantitative information. However, there are some open questions, formulated below.
+%% In particular, we are not sure how we can properly define trade-offs
+%% between increasing the range with respect to one dimension and decreasing the
+%% range with respect to another.
+%% 
+%% \begin{itemize}
+%%   \item In sequential/diverging palettes, there is a trade-off between high chroma and high luminance.
+%%         How should this be chosen in practice? Our impression is that when a small set of colors
+%%         (such as 3 or 4) are used, large differnces in chroma work well and large differences
+%%         in luminance are not necessary. However, when a larger set of colors is used (e.g., for 
+%%         heatmaps where extreme values should be identifiable) it is much more important to have a big
+%%         difference in luminance.
+%% 
+%%   \item How should the intensity $|i|$ be increased from 0 to 1?
+%%         Our experience is that for a small set of colors linear increase is sufficient ($p = 1$)
+%%         whereas in heatmaps where only very extreme regions are interesting a $p > 1$
+%%         should be used.
+%% 
+%%   \item For diverging palettes, how should pairs of colors be chosen? The hues $H = 0$
+%%         and $H = 260$ where chosen because they are on opposite sides of the color wheel
+%% 	but have a very similar chroma/luminance plane.
+%% 	
+%%   \item To allow more convenient navigation in HCL space, some more \proglang{R} infrastructure
+%%         could be helpful. We are not sure what would the best or most natural interface for this.
+%% 	A good start might be a function that computes the admissible value on one axis given 
+%% 	a particular value on the other two axes. Maybe you have already experimented with 
+%% 	something like this?
+%% \end{itemize}
 
-  \item To allow more convenient navigation in HCL space, some more \proglang{R} infrastructure
-        could be helpful. We are not sure what would the best or most natural interface for this.
-	A good start might be a function that computes the admissible value on one axis given 
-	a particular value on the other two axes. Maybe you have already experimented with 
-	something like this?
-\end{itemize}
+\section*{Acknowledgements}
 
-%FIXME% Acknowledgements
-%% Paul Murrell, Ross Ihaka, Brian D. Ripley, David Meyer, Thomas Kneib, Michael Hoehle
+We are thankful to David Meyer, Paul Murrell, Brian~D.~Ripley,
+Thomas Kneib, and Michael H\"ohle for feedback, suggestions and discussions.
 
@@
 
+\begin{appendix}
+
+\section[Supplementary R code]{Supplementary \proglang{R} code}
+\label{sec:R}
+
+All palettes discussed above are implemented in the \proglang{R} system for statistical
+computing and graphics \citep{vcd:R:2006} in the package 
+\pkg{vcd} \citep{vcd:Meyer+Zeileis+Hornik:2006}. The functions are based on the
+\code{hcl(h, c, l, ...)} function provedid by the \pkg{colorspace} package
+\citep{vcd:Ihaka:2004}. A simple convenience function for displaying a certain palette
+is
+\begin{Soutput}
+pal <- function(col, border = "light gray", ...)
+{
+  n <- length(col)
+  plot(0, 0, type = "n", xlim = c(0, 1), ylim = c(0, 1),
+       axes = FALSE, xlab = "", ylab = "", ...)
+  rect(0:(n-1)/n, 0, 1:n/n, 1, col = col, border = border)
+}
+\end{Soutput}
+
+The qualitative palettes in Figure~\ref{fig:pal-q-examples} are generated via
+\begin{Soutput}
+pal(rainbow_hcl(4, start = 30, end = 300))
+pal(rainbow_hcl(4, start = 60, end = 240))
+pal(rainbow_hcl(4, start = 270, end = 150))
+pal(rainbow_hcl(4, start = 90, end = -30))
+\end{Soutput}
+
+The sequential palettes in Figure~\ref{fig:pal-s-examples} are generated via
+\begin{Soutput}
+pal(sequential_hcl(12, c = 0, power = 2.2))
+pal(sequential_hcl(12, power = 2.2))
+pal(terrain_hcl(12, c = c(65, 0), l = c(45, 95), power = c(1/3, 1.5)))
+pal(heat_hcl(12, c = c(80, 30), l = c(30, 90), power = c(1/5, 1.5)))
+pal(heat_hcl(12, h = c(0, -100), l = c(75, 40), c = c(40, 80), power = 1))
+\end{Soutput}
+
+The diverging palettes in Figure~\ref{fig:pal-d-examples} are generated via
+\begin{Soutput}
+pal(diverge_hcl(7))
+pal(diverge_hcl(7, c = 100, l = c(50, 90), power = 1))
+pal(diverge_hcl(7, h = c(130, 43), c = 100, l = c(70, 90)))
+pal(diverge_hcl(7, h = c(180, 330), c = 59, l = c(75, 95)))
+\end{Soutput}
+
+Further information on all palettes is available from the corresponding
+online documentation. Examples that compare the new HCL-based palettes with
+those available in base \proglang{R} can be generated via \code{example("rainbow_hcl")}.
+
+\end{appendix}
+
 \end{document}

### 2006-10-16 trunk/vcd/paper/Color/draft.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@591 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 human color vision evolved. It has been hypothesized that it developed in three
-distinct stages: 1.~perception of \textit{light/dark} contrasts (monochrome only),
-2.~\textit{yellow/blue} contrasts (usually associated with our notion of warm/cold colors),
-3.~\textit{green/red} contrasts (helpful for assessing the ripeness of fruit). See 
+distinct stages: 1.~perception of \emph{light/dark} contrasts (monochrome only),
+2.~\emph{yellow/blue} contrasts (usually associated with our notion of warm/cold colors),
+3.~\emph{green/red} contrasts (helpful for assessing the ripeness of fruit). See 
 \cite{vcd:Ihaka:2003} for more details and references.
@@
 \begin{enumerate}
-  \item \textbf{hue} (dominant wavelength)
-  \item \textbf{chroma} (colorfulness, intensity of color as compared to gray)
-  \item \textbf{luminance} (brightness, amount of gray)
+  \item \emph{hue} (dominant wavelength)
+  \item \emph{chroma} (colorfulness, intensity of color as compared to gray)
+  \item \emph{luminance} (brightness, amount of gray)
 \end{enumerate}
@@
 A popular implementation of such a color space, available in many graphics and
-statistics software packages, are \textit{HSV} (hue, saturation, value) colors.
-They are a simple transformation of \textit{RGB} (red, green, blue) colors
+statistics software packages, are \emph{HSV} (hue, saturation, value) colors.
+They are a simple transformation of \emph{RGB} (red, green, blue) colors
 and are defined by a triplet $(H, S, V)$ with $H \in [0, 360]$ and $S, V \in [0, 100]$.
@@
 in statistical software packages. By taking polar coordinates in the UV plane of
-CIELUV, \textit{HCL} (hue, chroma, luminance) colors are obtained, defined by
+CIELUV, \emph{HCL} (hue, chroma, luminance) colors are obtained, defined by
 a triplet $(H, C, L)$ with $H \in [0, 360]$ and $C, L \in [0, 100]$. HCL space

### 2006-10-20 trunk/vcd/paper/Color/TAS.Rnw
draft almost finished, preparing submission version for TAS
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@595 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2006-10-16 trunk/vcd/paper/Color/draft.Rnw
@@
 \title{Choosing Color Palettes for Statistical Graphics}
-\author{Achim Zeileis \textnormal{and} Kurt Hornik\\Wirtschaftsuniversit\"at Wien}
+\author{Achim Zeileis \textnormal{and} Kurt Hornik\\Wirtschaftsuniversit\"at Wien, Austria}
 \Plainauthor{Achim Zeileis, Kurt Hornik}
@@
 source("functions.R")
-load("Bundestagswahl2005.rda")
 
@@
 and the basis for many palettes in statistical software packages. A notable
-exception is the \pkg{ColorBrewer.org} tool \citep{vcd:Harrower+Brewer:2003},
+exception is \pkg{ColorBrewer.org} \citep{vcd:Harrower+Brewer:2003},
 an online tool for selecting color palettes for maps, providing sets of colors
@@
 par(mar = rep(0.8, 4))
-pie(seats, clockwise = TRUE, col = hcl(c(240, 60, 300, 120, 0), 60, 75), radius = 1)
+seats <- structure(c(226, 61, 54, 51, 222),
+  .Names = c("CDU/CSU", "FDP",  "Linke", "Gruene", "SPD"))
+parties <- rainbow_hcl(6, c = 60, l = 75)[c(5, 2, 6, 3, 1)]
+names(parties) <- names(seats)
+pie(seats, clockwise = TRUE, col = parties, radius = 1)
 @
@@
 <<votes, fig=TRUE, height=9, width=11>>=
-Votes <- as.table(as.matrix(votes[c(1, 3, 4, 5, 9, 11, 13, 14, 15, 16, 2, 6, 7, 8, 10, 12),c("CDU/CSU", "FDP", "SPD", "Gruene", "Linke")]))
-names(dimnames(Votes)) <- c("", "")
-dimnames(Votes)[[2]][c(4, 5)] <- c("Gr", "Li")
-
-mosaic(Votes, labeling_args = list(rot_labels = c(0, 90, 0, 0), pos_labels = "center",
-  just_labels = c("center", "center", "center", "right")),
-  gp = gpar(fill = hcl(c(240, 60, 0, 120, 300), 60, 75)),
+data("Bundestag2005")
+votes <- Bundestag2005[c(1, 3:5, 9, 11, 13:16, 2, 6:8, 10, 12),
+                       c("CDU/CSU", "FDP", "SPD", "Gruene", "Linke")]
+parties <- parties[colnames(votes)]
+dimnames(votes)[[2]][c(4, 5)] <- c("Gr", "Li")
+mosaic(votes, gp = gpar(fill = parties),
   spacing = spacing_highlighting, labeling = labeling_left,
-  margins = unit(c(2.5, 1, 1, 12), "lines"), keep = FALSE)
+  labeling_args = list(rot_labels = c(0, 90, 0, 0), pos_labels = "center",
+  just_labels = c("center","center","center","right"), varnames = FALSE),
+  margins = unit(c(2.5, 1, 1, 12), "lines"),
+  keep_aspect_ratio = FALSE)
 @
@@
 luminance $L = 75$. The pie chart clearly shows that neither the governing coalition of SPD
-and Gruene nor the opposition of CDU and FDP could assemble a majority. Given that no party
-would enter a coalition with the leftists, this lead to a big coalition of CDU and SPD.
+and Gruene nor the opposition of CDU/CSU and FDP could assemble a majority. Given that no party
+would enter a coalition with the leftists, this lead to a big coalition of CDU/CSU and SPD.
 Figure~\ref{fig:votes} shows the distribution of votes in this election stratified by 
@@
 north to south, first for the 10 Western provinces, then for the 6 Eastern provinces. Clearly,
-the SPD performed better in the North and the CDU better in the South; furthermore,
+the SPD performed better in the North and the CDU/CSU better in the South; furthermore,
 Die Linke performed particularly well in the Eastern provinces and in Saarland.

### 2006-10-23 trunk/vcd/paper/Color/TAS.Rnw
incorporated KH's comments for TAS color paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@598 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 exception is \pkg{ColorBrewer.org} \citep{vcd:Harrower+Brewer:2003},
-an online tool for selecting color palettes for maps, providing sets of colors
-for coding various types of information. Here, we take a similar approach
+an online tool for selecting color schemes for maps. It provides a rich
+collection of prefabricated palettes (with a fixed maximal number
+of colors) and guides its users how to choose a suitable palette for coding
+various types of information. Here, we take a similar approach
 and describe strategies for the choice of color palettes for categorical and numerical
@@
 three types of palettes: qualitative, sequential and diverging. The first is tailored
-for coding categorical information and the latter two are aimed at numerical variables.
-Unlike \pkg{ColorBrewer.org}, we do not only provide fixed sets of colors (with a limited
-number of colors in each set) but describe paths through a perceptually-based color space
-so that the user can decide which path exactly should be taken and how many colors should
-be selected. The color space employed by us is the Hue-Chroma-Luminance (HCL) color space
-\citep[see][]{vcd:Ihaka:2003} and we derive explicit formulae describing the paths through
-this space that correspond to different types of palettes. Given a mapping from HCL to RGB
-coordinates, our formulas are extremely easy to implement and we provide such an implementation
+for coding categorical information and the latter two are aimed at numerical variables
+(or ordinal variables if some numerical coding for its levels is used).
+Unlike \pkg{ColorBrewer.org}, we do not only provide fixed sets of colors but suggest
+a general principle for selecting colors by traversing paths along perceptual axes in
+a suitable color space. Consequently, the user can decide which path exactly should be
+taken and how many colors should be selected. The color space employed by us is the
+Hue-Chroma-Luminance (HCL) color space \citep[see][]{vcd:Ihaka:2003} and we show how
+the general principles suggested can be turned into formulas describing paths of
+coordinates in this space. Given a mapping from HCL to RGB coordinates, our formulas are
+extremely easy to implement in software and we provide such an implementation
 in the \proglang{R} language \citep{vcd:R:2006} using the powerful \proglang{R} graphics
@@
 further information on HCL and other color spaces, contrasting in particular the properties
-of HSV and HCL colors. Sections~\ref{sec:qualitative}, \ref{sec:sequential}, and \ref{sec:diverging}
-introduce HCL-based qualitative, sequential, and diverging palettes, respectively.
-Section~\ref{sec:illustrations} shows some brief illustrations for each type of palette
-in statistical graphics before Section~\ref{sec:discussion} concludes the paper with a
-discussion. Some supplementary \proglang{R} code that facilitates experimentation with
-the new palettes is provided in Appendix~\ref{sec:R}.
+of HSV and HCL colors. Section~\ref{sec:palettes} introduces HCL-based qualitative,
+sequential, and diverging palettes and Section~\ref{sec:illustrations} shows some brief
+illustrations for each type of palette in statistical graphics. Section~\ref{sec:software}
+provides some general remarks on the implementation in statistical software packages
+as well as some details on our implementation in \proglang{R}. Section~\ref{sec:discussion}
+concludes the paper with a discussion.
 
@@
 
-  png(file = PNG(files[1]), height = 480, width = 720)
+  png(file = PNG(files[1]), height = 400, width = 600)
   hcl.wheel(luminance = seq(5, 95, length = 6))
   dev.off()
-  png(file = PNG(files[2]), height = 480, width = 720)
+  png(file = PNG(files[2]), height = 400, width = 600)
   hsv.wheel(value = seq(0, 1, length = 6), cone = TRUE)
@@
 A popular implementation of such a color space, available in many graphics and
-statistics software packages, are \emph{HSV} (hue, saturation, value) colors.
-They are a simple transformation of \emph{RGB} (red, green, blue) colors
+statistics software packages, are HSV colors. They are a simple transformation of RGB colors
 and are defined by a triplet $(H, S, V)$ with $H \in [0, 360]$ and $S, V \in [0, 100]$.
@@
 in statistical software packages. By taking polar coordinates in the UV plane of
-CIELUV, \emph{HCL} (hue, chroma, luminance) colors are obtained, defined by
+CIELUV, HCL colors are obtained, defined by
 a triplet $(H, C, L)$ with $H \in [0, 360]$ and $C, L \in [0, 100]$. HCL space
-has the shape of a distorted double cone: the admissable chroma and luminance values
-depend on the hue chosen. The lower panel of Figure~\ref{fig:hsv-hcl} shows
+has the shape of a distorted double cone: the admissible combinations of
+chroma and luminance coordinates (within the space's boundaries) depend on the hue chosen.
+The lower panel of Figure~\ref{fig:hsv-hcl} shows
 vertical sections through this space: each of the resulting hue/chroma planes (given
@@
 with increasing luminance) which becomes obvious when converting the colors to
-a gray scale is in Figure~\ref{fig:hsv-hcl-gray}. This balancing of HCL colors
+a gray scale as in Figure~\ref{fig:hsv-hcl-gray}. This balancing of HCL colors
 gives us the opportunity to conveniently choose color palettes which code categorical
@@
 
-All palettes described are 
-available in \proglang{R} \citep{vcd:R:2006} in the package \pkg{vcd} \citep{vcd:Meyer+Zeileis+Hornik:2006}
-using the HCL color implementation from \pkg{colorspace} \citep{vcd:Ihaka:2004}.
-Technical documentation to the \proglang{R} implementations along with a large
-collection of examples is available via \verb/help("rainbow_hcl")/ that provides 
-more comparisons between existing \proglang{R} palettes (based on HSV colors) and
-the HCL color palettes.
+\section{Color palettes} \label{sec:palettes}
 
-\section{Qualitative palettes} \label{sec:qualitative}
+\subsection{Qualitative palettes} \label{sec:qualitative}
 
@@
 
-In \pkg{vcd}, these palettes are available in the function
-\begin{Soutput}
-rainbow_hcl(n, c = 50, l = 70, start = 0, end = 360*(n-1)/n, ...)
-\end{Soutput}
-wher \code{n} controls the number of colors in the palette. The arguments
-\code{c} and \code{l} give the fixed chroma and luminance level, respectively,
-and \code{start} and \code{end} specify the range of hue angles. The function
-is named after the base \proglang{R} function \code{rainbow()} which has a similar
-interface but chooses colors in HSV coordinates. It computes the HCL coordinates
-and transforms them to RGB coordinates by calling the \code{hcl()} function 
-provided by the \pkg{colorspace} package. The \proglang{R} code for reproducing
-Figure~\ref{fig:pal-q-examples} is given in the appendix.
+\subsection{Sequential palettes} \label{sec:sequential}
 
-\section{Sequential palettes} \label{sec:sequential}
+Sequential palettes are used for coding numerical information that ranges
+in a certain interval where low values are considered to be uninteresting and
+high values are interesting. Suppose we need to visualize an intensity or
+interestingness $i$ which (without loss of generality) is scaled to the unit
+interval. A typical application in statistics are heatmaps (see Figure~\ref{fig:bkde}).
 
-Sequential palettes are used for coding numerical information that simply ranges
-in a certain interval where low values are considered to be uninteresting and
-high values are interesting. Without loss of generality, we assume that we want
-to visualize an intensity or interestingness $i \in [0, 1]$. A typical application
-in statistics are heatmaps (see Figure~\ref{fig:bkde}).
-
-The simplest solution to this task is to employ light/dark contrasts, i.e., employ the
-most basic perceptual axis. The interestingness is thus coded
-by an increasing amount of gray (i.e., decreasing luminance) 
+The simplest solution to this task is to employ light/dark contrasts, i.e., rely
+on the most basic perceptual axis. The interestingness can be coded
+by an increasing amount of gray corresponding to decreasing luminance in HCL space:
   \[ (H, 0, 90 - i \cdot 60), \]
@@
 could also be increased nonlinearly, e.g., by some function $f(i)$ that
-controls whether intensity/luminance is increased quickly or not. We found $f(i) = i^p$ 
+controls whether luminance is increased quickly with intensity or not. We found $f(i) = i^p$ 
 to be a convenient transformation where the power $p$ can be varied to achieve
@@
 pal(sequential_hcl(12, power = 2.2))
-pal(terrain_hcl(12, c = c(65, 0), l = c(45, 95), power = c(1/3, 1.5)))
-pal(heat_hcl(12, c = c(80, 30), l = c(30, 90), power = c(1/5, 1.5)))
+pal(heat_hcl(12, c = c(80, 30), l = c(30, 90), power = c(1/5, 2)))
+pal(terrain_hcl(12, c = c(65, 0), l = c(45, 90), power = c(1/2, 1.5)))
 pal(rev(heat_hcl(12, h = c(0, -100), l = c(75, 40), c = c(40, 80), power = 1)))
@@
 
-In \pkg{vcd}, this strategy is implemented in the function
-\begin{Soutput}
-sequential_hcl(n, h = 260, c = c(80, 0), l = c(30, 90), power = 1.5, ...) 
-\end{Soutput}
-where the first element of \code{c} and \code{l} give the starting chroma and luminance
-coordinate (by default colorful and dark) and the second element the ending coordinate
-(by default light gray). The \code{power} argument implements $p$ and defaults to $1.5$.
-
 To increase the contrast between the colors in the palette even further, the ideas from
@@
       L_{\max} - i^{p_2} \cdot (L_{\max} - L_{\min})). \]
-One application would be a HCL-based version of heat colors that increase from a light yellow
+One application would be an HCL-based version of ``heat colors'' that increase from a light yellow
 (e.g., $(90, 30, 90)$) to a full red (e.g., $(0, 100, 50)$). To make the change in hue
@@
 slowly for higher values of $i$. This can be achieved by choosing a power $p_1 < 1$.
+Such a strategy is adopted for the palettes shown in the lower three rows in
+Figure~\ref{fig:pal-s-examples} using different pairs of hues as well as different
+chroma and luminance contrasts. The exact parameter combinations are provided in the
+appendix.
 
-In \proglang{R}, these are available in the function
-\begin{Soutput}
-heat_hcl(n, h = c(0, 90), c = c(100, 30), l = c(50, 90), power = c(1/5, 1), ...)
-\end{Soutput}
-with which the lower three rows in Figure~\ref{fig:pal-s-examples} are produced. The
-\proglang{R} code reproducing this figure and stating the exact parameters used is again
-provided in the appendix.
-
-\clearpage
-
-\section{Diverging palettes} \label{sec:diverging}
+\subsection{Diverging palettes} \label{sec:diverging}
 
@@
 probabilities (with neutral value 0.5) that could be visualized in mosaic plots (see Figure~\ref{fig:arthritis})
-or classification maps (see Figure~\ref{fig:class}). Without loss of generality, we assume
-that we want to visualize an intensity or interestingness $i \in [-1, 1]$.
+or classification maps (see Figure~\ref{fig:class}). Analogously to the previous
+section, we suppose that we want to visualize an intensity or interestingness $i$
+from the interval $[-1, 1]$ (without loss of generality).
 
@@
 
-Given sequential palettes, deriving diverging palettes is easy: two different hues are
+Given useful sequential palettes, deriving diverging palettes is easy: two different hues are
 chosen for adding color to the same amount of `gray' at a given intensity $|i|$.
@@
 As Figure~\ref{fig:pal-s-construct} illustrates, the pair of hues should be chosen carefully
-because the admittable values of in the chrome/luminance plane differ across hues.
+because the admissible values of in the chrome/luminance plane differ across hues.
 Clearly, for deriving symmetric palettes, only colors from the intersection of the
-admittable chroma/luminance planes can be used. The particular hues $H = 0$ and $260$
-used in Figure~\ref{fig:pal-s-construct} were chosen because they have rather similar
-chroma/luminance planes, allowing for both large chroma and luminance contrasts.
-
-In \pkg{vcd}, the following function is provided:
-\begin{Soutput}
-diverge_hcl(n, h = c(260, 0), c = 80, l = c(30, 90), power = 1.5, ...) 
-\end{Soutput}
-which has the same arguments as \code{sequential_hcl()} but takes a pair of hues \code{h}.
+admissible chroma/luminance planes can be used. The particular hues $H = 0$ and $260$
+used in Figure~\ref{fig:pal-s-construct} were chosen because they correspond to similar
+geometric shapes in the chroma/luminance plane, allowing for both large chroma and
+luminance contrasts. If potential viewers of the resulting graphic might be color-blind,
+the pair of hues should be taken from the yellow/blue axis of the color wheel
+rather than the green/red axis as contrasts on the latter axis are more difficult to 
+distinguish for color-blind people.
 
@@
 In this section, we show a collection of examples for the various types of palettes applied
-to statistical graphics. The first illustration is a visualization of a bivariate
-density estimation for the Old Faithful geyser eruptions data \citep{vcd:Azzalini+Bowman:1990}.
-Figure~\ref{fig:bkde}
-shows heatmaps of a bivariate kernel density estimate of waiting times between
-and duration of geyser eruptions in Yellowstone National Park. Both use a sequential
-palette as derived in Section~\ref{sec:sequential} balanced towards the same gray levels
-with $L \in [30, 90]$ and $p_2 = 2$. The sequential palette in the left panel uses only gray
-colors (i.e., $C_{\max} = 0$) and the palette in the right panel additionally employs
-colors with $H \in [0, 90]$, $C \in [30, 80]$ and $p_1 = 1/5$.
-
-\setkeys{Gin}{width=\textwidth}
-\begin{figure}[b!]
-\begin{center}
-<<bkde, height=6, width=12>>=
-if(generate_png) {
-  png(file = "bkde.png", height = 420, width = 840)
-  par(mfrow = c(1, 2))
-  data("geyser", package = "MASS")
-  dens <- bkde2D(cbind(geyser$duration, geyser$waiting), bandwidth = c(0.2, 3), gridsize = c(201, 201))
-  image(dens$x1, dens$x2, sqrt(dens$fhat), xlab = "duration", ylab = "waiting", 
-    col = rev(sequential_hcl(33, c = 0, l = c(30, 90), power = c(1/5, 2))))
-  box()
-  image(dens$x1, dens$x2, sqrt(dens$fhat), xlab = "duration", ylab = "waiting", 
-    col = rev(heat_hcl(33, c = c(80, 30), l = c(30, 90), power = c(1/5, 2))))
-  box()
-  dev.off()
-}
-@
-\includegraphics{bkde}
-\caption{\label{fig:bkde} Bivariate density estimation for Old Faithful geyser eruptions.}
-\end{center}
-\end{figure}
+to statistical graphics. The first examples illustrates qualitative palettes
+and visualizes data from the 2005 election for the German parliament ``Bundestag''
+are employed. In this election, five parties were able to obtain enough votes to enter the
+Bundestag---the distribution of seats is depicted in a pie chart in Figure~\ref{fig:seats}.
+The colors used are rough metaphors for the political parties, using a red hue $H = 0$ for the
+social democrats SPD, a blue hue $H = 240$ for the conservative CDU/CSU, a yellow hue $H = 60$
+for the liberal FDP, a green hue $H = 120$ for the green party ``Die Gr\"unen'' and a purple 
+hue $H = 300$ for the leftist party ``Die Linke''. To obtain rather intense colors, we set
+chroma to $C = 60$ and luminance to $L = 75$ for all hues
+leading to a much better balanced palette compared to the fully saturated
+colors usually found in the (German) media (which instead of blue typically use the extreme
+``color'' black for the CDU/CSU). The pie chart clearly shows that neither the governing
+coalition of SPD and Gr\"une nor the opposition of CDU/CSU and FDP could assemble a majority.
+Given that no party would enter a coalition with the leftists, this lead to a big coalition
+of CDU/CSU and SPD. Figure~\ref{fig:votes} shows the distribution of votes in this election
+stratified by province (Bundesland) in a highlighted mosaic display. The order of provinces is from 
+north to south, first for the 10 western provinces (the former Federal Republic of Germany, FRG),
+then for the 6 eastern provinces (the former German Democratic Republic, GDR). 
+Clearly, the SPD performed better in the north and the CDU/CSU better in the south; furthermore,
+Die Linke performed particularly well in the eastern provinces and in Saarland.
 
@@
 names(parties) <- names(seats)
+names(seats)[4] <- "Grüne"
 pie(seats, clockwise = TRUE, col = parties, radius = 1)
@@
 \begin{center}
-<<votes, fig=TRUE, height=9, width=11>>=
+<<votes, fig=TRUE, height=7, width=8.6>>=
 data("Bundestag2005")
@@
 
-To illustrate qualitative palettes, data from the 2005 election for the German parliament `Bundestag'
-are employed. In that election, five parties were able to obtain enough votes to enter the
-Bundestag---the distribution of seats is depicted in a pie chart in Figure~\ref{fig:seats}.
-The colors used are rough metaphors for the political parties, using a red hue $H = 0$ for the
-social democrats SPD, a blue hue $H = 240$ for the conservative CDU/CSU, a yellow hue $H = 60$
-for the liberal FDP, a green hue $H = 120$ for the green party `Die Gruenen' and a purple 
-hue $H = 300$ for the leftist party `Die Linke'. All colors use the same chroma $C = 60$ and
-luminance $L = 75$. The pie chart clearly shows that neither the governing coalition of SPD
-and Gruene nor the opposition of CDU/CSU and FDP could assemble a majority. Given that no party
-would enter a coalition with the leftists, this lead to a big coalition of CDU/CSU and SPD.
-Figure~\ref{fig:votes} shows the distribution of votes in this election stratified by 
-province (Bundesland) in a highlighted mosaic display. The order of provinces is from 
-north to south, first for the 10 Western provinces, then for the 6 Eastern provinces. Clearly,
-the SPD performed better in the North and the CDU/CSU better in the South; furthermore,
-Die Linke performed particularly well in the Eastern provinces and in Saarland.
+To illustrate sequential palettes, a bivariate density estimation for the Old
+Faithful geyser eruptions data \citep{vcd:Azzalini+Bowman:1990} is visualized.
+The Old Faithful geyser is one of the most popular sites in Yellowstone National Park
+and it is of some interest to understand the relation ship between the duration
+of a geyser eruption and the waiting time until the next eruption.
+Figure~\ref{fig:bkde} shows (two versions of) a heatmap of a bivariate kernel density estimate of 
+duration and waiting time revealing a multi-modal bivariate distribution with more
+variation in the waiting time after a longer eruption. Both versions of the heatmap use a sequential
+palette as derived in Section~\ref{sec:sequential} balanced towards the same gray levels
+with $L \in [30, 90]$ and $p_2 = 1.5$. The sequential palette in the left panel uses only gray
+colors (i.e., $C_{\max} = 0$) and the palette in the right panel additionally employs
+colors with $H \in [0, 90]$, $C \in [30, 80]$ and $p_1 = 1/5$. The parameters have been 
+chosen such that the peaks in the density are brought out very clearly by using a rather
+large luminance contrast and $p_2 > 1$.
+
+\setkeys{Gin}{width=\textwidth}
+\begin{figure}[t!]
+\begin{center}
+<<bkde, height=6, width=12>>=
+if(generate_png) {
+  png(file = "bkde.png", height = 420, width = 840)
+  par(mfrow = c(1, 2))
+  data("geyser", package = "MASS")
+  dens <- bkde2D(cbind(geyser$duration, geyser$waiting), bandwidth = c(0.2, 3), gridsize = c(201, 201))
+  image(dens$x1, dens$x2, dens$fhat, xlab = "duration", ylab = "waiting", 
+    col = rev(sequential_hcl(33, c = 0, l = c(30, 90), power = c(1/5, 1.5))))
+  box()
+  image(dens$x1, dens$x2, dens$fhat, xlab = "duration", ylab = "waiting", 
+    col = rev(heat_hcl(33, c = c(80, 30), l = c(30, 90), power = c(1/5, 1.5))))
+  box()
+  dev.off()
+}
+@
+\includegraphics{bkde}
+\caption{\label{fig:bkde} Bivariate density estimation for Old Faithful geyser eruptions.}
+\end{center}
+\end{figure}
 
@@
 trial investigating a new treatment for rheumatoid arthritis. The mosaic rectangles alone
-signal that the treatment lead to higher improvement compared to the placebo group which are
-shown to be significant by the shading based on the Pearson residuals. Positive residuals, corresponding
+signal that the treatment lead to higher improvement compared to the placebo group; this effect
+is shown to be significant by the shading that codes the size of the Pearon residuals.
+Positive residuals, corresponding
 to more observations in the corresponding cell than expected under independence, are depicted
@@
 
-Figure~\ref{fig:class} shows the fit of a support vector machine (SVM) to an artificial 
-2-class supervised learning example: a mixture of two bivariate normal distributions.
+Figure~\ref{fig:class} shows a heatmap with the fit of a support vector machine (SVM) to an artificial 
+2-class supervised learning example: a mixture of two bivariate normal distributions with
+different means and covariance matrices.
 The circles and triangles show the original observations, solid symbols correspond to
@@
 red respectively.
+
+Both diverging palettes employ a smaller contrast on the luminance axis ($L \in [30, 90]$)
+and a larger chroma contrast ($C \in [0, 100]$) compared to the sequential palettes from
+the previous example. For Figure~\ref{fig:class}, we set $p_1$ and $p_2$ to the somewhat
+smaller value of $1.3$ to bring out all deviations from the decision boundary and not only the
+extreme decision values. For Figure~\ref{fig:arthritis}, the choice of
+$p_1$ and $p_2$ is not so important because only 5 colors are selected---we use values
+$p_1 = p_2 = 1$.
 
@@
 names(dimnames(art))[2] <- "Improvement"
-set.seed(1234)
-mosaic(art, gp = shading_max)
+set.seed(1071)
+mosaic(art, gp = shading_max, gp_args = list(n = 5000))
 @
@@
   plot(fm, data = ex1)
+  ## uses: diverge_hcl(n, c = c(100, 0), l = c(50, 90), power = 1.3)
   dev.off()
@@
 \end{figure}
+
+\section{Software} \label{sec:software}
+
+Implementing the different color palettes suggested in the previous section 
+is extremely easy if the software environment chosen already provides an implementation
+of HCL colors: from the formulas provided above the HCL coordinates for a
+palette can easily be computed. A bit more work is required if the sofware
+package does not yet provide an HCL implementation. In that case, additional
+functionality is needed for translating HCL coordinates to the software package's 
+color system which may vary between different packages, but standardized RGB (sRGB)
+is often used. The typical way of coordinate conversion is to go first from HCL to
+CIELUV by simply transforming the polar $H$ and $C$ coordinates back to the original
+$U$ and $V$. Subsequently CIELUV is converted to CIEXYZ which in turn is converted
+to sRGB (with the latter conversion depending on the device used for display).
+The details of these conversions are somewhat technical and tedious (and hence
+omitted here), however the conversion formulas are still straightfoward to
+implement and can for example be found in \cite{vcd:Wiki+HSV:2006} or
+\cite{vcd:Poynton:2000}.
+
+The \proglang{R} system for statistical computing \citep{vcd:R:2006} already comes
+with an open-source implementation of HCL (and other color spaces) in the package
+\pkg{colorspace} \citep{vcd:Ihaka:2004} which we build upon for implementing all
+palettes described above. The functions outlined below are provided by the package
+\pkg{vcd} \citep{vcd:Meyer+Zeileis+Hornik:2006}---technical documentation for
+the functions along with a large collection of examples is available via
+\verb/help("rainbow_hcl")/. These examples also include comparisons between
+existing \proglang{R} palettes (based on HSV colors) and the new HCL color palettes.
+
+Qualitative palettes are implemented in \pkg{vcd} in the function
+\begin{Soutput}
+rainbow_hcl(n, c = 50, l = 70, start = 0, end = 360*(n-1)/n, ...)
+\end{Soutput}
+where \code{n} controls the number of colors in the palette. The arguments
+\code{c} and \code{l} give the fixed chroma and luminance levels, respectively,
+and \code{start} and \code{end} specify the range of hue angles. The function
+is named after the base \proglang{R} function \code{rainbow()} which has a similar
+interface but chooses colors in HSV coordinates. It computes the HCL coordinates
+and transforms them to RGB coordinates by calling the \code{hcl()} function 
+provided by the \pkg{colorspace} package.
+
+Sequential palettes based on a single hue are implemented in the function
+\begin{Soutput}
+sequential_hcl(n, h = 260, c = c(80, 0), l = c(30, 90), power = 1.5, ...) 
+\end{Soutput}
+where the first element of \code{c} and \code{l} give the starting chroma and luminance
+coordinate (by default colorful and dark) and the second element the ending coordinate
+(by default light gray). The \code{power} argument implements the parameter $p$ 
+from the $i^p$ function. Sequential palettes using a range of hues are provided by
+\begin{Soutput}
+heat_hcl(n, h = c(0, 90), c = c(100, 30), l = c(50, 90), power = c(1/5, 1), ...)
+\end{Soutput}
+named after the HSV-based \proglang{R} function \code{heat.colors()} and
+by default starts from a red and going to a yellow hue. The defaults in \code{heat_hcl()}
+are set differently compared to \code{sequential_hcl()} as to make the default HCL heat
+colors more similar to the HSV version. The defaults of \code{sequential_hcl()}, on the
+other hand, are set as to achieve a large contrast on the luminance axis.
+
+Diverging palettes are implemented in the function
+\begin{Soutput}
+diverge_hcl(n, h = c(260, 0), c = 80, l = c(30, 90), power = 1.5, ...) 
+\end{Soutput}
+which has the same arguments as \code{sequential_hcl()} but takes a pair of hues \code{h}.
+
+The \proglang{R} code for reproducing
+Figures~\ref{fig:pal-q-examples}, \ref{fig:pal-s-examples} and \ref{fig:pal-d-examples}
+is given in the appendix. A vignette that additionally explains how the figures from the illustrations
+section can be reproduced is contained in \pkg{vcd} and be accessed via
+\code{vignette("hcl-colors", package = "vcd")}.
 
@@
 so that colors selected along these paths match perceptual dimensions.
-This gives the users the possibility to easily experiment with the HCL-based
-palettes by varying several degrees of freedom. For qualitative
+This gives the users the possibility to conveniently experiment with the HCL-based
+palettes by varying several graphical parameters. For qualitative
 palettes, these are the coordinates on the chroma and luminance axis, respectively,
@@
 in the chroma or luminance direction (or both) should be employed. In our
-experience, chroma contrasts work sufficiently well if a small set of colors
-is used. However, when a larger set of colors is used (e.g., for heatmaps
-where extreme values should be identifiable) it is much more important to
-have a big difference in luminance. Another degree of freedom in these 
+experience (as illustrated in Section~\ref{sec:illustrations}, chroma contrasts
+work sufficiently well if a small set of colors is used. However, when a larger
+set of colors is used (e.g., for heatmaps where extreme values should be identifiable)
+it is much more important to have a big difference in luminance. Another parameter in these 
 palettes is the rate at which the intensity $|i|$ is increased from $0$ to $1$:
-Small powers $p$ can be used if palettes with a lot of color should be 
+Employing a function $|i|^p$ for this, small
+powers $p$ can be used if palettes with a lot of color should be 
 constructed (e.g., to clearly separate regions in a classification map). On
@@
 Based on these conceputal guidelines and the computational tools readily
-provided in the \proglang{R} system (and easily implemented in other
+provided in the \proglang{R} system for statistical computing (and easily implemented in other
 statistical software packages), users can generate palettes varying these
-degrees of freedom and thus adapting the colors to their particular graphical display.
+graphical parameters and thus adapting the colors to their particular graphical display.
 
@@
 \pkg{vcd} \citep{vcd:Meyer+Zeileis+Hornik:2006}. The functions are based on the
-\code{hcl(h, c, l, ...)} function provedid by the \pkg{colorspace} package
+\code{hcl(h, c, l, ...)} function proveded by the \pkg{colorspace} package
 \citep{vcd:Ihaka:2004}. A simple convenience function for displaying a certain palette
@@
 pal(sequential_hcl(12, power = 2.2))
-pal(terrain_hcl(12, c = c(65, 0), l = c(45, 95), power = c(1/3, 1.5)))
-pal(heat_hcl(12, c = c(80, 30), l = c(30, 90), power = c(1/5, 1.5)))
+pal(heat_hcl(12, c = c(80, 30), l = c(30, 90), power = c(1/5, 2)))
+pal(terrain_hcl(12, c = c(65, 0), l = c(45, 90), power = c(1/2, 1.5)))
 pal(heat_hcl(12, h = c(0, -100), l = c(75, 40), c = c(40, 80), power = 1))
@@
 
-Further information on all palettes is available from the corresponding
-online documentation. Examples that compare the new HCL-based palettes with
-those available in base \proglang{R} can be generated via \code{example("rainbow_hcl")}.
+Further examples and technical information on all palettes is available from
+the corresponding online documentation.
 

### 2006-10-23 trunk/vcd/inst/doc/hcl-colors.Rnw
started vignette for HCL palettes, removed jss.bst and vcd.bib (are assumed to be in texmf)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@601 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2006-10-23 trunk/vcd/paper/Color/TAS.Rnw
@@
 in the \proglang{R} language \citep{vcd:R:2006} using the powerful \proglang{R} graphics
-system \citep[see][]{vcd:Murrell:2006}. A flexible implementation of all palettes described
-is provided in the package \pkg{vcd} \citep{vcd:Meyer+Zeileis+Hornik:2006} using the HCL
-color implementation from \pkg{colorspace} \citep{vcd:Ihaka:2004}. This gives the user
+system \citep[see][]{vcd:Murrell:2006} and the HCL
+color implementation from package \pkg{colorspace} \citep{vcd:Ihaka:2004}. This gives the user
 both a conceptual and computational tool box for experimenting with color palettes for
@@
 i.e., for coding a categorical variable. Usually, these should give the same
-perceptual weight to each category so that no group is perceived to be larger/more
+perceptual weight to each category so that no group is perceived to be larger or more
 important than any other one. Typical applications of qualitative palettes in
-statistics would be bar plots, pie charts (see Figure~\ref{fig:seats}) 
+statistics would be bar plots \citep[see][]{vcd:Ihaka:2003}, pie charts (see Figure~\ref{fig:seats}) 
 or highlighted mosaic displays (see Figure~\ref{fig:votes}).
@@
 As Figure~\ref{fig:pal-s-construct} illustrates, the pair of hues should be chosen carefully
-because the admissible values of in the chrome/luminance plane differ across hues.
+because the admissible values in the chrome/luminance plane differ across hues.
 Clearly, for deriving symmetric palettes, only colors from the intersection of the
 admissible chroma/luminance planes can be used. The particular hues $H = 0$ and $260$
-used in Figure~\ref{fig:pal-s-construct} were chosen because they correspond to similar
+used in Figure~\ref{fig:pal-d-construct} were chosen because they correspond to similar
 geometric shapes in the chroma/luminance plane, allowing for both large chroma and
@@
 In this section, we show a collection of examples for the various types of palettes applied
-to statistical graphics. The first examples illustrates qualitative palettes
-and visualizes data from the 2005 election for the German parliament ``Bundestag''
-are employed. In this election, five parties were able to obtain enough votes to enter the
+to statistical graphics. The first example illustrates qualitative palettes
+and visualizes data from the 2005 election for the German parliament ``Bundestag''.
+In this election, five parties were able to obtain enough votes to enter the
 Bundestag---the distribution of seats is depicted in a pie chart in Figure~\ref{fig:seats}.
@@
 and it is of some interest to understand the relation ship between the duration
-of a geyser eruption and the waiting time until the next eruption.
+of a geyser eruption and the waiting time for this eruption.
 Figure~\ref{fig:bkde} shows (two versions of) a heatmap of a bivariate kernel density estimate of 
-duration and waiting time revealing a multi-modal bivariate distribution with more
-variation in the waiting time after a longer eruption. Both versions of the heatmap use a sequential
+duration and waiting time revealing a multi-modal bivariate distribution: short waiting
+times (around 50 minutes) are typically followed by a long eruption (around 4 minutes)
+whereas long waiting times (around 80 minutes) can be followed by either a long
+or short eruption (around 4 minutes). Both versions of the heatmap use a sequential
 palette as derived in Section~\ref{sec:sequential} balanced towards the same gray levels
@@
   data("geyser", package = "MASS")
-  dens <- bkde2D(cbind(geyser$duration, geyser$waiting), bandwidth = c(0.2, 3), gridsize = c(201, 201))
-  image(dens$x1, dens$x2, dens$fhat, xlab = "duration", ylab = "waiting", 
+  ## geyser2 <- cbind(geyser$duration[-299], geyser$waiting[-1])
+  geyser2 <- cbind(geyser$duration, geyser$waiting)
+  dens <- bkde2D(geyser2, bandwidth = c(0.2, 3), gridsize = c(201, 201))
+  image(dens$x1, dens$x2, dens$fhat, xlab = "duration", ylab = "waiting time", 
     col = rev(sequential_hcl(33, c = 0, l = c(30, 90), power = c(1/5, 1.5))))
   box()
-  image(dens$x1, dens$x2, dens$fhat, xlab = "duration", ylab = "waiting", 
+  image(dens$x1, dens$x2, dens$fhat, xlab = "duration", ylab = "waiting time", 
     col = rev(heat_hcl(33, c = c(80, 30), l = c(30, 90), power = c(1/5, 1.5))))
@@
 
-Both diverging palettes employ a smaller contrast on the luminance axis ($L \in [30, 90]$)
+Both diverging palettes employ a smaller contrast on the luminance axis ($L \in [50, 90]$)
 and a larger chroma contrast ($C \in [0, 100]$) compared to the sequential palettes from
@@
 extreme decision values. For Figure~\ref{fig:arthritis}, the choice of
-$p_1$ and $p_2$ is not so important because only 5 colors are selected---we use values
+$p_1$ and $p_2$ is less important because only 5 colors are selected---we use values
 $p_1 = p_2 = 1$.
@@
 of HCL colors: from the formulas provided above the HCL coordinates for a
-palette can easily be computed. A bit more work is required if the sofware
+palette can be conveniently computed. A bit more work is required if the software
 package does not yet provide an HCL implementation. In that case, additional
@@
 CIELUV by simply transforming the polar $H$ and $C$ coordinates back to the original
-$U$ and $V$. Subsequently CIELUV is converted to CIEXYZ which in turn is converted
+$U$ and $V$. Subsequently, CIELUV is converted to CIEXYZ which in turn is converted
 to sRGB (with the latter conversion depending on the device used for display).
@@
 omitted here), however the conversion formulas are still straightfoward to
-implement and can for example be found in \cite{vcd:Wiki+HSV:2006} or
+implement and can, for example, be found in \cite{vcd:Wiki+LUV:2006} or
 \cite{vcd:Poynton:2000}.
@@
 interface but chooses colors in HSV coordinates. It computes the HCL coordinates
-and transforms them to RGB coordinates by calling the \code{hcl()} function 
+and transforms them to RGB by calling the \code{hcl()} function 
 provided by the \pkg{colorspace} package.
@@
 in the chroma or luminance direction (or both) should be employed. In our
-experience (as illustrated in Section~\ref{sec:illustrations}, chroma contrasts
+experience (as illustrated in Section~\ref{sec:illustrations}), chroma contrasts
 work sufficiently well if a small set of colors is used. However, when a larger

### 2006-10-25 trunk/vcd/inst/doc/strucplot.Rnw
new strucplot vignette added
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@608 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2005-10-16 trunk/vcd/inst/doc/strucplot.Rnw
@@
-\documentclass[a4paper]{article}
-\usepackage{hyperref, graphicx, color, alltt, a4wide}
-\usepackage{Sweave}
-\usepackage[round]{natbib}
-\definecolor{Red}{rgb}{0.7,0,0}
-\definecolor{Blue}{rgb}{0,0,0.8}
-\definecolor{hellgrau}{rgb}{0.55,0.55,0.55}
-\newcommand{\pkg}[1]{{\normalfont\fontseries{b}\selectfont #1}}
-\newcommand{\code}[1]{\texttt{#1}}
+\documentclass{Z}
+
+\usepackage{thumbpdf}
+
+%% almost as usual
+\author{David Meyer\\Wirtschaftsuniversit\"at Wien \And 
+        Achim Zeileis\\Wirtschaftsuniversit\"at Wien \And 
+        Kurt Hornik\\Wirtschaftsuniversit\"at Wien}
+\title{The Strucplot Framework:\\ Visualizing Multi-way Contingency
+  Tables with \pkg{vcd}}
+
+%% for pretty printing and a nice hypersummary also set:
+\Plainauthor{David Meyer, Achim Zeileis, Kurt Hornik} %% comma-separated
+\Shorttitle{The Strucplot Framework} %% a short title (if necessary)
+\Plaintitle{The Strucplot Framework: Visualizing Multi-way Contingency
+  Tables with vcd}
+
+%% an abstract and keywords
+\Abstract{
+This paper has been published in the Journal of Statistical Software
+\citep{vcd:Meyer+Zeileis+Hornik:2006b} and 
+describes the ``strucplot'' framework for the visualization of
+multi-way contingency tables. Strucplot displays include hierarchical
+conditional plots such as mosaic, association, and sieve plots, and
+can be combined into more complex, specialized plots for visualizing
+conditional independence, GLMs, and the results of
+independence tests. The framework's modular design allows
+flexible customization of the plots' graphical appearance, including 
+shading, labeling, spacing, and legend, by means of ``graphical
+appearance control'' functions. 
+The framework is provided by the \proglang{R} package \pkg{vcd}.
+}
+
+\Keywords{contingency tables, mosaic plots, association plots, sieve plots, 
+categorical data, independence, conditional independence, HSV, HCL,
+residual-based shading, \pkg{grid}, \proglang{R}}
+\Plainkeywords{contingency tables, mosaic plots, association plots,
+  sieve plots, categorical data, independence, 
+  conditional independence, HSV, HCL, residual-based shading, grid, R}
+
+\SweaveOpts{engine=R,eps=TRUE,height=6,width=7,results=hide,fig=FALSE,echo=TRUE}
+\setkeys{Gin}{width=0.7\textwidth}
+
+<<preliminaries,echo=FALSE,results=hide>>=
+set.seed(1071)
+library(vcd)
+data(Titanic)
+data(HairEyeColor)
+data(PreSex)
+data(Arthritis)
+art <- xtabs(~Treatment + Improved, data = Arthritis)
+@
+
+\newcommand{\var}[1]{\textit{\texttt{#1}}}
+\newcommand{\data}[1]{\texttt{#1}}
 \newcommand{\class}[1]{\textsf{#1}}
-\newenvironment{smallexample}{\begin{alltt}}{\end{alltt}}
-\newcommand{\email}[1]{\href{mailto:#1}{\normalfont\texttt{#1}}}
-\newcommand{\data}[1]{`\texttt{#1}'}
-
 %% \code without `-' ligatures
@@
 
+%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \begin{document}
 
-%\VignetteIndexEntry{The Strucplot Framework---Visualizing Multi-way Contingency Tables}
-%\VignetteDepends{vcd}
-%\VignetteKeywords{graphics, categorical data}
-%\VignettePackage{vcd}
-
-\SweaveOpts{engine=R,eps=TRUE,height=6,width=7,results=hide,fig=FALSE,echo=TRUE}
+%% include your article here, just as usual
+%% Note that you should use the \pkg{}, \proglang{} and \code{} commands.
+
+\section[Introduction]{Introduction}
+%% Note: If there is markup in \(sub)section, then it has to be escape as above.
+
+In order to explain multi-dimensional categorical data, statisticians
+typically look for (conditional) independence structures. Whether the
+task is purely exploratory or model-based, techniques such as mosaic and
+association plots offer good support for visualization.  Both visualize
+aspects of (possibly higher-dimensional) contingency tables, with
+several extensions introduced over the last two decades, and
+implementations available in many statistical environments.  A
+\emph{mosaic plot} \citep{vcd:Hartigan+Kleiner:1984} is basically an
+area-proportional visualization of (typically, observed) frequencies,
+composed of tiles (corresponding to the cells) created by recursive
+vertical and horizontal splits of a rectangle. Thus, the area of each tile
+is proportional to the corresponding cell entry \emph{given} the
+dimensions of previous splits. An \emph{association plot}
+\citep{vcd:Cohen:1980} visualizes the standardized deviations of
+observed frequencies from those expected under a certain independence
+hypothesis.  Each cell is represented by a rectangle that has (signed)
+height proportional to the residual and width proportional to the square
+root of the expected counts, so that the area of the box is proportional
+to the difference in observed and expected frequencies.
+
+Extensions to these techniques have mainly focused on the following
+aspects.
+
+\begin{enumerate}
+\item Varying the shape of bar plots and mosaic displays to yield, e.g.,
+  double-decker plots \citep{vcd:hofmann:2001}, spine plots, or
+  spinograms \citep{vcd:hofmann+theus}.
+\item Using residual-based shadings to visualize log-linear models
+  \citep{vcd:Friendly:1994,vcd:Friendly:2000} and significance of
+  statistical tests \citep{vcd:Meyer+Zeileis+Hornik:2003,vcd:Zeileis+Meyer+Hornik:2005}.
+\item Using pairs plots and trellis-like layouts for marginal, conditional and
+  partial views \citep{vcd:Friendly:1999}.
+\item Adding direct user interaction, allowing quick exploration and
+  modification of the visualized models 
+  \citep{vcd:Unwin+Hawkins+Hofmann:1996,vcd:Theus:2003}.
+\item Providing a modular and flexible implementation to easily allow
+  user extensions \citep{vcd:Meyer+Zeileis+Hornik:2003,vcd:Meyer+Zeileis+Hornik:2006b}.
+\end{enumerate}
+
+\noindent Current implementations of mosaic displays can be found,
+e.g., for \proglang{SAS} \citep{vcd:SAS:2005}, \pkg{ViSta} \citep{vcd:young:1996}, 
+\pkg{MANET} \citep{vcd:Unwin+Hawkins+Hofmann:1996},
+\pkg{Mondrian} \citep{vcd:Theus:2003},
+\proglang{R} \citep{vcd:R:2006}, and \proglang{S-PLUS}
+\citep{vcd:SPLUS:2005}. For \proglang{R}, currently three
+implementations do exist in the packages \pkg{graphics} (in base \proglang{R}), 
+\pkg{vcd} \citep{vcd:Meyer+Zeileis+Hornik:2006b},
+and \pkg{iplots} \citep{vcd:urbanek+wichtrey:2006}, respectively.
+Table \ref{tab:compare} gives an overview of the available functionality in these systems.
+Most environments are available on Windows, MacOS, and Linux/Unix
+variants, except \pkg{MANET} which is only available for the
+Macinthosh platforms.
+
+\begin{table}[h]
+  \centering
+  \begin{tabular}{|l|c|c|c|c|c|c|c|c|c|}
+    \hline
+                             &                &                   
+                             &\multicolumn{3}{c|}{}   &        & &\\
+                             & \proglang{SAS} & \proglang{S-PLUS} 
+                             &\multicolumn{3}{c|}{\proglang{R}} & \pkg{ViSta} & \pkg{MANET}
+                             & \pkg{Mondrian}\\
+                             &                &                   
+                             &\pkg{base}&\pkg{vcd}   &\pkg{iplots}&        & &\\\hline
+
+    Basic functionality      &  $\times$  & $\times$  & $\times$ &$\times$ &$\times$ &   $\times$   & $\times$& $\times$\\
+    Shape                    &            &           &          &$\times$ && $\times$  &  $\times$&\\
+    Res.-based shadings  &  $\times$  &        & $\times$ & $\times$ & ($\times$) &     &($\times$)& ($\times$)\\
+    Highlighting             &     &        &   &$\times$   &$\times$  & $\times$ & $\times$& $\times$\\
+    Conditional views        &  $\times$  &        &    &$\times$ &          & $\times$ &  $\times$&\\
+    Interaction              &     &        &   &  &$\times$  &   $\times$   & $\times$& $\times$\\
+    Linking                  &     &        &   &  &$\times$  & $\times$ & $\times$& $\times$\\
+    Extensible design        &     &        &  &$\times$ & &       &  &\\
+    Language                 & \proglang{SAS} & \proglang{S} &
+    \proglang{R} & \proglang{R} & \proglang{R}/\proglang{Java} & \proglang{XLisp} & \proglang{C++} & \proglang{Java}\\
+    \hline
+  \end{tabular}
+  \caption{Comparison of current software environments.}
+  \label{tab:compare}
+\end{table}
+
+Figures \ref{fig:arthritis} to \ref{fig:titanic} illustrate some of these
+extensions. Figure~\ref{fig:arthritis} shows the results from a
+double-blind clinical trial
+investigating a new treatment for rheumatoid arthritis, using an
+extended mosaic plot with residual-based shading based on the
+maximum statistic: clearly, the new treatment is effective. The dark
+blue cell indicates that the rate of treated patients showing marked improvement
+is significant at the 1\% level. Figure
+\ref{fig:ucbadmissions} visualizes the well-known UCB admissions data
+by means of a conditional association plot. The panels show the
+residuals from a conditional independence model (independence of gender
+and admission, given department), stratified by department. Clearly,
+the situation in department A (more women/less men accepted than
+would be expected under the null hypothesis) causes the rejection of the
+hypothesis of conditional independence. Figure~\ref{fig:presex}
+illustrates the conditional independence of premarital and
+extramarital sex, given gender and marital status. The $\chi^2$ test 
+of independence, based on the permutation distribution,  
+rejects the null hypothesis: possibly, because
+the tendency of people to have extramarital
+sex when they had premarital sex 
+is particularly marked among married people? The rate of such women and men
+ist significant at the 0.01 and 0.1 level, respectively.
+Finally, Figure~\ref{fig:titanic} visualizes the ``Survival on the Titanic'' 
+data using a double-decker plot. Here, a binary response (survival of the disaster)
+is to be explained by other factors (class, gender, and age). The gray
+boxes represent the proportion of survived passengers in a particular
+stratum. The proportions of saved women and children are indeed higher
+than those of men, but they clearly decrease from the 1st to the 3rd class. In
+addition, the proportion of saved men in the 1st class is higher than
+in the others.
+
 \setkeys{Gin}{width=0.7\textwidth}
-
-<<preliminaries,echo=FALSE,results=hide>>=
-set.seed(1071)
-library(vcd)
-data(Titanic)
-data(HairEyeColor)
-@
-
-\title{The Strucplot Framework---Visualizing Multi-way Contingency Tables}
-\author{by David Meyer, Achim Zeileis, and Kurt Hornik}
-\date{}
-
-\maketitle
-\sloppy
-
-\section{Framework Overview}
-
-The strucplot framework in the {\sf R} package \pkg{vcd} visualizes
-multi-way contingency tables and integrates techniques such as
+\begin{figure}[p]
+\begin{center}
+<<Arthritis,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
+mosaic(art, gp = shading_max, split_vertical = TRUE)
+@
+\caption{Mosaic plot for the \data{Arthritis} data.}
+\label{fig:arthritis}
+\end{center}
+\end{figure}
+
+\setkeys{Gin}{width=\textwidth}
+\begin{figure}[p]
+\begin{center}
+<<UCBAdmissions,fig=TRUE,echo=FALSE,height=6,width=10,results=hide>>=
+cotabplot(UCBAdmissions, panel = cotab_coindep, shade = TRUE, legend = FALSE, 
+          type = "assoc")
+@
+\caption{Conditional association plot for the \data{UCBAdmissions} data.}
+\label{fig:ucbadmissions}
+\end{center}
+\end{figure}
+
+\setkeys{Gin}{width=0.7\textwidth}
+\begin{figure}[p]
+\begin{center}
+<<PreSex,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
+presextest <- coindep_test(PreSex, margin = c(1,4), 
+                           indepfun = function(x) sum(x^2), n = 5000)
+mosaic(PreSex, condvars = c(1, 4), shade = TRUE, 
+       gp_args = list(p.value = presextest$p.value))
+@
+\caption{Mosaic plot for the \data{PreSex} data.}
+\label{fig:presex}
+\end{center}
+\end{figure}
+
+\setkeys{Gin}{width=0.8\textwidth}
+\begin{figure}[p]
+\begin{center}
+<<Titanic,fig=TRUE,echo=FALSE,height=6,width=10,results=hide>>=
+doubledecker(Survived ~ ., data = Titanic, labeling_args = 
+             list(set_varnames = c(Sex = "Gender")))
+@
+\caption{Double-decker plot for the \data{Titanic} data.}
+\label{fig:titanic}
+\end{center}
+\end{figure}
+
+This paper describes the strucplot framework provided by the \pkg{vcd}
+package for the \proglang{R} environment for statistical computing and
+graphics, available from the Comprehensive \proglang{R} Archive Network
+(\url{http://CRAN.R-project.org/}).  The framework integrates displays
+such as mosaic, association, and sieve plots by their unifying property
+of being flat representations of contingency tables.  These basic plots,
+as well as specialized displays for conditional independence, can be
+used both for exploratory visualization and model-based analysis.
+Exploratory techniques include specialized displays for the bivariate
+case, as well as pairs and trellis-type displays for
+higher-dimensional tables. Model-based tools include methods suitable
+for the visualization of conditional independence tests (including
+permutation tests), as well as for the visualization of particular GLMs
+(logistic regression, log-linear models). Additionally, two of the
+framework's further strengths are its flexibility and extensibility:
+graphical appearance aspects such as shading, labeling, and spacing are
+modularized by means of ``\underline{\vphantom{g}gr}aphical 
+\underline{\vphantom{g}ap}pearance \underline{\vphantom{g}con}trol''
+(\emph{grapcon}) functions, allowing
+fine-granular customization and user-level extensions.
+
+The remainder of the paper is organized as follows. In
+Section \ref{sec:strucplot}, we give an overview of the strucplot
+framework, describing the hierarchy of the main components and the
+basic functionality. 
+In Section \ref{sec:shading}, we demonstrate how
+(residual-based) shadings support the visualization of log-linear models and
+the results of independence tests. Also, we explain step-by-step how
+the concepts of generating and grapcon functions
+can be combined to provide a flexible customization of
+complex graphical displays as created by the strucplot framework.
+Sections \ref{sec:labeling} and \ref{sec:spacing} discuss in detail the labeling
+and spacing features, respectively. Section \ref{sec:example}
+exemplifies the framework in the analysis of a four-way data set.
+Section \ref{sec:conclusion} concludes the work.
+
+\section[The strucplot framework]{The strucplot framework}
+\label{sec:strucplot}
+
+The strucplot framework in the \proglang{R} package \pkg{vcd}, used for visualizing
+multi-way contingency tables, integrates techniques such as
 mosaic displays, association plots, and sieve plots. The main idea is to visualize
@@
 
-\noindent This document gives an introduction to the framework,
-whereas labeling and shading issues are described in separate vignettes.
-
-The strucplot framework is highly modularized: Figure \ref{fig:struc}
+The strucplot framework is highly modularized: Figure~\ref{fig:struc}
 shows the hierarchical relationship between the various components. 
@@
 parameter functions that directly or indirectly influence the final
-appearance of the plot. These are examples of `graphical appearance control'
-(`grapcon') functions. They are created by generating functions, allowing
-flexible parameterization and extensibility. 
-The first part of their names (\code{\emph{group\_foo()}}) 
-reflects the group they belong to (strucplot core, labeling,
+appearance of the plot (see Table \ref{tab:grapcons} for an overview). 
+These are examples of grapcon functions. They are created by generating functions
+(\emph{grapcon generators}), allowing
+flexible parameterization and extensibility (Figure~\ref{fig:struc}
+only shows the generators). The generator names
+follow the naming convention \code{\textit{group\_foo}()},  
+where \code{\textit{group}} reflects the group the
+generators belong to (strucplot core, labeling,
 legend, shading, or spacing). The workhorse functions (created by 
-\code{\emph{struc\_foo()}}, 
-\code{\emph{labeling\_foo()}}, and \code{\emph{legend\_foo()}}) 
-directly produce graphical output (``add ink to the canvas''), whereas
+\code{struc\_\textit{foo}()}, 
+\code{labeling\_\textit{foo}()}, and \code{legend\_\textit{foo}()}) 
+directly produce graphical output (i.e., ``add ink to the canvas''), whereas
 the parameter functions
-(created by \code{\emph{spacing\_foo()}} and \code{\emph{shading\_foo()}}) compute
+(created by \code{spacing\_\textit{foo}()} and \code{shading\_\textit{foo}()}) compute
 graphical parameters used by the others. The grapcon functions returned by 
-\code{\emph{struc\_foo()}} implement the core functionality, creating the tiles and their
+\code{struc\_\textit{foo}()} implement the core functionality, 
+creating the tiles and their
 content.  On the second level of the framework, a suitable combination
@@
 interface \codefun{strucplot} through sensible parameter defaults
-and support for model formulas. Finally, on the fourth
-level, there are `related' \pkg{vcd} functions (such as \codefun{cotabplot}
+and support for model formulae. Finally, on the fourth
+level, there are ``related'' \pkg{vcd} functions (such as \codefun{cotabplot}
 and the \codefun{pairs} methods for table objects) 
@@
 framework into more complex displays (e.g., by means of panel functions).
+
+\begin{table}
+  \begin{tabular}{|l|l|l|}
+    \hline
+    \textbf{Group} & \textbf{Grapcon generator} & \textbf{Description}\\\hline
+    strucplot & \codefun{struc\_assoc} & core function for association plots\\
+    core      & \codefun{struc\_mosaic} & core function for mosaic plots\\
+              & \codefun{struc\_sieve} & core function for sieve plots\\\hline\hline
+     labeling & \codefun{labeling\_border} & border labels\\
+              & \codefun{labeling\_cboxed} & centered labels with
+              boxes, all labels clipped,\\
+              && and on top and left border\\
+              & \codefun{labeling\_cells} & cell labels\\
+              & \codefun{labeling\_conditional} & border labels 
+                                                  for conditioning variables\\
+              && and cell labels for conditioned variables\\
+              & \codefun{labeling\_doubledecker} & draws labels for
+              doubledecker plot\\
+              & \codefun{labeling\_lboxed} & left-aligned labels with boxes\\
+              & \codefun{labeling\_left} & left-aligned border labels\\
+              & \codefun{labeling\_left2} & left-aligned border
+              labels, all labels on top and left border\\
+              & \codefun{labeling\_list} & draws a list of labels
+              under the plot\\\hline\hline
+     shading  & \codefun{shading\_binary} & visualizes the sign of the  residuals\\
+              & \codefun{shading\_Friendly} & implements Friendly
+              shading (based on HSV colors)\\
+              & \codefun{shading\_hcl} & shading based on HCL colors\\
+              & \codefun{shading\_hsv} & shading based on HSV colors\\
+              & \codefun{shading\_max} & shading visualizing the
+              maximum test statistic\\ && (based on HCL colors)\\
+              & \codefun{shading\_sieve} & implements Friendly
+              shading customized for sieve plots\\ && (based on HCL colors)\\\hline\hline
+     spacing  & \codefun{spacing\_conditional} & increasing spacing
+              for conditioning variables,\\&& equal spacing for
+              conditioned variables\\
+              & \codefun{spacing\_dimequal} & equal spacing for each dimension\\
+              & \codefun{spacing\_equal} & equal spacing for all dimensions\\
+              & \codefun{spacing\_highlighting} & increasing spacing,
+              last dimension set to zero\\
+              & \codefun{spacing\_increase} & increasing spacing\\\hline\hline
+     legend   & \codefun{legend\_fixed} & creates a fixed number of
+              bins (similar to \codefun{mosaicplot})\\
+              & \codefun{legend\_resbased} & suitable for an
+              arbitrary number of bins\\&& (also for continuous shadings)\\\hline
+
+  \end{tabular}
+  \caption{Available grapcon generators in the strucplot framework}
+  \label{tab:grapcons}
+\end{table}
 
@@
 
-\section{Mosaic, Association, and Sieve Plots}
+\subsection{Mosaic, association, and sieve plots}
 
@@
 data containing two polytomous variables (hair and eye color), 
-as well as one (artificial) dichotomous variable (sex, i.e., gender). The
-`flattened' contingency table can be obtained using the
+as well as one (artificial) dichotomous gender variable (\code{Sex}). The
+``flattened'' contingency table can be obtained using the
 \codefun{structable} function (quite similar to \codefun{ftable} in
-base {\sf R}, but allowing the specification of split directions):
-
-<<structable,results=verbatim>>=
-(hec <- structable(Eye ~ Sex + Hair, data = HairEyeColor))
-@
-
-Let us first visualize the contingency table by means of a mosaic plot
-\citep{vcd:Hartigan+Kleiner:1984} which is basically 
-an area-proportional visualization of (typically, observed) frequencies, composed
-of tiles (corresponding to the cells) created by recursive
-vertical and horizontal splits of a square. Thus, the area of each tile
-is proportional to the corresponding cell entry \emph{given} the
-dimensions of previous splits. Figure \ref{fig:observed} depicts 
-the effect of
+base \proglang{R}, but allowing the specification of split directions):
+
+<<structable,echo=TRUE,results=verbatim>>=
+(HEC <- structable(Eye ~ Sex + Hair, data = HairEyeColor))
+@
+
+Let us first visualize the contingency table by means of a mosaic plot.
+% \citep{vcd:Hartigan+Kleiner:1984} which is basically 
+% an area-proportional visualization of (typically, observed) frequencies, composed
+% of tiles (corresponding to the cells) created by recursive
+% vertical and horizontal splits of a square. Thus, the area of each tile
+% is proportional to the corresponding cell entry \emph{given} the
+% dimensions of previous splits. 
+The effect of
 
 <<Observed>>=
-mosaic(hec)
+mosaic(HEC)
 @
@@
 
+%\setkeys{Gin}{width=0.75\textwidth}
 \begin{figure}[p]
 \begin{center}
-<<Observedfig,fig=TRUE,echo=FALSE>>=
+<<Observedfig,fig=TRUE,echo=FALSE,height=6>>=
 <<Observed>>
@@
 
-\noindent The small bullets indicate zero entries in the corresponding
-cell. Note that in contrast to, e.g.,  \codefun{mosaicplot} in base
-{\sf R}, the default split direction and level ordering in all strucplot
-displays correspond to the textual representation.
+\noindent depicts the observed frequencies of the \code{HairEyeColor}
+data. If there are zero entries, tiles have zero area and are, additionally,
+marked by small bullets (see, e.g, Figure~\ref{fig:titanic}). By
+default, these cells are not split further. The bullets help
+distinguishing very small cells from zero entries, and are
+particularly useful when color shadings come into play 
+(see the example using the \data{Bundesliga} data in Section \ref{sec:overview}). 
+Note that in contrast to, e.g.,  \codefun{mosaicplot} in base
+\proglang{R}, the default split direction and level ordering in all strucplot
+displays correspond to the textual representation produced by the
+print methods.
 It is also possible to visualize the expected values instead of the
-observed values (see Figure \ref{fig:expected}):
+observed values (see Figure~\ref{fig:expected}):
 
 <<Expected>>=
-mosaic(hec, type = "expected")
-@
-
-\setkeys{Gin}{width=0.7\textwidth}
+mosaic(HEC, type = "expected")
+@
+
 \begin{figure}[p]
 \begin{center}
-<<Expectedfig,fig=TRUE,echo=FALSE>>=
+<<Expectedfig,fig=TRUE,echo=FALSE,height=6>>=
 <<Expected>>
@@
 \end{figure}
+%\setkeys{Gin}{width=0.7\textwidth}
 
@@
 
-\noindent Alternatively, we can directly inspect the residuals. 
+\noindent where \code{spacing\_dimequal} is used to set the spacing
+of the second and third dimension to zero. Alternatively, we can directly inspect the residuals. 
 The Pearson residuals (standardized deviations of observed from expected values) 
-are preferably visualized using association
+are conveniently visualized using association
 plots \citep{vcd:Cohen:1980}. In contrast to \codefun{assocplot} in
-base {\sf R}, \pkg{vcd}'s \codefun{assoc}
-function scales to more than two variables (see Figure \ref{fig:residuals}):
+base \proglang{R}, \pkg{vcd}'s \codefun{assoc}
+function scales to more than two variables (see Figure~\ref{fig:residuals}):
 
 <<Residuals>>=
-assoc(hec, compress = FALSE)
+assoc(HEC, compress = FALSE)
 @
@@
 the splitting of the tiles can be controlled using the
-\code{split\_vertical} argument (default: alternating splits starting
-with a vertical one).
+\code{split\_vertical} argument. The default is to alternate splits starting
+with a horizontal one (see Figure~\ref{fig:split}):
+
+<<echo=FALSE>>=
+options(width=60)
+@
 
 <<split1>>=
-mosaic(hec, split_vertical = c(TRUE, FALSE, TRUE))
-@
-
-\noindent For compatibility with \codefun{mosaicplot} in base {\sf R},
-the \codefun{mosaic} function also allows the use of a \code{"direction"} 
-argument taking a vector of \code{"h"} and
-\code{"v"} characters (see Figure \ref{fig:split}):
-
-<<split2>>=
-mosaic(hec, direction = c("v","h","v"))
+mosaic(HEC, split_vertical = c(TRUE, FALSE, TRUE), 
+labeling_args = list(abbreviate = c(Eye = 3)))
+@
+
+<<echo=FALSE>>=
+options(width=70)
 @
@@
 \end{figure}
+
+\noindent (Note that \code{HEC}, a \class{structable} object, already
+includes a splitting information which simply gets overloaded in this
+example.)  For compatibility with \codefun{mosaicplot} in base
+\proglang{R}, the \codefun{mosaic} function also allows the use of a
+\code{direction} argument taking a vector of \code{"h"} and \code{"v"}
+characters:
+
+<<split2>>=
+mosaic(HEC, direction = c("v","h","v"))
+@
 
@@
 can be customized in a quite flexible way. For example, 
-\codefun{doubledecker} is simply a wrapper for 
-\codefun{mosaic}, setting the right defaults. 
-Figure \ref{fig:doubledecker} shows a doubledecker plot of
-the \data{Titanic} data, explaining the probability of survival (`survived') by age, given
+the default method for \codefun{doubledecker} is simply a wrapper for 
+\codefun{strucplot}, setting the right defaults.
+Most default settings such as colors, spacing, and labeling are
+specified via the parameters and passed through to
+\codefun{strucplot}. The additional code just handles the dependent
+variable information, and in particular permutes the table to have the dependent
+variable as the last dimension as required for the doubledecker plot.
+Figure~\ref{fig:titanic} shows a doubledecker plot of
+the \data{Titanic} data, explaining the probability of survival (``survived'') by age, given
 sex, given class. It is created by:
@@
 
-\begin{figure}[h]
-\begin{center}
-<<doubledeckerfig,fig=TRUE,echo=FALSE>>=
-<<doubledecker1>>
-@
-\caption{Doubledecker plot for the \data{Titanic} data.}
-\label{fig:doubledecker}
-\end{center}
-\end{figure}
-
-\section{Conditional and partial views}
-
-So far, we have visualized full tables. For objects of class \class{table},
-conditioning on levels (i.e., choosing a table subset for fixed levels
-of the conditioning variable(s)) is simply done by indexing.
-However, subsetting \code{"structable"} objects is more restrictive 
-because of their inherent conditional structure. Since the variables
+\subsection{Conditional and partial views}
+
+So far, we have visualized either full or collapsed tables, as
+suggested by the analysis task at hand. Subtables can be selected in a
+similar way as for objects of class \class{table} using
+indexing. Note, however, that subsetting of \class{structable} objects 
+is more restrictive because of their inherent conditional structure. Since the variables
 on both the row and the columns side are nested, 
-conditioning is only possible ``outside-in'':
+subsetting is only possible ``outside-in'', that is, indexing operates
+on blocks defined by the variable levels. In the following, we use the
+Titanic data again, this time collapsed over \code{Survived} to investigate the
+structure of crew and passengers (and having the \code{Child} and
+\code{Age} labels of the \code{Age} variable swapped for optical clarity):
+
+<<echo=FALSE>>=
+options(width=75)
+@
 
 <<subsetting,results=verbatim>>=
-hec
-hec["Male",]
-hec[c("Male", "Brown"),]
-hec["Male","Green"]
+(STD <- structable(~ Sex + Class + Age, data = Titanic[,,2:1,]))
+STD["Male",]
+STD["Male", c("1st","2nd","3rd")]
+@
+
+<<echo=FALSE>>=
+options(width=70)
+@
+
+\noindent \emph{Conditioning} on levels (i.e., choosing a table subset for fixed levels
+of the conditioning variable(s)) is done using the \code{[[} operator. %]]
+Here again, the sequence of conditioning levels is restricted by the
+hierarchical structure of the \class{structable} object. In the
+following examples, note that compared to subsetting, the first
+dimension(s) are dropped:
+
+<<conditioning,results=verbatim>>=
+STD[["Male",]]
+STD[[c("Male", "Adult"),]]
+STD[["Male","1st"]]
 @
@@
 
-<<Variables>>=
+<<Variables1>>=
 pushViewport(viewport(layout = grid.layout(ncol = 2)))
-
+@
+
+<<Variables2>>=
 pushViewport(viewport(layout.pos.col = 1))
-mosaic(hec["Male"], margins = c(top = 2, 0), sub = "Male", newpage = FALSE)
+mosaic(STD[["Male"]], margins = c(left = 2.5, top = 2.5, 0), sub = "Male", newpage = FALSE)
 popViewport()
-
+@
+
+<<Variables3>>=
 pushViewport(viewport(layout.pos.col = 2))
-mosaic(hec["Female"], margins = c(top = 2, 0), sub = "Female", newpage = FALSE)
+mosaic(STD[["Female"]], margins = c(top = 2.5, 0), sub = "Female", newpage = FALSE)
 popViewport(2)
@@
 \begin{center}
-<<Variablesfig,fig=TRUE,echo=FALSE,height=5,width=12>>=
-<<Variables>>
-@
-\caption{Distribution of hair and eye color, given gender.}
+<<Variablesfig,fig=TRUE,echo=FALSE,height=6,width=12>>=
+<<Variables1>>
+<<Variables2>>
+<<Variables3>>
+@
+\caption{Two mosaic displays put side-by-side, visualizing the 
+distribution of class and age, given gender. The marginal
+distribution of gender cannot be seen.}
 \label{fig:parttable}
@@
 
-Since mosaic displays are ``conditional plots'' by definition, we can
-also use one single mosaic for stratified plots.
-The formula interface of \codefun{mosaic} allows the
-specification of conditioning variables (see Figure \ref{fig:conditioning}):
-
-<<Conditioning>>=
-mosaic(~ Hair + Eye | Sex, data = hec, split_vertical = TRUE, keep_aspect_ratio = FALSE)
-@
-
-\begin{figure}[p]
-\begin{center}
-<<Conditioningfig,fig=TRUE,echo=FALSE,width=12>>=
-<<Conditioning>>
-@
-\caption{Mosaic plot for conditional independence structures.}
-\label{fig:conditioning}
-\end{center}
-\end{figure}
-
-\noindent The effect of using this kind of formula is that
-conditioning variables are permuted ahead of the the conditioned variables
-in the table, and that \codefun{spacing\_conditional} is used as
-default to better
-distinguish conditioning from conditioned dimensions.
-This spacing uses equal space between tiles of conditioned variables,
-and increasing space between tiles of conditioning variables. In
-addition, we release the fixed aspect ratio to get less distorted margins.
-
 The \codefun{cotabplot} function does a much better job on this task:
@@
 in a lattice-like layout, conditioning on variable \emph{levels}.
-The plot in Figure \ref{fig:cotabplot} shows hair and eye color, given sex:
+The plot in Figure~\ref{fig:cotabplot} shows hair and eye color, given sex:
 
 <<cotabplot>>=
-cotabplot(~ Hair + Eye | Sex, data = hec, panel_args = list(margins = 3), labeling = labeling_left(clip = FALSE))
-#cotabplot(~ Hair + Eye | Sex, data = hec, labeling_args = list(abbreviate = c(Eye = 2)))
+cotabplot(~ Class + Age | Sex, data = STD, split_vertical = TRUE)
+@
+
+\setkeys{Gin}{width=\textwidth}
+\begin{figure}[p]
+\begin{center}
+<<cotabplotfig,fig=TRUE,echo=FALSE,height=6,width=11>>=
+<<cotabplot>>
+@
+\caption{Conditional table plot for the \data{Titanic} data,
+  again visualizing the distribution of age and class, given
+  gender, using separate mosaic displays like the ``manual'' plot 
+  in Figure~\ref{fig:parttable}.}
+\label{fig:cotabplot}
+\end{center}
+\end{figure}
+\setkeys{Gin}{width=0.7\textwidth}
+%\noindent The \code{labeling\_args} argument modifies the labels'
+%appearance: here, to be left-aligned and unclipped
+%(see Section \ref{sec:labeling}).
+
+\noindent Visualizing the strata separately ``hides'' the distribution of the
+conditioning variable(s) which may or may not be appropriate or
+sensible in a particular analysis step. If we wish to keep the
+information on the marginal distribution(s), we can use one single
+mosaic for the stratified plot since mosaic displays are 
+``conditional plots'' by definition. We just need to make sure that
+conditioning variables are used first for splitting. 
+Both the default and the formula interface of \codefun{mosaic} allow the
+specification of conditioning variables (see Figure~\ref{fig:conditioning}):
+
+<<Conditioning1>>=
+mosaic(STD, condvars = "Sex", split_vertical = c(TRUE, TRUE, FALSE))
+@ 
+<<Conditioning2>>=
+mosaic(~ Class + Age | Sex, data = STD, split_vertical = c(TRUE, TRUE, FALSE))
 @
@@
 \begin{center}
-<<cotabplotfig,fig=TRUE,echo=FALSE,height=6,width=11>>=
-<<cotabplot>>
-@
-\caption{Conditional table plot for the \data{HairEyeColor} data.}
-\label{fig:cotabplot}
-\end{center}
-\end{figure}
-\setkeys{Gin}{width=0.7\textwidth}
-
-\noindent The \code{labeling\_args} argument ensures that `eye' levels
-are abbreviated (see the vignette on labeling for detailed information).
-
-Another high-level function for visualizing conditional independence
-models are the \codefun{pairs} methods for table and structable objects.
+<<Conditioningfig,fig=TRUE,echo=FALSE,width=12>>=
+<<Conditioning2>>
+@
+\caption{Mosaic plot again visualizing the 
+distribution of class and age, given gender, this time using a
+single mosaic plot. In contrast to Figures~\ref{fig:parttable} and \ref{fig:cotabplot}, 
+this plot also visualizes the marginal distribution of gender.}
+\label{fig:conditioning}
+\end{center}
+\end{figure}
+\setkeys{Gin}{width=0.7}
+
+\noindent The effect of using this is that
+conditioning variables are permuted ahead of the the conditioned variables
+in the table, and that \codefun{spacing\_conditional} is used as default to better
+distinguish conditioning from conditioned dimensions.
+This spacing uses equal space between tiles of conditioned variables,
+and increasing space between tiles of conditioning variables (See 
+Section~\ref{sec:spacing}).
+
+Another set of high-level functions for visualizing conditional independence
+models are the \codefun{pairs} methods for \class{table} and \class{structable} objects.
 In contrast to \codefun{cotabplot} which conditions on variables,
 the \codefun{pairs} methods create pairwise views of the table. 
-The function produces, by default, a plot matrix 
+They produce, by default, a plot matrix 
 having strucplot displays in the off-diagonal panels, and
-the variable names (optionally, with univariate statistics) in the diagonal cells. 
-Figure \ref{fig:pairs} shows a pairs display with mosaic plots visualizing mutual 
-independence in the lower triangle,
-association plots for the same in the upper triangle, and bar charts in the diagonal.
+the variable names (optionally, with univariate displays) in the diagonal cells. 
+Figure~\ref{fig:pairs} shows a pairs display for the
+\data{Titanic} data with univariate mosaics in the diagonal,
+and mosaic plots visualizing the corresponding bivariate mosaics
+in the upper and lower triangles.
+Due to the inherent asymmetry of mosaic displays, the corresponding plots in the
+upper and lower triangle differ depending on which variable is used first
+for splitting---inspecting both views might help detecting patterns in a data set. 
+Additionally, 
+we are using a special spacing and shading normally used to `highlight' %'
+the second variable in the first (as will be discussed in Section \ref{sec:spacing}): here,
+the intention of the spacing 
+is to emphasize the conditional distributions of the second variable,
+given the first one, and the shading helps identifying the factor levels in
+the second variable.
 
 <<pairs>>=
-pairs(hec, lower_panel = pairs_assoc, space = 0.3, 
-      diag_panel_args = list(rot = -45, just_leveltext = c("left","bottom")))
+pairs(STD, highlighting = 2, diag_panel = pairs_diagonal_mosaic, 
+      diag_panel_args = list(fill = grey.colors))
 @
@@
 
-\noindent (The labels of the variables are to be read from left to
-right and from top to bottom.) In plots produced by \codefun{pairs}, 
+\noindent The labels of the variables are to be read from left to
+right and from top to bottom. In addition, the levels can be matched
+by position within the columns and by shading within the rows.
+In plots produced by \codefun{pairs}, 
 each panel's row and column define two variables $X$ and $Y$ used for the
@@
 higher dimensionality. Therefore, this plot is typically used with a suitable 
-residual-based shading (described in the vignette on ``Colors and
-Residual-based Shadings in the Strucplot Framework'').
-
-\section{Interactive plot modifications}
-
-All strucplot core functions are supposed to produce conditional hierarchical
-plots by the means of nested viewports, corresponding to the provided splitting
-information. Thus, at the end of the plotting, each tile is associated with
-a particular viewport. Each of those viewports has to be
-conventionally named, enabling other strucplot modules, in particular the
-labeling functions, to access specific tiles after they have been
-plotted. The naming convention for the viewports is:
-
-\begin{center}
-\code{cell:\emph{Variable1}=\emph{Level1},\emph{Variable2}=\emph{Level2}} \dots
+residual-based shading (see Section \ref{sec:shading}).
+
+\subsection{Interactive plot modifications}
+
+All strucplot core functions are supposed to produce conditional
+hierarchical plots by the means of nested viewports, corresponding to
+the provided splitting information. Thus, at the end of the plotting,
+each tile is associated with a particular viewport. Each of those
+viewports has to be conventionally named, enabling other strucplot
+modules, in particular the labeling functions, to access specific tiles
+after they have been plotted. The naming convention for the viewports
+is:
+
+\begin{center}
+\code{\emph{[Optional prefix]}cell:\emph{Variable1}=\emph{Level1},\emph{Variable2}=\emph{Level2}} \dots
 \end{center}
@@
 
-\bibliographystyle{jss}
+\subsection{Performance issues}
+\label{sec:performance}
+
+As stated above, the implementation of strucplot displays is based on
+creating and nesting \pkg{grid} viewports. The main time-consuming steps performed by
+the core functions are the following:
+
+\begin{enumerate}
+\item recursively, split the table until the individual cells are reached
+\item during the splits, add viewports to the plot
+\item for the individual cells, add plot-specific content 
+      (rectangles for mosaics, bars for association plots, etc.)
+\end{enumerate}
+
+\noindent All these operations scale linearly with the amount of
+created viewports. For a $d$-dimensional table with $k_i$
+levels, $i=1 \dots d$, the total number of needed viewports
+$T_d$ can roughly be estimated as
+
+\begin{equation}
+  \label{eq:numbervp}
+  T_d \quad = \quad k_1 + k_1k_2 + \cdots + k_1 \cdots k_d \quad =\quad \sum_{i=1}^d \prod_{j \le i} k_j
+\end{equation}
+
+\noindent since we first push the $k_1$ viewports for the levels of the
+first dimension, then, for \emph{each} of these, the $k_2$ levels of
+the second dimension, etc. If the number of levels is equal ($k$) for
+all dimensions, $T_d$ simplifies to
+
+\begin{equation}
+  \label{eq:equalvp}
+  T_d \quad = \quad \sum_{i=1}^d k^i = \frac{k(k^d-1)}{k-1}
+\end{equation}
+
+\noindent and so the time complexity for drawing a strucplot display is of order $k^d$.
+
+\section{Shadings}
+\label{sec:shading}
+
+Unlike other graphics functions in base \proglang{R}, the strucplot framework 
+allows almost full control over the graphical parameters of all plot elements. In
+particular, in association plots, mosaic plots, and sieve plots, 
+the user can modify the graphical appearance of each tile individually. 
+Built on top of this functionality, the
+framework supplies a set of shading functions choosing colors
+appropriate for the visualization of log-linear models.
+The tiles' graphical parameters are set using the \code{gp} argument
+of the functions of the strucplot framework. This argument basically expects an object
+of class \class{gpar} whose components are arrays of the same shape 
+(length and dimensionality) as the data table (see Section \ref{sec:gp}). 
+For convenience, however, the user can also
+supply a grapcon function that computes such an object given a vector of
+residuals, or, alternatively, a generating function that takes
+certain arguments and returns such a grapcon function (see Section
+\ref{sec:shadingcustom}). We provide several shading functions, including
+support for both HSV and HCL colors, and the
+visualization of significance tests (see Section \ref{sec:overview}).
+
+\subsection{Specifying graphical parameters of strucplot displays}
+\label{sec:gp}
+
+As an example, consider the \data{UCBAdmissions} data. 
+In the table aggregated over departments, we
+would like to highlight the (incidentally wrong) impression that there
+were too many male students accepted compared to the presumably
+discriminated female students (see Figure~\ref{fig:ucb}):
+
+<<ucb,results=verbatim>>=
+(ucb <- margin.table(UCBAdmissions, 1:2))
+
+(fill_colors <- matrix(c("dark cyan","gray","gray","dark magenta"), ncol = 2))
+
+mosaic(ucb, gp = gpar(fill = fill_colors, col = 0))
+@
+
+\begin{figure}[h]
+\begin{center}
+<<ucbfig,fig=TRUE,echo=FALSE>>=
+<<ucb>>
+@
+
+\caption{Mosaic plot for the \data{UCBAdmissions} data with highlighted cells.}
+\label{fig:ucb}
+\end{center}
+\end{figure}
+
+\noindent As the example shows, we create a fourfold table 
+with appropriate colors (dark cyan for admitted male students and dark
+magenta for rejected female students) and supply them to the \code{fill} component 
+of the \class{gpar} object passed to the \code{gp} argument of \codefun{mosaic}. 
+For visual clarity, we additionally hide the tiles' borders by setting the \code{col} 
+component to 0 (transparent).
+
+If the parameters specified in the \class{gpar} object are ``incomplete'',
+they will be recycled along the last splitting
+dimension. In the following example based on the \data{Titanic} data, 
+we will highlight all cells corresponding to survived
+passengers (see Figure~\ref{fig:recycling}):
+
+<<recycling>>=
+mosaic(Titanic, gp = gpar(fill = c("gray","dark magenta")),
+                spacing = spacing_highlighting,
+                labeling_args = list(abbreviate = c(Age = 3), rep = c(Survived = FALSE))
+      )
+@
+
+\noindent Note that \codefun{spacing\_highlighting} sets the spaces
+between tiles in the last dimension to 0. The \code{labeling\_args} 
+argument ensures that labels do not overlap (see Section \ref{sec:labeling}).
+
+\begin{figure}[h]
+\begin{center}
+<<recyclingfig,fig=TRUE,echo=FALSE>>=
+<<recycling>>
+@
+\caption{Recycling of parameters, used for highlighting the survived
+  passengers in the \data{Titanic} data.}
+\label{fig:recycling}
+\end{center}
+\end{figure}
+
+\subsection{Customizing residual-based shadings}
+\label{sec:shadingcustom}
+
+This flexible way of specifying graphical parameters is the basis for a suite of
+shading functions that modify the tiles' appearance with respect to a vector of 
+residuals, resulting from deviations of observed from expected
+frequencies under a given log-linear model. 
+The idea is to visualize at least sign and absolute size of the residuals, but some
+shadings, additionally, indicate overall significance. One particular
+shading, the maximum shading 
+\citep{vcd:Meyer+Zeileis+Hornik:2003,vcd:Zeileis+Meyer+Hornik:2005},
+even allows to identify the cells that cause the rejection of the null hypothesis.
+
+Conceptually, the strucplot framework offers three alternatives to
+add residual-based shading to plots:
+
+\begin{enumerate}
+\item Precomputing the graphical parameters (e.g., fill colors),
+  encapsulating them into an object of class
+  \class{gpar} as demonstrated in the previous section, and
+  passing this object to the \code{gp} argument.
+\item Providing a grapcon function to the \code{gp} argument that 
+  takes residuals as input and returns an object as described in
+  alternative 1.
+ \item Providing a grapcon generator taking parameters and 
+   returning a function as described in alternative~2.
+\end{enumerate}
+
+\noindent For each of these approaches, we will demonstrate the
+necessary steps to obtain a binary shading that visualizes the sign of
+the residuals by a corresponding 
+fill color (for simplicity, we will treat 0 as positive).
+
+\subsubsection*{Alternative 1: Precomputed \class{gpar} object}
+
+The first method is precomputing the graphical parameters ``by
+hand''. We will use \code{royalblue4} color for positive and \code{mediumorchid4}
+color for negative residuals (see Figure~\ref{fig:binary}):
+
+<<shading1,results=verbatim>>=
+expected <- independence_table(ucb)
+(x <- (ucb - expected)  / sqrt(expected))
+
+(shading1_obj <- ifelse(x > 0, "royalblue4", "mediumorchid4"))
+
+mosaic(ucb, gp = gpar(fill = shading1_obj))
+@
+
+\begin{figure}[h]
+\begin{center}
+<<shading1fig,fig=TRUE,echo=FALSE>>=
+<<shading1>>
+@ 
+\caption{Binary shading visualizing the sign of the residuals.}
+\label{fig:binary}
+\end{center}
+\end{figure}
+
+\subsubsection*{Alternative 2: Grapcon function}
+
+For implementing alternative 2, we need to create a ``shading function''
+that computes \class{gpar} objects from
+residuals. For that, we can just reuse the code from the previous step:
+
+<<shading2>>=
+shading2_fun <- function(x) gpar(fill = ifelse(x > 0, "royalblue4", "mediumorchid4"))
+@
+
+\noindent To create a mosaic display with binary shading, 
+it now suffices to specify the data table along with
+\codefun{shading2\_fun}:
+
+<<shading3>>=
+mosaic(ucb, gp = shading2_fun)
+@
+
+\noindent \codefun{mosaic} internally calls
+\codefun{strucplot} which computes the residuals from the specified
+independence model (total independence by default), passes them to
+\codefun{shading2\_fun}, and uses the \class{gpar} object returned 
+to finally create the plot.
+
+Our \codefun{shading2\_fun} function might be useful, 
+but can still be improved: the hard-wired colors should be
+customizable. We cannot
+simply extend the argument list to include, e.g., 
+a \code{fill = c("royalblue4", "mediumorchid4")} argument because
+\codefun{strucplot} will neither know how to handle it, nor let us
+change the defaults. In fact, the interface of shading functions is fixed,
+they are expected to take exactly one argument: a table of residuals. 
+This is where generating functions (alternative 3) come
+into play.
+
+\subsubsection*{Alternative 3: Grapcon generator}
+
+We simply wrap our grapcon shading function in another 
+function that takes all additional arguments it needs to use, possibly
+preprocesses them, and returns the actual shading function. This returned
+function will have access to the parameters since in \proglang{R}, nested
+functions are lexically scoped. Thus, the grapcon generator
+returns (``creates'') a 
+``parameterized'' shading function with the minimal standard interface
+\codefun{strucplot} requires. The following example shows the necessary
+extensions for our running example:
+
+<<shading3>>=
+shading3a_fun <- function(col = c("royalblue4", "mediumorchid4")) {
+  col <- rep(col, length.out = 2)
+  function(x) gpar(fill = ifelse(x > 0, col[1], col[2]))  
+}
+@
+
+\noindent The first statement just makes sure that exactly two colors
+are specified. In the call to \codefun{mosaic}, using the new
+\codefun{shading3a\_fun} function, we can now simply change the colors:
+
+<<shading4>>=
+mosaic(ucb, gp = shading3a_fun(c("royalblue4","mediumorchid4")))
+@
+
+\noindent (figure not shown).  The procedure described so far is a
+rather general concept, applicable to a wide family of user-level
+\pkg{grid} graphics. Indeed, the customization of other components of
+the strucplot framework (labeling, spacing, legend, and core functions)
+follows the same idea.  Now for the shading functions, more
+customization is needed. Note that \codefun{shading3a\_fun} needs to be
+evaluated by the user, even if the defaults are to be used. It is a
+better idea to let \codefun{strucplot} call the generating function,
+which, in particular, allows the passing of arguments that are computed
+by \codefun{strucplot}. Since shading functions can be used for
+visualizing significance (see Section \ref{sec:overview}), it makes
+sense for generating functions to have access to the model, i.e.,
+observed and expected values, residuals, and degrees of freedom. For
+example, the \codefun{shading\_max} generating function computes a
+permutation distribution of the maximum statistic and $p$ values for
+specified significance levels based on the observed table to create
+data-driven cut-off points. If this was done in the shading function
+itself, the permutation statistic would be recomputed every time the
+shading function is called, resulting in possibly severe performance
+loss and numerical inconsistencies. Therefore, generating functions for
+shadings are required to take at least the parameters \code{observed},
+\code{expected}, \code{residuals}, and \code{df} (these are provided by
+the strucplot framework), followed by other parameters controlling the
+shading appearance (to be specified by the user):
+
+<<shading4>>=
+shading3b_fun <- function(observed = NULL, residuals = NULL, expected = NULL, 
+    df = NULL, col = c("royalblue4", "mediumorchid4")) {
+  col <- rep(col, length.out = 2)
+  function(x) gpar(fill = ifelse(x > 0, col[1], col[2]))  
+}
+class(shading3b_fun) <- "grapcon_generator"
+@
+
+Note that in this simple binary shading example, the first four
+parameters are not used. In some sense, generating
+functions for shadings are parameterized both by the user and the
+strucplot framework. For shading functions that require model
+information, the user-specified parameters are 
+to be passed to the \code{gp\_args} argument instead, and for this to
+work, the generating function needs a class attribute to be distinguishable from
+the ``normal'' shading functions. For others
+(like our simple \codefun{shading3b\_fun}) this is optional, but
+recommended for consistency:
+
+<<shading5>>=
+mosaic(ucb, gp = shading3b_fun, gp_args = list(col = c("red","blue")))
+@
+
+\noindent The final \codefun{shading3b\_fun} pretty much resembles 
+\codefun{shading\_binary}, one of the standard shading
+functions provided by the \pkg{vcd} package.
+
+\subsection[An overview of the shading functions in vcd]{An overview of the shading functions in \pkg{vcd}}
+\label{sec:overview}
+
+\cite{vcd:Friendly:1994}
+suggested a residual-based shading for the mosaic tiles that can also be applied
+to the rectangles in association plots \citep{vcd:Meyer+Zeileis+Hornik:2003}.
+Apart from \codefun{shading\_binary}, there are currently two basic
+shadings available in \pkg{vcd}:
+\codefun{shading\_hcl} and \codefun{shading\_hsv}, as well
+as two derived functions: \codefun{shading\_Friendly} building upon
+\codefun{shading\_hsv}, and \codefun{shading\_max} building upon \codefun{shading\_hcl}.
+\codefun{shading\_hsv} and \codefun{shading\_hcl} provide
+the same conceptual tools, but use different color spaces: the
+Hue-Saturation-Value (HSV) and the Hue-Chroma-Luminance (HCL) scheme,
+respectively. We will first expose the basic concept of these shading
+functions using HSV space, and
+then briefly explain the differences to HCL space \citep[a detailed
+discussion can be found in][]{vcd:Zeileis+Meyer+Hornik:2005}. 
+Color palettes in HCL space are 
+preferable to palettes derived 
+from HSV space from a perceptual point of view. Functions for creating
+palettes (see, e.g., \codefun{diverge\_hcl}) are provided with the
+\pkg{vcd} package.
+
+In HSV space, colors are specified in three dimensions: Hue,
+Saturation (``colorfulness''), and Value (``lightness'', amount of gray). 
+These three dimensions are used by \codefun{shading\_hsv} to visualize information 
+about the residuals and the underlying independence model. The hue 
+indicates the residuals' sign: by default, blue for positive, and red
+for negative residuals. The saturation of a residual 
+is set according to its size: high saturation for large, and low
+saturation for small residuals. 
+Finally, the overall lightness is used to indicate the significance of
+a test statistic: light colors for significant, 
+and dark colors for non-significant results.
+
+As an example, we will visualize the association of hair and eye
+color in the \data{HairEyeColor} data set (see
+Figure~\ref{fig:haireye}, top):
+
+<<haireye1>>=
+haireye <- margin.table(HairEyeColor, 1:2)
+mosaic(haireye, gp = shading_hsv)
+@
+
+\noindent As introduced before, the default shading scheme is not
+\codefun{shading\_hsv} but \codefun{shading\_hcl} due to the better
+perceptual characteristics of HCL color space.
+The following example again illustrates the \data{HairEyeColor} data, 
+this time with HCL colors:
+
+<<haireye2>>=
+mosaic(haireye, gp = shading_hcl)
+@ 
+<<haireye3>>=
+mosaic(haireye, gp = shading_hcl, gp_args = list(h = c(130, 43), c = 100, l = c(90, 70)))
+@
+
+\noindent In Figure~\ref{fig:haireye}, the plot
+in the middle depicts the default palette, 
+and the bottom plot an alternative setting for Hue (\code{h}), Chroma (\code{c}), and
+Luminance (\code{l}).
+
+\setkeys{Gin}{width=0.5\textwidth}
+\begin{figure}[htbp]
+\begin{center}
+<<haireyefig1,fig=TRUE,echo=FALSE,height=6,width=8>>=
+mosaic(haireye, gp = shading_hsv, margin = c(bottom = 1), keep_aspect_ratio = FALSE)
+@
+<<haireyefig2,fig=TRUE,echo=FALSE,height=6,width=8>>=
+mosaic(haireye, gp = shading_hcl, margin = c(bottom = 1), keep_aspect_ratio = FALSE)
+@
+<<haireyefig3,fig=TRUE,echo=FALSE,height=6,width=8>>=
+mosaic(haireye, gp = shading_hcl, margin = c(bottom = 1), gp_args = list(h = c(130, 43), c = 100, l = c(90, 70)), keep_aspect_ratio = FALSE)
+@
+\caption{Three mosaic plots for the \data{HairEyeColor} data using
+  different color palettes. Top: default HSV color palette. Middle: default HCL
+ color palette. Bottom: a custom HCL color palette.}
+\label{fig:haireye}
+\end{center}
+\end{figure}
+\setkeys{Gin}{width=0.7\textwidth}
+
+\noindent Large positive residuals (greater than $4$) can be found for
+brown eyes/black hair and blue eyes/blond hair, and are colored in deep blue. 
+On the other hand, there is a large negative residual
+(less than $-4$) for brown eyes/blond hair, colored deep red. There are also three
+medium-sized positive (negative) residuals between 2 and 4 ($-2$ and
+$-4$): the colors for them are less saturated. Residuals between $-2$ and $2$ 
+are shaded in white (gray for HCL-shading).
+The heuristic for choosing the cut-off points $2$ and $4$ is that the Pearson residuals
+are approximately standard normal which implies that the highlighted cells are those with
+residuals \emph{individually} significant at approximately the $\alpha
+= 0.05$ and $\alpha = 0.0001$ levels, respectively. These default cut-off points can
+be changed to alternative values using the \code{interpolate}
+argument (see Figure~\ref{fig:interpolatecontinuous}):
+
+<<interpolate>>=
+mosaic(haireye, shade = TRUE, gp_args = list(interpolate = 1:4))
+@
+
+\noindent The elements of the numeric vector passed to
+\code{interpolate} define the knots of an interpolating step function
+used to map the absolute residuals to saturation levels. 
+The \code{interpolate} argument also accepts a user-defined function,
+which then is called with the absolute residuals to get a vector of
+cut-off points. Thus, it is possible to automatically choose the cut-off points 
+in a data-driven way. For example,
+one might think that the extension from four cut-off points 
+to a continuous shading---visualizing the whole range of 
+residuals---could be useful. We simply need a one-to-one
+mapping from the residuals to the saturation values:
+
+<<continuous1>>=
+ipol <- function(x) pmin(x/4, 1)
+@
+
+\noindent Note that this \codefun{ipol} 
+function maps residuals greater than 4 to a saturation
+level of 1. However, the resulting plot
+(Figure~\ref{fig:interpolatecontinuous}, right) is 
+deceiving:
+
+<<continuous2>>=
+mosaic(haireye, shade = TRUE, gp_args = list(interpolate = ipol),
+labeling_args = list(abbreviate = c(Sex = TRUE)))
+@
+
+\setkeys{Gin}{width=\textwidth}
+\begin{figure}[htbp]
+\begin{center}
+<<interpolatefig,fig=TRUE,echo=FALSE,width=13,height=6>>=
+pushViewport(viewport(layout = grid.layout(ncol = 2)))
+pushViewport(viewport(layout.pos.col = 1))
+mosaic(haireye, gp_args = list(interpolate = 1:4), margin = c(right = 1), keep_aspect_ratio= FALSE,newpage = FALSE,legend_width=5.5,shade = TRUE)
+popViewport(1)
+pushViewport(viewport(layout.pos.col = 2))
+mosaic(haireye, gp_args = list(interpolate = ipol), margin = c(left=3,right = 1), keep_aspect_ratio = FALSE, newpage = FALSE, shade = TRUE)
+popViewport(2)
+@
+\caption{\label{fig:interpolatecontinuous}The \data{HairEyeColor}
+  data. Left: shading with 4 cut-off points. Right: continuous shading.}
+
+\end{center}
+\end{figure} 
+\setkeys{Gin}{width=0.7\textwidth}
+
+\noindent Too much color makes it difficult to interpret the image, and the subtle color
+differences are hard to catch. Therefore, we only included shadings
+with discrete cut-off points.
+
+The third remaining dimension, the value, is used for
+visualizing the significance of a test statistic. The user can either
+directly specify the $p$ value, or, alternatively, a function that
+computes it, to the \code{p.value} argument. Such a function must take
+observed and expected values, residuals, and degrees of freedom (used
+by the independence model) as arguments. 
+If nothing is specified, the $p$ value is computed from 
+a $\chi^2$ distribution with \code{df} degrees of
+freedom. The \code{level} argument is used to specify the confidence
+level:  if \code{p.value} is smaller than \code{1 - level}, light colors are used, 
+otherwise dark colors are employed. 
+The following example using the \data{Bundesliga} data 
+shows the relationship of home goals and away
+goals of Germany's premier soccer league in 1995: although there are two
+``larger'' residuals (one greater than 2, one less then $-2$), the
+$\chi^2$ test does not reject the null hypothesis of
+independence. Consequently, the colors appear dark (see
+Figure~\ref{fig:bundesliga}, left):
+
+<<bundesliga>>=
+BL <- xtabs(~ HomeGoals + AwayGoals, data = Bundesliga, subset = Year == 1995)
+mosaic(BL, shade = TRUE)
+@
+
+\noindent Note that in extended mosaic plots, bullets drawn for zero
+cells are shaded, too, bringing out non-zero residuals, if any.
+
+A shading function building upon \codefun{shading\_hsv} is \codefun{shading\_Friendly},
+implementing the shading introduced by \cite{vcd:Friendly:1994}. In addition
+to the defaults of the HSV shading, it uses the border color and line type to 
+redundantly code the residuals' sign. The following example again uses 
+the \data{Bundesliga} data from above, this time using the Friendly 
+scheme and, in addition, an alternative legend (see
+Figure~\ref{fig:bundesliga}, right):
+
+<<friendly>>=
+mosaic(BL, gp = shading_Friendly, legend = legend_fixed, zero_size = 0)
+@
+
+\setkeys{Gin}{width=\textwidth}
+\begin{figure}[htbp]
+\begin{center}
+<<bundesligafig,fig=TRUE,echo=FALSE,height=5,width=12>>=
+pushViewport(viewport(layout = grid.layout(ncol = 2)))
+pushViewport(viewport(layout.pos.col = 1))
+mosaic(BL, margin = c(right = 1), keep_aspect_ratio= FALSE, newpage = FALSE, 
+       legend_width=5.5, shade = TRUE)
+popViewport(1)
+pushViewport(viewport(layout.pos.col = 2))
+mosaic(BL, gp = shading_Friendly, legend = legend_fixed, zero_size = 0, 
+       margin = c(right = 1), keep_aspect_ratio= FALSE, newpage = FALSE,  
+       legend_width=5.5)
+popViewport(2)
+@
+\caption{The \data{Bundesliga} data for 1995. Left: Non-significant
+  $\chi^2$ test. Right: using the Friendly
+  shading and a legend with fixed bins.}
+\label{fig:bundesliga}
+\end{center}
+\end{figure}
+\setkeys{Gin}{width=0.7\textwidth}
+
+\noindent (The \code{zero\_size = 0} argument removes the bullets indicating zero 
+observed values. This feature is not provided
+in the original \proglang{SAS} implementation of the Friendly mosaic plots.)
+
+% Figure~\ref{fig:shadingHSVHCL} depicts 
+% HSV space in the upper panel and HCL space in the lower panel.
+% On the left (right) side, we see the color scales for red (blue)
+% hue, respectively. The $x$-axis represents the colorfulness, and the
+% $y$-axis the brightness.
+% The boxes represent the diverging color palettes used for the shadings.
+% For HSV space, we can see that the effect of changing the 
+% level of brightness (`value') is not the same for different levels of 
+% saturation, and again not the same for the two different hues.
+% In fact, in HSV space all dimensions are confounded, which 
+% obviously is problematic for coding information. In contrast, HCL color
+% space offers perceptually uniform colors: as can be seen from the lower panel, 
+% the chroma is homogeneous for different levels of luminance. 
+% Unfortunately, this comes at the 
+% price of the space being irregularly shaped, making it difficult to automatically select 
+% diverging color palettes. 
+% <<auxiliary_functions,echo=FALSE,results=hide>>=
+% hue.slice <- function(hue, grid.n = 101, type = c("HCL", "HSV"), plot = TRUE, fixup = FALSE)
+% {
+%   type <- match.arg(type)
+%   if(type == "HCL") {
+%     chroma = seq(0, 100, length = grid.n)
+%     luminance = seq(0, 100, length = grid.n)
+%     nc <- length(chroma)
+%     nl <- length(luminance)
+%     color.slice <- outer(chroma, luminance, function(y, x) hcl(hue, x, y, fixup = fixup))
+%     xlab <- "chroma"
+%     ylab <- "luminance"
+%     main <- paste("hue =", round(hue, digits = 0))
+%   } else {
+%     chroma = seq(0, 1, length = grid.n)
+%     luminance = seq(0, 1, length = grid.n)
+%     nc <- length(chroma)
+%     nl <- length(luminance)
+%     color.slice <- outer(chroma, luminance, function(y, x) hsv(hue, x, y))
+%     xlab <- "saturation"
+%     ylab <- "value"
+%     main <- paste("hue =", round(hue, digits = 3))
+%   }
+%   if(plot) {
+%     plot(0.5, 0.5, xlim = range(chroma), ylim = range(luminance), type = "n", axes = FALSE,
+%          xlab = xlab, ylab = ylab, yaxs = "i", xaxs = "i", main = main)
+%     for(i in 1:(nc-1)) {
+%       rect(chroma[i], luminance[-nl], chroma[i] + 100/(nc-1), luminance[-1], border = color.slice[,i+1], col = color.slice[,i+1])
+%     }
+%     axis(1)
+%     axis(2)
+%     box()
+%   }
+%   colnames(color.slice) <- chroma
+%   rownames(color.slice) <- luminance
+%   attr(color.slice, "type") <- type
+%   class(color.slice) <- "slice"
+%   invisible(color.slice)
+% }
+% @
+
+% \setkeys{Gin}{width=.8\textwidth}
+% \begin{figure}[p]
+% \begin{center}
+% <<shading_HSV,fig=TRUE,echo=FALSE,height=7,width=8>>=
+% ## generate colors
+% hue23 <- hue.slice(2/3, grid.n = 101, plot = FALSE, type = "HSV")
+% hue0 <- hue.slice(0, grid.n = 101, plot = FALSE, type = "HSV")
+% saturation <- as.numeric(colnames(hue23))
+% value <- as.numeric(rownames(hue23))
+
+% ## select those with value >= 0.5
+% hue23 <- hue23[value >= .5, ]
+% hue0 <- hue0[value >= .5, ]
+% value <- value[value >= .5]
+% nl <- nrow(hue23)
+% nc <- ncol(hue23)
+
+% ## plot 2 slides from HSV space
+% plot(0.5, 0.5, xlim = c(-1, 1), ylim = c(0, 1), type = "n", axes = FALSE,
+%        xlab = "", ylab = "", yaxs = "i", xaxs = "i", main = "")
+% for(i in 1:(nc-1)) {
+%   rect(saturation[i], value[-nl], saturation[i] + 1/(nc-1), value[-1], border = hue23[,i+1], col = hue23[,i+1])
+% }
+% for(i in 1:(nc-1)) {
+%   rect(-saturation[i], value[-nl], -(saturation[i] + 1/(nc-1)), value[-1], border = hue0[,i+1], col = hue0[,i+1])
+% }
+% axis(2, at = c(50, 75, 100)/100, labels = c(0.5, 0.75, 1))
+% axis(4, at = c(50, 75, 100)/100, labels = c(0.5, 0.75, 1))
+% axis(3, at = -4:4*.25, labels=c(4:0*.25, 1:4*.25))
+% mtext(c("hue = 0", "hue = 2/3"), side = 3, at = c(-.5, .5), line = 3, cex = 1.2)
+% mtext("saturation", side = 3, at = 0, line = 2)
+% mtext("value", side = 2, at = .75, line = 2)
+% mtext("value", side = 4, at = .75, line = 2)
+% lines(c(-1, 1), c(.5, .5))
+
+% ## significant colors
+% rect(-1, 0.95, -.90, 1, col = hsv(0, 1, 1))
+% rect(-0.45, 0.95, -.55, 1, col = hsv(0, 0.5, 1))
+% rect(-.05, .95, .05, 1, col = hsv(2/3, 0, 1))
+% rect(0.45, 0.95, .55, 1, col = hsv(2/3, 0.5, 1))
+% rect(.90, .95, 1, 1, col = hsv(2/3, 1, 1))
+
+% text(-1, .33, "significant", pos = 4, cex = 1.2)
+% rect(-1, .20, -.80, .30, col = hsv(0, 1, 1))
+% rect(-.40, .20, -0.6, .30, col = hsv(0, 0.5, 1))
+% rect(-.20, .20, 0, .30, col = hsv(0, 0, 1))
+% rect(0, .20, .20, .30, col = hsv(2/3, 0, 1))
+% rect(0.4, .20, .60, .30, col = hsv(2/3, .5, 1))
+% rect(.80, .20, 1, .30, col = hsv(2/3, 1, 1))
+
+% lines(c(-.9, -.55), c(0.975, .975), lty = 2)
+% lines(c(-.45, -.05), c(0.975, .975), lty = 2)
+% lines(c(.45, .05), c(0.975, .975), lty = 2)
+% lines(c(.9, .55), c(0.975, .975), lty = 2)
+
+% ## non-significant colors
+% rect(-1, 0.5, -.90, 0.55, col = hsv(0, 1, 0.5))
+% rect(-0.45, 0.5, -.55, 0.55, col = hsv(0, 0.5, 0.5))
+% rect(-.05, .5, .05, 0.55, col = hsv(2/3, 0, 0.5))
+% rect(0.45, 0.5, .55, 0.55, col = hsv(2/3, 0.5, 0.5))
+% rect(.90, .5, 1, 0.55, col = hsv(2/3, 1, 0.5))
+
+% text(-1, .13, "non-significant", pos = 4, cex = 1.2)
+% rect(-1, 0, -.80, .10, col = hsv(0, 1, 0.5))
+% rect(-.60, 0, -.4, .10, col = hsv(0, 0.5, 0.5))
+% rect(-.20, 0, 0, .10, col = hsv(0, 0, 0.5))
+% rect(0, 0, .20, .10, col = hsv(2/3, 0, 0.5))
+% rect(0.4, 0, .60, .1, col = hsv(2/3, .5, 0.5))
+% rect(.80, 0, 1, .10, col = hsv(2/3, 1, 0.5))
+
+% lines(c(-.9, -.55), c(0.525, .525), lty = 2)
+% lines(c(-.45, -.05), c(0.525, .525), lty = 2)
+% lines(c(.45, .05), c(0.525, .525), lty = 2)
+% lines(c(.9, .55), c(0.525, .525), lty = 2)
+% @
+
+% <<shading_HCL,fig=TRUE,echo=FALSE,height=7,width=8>>=
+% ## generate colors
+% hue260 <- hue.slice(260, grid.n = 101, plot = FALSE)
+% hue360 <- hue.slice(360, grid.n = 101, plot = FALSE)
+% mychroma <- as.numeric(colnames(hue260))
+% luminance <- as.numeric(rownames(hue260))
+
+% ## select those with lumincance >= 50
+% hue260 <- hue260[luminance >= 50, ]
+% hue360 <- hue360[luminance >= 50, ]
+% luminance <- luminance[luminance >= 50]
+% nc <- ncol(hue260)
+% nl <- nrow(hue260)
+
+% ## plot 2 slides from HCL space
+% plot(0.5, 0.5, xlim = c(-100, 100), ylim = c(0, 100), type = "n", axes = FALSE,
+%        xlab = "", ylab = "", yaxs = "i", xaxs = "i", main = "")
+% for(i in 1:(nc-1)) {
+%   rect(mychroma[i], luminance[-nl], mychroma[i] + 100/(nc-1), luminance[-1], border = hue260[,i+1], col = hue260[,i+1])
+% }
+% for(i in 1:(nc-1)) {
+%   rect(-mychroma[i], luminance[-nl], -(mychroma[i] + 100/(nc-1)), luminance[-1], border = hue360[,i+1], col = hue360[,i+1])
+% }
+% axis(2, at = c(50, 70, 90, 100), labels = c(50, 70, 90, 100))
+% axis(4, at = c(50, 70, 90, 100), labels = c(50, 70, 90, 100))
+% axis(3, at = -4:4*25, labels=c(4:0*25, 1:4*25))
+% mtext(c("hue = 0", "hue = 260"), side = 3, at = c(-50, 50), line = 3, cex = 1.2)
+% mtext("chroma", side = 3, at = 0, line = 2)
+% mtext("luminance", side = 2, at = 75, line = 2)
+% mtext("luminance", side = 4, at = 75, line = 2)
+% lines(c(-100, 100), c(50, 50))
+
+% ## significant colors
+% rect(-100, 47.5, -90, 52.5, col = hcl(0, 100, 50))
+% rect(-55, 67.5, -45, 72.5, col = hcl(0, 50, 70))
+% rect(-5, 95, 5, 100, col = hcl(260, 0, 100))       ## grey vs. white
+% rect(-5, 87.5, 5, 92.5, col = hcl(260, 0, 90))     ## grey vs. white
+% rect(45, 67.5, 55, 72.5, col = hcl(260, 50, 70))
+% rect(90, 47.5, 100, 52.5, col = hcl(260, 100, 50))
+
+% text(-100, 33, "significant", pos = 4, cex = 1.2)
+% rect(-100, 20, -80, 30, col = hcl(0, 100, 50))
+% rect(-60, 20, -40, 30, col = hcl(0, 50, 70))
+% rect(-20, 20, 0, 30, col = hcl(0, 0, 90))       
+% rect(0, 20, 20, 30, col = hcl(260, 0, 90))
+% #white# rect(-20, 20, 0, 30, col = hcl(0, 0, 100))
+% #white# rect(0, 20, 20, 30, col = hcl(260, 0, 100))
+% rect(40, 20, 60, 30, col = hcl(260, 50, 70))
+% rect(80, 20, 100, 30, col = hcl(260, 100, 50))
+
+% lines(c(-45, -5), c(72.5, 87.5), lty = 2)
+% lines(c(45, 5), c(72.5, 87.5), lty = 2)
+% lines(c(-95, -55), c(52.5, 67.5), lty = 2)
+% lines(c(95, 55), c(52.5, 67.5), lty = 2)
+
+% ## non-significant colors
+% rect(-25, 47.5, -15, 52.5, col = hcl(0, 20, 50))
+% rect(-15, 67.5, -5, 72.5, col = hcl(0, 10, 70))
+% rect(5, 67.5, 15, 72.5, col = hcl(260, 10, 70))
+% rect(25, 47.5, 15, 52.5, col = hcl(260, 20, 50))
+
+% text(-100, 13, "non-significant", pos = 4, cex = 1.2)
+% rect(-60, 0, -40, 10, col = hcl(0, 20, 50))
+% rect(-40, 0, -20, 10, col = hcl(0, 10, 70))
+% rect(-20, 0, 0, 10, col = hcl(0, 0, 90))
+% rect(0, 0, 20, 10, col = hcl(260, 0, 90))
+% rect(20, 0, 40, 10, col = hcl(260, 10, 70))
+% rect(40, 0, 60, 10, col = hcl(260, 20, 50))
+
+% lines(c(-18.75, -11.25), c(52.5, 67.5), lty = 2)
+% lines(c(-8.75, -1.25), c(72.5, 87.5), lty = 2)
+% lines(c(18.75, 11.75), c(52.5, 67.5), lty = 2)
+% lines(c(8.75, 1.25), c(72.5, 87.5), lty = 2)
+% @ 
+% \caption{Residual-based shadings in HSV (upper) and HCL space (lower).}
+% \label{fig:shadingHSVHCL}
+% \end{center}
+% \end{figure}
+
+A more ``advanced'' function building upon \codefun{shading\_hcl} 
+is \codefun{shading\_max}, using the
+maximum statistic both to conduct the independence test and to 
+visualize significant \emph{cells} causing the
+rejection of the independence hypothesis 
+\citep{vcd:Meyer+Zeileis+Hornik:2003,vcd:Zeileis+Meyer+Hornik:2005}. The
+\code{level} argument of \codefun{shading\_max} 
+then can be used to specify several confidence
+levels from which the corresponding cut-off points are computed. 
+By default, two cut-off points are computed corresponding 
+to confidence levels of $90\%$ and $99\%$, respectively.
+In the following example, we investigate the effect of a new treatment
+for rheumatoid arthritis on a group of female patients using the
+maximum shading (see Figure~\ref{fig:maximum}):
+
+<<arthritis>>=
+set.seed(4711)
+mosaic(~ Treatment + Improved, data = Arthritis, subset = Sex == "Female",
+       gp = shading_max)
+@
+
+\begin{figure}[h]
+\begin{center}
+<<arthritisfig,fig=TRUE,echo=FALSE>>=
+<<arthritis>>
+@
+\caption{The \data{Arthritis} data (female patients) with significant maximum test.}
+\label{fig:maximum}
+\end{center}
+\end{figure}
+
+\noindent The maximum test is significant although the residuals are
+all in the $\left[-2,2\right]$ interval.
+The \codefun{shading\_hcl} function with default cut-off points would
+not have shown any color. In addition, since the test statistic is the
+maximum of the absolute Pearson residuals, \emph{each} colored
+residual violates the null hypotheses of independence, and thus, the
+``culprits'' can immediately be identified.
+
+\section[Labeling]{Labeling}
+\label{sec:labeling}
+
+One of the major enhancements in package \pkg{vcd} compared to
+\codefun{mosaicplot} and \codefun{assocplot} in
+base \proglang{R} is the labeling in the strucplot framework which 
+offers more features and greater flexibility. 
+Like shading, spacing, and drawing of legend and core plot, 
+labeling is now carried out by grapcon functions, rendering labeling completely
+modular.
+The user supplies either a labeling function, or, alternatively, 
+a generating function that parameterizes a labeling function, 
+to \codefun{strucplot} which then draws the labels.
+Labeling is well-separated from the actual plotting that occurs in the
+low-level core functions. It only relies on the viewport tree produced by them, and the
+\code{dimnames} attribute of the visualized table. Labeling
+functions are grapcons that ``add ink to the canvas'': the drawing of the
+labels happens after the actual plot has been drawn by the core function.
+Thus, it is possible to supply one's own labeling function, or
+to combine some of the basic functions to produce a more complex labeling.
+In the following, we describe the three basic modules
+(\codefun{labeling\_text}, \codefun{labeling\_list}, and \codefun{labeling\_cells}) and
+derived functions that build upon them.
+
+\subsection[Labels in the borders]{Labels in the borders: \texttt{labeling\_text()}}
+
+\codefun{labeling\_text} is the default for all strucplot displays. It
+plots labels in the borders similar to the \codefun{mosaicplot}
+function in base \proglang{R}, but is much more flexible: it is not limited
+to 4 dimensions, and the positioning and graphical parameters 
+of levels and variable names are customizable. In addition,
+the problem of overlapping labels can be handled in several ways.
+
+As an example, again consider the \data{Titanic} data:
+by default, the variable names and levels are plotted ``around'' the
+plot in a counter-clockwise way (see Figure~\ref{fig:labels1}, top left):
+
+<<default>>=
+mosaic(Titanic)
+@
+
+% \begin{figure}[p]
+% \begin{center}
+% <<defaultfig,fig=TRUE,echo=FALSE>>=
+% <<default>>
+% @
+% \caption{Mosaic plot for the \data{Titanic} data with default settings
+%   for labeling.}
+% \label{fig:defaults}
+% \end{center}
+% \end{figure}
+
+\noindent Note that the last two levels of the \code{survived} variable do
+overlap, as well as some adult and child labels of the \code{age} Variable.
+This issue can be addressed in several ways. The ``brute force''
+method is to enable clipping for these dimensions (see
+Figure~\ref{fig:labels1}, top right):
+
+<<clipping>>=
+mosaic(Titanic, labeling_args = list(clip = c(Survived = TRUE, Age = TRUE)))
+@
+
+% \begin{figure}[p]
+% \begin{center}
+% <<clippingfig,fig=TRUE,echo=FALSE>>=
+% <<clipping>>
+% @
+% \caption{The effect of clipping.}
+%  \label{fig:clipping}
+% \end{center}
+% \end{figure}
+
+\noindent The \code{clip} parameter is passed to the labeling function
+via the \code{labeling\_args} argument which takes a list of
+parameters. \code{clip} itself takes a vector of logicals (one for
+each dimension). 
+% as mentioned before
+Almost all vectorized arguments in the strucplot
+framework can be abbreviated in the following way: unnamed components
+(or the defaults, if there are none) are recycled as needed, but
+overridden by the named components. Here, the default is \code{FALSE},
+and therefore clipping is enabled only for the \code{survived} and \code{age} variables.
+A more sensible solution to the overlap problem is to abbreviate the 
+levels (see Figure~\ref{fig:labels1}, middle left):
+
+<<abbreviating>>=
+mosaic(Titanic, labeling_args = list(abbreviate = c(Survived = TRUE, Age = 3)))
+@
+
+% \begin{figure}[h]
+% \begin{center}
+% <<abbreviatingfig,fig=TRUE,echo=FALSE>>=
+% <<abbreviating>>
+% @
+% \caption{Abbreviating.}
+% \label{fig:abbreviating}
+% \end{center}
+% \end{figure}
+
+\noindent The \code{abbreviate} argument takes a vector of
+integers indicating the number of significant characters the levels should be
+abbreviated to (\code{TRUE} is interpreted as 1, obviously). Abbreviation
+is performed using the \codefun{abbreviate} function in base \proglang{R}. Another
+possibility is to rotate the levels (see Figure~\ref{fig:labels1}, bottom):
+
+<<rotate>>=
+mosaic(Titanic, labeling_args = list(rot_labels = c(bottom = 90, right = 0), 
+offset_varnames = c(right = 1), offset_labels = c(right = 0.3)), 
+margins = c(right = 4, bottom = 3))
+@
+
+% \begin{figure}[p]
+% \begin{center}
+% <<rotatefig,fig=TRUE,echo=FALSE,height=6>>=
+% <<rotate>>
+% @
+% \caption{Rotating labels.}
+% \label{fig:rotating}
+% \end{center}
+% \end{figure}
+
+\noindent Finally, we could also inhibit the output of repeated levels 
+(see Figure~\ref{fig:labels1}, middle right):
+
+<<repeat>>=
+mosaic(Titanic, labeling_args = list(rep = c(Survived = FALSE, Age = FALSE)))
+@
+
+\setkeys{Gin}{width=0.9\textwidth}
+\begin{figure}[p]
+\begin{center}
+<<label1fig,fig=TRUE,echo=FALSE,height=12,width=8>>=
+pushViewport(viewport(layout = grid.layout(ncol = 2,nrow=3)))
+pushViewport(viewport(layout.pos.col = 1, layout.pos.row = 1))
+mosaic(Titanic, newpage = FALSE, keep = TRUE, margin = c(right = 3), 
+       gp_labels = gpar(fontsize = 10))
+popViewport()
+pushViewport(viewport(layout.pos.col = 2, layout.pos.row = 1))
+mosaic(Titanic, labeling_args = list(clip = c(Survived = TRUE, Age = TRUE)), 
+       newpage = FALSE, keep = TRUE, margin = c(left = 3), gp_labels = gpar(fontsize = 10))
+popViewport()
+pushViewport(viewport(layout.pos.col = 1, layout.pos.row = 2))
+mosaic(Titanic, labeling_args = list(abbreviate = c(Survived = TRUE, Age = 2)), 
+       newpage = FALSE, keep = TRUE, margin = c(right = 3), gp_labels = gpar(fontsize = 10))
+popViewport()
+pushViewport(viewport(layout.pos.col = 2, layout.pos.row = 2))
+mosaic(Titanic, labeling_args = list(rep = c(Survived = FALSE, Age = FALSE)), 
+       newpage = FALSE, keep = TRUE, margin = c(left = 3), gp_labels = gpar(fontsize = 10))
+popViewport()
+pushViewport(viewport(layout.pos.col = 1:2, layout.pos.row = 3))
+pushViewport(viewport(width = 0.55))
+mosaic(Titanic, labeling_args = list(rot_labels = c(bottom = 90, right = 0), 
+                  offset_varnames = c(right = 1), offset_labels = c(right = 0.3)), 
+       margins = c(right = 4, bottom = 3), newpage = FALSE, keep = FALSE, gp_labels = gpar(fontsize = 10))
+popViewport(3)
+@
+\caption{Examples for possible labeling strategies for the Titanic data mosaic. 
+  Top left: default labeling (many labels overlap). 
+  Top right: with clipping turned on. Middle left: \texttt{Age} and
+  \texttt{Survived} labels abbreviated. Middle right: \texttt{Age} labels
+  not repeated. Bottom: \texttt{Age} and \texttt{Survived} labels rotated.}
+\label{fig:labels1}
+\end{center}
+\end{figure}
+\setkeys{Gin}{width=0.7\textwidth}
+
+We now proceed with a few more ``cosmetic'' features (which do not all
+produce satisfactory results for our sample data).
+A first simple, but effectful modification is to position
+all labels and variables left-aligned (see Figure~\ref{fig:labels2},
+top left):
+
+<<left>>=
+mosaic(Titanic, labeling_args = list(pos_varnames = "left", pos_labels = "left", 
+       just_labels = "left", rep = FALSE))
+@
+
+% \begin{figure}[h]
+% \begin{center}
+% <<leftfig,fig=TRUE,echo=FALSE>>=
+% <<left>>
+% @
+% \caption{Left-aligning.}
+% \label{fig:left}
+% \end{center}
+% \end{figure}
+
+\noindent Note that obviously we need to change the justification 
+to \code{"left"} as well. We can
+achieve the same effect by using the convenience function \codefun{labeling\_left}:
+
+<<left2>>=
+mosaic(Titanic, labeling = labeling_left)
+@
+
+\noindent Next, we show how to put all levels to the
+bottom and right margins, and all variable names to the top and left
+margins (see Figure~\ref{fig:labels2}, top right):
+
+<<margins>>=
+mosaic(Titanic, labeling_args = list(tl_labels = FALSE, tl_varnames = TRUE, abbreviate = c(Survived = 1, Age = 3)))
+@
+
+% \begin{figure}[p]
+% \begin{center}
+% <<marginsfig,fig=TRUE,echo=FALSE,height=6>>=
+% <<margins>>
+% @
+% \caption{Changes in the margins.}
+% \label{fig:margins}
+% \end{center}
+% \end{figure}
+
+\noindent The tl\_\var{foo} (``top left'') arguments are \code{TRUE} by default. 
+Now, we will add boxes to the labels and additionally
+enable clipping (see Figure~\ref{fig:labels2}, bottom left):
+
+<<boxes>>=
+mosaic(Titanic, labeling_args = list(tl_labels = FALSE, tl_varnames = TRUE, 
+       boxes = TRUE, clip = TRUE))
+@
+
+% \begin{figure}[p]
+% \begin{center}
+% <<boxesfig,fig=TRUE,echo=FALSE,height=6>>=
+% <<boxes>>
+% @
+% \caption{Boxes and Clipping.}
+% \label{fig:boxes}
+% \end{center}
+% \end{figure}
+
+\noindent The values to \code{boxes} and \code{clip} are recycled for
+all dimensions. The result is pretty close to what calling
+\codefun{mosaic} with the \codefun{labeling\_cboxed}
+wrapper does, except that variables and levels, by default, are put to the top and
+to the left of the plot:
+
+<<boxes2>>=
+mosaic(Titanic, labeling = labeling_cboxed)
+@
+
+\noindent Another variant is to put the variable names into the
+same line as the levels (see Figure~\ref{fig:labels2}, bottom
+right---clipping for \code{Survived} and \code{Age} is, additionally,
+disabled, and \code{Age} abbreviated):
+
+<<labbl>>=
+mosaic(Titanic, labeling_args = list(tl_labels = TRUE, boxes = TRUE,
+       clip = c(Survived = FALSE, Age = FALSE, TRUE), abbreviate = c(Age = 4),
+       labbl_varnames = TRUE), margins = c(left = 4, right = 1, 3))
+@
+
+% \begin{figure}[h]
+% \begin{center}
+% <<labblfig,fig=TRUE,echo=FALSE>>=
+% <<labbl>>
+% @
+% \caption{Variable names beneath levels, and clipping disabled for the
+%   survival variable.}
+% \label{fig:labbl}
+% \end{center}
+% \end{figure}
+
+\noindent \code{labbl\_varnames} (``variable names to the bottom/left of the
+labels'') is a vector of logicals indicating the side for the
+variable names. The resulting layout is close to what
+\codefun{labeling\_lboxed} produces, except that variables and levels,
+by default, are left-aligned and put to the bottom and to the right of the plot:
+
+<<labbl2>>=
+mosaic(Titanic, labeling = labeling_lboxed, margins = c(right = 4, left = 1, 3))
+@
+
+\noindent A similar design is used by the \codefun{doubledecker} function.
+
+\setkeys{Gin}{width=\textwidth}
+\begin{figure}[p]
+\begin{center}
+<<label2fig,fig=TRUE,echo=FALSE,height=12,width=12>>=
+pushViewport(viewport(layout = grid.layout(ncol = 2, nrow = 2)))
+pushViewport(viewport(layout.pos.col = 1, layout.pos.row = 1))
+mosaic(Titanic, labeling_args = list(pos_varnames = "left", pos_labels = "left", 
+       just_labels = "left", rep = FALSE), newpage = FALSE, keep = TRUE,
+       gp_labels = gpar(fontsize = 12))
+popViewport()
+pushViewport(viewport(layout.pos.col = 2, layout.pos.row = 1))
+mosaic(Titanic, labeling_args = list(tl_labels = FALSE, tl_varnames = TRUE, 
+                  abbreviate = c(Survived = 1, Age = 3)), newpage = FALSE, keep = TRUE,
+       margins = c(left = 4, right = 1, 3), gp_labels = gpar(fontsize = 12))
+popViewport()
+pushViewport(viewport(layout.pos.col = 1, layout.pos.row = 2))
+mosaic(Titanic, labeling_args = list(tl_labels = FALSE, tl_varnames = TRUE, 
+       boxes = TRUE, clip = TRUE), newpage = FALSE, keep = TRUE, 
+       gp_labels = gpar(fontsize = 12))
+popViewport()
+pushViewport(viewport(layout.pos.col = 2, layout.pos.row = 2))
+mosaic(Titanic, labeling_args = list(tl_labels = TRUE, boxes = TRUE,
+       clip = c(Survived = FALSE, Age = FALSE, TRUE), 
+       labbl_varnames = TRUE, abbreviate = c(Age = 4)), 
+       margins = c(left = 4, right = 1, 3), 
+       newpage = FALSE, keep = TRUE, gp_labels = gpar(fontsize = 12))
+popViewport(2)
+@
+\caption{Advanced strategies for labeling of the Titanic data. Top
+  left: left aligning of both variable names and labels. 
+  Top right: changes in the margins (all variable names are in the top
+  and left margins, and all labels in the bottom and right margins). 
+  Bottom left: clipping turned on, and boxes used. Bottom right: 
+  variable names beneath levels, clipping disabled for the
+  survival and age variables, and \texttt{Age} abbreviated.}
+\label{fig:labels2}
+\end{center}
+\end{figure}
+\setkeys{Gin}{width=0.7\textwidth}
+
+\subsection[Labels in the cells]{Labels in the cells: \texttt{labeling\_cells()}}
+
+This labeling draws both variable names and levels in the
+cells. As an example, we use the \data{PreSex} data on pre- and
+extramarital sex and divorce (see Figure~\ref{fig:labels3}, top left):
+
+<<cell>>=
+mosaic(~ MaritalStatus + Gender, data = PreSex, labeling = labeling_cells)
+@
+
+% \begin{figure}[h]
+% \begin{center}
+% <<cellfig,fig=TRUE,echo=FALSE,width=7,height=7>>=
+% <<cell>>
+% @
+% \caption{Cell labeling for the \data{PreSex} data.}
+% \label{fig:cell}
+% \end{center}
+% \end{figure}
+
+\noindent In the case of narrow cells, it might be useful to
+abbreviate labels and/or variable names and turn off clipping (see
+Figure~\ref{fig:labels3}, top right):
+
+<<cell2>>=
+mosaic(~ PremaritalSex + ExtramaritalSex, data = PreSex, 
+labeling = labeling_cells(abbreviate_labels = TRUE, abbreviate_varnames = TRUE, clip = FALSE))
+@
+
+% \begin{figure}[h]
+% \begin{center}
+% <<cell2fig,fig=TRUE,echo=FALSE>>=
+% <<cell2>>
+% @
+% \caption{Cell labeling for the \data{PreSex} data, labels abbreviated.}
+% \label{fig:cell2}
+% \end{center}
+% \end{figure}
+
+\noindent For some data, it might be convenient to combine cell
+labeling with border labeling as done by \codefun{labels\_conditional}
+(see Figure~\ref{fig:labels3}, bottom left):
+
+<<conditional>>=
+mosaic(~ PremaritalSex + ExtramaritalSex | MaritalStatus + Gender, data = PreSex, labeling = labeling_conditional(abbreviate_varnames = TRUE, abbreviate_labels = TRUE, clip = FALSE, gp_text = gpar(col = "red")))
+@
+
+% \begin{figure}[h]
+% \begin{center}
+% <<conditionalfig,fig=TRUE,echo=FALSE>>=
+% <<conditional>>
+% @
+% \caption{Conditional labeling for the \data{PreSex} data, labels (in
+%   red for clarity) abbreviated.}
+% \label{fig:conditional}
+% \end{center}
+% \end{figure}
+
+\noindent Additionally, the cell labeling allows the user to add
+arbitrary text to the cells by supplying a character array in the same
+shape as the data array to the \code{text} argument (cells with missing values
+are ignored). In the following example using the \code{Titanic} data, 
+this is used to add all observed values greater
+than 5 to the cells after the mosaic has been plotted (see
+Figure~\ref{fig:labels3}, bottom right):
+
+<<text>>=
+mosaic(Titanic, labeling_args = list(abbreviate = c(Survived = 1, Age = 4)), pop = FALSE)
+
+tab <- ifelse(Titanic < 6, NA, Titanic)
+labeling_cells(text = tab, clip = FALSE)(Titanic)
+@
+
+% \begin{figure}[h]
+% \begin{center}
+% <<textfig,fig=TRUE,echo=FALSE>>=
+% <<text>>
+% @
+% \caption{User-supplied text (observed frequencies exceeding 5) 
+% added to a mosaic display of the \data{Titanic} data.}
+% \label{fig:text}
+% \end{center}
+% \end{figure}
+
+\setkeys{Gin}{width=\textwidth}
+\begin{figure}[p]
+\begin{center}
+<<label3fig,fig=TRUE,echo=FALSE,height=8,width=8>>=
+grid.newpage()
+pushViewport(viewport(layout = grid.layout(ncol = 2, nrow = 2)))
+pushViewport(viewport(layout.pos.col = 1, layout.pos.row = 1))
+mosaic(~ MaritalStatus + Gender, data = PreSex, labeling = labeling_cells, newpage = FALSE, keep = TRUE, gp_labels = gpar(fontsize = 10))
+popViewport()
+pushViewport(viewport(layout.pos.col = 2, layout.pos.row = 1))
+mosaic(~ PremaritalSex + ExtramaritalSex, data = PreSex, 
+       labeling = labeling_cells(abbreviate_labels = TRUE, abbreviate_varnames = TRUE, 
+         clip = FALSE), newpage = FALSE, keep = TRUE, gp_labels = gpar(fontsize = 10))
+popViewport()
+pushViewport(viewport(layout.pos.col = 1, layout.pos.row = 2))
+mosaic(~ PremaritalSex + ExtramaritalSex | MaritalStatus + Gender, data = PreSex, 
+       labeling = labeling_conditional(abbreviate_varnames = TRUE, 
+         abbreviate_labels = TRUE, clip = FALSE, gp_text = gpar(col = "red")), 
+       newpage = FALSE, keep = TRUE, gp_labels = gpar(fontsize = 10))
+popViewport()
+pushViewport(viewport(layout.pos.col = 2, layout.pos.row = 2))
+mosaic(Titanic, labeling_args = list(abbreviate = c(Survived = 1, Age = 3)), 
+       pop = FALSE, newpage = FALSE, keep = TRUE, gp_labels = gpar(fontsize = 10))
+tab <- ifelse(Titanic < 6, NA, Titanic)
+labeling_cells(text = tab, clip = FALSE)(Titanic)
+@
+\caption{Cell labeling. Top left: default labeling using the
+  \data{PreSex} data. Top right: abbreviated labels. Bottom left:
+  conditional labeling (labels abbreviated and in red for clarity).
+  Bottom right: user-supplied text (observed frequencies exceeding 5) 
+  added to a mosaic display of the \data{Titanic} data. Note that
+  clipping is on by default (top left), and has explicitly been turned
+  off for the three other plots.}
+\label{fig:labels3}
+\end{center}
+\end{figure}
+\setkeys{Gin}{width=0.7\textwidth}
+
+\subsection[A simple list of labels]{A simple list of labels: \texttt{labeling\_list()}}
+
+If problems with overlapping labels cannot satisfactorily resolved,
+the last remedy could be to simply list the levels below the plot
+(see Figure~\ref{fig:list}):
+
+<<list>>=
+mosaic(Titanic, labeling = labeling_list, margins = c(bottom = 5))
+@
+
+\setkeys{Gin}{width=0.7\textwidth}
+\begin{figure}[p]
+\begin{center}
+<<listfig,fig=TRUE,echo=FALSE>>=
+mosaic(Titanic, labeling = labeling_list, margins = c(bottom = 5), keep = TRUE)
+@
+\caption{Labels indicated below the plot.}
+\label{fig:list}
+\end{center}
+\end{figure}
+
+\noindent The number of columns can be specified.
+
+\section{Spacing}
+\label{sec:spacing}
+
+Spacing of strucplot displays is customizable in a similar way than shading. The
+\code{spacing} argument of the \codefun{strucplot} function takes a
+list of \class{unit} vectors, one for each dimension, specifying the
+space between the tiles corresponding to the levels. 
+Consider again the introductory example of the \data{Arthritis} data 
+(Figure~\ref{fig:arthritis}). Since we are interested in the effect of
+the medicament in the placebo and treatment groups, a mosaic plot is
+certainly appropriate to visualize the three levels of \code{Improved} in
+the two \code{Treatment} strata. Another conceptual approach is to use
+spine plots with highlighting \citep{vcd:hummel:1996}. A spine plot is
+a variation of a bar plot where the heights of the bars are held
+constant, whereas the widths are used to represent the number of cases
+in each category. This is equivalent to a mosaic plot for a one-way table.
+If a second (indicator) variable is highlighted in a spine plot, we
+obtain a display equivalent to a simple mosaic display for a two-way
+table, except that no space between the levels of the highlighted
+variable is used.
+In the \data{Arthritis} example, we will highlight patients with \code{Marked} 
+improvement in both groups. To obtain such a display within 
+the strucplot framework, it suffices to set the space between the \code{Improved}
+tiles to 0 (see Figure~\ref{fig:artspine}):
+
+<<artspine,echo=TRUE,results=verbatim>>=
+(art <- structable(~Treatment + Improved, data = Arthritis, split_vertical = TRUE))
+(my_spacing <- list(unit(0.5, "lines"), unit(c(0, 0), "lines")))
+my_colors <- c("lightgray", "lightgray", "black")
+mosaic(art, spacing = my_spacing, gp = gpar(fill = my_colors, col = my_colors))
+@
+
+\begin{figure}[p]
+\begin{center}
+<<artspinefig,echo=FALSE,fig=TRUE>>=
+<<artspine>>
+@ 
+\caption{Spine plot for the \data{Arthritis} data using the strucplot framework.}
+\label{fig:artspine}
+\end{center}
+\end{figure}
+
+\noindent Note that the default and formula methods for 
+\codefun{mosaic} provide a
+convenience interface for highlighting. A similar plot (with slightly
+different shading) than the previous one can be obtained using:
+
+<<artspine>>=
+mosaic(Improved ~ Treatment, data = Arthritis, split_vertical = TRUE)
+@
+
+\noindent The strucplot framework also provides a set of spacing grapcon
+generators which compute suitable spacing objects for
+typical applications. The simplest spacing is \codefun{spacing\_equal}
+that uses the same space between all tiles (see
+Figure~\ref{fig:spacing}, top left):
+
+<<space1>>=
+mosaic(art, spacing = spacing_equal(unit(2, "lines")))
+@
+
+\noindent \codefun{spacing\_equal} is the default grapcon generator for
+two-dimensional tables. Slightly more flexible is \codefun{spacing\_dimequal} that
+allows an individual setting for each dimension (see
+Figure~\ref{fig:spacing}, top right):
+
+<<space2>>=
+mosaic(art, spacing = spacing_dimequal(unit(1:2, "lines")))
+@
+
+\noindent The default for multi-way contingency tables is
+\codefun{spacing\_increase} which uses increasing spaces for the
+dimensions. The user can specify a start value and the increase factor
+(see Figure~\ref{fig:spacing}, bottom left):
+
+<<space3>>=
+mosaic(art, spacing = spacing_increase(start = unit(0.5, "lines"), rate = 1.5))
+@
+
+\noindent For the arthritis example above, we could as well have used
+\codefun{spacing\_highlighting} which is similar to \codefun{spacing\_increase} but
+sets the spacing in the last splitting dimension to 0 (see
+Figure~\ref{fig:spacing}, bottom right):
+
+<<spine4>>=
+mosaic(art, spacing = spacing_highlighting, gp = my_colors)
+@
+
+\noindent Finally, \codefun{spacing\_conditional} can be used for
+visualizing conditional independence: it combines
+\codefun{spacing\_equal} (for the conditioned dimensions) and 
+\codefun{spacing\_increase} (for the conditioning dimensions). As an example, 
+consider Figure~\ref{fig:presex}: the spacing clearly allows to better 
+distinguish the conditioning variables 
+(\code{Gender} and \code{MaritalStatus}) from the conditioned 
+variables (\code{PremaritalSex} and \code{ExtramaritalSex}). This
+spacing is the default when conditional variables are specified for a
+strucplot display (see Section \ref{sec:strucplot}).
+
+\setkeys{Gin}{width=\textwidth}
+\begin{figure}[p]
+\begin{center}
+<<spacingfig,fig=TRUE,echo=FALSE,height=12,width=12>>=
+pushViewport(viewport(layout = grid.layout(ncol = 2, nrow = 2)))
+pushViewport(viewport(layout.pos.col = 1, layout.pos.row = 1))
+mosaic(art, spacing = spacing_equal(unit(2, "lines")), keep = TRUE, newpage = FALSE)
+popViewport()
+pushViewport(viewport(layout.pos.col = 2, layout.pos.row = 1))
+mosaic(art, spacing = spacing_dimequal(unit(c(0.5, 2), "lines")), keep = TRUE, newpage = FALSE)
+popViewport()
+pushViewport(viewport(layout.pos.col = 1, layout.pos.row = 2))
+mosaic(art, spacing = spacing_increase(start = unit(0.3, "lines"), rate = 2.5), keep = TRUE, newpage = FALSE)
+popViewport()
+pushViewport(viewport(layout.pos.col = 2, layout.pos.row = 2))
+mosaic(art, spacing = spacing_highlighting, keep = TRUE, newpage = FALSE)
+popViewport(2)
+@
+\caption{Varying spacing for the Arthritis  data. Top left: equal
+  spacing for all dimensions. Top right: different spacings for
+  individial dimensions. Bottom left: increasing spacing. Bottom right: 
+  spacing used for highlighting.}
+\label{fig:spacing}
+\end{center}
+\end{figure}
+\setkeys{Gin}{width=0.7\textwidth}
+
+\section{Example: Ovarian cancer survival}
+\label{sec:example}
+
+In the following, we demonstrate some of the described techniques in
+analyzing a data set originating from \citep{vcd:obel:1975}
+\cite[taken from][]{vcd:andersen:1991} about a retrospective study of ovary cancer
+carried out in 1973.  Information was obtained from 299 women, who
+were operated for ovary cancer 10 years before. The data consists of
+four binary variables: the \code{stage} of the cancer at the time of
+operation (levels: \code{early}, \code{advanced}), the type of \code{operation} performed
+(\code{radical}, \code{limited}), the \code{survival} status after 10 years (\code{yes}, 
+\code{no}), and \code{xray} indicating whether X-ray treatment was received (\code{yes},
+\code{no}).
+
+The dataset in \pkg{vcd} comes pretabulated in a data frame, so we
+first create the four-way table:
+
+<<oc1>>=
+tab <- xtabs(Freq ~ stage + operation + xray + survival, data = OvaryCancer)
+@
+
+\noindent A ``flattened'' textual representation can be obtained using \codefun{structable}:
+
+<<oc2,echo=TRUE,results=verbatim>>=
+structable(survival ~ ., data = tab)
+@
+
+\noindent A first overview can be obtained using a pairs plot (Figure~\ref{fig:ocpairs}):
+
+<<oc3>>=
+dpa <- list(var_offset = 1.2, rot = -30, just_leveltext= "left")
+pairs(tab, diag_panel_args = dpa)
+@
+
+\begin{figure}[h]
+\begin{center}
+<<ocpairs,echo=FALSE,fig=TRUE,width=8,height=8>>=
+<<oc3>>
+@ 
+\caption{Pairs plot for the \data{OvaryCancer} data showing mosaic displays for all pairwise distributions and bar plots for all marginal distributions.}
+\label{fig:ocpairs}
+\end{center}
+\end{figure}
+
+\noindent The pairs plot, by default, creates mosaic displays for all pairwise 
+variable combinations, and bar plots in the diagonal to visualize the absolute 
+frequencies of the variables. The \texttt{var\_offset} argument modifies the offset of the 
+(centered) variable names to avoid overlap with the bars. 
+Additionally, we use the \texttt{rot} and the
+\texttt{just\_leveltext} arguments to rotate the level names, again to avoid their overlap.
+First, we consider the marginal distributions. The study design involved 
+(nearly) the same number of survived (150) and deceased (149) patients. Similarly balanced,
+158 cases were in an advanced and 141 in an early stage. Most patients (251, 84\%) 
+were treated with a radical operation, and 186 (62\%) were submitted to
+X-ray treatment. Next, we inspect
+the two-way interaction of the influencing factors (\code{stage}, \code{operation}, 
+and \code{xray}): the
+corresponding mosaics exhibit symmetric, regular shapes with aligned tiles, 
+which indicate no marginal interaction between these variables. 
+The same is true for the interactions of \code{survival}
+with \code{operation} and \code{xray}, respectively. Only the stage seems to influence
+survival: here, the tiles are ``shifted''.
+
+A different view on the data, focused on the influence of the
+explanatory variables on \code{Survival},
+can be obtained using a doubledecker plot (Figure~\ref{fig:ocdoubledecker}):
+
+<<oc4>>=
+doubledecker(survival ~ stage + operation + xray, data = tab)
+@
+
+\begin{figure}[h]
+\begin{center}
+<<ocdoubledecker,echo=FALSE,fig=TRUE,width=8>>=
+<<oc4>>
+@ 
+\caption{Doubledecker plot for the \data{OvaryCancer} data showing the
+ conditional distribution of X-ray, given operation, given stage, and with
+ survival highlighted.}
+\label{fig:ocdoubledecker}
+\end{center}
+\end{figure}
+
+\noindent From a technical point of view, the display is constructed as a mosaic plot
+showing the conditional distribution of \code{survival}, given \code{xray}, 
+given \code{operation}, given \code{stage}, 
+with vertical splits for the conditioning variables and horizontal ones for
+\code{survival}. 
+Additionally, there is zero space between the tiles of the last dimension and
+a binary shading is used for survived and deceased patients.
+Conceptually, this plot is interpreted as a mosaic plot of just the influencing
+variables, with \code{survival} highlighted in the tiles. Thus, the plot really shows
+the influence of the explanatory variables on \code{survival}. Clearly, the survival
+rate is higher among patients in an early stage, but neither radical operation 
+nor X-ray treatment seem to improve the situation. From this exploratory
+phase, the survival rate seems to be slightly higher for patients who
+received a limited operation only, whereas the effect for X-ray
+treatment is less marked.
+
+To visualize inference results, we can make use of residual-based
+shadings, investigating log-linear models for the four-way table. 
+Figure~\ref{fig:ocmosaicnull} visualizes the null model, where
+survival is independent from the combined effect of operation, X-ray
+treatment, and stage:
+
+<<oc6>>=
+split <- c(TRUE, TRUE, TRUE, FALSE)
+mosaic(tab, expected = ~ survival + operation * xray * stage, split_vertical = split)
+@
+
+\begin{figure}[p]
+\begin{center}
+<<ocmosaicnull,echo=FALSE,fig=TRUE,width=10>>=
+<<oc6>>
+@ 
+\caption{Mosaic plot for the \data{OvaryCancer} data, with residual-based shading
+for the (clearly rejected) null model (survival)(operation, X-ray, stage).}
+\label{fig:ocmosaicnull}
+\end{center}
+\end{figure}
+
+\noindent The model is clearly rejected ($p$-value: 0.000). From the exploratory phase of
+our analysis, we (only) suspect \code{stage} to be influential on the survival rate. A 
+corresponding hypothesis is that \code{survival} be independent of \code{xray} 
+and \code{operation}, given
+\code{stage}. The model is specified using the \texttt{expected} argument, 
+either using the \codefun{loglin} interface or 
+the \codefun{loglm} formula interface (the resulting mosaic plot is shown in Figure
+\ref{fig:ocmosaicstage}):
+
+<<oc7>>=
+mosaic(tab, expected = ~ (survival + operation * xray) * stage, split_vertical = split)
+@
+
+\begin{figure}[p]
+\begin{center}
+<<ocmosaicstage,echo=FALSE,fig=TRUE,width=10>>=
+<<oc7>>
+@ 
+\caption{Mosaic plot for the \data{OvaryCancer} data, with residual-based shading
+for the hypothesis of survival being independent of X-ray and operation, given
+stage. The hypothesis is not rejected.}
+\label{fig:ocmosaicstage}
+\end{center}
+\end{figure}
+
+\noindent Thus, based on this data, only pre-diagnosis seems to matter in ovarian
+cancer therapy.
+
+\section{Conclusion}
+\label{sec:conclusion}
+
+In this paper, we describe the ``strucplot'' framework for the visualization of
+multi-way contingency tables. Strucplot displays include popular 
+basic plots such as mosaic, association, and sieve plots, integrated 
+in a unified framework: all can be seen as visualizations of
+hierarchical conditional flat tables. 
+Additionally, these core strucplot displays can be combined into more complex, 
+specialized plots, such as pairs and trellis-like displays for visualizing
+conditional independence. Residual-based shadings permit the
+visualization of log-linear models and the results of independence tests.
+The framework's modular design allows
+flexible customization of the plots' graphical appearance, including 
+shading, labeling, spacing, and legend, by means of graphical
+appearance control (``grapcon'') functions. These ``graphical
+hyperparameters'' are customized and created by
+generating functions. Our work includes a set of predefined grapcon generators for
+typical analysis tasks, and user-level extensions can easily be added.
+
 \bibliography{vcd}
 
+\begin{appendix}
+
+\section{Data sets}
+\label{sex:data}
+
+The data set names in the paper are those from the \proglang{R}
+system. In the following, we give a short description of each data set.
+
+\begin{description}
+\item[\texttt{Arthritis}] Data  from a 
+  double-blind clinical trial
+  investigating a new treatment for rheumatoid arthritis. 
+  Source: \cite{vcd:Koch+Edwards:1988}. Taken from: \cite{vcd:Friendly:2000}.
+  Package: \pkg{vcd}.
+\item[\texttt{Bundesliga}] Results from the first German soccer league in the
+  years 1995/6 \citep{vcd:Knorr-Held:1999} and 2001/2 (Collected by: Achim
+  Zeileis). Package: \pkg{vcd}.
+\item[\texttt{HairEyeColor}] Distribution of hair and eye color and 
+  gender in 592 statistics
+  students. The gender information is artificial. Source: \cite{vcd:Snee:1974}.
+  Taken from: \cite{vcd:Friendly:2000}. 
+  Package: \pkg{datasets} (included in base \proglang{R}).
+\item[\texttt{OvaryCancer}]  Data about a retrospective study of ovary cancer
+     carried out in 1973.  Information was obtained from 299 women, who
+     were operated for ovary cancer 10 years before. Source:
+     \cite{vcd:obel:1975}. Taken fromn: \cite{vcd:andersen:1991}. Package: \pkg{vcd}.
+\item[\texttt{PreSex}]  Data on pre- and extra-marital sex and divorce. 
+  Source: \cite{vcd:thornes+collard:1979}. Taken from \cite{vcd:gilbert:1981}.
+  Package: \pkg{vcd}.
+\item[\texttt{Titanic}] Information on the fate of passengers on
+  the fatal maiden voyage of the ocean liner ``Titanic'', summarized
+  according to economic status (class), gender (\code{Sex}), age and
+  survival. Data originally collected by the British Board of Trade
+  in their investigation of the sinking. Taken from: \cite{vcd:dawson:1995}.
+  Package: \pkg{datasets} (included in base \proglang{R}).
+\item[\texttt{UCBAdmissions}] 
+  Aggregate data on applicants to graduate school at Berkeley for
+  the six largest departments in 1973 classified by admission and
+  gender. Source: \cite{vcd:Bickel+Hammel+O'Connell:1975}. 
+  Taken from: \cite{vcd:Friendly:2000}.
+  Package: \pkg{datasets} (included in base \proglang{R}).
+\end{description}
+
+\end{appendix}
+
 \end{document}

### 2006-10-27 trunk/vcd/paper/Handbook/hsc.Rnw
fixed Figure 9 (used wrong HCL palette)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@616 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 plot.new()
-rect(0:4/5, 0.2, 1:5/5, 0.5, border = 0, col = diverge_hcl(5))
+rect(0:4/5, 0.2, 1:5/5, 0.5, border = 0, col = diverge_hcl(5, c = 100, l = c(50, 90)))
 rect(0, 0.2, 1, 0.5, border = 1, col = NULL)

### 2006-11-17 trunk/vcd/paper/Color/TAS.Rnw
updated color functionality to new colorspace
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@619 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
   hcl.wheel(luminance = 70, do.label = FALSE, inner.circle = TRUE)
-  rect(0.45, -0.05, 0.55, 0.05, col = hcl(0, 50, 70))
-  rect(-cos(pi/3)/2-0.05, sin(pi/3)/2-0.05, -cos(pi/3)/2+0.05, sin(pi/3)/2+0.05, col = hcl(120, 50, 70))
-  rect(-cos(pi/3)/2-0.05, -sin(pi/3)/2-0.05, -cos(pi/3)/2+0.05, -sin(pi/3)/2+0.05, col = hcl(240, 50, 70))
+  rect(0.45, -0.05, 0.55, 0.05, col = hcl2hex(0, 50, 70))
+  rect(-cos(pi/3)/2-0.05, sin(pi/3)/2-0.05, -cos(pi/3)/2+0.05, sin(pi/3)/2+0.05, col = hcl2hex(120, 50, 70))
+  rect(-cos(pi/3)/2-0.05, -sin(pi/3)/2-0.05, -cos(pi/3)/2+0.05, -sin(pi/3)/2+0.05, col = hcl2hex(240, 50, 70))
   dev.off()
@@
   lines(c(2.5, 2.5), c(90, 30), lty = 2)
-  rect(0, 87.5, 5, 92.5, col = hcl(260, 0, 90))
-  rect(47.5, 67.5, 52.5, 72.5, col = hcl(260, 50, 70))
-  rect(95, 47.5, 100, 52.5, col = hcl(260, 100, 50))
-  rect(37.5, 57.5, 42.5, 62.5, col = hcl(260, 40, 60))
-  rect(77.5, 27.5, 82.5, 32.5, col = hcl(260, 80, 30))
-  rect(0, 57.5, 5, 62.5, col = hcl(260, 0, 60))
-  rect(0, 27.5, 5, 32.5, col = hcl(260, 0, 30))
+  rect(0, 87.5, 5, 92.5, col = hcl2hex(260, 0, 90))
+  rect(47.5, 67.5, 52.5, 72.5, col = hcl2hex(260, 50, 70))
+  rect(95, 47.5, 100, 52.5, col = hcl2hex(260, 100, 50))
+  rect(37.5, 57.5, 42.5, 62.5, col = hcl2hex(260, 40, 60))
+  rect(77.5, 27.5, 82.5, 32.5, col = hcl2hex(260, 80, 30))
+  rect(0, 57.5, 5, 62.5, col = hcl2hex(260, 0, 60))
+  rect(0, 27.5, 5, 32.5, col = hcl2hex(260, 0, 30))
   hue.slice(120, grid.n = 301)
@@
   lines(c(2.5, 2.5), c(90, 30), lty = 2)
-  rect(0, 87.5, 5, 92.5, col = hcl(260, 0, 90))
-  rect(47.5, 80, 52.5, 85, col = hcl(120, 50, 82.5))
-  rect(95, 72.5, 100, 77.5, col = hcl(120, 100, 75))
-  rect(0, 57.5, 5, 62.5, col = hcl(260, 0, 60))
-  rect(0, 27.5, 5, 32.5, col = hcl(260, 0, 30))
+  rect(0, 87.5, 5, 92.5, col = hcl2hex(260, 0, 90))
+  rect(47.5, 80, 52.5, 85, col = hcl2hex(120, 50, 82.5))
+  rect(95, 72.5, 100, 77.5, col = hcl2hex(120, 100, 75))
+  rect(0, 57.5, 5, 62.5, col = hcl2hex(260, 0, 60))
+  rect(0, 27.5, 5, 32.5, col = hcl2hex(260, 0, 30))
   dev.off()
@@
   lines(c(-100, 0, 100), c(50, 90, 50), lty = 2)
-  rect(-100, 47.5, -90, 52.5, col = hcl(0, 100, 50))
-  rect(-55, 67.5, -45, 72.5, col = hcl(0, 50, 70))
-  rect(-5, 87.5, 5, 92.5, col = hcl(260, 0, 90))
-  rect(45, 67.5, 55, 72.5, col = hcl(260, 50, 70))
-  rect(90, 47.5, 100, 52.5, col = hcl(260, 100, 50))
+  rect(-100, 47.5, -90, 52.5, col = hcl2hex(0, 100, 50))
+  rect(-55, 67.5, -45, 72.5, col = hcl2hex(0, 50, 70))
+  rect(-5, 87.5, 5, 92.5, col = hcl2hex(260, 0, 90))
+  rect(45, 67.5, 55, 72.5, col = hcl2hex(260, 50, 70))
+  rect(90, 47.5, 100, 52.5, col = hcl2hex(260, 100, 50))
   lines(c(-80, 0, 80), c(30, 90, 30), lty = 2)
-  rect(-85, 27.5, -75, 32.5, col = hcl(0, 80, 30))
-  rect(-45, 57.5, -35, 62.5, col = hcl(0, 40, 60))
-  rect(-5, 87.5, 5, 92.5, col = hcl(260, 0, 90))
-  rect(35, 57.5, 45, 62.5, col = hcl(260, 40, 60))
-  rect(75, 27.5, 85, 32.5, col = hcl(260, 80, 30))
+  rect(-85, 27.5, -75, 32.5, col = hcl2hex(0, 80, 30))
+  rect(-45, 57.5, -35, 62.5, col = hcl2hex(0, 40, 60))
+  rect(-5, 87.5, 5, 92.5, col = hcl2hex(260, 0, 90))
+  rect(35, 57.5, 45, 62.5, col = hcl2hex(260, 40, 60))
+  rect(75, 27.5, 85, 32.5, col = hcl2hex(260, 80, 30))
   dev.off()
@@
 interface but chooses colors in HSV coordinates. It computes the HCL coordinates
-and transforms them to RGB coordinates by calling the \code{hcl()} function 
-provided by the \pkg{colorspace} package.
+and transforms them to RGB by using the \code{hcl2hex()} function which
+is a convenience wrapper to the function \code{polarLUV()} from
+the \pkg{colorspace} package.
 
@@
 \pkg{vcd} \citep{vcd:Meyer+Zeileis+Hornik:2006}. The functions are based on the
-\code{hcl(h, c, l, ...)} function proveded by the \pkg{colorspace} package
-\citep{vcd:Ihaka:2004}. A simple convenience function for displaying a certain palette
-is
+\code{polarLUV(l, c, h)} function proveded by the \pkg{colorspace} package
+\citep{vcd:Ihaka:2004}.
+A simple convenience function for displaying a certain palette is
 \begin{Soutput}

### 2006-11-17 trunk/vcd/paper/Shadings/JCGS.Rnw
finished draft version for colors paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@590 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
fully integrated new hcl2hex into papers
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@620 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
     nl <- length(luminance)
-    color.slice <- outer(chroma, luminance, function(y, x) hcl(hue, x, y, fixup = fixup))
+    color.slice <- outer(chroma, luminance, function(y, x) hcl2hex(hue, x, y, fixup = fixup))
     xlab <- "chroma"
@@
 \proglang{R} system for statistical computing and graphics
-\citep{vcd:R:2005} employs colors from HSV space. 
+\citep{vcd:R:2006} employs colors from HSV space. 
 Both spaces are rather similar transformations of RGB (Red-Green-Blue) space
@@
 ## full colors
-rect(-100, 47.5, -90, 52.5, col = hcl(0, 100, 50))
-rect(-55, 67.5, -45, 72.5, col = hcl(0, 50, 70))
-rect(-5, 95, 5, 100, col = hcl(260, 0, 100))       ## grey vs. white
-rect(-5, 87.5, 5, 92.5, col = hcl(260, 0, 90))     ## grey vs. white
-rect(45, 67.5, 55, 72.5, col = hcl(260, 50, 70))
-rect(90, 47.5, 100, 52.5, col = hcl(260, 100, 50))
+rect(-100, 47.5, -90, 52.5, col = hcl2hex(0, 100, 50))
+rect(-55, 67.5, -45, 72.5, col = hcl2hex(0, 50, 70))
+rect(-5, 95, 5, 100, col = hcl2hex(260, 0, 100))       ## grey vs. white
+rect(-5, 87.5, 5, 92.5, col = hcl2hex(260, 0, 90))     ## grey vs. white
+rect(45, 67.5, 55, 72.5, col = hcl2hex(260, 50, 70))
+rect(90, 47.5, 100, 52.5, col = hcl2hex(260, 100, 50))
 
 text(-100, 33, "full color", pos = 4, cex = 1.2)
-rect(-100, 20, -80, 30, col = hcl(0, 100, 50))
-rect(-60, 20, -40, 30, col = hcl(0, 50, 70))
-rect(-20, 20, 0, 30, col = hcl(0, 0, 90))       
-rect(0, 20, 20, 30, col = hcl(260, 0, 90))
-#white# rect(-20, 20, 0, 30, col = hcl(0, 0, 100))
-#white# rect(0, 20, 20, 30, col = hcl(260, 0, 100))
-rect(40, 20, 60, 30, col = hcl(260, 50, 70))
-rect(80, 20, 100, 30, col = hcl(260, 100, 50))
+rect(-100, 20, -80, 30, col = hcl2hex(0, 100, 50))
+rect(-60, 20, -40, 30, col = hcl2hex(0, 50, 70))
+rect(-20, 20, 0, 30, col = hcl2hex(0, 0, 90))       
+rect(0, 20, 20, 30, col = hcl2hex(260, 0, 90))
+#white# rect(-20, 20, 0, 30, col = hcl2hex(0, 0, 100))
+#white# rect(0, 20, 20, 30, col = hcl2hex(260, 0, 100))
+rect(40, 20, 60, 30, col = hcl2hex(260, 50, 70))
+rect(80, 20, 100, 30, col = hcl2hex(260, 100, 50))
 
@@
 ## reduced colors
-rect(-25, 47.5, -15, 52.5, col = hcl(0, 20, 50))
-rect(-15, 67.5, -5, 72.5, col = hcl(0, 10, 70))
-rect(5, 67.5, 15, 72.5, col = hcl(260, 10, 70))
-rect(25, 47.5, 15, 52.5, col = hcl(260, 20, 50))
+rect(-25, 47.5, -15, 52.5, col = hcl2hex(0, 20, 50))
+rect(-15, 67.5, -5, 72.5, col = hcl2hex(0, 10, 70))
+rect(5, 67.5, 15, 72.5, col = hcl2hex(260, 10, 70))
+rect(25, 47.5, 15, 52.5, col = hcl2hex(260, 20, 50))
 
 text(-100, 13, "reduced color", pos = 4, cex = 1.2)
-rect(-60, 0, -40, 10, col = hcl(0, 20, 50))
-rect(-40, 0, -20, 10, col = hcl(0, 10, 70))
-rect(-20, 0, 0, 10, col = hcl(0, 0, 90))
-rect(0, 0, 20, 10, col = hcl(260, 0, 90))
-rect(20, 0, 40, 10, col = hcl(260, 10, 70))
-rect(40, 0, 60, 10, col = hcl(260, 20, 50))
+rect(-60, 0, -40, 10, col = hcl2hex(0, 20, 50))
+rect(-40, 0, -20, 10, col = hcl2hex(0, 10, 70))
+rect(-20, 0, 0, 10, col = hcl2hex(0, 0, 90))
+rect(0, 0, 20, 10, col = hcl2hex(260, 0, 90))
+rect(20, 0, 40, 10, col = hcl2hex(260, 10, 70))
+rect(40, 0, 60, 10, col = hcl2hex(260, 20, 50))
 

### 2006-11-17 trunk/vcd/paper/Color/functions.R

@@ -53,7 +53,7 @@ hcl.wheel <-
         for(ang in hues) { # could do all this using outer() instead of for()...
             a. <- ang * pi/180
             z.a <- exp(1i * a.)
-            cols <- hcl(h = ang, c = chroma, l = luminance[i.c], fixup = fixup)
+            cols <- hcl2hex(h = ang, c = chroma, l = luminance[i.c], fixup = fixup)
             points(l.s * z.a, pch = 16, col = cols, cex = p.cex)
             if(do.label[2]) text(z.a*1.05, labels = ang, col = cols[length(cols)/2], srt = ang)
         }
@@ -74,7 +74,7 @@ hue.slice <- function(hue, grid.n = 251, plot = TRUE, fixup = FALSE, main = NULL
   luminance = seq(0, 100, length = grid.n)
   nc <- length(chroma)
   nl <- length(luminance)
-  color.slice <- outer(chroma, luminance, function(y, x) hcl(hue, x, y, fixup = fixup))
+  color.slice <- outer(chroma, luminance, function(y, x) hcl2hex(hue, x, y, fixup = fixup))
   xlab <- "chroma"
   ylab <- "luminance"
   if(is.null(main)) main <- paste("hue =", round(hue, digits = 0))

### 2006-12-07 trunk/vcd/paper/Shadings/JCGS.Rnw
incorporated changes for second revision of JCGS shadings paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@624 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 %% Packages:
-\usepackage{graphicx,color,a4wide,thumbpdf,hyperref}
+\usepackage{graphicx,color,a4wide,thumbpdf,rotating,hyperref}
 %% need no \usepackage{Sweave}
@@
 nrep_char <- "5,000"
-recompute <- TRUE
+recompute <- FALSE
 @
@@
   pun_cotab <- cotab_coindep(pun, condvars = 3:4, type = "assoc", n = nrep,
+    varnames = FALSE, margins = c(2, 1, 1, 2), test = "maxchisq", interpolate = 1:2)
+  set.seed(myseed)
+  pun_cotab2 <- cotab_coindep(pun, condvars = 3:4, type = "mosaic", n = nrep,
     varnames = FALSE, margins = c(2, 1, 1, 2), test = "maxchisq", interpolate = 1:2)
@@
        alz_dmax, alz_maxchisq, alz_sumchisq, alz_cotab,
-       pun_dmax, pun_maxchisq, pun_sumchisq, pun_cotab,
+       pun_dmax, pun_maxchisq, pun_sumchisq, pun_cotab, pun_cotab2,
        file = "JCGS-inference.rda")
@@
 However, the sum of squares is not the only plausible way of capturing deviations
-from zero in the residuals.
-There are many other conceivable functionals $\lambda(\cdot)$ which
-lead to reasonable test statistics $\lambda([r_{ij}])$: one which is particularly
-suitable for identifying the cells responsible for the `dependence' (i.e., significant
-departure from independence), if any, is the maximum of the absolute values
+from zero in the residuals. There are many other conceivable functionals $\lambda(\cdot)$ which
+lead to reasonable test statistics $\lambda([r_{ij}])$---and without further 
+specification of a certain pattern of dependence no functional $\lambda(\cdot)$
+uniformly dominates all others in terms of power of the resulting test procedure.
+Therefore, the choice of the functional is usually also guided by the data analysis problem at
+hand: one functional which is particularly suitable for identifying the cells responsible for
+the `dependence' (i.e., significant departure from independence), if any, is the maximum of
+the absolute values
 \begin{equation} \label{eq:MaxAbs}
@@
 \setkeys{Gin}{width=2\textwidth}
-\begin{figure}[p]
+\begin{sidewaysfigure}[p]
 \begin{center}
-<<shadings,fig=TRUE,echo=FALSE,height=12,width=16,results=hide,include=FALSE>>=
+<<shadings,fig=TRUE,echo=FALSE,height=12,width=18,results=hide,include=FALSE>>=
+##                                       width=16
 mymar <- c(1.5, 0.5, 0.5, 2.5)
@@
 @
-\includegraphics[height=\textwidth,angle=90,keepaspectratio]{JCGS-shadings}
+%%\includegraphics[height=\textwidth,angle=90,keepaspectratio]{JCGS-shadings}
+\includegraphics[width=.9\textwidth,keepaspectratio]{JCGS-shadings}
 \caption{Upper row: Mosaic plot for the arthritis data with Friendly shading (left),
@@
 \end{center}
-\end{figure}
+\end{sidewaysfigure}
 \setkeys{Gin}{width=\textwidth}
@@
 
+\setkeys{Gin}{width=\textwidth}
+\begin{figure}[t!]
+\begin{center}
+<<conditional_mosaic2,fig=TRUE,echo=FALSE,height=11,width=11>>=
+cotabplot(~ Memory + Attitude | Age + Education, data = pun, panel = pun_cotab2)
+@
+\caption{Conditional mosaic plot with maximum sum-of-squares shading for conditional
+independence of memory and attitude given age and education.}
+\label{fig:pun2}
+\end{center}
+\end{figure}
+
 As a 4-way example, we use the punishment data \citep{vcd:andersen:1991} 
@@
 interpreted as a diagnostic residual plot for the associated log-linear conditional
-independence model. It shows an association between memories and attitude
+independence model. 
+%FIXME: with mosaic%
+Alternatively, a conditional mosaic plot as in Figure~\ref{fig:pun2} can be used for
+visualizing the contingency table and its associated residuals from the conditional independence
+model using the same shading. Both reveal
+%FIXME: without mosaic% It shows 
+an association between memories and attitude
 for the lowest education group (first column) and highest age group (last row):
@@
 
-\clearpage
-
 \section*{Computational Details}
@@
 {\pkg{colorspace}~{\Sexpr{packageDescription("colorspace")$Version}}} \citep{vcd:Ihaka:2004}.
+
+\newpage
 
@@
   n = 5000, type = "assoc", test = "maxchisq", interpolate = 1:2)
+
+## Figure 6
+set.seed(rseed)
+cotabplot(~ memory + attitude | age + education, data = pun, panel = cotab_coindep,
+  n = 5000, type = "mosaic", test = "maxchisq", interpolate = 1:2)
 \end{verbatim}

### 2006-12-11 trunk/vcd/inst/doc/residual-shadings.Rnw
incorporated changes for JCGS shadings paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@625 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\documentclass{Z}
%% need no \usepackage{Sweave}
\usepackage{rotating}
\newcommand{\given}{\, | \,}

\title{Residual-based Shadings in \pkg{vcd}}
\Plaintitle{Residual-based Shadings in vcd}
\author{Achim Zeileis, David Meyer, \textnormal{and} Kurt Hornik\\Wirtschaftsuniversit\"at Wien, Austria}
\Plainauthor{Achim Zeileis, David Meyer, Kurt Hornik}

\Abstract{
  This vignette is a companion paper to \cite{vcd:Zeileis+Meyer+Hornik:2005}
  which introduces sever extensions to residual-based shadings for enhancing
  mosaic and association plots. The paper introduces (a)~perceptually uniform
  Hue-Chroma-Luminance (HCL) palettes and (b)~incorporates the result of an
  associated significance test into the shading. Here, we show how the
  examples can be easily reproduced using the \pkg{vcd} package.
}

\Keywords{association plots, conditional inference, contingency tables, HCL colors, HSV colors, mosaic plots}

\begin{document}

%\VignetteIndexEntry{Residual-based Shadings in vcd}
%\VignetteDepends{vcd,colorspace,MASS,grid}
%\VignetteKeywords{association plots, conditional inference, contingency tables, HCL colors, HSV colors, mosaic plots}
%\VignettePackage{vcd}

\SweaveOpts{engine=R,eps=FALSE}

\section{Introduction} \label{sec:intro}

In this vignette, we show how all empirical examples from \cite{vcd:Zeileis+Meyer+Hornik:2005}
can be reproduced in \proglang{R}\citep[\mbox{\url{http://www.R-project.org/}}]{vcd:R:2006},
in particular using the package \pkg{vcd} \citep{vcd:Meyer+Zeileis+Hornik:2006}. Additionally,
the pakcages \pkg{MASS} \citep[see][]{vcd:Venables+Ripley:2002},
\pkg{grid} \citep[see][]{vcd:Murrell:2002} and \pkg{colorspace} \citep{vcd:Ihaka:2004}
are employed. All are automatically loaded together with \pkg{vcd}:

<<preliminaries, results=hide>>=
library("vcd")
rseed <- 1071
@

Furthermore, we define a \code{rseed} which will be used as the random seed for making
the results of the permutation tests (conditional inference) below exactly reproducible.
In the following, we focus on the \proglang{R} code and output---for background information
on the methods and the data sets, please consult \cite{vcd:Zeileis+Meyer+Hornik:2005}.

\section{Arthritis data} \label{sec:arthritis}

First, we take a look at the association of treatment type and improvement in the
\code{Arthritis} data. The data set can be loaded and brought into tabular form
via:

<<Arthritis-data>>=
data("Arthritis", package = "vcd")
(art <- xtabs(~ Treatment + Improved, data = Arthritis, subset = Sex == "Female"))
@

Two basic explorative views of such a 2-way table are mosaic plots and association plots.
They can be generated via \code{mosaic()} and \code{assoc()} from \pkg{vcd}, respectively.
For technical documentation of these functions, please see \cite{vcd:Meyer+Zeileis+Hornik:2006b}.
When no further arguments are supplied as in

<<Arthritis-classic, eval=FALSE>>=
mosaic(art)
assoc(art)
@

this yields the plain plots without any color shading, see Figure~\ref{fig:classic}.
Both indicate that there are more patients in the treatment group with marked improvement
and less without improvement than would be expected under independence---and vice versa
in the placebo group.

\setkeys{Gin}{width=\textwidth}
\begin{figure}[b!]
\begin{center}
<<Arthritis-classic1, fig=TRUE, echo=FALSE, height=6, width=11, results=hide>>=
grid.newpage()
pushViewport(viewport(layout = grid.layout(1, 2)))
pushViewport(viewport(layout.pos.col=1, layout.pos.row=1))
mosaic(art, newpage = FALSE, margins = c(2.5, 4, 2.5, 3))
popViewport()
pushViewport(viewport(layout.pos.col=2, layout.pos.row=1))
assoc(art, newpage = FALSE, margins = c(5, 2, 5, 4))
popViewport(2)
@
\caption{Classic mosaic and association plot for the arthritis data.}
\label{fig:classic}
\end{center}
\end{figure}

For 2-way tables, \cite{vcd:Zeileis+Meyer+Hornik:2005} suggest to extend the
shading of \cite{vcd:Friendly:1994} to also visualize the outcome of an independence
test---either using the sum of squares of the Pearson residuals as the test statistic
or their absolute maximum. Both statistics and their corresponding (approximate) permutation
distribution can easily be computed using the function \code{coindep_test()}. Its arguments
are a contingency table, a specification of margins used for conditioning (only for conditional
independence models), a functional for aggregating the Pearson residuals (or alternatively the
raw counts) and the number of permutations that should be drawn. The conditional table needs
to be a 2-way table and the default is to compute the maximum statistic (absolute maximum of
Pearson residuals). For the Arthritis data, both, the maximum test

<<Arthritis-max>>=
set.seed(rseed)
(art_max <- coindep_test(art, n = 5000))
@

and the sum-of-squares test, indicate a significant departure from independence.

<<Arthritis-sumsq>>=
ss <- function(x) sum(x^2)
set.seed(rseed)
coindep_test(art, n = 5000, indepfun = ss)
@

Thus, it can be concluded that the treatment is effective and leads to significantly more
improvement than the placebo. The classic views from Figure~\ref{fig:classic} and the inference
above can also be combined, e.g., using the maximum shading that highlights the cells in
an association or mosaic plot when the associated residuals exceed critical values of the
maximum test (by default at levels 90\% and 99\%). To compare this shading (using either HSV
or HCL colors) with the Friendly shading (using HSV colors), we generate all three versions
of the mosaic plot:

<<Arthritis-extended, eval=FALSE>>=
mosaic(art, gp = shading_Friendly(lty = 1, eps = NULL))
mosaic(art, gp = shading_hsv, gp_args = list(
  interpolate = art_max$qdist(c(0.9, 0.99)), p.value = art_max$p.value))
set.seed(rseed)
mosaic(art, gp = shading_max, gp_args = list(n = 5000))
@

the  results are shown in the upper row of Figure~\ref{fig:shadings}. The last plot could
hae also been generated analogously to the second plot using \code{shading_hcl()} instead of
\code{shading_hsv()}---\code{shading_max()} is simply a wrapper function which performs the
inference and then visualizes it based on HCL colors.

\section{Piston rings data} \label{sec:arthritis}

Instead of bringing out the result of the maximum test in the shading, we could also use
a sum-of-squares shading that visualizes the result of the sum-of-squares test. As an illustration,
we use the \code{pistonrings} data from the \code{HSAUR} \citep{vcd:Everitt+Hothorn:2006}
package giving the number of piston ring failurs in different legs of different compressors
at an industry plant:

<<pistonrings-data>>=
data("pistonrings", package = "HSAUR")
pistonrings
@

\begin{sidewaysfigure}[p]
\begin{center}
<<shadings,fig=TRUE,echo=FALSE,height=12,width=18,results=hide,include=FALSE>>=
mymar <- c(1.5, 0.5, 0.5, 2.5)
grid.newpage()
pushViewport(viewport(layout = grid.layout(2, 3)))
pushViewport(viewport(layout.pos.row = 1, layout.pos.col = 1))
mosaic(art, margins = mymar, newpage = FALSE,
  gp = shading_Friendly(lty = 1, eps = NULL))
popViewport()
pushViewport(viewport(layout.pos.row = 1, layout.pos.col = 2))
mosaic(art, gp = shading_hsv, margins = mymar, newpage = FALSE,
  gp_args = list(interpolate = art_max$qdist(c(0.9, 0.99)), p.value = art_max$p.value))
popViewport()
pushViewport(viewport(layout.pos.row = 1, layout.pos.col = 3))
set.seed(rseed)
mosaic(art, gp = shading_max, margins = mymar, newpage = FALSE,
  gp_args = list(n = 5000))
popViewport()

pushViewport(viewport(layout.pos.row = 2, layout.pos.col = 1))
mosaic(pistonrings, margins = mymar, newpage = FALSE,
  gp = shading_Friendly(lty = 1, eps = NULL, interpolate = c(1, 1.5)))
popViewport()
pushViewport(viewport(layout.pos.row = 2, layout.pos.col = 2))
mosaic(pistonrings, gp = shading_hsv, margins = mymar, newpage = FALSE,
  gp_args = list(p.value = 0.069, interpolate = c(1, 1.5)))
popViewport()
pushViewport(viewport(layout.pos.row = 2, layout.pos.col = 3))
mosaic(pistonrings, gp = shading_hcl, margins = mymar, newpage = FALSE,
  gp_args = list(p.value = 0.069, interpolate = c(1, 1.5)))
popViewport(2)
@
\includegraphics[width=.9\textwidth,keepaspectratio]{residual-shadings-shadings}
\caption{Upper row: Mosaic plot for the arthritis data with Friendly shading (left),
HSV maximum shading (middle), HCL maximum shading (right). Lower row: Mosaic
plot for the piston rings data with fixed user-defined cut offs 1 and 1.5 and 
Friendly shading (left), HSV sum-of-squares shading (middle), HCL sum-of-squares shading
(right).}
\label{fig:shadings}
\end{center}
\end{sidewaysfigure}

Although there seems to be some slight association between the leg (especially center and South)
and the compressor (especially numbers 1 and 4), there is no significant deviation from
independence:

<<pistonrings-inference>>=
set.seed(rseed)
coindep_test(pistonrings, n = 5000)
set.seed(rseed)
(pring_ss <- coindep_test(pistonrings, n = 5000, indepfun = ss))
@

This can also be brought out graphically in a shaded mosaicplot by enhancing the Friendly
shading (based on the user-defined cut-offs 1 and 1.5, here) to use a less colorful palette,
either based on HSV or HCL colors:

<<pistonrings-plot, eval=FALSE>>=
mosaic(pistonrings, gp = shading_Friendly(lty = 1, eps = NULL, interpolate = c(1, 1.5)))
mosaic(pistonrings, gp = shading_hsv, gp_args = list(p.value = pring_ss$p.value, interpolate = c(1, 1.5)))
mosaic(pistonrings, gp = shading_hcl, gp_args = list(p.value = pring_ss$p.value, interpolate = c(1, 1.5)))
@

The resulting plots can be found in the lower row of Figure~\ref{fig:shadings}. The default
in \code{shading_hcl()} and \code{shading_hsv()} is to use the asymptotical $p$~value, hence
we set it explicitely to the permtuation-based $p$~value computed above.

\section{Alzheimer and smoking} \label{sec:alzheimer}

For illustrating that the same ideas can be employed for visualizing (conditional)
independence in multi-way tables, \cite{vcd:Zeileis+Meyer+Hornik:2005} use a
3-way and a 4-way table. The former is taken from a case-control study of smoking
and {A}lzheimer's disease (stratified by gender). The data set is available in \proglang{R}
in the package \pkg{coin} \cite{vcd:Hothorn+Hornik+VanDeWiel:2006}.

<<alzheimer-data>>=
data("alzheimer", package = "coin")
alz <- xtabs(~ smoking + disease + gender, data = alzheimer)
alz
@

\begin{figure}[b!]
\begin{center}
<<alzheimer-plot1, fig=TRUE, echo=FALSE, height=5, width=9>>=
set.seed(rseed)
cotabplot(~ smoking + disease | gender, data = alz, panel = cotab_coindep, n = 5000)
@
\caption{Conditional mosaic plot with double maximum shading for conditional
independence of smoking and disease given gender.}
\label{fig:alz}
\end{center}
\end{figure}

To assess whether smoking behaviour and disease status are conditionally independent
given gender, \cite{vcd:Zeileis+Meyer+Hornik:2005} use three different types of test
statistics: double maximum (maximum of maximum statistics in the two strata), 
maximum sum of squares (maximum of sum-of-squares statistics), and sum of squares
(sum of sum-of-squares statistics). All three can be computed and assessed via
permutation methods using the function \code{coindep_test()}:

<<alzheimer-inference>>=
set.seed(rseed)
coindep_test(alz, 3, n = 5000)
set.seed(rseed)
coindep_test(alz, 3, n = 5000, indepfun = ss)
set.seed(rseed)
coindep_test(alz, 3, n = 5000, indepfun = ss, aggfun = sum)
@

The conditional mosaic plot in Figure~\ref{fig:alz}
shows clearly that the association of smoking and disease is present only in
the group of male patients. The double maximum shading employed allows for
identification of the male heavy smokers as the cells `responsible' for the dependence:
other dementias are more frequent and Alzheimer's disease less frequent in this group than expected
under independence. Interestingly, there seems to be another large residual for
the light smoker group ($<$10 cigarettes) and Alzheimer's disease---however, this
is only significant at 10\% and not at the 1\% level as the other two cells.

<<alzheimer-plot, eval=FALSE>>=
<<alzheimer-plot1>>
@

\section{Corporal punishment of children}

As a 4-way example, data from a study of the Gallup Institute in Denmark in 1979 about
the attitude of a random sample of 783 persons towards corporal
punishment of children is used. The contingency table comprises four margins:
memory of punishments as a child (yes/no), attitude as a binary variable
(approval of ``moderate'' punishment or ``no'' approval), highest level
of education (elementary/secondary/high), and age group (15--24, 25--39, $\ge$40 years).

<<Punishment-data>>=
data("Punishment", package = "vcd")
pun <- xtabs(Freq ~ memory + attitude + age + education, data = Punishment)
ftable(pun, row.vars = c("age", "education", "memory"))
@

It is of interest whether there is an association between memories of corporal punishments
as a child and attitude towards punishment of children as an adult, controlling
for age and education. All three test statistics already used above confirm that 
memories and attitude are conditionally associated:

\setkeys{Gin}{width=\textwidth}
\begin{figure}[t!]
\begin{center}
<<Punishment-assoc1,fig=TRUE,echo=FALSE,height=11,width=11>>=
set.seed(rseed)
cotabplot(~ memory + attitude | age + education, data = pun, panel = cotab_coindep,
  n = 5000, type = "assoc", test = "maxchisq", interpolate = 1:2)
@
\caption{Conditional association plot with maximum sum-of-squares shading for conditional
independence of memory and attitude given age and education.}
\label{fig:pun}
\end{center}
\end{figure}

\setkeys{Gin}{width=\textwidth}
\begin{figure}[t!]
\begin{center}
<<Punishment-mosaic1,fig=TRUE,echo=FALSE,height=11,width=11>>=
set.seed(rseed)
cotabplot(~ memory + attitude | age + education, data = pun, panel = cotab_coindep,
  n = 5000, type = "mosaic", test = "maxchisq", interpolate = 1:2)
@
\caption{Conditional mosaic plot with maximum sum-of-squares shading for conditional
independence of memory and attitude given age and education.}
\label{fig:pun2}
\end{center}
\end{figure}

<<Punishment-inference>>=
set.seed(rseed)
coindep_test(pun, 3:4, n = 5000)
set.seed(rseed)
coindep_test(pun, 3:4, n = 5000, indepfun = ss)
set.seed(rseed)
coindep_test(pun, 3:4, n = 5000, indepfun = ss, aggfun = sum)
@

Graphically, this dependence can be brought out using conditional association
or mosaic plots as shown in Figure~\ref{fig:pun} and \ref{fig:pun2}, respectively.
Both reveal an association between memories and attitude
for the lowest education group (first column) and highest age group (last row):
experienced violence seems to engender violence again as there are less adults
that disapprove punishment in the group with memories of punishments than expected
under independence. For the remaining four age-education groups, there seems to be
no association: all residuals of the conditional independence model are very close to
zero in these cells. The figures employ the maximum sum-of-squares shading
with user-defined cut offs 1 and 2, chosen to be within the range of the residuals.
The full-color palette is used only for those strata associated with a
sum-of-squares statistic significant at (overall) 5\% level, the reduced-color
palette is used otherwise. This highlights that the dependence pattern is significant
only for the middle and high age group in the low education column. The other panels
in the first column and last row also show a similar dependence pattern, however, it is not
significant at 5\% level and hence graphically down-weighted by using reduced color.

<<Punishment-assoc, eval=FALSE>>=
<<Punishment-assoc1>>
@
<<Punishment-mosaic, eval=FALSE>>=
<<Punishment-mosaic1>>
@

\bibliography{vcd}

\end{document}

### 2006-12-11 trunk/vcd/paper/Shadings/JCGS.Rnw
incorporated changes for JCGS shadings paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@625 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
+\section*{Acknowledgements}
+
+We are thankful to Paul Murrell and Ross Ihaka---for providing tools, insights
+and feedback that helped to derive the results presented in this paper---and to
+two anonymous referees and editor Luke Tierney---for valuable comments and suggestions
+that lead to a substantial improvement of the paper.
+
 \section*{Computational Details}
@@
 {\pkg{colorspace}~{\Sexpr{packageDescription("colorspace")$Version}}} \citep{vcd:Ihaka:2004}.
+A vignette that demonstrates how all empirical examples can exactly be reproduced by the reader
+in \proglang{R} is provided in the package \pkg{vcd}, see
+\code{vignette("residual-shadings", package = "vcd")}
 
@@
 \bibliography{vcd}
+
+\end{document}
+
+%% deprecated stuff
 
@@
 \end{appendix}
-
-\end{document}
-
-%% deprecated stuff
 

### 2006-12-11 trunk/vcd/inst/doc/residual-shadings.Rnw
fixed 783 -> 1456 in all papers/vignettes
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@627 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 As a 4-way example, data from a study of the Gallup Institute in Denmark in 1979 about
-the attitude of a random sample of 783 persons towards corporal
+the attitude of a random sample of 1,456 persons towards corporal
 punishment of children is used. The contingency table comprises four margins:

### 2006-12-11 trunk/vcd/paper/Shadings/JCGS.Rnw
fixed 783 -> 1456 in all papers/vignettes
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@627 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 from a study of the Gallup Institute in Denmark in 1979 about
-the attitude of a random sample of 783 persons towards corporal
+the attitude of a random sample of 1,456 persons towards corporal
 punishment of children. The contingency table comprises four margins:

### 2006-12-11 trunk/vcd/paper/Handbook/hsc.Rnw
fixed 783 -> 1456 in all papers/vignettes
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@627 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 from a study of the Gallup Institute in Denmark in 1979 about
-the attitude of a random sample of 783 persons towards corporal
+the attitude of a random sample of 1,456 persons towards corporal
 punishment of children (see Table~\ref{tab:punish}).      

### 2006-12-11 trunk/vcd/paper/Shadings/JCGS.Rnw
resubmitted JCGS paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@628 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2006-12-11 trunk/vcd/paper/Shadings/JCGS.Rnw
@@
 nrep_char <- "5,000"
-recompute <- FALSE
+recompute <- TRUE
 @
@@
 
+\setkeys{Gin}{width=\textwidth}
+\begin{figure}[b!]
+\begin{center}
+<<conditional_mosaic,fig=TRUE,echo=FALSE,height=5,width=9>>=
+cotabplot(~ Smoking + Disease | Gender, data = alz, panel = alz_cotab)
+@
+\caption{Conditional mosaic plot with double maximum shading for conditional
+independence of smoking and disease given gender.}
+\label{fig:alz}
+\end{center}
+\end{figure}
+
 How the structure of the independence problem can be exploited for 
@@
 \setkeys{Gin}{width=\textwidth}
-\begin{figure}[b!]
+\begin{figure}[t!]
 \begin{center}
-<<conditional_mosaic,fig=TRUE,echo=FALSE,height=5,width=9>>=
-cotabplot(~ Smoking + Disease | Gender, data = alz, panel = alz_cotab)
-@
-\caption{Conditional mosaic plot with double maximum shading for conditional
-independence of smoking and disease given gender.}
-\label{fig:alz}
+<<conditional_assoc,fig=TRUE,echo=FALSE,height=11,width=11>>=
+cotabplot(~ Memory + Attitude | Age + Education, data = pun, panel = pun_cotab)
+@
+\caption{Conditional association plot with maximum sum-of-squares shading for conditional
+independence of memory and attitude given age and education.}
+\label{fig:pun}
+\end{center}
+\end{figure}
+
+\setkeys{Gin}{width=\textwidth}
+\begin{figure}[t!]
+\begin{center}
+<<conditional_mosaic2,fig=TRUE,echo=FALSE,height=11,width=11>>=
+cotabplot(~ Memory + Attitude | Age + Education, data = pun, panel = pun_cotab2)
+@
+\caption{Conditional mosaic plot with maximum sum-of-squares shading for conditional
+independence of memory and attitude given age and education.}
+\label{fig:pun2}
 \end{center}
@@
 
-\setkeys{Gin}{width=\textwidth}
-\begin{figure}[t!]
-\begin{center}
-<<conditional_assoc,fig=TRUE,echo=FALSE,height=11,width=11>>=
-cotabplot(~ Memory + Attitude | Age + Education, data = pun, panel = pun_cotab)
-@
-\caption{Conditional association plot with maximum sum-of-squares shading for conditional
-independence of memory and attitude given age and education.}
-\label{fig:pun}
-\end{center}
-\end{figure}
-
-\setkeys{Gin}{width=\textwidth}
-\begin{figure}[t!]
-\begin{center}
-<<conditional_mosaic2,fig=TRUE,echo=FALSE,height=11,width=11>>=
-cotabplot(~ Memory + Attitude | Age + Education, data = pun, panel = pun_cotab2)
-@
-\caption{Conditional mosaic plot with maximum sum-of-squares shading for conditional
-independence of memory and attitude given age and education.}
-\label{fig:pun2}
-\end{center}
-\end{figure}
-
 As a 4-way example, we use the punishment data \citep{vcd:andersen:1991} 
 from a study of the Gallup Institute in Denmark in 1979 about
-the attitude of a random sample of 783 persons towards corporal
+the attitude of a random sample of $1,456$ persons towards corporal
 punishment of children. The contingency table comprises four margins:
@@
 
+\clearpage
+
 \section*{Computational Details}
@@
 \citep[\mbox{\url{http://www.R-project.org/}}]{vcd:R:2006} and the packages
-\pkg{vcd}~\Sexpr{packageDescription("vcd")$Version} \citep{vcd:Meyer+Zeileis+Hornik:2006},
+\pkg{vcd}~\Sexpr{packageDescription("vcd")$Version} \citep{vcd:Meyer+Zeileis+Hornik:2006b},
 {\pkg{MASS}~\Sexpr{packageDescription("MASS")$Version}} \citep[see][]{vcd:Venables+Ripley:2002},
@@
 {\pkg{colorspace}~{\Sexpr{packageDescription("colorspace")$Version}}} \citep{vcd:Ihaka:2004}.
-A vignette that demonstrates how all empirical examples can exactly be reproduced by the reader
+A vignette that demonstrates how all empirical examples can exactly be reproduced
 in \proglang{R} is provided in the package \pkg{vcd}, see
-\code{vignette("residual-shadings", package = "vcd")}
-
-\newpage
+\code{vignette("residual-shadings", package = "vcd")}.
 

### 2006-12-12 trunk/vcd/paper/Shadings/JCGS.Rnw
JCGS shadings paper accepted
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@629 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
-\documentclass[10pt,a4paper]{article}
+\documentclass[10pt,a4paper,twoside]{article}
+\usepackage{fancyheadings}
 
@@
 
+\pagestyle{fancy}
+\thispagestyle{fancyplain}
+\headrulewidth0pt
+\lhead[]{}
+\chead[]{}
+\rhead[]{}
+\lfoot[This is a preprint of an article accepted for publication in \textit{Journal of Computational and Graphical Statistics}.
+Copyright {\copyright} 2007 American Statistical Association, Institute of Mathematical Statistics, and Interface Foundation of North America]%
+{This is a preprint of an article accepted for publication in \textit{Journal of Computational and Graphical Statistics}.
+Copyright {\copyright} 2007 American Statistical Association, Institute of Mathematical Statistics, and Interface Foundation of North America}
+\cfoot[]{}
+\rfoot[]{}
+
 \begin{center}
@@
 
-\SweaveOpts{engine=R,eps=FALSE}
+\SweaveOpts{engine=R,eps=FALSE,echo=FALSE,results=hide}
 \setkeys{Gin}{width=0.8\textwidth}
@@
 
+\headrulewidth0pt
+\lhead[\thepage]{}
+\chead[\textit{Residual-based Shadings for Visualizing (Conditional) Independence}]{\textit{Achim Zeileis, David Meyer, Kurt Hornik}}
+\rhead[]{\thepage}
+\lfoot[{\small Copyright {\copyright} 2007 American Statistical Association, Institute of Mathematical Statistics, and Interface Foundation of North America}]{{\small Copyright {\copyright} 2007 American Statistical Association, Institute of Mathematical Statistics, and Interface Foundation of North America}}
+\cfoot[]{}
+\rfoot[]{}
+
 \section{Independence in 2-way Tables}
@@
 \begin{center}
-<<shading_HSV,fig=TRUE,echo=FALSE,height=7,width=8>>=
+<<shading_HSV,echo=FALSE,results=hide>>=
+png(file = "shading_HSV.png", height = 525, width = 600)
 ## generate colors
@@
 lines(c(.9, .55), c(0.525, .525), lty = 2)
-@
-
-<<shading_HCL,fig=TRUE,echo=FALSE,height=7,width=8>>=
+dev.off()
+@
+\includegraphics{shading_HSV}
+
+%% height=7,width=8 for PDF
+
+<<shading_HCL,echo=FALSE,results=hide>>=
+png(file = "shading_HCL.png", height = 525, width = 600)
 ## generate colors
@@
 lines(c(8.75, 1.25), c(72.5, 87.5), lty = 2)
-@
+dev.off()
+@
+\includegraphics{shading_HCL}
 \caption{Residual-based shadings in HSV (upper) and HCL space (lower).}
@@
 status are conditionally independent given gender. All three statistics
-suggested above find evidence for departure from significance: the sum-of-squares
+suggested above find evidence for departure from independence: the sum-of-squares
 statistic from Equation~\ref{eq:sumchisq} is $\sum_k X_k^2 = \Sexpr{stat(alz_sumchisq)}$

### 2006-12-12 trunk/vcd/paper/Shadings/JCGS.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@630 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-<<load_data,echo=FALSE,results=hide>>=
+<<load_data>>=
 library("vcd")
@@
 
-<<inference,echo=FALSE,results=hide>>=
+<<inference>>=
 if(recompute) {
@@
 
-<<auxiliary_functions,echo=FALSE,results=hide>>=
+<<auxiliary_functions>>=
 stat <- function(x, digits = 3) as.vector(round(x$statistic, digits = digits))
@@
 \begin{center}
-<<classic_plots,fig=TRUE,echo=FALSE,height=6,width=11,results=hide>>=
+<<classic_plots,fig=TRUE,height=6,width=11>>=
 grid.newpage()
@@
 \begin{center}
-<<shading_HSV,echo=FALSE,results=hide>>=
+<<shading_HSV>>=
 png(file = "shading_HSV.png", height = 525, width = 600)
@@
 
-<<shading_HCL,echo=FALSE,results=hide>>=
+<<shading_HCL>>=
 png(file = "shading_HCL.png", height = 525, width = 600)
@@
 \begin{center}
-<<shadings,fig=TRUE,echo=FALSE,height=12,width=18,results=hide,include=FALSE>>=
+<<shadings,fig=TRUE,height=12,width=18,include=FALSE>>=
 ##                                       width=16
@@
 \begin{center}
-<<conditional_mosaic,fig=TRUE,echo=FALSE,height=5,width=9>>=
+<<conditional_mosaic,fig=TRUE,height=5,width=9>>=
 cotabplot(~ Smoking + Disease | Gender, data = alz, panel = alz_cotab)
@@
 \begin{center}
-<<conditional_assoc,fig=TRUE,echo=FALSE,height=11,width=11>>=
+<<conditional_assoc,fig=TRUE,height=11,width=11>>=
 cotabplot(~ Memory + Attitude | Age + Education, data = pun, panel = pun_cotab)
@@
 \begin{center}
-<<conditional_mosaic2,fig=TRUE,echo=FALSE,height=11,width=11>>=
+<<conditional_mosaic2,fig=TRUE,height=11,width=11>>=
 cotabplot(~ Memory + Attitude | Age + Education, data = pun, panel = pun_cotab2)

### 2006-12-30 trunk/vcd/paper/Color/TAS.Rnw
started TAS revision: non-technical with bad/good examples
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@634 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \SweaveOpts{engine=R, eps=FALSE, echo=FALSE, results=hide}
+\usepackage{amsfonts}
 
@@
 library("vcd")
-library("e1071")
 library("mvtnorm")
-library("ellipse")
-library("kernlab")
 library("KernSmooth")
@@
 source("functions.R")
+source("Nigeria/bnd.R")
 
@@
 of displays. However, more often than not, there is relatively little guidance about
-how to choose an appropriate palette for a certain visualization task. Here, we address
+how to choose a suitable palette for a certain visualization task. Consequently,
+sub-optimal color palettes are used for too many color graphics displayed in journal articles
+or on presentation slides which could easily be enhanced if the underlying information
+were captured appropriately in the color schemes. Here, we address
 this problem by suggesting a color selection strategy for visualizing both categorical
@@
 system \citep[see][]{vcd:Murrell:2006} and the HCL
-color implementation from package \pkg{colorspace} \citep{vcd:Ihaka:2004}. This gives the user
+color implementation from package \pkg{colorspace} \citep{vcd:Ihaka:2006}. This gives the user
 both a conceptual and computational tool box for experimenting with color palettes for
@@
 
-The remainder of the paper is organized as follows: Section~\ref{sec:spaces} provides
-further information on HCL and other color spaces, contrasting in particular the properties
-of HSV and HCL colors. Section~\ref{sec:palettes} introduces HCL-based qualitative,
-sequential, and diverging palettes and Section~\ref{sec:illustrations} shows some brief
-illustrations for each type of palette in statistical graphics. Section~\ref{sec:software}
-provides some general remarks on the implementation in statistical software packages
-as well as some details on our implementation in \proglang{R}. Section~\ref{sec:discussion}
-concludes the paper with a discussion.
+The remainder of the paper is organized as follows: Section~\ref{sec:motivation} provides
+several motivating examples, showing how typical HSV-based graphics can be enhanced
+by using HCL-based palettes. Section~\ref{sec:spaces} gives a brief introduction
+to the underlying HSV and HCL color spaces before Section~\ref{sec:palettes} 
+suggests strategies for deriving HCL-based qualitative, sequential, and diverging palettes.
+Section~\ref{sec:software} offers some general remarks on the implementation in statistical
+software packages as well as some details on our implementation in \proglang{R}.
+Section~\ref{sec:discussion} concludes the paper with a discussion.
 
-\section{Color spaces} \label{sec:spaces}
+\section{Motivation} \label{sec:motivation}
 
-\setkeys{Gin}{width=0.8\textwidth}
-\begin{figure}[p]
+To show what can be gained by selecting appropriate color schemes, we present a collection
+of illustrations with typical color graphics based on HSV palettes (always in the left panels)
+and more suitable HCL palettes (always in the right panels). All HSV color palettes (and some
+of the data sets) are taken from recent publications in statistical journals such as
+\textit{Journal of the American Statistical Association},
+\textit{Journal of Computational and Graphical Statistics} or
+\textit{Computational Statistics \& Data Analysis}. The examples have been selected to provide
+an overview of (typically colored) statistical graphics and their respective pitfalls (when
+it comes to color palette choice). All graphics are in principle very useful but can be even enhanced
+by using better color schemes.
+
+Our first illustration is a heatmap, a very popular display for visualizing a function of
+two arguments. Here, we use a bivariate kernel density estimate \citep[for the Old
+Faithful geyser eruptions data from][]{vcd:Azzalini+Bowman:1990}---other typical applications
+include objective functions with two arguments \citep[as in][]{vcd:Gneiting+Sevcikova+Percival:2006}
+or (physical) measurements on a 2-dimensional grid \citep[as in][]{vcd:Yang+Buckley+Dudoit:2002}.
+Figure~\ref{fig:bkde} shows such a heatmap bringing out the relationship between the duration
+of an eruption of the Old Faithful geyser in Yellowstone National Park 
+and the waiting time for this eruption. It reveals a multi-modal distribution: short waiting
+times (around 50 minutes) are typically followed by a long eruption (around 4 minutes)
+whereas long waiting times (around 80 minutes) can be followed by either a long
+or short eruption (around 4 minutes).
+
+\setkeys{Gin}{width=\textwidth}
+\begin{figure}[t!]
 \begin{center}
-<<hsv-hcl>>=
+<<bkde>>=
 if(generate_png) {
-  files <- c("hcl-dcone", "hsv-cylinder")
-  PNG <- function(x) paste(x, ".png", sep = "")
+  data("geyser", package = "MASS")
+  ## geyser2 <- cbind(geyser$duration[-299], geyser$waiting[-1])
+  geyser2 <- cbind(geyser$duration, geyser$waiting)
+  dens <- bkde2D(geyser2, bandwidth = c(0.2, 3), gridsize = c(201, 201))
+  plot_bkde2D <- function(x, col = NULL, xlab = "duration", ylab = "waiting time", legend = FALSE, ...) {
+    if(is.null(col)) col <- rev(grey.colors(33))
+    if(legend) {
+      filled.contour(x$x1, x$x2, x$fhat, col = col, nlevels = length(col), xlab = xlab, ylab = ylab, ...)
+    } else {
+      image(x$x1, x$x2, x$fhat,
+        breaks = seq(from = min(x$fhat), to = max(x$fhat), length.out = length(col) + 1),
+        xlab = xlab, ylab = ylab, col = col, axes = FALSE, cex.lab = 1.5, ...)
+      box()
+      axis(1, cex.axis = 1.5)
+      axis(2, cex.axis = 1.5)
+    }
+    invisible(x)
+  }
 
-  png(file = PNG(files[1]), height = 400, width = 600)
-  hcl.wheel(luminance = seq(5, 95, length = 6))
-  dev.off()
-  png(file = PNG(files[2]), height = 400, width = 600)
-  hsv.wheel(value = seq(0, 1, length = 6), cone = TRUE)
+  png(file = "bkde.png", height = 810, width = 840)
+  par(mfrow = c(2, 2))
+  plot_bkde2D(dens, col = hsv(seq(from = 2/3, to = 0, length = 33), 0.8, 0.9))
+  ## plot_bkde2D(dens, col = rainbow(33, start = 1/3, end = 1, v = 0.8, s = 0.9))
+  plot_bkde2D(dens, col = rev(sequential_hcl(33, c = 0, l = c(30, 90), power = c(1/5, 1.5))))
+  plot_bkde2D(dens, col = rev(heat.colors(33)))
+  plot_bkde2D(dens, col = rev(heat_hcl(33, c = c(80, 30), l = c(30, 90), power = c(1/5, 1.5))))
   dev.off()
@@
 @
-\includegraphics{hsv-cylinder}
-
-\vspace*{2.5cm}
-
-\includegraphics{hcl-dcone}
-\caption{\label{fig:hsv-hcl} HSV and HCL space. The circles show the hue/saturation
-and hue/chroma plane, respectively, for varying levels of value and luminance.}
+\includegraphics{bkde}
+\setkeys{Gin}{width=.47\textwidth}
+<<bkde-pal>>=
+if(generate_png) {
+  png(file = "bkde-pal.png", height = 160, width = 396)
+  par(mfrow = c(4, 1), mar = c(0, 0, 0, 0), oma = c(2, 1.5, 0, 1.5))
+  plot(0, 0, type="n", xlim = c(0, 0.03), ylim = c(0, 1),
+    xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
+  abline(v = seq(from = 0, to = 0.03, length.out = 99), lwd = 4, col = 
+    hsv(seq(from = 2/3, to = 0, length = 99), 0.8, 0.9))
+    ## rainbow(99, start = 1/3, end = 1, v = 0.8, s = 0.9))
+  box()
+  plot(0, 0, type="n", xlim = c(0, 0.03), ylim = c(0, 1),
+    xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
+  abline(v = seq(from = 0, to = 0.03, length.out = 99), lwd = 4, col = 
+    rev(heat.colors(99)))
+  box()
+  plot(0, 0, type="n", xlim = c(0, 0.03), ylim = c(0, 1),
+    xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
+  abline(v = seq(from = 0, to = 0.03, length.out = 99), lwd = 4, col = 
+    rev(heat_hcl(99, c = c(80, 30), l = c(30, 90), power = c(1/5, 1.5))))
+  box()
+  plot(0, 0, type="n", xlim = c(0, 0.03), ylim = c(0, 1),
+    xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
+  abline(v = seq(from = 0, to = 0.03, length.out = 99), lwd = 4, col = 
+    rev(sequential_hcl(99, c = 0, l = c(30, 90), power = c(1/5, 1.5))))
+  box()
+  axis(1, cex.axis = 1.5)
+  dev.off()
+}
+@
+\includegraphics{bkde-pal}
+\caption{\label{fig:bkde} Bivariate density estimation for Old Faithful geyser eruptions.}
 \end{center}
@@
 
+A simple and very effective palette for such a display is a set of gray colors
+as in the top right panel of Figure~\ref{fig:bkde}. This is often (appropriately) used
+in printed papers when the journal does not offer color graphics---however, in journals
+that support color graphics, on presentation slides and in interactive usage (in statistical
+software packages), many users prefer to have colored displays and most often HSV
+palettes (as in the left two panels) are used. These palettes code the variable of interest
+by varying hue in an HSV color wheel as done by \citet[Figure~4a]{vcd:Yang+Buckley+Dudoit:2002}
+or \citet[Figures~1--4]{vcd:Gneiting+Sevcikova+Percival:2006}. The palette in the upper left panel
+codes increasing density by going from a blue to a red hue (via green and yellow)---a similar
+strategy are the ``heat colors'' in the lower left panel that increase from yellow to red.
+The latter works a bit better than the former, however both palettes exhibit several drawbacks.
+The modes in the map look much more like ``rings'' than a smoothly increasing/decreasing 
+density. The heatmap looks very flashy which---although good for drawing attention to a plot---makes
+it hard to hold the attention of the viewer for a longer time because the large areas shaded
+with saturated colors can be distracting and produce after-image effects \citep{vcd:Ihaka:2003}.
+
+In contrast, the gray colors used in the top right panel do not exhibit the same disadvantages,
+coding the variable of interest much better and without flashy colors. If, however, the user wants
+to add some color to the plot, this could be done by using a better balanced version of the heat
+colors (as shown in the bottom right panel). These colors also increase from a yellow to a red
+hue while being balanced towards the same gray as in the grayscale palette (i.e., when converted
+to a grayscale or printed out on a grayscale printer, the upper and lower right panel would 
+essentially look identical). Both palettes have in common that they give increasing perceptual emphasis
+to regions with increasing density, resulting in a heatmap that highlights the (small)
+interesting high-density regions and not to the large low-density regions surrounding them.
+
+To illustrate sequential palettes,
+
+ Both versions of the heatmap use a sequential
+palette as derived in Section~\ref{sec:sequential} balanced towards the same gray levels
+with $L \in [30, 90]$ and $p_2 = 1.5$. The sequential palette in the left panel uses only gray
+colors (i.e., $C_{\max} = 0$) and the palette in the right panel additionally employs
+colors with $H \in [0, 90]$, $C \in [30, 80]$ and $p_1 = 1/5$. The parameters have been 
+chosen such that the peaks in the density are brought out very clearly by using a rather
+large luminance contrast and $p_2 > 1$.
+
+\setkeys{Gin}{width=\textwidth}
+\begin{figure}[t!]
+\begin{center}
+<<nigeria>>=
+if(generate_png) {
+  m <- read.bnd("Nigeria/nigeria37.bnd")
+  fit <- read.table("Nigeria/reml_purespatial_f_district_spatial.res", header = TRUE)
+  fit <- fit[order(fit$district),]
+
+  png(file = "nigeria.png", height = 400, width = 840)
+  par(mfrow = c(1, 2), mar = c(2.22, 0.22, 1.02, 0.22))
+  plot(m, z = fit$pmode, fill = rev(rainbow(99, start = 0, end = 1/3)),
+    zlim = c(-0.65, 0.65), legend = FALSE)
+  plot(m, z = fit$pmode, zlim = c(-0.65, 0.65), legend = FALSE,
+    fill = rev(diverge_hcl(99, h = c(0, 120), c = 60, l = c(45, 90), power = 1.2)))
+  dev.off()
+}
+@
+\includegraphics{nigeria}
+\setkeys{Gin}{width=.47\textwidth}
+<<nigeria-pal>>=
+if(generate_png) {
+  png(file = "nigeria-pal.png", height = 80, width = 396)
+  par(mfrow = c(2, 1), mar = c(0, 0, 0, 0), oma = c(2, 0.2, 0, 0.2))
+  plot(0, 0, type="n", xlim = c(-0.65, 0.65), ylim = c(0, 1),
+    xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
+  abline(v = seq(from = -0.65, to = 0.65, length.out = 99), lwd = 4, col = 
+    rev(rainbow(99, start = 0, end = 1/3)))
+  box()
+  plot(0, 0, type="n", xlim = c(-0.65, 0.65), ylim = c(0, 1),
+    xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="")
+  abline(v = seq(from = -0.65, to = 0.65, length.out = 99), lwd = 4, col = 
+    rev(diverge_hcl(99, h = c(0, 120), c = 60, l = c(45, 90), power = 1.2)))
+  box()
+  axis(1, cex = 1.5)
+  dev.off()
+}
+@
+\includegraphics{nigeria-pal}
+\caption{\label{fig:nigeria} Posterior mode estimates for childhood mortality in Nigeria.}
+\end{center}
+\end{figure}
+
+Figure~\ref{fig:nigeria} is taken from \citep[Figure~5, left]{vcd:Kneib:2006}
+
+\setkeys{Gin}{width=\textwidth}
 \begin{figure}[p]
 \begin{center}
-<<hsv-hcl-gray>>=
+<<seats, fig=TRUE, height=4, width=11>>=
+par(mar = rep(0.8, 4), mfrow = c(1, 2))
+seats <- structure(c(226, 61, 54, 51, 222),
+  .Names = c("CDU/CSU", "FDP",  "Linke", "Gruene", "SPD"))
+parties_hcl <- rainbow_hcl(6, c = 60, l = 75)[c(5, 2, 6, 3, 1)]
+parties_hsv <- hsv(c(0, 60, 300, 120, 0)/360, v = c(0, 1, 1, 1, 1))
+names(parties_hcl) <- names(parties_hsv) <- names(seats)
+names(seats)[4] <- "Grüne"
+pie(seats, clockwise = TRUE, col = parties_hsv, radius = 1)
+pie(seats, clockwise = TRUE, col = parties_hcl, radius = 1)
+@
+\caption{\label{fig:seats} Seats in the German parliament.}
+\end{center}
+\end{figure}
+
+\setkeys{Gin}{width=\textwidth}
+\begin{figure}[p]
+\begin{center}
+<<votes, fig=TRUE, height=6, width=14>>=
+data("Bundestag2005")
+votes <- Bundestag2005[c(1, 3:5, 9, 11, 13:16, 2, 6:8, 10, 12),
+                       c("CDU/CSU", "FDP", "SPD", "Gruene", "Linke")]
+parties_hcl <- parties_hcl[colnames(votes)]
+parties_hsv <- parties_hsv[colnames(votes)]
+dimnames(votes)[[2]][c(4, 5)] <- c("Gr", "Li")
+
+grid.newpage()
+pushViewport(viewport(layout = grid.layout(1, 2)))
+pushViewport(viewport(layout.pos.col=1, layout.pos.row=1))
+mosaic(votes, gp = gpar(fill = parties_hsv),
+  spacing = spacing_highlighting, labeling = labeling_left,
+  labeling_args = list(rot_labels = c(0, 90, 0, 0), pos_labels = "center",
+  just_labels = c("center","center","center","right"), varnames = FALSE),
+  margins = unit(c(2.5, 1, 1, 12), "lines"),
+  keep_aspect_ratio = FALSE, newpage = FALSE)
+popViewport()
+pushViewport(viewport(layout.pos.col=2, layout.pos.row=1))
+mosaic(votes, gp = gpar(fill = parties_hcl),
+  spacing = spacing_highlighting, labeling = labeling_left,
+  labeling_args = list(rot_labels = c(0, 90, 0, 0), pos_labels = "center",
+  just_labels = c("center","center","center","right"), varnames = FALSE),
+  margins = unit(c(2.5, 1, 1, 12), "lines"),
+  keep_aspect_ratio = FALSE, newpage = FALSE)
+popViewport(2)
+@
+\caption{\label{fig:votes} Votes in the German election 2005 (by province).}
+\end{center}
+\end{figure}
+
+The first example illustrates qualitative palettes
+and visualizes data from the 2005 election for the German parliament ``Bundestag''.
+In this election, five parties were able to obtain enough votes to enter the
+Bundestag---the distribution of seats is depicted in a pie chart in Figure~\ref{fig:seats}.
+The colors used are rough metaphors for the political parties, using a red hue $H = 0$ for the
+social democrats SPD, a blue hue $H = 240$ for the conservative CDU/CSU, a yellow hue $H = 60$
+for the liberal FDP, a green hue $H = 120$ for the green party ``Die Gr\"unen'' and a purple 
+hue $H = 300$ for the leftist party ``Die Linke''. To obtain rather intense colors, we set
+chroma to $C = 60$ and luminance to $L = 75$ for all hues
+leading to a much better balanced palette compared to the fully saturated
+colors usually found in the (German) media (which instead of blue typically use the extreme
+``color'' black for the CDU/CSU). The pie chart clearly shows that neither the governing
+coalition of SPD and Gr\"une nor the opposition of CDU/CSU and FDP could assemble a majority.
+Given that no party would enter a coalition with the leftists, this lead to a big coalition
+of CDU/CSU and SPD. Figure~\ref{fig:votes} shows the distribution of votes in this election
+stratified by province (Bundesland) in a highlighted mosaic display. The order of provinces is from 
+north to south, first for the 10 western provinces (the former Federal Republic of Germany, FRG),
+then for the 6 eastern provinces (the former German Democratic Republic, GDR). 
+Clearly, the SPD performed better in the north and the CDU/CSU better in the south; furthermore,
+Die Linke performed particularly well in the eastern provinces and in Saarland.
+
+\setkeys{Gin}{width=\textwidth}
+\begin{figure}[p]
+\begin{center}
+<<scatter>>=
 if(generate_png) {
-  for(i in files) {
-    file.copy(PNG(i), PNG(paste(i, "-gray", sep = "")), overwrite = TRUE)
-    system(paste("mogrify -type Grayscale", PNG(paste(i, "-gray", sep = ""))))
-  }
+  set.seed(123)
+  vc <- function(var1, var2, cov)
+    matrix(c(var1, cov * sqrt(var1 * var2), cov * sqrt(var1 * var2), var2), ncol = 2)
+  dat <- rbind(rmvnorm(1000, c(0, 0), vc(0.05, 0.3, -0.95)),
+               rmvnorm(1000, c(1.8, 2.5), vc(0.12, 0.25, 0)),
+               rmvnorm(1000, c(3.8, 3.5), vc(0.02, 0.15, -0.8)))
+  dat <- data.frame(y = dat[,2], x = dat[,1], id = rep(1:3, rep(1000, 3)))
+
+  png(file = "scatter.png", height = 350, width = 840)
+  par(mfrow = c(1, 2), mar = rep(0.5, 4))
+  plot(y ~ x, data = dat, axes = FALSE, xlab = "", ylab = "", pch = c(3, 22, 1)[id],
+    col = hsv(c(1/3, 1/6, 0))[id])
+  box()
+  plot(y ~ x, data = dat, axes = FALSE, xlab = "", ylab = "", pch = c(3, 22, 1)[id],
+    col = hcl2hex(c(120, 60, 0), 70, 65, fix = FALSE)[id])
+  box()
+  dev.off()
 }
 @
-\includegraphics{hsv-cylinder-gray}
-
-\vspace*{2.5cm}
-
-\includegraphics{hcl-dcone-gray}
-\caption{\label{fig:hsv-hcl-gray} HSV and HCL space (in gray levels). The circles
-show the hue/saturation and hue/chroma plane, respectively, for varying levels of
-value and luminance.}
+\includegraphics{scatter}
+\caption{\label{fig:scatter} Scatter plot with three clusters.}
 \end{center}
 \end{figure}
+
+similar to \citet[Figure~3]{vcd:Celeux+Hurn+Robert:2000}
+
+\setkeys{Gin}{width=\textwidth}
+\begin{figure}[p]
+\begin{center}
+<<haireye, fig=TRUE, height=5.5, width=12>>=
+he <- margin.table(HairEyeColor, 1:2)
+grid.newpage()
+pushViewport(viewport(layout = grid.layout(1, 2)))
+pushViewport(viewport(layout.pos.col=1, layout.pos.row=1))
+mosaic(he, gp = shading_hsv, split_vertical = TRUE, newpage = FALSE, legend = FALSE)
+popViewport()
+pushViewport(viewport(layout.pos.col=2, layout.pos.row=1))
+mosaic(he, gp = shading_hcl, split_vertical = TRUE, newpage = FALSE, legend = FALSE)
+popViewport(2)
+@
+\caption{\label{fig:haireye} Extended mosaic display for haireye data.}
+\end{center}
+\end{figure}
+
+taken from \citet[Figure~11, left/middle]{vcd:Friendly:2002}
+
+As pointed out in Section~\ref{sec:diverging}, diverging palettes are particularly useful
+when visualizing residuals or correlations (with natural neutral value 0) or probabilities 
+in 2-class supervised learning (with neutral value 0.5). Examples for both situations are
+provided here. Figure~\ref{fig:haireye} visualizes the outcome of a double-blind clinical
+trial investigating a new treatment for rheumatoid arthritis. The mosaic rectangles alone
+signal that the treatment lead to higher improvement compared to the placebo group; this effect
+is shown to be significant by the shading that codes the size of the Pearon residuals.
+Positive residuals, corresponding
+to more observations in the corresponding cell than expected under independence, are depicted
+in blue, negative residuals in red. Light colors signal significance at 10\% level, full colors
+significance at 1\% level. Hence, it can be concluded that there are significantly more
+marked improvements in the treated group and significantly fewer in the placebo group than
+would be expected under independence between treatment and improvement. More details can
+be found in \cite{vcd:Zeileis+Meyer+Hornik:2007}.
+
+Both diverging palettes employ a smaller contrast on the luminance axis ($L \in [50, 90]$)
+and a larger chroma contrast ($C \in [0, 100]$) compared to the sequential palettes from
+the previous example. 
+For Figure~\ref{fig:haireye}, the choice of
+$p_1$ and $p_2$ is less important because only 5 colors are selected---we use values
+$p_1 = p_2 = 1$.
+
+\section{Color spaces} \label{sec:spaces}
 
@@
 perceptual axes
-\begin{enumerate}
+\begin{itemize}
   \item \emph{hue} (dominant wavelength)
@@
   \item \emph{luminance} (brightness, amount of gray)
-\end{enumerate}
+\end{itemize}
 
@@
 and are defined by a triplet $(H, S, V)$ with $H \in [0, 360]$ and $S, V \in [0, 100]$.
-HSV space has the shape of a single regular cone (often inflated to a regular
-cylinder). Vertical sections through this space are shown in the upper panel
-of Figure~\ref{fig:hsv-hcl}, depicting hue and saturation given different value levels.
+%% HSV space has the shape of a single regular cone (often inflated to a regular
+%% cylinder). Vertical sections through this space are shown in the upper panel
+%% of Figure~\ref{fig:hsv-hcl}, depicting hue and saturation given different value levels.
 Although simple to specify and easily available in many computing environments,
@@
 dimensions of human color perception very poorly. The three dimensions are
-confounded which is most easily seen when converting the vertical sections to 
-gray scale images in Figure~\ref{fig:hsv-hcl-gray}. Clearly, the brightness of
+confounded:
+%% which is most easily seen when converting the vertical sections to 
+%% gray scale images in Figure~\ref{fig:hsv-hcl-gray}. Clearly,
+the brightness of
 colors is not uniform over hues and saturations (given value)---therefore,
@@
 CIELUV, HCL colors are obtained, defined by
-a triplet $(H, C, L)$ with $H \in [0, 360]$ and $C, L \in [0, 100]$. HCL space
-has the shape of a distorted double cone: the admissible combinations of
+a triplet $(H, C, L)$ with $H \in [0, 360]$ and $C, L \in [0, 100]$.
+%% HCL space has the shape of a distorted double cone: the admissible combinations of
+%% chroma and luminance coordinates (within the space's boundaries) depend on the hue chosen.
+%% The lower panel of Figure~\ref{fig:hsv-hcl} shows
+%% vertical sections through this space: each of the resulting hue/chroma planes (given
+%% luminance) is now properly balanced towards the same gray (going from black to white
+%% with increasing luminance) which becomes obvious when converting the colors to
+%% a gray scale as in Figure~\ref{fig:hsv-hcl-gray}. 
+Given a certain luminance $L$, all colors resulting from different combinations
+of hue $H$ and chroma $C$ are balanced towards the same gray (and look the same
+when converted to a gray scale). However, the admissible combinations of
 chroma and luminance coordinates (within the space's boundaries) depend on the hue chosen.
-The lower panel of Figure~\ref{fig:hsv-hcl} shows
-vertical sections through this space: each of the resulting hue/chroma planes (given
-luminance) is now properly balanced towards the same gray (going from black to white
-with increasing luminance) which becomes obvious when converting the colors to
-a gray scale as in Figure~\ref{fig:hsv-hcl-gray}. This balancing of HCL colors
+The reason for this is that some hues lead to light and others to dark colors,
+e.g., full chroma yellow is brighter (i.e., has higher luminance) than full chroma blue.
+
+The balancing of HCL colors
 gives us the opportunity to conveniently choose color palettes which code categorical
@@
 are chosen, given $C = 50$ and $L = 70$.
+
+\setkeys{Gin}{width=\textwidth}
+\begin{figure}[p]
+\begin{center}
+<<wheel, fig=TRUE, height=4, width=11>>=
+par(mar = rep(0.8, 4), mfrow = c(1, 2))
+par(mfrow = c(1,2), mar = c(1,1,1,1), oma = c(0,0,0,0))
+pie(rep(1,6), radius = 1, col = rainbow(6), labels = 360 * 0:5/6)
+pie(rep(1,6), radius = 1, col = rainbow_hcl(6, c = 50, l = 70), labels = 360 * 0:5/6)
+@
+\caption{\label{fig:wheel} HSV-based and HCL-based color wheel.}
+\end{center}
+\end{figure}
 
@@
 <<pal-q-examples, fig=TRUE, height=4, width=6>>=
-par(mfrow = c(2, 2), mar = c(0, 0, 3, 0))
+par(mfrow = c(2, 2), mar = c(0, 0, 3, 2))
 pal(rainbow_hcl(4, start = 30, end = 300), main = "dynamic [30, 300]")
@@
 <<pal-s-examples, fig=TRUE, height=5, width=6>>=
-par(mfrow = c(5, 1), mar = c(0, 0, 0, 0))
+par(mfrow = c(5, 1), mar = c(0.25, 0, 0.25, 0))
 pal(sequential_hcl(12, c = 0, power = 2.2))
@@
 residuals or correlations (both with the neutral value 0) or binary classification
-probabilities (with neutral value 0.5) that could be visualized in mosaic plots (see Figure~\ref{fig:arthritis})
+probabilities (with neutral value 0.5) that could be visualized in mosaic plots (see Figure~\ref{fig:haireye})
 or classification maps (see Figure~\ref{fig:class}). Analogously to the previous
@@
 <<pal-d-examples, fig=TRUE, height=4, width=6>>=
-par(mfrow = c(4, 1), mar = c(0, 0, 0, 0))
+par(mfrow = c(4, 1), mar = c(0.25, 0, 0.25, 0))
 pal(diverge_hcl(7))
@@
 rather than the green/red axis as contrasts on the latter axis are more difficult to 
-distinguish for color-blind people.
+distinguish for color-blind people \citep{vcd:Lumley:2006}.
+
+%%FIXME: comment on vcd:Lumley:2006
 
@@
 whereas the others mostly rely on chroma contrasts.
-
-\section{Illustrations} \label{sec:illustrations}
-
-In this section, we show a collection of examples for the various types of palettes applied
-to statistical graphics. The first example illustrates qualitative palettes
-and visualizes data from the 2005 election for the German parliament ``Bundestag''.
-In this election, five parties were able to obtain enough votes to enter the
-Bundestag---the distribution of seats is depicted in a pie chart in Figure~\ref{fig:seats}.
-The colors used are rough metaphors for the political parties, using a red hue $H = 0$ for the
-social democrats SPD, a blue hue $H = 240$ for the conservative CDU/CSU, a yellow hue $H = 60$
-for the liberal FDP, a green hue $H = 120$ for the green party ``Die Gr\"unen'' and a purple 
-hue $H = 300$ for the leftist party ``Die Linke''. To obtain rather intense colors, we set
-chroma to $C = 60$ and luminance to $L = 75$ for all hues
-leading to a much better balanced palette compared to the fully saturated
-colors usually found in the (German) media (which instead of blue typically use the extreme
-``color'' black for the CDU/CSU). The pie chart clearly shows that neither the governing
-coalition of SPD and Gr\"une nor the opposition of CDU/CSU and FDP could assemble a majority.
-Given that no party would enter a coalition with the leftists, this lead to a big coalition
-of CDU/CSU and SPD. Figure~\ref{fig:votes} shows the distribution of votes in this election
-stratified by province (Bundesland) in a highlighted mosaic display. The order of provinces is from 
-north to south, first for the 10 western provinces (the former Federal Republic of Germany, FRG),
-then for the 6 eastern provinces (the former German Democratic Republic, GDR). 
-Clearly, the SPD performed better in the north and the CDU/CSU better in the south; furthermore,
-Die Linke performed particularly well in the eastern provinces and in Saarland.
-
-\setkeys{Gin}{width=.6\textwidth}
-\begin{figure}[p]
-\begin{center}
-<<seats, fig=TRUE, height=5, width=6>>=
-par(mar = rep(0.8, 4))
-seats <- structure(c(226, 61, 54, 51, 222),
-  .Names = c("CDU/CSU", "FDP",  "Linke", "Gruene", "SPD"))
-parties <- rainbow_hcl(6, c = 60, l = 75)[c(5, 2, 6, 3, 1)]
-names(parties) <- names(seats)
-names(seats)[4] <- "Grüne"
-pie(seats, clockwise = TRUE, col = parties, radius = 1)
-@
-\caption{\label{fig:seats} Seats in the German parliament.}
-\end{center}
-\end{figure}
-
-\setkeys{Gin}{width=.8\textwidth}
-\begin{figure}[p]
-\begin{center}
-<<votes, fig=TRUE, height=7, width=8.6>>=
-data("Bundestag2005")
-votes <- Bundestag2005[c(1, 3:5, 9, 11, 13:16, 2, 6:8, 10, 12),
-                       c("CDU/CSU", "FDP", "SPD", "Gruene", "Linke")]
-parties <- parties[colnames(votes)]
-dimnames(votes)[[2]][c(4, 5)] <- c("Gr", "Li")
-mosaic(votes, gp = gpar(fill = parties),
-  spacing = spacing_highlighting, labeling = labeling_left,
-  labeling_args = list(rot_labels = c(0, 90, 0, 0), pos_labels = "center",
-  just_labels = c("center","center","center","right"), varnames = FALSE),
-  margins = unit(c(2.5, 1, 1, 12), "lines"),
-  keep_aspect_ratio = FALSE)
-@
-\caption{\label{fig:votes} Votes in the German election 2005.}
-\end{center}
-\end{figure}
-
-To illustrate sequential palettes, a bivariate density estimation for the Old
-Faithful geyser eruptions data \citep{vcd:Azzalini+Bowman:1990} is visualized.
-The Old Faithful geyser is one of the most popular sites in Yellowstone National Park
-and it is of some interest to understand the relation ship between the duration
-of a geyser eruption and the waiting time for this eruption.
-Figure~\ref{fig:bkde} shows (two versions of) a heatmap of a bivariate kernel density estimate of 
-duration and waiting time revealing a multi-modal bivariate distribution: short waiting
-times (around 50 minutes) are typically followed by a long eruption (around 4 minutes)
-whereas long waiting times (around 80 minutes) can be followed by either a long
-or short eruption (around 4 minutes). Both versions of the heatmap use a sequential
-palette as derived in Section~\ref{sec:sequential} balanced towards the same gray levels
-with $L \in [30, 90]$ and $p_2 = 1.5$. The sequential palette in the left panel uses only gray
-colors (i.e., $C_{\max} = 0$) and the palette in the right panel additionally employs
-colors with $H \in [0, 90]$, $C \in [30, 80]$ and $p_1 = 1/5$. The parameters have been 
-chosen such that the peaks in the density are brought out very clearly by using a rather
-large luminance contrast and $p_2 > 1$.
-
-\setkeys{Gin}{width=\textwidth}
-\begin{figure}[t!]
-\begin{center}
-<<bkde, height=6, width=12>>=
-if(generate_png) {
-  png(file = "bkde.png", height = 420, width = 840)
-  par(mfrow = c(1, 2))
-  data("geyser", package = "MASS")
-  ## geyser2 <- cbind(geyser$duration[-299], geyser$waiting[-1])
-  geyser2 <- cbind(geyser$duration, geyser$waiting)
-  dens <- bkde2D(geyser2, bandwidth = c(0.2, 3), gridsize = c(201, 201))
-  image(dens$x1, dens$x2, dens$fhat, xlab = "duration", ylab = "waiting time", 
-    col = rev(sequential_hcl(33, c = 0, l = c(30, 90), power = c(1/5, 1.5))))
-  box()
-  image(dens$x1, dens$x2, dens$fhat, xlab = "duration", ylab = "waiting time", 
-    col = rev(heat_hcl(33, c = c(80, 30), l = c(30, 90), power = c(1/5, 1.5))))
-  box()
-  dev.off()
-}
-@
-\includegraphics{bkde}
-\caption{\label{fig:bkde} Bivariate density estimation for Old Faithful geyser eruptions.}
-\end{center}
-\end{figure}
-
-As pointed out in Section~\ref{sec:diverging}, diverging palettes are particularly useful
-when visualizing residuals or correlations (with natural neutral value 0) or probabilities 
-in 2-class supervised learning (with neutral value 0.5). Examples for both situations are
-provided here. Figure~\ref{fig:arthritis} visualizes the outcome of a double-blind clinical
-trial investigating a new treatment for rheumatoid arthritis. The mosaic rectangles alone
-signal that the treatment lead to higher improvement compared to the placebo group; this effect
-is shown to be significant by the shading that codes the size of the Pearon residuals.
-Positive residuals, corresponding
-to more observations in the corresponding cell than expected under independence, are depicted
-in blue, negative residuals in red. Light colors signal significance at 10\% level, full colors
-significance at 1\% level. Hence, it can be concluded that there are significantly more
-marked improvements in the treated group and significantly fewer in the placebo group than
-would be expected under independence between treatment and improvement. More details can
-be found in \cite{vcd:Zeileis+Meyer+Hornik:2005}.
-
-Figure~\ref{fig:class} shows a heatmap with the fit of a support vector machine (SVM) to an artificial 
-2-class supervised learning example: a mixture of two bivariate normal distributions with
-different means and covariance matrices.
-The circles and triangles show the original observations, solid symbols correspond to
-the support vectors found. The shading underlying the plot visualizes the fitted decision
-values: values around 0 are on the decision boundary and are shaded in light gray, while
-regions that are firmly classified to one or the other class are shaded in full blue and
-red respectively.
-
-Both diverging palettes employ a smaller contrast on the luminance axis ($L \in [50, 90]$)
-and a larger chroma contrast ($C \in [0, 100]$) compared to the sequential palettes from
-the previous example. For Figure~\ref{fig:class}, we set $p_1$ and $p_2$ to the somewhat
-smaller value of $1.3$ to bring out all deviations from the decision boundary and not only the
-extreme decision values. For Figure~\ref{fig:arthritis}, the choice of
-$p_1$ and $p_2$ is less important because only 5 colors are selected---we use values
-$p_1 = p_2 = 1$.
-
-\setkeys{Gin}{width=.7\textwidth}
-\begin{figure}[p]
-\begin{center}
-<<arthritis, fig=TRUE, height=6, width=7>>=
-art <- xtabs(~ Treatment + Improved, data = Arthritis, subset = Sex == "Female")
-names(dimnames(art))[2] <- "Improvement"
-set.seed(1071)
-mosaic(art, gp = shading_max, gp_args = list(n = 5000))
-@
-\caption{\label{fig:arthritis} Extended mosaic display for arthritis data.}
-\end{center}
-\end{figure}
-
-\setkeys{Gin}{width=.7\textwidth}
-\begin{figure}[p]
-\begin{center}
-<<class, height=6, width=8>>=
-if(generate_png) {
-  png(file = "class.png", height = 480, width = 640)
-  set.seed(123)
-  x1 <- rmvnorm(75, mean = c(1.5, 1.5), sigma = matrix(c(1, 0.8, 0.8, 1), ncol = 2))
-  x2 <- rmvnorm(75, mean = c(-1, -1), sigma = matrix(c(1, -0.3, -0.3, 1), ncol = 2))
-  X <- rbind(x1, x2)
-  ex1 <- data.frame(class = factor(c(rep("a", 75), rep("b", 75))), x1 = X[,1], x2 = X[,2])
-  rm(x1, x2, X)
-  fm <- ksvm(class ~ ., data = ex1, C = 0.5)
-  plot(fm, data = ex1)
-  ## uses: diverge_hcl(n, c = c(100, 0), l = c(50, 90), power = 1.3)
-  dev.off()
-}
-@
-\includegraphics{class}
-\caption{\label{fig:class} SVM classification plot.}
-\end{center}
-\end{figure}
 
@@
 with an open-source implementation of HCL (and other color spaces) in the package
-\pkg{colorspace} \citep{vcd:Ihaka:2004} which we build upon for implementing all
+\pkg{colorspace} \citep{vcd:Ihaka:2006} which we build upon for implementing all
 palettes described above. The functions outlined below are provided by the package
@@
 in the chroma or luminance direction (or both) should be employed. In our
-experience (as illustrated in Section~\ref{sec:illustrations}), chroma contrasts
+experience (as illustrated in Section~\ref{sec:motivation}), chroma contrasts
 work sufficiently well if a small set of colors is used. However, when a larger
@@
 graphical parameters and thus adapting the colors to their particular graphical display.
-
-%% From our experience, the paths through HCL space described in the previous sections
-%% in principle work very well and can be used to effectively code qualitative and 
-%% quantitative information. However, there are some open questions, formulated below.
-%% In particular, we are not sure how we can properly define trade-offs
-%% between increasing the range with respect to one dimension and decreasing the
-%% range with respect to another.
-%% 
-%% \begin{itemize}
-%%   \item In sequential/diverging palettes, there is a trade-off between high chroma and high luminance.
-%%         How should this be chosen in practice? Our impression is that when a small set of colors
-%%         (such as 3 or 4) are used, large differnces in chroma work well and large differences
-%%         in luminance are not necessary. However, when a larger set of colors is used (e.g., for 
-%%         heatmaps where extreme values should be identifiable) it is much more important to have a big
-%%         difference in luminance.
-%% 
-%%   \item How should the intensity $|i|$ be increased from 0 to 1?
-%%         Our experience is that for a small set of colors linear increase is sufficient ($p = 1$)
-%%         whereas in heatmaps where only very extreme regions are interesting a $p > 1$
-%%         should be used.
-%% 
-%%   \item For diverging palettes, how should pairs of colors be chosen? The hues $H = 0$
-%%         and $H = 260$ where chosen because they are on opposite sides of the color wheel
-%% 	but have a very similar chroma/luminance plane.
-%% 	
-%%   \item To allow more convenient navigation in HCL space, some more \proglang{R} infrastructure
-%%         could be helpful. We are not sure what would the best or most natural interface for this.
-%% 	A good start might be a function that computes the admissible value on one axis given 
-%% 	a particular value on the other two axes. Maybe you have already experimented with 
-%% 	something like this?
-%% \end{itemize}
 
@@
 \code{polarLUV(l, c, h)} function proveded by the \pkg{colorspace} package
-\citep{vcd:Ihaka:2004}.
+\citep{vcd:Ihaka:2006}.
 A simple convenience function for displaying a certain palette is

### 2006-12-30 trunk/vcd/paper/Color/Nigeria/bnd.R
started TAS revision: non-technical with bad/good examples
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@634 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -1,0 +1,154 @@
+## changes:
+##   - readbndfile() changed to read.bnd()
+##     this returns a suitable object (of class "bnd")
+##     and does *not* yet assign it to some variable
+##   - added quiet argument (as in scan())
+##
+##   - added print() and range() method
+##
+##   - added plot() method (not unsimilar to drawmap())
+##   - changed interface to be similar to plot()
+##     support standard graphical parameters (xlim, xlab, main, ...)
+##     removed specialized color code (replaced by col argument)
+##     removed specialized x11() and postscript() code
+
+## comments:
+##  - better than grey(0:(n-1)/(n-1)) is grey.colors(n)
+##    which (by default) stays away from the extreme colors
+##    black and white and uses a gamma-correction
+##  - T, F should be better TRUE, FALSE
+##  - don't use if(foo == TRUE), just say if(foo)
+##  - don't use return(invisible()) just invisible() is enough
+
+
+read.bnd <- function(file, quiet = TRUE)
+{
+        owarn <- getOption("warn")
+	options(warn = -1)
+
+	data.raw <- scan(file, what = list("", ""), sep = ",", quote = "", quiet = quiet)
+	data.numeric <- list(as.numeric(data.raw[[1]]), as.numeric(data.raw[[2]]))
+
+	npolygon <- sum(is.na(data.numeric[[1]])) - sum(data.raw[[1]] == "is.in")
+	if(!quiet) cat("Note: map consists of", npolygon, "polygons\n")
+	if(!quiet) cat("Reading map ...\n")
+
+	map <- list()
+	i <- 1
+
+	for(k in 1:npolygon) {
+		j <- 1
+		npoints <- data.numeric[[2]][i]
+		if(is.na(data.numeric[[1]][i + 1]) && is.na(data.numeric[[2]][i + 1])) {
+			npoints <- npoints + 1
+		}
+		elem1 <- data.numeric[[1]][i+1:npoints]
+		elem2 <- data.numeric[[2]][i+1:npoints]
+		map[[k]] <- matrix(c(elem1, elem2), ncol = 2)
+		names(map)[k] <- substring(data.raw[[1]][i], 2, nchar(data.raw[[1]][i]) - 1)
+		i <- i + npoints + 1
+	}
+
+	if(sum(is.na(as.numeric(names(map)))) == 0) {
[…]

### 2006-12-30 trunk/vcd/paper/Color/functions.R

@@ -1,7 +1,8 @@
 pal <- function(col, border = "light gray", ...)
 {
   n <- length(col)
-  plot(0, 0, type="n", xlim = c(0, 1), ylim = c(0, 1), axes = FALSE, xlab = "", ylab="", ...)
+  plot(0, 0, type="n", xlim = c(0, 1), ylim = c(0, 1),
+    xaxs = "i", yaxs = "i", axes = FALSE, xlab = "", ylab="", ...)
   rect(0:(n-1)/n, 0, 1:n/n, 1, col = col, border = border)
 }
 

### 2007-01-04 trunk/vcd/paper/Color/TAS.Rnw
first draft for TAS revision finished
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@635 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \SweaveOpts{engine=R, eps=FALSE, echo=FALSE, results=hide}
-\usepackage{amsfonts}
+\usepackage{amsfonts,rotating}
 
@@
 \textit{Computational Statistics \& Data Analysis}. The examples have been selected to provide
-an overview of (typically colored) statistical graphics and their respective pitfalls (when
-it comes to color palette choice). All graphics are in principle very useful but can be even enhanced
-by using better color schemes.
+an overview of typically colored statistical graphics and their respective pitfalls (concerning
+color palette choice).
+%% All graphics are in principle very useful but can be even enhanced
+%% by using better color schemes.
 
-Our first illustration is a heatmap, a very popular display for visualizing a function of
+Our first illustration is a heatmap, a very popular display for visualizing a scalar function of
 two arguments. Here, we use a bivariate kernel density estimate \citep[for the Old
@@
 in printed papers when the journal does not offer color graphics---however, in journals
-that support color graphics, on presentation slides and in interactive usage (in statistical
-software packages), many users prefer to have colored displays and most often HSV
-palettes (as in the left two panels) are used. These palettes code the variable of interest
+that support color graphics (or on presentation slides and in interactive usage in statistical
+software packages), many users prefer to have colored displays and most often use HSV
+palettes (as in the two left two panels). These palettes code the variable of interest
 by varying hue in an HSV color wheel as done by \citet[Figure~4a]{vcd:Yang+Buckley+Dudoit:2002}
@@
 The latter works a bit better than the former, however both palettes exhibit several drawbacks.
-The modes in the map look much more like ``rings'' than a smoothly increasing/decreasing 
+The modes in the map look much more like ``rings'' rather than a smoothly increasing/decreasing 
 density. The heatmap looks very flashy which---although good for drawing attention to a plot---makes
@@
 interesting high-density regions and not to the large low-density regions surrounding them.
-
-To illustrate sequential palettes,
-
- Both versions of the heatmap use a sequential
-palette as derived in Section~\ref{sec:sequential} balanced towards the same gray levels
-with $L \in [30, 90]$ and $p_2 = 1.5$. The sequential palette in the left panel uses only gray
-colors (i.e., $C_{\max} = 0$) and the palette in the right panel additionally employs
-colors with $H \in [0, 90]$, $C \in [30, 80]$ and $p_1 = 1/5$. The parameters have been 
-chosen such that the peaks in the density are brought out very clearly by using a rather
-large luminance contrast and $p_2 > 1$.
 
@@
 
-Figure~\ref{fig:nigeria} is taken from \citep[Figure~5, left]{vcd:Kneib:2006}
+A similar example is presented in Figure~\ref{fig:nigeria}
+\citep[taken from][Figure~5, left]{vcd:Kneib:2006}, depicting posterior mode estimates
+for childhood mortality in different regions of Nigeria. Deviations from a model for
+childhoold mortality are brought out by shading a map according to the corresponding
+residuals, revealing decreased mortality in the south-west and increased mortality in
+the north-east. \cite{vcd:Kneib:2006} uses an HSV-based palette coding the residuals by
+the hue, going from green via yellow to red. Our HCL-based palette also employs green
+and red hues for negative and postive residuals respectively, but codes neutral values
+(around 0) by a neutral light gray. Compared to the HSV-based palette, this offers again
+a number of advantages: only the interesting areas are highlighted by high-chroma colors;
+flashy colors are avoided making it easier to look at the display for a longer time; positive
+and negative residuals with the same absolute size receive the same perceptual weight
+by being balanced towards the same gray.
 
 \setkeys{Gin}{width=\textwidth}
-\begin{figure}[p]
+\begin{figure}[t!]
 \begin{center}
@@
 
+Although problematic for many tasks, pie charts can be useful for visualizing
+whether a set of pie segments constitutes a majority. A typical application is shown
+in Figure~\ref{fig:seats}, visualizing the distribution of seats in the German parliament
+``Bundestag'' following the 2005 election. In this election, five parties were able to
+obtain enough votes to enter the Bundestag---however, neither the governing
+coalition of SPD and Gr\"une nor the opposition of CDU/CSU and FDP could assemble a majority.
+Given that no party would enter a coalition with the leftists ``Die Linke'', this lead to
+a big coalition of CDU/CSU and SPD. In graphical displays, the parties are usually matched
+by using colors as metaphors: red for the social democrats SPD, black for the conservative
+CDU/CSU, yellow for the liberal FDP, green for the green party ``Die Gr\"unen'' and purple 
+for the leftist party ``Die Linke''. The left panel shows fully saturated HSV colors as usually
+found in the (German) media whereas the right panel uses less flashy HCL colors with the
+same hues (except for the CDU/CSU where a blue hue instead of the extreme ``color'' black
+is used). The advantage of the latter is again that they are easier to hold in focus for
+a longer time. Furthermore, they are all balanced towards the same gray and have the same
+amount of color, resulting in a perceptually balanced palette that does not introduce undesired
+graphical distractions. While it could be argued that this pie chart is such a simple
+display that a well-balanced palette is not so important and that flashy colors are to
+be preferred, the balancing properties of the HCL-based palette are very important in more
+complex displays such as the mosaic display in Figure~\ref{fig:votes}. This re-uses the same
+colors and shows the distribution of votes in the 2005 election
+stratified by province (Bundesland). The order of provinces is from 
+north to south, first for the 10 western provinces (the former Federal Republic of Germany, FRG),
+then for the 6 eastern provinces (the former German Democratic Republic, GDR). 
+Clearly, the SPD performed better in the north and the CDU/CSU better in the south; furthermore,
+Die Linke performed particularly well in the eastern provinces and in Saarland.
+
 \setkeys{Gin}{width=\textwidth}
-\begin{figure}[p]
+\begin{sidewaysfigure}[p]
 \begin{center}
 <<votes, fig=TRUE, height=6, width=14>>=
-data("Bundestag2005")
+data("Bundestag2005", package = "vcd")
 votes <- Bundestag2005[c(1, 3:5, 9, 11, 13:16, 2, 6:8, 10, 12),
@@
 parties_hsv <- parties_hsv[colnames(votes)]
+dimnames(votes)[[1]][c(9, 16)] <- c("Baden-Württemberg", "Thüringen")
 dimnames(votes)[[2]][c(4, 5)] <- c("Gr", "Li")
@@
 \end{center}
-\end{figure}
+\end{sidewaysfigure}
 
-The first example illustrates qualitative palettes
-and visualizes data from the 2005 election for the German parliament ``Bundestag''.
-In this election, five parties were able to obtain enough votes to enter the
-Bundestag---the distribution of seats is depicted in a pie chart in Figure~\ref{fig:seats}.
-The colors used are rough metaphors for the political parties, using a red hue $H = 0$ for the
-social democrats SPD, a blue hue $H = 240$ for the conservative CDU/CSU, a yellow hue $H = 60$
-for the liberal FDP, a green hue $H = 120$ for the green party ``Die Gr\"unen'' and a purple 
-hue $H = 300$ for the leftist party ``Die Linke''. To obtain rather intense colors, we set
-chroma to $C = 60$ and luminance to $L = 75$ for all hues
-leading to a much better balanced palette compared to the fully saturated
-colors usually found in the (German) media (which instead of blue typically use the extreme
-``color'' black for the CDU/CSU). The pie chart clearly shows that neither the governing
-coalition of SPD and Gr\"une nor the opposition of CDU/CSU and FDP could assemble a majority.
-Given that no party would enter a coalition with the leftists, this lead to a big coalition
-of CDU/CSU and SPD. Figure~\ref{fig:votes} shows the distribution of votes in this election
-stratified by province (Bundesland) in a highlighted mosaic display. The order of provinces is from 
-north to south, first for the 10 western provinces (the former Federal Republic of Germany, FRG),
-then for the 6 eastern provinces (the former German Democratic Republic, GDR). 
-Clearly, the SPD performed better in the north and the CDU/CSU better in the south; furthermore,
-Die Linke performed particularly well in the eastern provinces and in Saarland.
+Color choice is usually much more important in graphical displays with
+shaded areas compared to displays with only points or lines. However, color choice
+becomes much more relevant also in scatter plots when there are many points. As an
+example, Figure~\ref{fig:scatter} depicts a scatter plot with three clusters similar to 
+the one shown in \citet[Figure~3]{vcd:Celeux+Hurn+Robert:2000}. To indicate cluster
+membership, three different colors (green, yellow, red) are employed. The HSV colors in
+the left panel are again very flashy and differ strongly with respect to luminance: the
+yellow is much lighter and hardly visible. In contrast, the HCL colors in the right
+panel use the same hues, but are balanced with respect to chroma and luminance, i.e.,
+the amount of color and gray.
+
+As our final example, Figure~\ref{fig:haireye} \citep[taken from][Figure~11, left/middle]{vcd:Friendly:2002}
+visualizes the cross-tabulation of hair end eye color of 592 students in a mosaic display.
+Clearly, hair and eye color are not independent and the pattern of association is highlighted
+by means of residual-based shading. Cells associated with Pearson residuals whose absolute value exceeds
+(2 or) 4 are shaded (light) blue/red. This shows that there are significantly more students
+with black hair and brown eyes, blond hair and blue eyes, red hair and green eyes than expected
+under independence. Conversely, fewer students than expected have blond hair and brown/hazel eyes
+or black hair and blue eyes. Comparing the HSV and HCL colors in Figure~\ref{fig:haireye},
+it is shown again that the HCL colors are better balanced (between the red and blue colors)
+and less flashy. Even the cells associated with small residuals (below 2) are a bit easier
+to read when shaded in a light grey rather than white. More details and extensions to residual-based
+shadings can be found in \cite{vcd:Zeileis+Meyer+Hornik:2007}.
 
 \setkeys{Gin}{width=\textwidth}
-\begin{figure}[p]
+\begin{figure}[h!]
 \begin{center}
@@
 
-similar to \citet[Figure~3]{vcd:Celeux+Hurn+Robert:2000}
-
 \setkeys{Gin}{width=\textwidth}
-\begin{figure}[p]
+\begin{figure}[h!]
 \begin{center}
@@
 \end{figure}
-
-taken from \citet[Figure~11, left/middle]{vcd:Friendly:2002}
-
-As pointed out in Section~\ref{sec:diverging}, diverging palettes are particularly useful
-when visualizing residuals or correlations (with natural neutral value 0) or probabilities 
-in 2-class supervised learning (with neutral value 0.5). Examples for both situations are
-provided here. Figure~\ref{fig:haireye} visualizes the outcome of a double-blind clinical
-trial investigating a new treatment for rheumatoid arthritis. The mosaic rectangles alone
-signal that the treatment lead to higher improvement compared to the placebo group; this effect
-is shown to be significant by the shading that codes the size of the Pearon residuals.
-Positive residuals, corresponding
-to more observations in the corresponding cell than expected under independence, are depicted
-in blue, negative residuals in red. Light colors signal significance at 10\% level, full colors
-significance at 1\% level. Hence, it can be concluded that there are significantly more
-marked improvements in the treated group and significantly fewer in the placebo group than
-would be expected under independence between treatment and improvement. More details can
-be found in \cite{vcd:Zeileis+Meyer+Hornik:2007}.
-
-Both diverging palettes employ a smaller contrast on the luminance axis ($L \in [50, 90]$)
-and a larger chroma contrast ($C \in [0, 100]$) compared to the sequential palettes from
-the previous example. 
-For Figure~\ref{fig:haireye}, the choice of
-$p_1$ and $p_2$ is less important because only 5 colors are selected---we use values
-$p_1 = p_2 = 1$.
 
@@
 %% gray scale images in Figure~\ref{fig:hsv-hcl-gray}. Clearly,
-the brightness of
+The brightness of
 colors is not uniform over hues and saturations (given value)---therefore,
@@
 
-Qualitative palettes are sets of colors for depicting different categories,
-i.e., for coding a categorical variable. Usually, these should give the same
-perceptual weight to each category so that no group is perceived to be larger or more
-important than any other one. Typical applications of qualitative palettes in
-statistics would be bar plots \citep[see][]{vcd:Ihaka:2003}, pie charts (see Figure~\ref{fig:seats}) 
-or highlighted mosaic displays (see Figure~\ref{fig:votes}).
-
-\cite{vcd:Ihaka:2003} describes a simple strategy for choosing such palettes: chroma
-and luminance are kept fixed and only the hue is varied for obtaining different colors
-which are consequently all balanced towards the same gray. As Figure~\ref{fig:hsv-hcl}
-illustrates, the range of hues available depends on the combination of chroma and
-luminance chosen. Figure~\ref{fig:pal-q-construct} depicts how three colors
-are chosen, given $C = 50$ and $L = 70$.
-
 \setkeys{Gin}{width=\textwidth}
-\begin{figure}[p]
+\begin{figure}[b!]
 \begin{center}
@@
 \end{figure}
-
-Various strategies for choosing the hues in a certain palette are conceivable.
-A simple and intuitive one is to use colors as metaphors for categories
-(e.g., for political parties), another approach would be to use segments from
-the color wheel corresponding to nearby or distant colors. The latter is shown
-in Figure~\ref{fig:pal-q-examples} which depicts
-examples for generating qualitative sets of colors $(H, 50, 70)$.
-In the upper left panel colors from the full spectrum are used ($H = 30, 120, 210, 300$)
-creating a `dynamic' set of colors. The upper right panel shows a `harmonic'
-set with $H = 60, 120, 180, 240$. Warm colors (from the blue/green part of the spectrum: 
-$H = 270, 230, 190, 150$) and cold colors (from the yellow/red part of the spectrum:
-$H = 90, 50, 10, 330$) are shown in the lower left and right panel, respectively.
 
@@
 
+Qualitative palettes are sets of colors for depicting different categories,
+i.e., for coding a categorical variable. Usually, these should give the same
+perceptual weight to each category so that no group is perceived to be larger or more
+important than any other one. Typical applications of qualitative palettes in
+statistics would be bar plots \citep[see][]{vcd:Ihaka:2003}, pie charts (see Figure~\ref{fig:seats}) 
+or highlighted mosaic displays (see Figure~\ref{fig:votes}).
+
+\cite{vcd:Ihaka:2003} describes a simple strategy for choosing such palettes: chroma
+and luminance are kept fixed and only the hue is varied for obtaining different colors
+which are consequently all balanced towards the same gray. This effect is illustrated
+in Figure~\ref{fig:wheel} which shows a color wheel obtained by varying the hue only in
+HSV coordinates $(H, 100, 100)$ and HCL coordinates $(H, 50, 70)$. Clearly, not only the
+hue but also the amount of chroma and luminance varies for the HSV wheel.
+
+Figure~\ref{fig:pal-q-construct} depicts how the HCL-based color wheel is constructed.
+It shows the hue/chroma plane of HCL space given a luminance of $L = 70$. Not all
+combinations of hue and chroma are admissible, however for a chroma of $C = 50$ a full
+color wheel can be obtained.
+For choosing the hues in a certain palette, various strategies are conceivable.
+A simple and intuitive one is to use colors as metaphors for categories
+(e.g., for political parties as in Figures~\ref{fig:seats} and \ref{fig:votes}),
+another approach would be to use segments from
+the color wheel corresponding to nearby or distant colors. The latter is shown
+in Figure~\ref{fig:pal-q-examples} which depicts
+examples for generating qualitative sets of colors $(H, 50, 70)$.
+In the upper left panel colors from the full spectrum are used ($H = 30, 120, 210, 300$)
+creating a `dynamic' set of colors. The upper right panel shows a `harmonic'
+set with $H = 60, 120, 180, 240$. Warm colors (from the blue/green part of the spectrum: 
+$H = 270, 230, 190, 150$) and cold colors (from the yellow/red part of the spectrum:
+$H = 90, 50, 10, 330$) are shown in the lower left and right panel, respectively.
+
 \subsection{Sequential palettes} \label{sec:sequential}
@@
 black ($L = 0$). Instead of going linearly from light to dark gray, luminance
-could also be increased nonlinearly, e.g., by some function $f(i)$ that
-controls whether luminance is increased quickly with intensity or not. We found $f(i) = i^p$ 
+could also be increased nonlinearly, e.g., by some function $i' = f(i)$ that
+controls whether luminance is increased quickly with intensity or not. We found $i' = f(i) = i^p$ 
 to be a convenient transformation where the power $p$ can be varied to achieve
@@
 Furthermore, the intensity $i$ could additionally be coded by colorfulness (chroma), e.g.,
-  \[ (H, 0 + i^p \cdot C_{\max}, L_{\max} - i^p \cdot (L_{\max} - L_{\min})). \]
+  \[ (H, 0 + i' \cdot C_{\max}, L_{\max} - i' \cdot (L_{\max} - L_{\min})). \]
 This strategy is depicted in the left panel of Figure~\ref{fig:pal-s-construct} for a blue
@@
 simultaneously varying the hue as well:
-  \[ (H_2 - i \cdot (H_1 - H_2), C_{\max} - i^{p_1} \cdot (C_{\max} - C_{\min}),
-      L_{\max} - i^{p_2} \cdot (L_{\max} - L_{\min})). \]
-One application would be an HCL-based version of ``heat colors'' that increase from a light yellow
+  \[ (H_2 - i \cdot (H_1 - H_2), C_{\max} - i' \cdot (C_{\max} - C_{\min}),
+      L_{\max} - i'' \cdot (L_{\max} - L_{\min})). \]
+One application is an HCL-based version of ``heat colors'' that increase from a light yellow
 (e.g., $(90, 30, 90)$) to a full red (e.g., $(0, 100, 50)$). To make the change in hue
 visible, the chroma needs to increase rather quickly for low values of $i$ and then only
-slowly for higher values of $i$. This can be achieved by choosing a power $p_1 < 1$.
+slowly for higher values of $i$. This can be achieved by choosing an appropriate transformation
+$i'$ for chroma and a different transformation $i''$ for the luminance.
 Such a strategy is adopted for the palettes shown in the lower three rows in
 Figure~\ref{fig:pal-s-examples} using different pairs of hues as well as different
-chroma and luminance contrasts. The exact parameter combinations are provided in the
-appendix.
+chroma and luminance contrasts.
 
@@
 residuals or correlations (both with the neutral value 0) or binary classification
-probabilities (with neutral value 0.5) that could be visualized in mosaic plots (see Figure~\ref{fig:haireye})
-or classification maps (see Figure~\ref{fig:class}). Analogously to the previous
+probabilities (with neutral value 0.5) that could be visualized in mosaic plots
+(see Figure~\ref{fig:haireye}), classification maps or model-based shading in maps
+such as Figure~\ref{fig:nigeria}. Analogously to the previous
 section, we suppose that we want to visualize an intensity or interestingness $i$
@@
 geometric shapes in the chroma/luminance plane, allowing for both large chroma and
-luminance contrasts. If potential viewers of the resulting graphic might be color-blind,
+luminance contrasts. 
+If potential viewers of the resulting graphic might be color-blind,
 the pair of hues should be taken from the yellow/blue axis of the color wheel
@@
 distinguish for color-blind people \citep{vcd:Lumley:2006}.
-
-%%FIXME: comment on vcd:Lumley:2006
 
@@
 with an open-source implementation of HCL (and other color spaces) in the package
-\pkg{colorspace} \citep{vcd:Ihaka:2006} which we build upon for implementing all
-palettes described above. The functions outlined below are provided by the package
-\pkg{vcd} \citep{vcd:Meyer+Zeileis+Hornik:2006}---technical documentation for
-the functions along with a large collection of examples is available via
-\verb/help("rainbow_hcl")/. These examples also include comparisons between
-existing \proglang{R} palettes (based on HSV colors) and the new HCL color palettes.
-
-Qualitative palettes are implemented in \pkg{vcd} in the function
-\begin{Soutput}
-rainbow_hcl(n, c = 50, l = 70, start = 0, end = 360*(n-1)/n, ...)
-\end{Soutput}
-where \code{n} controls the number of colors in the palette. The arguments
-\code{c} and \code{l} give the fixed chroma and luminance levels, respectively,
-and \code{start} and \code{end} specify the range of hue angles. The function
-is named after the base \proglang{R} function \code{rainbow()} which has a similar
-interface but chooses colors in HSV coordinates. It computes the HCL coordinates
-and transforms them to RGB by using the \code{hcl2hex()} function which
-is a convenience wrapper to the function \code{polarLUV()} from
-the \pkg{colorspace} package.
-
-Sequential palettes based on a single hue are implemented in the function
-\begin{Soutput}
-sequential_hcl(n, h = 260, c = c(80, 0), l = c(30, 90), power = 1.5, ...) 
-\end{Soutput}
-where the first element of \code{c} and \code{l} give the starting chroma and luminance
-coordinate (by default colorful and dark) and the second element the ending coordinate
-(by default light gray). The \code{power} argument implements the parameter $p$ 
-from the $i^p$ function. Sequential palettes using a range of hues are provided by
-\begin{Soutput}
-heat_hcl(n, h = c(0, 90), c = c(100, 30), l = c(50, 90), power = c(1/5, 1), ...)
-\end{Soutput}
-named after the HSV-based \proglang{R} function \code{heat.colors()} and
-by default starts from a red and going to a yellow hue. The defaults in \code{heat_hcl()}
-are set differently compared to \code{sequential_hcl()} as to make the default HCL heat
-colors more similar to the HSV version. The defaults of \code{sequential_hcl()}, on the
-other hand, are set as to achieve a large contrast on the luminance axis.
-
-Diverging palettes are implemented in the function
-\begin{Soutput}
-diverge_hcl(n, h = c(260, 0), c = 80, l = c(30, 90), power = 1.5, ...) 
-\end{Soutput}
-which has the same arguments as \code{sequential_hcl()} but takes a pair of hues \code{h}.
-
-The \proglang{R} code for reproducing
+\pkg{colorspace} \citep{vcd:Ihaka:2006}. The coordinate transformations mentioned
+above are contained in \proglang{C} code within \pkg{colorspace} that are easy to
+port to other statistical software systems. Based on \pkg{colorspace}, 
+the package \pkg{vcd} \citep{vcd:Meyer+Zeileis+Hornik:2006} provides implementations
+of all palettes discussed above. Qualitative palettes are provided by
+\code{rainbow_hcl()} (named after the HSV-based function \code{rainbow()} in base
+\proglang{R}). Sequential palettes based on a single hue are implemented in the function
+\code{sequential_hcl()} while \code{heat_hcl()} offers sequential palettes based on
+a range of hues. Diverging palettes can be obtained by \code{diverge_hcl()}.
+Technical documentation along with a large collection of examples is available via
+\verb/help("rainbow_hcl")/. \proglang{R} code for reproducing
 Figures~\ref{fig:pal-q-examples}, \ref{fig:pal-s-examples} and \ref{fig:pal-d-examples}
-is given in the appendix. A vignette that additionally explains how the figures from the illustrations
-section can be reproduced is contained in \pkg{vcd} and be accessed via
-\code{vignette("hcl-colors", package = "vcd")}.
+(and some illustrations) can be accessed via \code{vignette("hcl-colors", package = "vcd")}.
 
@@
 Many statistical graphics---especially when displayed on a computer screen,
-e.g., as in interactive usage, electronic papers or presentation slides---
-employ colors to code information about a certain variable. Despite this
+e.g., as in interactive usage, electronic papers or presentation slides---employ
+colors to code information about a certain variable. Despite this
 omnipresence of color, there is often only little guidance in statistical
@@
 set of colors is used (e.g., for heatmaps where extreme values should be identifiable)
-it is much more important to have a big difference in luminance. Another parameter in these 
-palettes is the rate at which the intensity $|i|$ is increased from $0$ to $1$:
-Employing a function $|i|^p$ for this, small
-powers $p$ can be used if palettes with a lot of color should be 
-constructed (e.g., to clearly separate regions in a classification map). On
-the other hand, large powers $p$ will result in palettes with less color
-(e.g., appropriate to highlight only extreme regions firmly classified in such a map).
+it is much more important to have a big difference in luminance.
+%% Another parameter in these 
+%% palettes is the rate at which the intensity $|i|$ is increased from $0$ to $1$:
+%% Employing a function $|i|^p$ for this, small
+%% powers $p$ can be used if palettes with a lot of color should be 
+%% constructed (e.g., to clearly separate regions in a classification map). On
+%% the other hand, large powers $p$ will result in palettes with less color
+%% (e.g., appropriate to highlight only extreme regions firmly classified in such a map).
 
@@
 
-\begin{appendix}
-
-\section[Supplementary R code]{Supplementary \proglang{R} code}
-\label{sec:R}
-
-All palettes discussed above are implemented in the \proglang{R} system for statistical
-computing and graphics \citep{vcd:R:2006} in the package 
-\pkg{vcd} \citep{vcd:Meyer+Zeileis+Hornik:2006}. The functions are based on the
-\code{polarLUV(l, c, h)} function proveded by the \pkg{colorspace} package
-\citep{vcd:Ihaka:2006}.
-A simple convenience function for displaying a certain palette is
-\begin{Soutput}
-pal <- function(col, border = "light gray", ...)
-{
-  n <- length(col)
-  plot(0, 0, type = "n", xlim = c(0, 1), ylim = c(0, 1),
-       axes = FALSE, xlab = "", ylab = "", ...)
-  rect(0:(n-1)/n, 0, 1:n/n, 1, col = col, border = border)
-}
-\end{Soutput}
-
-The qualitative palettes in Figure~\ref{fig:pal-q-examples} are generated via
-\begin{Soutput}
-pal(rainbow_hcl(4, start =  30, end = 300))
-pal(rainbow_hcl(4, start =  60, end = 240))
-pal(rainbow_hcl(4, start = 270, end = 150))
-pal(rainbow_hcl(4, start =  90, end = -30))
-\end{Soutput}
-
-The sequential palettes in Figure~\ref{fig:pal-s-examples} are generated via
-\begin{Soutput}
-pal(sequential_hcl(12, c = 0, power = 2.2))
-pal(sequential_hcl(12, power = 2.2))
-pal(heat_hcl(12, c = c(80, 30), l = c(30, 90), power = c(1/5, 2)))
-pal(terrain_hcl(12, c = c(65, 0), l = c(45, 90), power = c(1/2, 1.5)))
-pal(heat_hcl(12, h = c(0, -100), l = c(75, 40), c = c(40, 80), power = 1))
-\end{Soutput}
-
-The diverging palettes in Figure~\ref{fig:pal-d-examples} are generated via
-\begin{Soutput}
-pal(diverge_hcl(7))
-pal(diverge_hcl(7, c = 100, l = c(50, 90), power = 1))
-pal(diverge_hcl(7, h = c(130, 43), c = 100, l = c(70, 90)))
-pal(diverge_hcl(7, h = c(180, 330), c = 59, l = c(75, 95)))
-\end{Soutput}
-
-Further examples and technical information on all palettes is available from
-the corresponding online documentation.
-
-\end{appendix}
-
 \end{document}

### 2007-01-07 trunk/vcd/paper/Color/TAS.Rnw
small fixes
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@636 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 @
-\caption{\label{fig:haireye} Extended mosaic display for haireye data.}
+\caption{\label{fig:haireye} Extended mosaic display for hair and eye color data.}
 \end{center}
@@
 \end{document}
+
+<<barplot, eval=FALSE>>=
+x <- rbind(
+c(255, 255, 111),
+c(255, 255,  47),
+c(120,  50,   1),
+c(102, 204, 102),
+c( 50, 204, 204),
+c(203, 153, 204),
+c(102, 204, 102),
+c(255,  47,  47),
+c(102, 204, 102))/255
+
+y <- c(10, 20, 25, 18, 16, 13, 21, 15, 18) - 3
+
+par(mfrow = c(1, 2))
+barplot(y, col = hex(RGB(x[,1], x[,2], x[,3])), space = 0)
+## barplot(y, col = hcl(attr(as(RGB(x[,1], x[,2], x[,3]), "polarLUV"), "coords")[,"H"],
+##  c = 40, l = 80), space = 0)
+barplot(y, col = rainbow_hcl(7, c = 40, l = 80)[c(3, 5, 2, 4, 6, 1, 4, 7, 4)], space = 0)
+@

### 2007-01-07 trunk/vcd/paper/Color/Nigeria/bnd.R

@@ -152,3 +152,33 @@ plot.bnd <- function(x, z = NULL,
     invisible(x)
 }
 
+## EXAMPLES
+##
+## ## packages and source code
+## library("vcd")
+## source("bnd.R")
+## 
+## ## read map
+## m <- read.bnd("nigeria37.bnd")
+## 
+## ## read model fit align along district
+## fit <- read.table("reml_purespatial_f_district_spatial.res", header = TRUE)
+## fit <- fit[order(fit$district),]
+## 
+## ## simple maps
+## plot(m)
+## plot(m, fill = "white")
+## 
+## ## shaded maps
+## plot(m, z = fit$pmode)
+## plot(m, z = fit$pmode, fill = rainbow(100, start = 0, end = 1/3))
+## plot(m, z = fit$pmode, fill = diverge_hcl(100))
+## ## but for diverging palettes: use symmetrical zlim
+## 
+## ## compare HSV-based qualitative palette
+## ## with HCL-based diverging palette
+## par(mfrow = c(1, 2))
+## plot(m, z = fit$pmode, fill = rainbow(100, start = 0, end = 1/3),
+##   zlim = c(-0.65, 0.65))
+## plot(m, z = fit$pmode, zlim = c(-0.65, 0.65),
+##   fill = diverge_hcl(100, h = c(0, 120), c = 60, l = c(45, 90), power = 1.2))

### 2007-01-12 trunk/vcd/paper/Color/TAS.Rnw
first TAS revision finished
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@638 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 how to choose a suitable palette for a certain visualization task. Consequently,
-sub-optimal color palettes are used for too many color graphics displayed in journal articles
+sub-optimal color palettes are used quite frequently for color graphics displayed in journal articles
 or on presentation slides which could easily be enhanced if the underlying information
@@
 the general principles suggested can be turned into formulas describing paths of
-coordinates in this space. Given a mapping from HCL to RGB coordinates, our formulas are
-extremely easy to implement in software and we provide such an implementation
-in the \proglang{R} language \citep{vcd:R:2006} using the powerful \proglang{R} graphics
-system \citep[see][]{vcd:Murrell:2006} and the HCL
-color implementation from package \pkg{colorspace} \citep{vcd:Ihaka:2006}. This gives the user
-both a conceptual and computational tool box for experimenting with color palettes for
-a particular display.
+coordinates in this space.
+%% Given a mapping from HCL to RGB coordinates, our formulas are
+%% extremely easy to implement in software and we provide such an implementation
+%% in the \proglang{R} language \citep{vcd:R:2006} using the powerful \proglang{R} graphics
+%% system \citep[see][]{vcd:Murrell:2006} and the HCL
+%% color implementation from package \pkg{colorspace} \citep{vcd:Ihaka:2006}. This gives the user
+%% both a conceptual and computational tool box for experimenting with color palettes for
+%% a particular display.
 
@@
 Section~\ref{sec:software} offers some general remarks on the implementation in statistical
-software packages as well as some details on our implementation in \proglang{R}.
+software packages as well as some details on the implementation in \proglang{R}.
 Section~\ref{sec:discussion} concludes the paper with a discussion.
@@
 could also be increased nonlinearly, e.g., by some function $i' = f(i)$ that
-controls whether luminance is increased quickly with intensity or not. We found $i' = f(i) = i^p$ 
+controls whether luminance is increased quickly with intensity or not. We found $f(i) = i^p$ 
 to be a convenient transformation where the power $p$ can be varied to achieve
@@
 above are contained in \proglang{C} code within \pkg{colorspace} that are easy to
-port to other statistical software systems. Based on \pkg{colorspace}, 
+port to other statistical software systems. Based on \pkg{colorspace} and
+the \proglang{R} graphics system \citep[see][]{vcd:Murrell:2006},
 the package \pkg{vcd} \citep{vcd:Meyer+Zeileis+Hornik:2006} provides implementations
@@
 Technical documentation along with a large collection of examples is available via
-\verb/help("rainbow_hcl")/. \proglang{R} code for reproducing
+\verb/help("rainbow_hcl", package = "vcd")/. \proglang{R} code for reproducing
 Figures~\ref{fig:pal-q-examples}, \ref{fig:pal-s-examples} and \ref{fig:pal-d-examples}
@@
 
-We provide paths through HCL space along perceptual axes
-so that colors selected along these paths match perceptual dimensions.
+We provide paths through HCL space along perceptual axes of the human visual system
+so that colors selected along these paths match human perceptual dimensions.
 This gives the users the possibility to conveniently experiment with the HCL-based

### 2007-04-10 trunk/vcd/R/Bundesliga.R
Bundesliga.R deactivated
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@643 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -1,3 +1,4 @@
+if (FALSE) {
 ## collection of R functions to query http://www.dfb.de/
 ## for Bundesliga results
 
@@ -330,3 +331,4 @@ Bundesliga$Year <- as.integer(as.character(Bundesliga$Year))
 Bundesliga$Round <- as.integer(as.character(Bundesliga$Round))
 Bundesliga$AwayGoals <- as.integer(as.character(Bundesliga$AwayGoals))
 Bundesliga$HomeGoals <- as.integer(as.character(Bundesliga$HomeGoals))
+}

### 2007-10-02 trunk/vcd/paper/Handbook/greyscale/hsc.Rnw
grey palette version added
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@657 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2006-12-11 trunk/vcd/paper/Handbook/hsc.Rnw
@@
 hos_chisq <- coindep_test(hospital, n = 5000, indepfun = function(x) sum(x^2))
+
+## greyscale-Function
+togrey <- function(x) {
+  x <- as(hex2RGB(substr(x, 1, 7)), "polarLUV")
+  x@coords[,2] <- 0
+  hex(x)
+}
 @
@@
 \item Providing a modular and flexible implementation to easily allow
-  user extensions \cite{vcd:Meyer+Zeileis+Hornik:2005a}.
+  user extensions \cite{vcd:Meyer+Zeileis+Hornik:2006b}.
 \end{itemize}
@@
 categories for the visit frequency: regular (including
-%FIXME: lookup regular in Wing-paper.
-patients that were allowed to go home), less than monthly, and never.
+patients who were allowed to go home), less than monthly, and never.
 Wing \cite{vcd:wing:1962} concludes from this data
-that the longer the length of stay in hospital, the less frequent the
-visits, which can be seen from the column-standardized table 
+that the longer the patients stayed in hospital, the less frequent they
+are visited, which can be seen from the column-standardized table
 (see Table~\ref{tab:colhospital}).
@@
 obtained by recursive partitioning splits of a rectangle.
-In the following, we describe the main idea of mosaic plots, 
+In the following, we describe the main idea of mosaic plots;
 Chapter III-12 in this book provides more detailed information. Consider our
@@
 plot. In our example, if we use `Visit frequency' %'
-as the first splitting variable, the impression is very different compared to the previous
-mosaic (see Figure~\ref{fig:mosaic3fig}). 
-In this alternative display, we see the marginal distribution of `Visit frequency' %'
+as the first splitting variable, the impression is very different
+compared to the previous mosaic.
+In this alternative display (see Figure~\ref{fig:mosaic3fig}),
+we see the marginal distribution of `Visit frequency' %'
 in the rows: about half of the patients get visited regularly. This
@@
 of a contingency table by recursive conditional splits. If one variable is
-explanatory, it should be be used first for splitting: thus, the display shows
-the conditional distribution of the dependent variable given the explanatory one.
+explanatory, it should be be used first for splitting;
+the display then shows the conditional distribution of the
+dependent variable given the explanatory one.
 Sieve plots basically visualize the table of \emph{expected} frequencies, and
@@
 par(mfrow = c(1,2), mar = c(1,1,1,1), oma = c(0,0,0,0))
-pie(rep(1,9), radius = 1, col = rainbow(9), labels = 360 * 0:8/9)
-pie(rep(1,9), radius = 1, col = rainbow_hcl(9), labels = 360 * 0:8/9)
+pie(rep(1,9), radius = 1, col = togrey(rainbow(9)), labels = 360 * 0:8/9)
+pie(rep(1,9), radius = 1, col = togrey(rainbow_hcl(9)), labels = 360 * 0:8/9)
 @
@@
          in a monochrome version of this paper, all pies in the right
-         wheel will be shaded with the same gray, i.e., appear identical.}
+         wheel will be shaded with the same gray, i.e., they will
+         appear to be virtually identical.}
 \label{fig:colors}
@@
 plot.new()
-rect(0:4/5, 0.2, 1:5/5, 0.5, border = 0, col = diverge_hcl(5, c = 100, l = c(50, 90)))
+rect(0:4/5, 0.2, 1:5/5, 0.5, border = 0, col = togrey(diverge_hcl(5, c = 100, l = c(50, 90))))
 rect(0, 0.2, 1, 0.5, border = 1, col = NULL)
-rect(0:4/5, 0.55, 1:5/5, 0.85, border = 0, col = diverge_hsv(5))
+rect(0:4/5, 0.55, 1:5/5, 0.85, border = 0, col = togrey(diverge_hsv(5)))
 rect(0, 0.55, 1, 0.85, border = 1, col = NULL)
@@
   In a monochrome version of the paper, the right and left hand
-  sides of the HCL colors palette appear identical. }
+  sides of the HCL colors palette appear to be identical, unlike the
+  HSV color palette. }
 \label{fig:diverging}
@@
 Similarly, diverging HCL color palettes can be derived
-\cite{vcd:Zeileis+Meyer+Hornik:2005} by interpolating again between a
+\cite{vcd:Zeileis+Meyer+Hornik:2007} by interpolating again between a
 neutral color such as $(H, 0, 90)$ and two colors with full chroma
@@
 
-All plots introduced in Section~\ref{sec:twoway} are composed of empty
+The mosaic plots introduced in Section~\ref{sec:twoway} are composed of empty
 tiles. It seems intuitive to use filled tiles to highlight information
@@
 residuals \emph{individually} significant at approximately the $\alpha
-= 0.05$ and $\alpha = 0.0001$ levels. However, the main purpose of the
-Friendly shading is not to visualize significance but the \emph{pattern} of
+= 0.05$ and $\alpha = 0.0001$ levels. However, the main purpose of this
+shading is not to visualize significance but the \emph{pattern} of
 deviation from independence \cite{vcd:Friendly:2000}.
-In addition to the shading of the rectangles themselves, the Friendly shading also
+In addition to the shading of the rectangles themselves, the Friendly
+shading also
 encompasses a choice of line type and line color of the rectangle borders
-with similar ideas as described above, useful when no color can be used.
+with similar ideas as described above.
 
@@
 \end{figure}
-Clearly, the asymmetry for regular and never visited students,
+Clearly, the asymmetry for regular and never visited patients,
 and the pattern inversion for lengths of stay of 2--9 and more than 19
@@
 by using diverging color palettes to code the sign and size of a residual.
-In addition, the significance of test statistics can be visualized by using 
+In addition, the non-significance of test statistics can be visualized by using
 overall less colorful palettes. Using the maximum instead of the $\chi^2$ (or other) test
@@
 `flat' representation of the multi-way table that can be visualized in %'
-ways similar to a two-dimensional table.
+a way similar to a two-dimensional table.
 
@@
 among the departments (roughly speaking, the admission rate is high for A and B,
-low for F and in between for C to E). The rates 
+low for F, and in between for C to E). The rates
 are equal for male and female students, except for department A where
@@
 @
-\caption{Doubledecker plot for the \data{UCBAdmissions} data.}
+\caption{Doubledecker plot for the \data{UCB admissions} data.}
 \label{fig:doubleucb}
@@
 
-One approach is employing pairs plots (such as that in Figure~\ref{fig:pairs}) for model search:
+One approach is employing pairs plots (such as in Figure~\ref{fig:pairs}) for model search:
 we can use the position of the cells in the pairs matrix to select an
@@
 Another approach is to visualize (the deviations from) a particular fitted model
-with similar ideas as in Section~\ref{sec:extensions}. Thus, mosaic displays can
+with ideas similar to those described in Section~\ref{sec:extensions}.
+Thus, mosaic displays can
 be extended from purely explorative views to model-based views on the data by
@@
 In a first step, we create a mosaic plot for an exploratory view of the data:
-Attitude towards punishment clearly is the dependent variable 
+attitude towards punishment is clearly the response variable
 here and should be used as the last split via highlighting.
@@
 then vertically by education, and finally horizontically again by memory),
-and to have attitude, the dependent variable, 
+and to have attitude, the response variable,
 highlighted in the tiles (see Figure~\ref{fig:punishhigh}).
@@
 in the other age groups). Furthermore, it can be seen that the proportion
-of people with memories of punishments increases with age and that
+of people who recall being punished increases with age, and that
 the approval rate decreases with education (this, however, is better shown in
@@
 exploratory analyses (pairs plots, highlighting, doubledecker plots)
-or model-based analyses (residual-based shading, conditional plots).
+or model-based analyses (residual-based shadings, conditional plots).
 

### 2007-10-09 trunk/vcd/R/legends.R
* enhancement of legend (min/max in range can be set) * fixes from proof-reading added
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@659 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -3,25 +3,27 @@ legend_resbased <- function(fontsize = 12,
                             y = unit(0.1, "npc"),
                             height = unit(0.8, "npc"),
                             width = unit(0.7, "lines"),
-			    digits = 3, 
+			    digits = 3,
 			    check_overlap = TRUE,
                             text = NULL,
                             steps = 200,
                             ticks = 10,
-                            pvalue = TRUE) {
+                            pvalue = TRUE,
+                            minres = NULL,
+                            maxres = NULL) {
 
   if(!is.unit(x)) x <- unit(x, "native")
   if(!is.unit(y)) y <- unit(y, "npc")
   if(!is.unit(width)) width <- unit(width, "lines")
   if(!is.unit(height)) height <- unit(height, "npc")
-  
+
   function(residuals, shading, autotext) {
     res <- as.vector(residuals)
-    
+
     if(is.null(text)) text <- autotext
     p.value <- attr(shading, "p.value")
     legend <- attr(shading, "legend")
-    
+
     if (all(residuals == 0)) {
       pushViewport(viewport(x = x, y = y, just = c("left", "bottom"),
                             default.unit = "native",
@@ -29,19 +31,26 @@ legend_resbased <- function(fontsize = 12,
       grid.lines(y = 0.5)
       grid.text(0, x = unit(1, "npc") + unit(0.8, "lines"),  y = 0.5)
       warning("All residuals are zero.")
-      
+
     } else {
 
-      pushViewport(viewport(x = x, y = y, just = c("left", "bottom"),
-                            yscale = range(res), default.unit = "native",
-                            height = height, width = width))
+        if (is.null(minres))
+            minres <- min(res)
+        if (is.null(maxres))
+            maxres <- max(res)
+        rangeres <- c(minres, maxres)
+
+        pushViewport(viewport(x = x, y = y, just = c("left", "bottom"),
+                              yscale = rangeres, default.unit = "native",
+                              height = height, width = width))
+
 
       if(is.null(legend$col.bins)) {
-        col.bins <- seq(min(res), max(res), length = steps)
+        col.bins <- seq(minres, maxres, length = steps)
         at <- NULL
       } else {
-        col.bins <- sort(unique(c(legend$col.bins, range(res))))
[…]

### 2009-02-23 pkg/vcd/paper/Color/letter.tex
rearrange repository
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@690 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\documentclass{wuletter}
\signature{Achim Zeileis, Kurt Hornik}
\usepackage[american]{babel}
\begin{document}
\begin{letter}{
Peter Westfall, Editor\\
MS 2101\\
Texas Tech University\\
Lubbock, TX  79409\\
806-742-2174}

\concerning{Manuscript submission}

\opening{Dear Prof.~Westfall,}

please find attached our manuscript
\begin{quote}
\textit{Choosing Color Palettes for Statistical Graphics}
\end{quote}
as a submission for the ``Statistical Computing and Graphics''
section of \emph{The American Statistician} (TAS). We feel that the
choice of suitable colors coding statistical information in
graphical displays is a topic of great interest to TAS
readers because it is a problem all of us face in every-day
practice. The paper discusses some simple algorithms how 
suitable perceptually-based colors can be chosen. \textsf{R}
software implementing these color selection strategies is
also freely available.

A potential problem might be the inevitable use of color 
within the paper because it is, to the best of our knowledge,
not possible to have color pages in the printed version of TAS.
If it is possible, we would be happy to contribute to the extra
costs---if not, we could easily make an electronic version of
the graphics available to the readers and, of course, all \textsf{R}
users can easily reproduce the graphics from the paper using our
software.

We consider TAS to be the ideal outlet for discussing a topic of
such broad interest and would be very pleased if it were considered
for review as a TAS submission.

\closing{Best regards,}
\end{letter}
\end{document}

### 2009-02-23 pkg/vcd/paper/Strucplot/JSS.Rnw
rearrange repository
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@690 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2006-10-25 trunk/vcd/inst/doc/strucplot.Rnw
@@
-\documentclass{Z}
+\documentclass[article]{jss}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% declarations for jss.cls %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \Shorttitle{The Strucplot Framework} %% a short title (if necessary)
-\Plaintitle{The Strucplot Framework: Visualizing Multi-way Contingency
-  Tables with vcd}
 
@@
 \Abstract{
-This paper has been published in the Journal of Statistical Software
-\citep{vcd:Meyer+Zeileis+Hornik:2006b} and 
-describes the ``strucplot'' framework for the visualization of
+This paper describes the ``strucplot'' framework for the visualization of
 multi-way contingency tables. Strucplot displays include hierarchical
@@
   conditional independence, HSV, HCL, residual-based shading, grid, R}
+
+%\VignetteIndexEntry{The Strucplot Framework: Visualizing Multi-way Contingency Tables With package vcd}
+%\VignetteDepends{vcd}
+%\VignetteKeywords{contingency tables, mosaic plots, association plots, sieve plots, R, categorical data, independence, conditional independence, HSV, HCL, residual-based shading, grid}
+%\VignettePackage{vcd}
+
+%% publication information
+%% NOTE: This needs to filled out ONLY IF THE PAPER WAS ACCEPTED.
+%% If it was not (yet) accepted, leave them commented.
+%% \Volume{13}
+%% \Issue{9}
+%% \Month{September}
+%% \Year{2004}
+%% \Submitdate{2004-09-29}
+%% \Acceptdate{2004-09-29}
+
+%% The address of (at least) one author should be given
+%% in the following format:
+\Address{
+  David Meyer\\
+  Department of Information Systems and Operations\\
+  E-mail: \email{David.Meyer@wu-wien.ac.at}\\
+  URL: \url{http://wi.wu-wien.ac.at/~meyer/}\\
+
+  Achim Zeileis\\
+  Department of Statistics and Mathematics\\
+  E-mail: \email{Achim.Zeileis@wu-wien.ac.at}\\
+  URL: \url{http://statmath.wu-wien.ac.at/~zeileis/}\\
+
+  Kurt Hornik\\
+  Department of Statistics and Mathematics\\
+  E-mail: \email{Kurt.Hornik@wu-wien.ac.at}\\
+  URL: \url{http://statmath.wu-wien.ac.at/~hornik/}\\
+
+  Wirtschaftsuniversit\"at Wien\\
+  Augasse 2--6
+  1090 Wien, Austria\\
+}
+%% It is also possible to add a telephone and fax number
+%% before the e-mail in the following format:
+%% Telephone: +43/1/31336-5053
+%% Fax: +43/1/31336-734
+
+%% for those who use Sweave please include the following line (with % symbols):
+%% need no \usepackage{Sweave.sty}
 
@@
 \item Providing a modular and flexible implementation to easily allow
-  user extensions \citep{vcd:Meyer+Zeileis+Hornik:2003,vcd:Meyer+Zeileis+Hornik:2006b}.
+  user extensions \citep{vcd:Meyer+Zeileis+Hornik:2003}.
 \end{enumerate}
@@
 implementations do exist in the packages \pkg{graphics} (in base \proglang{R}), 
-\pkg{vcd} \citep{vcd:Meyer+Zeileis+Hornik:2006b},
+\pkg{vcd} \citep{vcd:Meyer+Zeileis+Hornik:2006},
 and \pkg{iplots} \citep{vcd:urbanek+wichtrey:2006}, respectively.
@@
 Figures \ref{fig:arthritis} to \ref{fig:titanic} illustrate some of these
-extensions. Figure~\ref{fig:arthritis} shows the results from a
-double-blind clinical trial
+extensions (the code is given in Appendix \ref{sec:code}). 
+Figure~\ref{fig:arthritis} shows the results from a double-blind clinical trial
 investigating a new treatment for rheumatoid arthritis, using an
@@
 
+\section{Code used in the introduction}
+\label{sec:code}
+
+\begin{itemize}
+
+\item Mosaic plot for the \data{Arthritis} data in Figure~\ref{fig:arthritis}:
+\begin{Sinput}
+mosaic(art, gp = shading_max, split_vertical = TRUE)
+\end{Sinput}
+
+\item Conditional association plot for the \data{UCBAdmissions} 
+data in Figure~\ref{fig:ucbadmissions}:
+\begin{Sinput}
+cotabplot(UCBAdmissions, panel = cotab_indep, shade = TRUE, legend = FALSE)
+\end{Sinput}
+
+\item Mosaic plot for the \data{PreSex} data in Figure~\ref{fig:presex}:
+\begin{Sinput}
+presextest <- coindep_test(PreSex, margin = c(1,4), 
+                           indepfun = function(x) sum(x^2), n = 5000)
+mosaic(PreSex, condvars = c(1, 4), shade = TRUE, 
+       gp_args = list(p.value = presextest$p.value)) %$
+\end{Sinput}
+
+\item Double-decker plot for the \data{Titanic} data in Figure~\ref{fig:titanic}:
+\begin{Sinput}
+doubledecker(Survived ~ ., data = Titanic, labeling_args = 
+             list(set_varnames = c(Sex = "Gender")))
+\end{Sinput}
+\end{itemize}
+
 \end{appendix}

### 2009-02-23 pkg/vcd/paper/Shadings/JCGS.Rnw
rearrange repository
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@690 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2006-12-12 trunk/vcd/paper/Shadings/JCGS.Rnw
@@
 \rhead[]{}
-\lfoot[This is a preprint of an article accepted for publication in \textit{Journal of Computational and Graphical Statistics}.
+\lfoot[This is a preprint of an article published in \textit{Journal of Computational and Graphical Statistics}, Volume 16, Number 3, Pages 507--525.
 Copyright {\copyright} 2007 American Statistical Association, Institute of Mathematical Statistics, and Interface Foundation of North America]%
-{This is a preprint of an article accepted for publication in \textit{Journal of Computational and Graphical Statistics}.
+{This is a preprint of an article published in \textit{Journal of Computational and Graphical Statistics}, Volume 16, Number 3, Pages 507--525.
 Copyright {\copyright} 2007 American Statistical Association, Institute of Mathematical Statistics, and Interface Foundation of North America}
@@
 {\pkg{grid}~{\Sexpr{packageDescription("grid")$Version}}} \citep[see][]{vcd:Murrell:2002} and
-{\pkg{colorspace}~{\Sexpr{packageDescription("colorspace")$Version}}} \citep{vcd:Ihaka:2004}.
+{\pkg{colorspace}~{\Sexpr{packageDescription("colorspace")$Version}}} \citep{vcd:Ihaka:2006}.
 A vignette that demonstrates how all empirical examples can exactly be reproduced
